### 2010-06-18 trunk/tex/naaclhlt2010.tex
adding tex directory

%
% File naaclhlt2010.tex
%
% Contact: nasmith@cs.cmu.edu

\documentclass[11pt,letterpaper]{article}
\usepackage{naaclhlt2010}
\usepackage{times}
\usepackage{latexsym}
\setlength\titlebox{6.5cm}    % Expanding the titlebox

\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This
    document has been adapted from the instructions for earlier ACL
    and NAACL proceedings, including those for NAACL-HLT-09 by Joakim
    Nivre and Noah Smith, for ACL-05 by Hwee Tou Ng and Kemal Oflazer,
    for ACL-02 by Eugene Charniak and Dekang Lin, and earlier ACL and
    EACL formats.  Those versions were written by several people,
    including John Chen, Henry S. Thompson and Donald Walker.
    Additional elements were taken from the formatting instructions of
    the {\em International Joint Conference on Artificial
      Intelligence}.}}

\author{Claudia Leacock\\
  Butler Hill Group\\
  100 Bleecker Street, 27A\\
  New York, NY 10012, USA\\
  {\tt claudia.leacock@gmail.com}
  \And
  Richard Wicentowski \\
  Swarthmore College \\
  500 College Avenue \\
  Swarthmore, PA 19081, USA\\
  {\tt richardw@cs.swarthmore.edu}}

\date{}

\begin{document}
\maketitle
\begin{abstract}
  This document contains the instructions for preparing a camera-ready
  manuscript for the proceedings of NAACL HLT 2010. The document itself conforms
  to its own specifications, and is therefore an example of what
  your manuscript should look like.  Authors are asked to conform to
  all the directions reported in this document.
\end{abstract}

\section{Introduction}

The following instructions are directed to authors of papers accepted
for publication in the NAACL HLT 2010 proceedings.  All authors are required
to adhere to these specifications. Authors are required to provide 
a Portable Document Format (PDF) version of
their papers.  The proceedings will be printed on US-Letter paper.
Authors from countries in which access to word-processing systems is
limited should contact the publication chairs as soon as possible.

\section{General Instructions}

Manuscripts must be in two-column format.  Exceptions to the
two-column format include the title, authors' names and complete
addresses, which must be centered at the top of the first page (see
the guidelines in Subsection~\ref{ssec:first}), and any full-width
figures or tables .  Type single-spaced.  Do not number the pages.
Start all pages directly under the top margin.  See the guidelines
later regarding formatting the first page.

%% If the paper is produced by a printer, make sure that the quality
%% of the output is dark enough to photocopy well.  It may be necessary
%% to have your laser printer adjusted for this purpose.  Papers that are too
%% faint to reproduce well may not be included.

%% {\bf Do not print page numbers on the manuscript.}  Write them lightly
%% on the back of each page in the upper left corner along with the
%% (first) author's name.

The maximum length of a manuscript is eight (8) pages for the main
conference, printed single-sided, plus one (1) page for references
(see Section~\ref{sec:length} for additional information on the
maximum number of pages).  Do not number the pages.

\subsection{Electronically-available resources}

NAACL HLT provides this description in \LaTeX2e{} ({\tt naaclhlt2010.tex}) and PDF
format ({\tt naaclhlt2010.pdf}), along with the \LaTeX2e{} style file used to
format it ({\tt naaclhlt2010.sty}) and an ACL bibliography style ({\tt naaclhlt2010.bst}).
These files are all available at
{\tt http://naaclhlt2010.isi.edu}.  A Microsoft Word
template file ({\tt naaclhlt2010.dot}) is also available at the same URL. We
strongly recommend the use of these style files, which have been
appropriately tailored for the NAACL HLT 2010 proceedings.

\subsection{Format of Electronic Manuscript}
\label{sect:pdf}

For the production of the electronic manuscript you must use Adobe's
Portable Document Format (PDF). This format can be generated from
postscript files: on Unix systems, you can use {\tt ps2pdf} for this
purpose; under Microsoft Windows, you can use Adobe's Distiller, or
if you have cygwin installed, you can use {\tt dvipdf} or
{\tt ps2pdf}.  Note 
that some word processing programs generate PDF which may not include
all the necessary fonts (esp. tree diagrams, symbols). When you print
or create the PDF file, there is usually an option in your printer
setup to include none, all or just non-standard fonts.  Please make
sure that you select the option of including ALL the fonts.  {\em
  Before sending it, test your {\/\em PDF} by printing it from a
  computer different from the one where it was created}. Moreover,
some word processor may generate very large postscript/PDF files,
where each page is rendered as an image. Such images may reproduce
poorly.  In this case, try alternative ways to obtain the postscript
and/or PDF.  One way on some systems is to install a driver for a
postscript printer, send your document to the printer specifying
``Output to a file'', then convert the file to PDF.

For reasons of uniformity, Adobe's {\bf Times Roman} font should be
used. In \LaTeX2e{} this is accomplished by putting

\begin{quote}
\begin{verbatim}
\usepackage{times}
\usepackage{latexsym}
\end{verbatim}
\end{quote}
in the preamble.

Additionally, it is of utmost importance to specify the {\bf
  US-Letter format} (8.5in $\times$ 11in) when formatting the paper.
When working with {\tt dvips}, for instance, one should specify {\tt
  -t letter}.

Print-outs of the PDF file on US-Letter paper should be identical to the
hardcopy version.  If you cannot meet the above requirements about the
production of your electronic submission, please contact the
publication chairs above  as soon as possible.

\subsection{Layout}
\label{ssec:layout}

Format manuscripts two columns to a page, in the manner these
instructions are formatted. The exact dimensions for a page on US-letter
paper are:

\begin{itemize}
\item Left and right margins: 1in
\item Top margin:1in
\item Bottom margin: 1in
\item Column width: 3.15in
\item Column height: 9in
\item Gap between columns: 0.2in
\end{itemize}

\noindent Papers should not be submitted on any other paper size. Exceptionally,
authors for whom it is \emph{impossible} to format on US-Letter paper,
may format for \emph{A4} paper. In this case, they should keep the \emph{top}
and \emph{left} margins as given above, use the same column width,
height and gap, and modify the bottom and right margins as necessary.
Note that the text will no longer be centered.

\subsection{The First Page}
\label{ssec:first}

Center the title, author's name(s) and affiliation(s) across both
columns. Do not use footnotes for affiliations.  Do not include the
paper ID number assigned during the submission process. 
Use the two-column format only when you begin the abstract.

{\bf Title}: Place the title centered at the top of the first page, in
a 15 point bold font.  (For a complete guide to font sizes and styles, see Table~\ref{font-table}.)
Long title should be typed on two lines without
a blank line intervening. Approximately, put the title at 1in from the
top of the page, followed by a blank line, then the author's names(s),
and the affiliation on the following line.  Do not use only initials
for given names (middle initials are allowed). Do not format surnames
in all capitals (e.g., ``Leacock,'' not ``LEACOCK'').  The affiliation should
contain the author's complete address, and if possible an electronic
mail address. Leave about 0.75in between the affiliation and the body
of the first page.

{\bf Abstract}: Type the abstract at the beginning of the first
column.  The width of the abstract text should be smaller than the
width of the columns for the text in the body of the paper by about
0.25in on each side.  Center the word {\bf Abstract} in a 12 point
bold font above the body of the abstract. The abstract should be a
concise summary of the general thesis and conclusions of the paper.
It should be no longer than 200 words.  The abstract text should be in 10 point font.

{\bf Text}: Begin typing the main body of the text immediately after
the abstract, observing the two-column format as shown in 
the present document.  Do not include page numbers.

{\bf Indent} when starting a new paragraph. For reasons of uniformity,
use Adobe's {\bf Times Roman} fonts, with 11 points for text and 
subsection headings, 12 points for section headings and 15 points for
the title.  If Times Roman is unavailable, use {\bf Computer Modern
  Roman} (\LaTeX2e{}'s default; see section \ref{sect:pdf} above).
Note that the latter is about 10\% less dense than Adobe's Times Roman
font.

\subsection{Sections}

{\bf Headings}: Type and label section and subsection headings in the
style shown on the present document.  Use numbered sections (Arabic
numerals) in order to facilitate cross references. Number subsections
with the section number and the subsection number separated by a dot,
in Arabic numerals.

{\bf Citations}: Citations within the text appear
in parentheses as~\cite{Gusfield:97} or, if the author's name appears in
the text itself, as Gusfield~\shortcite{Gusfield:97}. 
Append lowercase letters to the year in cases of ambiguities.  
Treat double authors as in~\cite{Aho:72}, but write as 
in~\cite{Chandra:81} when more than two authors are involved. 
Collapse multiple citations as in~\cite{Gusfield:97,Aho:72}.

\textbf{References}: Gather the full set of references together under
the heading {\bf References}; place the section before any Appendices,
unless they contain references. Arrange the references alphabetically
by first author, rather than by order of occurrence in the text.
Provide as complete a citation as possible, using a consistent format,
such as the one for {\em Computational Linguistics\/} or the one in the 
{\em Publication Manual of the American 
Psychological Association\/}~\cite{APA:83}.  Use of full names for
authors rather than initials is preferred.  A list of abbreviations
for common computer science journals can be found in the ACM 
{\em Computing Reviews\/}~\cite{ACM:83}.

The \LaTeX{} and Bib\TeX{} style files provided roughly fit the
American Psychological Association format, allowing regular citations, 
short citations and multiple citations as described above.

{\bf Appendices}: Appendices, if any, directly follow the text and the
references (but see above).  Letter them in sequence and provide an
informative title: {\bf Appendix A. Title of Appendix}.

\textbf{Acknowledgment} sections should go as a last (unnumbered) section immediately
before the references.

\subsection{Footnotes}

{\bf Footnotes}: Put footnotes at the bottom of the page. They may
be numbered or referred to by asterisks or other
symbols.\footnote{This is how a footnote should appear.} Footnotes
should be separated from the text by a line.\footnote{Note the
line separating the footnotes from the text.}  Footnotes should be in 9 point font.

\subsection{Graphics}

{\bf Illustrations}: Place figures, tables, and photographs in the
paper near where they are first discussed, rather than at the end, if
possible.  Wide illustrations may run across both columns and should be placed at
the top of a page. Color illustrations are discouraged, unless you have verified that 
they will be understandable when printed in black ink.

\begin{table}
\begin{center}
\begin{tabular}{|l|rl|}
\hline \bf Type of Text & \bf Font Size & \bf Style \\ \hline
paper title & 15 pt & bold \\
author names & 12 pt & bold \\
author affiliation & 12 pt & \\
the word ``Abstract'' & 12 pt & bold \\
section titles & 12 pt & bold \\
document text & 11 pt  &\\
abstract text & 10 pt & \\
captions & 10 pt & \\
bibliography & 10 pt & \\
footnotes & 9 pt & \\
\hline
\end{tabular}
\end{center}
\caption{\label{font-table} Font guide. }
\end{table}

{\bf Captions}: Provide a caption for every illustration; number each one
sequentially in the form:  ``Figure 1. Caption of the Figure.'' ``Table 1.
Caption of the Table.''  Type the captions of the figures and 
tables below the body, using 10 point text.

\section{Length of Submission}
\label{sec:length}

The NAACL HLT 2010 main conference accepts submissions of long papers
and short papers.  The maximum length of a long paper manuscript is
eight (8) pages of content and one (1) additional page of references
\emph{only} (appendices count against the eight pages, not the
additional one page).  The maximum length of a short paper manuscript
is four (4) pages including references.  For both long and short
papers, all illustrations, references, and appendices must be
accommodated within these page limits, observing the formatting
instructions given in the present document.  Papers that do not
conform to the specified length and formatting requirements are
subject to be rejected without review.

% Up to two (2) additional pages may be purchased from ACL at the
% price of \$250 per page; please contact the publication chairs above
% for more information about this option.

\section*{Acknowledgments}

Do not number the acknowledgment section.

\begin{thebibliography}{}

\bibitem[\protect\citename{Aho and Ullman}1972]{Aho:72}
Alfred~V. Aho and Jeffrey~D. Ullman.
\newblock 1972.
\newblock {\em The Theory of Parsing, Translation and Compiling}, volume~1.
\newblock Prentice-{Hall}, Englewood Cliffs, NJ.

\bibitem[\protect\citename{{American Psychological Association}}1983]{APA:83}
{American Psychological Association}.
\newblock 1983.
\newblock {\em Publications Manual}.
\newblock American Psychological Association, Washington, DC.

\bibitem[\protect\citename{{Association for Computing Machinery}}1983]{ACM:83}
{Association for Computing Machinery}.
\newblock 1983.
\newblock {\em Computing Reviews}, 24(11):503--512.

\bibitem[\protect\citename{Chandra \bgroup et al.\egroup }1981]{Chandra:81}
Ashok~K. Chandra, Dexter~C. Kozen, and Larry~J. Stockmeyer.
\newblock 1981.
\newblock Alternation.
\newblock {\em Journal of the Association for Computing Machinery},
  28(1):114--133.

\bibitem[\protect\citename{Gusfield}1997]{Gusfield:97}
Dan Gusfield.
\newblock 1997.
\newblock {\em Algorithms on Strings, Trees and Sequences}.
\newblock Cambridge University Press, Cambridge, UK.

\end{thebibliography}

\end{document}

### 2010-06-18 tex/sarf.tex
adding tex directory

%
% File naaclhlt2010.tex
%
% Contact: nasmith@cs.cmu.edu

\documentclass[11pt,letterpaper]{article}
\usepackage{naaclhlt2010}
\usepackage{times}
\usepackage{latexsym}
\setlength\titlebox{6.5cm}    % Expanding the titlebox

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
\title{Case based Arabic Morphological Analysis with 
    Compositional Non-deterministic Automata }

\author{Hamza Harkous \\
  \And
  Jad Makhlouta \\
  American University of Beirut \\
  {\tt \{hhh20, jem04, fz11\}@aub.edu.lb }
  \And 
      Fadi Zaraket 
}

\date{}

\begin{document}
\maketitle
\begin{abstract}
  Morphological analysis is key in current automated analysis techniques 
  for Arabic text. 
  Current morphological analyzers take an Arabic word as input and 
  enumerate all possible morphological solutions (stems or 
  derivations) using concatenation based analysis.
  The solutions suffer from accuracy due to the inherent difficulties
  of morphological analysis of the Arabic language. 
  The enumeration of all possible solutions may also hurt the performance
  in cases where not all solutions are needed.
  In this paper, we present a novel efficient morphological analyzer that 
  uses parallel compositional non-deterministic Automata 
  driven by a case based controller.
  The analyzer keeps alive possible analyses of the text 
  for the controller to intervene with a decision and thus 
  reduce false positives on case basis.

  Our concatenative analysis uses recursive affixes in order to
  retain better part of speech information and enhance the 
  efficiency of affix matching. 
  The analyzer takes as input a character from a full text stream 
  rather than a sequence of words and thus performs tokenization
  on the fly and based on morphological correctness. 
  This relieves the user from preprocessing the text and provides
  a more exact tokenization than delimiter based tokenization. 
  This is important for Arabic since two words can occur in a text
  without a delimiter in between.

  We used our analyzer to successfully automate the analysis of 
  three books of Islamic literature where we segmented each book
  into several narrations. In each narration, we detected 
  a chain of narrators, where each narrator is a complex sequence of
  proper names separated with name connector phrases. 
  This is a collection of chain of narrator structures where each
  chain is a structure with three levels of hierarchy.
  Our results show better accuracy and higher efficiency than current 
  morphological analyzers. 
\end{abstract}

\section{Introduction}

\subsection{Contributions}
% does not have to be a subsection, 
%just to highlight it as an item we should not miss

% paper structure

\section{Background}

\section{Related Work }

\section{Sarf}

\subsection{Recursive Affixes}
% smaller FSM
% retain POS info

\subsection{Motivating Example}
% the FSMs for the three words

\subsection{Affix Linear FSM}

\subsection{Stem Linear FSM}

\subsection{Non-deterministic Composition of FSMs}
% algorithm
% diagram of abstract machines

\section{Islamic Literature Case Study}

\subsection{Hadith Segmentation}

\subsection{Chain of Narrators}

\subsection{Controller}
% FSM for the controller

\section{Results}
\label{sec:results}

\subsection{Efficiency Comparison Against Buckwalter and SAMA}

\subsection{Case Study Accuracy and Efficiency Results}

% accuracy measure:
% on names: a word is a name or not
% chains: narrator is missing from a chain
% segmentation: two hadiths are merged

\section{Future Work}

%For reasons of uniformity, Adobe's {\bf Times Roman} font should be
%used. In \LaTeX2e{} this is accomplished by putting

%\begin{quote}
%\begin{verbatim}
%\usepackage{times}
%\usepackage{latexsym}
%\end{verbatim}
%\end{quote}
%in the preamble.

%Additionally, it is of utmost importance to specify the {\bf
%  US-Letter format} (8.5in $\times$ 11in) when formatting the paper.
%When working with {\tt dvips}, for instance, one should specify {\tt
%  -t letter}.

%{\bf Citations}: Citations within the text appear
%in parentheses as~\cite{Gusfield:97} or, if the author's name appears in
%the text itself, as Gusfield~\shortcite{Gusfield:97}. 
%Append lowercase letters to the year in cases of ambiguities.  
%Treat double authors as in~\cite{Aho:72}, but write as 
%in~\cite{Chandra:81} when more than two authors are involved. 
%Collapse multiple citations as in~\cite{Gusfield:97,Aho:72}.

\section*{Acknowledgments}

\begin{thebibliography}{}

\bibitem[\protect\citename{Aho and Ullman}1972]{Aho:72}
Alfred~V. Aho and Jeffrey~D. Ullman.
\newblock 1972.
\newblock {\em The Theory of Parsing, Translation and Compiling}, volume~1.
\newblock Prentice-{Hall}, Englewood Cliffs, NJ.

%\bibitem[\protect\citename{{American Psychological Association}}1983]{APA:83}
%{American Psychological Association}.
%\newblock 1983.
%\newblock {\em Publications Manual}.
%\newblock American Psychological Association, Washington, DC.
%
%\bibitem[\protect\citename{{Association for Computing Machinery}}1983]{ACM:83}
%{Association for Computing Machinery}.
%\newblock 1983.
%\newblock {\em Computing Reviews}, 24(11):503--512.
%
%\bibitem[\protect\citename{Chandra \bgroup et al.\egroup }1981]{Chandra:81}
%Ashok~K. Chandra, Dexter~C. Kozen, and Larry~J. Stockmeyer.
%\newblock 1981.
%\newblock Alternation.
%\newblock {\em Journal of the Association for Computing Machinery},
%  28(1):114--133.
%
%\bibitem[\protect\citename{Gusfield}1997]{Gusfield:97}
%Dan Gusfield.
%\newblock 1997.
%\newblock {\em Algorithms on Strings, Trees and Sequences}.
%\newblock Cambridge University Press, Cambridge, UK.

\end{thebibliography}

\end{document}

### 2010-06-24 trunk/tex/sarf1.tex
aaa

%
% File naaclhlt2010.tex
%
% Contact: nasmith@cs.cmu.edu

\documentclass[11pt,letterpaper]{article}
\usepackage{arabtex}
\usepackage{naaclhlt2010}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{multirow}
\usepackage{amsthm}

\usepackage{relsize}
\usepackage{fancyvrb}
%\usepackage{times}
\usepackage{latexsym}

\usepackage{utf8}
\setarab
\fullvocalize
%\transtrue
\arabtrue

\setlength\titlebox{6.5cm}    % Expanding the titlebox

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
\title{Case based Arabic Morphological Analysis with 
    Compositional Non-deterministic Automata }

%\author{ Jad Makhlouta \\
%  \And
%Hamza Harkous \\
%  American University of Beirut \\
%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
%  \And 
%  Fadi Zaraket 
%}

\date{}

\begin{document}
\maketitle

\begin{abstract}
Natural language processing uses morphological 
analysis to abstract and annotate text.
Often times, and in particular in the context
of the Arabic language, annotations resulting from 
an ongoing branch of morphological analysis may not be
appropriate for the case study at hand. 
In this paper we present Sarf, a case based morphological 
analyser for Arabic text. 
Sarf allows a case-based controller to control and refine
the state of the analysis on the fly. 
It uses compositional non-deterministic finite state machines 
to analyse a stream of text. 
Each alive machine corresponds to a valid analysis. 
Sarf uses concatenative analysis based on recursive affixes 
to better preserve part of speech information.
We automated the analysis of three books of Islamic
narrations (hadith) using Sarf where we
abstracted each book into a vector of complex structures.
Our results show high accuracy and efficiency compared to
state of the art analyzers. 
\end{abstract}

\section{Introduction}

Automated analysis of Arabic data sets, including texts, 
publications, records and digital media is essential
with the huge digital Arabic content available nowadays. 
Morphological analysis is key in current automated 
analysis techniques for Arabic text. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis
to analyze an Arabic word and
consider its internal structure composed of several
{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
The analysis of one word may lead to several possible
morphological solutions.
\vocalize
For instance the word \RL{'a.hmadH} 
may have two valid morphological analyses where in the first
\RL{'a} is a prefix and the word means "I praise him",
and in the second \RL{'a} is part of the stem 
\RL{'a.hmad} (a proper noun)
and the word means "his Ahmad".

\novocalize
The solutions suffer from accuracy due to inherent difficulties
of morphological analysis of the Arabic language. 
For example, it is common practice to write Arabic text
without short vowels. 
Arabic letters can also have up to four different shapes 
corresponding to their position in a word, i.e, begining of 
a word, middle of a word, end of word, and separated. 
This allows two consecutive words such as 
\RL{almdrsT}\RL{al_A}
to be visually recognizable
as two separate words \RL{al_A} and \RL{almdrsT} without 
the need of a delimiter such as a space in between. 
This specifically happens when the first word ends with
a non-connecting letter and is referred to as the ``run-o'' words
problem~\cite{Buckwalter:04}.
Such forms occur in text and greatly increase the
difficulty of tokenization.

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Xerox~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} exist.
They take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies as noted in~\cite{Maamouri:10}. 
Moreover, a white space delimited token may have 
more than one word.
Other morphological analyzers such as Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
also use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of perfomance.

We hypothesise that many NLP case studies need the 
morphological analyser to answer simple queries that do not need 
all this complexity and accuracy and thus most of the work can be ignored. 
For example, if the query is interested in proper names and the 
prefix in question now can only connect to a verb, we can provide an answer
without completing the analysis. 
We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
with features demanding resolution at an abstraction level
higher than the text itself
led to a refined analyzer~\cite{Kulick:10}.

In this paper, we present Sarf, a {\em novel case-based efficient morphological 
analyzer} that uses parallel compositional non-deterministic Automata 
driven by a case based controller.
Each alive machine in Sarf represents a valid morphological analysis. 
Sarf keeps alive all possible analyses of the text and gives opportunity for the 
controller to intervene at every single input character. 
The decisions of the case-based controller can thus prune false positives early 
in the run. 
Each alive machine in Sarf takes as input one character at a time 
from a text stream. Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness.
To out knowledge, This solves the "run-on" words problem.

Sarf uses a {\em recursive} concatenative analysis with a novel 
refinement. 
Sarf introduces recursive affixes in order to
retain better part of speech information and enhance the 
efficiency of affix matching.

We validate our hypothesis using a case study from the Islamic 
literature. 
A {\em hadith} is a narration related to the prophet Mohamad
through a {\em sanad} or a sequence of narrators. 
The authentication of a hadith higly depends on the credibility
of each of the narrators as reported in separate biography 
books. 
In this paper we consider the problem of automatically segmenting
a hadith book into narrations, then segmenting each narration into
its content or {\em matn} and its sanad.
We also would like to partition the sanad accurately into the 
separate narrators so that we can later look each one of them 
up in the biography books.

The hadith case study is interesting for Sarf since the controller
considers most of the analyses where we have a concentration
of proper names in the text, and ignores most of the analyses as long 
as they are valid where we do not have proper names. 
The controller is also interested in a small subset of words 
that relate persons to each other such as \RL{bin} or "the sun of" 
or words that mean narrate such as \RL{Kal} or "said". 
With Sarf, we successfully automated the analysis of 
three books of Islamic narrations~\cite{narrations1,narrations2, narrations3}
and extracted from each one of them a vector of a three level deep
complex structure.

\subsection{Contributions}
% does not have to be a subsection, 
%just to highlight it as an item we should not miss

% paper structure

\section{Background}

\section{Related Work }

\section{Sarf}

\subsection{Recursive Affixes}
% smaller FSM
% retain POS info

\subsection{Motivating Example}
% the FSMs for the three words

\begin{figure*}[tb]
\center{
\resizebox{1.8\columnwidth}{!}
{ \input{psfigs/exampleFSM1.pstex_t}}
\caption{Example FSM.}
\label{f:examples}
}
\end{figure*}

\subsection{Affix Linear FSM}

\subsection{Stem Linear FSM}

\subsection{Non-deterministic Composition of FSMs}
% algorithm
% diagram of abstract machines

\section{Islamic Literature Case Study}

\subsection{Hadith Segmentation}

\subsection{Chain of Narrators}

\subsection{Controller}
% FSM for the controller

\section{Results}
\label{sec:results}

\subsection{Efficiency Comparison Against Buckwalter and SAMA}

\subsection{Case Study Accuracy and Efficiency Results}

% accuracy measure:
% on names: a word is a name or not
% chains: narrator is missing from a chain
% segmentation: two hadiths are merged

\section{Future Work}

%For reasons of uniformity, Adobe's {\bf Times Roman} font should be
%used. In \LaTeX2e{} this is accomplished by putting

%\begin{quote}
%\begin{verbatim}
%\usepackage{times}
%\usepackage{latexsym}
%\end{verbatim}
%\end{quote}
%in the preamble.

%Additionally, it is of utmost importance to specify the {\bf
%  US-Letter format} (8.5in $\times$ 11in) when formatting the paper.
%When working with {\tt dvips}, for instance, one should specify {\tt
%  -t letter}.

%{\bf Citations}: Citations within the text appear
%in parentheses as~\cite{Gusfield:97} or, if the author's name appears in
%the text itself, as Gusfield~\shortcite{Gusfield:97}. 
%Append lowercase letters to the year in cases of ambiguities.  
%Treat double authors as in~\cite{Aho:72}, but write as 
%in~\cite{Chandra:81} when more than two authors are involved. 
%Collapse multiple citations as in~\cite{Gusfield:97,Aho:72}.

\section*{Acknowledgments}

\bibliographystyle{naaclhlt2010}
\bibliography{fzAr}

\end{document}

### 2010-06-24 tex/sarf.tex
Added FSM.fig with the arabic letters now working. Added code to latex to draw figure in the file.. I noticed that some part of the reference is above the figure...and the caption is not centered...so they need fixing later.
edited sarf.txe

@@
 \documentclass[11pt,letterpaper]{article}
+\usepackage{arabtex}
 \usepackage{naaclhlt2010}
+\usepackage{amsmath}
+\usepackage{amssymb}
 \usepackage{times}
+\usepackage{caption}
+\usepackage{epsfig}
+\usepackage{subfigure}
+\usepackage{color}
+\usepackage{rotate}
+\usepackage{rotating}
+\usepackage{multirow}
+\usepackage{amsthm}
+
+\usepackage{relsize}
+\usepackage{fancyvrb}
+%\usepackage{times}
 \usepackage{latexsym}
+
+\usepackage{utf8}
+\setarab
+\fullvocalize
+%\transtrue
+\arabtrue
+
 \setlength\titlebox{6.5cm}    % Expanding the titlebox
@@
 
-\author{Hamza Harkous \\
-  \And
-  Jad Makhlouta \\
-  American University of Beirut \\
-  {\tt \{hhh20, jem04, fz11\}@aub.edu.lb }
-  \And 
-      Fadi Zaraket 
-}
+%\author{ Jad Makhlouta \\
+%  \And
+%Hamza Harkous \\
+%  American University of Beirut \\
+%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
+%  \And 
+%  Fadi Zaraket 
+%}
 
@@
 \maketitle
+
 \begin{abstract}
-  Morphological analysis is key in current automated analysis techniques 
-  for Arabic text. 
-  Current morphological analyzers take an Arabic word as input and 
-  enumerate all possible morphological solutions (stems or 
-  derivations) using concatenation based analysis.
-  The solutions suffer from accuracy due to the inherent difficulties
-  of morphological analysis of the Arabic language. 
-  The enumeration of all possible solutions may also hurt the performance
-  in cases where not all solutions are needed.
-  In this paper, we present a novel efficient morphological analyzer that 
-  uses parallel compositional non-deterministic Automata 
-  driven by a case based controller.
-  The analyzer keeps alive possible analyses of the text 
-  for the controller to intervene with a decision and thus 
-  reduce false positives on case basis.
-
-  Our concatenative analysis uses recursive affixes in order to
-  retain better part of speech information and enhance the 
-  efficiency of affix matching. 
-  The analyzer takes as input a character from a full text stream 
-  rather than a sequence of words and thus performs tokenization
-  on the fly and based on morphological correctness. 
-  This relieves the user from preprocessing the text and provides
-  a more exact tokenization than delimiter based tokenization. 
-  This is important for Arabic since two words can occur in a text
-  without a delimiter in between.
-
-  We used our analyzer to successfully automate the analysis of 
-  three books of Islamic literature where we segmented each book
-  into several narrations. In each narration, we detected 
-  a chain of narrators, where each narrator is a complex sequence of
-  proper names separated with name connector phrases. 
-  This is a collection of chain of narrator structures where each
-  chain is a structure with three levels of hierarchy.
-  Our results show better accuracy and higher efficiency than current 
-  morphological analyzers. 
+Natural language processing uses morphological 
+analysis to abstract and annotate text.
+Often times, and in particular in the context
+of the Arabic language, annotations resulting from 
+an ongoing branch of morphological analysis may not be
+appropriate for the case study at hand. 
+In this paper we present Sarf, a case based morphological 
+analyser for Arabic text. 
+Sarf allows a case-based controller to control and refine
+the state of the analysis on the fly. 
+It uses compositional non-deterministic finite state machines 
+to analyse a stream of text. 
+Each alive machine corresponds to a valid analysis. 
+Sarf uses concatenative analysis based on recursive affixes 
+to better preserve part of speech information.
+We automated the analysis of three books of Islamic
+narrations (hadith) using Sarf where we
+abstracted each book into a vector of complex structures.
+Our results show high accuracy and efficiency compared to
+state of the art analyzers. 
 \end{abstract}
@@
 \section{Introduction}
+
+Automated analysis of Arabic data sets, including texts, 
+publications, records and digital media is essential
+with the huge digital Arabic content available nowadays. 
+Morphological analysis is key in current automated 
+analysis techniques for Arabic text. 
+Current morphological analyzers~\cite{Sughaiyer:04}
+use concatenative analysis
+to analyze an Arabic word and
+consider its internal structure composed of several
+{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
+An affix can be a {\em prefix, suffix, } or an {\em infix}.
+The analysis of one word may lead to several possible
+morphological solutions.
+\vocalize
+For instance the word \RL{'a.hmadH} 
+may have two valid morphological analyses where in the first
+\RL{'a} is a prefix and the word means ``I praise him'',
+and in the second \RL{'a} is part of the stem 
+\RL{'a.hmad} (a proper noun)
+and the word means ``his Ahmad''.
+
+\novocalize
+The solutions suffer from accuracy due to inherent difficulties
+of morphological analysis of the Arabic language. 
+For example, it is common practice to write Arabic text
+without short vowels. 
+Arabic letters can also have up to four different shapes 
+corresponding to their position in a word, i.e, begining of 
+a word, middle of a word, end of word, and separated. 
+This allows two consecutive words such as 
+\RL{almdrsT}\RL{al_A}
+to be visually recognizable
+as two separate words \RL{al_A} and \RL{almdrsT} without 
+the need of a delimiter such as a space in between. 
+This specifically happens when the first word ends with
+a non-connecting letter and is referred to as the ``run-o'' words
+problem~\cite{Buckwalter:04}.
+Such forms occur in text and greatly increase the
+difficulty of tokenization.
+
+Current morphological analyzers such as 
+Buckwalter~\shortcite{Buckwalter:02},
+Xerox~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+and ElixirFM~\cite{Otakar:07} exist.
+They take as input white space delimited tokens~\cite{Kulick:10},
+consider them as words,
+and enumerate all possible solutions. 
+This exhaustive enumeration may hurt performance and may
+not be necessary or appropriate
+in some case studies as noted in~\cite{Maamouri:10}. 
+Moreover, a white space delimited token may have 
+more than one word.
+Other morphological analyzers such as Amira~\cite{Diab:07,Benajiba:07},
+MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
+also use machine learning and support vector machines (SVM) 
+to enhance the accuracy of the morphological analysis at the expense 
+of perfomance.
+
+We hypothesise that many NLP case studies need the 
+morphological analyser to answer simple queries that do not need 
+all this complexity and accuracy and thus most of the work can be ignored. 
+For example, if the query is interested in proper names and the 
+prefix in question now can only connect to a verb, we can provide an answer
+without completing the analysis. 
+We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
+addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
+with features demanding resolution at an abstraction level
+higher than the text itself
+led to a refined analyzer~\cite{Kulick:10}.
+
+In this paper, we present Sarf, a {\em novel case-based efficient morphological 
+analyzer} that uses parallel compositional non-deterministic Automata 
+driven by a case based controller.
+Each alive machine in Sarf represents a valid morphological analysis. 
+Sarf keeps alive all possible analyses of the text and gives opportunity for the 
+controller to intervene at every single input character. 
+The decisions of the case-based controller can thus prune false positives early 
+in the run. 
+Each alive machine in Sarf takes as input one character at a time 
+from a text stream. Sarf does not assume that the word at hand is a token and
+performs tokenization on the fly based on morphological correctness.
+To out knowledge, This solves the ``run-on'' words problem.
+
+Sarf uses a {\em recursive} concatenative analysis with a novel 
+refinement. 
+Sarf introduces recursive affixes in order to
+retain better part of speech information and enhance the 
+efficiency of affix matching.
+
+We validate our hypothesis using a case study from the Islamic 
+literature. 
+A {\em hadith} is a narration related to the prophet Mohamad
+through a {\em sanad} or a sequence of narrators. 
+The authentication of a hadith higly depends on the credibility
+of each of the narrators as reported in separate biography 
+books. 
+In this paper we consider the problem of automatically segmenting
+a hadith book into narrations, then segmenting each narration into
+its content or {\em matn} and its sanad.
+We also would like to partition the sanad accurately into the 
+separate narrators so that we can later look each one of them 
+up in the biography books.
+
+The hadith case study is interesting for Sarf since the controller
+considers most of the analyses where we have a concentration
+of proper names in the text, and ignores most of the analyses as long 
+as they are valid where we do not have proper names. 
+The controller is also interested in a small subset of words 
+that relate persons to each other such as \RL{bin} or ``the sun of''
+or words that mean narrate such as \RL{qaal} or ``said''. 
+With Sarf, we successfully automated the analysis of 
+three books of Islamic narrations~\cite{narrations1,narrations2, narrations3}
+and extracted from each one of them a vector of a three level deep
+complex structure.
 
@@
 
+Many morphological analyzers exist~\cite{sWe limit our comparison to related wowour discussion of related work toLots of work
+
 \section{Sarf}
@@
 % the FSMs for the three words
+
+\begin{figure*}[tb]
+\center{
+\resizebox{1.8\columnwidth}{!}
+{ \input{psfigs/FSM.pstex_t}}
+\caption{Example FSM.}
+\label{f:examples}
+}
+\end{figure*}
 
@@
 
-\begin{thebibliography}{}
-
-\bibitem[\protect\citename{Aho and Ullman}1972]{Aho:72}
-Alfred~V. Aho and Jeffrey~D. Ullman.
-\newblock 1972.
-\newblock {\em The Theory of Parsing, Translation and Compiling}, volume~1.
-\newblock Prentice-{Hall}, Englewood Cliffs, NJ.
-
-%\bibitem[\protect\citename{{American Psychological Association}}1983]{APA:83}
-%{American Psychological Association}.
-%\newblock 1983.
-%\newblock {\em Publications Manual}.
-%\newblock American Psychological Association, Washington, DC.
-%
-%\bibitem[\protect\citename{{Association for Computing Machinery}}1983]{ACM:83}
-%{Association for Computing Machinery}.
-%\newblock 1983.
-%\newblock {\em Computing Reviews}, 24(11):503--512.
-%
-%\bibitem[\protect\citename{Chandra \bgroup et al.\egroup }1981]{Chandra:81}
-%Ashok~K. Chandra, Dexter~C. Kozen, and Larry~J. Stockmeyer.
-%\newblock 1981.
-%\newblock Alternation.
-%\newblock {\em Journal of the Association for Computing Machinery},
-%  28(1):114--133.
-%
-%\bibitem[\protect\citename{Gusfield}1997]{Gusfield:97}
-%Dan Gusfield.
-%\newblock 1997.
-%\newblock {\em Algorithms on Strings, Trees and Sequences}.
-%\newblock Cambridge University Press, Cambridge, UK.
-
-\end{thebibliography}
+\bibliographystyle{naaclhlt2010}
+\bibliography{fzAr}
 

### 2010-06-25 tex/sarf.tex
editing

@@
 morphological analyser to answer simple queries that do not need 
-all this complexity and accuracy and thus most of the work can be ignored. 
+high complexity and accuracy at the low morphological analysis level 
+and that can often compensate for tolerable inaccuracy at higher level. 
 For example, if the query is interested in proper names and the 
-prefix in question now can only connect to a verb, we can provide an answer
-without completing the analysis. 
+prefix in question in the analysis can only connect to a verb, 
+we can provide an answer without completing the analysis. 
+%TODO A lso if a word such as \RL{wasiim} was name was ambiguous and we were looking for a sequence of names, then 
 We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
@@
 led to a refined analyzer~\cite{Kulick:10}.  
+We also find evidence in~\cite{Habash:06} that different types of 
+morphological analyses behave differently on the same case study. 
+We strongly beleive and we find evidence in our results that a 
+case-based controller intervening at every decision is necessary to 
+guide, use, prune, and refine the morphological analysis.
 
-In this paper, we present Sarf, a {\em novel case-based efficient morphological 
-analyzer} that uses parallel compositional non-deterministic Automata 
-driven by a case based controller.
+\subsection{Sarf}
+\label{sec:intro:sarf}
+
+In this paper, we present Sarf, a {\em novel case-based efficient
+morphological analyzer} that uses parallel compositional 
+non-deterministic automata driven by a case based controller.
 Each alive machine in Sarf represents a valid morphological analysis. 
-Sarf keeps alive all possible analyses of the text and gives opportunity for the 
-controller to intervene at every single input character. 
-The decisions of the case-based controller can thus prune false positives early 
-in the run. 
+Sarf keeps alive all possible analyses of the text and gives 
+opportunity to the controller to intervene at every single input 
+character. 
+The decisions of the case-based controller can thus prune false 
+positives early in the run. 
 Each alive machine in Sarf takes as input one character at a time 
-from a text stream. Sarf does not assume that the word at hand is a token and
+from a text stream. 
+Sarf does not assume that the word at hand is a token and
 performs tokenization on the fly based on morphological correctness.
-To out knowledge, This solves the ``run-on'' words problem.
+To out knowledge, this is the first morphological analyzer that 
+handles the ``run-on'' words problem.
 
@@
 
-\subsection{Contributions}
+%\subsection{Contributions}
+
+We make the following key contributions. 
+\begin{enumerate}
+\item {\bf Case-based analysis:}  We deisgned and implemented
+a case-based morphological analyzer where a case-based controller
+machine controls and guides the analysis. 
+\item {\bf Exhaustive analysis:} Since we have a case-based 
+controller that can decide on the fly whether an analysis is 
+accepted or not, we can afford to keep all valid analyses modulo
+those pruned by the case controller. 
+We do that using non-deterministic FSAs. 
+\item {\bf Recursive affixes:}
+We refine the concatenative analysis of 
+Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
+This allowed Sarf to retain better compositional part of speech
+tags.
+\item {\bf Hadith case study:}
+We evaluated Sarf using the hadith case study. We were able to
+efficiently extract the sequence of narrators into a complex
+data structure with three levels of hierarchy with high accuracy. 
+\end{enumerate}
+
 % does not have to be a subsection, 
@@
 
-Many morphological analyzers exist~\cite{sWe limit our comparison to related wowour discussion of related work toLots of work
+Many morphological analyzers exist~\cite{Sughaiyer:04}. 
+Some use specialized part of speech tag sets~\cite{Khoja:01,Darwish:02}.
+
+Buckwalter: used lexicon, added proper nouns, word based, recursive affixes
+
+MAGEAD
+
+Beeseley: many compiled FSAs compared to compositions of three FSAs
+
+SAMA: word based, treebank specific,
+
+MADA+TOKAN: no controller, we do not need a tokenizer
+
+AMIRA: machine learning...
+
+ElixirFM:
 

### 2010-06-25 trunk/tex/sarf.tex
added related work

--- 2010-06-25 tex/sarf.tex
@@
 
-Many morphological analyzers exist~\cite{Sughaiyer:04}. 
-Some use specialized part of speech tag sets~\cite{Khoja:01,Darwish:02}.
+The survey work in~\cite{Sughaiyer:04} discusses and compares
+several morphological analyzers. 
+Analyzers such as~\cite{Khoja:01,Darwish:02} 
+target specific specific applications in the morphological 
+analyzer itself or use a specific set of part of speech tags
+as their reference.
+Sarf differs in that it is a general and rich morphological 
+analyzer that reports all possible solutions. 
+It is case-based in the sense that a controller that drives
+the morphological analyzers prunes these decisions. 
+Up to our knowledge, Sarf is the only morphological analyzer
+that addresses the ``run-on'' words problem.
 
-Buckwalter: used lexicon, added proper nouns, word based, recursive affixes
+Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
+It differs than Buckwalter where it defines a shorter list of affixes
+and a longer list of affix compatibility rules to allow recursive 
+affixes so that we can have prefix-prefix an suffix-suffix 
+concatenations.
+This allows Sarf to better maintain and propagate 
+the part of speech information
+tags associated with each prefix while these need to be highly 
+redundant with Buckwalter and that may lead to consistency issues. 
+Sarf keeps the possible analyses alive as a sequence corresponding
+to the text. Buckwalter produces a set of solutions for each token 
+and the several analysis threads of the whole text need to be 
+generated as a product of the token solutions.
+
+Sarf is similar to Beeseley~\shortcite{Beeseley:01} in that it uses
+finiste-state automata (FSA). 
+Beeseley reports that the number of the automata generated by a compiler
+from Xerox rules is not controlled and also reports a difficulty to 
+compose the FSAs into a single FSA framework. 
+Sarf solves the problem with a non-deterministic composition of three
+FSAs that are the prefix, stem, and suffix FSAs. 
+With Sarf it is easy to control the FSAs with a user custom case 
+controller, while with Beeseley the effort requires recompiling the
+Xerox rules.
 
 MAGEAD
-
-Beeseley: many compiled FSAs compared to compositions of three FSAs
 
@@
 
+Sarf extends the lexicon of Buckwalter~\shotcite{Buckwalter:02} with 
+proper names and location names extracted from different online 
+sources~\footnote{\hyperref{http://alasmaa.net/}{http://alasmaa.net/ }, 
+    \hyperref{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
+as well as books 
 \subsection{Recursive Affixes}

### 2010-06-25 trunk/tex/sarf.tex
finished related work

@@
 \usepackage{latexsym}
+\usepackage[colorlinks=false]{hyperref}
 
@@
 With Sarf, we successfully automated the analysis of 
-three books of Islamic narrations~\cite{narrations1,narrations2, narrations3}
+three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
 and extracted from each one of them a vector of a three level deep
@@
 
-Sarf is similar to Beeseley~\shortcite{Beeseley:01} in that it uses
+Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
 finiste-state automata (FSA). 
-Beeseley reports that the number of the automata generated by a compiler
+Beesley reports that the number of the automata generated by a compiler
 from Xerox rules is not controlled and also reports a difficulty to 
@@
 With Sarf it is easy to control the FSAs with a user custom case 
-controller, while with Beeseley the effort requires recompiling the
+controller, while with Beesley the effort requires recompiling the
 Xerox rules.
 
-MAGEAD
+Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
+the several valid morphological analyses as a richness that 
+should be exploited at a higher abstract level rather than
+an inaccuracy that should be corrected. 
+Sarf differs in that it provides an interface for the 
+case study to interfere in judging the analyses much earlier and
+at a finer granularity level. 
+Also in Sarf there is no need for a separate tokenizer such as
+TOKAN as each alive automata carries with it the positions in 
+text where a composition decision was made. 
+The case-baded controller can build the trace when needed.
 
-SAMA: word based, treebank specific,
+%MAGEAD 
+    %Maamouri, Mohamed and \bgroup et al.\egroup },
 
-MADA+TOKAN: no controller, we do not need a tokenizer
+We strongly feel that the refinemet of SAMA~\cite{Maamouri:10} 
+is the closest to our approach. SAMA was refined to interact with
+the ATB~\cite{Maamouri:04} project after the addition of a large 
+new corpus. We find evidence in this decision that our hypothesis
+about the utility of case-based analyzers. The refinements
+that resulted in algorithmic changes in SAMA were done manually
+and worked in integration with the ATB format only. 
+In Sarf we propse a case-based controller approach that can modify 
+the behaviour of the analyzer on the fly and that should interact
+with the case study. In Sarf the analyzer remains general, and the 
+case-based controller specializes the analyses.
 
-AMIRA: machine learning...
+The AMIRA~\cite{Diab:07} analyzer uses 
+a language independent SVM based analysis. 
+While the SVM analysis learns features from the text passed to the case
+study, it does not make use of the type oh the query the 
+case study targets in the text.
 
-ElixirFM:
+Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
+of the Buckwalter analyzer. 
+We also use deterministic parsing with tries or our analysis 
+to run the prefix, suffix and stem FSAs. 
+We think that their inferential-realizational approach 
+that is highly compatible with the Arabic linguistic 
+description~\cite{Badawi:04}
+can make benefit of many features unique to the Arabic language.
+Sarf leaves implementing that to the case-based controller
+since in many cases the abstraction needed may not need the 
+text to be grammar correct.
+We can compare the case-based controller in Sarf to a high level
+component in the ElixirFM domain sepcific language.
 
@@
 
-Sarf extends the lexicon of Buckwalter~\shotcite{Buckwalter:02} with 
+Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with 
 proper names and location names extracted from different online 
-sources~\footnote{\hyperref{http://alasmaa.net/}{http://alasmaa.net/ }, 
-    \hyperref{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
-as well as books 
+sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
+\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
+as well as biblical sources~\cite{}.
+
 \subsection{Recursive Affixes}

### 2010-06-25 trunk/tex/sarf.tex
added the example diagram and 2 sentences which are not yet proofread and working on to explain it in detail

@@
+%
+% File naaclhlt2010.tex
+%
+% Contact: nasmith@cs.cmu.edu
+
+\documentclass[11pt,letterpaper]{article}
+\usepackage{arabtex}
+\usepackage{naaclhlt2010}
+\usepackage{amsmath}
+\usepackage{amssymb}
+\usepackage{times}
+\usepackage{caption}
+\usepackage{epsfig}
+\usepackage{subfigure}
+\usepackage{color}
+\usepackage{rotate}
+\usepackage{rotating}
+\usepackage{multirow}
+\usepackage{amsthm}
+
+\usepackage{relsize}
+\usepackage{fancyvrb}
+%\usepackage{times}
+\usepackage{latexsym}
+
+\usepackage{utf8}
+\setarab
+\fullvocalize
+%\transtrue
+\arabtrue
+
+\setlength\titlebox{6.5cm}    % Expanding the titlebox
+
+%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
+\title{Case based Arabic Morphological Analysis with 
+    Compositional Non-deterministic Automata }
+
+%\author{ Jad Makhlouta \\
+%  \And
+%Hamza Harkous \\
+%  American University of Beirut \\
+%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
+%  \And 
+%  Fadi Zaraket 
+%}
+
+\date{}
+
+\begin{document}
+\maketitle
+
+\begin{abstract}
+Natural language processing uses morphological 
+analysis to abstract and annotate text.
+Often times, and in particular in the context
+of the Arabic language, annotations resulting from 
+an ongoing branch of morphological analysis may not be
+appropriate for the case study at hand. 
+In this paper we present Sarf, a case based morphological 
+analyser for Arabic text. 
+Sarf allows a case-based controller to control and refine
+the state of the analysis on the fly. 
+It uses compositional non-deterministic finite state machines 
+to analyse a stream of text. 
+Each alive machine corresponds to a valid analysis. 
+Sarf uses concatenative analysis based on recursive affixes 
+to better preserve part of speech information.
+We automated the analysis of three books of Islamic
+narrations (hadith) using Sarf where we
+abstracted each book into a vector of complex structures.
+Our results show high accuracy and efficiency compared to
+state of the art analyzers. 
+\end{abstract}
+
+\section{Introduction}
+
+Automated analysis of Arabic data sets, including texts, 
+publications, records and digital media is essential
+with the huge digital Arabic content available nowadays. 
+Morphological analysis is key in current automated 
+analysis techniques for Arabic text. 
+Current morphological analyzers~\cite{Sughaiyer:04}
+use concatenative analysis
+to analyze an Arabic word and
+consider its internal structure composed of several
+{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
+An affix can be a {\em prefix, suffix, } or an {\em infix}.
+The analysis of one word may lead to several possible
+morphological solutions.
+\vocalize
+For instance the word \RL{'a.hmadH} 
+may have two valid morphological analyses where in the first
+\RL{'a} is a prefix and the word means ``I praise him'',
+and in the second \RL{'a} is part of the stem 
+\RL{'a.hmad} (a proper noun)
+and the word means ``his Ahmad''.
+
+\novocalize
+The solutions suffer from accuracy due to inherent difficulties
+of morphological analysis of the Arabic language. 
+For example, it is common practice to write Arabic text
+without short vowels. 
+Arabic letters can also have up to four different shapes 
+corresponding to their position in a word, i.e, begining of 
+a word, middle of a word, end of word, and separated. 
+This allows two consecutive words such as 
+\RL{almdrsT}\RL{al_A}
+to be visually recognizable
+as two separate words \RL{al_A} and \RL{almdrsT} without 
+the need of a delimiter such as a space in between. 
+This specifically happens when the first word ends with
+a non-connecting letter and is referred to as the ``run-o'' words
+problem~\cite{Buckwalter:04}.
+Such forms occur in text and greatly increase the
+difficulty of tokenization.
+
+Current morphological analyzers such as 
+Buckwalter~\shortcite{Buckwalter:02},
+Xerox~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+and ElixirFM~\cite{Otakar:07} exist.
+They take as input white space delimited tokens~\cite{Kulick:10},
+consider them as words,
+and enumerate all possible solutions. 
+This exhaustive enumeration may hurt performance and may
+not be necessary or appropriate
+in some case studies as noted in~\cite{Maamouri:10}. 
+Moreover, a white space delimited token may have 
+more than one word.
+Other morphological analyzers such as Amira~\cite{Diab:07,Benajiba:07},
+MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
+also use machine learning and support vector machines (SVM) 
+to enhance the accuracy of the morphological analysis at the expense 
+of perfomance.
+
+We hypothesise that many NLP case studies need the 
+morphological analyser to answer simple queries that do not need 
+high complexity and accuracy at the low morphological analysis level 
+and that can often compensate for tolerable inaccuracy at higher level. 
+For example, if the query is interested in proper names and the 
+prefix in question in the analysis can only connect to a verb, 
+we can provide an answer without completing the analysis. 
+%TODO A lso if a word such as \RL{wasiim} was name was ambiguous and we were looking for a sequence of names, then 
+We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
+addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
+with features demanding resolution at an abstraction level
+higher than the text itself
+led to a refined analyzer~\cite{Kulick:10}.  
+We also find evidence in~\cite{Habash:06} that different types of 
+morphological analyses behave differently on the same case study. 
+We strongly beleive and we find evidence in our results that a 
+case-based controller intervening at every decision is necessary to 
+guide, use, prune, and refine the morphological analysis.
+
+\subsection{Sarf}
+\label{sec:intro:sarf}
+
+In this paper, we present Sarf, a {\em novel case-based efficient
+morphological analyzer} that uses parallel compositional 
+non-deterministic automata driven by a case based controller.
+Each alive machine in Sarf represents a valid morphological analysis. 
+Sarf keeps alive all possible analyses of the text and gives 
+opportunity to the controller to intervene at every single input 
+character. 
+The decisions of the case-based controller can thus prune false 
+positives early in the run. 
+Each alive machine in Sarf takes as input one character at a time 
+from a text stream. 
+Sarf does not assume that the word at hand is a token and
+performs tokenization on the fly based on morphological correctness.
+To out knowledge, this is the first morphological analyzer that 
+handles the ``run-on'' words problem.
+
+Sarf uses a {\em recursive} concatenative analysis with a novel 
+refinement. 
+Sarf introduces recursive affixes in order to
+retain better part of speech information and enhance the 
+efficiency of affix matching.
+
+We validate our hypothesis using a case study from the Islamic 
+literature. 
+A {\em hadith} is a narration related to the prophet Mohamad
+through a {\em sanad} or a sequence of narrators. 
+The authentication of a hadith higly depends on the credibility
+of each of the narrators as reported in separate biography 
+books. 
+In this paper we consider the problem of automatically segmenting
+a hadith book into narrations, then segmenting each narration into
+its content or {\em matn} and its sanad.
+We also would like to partition the sanad accurately into the 
+separate narrators so that we can later look each one of them 
+up in the biography books.
+
+The hadith case study is interesting for Sarf since the controller
+considers most of the analyses where we have a concentration
+of proper names in the text, and ignores most of the analyses as long 
+as they are valid where we do not have proper names. 
+The controller is also interested in a small subset of words 
+that relate persons to each other such as \RL{bin} or ``the sun of''
+or words that mean narrate such as \RL{qaal} or ``said''. 
+With Sarf, we successfully automated the analysis of 
+three books of Islamic narrations~\cite{narrations1,narrations2, narrations3}
+and extracted from each one of them a vector of a three level deep
+complex structure.
+
+%\subsection{Contributions}
+
+We make the following key contributions. 
+\begin{enumerate}
+\item {\bf Case-based analysis:}  We deisgned and implemented
+a case-based morphological analyzer where a case-based controller
+machine controls and guides the analysis. 
+\item {\bf Exhaustive analysis:} Since we have a case-based 
+controller that can decide on the fly whether an analysis is 
+accepted or not, we can afford to keep all valid analyses modulo
+those pruned by the case controller. 
+We do that using non-deterministic FSAs. 
+\item {\bf Recursive affixes:}
+We refine the concatenative analysis of 
+Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
+This allowed Sarf to retain better compositional part of speech
+tags.
+\item {\bf Hadith case study:}
+We evaluated Sarf using the hadith case study. We were able to
+efficiently extract the sequence of narrators into a complex
+data structure with three levels of hierarchy with high accuracy. 
+\end{enumerate}
+
+% does not have to be a subsection, 
+%just to highlight it as an item we should not miss
+
+% paper structure
+
+\section{Background}
+
+\section{Related Work }
+
+Many morphological analyzers exist~\cite{Sughaiyer:04}. 
+Some use specialized part of speech tag sets~\cite{Khoja:01,Darwish:02}.
+
+Buckwalter: used lexicon, added proper nouns, word based, recursive affixes
+
+MAGEAD
+
+Beeseley: many compiled FSAs compared to compositions of three FSAs
+
+SAMA: word based, treebank specific,
+
+MADA+TOKAN: no controller, we do not need a tokenizer
+
+AMIRA: machine learning...
+
+ElixirFM:
+
+\section{Sarf}
+
+\subsection{Recursive Affixes}
+% smaller FSM
+% retain POS info
+
+\subsection{Motivating Example}
+% the FSMs for the three words
+To help illustrate the strength of the method we imploy, we clarify it by walking through an example explaining the stages that lead to morphological analysis in SARF.
+Figure \ref{f:examples} shows a partial snapshot of the FSM's needed to analyze the following words: \RL{wsyal`bhaa}, \RL{syalbisohomaa} and \RL{al-laa`bown}.
+
+\begin{figure*}[tb]
+\center{
+\resizebox{1.8\columnwidth}{!}
+{ \input{psfigs/FSM.pstex_t}}
+\caption{Example FSM.}
+\label{f:examples}
+}
+\end{figure*}
+
+\subsection{Affix Linear FSM}
+
+\subsection{Stem Linear FSM}
+
+\subsection{Non-deterministic Composition of FSMs}
+% algorithm
+% diagram of abstract machines
+\begin{figure*}[tb]
+\center{
+\resizebox{1.8\columnwidth}{!}
+{ \input{psfigs/abstract_machine.pstex_t}}
+\caption{Abstract FSM.}
+\label{f:abstract}
+}
+\end{figure*}
+
+\section{Islamic Literature Case Study}
+
+\subsection{Hadith Segmentation}
+
+\subsection{Chain of Narrators}
+
+\subsection{Controller}
+% FSM for the controller
+
+\section{Results}
+\label{sec:results}
+
+\subsection{Efficiency Comparison Against Buckwalter and SAMA}
+
+\subsection{Case Study Accuracy and Efficiency Results}
+
+% accuracy measure:
+% on names: a word is a name or not
+% chains: narrator is missing from a chain
+% segmentation: two hadiths are merged
+
+\section{Future Work}
+
+%For reasons of uniformity, Adobe's {\bf Times Roman} font should be
+%used. In \LaTeX2e{} this is accomplished by putting
+
+%\begin{quote}
+%\begin{verbatim}
+%\usepackage{times}
+%\usepackage{latexsym}
+%\end{verbatim}
+%\end{quote}
+%in the preamble.
+
+%Additionally, it is of utmost importance to specify the {\bf
+%  US-Letter format} (8.5in $\times$ 11in) when formatting the paper.
+%When working with {\tt dvips}, for instance, one should specify {\tt
+%  -t letter}.
+
+%{\bf Citations}: Citations within the text appear
+%in parentheses as~\cite{Gusfield:97} or, if the author's name appears in
+%the text itself, as Gusfield~\shortcite{Gusfield:97}. 
+%Append lowercase letters to the year in cases of ambiguities.  
+%Treat double authors as in~\cite{Aho:72}, but write as 
+%in~\cite{Chandra:81} when more than two authors are involved. 
+%Collapse multiple citations as in~\cite{Gusfield:97,Aho:72}.
+
+\section*{Acknowledgments}
+
+\bibliographystyle{naaclhlt2010}
+\bibliography{fzAr}
+
+\end{document}
+=======
 %

### 2010-06-25 trunk/src/ATSarf/test-cases/hadith.h

@@ -8,7 +8,7 @@
 //#define STATS
 //#define HADITHDEBUG  //just for additional debugging statements displayed
 //#define TENTATIVE //are decisions that we are not sure about
-//#define REFINEMENTS
+#define REFINEMENTS
 int hadith(QString input_str,ATMProgressIFC *prgs);
 void hadith_initialize();
 

### 2010-06-25 tex/sarf.tex
added algorithm and fig

@@
 
+% nondeterministic finite-state automata
+
 \section{Related Work }
 
-Many morphological analyzers exist~\cite{Sughaiyer:04}. 
-Some use specialized part of speech tag sets~\cite{Khoja:01,Darwish:02}.
+The survey work in~\cite{Sughaiyer:04} discusses and compares
+several morphological analyzers. 
+Analyzers such as~\cite{Khoja:01,Darwish:02} 
+target specific specific applications in the morphological 
+analyzer itself or use a specific set of part of speech tags
+as their reference.
+Sarf differs in that it is a general and rich morphological 
+analyzer that reports all possible solutions. 
+It is case-based in the sense that a controller that drives
+the morphological analyzers prunes these decisions. 
+Up to our knowledge, Sarf is the only morphological analyzer
+that addresses the ``run-on'' words problem.
 
-Buckwalter: used lexicon, added proper nouns, word based, recursive affixes
+Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
+It differs than Buckwalter where it defines a shorter list of affixes
+and a longer list of affix compatibility rules to allow recursive 
+affixes so that we can have prefix-prefix an suffix-suffix 
+concatenations.
+This allows Sarf to better maintain and propagate 
+the part of speech information
+tags associated with each prefix while these need to be highly 
+redundant with Buckwalter and that may lead to consistency issues. 
+Sarf keeps the possible analyses alive as a sequence corresponding
+to the text. Buckwalter produces a set of solutions for each token 
+and the several analysis threads of the whole text need to be 
+generated as a product of the token solutions.
 
-MAGEAD
+Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
+finiste-state automata (FSA). 
+Beesley reports that the number of the automata generated by a compiler
+from Xerox rules is not controlled and also reports a difficulty to 
+compose the FSAs into a single FSA framework. 
+Sarf solves the problem with a non-deterministic composition of three
+FSAs that are the prefix, stem, and suffix FSAs. 
+With Sarf it is easy to control the FSAs with a user custom case 
+controller, while with Beesley the effort requires recompiling the
+Xerox rules.
 
-Beeseley: many compiled FSAs compared to compositions of three FSAs
+Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
+the several valid morphological analyses as a richness that 
+should be exploited at a higher abstract level rather than
+an inaccuracy that should be corrected. 
+Sarf differs in that it provides an interface for the 
+case study to interfere in judging the analyses much earlier and
+at a finer granularity level. 
+Also in Sarf there is no need for a separate tokenizer such as
+TOKAN as each alive automata carries with it the positions in 
+text where a composition decision was made. 
+The case-baded controller can build the trace when needed.
 
-SAMA: word based, treebank specific,
+%MAGEAD 
+    %Maamouri, Mohamed and \bgroup et al.\egroup },
 
-MADA+TOKAN: no controller, we do not need a tokenizer
+We strongly feel that the refinemet of SAMA~\cite{Maamouri:10} 
+is the closest to our approach. SAMA was refined to interact with
+the ATB~\cite{Maamouri:04} project after the addition of a large 
+new corpus. We find evidence in this decision that our hypothesis
+about the utility of case-based analyzers. The refinements
+that resulted in algorithmic changes in SAMA were done manually
+and worked in integration with the ATB format only. 
+In Sarf we propse a case-based controller approach that can modify 
+the behaviour of the analyzer on the fly and that should interact
+with the case study. In Sarf the analyzer remains general, and the 
+case-based controller specializes the analyses.
 
-AMIRA: machine learning...
+The AMIRA~\cite{Diab:07} analyzer uses 
+a language independent SVM based analysis. 
+While the SVM analysis learns features from the text passed to the case
+study, it does not make use of the type oh the query the 
+case study targets in the text.
 
-ElixirFM:
+Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
+of the Buckwalter analyzer. 
+We also use deterministic parsing with tries or our analysis 
+to run the prefix, suffix and stem FSAs. 
+We think that their inferential-realizational approach 
+that is highly compatible with the Arabic linguistic 
+description~\cite{Badawi:04}
+can make benefit of many features unique to the Arabic language.
+Sarf leaves implementing that to the case-based controller
+since in many cases the abstraction needed may not need the 
+text to be grammar correct.
+We can compare the case-based controller in Sarf to a high level
+component in the ElixirFM domain sepcific language.
 
 \section{Sarf}
+\label{sec:sarf}
 
-\subsection{Recursive Affixes}
-% smaller FSM
-% retain POS info
+Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with 
+proper names and location names extracted from different online 
+sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
+\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
+as well as biblical sources~\cite{}.
 
-\subsection{Motivating Example}
-% the FSMs for the three words
+\subsection{Sarf example }
+\label{sec:example}
 
@@
 { \input{psfigs/FSM.pstex_t}}
-\caption{Example FSM.}
-\label{f:examples}
+\caption{Example composition of the prefix, stem and suffix FSAs}
+\label{f:example}
 }
 \end{figure*}
+% the FSMs for the three words
+
+The compositional FSA diagram in Figure~\ref{f:example}
+illustrates the operation of Sarf when parsing the
+string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
+\footnote{ 
+\RL{n}%
+\RL{w}
+\RL{b}
+\RL{`}
+\RL{a} 
+\RL{l}
+\RL{l}
+\RL{a} %
+\RL{a}
+\RL{h}
+\RL{b}
+\RL{`}
+\RL{l}
+\RL{y}
+\RL{s}
+\RL{w}
+in separate form to ease following the example
+}
+The square and circle legends are states of the FSA
+and the directed edges are transitions driven by an input
+character that matches the label of the edge. 
+The square denotes an accept state, 
+the circle denotes a regular state, and the reject states
+are omitted for clarity. 
+Figure~\ref{f:example}a represents the prefix FSA and 
+Figures~\ref{f:example}b and~\ref{f:example}c represent the 
+stem and suffix FSAs respectively. 
+The symbol $\epsilon$ represents an empty string and is 
+the source of non-determinism in this example.
+
+We start at the root square state in the prefix FSA
+which is an accept state. 
+The edge labeled $\epsilon$ that connects the prefix 
+FSA to the stem FSA can transition from any accept state
+in the prefix FSA to the root state in the stem FSA.
+The edge that connects the stem FSA to the suffix FSA
+follows the same behaviour.
+
+When there are two valid transitions such as \RL{w} 
+and $\epsilon$ as is the start case, a non-deterministic 
+FSA $\Phi$ spawns an exact copy of itself $\Psi$. FSA $\Phi$
+makes one transition, and FSA $\Psi$ makes the other. 
+Each of the FSAs now represent a valid analysis so far. 
+When the FSA is at a state where the next input character
+leads to a reject state, the FSA dies.
+The reject states are omitted from Figure~\ref{f:example}
+and are denoted by the abscence of the corresponding 
+characters. 
+In our examples, if we assume there were no stems that start 
+with the letter \RL{w}, $\Psi$ will die then. 
+In reality probably $\Psi$ will die when the input string
+is at \RL{wsyl`}.
+
+Note that the prefix and affix FSAs allow recursive 
+prefixes. For example, \RL{w} will result in an accept state
+that transitions to the stem FSA. 
+When \RL{s} follows, we move to another accept state in the 
+prefix FSA corresponding to the suffix \RL{ws}. 
+The same applies to the suffix FSA.
+
+Lets consider FSA $\Phi$ after it consumed \RL{wsy} 
+and transitioned into the stem FSA root node. 
+Now $\Phi$ will transition with \RL{l`b} to reach an accept 
+state. 
+Before moving with the letter \RL{b} to the accept state,
+$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
+with the prefix \RL{wsy}. 
+We do that by associating a category value for each accept state. 
+We consider the value of the category as part of the state.
+Thus each state shown in Figure~\ref{f:example} actually represents
+more than one state. 
+If the category of \RL{l`b} is compatible with the category of
+\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
+Otherwise, it will move it to a regular or a reject state.
+
+Since $\Phi$ is now in an accept state in the stem FSA, it will
+spawn $\Sigma$ that transitions into the root of the 
+suffix FSA. When \RL{h} is consumed, $\Phi$ will die since 
+there is no edge from the current state labeled with \RL{h}. 
+Now $\Sigma$ represents a valid analysis and can proceed.
+
+Note that $\Sigma$ can report a full word at any accept state
+now by spawning a new automata using the $\epsilon$ transition
+to the prefix FSA.
+This is refined in Sarf to happen only with white space, delimiters, 
+or non-connecting characters. 
+We left that out in Figure~\ref{f:example} for clarity.
+
+Finally, The case-based controller can interfere at any point in the 
+analysis to make a decision like ignoring an automata that 
+is not interesting. 
+It may also decide to correct the analysis of one automata
+based on the decision of the others.
+
+\subsection{Recursive Affixes}
+% smaller FSM
+% retain POS info
 
@@
 \subsection{Non-deterministic Composition of FSMs}
+\label{sec:ndfsa}
+
+\begin{figure*}[tb]
+\center{
+\resizebox{1.8\columnwidth}{!}
+{ \input{psfigs/abstract_machine.pstex_t}}
+\caption{An abstraction of the non-deterministic composition of the prefix, stem and suffix FSAs}
+\label{f:example}
+}
+\end{figure*}
+
 % algorithm
 % diagram of abstract machines
+
+\begin{Verbatim}[fontsize=\relsize{-1},
+frame=topline,framesep=4mm,label=\fbox{NDFSA Sarf algorithm},
+commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
+Sarf(string $L$, Controller $A$) 
+  FSAVec $M$; -- collection of FSA automata
+  FSA $\Phi_1$;
+  $M$.insert($\Phi_1$);
+  foreach $c$ in $L$ \{
+    foreach $\Phi$ in $M$ \{
+      $A$.control($M$, $c$);
+      if ( $\Phi$.state.isAccept() ) \{
+        if ($\Phi$.isSuffix())
+          if ($c$.isWhiteORNonConnecting())
+            $A$.report();
+        $\Psi$ = $\Phi$.clone();
+        $M$.insert($\Psi$); \}
+
+      if ( $\Phi$.isWalkable($c$) ) \{
+        $\Phi$.transition($c$);
+      \} else \{
+        $M$.remove($\Phi$);
+        $\Phi$.die();
+      \} \} \}
+\end{Verbatim}
 

### 2010-06-25 tex/sarf.tex
added sarf section

@@
 \arabtrue
+
+\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
+\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
 
@@
 
-\section{Background}
-
-% nondeterministic finite-state automata
+%\section{Background}
+% TODO: nondeterministic finite-state automata
 
 \section{Related Work }
+\label{sec:related}
 
@@
 \href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
-as well as biblical sources~\cite{}.
+as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
 
-\subsection{Sarf example }
+Sarf encodes the prefix, suffix and stem lexicons into 
+three linear FSAs and uses a non-deterministic composition
+of the three FSAs to compute all valid morphological analyses.
+We first present a runing example of Sarf. 
+The running example explains
+how the non-deterministic composition works and 
+illustrates how Sarf implements recursive affixes and
+handles ``run-on'' words.
+Then we discuss how Sarf builds and composes the different FSAs.
+
+\subsection{Sarf example}
 \label{sec:example}
@@
 \center{
-\resizebox{1.8\columnwidth}{!}
+\resizebox{1.9\columnwidth}{!}
 { \input{psfigs/FSM.pstex_t}}
-\caption{Example composition of the prefix, stem and suffix FSAs}
+\caption{Example composition of the prefix, stem and suffix FSAs.}
 \label{f:example}
@@
 \subsection{Recursive Affixes}
-% smaller FSM
-% retain POS info
+\label{sec:recaffix}
 
-\subsection{Affix Linear FSM}
+Morphological analysis partitions a given Arabic word
+into a set of morphemes.
+A morpheme is either an affix or a stem. 
+We call two morphemes compatible if their concatenation
+forms a legal morpheme. 
+Buckwalter~\cite{Buckwalter:02} keeps separate lists 
+of prefixes, suffixes and stems and assigns a category
+to each morphemes. 
+The relation between morphemes and categories is one 
+to many.
 
-\subsection{Stem Linear FSM}
+A compatibility rule is a pair of categories 
+$\langle c_1, c_2\rangle$  stating that morphemes
+with category $c_1$ can be concatenated with morphemes
+of category $c_2$. 
+Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
+of compatibility rules relating
+prefixes to stems, stems to suffixes, and prefixes to suffixes
+respectively. 
+Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
+a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
+The string $\alpha\beta\gamma$ partition of the string is a 
+valid morphological analysis if  and only if we have
+$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
+$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
+$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$ where
+$c()$ returns the category of the morpheme.
+
+Many affixes are composed of other affixes. For example,
+the prefix \RL{wsy} is composed of three other prefixes
+namely \RL{w}, \RL{s}, and \RL{y}.
+Moreover, the POS tag for \RL{wsy} is a concatenation
+of the POS tags of each of the three morphemes. 
+Keeping these morphemes separate in a list
+is redundant and may produce consistency issues. 
+We add two lists of compatibilty rules $L_{pp}$ and
+$L_{xx}$ for prefix-prefix and suffix suffix compatibility
+respectively.
+This allows us to keep a shorter list of affixes. 
+We need one more additional modification to accomodate 
+a category for the morphemes accepted by the $L_{pp}$
+and the $L_{xx}$ rules. For that we introduce the concept
+of a {\em resulting category}. 
+The resulting category in all practical cases
+was the category of the second morpheme in the pair.
+
+\subsection{Affix Linear FSA}
+\label{sec:affixFSA}
+
+We analyzed the prefixes and suffixes of 
+Buckwalter~\cite{Buckwalter:02}
+and extracted our refined lists of recursives affixes.
+We encoded the lists into a FSA similar to that
+in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
+The two FSAs are directed acyclic graphs. 
+The abscence of cycles in these FSAs reduces the 
+traversal of the affix FSAs to a linear traversal.
+This is computationaly superior to the 
+approach of Buckwalter where the analyzer considers
+all possible substrings of the word in question
+and looks it up in the affix tables.
+
+The prefix and suffix FSAs encode the $L_{pp}$ and
+$L_{xx}$ compatibility lists.
+An accept state in the FSA corresponds to the last character
+of a prefix in the predix and suffix lists.
+The accept states hold category state information as well
+as POS and other tags.
+
+\subsection{Stem Linear FSA}
+\label{sec:stemFSA}
+
+We built our stem lexicon using the stem lexicon of 
+Buckwalter. 
+We augmented the lexicon with a set of proper names and
+a set of location names. 
+We obtained the set of proper names from online 
+and biblical sources.
+
+The stems share lots of substrings. We encoded them in
+an efficient double array trie structure~\cite{Aoe:89}. 
+The structure is also a linear FSA where the accept
+states are the terminal nodes corresponding to the last 
+letters in stems. 
+This approach is superior to Buckwalter since again it is
+a linear walk in the trie while with Buckwalter it is 
+several hash lookups.
 
@@
 \label{sec:ndfsa}
+
+The diagram in Figure~\ref{f:composition} shows an 
+abstraction of the non-deterministic composition 
+of the prefix, stem and suffix FSA. 
+The circle represents regular states, the square
+represents accept state and the triangle represents
+reject states. 
+Note that the concrete prefix, stem and suffix FSAs
+are all linear and do not have cycles as discussed 
+is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
+The cycles are introduced in the abstract machine
+for brevity and illustrative purposes.
 
@@
 { \input{psfigs/abstract_machine.pstex_t}}
-\caption{An abstraction of the non-deterministic composition of the prefix, stem and suffix FSAs}
-\label{f:example}
+\caption{An abstraction of the non-deterministic composition of 
+    the prefix, stem and suffix FSAs.}
+\label{f:composition}
 }
@@
 
-% algorithm
-% diagram of abstract machines
+The algorithm \CodeIn{NDSarf} below controls Sarf. 
+It takes as input a text string $L$ and a case-based controller
+$A$. 
+It starts with one FSA $\Phi_1$ similar to the one in 
+Figure~\ref{f:composition} and inserts it into a collection
+of FSAs $M$. 
+The algorithm reads a character $c$ at a time from $L$
+and moves on all the machines in $M$. 
+For each $\Phi$ in $M$ the algorithm checks whether
+$\Phi$ is in an accept state or not.
 
+If $\Phi$ is in an accpet state and it is in the suffix
+phase, then the algorithm checks whether it should report
+a full word. 
+This happens when $c$ is a white space or a delimiter, 
+or when the last character leading to the current state
+was a non-connecting character.
+
+In all cases, if $\Phi$ is in an accept state, 
+it spawns another FSA $\Psi$ and adds it to $M$. 
+If $c$ leads to a reject state, then $\Phi$ dies 
+and we remove it from $M$. 
+Otherwise, $\Phi$ transitions using the $c$ edge.
+We invoke the controller $A$ after all the FSAs in $M$ 
+have moved and allow it full access and control over
+all FSAs in $M$.
+
+The arrows in Figure~\ref{f:composition} that transition
+from prefix to stem, stem to suffix, and suffix to prefix
+show the compatibility constraints as a condition for 
+the transition.
+
+\begin{table}[tb]
+\centering
+\begin{tabular} {p{6cm}}
 \begin{Verbatim}[fontsize=\relsize{-1},
-frame=topline,framesep=4mm,label=\fbox{NDFSA Sarf algorithm},
+frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
 commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
-Sarf(string $L$, Controller $A$) 
+NDSarf(string $L$, Controller $A$) 
   FSAVec $M$; -- collection of FSA automata
@@
     foreach $\Phi$ in $M$ \{
-      $A$.control($M$, $c$);
       if ( $\Phi$.state.isAccept() ) \{
         if ($\Phi$.isSuffix())
-          if ($c$.isWhiteORNonConnecting())
+          if ($c$.isWhite()) or 
+            $A$.report();
+          if ($\Phi$.lastChar().isNonConnecting())
             $A$.report();
@@
         $\Phi$.die();
-      \} \} \}
+      \} \} 
+    $A$.control($M$, $c$); \}
 \end{Verbatim}
+\end{tabular}
+\label{a:ndsarf}
+\end{table}
 
 \section{Islamic Literature Case Study}
+\label{sec:islamic}
 
-\subsection{Hadith Segmentation}
+A hadith in Islamic literature is a narration from the prophet Mohamad
+related by multiple narrators.
+Establishing the authenticity of a hadith is an important task
+in Islamic studies. 
+%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
+%the analysis of the hadith literature. 
+In this paper we use Sarf to sucessfully automate the
+analysis of three books of hadith selected 
+arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
+  the digitized books from online sources such as 
+  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
+  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. 
+  Our understanding is that those books are historical documents with 
+  public IP.}.
 
-\subsection{Chain of Narrators}
+Our analysis accpets a book as a text stream input
+and segements it into a vector of individual hadiths. 
+We segment each hadith into two parts. The first part
+is the sanad and it
+contains the chain of narrators who related the hadith
+to Mohamad. 
+The second part is the matn and it constitutes the content
+of the hadith. 
+We are concerned in further exploring the sanad. 
+In particular, we detect the chain of narrators
+and the relation between each narrator, his ancestor, and 
+his predecessor in the chain. 
+We also detect the full name of each narrator that is
+usually composed of several proper names with connectors
+in between.
+
+We built our case-based controller to target
+the detection of proper names. 
+This includes the task of finding compound names 
+composed of several words such as \RL{`bd alr.hmn} often
+appearing as ``run-on'' words.
+The controller also targets words that mean narrate when
+they appear in the neighborhood of multiple proper names.
 
@@
 % FSM for the controller
+The FSA in Figure~\ref{f:hadith}
 

### 2010-06-25 tex/sarf.tex
spell check

@@
 In this paper we present Sarf, a case based morphological 
-analyser for Arabic text. 
+analyzer for Arabic text. 
 Sarf allows a case-based controller to control and refine
@@
 It uses compositional non-deterministic finite state machines 
-to analyse a stream of text. 
+to analyze a stream of text. 
 Each alive machine corresponds to a valid analysis. 
@@
 For instance the word \RL{'a.hmadH} 
-may have two valid morphological analyses where in the first
+may have two valid morphological analyzes where in the first
 \RL{'a} is a prefix and the word means ``I praise him'',
@@
 Arabic letters can also have up to four different shapes 
-corresponding to their position in a word, i.e, begining of 
+corresponding to their position in a word, i.e, beginning of 
 a word, middle of a word, end of word, and separated. 
@@
 to enhance the accuracy of the morphological analysis at the expense 
-of perfomance.
+of performance.
 
-We hypothesise that many NLP case studies need the 
-morphological analyser to answer simple queries that do not need 
+We hypothesize that many NLP case studies need the 
+morphological analyzer to answer simple queries that do not need 
 high complexity and accuracy at the low morphological analysis level 
@@
 morphological analyses behave differently on the same case study. 
-We strongly beleive and we find evidence in our results that a 
+We strongly believe and we find evidence in our results that a 
 case-based controller intervening at every decision is necessary to 
@@
 morphological analyzer} that uses parallel compositional 
-non-deterministic automata driven by a case based controller.
+non-deterministic Automata driven by a case based controller.
 Each alive machine in Sarf represents a valid morphological analysis. 
@@
 literature. 
-A {\em hadith} is a narration related to the prophet Mohamad
+A {\em hadith} is a narration related to the prophet Mohammad
 through a {\em sanad} or a sequence of narrators. 
-The authentication of a hadith higly depends on the credibility
+The authentication of a hadith highly depends on the credibility
 of each of the narrators as reported in separate biography 
@@
 \begin{enumerate}
-\item {\bf Case-based analysis:}  We deisgned and implemented
+\item {\bf Case-based analysis:}  We designed and implemented
 a case-based morphological analyzer where a case-based controller
@@
 Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
-finiste-state automata (FSA). 
-Beesley reports that the number of the automata generated by a compiler
+finite-state Automata (FSA). 
+Beesley reports that the number of the Automata generated by a compiler
 from Xerox rules is not controlled and also reports a difficulty to 
@@
 Also in Sarf there is no need for a separate tokenizer such as
-TOKAN as each alive automata carries with it the positions in 
+TOKAN as each alive Automata carries with it the positions in 
 text where a composition decision was made. 
-The case-baded controller can build the trace when needed.
+The case-based controller can build the trace when needed.
 
@@
 
-We strongly feel that the refinemet of SAMA~\cite{Maamouri:10} 
+We strongly feel that the refinement of SAMA~\cite{Maamouri:10} 
 is the closest to our approach. SAMA was refined to interact with
@@
 and worked in integration with the ATB format only. 
-In Sarf we propse a case-based controller approach that can modify 
-the behaviour of the analyzer on the fly and that should interact
+In Sarf we propose a case-based controller approach that can modify 
+the behavior of the analyzer on the fly and that should interact
 with the case study. In Sarf the analyzer remains general, and the 
@@
 We can compare the case-based controller in Sarf to a high level
-component in the ElixirFM domain sepcific language.
+component in the ElixirFM domain specific language.
 
@@
 of the three FSAs to compute all valid morphological analyses.
-We first present a runing example of Sarf. 
+We first present a running example of Sarf. 
 The running example explains
@@
 The edge that connects the stem FSA to the suffix FSA
-follows the same behaviour.
+follows the same behavior.
 
@@
 The reject states are omitted from Figure~\ref{f:example}
-and are denoted by the abscence of the corresponding 
+and are denoted by the absence of the corresponding 
 characters. 
@@
 Note that $\Sigma$ can report a full word at any accept state
-now by spawning a new automata using the $\epsilon$ transition
+now by spawning a new Automata using the $\epsilon$ transition
 to the prefix FSA.
@@
 Finally, The case-based controller can interfere at any point in the 
-analysis to make a decision like ignoring an automata that 
+analysis to make a decision like ignoring an Automata that 
 is not interesting. 
-It may also decide to correct the analysis of one automata
+It may also decide to correct the analysis of one Automata
 based on the decision of the others.
@@
 is redundant and may produce consistency issues. 
-We add two lists of compatibilty rules $L_{pp}$ and
+We add two lists of compatibility rules $L_{pp}$ and
 $L_{xx}$ for prefix-prefix and suffix suffix compatibility
@@
 This allows us to keep a shorter list of affixes. 
-We need one more additional modification to accomodate 
+We need one more additional modification to accommodate 
 a category for the morphemes accepted by the $L_{pp}$
@@
 Buckwalter~\cite{Buckwalter:02}
-and extracted our refined lists of recursives affixes.
+and extracted our refined lists of recursive affixes.
 We encoded the lists into a FSA similar to that
@@
 The two FSAs are directed acyclic graphs. 
-The abscence of cycles in these FSAs reduces the 
+The absence of cycles in these FSAs reduces the 
 traversal of the affix FSAs to a linear traversal.
-This is computationaly superior to the 
+This is computationally superior to the 
 approach of Buckwalter where the analyzer considers
@@
 An accept state in the FSA corresponds to the last character
-of a prefix in the predix and suffix lists.
+of a prefix in the prefix and suffix lists.
 The accept states hold category state information as well
@@
 
-\subsection{Non-deterministic Composition of FSMs}
+\subsection{Non-deterministic Composition of FSAs}
 \label{sec:ndfsa}
@@
 
-If $\Phi$ is in an accpet state and it is in the suffix
+If $\Phi$ is in an accept state and it is in the suffix
 phase, then the algorithm checks whether it should report
@@
 NDSarf(string $L$, Controller $A$) 
-  FSAVec $M$; -- collection of FSA automata
+  FSAVec $M$; -- collection of FSA Automata
   FSA $\Phi_1$;
@@
 
-A hadith in Islamic literature is a narration from the prophet Mohamad
+A hadith in Islamic literature is a narration from the prophet Mohammad
 related by multiple narrators.
@@
 %the analysis of the hadith literature. 
-In this paper we use Sarf to sucessfully automate the
+In this paper we use Sarf to successfully automate the
 analysis of three books of hadith selected 
@@
 
-Our analysis accpets a book as a text stream input
-and segements it into a vector of individual hadiths. 
+Our analysis accepts a book as a text stream input
+and segments it into a vector of individual hadiths. 
 We segment each hadith into two parts. The first part
@@
 contains the chain of narrators who related the hadith
-to Mohamad. 
+to Mohammad. 
 The second part is the matn and it constitutes the content

### 2010-06-26 tex/sarf.tex
table in

@@
 
+We report our results based on a manual validation of 
+a representative sample selected randomly from the text.
+
+\begin{table*}[bt]
+\centering
+\caption{Results of the hadith case study with Sarf.}
+%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
+\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
+ & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
+ & & Simple & Accurate & & Simple & Accurate & & Simple  & Accurate \\ \hline \hline
+Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
+ Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
+Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
+Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
+Narrator/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
+Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
+Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline \hline
+Segmentation accuracy & & 80\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
+Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
+%Narrator accuracy & & & & & &  & & & \\ 
+Name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
+Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
+Sarf running time & &a & & & & & & & \\ \hline
+\end{tabular}
+\normalsize
+\label{t:hadithresallresults}
+\end{table*}
+
 \subsection{Efficiency Comparison Against Buckwalter and SAMA}
@@
 
-\section*{Acknowledgments}
+%\section*{Acknowledgments}
+% this will go into the regular submission
 

### 2010-06-26 trunk/tex/sarf.tex
structure, correction, and future

--- 2010-06-25 trunk/tex/sarf.tex
@@
 
+\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
+\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
+
 \setlength\titlebox{6.5cm}    % Expanding the titlebox
@@
 abstracted each book into a vector of complex structures.
-Our results show high accuracy and efficiency compared to
-state of the art analyzers. 
+Our results show high accuracy and efficiency.
+%compared to
+%state of the art analyzers. 
 \end{abstract}
@@
 For instance the word \RL{'a.hmadH} 
-may have two valid morphological analyses where in the first
-\RL{'a} is a prefix and the word means ``I praise him'',
-and in the second \RL{'a} is part of the stem 
+may have two valid morphological analyses. In the first
+the letter \RL{'a} is a prefix and the word means ``I praise him'',
+and in the second the letter \RL{'a} is part of the stem 
 \RL{'a.hmad} (a proper noun)
@@
 \novocalize
-The solutions suffer from accuracy due to inherent difficulties
+The accuracy of the solutions suffer due to inherent difficulties
 of morphological analysis of the Arabic language. 
@@
 without short vowels. 
-Arabic letters can also have up to four different shapes 
-corresponding to their position in a word, i.e, begining of 
-a word, middle of a word, end of word, and separated. 
-This allows two consecutive words such as 
+This greatly increases the ambiguity of Arabic text. 
+Another difficulty is that  Arabic letters can have up to 
+four different forms
+corresponding to their position in a word, i.e, beginning,
+middle, end of word forms and separate forms. 
+This allows the phrase 
 \RL{almdrsT}\RL{al_A}
@@
 the need of a delimiter such as a space in between. 
-This specifically happens when the first word ends with
-a non-connecting letter and is referred to as the ``run-o'' words
-problem~\cite{Buckwalter:04}.
-Such forms occur in text and greatly increase the
-difficulty of tokenization.
+This happens because the first word \RL{almdrsT} ends with
+\RL{T} a non-connecting letter. 
+These words occur often in text and greatly increase the
+difficulty of tokenization and are referred to as 
+``run-on'' words~\cite{Buckwalter:04}.
 
@@
 Buckwalter~\shortcite{Buckwalter:02},
-Xerox~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+Beesely~\shortcite{Beesley:01}, SAMA~\cite{Kulick:10},
 and ElixirFM~\cite{Otakar:07} exist.
@@
 and enumerate all possible solutions. 
-This exhaustive enumeration may hurt performance and may
+This approach has several problems. 
+First, 
+a white space delimited token may have 
+more than one word.
+Second, the exhaustive enumeration may hurt performance and may
 not be necessary or appropriate
 in some case studies as noted in~\cite{Maamouri:10}. 
-Moreover, a white space delimited token may have 
-more than one word.
 Other morphological analyzers such as Amira~\cite{Diab:07,Benajiba:07},
@@
 to enhance the accuracy of the morphological analysis at the expense 
-of perfomance.
+of performance.
 
-We hypothesise that many NLP case studies need the 
-morphological analyser to answer simple queries that do not need 
+We hypothesize that many NLP case studies need the 
+morphological analyzer to answer simple queries that do not need 
 high complexity and accuracy at the low morphological analysis level 
-and that can often compensate for tolerable inaccuracy at higher level. 
-For example, if the query is interested in proper names and the 
+and that can often compensate for tolerable inaccuracy at higher levels. 
+For example, if the query concerns proper names and the 
 prefix in question in the analysis can only connect to a verb, 
@@
 morphological analyses behave differently on the same case study. 
-We strongly beleive and we find evidence in our results that a 
-case-based controller intervening at every decision is necessary to 
-guide, use, prune, and refine the morphological analysis.
+We strongly believe and we find evidence in our results that a 
+case-based morphological analyzer that allows a case-based
+controller to intervene at every decision to 
+guide, use, prune, and refine the morphological analysis
+is of high utility.
 
@@
 In this paper, we present Sarf, a {\em novel case-based efficient
-morphological analyzer} that uses parallel compositional 
-non-deterministic automata driven by a case based controller.
+morphological analyzer} that uses compositional 
+non-deterministic Automata driven by a case based controller.
 Each alive machine in Sarf represents a valid morphological analysis. 
@@
 considers most of the analyses where we have a concentration
-of proper names in the text, and ignores most of the analyses as long 
-as they are valid where we do not have proper names. 
-The controller is also interested in a small subset of words 
+of proper names in the text, and ignores most of the analyses 
+where we do not have proper names in text as long 
+as they are valid.
+The controller is also interested in discovering a subset of words 
 that relate persons to each other such as \RL{bin} or ``the sun of''
@@
 
-% paper structure
+The rest of this paper is structured as follows. In Section~\ref{sec:related}
+we compare Sarf to related work. Then in Section~\ref{sec:sarf}
+we discuss Sarf, illustrate the way it works with a running example
+that also explains how non-deterministic finite state automata
+works. In Section~\ref{sec:islamic} we present the hadith case
+study. We present our results in Section~\ref{sec:results}
+and we conclude with future work in Section~\ref{sec:future}.
 
-\section{Background}
+%\section{Background}
+% TODO: nondeterministic finite-state automata
 
 \section{Related Work }
+\label{sec:related}
 
@@
 We can compare the case-based controller in Sarf to a high level
-component in the ElixirFM domain sepcific language.
+component in the ElixirFM domain specific language.
 
 \section{Sarf}
+\label{sec:sarf}
 
@@
 \href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
-as well as biblical sources~\cite{}.
+as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
+
+Sarf encodes the prefix, suffix and stem lexicons into 
+three linear FSAs and uses a non-deterministic composition
+of the three FSAs to compute all valid morphological analyses.
+We first present a running example of Sarf. 
+The running example explains
+how the non-deterministic composition works and 
+illustrates how Sarf implements recursive affixes and
+handles ``run-on'' words.
+Then we discuss how Sarf builds and composes the different FSAs.
+
+\subsection{Sarf example}
+\label{sec:example}
+
+\begin{figure*}[tb]
+\center{
+\resizebox{1.9\columnwidth}{!}
+{ \input{psfigs/FSM.pstex_t}}
+\caption{Example composition of the prefix, stem and suffix FSAs.}
+\label{f:example}
+}
+\end{figure*}
+% the FSMs for the three words
+
+The compositional FSA diagram in Figure~\ref{f:example}
+illustrates the operation of Sarf when parsing the
+string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
+\footnote{ 
+\RL{n}%
+\RL{w}
+\RL{b}
+\RL{`}
+\RL{a} 
+\RL{l}
+\RL{l}
+\RL{a} %
+\RL{a}
+\RL{h}
+\RL{b}
+\RL{`}
+\RL{l}
+\RL{y}
+\RL{s}
+\RL{w}
+in separate form to ease following the example
+}
+The square and circle legends are states of the FSA
+and the directed edges are transitions driven by an input
+character that matches the label of the edge. 
+The square denotes an accept state, 
+the circle denotes a regular state, and the reject states
+are omitted for clarity. 
+Figure~\ref{f:example}a represents the prefix FSA and 
+Figures~\ref{f:example}b and~\ref{f:example}c represent the 
+stem and suffix FSAs respectively. 
+The symbol $\epsilon$ represents an empty string and is 
+the source of non-determinism in this example.
+
+We start at the root square state in the prefix FSA
+which is an accept state. 
+The edge labeled $\epsilon$ that connects the prefix 
+FSA to the stem FSA can transition from any accept state
+in the prefix FSA to the root state in the stem FSA.
+The edge that connects the stem FSA to the suffix FSA
+follows the same behavior.
+
+When there are two valid transitions such as \RL{w} 
+and $\epsilon$ as is the start case, a non-deterministic 
+FSA $\Phi$ spawns an exact copy of itself $\Psi$. FSA $\Phi$
+makes one transition, and FSA $\Psi$ makes the other. 
+Each of the FSAs now represent a valid analysis so far. 
+When the FSA is at a state where the next input character
+leads to a reject state, the FSA dies.
+The reject states are omitted from Figure~\ref{f:example}
+and are denoted by the absence of the corresponding 
+characters. 
+In our examples, if we assume there were no stems that start 
+with the letter \RL{w}, $\Psi$ will die then. 
+In reality probably $\Psi$ will die when the input string
+is at \RL{wsyl`}.
+
+Note that the prefix and affix FSAs allow recursive 
+prefixes. For example, \RL{w} will result in an accept state
+that transitions to the stem FSA. 
+When \RL{s} follows, we move to another accept state in the 
+prefix FSA corresponding to the suffix \RL{ws}. 
+The same applies to the suffix FSA.
+
+Lets consider FSA $\Phi$ after it consumed \RL{wsy} 
+and transitioned into the stem FSA root node. 
+Now $\Phi$ will transition with \RL{l`b} to reach an accept 
+state. 
+Before moving with the letter \RL{b} to the accept state,
+$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
+with the prefix \RL{wsy}. 
+We do that by associating a category value for each accept state. 
+We consider the value of the category as part of the state.
+Thus each state shown in Figure~\ref{f:example} actually represents
+more than one state. 
+If the category of \RL{l`b} is compatible with the category of
+\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
+Otherwise, it will move it to a regular or a reject state.
+
+Since $\Phi$ is now in an accept state in the stem FSA, it will
+spawn $\Sigma$ that transitions into the root of the 
+suffix FSA. When \RL{h} is consumed, $\Phi$ will die since 
+there is no edge from the current state labeled with \RL{h}. 
+Now $\Sigma$ represents a valid analysis and can proceed.
+
+Note that $\Sigma$ can report a full word at any accept state
+now by spawning a new Automata using the $\epsilon$ transition
+to the prefix FSA.
+This is refined in Sarf to happen only with white space, delimiters, 
+or non-connecting characters. 
+We left that out in Figure~\ref{f:example} for clarity.
+
+Finally, The case-based controller can interfere at any point in the 
+analysis to make a decision like ignoring an Automata that 
+is not interesting. 
+It may also decide to correct the analysis of one Automata
+based on the decision of the others.
 
 \subsection{Recursive Affixes}
-% smaller FSM
-% retain POS info
+\label{sec:recaffix}
 
-\subsection{Motivating Example}
-% the FSMs for the three words
+Morphological analysis partitions a given Arabic word
+into a set of morphemes.
+A morpheme is either an affix or a stem. 
+We call two morphemes compatible if their concatenation
+forms a legal morpheme. 
+Buckwalter~\cite{Buckwalter:02} keeps separate lists 
+of prefixes, suffixes and stems and assigns a category
+to each morphemes. 
+The relation between morphemes and categories is one 
+to many.
+
+A compatibility rule is a pair of categories 
+$\langle c_1, c_2\rangle$  stating that morphemes
+with category $c_1$ can be concatenated with morphemes
+of category $c_2$. 
+Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
+of compatibility rules relating
+prefixes to stems, stems to suffixes, and prefixes to suffixes
+respectively. 
+Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
+a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
+The string $\alpha\beta\gamma$ partition of the string is a 
+valid morphological analysis if  and only if we have
+$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
+$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
+$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$ where
+$c()$ returns the category of the morpheme.
+
+Many affixes are composed of other affixes. For example,
+the prefix \RL{wsy} is composed of three other prefixes
+namely \RL{w}, \RL{s}, and \RL{y}.
+Moreover, the POS tag for \RL{wsy} is a concatenation
+of the POS tags of each of the three morphemes. 
+Keeping these morphemes separate in a list
+is redundant and may produce consistency issues. 
+We add two lists of compatibility rules $L_{pp}$ and
+$L_{xx}$ for prefix-prefix and suffix suffix compatibility
+respectively.
+This allows us to keep a shorter list of affixes. 
+We need one more additional modification to accommodate 
+a category for the morphemes accepted by the $L_{pp}$
+and the $L_{xx}$ rules. For that we introduce the concept
+of a {\em resulting category}. 
+The resulting category in all practical cases
+was the category of the second morpheme in the pair.
+
+\subsection{Affix Linear FSA}
+\label{sec:affixFSA}
+
+We analyzed the prefixes and suffixes of 
+Buckwalter~\cite{Buckwalter:02}
+and extracted our refined lists of recursive affixes.
+We encoded the lists into a FSA similar to that
+in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
+The two FSAs are directed acyclic graphs. 
+The absence of cycles in these FSAs reduces the 
+traversal of the affix FSAs to a linear traversal.
+This is computationally superior to the 
+approach of Buckwalter where the analyzer considers
+all possible substrings of the word in question
+and looks it up in the affix tables.
+
+The prefix and suffix FSAs encode the $L_{pp}$ and
+$L_{xx}$ compatibility lists.
+An accept state in the FSA corresponds to the last character
+of a prefix in the prefix and suffix lists.
+The accept states hold category state information as well
+as POS and other tags.
+
+\subsection{Stem Linear FSA}
+\label{sec:stemFSA}
+
+We built our stem lexicon using the stem lexicon of 
+Buckwalter. 
+We augmented the lexicon with a set of proper names and
+a set of location names. 
+We obtained the set of proper names from online 
+and biblical sources.
+
+The stems share lots of substrings. We encoded them in
+an efficient double array trie structure~\cite{Aoe:89}. 
+The structure is also a linear FSA where the accept
+states are the terminal nodes corresponding to the last 
+letters in stems. 
+This approach is superior to Buckwalter since again it is
+a linear walk in the trie while with Buckwalter it is 
+several hash lookups.
+
+\subsection{Non-deterministic Composition of FSAs}
+\label{sec:ndfsa}
+
+The diagram in Figure~\ref{f:composition} shows an 
+abstraction of the non-deterministic composition 
+of the prefix, stem and suffix FSA. 
+The circle represents regular states, the square
+represents accept state and the triangle represents
+reject states. 
+Note that the concrete prefix, stem and suffix FSAs
+are all linear and do not have cycles as discussed 
+is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
+The cycles are introduced in the abstract machine
+for brevity and illustrative purposes.
 
@@
 \resizebox{1.8\columnwidth}{!}
-{ \input{psfigs/FSM.pstex_t}}
-\caption{Example FSM.}
-\label{f:examples}
+{ \input{psfigs/abstract_machine.pstex_t}}
+\caption{An abstraction of the non-deterministic composition of 
+    the prefix, stem and suffix FSAs.}
+\label{f:composition}
 }
@@
 
-\subsection{Affix Linear FSM}
+The algorithm \CodeIn{NDSarf} below controls Sarf. 
+It takes as input a text string $L$ and a case-based controller
+$A$. 
+It starts with one FSA $\Phi_1$ similar to the one in 
+Figure~\ref{f:composition} and inserts it into a collection
+of FSAs $M$. 
+The algorithm reads a character $c$ at a time from $L$
+and moves on all the machines in $M$. 
+For each $\Phi$ in $M$ the algorithm checks whether
+$\Phi$ is in an accept state or not.
 
-\subsection{Stem Linear FSM}
+If $\Phi$ is in an accept state and it is in the suffix
+phase, then the algorithm checks whether it should report
+a full word. 
+This happens when $c$ is a white space or a delimiter, 
+or when the last character leading to the current state
+was a non-connecting character.
 
-\subsection{Non-deterministic Composition of FSMs}
-% algorithm
-% diagram of abstract machines
+In all cases, if $\Phi$ is in an accept state, 
+it spawns another FSA $\Psi$ and adds it to $M$. 
+If $c$ leads to a reject state, then $\Phi$ dies 
+and we remove it from $M$. 
+Otherwise, $\Phi$ transitions using the $c$ edge.
+We invoke the controller $A$ after all the FSAs in $M$ 
+have moved and allow it full access and control over
+all FSAs in $M$.
+
+The arrows in Figure~\ref{f:composition} that transition
+from prefix to stem, stem to suffix, and suffix to prefix
+show the compatibility constraints as a condition for 
+the transition.
+
+\begin{table}[tb]
+\centering
+\begin{tabular} {p{6cm}}
+\begin{Verbatim}[fontsize=\relsize{-1},
+frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
+commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
+NDSarf(string $L$, Controller $A$) 
+  FSAVec $M$; -- collection of FSA Automata
+  FSA $\Phi_1$;
+  $M$.insert($\Phi_1$);
+  foreach $c$ in $L$ \{
+    foreach $\Phi$ in $M$ \{
+      if ( $\Phi$.state.isAccept() ) \{
+        if ($\Phi$.isSuffix())
+          if ($c$.isWhite()) or 
+            $A$.report();
+          if ($\Phi$.lastChar().isNonConnecting())
+            $A$.report();
+        $\Psi$ = $\Phi$.clone();
+        $M$.insert($\Psi$); \}
+
+      if ( $\Phi$.isWalkable($c$) ) \{
+        $\Phi$.transition($c$);
+      \} else \{
+        $M$.remove($\Phi$);
+        $\Phi$.die();
+      \} \} 
+    $A$.control($M$, $c$); \}
+\end{Verbatim}
+\end{tabular}
+\label{a:ndsarf}
+\end{table}
 
 \section{Islamic Literature Case Study}
+\label{sec:islamic}
 
-\subsection{Hadith Segmentation}
+A hadith in Islamic literature is a narration from the prophet Mohammad
+related by multiple narrators.
+Establishing the authenticity of a hadith is an important task
+in Islamic studies. 
+%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
+%the analysis of the hadith literature. 
+In this paper we use Sarf to successfully automate the
+analysis of three books of hadith selected 
+arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
+  the digitized books from online sources such as 
+  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
+  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. 
+  Our understanding is that those books are historical documents with 
+  public IP.}.
 
-\subsection{Chain of Narrators}
+Our analysis accepts a book as a text stream input
+and segments it into a vector of individual hadiths. 
+We segment each hadith into two parts. The first part
+is the sanad and it
+contains the chain of narrators who related the hadith
+to Mohammad. 
+The second part is the matn and it constitutes the content
+of the hadith. 
+We are concerned in further exploring the sanad. 
+In particular, we detect the chain of narrators
+and the relation between each narrator, his ancestor, and 
+his predecessor in the chain. 
+We also detect the full name of each narrator that is
+usually composed of several proper names with connectors
+in between.
+
+We built our case-based controller to target
+the detection of proper names. 
+This includes the task of finding compound names 
+composed of several words such as \RL{`bd alr.hmn} often
+appearing as ``run-on'' words.
+The controller also targets words that mean narrate when
+they appear in the neighborhood of multiple proper names.
 
@@
 \label{sec:results}
+
+We report our results based on a manual validation of 
+a representative sample selected randomly from the text.
+
+\begin{table*}[bt]
+\centering
+\caption{Results of the hadith case study with Sarf.}
+%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
+\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
+ & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
+ & & Simple & Accurate & & Simple & Accurate & & Simple  & Accurate \\ \hline \hline
+Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
+ Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
+Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
+Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
+Narrator/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
+Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
+Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline \hline
+Segmentation accuracy & & 80\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
+Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
+%Narrator accuracy & & & & & &  & & & \\ 
+Name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
+Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
+Sarf running time & &a & & & & & & & \\ \hline
+\end{tabular}
+\normalsize
+\label{t:hadithresallresults}
+\end{table*}
 
@@
 \section{Future Work}
+\label{sec:future}
+
+In the future we plan to complete the hadith case study 
+and automate narrator discovery and authentication in the 
+biography books. 
+We also plan on developing a domain specific language similar 
+that allows users to easily specify their case-based controller
+to interact with Sarf.
 

### 2010-06-26 tex/sarf.tex
submission ready

--- 2010-06-25 tex/sarf.tex
@@
 \vocalize
-For instance the word \RL{'a.hmadH} 
-may have two valid morphological analyzes where in the first
-\RL{'a} is a prefix and the word means ``I praise him'',
-and in the second \RL{'a} is part of the stem 
+For instance, the word \RL{'a.hmadH} 
+may have two valid morphological analyses. 
+The letter \RL{'a} may be prefix and the word means ``I praise him''.
+The letter \RL{'a} may also be part of the stem 
 \RL{'a.hmad} (a proper noun)
@@
 without short vowels. 
-Arabic letters can also have up to four different shapes 
-corresponding to their position in a word, i.e, beginning of 
-a word, middle of a word, end of word, and separated. 
-This allows two consecutive words such as 
+This greatly increases the ambiguity of Arabic text. 
+Another difficulty is that  Arabic letters can have up to 
+four different forms
+corresponding to their position in a word, i.e, beginning,
+middle, end of word forms and separate forms. 
+This allows the phrase 
 \RL{almdrsT}\RL{al_A}
@@
 high complexity and accuracy at the low morphological analysis level 
-and that can often compensate for tolerable inaccuracy at higher level. 
-For example, if the query is interested in proper names and the 
+and that can often compensate for tolerable inaccuracy at higher levels. 
+For example, if the query concerns proper names and the 
 prefix in question in the analysis can only connect to a verb, 
@@
 We strongly believe and we find evidence in our results that a 
-case-based controller intervening at every decision is necessary to 
-guide, use, prune, and refine the morphological analysis.
+case-based morphological analyzer that allows a case-based
+controller to intervene at every decision to 
+guide, use, prune, and refine the morphological analysis
+is of high utility.
 
@@
 In this paper, we present Sarf, a {\em novel case-based efficient
-morphological analyzer} that uses parallel compositional 
+morphological analyzer} that uses compositional 
 non-deterministic Automata driven by a case based controller.
@@
 opportunity to the controller to intervene at every single input 
-character. 
+letter. 
 The decisions of the case-based controller can thus prune false 
 positives early in the run. 
-Each alive machine in Sarf takes as input one character at a time 
+Each alive machine in Sarf takes as input one letter at a time 
 from a text stream. 
@@
 performs tokenization on the fly based on morphological correctness.
-To out knowledge, this is the first morphological analyzer that 
+Up to our knowledge, this is the first morphological analyzer that 
 handles the ``run-on'' words problem.
@@
 its content or {\em matn} and its sanad.
-We also would like to partition the sanad accurately into the 
+We also partition the sanad accurately into the 
 separate narrators so that we can later look each one of them 
@@
 
-The hadith case study is interesting for Sarf since the controller
+The hadith case study is interesting to Sarf since the controller
 considers most of the analyses where we have a concentration
-of proper names in the text, and ignores most of the analyses as long 
-as they are valid where we do not have proper names. 
-The controller is also interested in a small subset of words 
+of proper names in the text. It ignores the text where
+proper names occur rarely as long 
+as the text is valid.
+The controller is also interested in discovering a subset of words 
 that relate persons to each other such as \RL{bin} or ``the sun of''
@@
 complex structure. 
+We also showcase that the results of the accurate high level abstraction
+results of the case-based analysis was not affected by a major
+refinement to the analyzer that increased low level analysis accuracy.
 
@@
 Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
-This allowed Sarf to retain better compositional part of speech
+This allows Sarf to perform better and 
+better retain compositional part of speech
 tags.
@@
 
-% paper structure
+The rest of this paper is structured as follows. In Section~\ref{sec:related}
+we compare Sarf to related work. Then in Section~\ref{sec:sarf}
+we discuss Sarf, illustrate the way it works with a running example
+that also explains how non-deterministic finite state automata
+works. In Section~\ref{sec:islamic} we present the hadith case
+study. We present our results in Section~\ref{sec:results}
+and we conclude with future work in Section~\ref{sec:future}.
 
@@
 Analyzers such as~\cite{Khoja:01,Darwish:02} 
-target specific specific applications in the morphological 
+target specific applications in the morphological 
 analyzer itself or use a specific set of part of speech tags
@@
 It is case-based in the sense that a controller that drives
-the morphological analyzers prunes these decisions. 
+the morphological analyzer prunes these decisions. 
 Up to our knowledge, Sarf is the only morphological analyzer
@@
 Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
-It differs than Buckwalter where it defines a shorter list of affixes
+It differs from Buckwalter in that it defines a shorter list of affixes
 and a longer list of affix compatibility rules to allow recursive 
@@
 the part of speech information
-tags associated with each prefix while these need to be highly 
+tags associated with each prefix and suffix.
+This information is highly 
 redundant with Buckwalter and that may lead to consistency issues. 
-Sarf keeps the possible analyses alive as a sequence corresponding
-to the text. Buckwalter produces a set of solutions for each token 
-and the several analysis threads of the whole text need to be 
+Sarf keeps all the possible analyses alive where each analysis
+corresponds to a set of decisions and positions in the text. 
+Buckwalter produces a set of solutions for each token 
+and the several analysis threads of the text need to be 
 generated as a product of the token solutions.
@@
 Beesley reports that the number of the Automata generated by a compiler
-from Xerox rules is not controlled and also reports a difficulty to 
+from Xerox rules can not be controlled and also reports a difficulty to 
 compose the FSAs into a single FSA framework. 
@@
 FSAs that are the prefix, stem, and suffix FSAs. 
-With Sarf it is easy to control the FSAs with a user custom case 
-controller, while with Beesley the effort requires recompiling the
-Xerox rules.
+With Sarf it is easy to control the FSAs with a custom user case 
+controller, while with Beesley and up to our understanding the 
+effort requires recompiling the Xerox rules.
 
@@
 text where a composition decision was made. 
-The case-based controller can build the trace when needed.
+The case-based controller can build the trace 
+when needed.
 
@@
 the ATB~\cite{Maamouri:04} project after the addition of a large 
-new corpus. We find evidence in this decision that our hypothesis
-about the utility of case-based analyzers. The refinements
-that resulted in algorithmic changes in SAMA were done manually
-and worked in integration with the ATB format only. 
+new corpus. We find supporting evidence in this decision to our hypothesis
+about the utility of a case-based analyzer. 
+The refinements that resulted in algorithmic changes in SAMA were 
+done manually and worked in integration with the ATB format. 
 In Sarf we propose a case-based controller approach that can modify 
@@
 with the case study. In Sarf the analyzer remains general, and the 
-case-based controller specializes the analyses.
+case-based controller specializes the analysis.
 
@@
 While the SVM analysis learns features from the text passed to the case
-study, it does not make use of the type oh the query the 
+study, it does not make use of the type of the query the 
 case study targets in the text.
@@
 of the Buckwalter analyzer. 
-We also use deterministic parsing with tries or our analysis 
+We also use deterministic parsing with tries in our analysis 
 to run the prefix, suffix and stem FSAs. 
-We think that their inferential-realizational approach 
+We think that the inferential-realizational approach 
+of ElixierFM
 that is highly compatible with the Arabic linguistic 
 description~\cite{Badawi:04}
-can make benefit of many features unique to the Arabic language.
+can benefit from many features unique to the Arabic language.
 Sarf leaves implementing that to the case-based controller
@@
 \center{
-\resizebox{1.9\columnwidth}{!}
+\resizebox{1.7\columnwidth}{!}
 { \input{psfigs/FSM.pstex_t}}
@@
 and the directed edges are transitions driven by an input
-character that matches the label of the edge. 
+letter that matches the label of the edge. 
 The square denotes an accept state, 
@@
 are omitted for clarity. 
-Figure~\ref{f:example}a represents the prefix FSA and 
-Figures~\ref{f:example}b and~\ref{f:example}c represent the 
+Figure~\ref{f:example}(a) represents the prefix FSA and 
+Figures~\ref{f:example}(b) and~\ref{f:example}(c) represent the 
 stem and suffix FSAs respectively. 
@@
 which is an accept state. 
-The edge labeled $\epsilon$ that connects the prefix 
-FSA to the stem FSA can transition from any accept state
+The edge labeled $\epsilon$ connects the prefix 
+FSA to the stem FSA and transitions from any accept state
 in the prefix FSA to the root state in the stem FSA.
@@
 When there are two valid transitions such as \RL{w} 
-and $\epsilon$ as is the start case, a non-deterministic 
+and $\epsilon$ in the start case, the non-deterministic 
 FSA $\Phi$ spawns an exact copy of itself $\Psi$. FSA $\Phi$
@@
 Each of the FSAs now represent a valid analysis so far. 
-When the FSA is at a state where the next input character
+When the FSA is at a state where the next input letter
 leads to a reject state, the FSA dies.
@@
 and are denoted by the absence of the corresponding 
-characters. 
-In our examples, if we assume there were no stems that start 
+letters. 
+In our example, if we assume there were no stems that start 
 with the letter \RL{w}, $\Psi$ will die then. 
@@
 with the prefix \RL{wsy}. 
-We do that by associating a category value for each accept state. 
-We consider the value of the category as part of the state.
+It does that by associating a category value for each accept state. 
+Sarf considers the value of the category as part of the state.
 Thus each state shown in Figure~\ref{f:example} actually represents
@@
 This is refined in Sarf to happen only with white space, delimiters, 
-or non-connecting characters. 
+or non-connecting letters. 
 We left that out in Figure~\ref{f:example} for clarity.
@@
 
-\subsection{Recursive Affixes}
+\subsection{Recursive affixes}
 \label{sec:recaffix}
@@
 forms a legal morpheme. 
-Buckwalter~\cite{Buckwalter:02} keeps separate lists 
+Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
 of prefixes, suffixes and stems and assigns a category
-to each morphemes. 
+to each morpheme. 
 The relation between morphemes and categories is one 
@@
 a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
-The string $\alpha\beta\gamma$ partition of the string is a 
+The $\alpha\beta\gamma$ partition of the string is a 
 valid morphological analysis if  and only if we have
@@
 We add two lists of compatibility rules $L_{pp}$ and
-$L_{xx}$ for prefix-prefix and suffix suffix compatibility
+$L_{xx}$ for prefix-prefix and suffix-suffix compatibility
 respectively.
@@
 The resulting category in all practical cases
-was the category of the second morpheme in the pair.
+is the category of the second morpheme in the pair.
 
-\subsection{Affix Linear FSA}
+\subsection{Affix linear FSA}
 \label{sec:affixFSA}
@@
 We analyzed the prefixes and suffixes of 
-Buckwalter~\cite{Buckwalter:02}
+Buckwalter~\shortcite{Buckwalter:02}
 and extracted our refined lists of recursive affixes.
-We encoded the lists into a FSA similar to that
+We encoded the lists into an FSA similar to those
 in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
@@
 $L_{xx}$ compatibility lists.
-An accept state in the FSA corresponds to the last character
+An accept state in the FSA corresponds to the last letter 
 of a prefix in the prefix and suffix lists.
@@
 
-\subsection{Stem Linear FSA}
+\subsection{Stem linear FSA}
 \label{sec:stemFSA}
@@
 letters in stems. 
-This approach is superior to Buckwalter since again it is
-a linear walk in the trie while with Buckwalter it is 
-several hash lookups.
+This approach is superior to Buckwalter since it consists of
+a linear walk in the trie while with Buckwalter we need
+a number of hash lookups in the order of all possible partitions
+of the string.
 
-\subsection{Non-deterministic Composition of FSAs}
+\subsection{Non-deterministic composition of FSAs}
 \label{sec:ndfsa}
@@
 abstraction of the non-deterministic composition 
-of the prefix, stem and suffix FSA. 
+of the prefix, stem and suffix FSAs. 
 The circle represents regular states, the square
-represents accept state and the triangle represents
+represents accept states and the triangle represents
 reject states. 
@@
 of FSAs $M$. 
-The algorithm reads a character $c$ at a time from $L$
-and moves on all the machines in $M$. 
-For each $\Phi$ in $M$ the algorithm checks whether
-$\Phi$ is in an accept state or not.
-
-If $\Phi$ is in an accept state and it is in the suffix
+The algorithm reads a letter $c$ at a time from $L$
+and iterates over all machines in $M$. 
+For each $\Phi$ in $M$,
+if $\Phi$ is in an accept state and it is in the suffix
 phase, then the algorithm checks whether it should report
@@
 This happens when $c$ is a white space or a delimiter, 
-or when the last character leading to the current state
-was a non-connecting character.
+or when the last letter leading to the current state
+was a non-connecting letter.
 
@@
 Otherwise, $\Phi$ transitions using the $c$ edge.
-We invoke the controller $A$ after all the FSAs in $M$ 
-have moved and allow it full access and control over
+After iterating over all FSAs in $M$, we invoke the 
+controller $A$
+and grant it full access and control over
 all FSAs in $M$.
@@
 from prefix to stem, stem to suffix, and suffix to prefix
-show the compatibility constraints as a condition for 
+require compatibility constraints for 
 the transition.
@@
 
-Our analysis accepts a book as a text stream input
-and segments it into a vector of individual hadiths. 
-We segment each hadith into two parts. The first part
-is the sanad and it
+Our analysis accepts a book as input
+and segments it into a vector of hadiths. 
+We segment each individual hadith into two parts. 
+The first part is the sanad and it
 contains the chain of narrators who related the hadith
-to Mohammad. 
-The second part is the matn and it constitutes the content
+from Mohammad. 
+The second part is the matn or the content
 of the hadith. 
-We are concerned in further exploring the sanad. 
-In particular, we detect the chain of narrators
-and the relation between each narrator, his ancestor, and 
+We are concerned to further explore the sanad and
+we detect the chain of narrators and 
+the relation that links each narrator to his ancestor and 
 his predecessor in the chain. 
@@
 
-We built our case-based controller to target
+We built our hadith case-based controller to target
 the detection of proper names. 
@@
 % FSM for the controller
-The FSA in Figure~\ref{f:hadith}
+
+We illustrate the hadith case-based controller FSA is 
+in Figure~\ref{f:hadith}. 
+The transitions in the hadith FSA are excited
+by inputs from the Sarf morphological analyzer. 
+A NAME label means that Sarf encountered a proper name.
+An NRC label means that Sarf met a narrator connector such as
+\RL{`an} "on behalf of", \RL{.hada_t} "narrated", \RL{qaal} "said", 
+\RL{'a_hbar} "told" or one of their derivations. 
+The IBN label corresponds to the word \RL{ibn} "the son of" and is a name connector.
+The NISBA label corresponds to the an adjective that points to a person such 
+as \RL{alma.sriy} "the Egyptian".
+
+The symbol $\tau_{\mbox{NMC}}$ is a threshold
+that corresponds to the number of tolerated name connectors 
+that may occur between two names. 
+The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
+of name connectors collected since the controller
+started looping in the state NMC\_S. 
+The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
+that corresponds to a relaxed tolerance measure that
+the controller resorts to in case the words separating
+two names were longer than $\tau_{\mbox{NMC}}$ but 
+contained a name connector word such as IBN or NISBA.
+
+The controller has four states that correspond to 
+an abstract position in text relative to the next sanad. 
+State TEXT\_S is the initial state and denotes that
+the controller is outside the context of a sanad.
+The controller moves to the state NAME\_S whenever
+NAME is reported by Sarf.
+State NRC\_S means that the controller thinks it is in the context
+of a narrator name after Sarf reports an NRC.
+Similarly, state NMC\_S
+indicates that the controller expects a name to appear within 
+a tolerance threshold expressed by 
+$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
+
+\begin{figure}[tb!]
+\center{
+\resizebox{.9\columnwidth}{!}
+{ \input{psfigs/hadith.pstex_t}}
+\caption{FSA for the controller of the hadith case study.}
+\label{f:hadith}
+}
+\end{figure}
+
+Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
+before it gives up on its expectations.
+
+We reach the NAME\_S state only when a
+valid NAME is detected and we leave when no more NAME's are detected. 
+The NRC\_S state can only be reached if an NRC is detected.
+NMC\_S can only be reached from a NAME\_S state.
+
+The FSA reports a valid chain of narrators when a sequence of names
+connected by narrator connectors appears. 
+It marks the begining of the hadith with the begining of the sequence,
+and marks the end of the hadith by the begining of the next sequence. 
+It also marks the chain of narrators as the sequence itself.
+
+The definition of input labels such as NAME and IBN depends on the 
+morphological analyzer. 
+However, our case-based controller approach was able to perform well
+under both a good analyzer and a refined one.
 
@@
 
-\subsection{Efficiency Comparison Against Buckwalter and SAMA}
+\begin{table*}[bt]
+\centering
+\caption{Results of the hadith case study with Sarf.}
+%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
+\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
+ & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
+ & & Simple & Accurate & & Simple & Accurate & & Simple  & Accurate \\ \hline \hline
+Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
+ Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
+Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
+Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
+Narrator/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
+Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
+Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline \hline
+Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
+Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
+%Narrator accuracy & & & & & &  & & & \\ 
+Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
+Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
+Sarf running time (secs.)& & 24.5 & 22.67 & & 23.5 & 23.17 & & 2.25 & 1.67\\ \hline
+\end{tabular}
+\normalsize
+\label{t:hadithresallresults}
+\end{table*}
 
-\subsection{Case Study Accuracy and Efficiency Results}
+To evaluate our hypothesis of the utility of a case-based morphological analyzer,
+we ran our case study in two modes, simple and accurate.
+In the simple mode Sarf is controlled by the hadith controller.
+In the accurate mode we apply several refinements to Sarf.
+The first refinement removes female proper names since they do not occur often in
+hadith.
+We also add a refinement that detects compound names as they are common in narrator
+names.
+Finally, we consider a possessive form as a reference to a person and treat
+it as a narrator.
 
-% accuracy measure:
-% on names: a word is a name or not
-% chains: narrator is missing from a chain
-% segmentation: two hadiths are merged
+We report our results in Table~\ref{t:hadithresallresults} on the three hadith books.
+The word count row is an indicator of the size of data.
+The number of names, narrators, and chains reflect a hierarchical structure where a chain contains
+narrators, and a narrator contains names.
+In general, we observe that the refinements slightly improved performance and accuracy at
+the lowest level of abstraction by reducing false positives.
+However, the refienements had little to no effect on the high accuracy measures at higher
+levels of abstraction.
+We beleive that this is a strong supporting evidence to our hypothesis.
+
+The names row reports the total number of detected names. The names/narator row reports the average
+number of names per narrator structure. The narrator row reports the number of narrators detected.
+The narrator/chain reports the number of narrators per chain structure. And the chain row reports
+the number of detected chains in the book.
+The ignored names row reports the number of names detected outside the chains and thus irrelevant to
+the case study.
+The segmentation accuracy reports on the accuracy of segmentation of the book into separate hadiths.
+The metric is computed manually over a representative sample of each text. The same applies to all other accuracy measures.
+The chain accuracy reflects the percentage of correct narrators covered by extracted chains.
+The narrator name accuracy row reflects the percentage of correct proper names detected within
+narrator strutures.
+
+The name false positives row reports the percentage of words recognized as names while they should not.
+We notice that the low level metric of name false positives improved dramatically with the accurate
+analysis without substantially affecting the high abstract measures such as chain and segmentation
+accuracy.
+We notice that the controller in the simple analysis ignored more names and thus the case-based
+controller compensated for the inaccuracy of the low level analysis with a simple heuristic and
+without the need of the fancy refinements.
+
+We investigated the case of Al Kafi where the accuracy of the simple analysis was significantly lower than the accurate
+analysis and discovered that the chains detected by the accurate analysis were a subset of those detected by the simple
+analysis. 
+This means that the simple analysis results have false positives and did not miss important information. We hypothesize
+that a higher level of abstraction analysis such as intersecting the narrators with their biographies can 
+reduce false positives in the simple analysis.
 
 \section{Future Work}
+\label{sec:future}
+
+In the future we plan to complete the hadith case study 
+and automate narrator discovery and authentication in the 
+biography books. 
+We plan to make use of the name connector and narrator
+connector stop phrases that our case study extracted to create a 
+learning system that can learn more names. 
+We also plan on developing a domain specific language similar 
+that allows users to easily specify their case-based controller
+to interact with Sarf.
 
@@
 
-\section*{Acknowledgments}
+%\section*{Acknowledgments}
+% this will go into the regular submission
 
 \bibliographystyle{naaclhlt2010}
-\bibliography{fzAr}
+{\small \bibliography{fzAr}}
 

### 2010-06-26 tex/sarf.tex
spell

@@
 high complexity and accuracy at the low morphological analysis level 
-and that can often compensate for tolerable inaccuracy at higher levels. 
-For example, if the query concerns proper names and the 
-prefix in question in the analysis can only connect to a verb, 
-we can provide an answer without completing the analysis. 
-%TODO A lso if a word such as \RL{wasiim} was name was ambiguous and we were looking for a sequence of names, then 
+and that can often compensate at higher level for tolerable inaccuracy at low levels. 
+For example, if a query concerns proper names and the 
+prefix in question in the analysis connects only to a verb, 
+we do not need to further analyze the rest of the word 
+to provide an answer.
 We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
@@
 The controller is also interested in discovering a subset of words 
-that relate persons to each other such as \RL{bin} or ``the sun of''
+that relate persons to each other such as \RL{bin} or ``the son of''
 or words that mean narrate such as \RL{qaal} or ``said''. 
@@
 This allows Sarf to perform better and 
-better retain compositional part of speech
-tags.
+maintain compositional part of speech tags.
 \item {\bf Hadith case study:}
@@
 we compare Sarf to related work. Then in Section~\ref{sec:sarf}
-we discuss Sarf, illustrate the way it works with a running example
-that also explains how non-deterministic finite state automata
-works. In Section~\ref{sec:islamic} we present the hadith case
+we discuss Sarf and illustrate the way it works with a running example.
+We also explain how non-deterministic finite state Automata
+work. In Section~\ref{sec:islamic} we present the hadith case
 study. We present our results in Section~\ref{sec:results}
-and we conclude with future work in Section~\ref{sec:future}.
+and conclude with future work in Section~\ref{sec:future}.
 
@@
 
-The survey work in~\cite{Sughaiyer:04} discusses and compares
+The survey in~\cite{Sughaiyer:04} discusses and compares
 several morphological analyzers. 
@@
 the morphological analyzer prunes these decisions.
-Up to our knowledge, Sarf is the only morphological analyzer
-that addresses the ``run-on'' words problem.
 
@@
 With Sarf it is easy to control the FSAs with a custom user case 
-controller, while with Beesley and up to our understanding the 
-effort requires recompiling the Xerox rules.
+controller. Up to our understanding, Beesley  
+requires recompiling the Xerox rules.
 
@@
 done manually and worked in integration with the ATB format. 
-In Sarf we propose a case-based controller approach that can modify 
-the behavior of the analyzer on the fly and that should interact
-with the case study. In Sarf the analyzer remains general, and the 
+Sarf is a case-based controller approach that can modify 
+the behavior of the analyzer on the fly and can interact
+with the any case study. In Sarf, the analyzer remains general, and the 
 case-based controller specializes the analysis.
@@
 We think that the inferential-realizational approach 
-of ElixierFM
+of ElixirFM
 that is highly compatible with the Arabic linguistic 
@@
 since in many cases the abstraction needed may not need the 
-text to be grammar correct.
-We can compare the case-based controller in Sarf to a high level
-component in the ElixirFM domain specific language.
+text to be grammatically correct.
+%We can compare the case-based controller in Sarf to a high level
+%component in the ElixirFM domain specific language.
 
@@
 
-Finally, The case-based controller can interfere at any point in the 
+Finally, the case-based controller can interfere at any point in the 
 analysis to make a decision like ignoring an Automata that 
@@
   \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
-  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. 
-  Our understanding is that those books are historical documents with 
-  public IP.}.
+  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
+%  Our understanding is that those books are historical documents with 
+ % public IP.}.
 
@@
 connected by narrator connectors appears. 
-It marks the begining of the hadith with the begining of the sequence,
-and marks the end of the hadith by the begining of the next sequence. 
+It marks the beginning of the hadith with the beginning of the sequence,
+and marks the end of the hadith by the beginning of the next sequence. 
 It also marks the chain of narrators as the sequence itself.
@@
  & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
- & & Simple & Accurate & & Simple & Accurate & & Simple  & Accurate \\ \hline \hline
+ & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\ \hline \hline
 Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
@@
 Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
-Narrator/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
+Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
 Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
@@
 To evaluate our hypothesis of the utility of a case-based morphological analyzer,
-we ran our case study in two modes, simple and accurate.
-In the simple mode Sarf is controlled by the hadith controller.
-In the accurate mode we apply several refinements to Sarf.
+we ran our case study in two modes, simple and refined.
+In the simple mode, Sarf is controlled by the hadith controller.
+In the refined mode, we apply several refinements to Sarf.
 The first refinement removes female proper names since they do not occur often in
@@
 the lowest level of abstraction by reducing false positives.
-However, the refienements had little to no effect on the high accuracy measures at higher
+However, the refinements had little to no effect on the high accuracy measures at higher
 levels of abstraction.
-We beleive that this is a strong supporting evidence to our hypothesis.
+We believe that this is a strong supporting evidence to our hypothesis.
 
-The names row reports the total number of detected names. The names/narator row reports the average
-number of names per narrator structure. The narrator row reports the number of narrators detected.
-The narrator/chain reports the number of narrators per chain structure. And the chain row reports
+The names row reports the total number of detected names. The names/narrator row reports the average
+number of names per narrator structure. The narrators row reports the number of narrators detected.
+The narrators/chain row reports the number of narrators per chain structure. And the chain row reports
 the number of detected chains in the book.
@@
 The narrator name accuracy row reflects the percentage of correct proper names detected within
-narrator strutures.
+narrator structures.
 
 The name false positives row reports the percentage of words recognized as names while they should not.
-We notice that the low level metric of name false positives improved dramatically with the accurate
+We notice that the low level metric of name false positives improved dramatically with the refined
 analysis without substantially affecting the high abstract measures such as chain and segmentation
@@
 
-We investigated the case of Al Kafi where the accuracy of the simple analysis was significantly lower than the accurate
-analysis and discovered that the chains detected by the accurate analysis were a subset of those detected by the simple
+We investigated the case of Al Kafi where the accuracy of the simple analysis was significantly lower than that of the refined
+analysis. We  discovered that the chains detected by the refined analysis were a subset of those detected by the simple
 analysis. 
-This means that the simple analysis results have false positives and did not miss important information. We hypothesize
-that a higher level of abstraction analysis such as intersecting the narrators with their biographies can 
-reduce false positives in the simple analysis.
+The reason for the lower accuracy is false positive chains. 
+The simple analysis did not miss important information. 
+We hypothesize that a higher level of abstraction such as intersecting the narrators from the hadith books with the narrators 
+in the biographies can reduce the false positives in the chains.
 
@@
 
-In the future we plan to complete the hadith case study 
+In the future, we plan to complete the hadith case study 
 and automate narrator discovery and authentication in the 
@@
 We also plan on developing a domain specific language similar 
+to ElexirFM~\cite{Otakar:07}
 that allows users to easily specify their case-based controller

### 2010-06-26 trunk/tex/sarf.tex
9 pages

@@
 \usepackage{amsthm}
+\usepackage{booktabs}
 
@@
 \vocalize
-For instance the word \RL{'a.hmadH} 
-may have two valid morphological analyses. In the first
-the letter \RL{'a} is a prefix and the word means ``I praise him'',
-and in the second the letter \RL{'a} is part of the stem 
+For instance, the word \RL{'a.hmadH} 
+may have two valid morphological analyses. 
+The letter \RL{'a} may be prefix and the word means ``I praise him''.
+The letter \RL{'a} may also be part of the stem 
 \RL{'a.hmad} (a proper noun)
@@
 Buckwalter~\shortcite{Buckwalter:02},
-Beesely~\shortcite{Beesley:01}, SAMA~\cite{Kulick:10},
+Beesly~\shortcite{Beesley:01}, SAMA~\cite{Kulick:10},
 and ElixirFM~\cite{Otakar:07} exist.
@@
 
-The hadith case study is interesting for Sarf since the controller
+The hadith case study is interesting to Sarf since the controller
 considers most of the analyses where we have a concentration
-of proper names in the text, and ignores most of the analyses 
-where we do not have proper names in text as long 
-as they are valid.
+of proper names in the text. It ignores the text where
+proper names occur rarely as long 
+as the text is valid.
 The controller is also interested in discovering a subset of words 
-that relate persons to each other such as \RL{bin} or ``the sun of''
+that relate persons to each other such as \RL{bin} or ``the son of''
 or words that mean narrate such as \RL{qaal} or ``said''. 
@@
 Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
-This allowed Sarf to retain better compositional part of speech
-tags.
+This allows Sarf to perform better and 
+maintain compositional part of speech tags.
 \item {\bf Hadith case study:}
@@
 
-The survey work in~\cite{Sughaiyer:04} discusses and compares
+The survey in~\cite{Sughaiyer:04} discusses and compares
 several morphological analyzers. 
 Analyzers such as~\cite{Khoja:01,Darwish:02} 
-target specific specific applications in the morphological 
+target specific applications in the morphological 
 analyzer itself or use a specific set of part of speech tags
@@
 It is case-based in the sense that a controller that drives
-the morphological analyzers prunes these decisions. 
-Up to our knowledge, Sarf is the only morphological analyzer
-that addresses the ``run-on'' words problem.
+the morphological analyzer prunes these decisions.
 
 Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
-It differs than Buckwalter where it defines a shorter list of affixes
+It differs from Buckwalter in that it defines a shorter list of affixes
 and a longer list of affix compatibility rules to allow recursive 
@@
 FSAs that are the prefix, stem, and suffix FSAs. 
-With Sarf it is easy to control the FSAs with a user custom case 
-controller, while with Beesley the effort requires recompiling the
-Xerox rules.
+With Sarf it is easy to control the FSAs with a custom user case 
+controller. Up to our understanding, Beesley  
+requires recompiling the Xerox rules.
 
@@
 the ATB~\cite{Maamouri:04} project after the addition of a large 
-new corpus. We find evidence in this decision that our hypothesis
-about the utility of case-based analyzers. The refinements
-that resulted in algorithmic changes in SAMA were done manually
-and worked in integration with the ATB format only. 
-In Sarf we propose a case-based controller approach that can modify 
-the behavior of the analyzer on the fly and that should interact
-with the case study. In Sarf the analyzer remains general, and the 
-case-based controller specializes the analyses.
+new corpus. We find supporting evidence in this decision to our hypothesis
+about the utility of a case-based analyzer. 
+The refinements that resulted in algorithmic changes in SAMA were 
+done manually and worked in integration with the ATB format. 
+Sarf is a case-based controller approach that can modify 
+the behavior of the analyzer on the fly and can interact
+with the any case study. In Sarf, the analyzer remains general, and the 
+case-based controller specializes the analysis.
 
@@
 of the Buckwalter analyzer. 
-We also use deterministic parsing with tries or our analysis 
+We also use deterministic parsing with tries in our analysis 
 to run the prefix, suffix and stem FSAs. 
-We think that their inferential-realizational approach 
+We think that the inferential-realizational approach 
+of ElixirFM
 that is highly compatible with the Arabic linguistic 
 description~\cite{Badawi:04}
-can make benefit of many features unique to the Arabic language.
+can benefit from many features unique to the Arabic language.
 Sarf leaves implementing that to the case-based controller
 since in many cases the abstraction needed may not need the 
-text to be grammar correct.
-We can compare the case-based controller in Sarf to a high level
-component in the ElixirFM domain specific language.
+text to be grammatically correct.
+%We can compare the case-based controller in Sarf to a high level
+%component in the ElixirFM domain specific language.
 
@@
 
-Note that $\Sigma$ can report a full word at any accept state
-now by spawning a new Automata using the $\epsilon$ transition
+Note that $\Sigma$ reports a full word at any accept state
+by spawning a new Automata using the $\epsilon$ transition
 to the prefix FSA.
-This is refined in Sarf to happen only with white space, delimiters, 
-or non-connecting characters. 
+In Sarf, this only happens with white space, delimiters, 
+and non-connecting letters. 
 We left that out in Figure~\ref{f:example} for clarity.
 
-Finally, The case-based controller can interfere at any point in the 
+Finally, the case-based controller can interfere at any point in the 
 analysis to make a decision like ignoring an Automata that 
@@
 namely \RL{w}, \RL{s}, and \RL{y}.
-Moreover, the POS tag for \RL{wsy} is a concatenation
+The POS tag for \RL{wsy} is a concatenation
 of the POS tags of each of the three morphemes. 
@@
 The resulting category in all practical cases
-was the category of the second morpheme in the pair.
+is that of the second morpheme in the pair.
 
-\subsection{Affix Linear FSA}
+\subsection{Affix linear FSA}
 \label{sec:affixFSA}
@@
   \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
-  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. 
-  Our understanding is that those books are historical documents with 
-  public IP.}.
+  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
+%  Our understanding is that those books are historical documents with 
+ % public IP.}.
 
-Our analysis accepts a book as a text stream input
-and segments it into a vector of individual hadiths. 
-We segment each hadith into two parts. The first part
-is the sanad and it
+Our analysis accepts a book as input
+and segments it into a vector of hadiths. 
+We segment each individual hadith into two parts. 
+The first part is the sanad and it
 contains the chain of narrators who related the hadith
-to Mohammad. 
-The second part is the matn and it constitutes the content
+from Mohammad. 
+The second part is the matn or the content
 of the hadith. 
-We are concerned in further exploring the sanad. 
-In particular, we detect the chain of narrators
-and the relation between each narrator, his ancestor, and 
+We are concerned to further explore the sanad and
+we detect the chain of narrators and 
+the relation that links each narrator to his ancestor and 
 his predecessor in the chain. 
@@
 % FSM for the controller
-The FSA in Figure~\ref{f:hadith}
+
+We illustrate the hadith case-based controller FSA 
+in Figure~\ref{f:hadith}. 
+The transitions in the hadith FSA are excited
+by inputs from the Sarf morphological analyzer. 
+A NAME label means that Sarf encountered a proper name.
+An NRC label means that Sarf met a narrator connector such as
+\RL{`an} "on behalf of", \RL{.hada_t} "narrated", \RL{qaal} "said", 
+\RL{'a_hbar} "told" or one of their derivations. 
+The IBN label corresponds to the word \RL{ibn} "the son of" and is a name connector.
+The NISBA label corresponds to the an adjective that points to a person such 
+as \RL{alma.sriy} "the Egyptian".
+
+The symbol $\tau_{\mbox{NMC}}$ is a threshold
+that corresponds to the number of tolerated name connectors 
+that may occur between two names. 
+The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
+of name connectors collected since the controller
+started looping in the state NMC\_S. 
+The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
+that corresponds to a relaxed tolerance measure that
+the controller resorts to in case the words separating
+two names were longer than $\tau_{\mbox{NMC}}$ but 
+contained a name connector word such as IBN or NISBA.
+
+The controller has four states that correspond to 
+an abstract position in text relative to the next sanad. 
+State TEXT\_S is the initial state and denotes that
+the controller is outside the context of a sanad.
+The controller moves to the state NAME\_S whenever
+NAME is reported by Sarf.
+State NRC\_S means that the controller thinks it is in the context
+of a narrator name after Sarf reports an NRC.
+Similarly, state NMC\_S
+indicates that the controller expects a name to appear within 
+a tolerance threshold expressed by 
+$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
+
+\begin{figure}[tb!]
+\center{
+\resizebox{.9\columnwidth}{!}
+{ \input{psfigs/hadith.pstex_t}}
+\caption{FSA controller of the hadith case study.}
+\label{f:hadith}
+}
+\end{figure}
+
+Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
+before it gives up on its expectations.
+
+We reach the NAME\_S state only when a
+valid NAME is detected and we leave when no more NAME's are detected. 
+The NRC\_S state can only be reached if an NRC is detected.
+NMC\_S can only be reached from a NAME\_S state.
+
+The FSA reports a valid chain of narrators when a sequence of names
+connected by narrator connectors appears. 
+It marks the beginning of the hadith with the beginning of the current sequence,
+and marks the end of the hadith with the beginning of the next sequence. 
+It also marks the chain of narrators as the sequence itself.
+
+The definition of input labels such as NAME and IBN depends on the 
+morphological analyzer. 
+However, our case-based controller approach was able to perform well
+under both a good analyzer and a refined one.
 
@@
 \label{sec:results}
-
-We report our results based on a manual validation of 
-a representative sample selected randomly from the text.
 
@@
 \begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
- & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
- & & Simple & Accurate & & Simple & Accurate & & Simple  & Accurate \\ \hline \hline
+ & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
+ & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
 Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
@@
 Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
-Narrator/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
+Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
 Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
-Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline \hline
-Segmentation accuracy & & 80\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
+Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
+Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
 Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
 %Narrator accuracy & & & & & &  & & & \\ 
-Name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
+Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
 Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
-Sarf running time & &a & & & & & & & \\ \hline
+Sarf running time (secs.)& & 24.5 & 22.67 & & 23.5 & 23.17 & & 2.25 & 1.67\\ \hline
 \end{tabular}
@@
 
-\subsection{Efficiency Comparison Against Buckwalter and SAMA}
+To evaluate our hypothesis of the utility of a case-based morphological analyzer,
+we ran our case study in two modes, simple and refined.
+In the simple mode, Sarf is controlled by the hadith controller.
+In the refined mode, we apply several refinements to Sarf.
+The first refinement removes female proper names since they do not occur often in
+hadith.
+We also add a refinement that detects compound names as they are common in narrator
+names.
+Finally, we consider a possessive form as a reference to a person and treat
+it as a narrator.
 
-\subsection{Case Study Accuracy and Efficiency Results}
+We report our results in Table~\ref{t:hadithresallresults} on the three hadith books.
+The word count row is an indicator of the size of data.
+The number of names, narrators, and chains reflect a hierarchical structure where a chain contains
+narrators, and a narrator contains names.
+In general, we observe that the refinements slightly improved performance and accuracy at
+the lowest level of abstraction by reducing false positives.
+However, the refinements had little to no effect on the high accuracy measures at higher
+levels of abstraction.
+We believe that this is a strong supporting evidence to our hypothesis.
 
-% accuracy measure:
-% on names: a word is a name or not
-% chains: narrator is missing from a chain
-% segmentation: two hadiths are merged
+The names row reports the total number of detected names. The names/narrator row reports the average
+number of names per narrator structure. The narrators row reports the number of narrators detected.
+The narrators/chain row reports the number of narrators per chain structure. And the chain row reports
+the number of detected chains in the book.
+The ignored names row reports the number of names detected outside the chains and thus irrelevant to
+the case study.
+The segmentation accuracy reports on the accuracy of segmentation of the book into separate hadiths.
+The metric is computed manually over a representative sample of each text. The same applies to all other accuracy measures.
+The chain accuracy reflects the percentage of correct narrators covered by extracted chains.
+The narrator name accuracy row reflects the percentage of correct proper names detected within
+narrator structures.
+
+The name false positives row reports the percentage of words recognized as names while they should not.
+We notice that the low level metric of name false positives improved dramatically with the refined
+analysis without substantially affecting the high abstract measures such as chain and segmentation
+accuracy.
+We notice that the controller in the simple analysis ignored more names and thus the case-based
+controller compensated for the inaccuracy of the low level analysis with a simple heuristic and
+without the need of the fancy refinements.
+
+We investigated the case of Al Kafi where the accuracy of the simple analysis was significantly lower than that of the refined
+analysis. We  discovered that the chains detected by the refined analysis were a subset of those detected by the simple
+analysis. 
+The reason for the lower accuracy is false positive chains. 
+The simple analysis did not miss important information. 
+We hypothesize that a higher level of abstraction such as intersecting the narrators from the hadith books with the narrators 
+in the biographies can reduce the false positives in the chains.
 
@@
 
-In the future we plan to complete the hadith case study 
+In the future, we plan to complete the hadith case study 
 and automate narrator discovery and authentication in the 
 biography books. 
+We plan to make use of the name connector and narrator
+connector stop phrases that our case study extracted to create a 
+learning system that can learn more names. 
 We also plan on developing a domain specific language similar 
+to ElexirFM~\cite{Otakar:07}
 that allows users to easily specify their case-based controller
@@
 \bibliographystyle{naaclhlt2010}
-\bibliography{fzAr}
+{\small \bibliography{fzAr}}
 

### 2010-06-26 tex/sarf.tex
last one

@@
 
-Note that $\Sigma$ can report a full word at any accept state
-now by spawning a new Automata using the $\epsilon$ transition
+Note that $\Sigma$ reports a full word at any accept state
+by spawning a new Automata using the $\epsilon$ transition
 to the prefix FSA.
-This is refined in Sarf to happen only with white space, delimiters, 
-or non-connecting letters. 
+In Sarf, this only happens with white space, delimiters, 
+and non-connecting letters. 
 We left that out in Figure~\ref{f:example} for clarity.
@@
 The resulting category in all practical cases
-is the category of the second morpheme in the pair.
+is that of the second morpheme in the pair.
 
@@
 
+%\setcode{utf8}
+%\begin{arabtext}
+%حدثنا \emphasize{هشيم}، أخبرنا \emphasize{إسماعيل بن أبي خالد}،
+% عن أبي إسحاق، عن سعيد بن جبير، قال كنا
+%    مع ابن عمر رضي الله عنه حيث أفاض من
+%    عرفات إلى جمع فصلى بنا المغرب ومضى
+%\end{arabtext}
+%\setcode{standard}
+
 Our analysis accepts a book as input
@@
 \subsection{Controller}
+\label{sec:controller}
 % FSM for the controller
 
-We illustrate the hadith case-based controller FSA is 
+We illustrate the hadith case-based controller FSA 
 in Figure~\ref{f:hadith}. 
@@
 The IBN label corresponds to the word \RL{ibn} "the son of" and is a name connector.
-The NISBA label corresponds to the an adjective that points to a person such 
+The NISBA label corresponds to an adjective that points to a person such 
 as \RL{alma.sriy} "the Egyptian".
@@
 of a narrator name after Sarf reports an NRC.
-Similarly, state NMC\_S
+State NMC\_S
 indicates that the controller expects a name to appear within 
@@
 { \input{psfigs/hadith.pstex_t}}
-\caption{FSA for the controller of the hadith case study.}
+\caption{FSA controller of the hadith case study.}
 \label{f:hadith}
@@
 connected by narrator connectors appears. 
-It marks the beginning of the hadith with the beginning of the sequence,
-and marks the end of the hadith by the beginning of the next sequence. 
+It marks the beginning of the hadith with the beginning of the current sequence,
+and marks the end of the hadith with the beginning of the next sequence. 
 It also marks the chain of narrators as the sequence itself.
@@
 However, our case-based controller approach was able to perform well
-under both a good analyzer and a refined one.
+under both Sarf and a refined version of Sarf.
 
@@
 \begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
- & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ 
- & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\ \hline \hline
+ & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
+ & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
 Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
@@
 Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
-Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline \hline
+Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
 Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 

### 2010-09-07 tex/sarf.tex
added hadith example figure

@@
-%
 % File naaclhlt2010.tex
@@
 \fullvocalize
-%\transtrue
+\transtrue
 \arabtrue
@@
 \newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
+\newcommand{\frl}[1]{\fbox{\RL{#1}}}
 
@@
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Case based Arabic Morphological Analysis with 
-    Compositional Non-deterministic Automata }
+\title{Case-based Arabic Morphological Analysis }
 
@@
 \vocalize
-For instance, the word \RL{'a.hmadH} 
+For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
 may have two valid morphological analyses. 
@@
 middle, end of word forms and separate forms. 
-This allows the phrase 
-\RL{almdrsT}\RL{al_A}
+This allows the phrase \transfalse
+\RL{il_A\nospace almdrsT} \transtrue
 to be visually recognizable
-as two separate words \RL{al_A} and \RL{almdrsT} without 
-the need of a delimiter such as a space in between. 
+as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
+without the need of a delimiter such as a space in between. 
 This happens because the first word \RL{almdrsT} ends with
@@
 We hypothesize that many NLP case studies need the 
-morphological analyzer to answer simple queries that do not need 
-high complexity and accuracy at the low morphological analysis level 
-and that can often compensate at higher level for tolerable inaccuracy at low levels. 
+morphological analyzer to answer queries that do not need 
+high accuracy at the low morphological analysis level 
+and that can often compensate at higher levels for tolerable inaccuracy 
+at low levels. 
 For example, if a query concerns proper names and the 
@@
 efficiency of affix matching.
+
+\transfalse
+\begin{figure}[tb]
+\center{
+\resizebox{.9\columnwidth}{!}
+{ \input{psfigs/exhadith.pstex_t}}
+\caption{Hadith abstraction example.}
+\label{f:exhadith}
+}
+\end{figure}
+% the FSMs for the three words
+\transtrue
 
@@
 
+\transfalse
 \begin{figure*}[tb]
@@
 % the FSMs for the three words
+\transtrue
 

### 2010-11-04 tex/sarf.tex
why u r not there?

@@
 % File naaclhlt2010.tex
-%
 % Contact: nasmith@cs.cmu.edu
@@
 \newcommand{\frl}[1]{\fbox{\RL{#1}}} 
+\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
+\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}
 
@@
 
-%\author{ Jad Makhlouta \\
-%  \And
-%Hamza Harkous \\
-%  American University of Beirut \\
-%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
-%  \And 
-%  Fadi Zaraket 
-%}
+\author{ Jad Makhlouta \\
+  \And
+Hamza Harkous \\
+  American University of Beirut \\
+  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
+  \And 
+  Fadi Zaraket 
+}
 
@@
 \center{
-\resizebox{.9\columnwidth}{!}
+\resizebox{1.2\columnwidth}{!}
 { \input{psfigs/exhadith.pstex_t}}
@@
 literature. 
-A {\em hadith} is a narration related to the prophet Mohammad
-through a {\em sanad} or a sequence of narrators. 
-The authentication of a hadith highly depends on the credibility
+A \RL{.hady_t} is a narration related to the prophet Mohammad
+through a \RL{sanad} or a sequence of narrators. 
+Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
+transliteration and translation. 
+The authentication of a \noArRL{.hady_t} highly depends on the credibility
 of each of the narrators as reported in separate biography 
@@
 In this paper we consider the problem of automatically segmenting
-a hadith book into narrations, then segmenting each narration into
-its content or {\em matn} and its sanad.
+a \noArRL{.hady_t} book into narrations, then segmenting each narration into
+its content or \RL{matn} and its \noArRL{sanad}.
 We also partition the sanad accurately into the 
@@
 up in the biography books. 
+For example, the boxes in Figure~\ref{f:exhadith} are proper names and once
+connected toghether they form complex names of narrators. For example
+narrator $n_1$ has the first name \noTrRL{qtybT} and his father name is
+\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.
 
@@
 machine controls and guides the analysis. 
-\item {\bf Exhaustive analysis:} Since we have a case-based 
-controller that can decide on the fly whether an analysis is 
-accepted or not, we can afford to keep all valid analyses modulo
-those pruned by the case controller. 
-We do that using non-deterministic FSAs. 
+%\item {\bf Exhaustive analysis:} Since we have a case-based 
+%controller that can decide on the fly whether an analysis is 
+%accepted or not, we can afford to keep all valid analyses modulo
+%those pruned by the case controller. 
+%We do that using non-deterministic FSAs. 
 \item {\bf Recursive affixes:}
 We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02}  and use recursive affixes. 
+Buckwalter~\shortcite{Buckwalter:02}  and use define the affixes
+recursively. 
 This allows Sarf to perform better and 
@@
 This information is highly 
-redundant with Buckwalter and that may lead to consistency issues. 
+redundant with Buckwalter.% and that may lead to consistency issues. 
 Sarf keeps all the possible analyses alive where each analysis
@@
 compose the FSAs into a single FSA framework. 
-Sarf solves the problem with a non-deterministic composition of three
-FSAs that are the prefix, stem, and suffix FSAs. 
-With Sarf it is easy to control the FSAs with a custom user case 
+Sarf solves the problem with an FSA that is the non-deterministic 
+composition of a controller and three
+transducers that are the prefix, stem, and suffix tansducers. 
+With Sarf it is easy to control the FSA with a custom user case 
 controller. Up to our understanding, Beesley  
@@
 We also use deterministic parsing with tries in our analysis 
-to run the prefix, suffix and stem FSAs. 
+to run the prefix, suffix and stem transducers. 
 We think that the inferential-realizational approach 
@@
 Sarf encodes the prefix, suffix and stem lexicons into 
-three linear FSAs and uses a non-deterministic composition
-of the three FSAs to compute all valid morphological analyses.
+three linear transducers and uses a non-deterministic composition
+of the three to compute all valid morphological analyses.
 We first present a running example of Sarf. 
@@
 handles ``run-on'' words.
-Then we discuss how Sarf builds and composes the different FSAs.
+%Then we discuss how Sarf builds and composes the different FSAs.
 
@@
 { \input{psfigs/FSM.pstex_t}}
-\caption{Example composition of the prefix, stem and suffix FSAs.}
+\caption{Example composition of the prefix, stem and suffix transducers.}
 \label{f:example}
@@
 
-The compositional FSA diagram in Figure~\ref{f:example}
+The diagram in Figure~\ref{f:example}
 illustrates the operation of Sarf when parsing the
@@
 }
-The square and circle legends are states of the FSA
+The square and circle legends are states 
 and the directed edges are transitions driven by an input
@@
 are omitted for clarity. 
-Figure~\ref{f:example}(a) represents the prefix FSA and 
+Figure~\ref{f:example}(a) represents the prefix transducer and 
 Figures~\ref{f:example}(b) and~\ref{f:example}(c) represent the 
-stem and suffix FSAs respectively. 
+stem and suffix transducers respectively. 
 The symbol $\epsilon$ represents an empty string and is 
@@
 
-We start at the root square state in the prefix FSA
+We start at the root square state in the prefix transducer
 which is an accept state. 
 The edge labeled $\epsilon$ connects the prefix 
-FSA to the stem FSA and transitions from any accept state
-in the prefix FSA to the root state in the stem FSA.
-The edge that connects the stem FSA to the suffix FSA
+transducer to the stem transducer and transitions from any accept state
+in the prefix transducer to the root state in the stem transducer.
+The edge that connects the stem transducer to the suffix transducer
 follows the same behavior.
@@
 and $\epsilon$ in the start case, the non-deterministic 
-FSA $\Phi$ spawns an exact copy of itself $\Psi$. FSA $\Phi$
-makes one transition, and FSA $\Psi$ makes the other. 
-Each of the FSAs now represent a valid analysis so far. 
-When the FSA is at a state where the next input letter
-leads to a reject state, the FSA dies.
+composition $\Phi$ spawns an exact copy of itself $\Psi$. 
+$\Phi$ makes one transition, and $\Psi$ makes the other. 
+Each of $\Phi$ and $\Psi$ now represent a valid analysis so far. 
+When $\Phi$ or $\Psi$ is at a state where the next input letter
+leads to a reject state, it dies.
 The reject states are omitted from Figure~\ref{f:example}
@@
 
-Note that the prefix and affix FSAs allow recursive 
+Note that the prefix and affix transducers allow recursive 
 prefixes. For example, \RL{w} will result in an accept state
-that transitions to the stem FSA. 
+that transitions to the stem transducer. 
 When \RL{s} follows, we move to another accept state in the 
-prefix FSA corresponding to the suffix \RL{ws}. 
-The same applies to the suffix FSA.
+prefix transducer corresponding to the suffix \RL{ws}. 
+The same applies to the suffix transducer.
 
-Lets consider FSA $\Phi$ after it consumed \RL{wsy} 
-and transitioned into the stem FSA root node. 
+Lets consider $\Phi$ after it consumed \RL{wsy} 
+and transitioned into the stem transducer root node. 
 Now $\Phi$ will transition with \RL{l`b} to reach an accept 
@@
 
-Since $\Phi$ is now in an accept state in the stem FSA, it will
+Since $\Phi$ is now in an accept state in the stem transducer, it will
 spawn $\Sigma$ that transitions into the root of the 
-suffix FSA. When \RL{h} is consumed, $\Phi$ will die since 
+suffix transducer. When \RL{h} is consumed, $\Phi$ will die since 
 there is no edge from the current state labeled with \RL{h}. 
@@
 Note that $\Sigma$ reports a full word at any accept state
-by spawning a new Automata using the $\epsilon$ transition
-to the prefix FSA.
+by spawning a new composition using the $\epsilon$ transition
+to the prefix transducer.
 In Sarf, this only happens with white space, delimiters, 
@@
 Finally, the case-based controller can interfere at any point in the 
-analysis to make a decision like ignoring an Automata that 
+analysis to make a decision like ignoring a composition that 
 is not interesting. 
@@
 based on the decision of the others. 
+The controller uses the output of the transducers as input
+to make its decision. 
+It does not produce any output and therefore the composition
+of the three transducers with the controller is a finite 
+state Automaton (FSA).
 
@@
 namely \RL{w}, \RL{s}, and \RL{y}.
-The POS tag for \RL{wsy} is a concatenation
+The part of speech (POS) tag for \RL{wsy} is a concatenation
 of the POS tags of each of the three morphemes. 
 Keeping these morphemes separate in a list
-is redundant and may produce consistency issues. 
+is redundant.
+When users try to add their custom morphemes to these
+lists they may produce consistency issues. 
 We add two lists of compatibility rules $L_{pp}$ and
@@
 
-\subsection{Affix linear FSA}
+Formally we define an affix $f$ recursively to be either
+an atomic affix $a$ or an atomic affix concatenated
+to an affix and we write $f: a | a f$. 
+Atomic affixes are standalone affixes
+that are not the concatenation of other affixes. 
+Sarf only keeps atomic affixes and their categories.
+
+\subsection{Affix linear transducer}
 \label{sec:affixFSA}
@@
 and extracted our refined lists of recursive affixes.
-We encoded the lists into an FSA similar to those
+We encoded the lists into a structure similar to those
 in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
-The two FSAs are directed acyclic graphs. 
-The absence of cycles in these FSAs reduces the 
-traversal of the affix FSAs to a linear traversal.
+The two structures are directed acyclic graphs. 
+The absence of cycles reduces the 
+traversal of the affix transducer to a linear traversal.
 This is computationally superior to the 
@@
 
-The prefix and suffix FSAs encode the $L_{pp}$ and
+The prefix and suffix transducers encode the $L_{pp}$ and
 $L_{xx}$ compatibility lists.
-An accept state in the FSA corresponds to the last letter 
+An accept state in the transducer corresponds to the last letter 
 of a prefix in the prefix and suffix lists.
-The accept states hold category state information as well
-as POS and other tags.
+The accept states produce as output the POS and other tags
+associated with the state.
 
-\subsection{Stem linear FSA}
+\subsection{Stem linear transducer}
 \label{sec:stemFSA}
@@
 an efficient double array trie structure~\cite{Aoe:89}. 
-The structure is also a linear FSA where the accept
+The structure is also linear where the accept
 states are the terminal nodes corresponding to the last 
@@
 
-\subsection{Non-deterministic composition of FSAs}
+\subsection{Non-deterministic composition of transducers}
 \label{sec:ndfsa}
@@
 abstraction of the non-deterministic composition 
-of the prefix, stem and suffix FSAs. 
+of the prefix, stem and suffix transducers. 
 The circle represents regular states, the square
@@
 reject states. 
-Note that the concrete prefix, stem and suffix FSAs
+Note that the concrete prefix, stem and suffix transducers
 are all linear and do not have cycles as discussed 
@@
 The cycles are introduced in the abstract machine
-for brevity and illustrative purposes.
+for brevity and pedagogical purposes.
 
@@
 \caption{An abstraction of the non-deterministic composition of 
-    the prefix, stem and suffix FSAs.}
+    the prefix, stem and suffix transducers.}
 \label{f:composition}
@@
 $A$. 
-It starts with one FSA $\Phi_1$ similar to the one in 
+It starts with one machine $\Phi_1$ similar to the one in 
 Figure~\ref{f:composition} and inserts it into a collection
-of FSAs $M$. 
+$M$. 
 The algorithm reads a letter $c$ at a time from $L$
@@
 In all cases, if $\Phi$ is in an accept state, 
-it spawns another FSA $\Psi$ and adds it to $M$. 
+it spawns another machine $\Psi$ and adds it to $M$. 
 If $c$ leads to a reject state, then $\Phi$ dies 
@@
 Otherwise, $\Phi$ transitions using the $c$ edge.
-After iterating over all FSAs in $M$, we invoke the 
+After iterating over all machines in $M$, we invoke the 
 controller $A$
 and grant it full access and control over
-all FSAs in $M$.
+all machines in $M$.
 
@@
 NDSarf(string $L$, Controller $A$) 
-  FSAVec $M$; -- collection of FSA Automata
-  FSA $\Phi_1$;
+  MachineVec $M$; -- collection of machines
+  Machine $\Phi_1$;
   $M$.insert($\Phi_1$);
@@
 in Islamic studies. 
+Based on that Islamic scholars issue rules that affect the life
+of people around the world. 
+The literature lacks an exhaustive mechanism of authentication
+check as. This is currently manual work prone to error and almost
+impossible to complete in the life time of a scholar due to the
+large literature available. 
 %Several researchers~\cite{Hadithopaedia:08} attempted to automate 
@@
 
-We illustrate the hadith case-based controller FSA 
+We illustrate the hadith case-based controller 
 in Figure~\ref{f:hadith}. 
-The transitions in the hadith FSA are excited
+The transitions in the hadith controller are excited
 by inputs from the Sarf morphological analyzer. 
@@
 { \input{psfigs/hadith.pstex_t}}
-\caption{FSA controller of the hadith case study.}
+\caption{The controller of the hadith case study.}
 \label{f:hadith}
@@
 
-The FSA reports a valid chain of narrators when a sequence of names
+The controller reports a valid chain of narrators when a sequence of names
 connected by narrator connectors appears. 
@@
 Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
-Sarf running time (secs.)& & 24.5 & 22.67 & & 23.5 & 23.17 & & 2.25 & 1.67\\ \hline
+Sarf running time (secs.)& & 2.97 & 1.32 & & 2.93 & 1.31 & & .25 & .096\\ \hline \hline
+ & & Buckwalter& ElixirFM & & Buckwalter & ElixirFM & & Buckwalter & ElixirFM \\\bottomrule
+Names      & & 18,354 & 19,201 & & 17,934 &  18,202 & & 5,021 & 5,117 \\
+Names/Narrator& & 2.6 & 2.1 & & 2.47 &  2.07 & & 1.83 & 1.62\\
+Narrators & & 3,210 & 2,981 & & 6,425 &  6,208 & & 2,311 & 1,965 \\ 
+Narrators/Chain & & 5.57 & 5.32 & & 5.46 &  5.32 & & 5.12 & 5.01 \\
+Chains & & 627 & 601 &  & 1,318 & 1,250 & & 487 & 452 \\ 
+Ignored names & & 9,642 & 9,915 &  & 6,118 & 6,874 & & 1,812 & 1,904 \\ \hline
+Segmentation accuracy & & 76\%& 84\%& & 85\%& 89 \%& & 84\%& 85\%\\ 
+Chain accuracy & & 78\% & 87\%&  & 87\%& 90\%& & 85\% & 86\% \\ 
+%Narrator accuracy & & & & & &  & & & \\ 
+Narrator name accuracy & & 76\% & 86\%& & 84\%& 87\% & & 84\% & 84\% \\ \hline
+Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
+Running time (secs.)& & 6.65 & 12$\times$60& &4.55 & 14$\times$60 & & .66 & 6$\times$60 
 \end{tabular}
@@
 
-To evaluate our hypothesis of the utility of a case-based morphological analyzer,
-we ran our case study in two modes, simple and refined.
+In section we present results that compare Sarf to existing 
+morphological analyzers such as Buckwalter and ElixirFM. 
+We also evaluate our hypothesis of the utility of a case-based 
+morphological analyzer,
+by running our case study in two modes, simple and refined.
+We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
+with 4GB of memory using a Linux operating system.
+
 In the simple mode, Sarf is controlled by the hadith controller.
 In the refined mode, we apply several refinements to Sarf.
-The first refinement removes female proper names since they do not occur often in
-hadith.
-We also add a refinement that detects compound names as they are common in narrator
-names.
+The first refinement ignore female proper names except at the end of
+a chain since they do not occur often in narrator names in hadith while
+they can be common in early narrators such as the wives of the prophet.
+We also add a refinement that detects compound proper names 
+such as \RL{`bd alr.hmn} 
+as they are common in narrator names.
 Finally, we consider a possessive form as a reference to a person and treat
@@
 
-\section{Future Work}
+\subsection{Comparison with Buckwalter and ElixirFM}
+
+We implemented our case study with Buckwalter in two phases. 
+In phase one, we process the entire text with Buckwalter and
+obtain annotations to locations in text. 
+Then in the second phase we feed these annotations to 
+our controller as if they were coming from Sarf. 
+We measure the time it took to produce the annotations
+and the time it took the controller to process the 
+annotations whithout the time spent streaming the annotations.
+We used the same setup for ElixirFM.
+
+Sarf outperforms both Buckwalter and ElixirFM in runing time. 
+It also produces higher accuracy measures. 
+We think this is mainly due to the extended
+lexicon as well as to the fact that Sarf reduces false positive
+names by ignoring prefixes and suffixes that can not form
+names even if the stem was a name. 
+Doing the same with Buckwalter and ElixirFM requires  
+expanding the annotations resulting from one affix to reflect
+its multiple composing affixes.
+This is easily done with the way we defined affixes recursively.
+
+Buckwalter performed much better than ElixirFM. This is up to 
+our knowledge the first comparison of these two tools and 
+we think the big difference is due to the more exact 
+and grammar based functional morphological analysis ElixirFM 
+uses.
+
+\section{Conclusion and Future Work}
 \label{sec:future}
+
+In this paper we o
 

### 2010-12-05 trunk/tex/rp.tex
rp files

%\documentstyle{llncs}
%\documentclass[12pt,fullpage]{llncs}
\documentclass[12pt]{article}
%\usepackage[left=3cm,top=1.2cm,right=3cm,nohead,nofoot]{geometry}
%\usepackage[letter,left=.25in,right=.25in,top=.17in,textwidth=7.5in,textheight=9in]{geometry}
%\usepackage[left=0cm]{geometry}

%\documentclass[12pt]{report}
%\usepackage {utthesis2}              %% Preamble.

%\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{arabtex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{multirow}
%\usepackage[colorlinks=false]{color,hyperref}
\usepackage{color,hyperref}
%\usepackage{amsthm}
%\usepackage{booktabs}
\usepackage{natbib}
\bibpunct{[}{]}{,}{n}{}{;}

\usepackage{url}

\usepackage{relsize}
\usepackage{fancyvrb}
\usepackage{fancyhdr,lastpage}

\usepackage{utf8}
\setarab
\fullvocalize
\transtrue
\arabtrue

\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue} 
\newcommand{\noTrNoVocRL}[1]{\novocalize\transfalse\RL{#1}\transtrue\vocalize}

\usepackage{setspace}
%\doublespacing
\renewcommand{\baselinestretch}{1.15}
\setlength{\parindent}{0in}
%\parskip 6pt
%\parindent 0pt
%\setlength{\parskip}{.05in}
%\oddsidemargin 0in
%\evensidemargin 0in
\oddsidemargin .0in
\evensidemargin .0in
\hoffset -.75in
\voffset -.83in
\textwidth 7.5in
\textheight 9.5in

\topsep 0in
\topmargin 0.17in

%\usepackage{arabtex}
%\usepackage{utf8}

\begin{document}

\pagestyle{fancy}
%\lhead{}
\chead{}
\rhead{NPRP No.~:~~~~4-484-1-075}

\lfoot{QNRF Form}
\cfoot{}
\rfoot{Page \thepage~of~\pageref{LastPage} }
\renewcommand{\footrulewidth}{0.2pt}
\renewcommand{\headrulewidth}{0.2pt}

%\begin{titlepage}

\begin{center}
{\Large \bf Relational Arabic Text Mining Framework using 
    Morphological
    and Case-Based Analysis }

%\setlength{\unitlength}{1in}
%\setlength{\unitlength}{1in}
\vspace{1.5in}

\renewcommand{\arraystretch}{.6}
\begin{tabular}{cc}
Fadi Zaraket & Rehab Duwairi \\
Electrical and Computer Engineering &  Computer Science and Engineering Department \\
American University of Beirut & University of Qatar \\
{\tt fadi.zaraket@aub.edu.lb} & {\tt Rehab.Duwairi@qu.edu.qa}
\end{tabular}

\vspace{1.5in}

\renewcommand{\arraystretch}{.6}
\begin{tabular}{c}
{\small Research Plan Document } \\
{\small Proposal Submitted to }\\
\\
    Qatar National Research Foundation \\
    National Priorities Research Program \\
    4$^{th}$ Cycle 
\end{tabular}
\vspace{.5in}

\date{\today}
\pagebreak

\end{center}

\section{Background (max of 2 pages)}
\label{s:background}

Research in text mining started in the mid 1980s when Swanson 
realized that slicing and combining seemingly unrelated medical 
articles led to the discovery of new 
hypotheses~\cite{JNi06}.
Till present, most of the interest in text mining comes 
from biological sciences.
Lately we saw applications of text mining in the advertisement, 
political campaigning, and other businesses.
Text mining involves information retrieval (IR), natural language 
processing (NLP), information extraction (IE), 
and data mining~\cite{ASr09}.

Most of academic research in the field of Arabic text mining 
is covered in the almost comprehensive book 
``Arabic Computational Morphology''~\cite{Sou07}.
The book summarizes recent work, and presents strong evidence of 
shortage of historic research. It also highlights  an emerging 
interest in this area with a specific application to machine 
translation from Arabic to English after 2001.
The research community showed little interest in 
automating Arabic text analysis that may serve
the Arabic user.
While a non-Arabic user is interested in machine translation to 
help him understand an Arabic document, 
an Arabic user is more interested in 
information hard to discover manually such as 
the relation between clinical records, pharmaceutical 
prescriptions and sales, and a specific sickness. 
Another Arabic user may be interested in looking at the relation
between stock prices and political news or the 
relation between suspect profiles in interrogation reports 
and investigative reports from the crime scenes.

We propose to build a 
%{\em relational text mining 
%of Arabic using morphological and case-based analysis}. 
{\em case-based Arabic text analysis using relations and morphology}(CATARM) framework. 
CATARM will take as input sets of documents, each of a specific
type, and provides the user with a visual query language that
allows him to build a flow of Arabic text mining
modules and discover desired relations between documents. 
CATARM will translate the query into a netlist of computational 
components, it will also extract parameters from the query and 
the document sets and use them to simplify the computational 
components. 
%The main CATRAM computational components will be finite 
%state machines (FSM) that express a linguistic model of the Arabic 
%language and that can be parametrized and relaxed to fit a case 
%study.

%IR tools are similar to a Google search 
%engine where documents of interest are selected based on the 
%relevance to a set of keywords of interest to the user.
%NLP targets the automation of understanding human languages.
%This task is the oldest and most difficult task in the artificial 
%intelligence domain.
%The main difficulty behind NLP comes from the ambiguity of 
%natural languages~\cite{Osm08}.
%While NLP techniques cannot deliver their final target yet, they 
%can deliver analysis of sentences into nouns, verbs, and adjectives.
%They can also narrow the meaning of a word or a phrase based on 
%the context to one of its possible meanings.
%Finally they can parse a sentence into a relation between the 
%discovered entities using abstractions such as verb-name 
%abstractions~\cite{Osm08}.

%Information extraction (IE) targets forming data structures and 
%instances of these data structures out of a collection of documents.
%IE tries to fit the output of IR and NLP to templates of 
%interest to the user.
%Data mining can be applied at the end on the output of the 
%IE process to answer user queries about the input 
%documents~\cite{JHa05}.

%\subsection{Arabic text mining in academy }
%In the following we review the major works on Arabic text mining 
%in academy and industry.

Arabic morphological analyzers~\cite{Sughaiyer:04}
consider an Arabic word and its internal structure composed of 
several {\em morphemes}. 
A morpheme is a {\em stem} or an {\em affix}.
An affix is a {\em prefix, suffix,} or an {\em infix}.
The analysis of one word may lead to several possible
solutions.
\vocalize
For instance, the word 
\RL{'a.hmadH}~\footnote{In this document, we use the default 
ArabTeX transliteration style ZDMG.}
may have two valid morphological analyses. 
The word means ``I praise him'' when
the letter \RL{'a} is a prefix and  it means
``his Ahmad'' when 
the letter \RL{'a} is part of the proper noun stem 
\RL{'a.hmad}.
\novocalize

%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
It is common practice to write Arabic text
without short vowels. 
This greatly increases the ambiguity of Arabic text. 
Arabic letters can have up to 
four different forms
corresponding to their position in a word, i.e, beginning,
middle, end of word and separate forms. 
This allows the phrase
\noTrRL{il_A\nospace almdrsT}  with no delimiter within
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
This happens because the first word \RL{almdrsT} ends with
\RL{T} a non-connecting letter. 
These words occur often in text and greatly increase the
difficulty of tokenization and are referred to as 
``run-on'' words~\cite{Buckwalter:04}.
Such difficulties are inherrent to the 
morphological analysis of the Arabic language and
render the solutions inaccurate.

Current morphological analyzers such as 
Buckwalter~\cite{Buckwalter:02},
Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} exist.
These analyzers are used in several open source spell checkers as 
well as NLP frameworks~\cite{Col09}.
They take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, a white space delimited token may have 
more than one word.
Second, the exhaustive enumeration may hurt performance and may
not be necessary or appropriate
in some case studies~\cite{Maamouri:10}. 
Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy at the expense of efficiency.

{\bf Hypothesis 1.} We hypothesize that many case studies, in particular relational
case studies, do not require high accuracy from the 
low level morphological analysis.
The sophistication of the sought higher level query can eliminate 
many of the ambiguous low level solutions at a lower computational 
cost.
For example, if a query concerns proper names and the 
prefix in question in the analysis connects only to a verb, 
we do not need to further analyze the rest of the word 
to provide an answer.

We find evidence to our hypothesis in~\cite{Maamouri:10}. The 
addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
required the high NLP task to interact with the low 
level morphological analyzer.
This led to a refined analyzer~\cite{Kulick:10}.  
We also find evidence in~\cite{Habash:06} that different types of 
morphological analyses behave differently on the same case study. 
We strongly believe and we find evidence in our preliminary
results that a 
case-based morphological analyzer that allows a high level 
case-based controller to intervene at every decision to 
guide, use, prune, and refine the morphological analysis
is of high utility.

%Prior work in Arabic text mining addressed the classification and 
%categorization of Arabic text and made little use of features 
%unique to the Arabic language.
Researchers in~\cite{AEL07,Ham07,Abd07,MEl03} 
preprocessed the Arabic text to make it presentable to known text 
mining techniques that tend to Latin text and queries.
%Buckwalter~\cite{Bee89,Tim04} uses morphological analysis to enable 
%relaxed stemming of Arabic words.
El-Halees used statistical analysis to address the problem of 
classifying Arabic text documents with maximum 
entropy~\cite{AEL07}.
In~\cite{Abd07} Mesleh evaluated six different techniques to 
classify Arabic text using a support vector machine.
El Dost~\cite{MEl03} used root words in the Arabic language to 
speed up automated and hierarchical indexing.
Al-Zoghby~\cite{Ham07}
%addresses the problem of 
%entities extracted from Arabic text with similar content but 
%slightly different names via 
used the derivation feature in 
the Arabic language as a similarity measurement.
This work is limited to the names of the fields in the data 
structures extracted from Arabic texts.

{\bf Hypothesis 2.} This approach does not tend directly to 
the Arabic users who write their 
queries in Arabic and think in Arabic.
We hypothesize that Arabic queries use rules and 
relations similar to the Arabic text under investigation, 
and thus techniques native to the Arabic language will do better 
in Arabic text analysis.

We find evidence in the work of the CADIM \noTrRL{q-adim} 
project~\cite{Col09} at the University of Columbia to support our 
hypothesis. They observe that Arabic texts used as training 
data for common NLP techniques proved ineffective. 
They built semi-automatic models to address the problem of automatic 
speech recognition in Arabic.

%\subsection{\bf Arabic text mining in industry}
Few companies in the industry provide primitive Arabic text 
analysis tools.
Sakhr~\cite{Sak09} provides a categorization 
tool with a limited number of predefined categories, and a summary 
generator that selects sentences from text.
Basis technology~\cite{Bas09} provides Rosette, a lexical 
analysis tool that normalizes Arabic text as a pre-processing step 
for indexers and other general purpose data mining 
tools.  Basis also provides REX; an entity extractor.

ElixirFM~\cite{Otakar:07} uses an {\em inferential-realizational}
approach for morphological analysis. 
Realizational morphology considers the form of the word 
\noTrRL{.sarf}  instead of its segments. 
Formal rules and formulas express the relations between 
the root of the lexeme and the fully inflected word. 
While this is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04},
and it benefits from features unique to the Arabic language
captured by the formal rules, 
it is computationally expensive and enumerates all the possible
solutions.

{\bf Contributions.} Up to our knowledge, CATARM will be the 
first framework that tends to relational Arabic morphology 
and text mining. 
{\bf (1.)} CATARM will build a novel parametrized linguistic 
computational model of Arabic. 
The model will be implemented using efficient structures. 
 {\bf (2.)} It will be amenable to Arabic specific simplifications 
inferred from the correspondence between the fromal Arabic rules
and the relational user queries.
{\bf (3.)} The model will also be amenable to simplifications 
inferred from statistical features extracted from the document 
sets using language independent techniques such as latent semantic 
indexing~\cite{LSI89} and hidden Markov models (HMM).
%CATARM will augment current digital dictionaries to resolve 
%entity semantics and targeted qualificationsl.
{\bf (4.)} CATARM extracts semantic graphs from each text 
document and then uses set arithmetic and graph analysis techniques
to establish relations between these graphs. {\bf (5.)} CATARM will 
solve at least three interesting relational Arabic text case studies 
presented in Section~\ref{s:design:casestudies}.

\pagebreak

\section{Objectives and significance (max of 1 page) } 
\label{s:objectives}

This proposal aims at building  a relational Arabic text mining 
framework with the following objectives.
\begin{itemize}\itemsep0pt
\item CATARM tends to Arabic users and solves the
problem of finding relations between Arabic documents. 
\item CATARM builds a formal Arabic computational model
using efficient and configurable data structures such as 
finite state machines (FSM). 
\item The framework will provide the user with a visual query 
language that makes it easy to enter relational queries. 
The query language will also allow the framework to extract Arabic
specific features from the query to simplify the linguistic 
computational model. 
\item The framework will use statistical models and techniques
to infer language independent features from the text documents
that will further simplify the computational model. 
\item During the project, we will augment current Arabic digital 
dictionaries to resolve entity semantics and targeted 
qualifications such as adjectives.
\item The framework will allow the integration of library and 
custom computational components built on top of primitive
CATARM components. 
The components will be the linguistic computational model,
statistical, set arithmetic, logic,
and graph analysis techniques, and  text algorithms.
\end{itemize}

CATARM will be measured and validated against several 
case studies and will contain  
all the components needed to 
address the following case studies.

\begin{itemize}\itemsep0pt
\item {\bf The literature case study.}
We are given a set of books of historical accounts where
a chain of narrators precedes each account,
and a second set of books of biographies where
a biography includes an evaluation of the authenticity 
of the narrator therein.
CATARM will allow the user to automatically and exhaustively query 
all accounts related to a subject important to Islamic jurispudence 
(for example women and travel),
and extract a graph where nodes represent narrators and edges
are labeled with the authenticity of the narrators. 
A similar manual query takes ages to answer and is error prone.
\item {\bf The security case study.}
Given two sets of reports consisting of crime scene investigation 
and suspect interrogation reports, 
an officer may query CATARM 
for relations between suspects in terms of 
criminal action, locality, or third party persons unknown to the 
officer.
Such a query is very hard to answer manually.
\item {\bf The net traffic case study.}
Given a set of undesired internet content for minors
such as violent text, and a set of documents requested by internet
minor users, 
CATARM can discover whether there is a relation between 
the document and the violent content and flag the document if the
relation is strong enough. 
\end{itemize}

{\bf Significance.} 
CATARM will discover relations between Arabic text documents that 
are hard to discover by humans and will have direct live and
cost saving benefits in cases similar to the above case studies. 
CATARM will qualitatively enhance research in the cultural and 
literature fields. 
It may also help investors analyze Arabic news and data from Arab 
markets prior to making important decisions.

The use of Arabic text mining  will 
have a high impact on the prosperity and safety of the Arab world 
in particular and the world in general.
%For example, having the tools to check the authenticity of a hadith 
%at a click of a button may save youth from the hands of extreme 
%religious leaders.
For example, extracting a relation between two criminals may 
uncover a third criminal who is ready to commit a crime.
Finding inconsistencies in claims may curb down fraud and corruption 
in the insurance sector.

In addition, building the necessary academic 
infrastructure for Arabic text mining in Lebanon and Qatar will 
open the way for investments in the field and provide work 
opportunities for native workers.

\pagebreak
u
\section{Preliminary data or studies}
\label{s:prelim}

Our preliminary studies are promising. With a small manually entered dictionary, few programmed grammar rules and patterns, and two hardcoded queries we were able to solve selected samples from the literature and security case studies presented in Section "Significance". 
These preliminary results serve as a good proof of concept to our hypothesis. They also confirm that our approach performs better than currently existing techniques from Sakhr (Sakhr Inc. 2009) and Basis (Basis technology 2009) as well as from open source tools (Columbia's Arabic Dialect Modeling Group 2009).

The Co-Lead PI worked and published extensively in this domain. A selected list of her recent publications follows. 
    Rehab Duwairi, Mohammad Al-Refai, Natheer Khasawneh, "Feature Reduction Techniques for Arabic Text Categorization". Journal of the American Society for Information Science and Technology (JASIST), Volume 60, Issue 11, pages: 2347-2352, 2009. 
        R. M. Duwairi, "Arabic Text Categorization", International Arab Journal for Information Technology (IAJIT), Volume 4, Number 2, pages 125  132, 2007.
            R. M. Duwairi, "Machine Learning for Arabic Text Categorization". Journal of the American Society for Information Science and Technology (JASIST), Vol. 57, Issue 8, pages 1005-1010, 2006.

The Lead PI worked in the area of software arabization since 1996 as he worked on developing Arabization modules for Windows NT and an Arabic string manipulation library (Zaraket, Zbib and Khalil, Yaraah: Arabization module for Windows NT 1996). The Lead PI is also involved in working on an open source project that aims at comprehensive automation of authenticity checks against a corpus of literature heritage (Zaraket, Zeineddine, et al. 2006). The Lead PI is leading an effort at AUB to introduce Computational Arabic in the curriculum of the ECE program. He is also leading a team of students in an effort to place the infrastructure needed for this project to start. 
            The lead PI worked and published about relational logic and logic solvers applied to structured languages as follows.
                Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
                    Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007.

\section{Research design and methods}
\label{s:designmethods}

We see more and more Arabic documents such as texts,
 books,
 publications,
 hospital and governmental records emerging every day.
 Most of the newly generated documents are produced in digital textual form while also old paper documents are ported to digital form.
 These documents include huge amounts of information that is qualitatively different than structured information such as that contained in database entries.
 Text mining is the technology that automates the discovery of information in non structured text.

Text mining concerns the partitioning of text segments into classes,
 the clustering of text segments,
 the extraction of concepts and entities from text,
 and the discovery and modeling of relations between text entities and classes.
 Text mining techniques perform natural language processing (NLP) to analyze text and perform information extraction (IE) to extract information into data structures.
 Fundamental to IE are the name entity (NE) and the name entity relation (NER) extraction techniques which capture features from sentences and phrases.
 A key challenge to NE and NER is that sentence structures and words are often ambiguous in natural languages.
 While research to address ambiguity in Latin languages is still lagging behind~\cite{Red08},
 not much has been done to address NE and NER in the context of the Arabic language.

The application of text mining techniques to Arabic text documents will result in great benefits to many research fields such as Arabic literature,
 Islamic studies,
 Hadith authentication,
 and Arabic history and culture in general.
 Arabic text mining will also benefit sectors where Arabic text documents are key such as the security sector,
 the government personal records,
 the taxation department,
 the health sector,
 as well as the trading floor of a stock exchange.

The difficulty of applying text mining techniques to Arabic text documents lays in the absence of automated tools that understand the unique features of the Arabic language.
 We will make use of an example to illustrate the process.
 Given an Arabic text document we want a tool that can segment and isolate all names,
 dates,
 time of the day occurrences,
 tool names,
 and locations as entities.
 Then we want another tool to identify identities amongst these entities,
 and relate them in their order of appearance in the document.
 We want another tool to look at the verbs and actions in the sentences and try to build relations between the identified entities based on these verbs.
 If we were successful to do that,
 we now want to order all entities associated with dates based on a chronological order.

To do the above,
 we need tools that are able to automatically differentiate between names,
 verbs,
 adjectives and other vocabulary structures.
 This is not a simple task with the Arabic language.
 For example,
 Arabic grammar allows name-based sentences or sentences with no 
    verbs \noTrNoVocRL{^gml ismiyT} .
 Another unique feature is the possibility to have names in Arabic 
 that express action such as noun verbs 
 \noTrNoVocRL{ism f`l, f-a`l $\ldots$}.

 In addition to these structural differences,
 there are features in the Arabic language that need special 
 treatment such as derivatives and stems 
 \noTrNoVocRL{alm^staq-at w al^g_dwr}.
 We also need to build dictionary tools that associate words and phrases based on their meanings and their context.
 Often time,
 and depending on the application and the user of the tool,
 we need to allow the user to assign meanings and semantics for patterns in the text.

In the following we discuss our research methodology to develop the proposed Arabic text analysis tools.

Lexical analysis library: We need a library of routines that enables primitive lexical analysis of Arabic text.
 For example,
 simple primitive tasks such as comparing two strings need special attention in Arabic since each Arabic letter has four different forms depending on its position in the word; isolated form,
 start,
 middle and end of a word.
 Arabic diacritics should also be ignored most of the time when comparing two strings.
 These lexical analysis routines should also pay attention to the form of a letter when used in tokenization as a letter in and end of word form denotes the end of the word even if it was not followed with a space.
 Such a library can benefit from existing Unicode enabled string libraries such as Qt.
 It also can provide a layer of abstraction that hides the details of the encoding of Arabic letters used in the text documents.

Syntax analysis library: This library should allow sentence structure understanding.
 We will explore the best way to develop this library.
 Several options exist such as pattern based matching,
 tokenization and dictionary based lookups,
 or a language grammar that allows more than one matching rule.
 This library needs to be extendible to allow additions of new rules or patterns.

 (Note the ElixirFM does this)
 We will explore Arabic grammar books such as The Principles of Arabic~\cite{Sha73} and others~\cite{Abd00,Abd001}
 to augment current morphological analyzers~\cite{Tim04} and syntax analysis tools~\cite{Col09}.
 We will use relaxed NLP techniques to overapproximate results.
 Then and after a relational query,
 we will apply statistical analysis techniques to rank the matching results.

Dynamic modern Arabic dictionary: The dictionary will allow identities based on semantics,
 derivation rules,
 as well as user assigned rules.
 The dictionary will return modern grammatical categories for words classifying them as nouns,
 verbs,
 and adjectives instead of the classical Arabic categories that specify how the word is derived from an original radix.
 It remains to be explored how to represent these categories without exploding the size of the dictionaries with redundant words of multiple categories.
 This work will augment existing open source Arabic dictionaries and will allow for better morphological analysis and entity extraction.

An entity extraction library: We will explore building a library that allows extraction of entities from Arabic text.
 These entities can be proper names,
 dates,
 locations,
 and qualification adjectives to list few possibilities.
 The extractor will use the dynamic modern Arabic dictionary as well as the syntax library to extract entities from a sentence and to establish identity between entities across sentences.

A relation extraction library: We will explore building a library that allows extraction of relations from Arabic text.
 We will explore the use of pairing techniques to establish relations between the extracted entities.
 The relations will form graphs where the nodes are the entities and the previously extracted relations and the edges are the newly discovered relations.
 The routines in this library will take as input a template for an expected relation and will return a graph that describes whether this relation exists between the subject documents and where.
 These relations will be ranked using statistical analysis.
 The statistical criterion and the ranking methods remain to be decided.

A logic solver: This is a logical solver that will solve query expressions to resolve matching text and relations.
 We will explore the proposal of a logical query language that is restricted enough to be computable and expressive enough to allow interesting properties.
 We will use of the shelf solvers and matches from the entity and relation extractors to evaluate the expression.
 The choice of solvers remains to be decided after the query language is formed.

Tagger: We will explore the development of an automatic tagging interface that allows fast and guided inspection and correction mechanism.

We will integrate all the mentioned techniques in a scalable framework for Arabic text analysis.
 The framework as a whole will allow for exhausting Arabic text analysis and allow for extensions of the framework.
 A user should be able to iteratively and repetitively apply the abode techniques on his input text.

In the development cycle, we will use an Extreme programming methodology in the process of building the Arabic text mining framework.
 We will use open source tools, libraries, and frameworks to build our libraries.

Our research will concentrate on NLP techniques native to the Arabic language to analyze the Arabic text and we will avoid preprocessing that feeds common text and data mining techniques.
 We will build on existing open source programming and text mining tools as well as open source Arabic text analysis tools based on NLP such as MADA+TOKAN~\cite{Rot08}, NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.

We will use a dokuwiki based website~\cite{Dok09} to document the progress online and will use subversion~\cite{Sub09} as a source control tool.

\section{Anticipated results and evaluation criteria}
\label{s:results}

We anticipate that the research will lead to the production of an Arabic text analysis framework that allows relational queries between Arabic text documents.
 The components of the framework are listed in Section "Research Design and Methods" and are repeated for convenience as follows.
 \begin{enumerate}
 \item Lexical analysis library 
 \item Syntax analysis library
\item Dynamic modern Arabic dictionary
\item An entity extraction library
\item A relation extraction library
\item A logic solver 
\item Tagger
\end{enumerate}

We will use these tools to handle several case studies from different fields.
 In the following we discuss three case studies that we anticipate our proposed framework to solve.

{\bf The literature case study}
Assume we have two sets of texts.
 The first set is several books of historical accounts where each account is preceded by a chain of narrators who narrated the account.
 The second set is several books of biographies where each biography of a person may include an evaluation of the authenticity of that person.
 We want a tool that can answer a complex query such as relating all historical accounts in the first set that discuss women and vehicles (such as tamed animals) to the narrators of the accounts, and then relate these narrators to the second set of books of biographies with an emphasis on authenticity evaluations.

We need tools that can search and index the historical accounts to return a subset that relates to women and travel.
 Then we need tools that can identify names, and succession relations between names to build a chain of narrators for each historical account in the subset.
 We need tools that can identify the names in the biographies and detect evaluation adjectives.

Provided we have these tools we can exhaustively at the click of a button visualize the authenticity of all the historical accounts that may allow or forbid women from driving according to Islamic laws.
 We can go one step further to detect locations and dates and relations based on chronological orders and further question the consistency of the historical accounts.
 Such query takes currently huge amounts of manual effort that is also prone for error and far from being exhaustive.

We here envision that text mining applied to the authenticity of historical accounts will have a huge impact in the field.

Note that most of the Islamic history and Hadith books as well as a huge library of Arabic literature books are already digitized and available on the internet.

{\bf The security case study}

Given a huge set of investigation reports, an officer may throw a list of names of convicted people and check all the reports for possible relations between them in terms of criminal action, locality, or third party names unknown to the officer who happen to know the convicts.
 The query may return at the click of a button, two convicted people who had no direct relation in the reports, but who are related only through a third person unknown to the investigator.
 Such a query is very hard to answer manually.

{\bf The health case study}

Given a set of clinical reports gathered by the Order of Medical Doctors, the ministry of health may query to relate all reports on a specific sickness to the doctors treating it and evaluate the response to the different types of medications as classified in different areas in the country each with a different climate.
 Such query, while possible at a click of a button with text mining tools if the data is available in digital form, would take a huge manual effort to conduct surveys, collect and gather the data.
 In this case, digitizing the data may be a less expensive effort.

\subsection{Evaluation Criteria}

The Lead PI will conduct biweekly meetings to track progress on the project.
 The team will maintain a dokuwiki based website to continuously post updates on the progress of the project.
 In Section "Research Design and Methods" we presented a schedule for delivering the different products of the project.
 The schedule can be used as a timeline for measuring the progress of the project.

The lexical and syntax analysis libraries will be compared with current open source libraries against commonly used corpora of data such as a collection of books or a collection of news reports.
 The dictionary module will be an augmentation over existing digitized dictionaries such as the Buckwalter~\cite{Tim04} morphological analyzer.
 In the entity and relational extraction tools we will use techniques native to the Arabic language and we will compare against English entity and relational extraction tools of similar or translated texts.
 We will also compare techniques that pre-process Arabic and then pass it to English IE and RE techniques.
 The logic solver will build over the IE and RE tools and will use off-the shelf quantifier and propositional solvers such as MiniSat~\cite{Een03}.
 The tagger is a GUI interface that allows fast manual inspection and refinement of the automated results.
 Each tool in the framework can start on its own with a prototypical implementation of the other needed tools.

Each tool in the framework stands alone and can operate on a well defined input and generate the desired output.
 The input of one tool, such as the entity extraction, can be automatically generated by another tool, such as the dictionary and the syntax analyzer.
  However, the input can also be manually generated if desired using the straightforward GUI tagger facility.
 Thus the tagger facility stands as a desired utility on its own and at the same time represents a contingency plan in case any module in the framework failed to deliver.

The team will publish the results of the research in the form of refereed conference and journal papers as well as tool papers addressing each of the tools.
 The team will also hold workshops to educate possible users on the tools once the tools are available.

\section{Strategy for project continuation}
\label{s:continue}

Once completed the Arabic text analysis framework will be an open source contribution and developers from the Arabic world in the open source community are expected to join efforts to keep it going.
 Furthermore, once completed the Arabic text analysis framework can sustain itself by charging on support or expert use.
 The framework or service on the framework can easily be commercialized to tend to the security, health, insurance, news, media, or literature business sectors.

The area of Arabic text analysis will be a research area for a long time as NLP is one of the most complex computational problems.
 The Arabic text analysis framework will be used by NLP researchers in the Arabic language field and those are expected to maintain and advance the framework.

\section{Plans for disseminating research results}
\label{s:dissem}

The team will publish the results of the research in the form 
of refereed conference and journal papers as well as tool papers 
addressing each of the tools. 
The team will also hold workshops to educate possible users on the 
tools once the tools are available.
The Arabic text analysis framework will be partially based on 
existing work in the open source domain and will thus be available 
as an open source tool. 
Interested researchers and users will be able to download the tools 
for free and use them for non-profit non-commercial purposes.

\pagebreak
%\bibliography{adnan_refs}
%\bibliographystyle{ieeetr}
\bibliographystyle{abbrvnat}
%\bibliographystyle{ieeetr}
{\small
\bibliography{fzAr}  
}

\end{document}

### 2010-12-05 trunk/tex/rp.tex
monday changes

@@
 
-{\bf Hypothesis 1.} We hypothesize that many case studies, in particular relational
+{\bf Hypothesis 1.} We hypothesize that many case studies, 
+in particular relational
 case studies, do not require high accuracy from the 
@@
 
-{\bf Contributions.} Up to our knowledge, CATARM will be the 
+{\bf Goals.} Up to our knowledge, CATARM will be the 
 first framework that tends to relational Arabic morphology 
@@
 
-This proposal aims at building  a relational Arabic text mining 
-framework with the following objectives.
+The current prevailing direction in research 
+in Arabic text mining and NLP focuses on machine translation from 
+Arabic to English. 
+A little attention is given to native Arabic text mining 
+problems and applications. 
+CATARM is a relational Arabic text mining 
+framework that tends to Arabic users 
+with the following objectives.
+
 \begin{itemize}\itemsep0pt
-\item CATARM tends to Arabic users and solves the
-problem of finding relations between Arabic documents. 
-\item CATARM builds a formal Arabic computational model
+\item CATARM will build a formal Arabic computational model
 using efficient and configurable data structures such as 
 finite state machines (FSM). 
-\item The framework will provide the user with a visual query 
+\item CATARM will provide the user with a visual query 
 language that makes it easy to enter relational queries. 
@@
 computational model. 
-\item The framework will use statistical models and techniques
+\item CATARM will use statistical models and techniques
 to infer language independent features from the text documents
@@
 and graph analysis techniques, and  text algorithms.
+\item We will use CATARM to solve several realistic case studies.
+We will look at the hadith literature and build a system using
+CATARM to relate the hadith literature to the biographies of the
+narrators and extract an authenticity graph for the narrators.
+We will look at investigation and interrogation security reports
+and build a CATARM system that looks for relations between suspects
+in the reports. 
+We will look at clinical reports and relate them to pharmaceutical 
+records and find relations between sicknesses and prescribed medication. 
+Finally, we will build a system using CATARM to check the strength
+of the relations between internet documents viewed by a minor and
+a set of Arabic text documents containing unrated or unwanted 
+material. If the relation is strong, the system will flag access to
+the document. 
+All components and parts built during the development of solutions
+to the case studies will be generalized and added to CATARM as 
+library computational components. 
 \end{itemize}
 
-CATARM will be measured and validated against several 
-case studies and will contain  
-all the components needed to 
-address the following case studies.
+%\begin{itemize}\itemsep0pt
+%\item {\bf The literature case study.}
+%We are given a set of books of historical accounts where
+%a chain of narrators precedes each account,
+%and a second set of books of biographies where
+%a biography includes an evaluation of the authenticity 
+%of the narrator therein.
+%CATARM will allow the user to automatically and exhaustively query 
+%all accounts related to a subject important to Islamic jurispudence 
+%(for example women and travel),
+%and extract a graph where nodes represent narrators and edges
+%are labeled with the authenticity of the narrators. 
+%A similar manual query takes ages to answer and is error prone.
+%\item {\bf The security case study.}
+%Given two sets of reports consisting of crime scene investigation 
+%and suspect interrogation reports, 
+%an officer may query CATARM 
+%for relations between suspects in terms of 
+%criminal action, locality, or third party persons unknown to the 
+%officer.
+%Such a query is very hard to answer manually.
+%\item {\bf The net traffic case study.}
+%Given a set of undesired internet content for minors
+%such as violent text, and a set of documents requested by internet
+%minor users, 
+%CATARM can discover whether there is a relation between 
+%the document and the violent content and flag the document if the
+%relation is strong enough. 
+%\end{itemize}
 
-\begin{itemize}\itemsep0pt
-\item {\bf The literature case study.}
-We are given a set of books of historical accounts where
-a chain of narrators precedes each account,
-and a second set of books of biographies where
-a biography includes an evaluation of the authenticity 
-of the narrator therein.
-CATARM will allow the user to automatically and exhaustively query 
-all accounts related to a subject important to Islamic jurispudence 
-(for example women and travel),
-and extract a graph where nodes represent narrators and edges
-are labeled with the authenticity of the narrators. 
-A similar manual query takes ages to answer and is error prone.
-\item {\bf The security case study.}
-Given two sets of reports consisting of crime scene investigation 
-and suspect interrogation reports, 
-an officer may query CATARM 
-for relations between suspects in terms of 
-criminal action, locality, or third party persons unknown to the 
-officer.
-Such a query is very hard to answer manually.
-\item {\bf The net traffic case study.}
-Given a set of undesired internet content for minors
-such as violent text, and a set of documents requested by internet
-minor users, 
-CATARM can discover whether there is a relation between 
-the document and the violent content and flag the document if the
-relation is strong enough. 
-\end{itemize}
+\subsection{Significance}
 
-{\bf Significance.} 
 CATARM will discover relations between Arabic text documents that 
 are hard to discover by humans and will have direct live and
-cost saving benefits in cases similar to the above case studies. 
+cost saving benefits in cases similar to the above mentioned 
+case studies. 
 CATARM will qualitatively enhance research in the cultural and 
@@
 \pagebreak
-u
+
 \section{Preliminary data or studies}
@@
 
-Our preliminary studies are promising.
-With a small manually entered dictionary, few programmed grammar rules and patterns, and two hardcoded queries we were able to solve selected samples from the literature and security case studies presented in Section "Significance".
-These preliminary results serve as a good proof of concept to our hypothesis.
-They also confirm that our approach performs better than currently existing techniques from Sakhr (Sakhr Inc.
-2009) and Basis (Basis technology 2009) as well as from open source tools (Columbia's Arabic Dialect Modeling Group 2009).
+The PI and his team at the American University of Beirut (AUB)
+started work on an Arabic text mining framework~\cite{ATMine09}
+that is available now online.
+They have successfully developped ATSarf, a case-based morphological 
+analyzer, and applied it on the hadith literature case study.
 
-The Co-Lead PI worked and published extensively in this domain. A selected list of her recent publications follows. 
-    Rehab Duwairi, Mohammad Al-Refai, Natheer Khasawneh, "Feature Reduction Techniques for Arabic Text Categorization".
-Journal of the American Society for Information Science and Technology (JASIST), Volume 60, Issue 11, pages: 2347-2352, 2009.
-        R. M. Duwairi, "Arabic Text Categorization", International Arab Journal for Information Technology (IAJIT), Volume 4, Number 2, pages 125  132, 2007.
-            R. M. Duwairi, "Machine Learning for Arabic Text Categorization". Journal of the American Society for Information Science and Technology (JASIST), Vol. 57, Issue 8, pages 1005-1010, 2006.
+A \RL{.hady_t} is a narration related to the prophet Mohammad
+through a \RL{sanad} or a sequence of narrators. 
+Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
+transliteration and translation. 
+The authentication of a \noArRL{.hady_t} highly depends on the credibility
+of each of the narrators as reported in separate biography 
+books. 
+In this paper we consider the problem of automatically segmenting
+a \noArRL{.hady_t} book into narrations, then segmenting each 
+narration into
+its content or \RL{matn} and its \noArRL{sanad}.
+We also partition the sanad accurately into the 
+separate narrators so that we can later look each one of them 
+up in the biography books. 
+For example, the boxes in Figure~\ref{f:exhadith} are proper names 
+and once
+connected toghether they form complex names of narrators. 
+Narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
+name is \noTrRL{s`yd} as the word in between 
+\noTrRL{bn} (son of) indicates.
 
-The Lead PI worked in the area of software arabization since 1996 as he worked on developing Arabization modules for Windows NT and an Arabic string manipulation library (Zaraket, Zbib and Khalil, Yaraah: Arabization module for Windows NT 1996).
-The Lead PI is also involved in working on an open source project that aims at comprehensive automation of authenticity checks against a corpus of literature heritage (Zaraket, Zeineddine, et al.
-2006).
-The Lead PI is leading an effort at AUB to introduce Computational Arabic in the curriculum of the ECE program.
-He is also leading a team of students in an effort to place the infrastructure needed for this project to start.
-            The lead PI worked and published about relational logic and logic solvers applied to structured languages as follows.
-                Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
-                    Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007.
+\transfalse
+\begin{figure}[tb]
+\center{
+\resizebox{.7\columnwidth}{!}
+{ \input{figs/exhadith.pdftex_t}}
+\caption{Hadith abstraction example.}
+\label{f:exhadith}
+}
+\end{figure}
+\transtrue
+
+The framework now comprises a state of the art
+case-based morphological analyzer, ATSarf, 
+and computational components that allow the user to
+specify his queries as one or more FSM
+controlling the morphological analyser.
+We explain this approach in details in Section~\ref{s:design:fsm}.
+ATSarf along with other computational components, 
+extract the chains of narrators from a set of narrations
+with above 95 percent accuracy and in less than two minutes.
+
+\begin{figure}[tb]
+\center{
+\resizebox{.75\columnwidth}{!}
+{\includegraphics{figs/narrator_chain_output.pdf}
+\label{f:narrators} } 
+\caption{Directed acyclic graph representing a partial order 
+    relation between narrators extracted using ATSarf}
+}
+\end{figure}
+
+The directed acyclic graph (DAG) 
+in Figure~\ref{f:narrators} shows a partial order relation between
+narrators that was automatically extracted from hadith text 
+documents.
+Given a set of chains of narrators similar to 
+Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
+that merged equal names in the chains. 
+This partial order graph is instrumental to automatically
+locate narrators in biography documents and automatically
+segment biography documents.
+
+These preliminary results serve as a good proof of concept to our 
+hypothesis. They also confirm that our approach performs better than 
+currently existing techniques from Sakhr~\cite{Sak09},
+Basis~\cite{Bas09} and open source 
+tools~\cite{Col09,Otakar:07,Tim04}.
+
+We find evidence in our current results that features 
+of the Arabic language such as the hierarchical structure of
+names can be used to simplify the morphological analysis
+needed to extract the partial order graph. 
+Publications about these results are pending acceptance in 
+related NLP conferences.
+
+The Co-Lead PI published extensively in this domain. 
+A selected list of her recent publications follows. 
+\begin{itemize}\itemsep0pt
+\item Rehab Duwairi, Mohammad Al-Refai, Natheer Khasawneh, 
+"Feature Reduction Techniques for Arabic Text Categorization". 
+Journal of the American Society for Information Science and Technology 
+(JASIST), Volume 60, Issue 11, pages: 2347-2352, 2009. 
+\item Rehab M. Duwairi, "Arabic Text Categorization", 
+International Arab Journal for Information Technology (IAJIT), 
+Volume 4, Number 2, pages 125  132, 2007.
+\item Rehab M. Duwairi, "Machine Learning for Arabic Text Categorization". 
+Journal of the American Society for Information Science and Technology 
+(JASIST), Vol. 57, Issue 8, pages 1005-1010, 2006.
+\end{itemize}
+
+%The Lead PI worked in the area of software arabization since 1996 
+%as he worked on developing Arabization modules for Windows NT and 
+%an 
+%Arabic string manipulation library~\cite{Zar96}. 
+%The Lead PI is also involved in working on an open source 
+%project that aims at comprehensive automation of authenticity 
+%checks against a corpus of literature heritage~\cite{Zaraket06}.
+%The Lead PI is leading an effort at AUB to introduce 
+%Computational Arabic in the curriculum of the ECE program. 
+%He is also leading a team of students in an effort to place 
+%the infrastructure needed for this project to start. 
+%The lead PI worked and published about relational logic and 
+%logic solvers applied to structured languages as follows.
+%\begin{itemize}
+%\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
+%\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
+%\end{itemize}
 
@@
 
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.75\columnwidth}{!}
+{\includegraphics{figs/literature_case-study1.pdf}
+\label{f:literature}
+} 
+\caption{CATARM diagram for the literature case study }
+}
+\end{figure}
+
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.8\columnwidth}{!}
+{\includegraphics{figs/security_case-study.pdf}
+\label{f:security}
+}
+\caption{CATARM diagram for the security case study}
+}
+\end{figure}
+
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.8\columnwidth}{!}
+{\includegraphics{figs/detailed_submachine.pdf}
+\label{f:chain}
+}
+\caption{CATARM detailed diagram for the chain of narrators extractor}
+}
+\end{figure}
+
 We see more and more Arabic documents such as texts,
- books,
- publications,
- hospital and governmental records emerging every day.
- Most of the newly generated documents are produced in digital textual form while also old paper documents are ported to digital form.
- These documents include huge amounts of information that is qualitatively different than structured information such as that contained in database entries.
- Text mining is the technology that automates the discovery of information in non structured text.
+books, publications, hospital and governmental records emerging 
+every day.
+Most of the newly generated documents are produced in digital 
+textual form while also old paper documents are ported to digital 
+form.
+These documents include huge amounts of information that is 
+qualitatively different than structured information such as that 
+contained in database entries.
+Text mining is the technology that automates the discovery of 
+information in non structured text.
 
@@
 the extraction of concepts and entities from text,
- and the discovery and modeling of relations between text entities and classes.
- Text mining techniques perform natural language processing (NLP) to analyze text and perform information extraction (IE) to extract information into data structures.
- Fundamental to IE are the name entity (NE) and the name entity relation (NER) extraction techniques which capture features from sentences and phrases.
- A key challenge to NE and NER is that sentence structures and words are often ambiguous in natural languages.
- While research to address ambiguity in Latin languages is still lagging behind~\cite{Red08},
- not much has been done to address NE and NER in the context of the Arabic language.
+and the discovery and modeling of relations between text entities 
+and classes.
+Text mining techniques perform NLP to analyze text and perform 
+IE to extract information into data structures.
+Fundamental to IE are the name entity (NE) and the name entity 
+relation (NER) extraction techniques which capture features from 
+sentences and phrases.
+A key challenge to NE and NER is that sentence structures and 
+words are often ambiguous in natural languages.
+While research to address ambiguity in Latin languages is still 
+lagging behind~\cite{Red08},
+not much has been done to address NE and NER in the context of 
+the Arabic language.
 
-The application of text mining techniques to Arabic text documents will result in great benefits to many research fields such as Arabic literature,
- Islamic studies,
- Hadith authentication,
- and Arabic history and culture in general.
- Arabic text mining will also benefit sectors where Arabic text documents are key such as the security sector,
+The application of text mining techniques to Arabic text documents 
+will result in great benefits to many research fields such as 
+Arabic literature,
+Islamic studies, Hadith authentication, and Arabic history and 
+culture in general.
+Arabic text mining will also benefit sectors where Arabic text 
+documents are key such as the security sector,
 the government personal records,
@@
 
-The difficulty of applying text mining techniques to Arabic text documents lays in the absence of automated tools that understand the unique features of the Arabic language.
+The difficulty of applying text mining techniques to Arabic text 
+documents lays in the absence of automated tools that understand 
+the unique features of the Arabic language.
 We will make use of an example to illustrate the process.
- Given an Arabic text document we want a tool that can segment and isolate all names,
- dates,
- time of the day occurrences,
- tool names,
- and locations as entities.
- Then we want another tool to identify identities amongst these entities,
- and relate them in their order of appearance in the document.
+Given an Arabic text document we want a tool that can segment 
+and isolate all names, dates, time of the day occurrences,
+tool names, and locations as entities.
+Then we want another tool to identify relations amongst 
+these entities such as identities and orders.
+An order between two entities can be simply 
+their order of appearance in the document, or their
+numerical 
  We want another tool to look at the verbs and actions in the sentences and try to build relations between the identified entities based on these verbs.

### 2010-12-06 tex/rp.tex
method

--- 2010-12-05 trunk/tex/rp.tex
@@
 books. 
-In this paper we consider the problem of automatically segmenting
+We consider the problem of automatically segmenting
 a \noArRL{.hady_t} book into narrations, then segmenting each 
 narration into
-its content or \RL{matn} and its \noArRL{sanad}.
+its content or \RL{matn} and its \noArRL{sanad} or the
+chain of narrators.
 We also partition the sanad accurately into the 
@@
 The directed acyclic graph (DAG) 
-in Figure~\ref{f:narrators} shows a partial order relation between
-narrators that was automatically extracted from hadith text 
+in Figure~\ref{f:narrators} shows a partial order relation (POR) between
+narrators.
+We automatically extracted the POR from hadith text 
 documents.
+The nodes in boxes are the \RL{matn} of the hadith, 
+and the other nodes are the narrators.
 Given a set of chains of narrators similar to 
@@
 that merged equal names in the chains. 
-This partial order graph is instrumental to automatically
-locate narrators in biography documents and automatically
-segment biography documents.
+This partial order graph is instrumental to automate
+localizing narrators in biography documents and
+segmenting biography documents.
+
+Given that a biography will discuss a narrator and mention
+his professors and students,
+we can segment the biographies with a graph coloring algorithm 
+that traverses the text and colors the POR whenever
+a name is found. 
+We modify the color when we move in the graph 
+far away from the last name we colored.
+
+Once the biographies are analyzed, one can annotate
+the POR with qualifiers of the narrators that reflect
+their authenticity. 
+One can also annotate the POR with the locations and 
+the time they lived in. 
+Then One can compute a time and location overlap
+check to discover inconsistent narrations.
+We can also perform several interesting checks using 
+this POR on its own such as checking for the effect of
+one narrator on the hadith literature.
+
+We think that the ability to extract such orders
+between entities in documents can
 
@@
 IE to extract information into data structures.
-Fundamental to IE are the name entity (NE) and the name entity 
+Fundamental to IE are the named entity (NE) and the named entity 
 relation (NER) extraction techniques which capture features from 
@@
 words are often ambiguous in natural languages.
-While research to address ambiguity in Latin languages is still 
-lagging behind~\cite{Red08},
-not much has been done to address NE and NER in the context of 
-the Arabic language.
+%While research to address ambiguity in Latin languages is still 
+%lagging behind~\cite{Red08},
+%not much has been done to address NE and NER in the context of 
+%the Arabic language.
 
@@
 the unique features of the Arabic language.
-We will make use of an example to illustrate the process.
+%We will make use of an example to illustrate the process.
 Given an Arabic text document we want a tool that can segment 

### 2010-12-07 trunk/tex/rp.tex
wrote some description for the detailed statemachine diagram

@@
 \newcommand{\noTrNoVocRL}[1]{\novocalize\transfalse\RL{#1}\transtrue\vocalize} 
+%\newcommand{\drawline}{\begin{picture}(6,.1) \put(0,0) {\line(1,0){6.25}}\end{picture}}
 
@@
 
+%\drawline
 %\begin{figure}[tb]
@@
 \end{figure}
+Consider Figure~\ref{f:chain}, which depicts a possible rendering of a query that transforms a hadith text file into a list of chain of narrators.
+Note that this diagram does not faithfully translate the state machine we are currently implementing to generate our priliminary results which is represented in  Figure~\ref{f:statemachine}.
+Instead, it has been simplified in order not to clotter the diagram.
+We start first by defining the legend: we have used boxes to represent computational blocks provided by CATARM, while ovals represent structured or unstructured data which are mainly the inputs and the outputs of the computational blocks.
+When the data blocks are dotted they represent temporary values, while when they are solid and shaded they represent the initial input and the final output.
+When a data block represents an array of values, the oval will be reproduced three times to show that multiple values exist.
 
+This query generates the list of chains by first passing the hadith book text file into a Morphological analyzer which is controlled by an entity extraction controller which in this case detects Names (of people) and output such a word, its corresponding position in the text and its syntax category which must be a Name category.
+Then we "Iterate" over this list compoputing the distance between the positions of 2 consecutive words.
+This distance is inputed into a thresholding block which returns the appropriate values for the "Less" and "Bigger" outputs.
+Having "Less" as TRUE, means that the distance is less than the provided threshold $\theta_1$ (which is a modifiable parameter of the block).
+In that case, we add the word to a temporary data block representing a chain (i.e.
+a list of narrator names).
+Otherwise, we execute a sequential block which in order contains: (1) a "Table" block which tabulates the chain into a temporary table/array containing the list of chains.
+(2) a "Create" block which initializes a new "Chain" whith the output of the second iterator (i.e.
+the name whose distance from its previous one is larger than the threshold).
+Finally, the "Filter" block outputs only the chains whose cardinality is larger than some threshold  $\theta_2$.
+
+In this way, the above query adds names to the chain temporary variable until a name which is distant is detected.
+Then, we consider the valid chains as those which contain a minimum number of names.
+CATARM allows the librarization of this query into a user-defined computational block that can be used later in more complex queries as will be shown below.
+Thus, the details of the implementation of the block will be hidden and the block's interface will be only the shaded data blocks in the figure.
+In what follows, we assume that this query has been librarized into a "Narrator Chain Extractor" computational block to be used in Figure~\ref{f:literature}.
+
+\begin{figure}
+\center{
+\resizebox{.4\columnwidth}{!}
+{\input{figs/hadith.pdftex_t}
+\label{f:statemachine}
+}
+\caption{State-machine used to detect the chain of narrators}
+}
+\end{figure}
+
+\pagebreak
+
+%\drawline
 We see more and more Arabic documents such as texts,

### 2010-12-07 tex/rp.tex
good shape

@@
 
-We think that the ability to extract such orders
-between entities in documents can
-
 These preliminary results serve as a good proof of concept to our 
@@
 
-%\begin{figure}[tb]
-\begin{figure}
-\center{
-\resizebox{.75\columnwidth}{!}
-{\includegraphics{figs/literature_case-study1.pdf}
-\label{f:literature}
-} 
-\caption{CATARM diagram for the literature case study }
-}
-\end{figure}
-
-%\begin{figure}[tb]
-\begin{figure}
-\center{
-\resizebox{.8\columnwidth}{!}
-{\includegraphics{figs/security_case-study.pdf}
-\label{f:security}
-}
-\caption{CATARM diagram for the security case study}
-}
-\end{figure}
-
-%\begin{figure}[tb]
-\begin{figure}
-\center{
-\resizebox{.8\columnwidth}{!}
-{\includegraphics{figs/detailed_submachine.pdf}
-\label{f:chain}
-}
-\caption{CATARM detailed diagram for the chain of narrators extractor}
-}
-\end{figure}
-
 We see more and more Arabic documents such as texts,
@@
 
-The difficulty of applying text mining techniques to Arabic text 
-documents lays in the absence of automated tools that understand 
-the unique features of the Arabic language.
+%The difficulty of applying text mining techniques to Arabic text 
+%documents lays in the absence of automated tools that understand 
+%the unique features of the Arabic language.
 %We will make use of an example to illustrate the process.
@@
 these entities such as identities and orders.
-An order between two entities can be simply 
-their order of appearance in the document, or their
-numerical 
- We want another tool to look at the verbs and actions in the sentences and try to build relations between the identified entities based on these verbs.
+We want another tool to look at the verbs and actions in 
+the sentences and try to build relations between the identified 
+entities based on these verbs.
 If we were successful to do that,
- we now want to order all entities associated with dates based on a chronological order.
+we now want to order all entities associated with dates based 
+on a chronological order.
+
+An order between two entities can depend on 
+their simple order of appearance in the document, their 
+numerical values in case they happened to be numbers, or
+the distance between them in case they were cities. 
+The relations can also be more sophisticated and 
+based on templates provided by users or by 
+models extracted from training data sets.
 
 To do the above, 
- we need tools that are able to automatically differentiate between names,
- verbs,
- adjectives and other vocabulary structures.
+we need tools that automatically differentiate between 
+names, verbs, adjectives and other structures.
 This is not a simple task with the Arabic language.
- For example,
- Arabic grammar allows name-based sentences or sentences with no 
-    verbs \noTrNoVocRL{^gml ismiyT} .
+For example, Arabic grammar allows name-based sentences 
+or sentences with no verbs \noTrNoVocRL{^gml ismiyT} .
 Another unique feature is the possibility to have names in Arabic 
@@
 \noTrNoVocRL{alm^staq-at w al^g_dwr}.
- We also need to build dictionary tools that associate words and phrases based on their meanings and their context.
+Simple primitive tasks such as comparing two strings need 
+special attention in Arabic since each Arabic letter has 
+four different forms depending on its position in the word; 
+isolated form, start, middle and end of a word.
+Arabic diacritics should also be ignored most of the time 
+when comparing two strings.
+We also need to build dictionary tools that associate words 
+and phrases based on their meanings and their context.
 Often time,
 and depending on the application and the user of the tool,
- we need to allow the user to assign meanings and semantics for patterns in the text.
-
-In the following we discuss our research methodology to develop the proposed Arabic text analysis tools.
-
-Lexical analysis library: We need a library of routines that enables primitive lexical analysis of Arabic text.
- For example,
- simple primitive tasks such as comparing two strings need special attention in Arabic since each Arabic letter has four different forms depending on its position in the word; isolated form,
- start,
- middle and end of a word.
- Arabic diacritics should also be ignored most of the time when comparing two strings.
- These lexical analysis routines should also pay attention to the form of a letter when used in tokenization as a letter in and end of word form denotes the end of the word even if it was not followed with a space.
- Such a library can benefit from existing Unicode enabled string libraries such as Qt.
- It also can provide a layer of abstraction that hides the details of the encoding of Arabic letters used in the text documents.
-
-Syntax analysis library: This library should allow sentence structure understanding.
- We will explore the best way to develop this library.
- Several options exist such as pattern based matching,
- tokenization and dictionary based lookups,
- or a language grammar that allows more than one matching rule.
- This library needs to be extendible to allow additions of new rules or patterns.
-
- (Note the ElixirFM does this)
- We will explore Arabic grammar books such as The Principles of Arabic~\cite{Sha73} and others~\cite{Abd00,Abd001}
- to augment current morphological analyzers~\cite{Tim04} and syntax analysis tools~\cite{Col09}.
- We will use relaxed NLP techniques to overapproximate results.
- Then and after a relational query,
- we will apply statistical analysis techniques to rank the matching results.
-
-Dynamic modern Arabic dictionary: The dictionary will allow identities based on semantics,
- derivation rules,
- as well as user assigned rules.
- The dictionary will return modern grammatical categories for words classifying them as nouns,
- verbs,
- and adjectives instead of the classical Arabic categories that specify how the word is derived from an original radix.
- It remains to be explored how to represent these categories without exploding the size of the dictionaries with redundant words of multiple categories.
- This work will augment existing open source Arabic dictionaries and will allow for better morphological analysis and entity extraction.
-
-An entity extraction library: We will explore building a library that allows extraction of entities from Arabic text.
- These entities can be proper names,
- dates,
- locations,
- and qualification adjectives to list few possibilities.
- The extractor will use the dynamic modern Arabic dictionary as well as the syntax library to extract entities from a sentence and to establish identity between entities across sentences.
-
-A relation extraction library: We will explore building a library that allows extraction of relations from Arabic text.
- We will explore the use of pairing techniques to establish relations between the extracted entities.
- The relations will form graphs where the nodes are the entities and the previously extracted relations and the edges are the newly discovered relations.
- The routines in this library will take as input a template for an expected relation and will return a graph that describes whether this relation exists between the subject documents and where.
- These relations will be ranked using statistical analysis.
- The statistical criterion and the ranking methods remain to be decided.
-
-A logic solver: This is a logical solver that will solve query expressions to resolve matching text and relations.
- We will explore the proposal of a logical query language that is restricted enough to be computable and expressive enough to allow interesting properties.
- We will use of the shelf solvers and matches from the entity and relation extractors to evaluate the expression.
- The choice of solvers remains to be decided after the query language is formed.
-
-Tagger: We will explore the development of an automatic tagging interface that allows fast and guided inspection and correction mechanism.
-
-We will integrate all the mentioned techniques in a scalable framework for Arabic text analysis.
- The framework as a whole will allow for exhausting Arabic text analysis and allow for extensions of the framework.
- A user should be able to iteratively and repetitively apply the abode techniques on his input text.
-
-In the development cycle, we will use an Extreme programming methodology in the process of building the Arabic text mining framework.
- We will use open source tools, libraries, and frameworks to build our libraries.
-
-Our research will concentrate on NLP techniques native to the Arabic language to analyze the Arabic text and we will avoid preprocessing that feeds common text and data mining techniques.
- We will build on existing open source programming and text mining tools as well as open source Arabic text analysis tools based on NLP such as MADA+TOKAN~\cite{Rot08}, NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
-
-We will use a dokuwiki based website~\cite{Dok09} to document the progress online and will use subversion~\cite{Sub09} as a source control tool.
+we need to allow the user to assign meanings and semantics 
+for patterns in the text.
+
+\subsection{Literature authentication case study}
+\label{s:design:lit}
+
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.75\columnwidth}{!}
+{\includegraphics{figs/literature_case-study1.pdf}
+\label{f:literature}
+} 
+\caption{CATARM diagram for the literature case study }
+}
+\end{figure}
+
+The diagram in Figure~\ref{f:literature} describes our 
+high level vision for what the literature case study
+will look like in the CATARM framework. 
+The boxes represent 
+computational blocks provided by CATARM, 
+ovals represent data blocks which are mainly the 
+inputs and the outputs of the computational blocks. 
+Dotted data blocks represent intermediate results,
+and solid and shaded blocks represent the initial input 
+and the final output respectively.
+
+The computational components can be direct CATARM components
+or custom user components built from other CATARM components.
+We feed the hadith books to a user Narrator Chain Extractor
+component that we will discuss later in 
+Section~\ref{s:design:chainext}. We  obtain a set of chains
+of narrators that we feed to a morphological analyzer (MA) 
+component controlled by a merger controller. 
+The merger controller will look at the morphological analysis
+of the narrator names and will decide to merge two narrators in 
+one graph node in case they were morphologicaly equal. 
+It will also give feedback to the MA component to stop the
+analysis when a it reaches a satisfactory result.
+We consider this graph as a partial order relation between 
+narrators where a narrator precedes another narrator if 
+he narrated from him.
+
+The biography books are fed to another MA unit controlled by 
+a name and qualification pair controller that generates a 
+map consisting of pairs of narrator names associated with 
+qualifications. 
+Then the graph of narrators and the narrator qualification map
+are both processed through a locator component that computes
+a graph coloring routine to color the graph of narrators. 
+The color changes whener two consecutive narrators are far away
+from each other in the text. 
+The locator then looks at connected subgraphs of the same color,
+and finds a median node in each of them to be the narrator 
+associated with the qualifiers of the color. 
+Then the locator also annotates the narrator with the 
+corresponding qualifications. 
+The result is an annotated partial order graph of narrators
+with authentication qualification annotations.
+
+\subsection{Security case study}
+\label{s:design:sec}
+
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.8\columnwidth}{!}
+{\includegraphics{figs/security_case-study.pdf}
+\label{f:security}
+}
+\caption{CATARM diagram for the security case study}
+}
+\end{figure}
+
+The diagram in Figure~\ref{f:security} describes our 
+high level vision for what the solution of the 
+security case study
+will look like in the CATARM framework. 
+The inputs are two separate sets of documents. 
+The first set includes crime scene investigation reports, 
+and the second set includes a 
+interrogation reports with crime suspects as well as 
+a database of previous interrogation reports. 
+The CATARM system also takes as input a database of geographical
+locations and a database of personal records. 
+The aim is to check for inconsistencies in the iterrogation
+reports and to identify possible connections between the suspects
+and the crime scene.
+
+An MA with date, place, and name extraction controller extracts
+names, dates and places from the investigation reports. 
+We pass the dates to a sorting component to produce $G_D$ 
+a partial order graph between the dates. 
+We look up the names in the personal records database and find out
+$G_N$ a family dependency graph between the names. 
+We pass the places to a mapper that builds another $G_G$,
+a graph relating the places using a geographical database.
+
+We pass the interrogation reports to an MA with a relation 
+extraction controller that builds $G_R$. 
+The graph $G_R$ relates the interrogated suspects
+to each other, to places, and to dates based on a simple
+neighborhood threshold computed using a statistical feature.
+
+We feed $G_D, G_F, G_G,$ and $G_R$   
+to a consistency checker.
+The consistency checker considers the three dimensions 
+$N$ names, $G$ locations, and $D$ dates and projects $G_R$
+over each one of them.
+For each dimension $k$, it computes a local measure 
+$\alpha_k$ of how much $G_R|_k$ simulates $G_k$. 
+It flags the components in the graph where the accumelated 
+measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
+below a threshold $\theta$ as inconsistent.
+
+\subsection{CATRAM components}
+\label{s:design:components}
+
+To do the above we need CATARM to provide us with several 
+computational components such as a morphological analyzer, 
+statistical tools, graph algorithms, and other computational
+components. 
+In the following we discuss the components needed for 
+CATARM and we propose a methodology to 
+use and extend existing tools, and develop new tools to
+build the components of CATARM.
+
+\subsection{Case-based morphological analyzer}
+
+Morphological analyzers for the Arabic language
+such as Buckwalter~\cite{Buckwalter:02},
+Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+and ElixirFM~\cite{Otakar:07} take an Arabic word as input
+and produce several valid morpholgical solution for the word. 
+Each solution is a sequence of morphemes that form the word.
+Each morpheme is associated with several tags such as 
+morphological, part of speech, grammar, and semantic tags. 
+These analyzers are essential in several text analysis tools such 
+as spell checkers as well as NLP frameworks~\cite{Col09}.
+Other morphological analyzers such as 
+Amira~\cite{Diab:07,Benajiba:07},
+MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
+use machine learning and support vector machines (SVM) 
+to enhance the accuracy at the expense of efficiency.
+
+The current use of morphological analyzers suffers from 
+several problems. 
+Analyzers that enumerate all possible partitions of a word
+and match them against concatenative rules are computationaly 
+expensive. 
+Analyzers that exhaustively enumerate all possible solutions 
+for a word may hurt performance and may
+not be necessary or appropriate in some case studies~\cite{Maamouri:10}. 
+For example, the first solution may be the one needed and 
+we may not need to list the rest of the solutions. 
+A white space delimited token may have more than one word
+as is the case with ``run-on words''.
+
+We propose to make use of the current existing morphological
+analyzers to build a flexible and more efficient
+morphological analyzer. 
+The analyzer will be an essential
+component in the CATARM framework. 
+It will be a case-based analyzer as it will make guided decisions
+based on feedback from the NLP case study and it will prune
+invalid solutions on the fly.
+
+Case-based analysis can guide the morphological analysis
+to prune lots of the solutions and to bias the order of considered
+decisions. 
+For example, in the hadith literature case study pesented above, 
+amongst other things, we are intersted in checking whether a 
+word is a name.  
+We can bias the order of considered decisions to consider the 
+morphemes that can be names first. 
+If one of them matches, we can ignore the rest of the solutions. 
+In case a prefix is decided to be of a morphological category
+that does not connect with a name, then we can just ignore
+the analysis of the rest of the word. 
+This is more efficient than generating all possible partitionings
+of the word, checking for morphological correctness of each 
+partition, and then inspecting the part of speech tags of each
+correct solution. 
+We desire to build a morphological analyzer that allows a 
+case-based controller to intervene at every single decision
+without much overhead.
+
+\subsection{CATARM Query language}
+\label{s:design:query}
+
+We propose the CATARM query language with a user friendly
+visual representation similar to that in Figures~\ref{f:literature} and
+~\ref{f:security}. 
+The basic elements in the query language will be entities, relations
+and tags similar to those reported by a morphological analyzer. 
+The query language will also have connectives such as set operations, 
+relational operations, 
+scalar and Boolean arithmetic operations, 
+statistical computations, graph algorithms, and text algorithms
+form terms and predicates that specify the desired computation
+by the user. 
+We will specify the CATARM query language in a formal grammar.
+
+\subsection{Arabic linguistic computational model }
+\label{s:design:lcm}
+
+Several techniques~\cite{AEL07,Ham07,Abd07,MEl03} 
+preprocessed the Arabic text and then used known text 
+mining techniques that tend to Latin text and queries
+to solve classification, indexing, named entity 
+extraction and other problems.
+We think that these techniques do not make use 
+of Arabic language specific features.
+
+We think that we can discover such features by building
+a linguistic computational model of the Arabic language
+based on its formal grammar and formal morphology
+as defined in several Arabic linguistic 
+sources~\cite{Sha73,Abd00,Abd001}.
+We will build our morphological analyzer on top 
+of this linguistic model in addition to the current existing
+concatenative-based analysis techniques.
+
+Let $L_q$ be the query language grammar, 
+$L_A$ be the grammar that expresses the Arabic linguistic
+computational model, and let $q$ be a user query.
+We compute $\rho_q$, the grammar path that matches 
+the query $q$ in $L_q$. 
+We will explore methods that use $\rho_q$ to simplify 
+$L_A$ before we start the morphological analysis. 
+We think the possible simplifications that we can perform
+here are Language specific as they will reflect the 
+correspondence between the formal query language definition $L_q$, 
+the usage patterns of $L_q$  by Arabic users, 
+and $L_A$ the formal LCM for Arabic.
+
+\subsection{Methodology}
+
+\begin{figure}
+\center{
+\resizebox{.4\columnwidth}{!}
+{\input{figs/hadith.pdftex_t}
+\label{f:statemachine}
+}
+\caption{State-machine used to detect the chain of narrators}
+}
+\end{figure}
+
+{\bf Case-based morphological analyzer.}
+We experimented with several morphological analyzers.
+We think that using finite state machines for morphological analysis
+can be implemented in a more efficient way compared to other 
+approaches. 
+For example the concatenative rules used in Buckwalter~\cite{Tim04}
+can be encoded in three FSM with linear execution time using 
+the trie~\cite{datries} data structure.
+
+The use of FSMs also suits our case-based approach. 
+The controller can be a FSM machine on its own and its decision
+will be another input into the internal morphological FSM structures. 
+This also suits our choice of enriching the concatenative analysis
+with a formal linguistic model that can also be encoded into FSMs. 
+FSM can also be easily integrated with hidden Markov models (HMM)
+extracted from the sets of documents the MA will consider.
+
+The state machine in Firgure~\ref{f:statemachine} controls 
+the concatenation based state machines of the morphological
+analyzer.
+The transitions are excited
+by inputs from the morphological analyzer. 
+A NAME label represents a proper name, 
+an NRC label represents a narrator connector such as
+\RL{`an} "on behalf of", \RL{qaal} "said", 
+or one of their derivations. 
+The IBN label corresponds to the word \RL{ibn} 
+``the son of'' and represents a name connector.
+
+The symbol $\tau_{\mbox{NMC}}$ is a threshold
+that corresponds to the number of tolerated name connectors 
+that may occur between two names. 
+The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
+of name connectors collected since the controller
+started looping in the state NMC\_S. 
+The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
+that corresponds to a relaxed tolerance measure that
+the controller resorts to in case the words separating
+two names were longer than $\tau_{\mbox{NMC}}$ but 
+contained a name connector word such as IBN or NISBA.
+
+The controller has four states that correspond to 
+an abstract position in text relative to the next sanad. 
+State TEXT\_S is the initial state and denotes that
+the controller is outside the context of a sanad.
+The controller moves to the state NAME\_S whenever
+NAME is reported.
+State NRC\_S means that the controller thinks it is in the context
+of a narrator name after an NRC is reported.
+State NMC\_S
+indicates that the controller expects a name to appear within 
+a tolerance threshold expressed by 
+$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
+
+Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
+before it gives up on its expectations.
+
+We reach the NAME\_S state only when a
+valid NAME is detected and we leave when no more NAME's 
+are detected. 
+The NRC\_S state can only be reached if an NRC is detected.
+NMC\_S can only be reached from a NAME\_S state.
+
+The controller reports a valid chain of narrators when a 
+sequence of names
+connected by narrator connectors appears. 
+It marks the beginning of the hadith with the beginning of the 
+current sequence,
+and marks the end of the hadith with the beginning of the next 
+sequence. 
+It also marks the chain of narrators as the sequence itself. 
+The definition of input labels such as NAME and IBN depends on 
+the morphological analyzer. 
+We readily implemented the above approach and confirmed its 
+accuracy and efficiency compared to existing morphological
+analyzers.
+
+We will investigate enhancing our current case-based morphological
+analyzer.
+We will investigate methods to extend the internal structures 
+of our analyzer to be able to represent a formal linguistic
+computational model of the Arabic language.
+
+{\bf Linguistic computational model of Arabic.}
+We think that a linguistic computational model for the Arabic
+language will perform better in applications where
+relations between Arabic text documents are involved. 
+Such a model helps capture features unique and specific 
+to the Arabic language and make use of that in the analysis.
+
+ElixirFM~\cite{Otakar:07} builds on the lexicon
+of the Buckwalter analyzer and the annotations of the 
+Prague Arabic dependency databank to build a formal 
+computational model of the Arabic language.
+The models is a set of formulas and rules. 
+ElixirFM uses the computational model in a functional
+morphology framework that uses Haskell and functional
+programming. 
+The analyzer works by generating morphological trees of 
+decisions and then uses a disambiguation decision procedure 
+to annotate the trees with contextual elements.
+
+We will explore extracting our model from formal text book
+definitions of the Arabic language~\cite{Abd00,Abd001} and seek
+help from Arabic linguists to relax these formal rules to
+allow common mistakes. 
+We will encode the model in a grammar that we can synthesize
+into an FSM and incorporate that with our current FSM 
+morphological analyzer.
+
+{\bf User query language. }
+In Figure~\ref{f:statemachine} we showed an FSM 
+that controls the morphological analysis and generates a
+list of chains of narrators out of hadith text.
+We think that it is overwhelming to ask a user who is not 
+versed in FSM and in computational sciences in general 
+to design such a controller.
+
+The diagram in Figure~\ref{f:chain} depicts our vision of 
+an equivalent query.
+This query generates the list of chains by first passing the 
+hadith text into the morphological analyzer which 
+extracts names and their position in text. 
+We iterate over the extracted names and compute the distance.
+Whenever the distance is less than a threshold $\theta_1$
+we add the word to a temporary chain of names. 
+Otherwise, we execute a sequential block to 
+{\bf 1.} tabulate the chain into a temporary table containing 
+the list of chains. 
+{\bf 2.} Create a new chain whith the last element.
+Finally, we filter those chains whose cardinality 
+is less than some threshold  $\theta_2$.
+
+This query can be saved to a library of computational components
+and invoked directly in another query 
+as we saw in Figure~\ref{f:literature}.
+
+\begin{figure}
+\center{
+\resizebox{.8\columnwidth}{!}
+{\includegraphics{figs/detailed_submachine.pdf}
+\label{f:chain}
+}
+\caption{CATARM detailed diagram for the chain of narrators extractor}
+}
+\end{figure}
+
+The CATARM query language will be expressed in a grammar 
+where the atomic elements are user defined constants, or
+entities and tags extracted by the morphological analyzer. 
+The language will define terms, expressions, and predicates
+using several connectives that express set operations, 
+relational operations, Boolean operations, scalar operations, 
+statistical tools, text algorithms, and graph algorithms. 
+We need to provide a front end tool to support this query
+language.
+
+{\bf Simplify LCM based on the query.}
+Given
+L$L_q$ the query language, 
+$L_A$ the Arabic linguistic
+computational model, and  $q$ a user query
+we compute $\rho_q$, the grammar path that matches 
+the query $q$ in $L_q$. 
+We plan to inspect methods to overapproximate 
+the morphemes and the transitions in $L_A$ 
+that can be satisfied by morphemes and transitions
+in $\rho_q$. 
+This can simplify the number of states we need to 
+look at under $L_A$.
+
+This task is not trivial though, since $\rho_q$ may contain
+states and transitions that are not always morphemes and 
+transition between morphemes. 
+We will consider a hierarchical methodology that follows 
+the hierarchical structure of the Arabic language expressed
+in $L_A$. For example,
+we may consider sets of connected states in $\rho_q$ 
+containing a single morpheme as a state in one hierarchy level 
+in $L_A$.
+We will perform this simplification before starting the 
+morphological analysis since we already have to analyze
+the query $q$ and compute $\rho_q$ before we start
+the analysis.
+
+{\bf Simplify LCM and MA based on set of documents.}
+We will analyze the set of input documents given to a
+morphological analyzer component and use the analysis
+to further improve the morphological analysis. 
+We will extract statistical features from these documents 
+to help order the solutions considered by the analyzer
+to consider the most probable first.
+
+We will consider the use of HMM models 
+built on the fly while processing the documents.
+We will also consider the use of
+latent semantic indexing~\cite{LSI89} techniques. 
+We are considering these two techniques as they are 
+both language independent and can both be computed and
+adapted on the fly.
+
+{\bf Entity and relational extractor.}
+We will build a set of entity and relational extractors. 
+The entities can be proper names, dates, locations,
+and qualification adjectives. 
+We will explore the use of pairing techniques to establish 
+relations between the extracted entities.
+The relations will form graphs as in the case studies presented
+so far. The nodes represent the entities and the previously 
+extracted relations and the edges represent the newly discovered 
+relations.
+
+The entity and relational extraction components 
+may take as input a template for an expected relation and 
+may return a graph that describes whether and where 
+this relation exists between the subject documents.
+
+We will need to augment the tags in the lexicons of the 
+morphological analyzer to allow for better entity and relational
+extractors. 
+We plan to make use of existing annotated tree banks and of
+modern Arabic dictionaries to do that.
+
+{\bf Statistical components.}
+We will make use and build statistical components that can be used
+in CATARM. 
+We will need statistical components to rank solutions, compute 
+probable decisions, estimating good thresholds.
+
+{\bf Graph algorithms.}
+The relations extracted from the text documents will often 
+be expressed in terms of graphs.
+We will need textbook graph algorithms such as traversals, 
+topological sorts, spanning trees, strongly connected 
+components, minimum flow  and other algorithms.
+For example, in the literature case study, the discovery
+of articulation vertices may point to important narrators 
+who may have an insignificant number of narrations 
+but still connect narrator communities.
+
+{\bf Query solver.}
+This is a logical solver that will solve query expressions to \
+resolve matching text and relations.
+We will need to use existing off the shelf constraint solvers 
+and set theory satisfiability solvers to address our queries. 
+The solver should return and rank approximate matches.
+That is if a three entities out of four required in a relation are
+found, the solution is still worth reporting.
+
+The solver will call the morphological analyzer to compute 
+morphological matches to the atomic elements of the query. 
+Then each operation in the query such as a set intersection
+operation will correspond to a computational component. 
+The solver will call these components and integrate their
+results.
+
+The solver may make use of existing propositional~\cite{MiniSAT} or
+higher order logic~\cite{Z3} solvers if needed. 
+These solvers can solve satisfiability questions.
+We will explore the feasibility and utility 
+of extracting a necessary set of 
+constraints from the query and answering that with an off-the-shelf
+logic solver.
+
+{\bf Case studies.}
+We will work on developing the necessary components to complete 
+the solution for the hadith literature and the security case 
+studies.
+We will also work on the health and network traffic case studies.
+While working on the case studies, we will add the components
+we build to our library of components. 
+This will enrich CATARM with a useful set of computational 
+components.
+
+{\bf Development methodology.}
+We will integrate all the mentioned techniques in a scalable 
+framework for Arabic text analysis.
+The framework as a whole will allow for exhausting Arabic text 
+analysis and allow for extensions of the framework.
+A user should be able to iteratively and repetitively apply the 
+abode techniques on his input text.
+
+We will follow an eXtreme Programming 
+methodology in the process of building CATARM.
+We will use open source tools, libraries, and frameworks to 
+build our libraries.
+We will build on existing open source programming and text 
+mining tools as well as open source Arabic text analysis tools 
+based on NLP such as MADA+TOKAN~\cite{Rot08}, 
+NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
+
+We will use a dokuwiki based website~\cite{Dok09} 
+to document the progress online and will use 
+subversion~\cite{Sub09} as a source control tool.
+
+{\bf Visualization.} 
+We will build a visualization tool that allows fast 
+and guided inspection and correction mechanism. 
+The CATARM visualization aid will be a tag and relation color
+sensitive browser.
+The interface will allow a fast and guided inspection and 
+correction mechanism.
+
+\subsection{Research team}
+
+\subsection{Work plan}
+
+\subsection{Budget}
 
@@
 
-We anticipate that the research will lead to the production of an Arabic text analysis framework that allows relational queries between Arabic text documents.
- The components of the framework are listed in Section "Research Design and Methods" and are repeated for convenience as follows.
+We anticipate that the research will lead to the production of 
+CATARM, an Arabic text analysis framework that allows relational 
+queries between Arabic text documents.
+The components of the framework are listed and discussed
+in details in Section~\ref{s:design}.
+
+The following is a list of the anticipated components They are and are repeated for convenience as follows.
 \begin{enumerate}
- \item Lexical analysis library 
- \item Syntax analysis library
-\item Dynamic modern Arabic dictionary
-\item An entity extraction library
-\item A relation extraction library
-\item A logic solver 
-\item Tagger
+\item Case-based morphological analyzer 
+\item Linguistic computational model for Arabic
+\item User query language
+\item Optimization methodologies
+\begin{enumerate}
+\item Use of language specific features 
+\item Use of data in question specific features
 \end{enumerate}
-
-We will use these tools to handle several case studies from different fields.
- In the following we discuss three case studies that we anticipate our proposed framework to solve.
-
-{\bf The literature case study}
-Assume we have two sets of texts.
- The first set is several books of historical accounts where each account is preceded by a chain of narrators who narrated the account.
- The second set is several books of biographies where each biography of a person may include an evaluation of the authenticity of that person.
- We want a tool that can answer a complex query such as relating all historical accounts in the first set that discuss women and vehicles (such as tamed animals) to the narrators of the accounts, and then relate these narrators to the second set of books of biographies with an emphasis on authenticity evaluations.
-
-We need tools that can search and index the historical accounts to return a subset that relates to women and travel.
- Then we need tools that can identify names, and succession relations between names to build a chain of narrators for each historical account in the subset.
- We need tools that can identify the names in the biographies and detect evaluation adjectives.
-
-Provided we have these tools we can exhaustively at the click of a button visualize the authenticity of all the historical accounts that may allow or forbid women from driving according to Islamic laws.
- We can go one step further to detect locations and dates and relations based on chronological orders and further question the consistency of the historical accounts.
- Such query takes currently huge amounts of manual effort that is also prone for error and far from being exhaustive.
-
-We here envision that text mining applied to the authenticity of historical accounts will have a huge impact in the field.
-
-Note that most of the Islamic history and Hadith books as well as a huge library of Arabic literature books are already digitized and available on the internet.
-
-{\bf The security case study}
-
-Given a huge set of investigation reports, an officer may throw a list of names of convicted people and check all the reports for possible relations between them in terms of criminal action, locality, or third party names unknown to the officer who happen to know the convicts.
- The query may return at the click of a button, two convicted people who had no direct relation in the reports, but who are related only through a third person unknown to the investigator.
- Such a query is very hard to answer manually.
-
-{\bf The health case study}
-
-Given a set of clinical reports gathered by the Order of Medical Doctors, the ministry of health may query to relate all reports on a specific sickness to the doctors treating it and evaluate the response to the different types of medications as classified in different areas in the country each with a different climate.
- Such query, while possible at a click of a button with text mining tools if the data is available in digital form, would take a huge manual effort to conduct surveys, collect and gather the data.
- In this case, digitizing the data may be a less expensive effort.
+\item Entity and relational extractors
+\item Statistical tools
+\item Graph and text algorithms
+\item Query solver
+\item Case studies 
+\begin{enumerate}
+\item Hadith literature
+\item Security
+\item Health
+\item Network content
+\end{enumerate}
+\end{enumerate}
 

### 2010-12-09 trunk/tex/rp.tex
anticipated results

--- 2010-12-05 trunk/tex/rp.tex
@@
 that merged equal names in the chains. 
-This partial order graph is instrumental to automatically
-locate narrators in biography documents and automatically
-segment biography documents.
+This partial order graph is instrumental to automate
+localizing narrators in biography documents and
+segmenting biography documents.
+
+Given that a biography will discuss a narrator and mention
+his professors and students,
+we can segment the biographies with a graph coloring algorithm 
+that traverses the text and colors the POR whenever
+a name is found. 
+We modify the color when we move in the graph 
+far away from the last name we colored.
+
+Once the biographies are analyzed, one can annotate
+the POR with qualifiers of the narrators that reflect
+their authenticity. 
+One can also annotate the POR with the locations and 
+the time they lived in. 
+Then One can compute a time and location overlap
+check to discover inconsistent narrations.
+We can also perform several interesting checks using 
+this POR on its own such as checking for the effect of
+one narrator on the hadith literature.
 
@@
 
-The difficulty of applying text mining techniques to Arabic text 
-documents lays in the absence of automated tools that understand 
-the unique features of the Arabic language.
-We will make use of an example to illustrate the process.
+%The difficulty of applying text mining techniques to Arabic text 
+%documents lays in the absence of automated tools that understand 
+%the unique features of the Arabic language.
+%We will make use of an example to illustrate the process.
 Given an Arabic text document we want a tool that can segment 
@@
 \noTrNoVocRL{alm^staq-at w al^g_dwr}.
- We also need to build dictionary tools that associate words and phrases based on their meanings and their context.
+Simple primitive tasks such as comparing two strings need 
+special attention in Arabic since each Arabic letter has 
+four different forms depending on its position in the word; 
+isolated form, start, middle and end of a word.
+Arabic diacritics should also be ignored most of the time 
+when comparing two strings.
+We also need to build dictionary tools that associate words 
+and phrases based on their meanings and their context.
 Often time,
 and depending on the application and the user of the tool,
- we need to allow the user to assign meanings and semantics for patterns in the text.
-
-In the following we discuss our research methodology to develop the proposed Arabic text analysis tools.
-
-Lexical analysis library: We need a library of routines that enables primitive lexical analysis of Arabic text.
- For example,
- simple primitive tasks such as comparing two strings need special attention in Arabic since each Arabic letter has four different forms depending on its position in the word; isolated form,
- start,
- middle and end of a word.
- Arabic diacritics should also be ignored most of the time when comparing two strings.
- These lexical analysis routines should also pay attention to the form of a letter when used in tokenization as a letter in and end of word form denotes the end of the word even if it was not followed with a space.
- Such a library can benefit from existing Unicode enabled string libraries such as Qt.
- It also can provide a layer of abstraction that hides the details of the encoding of Arabic letters used in the text documents.
-
-Syntax analysis library: This library should allow sentence structure understanding.
- We will explore the best way to develop this library.
- Several options exist such as pattern based matching,
- tokenization and dictionary based lookups,
- or a language grammar that allows more than one matching rule.
- This library needs to be extendible to allow additions of new rules or patterns.
-
- (Note the ElixirFM does this)
- We will explore Arabic grammar books such as The Principles of Arabic~\cite{Sha73} and others~\cite{Abd00,Abd001}
- to augment current morphological analyzers~\cite{Tim04} and syntax analysis tools~\cite{Col09}.
- We will use relaxed NLP techniques to overapproximate results.
- Then and after a relational query,
- we will apply statistical analysis techniques to rank the matching results.
-
-Dynamic modern Arabic dictionary: The dictionary will allow identities based on semantics,
- derivation rules,
- as well as user assigned rules.
- The dictionary will return modern grammatical categories for words classifying them as nouns,
- verbs,
- and adjectives instead of the classical Arabic categories that specify how the word is derived from an original radix.
- It remains to be explored how to represent these categories without exploding the size of the dictionaries with redundant words of multiple categories.
- This work will augment existing open source Arabic dictionaries and will allow for better morphological analysis and entity extraction.
-
-An entity extraction library: We will explore building a library that allows extraction of entities from Arabic text.
- These entities can be proper names,
- dates,
- locations,
- and qualification adjectives to list few possibilities.
- The extractor will use the dynamic modern Arabic dictionary as well as the syntax library to extract entities from a sentence and to establish identity between entities across sentences.
-
-A relation extraction library: We will explore building a library that allows extraction of relations from Arabic text.
- We will explore the use of pairing techniques to establish relations between the extracted entities.
- The relations will form graphs where the nodes are the entities and the previously extracted relations and the edges are the newly discovered relations.
- The routines in this library will take as input a template for an expected relation and will return a graph that describes whether this relation exists between the subject documents and where.
- These relations will be ranked using statistical analysis.
- The statistical criterion and the ranking methods remain to be decided.
-
-A logic solver: This is a logical solver that will solve query expressions to resolve matching text and relations.
- We will explore the proposal of a logical query language that is restricted enough to be computable and expressive enough to allow interesting properties.
- We will use of the shelf solvers and matches from the entity and relation extractors to evaluate the expression.
- The choice of solvers remains to be decided after the query language is formed.
-
-Tagger: We will explore the development of an automatic tagging interface that allows fast and guided inspection and correction mechanism.
-
-We will integrate all the mentioned techniques in a scalable framework for Arabic text analysis.
- The framework as a whole will allow for exhausting Arabic text analysis and allow for extensions of the framework.
- A user should be able to iteratively and repetitively apply the abode techniques on his input text.
-
-In the development cycle, we will use an Extreme programming methodology in the process of building the Arabic text mining framework.
- We will use open source tools, libraries, and frameworks to build our libraries.
-
-Our research will concentrate on NLP techniques native to the Arabic language to analyze the Arabic text and we will avoid preprocessing that feeds common text and data mining techniques.
- We will build on existing open source programming and text mining tools as well as open source Arabic text analysis tools based on NLP such as MADA+TOKAN~\cite{Rot08}, NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
-
-We will use a dokuwiki based website~\cite{Dok09} to document the progress online and will use subversion~\cite{Sub09} as a source control tool.
+we need to allow the user to assign meanings and semantics 
+for patterns in the text.
+
+\subsection{Literature authentication case study}
+\label{s:design:lit}
+
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.75\columnwidth}{!}
+{\includegraphics{figs/literature_case-study1.pdf}
+\label{f:literature}
+} 
+\caption{CATARM diagram for the literature case study }
+}
+\end{figure}
+
+The diagram in Figure~\ref{f:literature} describes our 
+high level vision for what the literature case study
+will look like in the CATARM framework. 
+The boxes represent 
+computational blocks provided by CATARM, 
+ovals represent data blocks which are mainly the 
+inputs and the outputs of the computational blocks. 
+Dotted data blocks represent intermediate results,
+and solid and shaded blocks represent the initial input 
+and the final output respectively.
+
+The computational components can be direct CATARM components
+or custom user components built from other CATARM components.
+We feed the hadith books to a user Narrator Chain Extractor
+component that we will discuss later in 
+Section~\ref{s:design:chainext}. We  obtain a set of chains
+of narrators that we feed to a morphological analyzer (MA) 
+component controlled by a merger controller. 
+The merger controller will look at the morphological analysis
+of the narrator names and will decide to merge two narrators in 
+one graph node in case they were morphologicaly equal. 
+It will also give feedback to the MA component to stop the
+analysis when a it reaches a satisfactory result.
+We consider this graph as a partial order relation between 
+narrators where a narrator precedes another narrator if 
+he narrated from him.
+
+The biography books are fed to another MA unit controlled by 
+a name and qualification pair controller that generates a 
+map consisting of pairs of narrator names associated with 
+qualifications. 
+Then the graph of narrators and the narrator qualification map
+are both processed through a locator component that computes
+a graph coloring routine to color the graph of narrators. 
+The color changes whener two consecutive narrators are far away
+from each other in the text. 
+The locator then looks at connected subgraphs of the same color,
+and finds a median node in each of them to be the narrator 
+associated with the qualifiers of the color. 
+Then the locator also annotates the narrator with the 
+corresponding qualifications. 
+The result is an annotated partial order graph of narrators
+with authentication qualification annotations.
+
+\subsection{Security case study}
+\label{s:design:sec}
+
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.8\columnwidth}{!}
+{\includegraphics{figs/security_case-study.pdf}
+\label{f:security}
+}
+\caption{CATARM diagram for the security case study}
+}
+\end{figure}
+
+The diagram in Figure~\ref{f:security} describes our 
+high level vision for what the solution of the 
+security case study
+will look like in the CATARM framework. 
+The inputs are two separate sets of documents. 
+The first set includes crime scene investigation reports, 
+and the second set includes a 
+interrogation reports with crime suspects as well as 
+a database of previous interrogation reports. 
+The CATARM system also takes as input a database of geographical
+locations and a database of personal records. 
+The aim is to check for inconsistencies in the iterrogation
+reports and to identify possible connections between the suspects
+and the crime scene.
+
+An MA with date, place, and name extraction controller extracts
+names, dates and places from the investigation reports. 
+We pass the dates to a sorting component to produce $G_D$ 
+a partial order graph between the dates. 
+We look up the names in the personal records database and find out
+$G_N$ a family dependency graph between the names. 
+We pass the places to a mapper that builds another $G_G$,
+a graph relating the places using a geographical database.
+
+We pass the interrogation reports to an MA with a relation 
+extraction controller that builds $G_R$. 
+The graph $G_R$ relates the interrogated suspects
+to each other, to places, and to dates based on a simple
+neighborhood threshold computed using a statistical feature.
+
+We feed $G_D, G_F, G_G,$ and $G_R$   
+to a consistency checker.
+The consistency checker considers the three dimensions 
+$N$ names, $G$ locations, and $D$ dates and projects $G_R$
+over each one of them.
+For each dimension $k$, it computes a local measure 
+$\alpha_k$ of how much $G_R|_k$ simulates $G_k$. 
+It flags the components in the graph where the accumelated 
+measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
+below a threshold $\theta$ as inconsistent.
+
+\subsection{CATRAM components}
+\label{s:design:components}
+
+To do the above we need CATARM to provide us with several 
+computational components such as a morphological analyzer, 
+statistical tools, graph algorithms, and other computational
+components. 
+In the following we discuss the components needed for 
+CATARM and we propose a methodology to 
+use and extend existing tools, and develop new tools to
+build the components of CATARM.
+
+\subsection{Case-based morphological analyzer}
+
+Morphological analyzers for the Arabic language
+such as Buckwalter~\cite{Buckwalter:02},
+Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+and ElixirFM~\cite{Otakar:07} take an Arabic word as input
+and produce several valid morpholgical solution for the word. 
+Each solution is a sequence of morphemes that form the word.
+Each morpheme is associated with several tags such as 
+morphological, part of speech, grammar, and semantic tags. 
+These analyzers are essential in several text analysis tools such 
+as spell checkers as well as NLP frameworks~\cite{Col09}.
+Other morphological analyzers such as 
+Amira~\cite{Diab:07,Benajiba:07},
+MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
+use machine learning and support vector machines (SVM) 
+to enhance the accuracy at the expense of efficiency.
+
+The current use of morphological analyzers suffers from 
+several problems. 
+Analyzers that enumerate all possible partitions of a word
+and match them against concatenative rules are computationaly 
+expensive. 
+Analyzers that exhaustively enumerate all possible solutions 
+for a word may hurt performance and may
+not be necessary or appropriate in some case studies~\cite{Maamouri:10}. 
+For example, the first solution may be the one needed and 
+we may not need to list the rest of the solutions. 
+A white space delimited token may have more than one word
+as is the case with ``run-on words''.
+
+We propose to make use of the current existing morphological
+analyzers to build a flexible and more efficient
+morphological analyzer. 
+The analyzer will be an essential
+component in the CATARM framework. 
+It will be a case-based analyzer as it will make guided decisions
+based on feedback from the NLP case study and it will prune
+invalid solutions on the fly.
+
+Case-based analysis can guide the morphological analysis
+to prune lots of the solutions and to bias the order of considered
+decisions. 
+For example, in the hadith literature case study pesented above, 
+amongst other things, we are intersted in checking whether a 
+word is a name.  
+We can bias the order of considered decisions to consider the 
+morphemes that can be names first. 
+If one of them matches, we can ignore the rest of the solutions. 
+In case a prefix is decided to be of a morphological category
+that does not connect with a name, then we can just ignore
+the analysis of the rest of the word. 
+This is more efficient than generating all possible partitionings
+of the word, checking for morphological correctness of each 
+partition, and then inspecting the part of speech tags of each
+correct solution. 
+We desire to build a morphological analyzer that allows a 
+case-based controller to intervene at every single decision
+without much overhead.
+
+\subsection{CATARM Query language}
+\label{s:design:query}
+
+We propose the CATARM query language with a user friendly
+visual representation similar to that in Figures~\ref{f:literature} and
+~\ref{f:security}. 
+The basic elements in the query language will be entities, relations
+and tags similar to those reported by a morphological analyzer. 
+The query language will also have connectives such as set operations, 
+relational operations, 
+scalar and Boolean arithmetic operations, 
+statistical computations, graph algorithms, and text algorithms
+form terms and predicates that specify the desired computation
+by the user. 
+We will specify the CATARM query language in a formal grammar.
+
+\subsection{Arabic linguistic computational model }
+\label{s:design:lcm}
+
+Several techniques~\cite{AEL07,Ham07,Abd07,MEl03} 
+preprocessed the Arabic text and then used known text 
+mining techniques that tend to Latin text and queries
+to solve classification, indexing, named entity 
+extraction and other problems.
+We think that these techniques do not make use 
+of Arabic language specific features.
+
+We think that we can discover such features by building
+a linguistic computational model of the Arabic language
+based on its formal grammar and formal morphology
+as defined in several Arabic linguistic 
+sources~\cite{Sha73,Abd00,Abd001}.
+We will build our morphological analyzer on top 
+of this linguistic model in addition to the current existing
+concatenative-based analysis techniques.
+
+Let $L_q$ be the query language grammar, 
+$L_A$ be the grammar that expresses the Arabic linguistic
+computational model, and let $q$ be a user query.
+We compute $\rho_q$, the grammar path that matches 
+the query $q$ in $L_q$. 
+We will explore methods that use $\rho_q$ to simplify 
+$L_A$ before we start the morphological analysis. 
+We think the possible simplifications that we can perform
+here are Language specific as they will reflect the 
+correspondence between the formal query language definition $L_q$, 
+the usage patterns of $L_q$  by Arabic users, 
+and $L_A$ the formal LCM for Arabic.
+
+\subsection{Methodology}
+
+\begin{figure}
+\center{
+\resizebox{.4\columnwidth}{!}
+{\input{figs/hadith.pdftex_t}
+\label{f:statemachine}
+}
+\caption{State-machine used to detect the chain of narrators}
+}
+\end{figure}
+
+{\bf Case-based morphological analyzer.}
+We experimented with several morphological analyzers.
+We think that using finite state machines for morphological analysis
+can be implemented in a more efficient way compared to other 
+approaches. 
+For example the concatenative rules used in Buckwalter~\cite{Tim04}
+can be encoded in three FSM with linear execution time using 
+the trie~\cite{datries} data structure.
+
+The use of FSMs also suits our case-based approach. 
+The controller can be a FSM machine on its own and its decision
+will be another input into the internal morphological FSM structures. 
+This also suits our choice of enriching the concatenative analysis
+with a formal linguistic model that can also be encoded into FSMs. 
+FSM can also be easily integrated with hidden Markov models (HMM)
+extracted from the sets of documents the MA will consider.
+
+The state machine in Firgure~\ref{f:statemachine} controls 
+the concatenation based state machines of the morphological
+analyzer.
+The transitions are excited
+by inputs from the morphological analyzer. 
+A NAME label represents a proper name, 
+an NRC label represents a narrator connector such as
+\RL{`an} "on behalf of", \RL{qaal} "said", 
+or one of their derivations. 
+The IBN label corresponds to the word \RL{ibn} 
+``the son of'' and represents a name connector.
+
+The symbol $\tau_{\mbox{NMC}}$ is a threshold
+that corresponds to the number of tolerated name connectors 
+that may occur between two names. 
+The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
+of name connectors collected since the controller
+started looping in the state NMC\_S. 
+The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
+that corresponds to a relaxed tolerance measure that
+the controller resorts to in case the words separating
+two names were longer than $\tau_{\mbox{NMC}}$ but 
+contained a name connector word such as IBN or NISBA.
+
+The controller has four states that correspond to 
+an abstract position in text relative to the next sanad. 
+State TEXT\_S is the initial state and denotes that
+the controller is outside the context of a sanad.
+The controller moves to the state NAME\_S whenever
+NAME is reported.
+State NRC\_S means that the controller thinks it is in the context
+of a narrator name after an NRC is reported.
+State NMC\_S
+indicates that the controller expects a name to appear within 
+a tolerance threshold expressed by 
+$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
+
+Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
+before it gives up on its expectations.
+
+We reach the NAME\_S state only when a
+valid NAME is detected and we leave when no more NAME's 
+are detected. 
+The NRC\_S state can only be reached if an NRC is detected.
+NMC\_S can only be reached from a NAME\_S state.
+
+The controller reports a valid chain of narrators when a 
+sequence of names
+connected by narrator connectors appears. 
+It marks the beginning of the hadith with the beginning of the 
+current sequence,
+and marks the end of the hadith with the beginning of the next 
+sequence. 
+It also marks the chain of narrators as the sequence itself. 
+The definition of input labels such as NAME and IBN depends on 
+the morphological analyzer. 
+We readily implemented the above approach and confirmed its 
+accuracy and efficiency compared to existing morphological
+analyzers.
+
+We will investigate enhancing our current case-based morphological
+analyzer.
+We will investigate methods to extend the internal structures 
+of our analyzer to be able to represent a formal linguistic
+computational model of the Arabic language.
+
+{\bf Linguistic computational model of Arabic.}
+We think that a linguistic computational model for the Arabic
+language will perform better in applications where
+relations between Arabic text documents are involved. 
+Such a model helps capture features unique and specific 
+to the Arabic language and make use of that in the analysis.
+
+ElixirFM~\cite{Otakar:07} builds on the lexicon
+of the Buckwalter analyzer and the annotations of the 
+Prague Arabic dependency databank to build a formal 
+computational model of the Arabic language.
+The models is a set of formulas and rules. 
+ElixirFM uses the computational model in a functional
+morphology framework that uses Haskell and functional
+programming. 
+The analyzer works by generating morphological trees of 
+decisions and then uses a disambiguation decision procedure 
+to annotate the trees with contextual elements.
+
+We will explore extracting our model from formal text book
+definitions of the Arabic language~\cite{Abd00,Abd001} and seek
+help from Arabic linguists to relax these formal rules to
+allow common mistakes. 
+We will encode the model in a grammar that we can synthesize
+into an FSM and incorporate that with our current FSM 
+morphological analyzer.
+
+{\bf User query language. }
+In Figure~\ref{f:statemachine} we showed an FSM 
+that controls the morphological analysis and generates a
+list of chains of narrators out of hadith text.
+We think that it is overwhelming to ask a user who is not 
+versed in FSM and in computational sciences in general 
+to design such a controller.
+
+The diagram in Figure~\ref{f:chain} depicts our vision of 
+an equivalent query.
+This query generates the list of chains by first passing the 
+hadith text into the morphological analyzer which 
+extracts names and their position in text. 
+We iterate over the extracted names and compute the distance.
+Whenever the distance is less than a threshold $\theta_1$
+we add the word to a temporary chain of names. 
+Otherwise, we execute a sequential block to 
+{\bf 1.} tabulate the chain into a temporary table containing 
+the list of chains. 
+{\bf 2.} Create a new chain whith the last element.
+Finally, we filter those chains whose cardinality 
+is less than some threshold  $\theta_2$.
+
+This query can be saved to a library of computational components
+and invoked directly in another query 
+as we saw in Figure~\ref{f:literature}.
+
+\begin{figure}
+\center{
+\resizebox{.8\columnwidth}{!}
+{\includegraphics{figs/detailed_submachine.pdf}
+\label{f:chain}
+}
+\caption{CATARM detailed diagram for the chain of narrators extractor}
+}
+\end{figure}
+
+The CATARM query language will be expressed in a grammar 
+where the atomic elements are user defined constants, or
+entities and tags extracted by the morphological analyzer. 
+The language will define terms, expressions, and predicates
+using several connectives that express set operations, 
+relational operations, Boolean operations, scalar operations, 
+statistical tools, text algorithms, and graph algorithms. 
+We need to provide a front end tool to support this query
+language.
+
+{\bf Simplify LCM based on the query.}
+Given
+L$L_q$ the query language, 
+$L_A$ the Arabic linguistic
+computational model, and  $q$ a user query
+we compute $\rho_q$, the grammar path that matches 
+the query $q$ in $L_q$. 
+We plan to inspect methods to overapproximate 
+the morphemes and the transitions in $L_A$ 
+that can be satisfied by morphemes and transitions
+in $\rho_q$. 
+This can simplify the number of states we need to 
+look at under $L_A$.
+
+This task is not trivial though, since $\rho_q$ may contain
+states and transitions that are not always morphemes and 
+transition between morphemes. 
+We will consider a hierarchical methodology that follows 
+the hierarchical structure of the Arabic language expressed
+in $L_A$. For example,
+we may consider sets of connected states in $\rho_q$ 
+containing a single morpheme as a state in one hierarchy level 
+in $L_A$.
+We will perform this simplification before starting the 
+morphological analysis since we already have to analyze
+the query $q$ and compute $\rho_q$ before we start
+the analysis.
+
+{\bf Simplify LCM and MA based on set of documents.}
+We will analyze the set of input documents given to a
+morphological analyzer component and use the analysis
+to further improve the morphological analysis. 
+We will extract statistical features from these documents 
+to help order the solutions considered by the analyzer
+to consider the most probable first.
+
+We will consider the use of HMM models 
+built on the fly while processing the documents.
+We will also consider the use of
+latent semantic indexing~\cite{LSI89} techniques. 
+We are considering these two techniques as they are 
+both language independent and can both be computed and
+adapted on the fly.
+
+{\bf Entity and relational extractor.}
+We will build a set of entity and relational extractors. 
+The entities can be proper names, dates, locations,
+and qualification adjectives. 
+We will explore the use of pairing techniques to establish 
+relations between the extracted entities.
+The relations will form graphs as in the case studies presented
+so far. The nodes represent the entities and the previously 
+extracted relations and the edges represent the newly discovered 
+relations.
+
+The entity and relational extraction components 
+may take as input a template for an expected relation and 
+may return a graph that describes whether and where 
+this relation exists between the subject documents.
+
+We will need to augment the tags in the lexicons of the 
+morphological analyzer to allow for better entity and relational
+extractors. 
+We plan to make use of existing annotated tree banks and of
+modern Arabic dictionaries to do that.
+
+{\bf Statistical components.}
+We will make use and build statistical components that can be used
+in CATARM. 
+We will need statistical components to rank solutions, compute 
+probable decisions, estimating good thresholds.
+
+{\bf Graph algorithms.}
+The relations extracted from the text documents will often 
+be expressed in terms of graphs.
+We will need textbook graph algorithms such as traversals, 
+topological sorts, spanning trees, strongly connected 
+components, minimum flow  and other algorithms.
+For example, in the literature case study, the discovery
+of articulation vertices may point to important narrators 
+who may have an insignificant number of narrations 
+but still connect narrator communities.
+
+{\bf Query solver.}
+This is a logical solver that will solve query expressions to \
+resolve matching text and relations.
+We will need to use existing off the shelf constraint solvers 
+and set theory satisfiability solvers to address our queries. 
+The solver should return and rank approximate matches.
+That is if a three entities out of four required in a relation are
+found, the solution is still worth reporting.
+
+The solver will call the morphological analyzer to compute 
+morphological matches to the atomic elements of the query. 
+Then each operation in the query such as a set intersection
+operation will correspond to a computational component. 
+The solver will call these components and integrate their
+results.
+
+The solver may make use of existing propositional~\cite{MiniSAT} or
+higher order logic~\cite{Z3} solvers if needed. 
+These solvers can solve satisfiability questions.
+We will explore the feasibility and utility 
+of extracting a necessary set of 
+constraints from the query and answering that with an off-the-shelf
+logic solver.
+
+{\bf Case studies.}
+We will work on developing the necessary components to complete 
+the solution for the hadith literature and the security case 
+studies.
+We will also work on the health and network traffic case studies.
+While working on the case studies, we will add the components
+we build to our library of components. 
+This will enrich CATARM with a useful set of computational 
+components.
+
+{\bf Development methodology.}
+We will integrate all the mentioned techniques in a scalable 
+framework for Arabic text analysis.
+The framework as a whole will allow for exhausting Arabic text 
+analysis and allow for extensions of the framework.
+A user should be able to iteratively and repetitively apply the 
+abode techniques on his input text.
+
+We will follow an eXtreme Programming 
+methodology in the process of building CATARM.
+We will use open source tools, libraries, and frameworks to 
+build our libraries.
+We will build on existing open source programming and text 
+mining tools as well as open source Arabic text analysis tools 
+based on NLP such as MADA+TOKAN~\cite{Rot08}, 
+NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
+
+We will use a dokuwiki based website~\cite{Dok09} 
+to document the progress online and will use 
+subversion~\cite{Sub09} as a source control tool.
+
+{\bf Visualization.} 
+We will build a visualization tool that allows fast 
+and guided inspection and correction mechanism. 
+The CATARM visualization aid will be a tag and relation color
+sensitive browser.
+The interface will allow a fast and guided inspection and 
+correction mechanism.
+
+\subsection{Research team}
+
+The team will be formed by the two main PIs, Dr. Fadi Zaraket (LPI)
+and Dr. Rehab Duwairi (Co-LPI) and three research assistants.
+Currently the LPI is working with one research assistant 
+and two senior students on an Arabic morphological analyzer.
+
+The LPI offers a course on computational Arabic attended
+by excellent senior and graduate students at AUB. 
+Once funded, the team can be directly formed and work can start
+with productive team members.
+
+\subsection{Work plan}
+
+\subsection{Budget items}
+
+In the following we list the budget items. 
+Monatary values are provided in more details on the QNRF submission 
+website.
+
+\begin{itemize}
+\item One research assistant at QU
+\item Two research assistants at AUB, one of them will 
+work partially from QU. 
+\item Two month salary per year for the Co-LPI
+\item Two month salary per year for the LPI
+\item One course buy out per year for the Co-LPI
+\item Two weeks consultant for the second and the third years
+\item Licensing and membership fees to acquire digital libraries and corpora
+\item High performance computer and storage devices
+\item One conference for each PI per year
+\item One conference for one RA from QU per year
+\item One conference for one RA from AUB per year
+\item Site visit to AUB 
+\item Site visit to QU 
+\item Other miscellanous items
+\end{itemize}
 
@@
 
-We anticipate that the research will lead to the production of an Arabic text analysis framework that allows relational queries between Arabic text documents.
- The components of the framework are listed in Section "Research Design and Methods" and are repeated for convenience as follows.
+We anticipate that the research will lead to the production of 
+CATARM, an Arabic text analysis framework that allows relational 
+queries between Arabic text documents.
+The components of the framework are listed and discussed
+in details in Section~\ref{s:design}. 
+In the following we will discuss what to anticipate and 
+how to evaluate the different components of CATARM.
+
+\subsection{Case-based morphological analyzer}
+
+We will deliver a case-based morphological analyzer. 
+The analyzer will be compared to other available analyzers such as
+Buckwalter~\cite{Tim04} and ElixirFM~\cite{Otakar:07} in terms
+of accuracy and efficiency of analysis.
+We will adopt two comparison methodologies. 
+First, we will compare the analyzers in an exhaustive analysis mode
+which is the only way current analyzers perform. 
+Second, we will compare the analyzers under several case-studies
+and show how a case-based morphological analyzer can benefit
+from case-based optimizations.
+
+\subsection{Linguistic computational model for Arabic}
+
+We will build a linguistic computational model for Arabic 
+based on an understanding of the formal Arabic grammar 
+and on an abstraction of the current existing concatenation
+based lexicons.
+We will compare our LCM with that of ElixirGM~\cite{Otakar:07}.
+We will study how much such a model can affect the analysis
+in terms of accuracy and efficiency as well.
+
+\subsection{User query language}
+
+We will build a query language for relational Arabic text mining 
+and a front-end to compile a user query into an optimized
+computation. 
+We will check how user friendly and how expressive the language is.
+We will compare our query language with the sophisticated 
+interface of the MADA+TOKAN toolkit~\cite{Habash:09} and we will
+look for other attempts in the industry .
+
+\item Optimization methodologies
 \begin{enumerate}
- \item Lexical analysis library 
- \item Syntax analysis library
-\item Dynamic modern Arabic dictionary
-\item An entity extraction library
-\item A relation extraction library
-\item A logic solver 
-\item Tagger
+\item Use of language specific features 
+\item Use of data in question specific features
 \end{enumerate}
-
-We will use these tools to handle several case studies from different fields.
- In the following we discuss three case studies that we anticipate our proposed framework to solve.
-
-{\bf The literature case study}
-Assume we have two sets of texts.
- The first set is several books of historical accounts where each account is preceded by a chain of narrators who narrated the account.
- The second set is several books of biographies where each biography of a person may include an evaluation of the authenticity of that person.
- We want a tool that can answer a complex query such as relating all historical accounts in the first set that discuss women and vehicles (such as tamed animals) to the narrators of the accounts, and then relate these narrators to the second set of books of biographies with an emphasis on authenticity evaluations.
-
-We need tools that can search and index the historical accounts to return a subset that relates to women and travel.
- Then we need tools that can identify names, and succession relations between names to build a chain of narrators for each historical account in the subset.
- We need tools that can identify the names in the biographies and detect evaluation adjectives.
-
-Provided we have these tools we can exhaustively at the click of a button visualize the authenticity of all the historical accounts that may allow or forbid women from driving according to Islamic laws.
- We can go one step further to detect locations and dates and relations based on chronological orders and further question the consistency of the historical accounts.
- Such query takes currently huge amounts of manual effort that is also prone for error and far from being exhaustive.
-
-We here envision that text mining applied to the authenticity of historical accounts will have a huge impact in the field.
-
-Note that most of the Islamic history and Hadith books as well as a huge library of Arabic literature books are already digitized and available on the internet.
-
-{\bf The security case study}
-
-Given a huge set of investigation reports, an officer may throw a list of names of convicted people and check all the reports for possible relations between them in terms of criminal action, locality, or third party names unknown to the officer who happen to know the convicts.
- The query may return at the click of a button, two convicted people who had no direct relation in the reports, but who are related only through a third person unknown to the investigator.
- Such a query is very hard to answer manually.
-
-{\bf The health case study}
-
-Given a set of clinical reports gathered by the Order of Medical Doctors, the ministry of health may query to relate all reports on a specific sickness to the doctors treating it and evaluate the response to the different types of medications as classified in different areas in the country each with a different climate.
- Such query, while possible at a click of a button with text mining tools if the data is available in digital form, would take a huge manual effort to conduct surveys, collect and gather the data.
- In this case, digitizing the data may be a less expensive effort.
+\item Entity and relational extractors
+\item Statistical tools
+\item Graph and text algorithms
+\item Query solver
+\item Case studies 
+\begin{enumerate}
+\item Hadith literature
+\item Security
+\item Health
+\item Network content
+\end{enumerate}
+\end{enumerate}
 

### 2010-12-09 trunk/tex/rp.tex
evaluation

@@
 
+The Lead PI will conduct biweekly meetings to track progress 
+on the project.
+The team will maintain a dokuwiki based website to continuously 
+post updates on the progress of the project.
+
 \subsection{Budget items}
@@
 In the following we list the budget items. 
-Monatary values are provided in more details on the QNRF submission 
-website.
+Monatary values are provided in more details on the QNRF 
+submission website.
 
@@
 We will compare our query language with the sophisticated 
-interface of the MADA+TOKAN toolkit~\cite{Habash:09} and we will
-look for other attempts in the industry .
-
-\item Optimization methodologies
-\begin{enumerate}
-\item Use of language specific features 
-\item Use of data in question specific features
-\end{enumerate}
-\item Entity and relational extractors
-\item Statistical tools
-\item Graph and text algorithms
-\item Query solver
-\item Case studies 
+interface of the MADA+TOKAN toolkit~\cite{Habash:09}.
+We will look for other attempts to formalize Arabic text 
+mining queries and we will also compare agains 
+formal text mining queries from other text mining domains such
+as the gene ontolgy domain~\cite{GeneOntology10},
+
+\subsection{Optimizations}
+
+We will develop and evaluate our two optimization methodologies
+and look for other optimizations. 
+First we will use language specific features based on the 
+LCM of Arabic, the grammar of the query language, and 
+the Arabic query provided by the user to simplify 
+the LCM and the morphological analysis. 
+Second we will look at language independent techniques
+such as HMM and LSI models extracted from the 
+text documents and use that to simplify the analysis. 
+We will perform the analysis with and without the optimizations
+and compare the results in terms of accuracy and efficiency.
+
+%\subsection{ Entity and relational extractors}
+\subsection{CATARM and case studies }
+
+We will evaluate CATARM against at least three of the case studies
+that we discussed previously and that are listed below. 
 \begin{enumerate}
@@
 \end{enumerate}
-\end{enumerate}
-
-\subsection{Evaluation Criteria}
-
-The Lead PI will conduct biweekly meetings to track progress on the project.
- The team will maintain a dokuwiki based website to continuously post updates on the progress of the project.
- In Section "Research Design and Methods" we presented a schedule for delivering the different products of the project.
- The schedule can be used as a timeline for measuring the progress of the project.
-
-The lexical and syntax analysis libraries will be compared with current open source libraries against commonly used corpora of data such as a collection of books or a collection of news reports.
- The dictionary module will be an augmentation over existing digitized dictionaries such as the Buckwalter~\cite{Tim04} morphological analyzer.
- In the entity and relational extraction tools we will use techniques native to the Arabic language and we will compare against English entity and relational extraction tools of similar or translated texts.
- We will also compare techniques that pre-process Arabic and then pass it to English IE and RE techniques.
- The logic solver will build over the IE and RE tools and will use off-the shelf quantifier and propositional solvers such as MiniSat~\cite{Een03}.
- The tagger is a GUI interface that allows fast manual inspection and refinement of the automated results.
- Each tool in the framework can start on its own with a prototypical implementation of the other needed tools.
-
-Each tool in the framework stands alone and can operate on a well defined input and generate the desired output.
- The input of one tool, such as the entity extraction, can be automatically generated by another tool, such as the dictionary and the syntax analyzer.
-  However, the input can also be manually generated if desired using the straightforward GUI tagger facility.
- Thus the tagger facility stands as a desired utility on its own and at the same time represents a contingency plan in case any module in the framework failed to deliver.
-
-The team will publish the results of the research in the form of refereed conference and journal papers as well as tool papers addressing each of the tools.
- The team will also hold workshops to educate possible users on the tools once the tools are available.
+
+We anticipate CATARM to be able to successfully solve all the case
+studies with resonable computational resources available to the 
+average user. 
+We will also generalize the components we will develop to solve
+the case studies into a library of components that can be used
+in other case studies.
+
+\subsection{CATRAM evaluation }
+
+The team will seek the assistance and help of an independent 
+consultant in the middle of the second and third years of 
+the duration of the project to evaluate and validate the results.
 
@@
 
-Once completed the Arabic text analysis framework will be an open source contribution and developers from the Arabic world in the open source community are expected to join efforts to keep it going.
- Furthermore, once completed the Arabic text analysis framework can sustain itself by charging on support or expert use.
- The framework or service on the framework can easily be commercialized to tend to the security, health, insurance, news, media, or literature business sectors.
-
-The area of Arabic text analysis will be a research area for a long time as NLP is one of the most complex computational problems.
- The Arabic text analysis framework will be used by NLP researchers in the Arabic language field and those are expected to maintain and advance the framework.
+Once completed the Arabic text analysis framework will be 
+an open source contribution and developers from the Arabic 
+world in the open source community are expected to join 
+efforts to keep it going.
+Furthermore, once completed the Arabic text analysis framework 
+can sustain itself by charging on support or expert use.
+
+The framework or service on the framework can easily be 
+commercialized to tend to the security, health, insurance, market,
+news, media, or literature business sectors.
+
+The area of Arabic text analysis will be a research area for 
+a long time as NLP is one of the most complex computational 
+problems.
+The Arabic text analysis framework will be used by NLP 
+researchers in the Arabic language field and those are 
+expected to maintain and advance the framework.
+
+\subsection{Future work}
+
+Enhance morphological analysis
+Look for other optimizations
+Generalize the framework to other Arabic-like languages
+Generalize the case-based approach 
+Integrate other text mining techniques such as vocalization,
+suggestions...
 

### 2010-12-10 tex/rp.tex
modified the Future work section in the proposal, but it needs some proofreading.

@@
 
+The team will be formed by the two main PIs, Dr. Fadi Zaraket (LPI)
+and Dr. Rehab Duwairi (Co-LPI) and three research assistants.
+Currently the LPI is working with one research assistant 
+and two senior students on an Arabic morphological analyzer.
+
+The LPI offers a course on computational Arabic attended
+by excellent senior and graduate students at AUB. 
+Once funded, the team can be directly formed and work can start
+with productive team members.
+
 \subsection{Work plan}
 
-\subsection{Budget}
+The Lead PI will conduct biweekly meetings to track progress 
+on the project.
+The team will maintain a dokuwiki based website to continuously 
+post updates on the progress of the project.
+
+\subsection{Budget items}
+
+In the following we list the budget items. 
+Monatary values are provided in more details on the QNRF 
+submission website.
+
+\begin{itemize}
+\item One research assistant at QU
+\item Two research assistants at AUB, one of them will 
+work partially from QU. 
+\item Two month salary per year for the Co-LPI
+\item Two month salary per year for the LPI
+\item One course buy out per year for the Co-LPI
+\item Two weeks consultant for the second and the third years
+\item Licensing and membership fees to acquire digital libraries and corpora
+\item High performance computer and storage devices
+\item One conference for each PI per year
+\item One conference for one RA from QU per year
+\item One conference for one RA from AUB per year
+\item Site visit to AUB 
+\item Site visit to QU 
+\item Other miscellanous items
+\end{itemize}
 
@@
 in details in Section~\ref{s:design}. 
-
-The following is a list of the anticipated components They are and are repeated for convenience as follows.
-\begin{enumerate}
-\item Case-based morphological analyzer 
-\item Linguistic computational model for Arabic
-\item User query language
-\item Optimization methodologies
-\begin{enumerate}
-\item Use of language specific features 
-\item Use of data in question specific features
-\end{enumerate}
-\item Entity and relational extractors
-\item Statistical tools
-\item Graph and text algorithms
-\item Query solver
-\item Case studies 
+In the following we will discuss what to anticipate and 
+how to evaluate the different components of CATARM.
+
+\subsection{Case-based morphological analyzer}
+
+We will deliver a case-based morphological analyzer. 
+The analyzer will be compared to other available analyzers such as
+Buckwalter~\cite{Tim04} and ElixirFM~\cite{Otakar:07} in terms
+of accuracy and efficiency of analysis.
+We will adopt two comparison methodologies. 
+First, we will compare the analyzers in an exhaustive analysis mode
+which is the only way current analyzers perform. 
+Second, we will compare the analyzers under several case-studies
+and show how a case-based morphological analyzer can benefit
+from case-based optimizations.
+
+\subsection{Linguistic computational model for Arabic}
+
+We will build a linguistic computational model for Arabic 
+based on an understanding of the formal Arabic grammar 
+and on an abstraction of the current existing concatenation
+based lexicons.
+We will compare our LCM with that of ElixirGM~\cite{Otakar:07}.
+We will study how much such a model can affect the analysis
+in terms of accuracy and efficiency as well.
+
+\subsection{User query language}
+
+We will build a query language for relational Arabic text mining 
+and a front-end to compile a user query into an optimized
+computation. 
+We will check how user friendly and how expressive the language is.
+We will compare our query language with the sophisticated 
+interface of the MADA+TOKAN toolkit~\cite{Habash:09}.
+We will look for other attempts to formalize Arabic text 
+mining queries and we will also compare agains 
+formal text mining queries from other text mining domains such
+as the gene ontolgy domain~\cite{GeneOntology10},
+
+\subsection{Optimizations}
+
+We will develop and evaluate our two optimization methodologies
+and look for other optimizations. 
+First we will use language specific features based on the 
+LCM of Arabic, the grammar of the query language, and 
+the Arabic query provided by the user to simplify 
+the LCM and the morphological analysis. 
+Second we will look at language independent techniques
+such as HMM and LSI models extracted from the 
+text documents and use that to simplify the analysis. 
+We will perform the analysis with and without the optimizations
+and compare the results in terms of accuracy and efficiency.
+
+%\subsection{ Entity and relational extractors}
+\subsection{CATARM and case studies }
+
+We will evaluate CATARM against at least three of the case studies
+that we discussed previously and that are listed below. 
 \begin{enumerate}
@@
 
-Once completed the Arabic text analysis framework will be an open source contribution and developers from the Arabic world in the open source community are expected to join efforts to keep it going.
- Furthermore, once completed the Arabic text analysis framework can sustain itself by charging on support or expert use.
- The framework or service on the framework can easily be commercialized to tend to the security, health, insurance, news, media, or literature business sectors.
-
-The area of Arabic text analysis will be a research area for a long time as NLP is one of the most complex computational problems.
- The Arabic text analysis framework will be used by NLP researchers in the Arabic language field and those are expected to maintain and advance the framework.
+Once completed the Arabic text analysis framework will be 
+an open source contribution and developers from the Arabic 
+world in the open source community are expected to join 
+efforts to keep it going.
+Furthermore, once completed the Arabic text analysis framework 
+can sustain itself by charging on support or expert use.
+
+The framework or service on the framework can easily be 
+commercialized to tend to the security, health, insurance, market,
+news, media, or literature business sectors.
+
+The area of Arabic text analysis will be a research area for 
+a long time as NLP is one of the most complex computational 
+problems.
+The Arabic text analysis framework will be used by NLP 
+researchers in the Arabic language field and those are 
+expected to maintain and advance the framework.
+
+\subsection{Future work}
+
+Completion of the above goals would result in a significant contribution to the research community of NLP in general and Arabic Analysis in particular.
+However, as already noted, text mining is a wide problem that cannot be expected to saturate in the near future especially when it comes to the arabic language where much work is lacking.
+Consequently, we look forward for additional contributions to the field such as the following:
+\begin{itemize}
+\item Enhance morphological analysis: It is always the case, that the lexicon upon which the morphological analyzer can be augmented with useful categorization and entity tags that reflect semantical meanings.
+These modifications would give CATARM additional capabilities in terms of allowing users to extract and use a greater variety of entities when building instructive queries.
+The feedback of experts in the field of arabic linguistics can continually add accuracy to the morphological solutions.
+\item Look for other optimizations: From our priliminery research and results we were able to envision some optimizations to current tools; that's why it is expected that as we develop the complete framework to stumble at much more ideas and suggestions for improving the text mining process we are conducting expecially in terms of efficency and accuracy.
+\item Generalize the framework to other Arabic-like languages: Some sematic languages share many common features to the Arabic language such as the Amharic, Hebrew, etc...
+or even are written using the Arabic script like Persian, Kurdish, Urdu, Pashto.
+Accordingly, one may consider porting our techniques to them, or even generalizing the framework to support for languages other than Arabic.
+\item Generalize the case-based approach: ??
+\item Integrate other text mining techniques such as vocalization.
+Vocalization refers to the process of adding diacritics to an unvocalized text.
+The prerequisites of a successful vocalizer are the accuracy of the morphological analyzer in addition to grammatical analyzers.
+Such tools are especially useful in the academic sector where fully vocalized texts may be of high utility.
+\item Maintenance: We should never forget that for any software to remain useful, it must be continuously monitored to correct bugs, improve its interface, and enrich its capabilities.
+That's why CATARM should undergo extensive testing, fixes, and enhancements even after launching.
+\end{itemize}
 

### 2010-12-10 trunk/tex/rp.tex
future + prelim

@@
 %\usepackage{amsthm}
-%\usepackage{booktabs}
+\usepackage{booktabs}
+\usepackage{multirow}
 \usepackage{natbib}
@@
 %study.
-
-%IR tools are similar to a Google search 
-%engine where documents of interest are selected based on the 
-%relevance to a set of keywords of interest to the user.
-%NLP targets the automation of understanding human languages.
-%This task is the oldest and most difficult task in the artificial 
-%intelligence domain.
-%The main difficulty behind NLP comes from the ambiguity of 
-%natural languages~\cite{Osm08}.
-%While NLP techniques cannot deliver their final target yet, they 
-%can deliver analysis of sentences into nouns, verbs, and adjectives.
-%They can also narrow the meaning of a word or a phrase based on 
-%the context to one of its possible meanings.
-%Finally they can parse a sentence into a relation between the 
-%discovered entities using abstractions such as verb-name 
-%abstractions~\cite{Osm08}.
-
-%Information extraction (IE) targets forming data structures and 
-%instances of these data structures out of a collection of documents.
-%IE tries to fit the output of IR and NLP to templates of 
-%interest to the user.
-%Data mining can be applied at the end on the output of the 
-%IE process to answer user queries about the input 
-%documents~\cite{JHa05}.
-
-%\subsection{Arabic text mining in academy }
-%In the following we review the major works on Arabic text mining 
-%in academy and industry.
 
@@
 solve at least three interesting relational Arabic text case studies 
-presented in Section~\ref{s:design:casestudies}.
+presented in Sections~\ref{s:design:sec} and~\ref{s:design:lit}.
 
@@
 controlling the morphological analyser.
-We explain this approach in details in Section~\ref{s:design:fsm}.
+We explain this approach in details in 
+Section~\ref{s:design:method}.
 ATSarf along with other computational components, 
@@
 \resizebox{.75\columnwidth}{!}
-{\includegraphics{figs/narrator_chain_output.pdf}
-\label{f:narrators} } 
+{\includegraphics{figs/narrator_chain_output.pdf}}
 \caption{Directed acyclic graph representing a partial order 
     relation between narrators extracted using ATSarf}
+\label{f:narrators} 
 }
@@
 related NLP conferences.
+
+\subsection{PI related work}
+
+The Co-LPI, Rehab Duwairi, has several years of experience 
+working on data mining for Arabic documents. 
+She initially targeted Arabic text categorization where a 
+dataset of 1000 documents where used to train and test 
+classifiers~\cite{Duw06,Duw07}.
+Word morphology comes to the surface when extracting the 
+feature vectors of documents. 
+In that work, words were stemmed before using them as features. 
+Also, word frequency and inverted term frequency were used to 
+quantify the importance of features.
+
+In a recent work, Duwairi, Refae and Khasawneh~\cite{Duw09}
+have addressed text categorization again. 
+This research investigated possible word forms that can be 
+used to create feature vectors of documents in particular: 
+original word, stem of the word, light-stem (shallow stemming) 
+of the word, and word augmented with a list of synonyms. 
+Therefore, every document is represented using four different 
+feature vectors, the result revealed that light-stemming is the 
+best representation as it preserves the meaning of the words 
+and yet reduces the dimensions of feature vectors. 
+The dataset size was 15,000 documents that belong to three 
+categories.
+
+In recent research efforts, the CO-LPI is investigating 
+keyphrase extraction fro Arabic documents.
 
@@
 
-%The Lead PI worked in the area of software arabization since 1996 
-%as he worked on developing Arabization modules for Windows NT and 
-%an 
-%Arabic string manipulation library~\cite{Zar96}. 
-%The Lead PI is also involved in working on an open source 
-%project that aims at comprehensive automation of authenticity 
-%checks against a corpus of literature heritage~\cite{Zaraket06}.
-%The Lead PI is leading an effort at AUB to introduce 
-%Computational Arabic in the curriculum of the ECE program. 
-%He is also leading a team of students in an effort to place 
-%the infrastructure needed for this project to start. 
-%The lead PI worked and published about relational logic and 
-%logic solvers applied to structured languages as follows.
-%\begin{itemize}
-%\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
-%\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
-%\end{itemize}
+The Lead PI worked in the area of software arabization since 1996 
+as he worked on developing Arabization modules for Windows NT 
+and an Arabic string manipulation library~\cite{Zar96}. 
+The Lead PI is also involved in working on an open source 
+project that aims at comprehensive automation of authenticity 
+checks against a corpus of literature heritage~\cite{Zaraket06}.
+The Lead PI is leading an effort at AUB to introduce 
+Computational Arabic in the curriculum of the ECE program. 
+He is also leading a team of students in an effort to place 
+the infrastructure needed for this project to start.
+
+The lead PI holds several approved patents and has several 
+patents pending approval in the area of logic verification 
+and synthesis.
+The lead PI worked and published about relational logic and 
+logic solvers applied to structured languages as follows.
+\begin{itemize}
+\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
+\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
+\end{itemize}
 
@@
 
-Text mining concerns the partitioning of text segments into classes,
+Text mining concerns the partitioning of text segments 
+into classes,
 the clustering of text segments,
@@
 %the Arabic language.
+
+IR tools are similar to a Google search 
+engine where documents of interest are selected based on the 
+relevance to a set of keywords of interest to the user.
+NLP targets the automation of understanding human languages.
+This task is the oldest and most difficult task in the artificial 
+intelligence domain.
+The main difficulty behind NLP comes from the ambiguity of 
+natural languages~\cite{Osm08}.
+While NLP techniques cannot deliver their final target yet, they 
+can deliver analysis of sentences into nouns, verbs, 
+and adjectives.
+They can also narrow the meaning of a word or a phrase based on 
+the context to one of its possible meanings.
+Finally they can parse a sentence into a relation between the 
+discovered entities using abstractions such as verb-name 
+abstractions~\cite{Osm08}.
+
+IE targets forming data structures and instances of these 
+data structures out of a collection of documents.
+IE tries to fit the output of IR and NLP to templates of 
+interest to the user.
+Data mining can be applied at the end on the output of the 
+IE process to answer user queries about the input 
+documents~\cite{JHa05}.
 
@@
 {\includegraphics{figs/literature_case-study1.pdf}
-\label{f:literature}
 } 
 \caption{CATARM diagram for the literature case study }
+\label{f:literature}
 }
@@
 component that we will discuss later in 
-Section~\ref{s:design:chainext}. We  obtain a set of chains
+Section~\ref{s:design:method}. We  obtain a set of chains
 of narrators that we feed to a morphological analyzer (MA) 
@@
 {\includegraphics{figs/security_case-study.pdf}
-\label{f:security}
 }
 \caption{CATARM diagram for the security case study}
+\label{f:security}
 }
@@
 
-\subsection{CATRAM components}
+\subsection{CATARM components}
 \label{s:design:components}
@@
 \subsection{Methodology}
+\label{s:design:method}
 
@@
 {\input{figs/hadith.pdftex_t}
-\label{f:statemachine}
 }
 \caption{State-machine used to detect the chain of narrators}
+\label{f:statemachine}
 }
@@
 can be encoded in three FSM with linear execution time using 
-the trie~\cite{datries} data structure.
+the trie~\cite{Aoe:89} data structure.
 
@@
 {\includegraphics{figs/detailed_submachine.pdf}
-\label{f:chain}
 }
 \caption{CATARM detailed diagram for the chain of narrators extractor}
+\label{f:chain}
 }
@@
 
-The solver may make use of existing propositional~\cite{MiniSAT} or
-higher order logic~\cite{Z3} solvers if needed. 
+The solver may make use of existing 
+propositional~\cite{MiniSAT04} or
+higher order logic~\cite{Z308} solvers if needed. 
 These solvers can solve satisfiability questions.
@@
 
-{\bf Development methodology.}
-We will integrate all the mentioned techniques in a scalable 
-framework for Arabic text analysis.
-The framework as a whole will allow for exhausting Arabic text 
-analysis and allow for extensions of the framework.
-A user should be able to iteratively and repetitively apply the 
-abode techniques on his input text.
-
-We will follow an eXtreme Programming 
-methodology in the process of building CATARM.
-We will use open source tools, libraries, and frameworks to 
-build our libraries.
-We will build on existing open source programming and text 
-mining tools as well as open source Arabic text analysis tools 
-based on NLP such as MADA+TOKAN~\cite{Rot08}, 
-NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
-
-We will use a dokuwiki based website~\cite{Dok09} 
-to document the progress online and will use 
-subversion~\cite{Sub09} as a source control tool.
-
 {\bf Visualization.} 
@@
 Once funded, the team can be directly formed and work can start
-with productive team members.
+with readily productive team members.
 
 \subsection{Work plan}
+\label{s:design:plan}
+
+%\begin{table}[bt]
+\begin{table}[bt]
+\centering
+\caption{Three year quarter based work plan.}
+%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
+%\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
+\small
+\begin{tabular}{lp{.1cm}ccccp{.1cm}ccccp{.1cm}cccc} \\
+& & \multicolumn{4}{c}{Year 1} & & \multicolumn{4}{c}{Year 2} & & \multicolumn{4}{c}{Year 3} \\ \cline{3-6} \cline{8-11} \cline{13-16} %\\
+    %\cline{2-13} \\ 
+& & Q1 & Q2 & Q3 & Q4 & & Q1 & Q2 & Q3 & Q4 & & Q1 & Q2 & Q3 & Q4 \\ \bottomrule
+
+Case based & & 
+W & W & W & E & & E & E & E & E & & E & E & E & E \\ 
+morphological analyzer & & 
+ &  & D & DM & &   & M & M & M & & M & M & M & M \\  \hline
+
+\multirow{2}{*}{Arabic LCM} & & 
+P & W & W & W & & W & E & E & E & & E & E & E & E \\ 
+& & 
+ &  &  &  & & D & DM & M & M & & M & M & M & M \\ \hline
+
+\multirow{2}{*}{Query language} & & 
+P & P & W & W & & W & W & E & E & & E & E & E & E \\ 
+& & 
+  &  &  &  & &  & D & DM & M & & M & M & M & M \\ \hline
+
+Language & & 
+P & P & P & W & & W & W & W & W & & E & W & W & W \\ 
+optimizations & & 
+ &  &  &  & &  &  &  & D & & DM & M & M & D \\ \hline
+
+Data  & & 
+P & P & P & W & & W & W & W & W & & E & W & W & W \\ 
+optimizations & & 
+ &  &  &  & &  &  &  & D & & DM & M & M & D \\ \hline
+
+\multirow{2}{*}{E/R extractors} & & 
+W & W & W &  & & E & E & E & W & & W & W &  & E \\ 
+& & 
+ &  & D & DM & & M & M & M & M & & M & D & DM & M \\ \hline
+
+Statistical & & 
+W & W & W & E & & WE & WE & WE & E & & WE & WE & WE & WE \\ 
+components & & 
+ &  & D & DM & & M & M & M & DM & & M & M & M & M \\ \hline
+
+Graph  & & 
+W & W & W & E & & WE & WE & WE & E & & WE & WE & WE & WE \\ 
+components & & 
+ &  & D & DM & & M & M & M & DM & & M & M & M & M \\ \hline
+
+\multirow{2}{*}{Query solver} & & 
+P & P & W & W & & W & W & W & E & & E & E & E & E \\ 
+& & 
+ &  &  &  & &  &  & D & DM & & M & M & DM & M \\ \hline
+
+\multirow{2}{*}{Case studies} & & 
+W & W & W & W & & W & W & W & W & & W & W & WE & E \\ 
+& & 
+ &  & D$^1$ &  & &  &  & D$^2$ &  & &  & D$^3$ & D$^4$ M & M \\ \hline
+
+\multirow{2}{*}{Visualization} & & 
+P & W & W & W & & E & E & E & E & & E & E & E & E \\ 
+& & 
+ &  &  & D & & M & M & M & M & & M & M & M & M \\ \hline
+
+\multirow{2}{*}{Integration} & & 
+P & W & W & W & & W & W & W & W & & W & W & W & E \\ 
+& & 
+ &  &  & D & & D & D & D & D & & D & D & D & M \\  \hline
+
+Validation & & 
+W & W & W & W & & W & W & W & W & & W & W & W & W \\  \bottomrule
+        %\cline{2-13} \\
+\end{tabular}
+\normalsize
+\label{t:workplan}
+\end{table}
+
+We will follow an eXtreme Programming 
+methodology in the process of building CATARM.
+This means that we will define refined incremental items of 
+work that allows for more frequent deliverables. 
+This also means that the code base will always be subject
+to maintenance and refactoring to enhance reuse and 
+increase productivity. 
+While a item is delivered, it will still be subject to work,
+enhancements, and maintenance.
+
+We detailed an agressive work plan in Table~\ref{t:workplan}. 
+The plan spans over three years and projects a timeline
+of work per quarter. 
+The row headers describe work items, and the column headers
+describe quarters in the corresponding year of the project.
+The legends {\bf P, W, D, M, } 
+and {\bf E} 
+describe the status of the items 
+and stand for 
+{\bf P:} {\em planning phase}, 
+{\bf W:} {\em work in progress}, 
+{\bf D:} {\em deliverable}, 
+{\bf M:}  {\em maintainance}, 
+and 
+{\bf E: } {\em enhance} respectively.
+
+The $1,2,3,$ and $4$
+superscripts of the deliverables in the 
+in case studies row refer to the four 
+case studies described in the propoal.
+
+We will use open source tools, libraries, and frameworks to 
+build our libraries.
+We will build on existing open source programming and text 
+mining tools as well as open source Arabic text analysis tools 
+based on NLP such as MADA+TOKAN~\cite{Rot08}, 
+NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
+
+We will use a dokuwiki based website~\cite{Dok09} 
+to document the progress online and will use 
+subversion~\cite{Sub09} as a source control tool.
 
@@
 The components of the framework are listed and discussed
-in details in Section~\ref{s:design}. 
+in details in Section~\ref{s:designmethods}.
 In the following we will discuss what to anticipate and 
@@
 
-\subsection{CATRAM evaluation }
+\subsection{CATARM evaluation }
 
@@
 
-Enhance morphological analysis
-Look for other optimizations
-Generalize the framework to other Arabic-like languages
-Generalize the case-based approach 
-Integrate other text mining techniques such as vocalization,
-suggestions...
+The completion of the CATARM framework will result in a 
+significant contribution to the research community of 
+NLP in general and Arabic Analysis in particular. 
+The CATARM framework will act as vehicle to enhance and advance 
+research in the area. 
+We look forward for additional contributions to the field 
+such as the following.
+\begin{itemize}
+\item We will explore developing and integrating vocalization
+techniques. 
+Vocalization techniques assign diacritics to unvocalized text. 
+The prerequisites of a successful vocalizer are the accuracy of 
+the morphological analyzer in addition to grammatical analyzers. 
+\item We will explore the use of CATARM with more case studies
+such as market analysis and opinion mining.
+\item We will explore augmenting the lexicon of the
+morphological analyzer with useful categorization and entity 
+tags that reflect semantical meanings. 
+These modifications would give CATARM additional capabilities in 
+terms of allowing users to extract and use a greater variety of 
+entities when building queries. 
+The feedback of experts in the field of Arabic linguistics can 
+continually add accuracy to the morphological solutions.  
+\item We will explore further optimization techniques
+that can help simplify and the analysis. 
+\item We will explore generalizing our results to languages 
+that resemble the Arabic language. 
+Some semetic languages, such as the Amharic, cyrillic, and Hebrew,
+share common features with the Arabic 
+language. Some other languages are written using the 
+Arabic script like Persian, Kurdish, Urdu, and Pashto. 
+\item We will also investigate the case-based approach with
+Latin languages and check how that compares with current techniques
+in terms of accuracy and efficiency.
+\item We will maintain CATARM and provide support to users and 
+researchers of CATARM.
+\end{itemize}
 
@@
 as an open source tool. 
-Interested researchers and users will be able to download the tools 
-for free and use them for non-profit non-commercial purposes.
+Interested researchers and users will be able to download the 
+tools for free and use them for non-profit non-commercial 
+purposes.
 

### 2010-12-10 trunk/tex/rp.tex
spell check

@@
 ``run-on'' words~\cite{Buckwalter:04}.
-Such difficulties are inherrent to the 
+Such difficulties are inherent to the 
 morphological analysis of the Arabic language and
@@
  {\bf (2.)} It will be amenable to Arabic specific simplifications 
-inferred from the correspondence between the fromal Arabic rules
+inferred from the correspondence between the formal Arabic rules
 and the relational user queries.
@@
 Finally, we will build a system using CATARM to check the strength
-of the relations between internet documents viewed by a minor and
+of the relations between Internet documents viewed by a minor and
 a set of Arabic text documents containing unrated or unwanted 
@@
 that is available now online.
-They have successfully developped ATSarf, a case-based morphological 
+They have successfully developed ATSarf, a case-based morphological 
 analyzer, and applied it on the hadith literature case study.
@@
 and once
-connected toghether they form complex names of narrators. 
+connected together they form complex names of narrators. 
 Narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
@@
 specify his queries as one or more FSM
-controlling the morphological analyser.
+controlling the morphological analyzer.
 We explain this approach in details in 
@@
 She initially targeted Arabic text categorization where a 
-dataset of 1000 documents where used to train and test 
+data set of 1000 documents where used to train and test 
 classifiers~\cite{Duw06,Duw07}.
@@
 and yet reduces the dimensions of feature vectors. 
-The dataset size was 15,000 documents that belong to three 
+The data set size was 15,000 documents that belong to three 
 categories.
@@
 of the narrator names and will decide to merge two narrators in 
-one graph node in case they were morphologicaly equal. 
+one graph node in case they were morphologically equal. 
 It will also give feedback to the MA component to stop the
@@
 a graph coloring routine to color the graph of narrators. 
-The color changes whener two consecutive narrators are far away
+The color changes when two consecutive narrators are far away
 from each other in the text. 
@@
 locations and a database of personal records. 
-The aim is to check for inconsistencies in the iterrogation
+The aim is to check for inconsistencies in the interrogation
 reports and to identify possible connections between the suspects
@@
 $\alpha_k$ of how much $G_R|_k$ simulates $G_k$. 
-It flags the components in the graph where the accumelated 
+It flags the components in the graph where the accumulated 
 measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
@@
 and ElixirFM~\cite{Otakar:07} take an Arabic word as input
-and produce several valid morpholgical solution for the word. 
+and produce several valid morphological solution for the word. 
 Each solution is a sequence of morphemes that form the word.
@@
 Analyzers that enumerate all possible partitions of a word
-and match them against concatenative rules are computationaly 
+and match them against concatenative rules are computationally 
 expensive. 
@@
 decisions. 
-For example, in the hadith literature case study pesented above, 
-amongst other things, we are intersted in checking whether a 
+For example, in the hadith literature case study presented above, 
+amongst other things, we are interested in checking whether a 
 word is a name.  
@@
 the analysis of the rest of the word. 
-This is more efficient than generating all possible partitionings
+This is more efficient than generating all possible partitions
 of the word, checking for morphological correctness of each 
@@
 
-\subsection{CATARM Query language}
+\subsection{CATARM query language}
 \label{s:design:query}
@@
 
-The state machine in Firgure~\ref{f:statemachine} controls 
+The state machine in Figure~\ref{f:statemachine} controls 
 the concatenation based state machines of the morphological
@@
 the list of chains. 
-{\bf 2.} Create a new chain whith the last element.
+{\bf 2.} Create a new chain with the last element.
 Finally, we filter those chains whose cardinality 
@@
 the query $q$ in $L_q$. 
-We plan to inspect methods to overapproximate 
+We plan to inspect methods to over-approximate 
 the morphemes and the transitions in $L_A$ 
@@
 
-We detailed an agressive work plan in Table~\ref{t:workplan}. 
+We detailed an aggressive work plan in Table~\ref{t:workplan}. 
 The plan spans over three years and projects a timeline
@@
 {\bf D:} {\em deliverable}, 
-{\bf M:}  {\em maintainance}, 
+{\bf M:}  {\em maintenance}, 
 and 
@@
 in case studies row refer to the four 
-case studies described in the propoal.
+case studies described in the proposal.
 
@@
 In the following we list the budget items. 
-Monatary values are provided in more details on the QNRF 
+Monetary values are provided in more details on the QNRF 
 submission website.
@@
 \item Site visit to QU 
-\item Other miscellanous items
+\item Other miscellaneous items
 \end{itemize}
@@
 We will look for other attempts to formalize Arabic text 
-mining queries and we will also compare agains 
+mining queries and we will also compare against 
 formal text mining queries from other text mining domains such
-as the gene ontolgy domain~\cite{GeneOntology10},
+as the gene ontology domain~\cite{GeneOntology10},
 
@@
 We anticipate CATARM to be able to successfully solve all the case
-studies with resonable computational resources available to the 
+studies with reasonable computational resources available to the 
 average user. 
@@
 that resemble the Arabic language. 
-Some semetic languages, such as the Amharic, cyrillic, and Hebrew,
+Some Semitic languages, such as the Amharic, Cyrillic, and Hebrew,
 share common features with the Arabic 

### 2010-12-12 trunk/tex/pr.tex
added the progress report for LNCSR

--- 2010-12-09 trunk/tex/rp.tex
@@
 \chead{}
-\rhead{NPRP No.~:~~~~4-484-1-075}
+\rhead{}
 
-\lfoot{QNRF Form}
+\lfoot{LNCSR Progress Report}
 \cfoot{}
@@
 \begin{center}
-{\Large \bf Relational Arabic Text Mining Framework using 
-    Morphological
-    and Case-Based Analysis }
+{\Large \bf Relational Queries \\
+for Arabic Text Mining }
 
@@
 \begin{tabular}{cc}
-Fadi Zaraket & Rehab Duwairi \\
-Electrical and Computer Engineering &  Computer Science and Engineering Department \\
-American University of Beirut & University of Qatar \\
-{\tt fadi.zaraket@aub.edu.lb} & {\tt Rehab.Duwairi@qu.edu.qa}
+Fadi Zaraket \\
+Electrical and Computer Engineering  \\
+American University of Beirut \\
+{\tt fadi.zaraket@aub.edu.lb} 
 \end{tabular}
@@
 \begin{tabular}{c}
-{\small Research Plan Document } \\
-{\small Proposal Submitted to }\\
+{\small Research Progress Report } \\
+{\small Submitted to }\\
 \\
-    Qatar National Research Foundation \\
-    National Priorities Research Program \\
-    4$^{th}$ Cycle 
+    Lebanese National Council for Scientific Research \\
+    (LNCSR)
 \end{tabular}
@@
 
-\section{Background (max of 2 pages)}
-\label{s:background}
+This report describes the progress of the work carried out in the LNCSR funded project entitled: {\bf ``Relational Queries for Arabic Text Mining"}. 
+\\
 
-Research in text mining started in the mid 1980s when Swanson 
-realized that slicing and combining seemingly unrelated medical 
-articles led to the discovery of new 
-hypotheses~\cite{JNi06}.
-Till present, most of the interest in text mining comes 
-from biological sciences.
-Lately we saw applications of text mining in the advertisement, 
-political campaigning, and other businesses.
-Text mining involves information retrieval (IR), natural language 
-processing (NLP), information extraction (IE), 
-and data mining~\cite{ASr09}.
-
-Most of academic research in the field of Arabic text mining 
-is covered in the almost comprehensive book 
-``Arabic Computational Morphology''~\cite{Sou07}.
-The book summarizes recent work, and presents strong evidence of 
-shortage of historic research. It also highlights  an emerging 
-interest in this area with a specific application to machine 
-translation from Arabic to English after 2001.
-The research community showed little interest in 
-automating Arabic text analysis that may serve
-the Arabic user.
-While a non-Arabic user is interested in machine translation to 
-help him understand an Arabic document, 
-an Arabic user is more interested in 
-information hard to discover manually such as 
-the relation between clinical records, pharmaceutical 
-prescriptions and sales, and a specific sickness. 
-Another Arabic user may be interested in looking at the relation
-between stock prices and political news or the 
-relation between suspect profiles in interrogation reports 
-and investigative reports from the crime scenes.
-
-We propose to build a 
-%{\em relational text mining 
-%of Arabic using morphological and case-based analysis}. 
-{\em case-based Arabic text analysis using relations and morphology}(CATARM) framework. 
-CATARM will take as input sets of documents, each of a specific
-type, and provides the user with a visual query language that
-allows him to build a flow of Arabic text mining
-modules and discover desired relations between documents. 
-CATARM will translate the query into a netlist of computational 
-components, it will also extract parameters from the query and 
-the document sets and use them to simplify the computational 
-components. 
-%The main CATRAM computational components will be finite 
-%state machines (FSM) that express a linguistic model of the Arabic 
-%language and that can be parametrized and relaxed to fit a case 
-%study.
-
-%IR tools are similar to a Google search 
-%engine where documents of interest are selected based on the 
-%relevance to a set of keywords of interest to the user.
-%NLP targets the automation of understanding human languages.
-%This task is the oldest and most difficult task in the artificial 
-%intelligence domain.
-%The main difficulty behind NLP comes from the ambiguity of 
-%natural languages~\cite{Osm08}.
-%While NLP techniques cannot deliver their final target yet, they 
-%can deliver analysis of sentences into nouns, verbs, and adjectives.
-%They can also narrow the meaning of a word or a phrase based on 
-%the context to one of its possible meanings.
-%Finally they can parse a sentence into a relation between the 
-%discovered entities using abstractions such as verb-name 
-%abstractions~\cite{Osm08}.
-
-%Information extraction (IE) targets forming data structures and 
-%instances of these data structures out of a collection of documents.
-%IE tries to fit the output of IR and NLP to templates of 
-%interest to the user.
-%Data mining can be applied at the end on the output of the 
-%IE process to answer user queries about the input 
-%documents~\cite{JHa05}.
-
-%\subsection{Arabic text mining in academy }
-%In the following we review the major works on Arabic text mining 
-%in academy and industry.
-
-Arabic morphological analyzers~\cite{Sughaiyer:04}
-consider an Arabic word and its internal structure composed of 
-several {\em morphemes}. 
-A morpheme is a {\em stem} or an {\em affix}.
-An affix is a {\em prefix, suffix,} or an {\em infix}.
-The analysis of one word may lead to several possible
-solutions.
-\vocalize
-For instance, the word 
-\RL{'a.hmadH}~\footnote{In this document, we use the default 
-ArabTeX transliteration style ZDMG.}
-may have two valid morphological analyses. 
-The word means ``I praise him'' when
-the letter \RL{'a} is a prefix and  it means
-``his Ahmad'' when 
-the letter \RL{'a} is part of the proper noun stem 
-\RL{'a.hmad}.
-\novocalize
-
-%The accuracy of the solutions suffer due to inherent difficulties
-%of morphological analysis of the Arabic language. 
-It is common practice to write Arabic text
-without short vowels. 
-This greatly increases the ambiguity of Arabic text. 
-Arabic letters can have up to 
-four different forms
-corresponding to their position in a word, i.e, beginning,
-middle, end of word and separate forms. 
-This allows the phrase
-\noTrRL{il_A\nospace almdrsT}  with no delimiter within
-to be visually recognizable
-as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
-This happens because the first word \RL{almdrsT} ends with
-\RL{T} a non-connecting letter. 
-These words occur often in text and greatly increase the
-difficulty of tokenization and are referred to as 
-``run-on'' words~\cite{Buckwalter:04}.
-Such difficulties are inherrent to the 
-morphological analysis of the Arabic language and
-render the solutions inaccurate.
-
-Current morphological analyzers such as 
-Buckwalter~\cite{Buckwalter:02},
-Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
-and ElixirFM~\cite{Otakar:07} exist.
-These analyzers are used in several open source spell checkers as 
-well as NLP frameworks~\cite{Col09}.
-They take as input white space delimited tokens~\cite{Kulick:10},
-consider them as words,
-and enumerate all possible solutions. 
-This approach has several problems. 
-First, a white space delimited token may have 
-more than one word.
-Second, the exhaustive enumeration may hurt performance and may
-not be necessary or appropriate
-in some case studies~\cite{Maamouri:10}. 
-Other morphological analyzers such as 
-Amira~\cite{Diab:07,Benajiba:07},
-MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
-use machine learning and support vector machines (SVM) 
-to enhance the accuracy at the expense of efficiency.
-
-{\bf Hypothesis 1.} We hypothesize that many case studies, 
-in particular relational
-case studies, do not require high accuracy from the 
-low level morphological analysis.
-The sophistication of the sought higher level query can eliminate 
-many of the ambiguous low level solutions at a lower computational 
-cost.
-For example, if a query concerns proper names and the 
-prefix in question in the analysis connects only to a verb, 
-we do not need to further analyze the rest of the word 
-to provide an answer.
-
-We find evidence to our hypothesis in~\cite{Maamouri:10}. The 
-addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
-required the high NLP task to interact with the low 
-level morphological analyzer.
-This led to a refined analyzer~\cite{Kulick:10}.  
-We also find evidence in~\cite{Habash:06} that different types of 
-morphological analyses behave differently on the same case study. 
-We strongly believe and we find evidence in our preliminary
-results that a 
-case-based morphological analyzer that allows a high level 
-case-based controller to intervene at every decision to 
-guide, use, prune, and refine the morphological analysis
-is of high utility.
-
-%Prior work in Arabic text mining addressed the classification and 
-%categorization of Arabic text and made little use of features 
-%unique to the Arabic language.
-Researchers in~\cite{AEL07,Ham07,Abd07,MEl03} 
-preprocessed the Arabic text to make it presentable to known text 
-mining techniques that tend to Latin text and queries.
-%Buckwalter~\cite{Bee89,Tim04} uses morphological analysis to enable 
-%relaxed stemming of Arabic words.
-El-Halees used statistical analysis to address the problem of 
-classifying Arabic text documents with maximum 
-entropy~\cite{AEL07}.
-In~\cite{Abd07} Mesleh evaluated six different techniques to 
-classify Arabic text using a support vector machine.
-El Dost~\cite{MEl03} used root words in the Arabic language to 
-speed up automated and hierarchical indexing.
-Al-Zoghby~\cite{Ham07}
-%addresses the problem of 
-%entities extracted from Arabic text with similar content but 
-%slightly different names via 
-used the derivation feature in 
-the Arabic language as a similarity measurement.
-This work is limited to the names of the fields in the data 
-structures extracted from Arabic texts.
-
-{\bf Hypothesis 2.} This approach does not tend directly to 
-the Arabic users who write their 
-queries in Arabic and think in Arabic.
-We hypothesize that Arabic queries use rules and 
-relations similar to the Arabic text under investigation, 
-and thus techniques native to the Arabic language will do better 
-in Arabic text analysis.
-
-We find evidence in the work of the CADIM \noTrRL{q-adim} 
-project~\cite{Col09} at the University of Columbia to support our 
-hypothesis. They observe that Arabic texts used as training 
-data for common NLP techniques proved ineffective. 
-They built semi-automatic models to address the problem of automatic 
-speech recognition in Arabic.
-
-%\subsection{\bf Arabic text mining in industry}
-Few companies in the industry provide primitive Arabic text 
-analysis tools.
-Sakhr~\cite{Sak09} provides a categorization 
-tool with a limited number of predefined categories, and a summary 
-generator that selects sentences from text.
-Basis technology~\cite{Bas09} provides Rosette, a lexical 
-analysis tool that normalizes Arabic text as a pre-processing step 
-for indexers and other general purpose data mining 
-tools.  Basis also provides REX; an entity extractor.
-
-ElixirFM~\cite{Otakar:07} uses an {\em inferential-realizational}
-approach for morphological analysis. 
-Realizational morphology considers the form of the word 
-\noTrRL{.sarf}  instead of its segments. 
-Formal rules and formulas express the relations between 
-the root of the lexeme and the fully inflected word. 
-While this is highly compatible with the Arabic linguistic 
-description~\cite{Badawi:04},
-and it benefits from features unique to the Arabic language
-captured by the formal rules, 
-it is computationally expensive and enumerates all the possible
-solutions.
-
-{\bf Goals.} Up to our knowledge, CATARM will be the 
-first framework that tends to relational Arabic morphology 
-and text mining. 
-{\bf (1.)} CATARM will build a novel parametrized linguistic 
-computational model of Arabic. 
-The model will be implemented using efficient structures. 
- {\bf (2.)} It will be amenable to Arabic specific simplifications 
-inferred from the correspondence between the fromal Arabic rules
-and the relational user queries.
-{\bf (3.)} The model will also be amenable to simplifications 
-inferred from statistical features extracted from the document 
-sets using language independent techniques such as latent semantic 
-indexing~\cite{LSI89} and hidden Markov models (HMM).
-%CATARM will augment current digital dictionaries to resolve 
-%entity semantics and targeted qualificationsl.
-{\bf (4.)} CATARM extracts semantic graphs from each text 
-document and then uses set arithmetic and graph analysis techniques
-to establish relations between these graphs. {\bf (5.)} CATARM will 
-solve at least three interesting relational Arabic text case studies 
-presented in Section~\ref{s:design:casestudies}.
-
+This document is organized as follows:
+\begin{enumerate}
+\item Summary of work conducted during the first year of the project
+\item Goals for next year
+\item Budget for year 2011
+\item Submission to: ``ACL/HTL 2011"
+\end{enumerate}
 \pagebreak
 
-\section{Objectives and significance (max of 1 page) } 
-\label{s:objectives}
-
-The current prevailing direction in research 
-in Arabic text mining and NLP focuses on machine translation from 
-Arabic to English. 
-A little attention is given to native Arabic text mining 
-problems and applications. 
-CATARM is a relational Arabic text mining 
-framework that tends to Arabic users 
-with the following objectives.
-
-\begin{itemize}\itemsep0pt
-\item CATARM will build a formal Arabic computational model
-using efficient and configurable data structures such as 
-finite state machines (FSM). 
-\item CATARM will provide the user with a visual query 
-language that makes it easy to enter relational queries. 
-The query language will also allow the framework to extract Arabic
-specific features from the query to simplify the linguistic 
-computational model. 
-\item CATARM will use statistical models and techniques
-to infer language independent features from the text documents
-that will further simplify the computational model. 
-\item During the project, we will augment current Arabic digital 
-dictionaries to resolve entity semantics and targeted 
-qualifications such as adjectives.
-\item The framework will allow the integration of library and 
-custom computational components built on top of primitive
-CATARM components. 
-The components will be the linguistic computational model,
-statistical, set arithmetic, logic,
-and graph analysis techniques, and  text algorithms.
-\item We will use CATARM to solve several realistic case studies.
-We will look at the hadith literature and build a system using
-CATARM to relate the hadith literature to the biographies of the
-narrators and extract an authenticity graph for the narrators.
-We will look at investigation and interrogation security reports
-and build a CATARM system that looks for relations between suspects
-in the reports. 
-We will look at clinical reports and relate them to pharmaceutical 
-records and find relations between sicknesses and prescribed medication. 
-Finally, we will build a system using CATARM to check the strength
-of the relations between internet documents viewed by a minor and
-a set of Arabic text documents containing unrated or unwanted 
-material. If the relation is strong, the system will flag access to
-the document. 
-All components and parts built during the development of solutions
-to the case studies will be generalized and added to CATARM as 
-library computational components. 
-\end{itemize}
-
-%\begin{itemize}\itemsep0pt
-%\item {\bf The literature case study.}
-%We are given a set of books of historical accounts where
-%a chain of narrators precedes each account,
-%and a second set of books of biographies where
-%a biography includes an evaluation of the authenticity 
-%of the narrator therein.
-%CATARM will allow the user to automatically and exhaustively query 
-%all accounts related to a subject important to Islamic jurispudence 
-%(for example women and travel),
-%and extract a graph where nodes represent narrators and edges
-%are labeled with the authenticity of the narrators. 
-%A similar manual query takes ages to answer and is error prone.
-%\item {\bf The security case study.}
-%Given two sets of reports consisting of crime scene investigation 
-%and suspect interrogation reports, 
-%an officer may query CATARM 
-%for relations between suspects in terms of 
-%criminal action, locality, or third party persons unknown to the 
-%officer.
-%Such a query is very hard to answer manually.
-%\item {\bf The net traffic case study.}
-%Given a set of undesired internet content for minors
-%such as violent text, and a set of documents requested by internet
-%minor users, 
-%CATARM can discover whether there is a relation between 
-%the document and the violent content and flag the document if the
-%relation is strong enough. 
-%\end{itemize}
-
-\subsection{Significance}
-
-CATARM will discover relations between Arabic text documents that 
-are hard to discover by humans and will have direct live and
-cost saving benefits in cases similar to the above mentioned 
-case studies. 
-CATARM will qualitatively enhance research in the cultural and 
-literature fields. 
-It may also help investors analyze Arabic news and data from Arab 
-markets prior to making important decisions.
-
-The use of Arabic text mining  will 
-have a high impact on the prosperity and safety of the Arab world 
-in particular and the world in general.
-%For example, having the tools to check the authenticity of a hadith 
-%at a click of a button may save youth from the hands of extreme 
-%religious leaders.
-For example, extracting a relation between two criminals may 
-uncover a third criminal who is ready to commit a crime.
-Finding inconsistencies in claims may curb down fraud and corruption 
-in the insurance sector.
-
-In addition, building the necessary academic 
-infrastructure for Arabic text mining in Lebanon and Qatar will 
-open the way for investments in the field and provide work 
-opportunities for native workers.
-
-\pagebreak
-
-\section{Preliminary data or studies}
+\section{Summary of work conducted during the first year of the project}
 \label{s:prelim}
 
-The PI and his team at the American University of Beirut (AUB)
+At the American University of Beirut (AUB), we
 started work on an Arabic text mining framework~\cite{ATMine09}
-that is available now online.
-They have successfully developped ATSarf, a case-based morphological 
-analyzer, and applied it on the hadith literature case study.
+that is available now online. The source for this framework comprises code of the order of tens of thousands lines.
+We have successfully developed ``SARF", a case-based morphological 
+analyzer, which constitutes capabilities of both Lexical and Syntax Analysis.
+We have been able to verify that both in terms of speed and accuracy our analyzer outperformed 2 state-of-art analyzers: Buckwalter ~\cite{Buckwalter:02} and ElixirFM~\cite{Otakar:07}.
+Supporting statistics are present in the attached paper described in Section~\ref{s:paper}.
+In addition, we have applied SARF on a literature hadith case study.
 
@@
 The framework now comprises a state of the art
-case-based morphological analyzer, ATSarf, 
+case-based morphological analyzer, SARF, 
 and computational components that allow the user to
 specify his queries as one or more FSM
-controlling the morphological analyser.
-We explain this approach in details in Section~\ref{s:design:fsm}.
-ATSarf along with other computational components, 
+controlling the morphological analyzer.
+SARF along with other computational components, 
 extract the chains of narrators from a set of narrations
@@
 
-The Co-Lead PI published extensively in this domain. 
-A selected list of her recent publications follows. 
-\begin{itemize}\itemsep0pt
-\item Rehab Duwairi, Mohammad Al-Refai, Natheer Khasawneh, 
-"Feature Reduction Techniques for Arabic Text Categorization". 
-Journal of the American Society for Information Science and Technology 
-(JASIST), Volume 60, Issue 11, pages: 2347-2352, 2009. 
-\item Rehab M. Duwairi, "Arabic Text Categorization", 
-International Arab Journal for Information Technology (IAJIT), 
-Volume 4, Number 2, pages 125  132, 2007.
-\item Rehab M. Duwairi, "Machine Learning for Arabic Text Categorization". 
-Journal of the American Society for Information Science and Technology 
-(JASIST), Vol. 57, Issue 8, pages 1005-1010, 2006.
-\end{itemize}
+\section{Goals for next year}
+\label{s:nextyear}
 
-%The Lead PI worked in the area of software arabization since 1996 
-%as he worked on developing Arabization modules for Windows NT and 
-%an 
-%Arabic string manipulation library~\cite{Zar96}. 
-%The Lead PI is also involved in working on an open source 
-%project that aims at comprehensive automation of authenticity 
-%checks against a corpus of literature heritage~\cite{Zaraket06}.
-%The Lead PI is leading an effort at AUB to introduce 
-%Computational Arabic in the curriculum of the ECE program. 
-%He is also leading a team of students in an effort to place 
-%the infrastructure needed for this project to start. 
-%The lead PI worked and published about relational logic and 
-%logic solvers applied to structured languages as follows.
-%\begin{itemize}
-%\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
-%\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
-%\end{itemize}
-
-\section{Research design and methods}
-\label{s:designmethods}
-
-We see more and more Arabic documents such as texts,
-books, publications, hospital and governmental records emerging 
-every day.
-Most of the newly generated documents are produced in digital 
-textual form while also old paper documents are ported to digital 
-form.
-These documents include huge amounts of information that is 
-qualitatively different than structured information such as that 
-contained in database entries.
-Text mining is the technology that automates the discovery of 
-information in non structured text.
-
-Text mining concerns the partitioning of text segments into classes,
-the clustering of text segments,
-the extraction of concepts and entities from text,
-and the discovery and modeling of relations between text entities 
-and classes.
-Text mining techniques perform NLP to analyze text and perform 
-IE to extract information into data structures.
-Fundamental to IE are the named entity (NE) and the named entity 
-relation (NER) extraction techniques which capture features from 
-sentences and phrases.
-A key challenge to NE and NER is that sentence structures and 
-words are often ambiguous in natural languages.
-%While research to address ambiguity in Latin languages is still 
-%lagging behind~\cite{Red08},
-%not much has been done to address NE and NER in the context of 
-%the Arabic language.
-
-The application of text mining techniques to Arabic text documents 
-will result in great benefits to many research fields such as 
-Arabic literature,
-Islamic studies, Hadith authentication, and Arabic history and 
-culture in general.
-Arabic text mining will also benefit sectors where Arabic text 
-documents are key such as the security sector,
-the government personal records,
-the taxation department,
-the health sector,
-as well as the trading floor of a stock exchange.
-
-%The difficulty of applying text mining techniques to Arabic text 
-%documents lays in the absence of automated tools that understand 
-%the unique features of the Arabic language.
-%We will make use of an example to illustrate the process.
-Given an Arabic text document we want a tool that can segment 
-and isolate all names, dates, time of the day occurrences,
-tool names, and locations as entities.
-Then we want another tool to identify relations amongst 
-these entities such as identities and orders.
-We want another tool to look at the verbs and actions in 
-the sentences and try to build relations between the identified 
-entities based on these verbs.
-If we were successful to do that,
-we now want to order all entities associated with dates based 
-on a chronological order.
-
-An order between two entities can depend on 
-their simple order of appearance in the document, their 
-numerical values in case they happened to be numbers, or
-the distance between them in case they were cities. 
-The relations can also be more sophisticated and 
-based on templates provided by users or by 
-models extracted from training data sets.
-
-To do the above, 
-we need tools that automatically differentiate between 
-names, verbs, adjectives and other structures.
-This is not a simple task with the Arabic language.
-For example, Arabic grammar allows name-based sentences 
-or sentences with no verbs \noTrNoVocRL{^gml ismiyT} .
-Another unique feature is the possibility to have names in Arabic 
-that express action such as noun verbs 
-\noTrNoVocRL{ism f`l, f-a`l $\ldots$}.
-
-In addition to these structural differences,
-there are features in the Arabic language that need special 
-treatment such as derivatives and stems 
-\noTrNoVocRL{alm^staq-at w al^g_dwr}.
-Simple primitive tasks such as comparing two strings need 
-special attention in Arabic since each Arabic letter has 
-four different forms depending on its position in the word; 
-isolated form, start, middle and end of a word.
-Arabic diacritics should also be ignored most of the time 
-when comparing two strings.
-We also need to build dictionary tools that associate words 
-and phrases based on their meanings and their context.
-Often time,
-and depending on the application and the user of the tool,
-we need to allow the user to assign meanings and semantics 
-for patterns in the text.
+It is clear that our first year of research on this topic has been fruitful.
+For next year, we expect to complete the hadith literature case study and to start and complete the security case study.
+We will package the components built during the development of these case studies as independent text mining entity and relational extractors to be included in the framework to be completed.
+Moreover, we expect to be able to enhance the morphological analyzer in terms of speed and accuracy.
+We will also investigate methods to extend the internal structures of our analyzer to be able to represent a formal linguistic
+computational model of the Arabic language. In what follows we describe our expectations in terms of each case-study.
 
@@
 {\includegraphics{figs/literature_case-study1.pdf}
+} 
+\caption{Diagram for the literature case study }
 \label{f:literature}
-} 
-\caption{CATARM diagram for the literature case study }
 }
@@
 high level vision for what the literature case study
-will look like in the CATARM framework. 
+will look like in the framework to be developed. 
 The boxes represent 
-computational blocks provided by CATARM, 
+computational blocks provided by the framework, 
 ovals represent data blocks which are mainly the 
@@
 
-The computational components can be direct CATARM components
-or custom user components built from other CATARM components.
+The computational components can be direct components of the framework
+or custom user components built from other components.
 We feed the hadith books to a user Narrator Chain Extractor
-component that we will discuss later in 
-Section~\ref{s:design:chainext}. We  obtain a set of chains
+component that we have built using finite state machines (described in the attached paper).
+We  obtain a set of chains
 of narrators that we feed to a morphological analyzer (MA) 
@@
 {\includegraphics{figs/security_case-study.pdf}
+}
+\caption{Diagram for the security case study}
 \label{f:security}
-}
-\caption{CATARM diagram for the security case study}
 }
@@
 security case study
-will look like in the CATARM framework. 
+will look like in the framework. 
 The inputs are two separate sets of documents. 
@@
 a database of previous interrogation reports. 
-The CATARM system also takes as input a database of geographical
+The system also takes as input a database of geographical
 locations and a database of personal records. 
-The aim is to check for inconsistencies in the iterrogation
+The aim is to check for inconsistencies in the interrogation
 reports and to identify possible connections between the suspects
@@
 
-\subsection{CATRAM components}
-\label{s:design:components}
+\section{Budget for year 2011}
+\label{s:budget}
 
-To do the above we need CATARM to provide us with several 
-computational components such as a morphological analyzer, 
-statistical tools, graph algorithms, and other computational
-components. 
-In the following we discuss the components needed for 
-CATARM and we propose a methodology to 
-use and extend existing tools, and develop new tools to
-build the components of CATARM.
+In the table below we summarize our itemized budget for next year:
 
-\subsection{Case-based morphological analyzer}
+\begin{table}[ht]
+\begin{tabular}{l r}
 
-Morphological analyzers for the Arabic language
-such as Buckwalter~\cite{Buckwalter:02},
-Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
-and ElixirFM~\cite{Otakar:07} take an Arabic word as input
-and produce several valid morpholgical solution for the word. 
-Each solution is a sequence of morphemes that form the word.
-Each morpheme is associated with several tags such as 
-morphological, part of speech, grammar, and semantic tags. 
-These analyzers are essential in several text analysis tools such 
-as spell checkers as well as NLP frameworks~\cite{Col09}.
-Other morphological analyzers such as 
-Amira~\cite{Diab:07,Benajiba:07},
-MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
-use machine learning and support vector machines (SVM) 
-to enhance the accuracy at the expense of efficiency.
+ \hline
+ One part-time research assistant for 8 month: LL. 750K/month  & LL. 6000K  \\
+ Arabic Corpus Lisence & LL. 1500K \\
+Miscellaneous  & LL. 500K \\
+\hline
+ Total  & LL. 8000K \\
+ \hline
+\end{tabular} 
+\end{table}
 
-The current use of morphological analyzers suffers from 
-several problems. 
-Analyzers that enumerate all possible partitions of a word
-and match them against concatenative rules are computationaly 
-expensive. 
-Analyzers that exhaustively enumerate all possible solutions 
-for a word may hurt performance and may
-not be necessary or appropriate in some case studies~\cite{Maamouri:10}. 
-For example, the first solution may be the one needed and 
-we may not need to list the rest of the solutions. 
-A white space delimited token may have more than one word
-as is the case with ``run-on words''.
+\section{Submission to: ``ACL/HTL 2011"}
+\label{s:paper}
 
-We propose to make use of the current existing morphological
-analyzers to build a flexible and more efficient
-morphological analyzer. 
-The analyzer will be an essential
-component in the CATARM framework. 
-It will be a case-based analyzer as it will make guided decisions
-based on feedback from the NLP case study and it will prune
-invalid solutions on the fly.
+We have submitted the attached paper to the “The 49th Annual Meeting of the Association for Computational Linguistics: Human Language Technologies (ACL/HLT 2011)”.
+The paper is entitled: “Case-based Arabic Morphological Analysis”.
 
-Case-based analysis can guide the morphological analysis
-to prune lots of the solutions and to bias the order of considered
-decisions. 
-For example, in the hadith literature case study pesented above, 
-amongst other things, we are intersted in checking whether a 
-word is a name.  
-We can bias the order of considered decisions to consider the 
-morphemes that can be names first. 
-If one of them matches, we can ignore the rest of the solutions. 
-In case a prefix is decided to be of a morphological category
-that does not connect with a name, then we can just ignore
-the analysis of the rest of the word. 
-This is more efficient than generating all possible partitionings
-of the word, checking for morphological correctness of each 
-partition, and then inspecting the part of speech tags of each
-correct solution. 
-We desire to build a morphological analyzer that allows a 
-case-based controller to intervene at every single decision
-without much overhead.
-
-\subsection{CATARM Query language}
-\label{s:design:query}
-
-We propose the CATARM query language with a user friendly
-visual representation similar to that in Figures~\ref{f:literature} and
-~\ref{f:security}. 
-The basic elements in the query language will be entities, relations
-and tags similar to those reported by a morphological analyzer. 
-The query language will also have connectives such as set operations, 
-relational operations, 
-scalar and Boolean arithmetic operations, 
-statistical computations, graph algorithms, and text algorithms
-form terms and predicates that specify the desired computation
-by the user. 
-We will specify the CATARM query language in a formal grammar.
-
-\subsection{Arabic linguistic computational model }
-\label{s:design:lcm}
-
-Several techniques~\cite{AEL07,Ham07,Abd07,MEl03} 
-preprocessed the Arabic text and then used known text 
-mining techniques that tend to Latin text and queries
-to solve classification, indexing, named entity 
-extraction and other problems.
-We think that these techniques do not make use 
-of Arabic language specific features.
-
-We think that we can discover such features by building
-a linguistic computational model of the Arabic language
-based on its formal grammar and formal morphology
-as defined in several Arabic linguistic 
-sources~\cite{Sha73,Abd00,Abd001}.
-We will build our morphological analyzer on top 
-of this linguistic model in addition to the current existing
-concatenative-based analysis techniques.
-
-Let $L_q$ be the query language grammar, 
-$L_A$ be the grammar that expresses the Arabic linguistic
-computational model, and let $q$ be a user query.
-We compute $\rho_q$, the grammar path that matches 
-the query $q$ in $L_q$. 
-We will explore methods that use $\rho_q$ to simplify 
-$L_A$ before we start the morphological analysis. 
-We think the possible simplifications that we can perform
-here are Language specific as they will reflect the 
-correspondence between the formal query language definition $L_q$, 
-the usage patterns of $L_q$  by Arabic users, 
-and $L_A$ the formal LCM for Arabic.
-
-\subsection{Methodology}
-
-\begin{figure}
-\center{
-\resizebox{.4\columnwidth}{!}
-{\input{figs/hadith.pdftex_t}
-\label{f:statemachine}
-}
-\caption{State-machine used to detect the chain of narrators}
-}
-\end{figure}
-
-{\bf Case-based morphological analyzer.}
-We experimented with several morphological analyzers.
-We think that using finite state machines for morphological analysis
-can be implemented in a more efficient way compared to other 
-approaches. 
-For example the concatenative rules used in Buckwalter~\cite{Tim04}
-can be encoded in three FSM with linear execution time using 
-the trie~\cite{datries} data structure.
-
-The use of FSMs also suits our case-based approach. 
-The controller can be a FSM machine on its own and its decision
-will be another input into the internal morphological FSM structures. 
-This also suits our choice of enriching the concatenative analysis
-with a formal linguistic model that can also be encoded into FSMs. 
-FSM can also be easily integrated with hidden Markov models (HMM)
-extracted from the sets of documents the MA will consider.
-
-The state machine in Firgure~\ref{f:statemachine} controls 
-the concatenation based state machines of the morphological
-analyzer.
-The transitions are excited
-by inputs from the morphological analyzer. 
-A NAME label represents a proper name, 
-an NRC label represents a narrator connector such as
-\RL{`an} "on behalf of", \RL{qaal} "said", 
-or one of their derivations. 
-The IBN label corresponds to the word \RL{ibn} 
-``the son of'' and represents a name connector.
-
-The symbol $\tau_{\mbox{NMC}}$ is a threshold
-that corresponds to the number of tolerated name connectors 
-that may occur between two names. 
-The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
-of name connectors collected since the controller
-started looping in the state NMC\_S. 
-The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
-that corresponds to a relaxed tolerance measure that
-the controller resorts to in case the words separating
-two names were longer than $\tau_{\mbox{NMC}}$ but 
-contained a name connector word such as IBN or NISBA.
-
-The controller has four states that correspond to 
-an abstract position in text relative to the next sanad. 
-State TEXT\_S is the initial state and denotes that
-the controller is outside the context of a sanad.
-The controller moves to the state NAME\_S whenever
-NAME is reported.
-State NRC\_S means that the controller thinks it is in the context
-of a narrator name after an NRC is reported.
-State NMC\_S
-indicates that the controller expects a name to appear within 
-a tolerance threshold expressed by 
-$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
-
-Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
-before it gives up on its expectations.
-
-We reach the NAME\_S state only when a
-valid NAME is detected and we leave when no more NAME's 
-are detected. 
-The NRC\_S state can only be reached if an NRC is detected.
-NMC\_S can only be reached from a NAME\_S state.
-
-The controller reports a valid chain of narrators when a 
-sequence of names
-connected by narrator connectors appears. 
-It marks the beginning of the hadith with the beginning of the 
-current sequence,
-and marks the end of the hadith with the beginning of the next 
-sequence. 
-It also marks the chain of narrators as the sequence itself. 
-The definition of input labels such as NAME and IBN depends on 
-the morphological analyzer. 
-We readily implemented the above approach and confirmed its 
-accuracy and efficiency compared to existing morphological
-analyzers.
-
-We will investigate enhancing our current case-based morphological
-analyzer.
-We will investigate methods to extend the internal structures 
-of our analyzer to be able to represent a formal linguistic
-computational model of the Arabic language.
-
-{\bf Linguistic computational model of Arabic.}
-We think that a linguistic computational model for the Arabic
-language will perform better in applications where
-relations between Arabic text documents are involved. 
-Such a model helps capture features unique and specific 
-to the Arabic language and make use of that in the analysis.
-
-ElixirFM~\cite{Otakar:07} builds on the lexicon
-of the Buckwalter analyzer and the annotations of the 
-Prague Arabic dependency databank to build a formal 
-computational model of the Arabic language.
-The models is a set of formulas and rules. 
-ElixirFM uses the computational model in a functional
-morphology framework that uses Haskell and functional
-programming. 
-The analyzer works by generating morphological trees of 
-decisions and then uses a disambiguation decision procedure 
-to annotate the trees with contextual elements.
-
-We will explore extracting our model from formal text book
-definitions of the Arabic language~\cite{Abd00,Abd001} and seek
-help from Arabic linguists to relax these formal rules to
-allow common mistakes. 
-We will encode the model in a grammar that we can synthesize
-into an FSM and incorporate that with our current FSM 
-morphological analyzer.
-
-{\bf User query language. }
-In Figure~\ref{f:statemachine} we showed an FSM 
-that controls the morphological analysis and generates a
-list of chains of narrators out of hadith text.
-We think that it is overwhelming to ask a user who is not 
-versed in FSM and in computational sciences in general 
-to design such a controller.
-
-The diagram in Figure~\ref{f:chain} depicts our vision of 
-an equivalent query.
-This query generates the list of chains by first passing the 
-hadith text into the morphological analyzer which 
-extracts names and their position in text. 
-We iterate over the extracted names and compute the distance.
-Whenever the distance is less than a threshold $\theta_1$
-we add the word to a temporary chain of names. 
-Otherwise, we execute a sequential block to 
-{\bf 1.} tabulate the chain into a temporary table containing 
-the list of chains. 
-{\bf 2.} Create a new chain whith the last element.
-Finally, we filter those chains whose cardinality 
-is less than some threshold  $\theta_2$.
-
-This query can be saved to a library of computational components
-and invoked directly in another query 
-as we saw in Figure~\ref{f:literature}.
-
-\begin{figure}
-\center{
-\resizebox{.8\columnwidth}{!}
-{\includegraphics{figs/detailed_submachine.pdf}
-\label{f:chain}
-}
-\caption{CATARM detailed diagram for the chain of narrators extractor}
-}
-\end{figure}
-
-The CATARM query language will be expressed in a grammar 
-where the atomic elements are user defined constants, or
-entities and tags extracted by the morphological analyzer. 
-The language will define terms, expressions, and predicates
-using several connectives that express set operations, 
-relational operations, Boolean operations, scalar operations, 
-statistical tools, text algorithms, and graph algorithms. 
-We need to provide a front end tool to support this query
-language.
-
-{\bf Simplify LCM based on the query.}
-Given
-L$L_q$ the query language, 
-$L_A$ the Arabic linguistic
-computational model, and  $q$ a user query
-we compute $\rho_q$, the grammar path that matches 
-the query $q$ in $L_q$. 
-We plan to inspect methods to overapproximate 
-the morphemes and the transitions in $L_A$ 
-that can be satisfied by morphemes and transitions
-in $\rho_q$. 
-This can simplify the number of states we need to 
-look at under $L_A$.
-
-This task is not trivial though, since $\rho_q$ may contain
-states and transitions that are not always morphemes and 
-transition between morphemes. 
-We will consider a hierarchical methodology that follows 
-the hierarchical structure of the Arabic language expressed
-in $L_A$. For example,
-we may consider sets of connected states in $\rho_q$ 
-containing a single morpheme as a state in one hierarchy level 
-in $L_A$.
-We will perform this simplification before starting the 
-morphological analysis since we already have to analyze
-the query $q$ and compute $\rho_q$ before we start
-the analysis.
-
-{\bf Simplify LCM and MA based on set of documents.}
-We will analyze the set of input documents given to a
-morphological analyzer component and use the analysis
-to further improve the morphological analysis. 
-We will extract statistical features from these documents 
-to help order the solutions considered by the analyzer
-to consider the most probable first.
-
-We will consider the use of HMM models 
-built on the fly while processing the documents.
-We will also consider the use of
-latent semantic indexing~\cite{LSI89} techniques. 
-We are considering these two techniques as they are 
-both language independent and can both be computed and
-adapted on the fly.
-
-{\bf Entity and relational extractor.}
-We will build a set of entity and relational extractors. 
-The entities can be proper names, dates, locations,
-and qualification adjectives. 
-We will explore the use of pairing techniques to establish 
-relations between the extracted entities.
-The relations will form graphs as in the case studies presented
-so far. The nodes represent the entities and the previously 
-extracted relations and the edges represent the newly discovered 
-relations.
-
-The entity and relational extraction components 
-may take as input a template for an expected relation and 
-may return a graph that describes whether and where 
-this relation exists between the subject documents.
-
-We will need to augment the tags in the lexicons of the 
-morphological analyzer to allow for better entity and relational
-extractors. 
-We plan to make use of existing annotated tree banks and of
-modern Arabic dictionaries to do that.
-
-{\bf Statistical components.}
-We will make use and build statistical components that can be used
-in CATARM. 
-We will need statistical components to rank solutions, compute 
-probable decisions, estimating good thresholds.
-
-{\bf Graph algorithms.}
-The relations extracted from the text documents will often 
-be expressed in terms of graphs.
-We will need textbook graph algorithms such as traversals, 
-topological sorts, spanning trees, strongly connected 
-components, minimum flow  and other algorithms.
-For example, in the literature case study, the discovery
-of articulation vertices may point to important narrators 
-who may have an insignificant number of narrations 
-but still connect narrator communities.
-
-{\bf Query solver.}
-This is a logical solver that will solve query expressions to \
-resolve matching text and relations.
-We will need to use existing off the shelf constraint solvers 
-and set theory satisfiability solvers to address our queries. 
-The solver should return and rank approximate matches.
-That is if a three entities out of four required in a relation are
-found, the solution is still worth reporting.
-
-The solver will call the morphological analyzer to compute 
-morphological matches to the atomic elements of the query. 
-Then each operation in the query such as a set intersection
-operation will correspond to a computational component. 
-The solver will call these components and integrate their
-results.
-
-The solver may make use of existing propositional~\cite{MiniSAT} or
-higher order logic~\cite{Z3} solvers if needed. 
-These solvers can solve satisfiability questions.
-We will explore the feasibility and utility 
-of extracting a necessary set of 
-constraints from the query and answering that with an off-the-shelf
-logic solver.
-
-{\bf Case studies.}
-We will work on developing the necessary components to complete 
-the solution for the hadith literature and the security case 
-studies.
-We will also work on the health and network traffic case studies.
-While working on the case studies, we will add the components
-we build to our library of components. 
-This will enrich CATARM with a useful set of computational 
-components.
-
-{\bf Development methodology.}
-We will integrate all the mentioned techniques in a scalable 
-framework for Arabic text analysis.
-The framework as a whole will allow for exhausting Arabic text 
-analysis and allow for extensions of the framework.
-A user should be able to iteratively and repetitively apply the 
-abode techniques on his input text.
-
-We will follow an eXtreme Programming 
-methodology in the process of building CATARM.
-We will use open source tools, libraries, and frameworks to 
-build our libraries.
-We will build on existing open source programming and text 
-mining tools as well as open source Arabic text analysis tools 
-based on NLP such as MADA+TOKAN~\cite{Rot08}, 
-NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
-
-We will use a dokuwiki based website~\cite{Dok09} 
-to document the progress online and will use 
-subversion~\cite{Sub09} as a source control tool.
-
-{\bf Visualization.} 
-We will build a visualization tool that allows fast 
-and guided inspection and correction mechanism. 
-The CATARM visualization aid will be a tag and relation color
-sensitive browser.
-The interface will allow a fast and guided inspection and 
-correction mechanism.
-
-\subsection{Research team}
-
-The team will be formed by the two main PIs, Dr. Fadi Zaraket (LPI)
-and Dr. Rehab Duwairi (Co-LPI) and three research assistants.
-Currently the LPI is working with one research assistant 
-and two senior students on an Arabic morphological analyzer.
-
-The LPI offers a course on computational Arabic attended
-by excellent senior and graduate students at AUB. 
-Once funded, the team can be directly formed and work can start
-with productive team members.
-
-\subsection{Work plan}
-
-The Lead PI will conduct biweekly meetings to track progress 
-on the project.
-The team will maintain a dokuwiki based website to continuously 
-post updates on the progress of the project.
-
-\subsection{Budget items}
-
-In the following we list the budget items. 
-Monatary values are provided in more details on the QNRF 
-submission website.
-
-\begin{itemize}
-\item One research assistant at QU
-\item Two research assistants at AUB, one of them will 
-work partially from QU. 
-\item Two month salary per year for the Co-LPI
-\item Two month salary per year for the LPI
-\item One course buy out per year for the Co-LPI
-\item Two weeks consultant for the second and the third years
-\item Licensing and membership fees to acquire digital libraries and corpora
-\item High performance computer and storage devices
-\item One conference for each PI per year
-\item One conference for one RA from QU per year
-\item One conference for one RA from AUB per year
-\item Site visit to AUB 
-\item Site visit to QU 
-\item Other miscellanous items
-\end{itemize}
-
-\section{Anticipated results and evaluation criteria}
-\label{s:results}
-
-We anticipate that the research will lead to the production of 
-CATARM, an Arabic text analysis framework that allows relational 
-queries between Arabic text documents.
-The components of the framework are listed and discussed
-in details in Section~\ref{s:design}. 
-In the following we will discuss what to anticipate and 
-how to evaluate the different components of CATARM.
-
-\subsection{Case-based morphological analyzer}
-
-We will deliver a case-based morphological analyzer. 
-The analyzer will be compared to other available analyzers such as
-Buckwalter~\cite{Tim04} and ElixirFM~\cite{Otakar:07} in terms
-of accuracy and efficiency of analysis.
-We will adopt two comparison methodologies. 
-First, we will compare the analyzers in an exhaustive analysis mode
-which is the only way current analyzers perform. 
-Second, we will compare the analyzers under several case-studies
-and show how a case-based morphological analyzer can benefit
-from case-based optimizations.
-
-\subsection{Linguistic computational model for Arabic}
-
-We will build a linguistic computational model for Arabic 
-based on an understanding of the formal Arabic grammar 
-and on an abstraction of the current existing concatenation
-based lexicons.
-We will compare our LCM with that of ElixirGM~\cite{Otakar:07}.
-We will study how much such a model can affect the analysis
-in terms of accuracy and efficiency as well.
-
-\subsection{User query language}
-
-We will build a query language for relational Arabic text mining 
-and a front-end to compile a user query into an optimized
-computation. 
-We will check how user friendly and how expressive the language is.
-We will compare our query language with the sophisticated 
-interface of the MADA+TOKAN toolkit~\cite{Habash:09}.
-We will look for other attempts to formalize Arabic text 
-mining queries and we will also compare agains 
-formal text mining queries from other text mining domains such
-as the gene ontolgy domain~\cite{GeneOntology10},
-
-\subsection{Optimizations}
-
-We will develop and evaluate our two optimization methodologies
-and look for other optimizations. 
-First we will use language specific features based on the 
-LCM of Arabic, the grammar of the query language, and 
-the Arabic query provided by the user to simplify 
-the LCM and the morphological analysis. 
-Second we will look at language independent techniques
-such as HMM and LSI models extracted from the 
-text documents and use that to simplify the analysis. 
-We will perform the analysis with and without the optimizations
-and compare the results in terms of accuracy and efficiency.
-
-%\subsection{ Entity and relational extractors}
-\subsection{CATARM and case studies }
-
-We will evaluate CATARM against at least three of the case studies
-that we discussed previously and that are listed below. 
-\begin{enumerate}
-\item Hadith literature
-\item Security
-\item Health
-\item Network content
-\end{enumerate}
-
-We anticipate CATARM to be able to successfully solve all the case
-studies with resonable computational resources available to the 
-average user. 
-We will also generalize the components we will develop to solve
-the case studies into a library of components that can be used
-in other case studies.
-
-\subsection{CATRAM evaluation }
-
-The team will seek the assistance and help of an independent 
-consultant in the middle of the second and third years of 
-the duration of the project to evaluate and validate the results.
-
-\section{Strategy for project continuation}
-\label{s:continue}
-
-Once completed the Arabic text analysis framework will be 
-an open source contribution and developers from the Arabic 
-world in the open source community are expected to join 
-efforts to keep it going.
-Furthermore, once completed the Arabic text analysis framework 
-can sustain itself by charging on support or expert use.
-
-The framework or service on the framework can easily be 
-commercialized to tend to the security, health, insurance, market,
-news, media, or literature business sectors.
-
-The area of Arabic text analysis will be a research area for 
-a long time as NLP is one of the most complex computational 
-problems.
-The Arabic text analysis framework will be used by NLP 
-researchers in the Arabic language field and those are 
-expected to maintain and advance the framework.
-
-\subsection{Future work}
-
-Enhance morphological analysis
-Look for other optimizations
-Generalize the framework to other Arabic-like languages
-Generalize the case-based approach 
-Integrate other text mining techniques such as vocalization,
-suggestions...
-
-\section{Plans for disseminating research results}
-\label{s:dissem}
-
-The team will publish the results of the research in the form 
-of refereed conference and journal papers as well as tool papers 
-addressing each of the tools. 
-The team will also hold workshops to educate possible users on the 
-tools once the tools are available.
-The Arabic text analysis framework will be partially based on 
-existing work in the open source domain and will thus be available 
-as an open source tool. 
-Interested researchers and users will be able to download the tools 
-for free and use them for non-profit non-commercial purposes.
-
-\pagebreak
 %\bibliography{adnan_refs}

### 2010-12-12 trunk/tex/sarf.tex
added the progress report for LNCSR

@@
-%
 % File naaclhlt2010.tex
-%
 % Contact: nasmith@cs.cmu.edu
@@
 \newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
+\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
+\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
+\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}
 
@@
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Case based Arabic Morphological Analysis with 
-    Compositional Non-deterministic Automata }
+\title{Case-based Arabic Morphological Analysis }
 
-%\author{ Jad Makhlouta \\
-%  \And
-%Hamza Harkous \\
-%  American University of Beirut \\
-%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
-%  \And 
-%  Fadi Zaraket 
-%}
+\author{ Jad Makhlouta \\
+  \And
+Hamza Harkous \\
+  American University of Beirut \\
+  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
+  \And 
+  Fadi Zaraket 
+}
 
@@
 
+\transfalse
+\begin{figure}[tb]
+\center{
+\resizebox{1.2\columnwidth}{!}
+{ \input{psfigs/exhadith.pstex_t}}
+\caption{Hadith abstraction example.}
+\label{f:exhadith}
+}
+\end{figure}
+% the FSMs for the three words
+\transtrue
+
 We validate our hypothesis using a case study from the Islamic 
 literature. 
-A {\em hadith} is a narration related to the prophet Mohammad
-through a {\em sanad} or a sequence of narrators. 
-The authentication of a hadith highly depends on the credibility
+A \RL{.hady_t} is a narration related to the prophet Mohammad
+through a \RL{sanad} or a sequence of narrators. 
+Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
+transliteration and translation. 
+The authentication of a \noArRL{.hady_t} highly depends on the credibility
 of each of the narrators as reported in separate biography 
@@
 % the FSMs for the three words
+\transtrue
 
-The compositional FSA diagram in Figure~\ref{f:example}
+The diagram in Figure~\ref{f:example}
 illustrates the operation of Sarf when parsing the
@@
  % public IP.}.
+
+%\setcode{utf8}
+%\begin{arabtext}
+%حدثنا \emphasize{هشيم}، أخبرنا \emphasize{إسماعيل بن أبي خالد}،
+% عن أبي إسحاق، عن سعيد بن جبير، قال كنا
+%    مع ابن عمر رضي الله عنه حيث أفاض من
+%    عرفات إلى جمع فصلى بنا المغرب ومضى
+%\end{arabtext}
+%\setcode{standard}
 
@@
 \subsection{Controller}
+\label{sec:controller}
 % FSM for the controller
 
-We illustrate the hadith case-based controller FSA 
+We illustrate the hadith case-based controller 
 in Figure~\ref{f:hadith}. 
-The transitions in the hadith FSA are excited
+The transitions in the hadith controller are excited
 by inputs from the Sarf morphological analyzer. 
@@
 
-\section{Future Work}
+\subsection{Comparison with Buckwalter and ElixirFM}
+
+We implemented our case study with Buckwalter in two phases. 
+In phase one, we process the entire text with Buckwalter and
+obtain annotations to locations in text. 
+Then in the second phase we feed these annotations to 
+our controller as if they were coming from Sarf. 
+We measure the time it took to produce the annotations
+and the time it took the controller to process the 
+annotations whithout the time spent streaming the annotations.
+We used the same setup for ElixirFM.
+
+Sarf outperforms both Buckwalter and ElixirFM in runing time. 
+It also produces higher accuracy measures. 
+We think this is mainly due to the extended
+lexicon as well as to the fact that Sarf reduces false positive
+names by ignoring prefixes and suffixes that can not form
+names even if the stem was a name. 
+Doing the same with Buckwalter and ElixirFM requires  
+expanding the annotations resulting from one affix to reflect
+its multiple composing affixes.
+This is easily done with the way we defined affixes recursively.
+
+Buckwalter performed much better than ElixirFM. This is up to 
+our knowledge the first comparison of these two tools and 
+we think the big difference is due to the more exact 
+and grammar based functional morphological analysis ElixirFM 
+uses.
+
+\section{Conclusion and Future Work}
 \label{sec:future}
@@
 to interact with Sarf.
+
+\section{Acknoledgements}
+\label{sec:acc}
+We thank the Lebanese National Council for Scientific Research (LNCSR) for funding this research.
 

### 2010-12-16 trunk/tex/rp.tex
refinements+sort bib

@@
 
-Most of academic research in the field of Arabic text mining 
-is covered in the almost comprehensive book 
-``Arabic Computational Morphology''~\cite{Sou07}.
-The book summarizes recent work, and presents strong evidence of 
+The almost comprehensive book 
+``Arabic Computational Morphology''~\cite{Sou07}
+covers most academic research work on Arabic text mining.
+It summarizes recent work, and presents strong evidence of 
 shortage of historic research. It also highlights  an emerging 
-interest in this area with a specific application to machine 
+interest with a specific application to machine 
 translation from Arabic to English after 2001.
@@
 the Arabic user.
-While a non-Arabic user is interested in machine translation to 
-help him understand an Arabic document, 
+While machine translation is of gneral interest, 
 an Arabic user is more interested in 
-information hard to discover manually such as 
-the relation between clinical records, pharmaceutical 
-prescriptions and sales, and a specific sickness. 
-Another Arabic user may be interested in looking at the relation
-between stock prices and political news or the 
-relation between suspect profiles in interrogation reports 
+information hard to discover manually.
+Examples are {the relation between clinical records, pharmaceutical 
+prescriptions and sales, and a specific sickness,
+the relation
+between stock prices and political news, or the 
+relation between interrogation reports with suspects
 and investigative reports from the crime scenes.
 
-We propose to build a 
+We propose to build a framework for
 %{\em relational text mining 
 %of Arabic using morphological and case-based analysis}. 
-{\em case-based Arabic text analysis using relations and morphology}(CATARM) framework. 
-CATARM will take as input sets of documents, each of a specific
-type, and provides the user with a visual query language that
-allows him to build a flow of Arabic text mining
-modules and discover desired relations between documents. 
-CATARM will translate the query into a netlist of computational 
-components, it will also extract parameters from the query and 
-the document sets and use them to simplify the computational 
+{\em case-based Arabic text analysis using relations and morphology}(CATARM). 
+CATARM takes as input several sets of documents, each of a specific
+type, and a user query. % expressed in a visual language.
+The query, expressed in a visual language, allows the user 
+to build a flow of Arabic text mining
+components and discover desired relations between documents. 
+CATARM compiles the query into a netlist of computational 
+components. It also uses features extracted from the query and 
+the document sets to simplify the computational 
 components. 
@@
 %study.
+
+\novocalize
+%The accuracy of the solutions suffer due to inherent difficulties
+%of morphological analysis of the Arabic language. 
+It is common practice to write Arabic text
+without short vowels. 
+This greatly increases the ambiguity of Arabic text. 
+Arabic letters can have up to 
+four different forms
+corresponding to their position in a word, i.e, beginning,
+middle, end of word and separate forms. 
+This allows the phrase
+\noTrRL{il_A\nospace almdrsT}  with no delimiter within
+to be visually recognizable
+as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
+This happens because the first word \RL{il_A} ends with
+\RL{_A} a non-connecting letter. 
+These words occur often in text and greatly increase the
+difficulty of tokenization and are referred to as 
+``run-on'' words~\cite{Buckwalter:04}.
+Such difficulties are inherent to the 
+morphological analysis of the Arabic language and
+render the solutions inaccurate.
 
@@
 
-%The accuracy of the solutions suffer due to inherent difficulties
-%of morphological analysis of the Arabic language. 
-It is common practice to write Arabic text
-without short vowels. 
-This greatly increases the ambiguity of Arabic text. 
-Arabic letters can have up to 
-four different forms
-corresponding to their position in a word, i.e, beginning,
-middle, end of word and separate forms. 
-This allows the phrase
-\noTrRL{il_A\nospace almdrsT}  with no delimiter within
-to be visually recognizable
-as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
-This happens because the first word \RL{almdrsT} ends with
-\RL{T} a non-connecting letter. 
-These words occur often in text and greatly increase the
-difficulty of tokenization and are referred to as 
-``run-on'' words~\cite{Buckwalter:04}.
-Such difficulties are inherent to the 
-morphological analysis of the Arabic language and
-render the solutions inaccurate.
-
 Current morphological analyzers such as 
 Buckwalter~\cite{Buckwalter:02},
-Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
-and ElixirFM~\cite{Otakar:07} exist.
-These analyzers are used in several open source spell checkers as 
+Beesley~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+and ElixirFM~\cite{Otakar:07} exist
+and are used in several open source spell checkers as 
 well as NLP frameworks~\cite{Col09}.
@@
 {\bf Hypothesis 1.} We hypothesize that many case studies, 
-in particular relational
-case studies, do not require high accuracy from the 
+in particular relational ones, 
+do not require high accuracy from the 
 low level morphological analysis.
-The sophistication of the sought higher level query can eliminate 
-many of the ambiguous low level solutions at a lower computational 
+The sophistication of the higher level query can eliminate 
+many ambiguous low level solutions at a lower computational 
 cost.
@@
 prefix in question in the analysis connects only to a verb, 
-we do not need to further analyze the rest of the word 
-to provide an answer.
-
-We find evidence to our hypothesis in~\cite{Maamouri:10}. The 
-addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
-required the high NLP task to interact with the low 
-level morphological analyzer.
+we no further need to analyze the rest of the word.
+
+We find evidence to our hypothesis in~\cite{Maamouri:10} where
+a new corpus was added to the Arabic Tree Bank~\cite{Maamouri:04}.
+The addition requires compliance with 
+the DARPA GALE program including several NLP tasks such as 
+English translation annotations and word-level alignment
+of Arabic and English. 
+The new genre of data challenged the existing data processing 
+and annotation guidelines. 
+Alternative guidelines came to place where the 
+high level NLP task interacted with the 
+morphological analyzer.% along with other concurrent tasks. 
 This led to a refined analyzer~\cite{Kulick:10}.  
@@
 morphological analyses behave differently on the same case study. 
-We strongly believe and we find evidence in our preliminary
+For example, the analyzers may find different solutions, and may 
+present the same solutions in different orders.
+We find strong evidence in our preliminary
 results that a 
-case-based morphological analyzer that allows a high level 
+case-based morphological analyzer that allows a 
 case-based controller to intervene at every decision to 
-guide, use, prune, and refine the morphological analysis
+guide, use, and refine the morphological analysis
 is of high utility.
@@
 Researchers in~\cite{AEL07,Ham07,Abd07,MEl03} 
-preprocessed the Arabic text to make it presentable to known text 
-mining techniques that tend to Latin text and queries.
+preprocessed the Arabic text and presented it to known 
+techniques that tend to Latin text and queries.
 %Buckwalter~\cite{Bee89,Tim04} uses morphological analysis to enable 
@@
 the Arabic language as a similarity measurement.
-This work is limited to the names of the fields in the data 
-structures extracted from Arabic texts.
+%This work is limited to the names of the fields in the data 
+%structures extracted from Arabic texts.
 
@@
 %\subsection{\bf Arabic text mining in industry}
-Few companies in the industry provide primitive Arabic text 
+Few companies in the industry provide Arabic text 
 analysis tools.
 Sakhr~\cite{Sak09} provides a categorization 
-tool with a limited number of predefined categories, and a summary 
-generator that selects sentences from text.
+tool with a limited number of predefined categories.
+Sakhr also provides a summary 
+generator that selects important sentences from text.
 Basis technology~\cite{Bas09} provides Rosette, a lexical 
 analysis tool that normalizes Arabic text as a pre-processing step 
-for indexers and other general purpose data mining 
+for general purpose data mining 
 tools.  Basis also provides REX; an entity extractor.
@@
 
-{\bf Goals.} Up to our knowledge, CATARM will be the 
-first framework that tends to relational Arabic morphology 
-and text mining. 
-{\bf (1.)} CATARM will build a novel parametrized linguistic 
-computational model of Arabic. 
-The model will be implemented using efficient structures. 
- {\bf (2.)} It will be amenable to Arabic specific simplifications 
+{\bf Goals.} CATARM 
+tends to relational Arabic text mining using morphological analysis.
+{\bf (1.)} CATARM will build a novel parametrized 
+linguistic computational model (LCM) of Arabic. 
+The LCM will be implemented using efficient structures. 
+ {\bf (2.)} The LCM will be amenable to Arabic specific simplifications 
 inferred from the correspondence between the formal Arabic rules
 and the relational user queries.
-{\bf (3.)} The model will also be amenable to simplifications 
+{\bf (3.)} The LCM will be amenable to simplifications 
 inferred from statistical features extracted from the document 
-sets using language independent techniques such as latent semantic 
-indexing~\cite{LSI89} and hidden Markov models (HMM).
+sets using language independent techniques such as 
+hidden Markov models (HMM)  and 
+latent semantic indexing~\cite{LSI89}.
 %CATARM will augment current digital dictionaries to resolve 
 %entity semantics and targeted qualificationsl.
-{\bf (4.)} CATARM extracts semantic graphs from each text 
-document and then uses set arithmetic and graph analysis techniques
-to establish relations between these graphs. {\bf (5.)} CATARM will 
-solve at least three interesting relational Arabic text case studies 
-presented in Sections~\ref{s:design:sec} and~\ref{s:design:lit}.
+{\bf (4.)} CATARM extracts semantic graphs from text 
+documents and uses set arithmetic and graph analysis techniques
+to establish relations between the graphs. {\bf (5.)} CATARM will 
+solve at least three interesting case studies such as those
+presented in Sections~\ref{s:design:lit} and~\ref{s:design:sec}.
 
@@
 
+The Co-Lead PI published extensively in the 
+domain~\cite{Duw06, Duw07, Duw09}.
 In recent research efforts, the CO-LPI is investigating 
 keyphrase extraction fro Arabic documents.
-
-The Co-Lead PI published extensively in this domain. 
-A selected list of her recent publications follows. 
-\begin{itemize}\itemsep0pt
-\item Rehab Duwairi, Mohammad Al-Refai, Natheer Khasawneh, 
-"Feature Reduction Techniques for Arabic Text Categorization". 
-Journal of the American Society for Information Science and Technology 
-(JASIST), Volume 60, Issue 11, pages: 2347-2352, 2009. 
-\item Rehab M. Duwairi, "Arabic Text Categorization", 
-International Arab Journal for Information Technology (IAJIT), 
-Volume 4, Number 2, pages 125  132, 2007.
-\item Rehab M. Duwairi, "Machine Learning for Arabic Text Categorization". 
-Journal of the American Society for Information Science and Technology 
-(JASIST), Vol. 57, Issue 8, pages 1005-1010, 2006.
-\end{itemize}
 
@@
 project that aims at comprehensive automation of authenticity 
-checks against a corpus of literature heritage~\cite{Zaraket06}.
+checks against a corpus of literature heritage~\cite{Zar06}.
 The Lead PI is leading an effort at AUB to introduce 
@@
 The lead PI worked and published about relational logic and 
-logic solvers applied to structured languages as follows.
-\begin{itemize}
-\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
-\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
-\end{itemize}
+logic solvers applied to structured languages~\cite{seraICSE07,Zara05}
 

### 2010-12-16 trunk/tex/pr.tex
final

@@
 
-This report describes the progress of the work carried out in the LNCSR funded project entitled: {\bf ``Relational Queries for Arabic Text Mining"}. 
+In this report we discuss our progress and work on the 
+{\bf Relational Queries for Arabic Text Mining} project 
+funded by LNCSR.
 \\
@@
 \item Budget for year 2011
-\item Submission to: ``ACL/HTL 2011"
+\item Attached submission to ``ACL/HTL 2011"
 \end{enumerate}
@@
 started work on an Arabic text mining framework~\cite{ATMine09}
-that is available now online. The source for this framework comprises code of the order of tens of thousands lines.
+that is available now online. 
+ATMine is around 20K lines of C++, perl, and sql code and uses 
+middleware from the open source community such as MySQL, Qt, and 
+double array trie.
+
+Arabic morphological analyzers~\cite{Sughaiyer:04}
+consider an Arabic word and its internal structure composed of 
+several {\em morphemes}. 
+A morpheme is a {\em stem} or an {\em affix}.
+An affix is a {\em prefix, suffix,} or an {\em infix}.
+The analysis of one word may lead to several possible
+solutions.
+\vocalize
+For instance, the word 
+\RL{'a.hmadH}~\footnote{In this document, we use the default 
+ArabTeX transliteration style ZDMG.}
+may have two valid morphological analyses. 
+The word means ``I praise him'' when
+the letter \RL{'a} is a prefix and  it means
+``his Ahmad'' when 
+the letter \RL{'a} is part of the proper noun stem 
+\RL{'a.hmad}.
+
+Current morphological analyzers such as 
+Buckwalter~\cite{Buckwalter:02},
+Beesely~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+and ElixirFM~\cite{Otakar:07} exist.
+These analyzers are used in several open source spell checkers as 
+well as NLP frameworks~\cite{Col09}.
+They take as input white space delimited tokens~\cite{Kulick:10},
+consider them as words,
+and enumerate all possible solutions. 
+This approach has several problems. 
+First, a white space delimited token may have 
+more than one word.
+Second, the exhaustive enumeration may hurt performance and may
+not be necessary or appropriate
+in some case studies~\cite{Maamouri:10}. 
+Other morphological analyzers such as 
+Amira~\cite{Diab:07,Benajiba:07},
+MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
+use machine learning and support vector machines (SVM) 
+to enhance the accuracy at the expense of efficiency.
+
 We have successfully developed ``SARF", a case-based morphological 
-analyzer, which constitutes capabilities of both Lexical and Syntax Analysis.
-We have been able to verify that both in terms of speed and accuracy our analyzer outperformed 2 state-of-art analyzers: Buckwalter ~\cite{Buckwalter:02} and ElixirFM~\cite{Otakar:07}.
-Supporting statistics are present in the attached paper described in Section~\ref{s:paper}.
-In addition, we have applied SARF on a literature hadith case study.
+analyzer with lexical and syntax analysis capabilities. 
+SARF outperforms in terms of efficiency and accuracy 
+the Buckwalter~\cite{Tim04} and the ElixirFM~\cite{Otakar:07} 
+analyzers. 
+We report on the detailed results in the attached manuscript.
+We discuss the attached manuscript in Section~\ref{s:paper}. 
+In addition, we have applied SARF on a hadith literature case study.
 
@@
 through a \RL{sanad} or a sequence of narrators. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
-transliteration and translation. 
+Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with 
+its transliteration and translation. 
 The authentication of a \noArRL{.hady_t} highly depends on the credibility
@@
 
-It is clear that our first year of research on this topic has been fruitful.
-For next year, we expect to complete the hadith literature case study and to start and complete the security case study.
-We will package the components built during the development of these case studies as independent text mining entity and relational extractors to be included in the framework to be completed.
-Moreover, we expect to be able to enhance the morphological analyzer in terms of speed and accuracy.
-We will also investigate methods to extend the internal structures of our analyzer to be able to represent a formal linguistic
-computational model of the Arabic language. In what follows we describe our expectations in terms of each case-study.
+Our first year of research on 
+Arabic text mining proved to be 
+fruitful. 
+For the next year, we expect to complete the hadith literature case 
+study and to start and complete the security case study. 
+We will package the components built during the development of these 
+case studies as independent text mining entity and relational 
+extractors to be included in the framework.
+Moreover, we expect to be able to enhance the morphological analyzer in 
+terms of speed, accuracy, and semantic capabilities. 
+We will also investigate methods to extend the internal structures 
+of our analyzer to be able to represent a formal linguistic
+computational model of the Arabic language. 
+In what follows we describe our expectations in terms of each case-study.
 
@@
 a database of previous interrogation reports. 
+We will first try our system on dummy data with seeded relations.
+Once the system is ready, we will deliver it to local police
+and intelligence agencies to try it out on real digitized 
+reports.
+
 The system also takes as input a database of geographical
@@
 
-In the table below we summarize our itemized budget for next year:
+In the table below we summarize our itemized budget for year 2011.
 
@@
  One part-time research assistant for 8 month: LL. 750K/month  & LL. 6000K  \\
- Arabic Corpus Lisence & LL. 1500K \\
+ Arabic Corpus License & LL. 1500K \\
 Miscellaneous  & LL. 500K \\
@@
 
-We have submitted the attached paper to the “The 49th Annual Meeting of the Association for Computational Linguistics: Human Language Technologies (ACL/HLT 2011)”.
+We have submitted the attached paper to the “The 49h Annual Meeting of the Association for Computational Linguistics: Human Language Technologies (ACL/HLT 2011)”.
 The paper is entitled: “Case-based Arabic Morphological Analysis”.

### 2010-12-16 tex/sarf.aclhlt.tex
acl hlt

--- 2010-12-12 trunk/tex/sarf.tex
@@
 % Contact: nasmith@cs.cmu.edu
+\documentclass[11pt]{article}
+\usepackage{acl-hlt2011}
+\usepackage{times}
+\usepackage{latexsym}
+\usepackage{amsmath}
+\usepackage{multirow}
+\usepackage{url}
+\DeclareMathOperator*{\argmax}{arg\,max}
+\setlength\titlebox{6.5cm}    % Expanding the titlebox
 
-\documentclass[11pt,letterpaper]{article}
 \usepackage{arabtex}
-\usepackage{naaclhlt2010}
-\usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{times}
 \usepackage{caption}
@@
 \usepackage{rotating}
-\usepackage{multirow}
 \usepackage{amsthm}
@@
 \usepackage{fancyvrb}
-%\usepackage{times}
-\usepackage{latexsym}
 \usepackage[colorlinks=false]{hyperref}
@@
 
-\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
+%\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
 \newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
@@
 
-\setlength\titlebox{6.5cm}    % Expanding the titlebox
-
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
@@
 
-\author{ Jad Makhlouta \\
-  \And
-Hamza Harkous \\
-  American University of Beirut \\
-  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb }
-  \And 
-  Fadi Zaraket 
-}
+%\author{ Jad Makhlouta \\\And
+%Hamza Harkous \\
+%  American University of Beirut \\
+%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb } \\\And 
+%  Fadi Zaraket 
+%}
 
@@
 \maketitle
-
 \begin{abstract}
@@
 Our results show high accuracy and efficiency.
-%compared to
-%state of the art analyzers. 
+\end{quote}
 \end{abstract}
@@
 without the need of a delimiter such as a space in between. 
-This happens because the first word \RL{almdrsT} ends with
-\RL{T} a non-connecting letter. 
+This happens because the first word \RL{il_A} ends with
+\RL{_A} a non-connecting letter. 
 These words occur often in text and greatly increase the
@@
 \resizebox{1.2\columnwidth}{!}
-{ \input{psfigs/exhadith.pstex_t}}
+{ \input{figs/exhadith.pdftex_t}}
 \caption{Hadith abstraction example.}
@@
 \resizebox{1.7\columnwidth}{!}
-{ \input{psfigs/FSM.pstex_t}}
+{ \input{figs/FSM.pdftex_t}}
 \caption{Example composition of the prefix, stem and suffix transducers.}
@@
 \resizebox{1.8\columnwidth}{!}
-{ \input{psfigs/abstract_machine.pstex_t}}
+{ \input{figs/abstract_machine.pdftex_t}}
 \caption{An abstraction of the non-deterministic composition of 
@@
 \resizebox{.9\columnwidth}{!}
-{ \input{psfigs/hadith.pstex_t}}
+{ \input{figs/hadith.pdftex_t}}
 \caption{The controller of the hadith case study.}
@@
 
-\section{Acknoledgements}
-\label{sec:acc}
-We thank the Lebanese National Council for Scientific Research (LNCSR) for funding this research.
+%\section{Acknoledgements}
+%\label{sec:acc}
+%We thank the Lebanese National Council for Scientific Research (LNCSR) for funding this research.
 
@@
 
-\bibliographystyle{naaclhlt2010}
+%\bibliographystyle{naaclhlt2010}
+\bibliographystyle{acl}
 {\small \bibliography{fzAr}}

### 2010-12-16 tex/sarf.aclhlt.tex
intro 1

@@
 abstracted each book into a vector of complex structures.
-Our results show high accuracy and efficiency.
+Our results show higher accuracy and better efficiency
+compared to current analyzers.
 \end{quote}
@@
 corresponding to their position in a word, i.e, beginning,
-middle, end of word forms and separate forms. 
+middle, end of word and separate forms. 
 This allows the phrase \transfalse
@@
 \RL{_A} a non-connecting letter. 
-These words occur often in text and greatly increase the
-difficulty of tokenization and are referred to as 
-``run-on'' words~\cite{Buckwalter:04}.
+These words,
+referred to as ``run-on'' words~\cite{Buckwalter:04},
+occur often in text, and greatly increase the
+difficulty of tokenization.
 
@@
 Buckwalter~\shortcite{Buckwalter:02},
-Beesly~\shortcite{Beesley:01}, SAMA~\cite{Kulick:10},
+Beesley~\shortcite{Beesley:01},
+SAMA~\cite{Kulick:10},
 and ElixirFM~\cite{Otakar:07} exist.
@@
 This approach has several problems. 
-First, 
-a white space delimited token may have 
+First, a white space delimited token may have 
 more than one word.
@@
 in some case studies as noted in~\cite{Maamouri:10}. 
-Other morphological analyzers such as Amira~\cite{Diab:07,Benajiba:07},
+Other morphological analyzers such as 
+Amira~\cite{Diab:07,Benajiba:07},
 MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
@@
 of performance.
+Xerox rules can be compiled into specialized finite state
+machine (FSM) based analyzers as described in~\cite{Beesley:03}.
+The efficiency of the resulting analyzer highly depends on the
+way the Xerox rules are written. 
+This requires deep knowledge and insight from the user
+in morphological analysis as well as skills in writing 
+context free grammar rules that can be automatically compiled
+into efficient non-deterministic finite state Automata (NDFSA).
 
@@
 morphological analyzer to answer queries that do not need 
-high accuracy at the low morphological analysis level 
-and that can often compensate at higher levels for tolerable inaccuracy 
-at low levels. 
+high accuracy at the low morphological analysis level.
+The analysis at the higher levels can often compensate for 
+tolerable inaccuracy at low levels. 
 For example, if a query concerns proper names and the 
@@
 to provide an answer.
+
+We find evidence to our hypothesis in~\cite{Maamouri:10} where 
+a new corpus was added to the Arabic Tree Bank 
+(ATB)~\cite{Maamouri:04}. 
+The addition required compliance with the DARPA GALE program 
+including several NLP tasks such as English translation 
+annotations and word-level alignment of Arabic and English. 
+The new genre of data challenged the existing data processing and 
+annotation guidelines. 
+Alternative guidelines came to place where the high level NLP 
+task interacted with the morphological analyzer. 
+This led to refining SAMA~\cite{Kulick:10}. 
+The work of Habash~\shortcite{Habash:06} reports that different 
+types of morphological analyses behave differently on the same case 
+study. 
+For one, the analyzers may find different solutions. 
+The analyzers may present the solutions in different orders. 
+The presented solutions and the order of presentation may not be 
+appropriate for the case study. 
+We find strong evidence in our results 
+that a case-based morphological analyzer that allows a case-based 
+controller to intervene at every decision to guide, use, order, and 
+refine the morphological analysis is of high utility.
+
 We find evidence to our hypothesis in~\cite{Maamouri:10} where the 

### 2010-12-16 tex/sarf.aclhlt.tex
fig small 1

@@
 
-We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
-addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
-with features demanding resolution at an abstraction level
-higher than the text itself
-led to a refined analyzer~\cite{Kulick:10}.  
-We also find evidence in~\cite{Habash:06} that different types of 
-morphological analyses behave differently on the same case study. 
-We strongly believe and we find evidence in our results that a 
-case-based morphological analyzer that allows a case-based
-controller to intervene at every decision to 
-guide, use, prune, and refine the morphological analysis
-is of high utility.
-
 \subsection{Sarf}
@@
 Sarf introduces recursive affixes in order to
-retain better part of speech information and enhance the 
+betterr retain part of speech information and enhance the 
 efficiency of affix matching.
@@
 \center{
-\resizebox{1.2\columnwidth}{!}
+\resizebox{.9\columnwidth}{!}
 { \input{figs/exhadith.pdftex_t}}
@@
 For example, the boxes in Figure~\ref{f:exhadith} are proper names and once
-connected toghether they form complex names of narrators. For example
-narrator $n_1$ has the first name \noTrRL{qtybT} and his father name is
+connected toghether they form complex names of narrators. 
+For example,
+narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
+name is
 \noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.
@@
 complex structure. 
-We also showcase that the results of the accurate high level abstraction
+We also showcase that the results of the accurate high level 
+abstraction
 results of the case-based analysis was not affected by a major
-refinement to the analyzer that increased low level analysis accuracy.
+refinement to the analyzer that increased low level analysis 
+accuracy.
 
@@
 We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02}  and use define the affixes
+Buckwalter~\shortcite{Buckwalter:02}  and define the affixes
 recursively. 
@@
 we compare Sarf to related work. Then in Section~\ref{sec:sarf}
-we discuss Sarf and illustrate the way it works with a running example.
+we discuss Sarf and illustrate the way it works with a running 
+example.
 We also explain how non-deterministic finite state Automata
@@
 finite-state Automata (FSA). 
-Beesley reports that the number of the Automata generated by a compiler
-from Xerox rules can not be controlled and also reports a difficulty to 
+Beesley reports that the number of Automata generated by a compiler
+from Xerox rules can not be controlled and also reports a 
+difficulty to 
 compose the FSAs into a single FSA framework. 
-Sarf solves the problem with an FSA that is the non-deterministic 
-composition of a controller and three
-transducers that are the prefix, stem, and suffix tansducers. 
+Sarf solves the problem with an FSA that is constructed from 
+the controller and 
+the prefix, stem, and suffix tansducers. 
 With Sarf it is easy to control the FSA with a custom user case 
-controller. Up to our understanding, Beesley  
+controller. 
+Up to our understanding, Beesley  
 requires recompiling the Xerox rules.
@@
 at a finer granularity level. 
-Also in Sarf there is no need for a separate tokenizer such as
-TOKAN as each alive Automata carries with it the positions in 
-text where a composition decision was made. 
-The case-based controller can build the trace 
-when needed.
+In Sarf, there is no need for a separate tokenizer such as
+TOKAN since each solution keeps a stack of positions 
+that partition text into morphemes.
 
@@
 
-We strongly feel that the refinement of SAMA~\cite{Maamouri:10} 
-is the closest to our approach. SAMA was refined to interact with
+Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
+SAMA was refined to interact with
 the ATB~\cite{Maamouri:04} project after the addition of a large 
-new corpus. We find supporting evidence in this decision to our hypothesis
-about the utility of a case-based analyzer. 
-The refinements that resulted in algorithmic changes in SAMA were 
+new corpus. 
+The algorithmic changes in SAMA were 
 done manually and worked in integration with the ATB format. 
@@
 the behavior of the analyzer on the fly and can interact
-with the any case study. In Sarf, the analyzer remains general, and the 
+with any case study. 
+In Sarf, the analyzer remains general, and the 
 case-based controller specializes the analysis.
@@
 
-Like ElixirFM~\cite{Otakar:07} Sarf builds on the lexicon
+Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
 of the Buckwalter analyzer. 
-We also use deterministic parsing with tries in our analysis 
-to run the prefix, suffix and stem transducers. 
+Sarf also uses deterministic parsing with tries 
+to implement the affix and stem transducers. 
 We think that the inferential-realizational approach 
@@
 Sarf leaves implementing that to the case-based controller
-since in many cases the abstraction needed may not need the 
-text to be grammatically correct.
+since in many cases the abstraction needs only a partial 
+linguistic model of the Arabic language. 
 %We can compare the case-based controller in Sarf to a high level
@@
 Sarf encodes the prefix, suffix and stem lexicons into 
-three linear transducers and uses a non-deterministic composition
+three linear transducers and uses a non-deterministic construction
 of the three to compute all valid morphological analyses.
 We first present a running example of Sarf. 
-The running example explains
-how the non-deterministic composition works and 
-illustrates how Sarf implements recursive affixes and
+The running example explains how Sarf works, 
+implememts recursive affixes, and
 handles ``run-on'' words.
@@
 \transfalse
-\begin{figure*}[tb]
+\begin{figure}[tb]
 \center{
-\resizebox{1.7\columnwidth}{!}
+\resizebox{\columnwidth}{!}
 { \input{figs/FSM.pdftex_t}}
-\caption{Example composition of the prefix, stem and suffix transducers.}
+\caption{Example construction of the affix and stem transducers.}
 \label{f:example}
 }
-\end{figure*}
+\end{figure}
 % the FSMs for the three words
@@
 The diagram in Figure~\ref{f:example}
-illustrates the operation of Sarf when parsing the
+illustrates the operation of $\Phi$ the Sarf FSA 
+when parsing the
 string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
@@
 }
-The square and circle legends are states 
-and the directed edges are transitions driven by an input
-letter that matches the label of the edge. 
-The square denotes an accept state, 
-the circle denotes a regular state, and the reject states
-are omitted for clarity. 
-Figure~\ref{f:example}(a) represents the prefix transducer and 
-Figures~\ref{f:example}(b) and~\ref{f:example}(c) represent the 
-stem and suffix transducers respectively. 
+
+The box and circle denote accept and regular
+states respectively.
+The edges are transitions with labels corresponding to
+the input letters.
+The reject states are omitted for clarity and are denoted
+by the abscence of corresponding transitions.
+
+Subfigures (a), (b), and (c) in Figure~\ref{f:example}
+represent ${\cal P}$ the prefix ,
+${\cal S}$ the stem, and ${\cal X}$ the suffix
+transducers respectively. 
 The symbol $\epsilon$ represents an empty string and is 
@@
 
-We start at the root square state in the prefix transducer
-which is an accept state. 
-The edge labeled $\epsilon$ connects the prefix 
-transducer to the stem transducer and transitions from any accept state
-in the prefix transducer to the root state in the stem transducer.
-The edge that connects the stem transducer to the suffix transducer
+We start with $\Phi$ at the root square state in 
+${\cal P}$ which is an accept state. 
+The $\epsilon$-edge connects the ${\cal P}$
+to ${\cal S}$ and transitions from any accept state
+in ${\cal P}$ to the root state in ${\cal S}$.
+The $\epsilon$-edge that connects the ${\cal S}$ to ${\cal X}$
 follows the same behavior.
@@
 When there are two valid transitions such as \RL{w} 
-and $\epsilon$ in the start case, the non-deterministic 
-composition $\Phi$ spawns an exact copy of itself $\Psi$. 
+and $\epsilon$ in the start case, 
+$\Phi$ spawns an exact copy of itself $\Psi$. 
 $\Phi$ makes one transition, and $\Psi$ makes the other. 
 Each of $\Phi$ and $\Psi$ now represent a valid analysis so far. 
-When $\Phi$ or $\Psi$ is at a state where the next input letter
-leads to a reject state, it dies.
-The reject states are omitted from Figure~\ref{f:example}
-and are denoted by the absence of the corresponding 
-letters. 
-In our example, if we assume there were no stems that start 
-with the letter \RL{w}, $\Psi$ will die then. 
-In reality probably $\Psi$ will die when the input string
+An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
+In our example, if there were no stems that start 
+with the letter \RL{w}, $\Psi$ will die. 
+In reality, $\Psi$ will die when the input 
 is at \RL{wsyl`}.
 
-Note that the prefix and affix transducers allow recursive 
-prefixes. For example, \RL{w} will result in an accept state
-that transitions to the stem transducer. 
-When \RL{s} follows, we move to another accept state in the 
-prefix transducer corresponding to the suffix \RL{ws}. 
-The same applies to the suffix transducer.
+Note that the affix transducers allow recursive 
+affixes. 
+For example, \RL{w} will result in an accept state
+that transitions to the ${\cal S}$.
+When \RL{s} follows, we move to another accept state in 
+${\cal P}$ corresponding to the prefix \RL{ws}. 
+The same applies to ${\cal X}$.
 
 Lets consider $\Phi$ after it consumed \RL{wsy} 
-and transitioned into the stem transducer root node. 
+and transitioned into the root node of ${\cal S}$.
 Now $\Phi$ will transition with \RL{l`b} to reach an accept 
@@
 
-Since $\Phi$ is now in an accept state in the stem transducer, it will
-spawn $\Sigma$ that transitions into the root of the 
-suffix transducer. When \RL{h} is consumed, $\Phi$ will die since 
+Since $\Phi$ is now in an accept state in ${\cal S}$, it will
+spawn $\Xi$ that transitions into the root of ${\cal X}$. 
+When \RL{h} is consumed, $\Phi$ will die since 
 there is no edge from the current state labeled with \RL{h}. 
-Now $\Sigma$ represents a valid analysis and can proceed.
-
-Note that $\Sigma$ reports a full word at any accept state
-by spawning a new composition using the $\epsilon$ transition
-to the prefix transducer.
+Now $\Xi$ represents a valid analysis and can proceed.
+
+Note that $\Xi$ reports a full word at any accept state
+by spawning a new FSA using the $\epsilon$ transition
+to ${\cal P}$.
 In Sarf, this only happens with white space, delimiters, 

### 2010-12-16 trunk/tex/rp.tex
changed values for ELIXIR and some changes in abstract diagram

@@
 We find evidence to our hypothesis in~\cite{Maamouri:10} where 
-a new corpus was added to the Arabic Tree Bank~\cite{Maamouri:04}.
-The addition requires compliance with 
-the DARPA GALE program including several NLP tasks such as 
-English translation annotations and word-level alignment
-of Arabic and English. 
-The new genre of data challenged the existing data processing 
-and annotation guidelines. 
-Alternative guidelines came to place where the 
-high level NLP task interacted with the 
-morphological analyzer.% along with other concurrent tasks. 
-This led to a refined analyzer~\cite{Kulick:10}.  
-We also find evidence in~\cite{Habash:06} that different types of 
-morphological analyses behave differently on the same case study. 
-For example, the analyzers may find different solutions, and may 
-present the same solutions in different orders.
-We find strong evidence in our preliminary
-results that a 
-case-based morphological analyzer that allows a 
-case-based controller to intervene at every decision to 
-guide, use, and refine the morphological analysis
-is of high utility.
+a new corpus was added to the Arabic Tree Bank 
+(ATB)~\cite{Maamouri:04}. 
+The addition required compliance with the DARPA GALE program 
+including several NLP tasks such as English translation 
+annotations and word-level alignment of Arabic and English. 
+The new genre of data challenged the existing data processing and 
+annotation guidelines. 
+Alternative guidelines came to place where the high level NLP 
+task interacted with the morphological analyzer. 
+This led to refining SAMA~\cite{Kulick:10}. 
+The work of Habash~\shortcite{Habash:06} reports that different 
+types of morphological analyses behave differently on the same case 
+study. 
+For one, the analyzers may find different solutions. 
+The analyzers may present the solutions in different orders. 
+The presented solutions and the order of presentation may not be 
+appropriate for the case study. 
+We find strong evidence in our preliminary results 
+that a case-based morphological analyzer that allows a case-based 
+controller to intervene at every decision to guide, use, order, and 
+refine the morphological analysis is of high utility.
 

### 2010-12-16 trunk/tex/sarf.aclhlt.tex
changed values for ELIXIR and some changes in abstract diagram

--- 2010-12-12 trunk/tex/sarf.tex
@@
 % Contact: nasmith@cs.cmu.edu
-
-\documentclass[11pt,letterpaper]{article}
+\documentclass[11pt]{article}
+\usepackage{acl-hlt2011}
+\usepackage{times}
+\usepackage{latexsym}
+\usepackage{amsmath}
+\usepackage{multirow}
+\usepackage{url}
+\DeclareMathOperator*{\argmax}{arg\,max}
+\setlength\titlebox{6.5cm}    % Expanding the titlebox
+
 \usepackage{arabtex}
-\usepackage{naaclhlt2010}
-\usepackage{amsmath}
 \usepackage{amssymb}
-\usepackage{times}
 \usepackage{caption}
@@
 abstracted each book into a vector of complex structures.
-Our results show high accuracy and efficiency.
-%compared to
-%state of the art analyzers. 
+Our results show higher accuracy and better efficiency
+compared to current analyzers.
+\end{quote}
 \end{abstract}
@@
 without the need of a delimiter such as a space in between. 
-This happens because the first word \RL{almdrsT} ends with
-\RL{T} a non-connecting letter. 
-These words occur often in text and greatly increase the
-difficulty of tokenization and are referred to as 
-``run-on'' words~\cite{Buckwalter:04}.
+This happens because the first word \RL{il_A} ends with
+\RL{_A} a non-connecting letter. 
+These words,
+referred to as ``run-on'' words~\cite{Buckwalter:04},
+occur often in text, and greatly increase the
+difficulty of tokenization.
 
@@
 to provide an answer.
-We find evidence to our hypothesis in~\cite{Maamouri:10} where the 
-addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
-with features demanding resolution at an abstraction level
-higher than the text itself
-led to a refined analyzer~\cite{Kulick:10}.  
-We also find evidence in~\cite{Habash:06} that different types of 
-morphological analyses behave differently on the same case study. 
-We strongly believe and we find evidence in our results that a 
-case-based morphological analyzer that allows a case-based
-controller to intervene at every decision to 
-guide, use, prune, and refine the morphological analysis
-is of high utility.
+
+We find evidence to our hypothesis in~\cite{Maamouri:10} where 
+a new corpus was added to the Arabic Tree Bank 
+(ATB)~\cite{Maamouri:04}. 
+The addition required compliance with the DARPA GALE program 
+including several NLP tasks such as English translation 
+annotations and word-level alignment of Arabic and English. 
+The new genre of data challenged the existing data processing and 
+annotation guidelines. 
+Alternative guidelines came to place where the high level NLP 
+task interacted with the morphological analyzer. 
+This led to refining SAMA~\cite{Kulick:10}. 
+The work of Habash~\shortcite{Habash:06} reports that different 
+types of morphological analyses behave differently on the same case 
+study. 
+For one, the analyzers may find different solutions. 
+The analyzers may present the solutions in different orders. 
+The presented solutions and the order of presentation may not be 
+appropriate for the case study. 
+We find strong evidence in our results 
+that a case-based morphological analyzer that allows a case-based 
+controller to intervene at every decision to guide, use, order, and 
+refine the morphological analysis is of high utility.
 
@@
 Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
-Running time (secs.)& & 6.65 & 12$\times$60& &4.55 & 14$\times$60 & & .66 & 6$\times$60 
+Running time (secs.)& & 6.65 & 2.78$\times60^2$& &4.55 & 2.3$\times60^2$ & & .66 & 29.2$\times$60 
 \end{tabular}

### 2010-12-16 trunk/tex/sarf.aclhlt.tex
in the controller

@@
 
-\section{Related Work }
+\section{Related work }
 \label{sec:related}
@@
 compose the FSAs into a single FSA framework. 
-Sarf solves the problem with an FSA that is the non-deterministic 
-composition of a controller and three
-transducers that are the prefix, stem, and suffix tansducers. 
+Sarf solves the problem with an FSA that is constructed from 
+the controller and 
+the prefix, stem, and suffix deterministic tansducers. 
+Sarf differs in that it uses recursive affixes in the 
+affix transducers, and in that the transducers consider
+input from the controller. 
 With Sarf it is easy to control the FSA with a custom user case 
@@
 { \input{figs/FSM.pdftex_t}}
-\caption{Example composition of the prefix, stem and suffix transducers.}
+\caption{Example affix and stem transducers.}
 \label{f:example}
@@
 there is no edge from the current state labeled with \RL{h}. 
-Now $\Sigma$ represents a valid analysis and can proceed.
+$\Xi$ represents a valid analysis and can proceed.
 
@@
 
-Finally, the case-based controller can interfere at any point in the 
-analysis to make a decision like ignoring a composition that 
+The case-based controller can interfere at any point in the 
+analysis to make a decision like ignoring an FSA that 
 is not interesting. 
-It may also decide to correct the analysis of one Automata
+It may also decide to correct the analysis of one FSA
 based on the decision of the others. 
@@
 to make its decision. 
-It does not produce any output and therefore the composition
-of the three transducers with the controller is a finite 
-state Automaton (FSA).
+It does not produce any external output and therefore 
+the three transducers with the 
+controller compose an FSA.
 
@@
 
-Morphological analysis partitions a given Arabic word
-into a set of morphemes.
-A morpheme is either an affix or a stem. 
 We call two morphemes compatible if their concatenation
@@
 Buckwalter~\shortcite{Buckwalter:02} keeps separate lists 
-of prefixes, suffixes and stems and assigns a category
-to each morpheme. 
-The relation between morphemes and categories is one 
-to many.
-
+of affix and stem morphemes and assigns categories to
+morphemes. 
+%The relation between morphemes and categories is one 
+%to many. 
 A compatibility rule is a pair of categories 
@@
 a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
-The $\alpha\beta\gamma$ partition of the string is a 
-valid morphological analysis if  and only if we have
+The $\alpha\beta\gamma$ partition of $s$ is a 
+valid morphological analysis if and only if 
 $\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
 $\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
-$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$ where
+$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
 $c()$ returns the category of the morpheme.
@@
 of the POS tags of each of the three morphemes. 
-Keeping these morphemes separate in a list
-is redundant.
-When users try to add their custom morphemes to these
-lists they may produce consistency issues. 
-We add two lists of compatibility rules $L_{pp}$ and
-$L_{xx}$ for prefix-prefix and suffix-suffix compatibility
-respectively.
+Keeping all valid morphemes in a list
+is redundant and may lead to consistency issues when
+users try to add their custom morphemes.
+We introduce $L_{pp}$ and
+$L_{xx}$ as prefix-prefix and suffix-suffix 
+compatibility rules respectively.
 This allows us to keep a shorter list of affixes. 
-We need one more additional modification to accommodate 
-a category for the morphemes accepted by the $L_{pp}$
-and the $L_{xx}$ rules. For that we introduce the concept
-of a {\em resulting category}. 
+We also introduce a {\em resulting category}
+for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
 The resulting category in all practical cases
@@
 
-Formally we define an affix $f$ recursively to be either
-an atomic affix $a$ or an atomic affix concatenated
-to an affix and we write $f: a | a f$. 
 Atomic affixes are standalone affixes
 that are not the concatenation of other affixes. 
-Sarf only keeps atomic affixes and their categories.
-
-\subsection{Affix linear transducer}
+Formally, we define an affix $f$ to be either
+an atomic affix $a$ or an atomic affix concatenated
+with an affix. We write $f: a | a f$. 
+%Sarf only keeps atomic affixes and their categories.
+
+\subsection{Affix transducers}
 \label{sec:affixFSA}
@@
 Buckwalter~\shortcite{Buckwalter:02}
-and extracted our refined lists of recursive affixes.
-We encoded the lists into a structure similar to those
-in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
-The two structures are directed acyclic graphs. 
-The absence of cycles reduces the 
-traversal of the affix transducer to a linear traversal.
+and constructed the recursive affixes
+into directed acyclic graphs (DAG) similar to 
+those in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
+This transducer is deterministic and thus 
+its traversal is linear.
 This is computationally superior to the 
@@
 
-The prefix and suffix transducers encode the $L_{pp}$ and
-$L_{xx}$ compatibility lists.
+The affix transducers encode $L_{pp}$ and
+$L_{xx}$.
 An accept state in the transducer corresponds to the last letter 
-of a prefix in the prefix and suffix lists.
+of an affix in the affix lists.
 The accept states produce as output the POS and other tags
@@
 
-\subsection{Stem linear transducer}
+\subsection{Stem transducer}
 \label{sec:stemFSA}
@@
 We augmented the lexicon with a set of proper names and
-a set of location names. 
-We obtained the set of proper names from online 
-and biblical sources.
+a set of location names which we 
+obtained from online and biblical sources.
 
@@
 an efficient double array trie structure~\cite{Aoe:89}. 
-The structure is also linear where the accept
-states are the terminal nodes corresponding to the last 
+The structure is deterministic and free of cycles where the 
+accept states are the terminal nodes corresponding to the last 
 letters in stems. 
@@
 
-\subsection{Non-deterministic composition of transducers}
+\subsection{Construction of Sarf from transducers}
 \label{sec:ndfsa}
@@
 The diagram in Figure~\ref{f:composition} shows an 
-abstraction of the non-deterministic composition 
-of the prefix, stem and suffix transducers. 
+abstraction of the non-deterministic construction of Sarf
+from the prefix, stem and suffix transducers. 
 The circle represents regular states, the square
@@
 Note that the concrete prefix, stem and suffix transducers
-are all linear and do not have cycles as discussed 
+are all deterministic and cycle free as discussed 
 is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
 The cycles are introduced in the abstract machine
-for brevity and pedagogical purposes.
+for pedagogical purposes.
 
@@
 { \input{figs/abstract_machine.pdftex_t}}
-\caption{An abstraction of the non-deterministic composition of 
-    the prefix, stem and suffix transducers.}
+\caption{An abstraction of the non-deterministic construction of  
+    Sarf from the affix and stem transducers.}
 \label{f:composition}
@@
 
-The algorithm \CodeIn{NDSarf} below controls Sarf. 
-It takes as input a text string $L$ and a case-based controller
+The algorithm \CodeIn{NDSarf} 
+takes as input a text string $L$ and a case-based controller
 $A$. 
@@
 
-The arrows in Figure~\ref{f:composition} that transition
-from prefix to stem, stem to suffix, and suffix to prefix
-require compatibility constraints for 
-the transition.
+%The arrows in Figure~\ref{f:composition} that transition
+%from prefix to stem, stem to suffix, and suffix to prefix
+%require compatibility constraints for 
+%the transition.
 
@@
 
-\section{Islamic Literature Case Study}
+\section{Islamic literature case study}
 \label{sec:islamic}
@@
 The literature lacks an exhaustive mechanism of authentication
-check as. This is currently manual work prone to error and almost
+checks. This is currently manual work prone to error and almost
 impossible to complete in the life time of a scholar due to the
@@
 %the analysis of the hadith literature. 
-In this paper we use Sarf to successfully automate the
+In this paper, we use Sarf to successfully automate the
 analysis of three books of hadith selected 
@@
 of the hadith. 
-We are concerned to further explore the sanad and
-we detect the chain of narrators and 
+We further explore the sanad and
+detect the chain of narrators and 
 the relation that links each narrator to his ancestor and 
-his predecessor in the chain. 
+predecessor in the chain. 
 We also detect the full name of each narrator that is
@@
 
-We built our hadith case-based controller to target
+We built our hadith controller to target
 the detection of proper names. 
@@
 appearing as ``run-on'' words.
-The controller also targets words that mean narrate when
+The controller also targets words that mean ``narrate'' when
 they appear in the neighborhood of multiple proper names.
@@
 
-We illustrate the hadith case-based controller 
+We illustrate the hadith controller 
 in Figure~\ref{f:hadith}. 
-The transitions in the hadith controller are excited
-by inputs from the Sarf morphological analyzer. 
-A NAME label means that Sarf encountered a proper name.
-An NRC label means that Sarf met a narrator connector such as
-\RL{`an} "on behalf of", \RL{.hada_t} "narrated", \RL{qaal} "said", 
-\RL{'a_hbar} "told" or one of their derivations. 
-The IBN label corresponds to the word \RL{ibn} "the son of" and is a name connector.
+The transitions are labeled by outputs generated from 
+the Sarf transducers. 
+A NAME label denotes a proper name.
+An NRC label denotes a narrator connector such as
+\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
+\RL{'a_hbar} ``told'' or one of their derivations. 
+The IBN label corresponds to the word \RL{ibn} ``the son of'' and is a name connector.
 The NISBA label corresponds to an adjective that points to a person such 
-as \RL{alma.sriy} "the Egyptian".
+as \RL{alma.sriy} ``the Egyptian''.
 
@@
 The controller has four states that correspond to 
-an abstract position in text relative to the next sanad. 
+a position in text relative to the next sanad. 
 State TEXT\_S is the initial state and denotes that
 the controller is outside the context of a sanad.
-The controller moves to the state NAME\_S whenever
-NAME is reported by Sarf.
-State NRC\_S means that the controller thinks it is in the context
+The controller moves to state NAME\_S on
+NAME.
+State NRC\_S means that the controller is in the context
 of a narrator name after Sarf reports an NRC.
@@
 
-The controller reports a valid chain of narrators when a sequence of names
+The controller reports a valid chain of narrators when a 
+sequence of names
 connected by narrator connectors appears. 
-It marks the beginning of the hadith with the beginning of the current sequence,
-and marks the end of the hadith with the beginning of the next sequence. 
+It marks the beginning of the hadith with the beginning of 
+the current sequence,
+and marks the end of the hadith with the beginning of the 
+next sequence. 
 It also marks the chain of narrators as the sequence itself.

### 2010-12-16 tex/sarf.aclhlt.tex
save from diff

@@
 the state of the analysis on the fly. 
-It uses compositional non-deterministic finite state machines 
+It uses non-deterministic finite state Automata
 to analyze a stream of text. 
@@
 may have two valid morphological analyses. 
-The letter \RL{'a} may be prefix and the word means ``I praise him''.
-The letter \RL{'a} may also be part of the stem 
-\RL{'a.hmad} (a proper noun)
+The letter \RL{'a} may be a prefix and the word means 
+``I praise him'', or 
+%The letter \RL{'a} may also be 
+part of the stem \RL{'a.hmad} (a proper noun)
 and the word means ``his Ahmad''.
@@
 referred to as ``run-on'' words~\cite{Buckwalter:04},
-occur often in text, and greatly increase the
+occur often, and greatly increase the
 difficulty of tokenization.
@@
 This led to refining SAMA~\cite{Kulick:10}.
+
 The work of Habash~\shortcite{Habash:06} reports that different 
@@
 study. 
-For one, the analyzers may find different solutions. 
-The analyzers may present the solutions in different orders. 
+The analyzers may find different solutions or 
+may present the solutions in different orders. 
 The presented solutions and the order of presentation may not be 
@@
 In this paper, we present Sarf, a {\em novel case-based efficient
-morphological analyzer} that uses compositional 
-non-deterministic Automata driven by a case based controller.
-Each alive machine in Sarf represents a valid morphological analysis. 
+morphological analyzer} that uses 
+non-deterministic finite state Automata 
+driven by a case based controller.
+%Each alive machine in Sarf represents a valid morphological analysis.
 Sarf keeps alive all possible analyses of the text and gives 
@@
 
-Sarf uses a {\em recursive} concatenative analysis with a novel 
+Sarf uses 
+concatenative analysis with a novel 
+{\em recursive} 
 refinement. 
 Sarf introduces recursive affixes in order to
-betterr retain part of speech information and enhance the 
+better retain part of speech information and enhance the 
 efficiency of affix matching.
@@
 through a \RL{sanad} or a sequence of narrators. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
-transliteration and translation. 
-The authentication of a \noArRL{.hady_t} highly depends on the credibility
-of each of the narrators as reported in separate biography 
-books. 
+Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
+Arabic with its transliteration and translation. 
+The authentication of a \noArRL{.hady_t} depends on 
+the credibility of each of the narrators as reported in 
+separate biography books. 
 In this paper we consider the problem of automatically segmenting
-a \noArRL{.hady_t} book into narrations, then segmenting each narration into
-its content or \RL{matn} and its \noArRL{sanad}.
-We also partition the sanad accurately into the 
+a \noArRL{.hady_t} book into narrations, then segmenting each 
+narration into its content or \RL{matn} and its \noArRL{sanad}.
+We also partition the sanad into the 
 separate narrators so that we can later look each one of them 
 up in the biography books. 
-For example, the boxes in Figure~\ref{f:exhadith} are proper names and once
+For example, the boxes in 
+Figure~\ref{f:exhadith} are proper names and once
 connected toghether they form complex names of narrators. 
-For example,
-narrator $n_1$ has the first name \noTrRL{qtybT} and his father 
-name is
+For example, narrator $n_1$ has the first name 
+\noTrRL{qtybT} and his father name is
 \noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.
@@
 of proper names in the text. It ignores the text where
-proper names occur rarely as long 
-as the text is valid.
-The controller is also interested in discovering a subset of words 
+proper names occur rarely.% as long 
+%as the text is valid.
+We are also interested in discovering a subset of words 
 that relate persons to each other such as \RL{bin} or ``the son of''
@@
 three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
-and extracted from each one of them a vector of a three level deep
-complex structure. 
-We also showcase that the results of the accurate high level 
-abstraction
-results of the case-based analysis was not affected by a major
-refinement to the analyzer that increased low level analysis 
-accuracy.
+and extracted from each one of them a vector of complex structures 
+that are three level deep in hierarchy.
+%We also showcase that the results of the accurate high level 
+%abstraction
+%results of the case-based analysis was not affected by a major
+%refinement to the analyzer that increased low level analysis accuracy.
 
@@
 we compare Sarf to related work. Then in Section~\ref{sec:sarf}
-we discuss Sarf and illustrate the way it works with a running 
-example.
+we discuss Sarf and illustrate the way it works with a running example.
 We also explain how non-deterministic finite state Automata
@@
 finite-state Automata (FSA). 
-Beesley reports that the number of Automata generated by a compiler
-from Xerox rules can not be controlled and also reports a 
-difficulty to 
+Beesley reports that the number of the Automata generated by a compiler
+from Xerox rules can not be controlled and also reports a difficulty to 
 compose the FSAs into a single FSA framework. 
@@
 the controller and 
-the prefix, stem, and suffix tansducers. 
+the prefix, stem, and suffix deterministic tansducers. 
+Sarf differs in that it uses recursive affixes in the 
+affix transducers, and in that the transducers consider
+input from the controller. 
 With Sarf it is easy to control the FSA with a custom user case 
-controller. 
-Up to our understanding, Beesley  
+controller. Up to our understanding, Beesley  
 requires recompiling the Xerox rules.
@@
 that partition text into morphemes.
+%The case-based controller can build the trace 
+%when needed.
 
@@
 new corpus. 
+%We find supporting evidence in this decision to our hypothesis
+%about the utility of a case-based analyzer. 
 The algorithmic changes in SAMA were
@@
 { \input{figs/FSM.pdftex_t}}
-\caption{Example construction of the affix and stem transducers.}
+\caption{Example affix and stem transducers.}
 \label{f:example}
@@
 \transtrue
+
+%{\bf Sarf example.}
 
@@
 state.
+
 Before moving with the letter \RL{b} to the accept state,
@@
 Note that the concrete prefix, stem and suffix transducers
-are all linear and do not have cycles as discussed 
+are all deterministic and cycle free as discussed 
 is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
 The cycles are introduced in the abstract machine
-for brevity and pedagogical purposes.
-
-\begin{figure*}[tb]
+for pedagogical purposes.
+
+\begin{figure}[tb]
 \center{
-\resizebox{1.8\columnwidth}{!}
+\resizebox{\columnwidth}{!}
 { \input{figs/abstract_machine.pdftex_t}}
-\caption{An abstraction of the non-deterministic composition of 
-    the prefix, stem and suffix transducers.}
+\caption{Sarf NDFSA.}
 \label{f:composition}
 }
-\end{figure*}
-
-The algorithm \CodeIn{NDSarf} below controls Sarf. 
-It takes as input a text string $L$ and a case-based controller
+\end{figure}
+
+The algorithm \CodeIn{NDSarf} 
+takes as input a text string $L$ and a case-based controller
 $A$. 
@@
 
-The arrows in Figure~\ref{f:composition} that transition
-from prefix to stem, stem to suffix, and suffix to prefix
-require compatibility constraints for 
-the transition.
+%The arrows in Figure~\ref{f:composition} that transition
+%from prefix to stem, stem to suffix, and suffix to prefix
+%require compatibility constraints for 
+%the transition.
 
 \begin{table}[tb]
-\centering
+%\centering
+%\resizebox{.9\columnwidth}{!}{
 \begin{tabular} {p{6cm}}
-\begin{Verbatim}[fontsize=\relsize{-1},
+\begin{Verbatim}[fontsize=\relsize{-2},
 frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
@@
 \end{tabular}
+%}
 \label{a:ndsarf}
@@
 
+Intuitively, the controller reports a valid chain 
+of narrators when a sequence of names
+connected by narrator connectors appears. 
+It marks the beginning of the hadith with the beginning of 
+the current sequence,
+and marks the end of the hadith with the beginning of the 
+next sequence. 
+It also marks the chain of narrators as the sequence itself.
+
 The controller has four states that correspond to 
-an abstract position in text relative to the next sanad. 
+a position in text relative to the next sanad. 
 State TEXT\_S is the initial state and denotes that
 the controller is outside the context of a sanad.
-The controller moves to the state NAME\_S whenever
-NAME is reported by Sarf.
-State NRC\_S means that the controller thinks it is in the context
+The controller moves to state NAME\_S on
+NAME.
+State NRC\_S means that the controller is in the context
 of a narrator name after Sarf reports an NRC.
@@
 
+Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
+before it gives up on its expectations.
+
+We reach the NAME\_S state only when a
+valid NAME is detected and we leave when no more NAME's are detected.
+The NRC\_S state can only be reached if an NRC is detected.
+NMC\_S can only be reached from a NAME\_S state.
+
 \begin{figure}[tb!]
 \center{
-\resizebox{.9\columnwidth}{!}
+\resizebox{.8\columnwidth}{!}
 { \input{figs/hadith.pdftex_t}}
@@
 
-Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
-before it gives up on its expectations.
-
-We reach the NAME\_S state only when a
-valid NAME is detected and we leave when no more NAME's are detected. 
-The NRC\_S state can only be reached if an NRC is detected.
-NMC\_S can only be reached from a NAME\_S state.
-
-The controller reports a valid chain of narrators when a sequence of names
-connected by narrator connectors appears. 
-It marks the beginning of the hadith with the beginning of the current sequence,
-and marks the end of the hadith with the beginning of the next sequence. 
-It also marks the chain of narrators as the sequence itself.
-
-The definition of input labels such as NAME and IBN depends on the 
-morphological analyzer. 
-However, our case-based controller approach was able to perform well
-under both Sarf and a refined version of Sarf.
+%The definition of input labels such as NAME and IBN depends on the 
+%morphological analyzer. 
+%However, our case-based controller approach was able to perform well
+%under both Sarf and a refined version of Sarf.
 
@@
 %\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
+\resizebox{1.8\columnwidth}{!}{
 \begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
@@
 Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
-Running time (secs.)& & 6.65 & 12$\times$60& &4.55 & 14$\times$60 & & .66 & 6$\times$60 
+Running time (secs.)& & 6.65 & 2.78$\times60^2$& &4.55 & 2.3$\times60^2$ & & .66 & 29.2$\times$60 
 \end{tabular}
+}
 \normalsize

### 2010-12-17 tex/sarf.aclhlt.tex
results and conclusion

@@
 
-In section we present results that compare Sarf to existing 
+In this section we present results that compare Sarf to existing 
 morphological analyzers such as Buckwalter and ElixirFM. 
 We also evaluate our hypothesis of the utility of a case-based 
-morphological analyzer,
-by running our case study in two modes, simple and refined.
+morphological analyzer.
+We validate that the high-level analysis of the output 
+of low-level morphological analysis can compensate for accuracy.
+To acheive that we ran Sarf in two modes, simple and refined.
+In the simple mode, we used Sarf as discussed in 
+Section~\ref{sec:sarf} with the 
+controller of Section~\ref{sec:controller}.
+In the refined mode, we applied several low-level morphological 
+refinements to Sarf such as disambiguation of 
+female proper names so that the controller can ignore them. 
+We also added a refinement that detected compound proper names 
+such as \RL{`bd alr.hmn} 
+as they are common in narrator names.
+%Finally, we consider a possessive form as a reference to 
+%a person and treat
+%it as a narrator.
+
 We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
@@
 
-In the simple mode, Sarf is controlled by the hadith controller.
-In the refined mode, we apply several refinements to Sarf.
-The first refinement ignore female proper names except at the end of
-a chain since they do not occur often in narrator names in hadith while
-they can be common in early narrators such as the wives of the prophet.
-We also add a refinement that detects compound proper names 
-such as \RL{`bd alr.hmn} 
-as they are common in narrator names.
-Finally, we consider a possessive form as a reference to a person and treat
-it as a narrator.
-
-We report our results in Table~\ref{t:hadithresallresults} on the three hadith books.
-The word count row is an indicator of the size of data.
-The number of names, narrators, and chains reflect a hierarchical structure where a chain contains
+We report our results in Table~\ref{t:hadithresallresults} on the 
+three hadith books.
+In general, we observe that the refinements slightly 
+improved performance and accuracy at
+the lowest level of abstraction by reducing false positives.
+However, the refinements had little to no effect on the high 
+accuracy measures at higher levels of abstraction.
+We believe that this is a strong supporting evidence to our 
+hypothesis.
+
+The row for ``word count'' is an indicator of the size of the data.
+The %numbers in 
+rows ``names'', ``narrators'', and ``chains'' denote the total
+number of the corresponding entity per book.
+The relative magnitudes of these rows
+reflect the hierarchical structure where 
+a chain contains
 narrators, and a narrator contains names.
-In general, we observe that the refinements slightly improved performance and accuracy at
-the lowest level of abstraction by reducing false positives.
-However, the refinements had little to no effect on the high accuracy measures at higher
-levels of abstraction.
-We believe that this is a strong supporting evidence to our hypothesis.
-
-The names row reports the total number of detected names. The names/narrator row reports the average
-number of names per narrator structure. The narrators row reports the number of narrators detected.
-The narrators/chain row reports the number of narrators per chain structure. And the chain row reports
-the number of detected chains in the book.
-The ignored names row reports the number of names detected outside the chains and thus irrelevant to
+The row ``ignored names'' reports the number of names detected 
+outside the chains and thus are irrelevant to
 the case study.
-The segmentation accuracy reports on the accuracy of segmentation of the book into separate hadiths.
-The metric is computed manually over a representative sample of each text. The same applies to all other accuracy measures.
-The chain accuracy reflects the percentage of correct narrators covered by extracted chains.
-The narrator name accuracy row reflects the percentage of correct proper names detected within
+In addition, the rows ``names/narrator'' and ``narrators/chain'' 
+report the average
+number of names per narrator and 
+the average number of narrators per chain respectively.
+
+The segmentation accuracy reports on the accuracy of 
+segmentation of the book into separate hadiths.
+The metric is computed manually over a representative sample 
+of each book.
+The same applies to all other accuracy measures.
+The chain accuracy reflects the percentage of correct narrators 
+covered by extracted chains.
+The ``narrator name accuracy'' row reflects the percentage of 
+correct proper names detected within
 narrator structures.
 
-The name false positives row reports the percentage of words recognized as names while they should not.
-We notice that the low level metric of name false positives improved dramatically with the refined
-analysis without substantially affecting the high abstract measures such as chain and segmentation
+The row ``name false positives'' reports the percentage of words 
+recognized as names while they should not.
+We notice that this low level metric improved dramatically 
+with the refined analysis without substantially affecting the 
+high abstract measures such as chain and segmentation
 accuracy.
-We notice that the controller in the simple analysis ignored more names and thus the case-based
-controller compensated for the inaccuracy of the low level analysis with a simple heuristic and
-without the need of the fancy refinements.
-
-We investigated the case of Al Kafi where the accuracy of the simple analysis was significantly lower than that of the refined
-analysis. We  discovered that the chains detected by the refined analysis were a subset of those detected by the simple
+
+We notice that the controller in the simple analysis ignored 
+more names and thus the case-based
+controller compensated with a simple heuristic
+for the inaccuracy of the low level 
 analysis.
-The reason for the lower accuracy is false positive chains. 
-The simple analysis did not miss important information. 
-We hypothesize that a higher level of abstraction such as intersecting the narrators from the hadith books with the narrators 
-in the biographies can reduce the false positives in the chains.
+%and without the need of the low-level refinements.
+
+We investigated the case of Al-Kafi where the accuracy of 
+the simple analysis was significantly lower than that of the 
+refined analysis. 
+We discovered that the chains detected by the refined analysis 
+were a subset of those detected by the simple
+analysis. 
+The reason for the lower accuracy is false positive chains, 
+implying that the simple analysis did not miss important 
+information.
+
+We hypothesize that a higher level of abstraction such as 
+intersecting the narrators from the hadith books with the narrators 
+in the biographies can reduce the false positives in the chains
+in this case.
 
@@
 We implemented our case study with Buckwalter in two phases. 
-In phase one, we process the entire text with Buckwalter and
-obtain annotations to locations in text. 
-Then in the second phase we feed these annotations to 
-our controller as if they were coming from Sarf. 
-We measure the time it took to produce the annotations
+In phase one, we processed the entire text with Buckwalter and
+obtained annotations to text. 
+In the second phase, we fed the annotations to 
+the hadith controller as if they were coming from Sarf. 
+We measured the time it took to produce the annotations
 and the time it took the controller to process the 
@@
 
-Sarf outperforms both Buckwalter and ElixirFM in runing time. 
+Sarf outperforms both Buckwalter and ElixirFM in terms of 
+running time. 
 It also produces higher accuracy measures. 
@@
 names by ignoring prefixes and suffixes that can not form
-names even if the stem was a name. 
+names even if one possible alternative of the stem is a name. 
 Doing the same with Buckwalter and ElixirFM requires  
-expanding the annotations resulting from one affix to reflect
-its multiple composing affixes.
-This is easily done with the way we defined affixes recursively.
-
-Buckwalter performed much better than ElixirFM. This is up to 
-our knowledge the first comparison of these two tools and 
-we think the big difference is due to the more exact 
-and grammar based functional morphological analysis ElixirFM 
-uses.
+expanding the annotations resulting from one simple affix to be 
+reflected in all the composite affixes that contain it.
+%expanding the annotations resulting from one affix to reflect
+%its multiple composing affixes.
+%However, this was easily performed using our recursive definition 
+%of the affixes.
+
+Buckwalter performed much better than ElixirFM in 
+terms of running time, 
+while ElixirFM outperformed Buckwalter in terms of accuracy.
+%This is up to 
+%our knowledge the first comparison of these two tools and 
+We think the big difference is due to the sofistication of 
+grammar based functional morphological analysis ElixirFM 
+uses and its reliance on additional annotations extracted from both ATB~\cite{Maamouri:04} and the Pragbank~\cite{Prague04}.
 
@@
 
-In the future, we plan to complete the hadith case study 
+We presented Sarf, a novel case-based morphological analyzer,
+and validated its utility using the hadith literature
+case study. 
+Sarf uses non-deterministic finite state Automata with 
+concatenative analysis and recursive affixes to analyze 
+Arabic text. 
+Our results showed that Sarf with the hadith controller
+is faster and more accurate than
+current morphological analyzers.
+
+In the future, 
+we will try Sarf with other interesting case studies.
+We plan to complete the hadith case study 
 and automate narrator discovery and authentication in the 
@@
 learning system that can learn more names. 
-We also plan on developing a domain specific language similar 
-to ElexirFM~\cite{Otakar:07}
-that allows users to easily specify their case-based controller
-to interact with Sarf.
+We also plan on developing an Arabic linguistic computational model 
+similar to that of ElixirFM~\cite{Otakar:07} but that can be 
+explored partialy based on the case study.
 

### 2010-12-17 trunk/tex/sarf.aclhlt.tex
final before 8

@@
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Case-based Arabic Morphological Analysis }
+\title{Case-based Arabic Morphological Analyzer for Arabic Text Mining}
 
@@
 
-In this paper, we present Sarf, a {\em novel case-based efficient
-morphological analyzer} that uses compositional 
-non-deterministic Automata driven by a case based controller.
-Each alive machine in Sarf represents a valid morphological analysis. 
+In this paper, we present Sarf\footnote{Sarf (name modified for blind review purposes) 
+will be available online as an open source tool.},
+ a {\em novel case-based efficient
+morphological analyzer} that uses 
+non-deterministic finite state Automata 
+driven by a case based controller.
+%Each alive machine in Sarf represents a valid morphological analysis.
 Sarf keeps alive all possible analyses of the text and gives 
@@
 
-Sarf uses a {\em recursive} concatenative analysis with a novel 
-refinement. 
+Sarf uses concatenative analysis with a novel 
+{\em recursive affixes} refinement. 
 Sarf introduces recursive affixes in order to
-retain better part of speech information and enhance the 
+better retain part of speech information and enhance the 
 efficiency of affix matching.
@@
 through a \RL{sanad} or a sequence of narrators. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
-transliteration and translation. 
-The authentication of a \noArRL{.hady_t} highly depends on the credibility
-of each of the narrators as reported in separate biography 
-books. 
+Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
+Arabic with its transliteration and translation. 
+The authentication of a \noArRL{.hady_t} depends on 
+the credibility of each of the narrators as reported in 
+separate biography books. 
 In this paper we consider the problem of automatically segmenting
-a \noArRL{.hady_t} book into narrations, then segmenting each narration into
-its content or \RL{matn} and its \noArRL{sanad}.
-We also partition the sanad accurately into the 
+a \noArRL{.hady_t} book into narrations, then segmenting each 
+narration into its content or \RL{matn} and its \noArRL{sanad}.
+We also partition the sanad into the 
 separate narrators so that we can later look each one of them 
 up in the biography books. 
-For example, the boxes in Figure~\ref{f:exhadith} are proper names and once
-connected toghether they form complex names of narrators. For example
-narrator $n_1$ has the first name \noTrRL{qtybT} and his father name is
+For example, the boxes in 
+Figure~\ref{f:exhadith} are proper names and once
+connected together they form complex names of narrators. 
+For example, narrator $n_1$ has the first name 
+\noTrRL{qtybT} and his father name is
 \noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.
@@
 three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
-and extracted from each one of them a vector of a three level deep
-complex structure. 
-We also showcase that the results of the accurate high level abstraction
-results of the case-based analysis was not affected by a major
-refinement to the analyzer that increased low level analysis accuracy.
+and extracted from each one of them a vector of complex structures 
+that are three level deep in hierarchy.
+%We also showcase that the results of the accurate high level 
+%abstraction
+%results of the case-based analysis was not affected by a major
+%refinement to the analyzer that increased low level analysis accuracy.
 
@@
 We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02}  and use define the affixes
+Buckwalter~\shortcite{Buckwalter:02} and define the affixes
 recursively. 
@@
 the controller and 
-the prefix, stem, and suffix deterministic tansducers. 
+the prefix, stem, and suffix deterministic transducers. 
 Sarf differs in that it uses recursive affixes in the 
@@
 at a finer granularity level. 
-Also in Sarf there is no need for a separate tokenizer such as
-TOKAN as each alive Automata carries with it the positions in 
-text where a composition decision was made. 
-The case-based controller can build the trace 
-when needed.
+In Sarf, there is no need for a separate tokenizer such as
+TOKAN since each solution keeps a stack of positions
+that partition text into morphemes.
+%The case-based controller can build the trace 
+%when needed.
 
@@
 
-We strongly feel that the refinement of SAMA~\cite{Maamouri:10} 
-is the closest to our approach. SAMA was refined to interact with
+Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
+SAMA was refined to interact with
 the ATB~\cite{Maamouri:04} project after the addition of a large 
-new corpus. We find supporting evidence in this decision to our hypothesis
-about the utility of a case-based analyzer. 
-The refinements that resulted in algorithmic changes in SAMA were 
+new corpus. 
+%We find supporting evidence in this decision to our hypothesis
+%about the utility of a case-based analyzer. 
+The algorithmic changes in SAMA were
 done manually and worked in integration with the ATB format. 
@@
 Sarf encodes the prefix, suffix and stem lexicons into
-three linear transducers and uses a non-deterministic composition
+three linear transducers and uses a non-deterministic construction
 of the three to compute all valid morphological analyses.
 We first present a running example of Sarf.
-The running example explains
-how the non-deterministic composition works and 
-illustrates how Sarf implements recursive affixes and
+The running example explains how Sarf works,
+implements recursive affixes, and
 handles ``run-on'' words.
@@
 \transfalse
-\begin{figure*}[tb]
+\begin{figure}[tb]
 \center{
-\resizebox{1.7\columnwidth}{!}
+\resizebox{\columnwidth}{!}
 { \input{figs/FSM.pdftex_t}}
@@
 }
-\end{figure*}
+\end{figure}
 % the FSMs for the three words
@@
 
+%{\bf Sarf example.}
+
 The diagram in Figure~\ref{f:example}
-illustrates the operation of Sarf when parsing the
+illustrates the operation of $\Phi$ the Sarf FSA 
+when parsing the
 string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
@@
 }
-The square and circle legends are states 
-and the directed edges are transitions driven by an input
-letter that matches the label of the edge. 
-The square denotes an accept state, 
-the circle denotes a regular state, and the reject states
-are omitted for clarity. 
-Figure~\ref{f:example}(a) represents the prefix transducer and 
-Figures~\ref{f:example}(b) and~\ref{f:example}(c) represent the 
-stem and suffix transducers respectively. 
+
+The box and circle denote accept and regular
+states respectively.
+The edges are transitions with labels corresponding to
+the input letters.
+The reject states are omitted for clarity and are denoted
+by the absence of corresponding transitions.
+
+Subfigures (a), (b), and (c) in Figure~\ref{f:example}
+represent ${\cal P}$ the prefix ,
+${\cal S}$ the stem, and ${\cal X}$ the suffix
+transducers respectively. 
 The symbol $\epsilon$ represents an empty string and is 
@@
 When there are two valid transitions such as \RL{w} 
-and $\epsilon$ in the start case, the non-deterministic 
-composition $\Phi$ spawns an exact copy of itself $\Psi$. 
+and $\epsilon$ in the start case, 
+$\Phi$ spawns an exact copy of itself $\Psi$. 
 $\Phi$ makes one transition, and $\Psi$ makes the other. 
 Each of $\Phi$ and $\Psi$ now represent a valid analysis so far. 
-When $\Phi$ or $\Psi$ is at a state where the next input letter
-leads to a reject state, it dies.
-The reject states are omitted from Figure~\ref{f:example}
-and are denoted by the absence of the corresponding 
-letters. 
-In our example, if we assume there were no stems that start 
-with the letter \RL{w}, $\Psi$ will die then. 
-In reality probably $\Psi$ will die when the input string
+An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
+In our example, if there were no stems that start 
+with the letter \RL{w}, $\Psi$ will die. 
+In reality, $\Psi$ will die when the input 
 is at \RL{wsyl`}.
 
-Note that the prefix and affix transducers allow recursive 
-prefixes. For example, \RL{w} will result in an accept state
-that transitions to the stem transducer. 
-When \RL{s} follows, we move to another accept state in the 
-prefix transducer corresponding to the suffix \RL{ws}. 
-The same applies to the suffix transducer.
+Note that the affix transducers allow recursive 
+affixes. 
+For example, \RL{w} will result in an accept state
+that transitions to the ${\cal S}$.
+When \RL{s} follows, we move to another accept state in 
+${\cal P}$ corresponding to the prefix \RL{ws}. 
+The same applies to ${\cal X}$.
 
 Lets consider $\Phi$ after it consumed \RL{wsy} 
-and transitioned into the stem transducer root node. 
+and transitioned into the root node of ${\cal S}$.
 Now $\Phi$ will transition with \RL{l`b} to reach an accept 
 state.
+
 Before moving with the letter \RL{b} to the accept state,
@@
 
-Since $\Phi$ is now in an accept state in the stem transducer, it will
-spawn $\Sigma$ that transitions into the root of the 
-suffix transducer. When \RL{h} is consumed, $\Phi$ will die since 
+Since $\Phi$ is now in an accept state in ${\cal S}$, it will
+spawn $\Xi$ that transitions into the root of ${\cal X}$. 
+When \RL{h} is consumed, $\Phi$ will die since 
 there is no edge from the current state labeled with \RL{h}. 
-$\Xi$ represents a valid analysis and can proceed.
-
-Note that $\Sigma$ reports a full word at any accept state
-by spawning a new composition using the $\epsilon$ transition
-to the prefix transducer.
+Now $\Xi$ represents a valid analysis and can proceed.
+
+Note that $\Xi$ reports a full word at any accept state
+by spawning a new FSA using the $\epsilon$ transition
+to ${\cal P}$.
 In Sarf, this only happens with white space, delimiters, 
@@
 
-\begin{figure*}[tb]
+\begin{figure}[tb]
 \center{
-\resizebox{1.8\columnwidth}{!}
+\resizebox{\columnwidth}{!}
 { \input{figs/abstract_machine.pdftex_t}}
-\caption{An abstraction of the non-deterministic construction of  
-    Sarf from the affix and stem transducers.}
+\caption{Sarf NDFSA.}
 \label{f:composition}
 }
-\end{figure*}
+\end{figure}
 
@@
 \begin{table}[tb]
-\centering
+%\centering
+%\resizebox{.9\columnwidth}{!}{
 \begin{tabular} {p{6cm}}
-\begin{Verbatim}[fontsize=\relsize{-1},
+\begin{Verbatim}[fontsize=\relsize{-2},
 frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
@@
 
+Intuitively, the controller reports a valid chain 
+of narrators when a sequence of names
+connected by narrator connectors appears. 
+It marks the beginning of the hadith with the beginning of 
+the current sequence,
+and marks the end of the hadith with the beginning of the 
+next sequence. 
+It also marks the chain of narrators as the sequence itself.
+
 The controller has four states that correspond to 
@@
 
-Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
-before it gives up on its expectations.
-
-We reach the NAME\_S state only when a
-valid NAME is detected and we leave when no more NAME's are detected. 
-The NRC\_S state can only be reached if an NRC is detected.
-NMC\_S can only be reached from a NAME\_S state.
-
-The controller reports a valid chain of narrators when a 
-sequence of names
-connected by narrator connectors appears. 
-It marks the beginning of the hadith with the beginning of 
-the current sequence,
-and marks the end of the hadith with the beginning of the 
-next sequence. 
-It also marks the chain of narrators as the sequence itself.
-
-The definition of input labels such as NAME and IBN depends on the 
-morphological analyzer. 
-However, our case-based controller approach was able to perform well
-under both Sarf and a refined version of Sarf.
+%The definition of input labels such as NAME and IBN depends on the 
+%morphological analyzer. 
+%However, our case-based controller approach was able to perform well
+%under both Sarf and a refined version of Sarf.
 
@@
 \end{tabular}
+}
 \normalsize
@@
 
-In section we present results that compare Sarf to existing 
+In this section we present results that compare Sarf to existing 
 morphological analyzers such as Buckwalter and ElixirFM. 
 We also evaluate our hypothesis of the utility of a case-based 
-morphological analyzer,
-by running our case study in two modes, simple and refined.
+morphological analyzer.
+We validate that the high-level analysis of the output 
+of low-level morphological analysis can compensate for accuracy.
+To achieve that we ran Sarf in two modes, simple and refined.
+In the simple mode, we used Sarf as discussed in 
+Section~\ref{sec:sarf} with the 
+controller of Section~\ref{sec:controller}.
+In the refined mode, we applied several low-level morphological 
+refinements to Sarf such as disambiguation of 
+female proper names so that the controller can ignore them. 
+We also added a refinement that detected compound proper names 
+such as \RL{`bd alr.hmn} 
+as they are common in narrator names.
+%Finally, we consider a possessive form as a reference to 
+%a person and treat
+%it as a narrator.
+
 We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
@@
 We implemented our case study with Buckwalter in two phases. 
-In phase one, we process the entire text with Buckwalter and
-obtain annotations to locations in text. 
-Then in the second phase we feed these annotations to 
-our controller as if they were coming from Sarf. 
-We measure the time it took to produce the annotations
+In phase one, we processed the entire text with Buckwalter and
+obtained annotations to text. 
+In the second phase, we fed the annotations to 
+the hadith controller as if they were coming from Sarf. 
+We measured the time it took to produce the annotations
 and the time it took the controller to process the 
-annotations whithout the time spent streaming the annotations.
+annotations without the time spent streaming the annotations.
 We used the same setup for ElixirFM.
 
-Sarf outperforms both Buckwalter and ElixirFM in runing time. 
+Sarf outperforms both Buckwalter and ElixirFM in terms of 
+running time. 
 It also produces higher accuracy measures. 
@@
 names by ignoring prefixes and suffixes that can not form
-names even if the stem was a name. 
+names even if one possible alternative of the stem is a name. 
 Doing the same with Buckwalter and ElixirFM requires  
-expanding the annotations resulting from one affix to reflect
-its multiple composing affixes.
-This is easily done with the way we defined affixes recursively.
-
-Buckwalter performed much better than ElixirFM. This is up to 
-our knowledge the first comparison of these two tools and 
-we think the big difference is due to the more exact 
-and grammar based functional morphological analysis ElixirFM 
-uses.
+expanding the annotations resulting from one simple affix to be 
+reflected in all the composite affixes that contain it.
+%expanding the annotations resulting from one affix to reflect
+%its multiple composing affixes.
+%However, this was easily performed using our recursive definition 
+%of the affixes.
+
+Buckwalter performed much better than ElixirFM in 
+terms of running time, 
+while ElixirFM outperformed Buckwalter in terms of accuracy.
+%This is up to 
+%our knowledge the first comparison of these two tools and 
+We think the big difference is due to the sophistication of 
+grammar based functional morphological analysis ElixirFM 
+uses and its reliance on additional annotations extracted from both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.
 
@@
 learning system that can learn more names. 
-We also plan on developing a domain specific language similar 
-to ElexirFM~\cite{Otakar:07}
-that allows users to easily specify their case-based controller
-to interact with Sarf.
+We also plan on developing an Arabic linguistic computational model 
+similar to that of ElixirFM~\cite{Otakar:07} but that can be 
+explored partially based on the case study.
 

### 2010-12-17 trunk/tex/sarf.aclhlt.tex
reduce lines

@@
 The symbol $\epsilon$ represents an empty string and is 
-the source of non-determinism in this example.
+the source of non-determinism.
 
@@
 to make its decision. 
-It does not produce any external output and therefore 
+It does not produce external output and therefore 
 the three transducers with the 
@@
 is that of the second morpheme in the pair. 
-
 Atomic affixes are standalone affixes
 that are not the concatenation of other affixes. 
-Formally, we define an affix $f$ to be either
+Formally, we define an affix $f: a | a f$ to be either
 an atomic affix $a$ or an atomic affix concatenated
-with an affix. We write $f: a | a f$. 
+with an affix. 
 %Sarf only keeps atomic affixes and their categories.
@@
 approach of Buckwalter where the analyzer considers
-all possible substrings of the word in question
-and looks it up in the affix tables.
+all possible substrings %of the word in question
+and looks them up in the affix tables.
 
@@
 a full word. 
-This happens when $c$ is a white space or a delimiter, 
-or when the last letter leading to the current state
+This happens when $c$ is a white space, a delimiter, 
+or when the preceding letter %leading to the current state
 was a non-connecting letter.
@@
 it spawns another machine $\Psi$ and adds it to $M$. 
+Notice that spawning a new machine requires only keeping
+the state of the machine and all machines share the
+structures of the three transducers.
 If $c$ leads to a reject state, then $\Phi$ dies 
@@
 \RL{'a_hbar} ``told'' or one of their derivations. 
-The IBN label corresponds to the word \RL{ibn} ``the son of'' and is a name connector.
-The NISBA label corresponds to an adjective that points to a person such 
+An IBN label denotes the word \RL{ibn} ``the son of'' and is a name connector.
+A NISBA label is a possessive adjective that qualifies a person such 
 as \RL{alma.sriy} ``the Egyptian''.
@@
 the hadith controller as if they were coming from Sarf. 
-We measured the time it took to produce the annotations
-and the time it took the controller to process the 
-annotations without the time spent streaming the annotations.
-We used the same setup for ElixirFM.
+We measured the time it took to accomplish both tasks
+less the time spent streaming the annotations.
+We did the same for ElixirFM.
 
@@
 running time. 
-It also produces higher accuracy measures. 
-We think this is mainly due to the extended
+It also produces higher accuracy measures 
+mainly due to the extended
 lexicon as well as to the fact that Sarf reduces false positive
-names by ignoring prefixes and suffixes that can not form
-names even if one possible alternative of the stem is a name. 
+names by ignoring affixes that can not form
+names even if one possible solution of the stem is a name. 
 Doing the same with Buckwalter and ElixirFM requires  
@@
 We think the big difference is due to the sophistication of 
-grammar based functional morphological analysis ElixirFM 
-uses and its reliance on additional annotations extracted from both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.
+the formal functional morphological analysis of ElixirFM 
+and its reliance on additional annotations extracted from 
+both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.
 
@@
 Arabic text. 
-Our results showed that Sarf with the hadith controller
+Results show that Sarf with the hadith controller
 is faster and more accurate than
@@
 biography books. 
-We plan to make use of the name connector and narrator
-connector stop phrases that our case study extracted to create a 
-learning system that can learn more names. 
-We also plan on developing an Arabic linguistic computational model 
-similar to that of ElixirFM~\cite{Otakar:07} but that can be 
-explored partially based on the case study.
+%We plan to make use of the name connector and narrator
+%connector stop phrases that our case study extracted to create a 
+%learning system that can learn more names. 
+%We also plan on developing an Arabic linguistic computational model 
+%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
+%explored partially based on the case study.
 

### 2010-12-17 trunk/tex/sarf.aclhlt.tex
at 8 pages

@@
 analysis to abstract and annotate text.
-Often times, and in particular in the context
-of the Arabic language, annotations resulting from 
-an ongoing branch of morphological analysis may not be
+Often times, 
+%      and in particular in the context of the Arabic language, 
+annotations resulting from 
+a branch of morphological analysis may not be
 appropriate for the case study at hand. 
 In this paper we present Sarf, a case based morphological 
-analyzer for Arabic text. 
+analyzer for Arabic. 
 Sarf allows a case-based controller to control and refine
@@
 
-Automated analysis of Arabic data sets, including texts, 
-publications, records and digital media is essential
-with the huge digital Arabic content available nowadays. 
+%Automated analysis of Arabic data sets, including texts, 
+%publications, records and digital media is essential
+%with the huge digital Arabic content available nowadays. 
 Morphological analysis is key in current automated 
@@
 use concatenative analysis
-to analyze an Arabic word and
-consider its internal structure composed of several
-{\em morphemes}. A morpheme can be a {\em stem}, or an {\em affix}.
+consider the internal structure 
+of an Arabic word and
+compose it into several {\em morphemes}. 
+A morpheme can be a {\em stem}, or an {\em affix}.
 An affix can be a {\em prefix, suffix, } or an {\em infix}.
@@
 This greatly increases the ambiguity of Arabic text. 
-Another difficulty is that  Arabic letters can have up to 
+Arabic letters can have up to 
 four different forms
@@
 as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
-without the need of a delimiter such as a space in between. 
-This happens because the first word \RL{il_A} ends with
+without the need of a space in between. 
+The reason is the first word \RL{il_A} ends with
 \RL{_A} a non-connecting letter. 
@@
 SAMA~\cite{Kulick:10},
-and ElixirFM~\cite{Otakar:07} exist.
-They take as input white space delimited tokens~\cite{Kulick:10},
+and ElixirFM~\cite{Otakar:07} 
+take as input white space delimited tokens~\cite{Kulick:10},
 consider them as words,
@@
 MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
-also use machine learning and support vector machines (SVM) 
+use machine learning and support vector machines (SVM) 
 to enhance the accuracy of the morphological analysis at the expense 
@@
 machine (FSM) based analyzers as described in~\cite{Beesley:03}.
-The efficiency of the resulting analyzer highly depends on the
+The efficiency of the resulting analyzer depends on the
 way the Xerox rules are written. 
 This requires deep knowledge and insight from the user
-in morphological analysis as well as skills in writing 
-context free grammar rules that can be automatically compiled
-into efficient non-deterministic finite state Automata (NDFSA).
+in compilation techniques, context free grammars, 
+and morphological analysis.
+%context free grammar rules that can be automatically compiled
+%into efficient non-deterministic finite state Automata (NDFSA).
 
@@
 For example, if a query concerns proper names and the 
-prefix in question in the analysis connects only to a verb, 
-we do not need to further analyze the rest of the word 
-to provide an answer.
-
-We find evidence to our hypothesis in~\cite{Maamouri:10} where 
-a new corpus was added to the Arabic Tree Bank 
-(ATB)~\cite{Maamouri:04}. 
-The addition required compliance with the DARPA GALE program 
+analyzer is considering a 
+prefix that connects only to a verb,
+the analyzer can answer without considering the 
+rest of the word.
+
+We find evidence to our hypothesis in~\cite{Maamouri:10}.
+The addition of a new corpus to the Arabic Tree Bank 
+(ATB)~\cite{Maamouri:04}
+required compliance with the DARPA GALE program 
 including several NLP tasks such as English translation 
 annotations and word-level alignment of Arabic and English. 
-The new genre of data challenged the existing data processing and 
+The new data challenged the existing processing and 
 annotation guidelines. 
-Alternative guidelines came to place where the high level NLP 
+This lef to refining 
+SAMA~\cite{Kulick:10} with 
+alternative guidelines where the NLP 
 task interacted with the morphological analyzer.
-This led to refining SAMA~\cite{Kulick:10}.
 
@@
 
-\subsection{Sarf}
-\label{sec:intro:sarf}
-
-In this paper, we present Sarf\footnote{Sarf (name modified for blind review purposes) 
+%\subsection{Sarf}
+%\label{sec:intro:sarf}
+
+{\bf Sarf.~~~}
+In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
 will be available online as an open source tool.},
@@
 non-deterministic finite state Automata 
-driven by a case based controller.
-%Each alive machine in Sarf represents a valid morphological analysis.
-Sarf keeps alive all possible analyses of the text and gives 
-opportunity to the controller to intervene at every single input 
-letter. 
-The decisions of the case-based controller can thus prune false 
-positives early in the run. 
-Each alive machine in Sarf takes as input one letter at a time 
-from a text stream. 
+driven by a case-based controller.
+Each alive machine in Sarf represents a valid morphological analysis
+and takes as input one letter at a time.
+The case-based controller prunes false positives
+early in the run by making a decision at every input letter
+to control the alive machines.
+%Each alive machine in Sarf takes as input one letter at a time 
+%from a text stream. 
 Sarf does not assume that the word at hand is a token and
@@
 Arabic with its transliteration and translation. 
-The authentication of a \noArRL{.hady_t} depends on 
-the credibility of each of the narrators as reported in 
+The authenticity of a \noArRL{.hady_t} depends on 
+the credibility of the narrators as reported in 
 separate biography books. 
-In this paper we consider the problem of automatically segmenting
-a \noArRL{.hady_t} book into narrations, then segmenting each 
-narration into its content or \RL{matn} and its \noArRL{sanad}.
-We also partition the sanad into the 
-separate narrators so that we can later look each one of them 
+In this paper, we consider the problem of automatically segmenting
+a \noArRL{.hady_t} book into narrations, each 
+narration into its \RL{matn} (content) and its \noArRL{sanad},
+and each sanad into the 
+separate narrators. Now, we can look each narrator 
 up in the biography books. 
-For example, the boxes in 
-Figure~\ref{f:exhadith} are proper names and once
-connected together they form complex names of narrators. 
-For example, narrator $n_1$ has the first name 
-\noTrRL{qtybT} and his father name is
-\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.
+For example, 
+Figure~\ref{f:exhadith} shows proper names in boxes connected
+to form complex names of narrators. 
+For example, 
+\noTrRL{qtybT} is the first name 
+of narrator $n_1$, and 
+\noTrRL{s`yd} is the name of his father as 
+the word \noTrRL{bn} (son of) indicates.
 
 The hadith case study is interesting to Sarf since the controller
-considers most of the analyses where we have a concentration
-of proper names in the text. It ignores the text where
-proper names occur rarely.% as long 
-%as the text is valid.
-We are also interested in discovering a subset of words 
-that relate persons to each other such as \RL{bin} or ``the son of''
-or words that mean narrate such as \RL{qaal} or ``said''. 
+considers %most of the analyses where we have a concentration
+proper names and ignores text where
+proper names occur rarely.% as long %as the text is valid.
+We are also interested in discovering a words 
+that relate names to each other such as \RL{bin} or ``the son of''
+and words that mean narrate such as \RL{qaal} or ``said''.
+
 With Sarf, we successfully automated the analysis of 
@@
 %refinement to the analyzer that increased low level analysis accuracy. 
-
-%\subsection{Contributions}
-
 We make the following key contributions. 
@@
 \item {\bf Case-based analysis:}  We designed and implemented
-a case-based morphological analyzer where a case-based controller
-machine controls and guides the analysis. 
+a case-based morphological analyzer where a controller
+guides the analysis. 
 %\item {\bf Exhaustive analysis:} Since we have a case-based 
@@
 We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02} and define the affixes
+Buckwalter~\shortcite{Buckwalter:02} and define affixes
 recursively. 
 This allows Sarf to perform better and 
-maintain compositional part of speech tags.
+maintain compositional part of speech (POS) tags.
 \item {\bf Hadith case study:}
-We evaluated Sarf using the hadith case study. We were able to
-efficiently extract the sequence of narrators into a complex
-data structure with three levels of hierarchy with high accuracy. 
+%We evaluated Sarf using the hadith case study. 
+We were able to efficiently extract a complex three 
+level hierarchical 
+structure containing the chains of narrators 
+with high accuracy. 
 \end{enumerate}
@@
 The rest of this paper is structured as follows. In Section~\ref{sec:related}
-we compare Sarf to related work. Then in Section~\ref{sec:sarf}
-we discuss Sarf and illustrate the way it works with a running example.
-We also explain how non-deterministic finite state Automata
-work. In Section~\ref{sec:islamic} we present the hadith case
+we compare Sarf to related work. In Section~\ref{sec:sarf}
+we discuss Sarf and illustrate it with a running example.
+%We also explain how non-deterministic finite state Automata
+%work. 
+In Section~\ref{sec:islamic} we present the hadith case
 study. We present our results in Section~\ref{sec:results}
@@
 
-The survey in~\cite{Sughaiyer:04} discusses and compares
+The survey in~\cite{Sughaiyer:04} compares
 several morphological analyzers. 
 Analyzers such as~\cite{Khoja:01,Darwish:02} 
-target specific applications in the morphological 
-analyzer itself or use a specific set of part of speech tags
+target specific applications in the 
+analyzer itself or use a specific set of POS tags
 as their reference.
-Sarf differs in that it is a general and rich morphological 
+Sarf differs in that it is a general morphological 
 analyzer that reports all possible solutions. 
-It is case-based in the sense that a controller that drives
-the morphological analyzer prunes these decisions.
+It is case-based in the sense that a controller 
+prunes these solutions.
 
@@
 This allows Sarf to better maintain and propagate 
-the part of speech information
-tags associated with each prefix and suffix.
-This information is highly 
-redundant with Buckwalter.% and that may lead to consistency issues. 
+the POS tags associated with affixes. 
+%This information is highly 
+%redundant with Buckwalter.% and that may lead to consistency issues. 
 Sarf keeps all the possible analyses alive where each analysis
@@
 
-Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
-finite-state Automata (FSA). 
-Beesley reports that the number of the Automata generated by a compiler
-from Xerox rules can not be controlled and also reports a difficulty to 
-compose the FSAs into a single FSA framework. 
-Sarf solves the problem with an FSA that is constructed from 
-the controller and 
-the prefix, stem, and suffix deterministic transducers. 
+Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
+finite state Automata (FSA). 
+It reports that the number of Automata generated by a compiler
+for Xerox rules can not be controlled and also reports a difficulty 
+to compose the FSAs into a single framework. 
+Sarf constructs an NDFSA from
+the controller and the prefix, stem, and suffix deterministic 
+transducers to solve the problem.
 Sarf differs in that it uses recursive affixes in the 
 affix transducers, and in that the transducers consider
-input from the controller. 
-With Sarf it is easy to control the FSA with a custom user case 
-controller. Up to our understanding, Beesley  
-requires recompiling the Xerox rules.
-
-Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
+input from the controller to modify their behavior. 
+%With Sarf it is easy to control the FSA with a case based
+%controller. 
+Beesley requires recompilation.% of the Xerox rules.
+
+Similar to The MADA+TOKAN toolkit, Sarf thinks of
 the several valid morphological analyses as a richness that 
@@
 with any case study. 
-In Sarf, the analyzer remains general, and the 
-case-based controller specializes the analysis.
+%In Sarf, the analyzer remains general, and the 
+%case-based controller specializes the analysis.
 
@@
 a language independent SVM based analysis. 
-While the SVM analysis learns features from the text passed to the case
-study, it does not make use of the type of the query the 
-case study targets in the text.
+The SVM analysis learns features from the input text 
+and does not make use of any information 
+in the target query.
 
@@
 Sarf encodes the prefix, suffix and stem lexicons into
-three linear transducers and uses a non-deterministic construction
+three deterministic transducers and 
+uses a non-deterministic construction
 of the three to compute all valid morphological analyses.
@@
 The diagram in Figure~\ref{f:example}
-illustrates the operation of $\Phi$ the Sarf FSA 
+illustrates the operation of $\Phi$, the Sarf FSA,
 when parsing the
@@
 
-The box and circle denote accept and regular
+Boxes and circles denote accept and regular
 states respectively.
@@
 ${\cal P}$ which is an accept state. 
-The $\epsilon$-edge connects the ${\cal P}$
+The $\epsilon$-edge connects ${\cal P}$
 to ${\cal S}$ and transitions from any accept state
 in ${\cal P}$ to the root state in ${\cal S}$.
-The $\epsilon$-edge that connects the ${\cal S}$ to ${\cal X}$
+The $\epsilon$-edge that connects ${\cal S}$ to ${\cal X}$
 follows the same behavior.
@@
 $\Phi$ makes one transition, and $\Psi$ makes the other. 
-Each of $\Phi$ and $\Psi$ now represent a valid analysis so far. 
+Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
 An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
@@
 For example, \RL{w} will result in an accept state
-that transitions to the ${\cal S}$.
+that transitions to ${\cal S}$.
 When \RL{s} follows, we move to another accept state in 
@@
 with the prefix \RL{wsy}. 
-It does that by associating a category value for each accept state. 
-Sarf considers the value of the category as part of the state.
-Thus each state shown in Figure~\ref{f:example} actually represents
-more than one state. 
+Sarf keeps a compatibility category values as part
+of the accept states. 
+%Sarf considers the value of the category as part of the state.
+Thus each accecpt state in Figure~\ref{f:example} represents
+more than one concrete state. 
 If the category of \RL{l`b} is compatible with the category of
@@
 Since $\Phi$ is now in an accept state in ${\cal S}$, it will
-spawn $\Xi$ that transitions into the root of ${\cal X}$. 
-When \RL{h} is consumed, $\Phi$ will die since 
-there is no edge from the current state labeled with \RL{h}. 
-Now $\Xi$ represents a valid analysis and can proceed.
+spawn $\Xi$. $\Xi$ transitions to the root of ${\cal X}$. 
+$\Phi$ die when it consumes \RL{h} since
+no edge from the current state is labeled with \RL{h}. 
+$\Xi$ represents a valid analysis and can proceed.
 
@@
 is not interesting. 
-It may also decide to correct the analysis of one FSA
+It may also decide to correct one FSA
 based on the decision of the others. 
@@
 A compatibility rule is a pair of categories 
-$\langle c_1, c_2\rangle$  stating that morphemes
-with category $c_1$ can be concatenated with morphemes
-of category $c_2$. 
+$\langle c_1, c_2\rangle$  stating that $c_1$ morphemes
+can be concatenated with $c_2$ morphemes.
 Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
@@
 Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
-a prefix, $\beta$ is a stem, and $\gamma$ is a suffix. 
-The $\alpha\beta\gamma$ partition of $s$ is a 
-valid morphological analysis if and only if 
+a prefix, $\beta$ is a stem, and $\gamma$ is a suffix;
+$\alpha\beta\gamma$ is a 
+valid morphological analysis iff
 $\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
@@
 namely \RL{w}, \RL{s}, and \RL{y}.
-The part of speech (POS) tag for \RL{wsy} is a concatenation
+The POS tag for \RL{wsy} is a concatenation
 of the POS tags of each of the three morphemes. 
@@
 compatibility rules respectively.
-This allows us to keep a shorter list of affixes. 
+%This allows us to keep a shorter list of affixes. 
 We also introduce a {\em resulting category}
 for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
-The resulting category in all practical cases
-is that of the second morpheme in the pair. 
+%The resulting category in all practical cases
+%is that of the second morpheme in the pair.
+
 Atomic affixes are standalone affixes
@@
 into directed acyclic graphs (DAG) similar to 
-those in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
-This transducer is deterministic and thus 
-its traversal is linear.
+those in Figures~\ref{f:example}(a) and~\ref{f:example}(c).
+The transducers are deterministic and thus 
+their traversal is linear.
 This is computationally superior to the 
@@
 %\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
-\resizebox{1.8\columnwidth}{!}{
+\resizebox{1.9\columnwidth}{!}{
 \begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
@@
 
-\subsection{Comparison with Buckwalter and ElixirFM}
+\subsection{Buckwalter and ElixirFM}
 

### 2010-12-17 tex/sarf.aclhlt.tex
spell checker

@@
 For example, if a query concerns proper names and the 
-prefix in question in the analysis connects only to a verb, 
-we do not need to further analyze the rest of the word 
-to provide an answer.
-
-We find evidence to our hypothesis in~\cite{Maamouri:10} where 
-a new corpus was added to the Arabic Tree Bank 
-(ATB)~\cite{Maamouri:04}. 
-The addition required compliance with the DARPA GALE program 
+analyzer is considering a 
+prefix that connects only to a verb,
+the analyzer can answer without considering the 
+rest of the word.
+
+We find evidence to our hypothesis in~\cite{Maamouri:10}.
+The addition of a new corpus to the Arabic Tree Bank 
+(ATB)~\cite{Maamouri:04}
+required compliance with the DARPA GALE program 
 including several NLP tasks such as English translation 
 annotations and word-level alignment of Arabic and English. 
-The new genre of data challenged the existing data processing and 
+The new data challenged the existing processing and 
 annotation guidelines. 
-Alternative guidelines came to place where the high level NLP 
+This led to refining 
+SAMA~\cite{Kulick:10} with 
+alternative guidelines where the NLP 
 task interacted with the morphological analyzer.
-This led to refining SAMA~\cite{Kulick:10}.
 
@@
 
-\subsection{Sarf}
-\label{sec:intro:sarf}
-
-In this paper, we present Sarf, a {\em novel case-based efficient
+%\subsection{Sarf}
+%\label{sec:intro:sarf}
+
+{\bf Sarf.~~~}
+In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
+will be available online as an open source tool.},
+ a {\em novel case-based efficient
 morphological analyzer} that uses 
 non-deterministic finite state Automata 
-driven by a case based controller.
-%Each alive machine in Sarf represents a valid morphological analysis.
-Sarf keeps alive all possible analyses of the text and gives 
-opportunity to the controller to intervene at every single input 
-letter. 
-The decisions of the case-based controller can thus prune false 
-positives early in the run. 
-Each alive machine in Sarf takes as input one letter at a time 
-from a text stream. 
+driven by a case-based controller.
+Each alive machine in Sarf represents a valid morphological analysis
+and takes as input one letter at a time.
+The case-based controller prunes false positives
+early in the run by making a decision at every input letter
+to control the alive machines.
+%Each alive machine in Sarf takes as input one letter at a time 
+%from a text stream. 
 Sarf does not assume that the word at hand is a token and
@@
 
-Sarf uses 
-concatenative analysis with a novel 
-{\em recursive} 
-refinement. 
+Sarf uses concatenative analysis with a novel 
+{\em recursive affixes} refinement. 
 Sarf introduces recursive affixes in order to
@@
 Arabic with its transliteration and translation. 
-The authentication of a \noArRL{.hady_t} depends on 
-the credibility of each of the narrators as reported in 
+The authenticity of a \noArRL{.hady_t} depends on 
+the credibility of the narrators as reported in 
 separate biography books. 
-In this paper we consider the problem of automatically segmenting
-a \noArRL{.hady_t} book into narrations, then segmenting each 
-narration into its content or \RL{matn} and its \noArRL{sanad}.
-We also partition the sanad into the 
-separate narrators so that we can later look each one of them 
+In this paper, we consider the problem of automatically segmenting
+a \noArRL{.hady_t} book into narrations, each 
+narration into its \RL{matn} (content) and its \noArRL{sanad},
+and each sanad into the 
+separate narrators. Now, we can look each narrator 
 up in the biography books. 
-For example, the boxes in 
-Figure~\ref{f:exhadith} are proper names and once
-connected toghether they form complex names of narrators. 
-For example, narrator $n_1$ has the first name 
-\noTrRL{qtybT} and his father name is
-\noTrRL{s`yd} as the word in between \noTrRL{bn} (son of) indicates.
+For example, 
+Figure~\ref{f:exhadith} shows proper names in boxes connected
+to form complex names of narrators. 
+For example, 
+\noTrRL{qtybT} is the first name 
+of narrator $n_1$, and 
+\noTrRL{s`yd} is the name of his father as 
+the word \noTrRL{bn} (son of) indicates.
 
 The hadith case study is interesting to Sarf since the controller
-considers most of the analyses where we have a concentration
-of proper names in the text. It ignores the text where
-proper names occur rarely.% as long 
-%as the text is valid.
-We are also interested in discovering a subset of words 
-that relate persons to each other such as \RL{bin} or ``the son of''
-or words that mean narrate such as \RL{qaal} or ``said''. 
+considers %most of the analyses where we have a concentration
+proper names and ignores text where
+proper names occur rarely.% as long %as the text is valid.
+We are also interested in discovering a words 
+that relate names to each other such as \RL{bin} or ``the son of''
+and words that mean narrate such as \RL{qaal} or ``said''.
+
 With Sarf, we successfully automated the analysis of 
@@
 
-Sarf is similar to Beesley~\shortcite{Beesley:01} in that it uses
-finite-state Automata (FSA). 
-Beesley reports that the number of the Automata generated by a compiler
-from Xerox rules can not be controlled and also reports a difficulty to 
-compose the FSAs into a single FSA framework. 
-Sarf solves the problem with an FSA that is constructed from 
-the controller and 
-the prefix, stem, and suffix deterministic tansducers. 
+Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
+finite state Automata (FSA). 
+It reports that the number of Automata generated by a compiler
+for Xerox rules can not be controlled and also reports a difficulty 
+to compose the FSAs into a single framework. 
+Sarf constructs an NDFSA from
+the controller and the prefix, stem, and suffix deterministic 
+transducers to solve the problem.
 Sarf differs in that it uses recursive affixes in the 
 affix transducers, and in that the transducers consider
-input from the controller. 
-With Sarf it is easy to control the FSA with a custom user case 
-controller. Up to our understanding, Beesley  
-requires recompiling the Xerox rules.
-
-Sarf is similar to The MADA+TOKAN toolkit in that it thinks of
+input from the controller to modify their behavior. 
+%With Sarf it is easy to control the FSA with a case based
+%controller. 
+Beesley requires recompilation.% of the Xerox rules.
+
+Similar to The MADA+TOKAN toolkit, Sarf thinks of
 the several valid morphological analyses as a richness that 
@@
 The running example explains how Sarf works,
-implememts recursive affixes, and
+implements recursive affixes, and
 handles ``run-on'' words.
@@
 The reject states are omitted for clarity and are denoted
-by the abscence of corresponding transitions.
+by the absence of corresponding transitions.
 
@@
 with the prefix \RL{wsy}. 
-It does that by associating a category value for each accept state. 
-Sarf considers the value of the category as part of the state.
-Thus each state shown in Figure~\ref{f:example} actually represents
-more than one state. 
+Sarf keeps a compatibility category values as part
+of the accept states. 
+%Sarf considers the value of the category as part of the state.
+Thus each accept state in Figure~\ref{f:example} represents
+more than one concrete state. 
 If the category of \RL{l`b} is compatible with the category of
@@
 compatibility rules respectively.
-This allows us to keep a shorter list of affixes. 
+%This allows us to keep a shorter list of affixes. 
 We also introduce a {\em resulting category}
 for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
-The resulting category in all practical cases
-is that of the second morpheme in the pair.
+%The resulting category in all practical cases
+%is that of the second morpheme in the pair.
 
@@
 that are not the concatenation of other affixes. 
-Formally, we define an affix $f$ to be either
+Formally, we define an affix $f: a | a f$ to be either
 an atomic affix $a$ or an atomic affix concatenated
-with an affix. We write $f: a | a f$. 
+with an affix. 
 %Sarf only keeps atomic affixes and their categories.
@@
 into directed acyclic graphs (DAG) similar to 
-those in Figure~\ref{f:example}(a) and Figure~\ref{f:example}(c).
-This transducer is deterministic and thus 
-its traversal is linear.
+those in Figures~\ref{f:example}(a) and~\ref{f:example}(c).
+The transducers are deterministic and thus 
+their traversal is linear.
 This is computationally superior to the 
 approach of Buckwalter where the analyzer considers
-all possible substrings of the word in question
-and looks it up in the affix tables.
+all possible substrings %of the word in question
+and looks them up in the affix tables.
 
@@
 of low-level morphological analysis can compensate for accuracy.
-To acheive that we ran Sarf in two modes, simple and refined.
+To achieve that we ran Sarf in two modes, simple and refined.
 In the simple mode, we used Sarf as discussed in 
@@
 the hadith controller as if they were coming from Sarf. 
-We measured the time it took to produce the annotations
-and the time it took the controller to process the 
-annotations whithout the time spent streaming the annotations.
-We used the same setup for ElixirFM.
+We measured the time it took to accomplish both tasks
+less the time spent streaming the annotations.
+We did the same for ElixirFM.
 
@@
 %our knowledge the first comparison of these two tools and 
-We think the big difference is due to the sofistication of 
-grammar based functional morphological analysis ElixirFM 
-uses and its reliance on additional annotations extracted from both ATB~\cite{Maamouri:04} and the Pragbank~\cite{Prague04}.
+We think the big difference is due to the sophistication of 
+the formal functional morphological analysis of ElixirFM 
+and its reliance on additional annotations extracted from 
+both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.
 
@@
 biography books. 
-We plan to make use of the name connector and narrator
-connector stop phrases that our case study extracted to create a 
-learning system that can learn more names. 
-We also plan on developing an Arabic linguistic computational model 
-similar to that of ElixirFM~\cite{Otakar:07} but that can be 
-explored partialy based on the case study.
+%We plan to make use of the name connector and narrator
+%connector stop phrases that our case study extracted to create a 
+%learning system that can learn more names. 
+%We also plan on developing an Arabic linguistic computational model 
+%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
+%explored partially based on the case study.
 

### 2010-12-17 tex/sarf.aclhlt.tex
clarify alive, recursive, and controller early on

@@
 appropriate for the case study at hand. 
-In this paper we present Sarf, a case based morphological 
+In this paper we present Sarf, a case-based morphological 
 analyzer for Arabic. 
@@
 to analyze a stream of text. 
-Each alive machine corresponds to a valid analysis. 
+%Each machine corresponds to a valid analysis. 
 Sarf uses concatenative analysis based on recursive affixes 
@@
 driven by a case-based controller.
-Each alive machine in Sarf represents a valid morphological analysis
+Each machine in Sarf represents a valid morphological analysis
 and takes as input one letter at a time.
@@
 early in the run by making a decision at every input letter
-to control the alive machines.
+to control the machines.
+The controller we present in Section~\ref{sec:controller}
+is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
+it can be a logical entity of any type.
+
 %Each alive machine in Sarf takes as input one letter at a time 
@@
 {\em recursive affixes} refinement. 
+An atomic affix $a$ is a standalone affix
+that is not the concatenation of other affixes. 
+We define an affix $f: a | a f$ to be either
+an atomic affix $a$ or an atomic affix concatenated
+with an affix. 
 Sarf introduces recursive affixes in order to
@@
 input from the controller to modify their behavior. 
-%With Sarf it is easy to control the FSA with a case based
+%With Sarf it is easy to control the FSA with a case-based
 %controller. 
@@
 
-Since $\Phi$ is now in an accept state in ${\cal S}$, it will
-spawn $\Xi$. $\Xi$ transitions to the root of ${\cal X}$. 
-$\Phi$ die when it consumes \RL{h} since
-no edge from the current state is labeled with \RL{h}. 
-$\Xi$ represents a valid analysis and can proceed.
+Since $\Phi$ is now in an accept state in ${\cal S}$, it 
+spawns $\Xi$ which transitions to the root of ${\cal X}$. 
+After it consumes \RL{h--} $\Phi$ dies since there is no
+\noTrRL{h--}-edge from the current state.
+$\Xi$ represents a valid analysis and proceeds.
 
@@
 
-Atomic affixes are standalone affixes
-that are not the concatenation of other affixes. 
-Formally, we define an affix $f: a | a f$ to be either
-an atomic affix $a$ or an atomic affix concatenated
-with an affix. 
 %Sarf only keeps atomic affixes and their categories.
@@
 information. 
-
 We hypothesize that a higher level of abstraction such as 
-intersecting the narrators from the hadith books with the narrators 
-in the biographies can reduce the false positives in the chains
-in this case.
-
-\subsection{Buckwalter and ElixirFM}
-
+intersecting narrators from the hadith books with those from
+the biographies can reduce the false positives in the chains.
+
+%\subsection{Buckwalter and ElixirFM}
+{\bf Buckwalter and ElixirFM.~~~}
 We implemented our case study with Buckwalter in two phases. 
@@
 names even if one possible solution of the stem is a name. 
-Doing the same with Buckwalter and ElixirFM requires  
-expanding the annotations resulting from one simple affix to be 
-reflected in all the composite affixes that contain it.
+%Doing the same with Buckwalter and ElixirFM requires  
+%expanding the annotations resulting from one simple affix to be 
+%reflected in all the composite affixes that contain it.
 %expanding the annotations resulting from one affix to reflect

### 2010-12-19 tex/rp.tex
merge and rehab feedback

@@
 the Arabic user.
-While a non-Arabic user is interested in machine translation to 
-help him understand an Arabic document, 
+While machine translation is of general interest, 
 an Arabic user is more interested in 
-information hard to discover manually such as 
-the relation between clinical records, pharmaceutical 
-prescriptions and sales, and a specific sickness. 
-Another Arabic user may be interested in looking at the relation
-between stock prices and political news or the 
-relation between suspect profiles in interrogation reports 
+information hard to discover manually.
+Examples are the relation between clinical records, pharmaceutical 
+prescriptions and sales, and a specific sickness,
+the relation
+between stock prices and political news, or the 
+relation between interrogation reports with suspects
 and investigative reports from the crime scenes.
 
-We propose to build a 
+We propose to build a framework for
 %{\em relational text mining 
 %of Arabic using morphological and case-based analysis}. 
-{\em case-based Arabic text analysis using relations and morphology}(CATARM) framework. 
-CATARM will take as input sets of documents, each of a specific
-type, and provides the user with a visual query language that
-allows him to build a flow of Arabic text mining
-modules and discover desired relations between documents. 
-CATARM will translate the query into a netlist of computational 
-components, it will also extract parameters from the query and 
-the document sets and use them to simplify the computational 
+{\em case-based Arabic text analysis using relations and morphology} (CATARM). 
+CATARM will apply novel algorithms and tools
+to answer queries seeking relations amongst
+Arabic text documents.
+CATARM takes as input several sets of documents, each of a specific
+type, and a user query. % expressed in a visual language.
+The query, expressed in a visual language, allows the user 
+to build a flow of Arabic text mining
+components and discover desired relations between documents. 
+CATARM compiles the query into a netlist of computational 
+components. It also uses features extracted from the query and 
+the document sets to simplify the computational 
 components. 
@@
 
-%IR tools are similar to a Google search 
-%engine where documents of interest are selected based on the 
-%relevance to a set of keywords of interest to the user.
-%NLP targets the automation of understanding human languages.
-%This task is the oldest and most difficult task in the artificial 
-%intelligence domain.
-%The main difficulty behind NLP comes from the ambiguity of 
-%natural languages~\cite{Osm08}.
-%While NLP techniques cannot deliver their final target yet, they 
-%can deliver analysis of sentences into nouns, verbs, and adjectives.
-%They can also narrow the meaning of a word or a phrase based on 
-%the context to one of its possible meanings.
-%Finally they can parse a sentence into a relation between the 
-%discovered entities using abstractions such as verb-name 
-%abstractions~\cite{Osm08}.
-
-%Information extraction (IE) targets forming data structures and 
-%instances of these data structures out of a collection of documents.
-%IE tries to fit the output of IR and NLP to templates of 
-%interest to the user.
-%Data mining can be applied at the end on the output of the 
-%IE process to answer user queries about the input 
-%documents~\cite{JHa05}.
-
-%\subsection{Arabic text mining in academy }
-%In the following we review the major works on Arabic text mining 
-%in academy and industry.
+\novocalize
+%The accuracy of the solutions suffer due to inherent difficulties
+%of morphological analysis of the Arabic language. 
+It is common practice to write Arabic text
+without short vowels. 
+This greatly increases the ambiguity of Arabic text. 
+Arabic letters can have up to 
+four different forms
+corresponding to their position in a word, i.e. beginning,
+middle, end of word and separate forms. 
+This allows the phrase
+\noTrRL{il_A\nospace almdrsT}  with no delimiter within
+to be visually recognizable
+as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
+This happens because the first word \RL{il_A} ends with
+\RL{_A} a non-connecting letter. 
+These words occur often in text and greatly increase the
+difficulty of tokenization and are referred to as 
+``run-on'' words~\cite{Buckwalter:04}.
+Such difficulties are inherent to the 
+morphological analysis of the Arabic language and
+render the solutions inaccurate.
 
@@
 
-%The accuracy of the solutions suffer due to inherent difficulties
-%of morphological analysis of the Arabic language. 
-It is common practice to write Arabic text
-without short vowels. 
-This greatly increases the ambiguity of Arabic text. 
-Arabic letters can have up to 
-four different forms
-corresponding to their position in a word, i.e, beginning,
-middle, end of word and separate forms. 
-This allows the phrase
-\noTrRL{il_A\nospace almdrsT}  with no delimiter within
-to be visually recognizable
-as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school).
-This happens because the first word \RL{almdrsT} ends with
-\RL{T} a non-connecting letter. 
-These words occur often in text and greatly increase the
-difficulty of tokenization and are referred to as 
-``run-on'' words~\cite{Buckwalter:04}.
-Such difficulties are inherrent to the 
-morphological analysis of the Arabic language and
-render the solutions inaccurate.
-
 Current morphological analyzers such as 
 Buckwalter~\cite{Buckwalter:02},
-Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
-and ElixirFM~\cite{Otakar:07} exist.
-These analyzers are used in several open source spell checkers as 
+Beesley~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+and ElixirFM~\cite{Otakar:07} exist
+and are used in several open source spell checkers as 
 well as NLP frameworks~\cite{Col09}.
@@
 prefix in question in the analysis connects only to a verb, 
-we do not need to further analyze the rest of the word 
-to provide an answer.
-
-We find evidence to our hypothesis in~\cite{Maamouri:10}. The 
-addition of a new corpus to the Arabic Tree Bank~\cite{Maamouri:04}
-required the high NLP task to interact with the low 
-level morphological analyzer.
-This led to a refined analyzer~\cite{Kulick:10}.  
-We also find evidence in~\cite{Habash:06} that different types of 
-morphological analyses behave differently on the same case study. 
-We strongly believe and we find evidence in our preliminary
-results that a 
-case-based morphological analyzer that allows a high level 
-case-based controller to intervene at every decision to 
-guide, use, prune, and refine the morphological analysis
-is of high utility.
+we no further need to analyze the rest of the word.
+
+We find evidence to our hypothesis in~\cite{Maamouri:10} where 
+a new corpus was added to the Arabic Tree Bank 
+(ATB)~\cite{Maamouri:04}. 
+The addition required compliance with the DARPA GALE program 
+including several NLP tasks such as English translation 
+annotations and word-level alignment of Arabic and English. 
+The new genre of data challenged the existing data processing and 
+annotation guidelines. 
+Alternative guidelines came to place where the high level NLP 
+task interacted with the morphological analyzer. 
+This led to refining SAMA~\cite{Kulick:10}. 
+The work of Habash~\cite{Habash:06} reports that different 
+types of morphological analyses behave differently on the same case 
+study. 
+For one, the analyzers may find different solutions. 
+The analyzers may present the solutions in different orders. 
+The presented solutions and the order of presentation may not be 
+appropriate for the case study. 
+We find strong evidence in our preliminary results 
+that a case-based morphological analyzer that allows a case-based 
+controller to intervene at every decision to guide, use, order, and 
+refine the morphological analysis is of high utility.
 
@@
 the Arabic language as a similarity measurement.
-This work is limited to the names of the fields in the data 
-structures extracted from Arabic texts.
-
-{\bf Hypothesis 2.} This approach does not tend directly to 
+%This work is limited to the names of the fields in the data 
+%structures extracted from Arabic texts.
+
+{\bf Hypothesis 2.} 
+The preprocessing approach does not tend directly to 
 the Arabic users who write their 
@@
 We hypothesize that Arabic queries use rules and 
-relations similar to the Arabic text under investigation, 
-and thus techniques native to the Arabic language will do better 
+look for relations similar in nature to the structure of the Arabic language 
+and the text under investigation.
+Thus techniques native to the Arabic language will do better 
 in Arabic text analysis.
 
-We find evidence in the work of the CADIM \noTrRL{q-adim} 
-project~\cite{Col09} at the University of Columbia to support our 
+The work of the CADIM \noTrRL{q-adim} 
+project~\cite{Col09} at the University of Columbia supports our 
 hypothesis. They observe that Arabic texts used as training 
 data for common NLP techniques proved ineffective. 
-They built semi-automatic models to address the problem of automatic 
-speech recognition in Arabic.
+They built semi-automatic models to address the problem of
+Arabic speech recognition.
 
 %\subsection{\bf Arabic text mining in industry}
-Few companies in the industry provide primitive Arabic text 
+Few companies in the industry provide Arabic text 
 analysis tools.
 Sakhr~\cite{Sak09} provides a categorization 
-tool with a limited number of predefined categories, and a summary 
-generator that selects sentences from text.
+tool with a limited number of predefined categories.
+Sakhr also provides a summary 
+generator that selects important sentences from text.
 Basis technology~\cite{Bas09} provides Rosette, a lexical 
 analysis tool that normalizes Arabic text as a pre-processing step 
-for indexers and other general purpose data mining 
+for general purpose data mining 
 tools.  Basis also provides REX; an entity extractor.
@@
 ElixirFM~\cite{Otakar:07} uses an {\em inferential-realizational}
-approach for morphological analysis. 
-Realizational morphology considers the form of the word 
+approach for morphological analysis
+and considers the form of the word 
 \noTrRL{.sarf}  instead of its segments. 
-Formal rules and formulas express the relations between 
+Formal rules express the relations between 
 the root of the lexeme and the fully inflected word. 
-While this is highly compatible with the Arabic linguistic 
-description~\cite{Badawi:04},
-and it benefits from features unique to the Arabic language
-captured by the formal rules, 
-it is computationally expensive and enumerates all the possible
-solutions.
-
-{\bf Goals.} Up to our knowledge, CATARM will be the 
-first framework that tends to relational Arabic morphology 
-and text mining. 
-{\bf (1.)} CATARM will build a novel parametrized linguistic 
-computational model of Arabic. 
-The model will be implemented using efficient structures. 
- {\bf (2.)} It will be amenable to Arabic specific simplifications 
-inferred from the correspondence between the fromal Arabic rules
-and the relational user queries.
-{\bf (3.)} The model will also be amenable to simplifications 
-inferred from statistical features extracted from the document 
-sets using language independent techniques such as latent semantic 
-indexing~\cite{LSI89} and hidden Markov models (HMM).
+While the rules capture features unique to the 
+the Arabic language and its linguistic description~\cite{Badawi:04},
+the approach is computationally expensive and enumerates 
+all the possible solutions.
+
+{\bf Goals.} 
+CATARM tends to relational Arabic text mining using 
+{\bf (1.)} a novel case-based morphological analyzer.
+{\bf (2.)} CATARM will build a novel and efficient 
+linguistic computational model (LCM) of Arabic. 
+%The LCM will be implemented using efficient structures. 
+The LCM will be amenable to {\bf (3.)} Arabic specific simplifications 
+inferred from the correspondence between the formal Arabic rules
+and the relational user queries, and 
+{\bf (4.)} to simplifications 
+inferred from statistical features extracted from the documents 
+using language independent techniques such as 
+hidden Markov models (HMM)  and 
+latent semantic indexing~\cite{LSI89}.
 %CATARM will augment current digital dictionaries to resolve 
 %entity semantics and targeted qualificationsl.
-{\bf (4.)} CATARM extracts semantic graphs from each text 
-document and then uses set arithmetic and graph analysis techniques
-to establish relations between these graphs. {\bf (5.)} CATARM will 
-solve at least three interesting relational Arabic text case studies 
-presented in Section~\ref{s:design:casestudies}.
+{\bf (5.)} CATARM extracts semantic graphs 
+and uses set arithmetic and graph analysis techniques
+to establish relations between the graphs. {\bf (5.)} CATARM will 
+solve the case studies 
+presented in Sections~\ref{s:design:lit}--\ref{s:design:med}.
 
@@
 records and find relations between sicknesses and prescribed medication. 
-Finally, we will build a system using CATARM to check the strength
-of the relations between internet documents viewed by a minor and
-a set of Arabic text documents containing unrated or unwanted 
-material. If the relation is strong, the system will flag access to
-the document. 
-All components and parts built during the development of solutions
-to the case studies will be generalized and added to CATARM as 
-library computational components. 
+Finally, we will use CATARM to provide content-based parental control
+for Arabic text documents. We will look for 
+relations between documents requested by a minor and
+a set of undesired Arabic documents. 
+If the relation is strong, we will flag access to
+the documents.
+
+We further explain our approach with examples and discuss 
+how we will obtain the documents 
+in Sections~\ref{s:design:lit}--\ref{s:design:med}.
+We will generalize all components built during the development 
+of the case studies and we will add them to 
+the libraries of CATARM. 
+%library computational components. 
 \end{itemize}
@@
 For example, extracting a relation between two criminals may 
-uncover a third criminal who is ready to commit a crime.
-Finding inconsistencies in claims may curb down fraud and corruption 
+uncover a third criminal ready to commit a crime.
+Finding inconsistencies in claims may curb down fraud 
 in the insurance sector.
@@
 
+\transfalse
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.8\columnwidth}{!}
+{ \input{figs/exhadith.pdftex_t}}
+\caption{Hadith abstraction example.}
+\label{f:exhadith}
+}
+\end{figure}
+\transtrue
+
 The PI and his team at the American University of Beirut (AUB)
@@
 through a \RL{sanad} or a sequence of narrators. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in Arabic with its 
+Figure~\ref{f:exhadith} shows an example~\noArRL{.hady_t} in Arabic with its 
 transliteration and translation. 
@@
 \noTrRL{bn} (son of) indicates.
-
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{.7\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
-\label{f:exhadith}
-}
-\end{figure}
-\transtrue
 
@@
 specify his queries as one or more FSM
-controlling the morphological analyser.
-We explain this approach in details in Section~\ref{s:design:fsm}.
+controlling the morphological analyzer.
+We explain this approach in details in 
+Section~\ref{s:design:method}.
 ATSarf along with other computational components, 
@@
 \center{
-\resizebox{.75\columnwidth}{!}
-{\includegraphics{figs/narrator_chain_output.pdf}
-\label{f:narrators} } 
+\resizebox{.9\columnwidth}{!}
+{\includegraphics{figs/narrator_chain_output.pdf}}
 \caption{Directed acyclic graph representing a partial order 
     relation between narrators extracted using ATSarf}
+\label{f:narrators} 
 }
@@
 the time they lived in. 
-Then One can compute a time and location overlap
+Then one can compute a time and location overlap
 check to discover inconsistent narrations.
@@
 
-The Co-Lead PI published extensively in this domain. 
-A selected list of her recent publications follows. 
-\begin{itemize}\itemsep0pt
-\item Rehab Duwairi, Mohammad Al-Refai, Natheer Khasawneh, 
-"Feature Reduction Techniques for Arabic Text Categorization". 
-Journal of the American Society for Information Science and Technology 
-(JASIST), Volume 60, Issue 11, pages: 2347-2352, 2009. 
-\item Rehab M. Duwairi, "Arabic Text Categorization", 
-International Arab Journal for Information Technology (IAJIT), 
-Volume 4, Number 2, pages 125  132, 2007.
-\item Rehab M. Duwairi, "Machine Learning for Arabic Text Categorization". 
-Journal of the American Society for Information Science and Technology 
-(JASIST), Vol. 57, Issue 8, pages 1005-1010, 2006.
-\end{itemize}
-
-%The Lead PI worked in the area of software arabization since 1996 
-%as he worked on developing Arabization modules for Windows NT and 
-%an 
-%Arabic string manipulation library~\cite{Zar96}. 
-%The Lead PI is also involved in working on an open source 
-%project that aims at comprehensive automation of authenticity 
-%checks against a corpus of literature heritage~\cite{Zaraket06}.
-%The Lead PI is leading an effort at AUB to introduce 
-%Computational Arabic in the curriculum of the ECE program. 
-%He is also leading a team of students in an effort to place 
-%the infrastructure needed for this project to start. 
-%The lead PI worked and published about relational logic and 
-%logic solvers applied to structured languages as follows.
-%\begin{itemize}
-%\item Sequential circuits for relational analysis. F. Zaraket, A. Aziz, and S. Khurshid. International Conference on Software Engineering, Minneapolis MN, 2007. 
-%\item Sequential circuits for program analysis. F. Zaraket, A. Aziz, and S. Khurshid. Automated Software Engineering, Atlanta GA, 2007. 
-%\end{itemize}
+\subsection{Related work of principal investigators}
+
+The Co-LPI, Rehab Duwairi, has several years of experience 
+working on data mining for Arabic documents. 
+She initially targeted Arabic text categorization where a 
+data set of 1,000 documents was used to train and test 
+classifiers~\cite{Duw06,Duw07}.
+Word morphology comes to the surface when extracting the 
+feature vectors of documents. 
+In that work, words were stemmed before using them as features. 
+Also, word frequency and inverted term frequency were used to 
+quantify the importance of features.
+
+In a recent work, Duwairi, Refae and Khasawneh~\cite{Duw09}
+have addressed text categorization again. 
+This research investigated possible word forms that can be 
+used to create feature vectors of documents in particular: 
+original word, stem of the word, light-stem (shallow stemming) 
+of the word, and word augmented with a list of synonyms. 
+Therefore, every document is represented using four different 
+feature vectors, the result revealed that light-stemming is the 
+best representation as it preserves the meaning of the words 
+and yet reduces the dimensions of feature vectors. 
+The data set size was 15,000 documents that belong to three 
+categories.
+
+The Co-Lead PI published extensively in the 
+domain~\cite{Duw06, Duw07, Duw09}.
+In recent research efforts, the CO-LPI is investigating 
+keyphrase extraction fro Arabic documents.
+
+The Lead PI worked in the area of software Arabization since 1996 
+as he worked on developing Arabization modules for Windows NT 
+and an Arabic string manipulation library~\cite{Zar96}. 
+The Lead PI is also involved in working on an open source 
+project that aims at comprehensive automation of authenticity 
+checks against a corpus of literature heritage~\cite{Zar06}.
+The Lead PI is leading an effort at AUB to introduce 
+Computational Arabic in the curriculum of the ECE program. 
+He is also leading a team of students in an effort to place 
+the infrastructure needed for this project to start.
+
+The lead PI holds several approved patents and has several 
+patents pending approval in the area of logic verification 
+and synthesis.
+The lead PI worked and published about relational logic and 
+logic solvers applied to structured languages~\cite{seraICSE07,Zara05}
 
@@
 \label{s:designmethods}
+
+The specific aims of this project are:
+\begin{itemize}
+\item Build the CATARM framework while solving four interesting Arabic 
+text minining case studies
+  \begin{itemize}
+  \item Hadith literature case study
+  \item Security case study
+  \item Network content case study
+  \item Health case study
+  \end{itemize}
+\item Build and enhance a case-based morphological analyzer
+\item Build an Arabic linguistic computational model (LCM)
+\item Build entity and relation extractors
+\item Build a query language 
+  \begin{itemize}
+  \item Build optimization modules that use the query to reduce the LCM
+  \item Build a query solver that compiles all the components and run them 
+  \end{itemize}
+\item Build language independent LCM optimization modules
+\item Add statistical, graph, and text manipulation algorithms
+\item Build a visualization tool to showcase the results 
+\end{itemize}
+
+In the rest of this section, we will introduce Arabic text mining in 
+details, then we will discuss the case studies in 
+Sections~\ref{s:design:lit}--~\ref{s:design:med}. 
+We will discuss the morphological analyzer in Section~\ref{s:design:ma}, 
+the query language in Section~\ref{s:design:query}, and the linguistic
+computational model in Section~\ref{s:design:lcm}.
+In Section~\ref{s:design:method}, we will discuss our methodology
+to build the components, optimize and reduce the LCM, build the
+entity and relation extractors, solve the query, and visualize the 
+results. 
+Section~\ref{s:design:plan} will details the work plan and
+justify the proposed budget.
+
+\subsection{Arabic text mining}
+\label{s:design:atm}
 
@@
 \center{
-\resizebox{.75\columnwidth}{!}
+\resizebox{.9\columnwidth}{!}
 {\includegraphics{figs/literature_case-study1.pdf}
-\label{f:literature}
 } 
 \caption{CATARM diagram for the literature case study }
+\label{f:literature}
 }
@@
 \center{
-\resizebox{.8\columnwidth}{!}
+\resizebox{.9\columnwidth}{!}
 {\includegraphics{figs/security_case-study.pdf}
-\label{f:security}
 }
 \caption{CATARM diagram for the security case study}
+\label{f:security}
 }
@@
 
-An MA with date, place, and name extraction controller extracts
+A morphological analyzer with date, place, and name extraction controller extracts
 names, dates and places from the investigation reports. 
@@
 
-\subsection{CATRAM components}
+As such reports contain sensitive information and our contacts
+in the Lebanese security declined direct access, 
+we will start the security case study by studying dummy documents that 
+represent both interrogative and investigative types of documents. 
+We collected a small amount (in the order of tens)
+of similar documents published in the news in the last two years
+from cases that were brought to public court or from high profile
+cases that leaked to the media.
+We will work on augmenting this collection further to have a large
+enough set of reports (in the order of hundreds). 
+We will mutate the entities in these documents and seed the mutated
+set with relations we wish to discover and that are hard to discover
+manually. 
+Later on, we will use our contacts with the Lebanese and Qatari 
+security authorities
+to validate our system on their real data.
+
+\subsection{Network content case study}
+\label{s:design:net}
+
+Parental controls help parents allow their kids to benefit from the internet 
+without fearing the danger of violent and obscene material. 
+These controls partially rely on processing the content to detect and rate 
+the target material. 
+Control mechanisms for Arabic traffic as currently mostly manual. 
+In this case study we will detect the bad content. 
+Given a set of undesired internet content 
+and a set of requests by internet users, 
+CATARM can discover whether there is a relation between 
+the requested documents and the undesired content and flag the 
+requested document with strong enough relation.
+
+\subsection{Health case study}
+\label{s:design:med}
+
+Clinical reports contain diagnosis information, observations about 
+patient progress, prescriptions, and patient response to prescribed 
+medications. 
+In countries, like Lebanon, where medical doctors graduate and specialize 
+in several distinct practices around the world it is interesting to 
+check for common practices and emerging trends to gain insight
+about local and regional health and diagnosis features. 
+It is also interesting to relate 
+pharmaceutical 
+records to clinical reports to check on 
+market consistency and on self prescribed medicine 
+practices.
+Such a study can uncover serious tendencies in uncontrolled
+pharmaceutical markets.
+It can uncover what kind of sicknesses and what kind of medicines
+are used sensitive medicines are used without 
+appropriate medical supervision.
+
+We will work with our contacts at the American University
+Hospital (AUH) at Beirut to analyze this data and relate it
+to nearby pharmaceutical records.
+
+\subsection{CATARM components}
 \label{s:design:components}
@@
 \subsection{Case-based morphological analyzer}
+\label{s:design:ma}
 
@@
 such as Buckwalter~\cite{Buckwalter:02},
-Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+Beesley~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
 and ElixirFM~\cite{Otakar:07} take an Arabic word as input
-and produce several valid morpholgical solution for the word. 
+and produce several valid morphological solutions for the word. 
 Each solution is a sequence of morphemes that form the word.
@@
 \center{
-\resizebox{.4\columnwidth}{!}
+\resizebox{.5\columnwidth}{!}
 {\input{figs/hadith.pdftex_t}
-\label{f:statemachine}
 }
 \caption{State-machine used to detect the chain of narrators}
+\label{f:statemachine}
 }
@@
 computational model of the Arabic language.
-The models is a set of formulas and rules. 
+The model is a set of formulas and rules. 
 ElixirFM uses the computational model in a functional
@@
 \center{
-\resizebox{.8\columnwidth}{!}
+\resizebox{.9\columnwidth}{!}
 {\includegraphics{figs/detailed_submachine.pdf}
-\label{f:chain}
 }
 \caption{CATARM detailed diagram for the chain of narrators extractor}
+\label{f:chain}
 }
@@
 We will work on developing the necessary components to complete 
-the solution for the hadith literature and the security case 
-studies.
-We will also work on the health and network traffic case studies.
+the solution for the hadith literature  case study. 
+We will also work on the security, health and network traffic case studies.
 While working on the case studies, we will add the components
@@
 components.
-
-{\bf Development methodology.}
-We will integrate all the mentioned techniques in a scalable 
-framework for Arabic text analysis.
-The framework as a whole will allow for exhausting Arabic text 
-analysis and allow for extensions of the framework.
-A user should be able to iteratively and repetitively apply the 
-abode techniques on his input text.
-
-We will follow an eXtreme Programming 
-methodology in the process of building CATARM.
-We will use open source tools, libraries, and frameworks to 
-build our libraries.
-We will build on existing open source programming and text 
-mining tools as well as open source Arabic text analysis tools 
-based on NLP such as MADA+TOKAN~\cite{Rot08}, 
-NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
-
-We will use a dokuwiki based website~\cite{Dok09} 
-to document the progress online and will use 
-subversion~\cite{Sub09} as a source control tool.
 
@@
 Once funded, the team can be directly formed and work can start
-with productive team members.
+with readily productive team members.
 
 \subsection{Work plan}
+\label{s:design:plan}
+
+%\begin{table}[bt]
+\begin{table}[bt]
+\centering
+\caption{Three year quarter based work plan.}
+%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
+%\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
+\small
+\begin{tabular}{lp{.1cm}ccccp{.1cm}ccccp{.1cm}cccc} \\
+& & \multicolumn{4}{c}{Year 1} & & \multicolumn{4}{c}{Year 2} & & \multicolumn{4}{c}{Year 3} \\ \cline{3-6} \cline{8-11} \cline{13-16} %\\
+    %\cline{2-13} \\ 
+& & Q1 & Q2 & Q3 & Q4 & & Q1 & Q2 & Q3 & Q4 & & Q1 & Q2 & Q3 & Q4 \\ \bottomrule
+
+Case based & & 
+W & W & W & E & & E & E & E & E & & E & E & E & E \\ 
+morphological analyzer & & 
+ &  & D & DM & &   & M & M & M & & M & M & M & M \\  \hline
+
+\multirow{2}{*}{Arabic LCM} & & 
+P & W & W & W & & W & E & E & E & & E & E & E & E \\ 
+& & 
+ &  &  &  & & D & DM & M & M & & M & M & M & M \\ \hline
+
+\multirow{2}{*}{Query language} & & 
+P & P & W & W & & W & W & E & E & & E & E & E & E \\ 
+& & 
+  &  &  &  & &  & D & DM & M & & M & M & M & M \\ \hline
+
+Language & & 
+P & P & P & W & & W & W & W & W & & E & W & W & W \\ 
+optimizations & & 
+ &  &  &  & &  &  &  & D & & DM & M & M & D \\ \hline
+
+Data  & & 
+P & P & P & W & & W & W & W & W & & E & W & W & W \\ 
+optimizations & & 
+ &  &  &  & &  &  &  & D & & DM & M & M & D \\ \hline
+
+\multirow{2}{*}{E/R extractors} & & 
+W & W & W &  & & E & E & E & W & & W & W &  & E \\ 
+& & 
+ &  & D & DM & & M & M & M & M & & M & D & DM & M \\ \hline
+
+Statistical & & 
+W & W & W & E & & WE & WE & WE & E & & WE & WE & WE & WE \\ 
+components & & 
+ &  & D & DM & & M & M & M & DM & & M & M & M & M \\ \hline
+
+Graph  & & 
+W & W & W & E & & WE & WE & WE & E & & WE & WE & WE & WE \\ 
+components & & 
+ &  & D & DM & & M & M & M & DM & & M & M & M & M \\ \hline
+
+\multirow{2}{*}{Query solver} & & 
+P & P & W & W & & W & W & W & E & & E & E & E & E \\ 
+& & 
+ &  &  &  & &  &  & D & DM & & M & M & DM & M \\ \hline
+
+\multirow{2}{*}{Case studies} & & 
+W & W & W & W & & W & W & W & W & & W & W & WE & E \\ 
+& & 
+ &  & D$^1$ &  & &  &  & D$^2$ &  & &  & D$^3$ & D$^4$ M & M \\ \hline
+
+\multirow{2}{*}{Visualization} & & 
+P & W & W & W & & E & E & E & E & & E & E & E & E \\ 
+& & 
+ &  &  & D & & M & M & M & M & & M & M & M & M \\ \hline
+
+\multirow{2}{*}{Integration} & & 
+P & W & W & W & & W & W & W & W & & W & W & W & E \\ 
+& & 
+ &  &  & D & & D & D & D & D & & D & D & D & M \\  \hline
+
+Validation & & 
+W & W & W & W & & W & W & W & W & & W & W & W & W \\  \bottomrule
+        %\cline{2-13} \\
+\end{tabular}
+\normalsize
+\label{t:workplan}
+\end{table}
+
+We will follow an eXtreme Programming 
+methodology in the process of building CATARM.
+This means that we will define refined incremental items of 
+work that allows for more frequent deliverables. 
+This also means that the code base will always be subject
+to maintenance and refactoring to enhance reuse and 
+increase productivity. 
+While a item is delivered, it will still be subject to work,
+enhancements, and maintenance.
+
+We detailed an aggressive work plan in Table~\ref{t:workplan}. 
+The plan spans over three years and projects a timeline
+of work per quarter. 
+The row headers describe work items, and the column headers
+describe quarters in the corresponding year of the project.
+The legends {\bf P, W, D, M, } 
+and {\bf E} 
+describe the status of the items 
+and stand for 
+{\bf P:} {\em planning phase}, 
+{\bf W:} {\em work in progress}, 
+{\bf D:} {\em deliverable}, 
+{\bf M:}  {\em maintenance}, 
+and 
+{\bf E: } {\em enhance} respectively.
+
+The $1,2,3,$ and $4$
+superscripts of the deliverables in the 
+in case studies row refer to the four 
+case studies described in the proposal.
+
+We will use open source tools, libraries, and frameworks to 
+build our libraries.
+We will build on existing open source programming and text 
+mining tools as well as open source Arabic text analysis tools 
+based on NLP such as MADA+TOKAN~\cite{Rot08}, 
+NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
+
+We will use a dokuwiki based website~\cite{Dok09} 
+to document the progress online and will use 
+subversion~\cite{Sub09} as a source control tool.
 
@@
 \item Two month salary per year for the LPI
-\item One course buy out per year for the Co-LPI
+\item One course buyout per year for the Co-LPI
 \item Two weeks consultant for the second and the third years
@@
 
-We will evaluate CATARM against at least three of the case studies
-that we discussed previously and that are listed below. 
+We will evaluate CATARM against the case studies
+that we discussed in Section~\ref{s:design} and 
+that are listed below. 
 \begin{enumerate}
@@
 
-Once completed the Arabic text analysis framework will be 
+If granted, this project will initiate a long term
+research relationship between Qatar University
+and the American University of Beirut. 
+Once completed, the Arabic text analysis framework will be 
 an open source contribution and developers from the Arabic 
@@
 efforts to keep it going.
-Furthermore, once completed the Arabic text analysis framework 
+Furthermore, the Arabic text analysis framework 
 can sustain itself by charging on support or expert use.
@@
 
-Completion of the above goals would result in a significant contribution to the research community of NLP in general and Arabic Analysis in particular.
-However, as already noted, text mining is a wide problem that cannot be expected to saturate in the near future especially when it comes to the arabic language where much work is lacking.
-Consequently, we look forward for additional contributions to the field such as the following:
+The completion of the CATARM framework will result in a 
+significant contribution to the research community of 
+NLP in general and Arabic Analysis in particular. 
+The CATARM framework will act as vehicle to enhance and advance 
+research in the area. 
+We look forward for additional contributions to the field 
+such as the following.
 \begin{itemize}
-\item Enhance morphological analysis: It is always the case, that the lexicon upon which the morphological analyzer can be augmented with useful categorization and entity tags that reflect semantical meanings.
-These modifications would give CATARM additional capabilities in terms of allowing users to extract and use a greater variety of entities when building instructive queries.
-The feedback of experts in the field of arabic linguistics can continually add accuracy to the morphological solutions.
-\item Look for other optimizations: From our priliminery research and results we were able to envision some optimizations to current tools; that's why it is expected that as we develop the complete framework to stumble at much more ideas and suggestions for improving the text mining process we are conducting expecially in terms of efficency and accuracy.
-\item Generalize the framework to other Arabic-like languages: Some sematic languages share many common features to the Arabic language such as the Amharic, Hebrew, etc...
-or even are written using the Arabic script like Persian, Kurdish, Urdu, Pashto.
-Accordingly, one may consider porting our techniques to them, or even generalizing the framework to support for languages other than Arabic.
-\item Generalize the case-based approach: ??
-\item Integrate other text mining techniques such as vocalization.
-Vocalization refers to the process of adding diacritics to an unvocalized text.
-The prerequisites of a successful vocalizer are the accuracy of the morphological analyzer in addition to grammatical analyzers.
-Such tools are especially useful in the academic sector where fully vocalized texts may be of high utility.
-\item Maintenance: We should never forget that for any software to remain useful, it must be continuously monitored to correct bugs, improve its interface, and enrich its capabilities.
-That's why CATARM should undergo extensive testing, fixes, and enhancements even after launching.
+\item We will explore developing and integrating vocalization
+techniques. 
+Vocalization techniques assign diacritics to unvocalized text. 
+The prerequisites of a successful vocalizer are the accuracy of 
+the morphological analyzer in addition to grammatical analyzers. 
+\item We will explore the use of CATARM with more case studies
+such as market analysis and opinion mining.
+\item We will explore augmenting the lexicon of the
+morphological analyzer with useful categorization and entity 
+tags that reflect semantical meanings. 
+These modifications would give CATARM additional capabilities in 
+terms of allowing users to extract and use a greater variety of 
+entities when building queries. 
+The feedback of experts in the field of Arabic linguistics can 
+continually add accuracy to the morphological solutions.  
+\item We will explore further optimization techniques
+that can help simplify and the analysis. 
+\item We will explore generalizing our results to languages 
+that resemble the Arabic language. 
+Some Semitic languages, such as the Amharic, Cyrillic, and Hebrew,
+share common features with the Arabic 
+language. Some other languages are written using the 
+Arabic script like Persian, Kurdish, Urdu, and Pashto. 
+\item We will also investigate the case-based approach with
+Latin languages and check how that compares with current techniques
+in terms of accuracy and efficiency.
+\item We will maintain CATARM and provide support to users and 
+researchers of CATARM.
 \end{itemize}

### 2010-12-19 trunk/tex/rp.tex
comments and add rb

@@
 the Arabic user.
-While machine translation is of gneral interest, 
+While machine translation is of general interest, 
 an Arabic user is more interested in 
 information hard to discover manually.
-Examples are {the relation between clinical records, pharmaceutical 
+Examples are the relation between clinical records, pharmaceutical 
 prescriptions and sales, and a specific sickness,
@@
 %of Arabic using morphological and case-based analysis}. 
-{\em case-based Arabic text analysis using relations and morphology}(CATARM). 
+{\em case-based Arabic text analysis using relations and morphology} (CATARM). 
+CATARM will apply novel algorithms and tools
+to answer queries seeking relations amongst
+Arabic text documents.
 CATARM takes as input several sets of documents, each of a specific
@@
 four different forms
-corresponding to their position in a word, i.e, beginning,
+corresponding to their position in a word, i.e. beginning,
 middle, end of word and separate forms. 
@@
 This led to refining SAMA~\cite{Kulick:10}. 
-The work of Habash~\shortcite{Habash:06} reports that different 
+The work of Habash~\cite{Habash:06} reports that different 
 types of morphological analyses behave differently on the same case 
@@
 
-{\bf Hypothesis 2.} This approach does not tend directly to 
+{\bf Hypothesis 2.} 
+The preprocessing approach does not tend directly to 
 the Arabic users who write their 
@@
 We hypothesize that Arabic queries use rules and 
-relations similar to the Arabic text under investigation, 
-and thus techniques native to the Arabic language will do better 
+look for relations similar in nature to the structure of the Arabic language 
+and the text under investigation.
+Thus techniques native to the Arabic language will do better 
 in Arabic text analysis.
 
-We find evidence in the work of the CADIM \noTrRL{q-adim} 
-project~\cite{Col09} at the University of Columbia to support our 
+The work of the CADIM \noTrRL{q-adim} 
+project~\cite{Col09} at the University of Columbia supports our 
 hypothesis. They observe that Arabic texts used as training 
 data for common NLP techniques proved ineffective. 
-They built semi-automatic models to address the problem of automatic 
-speech recognition in Arabic.
+They built semi-automatic models to address the problem of
+Arabic speech recognition.
 
@@
 ElixirFM~\cite{Otakar:07} uses an {\em inferential-realizational}
-approach for morphological analysis. 
-Realizational morphology considers the form of the word 
+approach for morphological analysis
+and considers the form of the word 
 \noTrRL{.sarf}  instead of its segments. 
-Formal rules and formulas express the relations between 
+Formal rules express the relations between 
 the root of the lexeme and the fully inflected word. 
-While this is highly compatible with the Arabic linguistic 
-description~\cite{Badawi:04},
-and it benefits from features unique to the Arabic language
-captured by the formal rules, 
-it is computationally expensive and enumerates all the possible
-solutions.
-
-{\bf Goals.} CATARM 
-tends to relational Arabic text mining using morphological analysis.
-{\bf (1.)} CATARM will build a novel parametrized 
-linguistic computational model (LCM) of Arabic. 
-The LCM will be implemented using efficient structures. 
- {\bf (2.)} The LCM will be amenable to Arabic specific simplifications 
+While the rules capture features unique to the 
+the Arabic language and its linguistic description~\cite{Badawi:04},
+the approach is computationally expensive and enumerates 
+all the possible solutions.
+
+{\bf Goals.} 
+CATARM tends to relational Arabic text mining using 
+{\bf (1.)} a novel case-based morphological analyzer.
+{\bf (2.)} CATARM will build a novel and efficient 
+linguistic computational model (LCM) for Arabic. 
+%The LCM will be implemented using efficient structures. 
+The LCM will be amenable to novel 
+{\bf (3.)} Arabic specific simplifications 
 inferred from the correspondence between the formal Arabic rules
-and the relational user queries.
-{\bf (3.)} The LCM will be amenable to simplifications 
-inferred from statistical features extracted from the document 
-sets using language independent techniques such as 
-hidden Markov models (HMM)  and 
-latent semantic indexing~\cite{LSI89}.
+and the user queries, and 
+{\bf (4.)} language independent simplifications
+inferred from statistical features %extracted from the documents 
+using language independent techniques such as 
+hidden Markov models (HMM)~\cite{JanHMM06}
+and latent semantic indexing (LSI)~\cite{LSI89}.
 %CATARM will augment current digital dictionaries to resolve 
 %entity semantics and targeted qualificationsl.
-{\bf (4.)} CATARM extracts semantic graphs from text 
-documents and uses set arithmetic and graph analysis techniques
-to establish relations between the graphs. {\bf (5.)} CATARM will 
-solve at least three interesting case studies such as those
-presented in Sections~\ref{s:design:lit} and~\ref{s:design:sec}.
+{\bf (5.)} CATARM extracts semantic graphs 
+and uses set arithmetic and graph analysis techniques
+to establish relations between the graphs. 
+{\bf (6.)} CATARM will solve the case studies 
+presented in Sections~\ref{s:design:lit}--\ref{s:design:med}.
 
@@
 records and find relations between sicknesses and prescribed medication. 
-Finally, we will build a system using CATARM to check the strength
-of the relations between Internet documents viewed by a minor and
-a set of Arabic text documents containing unrated or unwanted 
-material. If the relation is strong, the system will flag access to
-the document. 
-All components and parts built during the development of solutions
-to the case studies will be generalized and added to CATARM as 
-library computational components. 
+Finally, we will use CATARM to provide content-based parental control
+for Arabic text documents. We will look for 
+relations between documents requested by a minor and
+a set of undesired Arabic documents. 
+If the relation is strong, we will flag access to
+the documents.
+
+We further explain our approach with examples and discuss 
+how we will obtain the documents 
+in Sections~\ref{s:design:lit}--\ref{s:design:med}.
+We will generalize all components built during the development 
+of the case studies and we will add them to 
+the libraries of CATARM. 
+%library computational components. 
 \end{itemize}
@@
 are hard to discover by humans and will have direct live and
-cost saving benefits in cases similar to the above mentioned 
-case studies. 
+cost saving benefits.% in cases similar to the above mentioned 
+%case studies. 
 CATARM will qualitatively enhance research in the cultural and 
@@
 markets prior to making important decisions.
+CATARM may be used to flag spammers 
+posting biased reviews in Arabic web forums.
 
@@
 \label{s:prelim}
+
+\transfalse
+%\begin{figure}[tb]
+\begin{figure}
+\center{
+\resizebox{.8\columnwidth}{!}
+{ \input{figs/exhadith.pdftex_t}}
+\caption{Hadith abstraction example.}
+\label{f:exhadith}
+}
+\end{figure}
+\transtrue
 
@@
 
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{.7\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
-\label{f:exhadith}
-}
-\end{figure}
-\transtrue
-
 The framework now comprises a state of the art
@@
 \center{
-\resizebox{.75\columnwidth}{!}
+\resizebox{.9\columnwidth}{!}
 {\includegraphics{figs/narrator_chain_output.pdf}}
@@
 The directed acyclic graph (DAG) 
-in Figure~\ref{f:narrators} shows a partial order relation (POR) between
-narrators.
-We automatically extracted the POR from hadith text 
-documents.
+in Figure~\ref{f:narrators} shows a partial order relation (POR) 
+between narrators.
+We automatically extracted the POR from three books of hadith 
+selected arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}
+\footnote{We obtained
+  the digitized books from online sources such as 
+  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
+  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
 The nodes in boxes are the \RL{matn} of the hadith, 
@@
 
-\subsection{PI related work}
+\subsection{Related work of principal investigators}
 
@@
 She initially targeted Arabic text categorization where a 
-data set of 1000 documents where used to train and test 
+data set of 1,000 documents was used to train and test 
 classifiers~\cite{Duw06,Duw07}.
@@
 
-The Lead PI worked in the area of software arabization since 1996 
+The Lead PI worked in the area of software Arabization since 1996 
 as he worked on developing Arabization modules for Windows NT 
@@
 project that aims at comprehensive automation of authenticity 
-checks against a corpus of literature heritage~\cite{Zar06}.
+checks against a corpus of hadith literature~\cite{Zar06}.
 The Lead PI is leading an effort at AUB to introduce 
@@
 
-\subsection{Literature authentication case study}
+We describe CATARM using four case studies and then move
+to describe the different tools needed to build
+CATARM.
+
+\subsection{Hadith authentication case study}
 \label{s:design:lit}
@@
 \center{
-\resizebox{.75\columnwidth}{!}
+\resizebox{.9\columnwidth}{!}
 {\includegraphics{figs/literature_case-study1.pdf}
 } 
-\caption{CATARM diagram for the literature case study }
+\caption{CATARM diagram for the hadith case study }
 \label{f:literature}
@@
 The diagram in Figure~\ref{f:literature} describes our 
-high level vision for what the literature case study
+high level vision for what the hadith case study
 will look like in the CATARM framework. 
@@
 \center{
-\resizebox{.8\columnwidth}{!}
+\resizebox{.9\columnwidth}{!}
 {\includegraphics{figs/security_case-study.pdf}
@@
 
+As such reports contain sensitive information and our contacts
+in the Lebanese security declined direct access, 
+we will start the security case study by studying dummy documents that 
+represent both interrogative and investigative types of documents. 
+We collected a small amount (in the order of tens)
+of similar documents published in the news in the last two years
+from cases that were brought to public court or from high profile
+cases that leaked to the media.
+We will work on augmenting this collection further to have a large
+enough set of reports (in the order of hundreds). 
+We will mutate the entities in these documents and seed the mutated
+set with relations we wish to discover and that are hard to discover
+manually. 
+Later on, we will use our contacts with the Lebanese and Qatari 
+security authorities
+to validate our system on their real data.
+
+\subsection{Network content case study}
+\label{s:design:net}
+
+Parental controls help parents allow their kids to benefit from the internet 
+without fearing the danger of violent and obscene material. 
+These controls partially rely on processing the content to detect and rate 
+the target material. 
+Control mechanisms for Arabic traffic as currently mostly manual. 
+In this case study we will detect the bad content. 
+Given a set of undesired internet content 
+and a set of requests by internet users, 
+CATARM can discover whether there is a relation between 
+the requested documents and the undesired content and flag the 
+requested document with strong enough relation.
+
+\subsection{Health case study}
+\label{s:design:med}
+
+Clinical reports contain diagnosis information, observations about 
+patient progress, prescriptions, and patient response to prescribed 
+medications. 
+In countries, like Lebanon, where medical doctors graduate and specialize 
+in several distinct practices around the world it is interesting to 
+check for common practices and emerging trends to gain insight
+about local and regional health and diagnosis features. 
+It is also interesting to relate 
+pharmaceutical 
+records to clinical reports to check on 
+market consistency and on self prescribed medicine 
+practices.
+Such a study can uncover serious tendencies in uncontrolled
+pharmaceutical markets.
+It can uncover what kind of sicknesses and what kind of medicines
+are used sensitive medicines are used without 
+appropriate medical supervision.
+
+We will work with our contacts at the American University
+Hospital (AUH) at Beirut to analyze this data and relate it
+to nearby pharmaceutical records.
+
 \subsection{CATARM components}
@@
 such as Buckwalter~\cite{Buckwalter:02},
-Beesly~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
+Beesley~\cite{Beesley:01}, SAMA~\cite{Kulick:10},
 and ElixirFM~\cite{Otakar:07} take an Arabic word as input
-and produce several valid morphological solution for the word. 
+and produce several valid morphological solutions for the word. 
 Each solution is a sequence of morphemes that form the word.
@@
 \center{
-\resizebox{.4\columnwidth}{!}
+\resizebox{.5\columnwidth}{!}
 {\input{figs/hadith.pdftex_t}
@@
 
-{\bf Case-based morphological analyzer.}
+{\bf Case-based morphological analyzer.~~}
 We experimented with several morphological analyzers.
@@
 
-{\bf Linguistic computational model of Arabic.}
+{\bf Linguistic computational model of Arabic.~~}
 We think that a linguistic computational model for the Arabic
@@
 
-{\bf User query language. }
+{\bf User query language.~~}
 In Figure~\ref{f:statemachine} we showed an FSM 
@@
 \center{
-\resizebox{.8\columnwidth}{!}
+\resizebox{.9\columnwidth}{!}
 {\includegraphics{figs/detailed_submachine.pdf}
@@
 
-{\bf Simplify LCM based on the query.}
-Given
-L$L_q$ the query language, 
+{\bf Simplify LCM based on the query.~~}
+Given $L_q$ the query language, 
 $L_A$ the Arabic linguistic
@@
 states and transitions that are not always morphemes and 
-transition between morphemes. 
+transitions between morphemes. 
 We will consider a hierarchical methodology that follows 
 the hierarchical structure of the Arabic language expressed
-in $L_A$. For example,
-we may consider sets of connected states in $\rho_q$ 
-containing a single morpheme as a state in one hierarchy level 
 in $L_A$. 
+Consider $\Phi$ the state machine representing the grammar 
+rules for $L_A$ and consider $\phi$ the parse graph 
+representing $\rho_q$, in particular consider the atomic nodes 
+representing lexical entries in $\phi$. 
+Consider $S$ to be the set of states in $\Phi$ that 
+correspond to these nodes. 
+Consider simplifying $\Phi$ using techniques that remove 
+states and transitions that do not lead to $S$. 
+An optimal reduction requires expensive reachability 
+analysis, however, we can apply simpler reductions such 
+as structural cone of influence reduction,
+constant propagation reduction that overapproximate
+reachability analysis. 
+%For example, we may consider sets of connected states in $\rho_q$ 
+%containing a single morpheme as a state in one hierarchy level 
+%in $L_A$.
 We will perform this simplification before starting the 
@@
 
-{\bf Simplify LCM and MA based on set of documents.}
+{\bf Simplify LCM and MA based on documents.~~}
 We will analyze the set of input documents given to a
@@
 to help order the solutions considered by the analyzer
-to consider the most probable first.
-
-We will consider the use of HMM models 
-built on the fly while processing the documents.
+and consider the most probable first.
+
+We will consider the use of HMM~\cite{JanHMM06}
+models built on the fly while processing the documents.
+A high level HMM model of the document builds an HMM state
+model where the HMM states are morphemes or partial 
+morphemes and the HMM edges represent transitions between
+the HMM states. 
+The HMM edges are labeled with weights that reflect the
+We will look at building a correspondence between 
+the HMM model and the LCM and then simplify the 
+LCM where the aggregated probabilities happen to be
+below a certain threshold. 
+The work in~\cite{JanHMM06} uses a similar approach
+to correspond English text to tags corresponding to the 
+Unified Medical Language System (UMLS).
+
 We will also consider the use of
@@
 
-{\bf Entity and relational extractor.}
+{\bf Entity and relational extractor.~~}
 We will build a set of entity and relational extractors. 
@@
 
-{\bf Statistical components.}
+{\bf Statistical components.~~}
 We will make use and build statistical components that can be used
@@
 
-{\bf Graph algorithms.}
+{\bf Graph algorithms.~~}
 The relations extracted from the text documents will often 
@@
 components, minimum flow  and other algorithms.
-For example, in the literature case study, the discovery
+For example, in the hadith literature case study, the discovery
 of articulation vertices may point to important narrators 
@@
 
-{\bf Query solver.}
-This is a logical solver that will solve query expressions to \
-resolve matching text and relations.
-We will need to use existing off the shelf constraint solvers 
-and set theory satisfiability solvers to address our queries. 
-The solver should return and rank approximate matches.
+{\bf Query solver.~~}
+The query solver considers query expressions and resolves them 
+into matching text and relations. 
+The solver returns and ranks approximate matches.
 That is if a three entities out of four required in a relation are
@@
 
-The solver will call the morphological analyzer to compute 
-morphological matches to the atomic elements of the query. 
-Then each operation in the query such as a set intersection
-operation will correspond to a computational component. 
-The solver will call these components and integrate their
-results.
-
+The solver passes the query to a front end that builds
+a parse graph of the query. 
+It uses the parse graph to simplify the LCM and the MA
+as discussed before. 
+Then it traverses the resulting parse graph. 
+When the solver meets an atomic element of the query, 
+it will call the morphological analyzer to compute 
+morphological matches to the atomic element.
+When the solver meets an operation of the query, 
+it will call the corresponding computational component 
+and passes to it the solutions computed so far from children
+nodes. 
+The solver will use existing off the shelf constraint solvers 
+and set theory satisfiability solvers to address the queries. 
 The solver may make use of existing 
@@
 These solvers can solve satisfiability questions.
-We will explore the feasibility and utility 
-of extracting a necessary set of 
-constraints from the query and answering that with an off-the-shelf
-logic solver.
-
-{\bf Case studies.}
+
+We explain the ranking aspect of the solver with a simple
+example. 
+Consider $q$ the parse graph of a query with $n$ nodes where 
+each node corresponds to either a detected entity or
+to an operation that can be computed with one of CATARM components. 
+We define the rank of a solution to be the ratio of  the 
+satisfied nodes in $q$ over the possible satisfied
+nodes in $q$.
+
+%We will explore the feasibility and utility 
+%of extracting a necessary set of 
+%constraints from the query and answering that with an off-the-shelf
+%logic solver.
+
+{\bf Case studies.~~}
 We will work on developing the necessary components to complete 
-the solution for the hadith literature and the security case 
-studies.
-We will also work on the health and network traffic case studies.
+the solution for the hadith literature  case study. 
+We will also work on the security, health and network traffic case studies.
 While working on the case studies, we will add the components
@@
 
-{\bf Visualization.} 
+{\bf Visualization.~~}
 We will build a visualization tool that allows fast 
@@
 We detailed an aggressive work plan in Table~\ref{t:workplan}. 
-The plan spans over three years and projects a timeline
-of work per quarter. 
+The plan spans over three years each with four quarters. 
 The row headers describe work items, and the column headers
@@
 and {\bf E} 
-describe the status of the items 
+describe the type of activity taking place 
 and stand for 
-{\bf P:} {\em planning phase}, 
-{\bf W:} {\em work in progress}, 
-{\bf D:} {\em deliverable}, 
-{\bf M:}  {\em maintenance}, 
+{\bf P:} {\em plan}, 
+{\bf W:} {\em work }, 
+{\bf D:} {\em deliver}, 
+{\bf M:}  {\em maintain}, 
 and 
@@
 superscripts of the deliverables in the 
-in case studies row refer to the four 
+row of case studies refer to the four 
 case studies described in the proposal.
 
+Work can happen on all items in parallel. 
+Planing and research work can start on items that depend
+on other items. 
+For example, the LCM optimization items will be under 
+active planing and do not need to be totally idle 
+while the LCM is being built. 
+That will allow better chances to build the LCM data structures
+in a flexible way that requires less modifications if any
+when the optimizations take place.
+
+That said, we describe the dependency of the items on 
+each other as follows. 
+The work on language optimizations and data optimizations 
+depend on the Arabic LCM item. 
+The work on language optimizations depends on the design
+of the query language as well. 
+The work on the query solver depends on the design of the 
+query language. 
+The entity and relation extractors depend on the morphological 
+analyzer. 
+However, they can start since we have already
+developed an working analyzer as a preliminary step. 
+The rest of the items can all work in parallel.
+
 We will use open source tools, libraries, and frameworks to 
-build our libraries.
+build our componenets.
 We will build on existing open source programming and text 
@@
 
-We will evaluate CATARM against at least three of the case studies
-that we discussed previously and that are listed below. 
+We will evaluate CATARM against the case studies
+that we discussed in Section~\ref{s:designmethods}
+and that are listed below. 
 \begin{enumerate}
@@
 
-\subsection{CATARM evaluation }
+\subsection{External evaluation }
 
@@
 tools once the tools are available.
+Graduate students such as Masters and PhD students are examplary t
+target audience for such workshops.
 The Arabic text analysis framework will be partially based on 

### 2010-12-20 tex/rb.tex
lit review and rebuttle 1

%\documentstyle{llncs}
%\documentclass[12pt,fullpage]{llncs}
\documentclass[12pt]{article}
%\usepackage[left=3cm,top=1.2cm,right=3cm,nohead,nofoot]{geometry}
%\usepackage[letter,left=.25in,right=.25in,top=.17in,textwidth=7.5in,textheight=9in]{geometry}
%\usepackage[left=0cm]{geometry}

%\documentclass[12pt]{report}
%\usepackage {utthesis2}              %% Preamble.

%\usepackage{amsmath}
%\usepackage{amssymb}
\usepackage{arabtex}
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{times}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{multirow}
%\usepackage[colorlinks=false]{color,hyperref}
\usepackage{color,hyperref}
%\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{natbib}
\bibpunct{[}{]}{,}{n}{}{;}

\usepackage{url}

\usepackage{relsize}
\usepackage{fancyvrb}
\usepackage{fancyhdr,lastpage}

\usepackage{utf8}
\setarab
\fullvocalize
\transtrue
\arabtrue

\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue} 
\newcommand{\noTrNoVocRL}[1]{\novocalize\transfalse\RL{#1}\transtrue\vocalize} 
%\newcommand{\drawline}{\begin{picture}(6,.1) \put(0,0) {\line(1,0){6.25}}\end{picture}}

\usepackage{setspace}
%\doublespacing
\renewcommand{\baselinestretch}{1.15}
\setlength{\parindent}{0in}
%\parskip 6pt
%\parindent 0pt
%\setlength{\parskip}{.05in}
%\oddsidemargin 0in
%\evensidemargin 0in
\oddsidemargin .0in
\evensidemargin .0in
\hoffset -.75in
\voffset -.83in
\textwidth 7.5in
\textheight 9.5in

\topsep 0in
\topmargin 0.17in

%\usepackage{arabtex}
%\usepackage{utf8}

\begin{document}

\pagestyle{fancy}
%\lhead{}
\chead{}
\rhead{NPRP No.~:~~~~4-484-1-075}

\lfoot{QNRF Form}
\cfoot{}
\rfoot{Page \thepage~of~\pageref{LastPage} }
\renewcommand{\footrulewidth}{0.2pt}
\renewcommand{\headrulewidth}{0.2pt}

%\begin{titlepage}

\begin{center}
{\Large \bf Relational Arabic Text Mining Framework using 
    Morphological
    and Case-Based Analysis }

%\setlength{\unitlength}{1in}
%\setlength{\unitlength}{1in}
%\vspace{1.5in}
\vspace{.1in}

\renewcommand{\arraystretch}{.6}
\begin{tabular}{cc}
Fadi Zaraket & Rehab Duwairi \\
%Electrical and Computer Engineering &  Computer Science and Engineering Department \\
American University of Beirut & University of Qatar \\
{\tt fadi.zaraket@aub.edu.lb} & {\tt Rehab.Duwairi@qu.edu.qa}
\end{tabular}

\vspace{.1in}

\renewcommand{\arraystretch}{.6}
\begin{tabular}{c}
{\small Response to PR comments } \\
\end{tabular}
\vspace{.1in}

\date{\today}
%\pagebreak
\end{center}

%\section{Background (max of 2 pages)}
%\label{s:background}
We detail in this document our responses to the comments of
the respected reviewers of our previous Arabic text mining
framework proposal.
We rewrote the proposal to reflect changes that address the 
comments of the reviewers and changes that reflect the substantial
progress we made after submitting the initial proposal in December 
2009.
We managed to build ATSarf~\cite{ATMine09}, 
a novel case-based morphological analyzer 
and used it to automate the analysis of 
hadith literature books. 
We extracted directed acyclic graphs that 
represent a partial order relation between narrators and narrations,
and we are looking at investigaive and interrogative police reports.
This work led us to modify our approach and build the framework 
thorugh solving several case studies.
For each case study we will visit the computational components we 
needed to make it work and generalize them as library components. 
We will refine these components to make them general enough
while working on the next case study.
We adopted a case-based approach to build a flexible
low level analysis infrastructure.
The research plan document explains the approach in more details. 
This work led to manuscripts submitted for publication in 
renowned conferences. 
We hope that this addresses the {\em concern of 
the reviewers about the experience of the LPI in the area
of Arabic text mining}.

%In the following we address the comments of the reviewers.
%We start by addressing the common concerns and
%The positive comments of most of the reviewers concerning the novelty
%of our approach encouraged us to continue 
{\bf Concern 1.Specific Arabic features.(PR1,PR2,PR3,PR4,PR5)~~}
Reviewers liked the idea that we are going to use 
features specific to the Arabic language in our analysis and had
a concern that we did not explain how we are going to do that in 
detail. 
We mentioned previously that we are going to build relaxed
Arabic grammar rules and use them in the analysis. 
We describe how to do that by building a grammar 
that formalizes the Arabic language similar to 
ElixirFM~\cite{Otakar:07}.
Our grammar will be novel in that it will follow a relational 
morphological approach rather than a functional morphological 
approach. 
That is the solutions will be expressed in a directed acyclic graph
representing a hierarchy of relations between the morphemes 
rather than moprhological trees mapping the morphemes to the 
solutions.
We think this is better since it provides a more succint
representation for the solutions and because interesting relations 
between morphemes can be computed statically before
the input is presented to the analyzer.
Morphological trees are traversals of these DAGs and are expensive 
to compute statically.

We will compile this formalization into finite state machines that
compute solutions of input text strings.
We will simplify these FSMs before starting the analysis 
based on the user input query.
We consider the parse graph of the user input query 
which has atomic elements corresonding to a state variables
in the FSM.
We use this correspondence and use simple structural analysis 
techniques such as constant propagation and cone of influence
reductions to simplify the FSM. 
We think that this simplification is Arabic specific since
the query language will be designed to match the
structure of question sentences in common standard Arabic,
and the query itself reflects the way an Arabic user
himself thinks and searches.

{\bf Concern 2. Details about case studies(PR1,PR2,.~~}
Peer reviewer 1 (PR1) shared a common concern with other reviewers
about missing details of the case studies. 
We addressed this concern by providing extensive details to the
hadith and security case studies and by explaining our
approach to solve the health and network content case studies.
We named the literature case study the hadith case study as 
PR1 suggested. 
We explained how the authenticity will be represented as annotations
to the graph of Figure 2 of the research document. 
We leave the quantification of the annotations to the expert
scholars %to avoid the controversy of hadith interpretation
and provide a partial order of the annotations of each narrator.
The subject scholar can decide where to draw his thresholds. 
We hope that addresses PR1's question on evaluation and quantification
of authenticity.
PR1 commented that a contradiction may not mean non authentication.
This is true and this is why the role of the tool is to flag 
inconsistencies and leave it to the expert to decide. 
As for additional sample queries, they are definitely welcome.
For example, the graph in Figure 2 allows 
for measuring the effect of one narrator 
against a certain subject covered in the hadith. 
PR1 asks for the details of how the solver will work, we address that 
in Section 4.10 as well as in the explanation of Figures 3-6 including
the diagrams of the case studies. 
PR1 also suggests involving experts in the project, the LPI is 
working with volunteers including experts in the hadith~\cite{Zar06},
health and security subjects. 
PR2 and PR5 asked how the data will be gathered, we 
addressed that in Sections
4.2--4.5 for each of the case studies. 
PR4 was concerned about the relevance and semantical intricacy 
of the case studies and suggests other case studies such as
web mining, economy, and intelligence. 
We kept our case studies as other PRs were favorable of them,
we added a web based case study and we explained how the same 
components can be used to analyze the market.
Intelligence (in the sense of machine learning) is on our future 
work list.
We understand the semantical intricacy concern of PR4 and we hope
that our preliminary results show that we can overcome 
the semantical complexities that concerned PR4.

{\bf Concern 3. Strengthen literature review (PR1,PR2,PR3,PR4,PR5).~~}
%@inproceedings{AlSham08,
%We hope this addresses the concern of PR4 on the shortcomings
%of previous approaches and the potentials of ours. 
We went over the literature suggested by the reviewers. 
We compared our approach to current and previous approaches in the
reseach plan document. 
Here we refer to the references cited by PR5 and illustrate how
we made great use of them.
We hope this addresses the concern of PR4 on the shortcomings
of previous approaches and the potentials of ours.

We validated our case-based approach by the observation of 
Daoud Daoud~\cite{Dao09} that the pipeline model is not 
adequate for Arabic. 
He suggests a synchronized syntactic and morphological
approach while we take the argument further to 
allow a tighter integration between the pipeline
levels up to the query in question by providing a case-based
morphological analyzer. 
%However, Mr. Daoud makes a mistake when he assumes that a noun
%cannot be followed by a preposition in Arabic to resolve
%an ambiguity in his example~\noTrNoVocRL{.s-a.hb-aa s-amy _dhb-aa 
%il_A alswq} (Sami's two companions went to the market) and decide 
%that \noTrNoVocRL{_dhb-aa} is a verb (they went)
%and not a noun since it is followed by \noTrNoVocRL{il_A}.
%Surely, one can say~\noTrNoVocRL{a_h_d  .s-a.hb-aa s-amy _dhb-aa 
%il_A alswq} (Sami's two companions took gold to the market) and 
%now \noTrNoVocRL{_dhb-aa} is a noun (gold stressed with double 
%diactric).
The work~\cite{AlSham08} reports a similar result where
the automated addition of syntax knowledge improved steeming
in terms of accuracy and efficiency.

The work of Rogati in~\cite{Rog03} builds a stemmer based on
a parallel corpus with a document containing Arabic and English
text and another document containing only English with the
translation of the Arabic embedded in place. 
The translation is done manually and an English stemmer is used. 
This approach works for parallel corpora which is not the case
for our data of interest. 
It is also highly dependent on the accuracy of the manual translation.
Several interpretations of a word may be omitted if the context
of the corpus happened to bias one meaning.

Toutanova in~\cite{Tou09} suggested that using 
a joint model to do lemmatization and POS tagging performs better
than the direct and the pipelined approach. 
We take the first argument further to suggest a full joint NLP task 
model via using the NLP query to simplify and render the analysis
more accurate at the low level morphological analysis.

Later, Toutanova, in joint work with Smith and others~\cite{Smi10},
suggested better extraction quality from comparable or related
corpora. 
They consider two sets of documents of the same nature from
two different languages and extract parallel structures. 
We are actually considering the same genre of analysis in 
our proposal and case studies only in the same language. 
For example, the hadith narrations are comparable to the hadith 
biographies as far as the researcher is concerned with authenticity.
The same applies to the clinical reports and the pharmaceutical 
records, and the investigative reports from the crime scenes
and the interrogative reports with suspects.

We benefited a lot from the work of Maamouri~\cite{Maamouri:10}
and the SAMA~\cite{Kulick:10} analyzer. 
In~\cite{Maamouri:10} Maamouri considering adding a new corpus
to the Arabic Tree Bank and the addition challenged existing
annotation guidelines. 
The solution was to refine SAMA and have it interact with the
high level NLP task.
We find strong evidence in that to support our approach of
a case-based morphological analyzer integrated
with the linguistic computational model.

%@inproceedings{AlSham08,
%We hope this addresses the concern of PR4 on the shortcomings
%of previous approaches and the potentials of ours.

{\bf Concern 4. Charging model.~~}

{\bf Concern 5. Budget issues (PR1,PR2,PR4)~~}

{\bf Concern 6. List of outcomes (PR1, PR2, ).~~}

{\bf Concern 7. Web based case studies (PR2, ).~~}
We addressed the concern of PR2 by adding a case study that 
analyzes web traffic. 
The opinion mining track suggested by PR2 is also of interest and 
can be another application we consider.
Opinion classification and detection of spam opinions
can make use of the traffic classification modules of the network
content casey study, and the span opinion detection can 
make use of the inconsistency checkers of the hadith case study.

{\bf Concern 8. Improve evaluation (PR3).~~}

{\bf Concern 9. Team involved (PR3).~~}

{\bf Concern 10. Novelty (PR4).~~}

{\bf Concern 11. Plan for dissemination.(PR4).~~}

we explain the term relational (PR3) in .. 
why extreme programming (PR4)
what existing techniques (PR5)

types of documents (PR5): comparable documents

logic solver (PR5)
size of documents (PR5)

\pagebreak
%\bibliography{adnan_refs}
%\bibliographystyle{ieeetr}
\bibliographystyle{abbrvnat}
%\bibliographystyle{ieeetr}
{\small
\bibliography{fzAr}  
}

\end{document}

### 2010-12-20 tex/rp.tex
lit review and rebuttle 1

@@
 {\bf (2.)} CATARM will build a novel and efficient 
-linguistic computational model (LCM) of Arabic. 
+linguistic computational model (LCM) for Arabic. 
 %The LCM will be implemented using efficient structures. 
-The LCM will be amenable to {\bf (3.)} Arabic specific simplifications 
+The LCM will be amenable to novel 
+{\bf (3.)} Arabic specific simplifications 
 inferred from the correspondence between the formal Arabic rules
-and the relational user queries, and 
-{\bf (4.)} to simplifications 
-inferred from statistical features extracted from the documents 
+and the user queries, and 
+{\bf (4.)} language independent simplifications
+inferred from statistical features %extracted from the documents 
 using language independent techniques such as 
-hidden Markov models (HMM)  and 
-latent semantic indexing~\cite{LSI89}.
+hidden Markov models (HMM)~\cite{JanHMM06}
+and latent semantic indexing (LSI)~\cite{LSI89}.
 %CATARM will augment current digital dictionaries to resolve 
@@
 and uses set arithmetic and graph analysis techniques
-to establish relations between the graphs. {\bf (5.)} CATARM will 
-solve the case studies 
+to establish relations between the graphs. 
+{\bf (6.)} CATARM will solve the case studies 
 presented in Sections~\ref{s:design:lit}--\ref{s:design:med}.
@@
 project that aims at comprehensive automation of authenticity 
-checks against a corpus of literature heritage~\cite{Zar06}.
+checks against a corpus of hadith literature~\cite{Zar06}.
+Volunteers including experts in the field of Islamic hadith
+are involved in the project.
 The Lead PI is leading an effort at AUB to introduce 
@@
 interest to the user.
-Data mining can be applied at the end on the output of the 
-IE process to answer user queries about the input 
-documents~\cite{JHa05}.
+%Data mining can be applied at the end on the output of the 
+%IE process to answer user queries about the input 
+%documents~\cite{JHa05}.
 
@@
 } 
-\caption{CATARM diagram for the literature case study }
+\caption{CATARM diagram for the hadith case study }
 \label{f:literature}
@@
 
+Toutanova in~\cite{Tou09} suggested that using 
+a joint model to do lemmatization and POS tagging performs better
+than the direct and the pipelined approach. 
+We take the first argument further to suggest a full joint NLP task 
+model via using the NLP query to simplify and render the analysis
+more accurate at the low level morphological analysis.
+
+Later, Toutanova, in joint work with Smith and others~\cite{Smi10},
+suggested better extraction quality from comparable or related
+corpora. 
+They consider two sets of documents of the same nature from
+two different languages and extract parallel structures. 
+We are actually considering the same genre of analysis in 
+our proposal and case studies only in the same language. 
+For example, the hadith narrations are comparable to the hadith 
+biographies as far as the researcher is concerned with authenticity.
+The same applies to the clinical reports and the pharmaceutical 
+records, and the investigative reports from the crime scenes
+and the interrogative reports with suspects.
+
 \subsection{CATARM query language}
@@
 We think that we can discover such features by building
-a linguistic computational model of the Arabic language
+a linguistic computational model (LCM) of the Arabic language
 based on its formal grammar and formal morphology
@@
 We will build our morphological analyzer on top 
-of this linguistic model in addition to the current existing
+of this LCM addition to the current existing
 concatenative-based analysis techniques.
+
+Our LCM will differ from ElixirFM~\cite{Otakar:07}
+in that it will follow a relational approach rather 
+than a functional morphological approach. 
+That is the solutions will be expressed in a directed acyclic graph 
+(DAG)
+representing a hierarchy of relations between the morphemes 
+rather than moprhological trees mapping the morphemes to the 
+solutions.
+We think this is better since it provides a more succint
+representation for the solutions and because interesting relations 
+between morphemes can be computed statically before
+the input is presented to the analyzer.
+Morphological trees are traversals of these DAGs, are expensive 
+to exhaustively compute and store statically, 
+and are expensive to compute dynamically. 
+Our intuition is that Arabic allows relational strcutures
+such as noun sentences and complex hierarchical entities.
 
@@
 We will explore methods that use $\rho_q$ to simplify 
-$L_A$ before we start the morphological analysis. 
+$L_A$ before we start the analysis. 
 We think the possible simplifications that we can perform
@@
 
+In ~\cite{Dao09}, Daoud observes that the pipeline model,
+where morphology is performed first and syntactic and other
+processes follow, is not adequate for Arabic. 
+He suggests a synchronized syntactic and morphological
+approach.  
+We take the argument further to 
+allow a tighter integration between the several pipeline
+levels up to the NLP query in question by providing a 
+case-based morphological analyzer, a linguistic computational
+model of Arabic, and a query language that can all be expressed
+as FSMs. 
+The work~\cite{AlSham08} reports a similar result where
+the automated addition of syntax knowledge improved steeming
+in terms of accuracy and efficiency.
+
+Here we point to a subtletie that shows how complex 
+low level Arabic morphology is. 
+The work in~\cite{Dao09} assumes that a noun
+cannot be followed by a preposition in Arabic and uses that 
+to resolve an ambiguity in the 
+example~\noTrNoVocRL{.s-a.hb-aa s-amy _dhb-aa il_A alswq} 
+(Sami's two companions went to the market) 
+    and decides that \noTrNoVocRL{_dhb-aa} 
+is a verb (they went)
+and not a noun since it is followed by \noTrNoVocRL{il_A}.
+One can 
+say~\noTrNoVocRL{a_h_d  .s-a.hb-aa s-amy _dhb-aa il_A alswq} 
+(Sami's two companions took gold to the market) and 
+now \noTrNoVocRL{_dhb-aa} is a noun (gold stressed with double 
+diactric).
+This shows that even complex and advanced techniques such as
+~\cite{Dao09} can be misled by ambiguous compatibility
+rules.
+
 We will explore extracting our model from formal text book
@@
 These solvers can solve satisfiability questions.
-We will explore the feasibility and utility 
-of extracting a necessary set of 
-constraints from the query and answering that with an off-the-shelf
-logic solver.
-
-{\bf Case studies.}
+
+We explain the ranking aspect of the solver with a simple
+example. 
+Consider $q$ the parse graph of a query with $n$ nodes where 
+each node corresponds to either a detected entity or
+to an operation that can be computed with one of CATARM components. 
+We define the rank of a solution to be the ratio of  the 
+satisfied nodes in $q$ over the possible satisfied
+nodes in $q$.
+
+%We will explore the feasibility and utility 
+%of extracting a necessary set of 
+%constraints from the query and answering that with an off-the-shelf
+%logic solver.
+
+{\bf Case studies.~~}
 We will work on developing the necessary components to complete 
@@
 We will evaluate CATARM against the case studies
-that we discussed in Section~\ref{s:design} and 
-that are listed below. 
+that we discussed in Section~\ref{s:designmethods}
+and that are listed below. 
 \begin{enumerate}

### 2010-12-20 tex/rb.tex
novelty, fine editing, new bibs

@@
 
-\lfoot{QNRF Form}
+\lfoot{QNRF Form -- response to PR comments}
 \cfoot{}
-\rfoot{Page \thepage~of~\pageref{LastPage} }
+%\rfoot{Page \thepage~of~\pageref{LastPage} }
+\rfoot{Page \thepage~of~3}
 \renewcommand{\footrulewidth}{0.2pt}
@@
 \begin{tabular}{c}
-{\small Response to PR comments } \\
+{\small Response to PR comments --- maximum of 3 pages} \\
 \end{tabular}
@@
 %\label{s:background}
-We detail in this document our responses to the comments of
-the respected reviewers of our previous Arabic text mining
+We highlight in this document the important responses to the 
+excellent comments of
+the respected peer reviewers (PR) of our previous Arabic text mining
 framework proposal.
-We rewrote the proposal to reflect changes that address the 
-comments of the reviewers and changes that reflect the substantial
-progress we made after submitting the initial proposal in December 
-2009.
-We managed to build ATSarf~\cite{ATMine09}, 
-a novel case-based morphological analyzer 
+More detailed responses can be found in the research plan document. 
+The excellent comments were instrumental in the 
+progress we made and helped us bring the proposal to a far better 
+quality.
+We start by the changes to the proposal 
+that reflect our progress after December 2009.
+With that we hope to address the concern of PR2, PR3, PR4,
+and PR5 on the adequacy of the research team.
+
+We built ATSarf~\cite{ATMine09}, 
+a novel case-based morphological analyzer,
 and used it to automate the analysis of 
 hadith literature books. 
-We extracted directed acyclic graphs that 
-represent a partial order relation between narrators and narrations,
-and we are looking at investigaive and interrogative police reports.
-This work led us to modify our approach and build the framework 
-thorugh solving several case studies.
-For each case study we will visit the computational components we 
-needed to make it work and generalize them as library components. 
-We will refine these components to make them general enough
-while working on the next case study.
+We extracted directed acyclic graphs (DAG) that 
+represent a partial order relation between narrators and narrations.
+We are currently looking at investigative and interrogative police 
+reports.
+We explain this work in details in Section 3 of the research document.
+
+Our experience led us to slightly modify our approach and 
+build the framework through solving several case studies.
+For each case study we will visit its computational components 
+and generalize them as library components. 
+We will refine the components built for other case studies
+and make them general enough to be used with the current case study.
 We adopted a case-based approach to build a flexible
-low level analysis infrastructure.
+low level morphological analysis infrastructure.
 The research plan document explains the approach in more details. 
@@
 renowned conferences.
-We hope that this addresses the {\em concern of 
-the reviewers about the experience of the LPI in the area
-of Arabic text mining}.
 
-%In the following we address the comments of the reviewers.
-%We start by addressing the common concerns and
-%The positive comments of most of the reviewers concerning the novelty
-%of our approach encouraged us to continue 
-{\bf Concern 1.Specific Arabic features.(PR1,PR2,PR3,PR4,PR5)~~}
+In the following we address the comments of the reviewers.
+
+{\bf Concern 1. Specific Arabic features.(PR1,PR2,PR3,PR4,PR5)~~}
 Reviewers liked the idea that we are going to use 
@@
 We mentioned previously that we are going to build relaxed
-Arabic grammar rules and use them in the analysis. 
-We describe how to do that by building a grammar 
-that formalizes the Arabic language similar to 
+Arabic grammar rules.
+We describe how to do that in Sections 4.7 and 4.9. 
+We build a grammar that formalizes the Arabic language similar to 
 ElixirFM~\cite{Otakar:07}.
-Our grammar will be novel in that it will follow a relational 
-morphological approach rather than a functional morphological 
-approach. 
-That is the solutions will be expressed in a directed acyclic graph
-representing a hierarchy of relations between the morphemes 
-rather than moprhological trees mapping the morphemes to the 
-solutions.
-We think this is better since it provides a more succint
-representation for the solutions and because interesting relations 
+Our grammar will be novel in that it will build relations between
+morphological entities rather than
+functional maps from morphological entities to solutions
+using morphological trees. 
+%In our approach the solutions will be expressed in a directed 
+%acyclic graph
+%representing a hierarchy of relations between the morphemes.
+We think this provides a more succinct
+representation for the solutions. Also interesting relations 
 between morphemes can be computed statically before
 the input is presented to the analyzer.
-Morphological trees are traversals of these DAGs and are expensive 
-to compute statically.
+%Morphological trees are traversals of these DAGs and are expensive 
+%to compute statically.
 
-We will compile this formalization into finite state machines that
-compute solutions of input text strings.
-We will simplify these FSMs before starting the analysis 
-based on the user input query.
-We consider the parse graph of the user input query 
-which has atomic elements corresonding to a state variables
-in the FSM.
+We explain further in Section 4.10 page 16 how we will
+use language specific features to simplify the linguistic model. 
+Briefly, we will compile the formalization into finite state 
+machines.
+We consider the parse graph of the user input query with
+elements that correspond to state variables in the FSMs of
+the linguistic model. 
+%The query has atomic elements corresonding to a state variables
+%in the FSM.
 We use this correspondence and use simple structural analysis 
 techniques such as constant propagation and cone of influence
-reductions to simplify the FSM. 
-We think that this simplification is Arabic specific since
-the query language will be designed to match the
-structure of question sentences in common standard Arabic,
-and the query itself reflects the way an Arabic user
-himself thinks and searches.
+reductions to simplify the FSMs. 
+This simplification is Arabic specific since
+the query language matches the structure of question sentences 
+in common standard Arabic.
+%and the query itself reflects the way an Arabic user
+%himself thinks and searches.
 
-{\bf Concern 2. Details about case studies(PR1,PR2,.~~}
-Peer reviewer 1 (PR1) shared a common concern with other reviewers
+{\bf Concern 2. Details about case studies (PR1,PR2,PR4,PR5).~~}
+PR1 shared a common concern with other reviewers
 about missing details of the case studies. 
 We addressed this concern by providing extensive details to the
-hadith and security case studies and by explaining our
-approach to solve the health and network content case studies.
-We named the literature case study the hadith case study as 
-PR1 suggested. 
-We explained how the authenticity will be represented as annotations
+hadith case study in Sections 3, 4.2 and 4.10.
+We also discussed the security case study with more details in 
+Section 4.3.
+We elaborated on the health case study in Section 4.4.
+
+We renamed the hadith case study study as PR1 suggested. 
+We explained how authenticity will be represented as annotations
 to the graph of Figure 2 of the research document. 
@@
 and provide a partial order of the annotations of each narrator.
-The subject scholar can decide where to draw his thresholds. 
-We hope that addresses PR1's question on evaluation and quantification
+The expert can decide where to draw his thresholds. 
+We hope this addresses PR1's question on evaluation and quantification
 of authenticity.
@@
 inconsistencies and leave it to the expert to decide. 
-As for additional sample queries, they are definitely welcome.
+PR1 also asked about additional sample queries, 
+they are definitely welcome.
 For example, the graph in Figure 2 allows 
@@
 against a certain subject covered in the hadith.
-PR1 asks for the details of how the solver will work, we address that 
-in Section 4.10 as well as in the explanation of Figures 3-6 including
-the diagrams of the case studies. 
+
+We address the question of PR1 about how the solver works, 
+in Section 4.10 as well as in the explanation of Figures 3-6 that
+include the computational diagrams. 
 PR1 also suggests involving experts in the project, the LPI is 
 working with volunteers including experts in the hadith~\cite{Zar06},
-health and security subjects. 
-PR2 and PR5 asked how the data will be gathered, we 
-addressed that in Sections
-4.2--4.5 for each of the case studies. 
-PR4 was concerned about the relevance and semantical intricacy 
-of the case studies and suggests other case studies such as
+health and security areas.
+
+We addressed the question of
+PR2 and PR5 about data gathering and data
+in Sections 4.2--4.5 for each case study.
+
+PR4 was concerned about the relevance and semantic intricacy 
+of the case studies and suggested other case studies such as
 web mining, economy, and intelligence. 
-We kept our case studies as other PRs were favorable of them,
-we added a web based case study and we explained how the same 
-components can be used to analyze the market.
+%We kept our case studies as other PRs were favorable of them,
+We added a web based case study in Section 4.5
+and we explained how the components from the case studies 
+can be used to analyze the market.
 Intelligence (in the sense of machine learning) is on our future 
 work list.
-We understand the semantical intricacy concern of PR4 and we hope
+We understand the semantic intricacy concern of PR4 and we hope
 that our preliminary results show that we can overcome 
-the semantical complexities that concerned PR4.
+the semantic complexities that concerned PR4.
 
@@
 We compared our approach to current and previous approaches in the
-reseach plan document. 
+research plan document. 
 Here we refer to the references cited by PR5 and illustrate how
@@
 
-We validated our case-based approach by the observation of 
-Daoud Daoud~\cite{Dao09} that the pipeline model is not 
+Daoud Daoud in~\cite{Dao09} observed that the pipeline model is not 
 adequate for Arabic. 
-He suggests a synchronized syntactic and morphological
-approach while we take the argument further to 
+He suggested a synchronized syntactic and morphological
+approach. 
+The work~\cite{AlSham08} reports a similar result where
+the automated addition of syntax knowledge improved stemming
+in terms of accuracy and efficiency. 
+We take their arguments further to 
 allow a tighter integration between the pipeline
-levels up to the query in question by providing a case-based
+levels up to the NLP query in question by providing a case-based
 morphological analyzer. 
@@
 %diactric).
-The work~\cite{AlSham08} reports a similar result where
-the automated addition of syntax knowledge improved steeming
-in terms of accuracy and efficiency.
 
 The work of Rogati in~\cite{Rog03} builds a stemmer based on
-a parallel corpus with a document containing Arabic and English
-text and another document containing only English with the
-translation of the Arabic embedded in place. 
-The translation is done manually and an English stemmer is used. 
+the English stemming of 
+a parallel corpus with documents containing Arabic and English
+text and parallel documents containing only English with the
+manual translation of Arabic. 
 This approach works for parallel corpora which is not the case
-for our data of interest. 
+for our case studies.
 It is also highly dependent on the accuracy of the manual translation.
-Several interpretations of a word may be omitted if the context
-of the corpus happened to bias one meaning.
+%Several interpretations of a word may be omitted if the context
+%of the corpus happened to bias one meaning.
 
 Toutanova in~\cite{Tou09} suggested that using 
-a joint model to do lemmatization and POS tagging performs better
+a joint model to do lemmatization and POS tagging performed better
 than the direct and the pipelined approach. 
-We take the first argument further to suggest a full joint NLP task 
+We take the argument further to suggest a full joint NLP task 
 model via using the NLP query to simplify and render the analysis
@@
 corpora. 
-They consider two sets of documents of the same nature from
-two different languages and extract parallel structures. 
-We are actually considering the same genre of analysis in 
-our proposal and case studies only in the same language. 
+%They consider two sets of documents of the same nature from
+%two different languages and extract parallel structures. 
+This is similar to our selection of the case studies
+except that our documents are in the same language.
 For example, the hadith narrations are comparable to the hadith 
-biographies as far as the researcher is concerned with authenticity.
-The same applies to the clinical reports and the pharmaceutical 
-records, and the investigative reports from the crime scenes
-and the interrogative reports with suspects.
+biographies.
+%as far as the researcher is concerned with authenticity.
+%The same applies to the clinical reports and the pharmaceutical 
+%records, and the investigative reports from the crime scenes
+%and the interrogative reports with suspects.
 
@@
 and the SAMA~\cite{Kulick:10} analyzer. 
-In~\cite{Maamouri:10} Maamouri considering adding a new corpus
-to the Arabic Tree Bank and the addition challenged existing
+In~\cite{Maamouri:10} Maamouri considers adding a new corpus
+to the Arabic Tree Bank. The addition challenged existing
 annotation guidelines. 
@@
 high level NLP task.
-We find strong evidence in that to support our approach of
-a case-based morphological analyzer integrated
-with the linguistic computational model.
+%We find strong evidence in that to support our approach of
+%a case-based morphological analyzer integrated
+%with the linguistic computational model.
 
@@
 
-{\bf Concern 4. Charging model.~~}
+%{\bf Concern 4. Continuity and charging model (PR1).~~}
+%We suggested to charge on support for the framework as a partial
+%means to sustain the work and not as a sole source of funds. 
+%This model works fine for many open source tools as the cost 
+%of open source contributions is not expensive. 
+%Note that the continuation of the project does not solely depend
+%on the charging model as it mostly depends on contributions to 
+%research in the area from around the world.
+%We will also charge on custom expertise to set up new applications.
 
-{\bf Concern 5. Budget issues (PR1,PR2,PR4)~~}
+{\bf Concern 4. Budget issues (PR1,PR2,PR4)~~}
+We addressed the budget issues by reducing the requested budget.
+Note that the travel and consultant expenses were noted inline 
+with the itemized expense guidelines at the American
+University of Beirut and at Qatar University as provided
+by the corresponding research offices.
 
-{\bf Concern 6. List of outcomes (PR1, PR2, ).~~}
+{\bf Concern 5. List of outcomes and novelty (PR1, PR2, PR4).~~}
+We refined our outcomes and detailed what is novel. 
+We claim the novelty of our case-based morphological analyzer, 
+a flexible linguistic computational analysis that can be loaded 
+partially, the language specific optimization techniques, 
+the application of language independent statistical techniques
+to the simplification of the linguistic model, and the 
+application of CATARM to the case studies considered.
 
-{\bf Concern 7. Web based case studies (PR2, ).~~}
-We addressed the concern of PR2 by adding a case study that 
+{\bf Concern 6. Web based case studies (PR2,PR4).~~}
+We addressed the concern of PR2 and PR4 by adding a case study that 
 analyzes web traffic. 
 The opinion mining track suggested by PR2 is also of interest and 
-can be another application we consider.
+can be another application we will consider.
 Opinion classification and detection of spam opinions
 can make use of the traffic classification modules of the network
-content casey study, and the span opinion detection can 
+content case study, and the spam opinion detection can 
 make use of the inconsistency checkers of the hadith case study.
 
-{\bf Concern 8. Improve evaluation (PR3).~~}
+{\bf Concern 7. Improve evaluation (PR3).~~}
+We improved the evaluation section to list how each component will 
+be evaluated and what it will be compared with.
 
-{\bf Concern 9. Team involved (PR3).~~}
+{\bf Concern 8. Team involved (PR3).~~}
+Our current team is the LPI, Co-LPI, one RA, 
+and two seniors. % undergraduate students.
+Once funded, the team will readily expand to 
+free the students from other work and 
+attract PhD and graduate students.
 
-{\bf Concern 10. Novelty (PR4).~~}
+%{\bf Concern 10. Novelty (PR4).~~}
 
-{\bf Concern 11. Plan for dissemination.(PR4).~~}
+{\bf Concern 9. Plan for dissemination.(PR2,PR4).~~}
+We devised a detailed plan to disseminate the results
+to address the comments of PR4 including 
+workshops and tutorial sessions to 
+interest graduate students as PR2 suggested.
 
-we explain the term relational (PR3) in .. 
-why extreme programming (PR4)
-what existing techniques (PR5)
+{\bf Concern 10. Types of documents. (PR5)~~}
+PR5 noted that analysis for documents differing in genre and 
+type requires different techniques. 
+We agree with that statement. 
+In our case studies we target comparable documents such as 
+hadith and biography and extract relations between them. 
+This is similar to the work of Al-Shammari~\cite{AlSham08} that
+takes as input comparable documents from two languages. 
+We differ in that we consider comparable documents in the same
+language.
 
-types of documents (PR5): comparable documents
+{\bf Concern 11. Logic solvers. (PR5)~~}
+PR5 asks about the role of the logic solver. 
+The logic solvers will work as constraint solvers to answer 
+satisfiability queries and return models from sets of solutions
+when set arithmetic or logic operations are applied over
+sets of morphological solutions. 
+The integration of the logic solver may be deferred to later 
+as PR5 suggested until the entity and relation extractors are 
+available. We note that when we discuss dependency between 
+work items in Section 4.12. 
+However, as PR5 noted, logic solvers are close to the expertise of
+the LPI and integrating them into the framework will not be
+a challenging task.
 
-logic solver (PR5)
-size of documents (PR5)
+{\bf Concern 12. Choice of eXtreme programming. (PR4)~~}
+We explained well in our work plan in Section 4.12 why we chose 
+eXtreme Programming as a methodology in response to PR4. 
+For one, pair programming, encouraged by XP, 
+works well in the context of academy and research to keep
+expertise in the team. 
+As senior students graduate, their junior partners in the pair
+will keep the knowledge.
+
+{\bf Concern 13. Use of existing techniques. (PR5)~~}
+We address this concern by citing the Buckwalter and ElixirFM 
+morphological analyzer we have worked with and used so far. 
+We also refer to possible use of the CADIM toolkit from Columbia
+university. 
+We cite all relevant techniques we plan to use including the use 
+of hidden Markov models in text mining and the use of latent semantic
+indexing.
+
+{\bf Concern 14. Relational text mining. (PR3)~~}
+We explain the term {\em relational text mining} as the task of 
+looking at sets of comparable documents with an expected set 
+of relations between the documents. 
+%Figures 1 and 2 are good examples of the relations.
+A user query expresses the expected relations and can
+%will be given as user queries and can 
+be a complex structure of other relations and entities. 
+%We gave explain that with the part
+
+We hope we have addressed most of the concerns of the PRs in this
+document and we highly appreciate their excellent and beneficial
+feedback that helped us bring the proposal to a far better quality
+as it stands now. 
+We look forward to their positive feedback.
 
 \pagebreak
+\rfoot{Page \thepage~of~\pageref{LastPage} }
 %\bibliography{adnan_refs}

### 2010-12-20 tex/rp.tex
novelty, fine editing, new bibs

@@
 \cfoot{}
-\rfoot{Page \thepage~of~\pageref{LastPage} }
+%\rfoot{Page \thepage~of~\pageref{LastPage} }
+\rfoot{Page \thepage~of~23}
 \renewcommand{\footrulewidth}{0.2pt}
@@
 ``Arabic Computational Morphology''~\cite{Sou07}
-covers most academic research work on Arabic text mining.
+covers substantial academic research work on Arabic text mining.
 It summarizes recent work, and presents strong evidence of 
@@
 Arabic text documents.
-CATARM takes as input several sets of documents, each of a specific
+CATARM takes as input several sets of comparable documents, 
+each of a specific
 type, and a user query. % expressed in a visual language.
@@
 components. It also uses features extracted from the query and 
-the document sets to simplify the computational 
+the document sets to simplify the %computational 
 components. 
@@
 the root of the lexeme and the fully inflected word. 
-While the rules capture features unique to the 
-the Arabic language and its linguistic description~\cite{Badawi:04},
+While the rules capture features unique to 
+the linguistic description of Arabic~\cite{Badawi:04},
 the approach is computationally expensive and enumerates 
@@
 CATARM.
+We chose our case studies to be documents of comparable
+content. 
+Our intuition is that this is more 
+adequate to a relational text mining approach.
 
@@
 with authentication qualification annotations. 
+The digitized hadith and biography books are available 
+online~ \footnote{We obtained
+  the digitized books from online sources such as 
+  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
+  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
+The books are considered open Islamic heritage and access to
+them is granted without considerable IP constraints. 
+We are looking at processing around 60 books of hadith and
+biographies each in several volumes. 
+The total data size is in the order of 30MB worth of digital 
+data.
 
@@
 manually. 
+We are looking at processing original documents in the order
+of hundreds, and at generating few tens of thousands of dummy 
+documents.
 Later on, we will use our contacts with the Lebanese and Qatari 
@@
 
-Parental controls help parents allow their kids to benefit from the internet 
+Parental controls help parents allow their kids to benefit 
+from the Internet 
 without fearing the danger of violent and obscene material. 
-These controls partially rely on processing the content to detect and rate 
-the target material. 
+These controls partially rely on processing the content to detect 
+and rate the target material. 
 Control mechanisms for Arabic traffic as currently mostly manual. 
 In this case study we will detect the bad content. 
-Given a set of undesired internet content 
-and a set of requests by internet users, 
+Given a set of undesired Internet content 
+and a set of requests by Internet users, 
 CATARM can discover whether there is a relation between 
@@
 requested document with strong enough relation. 
+The undesired content is available online and is in the 
+order of thousands of documents. 
+We will look at Arabic Internet traffic five times larger to
+validate our case study.
 
@@
 medications. 
-In countries, like Lebanon, where medical doctors graduate and specialize 
+In countries, like Lebanon, where medical doctors graduate and 
+specialize 
 in several distinct practices around the world it is interesting to 
@@
 to nearby pharmaceutical records. 
+The data available is in the order of few thousand clinical reports
+and comparable pharmaceutical records.
 
@@
 Toutanova in~\cite{Tou09} suggested that using 
-a joint model to do lemmatization and POS tagging performs better
+a joint model to do lemmatization and part of speech (POS
+tagging performs better
 than the direct and the pipelined approach. 
@@
 representing a hierarchy of relations between the morphemes 
-rather than moprhological trees mapping the morphemes to the 
+rather than morphological trees mapping the morphemes to the 
 solutions.
-We think this is better since it provides a more succint
-representation for the solutions and because interesting relations 
+We think this provides a more succinct
+representation for the solutions.
+Also interesting relations 
 between morphemes can be computed statically before
@@
 and are expensive to compute dynamically. 
-Our intuition is that Arabic allows relational strcutures
+Our intuition is that Arabic allows relational structures
 such as noun sentences and complex hierarchical entities.
@@
 The work~\cite{AlSham08} reports a similar result where
-the automated addition of syntax knowledge improved steeming
+the automated addition of syntax knowledge improved stemming
 in terms of accuracy and efficiency.
 
-Here we point to a subtletie that shows how complex 
+Here we point to a subtlety that shows how complex 
 low level Arabic morphology is. 
@@
 analysis, however, we can apply simpler reductions such 
-as structural cone of influence reduction,
-constant propagation reduction that overapproximate
+as structural cone of influence and
+constant propagation reductions that overapproximate
 reachability analysis. 
+This simplification is Arabic specific since
+the query language matches the structure of question sentences 
+in common standard Arabic.
+In addition, the query itself reflects the way an Arabic user
+himself thinks and searches. 
 %For example, we may consider sets of connected states in $\rho_q$ 
@@
 
-We will follow an eXtreme Programming 
+We will follow a relaxed  eXtreme Programming 
 methodology in the process of building CATARM.
@@
 increase productivity. 
-While a item is delivered, it will still be subject to work,
+While an item is delivered, it will still be subject to work,
 enhancements, and maintenance. 
+We selected eXtreme Programming as this fits well in the 
+context of research where the clients are at close 
+contact with the developers.
+It accomodates the high dependency amongst the 
+modules of CATARM. 
+The pair programming practice in eXtreme Programming 
+works well in the context of academy and research to keep
+expertise in the team. 
+As senior students graduate, their junior partners in the pair
+will keep the knowledge in.
 
@@
 The work on the query solver depends on the design of the 
-query language. 
+query language and on the availability of the entity and relation
+extractors. 
 The entity and relation extractors depend on the morphological 
@@
 We will use open source tools, libraries, and frameworks to 
-build our componenets.
+build our components.
 We will build on existing open source programming and text 
@@
 post updates on the progress of the project.
+
+\subsection{Novelty}
+
+We claim the novelty of the following items
+\begin{itemize}
+\item the case-based morphological analyzer,
+\item the flexible and partial linguistic computational model,
+\item the language specific LCM simplification techniques,
+\item the application of language independent statistical techniques
+to simplify the LCM,
+\item the solutions of the four proposed case studies,
+\item and the CATRAM framework as an integration of all the above techniques.
+\end{itemize}
 
@@
 
-The team will seek the assistance and help of an independent 
-consultant in the middle of the second and third years of 
+The team will seek the assistance and help of independent 
+consultants  in the middle of the second and third years of 
 the duration of the project to evaluate and validate the results.
@@
 efforts to keep it going.
+
+The PIs will continue soliciting funds for the development
+of extensions and enhancements of this work as well.
+
 Furthermore, the Arabic text analysis framework 
-can sustain itself by charging on support or expert use.
-
-The framework or service on the framework can easily be 
-commercialized to tend to the security, health, insurance, market,
+can partially sustain itself by charging on support or expert use
+similar to other similar projects such as the ATB.
+The framework or service of the framework can be 
+commercialized to tend to commercial usage from 
+the security, health, insurance, market,
 news, media, or literature business sectors.
@@
 morphological analyzer with useful categorization and entity 
-tags that reflect semantical meanings. 
+tags that reflect semantic meanings. 
 These modifications would give CATARM additional capabilities in 
@@
 
-The team will publish the results of the research in the form 
-of refereed conference and journal papers as well as tool papers 
-addressing each of the tools. 
-The team will also hold workshops to educate possible users on the 
-tools once the tools are available.
-Graduate students such as Masters and PhD students are examplary t
-target audience for such workshops.
 The Arabic text analysis framework will be partially based on 
@@
 purposes.  
+The tools we have developed so far are readily available 
+online under the {\em atmine} project~\cite{ATMine09}.
+
+The team will publish the results of the research in the form 
+of refereed conference and journal papers as well as tool papers 
+addressing each of the tools. 
+We already submitted one manuscript describing the case-based
+morphological analyzer. 
+We have a ready to publish manuscript on the hadith authentication
+case study. 
+We plan to have publications on each of the 
+linguistic computational model, the language specific techniques
+to simplify it, and the language independent techniques to simplify
+it. 
+We also plan to have a publication on each of the separate case 
+studies. 
+We plan to have technical reports, tool papers, and user
+guides describing
+the whole CATARM framework and its use.
+
+The team will also hold workshops to educate possible users of the
+tools at AUB, QU, and in other places of interest. 
+Masters,  PhD, and senior undergraduate students are 
+exemplary target audience for such workshops.
 
 \pagebreak
+\rfoot{Page \thepage~of~\pageref{LastPage} }
 %\bibliography{adnan_refs}

### 2010-12-20 tex/rp.tex
final version

@@
 political campaigning, and other businesses.
-Text mining involves information retrieval (IR), natural language 
-processing (NLP), information extraction (IE), 
-and data mining~\cite{ASr09}.
+%Text mining involves information retrieval (IR), natural language 
+%processing (NLP), information extraction (IE), 
+%and data mining~\cite{ASr09}.
 
@@
 statistical, set arithmetic, logic,
-and graph analysis techniques, and  text algorithms.
+graph analysis techniques, and  text algorithms.
 \item We will use CATARM to solve several realistic case studies.
@@
 are hard to discover by humans and will have direct live and
-cost saving benefits.% in cases similar to the above mentioned 
+cost saving benefits. % in cases similar to the above mentioned 
 %case studies. 
@@
 extract the chains of narrators from a set of narrations
-with above 95 percent accuracy and in less than two minutes.
+with above 95 percent accuracy and in less than 10 seconds.
 
@@
 This partial order graph is instrumental to automate
-localizing narrators in biography documents and
+localization of narrators in biography documents and
 segmenting biography documents.
@@
 One can also annotate the POR with the locations and 
-the time they lived in. 
+the time the narratores lived in. 
 Then one can compute a time and location overlap
@@
 
+The PI and his team at the American University of Beirut (AUB)
+started work on an Arabic text mining framework~\cite{ATMine09}
+that is available now online.
+They have successfully developed ATSarf, a case-based morphological 
+analyzer, and applied it on the hadith literature case study.
+
 The Lead PI worked in the area of software Arabization since 1996 
@@
 \begin{itemize}
-\item Build the CATARM framework while solving four interesting Arabic 
+\item Build the novel CATARM framework while solving four interesting Arabic 
 text minining case studies
@@
   \end{itemize}
-\item Build and enhance a case-based morphological analyzer
-\item Build an Arabic linguistic computational model (LCM)
+\item Build and enhance a novel case-based morphological analyzer
+\item Build a novel partial Arabic linguistic computational model (LCM)
 \item Build entity and relation extractors
@@
   \begin{itemize}
-  \item Build optimization modules that use the query to reduce the LCM
-  \item Build a query solver that compiles all the components and run them 
+  \item Build novel simplification modules that use the query to reduce the LCM
+  \item Build a query solver that compiles all the components, runs them, and ranks them
   \end{itemize}
-\item Build language independent LCM optimization modules
+\item Build novel language independent LCM simplification modules
 \item Add statistical, graph, and text manipulation algorithms
@@
 results. 
-Section~\ref{s:design:plan} will details the work plan and
+Section~\ref{s:design:plan} will detail the work plan and
 justify the proposed budget.
@@
 we need tools that automatically differentiate between 
-names, verbs, adjectives and other structures.
+names, verbs, adjectives and other elements.
 This is not a simple task with the Arabic language.
@@
 In addition to these structural differences,
+and to difficulties mentioned in Section~\ref{s:background},
 there are features in the Arabic language that need special 
@@
 It will also give feedback to the MA component to stop the
-analysis when a it reaches a satisfactory result.
+analysis when it reaches a satisfactory result.
 We consider this graph as a partial order relation between 
@@
 
+The diagram in Figure~\ref{f:security} describes our 
+high level vision for what the solution of the 
+security case study
+will look like in the CATARM framework. 
+The inputs are two separate sets of documents. 
+The first set includes crime scene investigation reports, 
+and the second set includes 
+interrogation reports with crime suspects as well as 
+a database of previous interrogation reports. 
+The CATARM system also takes as input a database of geographical
+locations and a database of personal records. 
+The aim is to check for inconsistencies in the interrogation
+reports and to identify possible connections between the suspects
+and the crime scene.
+
+A morphological analyzer with date, place, and name extraction controller extracts
+names, dates and places from the investigation reports. 
+We pass the dates to a sorting component to produce $G_D$ 
+a partial order graph between the dates. 
+We look up the names in the personal records database and find out
+$G_N$ a family dependency graph between the names. 
+We pass the places to a mapper that builds $G_G$,
+another graph relating the places using a geographical database.
+
+We pass the interrogation reports to an MA with a relation 
+extraction controller that builds $G_R$. 
+The graph $G_R$ relates the interrogated suspects
+to each other, to places, and to dates based on a simple
+neighborhood threshold computed using a statistical feature.
+
+We feed $G_D, G_F, G_G,$ and $G_R$   
+to a consistency checker.
+The consistency checker considers the three dimensions 
+$N$ names, $G$ locations, and $D$ dates and projects $G_R$
+over each one of them.
+For each dimension $k$, it computes a local measure 
+$\alpha_k$ of how much the projection $G_R|_k$ simulates $G_k$. 
+It flags the components in the graph where the accumulated 
+measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
+below a threshold $\theta$ as inconsistent.
+
 %\begin{figure}[tb]
@@
 \end{figure}
-
-The diagram in Figure~\ref{f:security} describes our 
-high level vision for what the solution of the 
-security case study
-will look like in the CATARM framework. 
-The inputs are two separate sets of documents. 
-The first set includes crime scene investigation reports, 
-and the second set includes a 
-interrogation reports with crime suspects as well as 
-a database of previous interrogation reports. 
-The CATARM system also takes as input a database of geographical
-locations and a database of personal records. 
-The aim is to check for inconsistencies in the interrogation
-reports and to identify possible connections between the suspects
-and the crime scene.
-
-A morphological analyzer with date, place, and name extraction controller extracts
-names, dates and places from the investigation reports. 
-We pass the dates to a sorting component to produce $G_D$ 
-a partial order graph between the dates. 
-We look up the names in the personal records database and find out
-$G_N$ a family dependency graph between the names. 
-We pass the places to a mapper that builds another $G_G$,
-a graph relating the places using a geographical database.
-
-We pass the interrogation reports to an MA with a relation 
-extraction controller that builds $G_R$. 
-The graph $G_R$ relates the interrogated suspects
-to each other, to places, and to dates based on a simple
-neighborhood threshold computed using a statistical feature.
-
-We feed $G_D, G_F, G_G,$ and $G_R$   
-to a consistency checker.
-The consistency checker considers the three dimensions 
-$N$ names, $G$ locations, and $D$ dates and projects $G_R$
-over each one of them.
-For each dimension $k$, it computes a local measure 
-$\alpha_k$ of how much $G_R|_k$ simulates $G_k$. 
-It flags the components in the graph where the accumulated 
-measure $\alpha= \Sigma_{k\in\{D,N,G\}} \alpha_k$ is
-below a threshold $\theta$ as inconsistent.
 
@@
 and rate the target material. 
-Control mechanisms for Arabic traffic as currently mostly manual. 
+Control mechanisms for Arabic traffic are currently mostly manual. 
 In this case study we will detect the bad content. 
@@
 medications. 
-In countries, like Lebanon, where medical doctors graduate and 
-specialize 
-in several distinct practices around the world it is interesting to 
+In countries, like Lebanon, where medical doctors graduate from and 
+specialize in 
+several distinct practices around the world it is interesting to 
 check for common practices and emerging trends to gain insight
@@
 pharmaceutical markets.
-It can uncover what kind of sicknesses and what kind of medicines
-are used sensitive medicines are used without 
+It can uncover what kind of sicknesses and what kind of
+sensitive medicines are used without 
 appropriate medical supervision.
@@
 Each morpheme is associated with several tags such as 
-morphological, part of speech, grammar, and semantic tags. 
+morphological, part of speech (POS), grammar, and semantic tags. 
 These analyzers are essential in several text analysis tools such 
@@
 of the word, checking for morphological correctness of each 
-partition, and then inspecting the part of speech tags of each
+partition, and then inspecting the POS tags of each
 correct solution. 
@@
 Toutanova in~\cite{Tou09} suggested that using 
-a joint model to do lemmatization and part of speech (POS
+a joint model to do lemmatization and part of speech 
 tagging performs better
@@
 
-We propose the CATARM query language with a user friendly
+We propose to build 
+the CATARM query language with a user friendly
 visual representation similar to that in Figures~\ref{f:literature} and
@@
 by the user. 
-We will specify the CATARM query language in a formal grammar.
+We will specify the CATARM query language in a formal grammar
+that will be structured to resemble the question sentence
+structure of the standard common Arabic.
+This will be instrumental in identifying features
+unique to the Arabic language in the analysis as we will
+discuss later.
 
@@
 We will build our morphological analyzer on top 
-of this LCM addition to the current existing
+of this LCM in addition to the current existing
 concatenative-based analysis techniques.
@@
 and are expensive to compute dynamically. 
-Our intuition is that Arabic allows relational structures
-such as noun sentences and complex hierarchical entities.
+Our intuition is that Arabic allows a higher order
+of relational semantics in structures
+such as noun sentences and complex hierarchical noun 
+entities.
 
@@
 $L_A$ be the grammar that expresses the Arabic linguistic
-computational model, and let $q$ be a user query.
+computational model, and let $q$ be a user query expressed
+in $L_q$.
 We compute $\rho_q$, the grammar path that matches 
@@
 We think the possible simplifications that we can perform
-here are Language specific as they will reflect the 
+here are Arabic language specific as they will reflect the 
 correspondence between the formal query language definition $L_q$, 
@@
 For example the concatenative rules used in Buckwalter~\cite{Tim04}
-can be encoded in three FSM with linear execution time using 
+can be encoded in three FSMs with linear execution time using 
 the trie~\cite{Aoe:89} data structure.
@@
 The use of FSMs also suits our case-based approach. 
-The controller can be a FSM machine on its own and its decision
+The controller can be an FSM machine on its own and its decision
 will be another input into the internal morphological FSM structures. 
@@
 FSM can also be easily integrated with hidden Markov models (HMM)
-extracted from the sets of documents the MA will consider.
+extracted from the sets of documents that the analyzer will consider.
 
@@
 
-In ~\cite{Dao09}, Daoud observes that the pipeline model,
+We will explore extracting our model from formal text book
+definitions of the Arabic language~\cite{Abd00,Abd001} and seek
+help from Arabic linguists to relax these formal rules to
+allow common mistakes. 
+We will encode the model in a grammar that we can synthesize
+into an FSM and incorporate that with our current FSM 
+morphological analyzer.
+
+Our LCM will differ from ElixirFM~\cite{Otakar:07}
+in that it will follow a relational approach.
+In a relation, an element can have several associates.
+In a function, an element is mapped to only one associate. 
+We will express our morphological solutions in 
+a directed acyclic graph (DAG)
+representing a hierarchy of relations between the morphemes.
+This is more succinct than morphological trees mapping the 
+morphemes to the solutions as in~\cite{Otakar:07}.
+Furthermore, interesting relations between the morphemes 
+can be computed statically before
+the input is presented to the analyzer.
+
+In~\cite{Dao09}, Daoud observes that the pipeline model,
 where morphology is performed first and syntactic and other
@@
 
-We will explore extracting our model from formal text book
-definitions of the Arabic language~\cite{Abd00,Abd001} and seek
-help from Arabic linguists to relax these formal rules to
-allow common mistakes. 
-We will encode the model in a grammar that we can synthesize
-into an FSM and incorporate that with our current FSM 
-morphological analyzer.
-
 {\bf User query language.~~}
@@
 $L_A$ the Arabic linguistic
-computational model, and  $q$ a user query
+computational model, and  $q$ a user query,
 we compute $\rho_q$, the grammar path that matches 
@@
 The HMM edges are labeled with weights that reflect the
+probability of the transitions.
 We will look at building a correspondence between 
@@
 so far. The nodes represent the entities and the previously 
-extracted relations and the edges represent the newly discovered 
+extracted relations and the edges represent the discovered 
 relations.
@@
 
-We will need to augment the tags in the lexicons of the 
-morphological analyzer to allow for better entity and relational
+We plan to make use of existing annotated tree banks and
+modern Arabic dictionaries
+to augment the tags in the lexicons of the 
+morphological analyzer and allow for better entity and relational
 extractors.
-We plan to make use of existing annotated tree banks and of
-modern Arabic dictionaries to do that.
 
 {\bf Statistical components.~~}
-We will make use and build statistical components that can be used
-in CATARM. 
+We will make use and build standard and classical technical 
+statistical components that can be used in CATARM. 
 We will need statistical components to rank solutions, compute 
-probable decisions, estimating good thresholds.
+probable decisions, and estimate good thresholds. 
+We can make use of current existing solutions such as the 
+${\cal R}$ statistical library from the finance field.
 
@@
 higher order logic~\cite{Z308} solvers if needed. 
-These solvers can solve satisfiability questions.
+These solvers can answer satisfiability questions.
 
@@
 This means that we will define refined incremental items of 
-work that allows for more frequent deliverables. 
+work that allow for more frequent deliverables. 
 This also means that the code base will always be subject
@@
 We detailed an aggressive work plan in Table~\ref{t:workplan}. 
-The plan spans over three years each with four quarters. 
+The plan spans over three years, each with four quarters. 
 The row headers describe work items, and the column headers
@@
 However, they can start since we have already
-developed an working analyzer as a preliminary step. 
+developed a working analyzer as a preliminary step. 
 The rest of the items can all work in parallel.
@@
 mining tools as well as open source Arabic text analysis tools 
-based on NLP such as MADA+TOKAN~\cite{Rot08}, 
+based on NLP such as ElixirFM~\cite{Otakar:07}, Buckwalter~\cite{Tim04},
+MADA+TOKAN~\cite{Rot08}, 
 NEMLAR~\cite{RAl09}, and IJAES~\cite{Int09}.
@@
 subversion~\cite{Sub09} as a source control tool.
-
-The Lead PI will conduct biweekly meetings to track progress 
+This will improve team communication across the two work sites.
+
+The LPI will conduct weekly meetings to track progress 
 on the project.
@@
 
-\subsection{Novelty}
-
-We claim the novelty of the following items
-\begin{itemize}
-\item the case-based morphological analyzer,
-\item the flexible and partial linguistic computational model,
-\item the language specific LCM simplification techniques,
-\item the application of language independent statistical techniques
-to simplify the LCM,
-\item the solutions of the four proposed case studies,
-\item and the CATRAM framework as an integration of all the above techniques.
-\end{itemize}
+%\subsection{Novelty}
+
+%We claim the novelty of the following items
+%\begin{itemize}
+%\item the case-based morphological analyzer,
+%\item the flexible and partial linguistic computational model,
+%\item the language specific LCM simplification techniques,
+%\item the application of language independent statistical techniques
+%to simplify the LCM,
+%\item the solutions of the four proposed case studies,
+%\item and the CATRAM framework as an integration of all the above techniques.
+%\end{itemize}
 
@@
 \item One course buyout per year for the Co-LPI
-\item Two weeks consultant for the second and the third years
+\item Two week consultants for the second and the third years
 \item Licensing and membership fees to acquire digital libraries and corpora
@@
 formal text mining queries from other text mining domains such
-as the gene ontology domain~\cite{GeneOntology10},
+as the gene ontology domain~\cite{GeneOntology10}.
 
@@
 
-We will develop and evaluate our two optimization methodologies
-and look for other optimizations. 
+We will develop and evaluate our two optimization 
+and simplification methodologies
+and look for other optimizations and simplifications. 
 First we will use language specific features based on the 
@@
 \item We will explore further optimization techniques
-that can help simplify and the analysis. 
+that can help simplify the analysis. 
 \item We will explore generalizing our results to languages 
@@
 We plan to have publications on each of the 
-linguistic computational model, the language specific techniques
-to simplify it, and the language independent techniques to simplify
-it. 
+linguistic computational model (LCM), the language specific techniques
+to simplify the LCM, and the language independent techniques to simplify
+the LCM. 
 We also plan to have a publication on each of the separate case 
@@
 
-The team will also hold workshops to educate possible users of the
+The team will also hold workshops and tutorials 
+to educate possible users of the
 tools at AUB, QU, and in other places of interest. 

### 2011-02-18 trunk/ATSarf/src/case/hadith.h
started working on sarf paper for short paper, and worked on introduction but will later merge literature review into the introduction or some part of the introduction into the literature review

@@ -14,7 +14,7 @@
 //#define SHOW_RANKS
 //#define SHOW_VERBOSE_RANKS
 //#define STATS
-#define HADITHDEBUG  //just for additional debugging statements displayed
+//#define HADITHDEBUG  //just for additional debugging statements displayed
 //#define TENTATIVE //are decisions that we are not sure about
 #define IBN_START
 #define REFINEMENTS

### 2011-02-18 tex/sarf.aclhlt.long.tex
started working on sarf paper for short paper, and worked on introduction but will later merge literature review into the introduction or some part of the introduction into the literature review

--- 2010-12-17 tex/sarf.aclhlt.tex
@@
 \begin{abstract}
-Natural language processing uses morphological 
+Natural language processing (NLP) uses morphological 
 analysis to abstract and annotate text.
@@
 Current morphological analyzers~\cite{Sughaiyer:04}
-use concatenative analysis
-consider the internal structure 
+use concatenative analysis when
+considering the internal structure 
 of an Arabic word and
-compose it into several {\em morphemes}. 
+composing it into several {\em morphemes}. 
 A morpheme can be a {\em stem}, or an {\em affix}.
@@
 machine (FSM) based analyzers as described in~\cite{Beesley:03}.
-The efficiency of the resulting analyzer depends on the
+However, the efficiency of the resulting analyzer depends on the
 way the Xerox rules are written. 
@@
 prefix that connects only to a verb,
-the analyzer can answer without considering the 
-rest of the word.
+%the analyzer can answer without considering the 
+%rest of the word.
+we do not need to further analyze the rest of the word 
+to provide an answer.
 
@@
 study. 
-The analyzers may find different solutions or 
+The analyzers may report different solutions or 
 may present the solutions in different orders. 
@@
 appropriate for the case study. 
-We find strong evidence in our results 
-that a case-based morphological analyzer that allows a case-based 
+Moreover, our results strongly support our claim that
+a case-based morphological analyzer that allows a case-based 
 controller to intervene at every decision to guide, use, order, and 
@@
 In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
-will be available online as an open source tool.},
+has been submitted with the paper and will be available online as an open source tool.},
  a {\em novel case-based efficient
@@
 driven by a case-based controller.
-Each machine in Sarf represents a valid morphological analysis
-and takes as input one letter at a time.
+Each machine in Sarf takes one letter at a time as input
+and represents a valid morphological analysis.
 The case-based controller prunes false positives
@@
 and each sanad into the 
-separate narrators. Now, we can look each narrator 
+separate narrators. Later, we can look each narrator 
 up in the biography books. 
@@
 proper names occur rarely.% as long %as the text is valid.
-We are also interested in discovering a words 
+We are also interested in discovering words 
 that relate names to each other such as \RL{bin} or ``the son of''
@@
 %refinement to the analyzer that increased low level analysis accuracy.
-We make the following key contributions. 
+
+We make the following key contributions: 
 \begin{enumerate}
@@
 It differs from Buckwalter in that it defines a shorter list of affixes
-and a longer list of affix compatibility rules to allow recursive 
-affixes so that we can have prefix-prefix an suffix-suffix 
+and a longer list of affix compatibility rules to allow 
+affixes to be defined recursively
+so that we can have prefix-prefix an suffix-suffix 
 concatenations.
@@
 to compose the FSAs into a single framework. 
-Sarf constructs an NDFSA from
+Sarf however constructs an NDFSA from
 the controller and the prefix, stem, and suffix deterministic 
 transducers to solve the problem.
-Sarf differs in that it uses recursive affixes in the 
+Sarf also differs in that it uses recursive affixes in the 
 affix transducers, and in that the transducers consider
@@
 %controller. 
-Beesley requires recompilation.% of the Xerox rules.
+Beesley requires recompilation of the Xerox rules.
 
@@
 when parsing the
-string \RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
+string \RL{wsyl`b-h-aa al-lA`bwn}.
+%\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
 \footnote{ 
@@
 with the prefix \RL{wsy}. 
-Sarf keeps a compatibility category values as part
+Sarf keeps compatibility category values as part
 of the accept states. 
@@
 are all deterministic and cycle free as discussed 
-is Sections~\ref{sec:stemFSA} and~\ref{sec:affixFSA}.
+is Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}.
 The cycles are introduced in the abstract machine
@@
 
-The algorithm \CodeIn{NDSarf} 
+The algorithm \CodeIn{NDSarf} (detailed below)
 takes as input a text string $L$ and a case-based controller
@@
 Notice that spawning a new machine requires only keeping
-the state of the machine and all machines share the
+the state of the machine since all machines share the
 structures of the three transducers.
@@
 in Islamic studies. 
-Based on that Islamic scholars issue rules that affect the life
+Based on them Islamic scholars issue rules that affect the life
 of people around the world. 
@@
 We segment each individual hadith into two parts. 
-The first part is the sanad and it
+The first part is the sanad which
 contains the chain of narrators who related the hadith
@@
 
-The segmentation accuracy reports on the accuracy of 
+The ``segmentation accuracy'' reports on the accuracy of 
 segmentation of the book into separate hadiths.
@@
 The same applies to all other accuracy measures.
-The chain accuracy reflects the percentage of correct narrators 
+The ``chain accuracy'' reflects the percentage of correct narrators 
 covered by extracted chains.
@@
 We measured the time it took to accomplish both tasks
-less the time spent streaming the annotations.
+ignoring the time spent streaming the annotations.
 We did the same for ElixirFM.

### 2011-02-21 trunk/tex/sarf.aclhlt.tex
made some structural changes, and added some needed sentences, although this did not reduce size, but helped creating a better structure of the paper, reduction should come next after we agree on what to keep and what to remove

@@
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Case-based Arabic Morphological Analyzer for Arabic Text Mining}
+\title{Application Specific Arabic Morphological Analyzer for Arabic Text Mining}
 
@@
 This greatly increases the ambiguity of Arabic text. 
-Arabic letters can have up to 
+Arabic letters can have up to %(maybe reduce this sentence)
 four different forms
@@
 
+{\bf Sarf.~~~}
+In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
+has been submitted with the paper and will be available online as an open source tool.},
+ a {\em novel case-based efficient
+morphological analyzer} that uses 
+non-deterministic finite state Automata 
+driven by an application-specific controller.
+Each machine in Sarf takes one letter at a time as input
+and represents a valid morphological analysis.
+The application-specific controller prunes false positives
+early in the run by making a decision at every input letter
+to control the machines.
+The controller we present in Section~\ref{sec:controller}
+is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
+it can be a logical entity of any type.
+
+%Each alive machine in Sarf takes as input one letter at a time 
+%from a text stream. 
+Sarf does not assume that the word at hand is a token and
+performs tokenization on the fly based on morphological correctness.
+Up to our knowledge, this is the first morphological analyzer that 
+handles the ``run-on'' words problem.
+
+Sarf uses concatenative analysis with a novel 
+{\em recursive affixes} refinement,
+where an affix can be a sequence of other affixes which are concatenatively compatible.
+%An atomic affix $a$ is a standalone affix
+%that is not the concatenation of other affixes. 
+%We define an affix $f: a | a f$ to be either
+%an atomic affix $a$ or an atomic affix concatenated
+%with an affix. 
+Sarf introduces recursive affixes in order to
+better retain part of speech information and enhance the 
+efficiency of affix matching.
+
+Our main focus when building Sarf, was not on improving the semantical rules and expanding the morphological alternatives, instead we worked on adding support for higher level control by the specific application utilizing the lower level morphological analyzer.
+In other words, application-specific controller, has full control over and directs the morphological analysis of Sarf.
+% (sentence should be reworded differently, but is needed to explain why we mention higher and lower accuracy hypothesis, which seems vague otherwise)
+
+We hypothesize that many NLP case studies need the 
+morphological analyzer to answer queries that do not need 
+high accuracy at the low morphological analysis level.
+The analysis at the higher levels 
+(i.e. answering the application-specific query)
+can often compensate for 
+tolerable inaccuracy at lower morphological levels. 
+For example, if a query concerns proper names and the 
+analyzer is considering a 
+prefix that connects only to a verb,
+%the analyzer can answer without considering the 
+%rest of the word.
+we do not need to further analyze the rest of the word 
+to provide a negative answer.
+
+We find evidence to our hypothesis in~\cite{Maamouri:10}.
+The addition of a new corpus to the Arabic Tree Bank 
+(ATB)~\cite{Maamouri:04}
+%required compliance with the DARPA GALE program 
+%including several NLP tasks such as English translation 
+%annotations and word-level alignment of Arabic and English. 
+%The new data 
+challenged the existing processing and 
+annotation guidelines. 
+This led to refining 
+SAMA~\cite{Kulick:10} with 
+alternative guidelines. 
+This refinement was achieved only after the NLP 
+task interacted with the morphological analyzer,
+a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
+In addition, 
+the work of Habash~\shortcite{Habash:06} reports that different 
+types of morphological analyses behave differently on the same %case study. 
+data set.
+The analyzers may report different solutions or 
+may present the solutions in different orders. 
+Clearly, some of 
+the presented solutions and their order of presentation may not be 
+appropriate for the %case study.
+application needs, which is an issue that Sarf tackles. % (maybe reworded)
+Moreover, our results strongly support our claim that
+an application-specific morphological analyzer that allows a application-specific 
+controller to intervene at every decision to guide, use, order, and 
+refine the morphological analysis is of high utility.
+We validate our hypothesis using a case study from the Islamic 
+literature, introduced later in Section~\ref{sec:islamic}.
+
+The rest of this paper is structured as follows. In Section~\ref{sec:related}
+we compare Sarf to related work. In Section~\ref{sec:sarf}
+we discuss Sarf and illustrate it with a running example.
+%We also explain how non-deterministic finite state Automata
+%work. 
+In Section~\ref{sec:islamic} we present the case
+study. We present our results in Section~\ref{sec:results}
+and conclude with future work in Section~\ref{sec:future}.
+
+% does not have to be a subsection, 
+%just to highlight it as an item we should not miss
+
+%\section{Background}
+% TODO: nondeterministic finite-state automata
+
+\section{Related work }
+\label{sec:related}
+
+The survey in~\cite{Sughaiyer:04} compares
+several morphological analyzers. 
+Analyzers such as~\cite{Khoja:01,Darwish:02} 
+target specific applications in the 
+analyzer itself or use a specific set of POS tags
+as their reference.
+Sarf differs in that it is a general morphological 
+analyzer that reports all possible solutions. 
+It is application-specific in the sense that a controller 
+prunes these solutions.
+
 Current morphological analyzers such as 
@@
 and enumerate all possible solutions. 
-This approach has several problems. 
+%This approach has several problems. 
+However, we note the following points that deserve consideration:
+%(We need not to mention especially point 2 as a problem according to reviewer comments)
 First, a white space delimited token may have 
 more than one word.
-Second, the exhaustive enumeration may hurt performance and may
-not be necessary or appropriate
-in some case studies as noted in~\cite{Maamouri:10}. 
+%Second, the exhaustive enumeration may hurt performance and may
+%not be necessary or appropriate
+%in some case studies as noted in~\cite{Maamouri:10}. 
+Second, in many applications, the exhaustive enumeration may not be needed or appropriate as noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.
 Other morphological analyzers such as 
@@
 MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
+% (reviewer wants to distinguish MADA as a tool for morphological disambiguation instead of analysis)
 use machine learning and support vector machines (SVM) 
 to enhance the accuracy of the morphological analysis at the expense 
-of performance.
+of performance. % (we should stress that we mean performance in time and not accuracy although evident since some reviewers did not get)
 Xerox rules can be compiled into specialized finite state
 machine (FSM) based analyzers as described in~\cite{Beesley:03}.
-The efficiency of the resulting analyzer depends on the
+However, the efficiency of the resulting analyzer depends on the
 way the Xerox rules are written. 
@@
 
-We hypothesize that many NLP case studies need the 
-morphological analyzer to answer queries that do not need 
-high accuracy at the low morphological analysis level.
-The analysis at the higher levels can often compensate for 
-tolerable inaccuracy at low levels. 
-For example, if a query concerns proper names and the 
-analyzer is considering a 
-prefix that connects only to a verb,
-the analyzer can answer without considering the 
-rest of the word.
-
-We find evidence to our hypothesis in~\cite{Maamouri:10}.
-The addition of a new corpus to the Arabic Tree Bank 
-(ATB)~\cite{Maamouri:04}
-required compliance with the DARPA GALE program 
-including several NLP tasks such as English translation 
-annotations and word-level alignment of Arabic and English. 
-The new data challenged the existing processing and 
-annotation guidelines. 
-This lef to refining 
-SAMA~\cite{Kulick:10} with 
-alternative guidelines where the NLP 
-task interacted with the morphological analyzer.
-
-The work of Habash~\shortcite{Habash:06} reports that different 
-types of morphological analyses behave differently on the same case 
-study. 
-The analyzers may find different solutions or 
-may present the solutions in different orders. 
-The presented solutions and the order of presentation may not be 
-appropriate for the case study. 
-We find strong evidence in our results 
-that a case-based morphological analyzer that allows a case-based 
-controller to intervene at every decision to guide, use, order, and 
-refine the morphological analysis is of high utility.
-
-%\subsection{Sarf}
-%\label{sec:intro:sarf}
-
-{\bf Sarf.~~~}
-In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
-will be available online as an open source tool.},
- a {\em novel case-based efficient
-morphological analyzer} that uses 
-non-deterministic finite state Automata 
-driven by a case-based controller.
-Each alive machine in Sarf represents a valid morphological analysis
-and takes as input one letter at a time.
-The case-based controller prunes false positives
-early in the run by making a decision at every input letter
-to control the alive machines.
-%Each alive machine in Sarf takes as input one letter at a time 
-%from a text stream. 
-Sarf does not assume that the word at hand is a token and
-performs tokenization on the fly based on morphological correctness.
-Up to our knowledge, this is the first morphological analyzer that 
-handles the ``run-on'' words problem.
-
-Sarf uses concatenative analysis with a novel 
-{\em recursive affixes} refinement. 
-Sarf introduces recursive affixes in order to
-better retain part of speech information and enhance the 
-efficiency of affix matching.
-
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{.9\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
-\label{f:exhadith}
-}
-\end{figure}
-% the FSMs for the three words
-\transtrue
-
-We validate our hypothesis using a case study from the Islamic 
-literature. 
-A \RL{.hady_t} is a narration related to the prophet Mohammad
-through a \RL{sanad} or a sequence of narrators. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
-Arabic with its transliteration and translation. 
-The authenticity of a \noArRL{.hady_t} depends on 
-the credibility of the narrators as reported in 
-separate biography books. 
-In this paper, we consider the problem of automatically segmenting
-a \noArRL{.hady_t} book into narrations, each 
-narration into its \RL{matn} (content) and its \noArRL{sanad},
-and each sanad into the 
-separate narrators. Now, we can look each narrator 
-up in the biography books. 
-For example, 
-Figure~\ref{f:exhadith} shows proper names in boxes connected
-to form complex names of narrators. 
-For example, 
-\noTrRL{qtybT} is the first name 
-of narrator $n_1$, and 
-\noTrRL{s`yd} is the name of his father as 
-the word \noTrRL{bn} (son of) indicates.
-
-The hadith case study is interesting to Sarf since the controller
-considers %most of the analyses where we have a concentration
-proper names and ignores text where
-proper names occur rarely.% as long %as the text is valid.
-We are also interested in discovering a words 
-that relate names to each other such as \RL{bin} or ``the son of''
-and words that mean narrate such as \RL{qaal} or ``said''.
-
-With Sarf, we successfully automated the analysis of 
-three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
-and extracted from each one of them a vector of complex structures 
-that are three level deep in hierarchy.
-%We also showcase that the results of the accurate high level 
-%abstraction
-%results of the case-based analysis was not affected by a major
-%refinement to the analyzer that increased low level analysis accuracy. 
-We make the following key contributions. 
-\begin{enumerate}
-\item {\bf Case-based analysis:}  We designed and implemented
-a case-based morphological analyzer where a controller
-guides the analysis. 
-%\item {\bf Exhaustive analysis:} Since we have a case-based 
-%controller that can decide on the fly whether an analysis is 
-%accepted or not, we can afford to keep all valid analyses modulo
-%those pruned by the case controller. 
-%We do that using non-deterministic FSAs. 
-\item {\bf Recursive affixes:}
-We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02} and define affixes
-recursively. 
-This allows Sarf to perform better and 
-maintain compositional part of speech (POS) tags.
-\item {\bf Hadith case study:}
-%We evaluated Sarf using the hadith case study. 
-We were able to efficiently extract a complex three 
-level hierarchical 
-structure containing the chains of narrators 
-with high accuracy. 
-\end{enumerate}
-
-% does not have to be a subsection, 
-%just to highlight it as an item we should not miss
-
-The rest of this paper is structured as follows. In Section~\ref{sec:related}
-we compare Sarf to related work. In Section~\ref{sec:sarf}
-we discuss Sarf and illustrate it with a running example.
-%We also explain how non-deterministic finite state Automata
-%work. 
-In Section~\ref{sec:islamic} we present the hadith case
-study. We present our results in Section~\ref{sec:results}
-and conclude with future work in Section~\ref{sec:future}.
-
-%\section{Background}
-% TODO: nondeterministic finite-state automata
-
-\section{Related work }
-\label{sec:related}
-
-The survey in~\cite{Sughaiyer:04} compares
-several morphological analyzers. 
-Analyzers such as~\cite{Khoja:01,Darwish:02} 
-target specific applications in the 
-analyzer itself or use a specific set of POS tags
-as their reference.
-Sarf differs in that it is a general morphological 
-analyzer that reports all possible solutions. 
-It is case-based in the sense that a controller 
-prunes these solutions.
-
 Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
 It differs from Buckwalter in that it defines a shorter list of affixes
-and a longer list of affix compatibility rules to allow recursive 
-affixes so that we can have prefix-prefix an suffix-suffix 
+and a longer list of affix compatibility rules to allow 
+affixes to be defined recursively
+so that we can have prefix-prefix an suffix-suffix 
 concatenations.
@@
 to compose the FSAs into a single framework. 
-Sarf constructs an NDFSA from
+Sarf however constructs an NDFSA from
 the controller and the prefix, stem, and suffix deterministic 
 transducers to solve the problem.
-Sarf differs in that it uses recursive affixes in the 
+Sarf also differs in that it uses recursive affixes in the 
 affix transducers, and in that the transducers consider
 input from the controller to modify their behavior. 
-%With Sarf it is easy to control the FSA with a case based
+%With Sarf it is easy to control the FSA with a case-based
 %controller. 
-Beesley requires recompilation.% of the Xerox rules.
+Beesley requires recompilation of the Xerox rules. % (don't we do also for any change in the controller code?)
 
@@
 done manually and worked in integration with the ATB format. 
-Sarf is a case-based controller approach that can modify 
+Sarf is an application-specific controller approach that can modify 
 the behavior of the analyzer on the fly and can interact
@@
 can benefit from many features unique to the Arabic language.
-Sarf leaves implementing that to the case-based controller
+Sarf leaves implementing that to the application-specific controller
 since in many cases the abstraction needs only a partial 
@@
 with the prefix \RL{wsy}. 
-Sarf keeps a compatibility category values as part
+Sarf keeps compatibility category values as part
 of the accept states. 
 %Sarf considers the value of the category as part of the state.
-Thus each accecpt state in Figure~\ref{f:example} represents
+Thus each accept state in Figure~\ref{f:example} represents
 more than one concrete state. 
@@
 
+% TODO: check how to merge this part here
+\transfalse
+\begin{figure}[tb]
+\center{
+\resizebox{.9\columnwidth}{!}
+{ \input{figs/exhadith.pdftex_t}}
+\caption{Hadith abstraction example.}
+\label{f:exhadith}
+}
+\end{figure}
+% the FSMs for the three words
+\transtrue
+
+We validate our hypothesis using a case study from the Islamic 
+literature. 
+A \RL{.hady_t} is a narration related to the prophet Mohammad
+through a \RL{sanad} or a sequence of narrators. 
+Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
+Arabic with its transliteration and translation. 
+The authenticity of a \noArRL{.hady_t} depends on 
+the credibility of the narrators as reported in 
+separate biography books. 
+In this paper, we consider the problem of automatically segmenting
+a \noArRL{.hady_t} book into narrations, each 
+narration into its \RL{matn} (content) and its \noArRL{sanad},
+and each sanad into the 
+separate narrators. Later, we can look each narrator 
+up in the biography books. 
+For example, 
+Figure~\ref{f:exhadith} shows proper names in boxes connected
+to form complex names of narrators. 
+For example, 
+\noTrRL{qtybT} is the first name 
+of narrator $n_1$, and 
+\noTrRL{s`yd} is the name of his father as 
+the word \noTrRL{bn} (son of) indicates.
+
+The hadith case study is interesting to Sarf since the controller
+considers %most of the analyses where we have a concentration
+proper names and ignores text where
+proper names occur rarely.% as long %as the text is valid.
+We are also interested in discovering words 
+that relate names to each other such as \RL{bin} or ``the son of''
+and words that mean narrate such as \RL{qaal} or ``said''.
+
+With Sarf, we successfully automated the analysis of 
+three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
+and extracted from each one of them a vector of complex structures 
+that are three level deep in hierarchy.
+%We also showcase that the results of the accurate high level 
+%abstraction
+%results of the case-based analysis was not affected by a major
+%refinement to the analyzer that increased low level analysis accuracy.
+
 A hadith in Islamic literature is a narration from the prophet Mohammad
@@
 current morphological analyzers.
+
+%TODO: merge into conclusion, taken from introduction
+We make the following key contributions: 
+\begin{enumerate}
+\item {\bf Case-based analysis:}  We designed and implemented
+a case-based morphological analyzer where a controller
+guides the analysis. 
+%\item {\bf Exhaustive analysis:} Since we have a case-based 
+%controller that can decide on the fly whether an analysis is 
+%accepted or not, we can afford to keep all valid analyses modulo
+%those pruned by the case controller. 
+%We do that using non-deterministic FSAs. 
+\item {\bf Recursive affixes:}
+We refine the concatenative analysis of 
+Buckwalter~\shortcite{Buckwalter:02} and define affixes
+recursively. 
+This allows Sarf to perform better and 
+maintain compositional part of speech (POS) tags.
+\item {\bf Hadith case study:}
+%We evaluated Sarf using the hadith case study. 
+We were able to efficiently extract a complex three 
+level hierarchical 
+structure containing the chains of narrators 
+with high accuracy. 
+\end{enumerate}
 

### 2011-02-21 tex/sarf.aclhlt.tex
some minimizations for Introduction

@@
 appropriate for the case study at hand. 
-In this paper we present Sarf, a case-based morphological 
-analyzer for Arabic. 
-Sarf allows a case-based controller to control and refine
+In this paper we present Sarf, an application-specific morphological analyzer for Arabic. 
+Sarf allows an application-specific controller to control and refine
 the state of the analysis on the fly. 
@@
 We automated the analysis of three books of Islamic
-narrations (hadith) using Sarf where we
+narrations (hadith) using Sarf, where we
 abstracted each book into a vector of complex structures.
@@
 analysis techniques for Arabic text. 
-Current morphological analyzers~\cite{Sughaiyer:04}
-use concatenative analysis
-consider the internal structure 
-of an Arabic word and
-compose it into several {\em morphemes}. 
-A morpheme can be a {\em stem}, or an {\em affix}.
-An affix can be a {\em prefix, suffix, } or an {\em infix}.
-The analysis of one word may lead to several possible
-morphological solutions.
-\vocalize
-For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
-may have two valid morphological analyses. 
-The letter \RL{'a} may be a prefix and the word means 
-``I praise him'', or 
-%The letter \RL{'a} may also be 
-part of the stem \RL{'a.hmad} (a proper noun)
-and the word means ``his Ahmad''.
-
-\novocalize
+%Current morphological analyzers~\cite{Sughaiyer:04}
+%use concatenative analysis when
+%considering the internal structure 
+%of an Arabic word and
+%composing it into several {\em morphemes}. 
+%A morpheme can be a {\em stem}, or an {\em affix}.
+%An affix can be a {\em prefix, suffix, } or an {\em infix}.
+%The analysis of one word may lead to several possible
+%morphological solutions.
+%\vocalize
+%For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
+%may have two valid morphological analyses. 
+%The letter \RL{'a} may be a prefix and the word means 
+%``I praise him'', or 
+%%The letter \RL{'a} may also be 
+%part of the stem \RL{'a.hmad} (a proper noun)
+%and the word means ``his Ahmad''.
+%
+%\novocalize
 The accuracy of the solutions suffer due to inherent difficulties
@@
 For example, it is common practice to write Arabic text
-without short vowels. 
+without short vowels / diacritics. 
 This greatly increases the ambiguity of Arabic text.
-Arabic letters can have up to 
+%(maybe reduce the couple of sentences below)
+Moreover, Arabic letters can have up to 
 four different forms
@@
 
+%{\bf Sarf.~~~}
+In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
+has been submitted with the paper and will be available online as an open source tool.},
+ a {\em novel application-specific efficient
+morphological analyzer} that uses 
+non-deterministic finite state Automata 
+driven by an application-specific controller.
+Each machine in Sarf takes one letter at a time as input
+and represents a valid morphological analysis.
+The application-specific controller prunes false positives
+early in the run by making a decision at every input letter
+to control the machines.
+The controller we present in Section~\ref{sec:controller}
+is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
+it can be a logical entity of any type.
+
+%Each alive machine in Sarf takes as input one letter at a time 
+%from a text stream. 
+Sarf does not assume that the word at hand is a token and
+performs tokenization on the fly based on morphological correctness.
+Up to our knowledge, this is the first morphological analyzer that 
+handles the ``run-on'' words problem.
+
+Sarf uses concatenative analysis with a novel %(maybe remove novel bc authors did not feel it is much)
+{\em recursive affixes} refinement,
+where an affix can be a sequence of other affixes which are concatenatively compatible.
+%An atomic affix $a$ is a standalone affix
+%that is not the concatenation of other affixes. 
+%We define an affix $f: a | a f$ to be either
+%an atomic affix $a$ or an atomic affix concatenated
+%with an affix. 
+Sarf introduces recursive affixes in order to
+better retain part of speech information and enhance the 
+efficiency of affix matching.
+
+%Our main focus when building Sarf, was not on improving the semantical rules and expanding the morphological alternatives, instead we worked on adding support for higher level control by the specific application utilizing the lower level morphological analyzer.
+In other words, the application-specific controller, has full control over and directs the morphological analysis of Sarf.
+% (sentence should be reworded differently, but is needed to explain why we mention higher and lower accuracy hypothesis, which seems vague otherwise)
+%
+%We hypothesize that many NLP case studies need the 
+%morphological analyzer to answer queries that do not need 
+%high accuracy at the low morphological analysis level.
+%The analysis at the higher levels 
+%(i.e. answering the application-specific query)
+%can often compensate for 
+%tolerable inaccuracy at lower morphological levels. 
+%For example, if a query concerns proper names and the 
+%analyzer is considering a 
+%prefix that connects only to a verb,
+%%the analyzer can answer without considering the 
+%%rest of the word.
+%we do not need to further analyze the rest of the word 
+%to provide a negative answer.
+%
+%We find evidence to our hypothesis in~\cite{Maamouri:10}.
+%%The addition of a new corpus to the Arabic Tree Bank 
+%%(ATB)~\cite{Maamouri:04}
+%%%required compliance with the DARPA GALE program 
+%%%including several NLP tasks such as English translation 
+%%%annotations and word-level alignment of Arabic and English. 
+%%%The new data 
+%%challenged the existing processing and 
+%%annotation guidelines. 
+%%This led to refining 
+%%SAMA~\cite{Kulick:10} with 
+%%alternative guidelines. 
+%%This refinement was achieved only after the NLP 
+%%task interacted with the morphological analyzer,
+%%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
+%It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when NLP task interacted with the morphological analyzer.
+%%a flexibility our analyzer is designed to provide for the application-specific controller on the run.
+%In addition, 
+%the work of Habash~\shortcite{Habash:06} reports that different 
+%types of morphological analyses behave differently on the same %case study. 
+%data set.
+%The analyzers may report different solutions or 
+%may present the solutions in different orders. 
+%Clearly, some of 
+%the presented solutions and their order of presentation may not be 
+%appropriate for the %case study.
+%application needs, which is an issue that Sarf tackles. % (maybe reworded)
+%Moreover, 
+Our results strongly support our claim that
+an application-specific morphological analyzer that allows a application-specific 
+controller to intervene at every decision to guide, use, order, and 
+refine the morphological analysis is of high utility.
+
+We validate our hypothesis using a case study from the Islamic 
+literature, introduced later in Section~\ref{sec:islamic}.
+
+%The rest of this paper is structured as follows. In Section~\ref{sec:related}
+%we compare Sarf to related work. In Section~\ref{sec:sarf}
+%we discuss Sarf and illustrate it with a running example.
+%%We also explain how non-deterministic finite state Automata
+%%work. 
+%In Section~\ref{sec:islamic} we present the case
+%study. We present our results in Section~\ref{sec:results}
+%and conclude with future work in Section~\ref{sec:future}.
+
+% does not have to be a subsection, 
+%just to highlight it as an item we should not miss
+
+%\section{Background}
+% TODO: nondeterministic finite-state automata
+
+\section{Related work }
+\label{sec:related}
+
+The survey in~\cite{Sughaiyer:04} compares
+several morphological analyzers. 
+Analyzers such as~\cite{Khoja:01,Darwish:02} 
+target specific applications in the 
+analyzer itself or use a specific set of POS tags
+as their reference.
+Sarf differs in that it is a general morphological 
+analyzer that reports all possible solutions. 
+It is application-specific in the sense that a controller 
+prunes these solutions.
+
 Current morphological analyzers such as 
@@
 
-We hypothesize that many NLP case studies need the 
-morphological analyzer to answer queries that do not need 
-high accuracy at the low morphological analysis level.
-The analysis at the higher levels can often compensate for 
-tolerable inaccuracy at low levels. 
-For example, if a query concerns proper names and the 
-analyzer is considering a 
-prefix that connects only to a verb,
-the analyzer can answer without considering the 
-rest of the word.
-
-We find evidence to our hypothesis in~\cite{Maamouri:10}.
-The addition of a new corpus to the Arabic Tree Bank 
-(ATB)~\cite{Maamouri:04}
-required compliance with the DARPA GALE program 
-including several NLP tasks such as English translation 
-annotations and word-level alignment of Arabic and English. 
-The new data challenged the existing processing and 
-annotation guidelines. 
-This led to refining 
-SAMA~\cite{Kulick:10} with 
-alternative guidelines where the NLP 
-task interacted with the morphological analyzer.
-
-The work of Habash~\shortcite{Habash:06} reports that different 
-types of morphological analyses behave differently on the same case 
-study. 
-The analyzers may find different solutions or 
-may present the solutions in different orders. 
-The presented solutions and the order of presentation may not be 
-appropriate for the case study. 
-We find strong evidence in our results 
-that a case-based morphological analyzer that allows a case-based 
-controller to intervene at every decision to guide, use, order, and 
-refine the morphological analysis is of high utility.
-
-%\subsection{Sarf}
-%\label{sec:intro:sarf}
-
-{\bf Sarf.~~~}
-In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
-will be available online as an open source tool.},
- a {\em novel case-based efficient
-morphological analyzer} that uses 
-non-deterministic finite state Automata 
-driven by a case-based controller.
-Each machine in Sarf represents a valid morphological analysis
-and takes as input one letter at a time.
-The case-based controller prunes false positives
-early in the run by making a decision at every input letter
-to control the machines.
-The controller we present in Section~\ref{sec:controller}
-is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
-it can be a logical entity of any type.
-
-%Each alive machine in Sarf takes as input one letter at a time 
-%from a text stream. 
-Sarf does not assume that the word at hand is a token and
-performs tokenization on the fly based on morphological correctness.
-Up to our knowledge, this is the first morphological analyzer that 
-handles the ``run-on'' words problem.
-
-Sarf uses concatenative analysis with a novel 
-{\em recursive affixes} refinement. 
-An atomic affix $a$ is a standalone affix
-that is not the concatenation of other affixes. 
-We define an affix $f: a | a f$ to be either
-an atomic affix $a$ or an atomic affix concatenated
-with an affix. 
-Sarf introduces recursive affixes in order to
-better retain part of speech information and enhance the 
-efficiency of affix matching.
-
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{.9\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
-\label{f:exhadith}
-}
-\end{figure}
-% the FSMs for the three words
-\transtrue
-
-We validate our hypothesis using a case study from the Islamic 
-literature. 
-A \RL{.hady_t} is a narration related to the prophet Mohammad
-through a \RL{sanad} or a sequence of narrators. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
-Arabic with its transliteration and translation. 
-The authenticity of a \noArRL{.hady_t} depends on 
-the credibility of the narrators as reported in 
-separate biography books. 
-In this paper, we consider the problem of automatically segmenting
-a \noArRL{.hady_t} book into narrations, each 
-narration into its \RL{matn} (content) and its \noArRL{sanad},
-and each sanad into the 
-separate narrators. Now, we can look each narrator 
-up in the biography books. 
-For example, 
-Figure~\ref{f:exhadith} shows proper names in boxes connected
-to form complex names of narrators. 
-For example, 
-\noTrRL{qtybT} is the first name 
-of narrator $n_1$, and 
-\noTrRL{s`yd} is the name of his father as 
-the word \noTrRL{bn} (son of) indicates.
-
-The hadith case study is interesting to Sarf since the controller
-considers %most of the analyses where we have a concentration
-proper names and ignores text where
-proper names occur rarely.% as long %as the text is valid.
-We are also interested in discovering a words 
-that relate names to each other such as \RL{bin} or ``the son of''
-and words that mean narrate such as \RL{qaal} or ``said''.
-
-With Sarf, we successfully automated the analysis of 
-three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
-and extracted from each one of them a vector of complex structures 
-that are three level deep in hierarchy.
-%We also showcase that the results of the accurate high level 
-%abstraction
-%results of the case-based analysis was not affected by a major
-%refinement to the analyzer that increased low level analysis accuracy. 
-We make the following key contributions. 
-\begin{enumerate}
-\item {\bf Case-based analysis:}  We designed and implemented
-a case-based morphological analyzer where a controller
-guides the analysis. 
-%\item {\bf Exhaustive analysis:} Since we have a case-based 
-%controller that can decide on the fly whether an analysis is 
-%accepted or not, we can afford to keep all valid analyses modulo
-%those pruned by the case controller. 
-%We do that using non-deterministic FSAs. 
-\item {\bf Recursive affixes:}
-We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02} and define affixes
-recursively. 
-This allows Sarf to perform better and 
-maintain compositional part of speech (POS) tags.
-\item {\bf Hadith case study:}
-%We evaluated Sarf using the hadith case study. 
-We were able to efficiently extract a complex three 
-level hierarchical 
-structure containing the chains of narrators 
-with high accuracy. 
-\end{enumerate}
-
-% does not have to be a subsection, 
-%just to highlight it as an item we should not miss
-
-The rest of this paper is structured as follows. In Section~\ref{sec:related}
-we compare Sarf to related work. In Section~\ref{sec:sarf}
-we discuss Sarf and illustrate it with a running example.
-%We also explain how non-deterministic finite state Automata
-%work. 
-In Section~\ref{sec:islamic} we present the hadith case
-study. We present our results in Section~\ref{sec:results}
-and conclude with future work in Section~\ref{sec:future}.
-
-%\section{Background}
-% TODO: nondeterministic finite-state automata
-
-\section{Related work }
-\label{sec:related}
-
-The survey in~\cite{Sughaiyer:04} compares
-several morphological analyzers. 
-Analyzers such as~\cite{Khoja:01,Darwish:02} 
-target specific applications in the 
-analyzer itself or use a specific set of POS tags
-as their reference.
-Sarf differs in that it is a general morphological 
-analyzer that reports all possible solutions. 
-It is case-based in the sense that a controller 
-prunes these solutions.
-
 Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
 It differs from Buckwalter in that it defines a shorter list of affixes
-and a longer list of affix compatibility rules to allow recursive 
-affixes so that we can have prefix-prefix an suffix-suffix 
+and a longer list of affix compatibility rules to allow 
+affixes to be defined recursively
+so that we can have prefix-prefix an suffix-suffix 
 concatenations.
@@
 %controller. 
-Beesley requires recompilation.% of the Xerox rules.
+Beesley requires recompilation of the Xerox rules. % (don't we do also for any change in the controller code?)
 
@@
 
+%TODO: merge into conclusion, taken from introduction
+We make the following key contributions: 
+\begin{enumerate}
+\item {\bf Case-based analysis:}  We designed and implemented
+a case-based morphological analyzer where a controller
+guides the analysis. 
+%\item {\bf Exhaustive analysis:} Since we have a case-based 
+%controller that can decide on the fly whether an analysis is 
+%accepted or not, we can afford to keep all valid analyses modulo
+%those pruned by the case controller. 
+%We do that using non-deterministic FSAs. 
+\item {\bf Recursive affixes:}
+We refine the concatenative analysis of 
+Buckwalter~\shortcite{Buckwalter:02} and define affixes
+recursively. 
+This allows Sarf to perform better and 
+maintain compositional part of speech (POS) tags.
+\item {\bf Hadith case study:}
+%We evaluated Sarf using the hadith case study. 
+We were able to efficiently extract a complex three 
+level hierarchical 
+structure containing the chains of narrators 
+with high accuracy. 
+\end{enumerate}
+
 In the future, 

### 2011-02-24 tex/sarf.aclhlt.tex
contains almost everything that needed to be added, but is about 1.5 pages off the limit till now

@@
 \newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}
+\newcommand{\noTrnoVocRL}[1]{\transfalse\novocalize\noTrRL{#1}\vocalize\transtrue}  
+\newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}
 
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Application Specific Arabic Morphological Analyzer for Arabic Text Mining}
+%\title{Application-Specific Arabic Morphological Analyzer for Arabic Text Mining}% maybe add "with Enhanced Capabilities"
+\title{Application-Specific Enhanced Morphological Analyzer \\for Arabic Text Mining}% maybe add "with Enhanced Capabilities"
 
@@
 to better preserve part of speech information.
-We automated the analysis of three books of Islamic
-narrations (hadith) using Sarf, where we
-abstracted each book into a vector of complex structures.
-Our results show higher accuracy and better efficiency
+%We automated the analysis of three books of Islamic
+%narrations (hadith) using Sarf, where we
+%abstracted each book into a vector of complex structures.
+Our results show better efficiency
 compared to current analyzers.
@@
 middle, end of word and separate forms. 
-This allows the phrase \transfalse
-\RL{il_A\nospace almdrsT} \transtrue
+This allows the phrase 
+\noTrnoVocRL{il_A\nospace almdrsT}
 to be visually recognizable
-as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
+as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school) 
 without the need of a space in between. 
-The reason is the first word \RL{il_A} ends with
-\RL{_A} a non-connecting letter. 
+The reason is the first word \noVocRL{il_A} ends with
+\noVocRL{_A} a non-connecting letter. 
 These words,
@@
 has been submitted with the paper and will be available online as an open source tool.},
- a {\em novel application-specific efficient
+ a {\em novel application-specific %efficient 
+enhanced %(instead of 'efficient' to engulf both efficiency and other refinements)
 morphological analyzer} that uses 
@@
 to control the machines.
-The controller we present in Section~\ref{sec:controller}
-is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
-it can be a logical entity of any type.
+%The controller we present in Section~\ref{sec:controller}
+%is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
+%it can be a logical entity of any type.
+The controller can be an FSA or any other %logical entity 
+programming entity of any type,
+since Automatas may be a very complex solution for some simple applications.
+  %(maybe re-word this sentence)
 
@@
 %from a text stream.
+In addition, to being application-specific, Sarf has introduced 
+many enhancements to the field of arabic morphological analysis. 
+For example,
 Sarf does not assume that the word at hand is a token and
@@
 
-%Our main focus when building Sarf, was not on improving the semantical rules and expanding the morphological alternatives, instead we worked on adding support for higher level control by the specific application utilizing the lower level morphological analyzer.
-In other words, the application-specific controller, has full control over and directs the morphological analysis of Sarf.
-% (sentence should be reworded differently, but is needed to explain why we mention higher and lower accuracy hypothesis, which seems vague otherwise)
+In addition, Sarf supports disambiguation using diacritics when they are available.
+This feature, to the best of our knowledge is the first attempt to benefit from 
+diacritics instead of ignoring them. Prior to Sarf, the trend was to preprocess 
+the text to remove any such `unwelcomed' letter. %Nonetheless, Sarf regards them as a richness to be utilized that should not be disregarded.
+
+%Moreover, since Sarf performs tokenization on the fly, Sarf is not just capable of solving
+%the problem of ``run-on'' words, but also of handling Multiword Expressions (MWE). 
+Moreover, since Sarf does not perform white-space delimited tokenization, Sarf is capable of 
+handling Multiword Expressions (MWEs).
+MWEs are defined in~\cite{MWE}
+``as idiosyncratic interpretations
+that cross word boundaries (or spaces)''. 
+%By this term, we refer to frequent cases when a group of words form together a single
+%item with a semantical meaning. In this paper, we focus on the example of compound names such as 
+%\noVocRL{`bid alkarym}. 
+%In current analyzers, this phrase is analyzed word by word and in best cases is 
+%considered to contain two names, instead of being understood to be a single but compound name.
+
+Our focus when building Sarf, was not on improving the semantical rules and 
+expanding the morphological alternatives, instead we worked on adding support 
+for higher level control by the specific application utilizing the lower level 
+morphological analyzer. %In other words, the application-specific controller, has 
+% full control over and directs the morphological analysis of Sarf. 
+% (sentence should be reworded differently, but is needed to explain why we mention 
+% higher and lower accuracy hypothesis, which seems vague otherwise)
 %
-%We hypothesize that many NLP case studies need the 
+%We have noted that many NLP case studies need the 
 %morphological analyzer to answer queries that do not need 
-%high accuracy at the low morphological analysis level.
+%high accuracy at the morphological analysis level.
 %The analysis at the higher levels 
@@
 %application needs, which is an issue that Sarf tackles. % (maybe reworded)
-%Moreover, 
-Our results strongly support our claim that
-an application-specific morphological analyzer that allows a application-specific 
+We strongly believe that by being
+an application-specific morphological analyzer that allows an application-specific 
 controller to intervene at every decision to guide, use, order, and 
-refine the morphological analysis is of high utility.
+refine the morphological analysis, Sarf is of high utility.
 
-We validate our hypothesis using a case study from the Islamic 
-literature, introduced later in Section~\ref{sec:islamic}.
+%We validate our hypothesis using a case study from the Islamic 
+%literature, introduced later in Section~\ref{sec:results}.
+In addition to all its enhanced analysis methods, Sarf still
+outperformed current morphological analyzers in terms of running time.
+These results will be presented in Section~\ref{sec:results}.
+
+We make the following key contributions: 
+\begin{enumerate}
+\item {\bf application-specific analysis:}  We designed and implemented
+a application-specific morphological analyzer where a controller
+guides the analysis. 
+%\item {\bf Exhaustive analysis:} Since we have an application-specific 
+%controller that can decide on the fly whether an analysis is 
+%accepted or not, we can afford to keep all valid analyses modulo
+%those pruned by the case controller. 
+%We do that using non-deterministic FSAs. 
+\item {\bf Recursive affixes:}
+We refine the concatenative analysis of 
+Buckwalter~\shortcite{Buckwalter:02} and define affixes
+recursively. 
+This allows Sarf to perform better and 
+maintain compositional part of speech (POS) tags.
+\item {\bf Diacritic-based Disambiguation:}
+Diacritics are used to disambiguate morphological
+alternatives using any existant diacritics.
+\end{enumerate}
 
@@
 
-\section{Related work }
+\subsection{Related work }
 \label{sec:related}
@@
 
-Current morphological analyzers such as 
+Current morphological analyzers such as %(maybe disambiguators also added here according to reviewer comments)
 Buckwalter~\shortcite{Buckwalter:02},
@@
 However, we note the following points that deserve consideration:
-%(We need not to mention especially point 2 as a problem according to reviewer comments)
+%(We need not to mention especially point 2 as a "problem" according to reviewer comments)
 First, a white space delimited token may have 
-more than one word.
+more than one word or just a part of a MWE.
 %Second, the exhaustive enumeration may hurt performance and may
@@
 to enhance the accuracy of the morphological analysis at the expense 
-of performance. % (we should stress that we mean performance in time and not accuracy although evident since some reviewers did not get)
+of efficiency.
 Xerox rules can be compiled into specialized finite state
@@
 
-Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
+Similar to Sarf, both Beesley~\shortcite{Beesley:01} and Attia~\shortcite{Attia:05} use
 finite state Automata (FSA). 
-It reports that the number of Automata generated by a compiler
+Beesley reports that the number of Automata generated by a compiler
 for Xerox rules can not be controlled and also reports a difficulty 
@@
 input from the controller to modify their behavior. 
-%With Sarf it is easy to control the FSA with a case-based
+%With Sarf it is easy to control the FSA with a application-specific
 %controller. 
-Beesley requires recompilation of the Xerox rules. % (don't we do also for any change in the controller code?)
+Beesley requires recompilation of the Xerox rules. 
+Attia has researched extensively MWEs handling methods~\cite{Attia:06} and 
+automated extending their dictionary coverage~\cite{Attia:10}. 
+% (I am not sure how clear this statement is)
 
@@
 Sarf differs in that it provides an interface for the 
-case study to interfere in judging the analyses much earlier and
+application to interfere in judging the analyses much earlier and
 at a finer granularity level. 
@@
 that partition text into morphemes.
-%The case-based controller can build the trace 
+%The application-specific controller can build the trace 
 %when needed.
@@
 %We find supporting evidence in this decision to our hypothesis
-%about the utility of a case-based analyzer. 
+%about the utility of a application-specific analyzer. 
 The algorithmic changes in SAMA were
@@
 the behavior of the analyzer on the fly and can interact
-with any case study. 
-%In Sarf, the analyzer remains general, and the 
-%case-based controller specializes the analysis.
+with any application as needed. 
+In Sarf, the analyzer remains general, and the 
+application-specific controller specializes the analysis.
 
@@
 linguistic model of the Arabic language. 
-%We can compare the case-based controller in Sarf to a high level
+%We can compare the application-specific controller in Sarf to a high level
 %component in the ElixirFM domain specific language.
@@
 % the FSMs for the three words
-\transtrue
+%\transtrue
 
@@
 when parsing the
-string \RL{wsyl`b-h-aa al-lA`bwn}.
+string \noTrnoVocRL{wsyl`b-h-aa\nospace al-lA`bwn}.
 %\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
 \footnote{ 
-\RL{n}%
-\RL{w}
-\RL{b}
-\RL{`}
-\RL{a} 
-\RL{l}
-\RL{l}
-\RL{a} %
-\RL{a}
-\RL{h}
-\RL{b}
-\RL{`}
-\RL{l}
-\RL{y}
-\RL{s}
-\RL{w}
+\noVocRL{n}%
+\noVocRL{w}
+\noVocRL{b}
+\noVocRL{`}
+\noVocRL{a} 
+\noVocRL{l}
+\noVocRL{l}
+\noVocRL{a} %
+\noVocRL{a}
+\noVocRL{h}
+\noVocRL{b}
+\noVocRL{`}
+\noVocRL{l}
+\noVocRL{y}
+\noVocRL{s}
+\noVocRL{w}
 in separate form to ease following the example
@@
 
-When there are two valid transitions such as \RL{w} 
+When there are two valid transitions such as \noTrnoVocRL{w} 
 and $\epsilon$ in the start case, 
@@
 In our example, if there were no stems that start 
-with the letter \RL{w}, $\Psi$ will die. 
+with the letter \noTrnoVocRL{w}, $\Psi$ will die. 
 In reality, $\Psi$ will die when the input 
-is at \RL{wsyl`}.
+is at \noTrnoVocRL{wsyl`}.
 
@@
 affixes. 
-For example, \RL{w} will result in an accept state
+For example, \noTrnoVocRL{w} will result in an accept state
 that transitions to ${\cal S}$.
-When \RL{s} follows, we move to another accept state in 
-${\cal P}$ corresponding to the prefix \RL{ws}. 
+When \noTrnoVocRL{s} follows, we move to another accept state in 
+${\cal P}$ corresponding to the prefix \noTrnoVocRL{ws}. 
 The same applies to ${\cal X}$.
 
-Lets consider $\Phi$ after it consumed \RL{wsy} 
+Lets consider $\Phi$ after it consumed \noTrnoVocRL{wsy} 
 and transitioned into the root node of ${\cal S}$.
-Now $\Phi$ will transition with \RL{l`b} to reach an accept 
+Now $\Phi$ will transition with \noTrnoVocRL{l`b} to reach an accept 
 state.
 
-Before moving with the letter \RL{b} to the accept state,
-$\Phi$ needs to make sure that the stem \RL{l`b} is compatible
-with the prefix \RL{wsy}. 
+Before moving with the letter \noTrnoVocRL{b} to the accept state,
+$\Phi$ needs to make sure that the stem \noTrnoVocRL{l`b} is compatible
+with the prefix \noTrnoVocRL{wsy}. 
 Sarf keeps compatibility category values as part
@@
 more than one concrete state. 
-If the category of \RL{l`b} is compatible with the category of
-\RL{wsy} then \RL{b} will move $\Phi$ to an accept state. 
+If the category of \noTrnoVocRL{l`b} is compatible with the category of
+\noTrnoVocRL{wsy} then \noTrnoVocRL{b} will move $\Phi$ to an accept state. 
 Otherwise, it will move it to a regular or a reject state.
@@
 spawns $\Xi$ which transitions to the root of ${\cal X}$. 
-After it consumes \RL{h--} $\Phi$ dies since there is no
-\noTrRL{h--}-edge from the current state.
+After it consumes \noTrnoVocRL{h--} $\Phi$ dies since there is no
+\noTrnoVocRL{h--}-edge from the current state.
 $\Xi$ represents a valid analysis and proceeds.
@@
 
-The case-based controller can interfere at any point in the 
+The application-specific controller can interfere at any point in the 
 analysis to make a decision like ignoring an FSA that 
@@
 accept states are the terminal nodes corresponding to the last 
-letters in stems. 
+letters in stems. Note that the trie implementation may not 
+necessary work on letter-by-letter states, but also compacts
+common substrings.
 This approach is superior to Buckwalter since it consists of
@@
 
+%Note that Sarf does not rely on implementing its machines as 
+%equivalent regular expressions due to  the following reasons ...
+% (not necessary)
+
 \begin{figure}[tb]
@@
 The algorithm \CodeIn{NDSarf} (detailed below)
-takes as input a text string $L$ and a case-based controller
+takes as input a text string $L$ and a application-specific controller
 $A$. 
@@
         if ($\Phi$.isSuffix())
-          if ($c$.isWhite()) or 
+          if ($c$.isWhite())  
             $A$.report();
@@
 
-\section{Islamic literature case study}
-\label{sec:islamic}
+\subsection{Disambiguation using diacritics}
+\label{sec:diacritics}
 
-% TODO: check how to merge this part here
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{.9\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
-\label{f:exhadith}
-}
-\end{figure}
-% the FSMs for the three words
-\transtrue
+Many Arabic documents, especially religious texts, are fully vocalized.
+Even in documents written without diacritics, we often stumble at a couple of words which are partially vocalized
+to reduce the difficulty of their disambiguation by the reader. Clearly, if native readers
+need the aid of short vowels, automatic morphological analyzers which are context-free should not lose this precious piece of information.
+Sarf is the first in the field to support diacritic-based disambiguation. Consequently, Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
+or \RL{`ilm} (science) with \RL{`lim} (knew), regardless of the specific letters which are vocalized.
 
-We validate our hypothesis using a case study from the Islamic 
-literature. 
-A \RL{.hady_t} is a narration related to the prophet Mohammad
-through a \RL{sanad} or a sequence of narrators. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
-Arabic with its transliteration and translation. 
-The authenticity of a \noArRL{.hady_t} depends on 
-the credibility of the narrators as reported in 
-separate biography books. 
-In this paper, we consider the problem of automatically segmenting
-a \noArRL{.hady_t} book into narrations, each 
-narration into its \RL{matn} (content) and its \noArRL{sanad},
-and each sanad into the 
-separate narrators. Later, we can look each narrator 
-up in the biography books. 
-For example, 
-Figure~\ref{f:exhadith} shows proper names in boxes connected
-to form complex names of narrators. 
-For example, 
-\noTrRL{qtybT} is the first name 
-of narrator $n_1$, and 
-\noTrRL{s`yd} is the name of his father as 
-the word \noTrRL{bn} (son of) indicates.
+The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
+On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
+to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.
 
-The hadith case study is interesting to Sarf since the controller
-considers %most of the analyses where we have a concentration
-proper names and ignores text where
-proper names occur rarely.% as long %as the text is valid.
-We are also interested in discovering words 
-that relate names to each other such as \RL{bin} or ``the son of''
-and words that mean narrate such as \RL{qaal} or ``said''.
+\subsection{Multiword Expressions}
+\label{sec:MWE}
 
-With Sarf, we successfully automated the analysis of 
-three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
-and extracted from each one of them a vector of complex structures 
-that are three level deep in hierarchy.
-%We also showcase that the results of the accurate high level 
-%abstraction
-%results of the case-based analysis was not affected by a major
-%refinement to the analyzer that increased low level analysis accuracy.
-
-A hadith in Islamic literature is a narration from the prophet Mohammad
-related by multiple narrators.
-Establishing the authenticity of a hadith is an important task
-in Islamic studies. 
-Based on them Islamic scholars issue rules that affect the life
-of people around the world. 
-The literature lacks an exhaustive mechanism of authentication
-checks. This is currently manual work prone to error and almost
-impossible to complete in the life time of a scholar due to the
-large literature available. 
-%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
-%the analysis of the hadith literature. 
-In this paper, we use Sarf to successfully automate the
-analysis of three books of hadith selected 
-arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
-  the digitized books from online sources such as 
-  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
-  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
-%  Our understanding is that those books are historical documents with 
- % public IP.}.
-
-%\setcode{utf8}
-%\begin{arabtext}
-%حدثنا \emphasize{هشيم}، أخبرنا \emphasize{إسماعيل بن أبي خالد}،
-% عن أبي إسحاق، عن سعيد بن جبير، قال كنا
-%    مع ابن عمر رضي الله عنه حيث أفاض من
-%    عرفات إلى جمع فصلى بنا المغرب ومضى
-%\end{arabtext}
-%\setcode{standard}
-
-Our analysis accepts a book as input
-and segments it into a vector of hadiths. 
-We segment each individual hadith into two parts. 
-The first part is the sanad which
-contains the chain of narrators who related the hadith
-from Mohammad. 
-The second part is the matn or the content
-of the hadith. 
-We further explore the sanad and
-detect the chain of narrators and 
-the relation that links each narrator to his ancestor and 
-predecessor in the chain. 
-We also detect the full name of each narrator that is
-usually composed of several proper names with connectors
-in between.
-
-We built our hadith controller to target
-the detection of proper names. 
-This includes the task of finding compound names 
-composed of several words such as \RL{`bd alr.hmn} often
-appearing as ``run-on'' words.
-The controller also targets words that mean ``narrate'' when
-they appear in the neighborhood of multiple proper names.
-
-\subsection{Controller}
-\label{sec:controller}
-% FSM for the controller
-
-We illustrate the hadith controller 
-in Figure~\ref{f:hadith}. 
-The transitions are labeled by outputs generated from 
-the Sarf transducers. 
-A NAME label denotes a proper name.
-An NRC label denotes a narrator connector such as
-\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
-\RL{'a_hbar} ``told'' or one of their derivations. 
-An IBN label denotes the word \RL{ibn} ``the son of'' and is a name connector.
-A NISBA label is a possessive adjective that qualifies a person such 
-as \RL{alma.sriy} ``the Egyptian''.
-
-The symbol $\tau_{\mbox{NMC}}$ is a threshold
-that corresponds to the number of tolerated name connectors 
-that may occur between two names. 
-The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
-of name connectors collected since the controller
-started looping in the state NMC\_S. 
-The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
-that corresponds to a relaxed tolerance measure that
-the controller resorts to in case the words separating
-two names were longer than $\tau_{\mbox{NMC}}$ but 
-contained a name connector word such as IBN or NISBA.
-
-Intuitively, the controller reports a valid chain 
-of narrators when a sequence of names
-connected by narrator connectors appears. 
-It marks the beginning of the hadith with the beginning of 
-the current sequence,
-and marks the end of the hadith with the beginning of the 
-next sequence. 
-It also marks the chain of narrators as the sequence itself.
-
-The controller has four states that correspond to 
-a position in text relative to the next sanad. 
-State TEXT\_S is the initial state and denotes that
-the controller is outside the context of a sanad.
-The controller moves to state NAME\_S on
-NAME.
-State NRC\_S means that the controller is in the context
-of a narrator name after Sarf reports an NRC.
-State NMC\_S
-indicates that the controller expects a name to appear within 
-a tolerance threshold expressed by 
-$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
-
-Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
-before it gives up on its expectations.
-
-We reach the NAME\_S state only when a
-valid NAME is detected and we leave when no more NAME's are detected.
-The NRC\_S state can only be reached if an NRC is detected.
-NMC\_S can only be reached from a NAME\_S state.
-
-\begin{figure}[tb!]
-\center{
-\resizebox{.8\columnwidth}{!}
-{ \input{figs/hadith.pdftex_t}}
-\caption{The controller of the hadith case study.}
-\label{f:hadith}
-}
-\end{figure}
-
-%The definition of input labels such as NAME and IBN depends on the 
-%morphological analyzer. 
-%However, our case-based controller approach was able to perform well
-%under both Sarf and a refined version of Sarf.
+The names which have been augmented to the lexicon of Sarf contain compound names such as 
+\noVocRL{`bid alkarym}. 
+In current analyzers with the exception of Attia, this expression is analyzed word by word and in best cases is 
+considered to contain two names, instead of being understood to be a single but compound name.
+Sarf enlists such expressions in the stem transducer (with one letter being the white space) and since tokenization
+is performed on the fly, such expressions are detected as one unit. These multi-word stems
+can accept also some compatible affixes such as \RL{w} (and) and \RL{f-} (but).
+This approach works perfectly for proper names but not for ``Syntactically-Flexible Expressions'' as noted
+in ~\cite{MWE}. %MWEs that can undergoe lexical inflections for their constituent words
 
@@
 
-\begin{table*}[bt]
+\begin{table}[tb]
 \centering
-\caption{Results of the hadith case study with Sarf.}
-%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
-\resizebox{1.9\columnwidth}{!}{
-\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
- & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
- & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
-Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
- Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
-Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
-Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
-Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
-Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
-Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
-Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
-Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
-%Narrator accuracy & & & & & &  & & & \\ 
-Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
-Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
-Sarf running time (secs.)& & 2.97 & 1.32 & & 2.93 & 1.31 & & .25 & .096\\ \hline \hline
- & & Buckwalter& ElixirFM & & Buckwalter & ElixirFM & & Buckwalter & ElixirFM \\\bottomrule
-Names      & & 18,354 & 19,201 & & 17,934 &  18,202 & & 5,021 & 5,117 \\
-Names/Narrator& & 2.6 & 2.1 & & 2.47 &  2.07 & & 1.83 & 1.62\\
-Narrators & & 3,210 & 2,981 & & 6,425 &  6,208 & & 2,311 & 1,965 \\ 
-Narrators/Chain & & 5.57 & 5.32 & & 5.46 &  5.32 & & 5.12 & 5.01 \\
-Chains & & 627 & 601 &  & 1,318 & 1,250 & & 487 & 452 \\ 
-Ignored names & & 9,642 & 9,915 &  & 6,118 & 6,874 & & 1,812 & 1,904 \\ \hline
-Segmentation accuracy & & 76\%& 84\%& & 85\%& 89 \%& & 84\%& 85\%\\ 
-Chain accuracy & & 78\% & 87\%&  & 87\%& 90\%& & 85\% & 86\% \\ 
-%Narrator accuracy & & & & & &  & & & \\ 
-Narrator name accuracy & & 76\% & 86\%& & 84\%& 87\% & & 84\% & 84\% \\ \hline
-Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
-Running time (secs.)& & 6.65 & 2.78$\times60^2$& &4.55 & 2.3$\times60^2$ & & .66 & 29.2$\times$60 
+\caption{Comparsion of Running Time}
+\resizebox{0.95\columnwidth}{!}{
+%\begin{tabular}{lp{.01cm}cp{.01cm}cp{.01cm}c}
+\begin{tabular}{lccc}
+\hline
+ & Al Kafi & Al Istibsar & Ibn Hanbal \\ \cline{2-4} 
+Sarf & 2.7002645 & 2.6884225 & 0.2356561 \\
+Buckwalter & 3.315950 & 5.406016 & 0.5472224 \\
+ElixirFM & 2.78$\times60^2$ & 2.3$\times60^2$ & 29.2$\times$60 \\
+\hline
 \end{tabular}
@@
 \normalsize
-\label{t:hadithresallresults}
-\end{table*}
+\label{t:efficeincyresults}
+\end{table}
 
-In this section we present results that compare Sarf to existing 
-morphological analyzers such as Buckwalter and ElixirFM. 
-We also evaluate our hypothesis of the utility of a case-based 
-morphological analyzer.
-We validate that the high-level analysis of the output 
-of low-level morphological analysis can compensate for accuracy.
-To achieve that we ran Sarf in two modes, simple and refined.
-In the simple mode, we used Sarf as discussed in 
-Section~\ref{sec:sarf} with the 
-controller of Section~\ref{sec:controller}.
-In the refined mode, we applied several low-level morphological 
-refinements to Sarf such as disambiguation of 
-female proper names so that the controller can ignore them. 
-We also added a refinement that detected compound proper names 
-such as \RL{`bd alr.hmn} 
-as they are common in narrator names.
-%Finally, we consider a possessive form as a reference to 
-%a person and treat
-%it as a narrator.
-
-We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
-with 4GB of memory using a Linux operating system.
-
-We report our results in Table~\ref{t:hadithresallresults} on the 
+In this section we compare Sarf to existing 
+morphological analyzers such as Buckwalter and ElixirFM in terms of running time.
+We ran the different analyzers on
+three books of Islamic narrations \RL{.hady_t} ~\cite{IbnHanbal,AlTousi,AlKulayni}
+We report our results in Table~\ref{t:efficeincyresults} on the 
 three hadith books.
-In general, we observe that the refinements slightly 
-improved performance and accuracy at
-the lowest level of abstraction by reducing false positives.
-However, the refinements had little to no effect on the high 
-accuracy measures at higher levels of abstraction.
-We believe that this is a strong supporting evidence to our 
-hypothesis.
-
-The row for ``word count'' is an indicator of the size of the data.
-The %numbers in 
-rows ``names'', ``narrators'', and ``chains'' denote the total
-number of the corresponding entity per book.
-The relative magnitudes of these rows
-reflect the hierarchical structure where 
-a chain contains
-narrators, and a narrator contains names.
-The row ``ignored names'' reports the number of names detected 
-outside the chains and thus are irrelevant to
-the case study.
-In addition, the rows ``names/narrator'' and ``narrators/chain'' 
-report the average
-number of names per narrator and 
-the average number of narrators per chain respectively.
-
-The ``segmentation accuracy'' reports on the accuracy of 
-segmentation of the book into separate hadiths.
-The metric is computed manually over a representative sample 
-of each book.
-The same applies to all other accuracy measures.
-The ``chain accuracy'' reflects the percentage of correct narrators 
-covered by extracted chains.
-The ``narrator name accuracy'' row reflects the percentage of 
-correct proper names detected within
-narrator structures.
-
-The row ``name false positives'' reports the percentage of words 
-recognized as names while they should not.
-We notice that this low level metric improved dramatically 
-with the refined analysis without substantially affecting the 
-high abstract measures such as chain and segmentation
-accuracy.
-
-We notice that the controller in the simple analysis ignored 
-more names and thus the case-based
-controller compensated with a simple heuristic
-for the inaccuracy of the low level 
-analysis.
-%and without the need of the low-level refinements.
-
-We investigated the case of Al-Kafi where the accuracy of 
-the simple analysis was significantly lower than that of the 
-refined analysis. 
-We discovered that the chains detected by the refined analysis 
-were a subset of those detected by the simple
-analysis. 
-The reason for the lower accuracy is false positive chains, 
-implying that the simple analysis did not miss important 
-information. 
-We hypothesize that a higher level of abstraction such as 
-intersecting narrators from the hadith books with those from
-the biographies can reduce the false positives in the chains.
-
-%\subsection{Buckwalter and ElixirFM}
-{\bf Buckwalter and ElixirFM.~~~}
-We implemented our case study with Buckwalter in two phases. 
-In phase one, we processed the entire text with Buckwalter and
-obtained annotations to text. 
-In the second phase, we fed the annotations to 
-the hadith controller as if they were coming from Sarf. 
-We measured the time it took to accomplish both tasks
-ignoring the time spent streaming the annotations.
-We did the same for ElixirFM.
 
@@
 running time. 
-It also produces higher accuracy measures 
-mainly due to the extended
-lexicon as well as to the fact that Sarf reduces false positive
-names by ignoring affixes that can not form
-names even if one possible solution of the stem is a name. 
-%Doing the same with Buckwalter and ElixirFM requires  
-%expanding the annotations resulting from one simple affix to be 
-%reflected in all the composite affixes that contain it.
-%expanding the annotations resulting from one affix to reflect
-%its multiple composing affixes.
-%However, this was easily performed using our recursive definition 
-%of the affixes.
-
 Buckwalter performed much better than ElixirFM in 
 terms of running time, 
-while ElixirFM outperformed Buckwalter in terms of accuracy.
-%This is up to 
-%our knowledge the first comparison of these two tools and 
 We think the big difference is due to the sophistication of 
@@
 
-We presented Sarf, a novel case-based morphological analyzer,
-and validated its utility using the hadith literature
-case study. 
-Sarf uses non-deterministic finite state Automata with 
-concatenative analysis and recursive affixes to analyze 
-Arabic text. 
-Results show that Sarf with the hadith controller
-is faster and more accurate than
+We presented Sarf, a novel application-specific morphological analyzer,
+which supports diacritic-based disambiguation and introduces recursive affixes.
+%Sarf uses non-deterministic finite state Automata with 
+%concatenative analysis and recursive affixes to analyze 
+%Arabic text. 
+Results show that Sarf is faster than
 current morphological analyzers.
 
-%TODO: merge into conclusion, taken from introduction
-We make the following key contributions: 
-\begin{enumerate}
-\item {\bf Case-based analysis:}  We designed and implemented
-a case-based morphological analyzer where a controller
-guides the analysis. 
-%\item {\bf Exhaustive analysis:} Since we have a case-based 
-%controller that can decide on the fly whether an analysis is 
-%accepted or not, we can afford to keep all valid analyses modulo
-%those pruned by the case controller. 
-%We do that using non-deterministic FSAs. 
-\item {\bf Recursive affixes:}
-We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02} and define affixes
-recursively. 
-This allows Sarf to perform better and 
-maintain compositional part of speech (POS) tags.
-\item {\bf Hadith case study:}
-%We evaluated Sarf using the hadith case study. 
-We were able to efficiently extract a complex three 
-level hierarchical 
-structure containing the chains of narrators 
-with high accuracy. 
-\end{enumerate}
+In a future paper, we will present our work of using Sarf to 
+parse Islamic Narrations from Prophet Muhammad commonly known commonly as hadith.
+We are also working on using Sarf to automate the problems of
+arabic email classification and
+arabic content-based parental control.
+Those use cases will benefit from the application-specific nature of Sarf.
 
-In the future, 
-we will try Sarf with other interesting case studies.
-We plan to complete the hadith case study 
-and automate narrator discovery and authentication in the 
-biography books. 
-%We plan to make use of the name connector and narrator
-%connector stop phrases that our case study extracted to create a 
-%learning system that can learn more names. 
-%We also plan on developing an Arabic linguistic computational model 
-%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
-%explored partially based on the case study.
+%In the future, 
+%we will try Sarf with other interesting case studies.
+%We plan to complete the hadith case study 
+%and automate narrator discovery and authentication in the 
+%biography books. 
+%%We plan to make use of the name connector and narrator
+%%connector stop phrases that our case study extracted to create a 
+%%learning system that can learn more names. 
+%%We also plan on developing an Arabic linguistic computational model 
+%%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
+%%explored partially based on the case study.
 

### 2011-02-25 trunk/tex/sarf.aclhlt.tex
left one page and 4 lines to reduce...

@@
 a branch of morphological analysis may not be
-appropriate for the case study at hand. 
-In this paper we present Sarf, a case-based morphological 
-analyzer for Arabic. 
-Sarf allows a case-based controller to control and refine
+appropriate for the application at hand. 
+In this paper we present Sarf, 
+%an application-specific morphological analyzer for Arabic. 
+an Arabic morphological analyzer, which
+%Sarf
+allows an application-specific controller to control and refine
 the state of the analysis on the fly. 
-It uses non-deterministic finite state Automata
+It uses non-deterministic finite state Automata % (NDFSA)
 to analyze a stream of text. 
 %Each machine corresponds to a valid analysis. 
-Sarf uses concatenative analysis based on recursive affixes 
-to better preserve part of speech information.
-We automated the analysis of three books of Islamic
-narrations (hadith) using Sarf where we
-abstracted each book into a vector of complex structures.
-Our results show higher accuracy and better efficiency
+Sarf uses concatenative analysis 
+%based on
+with  
+recursive affixes 
+to better preserve part of speech (POS) information.
+It solves the issue of ``run-on'' words 
+and utilizes diacritics for disambiguation.
+%We automated the analysis of three books of Islamic
+%narrations (hadith) using Sarf, where we
+%abstracted each book into a vector of complex structures.
+Our results show better efficiency
 compared to current analyzers.
@@
 analysis techniques for Arabic text. 
-Current morphological analyzers~\cite{Sughaiyer:04}
-use concatenative analysis when
-considering the internal structure 
-of an Arabic word and
-composing it into several {\em morphemes}. 
-A morpheme can be a {\em stem}, or an {\em affix}.
-An affix can be a {\em prefix, suffix, } or an {\em infix}.
-The analysis of one word may lead to several possible
-morphological solutions.
-\vocalize
-For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
-may have two valid morphological analyses. 
-The letter \RL{'a} may be a prefix and the word means 
-``I praise him'', or 
-%The letter \RL{'a} may also be 
-part of the stem \RL{'a.hmad} (a proper noun)
-and the word means ``his Ahmad''.
-
-\novocalize
-The accuracy of the solutions suffer due to inherent difficulties
+%Current morphological analyzers~\cite{Sughaiyer:04}
+%use concatenative analysis when
+%considering the internal structure 
+%of an Arabic word and
+%composing it into several {\em morphemes}. 
+%A morpheme can be a {\em stem}, or an {\em affix}.
+%An affix can be a {\em prefix, suffix, } or an {\em infix}.
+%The analysis of one word may lead to several possible
+%morphological solutions.
+%\vocalize
+%For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
+%may have two valid morphological analyses. 
+%The letter \RL{'a} may be a prefix and the word means 
+%``I praise him'', or 
+%%The letter \RL{'a} may also be 
+%part of the stem \RL{'a.hmad} (a proper noun)
+%and the word means ``his Ahmad''.
+%
+%\novocalize
+The accuracy of the solutions suffers due to inherent difficulties
 of morphological analysis of the Arabic language. 
 For example, it is common practice to write Arabic text
-without short vowels. 
-This greatly increases the ambiguity of Arabic text. 
-Arabic letters can have up to %(maybe reduce this sentence)
-four different forms
-corresponding to their position in a word, i.e, beginning,
-middle, end of word and separate forms. 
-This allows the phrase \transfalse
-\RL{il_A\nospace almdrsT} \transtrue
-to be visually recognizable
-as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
+without short vowels (diacritics) 
+which greatly increases its ambiguity.
+%This greatly increases the ambiguity of %Arabic text.
+%the text.
+%Moreover, Arabic letters can have up to 
+%four different forms
+%corresponding to their position in a word, i.e, beginning,
+%middle, end of word and separate forms. 
+%This allows the phrase
+%\noTrnoVocRL{il_A\nospace almdrsT}
+%to be visually recognizable
+Moreover, determining word boundaries is not always 
+%a trivial problem.
+trivial.
+For example,
+\noTrnoVocRL{il_A\nospace almdrsT}
+can be visually recognizable
+as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school) 
 without the need of a space in between. 
-The reason is the first word \RL{il_A} ends with
-\RL{_A} a non-connecting letter. 
-These words,
-referred to as ``run-on'' words~\cite{Buckwalter:04},
-occur often, and greatly increase the
+%The reason is the first word %\noVocRL{il_A} 
+%ends with
+%\noTrnoVocRL{_A} a non-connecting letter. 
+The reason is that
+\noTrnoVocRL{_A} is a non-connecting letter. 
+%These words, referred to as 
+Such 
+``run-on'' words~\cite{Buckwalter:04},
+%occur often, and greatly increase the
+occur often, %greatly
+increasing the
 difficulty of tokenization.
 
-{\bf Sarf.~~~}
+%{\bf Sarf.~~~}
 In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
-has been submitted with the paper and will be available online as an open source tool.},
- a {\em novel case-based efficient
+has been submitted with the paper and will be available 
+%online
+as an open source tool.}, 
+a {\em novel application-specific efficient 
+%(maybe 'enhanced' instead of 'efficient' to engulf both efficiency and other refinements)
 morphological analyzer} that uses 
-non-deterministic finite state Automata 
+non-deterministic finite state Automata (NDFSA)
 driven by an application-specific controller.
@@
 and represents a valid morphological analysis.
-The application-specific controller prunes false positives
+%The application-specific controller prunes false positives
+The controller prunes false positives
 early in the run by making a decision at every input letter
 to control the machines.
-The controller we present in Section~\ref{sec:controller}
-is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
-it can be a logical entity of any type.
+%The controller we present in Section~\ref{sec:controller}
+%is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
+%it can be a logical entity of any type.
+The controller can be a Finite State Automaton (FSA) or any other %logical entity 
+programming entity, % of any type,
+%since Automatas may be a %very complex
+%non-necessary solution for some simple applications.
+%(maybe re-word this sentence)
+depending on application needs.
+
+%(check if position of this paragraph seems in correct place)
+%Our focus when building Sarf, was not on improving the semantical rules and 
+%expanding the morphological alternatives, instead we worked on adding support 
+%for higher level control by the specific application utilizing the lower level 
+%morphological analyzer. %In other words, the application-specific controller, has 
+% full control over and directs the morphological analysis of Sarf. 
+% (sentence should be reworded differently, but is needed to explain why we mention 
+% higher and lower accuracy hypothesis, which seems vague otherwise)
+%
+%We have noted that many NLP case studies need the 
+%morphological analyzer to answer queries that do not need 
+%high accuracy at the morphological analysis level.
+%The analysis at the higher levels 
+%(i.e. answering the application-specific query)
+%can often compensate for 
+%tolerable inaccuracy at lower morphological levels. 
+%For example, if a query concerns proper names and the 
+%analyzer is considering a 
+%prefix that connects only to a verb,
+%%the analyzer can answer without considering the 
+%%rest of the word.
+%we do not need to further analyze the rest of the word 
+%to provide a negative answer.
+%
+%We find evidence to our hypothesis in~\cite{Maamouri:10}.
+%%The addition of a new corpus to the Arabic Tree Bank 
+%%(ATB)~\cite{Maamouri:04}
+%%%required compliance with the DARPA GALE program 
+%%%including several NLP tasks such as English translation 
+%%%annotations and word-level alignment of Arabic and English. 
+%%%The new data 
+%%challenged the existing processing and 
+%%annotation guidelines. 
+%%This led to refining 
+%%SAMA~\cite{Kulick:10} with 
+%%alternative guidelines. 
+%%This refinement was achieved only after the NLP 
+%%task interacted with the morphological analyzer,
+%%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
+%It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when NLP task interacted with the morphological analyzer.
+%%a flexibility our analyzer is designed to provide for the application-specific controller on the run.
+%In addition, 
+%the work of Habash~\shortcite{Habash:06} reports that different 
+%types of morphological analyses behave differently on the same %case study. 
+%data set.
+%The analyzers may report different solutions or 
+%may present the solutions in different orders. 
+%Clearly, some of 
+%the presented solutions and their order of presentation may not be 
+%appropriate for the %case study.
+%application needs, which is an issue that Sarf tackles. % (maybe reworded)
+We have noted that in many NLP applications, when the morphological analyzer
+benefits from the peculiarities and characteristics of the query at hand, 
+the analysis %can be easily 
+is greatly 
+enhanced. 
+Consequently, we strongly believe that by 
+%being 
+allowing
+an application-specific %morphological 
+%analyzer that allows %an application-specific 
+%the
+controller to intervene at every decision to guide, use, order, and 
+refine the morphological analysis, Sarf is of high utility.
 
@@
 %from a text stream.
+Apart from being application-specific, 
+Sarf %has
+%also  
+introduced 
+many enhancements. %to the field of Arabic morphological analysis. 
+%the following enhancements:
+%For example,
+First,
 Sarf does not assume that the word at hand is a token and
 performs tokenization on the fly based on morphological correctness.
-Up to our knowledge, this is the first morphological analyzer that 
-handles the ``run-on'' words problem.
-
-Sarf uses concatenative analysis with a novel 
+Up to our knowledge, this is the first %morphological 
+analyzer that 
+handles %the ``run-on'' words problem. 
+``run-on'' words. 
+%
+%Moreover, since Sarf performs tokenization on the fly, Sarf is not just capable of solving
+%the problem of ``run-on'' words, but also of handling Multiword Expressions (MWE). 
+Moreover,
+% since Sarf does not perform white-space delimited tokenization, 
+Sarf 
+%is capable of handling
+handles 
+Multiword Expressions (MWEs),
+%MWEs are defined in~\cite{MWE} as
+which are 
+``idiosyncratic interpretations
+that cross word boundaries (or spaces)''~\cite{MWE}. 
+%By this term, we refer to frequent cases when a group of words form together a single
+%item with a semantical meaning. In this paper, we focus on the example of compound names such as 
+%\noVocRL{`bid alkarym}. 
+%In current analyzers, this phrase is analyzed word by word and in best cases is 
+%considered to contain two names, instead of being understood to be a single but compound name.
+
+Second, 
+Sarf uses concatenative analysis with a novel %(maybe remove novel bc authors did not feel it is much)
 {\em recursive affixes} refinement,
-where an affix can be a sequence of other affixes which are concatenatively compatible.
+%where an affix can be a sequence of other affixes which are concatenatively compatible.
+where an affix can be a sequence of other concatenatively compatible affixes.
 %An atomic affix $a$ is a standalone affix
@@
 %with an affix. 
-Sarf introduces recursive affixes in order to
-better retain part of speech information and enhance the 
+%Sarf introduces recursive affixes in order to
+%This refinement is introduced to better retain
+Hence, it better retains
+%part of speech information
+POS information  
+and enhances the 
 efficiency of affix matching.
 
-Our main focus when building Sarf, was not on improving the semantical rules and expanding the morphological alternatives, instead we worked on adding support for higher level control by the specific application utilizing the lower level morphological analyzer.
-In other words, application-specific controller, has full control over and directs the morphological analysis of Sarf.
-% (sentence should be reworded differently, but is needed to explain why we mention higher and lower accuracy hypothesis, which seems vague otherwise)
-
-We hypothesize that many NLP case studies need the 
-morphological analyzer to answer queries that do not need 
-high accuracy at the low morphological analysis level.
-The analysis at the higher levels 
-(i.e. answering the application-specific query)
-can often compensate for 
-tolerable inaccuracy at lower morphological levels. 
-For example, if a query concerns proper names and the 
-analyzer is considering a 
-prefix that connects only to a verb,
-%the analyzer can answer without considering the 
-%rest of the word.
-we do not need to further analyze the rest of the word 
-to provide a negative answer.
-
-We find evidence to our hypothesis in~\cite{Maamouri:10}.
-The addition of a new corpus to the Arabic Tree Bank 
-(ATB)~\cite{Maamouri:04}
-%required compliance with the DARPA GALE program 
-%including several NLP tasks such as English translation 
-%annotations and word-level alignment of Arabic and English. 
-%The new data 
-challenged the existing processing and 
-annotation guidelines. 
-This led to refining 
-SAMA~\cite{Kulick:10} with 
-alternative guidelines. 
-This refinement was achieved only after the NLP 
-task interacted with the morphological analyzer,
-a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
-In addition, 
-the work of Habash~\shortcite{Habash:06} reports that different 
-types of morphological analyses behave differently on the same %case study. 
-data set.
-The analyzers may report different solutions or 
-may present the solutions in different orders. 
-Clearly, some of 
-the presented solutions and their order of presentation may not be 
-appropriate for the %case study.
-application needs, which is an issue that Sarf tackles. % (maybe reworded)
-Moreover, our results strongly support our claim that
-an application-specific morphological analyzer that allows a application-specific 
-controller to intervene at every decision to guide, use, order, and 
-refine the morphological analysis is of high utility.
-We validate our hypothesis using a case study from the Islamic 
-literature, introduced later in Section~\ref{sec:islamic}.
-
-The rest of this paper is structured as follows. In Section~\ref{sec:related}
-we compare Sarf to related work. In Section~\ref{sec:sarf}
-we discuss Sarf and illustrate it with a running example.
-%We also explain how non-deterministic finite state Automata
-%work. 
-In Section~\ref{sec:islamic} we present the case
-study. We present our results in Section~\ref{sec:results}
-and conclude with future work in Section~\ref{sec:future}.
+Third, Sarf supports disambiguation using 
+%diacritics when they are available.
+any available diacritics.
+%This feature, to the best of our knowledge is the first attempt to benefit from 
+%To the best of our knowledge,
+Up to our knowledge, this is the first attempt to benefit from
+diacritics instead of ignoring them. 
+Prior to Sarf, the trend was to preprocess 
+the text to remove  
+these %any such
+`unwelcomed' letters. 
+%Nonetheless, Sarf regards them as a richness to be utilized that should not be disregarded.
+
+%In addition to all its enhanced analysis methods, Sarf still
+%Sarf also
+%outperformed current morphological analyzers in terms of running time,
+%%These results will be presented 
+%as shown in Section~\ref{sec:results}.
+Sarf is also more efficent than current morphological analyzers, 
+as shown in Section~\ref{sec:results}.
+
+%%We make the following 
+%We summarize our key contributions: 
+%\begin{enumerate}
+%\item {\bf application-specific analysis:}  We designed and implemented
+%a application-specific morphological analyzer where a controller
+%guides the analysis. 
+%%\item {\bf Exhaustive analysis:} Since we have an application-specific 
+%%controller that can decide on the fly whether an analysis is 
+%%accepted or not, we can afford to keep all valid analyses modulo
+%%those pruned by the case controller. 
+%%We do that using non-deterministic FSAs. 
+%\item {\bf Recursive affixes:}
+%We refine the concatenative analysis of 
+%Buckwalter~\shortcite{Buckwalter:02} and define affixes
+%recursively. 
+%This allows Sarf to perform better and 
+%maintain compositional part of speech (POS) tags.
+%\item {\bf Diacritic-based Disambiguation:}
+%Diacritics are used to disambiguate morphological
+%alternatives using any existant diacritics.
+%\end{enumerate}
+
+%The rest of this paper is structured as follows. In Section~\ref{sec:related}
+%we compare Sarf to related work. In Section~\ref{sec:sarf}
+%we discuss Sarf and illustrate it with a running example.
+%%We also explain how non-deterministic finite state Automata
+%%work. 
+%In Section~\ref{sec:islamic} we present the case
+%study. We present our results in Section~\ref{sec:results}
+%and conclude with future work in Section~\ref{sec:future}.
 
@@
 Analyzers such as~\cite{Khoja:01,Darwish:02} 
-target specific applications in the 
-analyzer itself or use a specific set of POS tags
-as their reference.
-Sarf differs in that it is a general morphological 
+target specific applications 
+%in the analyzer itself or use a specific set of POS tags as their reference
+or limit themselves to a specific set of POS tags. 
+Sarf differs in that it is a {\em general} morphological 
 analyzer that reports all possible solutions. 
-It is application-specific in the sense that a controller 
+It is {\em application-specific} in the sense that a controller 
 prunes these solutions.
 
-Current morphological analyzers such as 
+Current morphological analyzers such as %(maybe disambiguators also added here according to reviewer comments)
 Buckwalter~\shortcite{Buckwalter:02},
@@
 and ElixirFM~\cite{Otakar:07} 
-take as input white space delimited tokens~\cite{Kulick:10},
+take as input white-space delimited tokens~\cite{Kulick:10},
 consider them as words,
 and enumerate all possible solutions. 
-%This approach has several problems. 
-However, we note the following points that deserve consideration:
-%(We need not to mention especially point 2 as a problem according to reviewer comments)
-First, a white space delimited token may have 
-more than one word.
+This approach has several problems. 
+%However, we note the following: %(I thought of this change not to include point 2 as a "problem" according to reviewer comments)
+First, a white-space delimited token may have 
+more than one word or just part of a MWE.
 %Second, the exhaustive enumeration may hurt performance and may
@@
 %in some case studies as noted in~\cite{Maamouri:10}. 
-Second, in many applications, the exhaustive enumeration may not be needed or appropriate as noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.
+Second, in many applications, exhaustive enumeration may not be needed or appropriate
+~\cite{Maamouri:10}; consequently, adding unnecessary running time.
+
 Other morphological analyzers such as 
@@
 to enhance the accuracy of the morphological analysis at the expense 
-of performance. % (we should stress that we mean performance in time and not accuracy although evident since some reviewers did not get)
-Xerox rules can be compiled into specialized finite state
-machine (FSM) based analyzers as described in~\cite{Beesley:03}.
+of efficiency.
+Xerox rules can be compiled into specialized 
+%finite state machine (FSM) based analyzers as described in~\cite{Beesley:03}.
+FSA-based analyzers~\cite{Beesley:03}.
 However, the efficiency of the resulting analyzer depends on the
@@
 %into efficient non-deterministic finite state Automata (NDFSA). 
-
-Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
-It differs from Buckwalter in that it defines a shorter list of affixes
-and a longer list of affix compatibility rules to allow 
+Sarf, however, allows users to write directly the code for the needed controller. %(maybe not clear)
+
+Sarf builds upon the lexicon of Buckwalter. %\shortcite{Buckwalter:02}, %(already cited above)
+%It differs from Buckwalter in that it defines a shorter list of affixes
+It defines a shorter list of affixes
+and a longer list of 
+%affix 
+compatibility rules to allow 
 affixes to be defined recursively
-so that we can have prefix-prefix an suffix-suffix 
+so that we can have prefix-prefix and suffix-suffix 
 concatenations.
-This allows Sarf to better maintain and propagate 
+%This allows Sarf to better maintain and propagate 
+Hence, it better maintains and propagates 
 the POS tags associated with affixes. 
@@
 %redundant with Buckwalter.% and that may lead to consistency issues. 
-Sarf keeps all the possible analyses alive where each analysis
-corresponds to a set of decisions and positions in the text. 
-Buckwalter produces a set of solutions for each token 
-and the several analysis threads of the text need to be 
-generated as a product of the token solutions.
-
-Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
-finite state Automata (FSA). 
-It reports that the number of Automata generated by a compiler
+% (check if necessary to keep the 2 sentences below)
+%Sarf keeps all the possible analyses alive where each analysis
+%corresponds to a set of decisions and positions in the text. 
+%Buckwalter produces a set of solutions for each token 
+%and the several analysis threads of the text need to be 
+%generated as a product of the token solutions.
+
+Similar to Sarf, both Beesley %~\shortcite{Beesley:01} (already cited previously)
+and Attia~\shortcite{Attia:05} use
+%finite state Automata (FSA). 
+FSA.
+Beesley reports that the number of Automata generated by a compiler
 for Xerox rules can not be controlled and also reports a difficulty 
@@
 Sarf however constructs an NDFSA from 
-the controller and the prefix, stem, and suffix deterministic 
+%the controller and 
+the prefix, stem, and suffix deterministic 
 transducers to solve the problem.
-Sarf also differs in that it uses recursive affixes in the 
-affix transducers, and in that the transducers consider
+Sarf also differs in that 
+%it uses recursive affixes in the affix transducers, and in that 
+the transducers consider
 input from the controller to modify their behavior. 
-%With Sarf it is easy to control the FSA with a case-based
+%With Sarf it is easy to control the FSA with a application-specific
 %controller. 
-Beesley requires recompilation of the Xerox rules. % (don't we do also for any change in the controller code?)
+%Beesley requires recompilation of the Xerox rules. 
+% (I am not sure if it is ok to have the above removed)
+Similar to Sarf, Attia handles MWEs, but with a more extensive coverage
+~\cite{Attia:06,Attia:10}.
+%Attia has researched extensively MWEs handling methods~\cite{Attia:06} and 
+%automated extending their dictionary coverage~\cite{Attia:10}. 
+% (I am not sure how clear this statement is)
 
@@
 when parsing the
-string \RL{wsyl`b-h-aa al-lA`bwn}.
+string \noTrnoVocRL{wsyl`b-h-aa\nospace al--lA`bwn}.
 %\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
-\footnote{ 
-\RL{n}%
-\RL{w}
-\RL{b}
-\RL{`}
-\RL{a} 
-\RL{l}
-\RL{l}
-\RL{a} %
-\RL{a}
-\RL{h}
-\RL{b}
-\RL{`}
-\RL{l}
-\RL{y}
-\RL{s}
-\RL{w}
-in separate form to ease following the example
-}
+%\footnote{ 
+%%\noVocRL{n}%
+%%\noVocRL{w}
+%%\noVocRL{b}
+%%\noVocRL{`}
+%%\noVocRL{a} 
+%%\noVocRL{l}
+%%\noVocRL{l}
+%%\noVocRL{a} %
+%%\noVocRL{a}
+%%\noVocRL{h}
+%%\noVocRL{b}
+%%\noVocRL{`}
+%%\noVocRL{l}
+%%\noVocRL{y}
+%%\noVocRL{s}
+%%\noVocRL{w}
+%\noTrnoVocRL{n}%
+%\noTrnoVocRL{w}
+%\noTrnoVocRL{b}
+%\noTrnoVocRL{`}
+%\noTrnoVocRL{a} 
+%\noTrnoVocRL{l}
+%\noTrnoVocRL{l}
+%\noTrnoVocRL{a} %
+%\noTrnoVocRL{a}
+%\noTrnoVocRL{h}
+%\noTrnoVocRL{b}
+%\noTrnoVocRL{`}
+%\noTrnoVocRL{l}
+%\noTrnoVocRL{y}
+%\noTrnoVocRL{s}
+%\noTrnoVocRL{w}
+%in separate form% to ease following the example
+%}
 
@@
 are all deterministic and cycle free as discussed 
-is Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}.
+in Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}.
 The cycles are introduced in the abstract machine
 for pedagogical purposes.
+
+%Note that Sarf does not rely on implementing its machines as 
+%equivalent regular expressions due to  the following reasons ...
+% (not necessary)
 
@@
 
-\section{Islamic literature case study}
-\label{sec:islamic}
-
-% TODO: check how to merge this part here
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{.9\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
-\label{f:exhadith}
-}
-\end{figure}
-% the FSMs for the three words
-\transtrue
-
-We validate our hypothesis using a case study from the Islamic 
-literature. 
-A \RL{.hady_t} is a narration related to the prophet Mohammad
-through a \RL{sanad} or a sequence of narrators. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
-Arabic with its transliteration and translation. 
-The authenticity of a \noArRL{.hady_t} depends on 
-the credibility of the narrators as reported in 
-separate biography books. 
-In this paper, we consider the problem of automatically segmenting
-a \noArRL{.hady_t} book into narrations, each 
-narration into its \RL{matn} (content) and its \noArRL{sanad},
-and each sanad into the 
-separate narrators. Later, we can look each narrator 
-up in the biography books. 
-For example, 
-Figure~\ref{f:exhadith} shows proper names in boxes connected
-to form complex names of narrators. 
-For example, 
-\noTrRL{qtybT} is the first name 
-of narrator $n_1$, and 
-\noTrRL{s`yd} is the name of his father as 
-the word \noTrRL{bn} (son of) indicates.
-
-The hadith case study is interesting to Sarf since the controller
-considers %most of the analyses where we have a concentration
-proper names and ignores text where
-proper names occur rarely.% as long %as the text is valid.
-We are also interested in discovering words 
-that relate names to each other such as \RL{bin} or ``the son of''
-and words that mean narrate such as \RL{qaal} or ``said''.
-
-With Sarf, we successfully automated the analysis of 
-three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
-and extracted from each one of them a vector of complex structures 
-that are three level deep in hierarchy.
-%We also showcase that the results of the accurate high level 
-%abstraction
-%results of the case-based analysis was not affected by a major
-%refinement to the analyzer that increased low level analysis accuracy.
-
-A hadith in Islamic literature is a narration from the prophet Mohammad
-related by multiple narrators.
-Establishing the authenticity of a hadith is an important task
-in Islamic studies. 
-Based on them Islamic scholars issue rules that affect the life
-of people around the world. 
-The literature lacks an exhaustive mechanism of authentication
-checks. This is currently manual work prone to error and almost
-impossible to complete in the life time of a scholar due to the
-large literature available. 
-%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
-%the analysis of the hadith literature. 
-In this paper, we use Sarf to successfully automate the
-analysis of three books of hadith selected 
-arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
-  the digitized books from online sources such as 
-  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
-  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
-%  Our understanding is that those books are historical documents with 
- % public IP.}.
-
-%\setcode{utf8}
-%\begin{arabtext}
-%حدثنا \emphasize{هشيم}، أخبرنا \emphasize{إسماعيل بن أبي خالد}،
-% عن أبي إسحاق، عن سعيد بن جبير، قال كنا
-%    مع ابن عمر رضي الله عنه حيث أفاض من
-%    عرفات إلى جمع فصلى بنا المغرب ومضى
-%\end{arabtext}
-%\setcode{standard}
-
-Our analysis accepts a book as input
-and segments it into a vector of hadiths. 
-We segment each individual hadith into two parts. 
-The first part is the sanad which
-contains the chain of narrators who related the hadith
-from Mohammad. 
-The second part is the matn or the content
-of the hadith. 
-We further explore the sanad and
-detect the chain of narrators and 
-the relation that links each narrator to his ancestor and 
-predecessor in the chain. 
-We also detect the full name of each narrator that is
-usually composed of several proper names with connectors
-in between.
-
-We built our hadith controller to target
-the detection of proper names. 
-This includes the task of finding compound names 
-composed of several words such as \RL{`bd alr.hmn} often
-appearing as ``run-on'' words.
-The controller also targets words that mean ``narrate'' when
-they appear in the neighborhood of multiple proper names.
-
-\subsection{Controller}
-\label{sec:controller}
-% FSM for the controller
-
-We illustrate the hadith controller 
-in Figure~\ref{f:hadith}. 
-The transitions are labeled by outputs generated from 
-the Sarf transducers. 
-A NAME label denotes a proper name.
-An NRC label denotes a narrator connector such as
-\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
-\RL{'a_hbar} ``told'' or one of their derivations. 
-An IBN label denotes the word \RL{ibn} ``the son of'' and is a name connector.
-A NISBA label is a possessive adjective that qualifies a person such 
-as \RL{alma.sriy} ``the Egyptian''.
-
-The symbol $\tau_{\mbox{NMC}}$ is a threshold
-that corresponds to the number of tolerated name connectors 
-that may occur between two names. 
-The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
-of name connectors collected since the controller
-started looping in the state NMC\_S. 
-The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
-that corresponds to a relaxed tolerance measure that
-the controller resorts to in case the words separating
-two names were longer than $\tau_{\mbox{NMC}}$ but 
-contained a name connector word such as IBN or NISBA.
-
-Intuitively, the controller reports a valid chain 
-of narrators when a sequence of names
-connected by narrator connectors appears. 
-It marks the beginning of the hadith with the beginning of 
-the current sequence,
-and marks the end of the hadith with the beginning of the 
-next sequence. 
-It also marks the chain of narrators as the sequence itself.
-
-The controller has four states that correspond to 
-a position in text relative to the next sanad. 
-State TEXT\_S is the initial state and denotes that
-the controller is outside the context of a sanad.
-The controller moves to state NAME\_S on
-NAME.
-State NRC\_S means that the controller is in the context
-of a narrator name after Sarf reports an NRC.
-State NMC\_S
-indicates that the controller expects a name to appear within 
-a tolerance threshold expressed by 
-$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
-
-Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
-before it gives up on its expectations.
-
-We reach the NAME\_S state only when a
-valid NAME is detected and we leave when no more NAME's are detected.
-The NRC\_S state can only be reached if an NRC is detected.
-NMC\_S can only be reached from a NAME\_S state.
-
-\begin{figure}[tb!]
-\center{
-\resizebox{.8\columnwidth}{!}
-{ \input{figs/hadith.pdftex_t}}
-\caption{The controller of the hadith case study.}
-\label{f:hadith}
-}
-\end{figure}
-
-%The definition of input labels such as NAME and IBN depends on the 
-%morphological analyzer. 
-%However, our case-based controller approach was able to perform well
-%under both Sarf and a refined version of Sarf.
+\subsection{Disambiguation using diacritics}
+\label{sec:diacritics}
+
+Many Arabic documents, especially religious texts, are fully vocalized.
+Even in documents written without diacritics, we often stumble at a couple of words which are partially vocalized
+to reduce the difficulty of their disambiguation by the reader. Clearly, if native readers
+need the aid of short vowels, automatic morphological analyzers which are context-free should not lose this precious piece of information.
+Sarf is the first in the field to support diacritic-based disambiguation. Consequently, Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
+or \RL{`ilm} (science) with \RL{`lim} (knew), regardless of the specific letters which are vocalized.
+
+The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
+On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
+to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.
+
+\subsection{Multiword Expressions}
+\label{sec:MWE}
+
+The names which have been augmented to the lexicon of Sarf contain compound names such as 
+\noVocRL{`bid alkarym}. 
+In current analyzers with the exception of Attia, this expression is analyzed word by word and in best cases is 
+considered to contain two names, instead of being understood to be a single but compound name.
+Sarf enlists such expressions in the stem transducer (with one letter being the white space) and since tokenization
+is performed on the fly, such expressions are detected as one unit. These multi-word stems
+can accept also some compatible affixes such as \RL{w} (and) and \RL{f-} (but).
+This approach works perfectly for proper names but not for ``Syntactically-Flexible Expressions'' as noted
+in ~\cite{MWE}. %MWEs that can undergoe lexical inflections for their constituent words
 
@@
 
-\begin{table*}[bt]
+\begin{table}[tb]
 \centering
-\caption{Results of the hadith case study with Sarf.}
-%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
-\resizebox{1.9\columnwidth}{!}{
-\begin{tabular}{lp{.2cm}ccp{.2cm}ccp{.2cm}cc} %\cline{2-10}
- & & \multicolumn{2}{c}{Al Kafi} & & \multicolumn{2}{c}{Al Istibsar} & &\multicolumn{2}{c}{Ibn Hanbal} \\ \cline{3-10}
- & & Simple & Refined & & Simple & Refined & & Simple  & Refined \\\bottomrule  %\hline \hline
-Word count & &\multicolumn{2}{c}{98,943} & & \multicolumn{2}{c}{103,835} & & \multicolumn{2}{c}{20,354} \\ 
- Names      & & 20,473 & 12,060  & &20,646 &  14,613& & 4,762 & 3,013\\
-Names/Narrator& &2.48 & 1.97 & & 2.17 &  1.84& & 1.49 & 1.25 \\
-Narrators & &3,080 & 2,623 & & 6,164 &  5,767& & 2,082 & 1,755 \\ 
-Narrators/Chain & & 5.28 & 4.84 & & 4.98&  4.76 & & 4.8 &4.05 \\
-Chains & & 583 & 542 &  & 1,238& 1,211& & 433 & 433 \\ 
-Ignored names & & 11,801 & 6,400 &  & 5,845 & 3,348 & & 1,386 & 642 \\ \hline
-Segmentation accuracy & & 85\% & 96\%& & 96\%& 96\%& & 92\%& 92\%\\ 
-Chain accuracy & & 99\% & 99\%&  & 98\%& 99\%& & 99\% & 97\% \\ 
-%Narrator accuracy & & & & & &  & & & \\ 
-Narrator name accuracy & & 93\% & 91\%& &  92\%& 90\% & & 90\% & 90\% \\ \hline
-Name false positives & & 28\% & 7\%&  & 9\%& 4\% & & 34\% & 4\% \\ \hline
-Sarf running time (secs.)& & 2.97 & 1.32 & & 2.93 & 1.31 & & .25 & .096\\ \hline \hline
- & & Buckwalter& ElixirFM & & Buckwalter & ElixirFM & & Buckwalter & ElixirFM \\\bottomrule
-Names      & & 18,354 & 19,201 & & 17,934 &  18,202 & & 5,021 & 5,117 \\
-Names/Narrator& & 2.6 & 2.1 & & 2.47 &  2.07 & & 1.83 & 1.62\\
-Narrators & & 3,210 & 2,981 & & 6,425 &  6,208 & & 2,311 & 1,965 \\ 
-Narrators/Chain & & 5.57 & 5.32 & & 5.46 &  5.32 & & 5.12 & 5.01 \\
-Chains & & 627 & 601 &  & 1,318 & 1,250 & & 487 & 452 \\ 
-Ignored names & & 9,642 & 9,915 &  & 6,118 & 6,874 & & 1,812 & 1,904 \\ \hline
-Segmentation accuracy & & 76\%& 84\%& & 85\%& 89 \%& & 84\%& 85\%\\ 
-Chain accuracy & & 78\% & 87\%&  & 87\%& 90\%& & 85\% & 86\% \\ 
-%Narrator accuracy & & & & & &  & & & \\ 
-Narrator name accuracy & & 76\% & 86\%& & 84\%& 87\% & & 84\% & 84\% \\ \hline
-Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
-Running time (secs.)& & 6.65 & 2.78$\times60^2$& &4.55 & 2.3$\times60^2$ & & .66 & 29.2$\times$60 
+\caption{Comparsion of Running Time}
+\resizebox{0.95\columnwidth}{!}{
+%\begin{tabular}{lp{.01cm}cp{.01cm}cp{.01cm}c}
+\begin{tabular}{lccc}
+\hline
+ & Al Kafi & Al Istibsar & Ibn Hanbal \\ \cline{2-4} 
+Sarf & 2.7002645 & 2.6884225 & 0.2356561 \\
+Buckwalter & 3.315950 & 5.406016 & 0.5472224 \\
+ElixirFM & 166.8$\times60$ & 138$\times60$ & 29.2$\times$60 \\
+\hline
 \end{tabular}
@@
 \normalsize
-\label{t:hadithresallresults}
-\end{table*}
-
-In this section we present results that compare Sarf to existing 
-morphological analyzers such as Buckwalter and ElixirFM. 
-We also evaluate our hypothesis of the utility of a case-based 
-morphological analyzer.
-We validate that the high-level analysis of the output 
-of low-level morphological analysis can compensate for accuracy.
-To achieve that we ran Sarf in two modes, simple and refined.
-In the simple mode, we used Sarf as discussed in 
-Section~\ref{sec:sarf} with the 
-controller of Section~\ref{sec:controller}.
-In the refined mode, we applied several low-level morphological 
-refinements to Sarf such as disambiguation of 
-female proper names so that the controller can ignore them. 
-We also added a refinement that detected compound proper names 
-such as \RL{`bd alr.hmn} 
-as they are common in narrator names.
-%Finally, we consider a possessive form as a reference to 
-%a person and treat
-%it as a narrator.
-
-We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
-with 4GB of memory using a Linux operating system.
-
-We report our results in Table~\ref{t:hadithresallresults} on the 
+\label{t:efficeincyresults}
+\end{table}
+
+In this section we compare Sarf to existing 
+morphological analyzers such as Buckwalter and ElixirFM in terms of running time.
+We ran the different analyzers on
+three books of Islamic narrations \RL{.hady_t} ~\cite{IbnHanbal,AlTousi,AlKulayni}
+We report our results in Table~\ref{t:efficeincyresults} on the 
 three hadith books.
-In general, we observe that the refinements slightly 
-improved performance and accuracy at
-the lowest level of abstraction by reducing false positives.
-However, the refinements had little to no effect on the high 
-accuracy measures at higher levels of abstraction.
-We believe that this is a strong supporting evidence to our 
-hypothesis.
-
-The row for ``word count'' is an indicator of the size of the data.
-The %numbers in 
-rows ``names'', ``narrators'', and ``chains'' denote the total
-number of the corresponding entity per book.
-The relative magnitudes of these rows
-reflect the hierarchical structure where 
-a chain contains
-narrators, and a narrator contains names.
-The row ``ignored names'' reports the number of names detected 
-outside the chains and thus are irrelevant to
-the case study.
-In addition, the rows ``names/narrator'' and ``narrators/chain'' 
-report the average
-number of names per narrator and 
-the average number of narrators per chain respectively.
-
-The ``segmentation accuracy'' reports on the accuracy of 
-segmentation of the book into separate hadiths.
-The metric is computed manually over a representative sample 
-of each book.
-The same applies to all other accuracy measures.
-The ``chain accuracy'' reflects the percentage of correct narrators 
-covered by extracted chains.
-The ``narrator name accuracy'' row reflects the percentage of 
-correct proper names detected within
-narrator structures.
-
-The row ``name false positives'' reports the percentage of words 
-recognized as names while they should not.
-We notice that this low level metric improved dramatically 
-with the refined analysis without substantially affecting the 
-high abstract measures such as chain and segmentation
-accuracy.
-
-We notice that the controller in the simple analysis ignored 
-more names and thus the case-based
-controller compensated with a simple heuristic
-for the inaccuracy of the low level 
-analysis.
-%and without the need of the low-level refinements.
-
-We investigated the case of Al-Kafi where the accuracy of 
-the simple analysis was significantly lower than that of the 
-refined analysis. 
-We discovered that the chains detected by the refined analysis 
-were a subset of those detected by the simple
-analysis. 
-The reason for the lower accuracy is false positive chains, 
-implying that the simple analysis did not miss important 
-information. 
-We hypothesize that a higher level of abstraction such as 
-intersecting narrators from the hadith books with those from
-the biographies can reduce the false positives in the chains.
-
-%\subsection{Buckwalter and ElixirFM}
-{\bf Buckwalter and ElixirFM.~~~}
-We implemented our case study with Buckwalter in two phases. 
-In phase one, we processed the entire text with Buckwalter and
-obtained annotations to text. 
-In the second phase, we fed the annotations to 
-the hadith controller as if they were coming from Sarf. 
-We measured the time it took to accomplish both tasks
-ignoring the time spent streaming the annotations.
-We did the same for ElixirFM.
 
@@
 
-We presented Sarf, a novel case-based morphological analyzer,
-and validated its utility using the hadith literature
-case study. 
-Sarf uses non-deterministic finite state Automata with 
-concatenative analysis and recursive affixes to analyze 
-Arabic text. 
-Results show that Sarf with the hadith controller
-is faster and more accurate than
+We presented Sarf, a novel application-specific morphological analyzer,
+which supports diacritic-based disambiguation and introduces recursive affixes.
+%Sarf uses non-deterministic finite state Automata with 
+%concatenative analysis and recursive affixes to analyze 
+%Arabic text. 
+Results show that Sarf is faster than
 current morphological analyzers.
 
-%TODO: merge into conclusion, taken from introduction
-We make the following key contributions: 
-\begin{enumerate}
-\item {\bf Case-based analysis:}  We designed and implemented
-a case-based morphological analyzer where a controller
-guides the analysis. 
-%\item {\bf Exhaustive analysis:} Since we have a case-based 
-%controller that can decide on the fly whether an analysis is 
-%accepted or not, we can afford to keep all valid analyses modulo
-%those pruned by the case controller. 
-%We do that using non-deterministic FSAs. 
-\item {\bf Recursive affixes:}
-We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02} and define affixes
-recursively. 
-This allows Sarf to perform better and 
-maintain compositional part of speech (POS) tags.
-\item {\bf Hadith case study:}
-%We evaluated Sarf using the hadith case study. 
-We were able to efficiently extract a complex three 
-level hierarchical 
-structure containing the chains of narrators 
-with high accuracy. 
-\end{enumerate}
-
-In the future, 
-we will try Sarf with other interesting case studies.
-We plan to complete the hadith case study 
-and automate narrator discovery and authentication in the 
-biography books. 
-%We plan to make use of the name connector and narrator
-%connector stop phrases that our case study extracted to create a 
-%learning system that can learn more names. 
-%We also plan on developing an Arabic linguistic computational model 
-%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
-%explored partially based on the case study.
+In a future paper, we will present our work of using Sarf to 
+parse Islamic Narrations from Prophet Muhammad commonly known commonly as hadith.
+We are also working on using Sarf to automate the problems of
+arabic email classification and
+arabic content-based parental control.
+Those use cases will benefit from the application-specific nature of Sarf.
+
+%In the future, 
+%we will try Sarf with other interesting case studies.
+%We plan to complete the hadith case study 
+%and automate narrator discovery and authentication in the 
+%biography books. 
+%%We plan to make use of the name connector and narrator
+%%connector stop phrases that our case study extracted to create a 
+%%learning system that can learn more names. 
+%%We also plan on developing an Arabic linguistic computational model 
+%%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
+%%explored partially based on the case study.
 

### 2011-02-25 trunk/tex/hadith.short.tex
new file

% File naaclhlt2010.tex
% Contact: nasmith@cs.cmu.edu
\documentclass[11pt]{article}
\usepackage{acl-hlt2011}
\usepackage{times}
\usepackage{latexsym}
\usepackage{amsmath}
\usepackage{multirow}
\usepackage{url}
\DeclareMathOperator*{\argmax}{arg\,max}
\setlength\titlebox{6.5cm}    % Expanding the titlebox

\usepackage{arabtex}
\usepackage{amssymb}
\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{amsthm}
\usepackage{booktabs}

\usepackage{relsize}
\usepackage{fancyvrb}
\usepackage[colorlinks=false]{hyperref}

\usepackage{utf8}
\setarab
\fullvocalize
\transtrue
\arabtrue

%\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}

%\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
\title{Hadith Narrator Chain Extraction using Arabic Morphological Analysis}

%\author{ Jad Makhlouta \\\And
%Hamza Harkous \\
%  American University of Beirut \\
%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb } \\\And 
%  Fadi Zaraket 
%}

\date{}

\begin{document}
\maketitle
\begin{abstract}
A hadith in Islamic literature is a narration from the 
prophet Mohammad related by multiple narrators.
Establishing the authenticity of a hadith is an important task
in Islamic studies. 
Based on them Islamic scholars issue rules that affect the life
of people around the world. 
The literature lacks an automated exhaustive mechanism of 
authentication checks. 
This is currently manual work prone to error and almost
impossible to complete in the life time of a scholar due to the
large literature available. 
In this paper, we present a tool that uses 
an Arabic application-specific morphological analyzer 
to successfully automate the
analysis of three books of hadith selected 
arbitrarily and abstract 
each book into a graph of narrators related to each other. 
\end{quote}
\end{abstract}

\section{Introduction}

\transfalse
\begin{figure}[tb]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue

A \RL{.hady_t}~\footnote{In this document, 
we use the default ArabTeX transliteration style ZDMG.}
is a narration related to the prophet Mohammad
through a \RL{sanad} or a sequence of narrators. 
The collections of traditions are the second source of
jurisprudence after the \RL{qor'An} for all Islamic schools of thought. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
We show proper names in boxes connected
to form complex names of narrators. 
For example, 
\noTrRL{qtybT} is the first name 
of narrator $n_1$, and 
\noTrRL{s`yd} is the name of his father as 
the word \noTrRL{bn} (son of) indicates. 
The sequence of names from $n_1$ to $n_5$ 
constitutes the \noArRL{sanad} 
of the \noArRL{.hady_t}. 
The second part of the \noArRL{.hady_t} is the 
\RL{matn} (content) and constitutes the actual content 
and tradition.

Due to religious and political reasons, 
writing the traditions was forbidden 
until the days of the eighth Umayad Calif,
\novocalize
\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after 
the death of the prophet~\cite{AlAskari}. 
Consequently, many inconsistencies were
introduced to the literature which necessitated 
a thorough authentication study of a tradition before its use in 
jurisprudence.
\vocalize
%An Islamic hadith scholar studies the authenticity of 
%the \noArRL{sanad} of a 
%set of related traditions before using these traditions 
%jurisprudence. 
While different Islamic schools of thought differ on 
how to interpret the content, they almost all agree
that if a \noArRL{sanad} lacks authenticity, 
then scholars can not use the tradition for jurisprudence.

The authenticity of a \noArRL{.hady_t} depends on 
the credibility of the narrators as reported in 
separate biography books. 
The study of \noArRL{.hady_t} authentication is 
currently manual and error prone due to the huge number
of existing traditions and tradition books. 
Hadithopedia~\cite{Hadithopaedia:08}
reports that the tradition and biography
books for one of the Islamic sects amounts to more than 
300 thousand lines of text. 
Al-Azami\shortcite{Al-Azami-91} cites more than eleven books
of digitized tradition books each of several volumes, and a dozen
other biography and secondary authentication books such
as a geographical dictionary of places in hadith.

In this paper, we present the hadith {narrator chain 
extractor using morphological analysis} (NaCEMA), 
a novel technique that automates
extracting the \noArRL{sanad} from the books of traditions 
into a diagram of chains of narrators. NaCEMA segments
a book into narrations, each narration into its \noArRL{matn} 
and its \noArRL{sanad}, 
and each sanad into its separate narrators. 
This work will lead to 
automate an exhaustive automated \noArRL{.hady_t} authentication 
effort in the future as we plan to 
automatically analyze the biography books.

%Automated analysis of Arabic data sets, including texts, 
%publications, records and digital media is essential
%with the huge digital Arabic content available nowadays. 
\subsection{ Arabic morphological analysis}
Arabic morphological analysis is key to our analysis. 
Current morphological analyzers~\cite{Sughaiyer:04}
use concatenative analysis when
considering the internal structure 
of an Arabic word and
composing it into several {\em morphemes}. 
A morpheme can be a {\em stem}, or an {\em affix}.
An affix can be a {\em prefix, suffix, } or an {\em infix}.
%The analysis of one word may lead to several possible
%morphological solutions.
\vocalize
The word \RL{'a.hmadH}
may have two valid morphological analyses. 
The letter \RL{'a} may be a prefix and the word means 
``I praise him'', or 
%The letter \RL{'a} may also be 
part of the stem \RL{'a.hmad} (a proper noun)
and the word means ``his Ahmad''.

\novocalize
%The accuracy of the solutions suffer due to inherent difficulties
%of morphological analysis of the Arabic language. 
%For example, it is common practice to write Arabic text
%without short vowels. 
%This greatly increases the ambiguity of Arabic text. 
%Arabic letters can have up to 
%four different forms
%corresponding to their position in a word, i.e, beginning,
%middle, end of word and separate forms. 
%This allows the phrase \transfalse
%\RL{il_A\nospace almdrsT} \transtrue
%to be visually recognizable
%as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
%without the need of a space in between. 
%The reason is the first word \RL{il_A} ends with
%\RL{_A} a non-connecting letter. 
%These words,
%referred to as ``run-on'' words~\cite{Buckwalter:04},
%occur often, and greatly increase the
%difficulty of tokenization.

Current morphological analyzers such as 
Buckwalter~\shortcite{Buckwalter:02},
Beesley~\shortcite{Beesley:01},
SAMA~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
This approach has several problems. 
First, a white space delimited token may have 
more than one word, referred to as ``run-on'' 
words~\cite{Buckwalter:04}.
Second, the exhaustive enumeration may not be efficient and may
not be necessary or appropriate
in some applications as noted in~\cite{Maamouri:10}. 
%Other morphological analyzers such as 
%Amira~\cite{Diab:07,Benajiba:07},
%MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
%use machine learning and support vector machines (SVM) 
%to disambiguate the morphological analysis at the expense 
%of efficiency. 
%Xerox rules can be compiled into specialized finite state
%machine (FSM) based analyzers as described in~\cite{Beesley:03}.
%However, the efficiency of the resulting analyzer depends on the
%way the Xerox rules are written. 
%This requires deep knowledge and insight from the user
%in compilation techniques, context free grammars, 
%and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA).

%\subsection{Sarf}
%\label{sec:intro:sarf}

{\bf Sarf.~~~}
In this paper, 
we use Sarf~\footnote{Sarf (name modified for blind review) }
an {\em application-specific efficient
morphological analyzer} that uses 
non-deterministic finite state Automata 
driven by an application-specific controller customized 
by the user. 
%Each machine in Sarf takes one letter at a time as input
%and represents a valid morphological analysis.
The application-specific controller prunes false positives
early in the run by making a decision at every input letter
to control the machines.
Sarf builds on and extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} 
with proper and location names extracted from different online 
sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
%as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
as well as biblical sources.
%~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.

%Each alive machine in Sarf takes as input one letter at a time 
%from a text stream. 
Sarf does not assume that the word at hand is a token and
performs tokenization on the fly based on morphological correctness
and thus can handle 
the ``run-on'' words problem. 
Sarf also uses diacritics if they exist to disambiguate 
morphological solutions.

%The hadith 
%application specific controller that controls Sarf considers 
%proper names and ignores text where
%proper names occur rarely.
%It also considers words that relate names to each other 
%such as \RL{bin} or ``the son of''
%and words that mean narrate such as \RL{qaal} or ``said''.

%With Sarf, we successfully automated the analysis of 
%three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
%and extracted from each one of them a vector of complex structures 
%that are three level deep in hierarchy.
%We also merged these vectors into a graph of narrators where 
%we represented identical narrators from separate traditions 
%in one single node.

%The rest of this paper is structured as follows. 
%In Section~\ref{sec:sarf}
%we discuss Sarf and illustrate it with a running example.
%We also explain how non-deterministic finite state Automata
%work. 
%In Section~\ref{sec:islamic} we present the hadith application. 
%We present our results in Section~\ref{sec:results}.
%In Section~\ref{sec:related}
%we compare NaCEMA to related work. 
%We conclude with future work in Section~\ref{sec:future}.

%\section{Background}
% TODO: nondeterministic finite-state automata

%The survey in~\cite{Sughaiyer:04} compares
%several morphological analyzers. 
%Analyzers such as~\cite{Khoja:01,Darwish:02} 
%target specific applications in the 
%analyzer itself or use a specific set of POS tags
%as their reference.
%Sarf differs in that it is a general morphological 
%analyzer that reports all possible solutions. 
%It is case-based in the sense that a controller 
%prunes these solutions. 
%
%Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
%It differs from Buckwalter in that it defines a shorter list of affixes
%and a longer list of affix compatibility rules to allow 
%affixes to be defined recursively
%so that we can have prefix-prefix an suffix-suffix 
%concatenations.
%This allows Sarf to better maintain and propagate 
%the POS tags associated with affixes. 
%%This information is highly 
%%redundant with Buckwalter.% and that may lead to consistency issues. 
%Sarf keeps all the possible analyses alive where each analysis
%corresponds to a set of decisions and positions in the text. 
%Buckwalter produces a set of solutions for each token 
%and the several analysis threads of the text need to be 
%generated as a product of the token solutions. 
%
%Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
%finite state Automata (FSA). 
%It reports that the number of Automata generated by a compiler
%for Xerox rules can not be controlled and also reports a difficulty 
%to compose the FSAs into a single framework. 
%Sarf however constructs an NDFSA from
%the controller and the prefix, stem, and suffix deterministic 
%transducers to solve the problem.
%Sarf also differs in that it uses recursive affixes in the 
%affix transducers, and in that the transducers consider
%input from the controller to modify their behavior. 
%%With Sarf it is easy to control the FSA with a case-based
%%controller. 
%Beesley requires recompilation of the Xerox rules.
%
%Similar to The MADA+TOKAN toolkit, Sarf thinks of
%the several valid morphological analyses as a richness that 
%should be exploited at a higher abstract level rather than
%an inaccuracy that should be corrected. 
%Sarf differs in that it provides an interface for the 
%case study to interfere in judging the analyses much earlier and
%at a finer granularity level. 
%In Sarf, there is no need for a separate tokenizer such as
%TOKAN since each solution keeps a stack of positions
%that partition text into morphemes.
%%The case-based controller can build the trace 
%%when needed. 
%
%%MAGEAD 
%%Maamouri, Mohamed and \bgroup et al.\egroup },
%
%Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
%SAMA was refined to interact with
%the ATB~\cite{Maamouri:04} project after the addition of a large 
%new corpus. 
%%We find supporting evidence in this decision to our hypothesis
%%about the utility of a case-based analyzer. 
%The algorithmic changes in SAMA were
%done manually and worked in integration with the ATB format. 
%Sarf is a case-based controller approach that can modify 
%the behavior of the analyzer on the fly and can interact
%with any case study. 
%%In Sarf, the analyzer remains general, and the 
%%case-based controller specializes the analysis.
%
%The AMIRA~\cite{Diab:07} analyzer uses 
%a language independent SVM based analysis. 
%The SVM analysis learns features from the input text 
%and does not make use of any information 
%in the target query.
%
%Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
%of the Buckwalter analyzer. 
%Sarf also uses deterministic parsing with tries 
%to implement the affix and stem transducers. 
%We think that the inferential-realizational approach 
%of ElixirFM
%that is highly compatible with the Arabic linguistic 
%description~\cite{Badawi:04}
%can benefit from many features unique to the Arabic language.
%Sarf leaves implementing that to the case-based controller
%since in many cases the abstraction needs only a partial 
%linguistic model of the Arabic language. 
%%We can compare the case-based controller in Sarf to a high level
%%component in the ElixirFM domain specific language. 
%

%Sarf encodes the prefix, suffix and stem lexicons into
%three deterministic transducers and 
%uses a non-deterministic construction
%of the three to compute all valid morphological analyses.
%We first present a running example of Sarf.
%The running example explains how Sarf works,
%implements recursive affixes, and
%handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

\section{Narrator chain extraction }
\label{sec:islamic}

%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
%the analysis of the hadith literature. 
In this paper, we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
%  Our understanding is that those books are historical documents with 
 % public IP.}.
%\setcode{utf8}
%\begin{arabtext}
%حدثنا \emphasize{هشيم}، أخبرنا \emphasize{إسماعيل بن أبي خالد}،
% عن أبي إسحاق، عن سعيد بن جبير، قال كنا
%    مع ابن عمر رضي الله عنه حيث أفاض من
%    عرفات إلى جمع فصلى بنا المغرب ومضى
%\end{arabtext}
%\setcode{standard}
NaCEMA accepts a book as input
and segments it into a vector of hadiths. 
NaCEMA segments each hadith into its
sanad and matn parts. 
NaCEMA 
detects the chain of narrators in the sanad and 
the relation that links each narrator to his ancestor and 
predecessor in the chain. 
It also detects the full name of each narrator that is
composed of several proper names with connectors
in between.

\subsection{The hadith controller}
\label{sec:controller}
% FSM for the controller

Intuitively, the controller reports a valid chain 
of narrators when a sequence of names
connected by narrator connectors appears. 
This includes the task of finding multi-word names
such as \RL{`bd alr.hmn} often appearing as ``run-on'' words.
The controller marks the beginning of the hadith with the beginning of 
the current proper name sequence,
and marks the end of the hadith with the beginning of the 
next proper name sequence. 
It detects the chain of narrators as the sequence itself.

%The controller also targets words that mean ``narrate'' when
%they appear between proper names. 
We illustrate the hadith controller 
in Figure~\ref{f:hadith}. 
The transitions are labeled by outputs generated from 
the Sarf morphological analyzer that correspond to 
part of speech (POS) tags. 
A NAME label denotes a proper name.
An NRC label denotes a narrator connector such as
\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
\RL{'a_hbar} ``told'' or one of their morphological equivalents. 
An IBN label denotes the word \RL{ibn} ``the son of'' and is a name connector.
A NISBA is a possessive adjective that qualifies a person such 
as \RL{alma.sriy} ``the Egyptian''.

The symbol $\tau_{\mbox{NMC}}$ is a threshold
that corresponds to the number of tolerated name connectors 
that may occur between two names. %3
The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
of name connectors collected since the controller
started looping in the state NMC\_S. 
The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
that corresponds to a relaxed tolerance measure that
the controller resorts to in case the words separating
two names were longer than $\tau_{\mbox{NMC}}$ but 
contained a name connector word such as IBN or NISBA.

The controller has four states that correspond to 
a position in text relative to the next sanad. 
State TEXT\_S is the initial state and denotes that
the controller is outside the context of a sanad.
The controller moves to state NAME\_S on
NAME.
State NRC\_S means that the controller is in the context
of a narrator name after Sarf reports an NRC.
State NMC\_S
indicates that the controller expects a name to appear within 
a tolerance threshold expressed by 
$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.

Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
before it gives up on its expectations. 
We reach the NAME\_S state only when a
valid NAME is detected and we leave when no more NAME's are detected.
The NRC\_S state can only be reached if an NRC is detected.
NMC\_S can only be reached from a NAME\_S state.
We used the values of 3, 5 and 5 for $\tau_{\mbox{NMC}}$, 
$\lambda_{\mbox{NMC}}$, and $\tau_{\mbox{NRC}}$ respectively.

\begin{figure}[tb!]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/hadith.pdftex_t}}
\caption{The controller of the hadith application.}
\label{f:hadith}
}
\end{figure}

%The definition of input labels such as NAME and IBN depends on the 
%morphological analyzer. 
%However, our case-based controller approach was able to perform well
%under both Sarf and a refined version of Sarf.

\section{The narrator graph}

\begin{figure}[tb]
\center{
\resizebox{1.1\columnwidth}{!}
%{\includegraphics{figs/narrator_chain_output.pdf}}
{\includegraphics{figs/DAG_narrator.pdf}}
\caption{Narrator directed acyclic graph 
    extracted using NaCEMA }
\label{f:narrators} 
}
\end{figure}

The directed acyclic graph (DAG) 
in Figure~\ref{f:narrators} shows a partial order relation (POR) 
between narrators.
We automatically extracted the POR from three books of hadith 
selected arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}.
The nodes in boxes are the \RL{matn} of the hadith, 
and the other nodes are the narrators.
Given a set of chains of narrators similar to 
Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
that merged equal names in the chains. 
We compared the narrators together using a morphological
distance function. 
We considered each narrator to have several morphological 
components and we rewarded every morphological match with 
a small $\delta$. 
Then we apply a threshold $\tau$ to the sum of rewards
decide the equality of the two narrators. 
We omitted the details of the distance function for brevity.

\section{Results}
\label{sec:results}

\begin{table}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\resizebox{1.1\columnwidth}{!}{
\begin{tabular}{lcp{.2cm}cp{.2cm}c} %\cline{2-10}
 &  AlKafi & & AlIstibsar & &IbnHanbal \\ \cline{1-6}
Word count  &98,943 & & 103,835 & & 20,354 \\ 
 Names       & 12,060  & & 14,613& & 3,013\\
%Names/Narrator & 1.97 & & 1.84& & 1.25 \\
Narrators  & 2,623 & & 5,767& & 1,755 \\ 
%Narrators/Chain  & 4.84 & & 4.76 & &4.05 \\
Chains  & 542 &  & 1,211& & 433 \\ 
Ignored names  & 6,400 &  & 3,348 & & 642 \\ \hline
Segmentation accuracy  & 96\%& & 96\%& & 92\%\\ 
Chain accuracy & 99\%&  & 99\%& & 97\% \\ 
%Narrator accuracy  & & & & &  & & & \\ 
Narrator accuracy  & 91\%& & 90\% & & 90\% \\ \hline
%Name false positives  & 7\%&  & 4\% & & 4\% \\ \hline
Running time (secs.)& 1.32 & & 1.31 & & .096\\ \hline 
\end{tabular}
}
\normalsize
\label{t:hadithresallresults}
\end{table}

We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
with 4GB of memory using a Linux operating system. 
We report our results in Table~\ref{t:hadithresallresults} on the 
three hadith books.
NaCEMA detected narrator chains and segmented the books 
correctly with an accuracy above 90\%
in all cases and across all the traditions in the three books.

The row for ``word count'' is an indicator of the size of the data.
The %numbers in 
rows ``names'', ``narrators'', and ``chains'' denote the total
number of the corresponding entities per book.
The row ``ignored names'' reports the number of names detected 
outside the chains and thus are irrelevant to
the application. 
%In addition, the rows ``names/narrator'' and ``narrators/chain'' 
%report the average
%number of names per narrator and 
%the average number of narrators per chain respectively.

The ``segmentation accuracy'' reports on the accuracy of 
segmentation of the book into separate hadiths.
We computed the metric manually by marking the beginning 
and end of a representative sample of hadiths in each 
book and compared that to the NaCEMA results. 
We used the same approach to compute all other accuracy measures.
The ``chain accuracy'' reflects the percentage of correct narrators 
covered by extracted chains.
The ``narrator name accuracy'' row reflects the percentage of 
correct proper names detected within
narrator structures.

%The row ``name false positives'' reports the percentage of words 
%recognized as names while they should not.
%We notice that this metric did not affect the accuracy
%of narrator and chain segmentation since detecting
%a narrator or a chain with the refined analysis without substantially affecting the 
%high abstract measures such as chain and segmentation
%accuracy.

\section{Related work }
\label{sec:related}

The closest to our work is e-Narrator~\cite{Azmi2010}, a 
tool that uses a grammar based approach to solve 
the narrator extraction problem. 
The e-Narrator tool preprocesses the text to remove 
punctuation and diacritics and to normalize white
spaces. 
Then e-Narrator uses a similarity-based approach
to memory based learning~\cite{Azmi2010} in order to perform 
shallow parsing of the preprocessed text. 
The e-Narrator tool uses a parser based on a context
free grammar that enumerates several ways to express
the name of the prophet, several possible words that 
might precede the name of the prophet, several possible
words that might separate two narrators, and several
words that might separate parts of the name of the narrators.
The e-Narrator tool fails when a variation of one of the 
listed words and not an exact form of it
happens in the text. 
In their grammar they consider a name to be any sequence of 
legal Arabic characters. 
The e-Narrator tool generates successfully 86.7 percent of
the narration chains of 90 selected traditions with 
34 simple cases and 56 hard cases.

NaCEMA does not need a preprocessing step since 
it uses morphological analysis to process the text. 
While e-Narrator bases its analysis on the patterns 
of words that surround the narrator names, 
NaCEMA bases its analysis on the fact that narrator
names are likely to happen in localities with a high
number of proper names. 
NaCEMA uses the narrator connectors as helpers and
uses morphological analysis to compare the stems of
those words and thus NaCEMA avoids the problem of 
``parser noise words'' that e-Narrator faces. 
Even though we do not have access to the selected
90 traditions e-Narrator reports on, we claim
that NaCEMA outperforms e-Narrator as it works on
the full text of the presented hadith books
and reports a higher success rate.

\section{Conclusion and future work}
\label{sec:future}

We presented NaCEMA, a narrator chain extractor
using morphological analysis. 
NaCEMA extracted narrator chains with above 90\% accuracy
from three different hadith books and generated 
directed acyclic graphs relating the narrators 
in a partial order relation.

This partial order graph is instrumental to automate
localization of narrators in biography documents and
segmenting biography documents in the future.
Given that a biography discusses a narrator 
and mentions his professors and students,
we plan to segment the biographies with a graph 
coloring algorithm
that traverses the text and colors the POR whenever
a name is found.

Once the biographies are analyzed, we can annotate
the POR with qualifiers of the narrators that reflect
their authenticity. 
%We can also annotate the POR with the locations and 
%the time the narrators lived in. 
%Then we can compute a time and location overlap
%check to discover inconsistent narrations.
%We can also perform several interesting checks using 
%this POR on its own such as checking for the effect of
%one narrator on the hadith literature.

We also plan to 
enhance the accuracy of hadith detection and the 
narrator merging algorithms.

%We plan to make use of the name connector and narrator
%connector stop phrases that our case study extracted to create a 
%learning system that can learn more names. 
%We also plan on developing an Arabic linguistic computational model 
%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
%explored partially based on the case study.

%\section{Acknoledgements}
%\label{sec:acc}
%We thank the Lebanese National Council for Scientific Research (LNCSR) for funding this research.

%For reasons of uniformity, Adobe's {\bf Times Roman} font should be
%used. In \LaTeX2e{} this is accomplished by putting

%\begin{quote}
%\begin{verbatim}
%\usepackage{times}
%\usepackage{latexsym}
%\end{verbatim}
%\end{quote}
%in the preamble.

%Additionally, it is of utmost importance to specify the {\bf
%  US-Letter format} (8.5in $\times$ 11in) when formatting the paper.
%When working with {\tt dvips}, for instance, one should specify {\tt
%  -t letter}.

%{\bf Citations}: Citations within the text appear
%in parentheses as~\cite{Gusfield:97} or, if the author's name appears in
%the text itself, as Gusfield~\shortcite{Gusfield:97}. 
%Append lowercase letters to the year in cases of ambiguities.  
%Treat double authors as in~\cite{Aho:72}, but write as 
%in~\cite{Chandra:81} when more than two authors are involved. 
%Collapse multiple citations as in~\cite{Gusfield:97,Aho:72}.

%\section*{Acknowledgments}
% this will go into the regular submission

%\bibliographystyle{naaclhlt2010}
\bibliographystyle{acl}
{\small \bibliography{fzAr}}

\end{document}

### 2011-02-25 tex/hadith.short.tex
last change

--- 2011-02-25 trunk/tex/hadith.short.tex
@@
 in Islamic studies. 
-Based on them Islamic scholars issue rules that affect the life
+Based on that Islamic scholars issue rules that affect the life
 of people around the world. 
@@
 \section{Introduction}
+\novocalize
 
@@
 \transtrue
+\vocalize
 
@@
 For example, 
+\novocalize
 \noTrRL{qtybT} is the first name 
@@
 \RL{matn} (content) and constitutes the actual content 
-and tradition.
+of the tradition.
 
@@
 This work will lead to 
-automate an exhaustive automated \noArRL{.hady_t} authentication 
+automate an exhaustive \noArRL{.hady_t} authentication 
 effort in the future as we plan to 
@@
   \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
-  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }
+  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
 %  Our understanding is that those books are historical documents with 
@@
 between narrators.
-We automatically extracted the POR from three books of hadith 
-selected arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}.
+We automatically extracted the POR from the three books of 
+hadith~\cite{IbnHanbal,AlKulayni,AlTousi}.
 The nodes in boxes are the \RL{matn} of the hadith, 
@@
 a small $\delta$. 
-Then we apply a threshold $\tau$ to the sum of rewards
-decide the equality of the two narrators. 
+Then we applied a threshold $\tau$ to the sum of rewards
+and decided the equality of the narrators. 
 We omitted the details of the distance function for brevity.
@@
 NaCEMA detected narrator chains and segmented the books 
-correctly with an accuracy above 90\%
-in all cases and across all the traditions in the three books.
+correctly with a chain accuracy above 97\% and a narrator name
+accuracy above 90\%
+in all cases and across all the 1986 traditions in the three 
+different books with different formats.
 
-The row for ``word count'' is an indicator of the size of the data.
+The row ``word count'' is an indicator of the size of the data.
 The %numbers in 
@@
 coloring algorithm
-that traverses the text and colors the POR whenever
+that traverses the text and colors the POR node whenever
 a name is found.

### 2011-02-26 tex/sarf.aclhlt.tex
presumably final version of Sarf

@@
 Natural language processing (NLP) uses morphological 
-analysis to abstract and annotate text.
+analysis to annotate text.
 Often times, 
@@
 a branch of morphological analysis may not be
-appropriate for the case study at hand. 
-In this paper we present Sarf, an application-specific morphological analyzer for Arabic. 
-Sarf allows an application-specific controller to control and refine
-the state of the analysis on the fly. 
-It uses non-deterministic finite state Automata
+appropriate for the NLP application. 
+We present Sarf, 
+%an application-specific morphological analyzer for Arabic. 
+an Arabic application-specific morphological analyzer, which
+%Sarf
+allows a controller to refine
+the analysis on the fly. 
+It uses non-deterministic finite state Automata % (NDFSA)
 to analyze a stream of text. 
 %Each machine corresponds to a valid analysis. 
-Sarf uses concatenative analysis based on recursive affixes 
-to better preserve part of speech information.
+Sarf uses concatenative analysis 
+%based on
+with  
+recursive affixes 
+to better preserve part of speech (POS) information.
+It solves the issue of ``run-on'' words 
+and utilizes diacritics for disambiguation.
 %We automated the analysis of three books of Islamic
@@
 
-%Automated analysis of Arabic data sets, including texts, 
-%publications, records and digital media is essential
-%with the huge digital Arabic content available nowadays. 
 Morphological analysis is key in current automated 
 analysis techniques for Arabic text. 
-%Current morphological analyzers~\cite{Sughaiyer:04}
-%use concatenative analysis when
-%considering the internal structure 
-%of an Arabic word and
-%composing it into several {\em morphemes}. 
-%A morpheme can be a {\em stem}, or an {\em affix}.
-%An affix can be a {\em prefix, suffix, } or an {\em infix}.
-%The analysis of one word may lead to several possible
-%morphological solutions.
-%\vocalize
-%For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
-%may have two valid morphological analyses. 
-%The letter \RL{'a} may be a prefix and the word means 
-%``I praise him'', or 
-%%The letter \RL{'a} may also be 
-%part of the stem \RL{'a.hmad} (a proper noun)
-%and the word means ``his Ahmad''.
-%
-%\novocalize
-The accuracy of the solutions suffer due to inherent difficulties
-of morphological analysis of the Arabic language. 
+The accuracy of the solutions suffers due to inherent difficulties
+of morphological analysis of Arabic. 
 For example, it is common practice to write Arabic text
-without short vowels / diacritics. 
-This greatly increases the ambiguity of Arabic text.
-%(maybe reduce the couple of sentences below)
-Moreover, Arabic letters can have up to 
-four different forms
-corresponding to their position in a word, i.e, beginning,
-middle, end of word and separate forms. 
-This allows the phrase 
+without short vowels (diacritics) 
+which greatly increases ambiguity.
+Moreover, tokenization is not always 
+trivial.
+For example,
 \noTrnoVocRL{il_A\nospace almdrsT}
-to be visually recognizable
+can be visually recognizable
 as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school) 
 without the need of a space in between. 
-The reason is the first word \noVocRL{il_A} ends with
-\noVocRL{_A} a non-connecting letter. 
-These words,
-referred to as ``run-on'' words~\cite{Buckwalter:04},
-occur often, and greatly increase the
-difficulty of tokenization.
+The reason is that
+\noTrnoVocRL{_A} is a non-connecting letter. 
+Such 
+``run-on'' words~\cite{Buckwalter:04},
+occur often.
 
@@
 In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
-has been submitted with the paper and will be available online as an open source tool.},
- a {\em novel application-specific %efficient 
-enhanced %(instead of 'efficient' to engulf both efficiency and other refinements)
+has been submitted with the paper and will be available 
+as an open source tool.}, 
+a {\em novel application-specific efficient 
+%(maybe 'enhanced' instead of 'efficient' to engulf both efficiency and other refinements)
 morphological analyzer} that uses 
-non-deterministic finite state Automata 
+non-deterministic finite state Automata (NDFSA)
 driven by an application-specific controller.
@@
 and represents a valid morphological analysis.
-The application-specific controller prunes false positives
+%The application-specific controller prunes false positives
+The controller prunes false positives
 early in the run by making a decision at every input letter
@@
 %it can be a logical entity of any type.
-The controller can be an FSA or any other %logical entity 
-programming entity of any type,
-since Automatas may be a very complex solution for some simple applications.
+%since Automatas may be a %very complex
+%non-necessary solution for some simple applications.
 %(maybe re-word this sentence)
+%depending on application needs.
 
-%Each alive machine in Sarf takes as input one letter at a time 
-%from a text stream.
-In addition, to being application-specific, Sarf has introduced 
-many enhancements to the field of arabic morphological analysis. 
-For example,
-Sarf does not assume that the word at hand is a token and
-performs tokenization on the fly based on morphological correctness.
-Up to our knowledge, this is the first morphological analyzer that 
-handles the ``run-on'' words problem.
-
-Sarf uses concatenative analysis with a novel %(maybe remove novel bc authors did not feel it is much)
-{\em recursive affixes} refinement,
-where an affix can be a sequence of other affixes which are concatenatively compatible.
-%An atomic affix $a$ is a standalone affix
-%that is not the concatenation of other affixes. 
-%We define an affix $f: a | a f$ to be either
-%an atomic affix $a$ or an atomic affix concatenated
-%with an affix. 
-Sarf introduces recursive affixes in order to
-better retain part of speech information and enhance the 
-efficiency of affix matching.
-
-In addition, Sarf supports disambiguation using diacritics when they are available.
-This feature, to the best of our knowledge is the first attempt to benefit from 
-diacritics instead of ignoring them. Prior to Sarf, the trend was to preprocess 
-the text to remove any such `unwelcomed' letter. %Nonetheless, Sarf regards them as a richness to be utilized that should not be disregarded.
-
-%Moreover, since Sarf performs tokenization on the fly, Sarf is not just capable of solving
-%the problem of ``run-on'' words, but also of handling Multiword Expressions (MWE). 
-Moreover, since Sarf does not perform white-space delimited tokenization, Sarf is capable of 
-handling Multiword Expressions (MWEs).
-MWEs are defined in~\cite{MWE}
-``as idiosyncratic interpretations
-that cross word boundaries (or spaces)''. 
-%By this term, we refer to frequent cases when a group of words form together a single
-%item with a semantical meaning. In this paper, we focus on the example of compound names such as 
-%\noVocRL{`bid alkarym}. 
-%In current analyzers, this phrase is analyzed word by word and in best cases is 
-%considered to contain two names, instead of being understood to be a single but compound name.
-
-Our focus when building Sarf, was not on improving the semantical rules and 
-expanding the morphological alternatives, instead we worked on adding support 
-for higher level control by the specific application utilizing the lower level 
-morphological analyzer. %In other words, the application-specific controller, has 
+%(check if position of this paragraph seems in correct place)
+%Our focus when building Sarf, was not on improving the semantical rules and 
+%expanding the morphological alternatives, instead we worked on adding support 
+%for higher level control by the specific application utilizing the lower level 
+%morphological analyzer. %In other words, the application-specific controller, has 
 % full control over and directs the morphological analysis of Sarf. 
@@
 %application needs, which is an issue that Sarf tackles. % (maybe reworded)
-We strongly believe that by being
-an application-specific morphological analyzer that allows an application-specific 
-controller to intervene at every decision to guide, use, order, and 
-refine the morphological analysis, Sarf is of high utility.
+We hypothesize that in many NLP applications, the morphological analyzer may
+benefit from the characteristics of the query at hand, 
+to greatly enhance the analysis.
+Consequently, %we believe that by 
+%being 
+%allowing
+an application-specific %morphological 
+%analyzer that allows %an application-specific 
+%the
+controller that intervenes at every decision to guide, use, order, and 
+refine the morphological analysis is of high utility.
 
-%We validate our hypothesis using a case study from the Islamic 
-%literature, introduced later in Section~\ref{sec:results}.
-In addition to all its enhanced analysis methods, Sarf still
-outperformed current morphological analyzers in terms of running time.
-These results will be presented in Section~\ref{sec:results}.
+%Each alive machine in Sarf takes as input one letter at a time 
+%from a text stream.
+%Apart from being application-specific, 
+Sarf %has
+also  
+introduces 
+several enhancements. %to the field of Arabic morphological analysis. 
+%the following enhancements:
+%For example,
+First,
+Sarf %does not assume that the word at hand is a token and
+performs tokenization on the fly based on morphological correctness.
+Up to our knowledge, this is the first %morphological 
+analyzer that 
+handles %the ``run-on'' words problem. 
+``run-on'' words. 
+%
+%Moreover, since Sarf performs tokenization on the fly, Sarf is not just capable of solving
+%the problem of ``run-on'' words, but also of handling Multiword Expressions (MWE). 
+On the fly tokenization also handles
+Multiword Expressions (MWEs)~\cite{MWE} such as \noVocRL{`bid alkarym}.
+%MWEs are defined in~\cite{MWE} as
+%which are 
+%``idiosyncratic interpretations
+%that cross word boundaries (or spaces)''. 
+%By this term, we refer to frequent cases when a group of words form together a single
+%item with a semantical meaning. In this paper, we focus on the example of compound names such as 
+%\noVocRL{`bid alkarym}. 
+%In current analyzers, this phrase is analyzed word by word and in best cases is 
+%considered to contain two names, instead of being understood to be a single but compound name.
 
-We make the following key contributions: 
-\begin{enumerate}
-\item {\bf application-specific analysis:}  We designed and implemented
-a application-specific morphological analyzer where a controller
-guides the analysis. 
-%\item {\bf Exhaustive analysis:} Since we have an application-specific 
-%controller that can decide on the fly whether an analysis is 
-%accepted or not, we can afford to keep all valid analyses modulo
-%those pruned by the case controller. 
-%We do that using non-deterministic FSAs. 
-\item {\bf Recursive affixes:}
-We refine the concatenative analysis of 
-Buckwalter~\shortcite{Buckwalter:02} and define affixes
-recursively. 
-This allows Sarf to perform better and 
-maintain compositional part of speech (POS) tags.
-\item {\bf Diacritic-based Disambiguation:}
-Diacritics are used to disambiguate morphological
-alternatives using any existant diacritics.
-\end{enumerate}
+Second, 
+Sarf uses concatenative analysis with a novel %(maybe remove novel bc authors did not feel it is much)
+{\em recursive affixes} refinement,
+%where an affix can be a sequence of other affixes which are concatenatively compatible.
+where an affix can be a sequence of other compatible affixes.
+%An atomic affix $a$ is a standalone affix
+%that is not the concatenation of other affixes. 
+%We define an affix $f: a | a f$ to be either
+%an atomic affix $a$ or an atomic affix concatenated
+%with an affix. 
+%Sarf introduces recursive affixes in order to
+%This refinement is introduced to better retain
+Hence, it better retains
+%part of speech information
+POS information  
+and enhances the 
+efficiency of affix matching. 
+Third, Sarf supports disambiguation using 
+%diacritics when they are available.
+available diacritics, instead of the traditional approach that ignores them.
+%This feature, to the best of our knowledge is the first attempt to benefit from 
+%To the best of our knowledge,
+%Up to our knowledge, this is the first attempt to benefit from
+%diacritics instead of ignoring them. 
+%Nonetheless, Sarf regards them as a richness to be utilized that should not be disregarded.
+
+%In addition to all its enhanced analysis methods, Sarf still
+%Sarf also
+%outperformed current morphological analyzers in terms of running time,
+%%These results will be presented 
+%as shown in Section~\ref{sec:results}.
+Sarf is also more efficient than current morphological analyzers, 
+as shown in Section~\ref{sec:results}.
+
+%%We make the following 
+%We summarize our key contributions: 
+%\begin{enumerate}
+%\item {\bf application-specific analysis:}  We designed and implemented
+%a application-specific morphological analyzer where a controller
+%guides the analysis. 
+%%\item {\bf Exhaustive analysis:} Since we have an application-specific 
+%%controller that can decide on the fly whether an analysis is 
+%%accepted or not, we can afford to keep all valid analyses modulo
+%%those pruned by the case controller. 
+%%We do that using non-deterministic FSAs. 
+%\item {\bf Recursive affixes:}
+%We refine the concatenative analysis of 
+%Buckwalter~\shortcite{Buckwalter:02} and define affixes
+%recursively. 
+%This allows Sarf to perform better and 
+%maintain compositional part of speech (POS) tags.
+%\item {\bf Diacritic-based Disambiguation:}
+%Diacritics are used to disambiguate morphological
+%alternatives using any existant diacritics.
+%\end{enumerate}
 
@@
 Analyzers such as~\cite{Khoja:01,Darwish:02} 
-target specific applications in the 
-analyzer itself or use a specific set of POS tags
-as their reference.
+target specific applications 
+%in the analyzer itself or use a specific set of POS tags as their reference
+or limit themselves to a specific set of POS tags. 
 Sarf differs in that it is a general morphological 
@@
 
-Current morphological analyzers such as %(maybe disambiguators also added here according to reviewer comments)
 Buckwalter~\shortcite{Buckwalter:02},
@@
 and ElixirFM~\cite{Otakar:07} 
-take as input white space delimited tokens~\cite{Kulick:10},
-consider them as words,
-and enumerate all possible solutions. 
-%This approach has several problems. 
-However, we note the following points that deserve consideration:
-%(We need not to mention especially point 2 as a "problem" according to reviewer comments)
-First, a white space delimited token may have 
-more than one word or just a part of a MWE.
+consider white-space delimited tokens~\cite{Kulick:10}
+as words.
+%This suffers has several problems. 
+%However, we note the following: %(I thought of this change not to include point 2 as a "problem" according to reviewer comments)
+This fails when a white-space delimited token has
+more than one word or is part of an MWE.
 %Second, the exhaustive enumeration may hurt performance and may
@@
 %in some case studies as noted in~\cite{Maamouri:10}. 
-Second, in many applications, the exhaustive enumeration may not be needed or appropriate as noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.
-Other morphological analyzers such as 
-Amira~\cite{Diab:07,Benajiba:07},
+These analyzers also exhaustively enumerate all 
+possible morphological solutions 
+which may not be appropriate in many applications
+~\cite{Maamouri:10}. 
+%consequently, adding unnecessary running time.
+
+Others such as 
+AMIRA~\cite{Diab:07,Benajiba:07},
 MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
-% (reviewer wants to distinguish MADA as a tool for morphological disambiguation instead of analysis)
 use machine learning and support vector machines (SVM) 
-to enhance the accuracy of the morphological analysis at the expense 
+to disambiguate the morphological analysis at the expense 
 of efficiency.
-Xerox rules can be compiled into specialized finite state
-machine (FSM) based analyzers as described in~\cite{Beesley:03}.
-However, the efficiency of the resulting analyzer depends on the
-way the Xerox rules are written. 
+Xerox rules can be compiled into specialized 
+finite state Automata (FSA)~\cite{Beesley:03}.
+%FSA-based analyzers~\cite{Beesley:03}.
+The efficiency of the resulting analyzer depends on the
+way the rules are written. 
 This requires deep knowledge and insight from the user
@@
 %into efficient non-deterministic finite state Automata (NDFSA). 
-
-Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
-It differs from Buckwalter in that it defines a shorter list of affixes
-and a longer list of affix compatibility rules to allow 
-affixes to be defined recursively
-so that we can have prefix-prefix an suffix-suffix 
-concatenations.
-This allows Sarf to better maintain and propagate 
-the POS tags associated with affixes. 
-%This information is highly 
-%redundant with Buckwalter.% and that may lead to consistency issues. 
-Sarf keeps all the possible analyses alive where each analysis
-corresponds to a set of decisions and positions in the text. 
-Buckwalter produces a set of solutions for each token 
-and the several analysis threads of the text need to be 
-generated as a product of the token solutions.
-
-Similar to Sarf, both Beesley~\shortcite{Beesley:01} and Attia~\shortcite{Attia:05} use
-finite state Automata (FSA). 
 Beesley reports that the number of Automata generated by a compiler
@@
 to compose the FSAs into a single framework. 
+%Beesley requires recompilation of the Xerox rules. 
+% (I am not sure if it is ok to have the above removed)
 Sarf however constructs an NDFSA from 
-the controller and the prefix, stem, and suffix deterministic 
+%the controller and 
+the prefix, stem, and suffix deterministic 
 transducers to solve the problem.
-Sarf also differs in that it uses recursive affixes in the 
-affix transducers, and in that the transducers consider
+Sarf also differs in that 
+%it uses recursive affixes in the affix transducers, and in that 
+the transducers consider
 input from the controller to modify their behavior.
-%With Sarf it is easy to control the FSA with a application-specific
-%controller. 
-Beesley requires recompilation of the Xerox rules. 
-Attia has researched extensively MWEs handling methods~\cite{Attia:06} and 
-automated extending their dictionary coverage~\cite{Attia:10}. 
+The user can express the controller,not necessarily an FSA, in a programming language such as C++.
+
+Sarf builds upon the lexicon of Buckwalter. %\shortcite{Buckwalter:02}, %(already cited above)
+%It differs from Buckwalter in that it defines a shorter list of affixes
+It defines a shorter list of affixes
+and a longer list of 
+%affix 
+compatibility rules to define 
+affixes recursively and allow 
+prefix-prefix and suffix-suffix 
+concatenations.
+%This allows Sarf to better maintain and propagate 
+This better maintains and propagates 
+the tags associated with affixes. 
+%This information is highly 
+%redundant with Buckwalter.% and that may lead to consistency issues. 
+% (check if necessary to keep the 2 sentences below)
+%Sarf keeps all the possible analyses alive where each analysis
+%corresponds to a set of decisions and positions in the text. 
+%Buckwalter produces a set of solutions for each token 
+%and the several analysis threads of the text need to be 
+%generated as a product of the token solutions. 
+Similar to Sarf, Attia uses FSAs and handles more and qualitatively different MWEs
+~\cite{Attia:06,Attia:10}.
+%Attia has researched extensively MWEs handling methods~\cite{Attia:06} and 
+%automated extending their dictionary coverage~\cite{Attia:10}. 
 % (I am not sure how clear this statement is)
 
-Similar to The MADA+TOKAN toolkit, Sarf thinks of
-the several valid morphological analyses as a richness that 
-should be exploited at a higher abstract level rather than
-an inaccuracy that should be corrected. 
-Sarf differs in that it provides an interface for the 
+%Similar to %the 
+%MADA+TOKAN, Sarf uses
+%the several valid morphological analyses as 
+%%a richness that should 
+%richness at a higher abstract level, and does not view them as 
+%inaccuracies to be corrected. 
+Sarf differs from MADA+TOKAN in that it provides an interface for the 
 application to interfere in judging the analyses much earlier and
 at a finer granularity level. 
-In Sarf, there is no need for a separate tokenizer such as
+%In Sarf, there is no need for a separate tokenizer such as
+Sarf does not need a separate tokenizer such as
 TOKAN since each solution keeps a stack of positions
@@
 
-Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
-SAMA was refined to interact with
-the ATB~\cite{Maamouri:04} project after the addition of a large 
-new corpus. 
-%We find supporting evidence in this decision to our hypothesis
-%about the utility of a application-specific analyzer. 
-The algorithmic changes in SAMA were
-done manually and worked in integration with the ATB format. 
-Sarf is an application-specific controller approach that can modify 
-the behavior of the analyzer on the fly and can interact
-with any application as needed. 
-In Sarf, the analyzer remains general, and the 
-application-specific controller specializes the analysis.
+SAMA~\cite{Maamouri:10} was refined after the addition of a new corpus to 
+the ATB~\cite{Maamouri:04}, since the new data challenged the existing processing and 
+annotation guidelines.
+The modification involved interaction between SAMA and the NLP 
+task and resulted in alternative guidelines. 
+Sarf supports such an interaction on the fly.
 
-The AMIRA~\cite{Diab:07} analyzer uses 
-a language independent SVM based analysis. 
-The SVM analysis learns features from the input text 
-and does not make use of any information 
-in the target query.
+%The AMIRA %~\cite{Diab:07}
+%analyzer uses 
+%a language independent SVM-based analysis. 
+%%The SVM 
+%This 
+%analysis learns features from the input text 
+%but does not make use of any information 
+%in the target query.
 
 Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
-of the Buckwalter analyzer. 
+of Buckwalter. %analyzer. 
 Sarf also uses deterministic parsing with tries 
 to implement the affix and stem transducers. 
-We think that the inferential-realizational approach 
+The inferential-realizational approach 
 of ElixirFM
-that is highly compatible with the Arabic linguistic 
-description~\cite{Badawi:04}
-can benefit from many features unique to the Arabic language.
-Sarf leaves implementing that to the application-specific controller
-since in many cases the abstraction needs only a partial 
-linguistic model of the Arabic language. 
+is highly compatible with the Arabic linguistic 
+description~\cite{Badawi:04}.
+%can benefit from many features unique to the Arabic language.
+Sarf leaves implementing that to the %application-specific 
+controller
+since in many applications the abstraction needs only a partial 
+linguistic model. 
 %We can compare the application-specific controller in Sarf to a high level
@@
 
-Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with
-proper names and location names extracted from different online 
-sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
-\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
-as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
+%Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with
+%proper names and location names extracted from different online 
+%sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
+%\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
+%as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
 
@@
 of the three to compute all valid morphological analyses.
-We first present a running example of Sarf.
-The running example explains how Sarf works,
+We first present a running example 
+%of Sarf. The running example 
+that 
+explains how Sarf works,
 implements recursive affixes, and
@@
 
-\subsection{Sarf example}
-\label{sec:example}
+%\subsection{Sarf example}
+%\label{sec:example}
 
@@
 
-%{\bf Sarf example.}
-
+{\bf Sarf example.~~~}
 The diagram in Figure~\ref{f:example}
@@
 when parsing the
-string \noTrnoVocRL{wsyl`b-h-aa\nospace al-lA`bwn}.
+string \noTrnoVocRL{wsyl`b-h-aa\nospace al--lA`bwn} \noArRL{wsyl`b-h-aa al--lA`bwn}.
 %\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
-\footnote{ 
-\noVocRL{n}%
-\noVocRL{w}
-\noVocRL{b}
-\noVocRL{`}
-\noVocRL{a} 
-\noVocRL{l}
-\noVocRL{l}
-\noVocRL{a} %
-\noVocRL{a}
-\noVocRL{h}
-\noVocRL{b}
-\noVocRL{`}
-\noVocRL{l}
-\noVocRL{y}
-\noVocRL{s}
-\noVocRL{w}
-in separate form to ease following the example
-}
+%\footnote{ 
+%%\noVocRL{n}%
+%%\noVocRL{w}
+%%\noVocRL{b}
+%%\noVocRL{`}
+%%\noVocRL{a} 
+%%\noVocRL{l}
+%%\noVocRL{l}
+%%\noVocRL{a} %
+%%\noVocRL{a}
+%%\noVocRL{h}
+%%\noVocRL{b}
+%%\noVocRL{`}
+%%\noVocRL{l}
+%%\noVocRL{y}
+%%\noVocRL{s}
+%%\noVocRL{w}
+%\noTrnoVocRL{n}%
+%\noTrnoVocRL{w}
+%\noTrnoVocRL{b}
+%\noTrnoVocRL{`}
+%\noTrnoVocRL{a} 
+%\noTrnoVocRL{l}
+%\noTrnoVocRL{l}
+%\noTrnoVocRL{a} %
+%\noTrnoVocRL{a}
+%\noTrnoVocRL{h}
+%\noTrnoVocRL{b}
+%\noTrnoVocRL{`}
+%\noTrnoVocRL{l}
+%\noTrnoVocRL{y}
+%\noTrnoVocRL{s}
+%\noTrnoVocRL{w}
+%in separate form% to ease following the example
+%}
 
@@
 by the absence of corresponding transitions. 
-
 Subfigures (a), (b), and (c) in Figure~\ref{f:example}
@@
 
-We start with $\Phi$ at the root square state in 
-${\cal P}$ which is an accept state. 
-The $\epsilon$-edge connects ${\cal P}$
-to ${\cal S}$ and transitions from any accept state
-in ${\cal P}$ to the root state in ${\cal S}$.
-The $\epsilon$-edge that connects ${\cal S}$ to ${\cal X}$
-follows the same behavior.
+We start with $\Phi$ at the root accept state 
+in ${\cal P}$. 
+%The $\epsilon$-edge connects ${\cal P}$
+%to ${\cal S}$ and transitions from any accept state
+%in ${\cal P}$ to the root state in ${\cal S}$.
+%The $\epsilon$-edge that connects ${\cal S}$ to ${\cal X}$
+%follows the same behavior. 
+The $\epsilon$-edges connect ${\cal P}$
+to ${\cal S}$  and ${\cal S}$ to ${\cal X}$.
+They transition from any accept state
+in the current transducer to the root state of the target transducer.
 
@@
 Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
-An FSA ($\Phi$ or $\Psi$) dies when it reaches a reject state.
+An FSA %($\Phi$ or $\Psi$) 
+dies when it reaches a reject state.
 In our example, if there were no stems that start 
@@
 affixes. 
-For example, \noTrnoVocRL{w} will result in an accept state
+For example, \noTrnoVocRL{w} 
+%will result 
+results 
+in an accept state
 that transitions to ${\cal S}$.
 When \noTrnoVocRL{s} follows, we move to another accept state in 
-${\cal P}$ corresponding to the prefix \noTrnoVocRL{ws}. 
-The same applies to ${\cal X}$.
+${\cal P}$ corresponding to 
+%the 
+prefix \noTrnoVocRL{ws}. 
+%The same applies to ${\cal X}$.
 
@@
 Before moving with the letter \noTrnoVocRL{b} to the accept state,
-$\Phi$ needs to make sure that the stem \noTrnoVocRL{l`b} is compatible
-with the prefix \noTrnoVocRL{wsy}. 
-Sarf keeps compatibility category values as part
+$\Phi$ needs to make sure that %the 
+stem \noTrnoVocRL{l`b} is compatible
+with %the 
+prefix \noTrnoVocRL{wsy}. 
+Sarf keeps compatibility information as part
 of the accept states. 
@@
 more than one concrete state. 
-If the category of \noTrnoVocRL{l`b} is compatible with the category of
-\noTrnoVocRL{wsy} then \noTrnoVocRL{b} will move $\Phi$ to an accept state. 
+If the categories of \noTrnoVocRL{l`b} and \noTrnoVocRL{wsy} are compatible
+then \noTrnoVocRL{b} will move $\Phi$ to an accept state. 
 Otherwise, it will move it to a regular or a reject state.
@@
 
-The application-specific controller can interfere at any point in the 
-analysis to make a decision like ignoring an FSA that 
-is not interesting. 
-It may also decide to correct one FSA
-based on the decision of the others. 
-The controller uses the output of the transducers as input
-to make its decision. 
-It does not produce external output and therefore 
-the three transducers with the 
-controller compose an FSA.
+%The application-specific controller can interfere at any point in the 
+%analysis to make a decision like ignoring a solution that 
+%is not interesting. 
+%It may also decide to correct one solution
+%based on the decision of the others. 
+%The controller uses the output of the transducers as input
+%to make 
+%%its decision. 
+%refinement decisions.
+%%It does not produce external output and therefore 
+%%the three transducers with the 
+%%controller compose an FSA.
 
-\subsection{Recursive affixes}
-\label{sec:recaffix}
-
+%\subsection{Recursive affixes}
+%\label{sec:recaffix}
+{\bf Recursive affixes.~~~}
 We call two morphemes compatible if their concatenation
@@
 Many affixes are composed of other affixes. For example,
-the prefix \RL{wsy} is composed of three other prefixes
-namely \RL{w}, \RL{s}, and \RL{y}.
-The POS tag for \RL{wsy} is a concatenation
-of the POS tags of each of the three morphemes. 
+the prefix \noTrnoVocRL{wsy} is composed of three other prefixes
+namely \noTrnoVocRL{w}, \noTrnoVocRL{s}, and \noTrnoVocRL{y}.
+The POS tag for \noTrnoVocRL{wsy} is a concatenation
+%of the POS tags of each of the three morphemes. 
+of those of the three morphemes. 
 Keeping all valid morphemes in a list
@@
 
-The affix transducers encode $L_{pp}$ and
-$L_{xx}$.
-An accept state in the transducer corresponds to the last letter 
-of an affix in the affix lists.
-The accept states produce as output the POS and other tags
-associated with the state.
+%The affix transducers encode $L_{pp}$ and
+%$L_{xx}$.
+%An accept state in the transducer corresponds to the last letter 
+%of an affix in the affix lists.
+%The accept states produce as output the POS and other tags
+%associated with the state.
+%(check if it is not a loss of information to remove it)
 
@@
 
-We built our stem lexicon using the stem lexicon of 
-Buckwalter. 
-We augmented the lexicon with a set of proper names and
-a set of location names which we 
-obtained from online and biblical sources.
+%We built our stem lexicon using the stem lexicon of 
+%Buckwalter. 
+%We augmented the lexicon with a set of proper names and
+%a set of location names which we 
+%obtained from online and biblical sources.
 
-The stems share lots of substrings. We encoded them in
-an efficient double array trie structure~\cite{Aoe:89}. 
+%The stems share lots of substrings. 
+Since stems share lots of substrings. 
+We encoded them in
+an efficient double array trie (DAT) structure~\cite{Aoe:89}. 
 The structure is deterministic and free of cycles where the 
 accept states are the terminal nodes corresponding to the last 
-letters in stems. Note that the trie implementation may not 
-necessary work on letter-by-letter states, but also compacts
-common substrings.
+letters in stems. 
+%Note that DAT is a suffix trie implementation which may not 
+%necessary work on letter-by-letter states, but also compacts
+%common substrings.
 This approach is superior to Buckwalter since it consists of
@@
 
-The diagram in Figure~\ref{f:composition} shows an 
-abstraction of the non-deterministic construction of Sarf
-from the prefix, stem and suffix transducers. 
-The circle represents regular states, the square
-represents accept states and the triangle represents
-reject states. 
-Note that the concrete prefix, stem and suffix transducers
-are all deterministic and cycle free as discussed 
-is Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}.
-The cycles are introduced in the abstract machine
-for pedagogical purposes.
-
-%Note that Sarf does not rely on implementing its machines as 
-%equivalent regular expressions due to  the following reasons ...
-% (not necessary)
-
-\begin{figure}[tb]
-\center{
-\resizebox{\columnwidth}{!}
-{ \input{figs/abstract_machine.pdftex_t}}
-\caption{Sarf NDFSA.}
-\label{f:composition}
-}
-\end{figure}
+%The diagram in Figure~\ref{f:composition} shows an 
+%abstraction of the non-deterministic construction of Sarf
+%from the prefix, stem and suffix transducers. 
+%The circle represents regular states, the square
+%represents accept states and the triangle represents
+%reject states. 
+%Note that the concrete prefix, stem and suffix transducers
+%are all deterministic and cycle free as discussed 
+%in Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}.
+%The cycles are introduced in the abstract machine
+%for pedagogical purposes.
+%
+%%Note that Sarf does not rely on implementing its machines as 
+%%equivalent regular expressions due to  the following reasons ...
+%% (not necessary)
+%
+%\begin{figure}[tb]
+%\center{
+%\resizebox{\columnwidth}{!}
+%{ \input{figs/abstract_machine.pdftex_t}}
+%\caption{Sarf NDFSA.}
+%\label{f:composition}
+%}
+%\end{figure}
 
@@
 It starts with one machine $\Phi_1$ similar to the one in 
-Figure~\ref{f:composition} and inserts it into a collection
+Figure~\ref{f:example} and inserts it into a collection
 $M$. 
@@
 
-\subsection{Disambiguation using diacritics}
-\label{sec:diacritics}
-
+%\subsection{Disambiguation using diacritics}
+{\bf Disambiguation using diacritics.~~~}
+%\label{sec:diacritics}
 Many Arabic documents, especially religious texts, are fully vocalized.
-Even in documents written without diacritics, we often stumble at a couple of words which are partially vocalized
+Even in documents written without diacritics, we often %stumble at 
+notice 
+a couple of words which are partially vocalized
 to reduce the difficulty of their disambiguation by the reader. Clearly, if native readers
 need the aid of short vowels, automatic morphological analyzers which are context-free should not lose this precious piece of information.
-Sarf is the first in the field to support diacritic-based disambiguation. Consequently, Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
+%Sarf is the first in the field to support diacritic-based disambiguation. Consequently, 
+Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
 or \RL{`ilm} (science) with \RL{`lim} (knew), regardless of the specific letters which are vocalized.
 
-The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
-On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
-to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.
+%The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
+%On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
+%to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.
 
-\subsection{Multiword Expressions}
-\label{sec:MWE}
-
-The names which have been augmented to the lexicon of Sarf contain compound names such as 
-\noVocRL{`bid alkarym}. 
-In current analyzers with the exception of Attia, this expression is analyzed word by word and in best cases is 
-considered to contain two names, instead of being understood to be a single but compound name.
-Sarf enlists such expressions in the stem transducer (with one letter being the white space) and since tokenization
-is performed on the fly, such expressions are detected as one unit. These multi-word stems
-can accept also some compatible affixes such as \RL{w} (and) and \RL{f-} (but).
-This approach works perfectly for proper names but not for ``Syntactically-Flexible Expressions'' as noted
-in ~\cite{MWE}. %MWEs that can undergoe lexical inflections for their constituent words
+%\subsection{Multiword Expressions}
+%\label{sec:MWE}
+{\bf Multiword Expressions.~~~}
+%The names which have been augmented to the lexicon of Sarf contain compound names such as 
+%Some of the names which have been augmented to the lexicon of Sarf are compound such as
+%
+Current analyzers, with the exception of Attia, fail to recognize MWE such as \noTrnoVocRL{`bid alkarym}, and 
+report it as two consecutive names. 
+Sarf includes such expressions in the stem transducer (with one letter being the space) and 
+detects them as one unit since tokenization
+is performed on the fly. The multi-word stems
+can accept some compatible affixes such as \RL{w} (and) and \RL{f-} (but).
+This approach works for proper names but not for ``Syntactically-Flexible Expressions''~\cite{MWE}.
+%MWEs that can undergoe lexical inflections for their constituent words
 
@@
 \centering
-\caption{Comparsion of Running Time}
-\resizebox{0.95\columnwidth}{!}{
+\caption{Results of Sarf, Buckwalter and ElixirFM.}
+\resizebox{0.9\columnwidth}{!}{
 %\begin{tabular}{lp{.01cm}cp{.01cm}cp{.01cm}c}
@@
  & Al Kafi & Al Istibsar & Ibn Hanbal \\ \cline{2-4} 
-Sarf & 2.7002645 & 2.6884225 & 0.2356561 \\
-Buckwalter & 3.315950 & 5.406016 & 0.5472224 \\
-ElixirFM & 2.78$\times60^2$ & 2.3$\times60^2$ & 29.2$\times$60 \\
+Word Count & 98,943 & 103,835 & 20,354 \\
+Sarf (sec) & 2.70 & 2.69 & 0.236 \\
+Sarf/Controller (sec) & 1.32 & 1.31 & 0.0960 \\
+Buckwalter (sec) & 3.32 & 5.41 & 0.547 \\
+ElixirFM (sec) & 167$\times60$ & 138$\times60$ & 29.2$\times$60 \\
 \hline
@@
 
-In this section we compare Sarf to existing 
-morphological analyzers such as Buckwalter and ElixirFM in terms of running time.
-We ran the different analyzers on
-three books of Islamic narrations \RL{.hady_t} ~\cite{IbnHanbal,AlTousi,AlKulayni}
-We report our results in Table~\ref{t:efficeincyresults} on the 
-three hadith books.
+%In this section we compare Sarf to existing 
+%morphological analyzers such as Buckwalter and ElixirFM in terms of running time.
+We ran Sarf without a controller then with a controller that only 
+focused on proper names and words that are related to the word ``narrated".
+We compared the running times to those of Buckwalter and ElixirFM.
+We used three books of Islamic narrations \RL{.hady_t} ~\cite{IbnHanbal,AlTousi,AlKulayni} as our data set.
+We report our results in Table~\ref{t:efficeincyresults}.
 
-Sarf outperforms both Buckwalter and ElixirFM in terms of 
-running time. 
+With a controller, Sarf performed much better since the controller pruned several solutions irrelevant to the application. 
+Sarf, with and without a controller, was more efficient than both Buckwalter and ElixirFM. 
 Buckwalter performed much better than ElixirFM in 
-terms of running time, 
-We think the big difference is due to the sophistication of 
-the formal functional morphological analysis of ElixirFM 
-and its reliance on additional annotations extracted from 
-both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.
+terms of running time. 
+The difference may be due to the sophisticated formal 
+functional morphological analysis of ElixirFM. 
+%and its reliance on additional annotations extracted from 
+%both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.
 
@@
 
-In a future paper, we will present our work of using Sarf to 
-parse Islamic Narrations from Prophet Muhammad commonly known commonly as hadith.
-We are also working on using Sarf to automate the problems of
-arabic email classification and
-arabic content-based parental control.
-Those use cases will benefit from the application-specific nature of Sarf.
+In the future, we will use Sarf to fully automate the analysis of Islamic Narrations 
+from Prophet Muhammad commonly known as hadith.
+We will also use Sarf to automate the classification of
+Arabic emails. Another application where we are looking to use Sarf is
+Arabic web content analysis to flag violent and inappropriate material for parental control.
+Those applications will benefit from the application-specific nature of Sarf.
 

### 2011-02-26 trunk/tex/sarf.aclhlt.tex
changed title+ swapped 2 paragraphs

@@
 %\title{Application-Specific Arabic Morphological Analyzer for Arabic Text Mining}% maybe add "with Enhanced Capabilities"
-\title{Application-Specific Enhanced Morphological Analyzer \\for Arabic Text Mining}% maybe add "with Enhanced Capabilities"
+\title{Application-Specific Morphological Analyzer for Arabic Text Mining}% maybe add "with Enhanced Capabilities"
 
@@
 a branch of morphological analysis may not be
-appropriate for the application at hand. 
-In this paper we present Sarf, 
+appropriate for the NLP application. 
+We present Sarf, 
 %an application-specific morphological analyzer for Arabic. 
-an Arabic morphological analyzer, which
+an Arabic application-specific morphological analyzer, which
 %Sarf
-allows an application-specific controller to control and refine
-the state of the analysis on the fly. 
+allows a controller to refine
+the analysis on the fly. 
 It uses non-deterministic finite state Automata % (NDFSA)
@@
 
-%Automated analysis of Arabic data sets, including texts, 
-%publications, records and digital media is essential
-%with the huge digital Arabic content available nowadays. 
 Morphological analysis is key in current automated 
 analysis techniques for Arabic text. 
-%Current morphological analyzers~\cite{Sughaiyer:04}
-%use concatenative analysis when
-%considering the internal structure 
-%of an Arabic word and
-%composing it into several {\em morphemes}. 
-%A morpheme can be a {\em stem}, or an {\em affix}.
-%An affix can be a {\em prefix, suffix, } or an {\em infix}.
-%The analysis of one word may lead to several possible
-%morphological solutions.
-%\vocalize
-%For instance, the word \RL{'a.hmadH}~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
-%may have two valid morphological analyses. 
-%The letter \RL{'a} may be a prefix and the word means 
-%``I praise him'', or 
-%%The letter \RL{'a} may also be 
-%part of the stem \RL{'a.hmad} (a proper noun)
-%and the word means ``his Ahmad''.
-%
-%\novocalize
 The accuracy of the solutions suffers due to inherent difficulties
-of morphological analysis of the Arabic language. 
+of morphological analysis of Arabic. 
 For example, it is common practice to write Arabic text
 without short vowels (diacritics) 
-which greatly increases its ambiguity.
-%This greatly increases the ambiguity of %Arabic text.
-%the text.
-%Moreover, Arabic letters can have up to 
-%four different forms
-%corresponding to their position in a word, i.e, beginning,
-%middle, end of word and separate forms. 
-%This allows the phrase
-%\noTrnoVocRL{il_A\nospace almdrsT}
-%to be visually recognizable
-Moreover, determining word boundaries is not always 
-%a trivial problem.
+which greatly increases ambiguity.
+Moreover, tokenization is not always 
 trivial.
@@
 without the need of a space in between. 
-%The reason is the first word %\noVocRL{il_A} 
-%ends with
-%\noTrnoVocRL{_A} a non-connecting letter. 
 The reason is that
 \noTrnoVocRL{_A} is a non-connecting letter. 
-%These words, referred to as 
 Such 
 ``run-on'' words~\cite{Buckwalter:04},
-%occur often, and greatly increase the
-occur often, %greatly
-increasing the
-difficulty of tokenization.
+occur often.
 
 %{\bf Sarf.~~~}
-In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
-has been submitted with the paper and will be available 
-%online
-as an open source tool.}, 
-a {\em novel application-specific efficient 
-%(maybe 'enhanced' instead of 'efficient' to engulf both efficiency and other refinements)
-morphological analyzer} that uses 
-non-deterministic finite state Automata (NDFSA)
-driven by an application-specific controller.
-Each machine in Sarf takes one letter at a time as input
-and represents a valid morphological analysis.
-%The application-specific controller prunes false positives
-The controller prunes false positives
-early in the run by making a decision at every input letter
-to control the machines.
-%The controller we present in Section~\ref{sec:controller}
-%is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
-%it can be a logical entity of any type.
-The controller can be a Finite State Automaton (FSA) or any other %logical entity 
-programming entity, % of any type,
-%since Automatas may be a %very complex
-%non-necessary solution for some simple applications.
-%(maybe re-word this sentence)
-depending on application needs.
-
 %(check if position of this paragraph seems in correct place)
@@
 %application needs, which is an issue that Sarf tackles. % (maybe reworded)
-We have noted that in many NLP applications, when the morphological analyzer
-benefits from the peculiarities and characteristics of the query at hand, 
-the analysis %can be easily 
-is greatly 
-enhanced. 
-Consequently, we strongly believe that by 
+We hypothesize that in many NLP applications, the morphological analyzer may
+benefit from the characteristics of the query at hand, 
+to greatly enhance the analysis.
+Consequently, %we believe that by 
 %being 
-allowing
+%allowing
 an application-specific %morphological 
@@
 %the
-controller to intervene at every decision to guide, use, order, and 
-refine the morphological analysis, Sarf is of high utility.
+controller that intervenes at every decision to guide, use, order, and 
+refine the morphological analysis is of high utility.
+
+In this paper, we present Sarf~\footnote{Sarf (name modified for blind review) 
+has been submitted with the paper and will be available 
+as an open source tool.}, 
+a {\em novel application-specific efficient 
+%(maybe 'enhanced' instead of 'efficient' to engulf both efficiency and other refinements)
+morphological analyzer} that uses 
+non-deterministic finite state Automata (NDFSA)
+driven by an application-specific controller.
+Each machine in Sarf takes one letter at a time as input
+and represents a valid morphological analysis.
+%The application-specific controller prunes false positives
+The controller prunes false positives
+early in the run by making a decision at every input letter
+to control the machines.
+%The controller we present in Section~\ref{sec:controller}
+%is an FSA. However, as we explain in Section~\ref{sec:ndfsa},
+%it can be a logical entity of any type.
+%since Automatas may be a %very complex
+%non-necessary solution for some simple applications.
+%(maybe re-word this sentence)
+%depending on application needs.
 
@@
 %from a text stream.
-Apart from being application-specific, 
+%Apart from being application-specific, 
 Sarf %has
-%also  
-introduced 
-many enhancements. %to the field of Arabic morphological analysis. 
+also  
+introduces 
+several enhancements. %to the field of Arabic morphological analysis. 
 %the following enhancements:
@@
 First,
-Sarf does not assume that the word at hand is a token and
+Sarf %does not assume that the word at hand is a token and
 performs tokenization on the fly based on morphological correctness.
@@
 %the problem of ``run-on'' words, but also of handling Multiword Expressions (MWE). 
-Moreover,
-% since Sarf does not perform white-space delimited tokenization, 
-Sarf 
-%is capable of handling
-handles 
-Multiword Expressions (MWEs),
+On the fly tokenization also handles
+Multiword Expressions (MWEs)~\cite{MWE} such as \noVocRL{`bid alkarym}.
 %MWEs are defined in~\cite{MWE} as
-which are 
-``idiosyncratic interpretations
-that cross word boundaries (or spaces)''~\cite{MWE}. 
+%which are 
+%``idiosyncratic interpretations
+%that cross word boundaries (or spaces)''. 
 %By this term, we refer to frequent cases when a group of words form together a single
@@
 %where an affix can be a sequence of other affixes which are concatenatively compatible.
-where an affix can be a sequence of other concatenatively compatible affixes.
+where an affix can be a sequence of other compatible affixes.
 %An atomic affix $a$ is a standalone affix
@@
 efficiency of affix matching. 
-
 Third, Sarf supports disambiguation using 
 %diacritics when they are available.
-any available diacritics.
+available diacritics, instead of the traditional approach that ignores them.
 %This feature, to the best of our knowledge is the first attempt to benefit from 
 %To the best of our knowledge,
-Up to our knowledge, this is the first attempt to benefit from
-diacritics instead of ignoring them. 
-Prior to Sarf, the trend was to preprocess 
-the text to remove  
-these %any such
-`unwelcomed' letters. 
+%Up to our knowledge, this is the first attempt to benefit from
+%diacritics instead of ignoring them. 
 %Nonetheless, Sarf regards them as a richness to be utilized that should not be disregarded.
@@
 %as shown in Section~\ref{sec:results}.
-Sarf is also more efficent than current morphological analyzers, 
+Sarf is also more efficient than current morphological analyzers, 
 as shown in Section~\ref{sec:results}.
@@
 or limit themselves to a specific set of POS tags. 
-Sarf differs in that it is a {\em general} morphological 
+Sarf differs in that it is a general morphological 
 analyzer that reports all possible solutions. 
-It is {\em application-specific} in the sense that a controller 
+It is application-specific in the sense that a controller 
 prunes these solutions.
 
-Current morphological analyzers such as %(maybe disambiguators also added here according to reviewer comments)
 Buckwalter~\shortcite{Buckwalter:02},
@@
 and ElixirFM~\cite{Otakar:07} 
-take as input white-space delimited tokens~\cite{Kulick:10},
-consider them as words,
-and enumerate all possible solutions. 
-This approach has several problems. 
+consider white-space delimited tokens~\cite{Kulick:10}
+as words.
+%This suffers has several problems. 
 %However, we note the following: %(I thought of this change not to include point 2 as a "problem" according to reviewer comments)
-First, a white-space delimited token may have 
-more than one word or just part of a MWE.
+This fails when a white-space delimited token has
+more than one word or is part of an MWE.
 %Second, the exhaustive enumeration may hurt performance and may
@@
 %in some case studies as noted in~\cite{Maamouri:10}. 
-Second, in many applications, exhaustive enumeration may not be needed or appropriate
-~\cite{Maamouri:10}; consequently, adding unnecessary running time.
+These analyzers also exhaustively enumerate all 
+possible morphological solutions 
+which may not be appropriate in many applications
+~\cite{Maamouri:10}. 
+%consequently, adding unnecessary running time.
 
-Other morphological analyzers such as 
-Amira~\cite{Diab:07,Benajiba:07},
+Others such as 
+AMIRA~\cite{Diab:07,Benajiba:07},
 MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
-% (reviewer wants to distinguish MADA as a tool for morphological disambiguation instead of analysis)
 use machine learning and support vector machines (SVM) 
-to enhance the accuracy of the morphological analysis at the expense 
+to disambiguate the morphological analysis at the expense 
 of efficiency.
 Xerox rules can be compiled into specialized 
-%finite state machine (FSM) based analyzers as described in~\cite{Beesley:03}.
-FSA-based analyzers~\cite{Beesley:03}.
-However, the efficiency of the resulting analyzer depends on the
-way the Xerox rules are written. 
+finite state Automata (FSA)~\cite{Beesley:03}.
+%FSA-based analyzers~\cite{Beesley:03}.
+The efficiency of the resulting analyzer depends on the
+way the rules are written. 
 This requires deep knowledge and insight from the user
@@
 %into efficient non-deterministic finite state Automata (NDFSA). 
-Sarf, however, allows users to write directly the code for the needed controller. %(maybe not clear)
+Beesley reports that the number of Automata generated by a compiler
+for Xerox rules can not be controlled and also reports a difficulty 
+to compose the FSAs into a single framework. 
+%Beesley requires recompilation of the Xerox rules. 
+% (I am not sure if it is ok to have the above removed)
+Sarf however constructs an NDFSA from 
+%the controller and 
+the prefix, stem, and suffix deterministic 
+transducers to solve the problem.
+Sarf also differs in that 
+%it uses recursive affixes in the affix transducers, and in that 
+the transducers consider
+input from the controller to modify their behavior.
+The user can express the controller,not necessarily an FSA, in a programming language such as C++.
 
@@
 %affix 
-compatibility rules to allow 
-affixes to be defined recursively
-so that we can have prefix-prefix and suffix-suffix 
+compatibility rules to define 
+affixes recursively and allow 
+prefix-prefix and suffix-suffix 
 concatenations.
 %This allows Sarf to better maintain and propagate 
-Hence, it better maintains and propagates 
-the POS tags associated with affixes. 
+This better maintains and propagates 
+the tags associated with affixes. 
 %This information is highly 
@@
 %generated as a product of the token solutions. 
-
-Similar to Sarf, both Beesley %~\shortcite{Beesley:01} (already cited previously)
-and Attia~\shortcite{Attia:05} use
-%finite state Automata (FSA). 
-FSA.
-Beesley reports that the number of Automata generated by a compiler
-for Xerox rules can not be controlled and also reports a difficulty 
-to compose the FSAs into a single framework. 
-Sarf however constructs an NDFSA from 
-%the controller and 
-the prefix, stem, and suffix deterministic 
-transducers to solve the problem.
-Sarf also differs in that 
-%it uses recursive affixes in the affix transducers, and in that 
-the transducers consider
-input from the controller to modify their behavior. 
-%With Sarf it is easy to control the FSA with a application-specific
-%controller. 
-%Beesley requires recompilation of the Xerox rules. 
-% (I am not sure if it is ok to have the above removed)
-Similar to Sarf, Attia handles MWEs, but with a more extensive coverage
+Similar to Sarf, Attia uses FSAs and handles more and qualitatively different MWEs
 ~\cite{Attia:06,Attia:10}.
@@
 
-Similar to The MADA+TOKAN toolkit, Sarf thinks of
-the several valid morphological analyses as a richness that 
-should be exploited at a higher abstract level rather than
-an inaccuracy that should be corrected. 
-Sarf differs in that it provides an interface for the 
+%Similar to %the 
+%MADA+TOKAN, Sarf uses
+%the several valid morphological analyses as 
+%%a richness that should 
+%richness at a higher abstract level, and does not view them as 
+%inaccuracies to be corrected. 
+Sarf differs from MADA+TOKAN in that it provides an interface for the 
 application to interfere in judging the analyses much earlier and
 at a finer granularity level. 
-In Sarf, there is no need for a separate tokenizer such as
+%In Sarf, there is no need for a separate tokenizer such as
+Sarf does not need a separate tokenizer such as
 TOKAN since each solution keeps a stack of positions
@@
 when parsing the
-string \noTrnoVocRL{wsyl`b-h-aa\nospace al--lA`bwn}.
+string \noTrnoVocRL{wsyl`b-h-aa\nospace al--lA`bwn} \noArRL{wsyl`b-h-aa al--lA`bwn}.
 %\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
@@
 
-The diagram in Figure~\ref{f:composition} shows an 
-abstraction of the non-deterministic construction of Sarf
-from the prefix, stem and suffix transducers. 
-The circle represents regular states, the square
-represents accept states and the triangle represents
-reject states. 
-Note that the concrete prefix, stem and suffix transducers
-are all deterministic and cycle free as discussed 
-in Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}.
-The cycles are introduced in the abstract machine
-for pedagogical purposes.
-
-%Note that Sarf does not rely on implementing its machines as 
-%equivalent regular expressions due to  the following reasons ...
-% (not necessary)
-
-\begin{figure}[tb]
-\center{
-\resizebox{\columnwidth}{!}
-{ \input{figs/abstract_machine.pdftex_t}}
-\caption{Sarf NDFSA.}
-\label{f:composition}
-}
-\end{figure}
+%The diagram in Figure~\ref{f:composition} shows an 
+%abstraction of the non-deterministic construction of Sarf
+%from the prefix, stem and suffix transducers. 
+%The circle represents regular states, the square
+%represents accept states and the triangle represents
+%reject states. 
+%Note that the concrete prefix, stem and suffix transducers
+%are all deterministic and cycle free as discussed 
+%in Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}.
+%The cycles are introduced in the abstract machine
+%for pedagogical purposes.
+%
+%%Note that Sarf does not rely on implementing its machines as 
+%%equivalent regular expressions due to  the following reasons ...
+%% (not necessary)
+%
+%\begin{figure}[tb]
+%\center{
+%\resizebox{\columnwidth}{!}
+%{ \input{figs/abstract_machine.pdftex_t}}
+%\caption{Sarf NDFSA.}
+%\label{f:composition}
+%}
+%\end{figure}
 
@@
 
-\subsection{Disambiguation using diacritics}
-\label{sec:diacritics}
-
+%\subsection{Disambiguation using diacritics}
+{\bf Disambiguation using diacritics.~~~}
+%\label{sec:diacritics}
 Many Arabic documents, especially religious texts, are fully vocalized.
-Even in documents written without diacritics, we often stumble at a couple of words which are partially vocalized
+Even in documents written without diacritics, we often %stumble at 
+notice 
+a couple of words which are partially vocalized
 to reduce the difficulty of their disambiguation by the reader. Clearly, if native readers
 need the aid of short vowels, automatic morphological analyzers which are context-free should not lose this precious piece of information.
-Sarf is the first in the field to support diacritic-based disambiguation. Consequently, Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
+%Sarf is the first in the field to support diacritic-based disambiguation. Consequently, 
+Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
 or \RL{`ilm} (science) with \RL{`lim} (knew), regardless of the specific letters which are vocalized.
 
-The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
-On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
-to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.
+%The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
+%On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
+%to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.
 
-\subsection{Multiword Expressions}
-\label{sec:MWE}
-
-The names which have been augmented to the lexicon of Sarf contain compound names such as 
-\noVocRL{`bid alkarym}. 
-In current analyzers with the exception of Attia, this expression is analyzed word by word and in best cases is 
-considered to contain two names, instead of being understood to be a single but compound name.
-Sarf enlists such expressions in the stem transducer (with one letter being the white space) and since tokenization
-is performed on the fly, such expressions are detected as one unit. These multi-word stems
-can accept also some compatible affixes such as \RL{w} (and) and \RL{f-} (but).
-This approach works perfectly for proper names but not for ``Syntactically-Flexible Expressions'' as noted
-in ~\cite{MWE}. %MWEs that can undergoe lexical inflections for their constituent words
+%\subsection{Multiword Expressions}
+%\label{sec:MWE}
+{\bf Multiword Expressions.~~~}
+%The names which have been augmented to the lexicon of Sarf contain compound names such as 
+%Some of the names which have been augmented to the lexicon of Sarf are compound such as
+%
+Current analyzers, with the exception of Attia, fail to recognize MWE such as \noTrnoVocRL{`bid alkarym}, and 
+report it as two consecutive names. 
+Sarf includes such expressions in the stem transducer (with one letter being the space) and 
+detects them as one unit since tokenization
+is performed on the fly. The multiword stems
+can accept some compatible affixes such as \RL{w} (and) and \RL{f-} (but).
+This approach works for proper names but not for ``Syntactically-Flexible Expressions''~\cite{MWE}.
+%MWEs that can undergoe lexical inflections for their constituent words
 
@@
  & Al Kafi & Al Istibsar & Ibn Hanbal \\ \cline{2-4} 
-Sarf & 2.7002645 & 2.6884225 & 0.2356561 \\
-Buckwalter & 3.315950 & 5.406016 & 0.5472224 \\
-ElixirFM & 166.8$\times60$ & 138$\times60$ & 29.2$\times$60 \\
+Word Count & 98,943 & 103,835 & 20,354 \\
+Sarf (sec) & 2.70 & 2.69 & 0.236 \\
+Sarf/Controller (sec) & 1.32 & 1.31 & 0.0960 \\
+Buckwalter (sec) & 3.32 & 5.41 & 0.547 \\
+ElixirFM (sec) & 167$\times60$ & 138$\times60$ & 29.2$\times$60 \\
 \hline

### 2011-02-26 trunk/ATSarf/src/gui/mainwindow.cpp

@@ -14,6 +14,7 @@ MainWindow::MainWindow(QWidget *parent) :
 {
 	m_ui->setupUi(this);
 	m_ui->pushButton->setVisible(false);
+	m_ui->chk_testing->setVisible(false);
 }
 
 MainWindow::~MainWindow()
@@ -72,6 +73,7 @@ void MainWindow::on_pushButton_clicked()
 			try{
 				system("dot -Tsvg graph.dot -o graph.svg");
 				QMainWindow * mw =new QMainWindow(NULL);
+				mw->setWindowTitle("Sarf Graph");
 				QScrollArea * sa=new QScrollArea(mw);
 				mw->setCentralWidget(sa);
 				QLabel *pic=new QLabel(sa);
@@ -97,7 +99,7 @@ void MainWindow::on_fill_clicked()
 void MainWindow::on_cmd_browse_clicked()
 {
 	QString fileName = QFileDialog::getOpenFileName(this,
-		tr("Open File"), "/home/jad/Desktop/linux", tr("All Files (*)"));
+		tr("Open File"), "", tr("All Files (*)")); // /home/jad/Desktop/linux
 	if (!fileName.isEmpty())
 		m_ui->input->setText(fileName);
 }

### 2011-04-01 trunk/ATSarf/src/common/common.h
some structural code refactorment to sate machine of Hadith,
still have many false possitives due to runon words

@@ -89,7 +89,7 @@ public:
 typedef struct text_info_
 {
 	QString *text;
-	long long start,finish;
+	long start,finish;
 }text_info;
 
 class multiply_params

### 2011-04-01 ATSarf/src/case/hadith_general.cpp
some structural code refactorment to sate machine of Hadith,
still have many false possitives due to runon words

@@ -35,7 +35,6 @@ QString PhrasesFileName=".phrases";
 QString StopwordsFileName=".stop_words";
 #endif
 
-
 #ifndef REFINEMENTS
 	enum wordType { NAME, NRC,NMC};
 	enum stateType { TEXT_S , NAME_S, NMC_S , NRC_S};
@@ -44,6 +43,22 @@ QString StopwordsFileName=".stop_words";
 	enum stateType { TEXT_S , NAME_S, NMC_S , NRC_S, STOP_WORD_S};
 	QStringList rasoul_words;
 #endif
+	typedef struct  {
+		long startPos;
+		long endPos;
+		long nextPos;
+		wordType currentType:3;
+		stateType currentState:4;
+		stateType nextState:4;
+		bool ibn:1;
+		bool _3abid:1;
+		bool possessivePlace:1;
+		bool followedByPunctuation:1;
+		int unused:21;
+		void resetCurrentWordInfo()	{ibn=false;_3abid=false;possessivePlace=false;followedByPunctuation=false;}
+		bool ibnOr3abid() { return ibn || _3abid;}
+		bool isIbnOrPossessivePlace(){return ibn || possessivePlace;}
+	} StateInfo;
 
 	QStringList compound_words;
 	QString hadath,abid,alrasoul,abihi;
@@ -59,7 +74,7 @@ QString StopwordsFileName=".stop_words";
 	QTextStream * myoutPtr;
 #endif
 #ifdef COUNT_AVERAGE_SOLUTIONS
-	long long total_solutions=0;
+	long total_solutions=0;
 	long stemmings=0;
 #endif
 #ifdef HADITHDEBUG
@@ -133,8 +148,8 @@ typedef struct chainData_ {
 
 } chainData;
 typedef struct stateData_ {
-	long long  sanadStartIndex, narratorCount,narratorStartIndex,narratorEndIndex,nrcStartIndex,nrcEndIndex;
-	long long  nmcCount, nrcCount,nameStartIndex,nmcStartIndex;
+	long  sanadStartIndex, narratorCount,narratorStartIndex,narratorEndIndex,nrcStartIndex,nrcEndIndex;
+	long  nmcCount, nrcCount,nameStartIndex,nmcStartIndex;
 	bool nmcValid;
 	bool ibn_or_3abid;
 	bool nrcPunctuation;
@@ -321,8 +336,8 @@ private:
 	bool finished;
 #endif
 public:
[…]

### 2011-04-19 trunk/tex/tp.tex
created tp

\documentclass[a4,12pt]{report}
\usepackage[Bjornstrup]{fncychap}
\usepackage{graphicx}
\usepackage{epsf}
\usepackage{a4}
\usepackage{fancyhdr}
\usepackage{setspace}
\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{enumerate}

\usepackage{caption}
\usepackage{epsfig}
\usepackage{subfigure}
\usepackage{colortbl}
\usepackage{color}
\usepackage{rotate}
\usepackage{rotating}
\usepackage{amsthm}
\usepackage{booktabs}
\usepackage{slashbox}

\usepackage{relsize}
\usepackage{fancyvrb}
\usepackage[colorlinks=false]{hyperref}

\usepackage{url}
\usepackage{arabtex}

\usepackage{utf8}
\setarab
\fullvocalize
\transtrue
\arabtrue

%\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}
\newcommand{\noTrnoVocRL}[1]{\transfalse\novocalize\noTrRL{#1}\vocalize\transtrue}  
\newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}

\newcommand{\X}{\cellcolor[gray]{0.5}}
\newcommand{\x}{\cellcolor[gray]{0.8}}

%Header and Footer
\fancyhead{} \fancyhead[LO,RE]{\scriptsize\sc\leftmark}
%\fancyhead[R]{Jad Makhlouta}
\fancyfoot[l]{ \scriptsize \textit{AUB, Spring 2010-2011}}
\fancyfoot[R]{ \scriptsize \textit{ Thesis Proposal Report}}

%-------BEGIN----------------------------------------------------
\begin{document}

%----------------Report Cover--------------------------------------------------
%\include{parts/cover}
\begin{titlepage}
%\vspace{1cm}
\begin{center}

% \vspace{1cm}

\large{Department of Electrical and Computer Engineering\\
Faculty of Engineering and Architecture \\
American University of Beirut\\}
\vspace{3cm}
%\begin{figure}[h]
%    \begin{center}
%       \includegraphics[scale=0.25]{Fig/aublogo.eps}
%    \end{center}
%\end{figure}
 \LARGE \textbf{Application-Specific Arabic Morphological Analyzer and \\Automatic Hadith Authentication\\}
\vspace{1.5cm}
\LARGE {Thesis Proposal \\ Spring 2010-2011}\\
\end{center}
\vspace{3.5cm}
\begin{tabbing}

Submitted by: \quad\quad \quad \quad \=Jad Makhlouta, 200700856 \\
\\
Thesis Supervisor:  \> Dr. Fadi Zaraket \\

Committee Member: \> Dr. Hasan Artail  \\

Committee Member:  \> Dr. Ibrahim Abou Faycal \\

\\
\\
\\
Submitted on: \> \today
\end{tabbing}
\end{titlepage}

% Select arabic numbering style
\pagestyle{fancy}
 \pagenumbering{arabic}
\setcounter{page}{1}

\begin{abstract}
Text Mining 
%can be considered as a subfield of both Natural Language Processing (NLP) and Data Mining.
is interested in automatically extracting useful information and relations from text documents.
%Extracted information is hard to discover manually or previously unknown
%Natural Language Processing applications perform 
%to discover useful information that is otherwise unknown.
The complexity of natural languages introduces interesting challenges 
to automatic analysis of underlying textual data.
Morphological analysis is key to many Natural Language Processing Applications.
General morphological analyzers such as Buckwalter and ElixirFM 
generate all morphological solutions.
However, this exhaustive enumeration may not be needed or appropriate for the
application at hand.
We propose developing Sarf, an Arabic application-specific morphological analyzer whose functionality can be customized 
based on the specific needs of the application. 
%This flexibility has been motivated by 
%the broad spectrum of text mining applications, each with its specific characteristics and corpora, which
%necessitate a special treatment.
Sarf encodes its lexicon into a non-deterministic composition of three finite automata.
It also solves the issue of `run-on' words, supports disambiguation using diacritics, 
handles multi-word expressions and performs online tokenization.
In our thesis, we will use Sarf to solve two useful Natural Language Processing applications: 
(1) the extraction of time entities from Arabic documents, and 
(2) the automation of the process of hadith authentication.

\end{abstract}

\chapter{Introduction}
\label{s:intro}

The world we live in overwhelms us with electronic data. There is an ever-increasing gap between the amount
of information available and our ability to understand and analyze it. ``Omnipresent computers make
it too easy to save things that previously we would have trashed. Inexpensive disks
and online storage make it too easy to postpone decisions about what to do with all
this stuff -- we simply get more memory and keep it all.''~\cite{Witten:11}
This explosion in available data has motivated the automation of the task 
of extracting information of interest for a certain usage. 
Manual extraction is practically unfeasible due to
our limited precious time recourses compared to the magnitude of the task.
In addition, it was shown that automatic methods have been able to discover previously unknown 
information through relating separate datasets~\cite{JNi06}.

%Consequently, \textit{Data Mining} as a field of study emerged
%as an attempt to tackle this problem which is manually unfeasible.
%Data Mining in general is defined as the task of `extraction of interesting 
%(non-trivial, implicit, previously unknown and potentially useful) patterns 
%or knowledge from huge amounts of data'~\cite{Witten:00}.

Clearly, a small percentage of the data available worldwide 
is stored in databases in structured form; instead, the greater amount of information is available in the 
form of electronic text files and documents such as news articles, publications, governmental records, etc. ...
Even old paper documents are being ported to digital form to facilitate their use. 
Text Mining emerged as an attempt to automate the extraction of interesting, 
previously unknown and potentially useful information
from the huge amounts of text documents~\cite{Witten:08}.
Information Extraction (IE) from natural languages, however, is challenged
by the difficulty of getting structured 
entities from the underlying unstructured textual data.

%Information Extraction (IE) from Natural Language Texts has always been an important 
%problem in Data mining. What distinguishes text mining from the 
%other subfields of data mining is mainly due to the challenge of getting structured 
%entities from the underlying unstructured textual data. Clearly, a small percentage of the data available worldwide 
%is stored in databases in structured form; instead the greater amount of information is available in the 
%form of electronic text files and documents such as news articles, publications, governmental records, etc. ...
%Even old paper documents are being ported to digital form to facilitate their use.

Natural Language Processing (NLP)
as a field of research is interested in achieving a computational model of the language
and understanding its underlying meaning~\cite{sharp:01}. It is known that this task
is very complex and possibly not completely achievable. However, in most applications it 
is not needed to grasp the whole meaning of the text but to extract some useful entities.
This Information Extraction(IE) problem can be performed through some statistical approaches 
and/or partial NLP techniques.

A subproblem of NLP is morphology.
Morphological analyzers%~\cite{Sughaiyer:04}
are interested in understanding the internal structure of a word;
It considers words as being composed of 
several {\em morphemes}. 
A morpheme is a {\em stem} or an {\em affix}.
An affix is a {\em prefix, suffix,} or an {\em infix}.
Prefixes and suffixes are attached to the beginning and end of the word respectively; 
while infixes introduce changes in the stem itself. 
Morphological analyzers have been extensively used as building blocks 
in many text mining applications~\cite{Sou07}.  In Arabic, many morphological analyzers exist including:  
Buckwalter~\cite{Buckwalter:02},
Beesley~\cite{Beesley:01},
SAMA ~\cite{Kulick:10},
ElixirFM~\cite{Otakar:07}, 
AMIRA~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, 
and MADA+TOKAN~\cite{Habash:09}.
%Arabic however has distinctive characteristics that
%complicate its analysis; and consequently, much remains to be performed in the field.

In this thesis, we propose to develop Sarf, an application-specific 
morphological analyzer for the Arabic language 
which is capable of benefiting
from the specific characteristics of the application at hand. 
In addition, it supports disambiguation using diacritics and 
handles `run-on' words. We also propose to use Sarf to solve two interesting .
text mining applications: time entity extraction and hadith authentication.

\section{Applications of Text Mining}
\label{sec:Applications}

%Text Mining has gained much interest lately especially due to its wide range of applications.
%However, these applications have substantially differing corpora (terminology, 
%sentence structure, etc...). In this section, we present some of the 
%previously explored applications of text mining.
%
%Research in text mining started in the mid 1980s when Swanson 
%realized that slicing and combining seemingly unrelated medical 
%articles led to the discovery of new 
%hypotheses~\cite{JNi06}. 
%Lately we saw applications of text mining in the advertisement, 
%political campaigning, and other businesses.

Research in text mining started in the mid 1980s when Swanson 
realized that slicing and combining seemingly unrelated medical 
articles led to the discovery of new 
hypotheses~\cite{JNi06}. 
Text Mining has gained much interest lately due to its wide range of applications,
including its use in the areas of advertisement, 
political campaigning, and other businesses.
However, these applications have substantially differing corpora (terminology, 
sentence structure, etc...). In this section, we present some of the 
previously explored applications of text mining.

\subsection{Sentiment Analysis} 
	This field -- also known as Opinion Mining' -- attempts classifying the polarity of sentiments of people 
	or even their emotional state ~\cite{Pang:08}. The inputs for such analyzers include product and movie reviews, 
	Internet blogs, electronic social networks and possibly newspapers. Early research was conducted on review sites
	to automate the task classifying whether the review was positive and negative. Consequently, users can get concise helpful
	sentiment summaries about the product, movie, etc...~\cite{Pang:02, Turney:02}. Such abilities motivated research targeting 
	the following applications:
	\subsubsection{Marketing Applications} 
		The strength of blogs in disseminating opinions is a two-edged sword that can 
		``make or break a company's reputation''~\cite{hoffman:08a}. Consequently, analyzing customers' opinions as portrayed 
		on online discussion forums should be detected as soon as possible. This input is vital in the company's policy for its
		reputation management. It can accordingly take actions to retain current customers, attract new ones and win back former 
		customers~\cite{hoffman:08b}. Sentiment Mining has also been used to predict customer churns~\cite{marketing:07,marketing:08}
		to take appropriate actions. It can help also in producing more effective advertisements based on the interests of the market.
	\subsubsection{Political Applications} 
		Political experts and analysts can benefit also. Since opinions extracted from textual data 
		can reveal much even without the need of performing any explicit surveys. According to ~\cite{Gryc:10}, it ``opens up many avenues
		for more accurate and naturalistic large-scale political analysis.'' Some research has been performed on classifying and predicting
		political inclinations~\cite{Durant:07, Yu:08}. This field can help political leaders to better understand the views of their people 
		and possibly perform better campaigns.

\subsection{Academic Applications} 
	In scientific disciplines where written papers and publications contain technical information, the need to retrieve
	specific information that a user needs, aids in \textbf{\em developing advanced search engines}. 
	Text mining plays a great role toward this end through the following:
	\subsubsection{Text Summarization} 
		When an application is interested in mining a text for the most important and informative portions,
		the application is considered to be tackling the field of `Text Summarization'. Summaries may have different forms such as being 
		a group of representative words, to a coherent abstraction of the text in novel terms and phrases that capture its main meaning. 
		The main motivation
		for such a research is to enhance search engines. By capturing the most informative concepts in a document, it aids determining a 
		better metric for a document's relevance to a search query. It also helps search engines display a shortened preview of the document for returned
		results. This reduces the overwhelming amount of information a user needs to read before deciding weather the whole document deserves
		comprehensive study. It is also helpful when the editorial staff of a magazine or newspaper need to present an article in the limited space.
		Similarly, summarization is vital when the content of 
		news articles needs to be dramatically shortened to be sent in an SMS form to subscribed PDA users. 
		Useful surveys of current research on text Summarization can be
		found in ~\cite{Radev:02,Das:07}.
	\subsubsection{Text Categorization and Clustering} 
		In many cases, it is very useful to be able to search for documents by theme, topic, domain, etc...
		This ability is achieved if documents can be automatically clustered into different categories over which a user can search. It has been also 
		proposed to cluster documents based on biographical data~\cite{Kostoff:99}. Technically speaking the difference between Categorization and
		Clustering is that the former organizes documents into predefined categories where in the latter the categories arise from the document content
		themselves~\cite{sharp:01}. Many approaches exist in this field as described in~\cite{Berry:03, Aas:99, MUNTEANU:05}.
		\begin{enumerate}
		\item \textbf{Email Classification:} It is noteworthy to mention that it is text categorization can be used for emails 
			-- such as urgent, work, friends, spam, etc... Consequently, it can be used for spam detection and email 
			filtering~\cite{Mock:01,bekkerman:04}.
		\item \textbf{Parental Control:} It has also been proposed to use categorization for parental control by
			automatically detecting inappropriate content for their children~\cite{Kontostathis:09}.
		\end{enumerate}
	\subsubsection{Entity Extraction:} Named Entity Extraction/Recognition (NER) targets being able to detect phrases that convey some general concept
		and/or belong to a predefined set of categories such as persons, places, time, etc... It has been argued that having the luxury of searching
		for entities enhances search abilities instead of being restricted to a list of words appearing in a text~\cite{basis:06}. 
		It seems that humans think of their
		queries in terms of entities/concepts and then try translating them to some list of words which are most likely to appear in a document that
		is relevant to their search query. In general, a user does not ``know all the search terms that will return all instances of the desired
		information''~\cite{basis:06}. A survey of research in the field is presented in ~\cite{Nadeau:07}.

\subsection{Security Applications} 
	It has been proposed that text mining of emails can be used for detecting money laundering and 
	terrorist plans. However the effectiveness of data mining in this context is still debatable~\cite{Jonas:06,Federici:07}.

	We also propose another security application which can be used to be used when investigating in criminal cases:
	Given two sets of reports consisting of crime scene investigation and suspect interrogation reports, 
	an officer may query use a specialized text mining application to query for relations between suspects in terms of 
	criminal action, locality, or third party persons unknown to the officer.
	It can be envisioned that such an application will return two convicted people who had no direct relation
	in the reports, but who are related only through a third person unknown to the investigator. 
	Such a query is very hard to answer manually.

\subsection{Medical Applications} 
	Medical records and clinical reports are studied to detect
	patterns of occurrence of certain symptoms and diseases. It attempts relate those patterns with
	other biological factors such as certain 
	genes or protein interaction~\cite{blaschke:99, Krallinger:05, Rebholz-Schuhmann:07, Cohen:08, Tsuruoka:08}.
	They are also used to analyze factors correlating with medical accidents to minimize their occurrence~\cite{Kimura:08}.
	Till present, most of the interest in text mining comes from biological sciences~\cite{Giannopoulou:08}.

\subsection{Hadith Authentication}

In this thesis, we will present and propose to solve an interesting application 
for text mining in the field of Islamic Studies. 
A \noVocRL{.hady_t}~\footnote{In this document, 
we use the default ArabTeX transliteration style ZDMG.}
is a narration related to the prophet Mohammad
through a \noVocRL{sanad} or a sequence of narrators. 
The collections of traditions are the second source of
jurisprudence after the \noVocRL{qor'An} for all Islamic schools of thought. 
Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
Arabic with its transliteration and translation. 
We show proper names in boxes connected
to form complex names of narrators. 
For example, 
\novocalize
\noTrRL{qtybT} is the first name 
of narrator $n_1$, and 
\noTrRL{s`yd} is the name of his father as 
the word \noTrRL{bn} (son of) indicates. 
The sequence of names from $n_1$ to $n_5$ 
constitutes the \noArRL{sanad} 
of the \noArRL{.hady_t}. 
The second part of the \noArRL{.hady_t} is the 
\RL{matn} (content) and constitutes the actual content 
of the tradition.

Due to religious and political reasons, 
writing the traditions was forbidden 
until the days of the eighth Umayad Calif,
\novocalize
\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after 
the death of the prophet~\cite{AlAskari}. 
Consequently, many inconsistencies were
introduced to the literature which necessitated 
a thorough authentication study of a tradition before its use in 
jurisprudence.
\vocalize
An Islamic hadith scholar studies the authenticity of 
the \noArRL{sanad} of a 
set of related traditions before using these traditions 
jurisprudence. 
While different Islamic schools of thought differ on 
how to interpret the content, they almost all agree
that if a \noArRL{sanad} lacks authenticity, 
then scholars can not use the tradition for jurisprudence.

\novocalize

\transfalse
\begin{figure}[tb]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/exhadith.pdftex_t}}
\caption{Hadith abstraction example.}
\label{f:exhadith}
}
\end{figure}
% the FSMs for the three words
\transtrue
\vocalize

The authenticity of a \noArRL{.hady_t} depends on 
the credibility of the narrators as reported in 
separate biography books. 
The study of \noArRL{.hady_t} authentication is 
currently manual and error prone due to the huge number
of existing traditions and tradition books. 
Hadithopedia~\cite{Hadithopaedia:08}
reports that the tradition and biography
books for one of the Islamic sects amounts to more than 
300 thousand lines of text. 
Al-Azami\cite{Al-Azami-91} cites more than eleven books
of digitized tradition books each of several volumes, and a dozen
other biography and secondary authentication books such
as a geographical dictionary of places in hadith.

In this our thesis, we present an application
for authenticating inputted hadith based on inputted biography books.

\section{Arabic Text Mining}

Morphological analysis is key in current automated 
analysis techniques for Arabic text.
The almost comprehensive book 
``Arabic Computational Morphology''~\cite{Sou07}
covers substantial academic research work on Arabic text mining.
It summarizes recent work, and presents strong evidence of 
shortage of historic research. It also highlights  an emerging 
interest with a specific application to machine 
translation from Arabic to English after 2001.
The research community showed little interest in 
automating Arabic text analysis that may serve
the Arabic user.
While machine translation is of general interest, 
an Arabic user is more interested in 
information hard to discover manually.
Examples are the relation between clinical records, pharmaceutical 
prescriptions and sales, and a specific sickness,
the relation
between stock prices and political news, or the 
relation between interrogation reports with suspects
and investigative reports from the crime scenes.

\subsection{Distinctive Features of the Arabic Language}
\label{sec:ArabicDistinctive}

In our work, we will be focus on morphological analyzers. 
However it is noteworthy to mention that 
Arabic analyzers face many difficulties
inherent to the morphological analysis of Arabic. 
In what follows, we consider some distinctive features that
render Arabic morphology more complicated than that of Latin languages
such as English.

\subsubsection{\ref*{sec:ArabicDistinctive}.1 High degree of Ambiguity}

	Morphological ambiguity is a `notorious' problem for Arabic language~\cite{Kiraz:98}.
	Ambiguity refers to the fact that one word may lead to several possible solutions.
	As discussed in~\cite{Attia:08a}, the following are the reasons of genuine ambiguity\footnote{
	Here we are discussing ambiguity that is inherent to the language. Other sources of ambiguity
	may be due to over-approximations and relaxed compatibility rules in the morphological analyzer.} 
	in the language:
	\begin{enumerate}
		\item \textbf{Orthographic alternation operations:} These refer to alterations from the stem
			original letters upon inflection, which can make it similar to other distinct words.
			For instance, \noVocRL{ya`id} can result in 5 different solutions, 2 of which are
			inflections of \RL{A`id} (bring back) and \RL{`od} (come back) as in \RL{lam ya`od}.
			~\cite{Attia:08a}
		\item \textbf{Prefixes and Suffixes that are homomorphic:} For example, 
			``the prefix {\em ta-} can indicate 3rd person feminine or 2nd person masculine''~\cite{Attia:08a}.
			Consider \noVocRL{tadros} which can mean `she studies' or `you (masculine) study'.
		\item \textbf{``Coincidental Identity'':} In this case, complex forms (i.e. containing affixes) coincidentally
			are homomorphic to a simple full word ~\cite{Kamir:02}. 
			%An example is in the homomorphism of \noVocRL{Akala}
			%he ate) and \noVocRL{Akalla} [\noVocRL{A} + \noVocRL{kalla}] (he made tired).
			\vocalize 
			For instance, the word \RL{'a.hmadH} may have two valid morphological analyses. 
			The word means ``I praise him'' when the letter \RL{'a} is a prefix and  it means
			``his Ahmad'' when the letter \RL{'a} is part of the proper noun stem \RL{'a.hmad}.
		\item \textbf{Natural Homomorphism in Un-inflected Forms:} This occurs in \noVocRL{_dahaba} which can mean
			`went' as a verb and `gold' as a noun.~\cite{Attia:08a}
		\item \textbf{Omitted Diacritizations} (will be discussed next)
	\end{enumerate}

	The statistical measure of ambiguity is the average count of valid solutions per word. According to our results performed
	on 3 books of hadith~\cite{IbnHanbal,AlKulayni,AlTousi}, the resulting ambiguity score was 2.62 using an extended corpus to that of
	Buckwalter Analyzer~\cite{Buckwalter:02}\footnote{This analyzer will be considered in Section \ref{sec:literature}.}.
	Those results where verified by ~\cite{Attia:06b} which reported a similar score (2.6) and a score of 1.76 for English.
	This difference can be used to argue that Arabic is more ambiguous than English. Still however, the author of ~\cite{Attia:06b}, Attia,
	claims that this ambiguity resulted from some grammatical over-approximations and inaccuracies in the lexicon of Buckwalter along with
	its inclusion of classical terms that are rarely used. By this he means that the score does not reflect the intrinsic Arabic ambiguity and 
	they report a score of 1.75 for their tool which used tighter grammar and lexical rules and omitted classical words. Although, Dr. Attia
	has a point relating to exaggerated numbers due to over-approximations, the reduction of the lexicon is not a correct approach since it does
	not reflect the ambiguity of Arabic but of the subset of Arabic needed for some applications (mainly modern text documents such as news).
	In all cases, this is a question open to future investigations by NLP researchers and Arabic linguists...

\subsubsection{\ref*{sec:ArabicDistinctive}.2 Diacritics}

	It is common practice to write Arabic text without short vowels (diacritics) 
	which greatly increases its ambiguity. For instance, \noVocRL{jzr} can have very different interpretations based on 
	the fashion in which it is vocalized. Consequently, 
	\RL{jazar} is a noun that means carrots; 
	\RL{jozor} is a noun that means islands; 
	\RL{jaz"r} is a noun that means sink/ebb typically referring to the state of the sea as in \noVocRL{Al-mad wAl-jazr};
	\RL{jazara} is a verb in the past tense that means slaughtered/butchered.

	Since most Arabic documents are unvocalized, the ambiguity of the morphological analyses is significantly increased.
	For instance, in many cases, it is impossible to distinguish between active and passive verbs on the word level without
	appropriate diacritics such as the undiacriticized versions of \RL{`abada} ($<$subject$>$ worshiped) and \RL{`obida} ($<$subject$>$ was worshiped).

	Also, removing the shadda which is commonly assumed to be a diacritization removes practically a repeated letter as in \RL{`allama} (taught)
	and results in ambiguation with other words having a totally different meaning such as \RL{`alima} (learned)~\cite{Attia:08a}.

\subsubsection{\ref*{sec:ArabicDistinctive}.3 Run-on words}
\label{sec:ArabicDistinctive:runon}

	Arabic letters can have up to 
	four different forms
	corresponding to their position in a word, i.e. beginning,
	middle, end of word and separate forms. 
	This allows the phrase
	\noTrnoVocRL{il_A\nospace almdrsT}  with no delimiter within
	to be visually recognizable
	as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school).
	This happens because the first word \noVocRL{il_A} ends with
	\noVocRL{_A} a non-connecting letter. 
	The non-connecting letters are \RL{A}, \noVocRL{'a}, \noVocRL{'i},
		\noVocRL{'A},\noVocRL{_A}, \noVocRL{w}, \noVocRL{|u'|}, \noVocRL{|i'|},	
		\noVocRL{d}, \noVocRL{_d}, \noVocRL{r}, \noVocRL{z} and \noVocRL{T}.
	These words occur often in text and greatly increase the
	difficulty of tokenization and are referred to as 
	``run-on'' words~\cite{Buckwalter:04}.
	Such difficulties are inherent in the 
	morphological analysis of the Arabic language and
	render the solutions inaccurate.
	Consequently, tokenization of Arabic text is not a trivial task of 
	detecting delimiters as is the case in other languages.

\subsubsection{\ref*{sec:ArabicDistinctive}.4 Multi-word Expressions} % not totally distinctive

	Multiword Expressions (MWEs) are defined in~\cite{MWE} as
	``idiosyncratic interpretations
	that cross word boundaries (or spaces)''. 
	By this term, we refer to frequent cases when a group of words form together a single
	item with a semantical meaning not captured by its individuals words compositionally. 
	We focus on the example of compound names such as 
	\noVocRL{`bid alkarym}. 
	In current analyzers, this phrase is analyzed word by word and in best cases is 
	considered to contain two names, instead of being understood to be a single compound name.

\subsubsection{\ref*{sec:ArabicDistinctive}.5 Inflectional Nature}

	It is known that ``Arabic is a highly inflectional language, which makes the morphological
	analysis complicated'' ~\cite{Attia:08a}. In other words, instead of forming words from stems, the
	structure of the Arabic language is built from roots. That's why Arabic dictionaries are indexed by the roots.
	The Arabic stems are formed by applying some inflectional patterns that modify the meaning of the word.
	For instance, the root \noVocRL{drs} can derive the following stems: \RL{darras}, \RL{dAris}, 
	\RL{modarris}, \RL{madrows}, \RL{madrasaT}, etc..
	However, in Latin languages the inflectional forms don't transform the meaning completely and can generally be treated
	as prefixes and suffixes. In such inflectional languages such as Arabic, concatenative analysis sometimes fails to capture the
	word structure especially when occur inside the stem itself caused by infixes. In this our thesis, we will not focus on this problem,
	however we have mentioned it for completeness.

\section{Proposal Organization} %TODO:

The rest of the proposal is organized as follows. In Section 2, we present the problem(s) to be attempted. 
In Section 3, we review related literature. 
Then, in Section 4, we describe our preliminary results. 
Finally, in Section 5, a time-schedule and a research plan are presented.

\chapter{Problem Description}

In what follows we describe the problems that we would like to tackle in our thesis:

\section{Enhance Morphological Analysis Methods}

We propose developing `Sarf', a novel morphological analyzer that introduces many enhancements to the
field of morphological analysis. As noted in Section~\ref{sec:ArabicDistinctive}, 
Arabic poses many difficulties which currently 
have not been tackled in a satisfactory way. Of these we focus on the issues of `run-on' words and
the use of diacritics in morphological analysis -- two issues which have been under-looked by researchers in the field.
We will also work on speeding morphological analysis and 
enhancing affix representations to better reflect the semantics of underlying word structure.
Morover, we target supporting for interaction between the different text mining applications 
and the morphological analyzer, 
allowing the applications to drive the analysis and prune it on the fly. 
We will start first by justifying the need for the latter flexibility.

\subsection{Application-Specific Analysis}

Current Analyzers are unable to benefit from the characteristics of the application in which the Analyzer is used.
Their output is the same regardless of the needs of the application used and the characteristics of the corpus
being analyzed. We noted in Section~\ref{sec:Applications}, that
there is a broad spectrum of applications to which Text Mining is applied. In most cases, morphology 
``plays an essential role in any higher-level understanding and processing of Arabic text"~\cite{Sou07}.

Our focus when building Sarf, was not on improving the semantical rules and 
expanding the morphological alternatives, instead we worked on adding support 
for higher level control by the specific application utilizing the lower level 
morphological analyzer. In other words, the application-specific controller, has 
full control over and directs the morphological analysis of Sarf. 
% (sentence should be reworded differently, but is needed to explain why we mention 
% higher and lower accuracy hypothesis, which seems vague otherwise)

We justify the above enhancement as follows:

\subsubsection{Reduction in the Required Analysis}

We have noted that many NLP case studies need the 
morphological analyzer to answer queries that do not need 
high accuracy at the morphological analysis level.
The analysis at the higher levels 
(i.e. answering the application-specific query)
can often compensate for 
tolerable inaccuracy at lower morphological levels. 
For example, if a query concerns proper names and the 
analyzer is considering a 
prefix that connects only to a verb,
%the analyzer can answer without considering the 
%rest of the word.
the analyzer does not need to further analyze the rest of the word 
to provide a negative answer.
If another query is interested in extracting time entities, the only 
information that it needs to extract from the morphological analyzer 
is whether the word under consideration has a stem which reflects 
time structure; and whenever a positive answer is verified by one of the analysis,
the analyzer can stop without computing the other alternative solutions.

In other words, if the analyzer can be controlled at every step by an application-specific
controller, the analysis can be enhanced and reduced significantly.

\subsubsection{Evidence from Related Works}

\begin{enumerate}[a)]

\item We find evidence to our hypothesis in~\cite{Maamouri:10}.
The addition of a new corpus to the Arabic Tree Bank 
(ATB)~\cite{Maamouri:04}
%required compliance with the DARPA GALE program 
%including several NLP tasks such as English translation 
%annotations and word-level alignment of Arabic and English. 
%The new data 
challenged the existing processing and 
annotation guidelines. 
This led to refining 
the analyzer SAMA~\cite{Kulick:10} with 
alternative guidelines. 
%This refinement was achieved only after the NLP 
%task interacted with the morphological analyzer,
%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when the NLP 
task interacted with the morphological analyzer, which is
a flexibility our analyzer is designed to provide for the application-specific controller on the run.

\item In addition, 
the work of Habash~\cite{Habash:06} reports that different 
types of morphological analyses behave differently on the same %case study. 
data set.
The analyzers may report different solutions or 
may present the solutions in different orders. 
Clearly, some of 
the presented solutions and their order of presentation may not be 
appropriate for the %case study.
application needs, which is an issue that Sarf tackles. % (maybe reworded)
\end{enumerate}

\subsubsection{Possible Reorder of Analyzer Structures}

Analyzers provide different levels of features for the alternative solutions.
Commonly, those features include: the vocalized version corresponding to the analysis,
its Part of Speech (POS), its gloss (i.e. semantical meaning) and other abstract categories to which
it belongs (e.g. Proper Name, Place, Time, etc...)\footnote{For example, the analyzer Attia 
provides the following information for Nouns: Gender, Number, Case, Humanness, its possible
categorization as proper noun (subdivided into person, place, organization), etc..~\cite{Attia:08a}.}.
If it is known beforehand by the application if one feature is its primary interest, using which
early termination of a word analysis can be interpreted, the analyzer can reorder its structures to support for
faster lookup for this feature.

For example, by default in Sarf, the features are ordered in the following hierarchy: 
vocalized entry, abstract categories, gloss, then POS. In this way, Sarf will store corresponding vocalized entries per 
solution and for each of them will report the different abstract categories that pertain to this (solution, vocalization) tuple,
and so on...
In this case, if some analysis is interested in solely the part of speech of the words considered, an advanced 
application-specific analyzer can re-order its internal maps, trees, and lookup tables to report 
that information first, which will significantly speed up its retrieval.

\subsubsection{Refine Reported Solutions}

We argue also that by introducing an application specific controller that is capable
of intervening with the morphological analyzer at every decision, the reported query results 
can be refined significantly. This task can be much more difficult otherwise.

For example, assume we are
building an application that verifies that the output of an OCR document is correct by checking
that the scanned words are morphologically correct and exist in an exhaustive Arabic lexicon.
To make the problem more interesting, we will assume that our application can benefit from the morphological analyzer to 
improve its interpretation of the scanned glitch by refining its interpretation on the fly, restricting its output
to words that exist in the lexicon.\footnote{This application was inspired from the works of~\cite{Badr:95,Obaid:98} that
used benefited from the lexicon while performing the OCR; their methods, however, is not totally applicable to our
hypothetical application.}
We can envision that for a particular font used in some documents, 
the default OCR engine misrepresents a character 
by another with a high probability. Also it may be the case that word 
boundaries are confused in some other way than `run-on' words, possibly splitting a word into two \cite{Kukich:92}.

In such cases, if the analyzer's actual functionality cannot be altered, the application cannot 
benefit much from the output of the analyzer. On the other hand, if an application-specific controller
can be written to represent those peculiarities in the application at hand, the analysis will become 
of high utility. The interventions need not be complicated. For instance, in the above examples, all that might 
be needed is to intervene whenever any of the frequently confused letters is encountered in the analysis 
to continue the analysis with either of the possibly confused letters assumed as matching.
%In other words, refine the state of the machines to assume continue as if the alternative letter is encountered.
% (reword this example, very bad structure)
Another check might be needed when a space is encountered to solve the issue of a split word.
In this case, the controller upon detecting a white-space can drive another analysis of the expression by directing 
the analyzer to ignore the white-space and check if a valid solution will be generated.

In some cases, the application would consider the high number of run-on words that can be generated as possible solutions
as an unnecessary over-approximation. In that case, it can modify its controller to report only the `run-on' words it considers
interesting or most frequent. This can be done by possibly applying some intelligent heuristics such as ignoring all run-on words
which contains a sub-word of size: equal to one except for the clitic \noVocRL{w}; equal to two except for \noVocRL{mA} 
and \noVocRL{lA} which appear frequently in run-on words and so on... Such statistical refinements, when handled early on during
the analysis of the subparts of the word through the application-specific controller, are clearly much more affective than 
a later post-processing step.

~\\

We hypothesize that in many NLP applications, the morphological analyzer may
benefit from the characteristics of the query at hand, 
to greatly enhance the analysis.
Consequently, we believe that 
an application-specific 
controller that intervenes at every decision to guide, use, order, and 
refine the morphological analysis is of high utility.

\subsection{Solving the Issue of `Run-on' Words}

`Run-on' words as described in Section~\ref{sec:ArabicDistinctive:runon} represent multiple words
that are not separated by a delimiter but are visually conceived as non-connected separate words.

This problem is especially important due to the fact that many of the valuable 
Arabic texts available for text mining have been originally ported from
paper documents. OCR is often used to automate this task; however, this task
results in mistakes that are often escape manual inspection.
In particular, run-on words are admitted 
to be hard to detect especially when the `glyph design (i.e., the
perceived width) of the non-connecting letter' seems comparable
to that of a white-space~\cite{Sou07}. In his PhD thesis, Mohammad S. M. Khorsheed
explained the OCR difficulty that arises: ``Arabic scripts are inherently cursive...
Since not all characters connect, word boundary location becomes an interesting problem, 
as spacing may not only separate words but also certain characters within a word.''~\cite{Khorsheed:00}

OCR and manual errors of typing are not the only source for such mistakes. As the field
of speech mining gains interest and many systems that translate speech to text for parsing it
are getting prevalent, such errors are getting more prevalent. For instance, a study in~\cite{Kukich:92}
has shown that 13\% of nonword-spelling errors which have been generated when transforming
conversations to text are `run-on' words. Although this study was performed on English-like concept of
`run-on' words\footnote{such as `{\em willsee, weare, iam, onthe, yesthisis}'~\cite{Kukich:92}} 
which are unarguably assumed as errors, in Arabic readers generally don't assume such
expressions as being spelling mistakes~\cite{Buckwalter:04}.

The importance of this characteristic in Arabic morphology can be appreciated 
through the frequency statistics reported in~\cite{Sou07}. It reports some statistics
from Google Search Engine. Obviously, many ``run-on'' words have a high frequency in online 
Arabic texts exceeding the count of thousands. Another noteworthy observation that is reflected
in the collected numbers is that those frequencies increased in magnitudes from 2004 to 2006; this 
suggests that the solving this difficulty is a problem that deserves research attention, instead of 
one that is related to some old restricted group of documents. Also as noted in~\cite{Buckwalter:04},
lexicalized collocations which are spelled without intervening space -- such as \noVocRL{lAyzAl}, 
\noVocRL{lAbod}, etc... -- are ``just as frequent as their spelling in concatenated form".

This difficulty in tokenization has been reported by Buckwalter in~\cite{Buckwalter:04} because of its 
frequency in three corpus of the Arabic Tree Bank on which some experiments were performed. 
That paper was concluded that ``the problem of run-on words in Arabic calls for
a reassessment of current tokenization strategies,
including the definition of `word token' itself. It
should be assumed that each input string represents
one or more potential word tokens, each of which
needs to be submitted individually for morphology
analysis.'' In this thesis, we reassess current tokenization strategies and solve the issue
of run-on words in a very efficient way to be described later in Section~\ref{sec:runon}.

\subsection{Disambiguation using Diacritics}

Up to our knowledge, currently all Arabic morphological analyzers have ignored diacritics due to the complexity 
of their analysis and the tendency of Arabic writers to omit them.
Nonetheless, many Arabic documents especially religious texts are fully vocalized.
Even in documents written without diacritics, we often notice 
a couple of words which are partially vocalized
to reduce the difficulty of their disambiguation by the reader. Clearly, if native readers
need the aid of short vowels, automatic morphological analyzers which are context-free should not lose this precious piece of information.

\subsection{Support for Multi-Word Expressions} % maybe ommit %TODO

Many Arabic proper names are compound names which are composed of multiple words such as 
\noVocRL{bayt la.hem}, \noVocRL{`bd al-razAq}, etc...
It is important for the morphological analyzer to analyze them as one unit to be able to
correctly reflect their part of speech information. Especially for the hadith case study,
where the extraction of names of people is an essential task, having Sarf supportive of
multi-word expressions will significantly enhance the accuracy of the results.

\subsection{Enhancing Affix Representation}

\begin{table}[tb]
\centering
\caption{Illustrative Subset of the Prefix Lexicon of Buckwalter's Analyzer}
\resizebox{1\columnwidth}{!}{
\begin{tabular}{|c|c|c|c|c|} \hline
 ~ & \textbf{Vocalized} & \textbf{Compatibility}  & ~ & ~ \\
\textbf{Prefix} & \textbf{Prefix} & \textbf{Category} & \textbf{Gloss} & \textbf{POS data} \\ \hline \hline
\RL{w} & \RL{wa} & Pref-Wa & and & wa/CONJ+ \\
\RL{y} & \RL{ya} & IVPref-hw-ya & he/it & ya/IV3MS+ \\
\RL{wy} & \RL{waya} & IVPref-hw-ya & and + he/it & wa/CONJ+ya/IV3MS+ \\
\RL{fy} & \RL{faya} & IVPref-hw-ya & and/so + he/it & fa/CONJ+ya/IV3MS+ \\
\RL{sy} & \RL{saya} & IVPref-hw-ya & will + he/it & sa/FUT+ya/IV3MS+ \\
\RL{wsy} & \RL{wasaya} & IVPref-hw-ya & and + will + he/it & wa/CONJ+sa/FUT+ya/IV3MS+ \\
\RL{fsy} & \RL{fasaya} & IVPref-hw-ya & and/so + will + he/it & fa/CONJ+sa/FUT+ya/IV3MS+ \\

\RL{y} & \RL{ya} & IVPref-hmA-ya & they (both) & ya/IV3MD+ \\
\RL{wy} & \RL{waya} & IVPref-hmA-ya & and + they (both) & wa/CONJ+ya/IV3MD+ \\
\RL{fy} & \RL{faya} & IVPref-hmA-ya & and/so + they (both) & fa/CONJ+ya/IV3MD+ \\
\RL{sy} & \RL{saya} & IVPref-hmA-ya & will + they (both) & sa/FUT+ya/IV3MD+ \\
\RL{wsy} & \RL{wasaya} & IVPref-hmA-ya & and + will + they (both) & wa/CONJ+sa/FUT+ya/IV3MD+ \\
\RL{fsy} & \RL{fasaya} & IVPref-hmA-ya & and/so + will + they (both) & fa/CONJ+sa/FUT+ya/IV3MD+ \\ \hline
\end{tabular}
}
\label{t:affixes}
\end{table}

Current affix representations are compound and contain highly redundant information. 
These issues can be seen in Table~\ref{t:affixes}
which contains a subset of the Prefix Lexicon of Buckwalter's Analyzer.
The column for `Compatibility Category' is used to later validate
if the affix is compatible with other parts of the word. The `Gloss'
reflects the possibly compound description of the prefix. The `POS data'
conveys the part of speech information.

We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\cite{Buckwalter:02} keeps separate lists 
of affix and stem morphemes and assigns categories to
morphemes. 
The relation between morphemes and categories is one 
to many. 
A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that $c_1$ morphemes
can be concatenated with $c_2$ morphemes.
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix;
$\alpha\beta\gamma$ is a 
valid morphological analysis iff
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
$c()$ returns the category of the morpheme.

It is clear that in current analyzers, the list of affixes is much larger than needed
due to the explosion of possible valid combinations that result from the most basic 
prefixes. Note also that in current analyzers, sometimes to add just one prefix to the lexicon, it is required to
manually generate all suitable combinations with those previously in the dictionary.
For instance in Table~\ref{t:affixes}, the two syntactically and semantically different but homomorphic prefixes for \RL{ya}
repeat the same combinations with other prefixes.

Consider the case of the missing prefix from the lexicon of Buckwalter detected by Prof Attia. It is the 
``clitic question morpheme which can be prefixed to verbs and nouns''(\RL{'a})~\cite{Attia:06b}.
If we are to modify Buckwalter to support for its insertion that would result in reproducing most of the current
prefix entries to concatenate to them this possible clitic. Having prefixes been recursive as in Sarf, just a single 
prefix will be inserted along with some compatibility rules.

As noted in Table~\ref{t:affixes}, many affixes are composed of other affixes. For example,
the prefix \noTrnoVocRL{wsy} is composed of three other prefixes
namely \noTrnoVocRL{w}, \noTrnoVocRL{s}, and \noTrnoVocRL{y}.
The POS tag for \noTrnoVocRL{wsy} is a concatenation
%of the POS tags of each of the three morphemes. 
of those of the three morphemes. 
Keeping all valid morphemes in a list
is redundant and may lead to consistency issues when
users try to add their custom morphemes.
We introduce $L_{pp}$ and
$L_{xx}$ as prefix-prefix and suffix-suffix 
compatibility rules respectively.
This allows us to keep a shorter list of affixes. 
We also introduce a {\em resulting category}
for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
The resulting category in all practical cases
is that of the second morpheme in the pair.

Sarf only keeps atomic affixes and their categories.
By doing so, the affix representation better reflects the rules
which govern their formation. It also is able to better capture what each part of the compound prefix
contributed to the gloss of the word and its POS. For instance, it is much more instructive to analyze
and classify words based on their atomic affixes related POS and gloss instead of parsing string
delimited entries.

More than that, the fine scale of POS tagging and word division better suits any higher level attempt of semantical
parsing at the sentence structure and better aligns with tagged corpora which commonly tag atomic affixes such as clitic 
parts which are generally attached as prefixes. 
%(check for some citation or paper evidence?)

\subsection{Speeding the Analysis} %TODO:

As will be shown later in Section~\ref{sec:timeefficiency}, Sarf has outperformed commonly used
morphological analyzers. That has been made possible through many optimizations to the functioning
of the analysis methods and implementation, of which we enumerate the following:

\begin{enumerate}
\item \textbf{Linear Traversal of the input text through the use of finite state automata:}
	This will be described in Section~\ref{sec:Automata}.
\item \textbf{Efficient use of data structures:} Lexicon entries are loaded into main memory 
	when Sarf is started in appropriate structures that support efficient look-up. For instance,
	for stem lookup, we used a double array trie (datrie)~\cite{Aoe:89}. We also extensively used
	maps and hash-tables when needed. Also whenever an array reverse lookup was needed we stored
	that lookup array for constant time access instead of a linear search in the original array.
	Also when using a bit vector to access a specific bit, binary operations are used instead of
	multiplication and division...
\item \textbf{Reduced analysis and computations when not needed:} As discussed above, the 
	application-specific controller can enhance performance by some early analysis aborting, and
	non-complete retrieval of needed information for the analysis. For example, if for the application 
	at hand, only stems are needed, affix information such as gloss and POS need not be retrieved.
	Many other reductions can be envisioned suitable for particular applications of Sarf. 
\end{enumerate}
\section{Building a Time Entity Extractor}

One of the applications we propose to build using Sarf is a time entity extractor. This application
recognizes time expressions in text document and transforms them to time information of the default format
that can be stored in a database and on which computations and analysis can be performed.

Time structures which are conveyed in the content of a document instead of its meta-data 
have always been an interesting problem for the research community \cite{GROSS:93,Llido:01,DSTO:05,patent:09}. 
Li et al. working at the 
Microsoft have been awarded a patent for having devised an application for extracting authorship 
dates from documents~\cite{patent:09}. The importance of their work is justified by the fact that 
in many cases the modification date of a document is not representative of its date of generation due 
to many reasons. For example, ``when documents are uploaded or copied to
collaborative websites, the date meta-data gets changed from the last modification date to the 
upload date, which is rarely a significant or helpful date''~\cite{patent:09}. It is pointed also that any change in
other meta-data fields results in automatically changing the modification date of the document although its content would have 
remained intact. For such reasons, it is important to better approximate the authorship date 
using the actual content of the document. This date can be used later in the search engines, when the user is interested 
in documents that have been written in some time span. Obviously, a necessary capability for performing this analysis, 
is being able 
to extract temporal entities from documents.

Apart from authorship dates,~\cite{Llido:01} considers the problem of automatic assignment of event-time periods.
For many documents such as newspaper articles, medical reports and legal texts which contain many temporal 
references, time periods of the events under consideration in the document may be of much higher importance than 
the date of the document writing. This extracted information would facilitate the task of document retrieval.
As proposed and partially implemented in~\cite{Koen:00}, extraction of time references from news articles can help
augment those articles with personalized visualizations. Of those visualizations is a timeline containing all dates
extracted from the article with the ability of to navigate to sentences from which a selected date from the timeline 
is extracted. This tool also supports for augmentations which relate the events under consideration to
other news articles about incidents that are temporally
related to the one being read. Additionally, it provides an ability to display the historical context of an event and its relation
to the personal context of the user reading the article. For instance, it is proposed that the tool
be able to providing answers to questions like: `Where was I when that event took place?', `What other events took place
at the same time worldwide or in my neighborhood?', etc...

An interesting application based on time entity extraction was developed by a group of German researchers~\cite{Busemann:97}.
They developed a system for Appointment Scheduling via Emails (COSMA). This system automates the time-consuming task of
scheduling meeting between different parties having their own free/busy electronic schedule. The process of agreeing on
a suitable time for all corresponding participants takes more than one round of communication. The automation of this task
saves this time which is lost on such a routinely task.

In addition, when time entities are extracted, other data mining techniques can help discover useful relationships.
\cite{Berlanga:98} considers the problem of recognizing frequent temporal patterns in newspapers. While~\cite{Swan:99}
considered extracting time varying features from the news corpus to discover interesting events. Also, the problem of detecting
new evolving events in the news and tracking them was considered by~\cite{Allan:98}. One can also envision other usages in 
sentiment mining such as the development and advancement of the opinions of people living in some country as conveyed by the 
news articles across time.

Other applications also include comparison of parallel accounts and narrations of the same events. For example, many history 
books have been written to describe events that took place in the same periods of time. With event-time extraction, correspondence
between such events can be achieved. Afterwards detection of inconsistencies can be facilitated. Also, sorting events from the 
different accounts in partial-order time-line can help merging complementary accounts where one may describe events which are 
missed in another and vise versa. This analysis used to be performed all manually -- 
such as manually merging the different narrations of 
the Bible.

Finally, We would like to mention that the main motivation of this project is its usefulness in hadith authentication. 
In this field, one criterion for criticizing a given narrator
in the sequence of those having relayed the narration is based on 
time consistency. In other words, there should exist an overlap in the lifespans 
between a student narrating from his presumed 
teacher~\cite{Al-Azami-91}. In the absence of such an overlap, the relayed hadith loses its credibility.
A prerequisite to such consistency checks is being able to extract dates of birth and death from biography books and relate
them to individual narrators.

\section{Automating the Process of Hadith Authentication}

We also consider the case study of hadith authentication, a very important problem
that has historically been a totally manual work; however, authentication can clearly be 
enhanced by automatic aids.

We propose the following process:
\begin{enumerate}
\item \textbf{Segmenting Hadith Literature into Chains of Narrators:}

	A hadith book is divided into its different narrations and locating the sanad of each,
	subdivided into its chain of narrators. This can be performed by detecting sequences of narrators
	connected by narration connectors. This task has been completed and is presented in detail in Secion~
	\ref{sec:hadithSegmentation}.
\item \textbf{Building a Partial Order Graph of Narrators:}

	At this stage, the different extracted chains of Narrators are merged into a 
	partial order graph which shows the relation
	between the narrators and their contribution in relaying the hadith literatures. An example of such a 
	partial order graph
	is depicted in Figure~\ref{f:narrators}. 
	It has been extracted by our preliminary tool for the hadith case study introduced in Section~\ref{sec:hadith}.

\item \textbf{Localization of Narrators in Biograph Books:}

	This partial order graph is instrumental to automate
	localization of narrators in biography documents and
	segmenting biography documents in the future.
	Given that a biography discusses a narrator 
	and mentions his professors and students,
	we plan to segment the biographies with a graph 
	coloring algorithm
	that traverses the text and colors the graph node whenever
	a name is found.

\item \textbf{Graph Annotation and Consistency Checks:}

	Once the biographies are analyzed, we can annotate
	the POR with qualifiers of the narrators that reflect
	their authenticity. 
	We can also annotate the POR with the locations and 
	the time the narrators lived in. 
	Then we can compute a time and location overlap
	check to discover inconsistent narrations.
	We can also perform several interesting checks using 
	this POR on its own, such as checking for the effect of
	one narrator on the hadith literature.
\end{enumerate}

\chapter{Literature Review}
\label{sec:literature}

The survey in~\cite{Sughaiyer:04} compares
several morphological analyzers. 
Analyzers such as~\cite{Khoja:01,Darwish:02} 
target specific applications 
%in the analyzer itself or use a specific set of POS tags as their reference
or limit themselves to a specific set of POS tags. 
Sarf differs in that it is a general morphological 
analyzer that reports all possible solutions. 
It is application-specific in the sense that a controller 
prunes these solutions.

Buckwalter~\cite{Buckwalter:02},
Beesley~\cite{Beesley:01},
SAMA ~\cite{Kulick:10},
and ElixirFM~\cite{Otakar:07} 
take as input white space delimited tokens~\cite{Kulick:10},
consider them as words,
and enumerate all possible solutions. 
%This approach has several problems. 
However, we note the following points that deserve consideration:
%(We need not to mention especially point 2 as a "problem" according to reviewer comments)
First, a white space delimited token may have 
more than one word or just a part of a MWE.
%Second, the exhaustive enumeration may hurt performance and may
%not be necessary or appropriate
%in some case studies as noted in~\cite{Maamouri:10}. 
Second, in many applications, the exhaustive enumeration may not be needed or appropriate as 
noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.

Other morphological analyzers such as 
Amira~\cite{Diab:07,Benajiba:07},
MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
use machine learning and support vector machines (SVM) 
to enhance the accuracy of the morphological analysis at the expense 
of efficiency.
Xerox rules can be compiled into specialized finite state
finite state Automata (FSA)~\cite{Beesley:03}.
However, the efficiency of the resulting analyzer depends on the
way the Xerox rules are written. 
This requires deep knowledge and insight from the user
in compilation techniques, context free grammars, 
and morphological analysis.
%context free grammar rules that can be automatically compiled
%into efficient non-deterministic finite state Automata (NDFSA). 
Beesley reports that the number of Automata generated by a compiler
for Xerox rules can not be controlled and also reports a difficulty 
to compose the FSAs into a single framework. 
%Beesley requires recompilation of the Xerox rules. 
% (I am not sure if it is ok to have the above removed)
Sarf however constructs an NDFSA from 
%the controller and 
the prefix, stem, and suffix deterministic 
transducers to solve the problem.
Sarf also differs in that 
%it uses recursive affixes in the affix transducers, and in that 
the transducers consider
input from the controller to modify their behavior.
The user can express the controller,not necessarily an FSA, in a programming language such as C++.

Sarf builds upon the lexicon of Buckwalter\cite{Buckwalter:02}.
It differs from Buckwalter in that it defines a shorter list of affixes
and a longer list of 
compatibility rules to define 
affixes recursively and allow 
prefix-prefix and suffix-suffix 
concatenations.
This better maintains and propagates 
the tags associated with affixes.  
This information is highly 
redundant with Buckwalter and may lead to consistency issues. 
%Sarf keeps all the possible analyses alive where each analysis
%corresponds to a set of decisions and positions in the text. 
%Buckwalter produces a set of solutions for each token 
%and the several analysis threads of the text need to be 
%generated as a product of the token solutions. 
Similar to Sarf, Attia uses FSAs and handles more and qualitatively different MWEs
~\cite{Attia:06,Attia:10}.

%Similar to %the 
%MADA+TOKAN, Sarf uses
%the several valid morphological analyses as 
%%a richness that should 
%richness at a higher abstract level, and does not view them as 
%inaccuracies to be corrected. 
Sarf differs from MADA+TOKAN in that it provides an interface for the 
application to interfere in judging the analyses much earlier and
at a finer granularity level. 
%In Sarf, there is no need for a separate tokenizer such as
Sarf does not need a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The application-specific controller can build the trace 
%when needed.

Similar to The MADA+TOKAN toolkit, Sarf thinks of
the several valid morphological analyses as a richness that 
should be exploited at a higher abstract level rather than
an inaccuracy that should be corrected. 
Sarf differs in that it provides an interface for the 
application to interfere in judging the analyses much earlier and
at a finer granularity level. 
In Sarf, there is no need for a separate tokenizer such as
TOKAN since each solution keeps a stack of positions
that partition text into morphemes.
%The application-specific controller can build the trace 
%when needed.

%MAGEAD 
%Maamouri, Mohamed and \bgroup et al.\egroup },

Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
SAMA was refined to interact with
the ATB~\cite{Maamouri:04} project after the addition of a large 
new corpus. 
We find supporting evidence in this decision to our hypothesis
about the utility of a application-specific analyzer. 
The algorithmic changes in SAMA were
done manually and worked in integration with the ATB format. 
Sarf is an application-specific controller approach that can modify 
the behavior of the analyzer on the fly and can interact
with any application as needed. 
In Sarf, the analyzer remains general, and the 
application-specific controller specializes the analysis.

The AMIRA~\cite{Diab:07} analyzer uses 
a language independent SVM based analysis. 
The SVM analysis learns features from the input text 
and does not make use of any information 
in the target query.

Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
of the Buckwalter analyzer. 
Sarf also uses deterministic parsing with tries 
to implement the affix and stem transducers. 
We think that the inferential-realizational approach 
of ElixirFM
that is highly compatible with the Arabic linguistic 
description~\cite{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the application-specific controller
since in many cases the abstraction needs only a partial 
linguistic model of the Arabic language. 
We can compare the application-specific controller in Sarf to a high level
component in the ElixirFM domain specific language.

\chapter{Preliminary Results}

In what follows, we present our preliminary results which have been performed throughout the course of the past 2 years
which support our hypothesis and provide evidence for the effectiveness of our methods. We will divide our presentation into
results pertaining to the functionality of Sarf and results pertaining to the hadith case study in which Saf was a fundamental component.

\section{Application-Specific Morphological Analyzer (Sarf)}

Sarf extends the lexicon of Buckwalter~\cite{Buckwalter:02} with
proper names and location names extracted from different online 
sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.

Sarf encodes the prefix, suffix and stem lexicons into
three deterministic transducers and 
uses a non-deterministic construction
of the three to compute all valid morphological analyses.
We first present a running example 
%of Sarf. The running example 
that 
explains how Sarf works,
implements recursive affixes, and
handles ``run-on'' words.
%Then we discuss how Sarf builds and composes the different FSAs.

%\subsection{Sarf example}
%\label{sec:example}

\transfalse
\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/FSM.pdftex_t}}
\caption{Example affix and stem transducers.}
\label{f:example}
}
\end{figure}
% the FSMs for the three words
%\transtrue

\subsubsection{Sarf example}
The diagram in Figure~\ref{f:example}
illustrates the operation of $\Phi$, the Sarf FSA,
when parsing the
string \noTrnoVocRL{wsyl`b-h-aa\nospace al--lA`bwn} \noArRL{wsyl`b-h-aa al--lA`bwn}.
%\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
\footnote{ 
%\noVocRL{n}%
%\noVocRL{w}
%\noVocRL{b}
%\noVocRL{`}
%\noVocRL{a} 
%\noVocRL{l}
%\noVocRL{l}
%\noVocRL{a} %
%\noVocRL{a}
%\noVocRL{h}
%\noVocRL{b}
%\noVocRL{`}
%\noVocRL{l}
%\noVocRL{y}
%\noVocRL{s}
%\noVocRL{w}
\noTrnoVocRL{n}%
\noTrnoVocRL{w}
\noTrnoVocRL{b}
\noTrnoVocRL{`}
\noTrnoVocRL{a} 
\noTrnoVocRL{l}
\noTrnoVocRL{l}
\noTrnoVocRL{a} %
\noTrnoVocRL{a}
\noTrnoVocRL{h}
\noTrnoVocRL{b}
\noTrnoVocRL{`}
\noTrnoVocRL{l}
\noTrnoVocRL{y}
\noTrnoVocRL{s}
\noTrnoVocRL{w}
in separate form to ease following the example
}

Boxes and circles denote accept and regular
states respectively.
The edges are transitions with labels corresponding to
the input letters.
The reject states are omitted for clarity and are denoted
by the absence of corresponding transitions. 
Subfigures (a), (b), and (c) in Figure~\ref{f:example}
represent ${\cal P}$ the prefix ,
${\cal S}$ the stem, and ${\cal X}$ the suffix
transducers respectively. 
The symbol $\epsilon$ represents an empty string and is 
the source of non-determinism.

We start with $\Phi$ at the root accept state 
in ${\cal P}$. 
%The $\epsilon$-edge connects ${\cal P}$
%to ${\cal S}$ and transitions from any accept state
%in ${\cal P}$ to the root state in ${\cal S}$.
%The $\epsilon$-edge that connects ${\cal S}$ to ${\cal X}$
%follows the same behavior. 
The $\epsilon$-edges connect ${\cal P}$
to ${\cal S}$  and ${\cal S}$ to ${\cal X}$.
They transition from any accept state
in the current transducer to the root state of the target transducer.

When there are two valid transitions such as \noTrnoVocRL{w} 
and $\epsilon$ in the start case, 
$\Phi$ spawns an exact copy of itself $\Psi$. 
$\Phi$ makes one transition, and $\Psi$ makes the other. 
Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
An FSA %($\Phi$ or $\Psi$) 
dies when it reaches a reject state.
In our example, if there were no stems that start 
with the letter \noTrnoVocRL{w}, $\Psi$ will die. 
In reality, $\Psi$ will die when the input 
is at \noTrnoVocRL{wsyl`}.

Note that the affix transducers allow recursive 
affixes. 
For example, \noTrnoVocRL{w} 
%will result 
results 
in an accept state
that transitions to ${\cal S}$.
When \noTrnoVocRL{s} follows, we move to another accept state in 
${\cal P}$ corresponding to 
%the 
prefix \noTrnoVocRL{ws}. 
%The same applies to ${\cal X}$.

Lets consider $\Phi$ after it consumed \noTrnoVocRL{wsy} 
and transitioned into the root node of ${\cal S}$.
Now $\Phi$ will transition with \noTrnoVocRL{l`b} to reach an accept 
state.

Before moving with the letter \noTrnoVocRL{b} to the accept state,
$\Phi$ needs to make sure that %the 
stem \noTrnoVocRL{l`b} is compatible
with %the 
prefix \noTrnoVocRL{wsy}. 
Sarf keeps compatibility information as part
of the accept states. 
%Sarf considers the value of the category as part of the state.
Thus each accept state in Figure~\ref{f:example} represents
more than one concrete state. 
If the categories of \noTrnoVocRL{l`b} and \noTrnoVocRL{wsy} are compatible
then \noTrnoVocRL{b} will move $\Phi$ to an accept state. 
Otherwise, it will move it to a regular or a reject state.

Since $\Phi$ is now in an accept state in ${\cal S}$, it 
spawns $\Xi$ which transitions to the root of ${\cal X}$. 
After it consumes \noTrnoVocRL{h--} $\Phi$ dies since there is no
\noTrnoVocRL{h--}-edge from the current state.
$\Xi$ represents a valid analysis and proceeds.

Note that $\Xi$ reports a full word at any accept state
at ${\cal X}$, whenever a white space or delimiter is encountered.
We left that out in Figure~\ref{f:example} for clarity. 
However, non-connecting letters result in spawning a new FSA 
using the $\epsilon$ transition
to ${\cal P}$, to be able to analyze the other `sub-word' of 
the possible `run-on' word.

\subsection{Recursive affixes}
\label{sec:recaffix}

We call two morphemes compatible if their concatenation
forms a legal morpheme. 
Buckwalter~\cite{Buckwalter:02} keeps separate lists 
of affix and stem morphemes and assigns categories to
morphemes. 
%The relation between morphemes and categories is one 
%to many. 
A compatibility rule is a pair of categories 
$\langle c_1, c_2\rangle$  stating that $c_1$ morphemes
can be concatenated with $c_2$ morphemes.
Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
of compatibility rules relating
prefixes to stems, stems to suffixes, and prefixes to suffixes
respectively. 
Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
a prefix, $\beta$ is a stem, and $\gamma$ is a suffix;
$\alpha\beta\gamma$ is a 
valid morphological analysis iff
$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
$c()$ returns the category of the morpheme.

Many affixes are composed of other affixes. For example,
the prefix \noTrnoVocRL{wsy} is composed of three other prefixes
namely \noTrnoVocRL{w}, \noTrnoVocRL{s}, and \noTrnoVocRL{y}.
The POS tag for \noTrnoVocRL{wsy} is a concatenation
%of the POS tags of each of the three morphemes. 
of those of the three morphemes. 
Keeping all valid morphemes in a list
is redundant and may lead to consistency issues when
users try to add their custom morphemes.
We introduce $L_{pp}$ and
$L_{xx}$ as prefix-prefix and suffix-suffix 
compatibility rules respectively.
%This allows us to keep a shorter list of affixes. 
We also introduce a {\em resulting category}
for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
The resulting category in all practical cases
is that of the second morpheme in the pair.

Consequently, the above representation allows Sarf to keep 
only atomic affixes, their categories and compatibility rules.

\subsection{Finite State Automata Implementation}
\label{sec:Automata}

In what follows, we detail the implementation and functionality of Sarf's
automata for both stems and affixes.

\subsubsection{Affix transducers}
\label{sec:affixFSA}

We analyzed the prefixes and suffixes of 
Buckwalter~\cite{Buckwalter:02}
and constructed the recursive affixes
into directed acyclic graphs (DAG) similar to 
those in Figures~\ref{f:example}(a) and~\ref{f:example}(c).
The transducers are deterministic and thus 
their traversal is linear.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings %of the word in question
and looks them up in the affix tables.

The affix transducers encode $L_{pp}$ and
$L_{xx}$.
An accept state in the transducer corresponds to the last letter 
of an affix in the affix lists.
The accept states produce as output the POS and other tags
associated with the state.

\subsubsection{Stem transducer}
\label{sec:stemFSA}

We built our stem lexicon using the stem lexicon of 
Buckwalter. 
We augmented the lexicon with a set of proper names and
a set of location names which we 
obtained from online and biblical sources.

%The stems share lots of substrings. 
Since stems share lots of substrings. 
We encoded them in
an efficient double array trie (DAT) structure~\cite{Aoe:89}. 
The structure is deterministic and free of cycles where the 
accept states are the terminal nodes corresponding to the last 
letters in stems. 
Note that DAT is a suffix trie implementation which may not 
necessary work on letter-by-letter states, but also compacts
common substrings.
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need
a number of hash lookups in the order of all possible partitions
of the string.

\subsubsection{Construction of Sarf from transducers}
\label{sec:ndfsa}

The diagram in Figure~\ref{f:composition} shows an 
abstraction of the non-deterministic construction of Sarf
from the prefix, stem and suffix transducers. 
The circle represents regular states, the square
represents accept states and the triangle represents
reject states. 
Note that the concrete prefix, stem and suffix transducers
are all deterministic and cycle free as discussed 
in previous subsections.
%in Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}. (bc will result in same refernce bc are subsubsections
The cycles are introduced in the abstract machine
for pedagogical purposes.

%Note that Sarf does not rely on implementing its machines as 
%equivalent regular expressions due to  the following reasons ...
% (not necessary)

\begin{figure}[tb]
\center{
\resizebox{\columnwidth}{!}
{ \input{figs/abstract_machine.pdftex_t}}
\caption{Sarf NDFSA.}
\label{f:composition}
}
\end{figure}

The algorithm \CodeIn{NDSarf} (detailed below)
takes as input a text string $L$ and a application-specific controller
$A$. 
It starts with one machine $\Phi_1$ similar to the one in 
Figure~\ref{f:example} and inserts it into a collection
$M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all machines in $M$. 
For each $\Phi$ in $M$,
if $\Phi$ is in an accept state and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space, a delimiter, 
or when the preceding letter %leading to the current state
was a non-connecting letter.

In all cases, if $\Phi$ is in an accept state, 
it spawns another machine $\Psi$ and adds it to $M$. 
Notice that spawning a new machine requires only keeping
the state of the machine since all machines share the
structures of the three transducers.
If $c$ leads to a reject state, then $\Phi$ dies 
and we remove it from $M$. 
Otherwise, $\Phi$ transitions using the $c$ edge.
After iterating over all machines in $M$, we invoke the 
controller $A$
and grant it full access and control over
all machines in $M$.

The arrows in Figure~\ref{f:composition} that transition
from prefix to stem, stem to suffix, and suffix to prefix
require compatibility constraints for 
the transition.

\begin{table}[tb]
\centering
%\resizebox{.9\columnwidth}{!}{
\begin{tabular} {p{8.5cm}}
\begin{Verbatim}[fontsize=\relsize{-2},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}] 
$$
NDSarf(string $L$, Controller $A$) 
  MachineVec $M$; -- collection of machines
  Machine $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.state.isAccept() ) \{
        if ( $\Phi$.isSuffix() ) \{
          if ( $c$.isWhite() )  
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting()) -- run-on
            $\Phi$.transition($\epsilon$); -- back to Prefix Machine
        \}
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); 
      \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.transition($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} 
    \} 
    $A$.control($M$, $c$); 
  \}
\end{Verbatim}
\end{tabular}
%}
\label{a:ndsarf}
\end{table}

\subsection{Application-Specific Controller}

A valuable aspect of Sarf is that its machines are driven by 
an application-specific controller that can refine the computed
analyses as needed.

The controller can interfere at any point in the 
analysis to make a decision like ignoring a solution that 
is not interesting. 
It may also decide to correct one solution
based on the decision of the others. 
The controller uses the output of the transducers as input
to make 
%its decision. 
refinement decisions.
%%It does not produce external output and therefore 
%%the three transducers with the 
%%controller compose an FSA.

In a sense, Sarf can be treated as a customizable morphological analyzer which can be
programmed to take into account the characteristics of the application at hand.

\subsection{Disambiguation Using Diacritics}
\label{sec:diacritics}

Sarf is the first in the field to support diacritic-based disambiguation. Consequently, 
Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
or \RL{`ilm} (science) with \RL{`lim} (knew), regardless of the specific letters which are vocalized.
The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.

The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.

\subsection{Handling Run-on Words}
\label{sec:runon}

As depicted in Figure~\ref{f:example}, there exists an $\epsilon$ 
transition from ${\cal X}$ to ${\cal P}$.
A cloned state of the machines spawns this transition 
only when a non-connection letter is the last encountered letter.
In this way, the remaining part of the word is analyzed as if a 
complete word consisting of Prefix, Stem, Suffix concatenated constituents.
Note that upon performing this transition, the analysis of the previous subpart is
not reported since it may turn that this structure does not result in a valid run-on word.
Results are only reported when a white space or delimiter is encountered
and the word is at some accept state in ${\cal X}$; otherwise, the corresponding
analysis dies.

If results were reported at at any connecting letter, many false positive 
run-on words will be generated. For instance, \noVocRL{al-madrasaT} would have its 
\noVocRL{al-mad} subpart reported as a valid run-on constituent although the analysis
will die finally since the remaining part \noVocRL{rasaT} is invalid. Such errors are 
avoided by restricting the result reportage to the encounter of a white space or delimiter.
In the case of a run-on word, the complete analysis is reported consisting of a valid analysis 
to multiple `words'.

\subsection{Support for Multi-word Expressions}
\label{sec:MWE}

Current analyzers, with the exception of Attia, fail to recognize MWE such as \noTrnoVocRL{`bid alkarym}, and 
report it as two consecutive names. 
Sarf includes such expressions in the stem transducer (with one letter being the space) and 
detects them as one unit since tokenization
is performed on the fly. The multiword stems
can accept some compatible affixes such as \RL{w} (and) and \RL{f-} (but).
This approach works for proper names but not for ``Syntactically-Flexible Expressions'' which are
MWEs that can undergo lexical inflections for their constituent words~\cite{MWE}.

The strength of our multi-word expression Support is that it can be designed to support for
multi-word expressions which are written as one word in a similar fashion to run-on words such as
\noVocRL{`abidAl-ra.hyim}. A correct analysis can be achieved since spaces that exist inside stems 
that exist in the lexicon are allowed to match with an empty string when the previous letter is a
non-connecting one. If such a flexibility was not supported, it would have been required to add
space-containing multi-word versions along with run-on-like multi-word versions of those expressions,
which would be highly redundant.

\subsection{Time Efficiency Comparison Results}
\label{sec:timeefficiency}

\begin{table}[tb]
\centering
\caption{Results of Sarf, Buckwalter and ElixirFM.}
\resizebox{0.9\columnwidth}{!}{
%\begin{tabular}{lp{.01cm}cp{.01cm}cp{.01cm}c}
\begin{tabular}{lccc}
\hline
 & Al Kafi & Al Istibsar & Ibn Hanbal \\ \cline{2-4} 
Word Count & 98,943 & 103,835 & 20,354 \\
Sarf (sec) & 2.70 & 2.69 & 0.236 \\
Sarf/Controller (sec) & 1.32 & 1.31 & 0.0960 \\
Buckwalter (sec) & 3.32 & 5.41 & 0.547 \\
ElixirFM (sec) & 167$\times60$ & 138$\times60$ & 29.2$\times$60 \\
\hline
\end{tabular}
}
\normalsize
\label{t:efficeincyresults}
\end{table}

%In this section we compare Sarf to existing 
%morphological analyzers such as Buckwalter and ElixirFM in terms of running time.
We ran Sarf without a controller then with a controller that only 
focused on proper names and words that are related to the word ``narrated".
We compared the running times to those of Buckwalter and ElixirFM.
We used three books of Islamic narrations \RL{.hady_t} ~\cite{IbnHanbal,AlTousi,AlKulayni} as our data set.
We report our results in Table~\ref{t:efficeincyresults}.

With a controller, Sarf performed much better since the controller pruned several solutions irrelevant to the application. 
Sarf, with and without a controller, was more efficient than both Buckwalter and ElixirFM. 
Buckwalter performed much better than ElixirFM in 
terms of running time. 
The difference may be due to the sophisticated formal 
functional morphological analysis of ElixirFM. 
%and its reliance on additional annotations extracted from 
%both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.

\section{Hadith Case Study}
\label{sec:hadith}

Several researchers~\cite{Hadithopaedia:08,Azmi-2010} attempted to automate 
the analysis of the hadith literature. 
In this our thesis, we use Sarf to successfully automate the
analysis of three books of hadith selected 
arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
  the digitized books from online sources such as 
  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.

In what follows, we present our preliminary tool:
the hadith {narrator chain 
extractor using morphological analysis} (NaCEMA).
It employs novel techniques for automating
the extraction of the \noArRL{sanad} from the books of traditions 
into a diagram of chains of narrators. NaCEMA segments
a book into narrations, each narration into its \noArRL{matn} 
and its \noArRL{sanad}, 
and each sanad into its separate narrators. 
This work will lead to 
automate an exhaustive \noArRL{.hady_t} authentication 
effort in the future as we plan to 
automatically analyze the biography books.

\subsection{Hadith Segmentation and Abstraction}
\label{sec:hadithSegmentation}

NaCEMA accepts a book as input
and segments it into a vector of hadiths. 
NaCEMA segments each hadith into its
sanad and matn parts. 
NaCEMA 
detects the chain of narrators in the sanad and 
the relation that links each narrator to his ancestor and 
predecessor in the chain. 
It also detects the full name of each narrator that is
composed of several proper names with connectors
in between.

\subsection{The hadith controller}
\label{sec:controller}
% FSM for the controller

Intuitively, the controller reports a valid chain 
of narrators when a sequence of names
connected by narrator connectors appears. 
This includes the task of finding multi-word names
such as \RL{`bd alr.hmn} often appearing as ``run-on'' words.
The controller marks the beginning of the hadith with the beginning of 
the current proper name sequence,
and marks the end of the hadith with the beginning of the 
next proper name sequence. 
It detects the chain of narrators as the sequence itself.

%The controller also targets words that mean ``narrate'' when
%they appear between proper names. 
We illustrate the hadith controller 
in Figure~\ref{f:hadith}. 
The transitions are labeled by outputs generated from 
the Sarf morphological analyzer that correspond to 
part of speech (POS) tags. 
A NAME label denotes a proper name.
An NRC label denotes a narrator connector such as
\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
\RL{'a_hbar} ``told'' or one of their morphological equivalents. 
An IBN label denotes the word \RL{ibn} ``the son of'' and is a name connector.
A NISBA is a possessive adjective that qualifies a person such 
as \RL{alma.sriy} ``the Egyptian''.

The symbol $\tau_{\mbox{NMC}}$ is a threshold
that corresponds to the number of tolerated name connectors 
that may occur between two names. %3
The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
of name connectors collected since the controller
started looping in the state NMC\_S. 
The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
that corresponds to a relaxed tolerance measure that
the controller resorts to in case the words separating
two names were longer than $\tau_{\mbox{NMC}}$ but 
contained a name connector word such as IBN or NISBA.

The controller has four states that correspond to 
a position in text relative to the next sanad. 
State TEXT\_S is the initial state and denotes that
the controller is outside the context of a sanad.
The controller moves to state NAME\_S on
NAME.
State NRC\_S means that the controller is in the context
of a narrator name after Sarf reports an NRC.
State NMC\_S
indicates that the controller expects a name to appear within 
a tolerance threshold expressed by 
$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.

Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
before it gives up on its expectations. 
We reach the NAME\_S state only when a
valid NAME is detected and we leave when no more NAME's are detected.
The NRC\_S state can only be reached if an NRC is detected.
NMC\_S can only be reached from a NAME\_S state.
We used the values of 3, 5 and 5 for $\tau_{\mbox{NMC}}$, 
$\lambda_{\mbox{NMC}}$, and $\tau_{\mbox{NRC}}$ respectively.

\begin{figure}[tb!]
\center{
\resizebox{.9\columnwidth}{!}
{ \input{figs/hadith.pdftex_t}}
\caption{The controller of the hadith application.}
\label{f:hadith}
}
\end{figure}

%The definition of input labels such as NAME and IBN depends on the 
%morphological analyzer. 
%However, our case-based controller approach was able to perform well
%under both Sarf and a refined version of Sarf.

\subsection{Building Narrator Partial Order Graphs}
\label{sec:POR}

\begin{figure}[tb]
\center{
\resizebox{1.1\columnwidth}{!}
%{\includegraphics{figs/narrator_chain_output.pdf}}
{\includegraphics{figs/DAG_narrator.pdf}}
\caption{Narrator directed acyclic graph 
    extracted using NaCEMA }
\label{f:narrators} 
}
\end{figure}

The directed acyclic graph (DAG) 
in Figure~\ref{f:narrators} shows a partial order relation (POR) 
between narrators.
We automatically extracted the POR from the three books of 
hadith~\cite{IbnHanbal,AlKulayni,AlTousi}.
The nodes in boxes are the \RL{matn} of the hadith, 
and the other nodes are the narrators.
Given a set of chains of narrators similar to 
Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
that merged equal names in the chains. 
We compared the narrators together using a morphological
distance function. 
We considered each narrator to have several morphological 
components and we rewarded every morphological match with 
a small $\delta$. 
Then we applied a threshold $\tau$ to the sum of rewards
and decided the equality of the narrators. 
We omitted the details of the distance function for brevity; 
%
in all cases, we are considering their revisit and re-implementation
in ways more suitable to the hadith case study.

\subsection{Accuracy Results}

\begin{table}[bt]
\centering
\caption{Results of the hadith case study with Sarf.}
%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
\resizebox{1.1\columnwidth}{!}{
\begin{tabular}{lcp{.2cm}cp{.2cm}c} %\cline{2-10}
 &  AlKafi & & AlIstibsar & &IbnHanbal \\ \cline{1-6}
Word count  &98,943 & & 103,835 & & 20,354 \\ 
 Names       & 12,060  & & 14,613& & 3,013\\
%Names/Narrator & 1.97 & & 1.84& & 1.25 \\
Narrators  & 2,623 & & 5,767& & 1,755 \\ 
%Narrators/Chain  & 4.84 & & 4.76 & &4.05 \\
Chains  & 542 &  & 1,211& & 433 \\ 
Ignored names  & 6,400 &  & 3,348 & & 642 \\ \hline
Segmentation accuracy  & 96\%& & 96\%& & 92\%\\ 
Chain accuracy & 99\%&  & 99\%& & 97\% \\ 
%Narrator accuracy  & & & & &  & & & \\ 
Narrator accuracy  & 91\%& & 90\% & & 90\% \\ \hline
%Name false positives  & 7\%&  & 4\% & & 4\% \\ \hline
Running time (secs.)& 1.32 & & 1.31 & & .096\\ \hline 
\end{tabular}
}
\normalsize
\label{t:hadithresallresults}
\end{table}

We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
with 4GB of memory using a Linux operating system. 
We report our results in Table~\ref{t:hadithresallresults} on the 
three hadith books.
NaCEMA detected narrator chains and segmented the books 
correctly with a chain accuracy above 97\% and a narrator name
accuracy above 90\%
in all cases and across all the 1986 traditions in the three 
different books with different formats.

The row ``word count'' is an indicator of the size of the data.
The %numbers in 
rows ``names'', ``narrators'', and ``chains'' denote the total
number of the corresponding entities per book.
The row ``ignored names'' reports the number of names detected 
outside the chains and thus are irrelevant to
the application. 
%In addition, the rows ``names/narrator'' and ``narrators/chain'' 
%report the average
%number of names per narrator and 
%the average number of narrators per chain respectively.

The ``segmentation accuracy'' reports on the accuracy of 
segmentation of the book into separate hadiths.
We computed the metric manually by marking the beginning 
and end of a representative sample of hadiths in each 
book and compared that to the NaCEMA results. 
We used the same approach to compute all other accuracy measures.
The ``chain accuracy'' reflects the percentage of correct narrators 
covered by extracted chains.
The ``narrator name accuracy'' row reflects the percentage of 
correct proper names detected within
narrator structures.

%The row ``name false positives'' reports the percentage of words 
%recognized as names while they should not.
%We notice that this metric did not affect the accuracy
%of narrator and chain segmentation since detecting
%a narrator or a chain with the refined analysis without substantially affecting the 
%high abstract measures such as chain and segmentation
%accuracy.

%\subsection{Future Suggestions}

Although our results reflect a high accuracy, we plan to 
enhance the accuracy of hadith detection and the 
narrator merging algorithms. We also plan to make use of the name connector and narrator
connector stop phrases that our case study extracted to create a 
learning system that can learn more names. 
%We also plan on developing an Arabic linguistic computational model 
%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
%explored partially based on the case study.

\chapter{Thesis Plan}

The main tasks of the thesis are listed below: \\
\begin{enumerate}[1)]
\item \textbf{Build and Enhance Sarf}
	\begin{enumerate}[\ref*{s:intro}.1)]
	\item Port Buckwalter into C++ and mysql
	\item Implement the Finite State Automata Framework
	\item Implement an application-specific controller interface
	\item Support for disambiguation using diacritics
	\item Support for multi-word expressions
	\item Solve the Issue of `run-on' words
	\item Optimize performance
	\item Support for Running Sarf Under the Windows Operating System
	\item Refactor the code
	\item Write a user Manual
	\item Write a developer manual and documentation
	\end{enumerate}
\item \textbf{Build a time entity extraction application}
\item \textbf{Automate the Hadith Authentication Case Study}
	\begin{enumerate}[3.1)]
	\item Extend the Lexicon of Buckwalter with proper names 
	\item Implement the hadith segmentation tool
	\item Write the Narrator Equality Function
	\item Build Parial Order Graphs
	\item Correct Graph mistakes that result from Ambiguous narrator names which introduce false cycles to the graph
	\item Segment biography books using a graph coloring algorithm
	\item Extract Annotations from the Biography books
	\item Assign annotations to graph nodes
	\item Perform time and possibly location consistency checks
	\item Perform Experiments on the effects of different narrators in Hadith Literature
	\item Write a user Manual
	\item Distribute the tool to willing religious scientists and report their findings
	\end{enumerate}
\item \textbf{Write thesis Report and defend the thesis}
\end{enumerate}

\section{Time Table}

We anticipate that the work can be performed according to the time table presented in
Table~\ref{t:timetable}. \\

~\\

\begin{table}[bt]
\centering
\caption{Thesis Time Table}
\resizebox{1\columnwidth}{!}{
\begin{tabular}{|c|c||c|c||c|c||c||c|c|c|c||c|c|c|c||c|} \bottomrule[0.1em]
\textbf{Year} & \multicolumn{9}{|c|}{\textbf{2011}} & \multicolumn{6}{|c|}{\textbf{2012}} \\ \hline
\textbf{Semester} & ~ & \multicolumn{2}{|c||}{\textbf{\tiny Spring}} & \multicolumn{2}{|c||}{\textbf{\tiny Summer}} 
		& ~ & \multicolumn{4}{|c||}{\textbf{\tiny Fall}} & \multicolumn{4}{|c||}{\textbf{ \tiny Spring}} & ~ \\ \hline
\backslashbox{Task}{Month}
			& \textbf{Done} & \textbf{5} & \textbf{6} & \textbf{7} & \textbf{8} & \textbf{9} & \textbf{\small 10} & 
			 \textbf{\small 11} & \textbf{\small 12} &
			 \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} &
				\begin{tabular}{c}
				\textbf{\small Future} \\
				\textbf{\small Work}
				\end{tabular}
			 \\ \toprule[0.1em] \bottomrule[0.1em]
\textbf{1.1)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{1.2)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
\textbf{1.3)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{1.4)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~  & ~ &\\ \hline
\textbf{1.5)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{1.6)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~  & ~ &\\ \hline
\textbf{1.7)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{1.8)} & \x & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
\textbf{1.9)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & ~ & \\ \hline
\textbf{1.10)}& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X  & ~ & \\ \hline
\textbf{1.11)}& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \toprule[0.1em] \bottomrule[0.1em]
\textbf{2)}   & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \toprule[0.1em] \bottomrule[0.1em]
\textbf{3.1)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{3.2)} & \x & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{3.3)} & \x & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{3.4)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{3.5)} & ~ & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{3.6)} & ~ & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{3.7)} & ~ & ~ & ~ & ~ & ~ & \X & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{3.8)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
\textbf{3.9)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & \X & ~ & ~ & ~ & \X \\ \hline
\textbf{3.10)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & ~ & ~ & \X \\ \hline
\textbf{3.11)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \hline
\textbf{3.12)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X \\ \toprule[0.1em] \bottomrule[0.1em]
\textbf{4)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X &  \\ \hline
\end{tabular}
}
\normalsize
\label{t:timetable}
\end{table}

\large\underline{\textbf{Important Dates}}\\\\
\normalsize
The important dates are as follows:
\begin{itemize}
\item \textbf{Comprehensive exam:} Mid May 2011
\item\textbf{Thesis defense:} Mid May 2012
\end{itemize}

% Include Bibliography
\bibliographystyle{IEEEtran}
{\small \bibliography{jm}}
%\nocite{*}

\end{document}
%-------END----------------------------------------------------

### 2011-04-30 trunk/tex/tp.tex
wrote automatic code for splitting prefixes, not useful yet for suffixes.
Did much of the work and splitted prefixes except for some documented issues due to subject/object modifications and discovered some minor inconsistencies in Buckwalter's work

@@
 \fancyfoot[R]{ \scriptsize \textit{ Thesis Proposal Report}}
+
+\renewcommand{\labelenumi}{\arabic{enumi}.}
+\renewcommand{\labelenumii}{\labelenumi\arabic{enumii}.}
+\renewcommand{\labelenumiii}{\labelenumii\arabic{enumiii}.}
 
@@
 Text Mining 
-%can be considered as a subfield of both Natural Language Processing (NLP) and Data Mining.
 is interested in automatically extracting useful information and relations from text documents.
-%Extracted information is hard to discover manually or previously unknown
-%Natural Language Processing applications perform 
-%to discover useful information that is otherwise unknown.
-The complexity of natural languages introduces interesting challenges 
-to automatic analysis of underlying textual data.
-Morphological analysis is key to many Natural Language Processing Applications.
-General morphological analyzers such as Buckwalter and ElixirFM 
-generate all morphological solutions.
-However, this exhaustive enumeration may not be needed or appropriate for the
-application at hand.
-We propose developing Sarf, an Arabic application-specific morphological analyzer whose functionality can be customized 
-based on the specific needs of the application. 
-%This flexibility has been motivated by 
-%the broad spectrum of text mining applications, each with its specific characteristics and corpora, which
-%necessitate a special treatment.
-Sarf encodes its lexicon into a non-deterministic composition of three finite automata.
+This information is hard to discover manually or is previously unknown.
+This thesis is concerned with Arabic Text Mining where
+%The complexity of natural languages introduces interesting challenges 
+%to automatic analysis of underlying textual data.
+morphological analysis is key.%to many Natural Language Processing Applications.
+General-purpose morphological analyzers such as Buckwalter and ElixirFM exist for the Arabic language and
+generate all possible morphological solutions.
+Many Natural Language Processing applications need only few of these solutions.
+We propose building Sarf, an Arabic application-specific morphological analyzer. 
+Natural Language Processing applications can easily customize Sarf 
+to prune unwanted solutions. This refinement allows applications to run more efficiently. 
+%The flexible API of Sarf enables developers without expertise in morphology to have access rich NLP tools. (TODO: re-write)
+Sarf encodes its lexicon into a non-deterministic composition of three finite transducers.
 It also solves the issue of `run-on' words, supports disambiguation using diacritics, 
 handles multi-word expressions and performs online tokenization.
-In our thesis, we will use Sarf to solve two useful Natural Language Processing applications: 
-(1) the extraction of time entities from Arabic documents, and 
-(2) the automation of the process of hadith authentication.
+We will apply Sarf to  
+(1) extract time entities from Arabic documents,
+(2) extract narrators from hadith documents,
+(3) relate narrators to their biographies in separate documents, and 
+ thus automate the process of hadith authentication.
 
@@
 this stuff -- we simply get more memory and keep it all.''~\cite{Witten:11}
+Even old paper documents are being ported to digital form to facilitate their use. 
 This explosion in available data has motivated the automation of the task 
@@
 
-Clearly, a small percentage of the data available worldwide 
-is stored in databases in structured form; instead, the greater amount of information is available in the 
-form of electronic text files and documents such as news articles, publications, governmental records, etc. ...
-Even old paper documents are being ported to digital form to facilitate their use. 
-Text Mining emerged as an attempt to automate the extraction of interesting, 
+While a small percentage of data %available worldwide 
+is available in structured form, the greater amount is stored in unstructured form such as  
+text files.
+Examples are news articles, publications, governmental records, etc. ...
+{\em Text Mining} emerged as an attempt to automate the extraction of interesting, 
 previously unknown and potentially useful information
 from the huge amounts of text documents~\cite{Witten:08}.
-Information Extraction (IE) from natural languages, however, is challenged
-by the difficulty of getting structured 
-entities from the underlying unstructured textual data.
-
-%Information Extraction (IE) from Natural Language Texts has always been an important 
-%problem in Data mining. What distinguishes text mining from the 
-%other subfields of data mining is mainly due to the challenge of getting structured 
-%entities from the underlying unstructured textual data. Clearly, a small percentage of the data available worldwide 
-%is stored in databases in structured form; instead the greater amount of information is available in the 
-%form of electronic text files and documents such as news articles, publications, governmental records, etc. ...
-%Even old paper documents are being ported to digital form to facilitate their use.
-
-Natural Language Processing (NLP)
-as a field of research is interested in achieving a computational model of the language
-and understanding its underlying meaning~\cite{sharp:01}. It is known that this task
-is very complex and possibly not completely achievable. However, in most applications it 
-is not needed to grasp the whole meaning of the text but to extract some useful entities.
-This Information Extraction(IE) problem can be performed through some statistical approaches 
-and/or partial NLP techniques.
-
-A subproblem of NLP is morphology.
-Morphological analyzers%~\cite{Sughaiyer:04}
-are interested in understanding the internal structure of a word;
-It considers words as being composed of 
+%Information Extraction (IE) from natural languages, however, is challenged
+%by the difficulty of getting structured 
+%entities from the underlying unstructured textual data.
+Natural languages are inherently ambiguous which renders the problem of 
+Information Extraction (IE) from natural language texts harder than 
+its counterpart in structured data.
+
+Natural Language Processing (NLP), one of the hardest fields in Artificial Intelligence,
+is interested in building a computational model of text or speech
+and their underlying meaning~\cite{sharp:01}. 
+%It is known that this task
+%is very complex and possibly not completely achievable. 
+Many applications target entities and relations that can be 
+easily extracted with partial computational models derived from
+statistical and morphological analysis techniques.
+%needed to grasp the whole meaning of the text but to extract some useful entities.
+%This IE problem can be performed through some statistical approaches 
+%and/or partial NLP techniques.
+
+%A subproblem of NLP is morphology.
+Morphological analyzers %~\cite{Sughaiyer:04}
+consider the internal structure of a word and compose it from
 several {\em morphemes}. 
@@
 Morphological analyzers have been extensively used as building blocks 
-in many text mining applications~\cite{Sou07}.  In Arabic, many morphological analyzers exist including:  
+in many text mining applications~\cite{Sou07}.  In Arabic, morphological analyzers such as  
 Buckwalter~\cite{Buckwalter:02},
@@
 MAGEAD~\cite{Habash:05}, 
-and MADA+TOKAN~\cite{Habash:09}.
+and MADA+TOKAN~\cite{Habash:09}
+exist.
+They preprocess text to remove diacritics, short vowels, such as 
+\RL{ha}, \RL{hu}, \RL{hi}, \RL{TaN}, \RL{huN}, \RL{hiN}.~\footnote{In this document, 
+we use the default ArabTeX transliteration style ZDMG.} Diacritics help to disambiguate the
+meaning of a word. For example, \noVocRL{wrd} may mean ``rose"(\RL{ward}), ``came"(\fullvocalize\RL{warada}) 
+or ``and he replied"(\RL{waradda}). We think that using diacritics in the analysis can highly increase accuracy.
+Current analyzers take as input whitespace-delimited tokens. However, in Arabic word boundaries can be more complex.
+Arabic letters can have up to 
+four different forms
+corresponding to their position in a word, i.e. beginning,
+middle, end of word and separate forms. 
+This allows the phrase
+\noTrnoVocRL{il_A\nospace almdrsT}  with no delimiter within
+to be visually recognizable
+as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school).
+This happens because the first word \noVocRL{il_A} ends with
+\noVocRL{_A} a non-connecting letter. We propose to solve this problem, commonly known as the run-on word problem, using online tokenization.
+This also solves the issue of multi-word expressions such as \noVocRL{`abid al-la.tiyf} being analyzed as one unit -- here a compound proper name.
+
 %Arabic however has distinctive characteristics that
-%complicate its analysis; and consequently, much remains to be performed in the field.
-
-In this thesis, we propose to develop Sarf, an application-specific 
+%complicate its analysis. 
+%Arabic text is composed of
+
+In this thesis, we propose to build Sarf, an application-specific 
 morphological analyzer for the Arabic language 
@@
 from the specific characteristics of the application at hand.
-In addition, it supports disambiguation using diacritics and 
-handles `run-on' words. We also propose to use Sarf to solve two interesting .
-text mining applications: time entity extraction and hadith authentication.
-
-\section{Applications of Text Mining}
-\label{sec:Applications}
-
-%Text Mining has gained much interest lately especially due to its wide range of applications.
-%However, these applications have substantially differing corpora (terminology, 
-%sentence structure, etc...). In this section, we present some of the 
-%previously explored applications of text mining.
-%
-%Research in text mining started in the mid 1980s when Swanson 
-%realized that slicing and combining seemingly unrelated medical 
-%articles led to the discovery of new 
-%hypotheses~\cite{JNi06}. 
-%Lately we saw applications of text mining in the advertisement, 
-%political campaigning, and other businesses.
-
-Research in text mining started in the mid 1980s when Swanson 
-realized that slicing and combining seemingly unrelated medical 
-articles led to the discovery of new 
-hypotheses~\cite{JNi06}. 
-Text Mining has gained much interest lately due to its wide range of applications,
-including its use in the areas of advertisement, 
-political campaigning, and other businesses.
-However, these applications have substantially differing corpora (terminology, 
-sentence structure, etc...). In this section, we present some of the 
-previously explored applications of text mining.
-
-\subsection{Sentiment Analysis} 
-	This field -- also known as Opinion Mining' -- attempts classifying the polarity of sentiments of people 
-	or even their emotional state ~\cite{Pang:08}. The inputs for such analyzers include product and movie reviews, 
-	Internet blogs, electronic social networks and possibly newspapers. Early research was conducted on review sites
-	to automate the task classifying whether the review was positive and negative. Consequently, users can get concise helpful
-	sentiment summaries about the product, movie, etc...~\cite{Pang:02, Turney:02}. Such abilities motivated research targeting 
-	the following applications:
-	\subsubsection{Marketing Applications} 
-		The strength of blogs in disseminating opinions is a two-edged sword that can 
-		``make or break a company's reputation''~\cite{hoffman:08a}. Consequently, analyzing customers' opinions as portrayed 
-		on online discussion forums should be detected as soon as possible. This input is vital in the company's policy for its
-		reputation management. It can accordingly take actions to retain current customers, attract new ones and win back former 
-		customers~\cite{hoffman:08b}. Sentiment Mining has also been used to predict customer churns~\cite{marketing:07,marketing:08}
-		to take appropriate actions. It can help also in producing more effective advertisements based on the interests of the market.
-	\subsubsection{Political Applications} 
-		Political experts and analysts can benefit also. Since opinions extracted from textual data 
-		can reveal much even without the need of performing any explicit surveys. According to ~\cite{Gryc:10}, it ``opens up many avenues
-		for more accurate and naturalistic large-scale political analysis.'' Some research has been performed on classifying and predicting
-		political inclinations~\cite{Durant:07, Yu:08}. This field can help political leaders to better understand the views of their people 
-		and possibly perform better campaigns.
-
-\subsection{Academic Applications} 
-	In scientific disciplines where written papers and publications contain technical information, the need to retrieve
-	specific information that a user needs, aids in \textbf{\em developing advanced search engines}. 
-	Text mining plays a great role toward this end through the following:
-	\subsubsection{Text Summarization} 
-		When an application is interested in mining a text for the most important and informative portions,
-		the application is considered to be tackling the field of `Text Summarization'. Summaries may have different forms such as being 
-		a group of representative words, to a coherent abstraction of the text in novel terms and phrases that capture its main meaning. 
-		The main motivation
-		for such a research is to enhance search engines. By capturing the most informative concepts in a document, it aids determining a 
-		better metric for a document's relevance to a search query. It also helps search engines display a shortened preview of the document for returned
-		results. This reduces the overwhelming amount of information a user needs to read before deciding weather the whole document deserves
-		comprehensive study. It is also helpful when the editorial staff of a magazine or newspaper need to present an article in the limited space.
-		Similarly, summarization is vital when the content of 
-		news articles needs to be dramatically shortened to be sent in an SMS form to subscribed PDA users. 
-		Useful surveys of current research on text Summarization can be
-		found in ~\cite{Radev:02,Das:07}.
-	\subsubsection{Text Categorization and Clustering} 
-		In many cases, it is very useful to be able to search for documents by theme, topic, domain, etc...
-		This ability is achieved if documents can be automatically clustered into different categories over which a user can search. It has been also 
-		proposed to cluster documents based on biographical data~\cite{Kostoff:99}. Technically speaking the difference between Categorization and
-		Clustering is that the former organizes documents into predefined categories where in the latter the categories arise from the document content
-		themselves~\cite{sharp:01}. Many approaches exist in this field as described in~\cite{Berry:03, Aas:99, MUNTEANU:05}.
-		\begin{enumerate}
-		\item \textbf{Email Classification:} It is noteworthy to mention that it is text categorization can be used for emails 
-			-- such as urgent, work, friends, spam, etc... Consequently, it can be used for spam detection and email 
-			filtering~\cite{Mock:01,bekkerman:04}.
-		\item \textbf{Parental Control:} It has also been proposed to use categorization for parental control by
-			automatically detecting inappropriate content for their children~\cite{Kontostathis:09}.
-		\end{enumerate}
-	\subsubsection{Entity Extraction:} Named Entity Extraction/Recognition (NER) targets being able to detect phrases that convey some general concept
-		and/or belong to a predefined set of categories such as persons, places, time, etc... It has been argued that having the luxury of searching
-		for entities enhances search abilities instead of being restricted to a list of words appearing in a text~\cite{basis:06}. 
-		It seems that humans think of their
-		queries in terms of entities/concepts and then try translating them to some list of words which are most likely to appear in a document that
-		is relevant to their search query. In general, a user does not ``know all the search terms that will return all instances of the desired
-		information''~\cite{basis:06}. A survey of research in the field is presented in ~\cite{Nadeau:07}.
-
-\subsection{Security Applications} 
-	It has been proposed that text mining of emails can be used for detecting money laundering and 
-	terrorist plans. However the effectiveness of data mining in this context is still debatable~\cite{Jonas:06,Federici:07}.
-
-	We also propose another security application which can be used to be used when investigating in criminal cases:
-	Given two sets of reports consisting of crime scene investigation and suspect interrogation reports, 
-	an officer may query use a specialized text mining application to query for relations between suspects in terms of 
-	criminal action, locality, or third party persons unknown to the officer.
-	It can be envisioned that such an application will return two convicted people who had no direct relation
-	in the reports, but who are related only through a third person unknown to the investigator. 
-	Such a query is very hard to answer manually.
-
-\subsection{Medical Applications} 
-	Medical records and clinical reports are studied to detect
-	patterns of occurrence of certain symptoms and diseases. It attempts relate those patterns with
-	other biological factors such as certain 
-	genes or protein interaction~\cite{blaschke:99, Krallinger:05, Rebholz-Schuhmann:07, Cohen:08, Tsuruoka:08}.
-	They are also used to analyze factors correlating with medical accidents to minimize their occurrence~\cite{Kimura:08}.
-	Till present, most of the interest in text mining comes from biological sciences~\cite{Giannopoulou:08}.
-
-\subsection{Hadith Authentication}
+In addition , Sarf solves the issue of `run-on' words, supports disambiguation using diacritics, 
+handles multi-word expressions and performs online tokenization. 
+We will apply Sarf to  
+(1) extract time entities from Arabic documents,
+(2) extract narrators from hadith documents,
+(3) relate narrators to their biographies in separate documents, and 
+ thus automate the process of hadith authentication.
+
+\subsubsection{Time Entity Extraction}
+
+%One of the applications we propose to build using Sarf is a time entity extractor. This application
+%recognizes time expressions in text document and transforms them to time information of the default format
+%that can be stored in a database and on which computations and analysis can be performed.
+Consider the following phrase: \noVocRL{al-'i_tnayn fy al-sAbi` min ^sbA.t lsanaT 2011} (Monday the seventh of February 2011).
+We are interested in transforming such phrases into a time structure of the form: 05/02/2011 that can be used in further analysis.
+Many applications can benefit from temporal entities extracted from text documents.
+For example, %applications that sort the documents chronologically.
+Li et al. from 
+Microsoft hold a patent for a tool that extracts authorship 
+dates from Latin documents~\cite{patent:09}. We will use Sarf to automate time entity extraction for Arabic.
+
+\subsubsection{Hadith Authentication}
 
@@
 for text mining in the field of Islamic Studies. 
-A \noVocRL{.hady_t}~\footnote{In this document, 
-we use the default ArabTeX transliteration style ZDMG.}
+A \noVocRL{.hady_t}
 is a narration related to the prophet Mohammad
@@
 for authenticating inputted hadith based on inputted biography books.
+
+\section{Other Applications of Text Mining}
+\label{sec:Applications}
+
+%Text Mining has gained much interest lately especially due to its wide range of applications.
+%However, these applications have substantially differing corpora (terminology, 
+%sentence structure, etc...). In this section, we present some of the 
+%previously explored applications of text mining.
+%
+%Research in text mining started in the mid 1980s when Swanson 
+%realized that slicing and combining seemingly unrelated medical 
+%articles led to the discovery of new 
+%hypotheses~\cite{JNi06}. 
+%Lately we saw applications of text mining in the advertisement, 
+%political campaigning, and other businesses.
+
+Research in text mining started in the mid 1980s when Swanson 
+realized that slicing and combining seemingly unrelated medical 
+articles led to the discovery of new 
+hypotheses~\cite{JNi06}. 
+Text Mining has gained much interest lately due to its wide range of applications,
+including its use in the areas of advertisement, 
+political campaigning, and other businesses.
+These applications consider corpora that substantially differ in terminology, 
+sentence structure and morphology. Thus, an application-specific analyzer such as Sarf is of great benefit.
+
+In this section, we present previously explored applications of text mining. %TODO: and focus on their discrimenent morphological features.
+
+\subsection{Sentiment Analysis} 
+	This field -- also known as `Opinion Mining' -- attempts classifying the polarity of sentiments of people 
+	or even their emotional state ~\cite{Pang:08}. The inputs for such analyzers include product and movie reviews, 
+	Internet blogs, electronic social networks and possibly newspapers. Early research was conducted on review sites
+	to automate the task classifying whether the review was positive and negative. Consequently, users can get concise helpful
+	sentiment summaries about the product, movie, etc...~\cite{Pang:02, Turney:02}. Such abilities motivated research targeting 
+	the following applications:
+	\subsubsection{Marketing Applications} 
+		The strength of blogs in disseminating opinions is a two-edged sword that can 
+		``make or break a company's reputation''~\cite{hoffman:08a}. Consequently, analyzing customers' opinions as portrayed 
+		on online discussion forums should be detected as soon as possible. This input is vital in the company's policy for its
+		reputation management. It can accordingly take actions to retain current customers, attract new ones and win back former 
+		customers~\cite{hoffman:08b}. Sentiment Mining has also been used to predict customer churns~\cite{marketing:07,marketing:08}% maybe explain churn
+		to take appropriate actions. It can help also in producing more effective advertisements based on the interests of the market.
+	\subsubsection{Political Applications} 
+		Political experts and analysts can also benefit, since opinions extracted from textual data 
+		can reveal much even without the need of performing any explicit surveys. According to ~\cite{Gryc:10}, it ``opens up many avenues
+		for more accurate and naturalistic large-scale political analysis.'' Some research has been performed on classifying and predicting
+		political inclinations~\cite{Durant:07, Yu:08}. This field can help political leaders to better understand the views of their people 
+		and possibly perform better campaigns.
+
+\subsection{Academic Applications} 
+	In scientific disciplines where written papers and publications contain technical information, the need to retrieve
+	specific information that a user needs, aids in \textbf{\em developing advanced search engines}. 
+	Text mining plays a great role toward this end through the following:
+	\subsubsection{Text Summarization} 
+		When an application is interested in mining a text for the most important and informative portions,
+		the application is considered to be tackling the field of `Text Summarization'. Summaries may have different forms such as being 
+		a group of representative words, or even a coherent abstraction of the text in novel terms and phrases that capture its meaning. 
+		The main motivation
+		for such a research is to enhance search engines. By capturing the most informative concepts in a document, it aids determining a 
+		better metric for a document's relevance to a search query. It also helps search engines display a shortened preview of the document for returned
+		results. This reduces the overwhelming amount of information a user needs to read before deciding whether the whole document deserves
+		comprehensive study. It is also helpful when the editorial staff of a magazine or newspaper need to present a large article in the limited space.
+		Similarly, summarization is vital when the content of 
+		news articles needs to be dramatically shortened to be sent in an SMS form to subscribed PDA users. 
+		Useful surveys of current research on text Summarization can be
+		found in ~\cite{Radev:02,Das:07}.
+	\subsubsection{Text Categorization and Clustering} 
+		In many cases, it is very useful to be able to search for documents by theme, topic, domain, etc...
+		This ability is achieved if documents can be automatically clustered into different categories over which a user can search. It has been also 
+		proposed to cluster documents based on biographical data~\cite{Kostoff:99}. Technically speaking the difference between categorization and
+		clustering is that the former organizes documents into predefined categories where in the latter the categories arise from the document content
+		themselves~\cite{sharp:01}. Many approaches exist in this field as described in~\cite{Berry:03, Aas:99, MUNTEANU:05}.
+		\begin{enumerate}
+		\item \textbf{Email Classification:} It is noteworthy to mention that categorization can be used to classify emails into different subgroups
+			-- such as urgent, work, friends, spam, etc... Consequently, it can be used for spam detection and email 
+			filtering~\cite{Mock:01,bekkerman:04}.
+		\item \textbf{Parental Control:} It has also been proposed to use categorization for parental control by
+			automatically detecting inappropriate content for children~\cite{Kontostathis:09}.
+		\end{enumerate}
+	\subsubsection{Entity Extraction:} Named Entity Extraction/Recognition (NER) targets being able to detect phrases that convey some general concept
+		and/or belong to a predefined set of categories such as persons, places, time, etc... It has been argued that having the luxury of searching
+		for entities enhances search abilities instead of being restricted to search for a list of words appearing in a document~\cite{basis:06}. 
+		It seems that humans think of their
+		queries in terms of entities/concepts and then try translating them to some list of words which are most likely to appear in a document that
+		is relevant to their search query. In general, a user does not ``know all the search terms that will return all instances of the desired
+		information''~\cite{basis:06}. A survey of research in the field is presented in ~\cite{Nadeau:07}.
+
+\subsection{Security Applications} 
+	It has been proposed that text mining of emails can be used for detecting money laundering and 
+	terrorist plans. However the effectiveness of mining techniques in this context is still debatable~\cite{Jonas:06,Federici:07}.
+
+	We also propose another security application which can be used when investigating in criminal cases:
+	Given two sets of reports consisting of crime scene investigation and suspect interrogation reports, 
+	an officer may use a specialized text mining application to query for relations between suspects in terms of 
+	criminal action, locality, or third party persons. %common between suspects
+	It can be envisioned that such an application will return two convicted people who had no direct relation
+	in the reports, but who are related only through a third person unknown to the investigator. 
+	Such a query is very hard to answer manually.
+
+\subsection{Medical Applications} 
+	Medical records and clinical reports are studied to detect
+	patterns of occurrence of certain symptoms and diseases. It attempts to relate those patterns with
+	other biological factors such as certain 
+	genes or protein interaction~\cite{blaschke:99, Krallinger:05, Rebholz-Schuhmann:07, Cohen:08, Tsuruoka:08}.
+	They are also used to analyze factors correlating with medical accidents to minimize their occurrence~\cite{Kimura:08}.
+	Till present, most of the interest in text mining comes from biological sciences~\cite{Giannopoulou:08}.
 
@@
 
-	Morphological ambiguity is a `notorious' problem for Arabic language~\cite{Kiraz:98}.
+	Morphological ambiguity is a `notorious' problem for the Arabic language~\cite{Kiraz:98}.
 	Ambiguity refers to the fact that one word may lead to several possible solutions.
@@
 	may be due to over-approximations and relaxed compatibility rules in the morphological analyzer.} 
-	in the language:
+	in the language.
 	\begin{enumerate}
@@
 	its inclusion of classical terms that are rarely used. By this he means that the score does not reflect the intrinsic Arabic ambiguity and 
-	they report a score of 1.75 for their tool which used tighter grammar and lexical rules and omitted classical words. Although, Dr. Attia
-	has a point relating to exaggerated numbers due to over-approximations, the reduction of the lexicon is not a correct approach since it does
-	not reflect the ambiguity of Arabic but of the subset of Arabic needed for some applications (mainly modern text documents such as news).
+	he reports a score of 1.75 for his tool which used tighter grammar and lexical rules and omitted classical words. Although, Dr. Attia
+	has a point relating to exaggerated numbers due to over-approximations, the omission of some non-frequent words from the lexicon 
+	does not a correct approach for reporting ambiguity since it does
+	not reflect the ambiguity of Arabic but of the subset of Arabic needed for some applications.\footnote{
+	In his case, he focused mainly on modern text documents such as news articles.}
 	In all cases, this is a question open to future investigations by NLP researchers and Arabic linguists...
@@
 	For instance, in many cases, it is impossible to distinguish between active and passive verbs on the word level without
-	appropriate diacritics such as the undiacriticized versions of \RL{`abada} ($<$subject$>$ worshiped) and \RL{`obida} ($<$subject$>$ was worshiped).
+	appropriate diacritics. For instance the undiacriticized versions of \RL{`abada} ($<$subject$>$ worshiped) and \RL{`obida} ($<$subject$>$ was worshiped) are homomorphic.
 
@@
 
-	Arabic letters can have up to 
-	four different forms
-	corresponding to their position in a word, i.e. beginning,
-	middle, end of word and separate forms. 
-	This allows the phrase
-	\noTrnoVocRL{il_A\nospace almdrsT}  with no delimiter within
-	to be visually recognizable
-	as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school).
-	This happens because the first word \noVocRL{il_A} ends with
-	\noVocRL{_A} a non-connecting letter. 
+%	Arabic letters can have up to 
+%	four different forms
+%	corresponding to their position in a word, i.e. beginning,
+%	middle, end of word and separate forms. 
+%	This allows the phrase
+	The example provided in the introduction,
+	\noTrnoVocRL{il_A\nospace almdrsT} provides evidence to the issues 
+	that can arise due to the presence of non-connecting letters. 
 	The non-connecting letters are \RL{A}, \noVocRL{'a}, \noVocRL{'i},
@@
 		\noVocRL{d}, \noVocRL{_d}, \noVocRL{r}, \noVocRL{z} and \noVocRL{T}.
-	These words occur often in text and greatly increase the
+	Words that are not separated by white-space occur often in text and 
+	greatly increase the
 	difficulty of tokenization and are referred to as 
@@
 	analysis complicated'' ~\cite{Attia:08a}. In other words, instead of forming words from stems, the
-	structure of the Arabic language is built from roots. That's why Arabic dictionaries are indexed by the roots.
+	structure of the Arabic language is built from roots. That's why Arabic dictionaries are indexed by the roots of words.
 	The Arabic stems are formed by applying some inflectional patterns that modify the meaning of the word.
@@
 	However, in Latin languages the inflectional forms don't transform the meaning completely and can generally be treated
-	as prefixes and suffixes. In such inflectional languages such as Arabic, concatenative analysis sometimes fails to capture the
-	word structure especially when occur inside the stem itself caused by infixes. In this our thesis, we will not focus on this problem,
+	as prefixes and suffixes. In inflectional languages such as Arabic, concatenative analysis sometimes fails to capture the
+	word structure especially when infixes occurs inside the stem itself. In our thesis, we will not focus on this problem,
 	however we have mentioned it for completeness.
@@
 
-The rest of the proposal is organized as follows. In Section 2, we present the problem(s) to be attempted. 
+The rest of the proposal is organized as follows. In Section 2, we present the problems to be attempted. 
 In Section 3, we review related literature. 
@@
 Arabic poses many difficulties which currently 
-have not been tackled in a satisfactory way. Of these we focus on the issues of `run-on' words and
+have not been tackled in a satisfactory way. Of these, we focus on the issues of `run-on' words and
 the use of diacritics in morphological analysis -- two issues which have been under-looked by researchers in the field.
@@
 enhancing affix representations to better reflect the semantics of underlying word structure.
-Morover, we target supporting for interaction between the different text mining applications 
+Moreover, we target supporting for interaction between the different text mining applications 
 and the morphological analyzer, 
@@
 
-Current Analyzers are unable to benefit from the characteristics of the application in which the Analyzer is used.
-Their output is the same regardless of the needs of the application used and the characteristics of the corpus
+Current analyzers are unable to benefit from the characteristics of the application in which they are used.
+Their output is the same regardless of the needs of the application and the characteristics of the corpus
 being analyzed. We noted in Section~\ref{sec:Applications}, that
@@
 for higher level control by the specific application utilizing the lower level 
-morphological analyzer. In other words, the application-specific controller, has 
-full control over and directs the morphological analysis of Sarf. 
+morphological analyzer. In other words, the application-specific controller has 
+full control in directing the general morphological analysis of Sarf. 
 % (sentence should be reworded differently, but is needed to explain why we mention 
@@
 
-We justify the above enhancement as follows:
+We justify the above enhancement as follows.
 
@@
 In other words, if the analyzer can be controlled at every step by an application-specific
-controller, the analysis can be enhanced and reduced significantly.
+controller, the analysis can be enhanced significantly.
 
@@
 %a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
-It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when the NLP 
+It reports that a refined version of the analyzer was achieved when the NLP 
 task interacted with the morphological analyzer, which is
-a flexibility our analyzer is designed to provide for the application-specific controller on the run.
+a flexibility our analyzer is designed to provide on the run 
+through the application-specific controller.
 
@@
 may present the solutions in different orders. 
-Clearly, some of 
+It can be argued that some of 
 the presented solutions and their order of presentation may not be 
@@
 
-Analyzers provide different levels of features for the alternative solutions.
-Commonly, those features include: the vocalized version corresponding to the analysis,
+Analyzers provide different levels of features for presented solutions.
+Commonly, those features include: the vocalized version of the corresponding analysis,
 its Part of Speech (POS), its gloss (i.e. semantical meaning) and other abstract categories to which
@@
 categorization as proper noun (subdivided into person, place, organization), etc..~\cite{Attia:08a}.}.
-If it is known beforehand by the application if one feature is its primary interest, using which
-early termination of a word analysis can be interpreted, the analyzer can reorder its structures to support for
-faster lookup for this feature.
+If it is known beforehand by the application that some features are of primary interest, using which
+early termination of word analysis can be achieved, the analyzer can reorder its structures to support for
+faster lookup for those feature.
 
@@
 
-We argue also that by introducing an application specific controller that is capable
+We argue also that by introducing an application-specific controller that is capable
 of intervening with the morphological analyzer at every decision, the reported query results 
@@
 to words that exist in the lexicon.\footnote{This application was inspired from the works of~\cite{Badr:95,Obaid:98} that 
-used benefited from the lexicon while performing the OCR; their methods, however, is not totally applicable to our
+utilized the lexicon while performing the OCR; their methods, however, are not applicable to our
 hypothetical application.}
@@
 and \noVocRL{lA} which appear frequently in run-on words and so on... Such statistical refinements, when handled early on during
-the analysis of the subparts of the word through the application-specific controller, are clearly much more affective than 
+the analysis of the subparts of the word through the application-specific controller, are clearly much more effective than 
 a later post-processing step.
@@
 paper documents. OCR is often used to automate this task; however, this task
-results in mistakes that are often escape manual inspection.
+results in mistakes that often escape manual inspection.
 In particular, run-on words are admitted 
@@
 frequency in three corpus of the Arabic Tree Bank on which some experiments were performed. 
-That paper was concluded that ``the problem of run-on words in Arabic calls for
+That paper concluded that ``the problem of run-on words in Arabic calls for
 a reassessment of current tokenization strategies,
@@
 
-It is clear that in current analyzers, the list of affixes is much larger than needed
-due to the explosion of possible valid combinations that result from the most basic 
-prefixes. Note also that in current analyzers, sometimes to add just one prefix to the lexicon, it is required to
-manually generate all suitable combinations with those previously in the dictionary.
-For instance in Table~\ref{t:affixes}, the two syntactically and semantically different but homomorphic prefixes for \RL{ya}
-repeat the same combinations with other prefixes.
-
-Consider the case of the missing prefix from the lexicon of Buckwalter detected by Prof Attia. It is the 
-``clitic question morpheme which can be prefixed to verbs and nouns''(\RL{'a})~\cite{Attia:06b}.
-If we are to modify Buckwalter to support for its insertion that would result in reproducing most of the current
-prefix entries to concatenate to them this possible clitic. Having prefixes been recursive as in Sarf, just a single 
-prefix will be inserted along with some compatibility rules.
-
 As noted in Table~\ref{t:affixes}, many affixes are composed of other affixes. For example,
-the prefix \noTrnoVocRL{wsy} is composed of three other prefixes
-namely \noTrnoVocRL{w}, \noTrnoVocRL{s}, and \noTrnoVocRL{y}.
-The POS tag for \noTrnoVocRL{wsy} is a concatenation
+the prefix \noVocRL{wsy} is composed of three other prefixes
+namely \noVocRL{w}, \noVocRL{s}, and \noVocRL{y}.
+The POS tag for \noVocRL{wsy} is a concatenation
 %of the POS tags of each of the three morphemes. 
@@
 By doing so, the affix representation better reflects the rules
-which govern their formation. It also is able to better capture what each part of the compound prefix
-contributed to the gloss of the word and its POS. For instance, it is much more instructive to analyze
-and classify words based on their atomic affixes related POS and gloss instead of parsing string
-delimited entries.
+which govern their formation. This representation also better captures what each part of the compound prefix
+contributes to the gloss of the word and its POS. For instance, it is much more instructive to analyze
+and classify words based on the POS and gloss of their atomic affixes, instead of parsing string
+delimited entries per compound affixes.
+
+It is clear that in current analyzers, the list of affixes is much larger than needed
+due to the explosion of possible valid combinations that result from composing the atomic 
+affixes. Note also that in current analyzers, sometimes to add just one affix to the lexicon, it is required to
+manually generate all suitable combinations with those previously in the dictionary.
+For instance in Table~\ref{t:affixes}, the two syntactically and semantically different but homomorphic prefixes for \RL{ya}
+repeat the same combinations with other prefixes.
+
+Consider the case of the missing prefix from the lexicon of Buckwalter detected by Dr. Attia: the 
+``clitic question morpheme which can be prefixed to verbs and nouns''(\RL{'a})~\cite{Attia:06b}.
+If we are to modify Buckwalter to support for its insertion, this would result in reproducing most of the current
+prefix entries with the clitic concatenated to them. Since prefixes are defined recursively in Sarf, just a single 
+prefix will be inserted along with some compatibility rules to support for the clitic under consideration.
 
 More than that, the fine scale of POS tagging and word division better suits any higher level attempt of semantical
-parsing at the sentence structure and better aligns with tagged corpora which commonly tag atomic affixes such as clitic 
+parsing at the sentence structure. For instance an Arabic word can make up a sentence where each of its morphemes 
+represents a sentence subpart. This refinement results in better alignment with tagged corpora which commonly tag atomic affixes such as clitic 
 parts which are generally attached as prefixes. 
@@
 morphological analyzers. That has been made possible through many optimizations to the functioning
-of the analysis methods and implementation, of which we enumerate the following:
+of the analysis methods and their implementation, of which we enumerate the following.
 
@@
 	for stem lookup, we used a double array trie (datrie)~\cite{Aoe:89}. We also extensively used
-	maps and hash-tables when needed. Also whenever an array reverse lookup was needed we stored
-	that lookup array for constant time access instead of a linear search in the original array.
+	maps and hash-tables when needed. Also whenever a reverse-array lookup was needed, we stored
+	reverse lookup arrays for constant time access instead of a linear search in the original array.
 	Also when using a bit vector to access a specific bit, binary operations are used instead of
-	multiplication and division...
+	decimal multiplication and division...
 \item \textbf{Reduced analysis and computations when not needed:} As discussed above, the 
 	application-specific controller can enhance performance by some early analysis aborting, and
-	non-complete retrieval of needed information for the analysis. For example, if for the application 
+	non-complete retrieval of analysis features when not needed. For example, if for the application 
 	at hand, only stems are needed, affix information such as gloss and POS need not be retrieved.
-	Many other reductions can be envisioned suitable for particular applications of Sarf. 
+	Many other reductions can be envisioned depending on the characteristics of the
+	specific applications of Sarf. 
 \end{enumerate}
+
 \section{Building a Time Entity Extractor}
@@
 Li et al. working at the 
-Microsoft have been awarded a patent for having devised an application for extracting authorship 
+Microsoft have been filed a patent for having devised an application for extracting authorship 
 dates from documents~\cite{patent:09}. The importance of their work is justified by the fact that 
@@
 	\ref{sec:hadithSegmentation}.
+
 \item \textbf{Building a Partial Order Graph of Narrators:}
@@
 
-\item \textbf{Localization of Narrators in Biograph Books:}
+\item \textbf{Localization of Narrators in Biography Books:}
 
@@
 The main tasks of the thesis are listed below: \\
-\begin{enumerate}[1)]
-\item \textbf{Build and Enhance Sarf}
-	\begin{enumerate}[\ref*{s:intro}.1)]
-	\item Port Buckwalter into C++ and mysql
-	\item Implement the Finite State Automata Framework
-	\item Implement an application-specific controller interface
-	\item Support for disambiguation using diacritics
-	\item Support for multi-word expressions
-	\item Solve the Issue of `run-on' words
-	\item Optimize performance
-	\item Support for Running Sarf Under the Windows Operating System
+\begin{enumerate}
+\item \textbf{Build and Enhance Sarf} \label{I:Sarf}
+	\begin{enumerate}
+	\item Port Buckwalter into C++ and mysql \label{I:portBuckwalter}
+	\item Design and implement the Non-deterministic Finite State Automata framework \label{I:ndfsa}
+	\item Implement an application-specific controller interface \label{I:applicationController}
+	\item Support for disambiguation using diacritics \label{I:diacritics}
+	\item Support for multi-word expressions \label{I:multiword}
+	\item Solve the issue of `run-on' words \label{I:runon}
+	\item Validate Sarf \label{I:validateSarf}
+	\item Benchmark against existing corpora \label{I:benchMark}
+	\end{enumerate}
+\item \textbf{Support Sarf}
+	\begin{enumerate}
+	\item Optimize performance \label{I:optimizeSarf}
+	\item Support for running Sarf under Windows \label{I:Sarfo}
 	\item Refactor the code
@@
 \item \textbf{Build a time entity extraction application}
-\item \textbf{Automate the Hadith Authentication Case Study}
-	\begin{enumerate}[3.1)]
+	\begin{enumerate}
+	\item Recognize time entities
+	\item Normalize extracted time entities
+	\item Validate with a sorting application
+	\end{enumerate}
+\item \textbf{Automate Hadith Authentication}
+	\begin{enumerate}
 	\item Extend the Lexicon of Buckwalter with proper names 
@@
 	\item Segment biography books using a graph coloring algorithm
-	\item Extract Annotations from the Biography books
+	\item Extract Annotations from the biography books
 	\item Assign annotations to graph nodes
-	\item Perform time and possibly location consistency checks
-	\item Perform Experiments on the effects of different narrators in Hadith Literature
-	\item Write a user Manual
-	\item Distribute the tool to willing religious scientists and report their findings
+	\item Enable and perform consistency checks between narrator chains
+	\item Perform experiments on the effects of different narrators in hadith literature
+	\item Write a user manual
+	\item Distribute the tool to volunteer scholars and report their findings
 	\end{enumerate}

### 2011-04-30 trunk/ATSarf/src/case/split_affixes.cpp

@@ -1,0 +1,62 @@
+#include "split_affixes.h"
+
+void SplitDialog::split_action() {
+	QList<QTableWidgetSelectionRange>  selection=originalAffixList->selectedRanges();
+	if (selection.size()==1 && selection[0].rowCount()==1){
+		if (source_id<0)
+			source_id=insert_source("Manual Work by Jad", "Splitting Affixes by an automatic process","Jad Makhlouta");
+		//select the items to be splitted
+		originalAffixList->setSelectionMode(QAbstractItemView::MultiSelection);
+		int rowIndex=selection[0].topRow();
+		//long affix_id=originalAffixList->item(rowIndex,0)->text().toLongLong();
+		QString affix=originalAffixList->item(rowIndex,1)->text();
+		QString category=originalAffixList->item(rowIndex,2)->text();
+		QString raw_data=originalAffixList->item(rowIndex,3)->text();
+		QString pos=originalAffixList->item(rowIndex,4)->text();
+		QString description=originalAffixList->item(rowIndex,5)->text();
+		originalAffixList->clearSelection();
+		for (int i=0;i<originalAffixList->rowCount();i++) {
+			long affix_id1=originalAffixList->item(i,0)->text().toLongLong();
+			QString affix1=originalAffixList->item(i,1)->text();
+			QString category1=originalAffixList->item(i,2)->text();
+			QString raw_data1=originalAffixList->item(i,3)->text();
+			QString pos1=originalAffixList->item(i,4)->text();
+			QString description1=originalAffixList->item(i,5)->text();
+			if (	affix1.startsWith(affix) &&
+					raw_data1.startsWith(raw_data) &&
+					pos1.startsWith(pos) &&
+					description1.startsWith(description)	) {
+				originalAffixList->selectRow(i);
+				//split the item selected
+				item_types t=(item_types)affixType->itemData(affixType->currentIndex()).toInt();
+				QString affix2=affix1.mid(affix.size());
+				QString raw_data2=raw_data1.mid(raw_data.size());
+				QString pos2=pos1.mid(pos.size());
+				QString description2=description1.mid(description.size()+ (t==PREFIX?3:1)); //' + ' or '+'
+				int rowIndex2=getRow(affix2,raw_data2,pos2,description2);
+				if (rowIndex2<0) {
+					if (!affix2.isEmpty())
[…]

### 2011-04-30 trunk/ATSarf/src/util/text_handling.h

@@ -47,7 +47,7 @@ inline bool equal(const QChar & c1, const QChar & c2)
 		return true;
 	return false;
 }
-inline bool equal_ignore_diacritics(const QString &word1,const QString &word2)
+inline bool equal_ignore_diacritics(const QString &word1,const QString &word2)//TODO: solve bug in index boundaries here
 {
 	int length1=word1.length(),length2=word2.length();
 	int i=0,j=0;

### 2011-04-30 ATSarf/src/gui/test.cpp
wrote automatic code for splitting prefixes, not useful yet for suffixes.
Did much of the work and splitted prefixes except for some documented issues due to subject/object modifications and discovered some minor inconsistencies in Buckwalter's work

@@ -12,6 +12,8 @@
 #include "diacritics.h"
 #include <sys/time.h>
 
+extern void splitRecursiveAffixes();
+
 int word_sarf_test(QString input_str)
 {
 	QString line=input_str.split('\n')[0];
@@ -89,6 +91,7 @@ int start(QString input_str, QString &output_str, QString &error_str, QString &h
 
 int test(QString input_str, QString &output_str, QString &error_str, QString & hadith_str, bool /*hadith*/,ATMProgressIFC * /*m_ui*/)
 {
+#if 0
 	out.setString(&output_str);
 	out.setCodec("utf-8");
 	in.setString(&input_str);
@@ -143,6 +146,10 @@ int test(QString input_str, QString &output_str, QString &error_str, QString & h
 	out <<QString("\nSuccessfully processed all %1 %2 entries\n").arg(line_num).arg(file_name);
 	input.close();
 	return 0;
+#else
+	splitRecursiveAffixes();
+	return 0;
+#endif
 }
 
 

### 2011-04-30 ATSarf/src/sql/sql_queries.cpp

@@ -780,7 +780,7 @@ long insert_item(item_types type,QString name, QString raw_data, QString categor
 		raw_data=name;
 	int i=0;
 	QString reducedName, reducedRawData;
-	if (name.length()>0)
+	if (type==STEM && name.length()>0)
 		do
 		{
 			//qDebug()<<"o:"<<name;
@@ -965,6 +965,37 @@ long insert_item(item_types type,QString name, QString raw_data, QString categor
 	perform_query(stmt);
 	update_dates(source_id);
 	return item_id;
+}
+int remove_item(item_types type,long item_id, QString raw_data, long category_id, long description_id, QString POS)
+{
+	QString stmt( QString("DELETE FROM %1_category ")+
+				  "WHERE %1_id=%2 AND category_id=%3 AND raw_data=\"%4\" "+
+						"AND description_id=%5 AND POS=\"%6\"");
+	stmt=stmt.arg(interpret_type(type)).arg(item_id).arg(category_id)
+			 .arg(raw_data).arg(description_id).arg(POS);
+	perform_query(stmt);
+	if (query.numRowsAffected()==0)
+	{
+		error << "No entry removed in "+interpret_type(type)+"_category!\n";
+		return -2;
+	} else if (query.numRowsAffected()>1) {
+		error << "More than one entry in "+interpret_type(type)+"_category removed!\n";
+		return -2;
+	}
+	stmt=QString( "SELECT COUNT(*)")+
+				  "FROM %1_category "+
+				  "WHERE %1_id=%2 AND category_id=%3 AND raw_data=\"%4\" "+
+						"AND description_id=%5 AND POS=\"%6\"";
+	stmt=stmt.arg(interpret_type(type)).arg(item_id).arg(category_id)
+			 .arg(raw_data).arg(description_id).arg(POS);
+	perform_query(stmt);
+	assert(query.next());
+	if (query.record().value(1)==0) {
+		QString stmt( "DELETE FROM %1 WHERE id=%2");
+		stmt=stmt.arg(interpret_type(type)).arg(item_id);
+		perform_query(stmt);
+	}
+	return 0;
 }
 //let dispay table return the number of rows in the table
 long display_table(QString table) //TODO: has some error in producing sources for example may result in "3,0,0" and also in rules type may result in "AA" always

### 2011-06-02 tex/hadith.ijcnlp.tex
added tex file and modified fig

--- 2011-02-25 tex/hadith.short.tex
@@
 \documentclass[11pt]{article}
-\usepackage{acl-hlt2011}
+\usepackage{ijcnlp2011}
 \usepackage{times}
+\usepackage{url}
 \usepackage{latexsym}
@@
 \usepackage{url}
-\DeclareMathOperator*{\argmax}{arg\,max}
-\setlength\titlebox{6.5cm}    % Expanding the titlebox
+%\DeclareMathOperator*{\argmax}{arg\,max}
+%\setlength\titlebox{6.5cm}    % Expanding the titlebox
 
@@
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Hadith Narrator Chain Extraction using Arabic Morphological Analysis}
+\title{Automatic Hadith Graphs using Arabic Morphological Analysis}
 
@@
 \begin{document}
+
 \maketitle
+
 \begin{abstract}
+\end{list}
 A hadith in Islamic literature is a narration from the 
@@
 Establishing the authenticity of a hadith is an important task
-in Islamic studies. 
-Based on that Islamic scholars issue rules that affect the life
+based on which Islamic scholars issue rules that affect the life
 of people around the world. 
-The literature lacks an automated exhaustive mechanism of 
+The literature lacks an automated and exhaustive mechanism of 
 authentication checks. 
@@
 each book into a graph of narrators related to each other. 
-\end{quote}
 \end{abstract}
@@
 \section{Introduction}
-\novocalize
-
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{.9\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
-\label{f:exhadith}
-}
-\end{figure}
-% the FSMs for the three words
-\transtrue
-\vocalize
 
@@
 we use the default ArabTeX transliteration style ZDMG.}
+(tradition) 
 is a narration related to the prophet Mohammad
@@
 
-In this paper, we present the hadith {narrator chain 
-extractor using morphological analysis} (NaCEMA), 
+\novocalize
+
+\transfalse
+\begin{figure}[tb]
+\center{
+\resizebox{.9\columnwidth}{!}
+{ \input{figs/exhadith.pdftex_t}}
+\caption{Hadith abstraction example.}
+\label{f:exhadith}
+}
+\end{figure}
+% the FSMs for the three words
+\transtrue
+\vocalize
+
+In this paper, we present the hadith {narrator graph
+extractor using morphological analysis} (NaGEMA), 
 a novel technique that automates
 extracting the \noArRL{sanad} from the books of traditions 
-into a diagram of chains of narrators. NaCEMA segments
+into a diagram of chains of narrators. NaGEMA segments
 a book into narrations, each narration into its \noArRL{matn} 
@@
 and the word means ``his Ahmad''.
+
+The morphological analysis of the word \RL{.hd_tnA} 
+``narrated to us'' that starts the hadith in Figure~\ref{f:exhadith}
+returns the stem \RL{.hd_t}.
+The stem  \RL{.hd_t} is also the stem of other 
+words such as \RL{.hd_tny} and \RL{.hd_thm}.
+The stem also shares similar part of speech (POS)
+and English meaning gloss annotations with words
+such as \RL{qaal} ``said'' and \RL{'a_hbar} ``told''. 
+The use of morphological analysis allows us to 
+subsume listing these stems and all of their morphological
+derivations with one rule that matches the gloss and 
+POS tags. 
+Similarly the words \RL{ibn} ``the son of'' and \RL{bn} 
+return the same morphological analysis.
 
@@
 to control the machines.
-Sarf builds on and extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} 
+Sarf builds on and extends the lexicon of 
+Buckwalter~\shortcite{Buckwalter:02} 
 with proper and location names extracted from different online 
@@
 %In Section~\ref{sec:related}
-%we compare NaCEMA to related work. 
+%we compare NaGEMA to related work. 
 %We conclude with future work in Section~\ref{sec:future}.
@@
 %the analysis of the hadith literature. 
-In this paper, we use Sarf to successfully automate the
+In this paper, we use Sarf to automate the
 analysis of three books of hadith selected 
@@
 %\setcode{standard}
-NaCEMA accepts a book as input
+NaGEMA accepts a book as input
 and segments it into a vector of hadiths. 
-NaCEMA segments each hadith into its
+NaGEMA segments each hadith into its
 sanad and matn parts. 
-NaCEMA 
+NaGEMA 
 detects the chain of narrators in the sanad and 
@@
 
+To do that NaGEMA queries Sarf for the following morphological 
+categories.
+\begin{itemize}
+\item $\mathit{NAME}$: a proper name of a person such as \RL{'a.hmad} .
+\item $\mathit{IBN}$: a special name connector meaning the son of such as \RL{ibn}.
+\item $\mathit{NMC}$: other name connectors. 
+\item $\mathit{NISBA}$: a possessive adjective that qualifies a person such 
+as \RL{alma.sriy} ``the Egyptian''. 
+\item $\mathit{NRC}$: a narrator connector such as
+\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
+\RL{'a_hbar} ``told'' or one of their morphological equivalents. 
+\end{itemize}
+
+\begin{figure*}
+\center{
+\resizebox{1.9\columnwidth}{!}
+{\includegraphics{figs/detailed_submachine.pdf}
+}
+\caption{The chain of narrators extractor diagram}
+\label{f:chain}
+}
+\end{figure*}
+
+The diagram in Figure~\ref{f:chain} describes 
+the chain extraction process. 
+We pass the hadith to the morphological analyzer 
+and check the returned words and category pairs 
+for name categories.
+Next we compute the number of words between every two names
+and compare that to a threshold $\theta_1$.
+If the name is within the threshold we add it to a temporary 
+chain of names. 
+Otherwise, we 
+{\bf 1.} tabulate the chain into a temporary table containing 
+the list of chains and
+{\bf 2.} Create a new chain with the last element.
+Finally, we filter out those chains whose cardinality 
+is less than some threshold  $\theta_2$.
+
+The process also benefits from name connector 
+words and narrator connector words to enhance the
+accuracy of the chain extraction and to segment the 
+chain accurately into narrators.
+Once we meet a name connector or a narrator 
+connector word we reset the word counters that we
+compare against the thresholds. 
+that directly interacted with Sarf.
+
 \subsection{The hadith controller}
@@
 
+We implemented our hadith chain extraction 
+in an finite state machine
+controller that interfaces with Sarf.
 Intuitively, the controller reports a valid chain 
@@
 connected by narrator connectors appears. 
-This includes the task of finding multi-word names
-such as \RL{`bd alr.hmn} often appearing as ``run-on'' words.
-The controller marks the beginning of the hadith with the beginning of 
+%This includes the task of finding multi-word names
+%such as \RL{`bd alr.hmn} often appearing as ``run-on'' words.
+The controller marks the beginning of the hadith with 
+the beginning of 
 the current proper name sequence,
@@
 in Figure~\ref{f:hadith}. 
-The transitions are labeled by outputs generated from 
-the Sarf morphological analyzer that correspond to 
-part of speech (POS) tags. 
-A NAME label denotes a proper name.
-An NRC label denotes a narrator connector such as
-\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
-\RL{'a_hbar} ``told'' or one of their morphological equivalents. 
-An IBN label denotes the word \RL{ibn} ``the son of'' and is a name connector.
-A NISBA is a possessive adjective that qualifies a person such 
-as \RL{alma.sriy} ``the Egyptian''.
-
-The symbol $\tau_{\mbox{NMC}}$ is a threshold
-that corresponds to the number of tolerated name connectors 
-that may occur between two names. %3
-The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
-of name connectors collected since the controller
-started looping in the state NMC\_S. 
-The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
-that corresponds to a relaxed tolerance measure that
-the controller resorts to in case the words separating
-two names were longer than $\tau_{\mbox{NMC}}$ but 
-contained a name connector word such as IBN or NISBA.
+The transitions are labeled using expressions
+that operate on the output tags 
+generated from the Sarf morphological analyzer.
 
@@
 a position in text relative to the next sanad. 
-State TEXT\_S is the initial state and denotes that
-the controller is outside the context of a sanad.
-The controller moves to state NAME\_S on
-NAME.
-State NRC\_S means that the controller is in the context
-of a narrator name after Sarf reports an NRC.
-State NMC\_S
+State $\mathit{TEXT\_S}$ is the initial state and denotes 
+that the controller is outside the context of a sanad.
+State $\mathit{NAME\_S}$ denotes that the controller is in
+the middle of a chain and has last met a name. 
+State $\mathit{NMC\_S}$ denotes that the controller is in
+the middle of a narrator name.
+State $\mathit{NRC\_S}$ denotes that the controller is 
+in the context of a narrator name and the controller 
+has met a narrator connector and is ready to meet 
+a new narrator.
+
+The threshold $\theta_{\mathit{nmc}}$ 
+corresponds to the number of tolerated name connectors 
+that may occur between two names. %3
+The symbol $\mathit{LIST}_{\mathit{nmc}}$ corresponds to the list 
+of name connectors collected since the controller
+started looping in the state $\mathit{NMC\_S}$.
+The symbol $\lambda_{\mathit{nmc}}$ is a parameter 
+that corresponds to a relaxed tolerance measure that
+the controller resorts to in case the words separating
+two names were longer than $\theta_{\mathit{nmc}}$ but 
+contained a name connector word such as $\mathit{IBN}$ 
+or $\mathit{NISBA}$.
+
+The controller moves to state $\mathit{NAME\_S}$ on
+$\mathit{NAME}$.
+It moves to state $\mathit{NRC\_S}$ 
+when Sarf reports an $\mathit{NRC}$.
+State $\mathit{NMC\_S}$
 indicates that the controller expects a name to appear within 
 a tolerance threshold expressed by 
-$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
+$\theta_{\mathit{nmc}}$ and $\lambda_{\mathit{bmc}}$.
+It returns to $\mathit{NAME\_S}$ when a name is met, 
+loops when still within the threshold, and 
+resets to $\mathit{TEXT\_S}$ when the threshold is exceeded 
+signaling that the last collection of names met do not qualify
+as a narrator or a chain or narrators.
 
-Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
+Similarly, $\mathit{NRC\_S}$ tolerates $\theta_{\mathit{nrc}}$ words 
 before it gives up on its expectations. 
-We reach the NAME\_S state only when a
-valid NAME is detected and we leave when no more NAME's are detected.
-The NRC\_S state can only be reached if an NRC is detected.
-NMC\_S can only be reached from a NAME\_S state.
-We used the values of 3, 5 and 5 for $\tau_{\mbox{NMC}}$, 
-$\lambda_{\mbox{NMC}}$, and $\tau_{\mbox{NRC}}$ respectively.
+Note that we reach the $\mathit{NAME\_S}$
+state only when a
+valid $\mathit{NAME}$ is detected and we leave when no 
+more $\mathit{NAME}$'s are detected.
+The $\mathit{NRC\_S}$ state can only be reached if an 
+$\mathit{NRC}$ is detected.
+The state $\mathit{NMC\_S}$ can only be reached from a 
+$\mathit{NAME\_S}$ state.
+We used the values of 3, 5 and 5 for $\theta_{\mathit{nmc}}$, 
+$\lambda_{\mathit{nmc}}$, and 
+$\theta_{\mathit{nrc}}$ respectively to run the experiments 
+we report in Section~\ref{sec:results}.
 
@@
 \caption{Narrator directed acyclic graph 
-    extracted using NaCEMA }
+    extracted using NaGEMA }
 \label{f:narrators} 
@@
 a small $\delta$. 
-Then we applied a threshold $\tau$ to the sum of rewards
+Then we applied a threshold $\theta$ to the sum of rewards
 and decided the equality of the narrators. 
-We omitted the details of the distance function for brevity.
+%We omitted the details of the distance function for brevity.
+
+\begin{table}[tb]
+%\centering
+%\resizebox{.9\columnwidth}{!}{
+\begin{tabular} {p{6cm}}
+\begin{Verbatim}[fontsize=\relsize{-2},
+frame=topline,framesep=4mm,label=\fbox{MergeChains algorithm},
+commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
+Merge(Chain $C$[])
+  Chain $C_c$[]=\{\}
+  foreach Chain $c_1$ in $C$ 
+    $C_c = C_c \cup c_1$
+    foreach Narrator $n_1$ in $c_1$
+      foreach Chain $c_2$ in $C-C_c$
+        foreach narrator $n_2$ in $c_2$
+
+    \}
+  \}
+\end{Verbatim}
+\end{tabular}
+%}
+\label{a:merge}
+\end{table}
+
+\subsection{Graph merge Algorithm}
+
+\begin{table}[tb]
+%\centering
+%\resizebox{.9\columnwidth}{!}{
+\begin{tabular} {p{6cm}}
+\begin{Verbatim}[fontsize=\relsize{-2},
+frame=topline,framesep=4mm,label=\fbox{Narrator distance metric },
+commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
+Distance(Narrator n1, Narrator n2)
+  $N_1$ = stems(n1.names)
+  $N_2$ = stems(n2.names)
+
+  $I_1$ = extractIbn( stems(n1.connectors) )
+  $I_2$ = extractIbn( stems(n2.connectors) )
+
+  $C_1$ = stems(n1.connectors - $I_1$)
+  $C_2$ = stems(n2.connectors - $I_2$)
+
+  $E_n$ = getEqualNames($N_1$, $N_2$)
+  if ( $E_n$.isEmpty ) 
+      return $\infty$
+  if ( $E_n$.size $\ge$ min($N_1$.size,$N_2$.size))
+    dist -= $\delta_1$
+  if ($E_n$.equalNamesInOrder())  
+    dist -= $\delta_2$
+
+  if ($C_1 \cap$ Places $\not= C_2 \cap$ Places)
+    dist += $\delta_3$
+  else
+    dist -= $\delta_3$
+
+  $E_c$ = getEqualConnectors($C_1$, $C_2$)
+  dist -= $E_c$.size * $\delta_4$
+
+  if (correctIbnCorrespondence($I_1$, $I_2$))
+      dist -= $\delta_5$
+  return (MAX - dist)
+\end{Verbatim}
+\end{tabular}
+%}
+\label{a:distance}
+\end{table}
+
+\subsection{Narrator morphological distance metric}
+
+\subsection{Use of the POR graph}
+
+inconsistency check 
+consistency checks
+annotate with information extracted from biographies
+what if analysis
 
@@
 three hadith books.
-NaCEMA detected narrator chains and segmented the books 
+NaGEMA detected narrator chains and segmented the books 
 correctly with a chain accuracy above 97\% and a narrator name
@@
 and end of a representative sample of hadiths in each 
-book and compared that to the NaCEMA results. 
+book and compared that to the NaGEMA results. 
 We used the same approach to compute all other accuracy measures.
@@
 
-NaCEMA does not need a preprocessing step since 
+NaGEMA does not need a preprocessing step since 
 it uses morphological analysis to process the text. 
@@
 of words that surround the narrator names, 
-NaCEMA bases its analysis on the fact that narrator
+NaGEMA bases its analysis on the fact that narrator
 names are likely to happen in localities with a high
 number of proper names. 
-NaCEMA uses the narrator connectors as helpers and
+NaGEMA uses the narrator connectors as helpers and
 uses morphological analysis to compare the stems of
-those words and thus NaCEMA avoids the problem of 
+those words and thus NaGEMA avoids the problem of 
 ``parser noise words'' that e-Narrator faces. 
@@
 90 traditions e-Narrator reports on, we claim
-that NaCEMA outperforms e-Narrator as it works on
+that NaGEMA outperforms e-Narrator as it works on
 the full text of the presented hadith books
@@
 
-We presented NaCEMA, a narrator chain extractor
+We presented NaGEMA, a narrator chain extractor
 using morphological analysis. 
-NaCEMA extracted narrator chains with above 90\% accuracy
+NaGEMA extracted narrator chains with above 90\% accuracy
 from three different hadith books and generated 

### 2011-06-02 trunk/tex/hadith.ijcnlp.tex
added tables

--- 2011-06-02 tex/hadith.ijcnlp.tex
@@
     $C_c = C_c \cup c_1$
+    foreach Chain $c_2$ in $C-C_c$
       foreach Narrator $n_1$ in $c_1$
-      foreach Chain $c_2$ in $C-C_c$
-        foreach narrator $n_2$ in $c_2$
-
-    \}
-  \}
+        start = max( $n_1$.index - radius, 0)
+        end = min($n_1$.index + radius, $n_2$.size)
+        $n_2$ = $c_2$[start]
+        while (true)
+            if (Distance($n_1$, $n_2$) < $\theta$)
+                mergeNodes($n_1$, $n_2$)
+            if ($n_2$ = $c_2$[end])
+                    break
+            $n_2$ = $c_2$.next($n_2$)        
 \end{Verbatim}
@@
 \subsection{Use of the POR graph}
-
 inconsistency check 
@@
 
-\begin{table}[bt]
+%\begin{table}[bt]
+%\centering
+%\caption{Results of the hadith case study with Sarf.}
+%%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
+%\resizebox{1.1\columnwidth}{!}{
+%\begin{tabular}{lcp{.2cm}cp{.2cm}c} %\cline{2-10}
+% &  AlKafi & & AlIstibsar & &IbnHanbal \\ \cline{1-6}
+%Word count  &98,943 & & 103,835 & & 20,354 \\ 
+% Names       & 12,060  & & 14,613& & 3,013\\
+%%Names/Narrator & 1.97 & & 1.84& & 1.25 \\
+%Narrators  & 2,623 & & 5,767& & 1,755 \\ 
+%%Narrators/Chain  & 4.84 & & 4.76 & &4.05 \\
+%Chains  & 542 &  & 1,211& & 433 \\ 
+%Ignored names  & 6,400 &  & 3,348 & & 642 \\ \hline
+%Segmentation accuracy  & 96\%& & 96\%& & 92\%\\ 
+%Chain accuracy & 99\%&  & 99\%& & 97\% \\ 
+%%Narrator accuracy  & & & & &  & & & \\ 
+%Narrator accuracy  & 91\%& & 90\% & & 90\% \\ \hline
+%%Name false positives  & 7\%&  & 4\% & & 4\% \\ \hline
+%Running time (secs.)& 1.32 & & 1.31 & & .096\\ \hline 
+%\end{tabular}
+%}
+%\normalsize
+%\label{t:hadithresallresults}
+%\end{table}
+%
+
+\begin{table*}[bt]                                     
 \centering                                            
-\caption{Results of the hadith case study with Sarf.}
-%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
-\resizebox{1.1\columnwidth}{!}{
-\begin{tabular}{lcp{.2cm}cp{.2cm}c} %\cline{2-10}
- &  AlKafi & & AlIstibsar & &IbnHanbal \\ \cline{1-6}
-Word count  &98,943 & & 103,835 & & 20,354 \\ 
- Names       & 12,060  & & 14,613& & 3,013\\
-%Names/Narrator & 1.97 & & 1.84& & 1.25 \\
-Narrators  & 2,623 & & 5,767& & 1,755 \\ 
-%Narrators/Chain  & 4.84 & & 4.76 & &4.05 \\
-Chains  & 542 &  & 1,211& & 433 \\ 
-Ignored names  & 6,400 &  & 3,348 & & 642 \\ \hline
-Segmentation accuracy  & 96\%& & 96\%& & 92\%\\ 
-Chain accuracy & 99\%&  & 99\%& & 97\% \\ 
-%Narrator accuracy  & & & & &  & & & \\ 
-Narrator accuracy  & 91\%& & 90\% & & 90\% \\ \hline
-%Name false positives  & 7\%&  & 4\% & & 4\% \\ \hline
-Running time (secs.)& 1.32 & & 1.31 & & .096\\ \hline 
+\caption{Results comparing between NaGEMA and iTree.}
+\resizebox{2.1\columnwidth}{!}{
+\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+%\multicolumn{13}{c}{NaGEMA with all refinements} \\ \cline{3-13}
+& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{NaGEMA }&musnad & & 1.000 & 1.000 & 1.000 & & 0.993 & 0.990 & 0.991 & & 0.999 & 0.998 & 0.999 \\
+&kafi & & 1.000 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.988 & 0.990 \\
+&istibsar & & 1.000 & 0.973 & 0.986 & & 0.980 & 0.950 & 0.965 & & 0.996 & 0.998 & 0.997 \\ \cline{2-14}
+& & & 1.000 & 0.964 & 0.982 & & 0.970 & 0.970 & 0.970 & & 0.996 & 0.995 & 0.995 \\ \hline \hline
+%& & &  &  &  & &  &  &  & &  &  & \\
+%&\multicolumn{13}{c}{iTree} \\ \cline{3-13}
+& & & \multicolumn{3}{c}{accept as hadith} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{iTree} &musnad & & 1.000 & NA & 1.000 & & 0.997 & 0.977 & 0.987 & & 0.998 & 1.000 & 0.999  \\
+&kafi & & 0.043 & NA & 0.083 & & 1.000 & 0.313 & 0.476 & & 1.000 & 1.000 & 1.000 \\
+&istibsar & & 0.028 & NA & 0.054 & & 0.875 & 0.583 & 0.700 & & 1.000 & 1.000 & 1.000 \\ \cline{2-14}
+& & & 0.357 & NA & 0.379 & & 0.957 & 0.624 & 0.721 & & 0.999 & 1.000 & 1.000 \\ 
  \end{tabular}                 
@@
 \normalsize                   
-\label{t:hadithresallresults}
-\end{table}
+\label{t:hadithresults} 
+\end{table*}
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Results without punctuation and without learning. }
+\resizebox{2.1\columnwidth}{!}{
+\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{no punctuation} 
+& musnad  & & 1.000 & 1.000 & 1.000 & & 0.982 & 0.955 & 0.968 & & 1.000 & 0.998 & 0.999 \\ 
+& kafi & & 1.000 & 0.793 & 0.885 & & 0.944 & 0.967 & 0.955 & & 0.996 & 0.938 & 0.966 \\
+& istibsar & & 1.000 & 1.000 & 1.000 & & 0.980 & 0.931 & 0.955 & & 0.994 & 0.985 & 0.990 \\  \cline{2-14}
+& &  & 1.000 & 0.931 & 0.962 & & 0.969 & 0.951 & 0.960 & & 0.997 & 0.974 & 0.985 \\ \hline \hline
+%& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{no learning}
+& musnad & & 1.000 & 1.000 & 1.000 & & 0.894 & 0.985 & 0.937 & & 0.995 & 1.000 & 0.998\\
+& kafi & & 1.000 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.991 & 0.992\\
+& istibsar & & 1.000 & 0.973 & 0.986 & & 0.977 & 0.950 & 0.963 & & 0.996 & 0.998 & 0.997\\ \cline{2-14}
+ & &  & 1.000 & 0.964 & 0.982 & & 0.936 & 0.968 & 0.951 & & 0.995 & 0.996 & 0.995 \\
+ \end{tabular}                 
+}                             
+\normalsize                   
+\label{t:nopunctlearnresults} 
+\end{table*}
 

### 2011-06-02 tex/hadith.ijcnlp.tex
merged and added small part to results section

@@
 \normalsize                   
-\label{t:hadithresallresults}
-\end{table}
+\label{t:hadithresults} 
+\end{table*}
+
+To evaluate our approach, we performed experiments on three books of 
+hadith~\cite{IbnHanbal,AlTousi,AlKulayni} whose accuracy results are detailed in 
+Tables~\ref{t:hadithresults} and \ref{t:nopunctlearnresults}. We used recall and precision as our evaluation metrics.
+We report results on three levels of accuracy: segmentation, sanad boundary and
+narrator boundary. Segmentation accuracy refers to the success of NaGEMA in
+detecting in the plain text of a hadith book the locations of the different
+narrations that it contains. While the sanad boundary is a measure of the accuracy
+in reporting the correct number of narrators in the sanad. Finally, the narrator boundary refers
+to the accuracy in reporting the correct number of words inside each narrator.
+
+Precisely defined, segmentation recall refers to the fraction of extracted narrations from 
+those which are correct; and segmentation precision refers to the fraction of correct narrations
+from those extracted. Similarly, sanad boundary recall is a measure whether we detected all narrators
+in the sand; and precision is a measure whether we did so without introducing much false positives.
+
+iTree does not tackle the problem hadith segmentation. In contrast to NaGEMA, it does not work on a book level, 
+instead it assumes that the knowledge of what part of the text of a hadith book is a narration is available.
+Consequently, for the narrators to be extracted, the hadith must be manually extracted
+from the book and inputted on its own to the tool. Still however, in many cases it does not recognize the structure of
+the inputted hadith and is incapable of extracting any useful information from the narration. 
+In table~\ref{t:hadithresults}, we report the fraction of narrations which iTree recognizes from those inputted, instead
+of the segmentation measure used for NaGEMA. This measure is equivalent to the concept of recall; nonetheless, precision
+in this context is not applicable since by default the input is required to be represent correctly extracted narrations.
+
+Now we move to explain the significance of the experiments performed.
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Results without punctuation and without learning. }
+\resizebox{2.1\columnwidth}{!}{
+\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{no punctuation} 
+& musnad  & & 1.000 & 1.000 & 1.000 & & 0.982 & 0.955 & 0.968 & & 1.000 & 0.998 & 0.999 \\ 
+& kafi & & 1.000 & 0.793 & 0.885 & & 0.944 & 0.967 & 0.955 & & 0.996 & 0.938 & 0.966 \\
+& istibsar & & 1.000 & 1.000 & 1.000 & & 0.980 & 0.931 & 0.955 & & 0.994 & 0.985 & 0.990 \\  \cline{2-14}
+& &  & 1.000 & 0.931 & 0.962 & & 0.969 & 0.951 & 0.960 & & 0.997 & 0.974 & 0.985 \\ \hline \hline
+%& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{no learning}
+& musnad & & 1.000 & 1.000 & 1.000 & & 0.894 & 0.985 & 0.937 & & 0.995 & 1.000 & 0.998\\
+& kafi & & 1.000 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.991 & 0.992\\
+& istibsar & & 1.000 & 0.973 & 0.986 & & 0.977 & 0.950 & 0.963 & & 0.996 & 0.998 & 0.997\\ \cline{2-14}
+ & &  & 1.000 & 0.964 & 0.982 & & 0.936 & 0.968 & 0.951 & & 0.995 & 0.996 & 0.995 \\
+ \end{tabular}                 
+}                             
+\normalsize                   
+\label{t:nopunctlearnresults} 
+\end{table*}
 

### 2011-06-02 tex/hadith.ijcnlp.tex
results section draft written

@@
 \newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue} 
+\newcommand{\noTrnoVocRL}[1]{\transfalse\novocalize\noTrRL{#1}\vocalize\transtrue}  
+\newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}
 
@@
 hadith~\cite{IbnHanbal,AlTousi,AlKulayni} whose accuracy results are detailed in 
-Tables~\ref{t:hadithresults} and \ref{t:nopunctlearnresults}. We used recall and precision as our evaluation metrics.
+Tables~\ref{t:hadithresults} and \ref{t:nopunctlearnresults}. 
+The numbers we present in the tables are the result of a manual verification of the correctness of the outputs of the first
+10\% of the narrations in the books under consideration.
+
+We used recall and precision as our evaluation metrics.
 We report results on three levels of accuracy: segmentation, sanad boundary and
@@
 narrations that it contains. While the sanad boundary is a measure of the accuracy
-in reporting the correct number of narrators in the sanad. Finally, the narrator boundary refers
-to the accuracy in reporting the correct number of words inside each narrator.
+in correctly reporting the individual narrators inside the sanad. Finally, the narrator boundary refers
+to the accuracy in correctly reporting the words that constitute the narrators.
 
-Precisely defined, segmentation recall refers to the fraction of extracted narrations from 
-those which are correct; and segmentation precision refers to the fraction of correct narrations
+Precisely defined, segmentation recall refers to the fraction of the narrations correctly detected to the
+total number of narrations available; while segmentation precision refers to the fraction of correctly detected narrations
 from those extracted. Similarly, sanad boundary recall is a measure whether we detected all narrators
-in the sand; and precision is a measure whether we did so without introducing much false positives.
+in the sand; while precision is a measure whether we did so without introducing much false positives. 
+The same concepts apply to the level of narrator boundary where the count is of the number of words which constitute a valid narrator.
 
 iTree does not tackle the problem hadith segmentation. In contrast to NaGEMA, it does not work on a book level, 
-instead it assumes that the knowledge of what part of the text of a hadith book is a narration is available.
-Consequently, for the narrators to be extracted, the hadith must be manually extracted
+instead it assumes that the knowledge of which part of the text of a hadith book is a narration is available.
+Consequently, for the narrators to be extracted from the sanad of a narration, the hadith must be manually extracted
 from the book and inputted on its own to the tool. Still however, in many cases it does not recognize the structure of
@@
 
-Now we move to explain the significance of the experiments performed.
+In those experiments, we were interested in evaluating the approaches used by iTree and NaGEMA, without any bias that results
+from the difference in the coverage of the lexicons used. For this reason, we appended the list of names used by iTree with those
+available in Sarf's lexicon on top of which NaGEMA operates.
+
+Now we move to explain the significance of the experiments performed. First, we compare the recall of the segmentation 
+in NaGEMA and that of hadith acceptance rates in iTree. Although the task of segmentation is more complex than that of
+hadith recognition/acceptance, we still notice that iTree has on the average 35.7\% recall rate compared to 99.9\% for 
+NaGEMA. However, we notice that the difference in these numbers depended highly on the books under consideration. For instance,
+for the book of musnad, both approaches correctly detected  practically all narrations under consideration; 
+however iTree suffers a lot in 
+the books of both Kafi and Istibsar , resulting in a recall of less than 5\%. This is due to the less structured nature of the narrations
+those books contain. For example, iTree's grammar is defined such that a hadith starts by an Ikhbar word (\noVocRL{AdaT al-A_hbAar}) such as
+\noVocRL{qAl}, \noVocRL{.hadda_t}, etc...; 
+however, those books had a more flexible style of expressing the narration 
+and most of them started directly by narrator name without a preceeding Ikhbar word. This is a strong evidence of the supperiority of 
+our tolerance-based state-machine compared to the strict grammar approach used in iTree. 
+Clearly, our machine does not fail on such cases, since what it detects as a hadith is based on a concentration of names intervened by narrator 
+connectors. Similarly, iTree had a lower sanad boundary precision (62.4\% on the average) due to the fact that the its grammar has no method
+of detecting that the hadith ended except if it reaches a prophet indicative phrase such as \noVocRL{rasowl al-lah}, 
+\noVocRL{.sal_A `alayh al-lah wasalam}, etc.. However, especially in the books of Kafi and Istibsar, the prophet is not explicitly mentioned but 
+implicitly understood. In those cases, iTree will continue reporting narrators whenever it reaches a word that can act as a name even if it appears
+in the matn of the hadith. On the other hand, NaGEMA considers that a sanad ends whenever a tolerance state is broken even if it does not reach a prophet
+indicative phrase. Narrator boundary accuracy was similar however for both approaches because they relied on the same name lexicons.
 
@@
 
-We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
-with 4GB of memory using a Linux operating system. 
-We report our results in Table~\ref{t:hadithresallresults} on the 
-three hadith books.
-NaGEMA detected narrator chains and segmented the books 
-correctly with a chain accuracy above 97\% and a narrator name
-accuracy above 90\%
-in all cases and across all the 1986 traditions in the three 
-different books with different formats.
+After having provided evidence to the superiority of our approach compared to that of iTree, 
+we move to inspect the effect of two of the optimizations we implemented in NaGEMA.
+Those optimizations are: utilizing punctuation in disambiguating narration detection and narrator boundaries, 
+and learning names of narrators based on their local context.
+In Table~\ref{t:nopunctlearnresults}, we report the results of running NaGEMA, 
+disabling each of the optimizations at a time.
 
-The row ``word count'' is an indicator of the size of the data.
-The %numbers in 
-rows ``names'', ``narrators'', and ``chains'' denote the total
-number of the corresponding entities per book.
-The row ``ignored names'' reports the number of names detected 
-outside the chains and thus are irrelevant to
-the application. 
-%In addition, the rows ``names/narrator'' and ``narrators/chain'' 
-%report the average
-%number of names per narrator and 
-%the average number of narrators per chain respectively.
+We notice that punctuations are useful for increasing all precision measures: hadith segmentation, sanad boundary and
+narrator boundary. Those measures are enhanced by about 3\%, 2\%, 1\% respectively as punctuation related information is 
+utilized.  Clearly, punctuation can help eliminate any confusion that may result from a concentration of names and words that are
+can be interpreted shallowly as narrator connectors. It also helps cleaning the sanad of a narration from names that coincidently appear at the 
+end of the matn of a previous narration. If close enough those names can be assumed as narrators in the sanad at hand; however, punctuation marks
+reduce such confusions.
 
-The ``segmentation accuracy'' reports on the accuracy of 
-segmentation of the book into separate hadiths.
-We computed the metric manually by marking the beginning 
-and end of a representative sample of hadiths in each 
-book and compared that to the NaGEMA results. 
-We used the same approach to compute all other accuracy measures.
-The ``chain accuracy'' reflects the percentage of correct narrators 
-covered by extracted chains.
-The ``narrator name accuracy'' row reflects the percentage of 
-correct proper names detected within
-narrator structures.
-
-%The row ``name false positives'' reports the percentage of words 
-%recognized as names while they should not.
-%We notice that this metric did not affect the accuracy
-%of narrator and chain segmentation since detecting
-%a narrator or a chain with the refined analysis without substantially affecting the 
-%high abstract measures such as chain and segmentation
-%accuracy.
+Concerning the ability to learn names, our results show that it enhances mainly the recall of sanad boundaries by about 3.4\% on the average.
+Such an increased accuracy although small is not insignificant, since the task of hadith authentication requires a high accuracy. In particular,
+sometimes the presence of a specific narrator in the chain can render a hadith as not trustworthy since the narrator is not considered credible.
+That's why it is important not to lose any of the narrators that appear in the sanad.
 
@@
 
-The closest to our work is e-Narrator~\cite{Azmi2010}, a 
+The closest to our work is e-Narrator~\cite{eNarrator}, a 
 tool that uses a grammar based approach to solve 
@@
 Then e-Narrator uses a similarity-based approach
-to memory based learning~\cite{Azmi2010} in order to perform 
+to memory based learning~\cite{eNarrator} in order to perform 
 shallow parsing of the preprocessed text. 

### 2011-06-16 trunk/tex/tp.tex
corrected a bug resulting in not detecting آب
also added some words to lexicon of time.

@@
 \noVocRL{_A} a non-connecting letter. We propose to solve this problem, commonly known as the run-on word problem, using online tokenization.
-This also solves the issue of multi-word expressions such as \noVocRL{`abid al-la.tiyf} being analyzed as one unit -- here a compound proper name.
+Online tokenization also solves the issue of multi-word expressions such as \noVocRL{`abid al-la.tiyf} and 
+allows them to be analyzed as one unit -- here a compound proper name.
 
@@
 
-%Similar to %the 
-%MADA+TOKAN, Sarf uses
-%the several valid morphological analyses as 
-%%a richness that should 
-%richness at a higher abstract level, and does not view them as 
-%inaccuracies to be corrected. 
-Sarf differs from MADA+TOKAN in that it provides an interface for the 
-application to interfere in judging the analyses much earlier and
-at a finer granularity level. 
-%In Sarf, there is no need for a separate tokenizer such as
-Sarf does not need a separate tokenizer such as
-TOKAN since each solution keeps a stack of positions
-that partition text into morphemes.
-%The application-specific controller can build the trace 
-%when needed.
+%%Similar to %the 
+%%MADA+TOKAN, Sarf uses
+%%the several valid morphological analyses as 
+%%%a richness that should 
+%%richness at a higher abstract level, and does not view them as 
+%%inaccuracies to be corrected. 
+%Sarf differs from MADA+TOKAN in that it provides an interface for the 
+%application to interfere in judging the analyses much earlier and
+%at a finer granularity level. 
+%%In Sarf, there is no need for a separate tokenizer such as
+%Sarf does not need a separate tokenizer such as
+%TOKAN since each solution keeps a stack of positions
+%that partition text into morphemes.
+%%The application-specific controller can build the trace 
+%%when needed.
 
@@
 The main tasks of the thesis are listed below: \\
-\begin{enumerate}
-\item \textbf{Build and Enhance Sarf} \label{I:Sarf}
-	\begin{enumerate}
+\begin{enumerate}[1.]
+\item \textbf{Build and Enhance Sarf} \label{I:BuildSarf}
+	\begin{enumerate}[\ref*{I:BuildSarf}.1.]
 	\item Port Buckwalter into C++ and mysql \label{I:portBuckwalter}
@@
 	\end{enumerate}
-\item \textbf{Support Sarf}
-	\begin{enumerate}
+\item \textbf{Support Sarf} \label{I:SupportSarf}
+	\begin{enumerate}[\ref*{I:SupportSarf}.1.]
 	\item Optimize performance \label{I:optimizeSarf}
-	\item Support for running Sarf under Windows \label{I:Sarfo}
-	\item Refactor the code
-	\item Write a user Manual
-	\item Write a developer manual and documentation
+	\item Support for running Sarf under Windows \label{I:SarfonWindows}
+	\item Refactor the code \label{I:RefactorSarf}
+	\item Write a user Manual \label{I:SarfUserManual}
+	\item Write a developer manual and documentation \label{I:SarfDeveloperManual}
 	\end{enumerate}
-\item \textbf{Build a time entity extraction application}
-	\begin{enumerate}
-	\item Recognize time entities
-	\item Normalize extracted time entities
-	\item Validate with a sorting application
+\item \textbf{Build a time entity extraction application} \label{I:TimeEntityExtractor}
+	\begin{enumerate}[\ref*{I:TimeEntityExtractor}.1.]
+	\item Recognize time entities \label{I:TimeEntityRecognition}
+	\item Normalize extracted time entities \label{I:TimeEntityNormalization}
+	\item Validate with a sorting application \label{I:TimeEntityValidation}
 	\end{enumerate}
-\item \textbf{Automate Hadith Authentication}
-	\begin{enumerate}
-	\item Extend the Lexicon of Buckwalter with proper names 
-	\item Implement the hadith segmentation tool
-	\item Write the Narrator Equality Function
-	\item Build Parial Order Graphs
-	\item Correct Graph mistakes that result from Ambiguous narrator names which introduce false cycles to the graph
-	\item Segment biography books using a graph coloring algorithm
-	\item Extract Annotations from the biography books
-	\item Assign annotations to graph nodes
-	\item Enable and perform consistency checks between narrator chains
-	\item Perform experiments on the effects of different narrators in hadith literature
-	\item Write a user manual
-	\item Distribute the tool to volunteer scholars and report their findings
+\item \textbf{Automate Hadith Authentication} \label{I:Hadith}
+	\begin{enumerate}[\ref*{I:Hadith}.1.]
+	\item Extend the Lexicon of Buckwalter with proper names \label{I:HadithNames}
+	\item Implement the hadith segmentation tool \label{I:HadithSegmentation}
+	\item Write the Narrator Equality Function \label{I:HadithEquality}
+	\item Build Parial Order Graphs \label{I:HadithPOR}
+	\item Correct Graph mistakes that result from Ambiguous narrator names which introduce false cycles to the graph \label{I:HadithBreakCycles}
+	\item Segment biography books using a graph coloring algorithm \label{I:BiographySegmentation}
+	\item Extract Annotations from the biography books \label{I:BiographyAnnotationExtraction}
+	\item Assign annotations to graph nodes \label{I:HadithPORAnnotation}
+	\item Enable and perform consistency checks between narrator chains \label{I:HadithChecks}
+	\item Perform experiments on the effects of different narrators in hadith literature \label{I:HadithExperiments}
+	\item Write a user manual \label{I:HadithManual}
+	\item Distribute the tool to volunteer scholars and report their findings \label{I:HadithScolars}
 	\end{enumerate}
-\item \textbf{Write thesis Report and defend the thesis}
+\item \textbf{Write thesis Report and defend the thesis} \label{I:WriteReport}
 \end{enumerate}
@@
 			 \\ \toprule[0.1em] \bottomrule[0.1em]
-\textbf{1.1)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{1.2)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
-\textbf{1.3)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{1.4)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~  & ~ &\\ \hline
-\textbf{1.5)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{1.6)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~  & ~ &\\ \hline
-\textbf{1.7)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{1.8)} & \x & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
-\textbf{1.9)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & ~ & \\ \hline
-\textbf{1.10)}& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X  & ~ & \\ \hline
-\textbf{1.11)}& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \toprule[0.1em] \bottomrule[0.1em]
-\textbf{2)}   & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \toprule[0.1em] \bottomrule[0.1em]
-\textbf{3.1)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.2)} & \x & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.3)} & \x & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.4)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.5)} & ~ & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.6)} & ~ & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.7)} & ~ & ~ & ~ & ~ & ~ & \X & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.8)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.9)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & \X & ~ & ~ & ~ & \X \\ \hline
-\textbf{3.10)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & ~ & ~ & \X \\ \hline
-\textbf{3.11)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \hline
-\textbf{3.12)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X \\ \toprule[0.1em] \bottomrule[0.1em]
-\textbf{4)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X &  \\ \hline
+\textbf{\ref{I:portBuckwalter}} 				& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:ndfsa}} 							& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
+\textbf{\ref{I:applicationController}} 			& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:diacritics}}						& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
+\textbf{\ref{I:multiword}} 						& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:runon}} 							& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:benchMark}} 						& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
+\textbf{\ref{I:validateSarf}} 					& ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \toprule[0.1em] \bottomrule[0.1em]
+\textbf{\ref{I:optimizeSarf}} 					& \x & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\X & ~ &\X &\\ \hline
+\textbf{\ref{I:SarfonWindows}} 					& \x & ~ &\X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:RefactorSarf}} 					& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X &\X & ~ & \\ \hline
+\textbf{\ref{I:SarfUserManual}}					& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \hline
+\textbf{\ref{I:SarfDeveloperManual}}			& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \toprule[0.1em] \bottomrule[0.1em]
+\textbf{\ref{I:TimeEntityRecognition}}  		& \x &\X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:TimeEntityNormalization}}		& ~ & \X &\X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline 
+\textbf{\ref{I:TimeEntityValidation}}			& ~ & ~ &\X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \toprule[0.1em] \bottomrule[0.1em]
+\textbf{\ref{I:HadithNames}} 					& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline 
+\textbf{\ref{I:HadithSegmentation}} 			& \x & ~ & ~ & ~ &\X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:HadithEquality}} 				& \x & ~ & ~ & ~ &\X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:HadithPOR}} 						& \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:HadithBreakCycles}} 				& \x & ~ & ~ & ~ & ~ &\X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:BiographySegmentation}} 			& ~ & ~ & ~ & ~ & ~ &\X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:BiographyAnnotationExtraction}} 	& ~ & ~ & ~ & ~ & ~ & ~ &\X &\X & \X & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:HadithPORAnnotation}} 			& ~ & ~ & ~ & ~ & ~ & ~ & ~ &\X & \X & ~ & ~ & ~ & ~ & ~ & \\ \hline
+\textbf{\ref{I:HadithChecks}} 					& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\X &\X & \X & ~ & ~ & ~ & \X \\ \hline
+\textbf{\ref{I:HadithExperiments}} 				& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\X & \X & ~ & ~ & \X \\ \hline
+\textbf{\ref{I:HadithManual}} 					& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \hline
+\textbf{\ref{I:HadithScolars}} 					& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~  & ~ & \X \\ \toprule[0.1em] \bottomrule[0.1em]
+\textbf{\ref{I:WriteReport}} 					& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X &  \\ \hline
 \end{tabular}

### 2011-06-16 trunk/ATSarf/src/case/graph.h

@@ -622,7 +622,7 @@ private:
 						if (eq_val>=threshold)
 						{
 						#ifdef DEBUG_BUILDGRAPH
-							out<<n1_ref.getString()<<"["<<i<<","<<n1->getIndex()<<"]Versus["<<n2_ref.getString()<<"["<<k<<","<<n2->getIndex()<<"]\n";
+							out<<n1_ref.getString()<<"["<<i<<","<<n1->getIndex()<<"]Versus["<<n2_ref.getString()<<"["<<k<<","<<n2->getIndex()<<"]";
 						#endif
 							mergeNodes(*n1,*n2);
 							offset=u;	//this is matched, we must skip it in search for match for next node in c1

### 2011-06-16 trunk/ATSarf/src/case/hadith_general.cpp

@@ -46,6 +46,11 @@ QString StopwordsFileName=".stop_words";
 	enum stateType { TEXT_S , NAME_S, NMC_S , NRC_S, STOP_WORD_S};
 	QStringList rasoul_words;
 #endif
+//#define COUNT_RUNON
+#ifdef COUNT_RUNON
+	bool runon;
+#endif
+
 	typedef struct  {
 		long lastEndPos;
 		long startPos;
@@ -268,7 +273,7 @@ void hadith_initialize()
 	long abstract_COMPOUND_NAMES=database_info.comp_rules->getAbstractCategoryID("Compound Names");
 	int bit_COMPOUND_NAMES=database_info.comp_rules->getAbstractCategoryBitIndex(abstract_COMPOUND_NAMES);
 	bits_NAME.append(bit_COMPOUND_NAMES);
-	long abstract_NOUN_PROP=database_info.comp_rules->getAbstractCategoryID("NOUN_PROP");
+	long abstract_NOUN_PROP=database_info.comp_rules->getAbstractCategoryID("Female Names");//"NOUN_PROP"
 	bit_NOUN_PROP=database_info.comp_rules->getAbstractCategoryBitIndex(abstract_NOUN_PROP);
 #ifdef ADD_ENARRATOR_NAMES
 	long abstract_ENARRATOR_NAMES=database_info.comp_rules->getAbstractCategoryID("eNarrator Names");
@@ -567,6 +572,10 @@ public:
 						)
 					#endif
 				{
+				#ifdef REFINEMENTS
+					if (removeDiacritics(Stem->info.getString()).count()<3) //bit_NOUN_PROP==bits_NAME[i] &&
+						continue;
+				#endif
 					name=true;
 					if (info.finish>finish_pos)
 					{
@@ -578,6 +587,13 @@ public:
 							learnedName=false;
 						finishStem=Stem->info.finish;
 						startStem=Stem->info.start;
+					#ifdef COUNT_RUNON
+						runon=false;
+						if (equal(Stem->info.getString(),stem_info->raw_data)) {
+							if (runon)
+								error<<Stem->info.getString()<<"-"<<stem_info->raw_data<<"\n";
+						}
+					#endif
 					#endif
 					#ifdef GET_WAW
 						checkForWaw();
@@ -697,6 +713,11 @@ bool getNextState(StateInfo &  stateInfo,chainData *currentChain)
 			display("<ending Punctuation>");
 		}
 	}
[…]

### 2011-06-16 trunk/ATSarf/src/case/narrator_abstraction.cpp

@@ -86,6 +86,7 @@ void Narrator::serialize(QDataStream &chainOut) const{
 	chainOut<<getType((const Narrator*)this);
 	chainOut<<isRasoul;
 	int size=m_narrator.size();
+	//assert(size>0);
 	chainOut<<(qint32)size;
 	for (int i=0;i<size;i++)
 	{
@@ -500,7 +501,7 @@ inline double getdistance(const Narrator & n1,const Narrator & n2) //TODO: use p
 	display(n1.getString());
 	display(" VS ");
 	display(n2.getString());
-	display(">\n");
+	display("> ");
 #if 0
 	for (int i=0;i<Names1.count();i++)
 		display(Names1[i].getString()+" - ");
@@ -540,6 +541,7 @@ inline double getdistance(const Narrator & n1,const Narrator & n2) //TODO: use p
 		return dist;
 	if (equal_names.count()==min(Names1.count(),Names2.count()))
 	{
+		display(QString("%1,%2,%3").arg(equal_names.count()).arg(Names1.count()).arg(Names2.count()));
 		display("equal names \\ ");//may be wrong will be refined later when we used EQUALITY_REFINEMENTS
 		dist-=delta;
 	}
@@ -646,7 +648,7 @@ inline double getdistance(const Narrator & n1,const Narrator & n2) //TODO: use p
 	}
 #ifdef EQUALITY_REFINEMENTS
 	int min_names=min(Names1.count(),Names2.count()),
-		min_ibn_cnt=min(num_ibn1,num_ibn2);
+	min_ibn_cnt=min(num_ibn1,num_ibn2);
 	bool one_ibn_first=((!first_ibn1 && first_ibn2)|| (!first_ibn2 && first_ibn1));
 	bool correctly_skipped=one_ibn_first && ((first_ibn1 &&  num_ibn1==min_ibn_cnt && Names1.count()==min_names) ||
 						   (first_ibn2 &&  num_ibn2==min_ibn_cnt && Names2.count()==min_names));
@@ -692,7 +694,9 @@ inline double getdistance(const Narrator & n1,const Narrator & n2) //TODO: use p
 }
 double equal(const Narrator & n1,const  Narrator  & n2)
 {
-	return max_distance - getdistance(n1,n2);
+	double val= max_distance - getdistance(n1,n2);
+	//out<<n1.getString()<<"{"<<val<<"}"<<n2.getString()<<"\n";
+	return val;
 }
 
 double Narrator::equals(const Narrator & rhs) const

### 2011-06-16 trunk/ATSarf/src/case/timeRecognizer.h

@@ -3,7 +3,7 @@
 
 #include "ATMProgressIFC.h"
 
-//#define TIMEDEBUG
+#define TIMEDEBUG
 #define GET_AFFIXES_ALSO
 
 class TimeParameters
@@ -14,7 +14,7 @@ public:
 	unsigned int unused_int:22;
 	TimeParameters() {
 		num_before_absolute=1;
-		num_after_absolute =2;
+		num_after_absolute =1;
 	}
 };
 

### 2011-06-16 trunk/ATSarf/src/common/atmTrie.cpp

@@ -223,7 +223,7 @@ const StemNode *
 ATTrie :: getData(ATTrie::Position pos)
 {
 	int index=getData(pos,true);
-	if (index>0)
+	if (index>=0)
 	{
 		return &(database_info.trie_nodes->at(index));
 	}

### 2011-06-16 trunk/ATSarf/src/sarf/stemmer.cpp

@@ -37,7 +37,7 @@ bool StemMachine::onMatch()
 	if (controller->called_everything)
 	{
 	#ifdef DEBUG
-		qDebug()<<"s:"<<info.text->mid(info.start,currentMatchPos-info.start+1)<<"\n";
+		out<<"s:"<<info.text->mid(info.start,currentMatchPos-info.start+1)<<"\n";
 	#endif
 		if (controller->Suffix!=NULL)
 			delete controller->Suffix;//TODO: change this allocation, deallocation to clear

### 2011-06-16 trunk/ATSarf/src/sql/sql_queries.cpp

@@ -783,7 +783,7 @@ long insert_item(item_types type,QString name, QString raw_data, QString categor
 	if (raw_data.length()<name.length())
 		raw_data=name;
 	int i=0;
-	QString reducedName;
+	//QString reducedName;
 	if (type==STEM && name.length()>0)
 		do
 		{
@@ -799,11 +799,11 @@ long insert_item(item_types type,QString name, QString raw_data, QString categor
 					name[i]=corresponding_part[0];
 				}
 			}
-			if (i>0 && isNonConnectingLetter(name[i-1]))
+			/*if (i>0 && isNonConnectingLetter(name[i-1]))
 			{
 				reducedName=name.remove(i,1);
 				//TODO: remove also from raw_data then later call insert_item again
-			}
+			}*/
 			//qDebug()<<"n:"<<name;
 		}while ((i=name.indexOf(' ',i)+1)>0 && i<name.length());
 #endif

### 2011-06-16 trunk/ATSarf/src/util/textParsing.h

@@ -73,6 +73,10 @@ inline long getLastLetter_IN_previousWord(QString * text,long start_letter_curre
 
 inline long getLastLetter_IN_currentWord(QString * text,long start_letter_current_word) {
 	int size=text->length();
+#if 0
+	if (!isDelimiter(text->at(start_letter_current_word)))
+		start_letter_current_word++;
+#endif
 	while(start_letter_current_word<size)
 	{
 		if(!isDelimiter(text->at(start_letter_current_word)))

### 2011-06-16 trunk/ATSarf/src/util/text_handling.h

@@ -57,7 +57,7 @@ inline bool equal_ignore_diacritics(const QString &word1,const QString &word2)//
 			i++;
 		if (isDiacritic(word2[j]))
 			j++;
-		if (i<length1 && j<length2 && word1[i]!=word2[j]) {
+		if (i<length1 && j<length2 && !equal(word1[i],word2[j])) {
 		#ifdef ENABLE_RUNON_WORD_INSIDE_COMPOUND_WORD
 			if (word1[i]==' ') {
 				i++;
@@ -95,7 +95,8 @@ inline bool equal(const QStringRef &word1,const QStringRef &word2) // is diacrit
 	int i1,i2;
 	int length1=word1.length(),
 		length2=word2.length();
-	checkIfSmallestIsPrefixOfLargest(word1,word2,i1,i2);
+	if (!checkIfSmallestIsPrefixOfLargest(word1,word2,i1,i2))
+		return false;
 	if (length1-(i1+1)<=0) {
 		for (int i=i2+1;i<length2;i++)
 			if (!isDiacritic(word2.at(i)) && !isPunctuationMark(word2.at(i)))

### 2011-06-16 ATSarf/src/case/hadith.h
corrected a bug resulting in not detecting آب
also added some words to lexicon of time.

@@ -24,15 +24,15 @@
 //#define HADITHDEBUG  //just for additional debugging statements displayed
 #define PUNCTUATION
 #define TRYTOLEARN
-//#define ADD_ENARRATOR_NAMES
+#define ADD_ENARRATOR_NAMES
 #define IBN_START
 #define REFINEMENTS
 #define PREPROCESS_DESCRIPTIONS
 #define DISPLAY_HADITH_OVERVIEW
 #define CHAIN_BUILDING
-//#define TEST_NARRATOR_GRAPH
+#define TEST_NARRATOR_GRAPH
 #define EQUALITY_REFINEMENTS
-//#define EQUALITYDEBUG
+#define EQUALITYDEBUG
 //#define COMPARE_TO_BUCKWALTER
 //#define COMPARE_WITHOUT_ABSCAT
 #define PROGRESSBAR

### 2011-06-16 ATSarf/src/util/text_handling.cpp

@@ -4,7 +4,7 @@
 #include <QChar>
 #include <QList>
 //#include <QVector>
-//#include <QDebug>
+#include <QDebug>
 
 /* (rules for possessive)
 alef 				+consonant before			-wey					sayda --> saydawey		or		7alba --> 7albawey
@@ -18,6 +18,10 @@ Notes:
 -remove ta3reef if it is there, except for almanya, albanya, ...
 -asma2 el mourakaba according to last word
   */
+//#define COUNT_RUNON
+#ifdef COUNT_RUNON
+	extern bool runon;
+#endif
 
 inline bool equal_strict(QList<QChar> & list1,QList<QChar> & list2)
 {
@@ -77,10 +81,16 @@ bool checkIfSmallestIsPrefixOfLargest(const QStringRef &word1,const QStringRef &
 			#ifdef ENABLE_RUNON_WORD_INSIDE_COMPOUND_WORD
 				if (letter1==' ') {
 					i2--;//bc will be incremented later so this in effect only moves i1 and keeps i2 in its place
+				#ifdef COUNT_RUNON
+					runon=true;
+				#endif
 					continue;
 				}
 				if (letter2==' ') {
 					i1--;//bc will be incremented later so this in effect only moves i2 and keeps i1 in its place
+				#ifdef COUNT_RUNON
+					runon=true;
+				#endif
 					continue;
 				}
 			#endif

### 2011-06-22 ATSarf/src/case/hadith.h
some refactoring and correcting a bug in the rasoul words
also committing the tex file which had not been previously been so

@@ -2,8 +2,6 @@
 #define HADITH_H
 
 #include <QString>
-#include <QStringList>
-#include <QHash>
 #include "common.h"
 #include "ATMProgressIFC.h"
 
@@ -12,7 +10,7 @@
 
 //#define DEBUG_BFS_TRAVERSAL
 //#define DEBUG_DFS_TRAVERSAL
-//#define DEBUG_BUILDGRAPH
+#define DEBUG_BUILDGRAPH
 #define GENERAL_HADITH
 //#define FORCE_RANKS
 #define DISPLAY_GRAPHNODES_CONTENT
@@ -24,7 +22,7 @@
 //#define HADITHDEBUG  //just for additional debugging statements displayed
 #define PUNCTUATION
 #define TRYTOLEARN
-#define ADD_ENARRATOR_NAMES
+//#define ADD_ENARRATOR_NAMES
 #define IBN_START
 #define REFINEMENTS
 #define PREPROCESS_DESCRIPTIONS
@@ -32,7 +30,7 @@
 #define CHAIN_BUILDING
 #define TEST_NARRATOR_GRAPH
 #define EQUALITY_REFINEMENTS
-#define EQUALITYDEBUG
+//#define EQUALITYDEBUG
 //#define COMPARE_TO_BUCKWALTER
 //#define COMPARE_WITHOUT_ABSCAT
 #define PROGRESSBAR
@@ -43,47 +41,10 @@
 //#define BUCKWALTER_INTERFACE
 //#define OPTIMIZED_BUCKWALTER_TEST_CASE
 //#define IMAN_CODE
-
-#ifdef PREPROCESS_DESCRIPTIONS
-extern QHash<long,bool> IBN_descriptions;
-#endif
-
-#ifdef REFINEMENTS
-extern QStringList rasoul_words,suffixNames;
-extern QString alrasoul,abyi;
-#endif
-
-extern int bit_POSSESSIVE, bit_PLACE,bit_CITY,bit_COUNTRY;
+//#define COUNT_RUNON
 
 int hadithHelper(QString input_str,ATMProgressIFC *prgs);
 void hadith_initialize();
 
-class HadithParameters
-{
-public:
-	unsigned int nmc_max:5;
-	unsigned int nrc_max:5;
-	unsigned int narr_min:5;
-	unsigned int equality_radius:5;
-	bool display_chain_num:1;
-	bool break_cycles:1;
-	unsigned int unused_int:10;
-	double equality_delta;
[…]

### 2011-06-22 ATSarf/src/case/hadithCommon.h

@@ -1,0 +1,545 @@
+#ifndef HADITHCOMMON_H
+#define HADITHCOMMON_H
+
+#include <assert.h>
+#include <QFile>
+#include <QRegExp>
+#include <QString>
+#include <QStringList>
+
+#include "hadith.h"
+#include "stemmer.h"
+#include "letters.h"
+#include "diacritics.h"
+#include "text_handling.h"
+#include "textParsing.h"
+#include "common.h"
+#include "Retrieve_Template.h"
+#include "Search_by_item_locally.h"
+#include "narrator_abstraction.h"
+
+
+
+	class HadithParameters
+	{
+	public:
+		unsigned int nmc_max:5;
+		unsigned int nrc_max:5;
+		unsigned int narr_min:5;
+		unsigned int equality_radius:5;
+		bool display_chain_num:1;
+		bool break_cycles:1;
+		unsigned int unused_int:10;
+		double equality_delta;
+		double equality_threshold;
+		HadithParameters()
+		{
+			nmc_max=3;
+			narr_min=3;
+			nrc_max=5;
+			equality_radius=3;
+			equality_delta=0.4;
+			equality_threshold=1;
+			display_chain_num=true;
+			break_cycles=true;
+		}
+	};
+
+	extern HadithParameters hadithParameters;
+	extern QString chainDataStreamFileName;
+
+	extern QString chainDataStreamFileName;
+#ifdef PREPROCESS_DESCRIPTIONS
+	extern QString preProcessedDescriptionsFileName;
+#endif
+	extern QString PhrasesFileName;
+	extern QString StopwordsFileName;
+
+#ifndef REFINEMENTS
+	enum wordType { NAME, NRC,NMC};
+	enum stateType { TEXT_S , NAME_S, NMC_S , NRC_S};
+#else
+	#define GET_WAW
+
+	enum wordType { NAME, NRC,NMC,STOP_WORD};
+	enum stateType { TEXT_S , NAME_S, NMC_S , NRC_S, STOP_WORD_S};
+#endif
+
+#ifdef COUNT_RUNON
+	extern bool runon;
+#endif
+
+	typedef struct  {
+		long lastEndPos;
+		long startPos;
+		long endPos;
+		long nextPos;
+		wordType currentType:3;
+		stateType currentState:4;
+		stateType nextState:4;
+		bool ibn:1;
+		bool _3abid:1;
+		bool possessivePlace:1;
+		bool number:1;
[…]

### 2011-06-22 ATSarf/src/case/hadith_utilities.h

@@ -1,7 +1,7 @@
 #ifndef HADITH_UTILITIES_H
 #define HADITH_UTILITIES_H
 
-#include "hadith.h"
+#include "hadithCommon.h"
 
 #ifdef REFINEMENTS
 #include <QString>

### 2011-06-22 ATSarf/src/case/narratordetector.cpp

@@ -1,0 +1,5 @@
+#include "narratordetector.h"
+
+NarratorDetector::NarratorDetector()
+{
+}

### 2011-06-22 ATSarf/src/case/narratordetector.h

@@ -1,0 +1,10 @@
+#ifndef NARRATORDETECTOR_H
+#define NARRATORDETECTOR_H
+
+class NarratorDetector
+{
+public:
+    NarratorDetector();
+};
+
+#endif // NARRATORDETECTOR_H

### 2011-06-22 ATSarf/src/gui/mainwindow.cpp

@@ -5,7 +5,7 @@
 #include "sql_queries.h"
 #include "ui_mainwindow.h"
 #include "database_info_block.h"
-#include "hadith.h"
+#include "hadithCommon.h"
 #include "stemmer.h"
 #include "timeRecognizer.h"
 #include <QFileDialog>

### 2011-06-22 trunk/tex/hadith.ijcnlp.tex
some refactoring and correcting a bug in the rasoul words
also committing the tex file which had not been previously been so

@@
 In this paper, we present a tool that uses 
-an Arabic application-specific morphological analyzer 
+an Arabic morphological analyzer 
 to successfully automate the
@@
 
-In this paper, we present the hadith {narrator graph
+In this paper, we present the hadith {\em narrator graph
 extractor using morphological analysis} (NaGEMA), 
@@
 {\bf Sarf.~~~}
-In this paper, 
-we use Sarf~\footnote{Sarf (name modified for blind review) }
+Sarf~\footnote{Sarf (name modified for blind review) }
+is 
 an {\em application-specific efficient
-morphological analyzer} that uses 
+morphological analyzer} based on
 non-deterministic finite state Automata 
@@
 morphological solutions.
+
+In this paper we make the following contributions. 
+\begin{itemize}
+\item We automate the segmentation of hadith books into separate narrations and extract chains of narrators into a partial order relation graph using a knowledge based approach with morphological analysis.
+\item We introduce a specialized distance metric for Arabic names used in Arabic literature. 
+\item We provide a working and efficient implementation of
+our work. 
+\end{itemize}
+
+The rest of this paper is organized as follows. 
+We describe our approach to narrator chain extraction
+in Section~\ref{sec:islamic}.
+We describe how we build the partial order relation
+graph in Section~\ref{sec:graph}.
+We discuss related work in Section~\ref{sec:related}.
+We present our results in Section~\ref{sec:results} and 
+we conclude in Section~\ref{sec:future}.
 
@@
 
+\begin{figure*}
+\center{
+\resizebox{1.9\columnwidth}{!}
+{\includegraphics{figs/detailed_submachine.pdf}
+}
+\caption{The chain of narrators extractor diagram}
+\label{f:chain}
+}
+\end{figure*}
+
 %Several researchers~\cite{Hadithopaedia:08} attempted to automate 
 %the analysis of the hadith literature. 
-In this paper, we use Sarf to automate the
+NaGEMA uses Sarf to automate the
 analysis of three books of hadith selected 
@@
 \item $\mathit{IBN}$: a special name connector meaning the son of such as \RL{ibn}.
-\item $\mathit{NMC}$: other name connectors. 
+\item $\mathit{NMC}$: other name connectors that may be any word appearing amongst the other interesting categories. 
 \item $\mathit{NISBA}$: a possessive adjective that qualifies a person such 
@@
 \end{itemize}
-
-\begin{figure*}
-\center{
-\resizebox{1.9\columnwidth}{!}
-{\includegraphics{figs/detailed_submachine.pdf}
-}
-\caption{The chain of narrators extractor diagram}
-\label{f:chain}
-}
-\end{figure*}
 
@@
 compare against the thresholds.
-that directly interacted with Sarf.
 
@@
 We implemented our hadith chain extraction 
-in an finite state machine
+in a finite state machine
 controller that interfaces with Sarf.
@@
 signaling that the last collection of names met do not qualify
-as a narrator or a chain or narrators.
+as a narrator or a chain of narrators.
 
@@
 
+\subsection{Optimizations}
+
+In addition to the controller, 
+NaGEMA uses punctuation marks to refine its analysis and 
+increase the accuracy of name detection, narration detection and narrator boundaries. 
+It also learns narrator names from simple patterns that 
+happen within narrator connectors and that are not 
+reported by Sarf as names. 
+NaGEMA learns names of narrators that consist of words missing from 
+the Sarf lexicon.
+NaGEMA learns the narrator names based on their context as they occurred 
+between narrator connectors or contained name connectors.
+
 \section{The narrator graph}
+\label{sec:graph}
 
@@
 The directed acyclic graph (DAG) 
-in Figure~\ref{f:narrators} shows a partial order relation (POR) 
-between narrators.
+in Figure~\ref{f:narrators} shows a partial order 
+relation (POR) between narrators.
 We automatically extracted the POR from the three books of 
@@
 that merged equal names in the chains.
-We compared the narrators together using a morphological
-distance function. 
-We considered each narrator to have several morphological 
-components and we rewarded every morphological match with 
-a small $\delta$. 
-Then we applied a threshold $\theta$ to the sum of rewards
-and decided the equality of the narrators. 
-%We omitted the details of the distance function for brevity.
+
+\subsection{Motivation}
+
+The hadith POR is instrumental to automate
+localization of narrators in biography documents and
+to segment biography documents in future work.
+Consider a biography that discusses a narrator.
+It mentions his professors and students.
+We plan to segment the biographies with a graph 
+coloring algorithm
+that parses the text looking for names. 
+Once a name is found it locates it in the POR graph
+and colors the POR node corresponding
+to it.
+The narrator predecessors in the POR are supposed to be
+his professors and the narrator successors in the POR 
+are supposed to be his students. 
+Consequently, whenever a name moves away from the 
+colored local cluster in the POR graph, 
+we know that we are within a new biography.
+
+Once the biographies are analyzed, we can annotate
+the POR with qualifiers of the narrators that reflect
+their authenticity. 
+We can also annotate the POR with the locations and 
+the time the narrators lived in. 
+Then we can check for time and location overlap
+to discover inconsistent narrations.
+This motivates our interest to automatically generate 
+the POR graph.
+
+\subsection{Graph merge Algorithm}
 
@@
 \begin{Verbatim}[fontsize=\relsize{-2},
-frame=topline,framesep=4mm,label=\fbox{MergeChains algorithm},
+frame=topline,framesep=4mm,label=\fbox{Merge Algorithm},
 commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
@@
 %}
-\label{a:merge}
+\label{t:merge}
 \end{table}
 
-\subsection{Graph merge Algorithm}
+The \CodeIn{Merge} Algorithm 
+iterates over the chains of narrators looking 
+for equivalent narrators and merges the equivalent
+narrators.
+The \CodeIn{radius} parameter refines the search 
+to only consider narrators who belong to the same 
+generation. 
+For example, a narrator relating directly to the prophet
+is definitely not equivalent to a narrator separated
+from the prophet by ten or more narrators.
 
@@
 
-  $I_1$ = extractIbn( stems(n1.connectors) )
-  $I_2$ = extractIbn( stems(n2.connectors) )
-
-  $C_1$ = stems(n1.connectors - $I_1$)
-  $C_2$ = stems(n2.connectors - $I_2$)
-
   $E_n$ = getEqualNames($N_1$, $N_2$)
@@
 
+  $I_1$ = extractIbn( stems(n1.connectors) )
+  $I_2$ = extractIbn( stems(n2.connectors) )
+  $C_1$ = stems(n1.connectors - $I_1$)
+  $C_2$ = stems(n2.connectors - $I_2$)
   if ($C_1 \cap$ Places $\not= C_2 \cap$ Places)
@@
 %}
-\label{a:distance}
 \end{table}
@@
 
-\subsection{Use of the POR graph}
-inconsistency check 
-consistency checks
-annotate with information extracted from biographies
-what if analysis
+The \CodeIn{Distance} Algorithm 
+takes as input two narrators and returns 
+their morphological and order distance
+from each other. 
+The lower the distance is, the more
+equivalent the narrators are.
+
+Each narrator has several names and several name and 
+narrator connectors. 
+\CodeIn{Distance} computes the names of the two narrators
+and stores them in $N_1$ and $N_2$ 
+and checks whether they are morphologically equivalent. 
+If they are not then the distance is set to $\infty$. 
+Otherwise the distance is decremented by $\delta_1$. 
+Then \CodeIn{Distance} checks whether the morphologically 
+equivalent names appear in the same order and 
+decrements the distance if so.
+
+\CodeIn{Distance}  pays attention to the $\mathit{NISBA}$ 
+words that relate to places and checks on that to make sure
+that the two narrators considered equivalent are actually 
+from the same place. 
+It also checks whether one place such as Baghdad 
+is geographically 
+contained in the other such as Iraq 
+(we skip that in the Algorithm for clarity). 
+\CodeIn{Distance} also checks for the equivalency of the
+name connectors and the correspondence of the \RL{ibn}
+connectors.
+
+The distance metric of NaGEMA performs better than       
+existing metrics~\cite{Azmi-2010} as it can detect that  
+\RL{`bd al-lah bn `mr} and \RL{ibn `mr} are the same person
+while other distance metrics based on simple
+string similarity techniques can not.
+
+\section{Related work }
+\label{sec:related}
+
+\transfalse
+\novocalize
+\begin{figure*}[tb]
+\center{
+\resizebox{1.8\columnwidth}{!}
+%{ \input{iTreeMohamada}}
+{\includegraphics{iTreeMohamada}}
+\caption{Output of iTree fails to detect \RL{m_hmdA}.}
+\label{f:mohamada}
+}
+\end{figure*}
+\transtrue
+
+The closest work to NaGEMA is 
+iTree~\cite{Azmi-2010,iTree}, a 
+tool that uses a grammar based approach to solve 
+the narrator extraction problem. 
+The iTree tool preprocesses the text to remove 
+punctuation and diacritics and to normalize white
+spaces. 
+Then iTree uses a similarity-based approach
+to memory based learning~\cite{Azmi-2010} 
+in order to perform 
+shallow parsing of the preprocessed text. 
+The iTree tool uses a parser based on a context
+free grammar that enumerates several ways to express
+the name of the prophet, several possible words that 
+might precede the name of the prophet, several possible
+words that might separate two narrators, and several
+words that might separate parts of the name of 
+the narrators.
+The iTree tool fails when a morphological variation of 
+one of the 
+listed words and not an exact form of it
+happens in the text. 
+For example, Figure~\ref{f:mohamada} extracted 
+using iTree fails to detect \RL{m_hmdA} as a name of the
+prophet and does not stop at it when computing the chain 
+of the hadith
+\transfalse
+\setcode{utf8}
+%\begin{arabtext}
+\RL{
+حدثنا وكيع، حدثني سعيد بن السائب، عن داود بن أبي عاصم الثقفي، قال
+سألت ابن عمر عن الصلاة، بمنى فقال هل سمعت {\bf محمدا}، صلى الله عليه وسلم قلت
+نعم وآمنت فاهتديت به قال فإنه كان يصلي بمنى ركعتين.
+%\end{arabtext}
+}
+\setcode{standard}
+\transtrue
+\vocalize
+
+%In their grammar they consider a name to be any sequence 
+%of 
+%legal Arabic characters.
+
+NaGEMA does not need a prepossessing step since 
+it uses morphological analysis to process the text. 
+While iTree bases its analysis on the patterns 
+of words that surround the narrator names, 
+NaGEMA bases its analysis on the fact that narrator
+names are likely to happen in localities with a high
+number of proper names. 
+NaGEMA uses the narrator connectors as helpers and
+uses morphological analysis to compare the stems of
+those words and thus NaGEMA avoids the problem of 
+``parser noise words'' that iTree faces.
+
+The iTree tool extracts successfully 86.7 percent of
+the narration chains of 90 selected traditions with 
+34 simple cases and 56 hard cases. 
+We ran NaGEMA against the five hadith that ship with
+the tool and the two hadith that are listed in the iTree
+paper and NaGEMA was able to extract all narration
+chains with no mistakes even on the examples where iTree
+reports a ``parser noise'' problem.
+
+As we do not have access to the selected
+90 traditions iTree reports on, we claim
+that NaGEMA outperforms iTree as it works on
+the full text of the presented hadith books
+and reports a higher success rate.
+
+Our approach to detect and relate names to verbs
+is similar to a local grammar based 
+approach~\cite{Traboulsi:09} that detects names by
+associating them with verbs 
+such as \RL{qaal} ``said'' and \RL{'a_hbar} ``told''. 
+The local grammars in~\cite{Traboulsi:09} are also
+built into transducers that generate the names and
+the learning phase benefits from morphological analysis. 
+We differ in that we do not use a local grammar, rather
+our tolerance thresholds allow a flexible pattern that is
+not easy to capture in a typical regular expression.
+
+Another attempt to Arabic named entity recognition
+uses an n-gram approach and maximum 
+entropy~\cite{ANERSys:07}. 
+The authors use their own corpora and gazetteers to boost 
+and train the system and ignore POS tags. 
+They achieve major improvements when they use lexical and
+morphosyntactic features and a parallel Arabic and English 
+corpora to bootstrap noisy features~\cite{Benajiba:2010}.
 
@@
 & & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{NaGEMA }&musnad & & 1.000 & 1.000 & 1.000 & & 0.993 & 0.990 & 0.991 & & 0.999 & 0.998 & 0.999 \\
-&kafi & & 1.000 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.988 & 0.990 \\
-&istibsar & & 1.000 & 0.973 & 0.986 & & 0.980 & 0.950 & 0.965 & & 0.996 & 0.998 & 0.997 \\ \cline{2-14}
-& & & 1.000 & 0.964 & 0.982 & & 0.970 & 0.970 & 0.970 & & 0.996 & 0.995 & 0.995 \\ \hline \hline
+\multirow{3}{*}{NaGEMA }&musnad & & 0.999 & 0.999 & 0.999 & & 0.993 & 0.990 & 0.991 & & 0.999 & 0.998 & 0.999 \\
+&kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.988 & 0.990 \\
+&istibsar & & 0.999 & 0.973 & 0.986 & & 0.980 & 0.950 & 0.965 & & 0.996 & 0.998 & 0.997 \\ \cline{2-14}
+& & & 0.999 & 0.964 & 0.982 & & 0.970 & 0.970 & 0.970 & & 0.996 & 0.995 & 0.995 \\ \hline \hline
 %& & &  &  &  & &  &  &  & &  &  & \\
@@
 & & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{iTree} &musnad & & 1.000 & NA & 1.000 & & 0.997 & 0.977 & 0.987 & & 0.998 & 1.000 & 0.999  \\
-&kafi & & 0.043 & NA & 0.083 & & 1.000 & 0.313 & 0.476 & & 1.000 & 1.000 & 1.000 \\
-&istibsar & & 0.028 & NA & 0.054 & & 0.875 & 0.583 & 0.700 & & 1.000 & 1.000 & 1.000 \\ \cline{2-14}
-& & & 0.357 & NA & 0.379 & & 0.957 & 0.624 & 0.721 & & 0.999 & 1.000 & 1.000 \\ 
+\multirow{3}{*}{iTree} &musnad & & 0.999 & NA & 0.999 & & 0.997 & 0.977 & 0.987 & & 0.998 & 0.999 & 0.999  \\
+&kafi & & 0.043 & NA & 0.083 & & 0.999 & 0.313 & 0.476 & & 0.999 & 0.999 & 0.999 \\
+&istibsar & & 0.028 & NA & 0.054 & & 0.875 & 0.583 & 0.700 & & 0.999 & 0.999 & 0.999 \\ \cline{2-14}
+& & & 0.357 & NA & 0.379 & & 0.957 & 0.624 & 0.721 & & 0.999 & 0.999 & 0.999 \\ 
  \end{tabular}                 
@@
 \end{table*}
+
+To evaluate our approach, we ran NaGEMA against three books of 
+hadith~\cite{IbnHanbal,AlTousi,AlKulayni} and compared the
+results to iTree. 
+The results are detailed in 
+Tables~\ref{t:hadithresults} and \ref{t:nopunctlearnresults}. 
+The numbers we present in the tables are the result of a 
+manual verification of the correctness of the outputs of 
+10\% of the narrations in the books under consideration.
+
+We used recall and precision as our evaluation metrics.
+We report results on three levels of accuracy: segmentation, sanad boundary and
+narrator boundary. 
+Segmentation accuracy refers to the success of NaGEMA in
+detecting in the plain text of a hadith book the locations of the different
+narrations therein.
+The sanad boundary accuracy measures the
+correctness of individual narrators inside the sanad. 
+Finally, the narrator boundary accuracy measures
+the correctness of narrator full names.
+
+Precisely defined, segmentation recall refers to the ratio of the narrations correctly detected against the
+total number of narrations present. 
+Segmentation precision refers to the fraction of correctly detected narration chains compared to all the extracted
+narration chains. 
+Similarly, sanad boundary recall measures whether we detected all narrators
+in the sanad; while sanad boundary precision measures whether we did so without introducing false positives. 
+The same concept applies to the narrator boundary where we count the number of words constituting 
+a valid narrator.
+
+\subsection{NaGEMA versus iTree}
+
+iTree does not tackle the problem of hadith segmentation. 
+In contrast to NaGEMA, it does not work on a book level, 
+instead it takes as input an isolated hadith. 
+It assumes the hadith book is already segmented.
+
+Consequently, for iTree to extract the narrators from the sanad of a narration, 
+the user has to extract and pass the hadith manually.
+Still in many cases iTree fails to recognize the structure of
+the hadith and does not extract any useful information from the narration. 
+This happens for example when a slight variation of the iTree grammar rules 
+happens first in the hadith.
+In Table~\ref{t:hadithresults}, instead of the segmentation columns 
+we report the percentage of times iTree succeeded to recognize the text passed
+as a hadith.
+This measure is equivalent to the concept of recall; nonetheless, precision
+in this context is not applicable since by default the input is required 
+to represent correctly extracted narrations.
+
+In order to evaluate the approaches used by iTree and NaGEMA without bias,
+we augmented the list of names used by 
+iTree with those available in the lexicon of Sarf.
+%'s lexicon on top of which NaGEMA operates.
+
+First, we compare the recall of the segmentation 
+in NaGEMA to that of hadith acceptance rates in iTree. 
+Although the task of segmentation is more complex than that of
+hadith recognition/acceptance, we still notice that iTree has on the average 35.7\% recall rate compared to 99.9\% for 
+NaGEMA. 
+However, we notice that the difference in these numbers depended highly on the books under consideration. 
+For instance,
+for the book of musnad, both approaches correctly detected  practically all narrations under consideration; 
+however iTree suffers a lot in 
+the books of both Kafi and Istibsar, resulting in a recall of less than 5\%. 
+This is due to the less structured nature of the narrations of those books.
+For example, iTree's grammar is defined such that a hadith starts by 
+an Ikhbar word (\noVocRL{AdaT al-A_hbAar}) such as
+\noVocRL{qAl}, \noVocRL{.hadda_t}, etc...; 
+however, the Kafi and Istibsar books followed a flexible style to express the narration 
+and most narrations started directly by a narrator name without a preceding Ikhbar word. 
+This is a strong evidence of the superiority and tolerance of NaGEMA 
+and its threshold based approach
+compared to the strict grammar approach used in iTree. 
+Clearly, NaGEMA does not fail on such cases, since what it detects as a hadith is based 
+on a concentration of names.
+Similarly, iTree had a lower sanad boundary precision (62.4\% on the average).
+This is mainly due to the grammar that ends a sanad only when a predefined prophet phrase is met.
+Prophet phrases are phrases that point to the prophet such as 
+\noVocRL{rasowl al-lah} and 
+\noVocRL{.sal_A `alayh al-lah wasalam}.
+
+However, especially in the books of Kafi and Istibsar, the prophet is not explicitly mentioned but 
+implicitly understood. In those cases, iTree will continue reporting narrators whenever it reaches 
+a word that it predefines as a name even if it appears
+in the matn of the hadith. 
+On the other hand, NaGEMA considers that a sanad ends whenever names become less frequent and measures 
+that with a threshold even if it does not reach a prophet
+indicative phrase.
+
+Narrator boundary accuracy showed similar results between iTree and 
+NaGEMA since we augmented iTree with the names from the lexicon of Sarf.
+
+\subsection{Effect of punctuation and learning}
 
@@
 \multirow{3}{*}{no punctuation} 
-& musnad  & & 1.000 & 1.000 & 1.000 & & 0.982 & 0.955 & 0.968 & & 1.000 & 0.998 & 0.999 \\ 
-& kafi & & 1.000 & 0.793 & 0.885 & & 0.944 & 0.967 & 0.955 & & 0.996 & 0.938 & 0.966 \\
-& istibsar & & 1.000 & 1.000 & 1.000 & & 0.980 & 0.931 & 0.955 & & 0.994 & 0.985 & 0.990 \\  \cline{2-14}
-& &  & 1.000 & 0.931 & 0.962 & & 0.969 & 0.951 & 0.960 & & 0.997 & 0.974 & 0.985 \\ \hline \hline
+& musnad  & & 0.999 & 0.999 & 0.999 & & 0.982 & 0.955 & 0.968 & & 0.999 & 0.998 & 0.999 \\ 
+& kafi & & 0.999 & 0.793 & 0.885 & & 0.944 & 0.967 & 0.955 & & 0.996 & 0.938 & 0.966 \\
+& istibsar & & 0.999 & 0.999 & 0.999 & & 0.980 & 0.931 & 0.955 & & 0.994 & 0.985 & 0.990 \\  \cline{2-14}
+& &  & 0.999 & 0.931 & 0.962 & & 0.969 & 0.951 & 0.960 & & 0.997 & 0.974 & 0.985 \\ \hline \hline
 %& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
@@
 \multirow{3}{*}{no learning}
-& musnad & & 1.000 & 1.000 & 1.000 & & 0.894 & 0.985 & 0.937 & & 0.995 & 1.000 & 0.998\\
-& kafi & & 1.000 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.991 & 0.992\\
-& istibsar & & 1.000 & 0.973 & 0.986 & & 0.977 & 0.950 & 0.963 & & 0.996 & 0.998 & 0.997\\ \cline{2-14}
- & &  & 1.000 & 0.964 & 0.982 & & 0.936 & 0.968 & 0.951 & & 0.995 & 0.996 & 0.995 \\
+& musnad & & 0.999 & 0.999 & 0.999 & & 0.894 & 0.985 & 0.937 & & 0.995 & 0.999 & 0.998\\
+& kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.991 & 0.992\\
+& istibsar & & 0.999 & 0.973 & 0.986 & & 0.977 & 0.950 & 0.963 & & 0.996 & 0.998 & 0.997\\ \cline{2-14}
+ & &  & 0.999 & 0.964 & 0.982 & & 0.936 & 0.968 & 0.951 & & 0.995 & 0.996 & 0.995 \\
  \end{tabular}                 
@@
 
-We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
-with 4GB of memory using a Linux operating system. 
-We report our results in Table~\ref{t:hadithresallresults} on the 
-three hadith books.
-NaGEMA detected narrator chains and segmented the books 
-correctly with a chain accuracy above 97\% and a narrator name
-accuracy above 90\%
-in all cases and across all the 1986 traditions in the three 
-different books with different formats.
-
-The row ``word count'' is an indicator of the size of the data.
-The %numbers in 
-rows ``names'', ``narrators'', and ``chains'' denote the total
-number of the corresponding entities per book.
-The row ``ignored names'' reports the number of names detected 
-outside the chains and thus are irrelevant to
-the application. 
-%In addition, the rows ``names/narrator'' and ``narrators/chain'' 
-%report the average
-%number of names per narrator and 
-%the average number of narrators per chain respectively.
-
-The ``segmentation accuracy'' reports on the accuracy of 
-segmentation of the book into separate hadiths.
-We computed the metric manually by marking the beginning 
-and end of a representative sample of hadiths in each 
-book and compared that to the NaGEMA results. 
-We used the same approach to compute all other accuracy measures.
-The ``chain accuracy'' reflects the percentage of correct narrators 
-covered by extracted chains.
-The ``narrator name accuracy'' row reflects the percentage of 
-correct proper names detected within
-narrator structures.
-
-%The row ``name false positives'' reports the percentage of words 
-%recognized as names while they should not.
-%We notice that this metric did not affect the accuracy
-%of narrator and chain segmentation since detecting
-%a narrator or a chain with the refined analysis without substantially affecting the 
-%high abstract measures such as chain and segmentation
-%accuracy.
-
-\section{Related work }
-\label{sec:related}
-
-The closest to our work is e-Narrator~\cite{Azmi2010}, a 
-tool that uses a grammar based approach to solve 
-the narrator extraction problem. 
-The e-Narrator tool preprocesses the text to remove 
-punctuation and diacritics and to normalize white
-spaces. 
-Then e-Narrator uses a similarity-based approach
-to memory based learning~\cite{Azmi2010} in order to perform 
-shallow parsing of the preprocessed text. 
-The e-Narrator tool uses a parser based on a context
-free grammar that enumerates several ways to express
-the name of the prophet, several possible words that 
-might precede the name of the prophet, several possible
-words that might separate two narrators, and several
-words that might separate parts of the name of the narrators.
-The e-Narrator tool fails when a variation of one of the 
-listed words and not an exact form of it
-happens in the text. 
-In their grammar they consider a name to be any sequence of 
-legal Arabic characters. 
-The e-Narrator tool generates successfully 86.7 percent of
-the narration chains of 90 selected traditions with 
-34 simple cases and 56 hard cases.
-
-NaGEMA does not need a preprocessing step since 
-it uses morphological analysis to process the text. 
-While e-Narrator bases its analysis on the patterns 
-of words that surround the narrator names, 
-NaGEMA bases its analysis on the fact that narrator
-names are likely to happen in localities with a high
-number of proper names. 
-NaGEMA uses the narrator connectors as helpers and
-uses morphological analysis to compare the stems of
-those words and thus NaGEMA avoids the problem of 
-``parser noise words'' that e-Narrator faces. 
-Even though we do not have access to the selected
-90 traditions e-Narrator reports on, we claim
-that NaGEMA outperforms e-Narrator as it works on
-the full text of the presented hadith books
-and reports a higher success rate.
+We are interested in studying the effect of the punctuation 
+and the learning optimizations we implemented in NaGEMA.
+We utilized punctuation marks to increase the accuracy of narration detection and narrator boundaries. 
+We also learned names of narrators that consisted of words missing from 
+our lexicon.
+We learned the narrator names based on their context as they occurred 
+between narrator connectors or contained name connectors.
+
+In Table~\ref{t:nopunctlearnresults}, we report the results of running NaGEMA, 
+disabling each of the optimizations at a time.
+
+We notice that punctuations are useful for increasing all precision measures: hadith segmentation, sanad boundary and
+narrator boundary. Those measures are enhanced by about 
+3\%, 2\%, and 1\% respectively as punctuation related 
+information is utilized.  
+We observed that punctuation marks helped eliminate confusion that resulted from a 
+concentration of names and words that 
+can be interpreted falsely as narrator connectors. 
+A period or a paragraph mark separated some name concentrations 
+dropping each part below the acceptable sanad threshold. 
+Punctuation marks also help cleaning the sanad of a narration from 
+names that coincidentally appear at the 
+end of the matn of a previous narration. 
+Names from a previous narration maybe mistaken to belong to the sanad of the current 
+narration if they were close enough. 
+Punctuation marks reduce such confusions.
+
+Although simple, our name learning optimization enhanced the recall 
+of sanad boundaries by about 3.4\% on the average.
+This means that we correctly report more narrators.
+This is a significant improvement as the task of hadith authentication requires high accuracy. 
+In particular, sometimes the presence of a specific narrator in the chain can render a 
+hadith as not trustworthy since the narrator is not considered credible.
+%That's why it is important not to lose any of the narrators that appear in the sanad.
+
+\subsection{Distance metric}
+Finally, the distance metric had 0.993 precision,
+0.999 recall and an F-score of 0.997.
 
@@
 using morphological analysis. 
-NaGEMA extracted narrator chains with above 90\% accuracy
+NaGEMA automatically extracted narrator chains with 
+above 90\% accuracy
 from three different hadith books and generated 
@@
 
-This partial order graph is instrumental to automate
+Up to our knowledge, NaGEMA is the first 
+tool that automates hadith segmentation and builds
+a comprehensive POR graph.
+It uses morphological analysis coupled with 
+and a knowledge based FSM to  detect the hadith 
+sanad.
+
+The partial order graph is instrumental to automate
 localization of narrators in biography documents and
 segmenting biography documents in the future.
-Given that a biography discusses a narrator 
-and mentions his professors and students,
-we plan to segment the biographies with a graph 
-coloring algorithm
-that traverses the text and colors the POR node whenever
-a name is found.
-
 Once the biographies are analyzed, we can annotate
@@
 their authenticity. 
+We also plan to 
+enhance the accuracy of hadith detection and the 
+narrator merging algorithms. 
 %We can also annotate the POR with the locations and 
@@
 
-We also plan to 
-enhance the accuracy of hadith detection and the 
-narrator merging algorithms.
-
 %We plan to make use of the name connector and narrator

### 2011-06-22 trunk/ATSarf/src/case/graph.h

@@ -613,16 +613,14 @@ private:
 					{
 						Narrator & n1_ref=n1->getNarrator();
 						Narrator & n2_ref=n2->getNarrator();
-					#ifdef DEBUG_BUILDGRAPH
-					#if 1
-						qDebug()<<n1_ref.getString()<<"["<<i<<","<<n1->getIndex()<<"]Versus["<<n2_ref.getString()<<"["<<k<<","<<n2->getIndex()<<"]\n";
-					#endif
-					#endif
 						double eq_val=equal(n1_ref,n2_ref);
+					#ifdef DEBUG_BUILDGRAPH
+						out<<n1_ref.getString()<<"["<<i<<","<<n1->getIndex()<<"]Versus["<<n2_ref.getString()<<"["<<k<<","<<n2->getIndex()<<"]\t"<<eq_val<<"\n";
+					#endif
 						if (eq_val>=threshold)
 						{
 						#ifdef DEBUG_BUILDGRAPH
-							out<<n1_ref.getString()<<"["<<i<<","<<n1->getIndex()<<"]Versus["<<n2_ref.getString()<<"["<<k<<","<<n2->getIndex()<<"]";
+							out<<"\t=>merge\n";
 						#endif
 							mergeNodes(*n1,*n2);
 							offset=u;	//this is matched, we must skip it in search for match for next node in c1

### 2011-06-22 trunk/ATSarf/src/case/hadithCommon.cpp

@@ -1,0 +1,166 @@
+#include "hadithCommon.h"
+#include <QStringList>
+
+HadithParameters hadithParameters;
+
+#ifdef REFINEMENTS
+	QStringList rasoul_words;
+	QStringList compound_words,suffixNames;
+#endif
+
+	QString chainDataStreamFileName= ".chainOutput";
+#ifdef PREPROCESS_DESCRIPTIONS
+	QString preProcessedDescriptionsFileName= ".HadithPreProcessedDescriptions";
+#endif
+#ifndef SUBMISSION
+	QString PhrasesFileName="../src/case/phrases";
+	QString StopwordsFileName="../src/case/stop_words";
+#else
+	QString PhrasesFileName=".phrases";
+	QString StopwordsFileName=".stop_words";
+#endif
+
+#ifdef COUNT_RUNON
+	bool runon;
+#endif
+
+	QString hadath,abid,alrasoul,abyi,_3an;
+	int bit_POSSESSIVE, bit_PLACE,bit_CITY,bit_COUNTRY,bit_NOUN_PROP;
+	QList<int> bits_NAME;
+
+#ifdef PREPROCESS_DESCRIPTIONS
+	QHash<long,bool> NMC_descriptions;
+	QHash<long,bool> NRC_descriptions;
+	QHash<long,bool> IBN_descriptions;
+#endif
+#ifdef COMPARE_TO_BUCKWALTER
+	QTextStream * myoutPtr;
+#endif
+#ifdef COUNT_AVERAGE_SOLUTIONS
+	long total_solutions=0;
+	long stemmings=0;
+#endif
+
+
+#ifdef PREPROCESS_DESCRIPTIONS
+void readFromDatabasePreProcessedDescriptions() {
+	Retrieve_Template nrc_s("description","id","name='said' OR name='say' OR name='notify/communicate' OR name LIKE '%/listen' OR name LIKE 'listen/%' OR name LIKE 'listen %' OR name LIKE '% listen' OR name = 'listen' OR name LIKE '%/inform' OR name LIKE 'inform/%' OR name LIKE 'inform %' OR name LIKE '% inform' OR name = 'inform' OR name LIKE '%from/about%' OR name LIKE '%narrate%'");
+	while (nrc_s.retrieve())
+		NRC_descriptions.insert(nrc_s.get(0).toULongLong(),true);
[…]

### 2011-06-22 trunk/ATSarf/src/case/hadith_general.cpp

@@ -1,99 +1,20 @@
 #include "hadith.h"
 
 #ifdef GENERAL_HADITH
-#include <QFile>
-#include <QRegExp>
-#include <QString>
-#include <QStringList>
 #include <QTextBrowser>
+#include <assert.h>
 
-#include "stemmer.h"
-#include "letters.h"
-#include "text_handling.h"
-#include "diacritics.h"
-#include "narrator_abstraction.h"
-#include "common.h"
 #include "ATMProgressIFC.h"
 #include "Math_functions.h"
-#include "Retrieve_Template.h"
-#include "Search_by_item_locally.h"
 #include "graph.h"
 #include "mainwindow.h"
 #include "ui_mainwindow.h"
-#include "textParsing.h"
-#include <assert.h>
 
-HadithParameters hadithParameters;
-QString chainDataStreamFileName= ".chainOutput";
-#ifdef PREPROCESS_DESCRIPTIONS
-QString preProcessedDescriptionsFileName= ".HadithPreProcessedDescriptions";
-#endif
-#ifndef SUBMISSION
-QString PhrasesFileName="../src/case/phrases";
-QString StopwordsFileName="../src/case/stop_words";
-#else
-QString PhrasesFileName=".phrases";
-QString StopwordsFileName=".stop_words";
-#endif
+#include "hadithCommon.h"
 
-#ifndef REFINEMENTS
-	enum wordType { NAME, NRC,NMC};
-	enum stateType { TEXT_S , NAME_S, NMC_S , NRC_S};
-#else
-	#define GET_WAW
 
-	enum wordType { NAME, NRC,NMC,STOP_WORD};
-	enum stateType { TEXT_S , NAME_S, NMC_S , NRC_S, STOP_WORD_S};
-	QStringList rasoul_words;
-#endif
-//#define COUNT_RUNON
-#ifdef COUNT_RUNON
-	bool runon;
-#endif
-
-	typedef struct  {
-		long lastEndPos;
-		long startPos;
-		long endPos;
-		long nextPos;
-		wordType currentType:3;
-		stateType currentState:4;
-		stateType nextState:4;
-		bool ibn:1;
-		bool _3abid:1;
-		bool possessivePlace:1;
-		bool number:1;
-		bool isWaw:1;
-		bool _3an:1;
-		bool learnedName:1;
-		bool nrcIsPunctuation:1; //if state is NRC and that is caused soley by punctuation
-		int unused:18;
[…]

### 2011-06-22 trunk/ATSarf/src/case/narrator_abstraction.cpp

@@ -1,7 +1,8 @@
-#include "hadith.h"
+#include "hadithCommon.h"
 #include "narrator_abstraction.h"
 #include <QStringList>
 #include <QDataStream>
+#include <QString>
 #include <assert.h>
 #include "Triplet.h"
 #include "stemmer.h"
@@ -9,13 +10,12 @@
 #include "hadith_utilities.h"
 
 #ifdef EQUALITYDEBUG
-inline void display(QString t)
-{
-	out<<t;
-	//qDebug() <<t;
-}
+	inline void display(QString t) {
+		out<<t;
+		//qDebug() <<t;
+	}
 #else
-#define display(c) ;
+	#define display(c) ;
 #endif
 
 qint8 getType(const NarratorPrim *)
@@ -501,7 +501,7 @@ inline double getdistance(const Narrator & n1,const Narrator & n2) //TODO: use p
 	display(n1.getString());
 	display(" VS ");
 	display(n2.getString());
-	display("> ");
+	display(">\n");
 #if 0
 	for (int i=0;i<Names1.count();i++)
 		display(Names1[i].getString()+" - ");

### 2011-07-28 trunk/tex/hadith.ijcnlp.tex
corrected a bug in hadith state machine
corrected bug in the traversal with caching

started editting the hadith paper

@@
 
+(modify according to hash)
+
 \begin{table}[tb]
@@
 
+\subsubsection{On Test-suite Of iTree}
+
+(report on the 90 hadith used in paper of iTree)
+
+\subsubsection{On More Representative Test-suite}
+
 In order to evaluate the approaches used by iTree and NaGEMA without bias,
@@
 
+\subsection{On Unorganized hadith books}
+
+(report results on such hadith books)
+
 \subsection{Effect of punctuation and learning}
@@
 
+(explain why successful on this problem)
+
+(show detailed statistics and compare against that of eNarrator)
+
+\subsection{POR Correctness}
+
+(report recall and precision of nodes missing or those added, bc edges cannot be mistaken other than that)
+
+(report recall and precision on the chainnodes per graphnode, that is all what retains to correctness resulting from algorithm)
+
 \section{Conclusion and future work}

### 2011-07-28 trunk/ATSarf/src/case/graph.h

@@ -83,45 +83,37 @@ public:
 
 class GraphVisitorController
 {
+public:
+	typedef QStack< NarratorNodeIfc*> ParentStack;
 private:
+	typedef QPair<NarratorNodeIfc *,NarratorNodeIfc*> Edge;
+	typedef QMap<Edge, bool> EdgeMap;
 	typedef enum{NOT_VISITED,VISITED,FINISHED} Status;
 	typedef QMap<NarratorNodeIfc *, Status>  ColorMap;
 
-	bool firstCall;
 	ColorMap * visitedMap;
-public:
-	typedef QStack< NarratorNodeIfc*> ParentStack;
-protected:
-	typedef QPair<NarratorNodeIfc *,NarratorNodeIfc*> Edge;
-	typedef QMap<Edge, bool> EdgeMap;
+	bool firstCall:1;
+	bool direction:1;
+	bool graphBuilt:1;
+	bool keep_track_of_edges:1, keep_track_of_nodes:1,merged_edges_as_one:1;
+
+	friend bool GraphNarratorNode::fillChildren();
+	friend bool GraphNarratorNode::fillParents();
+	bool disableFillChildren:1;
+
+	int unused:26;
+
 	EdgeMap edgeMap;
 	ParentStack parentStack;
 	unsigned int visitIndex, finishIndex ;
 
 	NarratorGraph * graph;
-	bool keep_track_of_edges, keep_track_of_nodes,merged_edges_as_one, detect_cycles;
 	NodeVisitor * visitor;
-	bool stop_searching_for_cycles;
 
 	friend class NarratorGraph;
 
 	void init();
-	void construct(NodeVisitor * visitor,NarratorGraph * graph,bool keep_track_of_edges,bool keep_track_of_nodes, bool merged_edges_as_one)
-	{
-		assert(!(!keep_track_of_edges && !keep_track_of_nodes)); //i dont want both to be false at same time=> no exponential traversal
-		assert(!(keep_track_of_edges && !merged_edges_as_one && graph==NULL));//bc we cannot clear them after traversal if so
-		firstCall=true;
-		this->visitor=visitor;
-		assert(visitor!=NULL);
-		this->visitor->controller=this;
-		this->keep_track_of_edges=keep_track_of_edges || merged_edges_as_one;//if merged_edges_as_one is needed we must keep_track_of_edges
[…]

### 2011-07-28 trunk/ATSarf/src/gui/mainwindow.cpp

@@ -211,7 +211,7 @@ void MainWindow::on_fill_clicked()
 
 void MainWindow::on_cmd_browse_clicked()
 {
-	QString fileName=getFileName(browseFileDlg);
+	QString fileName=getFileName(&browseFileDlg);
 	if (!fileName.isEmpty())
 		m_ui->input->setText(fileName);
 }

### 2011-07-28 trunk/ATSarf/src/gui/test.cpp

@@ -83,11 +83,12 @@ int test(QString inputString,ATMProgressIFC * prg) {
 #elif 0
 	if (biographyHelper(inputString,prg))
 		return -1;
-#elif 1
+#elif 0
 	if (deserializeGraph(inputString,prg))
 		return -1;
 #else
-	QString fileName2=getFileName(NULL);
+	QFileDialog * browseFileDlg=NULL;
+	QString fileName2=getFileName(&browseFileDlg);
 	if (fileName2=="")
 		return -1;
 	if (mergeGraphs(inputString,fileName2,prg))

### 2011-07-28 ATSarf/src/case/biographies.h
corrected a bug in hadith state machine
corrected bug in the traversal with caching

started editting the hadith paper

@@ -225,7 +225,7 @@ public slots:
 	#endif
 	}
 	void browse_clicked() {
-		QString fileName=getFileName(browseFileDlg);
+		QString fileName=getFileName(&browseFileDlg);
 		if (!fileName.isEmpty())
 			input->setText(fileName);
 	}

### 2011-07-28 ATSarf/src/case/graph.cpp

@@ -47,6 +47,23 @@ GraphVisitorController::GraphVisitorController(NodeVisitor * visitor,NarratorGra
 		graph->colorGuard.use(finishIndex);
 	}
 }
+void GraphVisitorController::construct(NodeVisitor * visitor,NarratorGraph * graph,bool keep_track_of_edges,bool keep_track_of_nodes, bool merged_edges_as_one)
+{
+	assert(!(!keep_track_of_edges && !keep_track_of_nodes)); //i dont want both to be false at same time=> no exponential traversal
+	assert(!(keep_track_of_edges && !merged_edges_as_one && graph==NULL));//bc we cannot clear them after traversal if so
+	graphBuilt=(graph!=NULL?graph->isBuilt():true); //we are assuming if we are doing a traversal without a graph then graph has been built
+	firstCall=true;
+	disableFillChildren=false;
+	this->visitor=visitor;
+	assert(visitor!=NULL);
+	this->visitor->controller=this;
+	this->keep_track_of_edges=keep_track_of_edges || merged_edges_as_one;//if merged_edges_as_one is needed we must keep_track_of_edges
+	this->keep_track_of_nodes=keep_track_of_nodes;
+	this->merged_edges_as_one=merged_edges_as_one;
+	this->graph=graph;
+	init();
+}
+
 void GraphVisitorController::init()
 {
 	if (firstCall) {
@@ -66,9 +83,10 @@ void GraphVisitorController::init()
 		edgeMap.clear();
 	parentStack.clear();
 }
-void GraphVisitorController::initialize()
+void GraphVisitorController::initialize(int direction)
 {
 	init();
+	this->direction=direction;
 	visitor->initialize();
 }
 void GraphVisitorController::finish()
@@ -208,7 +226,7 @@ int mergeGraphs(QString fileName1,QString fileName2,ATMProgressIFC * prg) {
 
 	graph1->mergeWith(graph2);
 	delete graph2;
-#if 1
+#if 0
 	QFile file(fileName1.remove(".por")+"Merged.por");
 	if (!file.open(QIODevice::ReadWrite))
 		return -1;

### 2011-07-28 ATSarf/src/case/graph_nodes.cpp

@@ -9,7 +9,12 @@ NULLNarratorNodeIfc nullNarratorNodeIfc;
 NULLGraphNarratorNode nullGraphNarratorNode;
 ChainsContainer chains;
 NodeAddress nullNodeAddress((NarratorNodeIfc &)nullNarratorNodeIfc,(ChainNarratorNode &)nullChainNarratorNode);
-ChainNodeIterator ChainNodeIterator::null(NULL,-1,-1);
+#ifdef ITERATORS_SEPERATED
+ChainNodeIterator ChainNodeIterator::null(NULL);
+FilledNodeIterator FilledNodeIterator::null(NULL);
+#else
+NodeIterator NodeIterator::null(NULL);
+#endif
 
 NarratorNodeIfc::NarratorNodeIfc(NarratorGraph & g)
 	: id(g.nodesCount){
@@ -18,8 +23,36 @@ NarratorNodeIfc::NarratorNodeIfc(NarratorGraph & g)
 }
 
 
-ChainNodeIterator NarratorNodeIfc::begin() {
-	return ChainNodeIterator(this,0,0);
+NodeIterator NarratorNodeIfc::begin() {
+	assert(!isNull());
+	return NodeIterator(this);
+}
+
+NodeIterator NarratorNodeIfc::childrenBegin() {
+	assert(!isNull());
+	return NodeIterator(this);
+}
+
+NodeIterator NarratorNodeIfc::parentsBegin() {
+	assert(!isNull());
+	return NodeIterator(this);
+}
+
+NodeIterator GraphNarratorNode::childrenBegin() {
+	assert(!isNull());
+	if (children.isEmpty()) {
+		return NodeIterator(this);
+	} else {
+		return NodeIterator(this,true);
+	}
+}
+NodeIterator GraphNarratorNode::parentsBegin() {
+	assert(!isNull());
+	if (parents.isEmpty()) {
+		return NodeIterator(this);
+	} else {
+		return NodeIterator(this,false);
+	}
 }
 
 void NarratorNodeIfc::BFS_traverse(GraphVisitorController & visitor, int maxLevels,int direction) {
@@ -160,6 +193,17 @@ void ChainNarratorNode::deserializeHelper(QDataStream &streamIn,NarratorGraph & 
 	chainContext.set(index,chainNum);
 }
 
+void GraphNarratorNode::serializeCache(QDataStream &chainOut, NarratorGraph & graph,const NodeList & list) const {
+	int size=list.size();
+	chainOut<<size;
[…]

### 2011-07-28 ATSarf/src/case/narrator_abstraction.cpp

@@ -845,8 +845,10 @@ bool Biography::isRealNarrator(Narrator * n) {
 	} else {
 		v.resetFound();
 		graph->performActionToAllCorrespondingNodes(n,v);
-		if (!v.isFound())
+		if (!v.isFound()) {
 			return false;
+			nodeGroups.removeLast();
+		}
 		else
 			return true;
 	}

### 2011-07-28 ATSarf/src/util/browseDialog.h

@@ -4,16 +4,17 @@
 #include <QFileDialog>
 #include <QString>
 
-inline QString getFileName(QFileDialog * browseFileDlg) {
-	if (browseFileDlg == NULL) {
+inline QString getFileName(QFileDialog ** browseFileDlg) {
+
+	if (*browseFileDlg == NULL) {
 		QString dir = QDir::currentPath();
-		browseFileDlg = new QFileDialog(NULL, QString("Open File"), dir, QString("All Files (*)"));
-		browseFileDlg->setOptions(QFileDialog::DontUseNativeDialog);
-		browseFileDlg->setFileMode(QFileDialog::ExistingFile);
-		browseFileDlg->setViewMode(QFileDialog::Detail);
+		*browseFileDlg = new QFileDialog(NULL, QString("Open File"), dir, QString("All Files (*)"));
+		(*browseFileDlg)->setOptions(QFileDialog::DontUseNativeDialog);
+		(*browseFileDlg)->setFileMode(QFileDialog::ExistingFile);
+		(*browseFileDlg)->setViewMode(QFileDialog::Detail);
 	}
-	if (browseFileDlg->exec()){
-		QStringList files = browseFileDlg->selectedFiles();
+	if ((*browseFileDlg)->exec()){
+		QStringList files = (*browseFileDlg)->selectedFiles();
 		QString fileName = files[0];
 		return fileName;
 	}

### 2011-08-01 tex/hadith.ijcnlp.tex
some changes in tex file
corrected bugs in satemachine of hadith

corrected bugs in mergeGraphs and buildGraph using hash

--- 2011-06-02 tex/hadith.ijcnlp.tex
@@
 ``narrated to us'' that starts the hadith in Figure~\ref{f:exhadith}
-returns the stem \RL{.hd_t}.
+returns the stem \RL{.hd_t} ``narrated''.
 The stem  \RL{.hd_t} is also the stem of other 
-words such as \RL{.hd_tny} and \RL{.hd_thm}.
+words such as \RL{.hd_tny} ``narrated to me'' and \RL{.hd_thm} ``narrated to them''.
 The stem also shares similar part of speech (POS)
@@
 \href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
-%as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
-as well as biblical sources.
-%~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
+as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
 
@@
 morphological solutions.
+
+In this paper we make the following contributions. 
+\begin{itemize}
+\item We automate the segmentation of hadith books into separate narrations and extract chains of narrators 
+%into a partial order relation graph 
+using a knowledge based approach with morphological analysis.
+\item We optimize our analysis through utilizing punctuation and learning previously unknown names.
+\item We build from different chains of narrators a partial order relation graph based on merging common narrators.
+\item We introduce a specialized distance metric for Arabic names used in Arabic literature. 
+\item We highly outperform the currently available narrator extraction tool. 
+\item We provide a working and efficient implementation of our work. 
+\end{itemize}
+
+The rest of this paper is organized as follows. 
+We describe our approach to narrator chain extraction
+in Section~\ref{sec:islamic}.
+We describe how we build the partial order relation
+graph in Section~\ref{sec:graph}.
+We discuss related work in Section~\ref{sec:related}.
+We present our results in Section~\ref{sec:results} and 
+we conclude in Section~\ref{sec:future}.
 
@@
 compare against the thresholds.
-that directly interacted with Sarf.
+
+(Figure~\ref{f:chain} is confusing since it misses some information uses other symbols like 
+those of $\theta$ and is redundant anyways)
 
@@
 
+\subsection{Optimizations}
+
+(explain better and merged information present about them in "Results" section)
+
+In addition to the controller, 
+NaGEMA uses punctuation marks to refine its analysis and 
+increase the accuracy of name detection, narration detection and narrator boundaries. 
+It also learns narrator names from simple patterns that 
+happen within narrator connectors and that are not 
+reported by Sarf as names. 
+NaGEMA learns names of narrators that consist of words missing from 
+the Sarf lexicon.
+NaGEMA learns the narrator names based on their context as they occurred 
+between narrator connectors or contained name connectors.
+
 \section{The narrator graph}
+\label{sec:graph}
 
@@
 that merged equal names in the chains.
-We compared the narrators together using a morphological
-distance function. 
-We considered each narrator to have several morphological 
-components and we rewarded every morphological match with 
-a small $\delta$. 
-Then we applied a threshold $\theta$ to the sum of rewards
-and decided the equality of the narrators. 
-%We omitted the details of the distance function for brevity.
+
+\subsection{Motivation}
+
+The hadith POR is instrumental to automate
+localization of narrators in biography documents and
+to segment biography documents in future work.
+Consider a biography that discusses a narrator.
+It mentions his professors and students.
+We plan to segment the biographies with a graph 
+coloring algorithm
+that parses the text looking for names. 
+Once a name is found it locates it in the POR graph
+and colors the POR node corresponding
+to it.
+The narrator predecessors in the POR are supposed to be
+his professors and the narrator successors in the POR 
+are supposed to be his students. 
+Consequently, whenever a name moves away from the 
+colored local cluster in the POR graph, 
+we know that we are within a new biography.
+
+Once the biographies are analyzed, we can annotate
+the POR with qualifiers of the narrators that reflect
+their authenticity. 
+We can also annotate the POR with the locations and 
+the time the narrators lived in. 
+Then we can check for time and location overlap
+to discover inconsistent narrations.
+This motivates our interest to automatically generate 
+the POR graph.
+
+\subsection{Graph merge Algorithm}
+
+(modify according to hash)
 
@@
     $C_c = C_c \cup c_1$
+    foreach Chain $c_2$ in $C-C_c$
       foreach Narrator $n_1$ in $c_1$
-      foreach Chain $c_2$ in $C-C_c$
-        foreach narrator $n_2$ in $c_2$
-
-    \}
-  \}
+        start = max( $n_1$.index - radius, 0)
+        end = min($n_1$.index + radius, $n_2$.size)
+        $n_2$ = $c_2$[start]
+        while (true)
+          if (Distance($n_1$, $n_2$) < $\theta$)
+            mergeNodes($n_1$, $n_2$)
+          if ($n_2$ = $c_2$[end])
+            break
+          $n_2$ = $c_2$.next($n_2$)        
 \end{Verbatim}
@@
 
-\subsection{Use of the POR graph}
-
-inconsistency check 
-consistency checks
-annotate with information extracted from biographies
-what if analysis
-
-\section{Results}
-\label{sec:results}
-
-\begin{table}[bt]
-\centering
-\caption{Results of the hadith case study with Sarf.}
-%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
-\resizebox{1.1\columnwidth}{!}{
-\begin{tabular}{lcp{.2cm}cp{.2cm}c} %\cline{2-10}
- &  AlKafi & & AlIstibsar & &IbnHanbal \\ \cline{1-6}
-Word count  &98,943 & & 103,835 & & 20,354 \\ 
- Names       & 12,060  & & 14,613& & 3,013\\
-%Names/Narrator & 1.97 & & 1.84& & 1.25 \\
-Narrators  & 2,623 & & 5,767& & 1,755 \\ 
-%Narrators/Chain  & 4.84 & & 4.76 & &4.05 \\
-Chains  & 542 &  & 1,211& & 433 \\ 
-Ignored names  & 6,400 &  & 3,348 & & 642 \\ \hline
-Segmentation accuracy  & 96\%& & 96\%& & 92\%\\ 
-Chain accuracy & 99\%&  & 99\%& & 97\% \\ 
-%Narrator accuracy  & & & & &  & & & \\ 
-Narrator accuracy  & 91\%& & 90\% & & 90\% \\ \hline
-%Name false positives  & 7\%&  & 4\% & & 4\% \\ \hline
-Running time (secs.)& 1.32 & & 1.31 & & .096\\ \hline 
-\end{tabular}
-}
-\normalsize
-\label{t:hadithresallresults}
-\end{table}
-
-We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
-with 4GB of memory using a Linux operating system. 
-We report our results in Table~\ref{t:hadithresallresults} on the 
-three hadith books.
-NaGEMA detected narrator chains and segmented the books 
-correctly with a chain accuracy above 97\% and a narrator name
-accuracy above 90\%
-in all cases and across all the 1986 traditions in the three 
-different books with different formats.
-
-The row ``word count'' is an indicator of the size of the data.
-The %numbers in 
-rows ``names'', ``narrators'', and ``chains'' denote the total
-number of the corresponding entities per book.
-The row ``ignored names'' reports the number of names detected 
-outside the chains and thus are irrelevant to
-the application. 
-%In addition, the rows ``names/narrator'' and ``narrators/chain'' 
-%report the average
-%number of names per narrator and 
-%the average number of narrators per chain respectively.
-
-The ``segmentation accuracy'' reports on the accuracy of 
-segmentation of the book into separate hadiths.
-We computed the metric manually by marking the beginning 
-and end of a representative sample of hadiths in each 
-book and compared that to the NaGEMA results. 
-We used the same approach to compute all other accuracy measures.
-The ``chain accuracy'' reflects the percentage of correct narrators 
-covered by extracted chains.
-The ``narrator name accuracy'' row reflects the percentage of 
-correct proper names detected within
-narrator structures.
-
-%The row ``name false positives'' reports the percentage of words 
-%recognized as names while they should not.
-%We notice that this metric did not affect the accuracy
-%of narrator and chain segmentation since detecting
-%a narrator or a chain with the refined analysis without substantially affecting the 
-%high abstract measures such as chain and segmentation
-%accuracy.
+The \CodeIn{Distance} Algorithm 
+takes as input two narrators and returns 
+their morphological and order distance
+from each other. 
+The lower the distance is, the more
+equivalent the narrators are.
+
+Each narrator has several names and several name and 
+narrator connectors. 
+\CodeIn{Distance} computes the names of the two narrators
+and stores them in $N_1$ and $N_2$ 
+and checks whether they are morphologically equivalent. 
+If they are not then the distance is set to $\infty$. 
+Otherwise the distance is decremented by $\delta_1$. 
+Then \CodeIn{Distance} checks whether the morphologically 
+equivalent names appear in the same order and 
+decrements the distance if so.
+
+\CodeIn{Distance}  pays attention to the $\mathit{NISBA}$ 
+words that relate to places and checks on that to make sure
+that the two narrators considered equivalent are actually 
+from the same place. 
+It also checks whether one place such as Baghdad 
+is geographically 
+contained in the other such as Iraq 
+(we skip that in the Algorithm for clarity). 
+\CodeIn{Distance} also checks for the equivalency of the
+name connectors and the correspondence of the \RL{ibn}
+connectors.
+
+The distance metric of NaGEMA performs better than       
+existing metrics~\cite{Azmi-2010} as it can detect that  
+\RL{`bd al-lah bn `mr} and \RL{ibn `mr} are the same person
+while other distance metrics based on simple
+string similarity techniques can not.
+
+\subsection{Speedup Optimizations for Large Datasets}
+
+\subsubsection{MultiKey Hashing}
+
+(explain why suitable)
+
+\subsubsection{Three-Level Hierarchy Structure for Graph Nodes}
+
+(possibly omit this info, since is complicated)
 
@@
 
-The closest to our work is e-Narrator~\cite{Azmi2010}, a 
+%\transfalse
+%\novocalize
+%\begin{figure*}[tb]
+%\center{
+%\resizebox{1.8\columnwidth}{!}
+%%{ \input{iTreeMohamada}}
+%{\includegraphics{iTreeMohamada}}
+%\caption{Output of iTree fails to detect \RL{m_hmdA}.}
+%\label{f:mohamada}
+%}
+%\end{figure*}
+%\transtrue
+
+The closest work to NaGEMA is 
+iTree~\cite{Azmi-2010,iTree}, a 
 tool that uses a grammar based approach to solve 
 the narrator extraction problem. 
-The e-Narrator tool preprocesses the text to remove 
+The iTree tool preprocesses the text to remove 
 punctuation and diacritics and to normalize white
 spaces. 
-Then e-Narrator uses a similarity-based approach
-to memory based learning~\cite{Azmi2010} in order to perform 
+Then iTree uses a similarity-based approach
+to memory based learning~\cite{Azmi-2010} 
+in order to perform 
 shallow parsing of the preprocessed text. 
-The e-Narrator tool uses a parser based on a context
+The iTree tool uses a parser based on a context
 free grammar that enumerates several ways to express
@@
 words that might separate two narrators, and several
-words that might separate parts of the name of the narrators.
-The e-Narrator tool fails when a variation of one of the 
+words that might separate parts of the name of 
+the narrators.
+The iTree tool fails when a morphological variation of 
+one of the 
 listed words and not an exact form of it
 happens in the text. 
-In their grammar they consider a name to be any sequence of 
-legal Arabic characters. 
-The e-Narrator tool generates successfully 86.7 percent of
-the narration chains of 90 selected traditions with 
-34 simple cases and 56 hard cases.
-
-NaGEMA does not need a preprocessing step since 
+For example, Figure~\ref{f:mohamada} extracted 
+using iTree fails to detect \RL{m_hmdA} as a name of the
+prophet and does not stop at it when computing the chain 
+of the hadith
+\transfalse
+\setcode{utf8}
+%\begin{arabtext}
+\RL{
+حدثنا وكيع، حدثني سعيد بن السائب، عن داود بن أبي عاصم الثقفي، قال
+سألت ابن عمر عن الصلاة، بمنى فقال هل سمعت {\bf محمدا}، صلى الله عليه وسلم قلت
+نعم وآمنت فاهتديت به قال فإنه كان يصلي بمنى ركعتين.
+%\end{arabtext}
+}
+\setcode{standard}
+\transtrue
+\vocalize
+
+%In their grammar they consider a name to be any sequence 
+%of 
+%legal Arabic characters.
+
+NaGEMA does not need a prepossessing step since 
 it uses morphological analysis to process the text. 
-While e-Narrator bases its analysis on the patterns 
+While iTree bases its analysis on the patterns 
 of words that surround the narrator names, 
@@
 those words and thus NaGEMA avoids the problem of 
-``parser noise words'' that e-Narrator faces. 
-Even though we do not have access to the selected
-90 traditions e-Narrator reports on, we claim
-that NaGEMA outperforms e-Narrator as it works on
+``parser noise words'' that iTree faces.
+
+The iTree tool extracts successfully 86.7 percent of
+the narration chains of 90 selected traditions with 
+34 simple cases and 56 hard cases. 
+We ran NaGEMA against the five hadith that ship with
+the tool and the two hadith that are listed in the iTree
+paper and NaGEMA was able to extract all narration
+chains with no mistakes even on the examples where iTree
+reports a ``parser noise'' problem.
+
+As we do not have access to the selected
+90 traditions iTree reports on, we claim
+that NaGEMA outperforms iTree as it works on
 the full text of the presented hadith books
@@
 
+Our approach to detect and relate names to verbs
+is similar to a local grammar based 
+approach~\cite{Traboulsi:09} that detects names by
+associating them with verbs 
+such as \RL{qaal} ``said'' and \RL{'a_hbar} ``told''. 
+The local grammars in~\cite{Traboulsi:09} are also
+built into transducers that generate the names and
+the learning phase benefits from morphological analysis. 
+We differ in that we do not use a local grammar, rather
+our tolerance thresholds allow a flexible pattern that is
+not easy to capture in a typical regular expression.
+
+Another attempt to Arabic named entity recognition
+uses an n-gram approach and maximum 
+entropy~\cite{ANERSys:07}. 
+The authors use their own corpora and gazetteers to boost 
+and train the system and ignore POS tags. 
+They achieve major improvements when they use lexical and
+morphosyntactic features and a parallel Arabic and English 
+corpora to bootstrap noisy features~\cite{Benajiba:2010}.
+
+\section{Results}
+\label{sec:results}
+
+%\begin{table}[bt]
+%\centering
+%\caption{Results of the hadith case study with Sarf.}
+%%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
+%\resizebox{1.1\columnwidth}{!}{
+%\begin{tabular}{lcp{.2cm}cp{.2cm}c} %\cline{2-10}
+% &  AlKafi & & AlIstibsar & &IbnHanbal \\ \cline{1-6}
+%Word count  &98,943 & & 103,835 & & 20,354 \\ 
+% Names       & 12,060  & & 14,613& & 3,013\\
+%%Names/Narrator & 1.97 & & 1.84& & 1.25 \\
+%Narrators  & 2,623 & & 5,767& & 1,755 \\ 
+%%Narrators/Chain  & 4.84 & & 4.76 & &4.05 \\
+%Chains  & 542 &  & 1,211& & 433 \\ 
+%Ignored names  & 6,400 &  & 3,348 & & 642 \\ \hline
+%Segmentation accuracy  & 96\%& & 96\%& & 92\%\\ 
+%Chain accuracy & 99\%&  & 99\%& & 97\% \\ 
+%%Narrator accuracy  & & & & &  & & & \\ 
+%Narrator accuracy  & 91\%& & 90\% & & 90\% \\ \hline
+%%Name false positives  & 7\%&  & 4\% & & 4\% \\ \hline
+%Running time (secs.)& 1.32 & & 1.31 & & .096\\ \hline 
+%\end{tabular}
+%}
+%\normalsize
+%\label{t:hadithresallresults}
+%\end{table}
+%
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Results comparing between NaGEMA and iTree.}
+\resizebox{2.1\columnwidth}{!}{
+\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+%\multicolumn{13}{c}{NaGEMA with all refinements} \\ \cline{3-13}
+& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{NaGEMA }&musnad & & 0.999 & 0.999 & 0.999 & & 0.993 & 0.990 & 0.991 & & 0.999 & 0.998 & 0.999 \\
+&kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.988 & 0.990 \\
+&istibsar & & 0.999 & 0.973 & 0.986 & & 0.980 & 0.950 & 0.965 & & 0.996 & 0.998 & 0.997 \\ \cline{2-14}
+& & & 0.999 & 0.964 & 0.982 & & 0.970 & 0.970 & 0.970 & & 0.996 & 0.995 & 0.995 \\ \hline \hline
+%& & &  &  &  & &  &  &  & &  &  & \\
+%&\multicolumn{13}{c}{iTree} \\ \cline{3-13}
+& & & \multicolumn{3}{c}{accept as hadith} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{iTree} &musnad & & 0.999 & NA & 0.999 & & 0.997 & 0.977 & 0.987 & & 0.998 & 0.999 & 0.999  \\
+&kafi & & 0.043 & NA & 0.083 & & 0.999 & 0.313 & 0.476 & & 0.999 & 0.999 & 0.999 \\
+&istibsar & & 0.028 & NA & 0.054 & & 0.875 & 0.583 & 0.700 & & 0.999 & 0.999 & 0.999 \\ \cline{2-14}
+& & & 0.357 & NA & 0.379 & & 0.957 & 0.624 & 0.721 & & 0.999 & 0.999 & 0.999 \\ 
+ \end{tabular}                 
+}                             
+\normalsize                   
+\label{t:hadithresults} 
+\end{table*}
+
+To evaluate our approach, we ran NaGEMA against three books of 
+hadith~\cite{IbnHanbal,AlTousi,AlKulayni} and compared the
+results to iTree. 
+The results are detailed in 
+Tables~\ref{t:hadithresults} and \ref{t:nopunctlearnresults}. 
+The numbers we present in the tables are the result of a 
+manual verification of the correctness of %the outputs of 
+10\% of the narrations in the books under consideration.
+
+(discuss annotation and verification process and mention smg about inter-annotator agreement)
+
+We used recall and precision as our evaluation metrics.
+We report results on three levels of accuracy: segmentation, sanad boundary and
+narrator boundary. 
+Segmentation accuracy refers to the success of NaGEMA in
+detecting in the plain text of a hadith book the locations of the different
+narrations therein.
+The sanad boundary accuracy measures the
+correctness of individual narrators inside the sanad. 
+Finally, the narrator boundary accuracy measures
+the correctness of narrator full names.
+
+Precisely defined, segmentation recall refers to the ratio of the narrations correctly detected against the
+total number of narrations present. 
+Segmentation precision refers to the fraction of correctly detected narration chains compared to all the extracted
+narration chains. 
+Similarly, sanad boundary recall measures whether we detected all narrators
+in the sanad; while sanad boundary precision measures whether we did so without introducing false positives. 
+The same concept applies to the narrator boundary where we count the number of words constituting 
+a valid narrator.
+
+\subsection{NaGEMA versus iTree}
+
+iTree does not tackle the problem of hadith segmentation. 
+In contrast to NaGEMA, it does not work on a book level, 
+instead it takes as input an isolated hadith. 
+It assumes the hadith book is already segmented.
+
+Consequently, for iTree to extract the narrators from the sanad of a narration, 
+the user has to extract and pass the hadith manually.
+Still in many cases iTree fails to recognize the structure of
+the hadith and does not extract any useful information from the narration. 
+This happens for example when a slight variation of the iTree grammar rules 
+happens first in the hadith.
+In Table~\ref{t:hadithresults}, instead of the segmentation columns 
+we report the percentage of times iTree succeeded to recognize the text passed
+as a hadith.
+This measure is equivalent to the concept of recall; nonetheless, precision
+in this context is not applicable since by default the input is required 
+to represent correctly extracted narrations.
+
+\subsubsection{On Test-suite Of iTree}
+
+(report on the 90 hadith used in paper of iTree)
+
+\subsubsection{On More Representative Test-suite}
+
+In order to evaluate the approaches used by iTree and NaGEMA without bias,
+we augmented the list of names used by 
+iTree with those available in the lexicon of Sarf.
+%'s lexicon on top of which NaGEMA operates.
+
+First, we compare the recall of the segmentation 
+in NaGEMA to that of hadith acceptance rates in iTree. 
+Although the task of segmentation is more complex than that of
+hadith recognition/acceptance, we still notice that iTree has on the average 35.7\% recall rate compared to 99.9\% for 
+NaGEMA. 
+However, we notice that the difference in these numbers depended highly on the books under consideration. 
+For instance,
+for the book of musnad, both approaches correctly detected  practically all narrations under consideration; 
+however iTree suffers a lot in 
+the books of both Kafi and Istibsar, resulting in a recall of less than 5\%. 
+This is due to the less structured nature of the narrations of those books.
+For example, iTree's grammar is defined such that a hadith starts by 
+an Ikhbar word (\noVocRL{AdaT al-A_hbAar}) such as
+\noVocRL{qAl} ``said'', \noVocRL{.hadda_t} ``narrated'', etc...; 
+however, the Kafi and Istibsar books followed a flexible style to express the narration 
+and most narrations started directly by a narrator name without a preceding Ikhbar word. 
+This is a strong evidence of the superiority and tolerance of NaGEMA 
+and its threshold based approach
+compared to the strict grammar approach used in iTree. 
+Clearly, NaGEMA does not fail on such cases, since what it detects as a hadith is based 
+on a concentration of names.
+Similarly, iTree had a lower sanad boundary precision (62.4\% on the average).
+This is mainly due to the grammar that ends a sanad only when a predefined prophet phrase is met.
+Prophet phrases are phrases that point to the prophet such as 
+\noVocRL{rasowl al-lah} and 
+\noVocRL{.sal_A `alayh al-lah wasalam}.
+
+However, especially in the books of Kafi and Istibsar, the prophet is not explicitly mentioned but 
+implicitly understood. In those cases, iTree will continue reporting narrators whenever it reaches 
+a word that it predefines as a name even if it appears
+in the matn of the hadith. 
+On the other hand, NaGEMA considers that a sanad ends whenever names become less frequent and measures 
+that with a threshold even if it does not reach a prophet
+indicative phrase.
+
+Narrator boundary accuracy showed similar results between iTree and 
+NaGEMA since we augmented iTree with the names from the lexicon of Sarf.
+
+\subsection{On Unorganized hadith books}
+
+(report results on such hadith books)
+
+\subsection{Effect of punctuation and learning}
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Results without punctuation and without learning. }
+\resizebox{2.1\columnwidth}{!}{
+\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{no punctuation} 
+& musnad  & & 0.999 & 0.999 & 0.999 & & 0.982 & 0.955 & 0.968 & & 0.999 & 0.998 & 0.999 \\ 
+& kafi & & 0.999 & 0.793 & 0.885 & & 0.944 & 0.967 & 0.955 & & 0.996 & 0.938 & 0.966 \\
+& istibsar & & 0.999 & 0.999 & 0.999 & & 0.980 & 0.931 & 0.955 & & 0.994 & 0.985 & 0.990 \\  \cline{2-14}
+& &  & 0.999 & 0.931 & 0.962 & & 0.969 & 0.951 & 0.960 & & 0.997 & 0.974 & 0.985 \\ \hline \hline
+%& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+\multirow{3}{*}{no learning}
+& musnad & & 0.999 & 0.999 & 0.999 & & 0.894 & 0.985 & 0.937 & & 0.995 & 0.999 & 0.998\\
+& kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.991 & 0.992\\
+& istibsar & & 0.999 & 0.973 & 0.986 & & 0.977 & 0.950 & 0.963 & & 0.996 & 0.998 & 0.997\\ \cline{2-14}
+ & &  & 0.999 & 0.964 & 0.982 & & 0.936 & 0.968 & 0.951 & & 0.995 & 0.996 & 0.995 \\
+ \end{tabular}                 
+}                             
+\normalsize                   
+\label{t:nopunctlearnresults} 
+\end{table*}
+
+We are interested in studying the effect of the punctuation 
+and the learning optimizations we implemented in NaGEMA.
+We utilized punctuation marks to increase the accuracy of narration detection and narrator boundaries. 
+We also learned names of narrators that consisted of words missing from 
+our lexicon.
+We learned the narrator names based on their context as they occurred 
+between narrator connectors or contained name connectors.
+
+In Table~\ref{t:nopunctlearnresults}, we report the results of running NaGEMA, 
+disabling each of the optimizations at a time.
+
+We notice that punctuations are useful for increasing all precision measures: hadith segmentation, sanad boundary and
+narrator boundary. Those measures are enhanced by about 
+3\%, 2\%, and 1\% respectively as punctuation related 
+information is utilized.  
+We observed that punctuation marks helped eliminate confusion that resulted from a 
+concentration of names and words that 
+can be interpreted falsely as narrator connectors. 
+A period or a paragraph mark separated some name concentrations 
+dropping each part below the acceptable sanad threshold. 
+Punctuation marks also help cleaning the sanad of a narration from 
+names that coincidentally appear at the 
+end of the matn of a previous narration. 
+Names from a previous narration maybe mistaken to belong to the sanad of the current 
+narration if they were close enough. 
+Punctuation marks reduce such confusions.
+
+Although simple, our name learning optimization enhanced the recall 
+of sanad boundaries by about 3.4\% on the average.
+This means that we correctly report more narrators.
+This is a significant improvement as the task of hadith authentication requires high accuracy. 
+In particular, sometimes the presence of a specific narrator in the chain can render a 
+hadith as not trustworthy since the narrator is not considered credible.
+%That's why it is important not to lose any of the narrators that appear in the sanad.
+
+\subsection{Distance metric}
+Finally, the distance metric had 0.993 precision,
+0.999 recall and an F-score of 0.997.
+
+(explain why successful on this problem)
+
+(show detailed statistics and compare against that of eNarrator)
+
+\subsection{POR Correctness}
+
+(report recall and precision of nodes missing or those added, bc edges cannot be mistaken other than that)
+
+(report recall and precision on the chainnodes per graphnode, that is all what retains to correctness resulting from algorithm)
+
+\section{Significance \& Extensions -- Named Entity Extraction in Other Contexts}
+
+\subsection{Genealogical Lists in the Bible}
+
+(explain what will be equivalent of NMC and NRC and how it will work
+
+\subsection{Place Names in Travel Itineraries}
+
+(check if it can be applicable)
+
+\subsection{Extending Arabic People Name Gazetteers}
+
+(briefly explain how this specific application has general value through contributing to the general task of entity extraction [learned names])
+
 \section{Conclusion and future work}

### 2011-08-01 ATSarf/src/case/graph_nodes.cpp

@@ -116,6 +116,31 @@ NarratorNodeIfc & GroupNode::getCorrespondingNarratorNode() {
 	}*/
 }
 
+void GroupNode::addChainNode(NarratorGraph */*g*/,ChainNarratorNode & nar) { //we dont check for duplicates here
+#if 0
+	for (int i=0;i<size();i++)
+	{
+		ChainNarratorNode * n2=&(*this)[i];
+		ChainNarratorNode * n1=&nar;
+		if (n1==n2)
+			return;
+		//assert (n1!=n2); //TODO...
+	}
+#endif
+	//assert(nar.graphNode==NULL);
+	list.append(&nar);
+	nar.setCorrespondingNarratorNodeGroup(this);
+	int index=nar.getIndex();
+	if (lowestIndex<0 || lowestIndex>index)
+		lowestIndex=index;
+	if (highestIndex<index)
+		highestIndex=index;
+	if (graphNode!=NULL && graphNode->isGraphNode()) {
+		graphNode->lowestIndex=max(graphNode->lowestIndex,lowestIndex);
+		graphNode->highestIndex=max(graphNode->highestIndex,highestIndex);
+	}
+}
+
 NarratorNodeIfc * NarratorNodeIfc::deserialize(QDataStream &chainIn, NarratorGraph & graph) {
 	bool isNull,isGraph,isChain;
 	chainIn >>isNull
@@ -287,8 +312,7 @@ void GroupNode::deserializeHelper(QDataStream &chainIn,NarratorGraph & graph) {
 		assert(c!=NULL);
 		assert(!c->isGraphNode());
 		ChainNarratorNode* n=(ChainNarratorNode*)c;
-		list.append(n);
-		n->setCorrespondingNarratorNodeGroup(this);
+		addChainNode(NULL,*n);
 	}
 
 	chainIn >>key

### 2011-08-01 ATSarf/src/case/hadith.h

@@ -14,11 +14,10 @@
 #define GENERAL_HADITH
 //#define FORCE_RANKS
 #define DISPLAY_GRAPHNODES_CONTENT
-//#define DISPLAY_NODES_BEING_BROKEN
+#define DISPLAY_NODES_BEING_BROKEN
 //#define SHOW_RANKS
 //#define SHOW_VERBOSE_RANKS
 //#define STATS
-#define TAG_HADITH
 //#define HADITHDEBUG  //just for additional debugging statements displayed
 #define PUNCTUATION
 #define TRYTOLEARN
@@ -53,8 +52,10 @@
 #define EQUAL_NEW
 #define PROGRESS_SERIALZATION
 #define SEGMENT_BIOGRAPHY_USING_POR
-//#define DONT_DISPLAY_BIOGRAPHY_GRAPHY
 #define HASH_GRAPH_BUILDING
+
+//#define DISPLAY_BIOGRAPHY_GRAPH
+//#define TAG_HADITH
 
 int hadithHelper(QString input_str,ATMProgressIFC *prgs);
 void hadith_initialize();

### 2011-08-01 ATSarf/src/case/hadith_1.cpp

@@ -1,266 +1,0 @@
-#include "hadith.h"
-
-#ifndef GENERAL_HADITH
-
-#include <QFile>
-#include <QRegExp>
-#include <QStringList>
-#include "logger.h"
-#include "Search_by_item.h"
-#include "stemmer.h"
-#include "letters.h"
-#include "common.h"
-
-enum wordType { IKHBAR, KAWL, AAN, NAME, TERMINAL_NAME,OTHER};
-//QString delimiters(" :.,()");
-QString a5barani,a5barana,sami3to,hadathana,hadathani,Aan,qal,yaqool,_bin,_ibin;
-
-void hadith_initialize()
-{
-	a5barani.append(alef_hamza_above).append(kha2).append(ba2).append(ra2).append(noon).append(ya2);
-	a5barana.append(alef_hamza_above).append(kha2).append(ba2).append(ra2).append(noon).append(alef);
-	sami3to.append(seen).append(meem).append(ayn).append(ta2);
-	hadathana.append(_7a2).append(dal).append(tha2).append(noon).append(alef);
-	hadathani.append(_7a2).append(dal).append(tha2).append(noon).append(ya2);
-	Aan.append(ayn).append(noon);
-	qal.append(qaf).append(alef).append(lam);
-	yaqool.append(ya2).append(qaf).append(waw).append(lam);
-
-	//delimiters="["+delimiters+fasila+"]";
-}
-
-class mystemmer: public Stemmer
-{
-public:
-	bool name;
-
-	mystemmer(QString word):Stemmer(word)
-	{
-		name=false;
-	}
-	bool on_match()
-	{
-#if !defined (REDUCE_THRU_DIACRITICS)
-				Stem->fill_details();
-#endif
-				minimal_item_info stem_info;
-				//out<<Suffix->startingPos-1<<" "<<getColumn("category","name",Stem->category_of_currentmatch)<<" --- ";
-				Search_by_item s(STEM,Stem->id_of_currentmatch);
-				while(s.retrieve(stem_info))
-				{
-#ifdef REDUCE_THRU_DIACRITICS
-						if (stem_info.category_id==Stem->category_of_currentmatch && stem_info.raw_data==Stem->raw_data_of_currentmatch)
-#else
-						if (stem_info.category_id==Stem->category_of_currentmatch)
-#endif
-						{
-							for (unsigned int i=0;i<stem_info.abstract_categories.size();i++)
[…]

### 2011-08-01 ATSarf/src/case/hadith_general.cpp

@@ -1,1957 +1,0 @@
-#include "hadith.h"
-
-#if 0
-#ifdef GENERAL_HADITH
-#include <QTextBrowser>
-#include <assert.h>
-
-#include "ATMProgressIFC.h"
-#include "Math_functions.h"
-#include "graph.h"
-#include "mainwindow.h"
-#include "ui_mainwindow.h"
-
-#include "hadithCommon.h"
-
-
-#ifdef HADITHDEBUG
-inline QString type_to_text(wordType t) {
-	switch(t)
-	{
-		case NAME:
-			return "NAME";
-		case NRC:
-			return "NRC";
-		case NMC:
-			return "NMC";
-	#ifdef REFINEMENTS
-		case STOP_WORD:
-			return "STOP_WORD";
-	#endif
-		default:
-			return "UNDEFINED-TYPE";
-	}
-}
-inline QString type_to_text(stateType t) {
-	switch(t)
-	{
-		case TEXT_S:
-			return "TEXT_S";
-		case NAME_S:
-			return "NAME_S";
-		case NMC_S:
-			return "NMC_S";
-		case NRC_S:
-			return "NRC_S";
-	#ifdef REFINEMENTS
-		case STOP_WORD_S:
-			return "STOP_WORD_S";
-	#endif
-		default:
-			return "UNDEFINED-TYPE";
-	}
-}
-inline void display(wordType t) {
-	out<<type_to_text(t)<<" ";
-	//qDebug() <<type_to_text(t)<<" ";
-}
-inline void display(stateType t) {
-	out<<type_to_text(t)<<" ";
-	//qDebug() <<type_to_text(t);
-}
-inline void display(QString t) {
-	out<<t;
-	//qDebug() <<t;
-}
-#else
-	#define display(c)
-#endif
-
-class HadithSegmentor {
-private:
-	typedef struct chainData_ {
-		NamePrim *namePrim;
-		NameConnectorPrim *nameConnectorPrim;
-		NarratorConnectorPrim *narratorConnectorPrim;
-		TempConnectorPrimList * temp_nameConnectors;
-		Narrator *narrator;
-		Chain *chain;
-
-		void initialize(QString * text) {
-			delete namePrim;
-			delete nameConnectorPrim;
-			delete narratorConnectorPrim;
-			delete narrator;
-			delete chain;
-			/*for (int i=0;i<temp_nameConnectors->count()-1;i++)
-				delete (*temp_nameConnectors)[i];*/
[…]

### 2011-08-01 ATSarf/src/common/common.h

@@ -153,4 +153,6 @@ static const QString stem_info_path=".stem_info.dat";
 #define resetReverseDirection(abstractCat) abstractCat.setBit(REVERSEDESCBIT,false)
 #define isReverseDirection(abstractCat) abstractCat.getBit(REVERSEDESCBIT)
 
+#define absVal(v) (v>=0?v:-v)
+
 #endif

### 2011-08-01 trunk/ATSarf/src/case/biographies.h
some changes in tex file
corrected bugs in satemachine of hadith

corrected bugs in mergeGraphs and buildGraph using hash

@@ -169,7 +169,7 @@ public slots:
 					narratorListDisplay->setItem(count,1,new QTableWidgetItem(QString("%1").arg(i)));
 					narratorList.append(n);
 					count++;
-				#ifndef DONT_DISPLAY_BIOGRAPHY_GRAPHY
+				#ifdef DISPLAY_BIOGRAPHY_GRAPHY
 					ColorBiographiesAction c(i);
 					graph->performActionToAllCorrespondingNodes(n,c);
 				#endif
@@ -182,7 +182,7 @@ public slots:
 	#endif
 	}
 	void colorBiography_clicked() {
-	#ifndef DONT_DISPLAY_BIOGRAPHY_GRAPHY
+	#ifdef DISPLAY_BIOGRAPHY_GRAPH
 		int num=biographyNum->currentText().toInt();
 		setCurrentAction("Display Graph");
 		report(0);
@@ -199,7 +199,7 @@ public slots:
 	#endif
 	}
 	void colorNarrators_clicked() {
-	#ifndef DONT_DISPLAY_BIOGRAPHY_GRAPHY
+	#ifdef DISPLAY_BIOGRAPHY_GRAPH
 		QList<QTableWidgetSelectionRange>  selection=narratorListDisplay->selectedRanges();
 		ColorNarratorsAction::DetectedNodesMap map;
 		ColorNarratorsAction action(map);
@@ -239,7 +239,7 @@ private:
 	virtual void resetActionDisplay();
 
 	void displayUncoloredGraph(){
-	#ifndef DONT_DISPLAY_BIOGRAPHY_GRAPHY
+	#ifdef DISPLAY_BIOGRAPHY_GRAPH
 		DisplayNodeVisitor visitor;
 		GraphVisitorController c(&visitor,graph,true,true);
 		graph->DFS_traverse(c);

### 2011-08-01 trunk/ATSarf/src/case/graph.cpp

@@ -104,48 +104,50 @@ void LoopBreakingVisitor::reMergeNodes(NarratorNodeIfc * n)
 {
 	if (n==NULL)
 		return;
-	if (!n->isGraphNode())
+	if (n->isChainNode())
 		return; //unable to resolve
-	assert(!n->isChainNode());
-	GraphNarratorNode * g=(GraphNarratorNode *)n;
+	assert(n->isGraphNode());
+	GraphNarratorNode * g=dynamic_cast<GraphNarratorNode *>(n);
 #ifdef DISPLAY_NODES_BEING_BROKEN
-	qDebug()<<g->CanonicalName();
+	qDebug()<<g->toString();
 #endif
-	QList<GroupNode *> narrators;
+	QList<GroupNode *> groups;
 	int size=g->size();
 	for (int i=0;i<size;i++) {
 		GroupNode & c=(*g)[i];
-		narrators.append(&c);
+		assert (c.size()>0); //otherwise means has been deleted which must not happen
+		groups.append(&c);
 		c.setGraphNode(NULL);
 	}
 	g->groupList.clear();
 	NarratorGraph* graph=controller->getGraph();
 	QList<NarratorNodeIfc *> new_nodes;
 	QList<GroupNode *> unmatchedGroups;
-	size=narrators.size();
+	size=groups.size();
 	for (int j=0;j<size;j++) {
-		unmatchedGroups.append(narrators[j]);
+		unmatchedGroups.append(groups[j]);
 	}
 	for (int j=0;j<size;j++) {
-		assert(narrators[j]->size()>0);
-		Narrator & n1=(*narrators[j])[0].getNarrator();
+		assert(groups[j]->size()>0);
+		Narrator & n1=(*groups[j])[0].getNarrator();
 		for (int k=j+1;k<size;k++) {
-			assert(narrators[k]->size()>0);
-			Narrator & n2=(*narrators[k])[0].getNarrator();
+			assert (groups[k]->size()>0);
+			Narrator & n2=(*groups[k])[0].getNarrator();
 			double eq_val=equal(n1,n2);
 			if (eq_val>threshold) {
-				NarratorNodeIfc * n1=&narrators[j]->getCorrespondingNarratorNode();
-				NarratorNodeIfc * n2=&narrators[k]->getCorrespondingNarratorNode();
-				NarratorNodeIfc * n_new=controller->getGraph()->mergeNodes((*narrators[j]),(*narrators[k]),&toDelete);
[…]

### 2011-08-01 trunk/ATSarf/src/case/graph_nodes.h

@@ -409,6 +409,7 @@ private:
 	GraphNarratorNode * graphNode;
 	QList<ChainNarratorNode *> list;
 	KeyList allKeys;
+	int lowestIndex:16, highestIndex:16;
 private:
 	friend class GraphNarratorNode;
 	friend class ChainNarratorNode;
@@ -419,18 +420,24 @@ protected:
 	friend class NarratorGraph;
 	GroupNode():GraphNodeItem() {
 		graphNode=NULL;
+		lowestIndex=-1;
+		highestIndex=-1;
 	}
 	GroupNode(NarratorGraph&g,GraphNarratorNode * gNode, ChainNarratorNode * cNode,QString key)
 		:GraphNodeItem(g),graphNode(gNode) {
 		list.append(cNode);
 		this->key=key;
 		cNode->setCorrespondingNarratorNodeGroup(this);
+		lowestIndex=cNode->getIndex();
+		highestIndex=lowestIndex;
 	}
 	GroupNode(NarratorGraph&g,GraphNarratorNode * gNode, ChainNarratorNode * cNode)
 		:GraphNodeItem(g),graphNode(gNode) {
 		list.append(cNode);
 		this->key=cNode->getKey();//cNode->getNarrator().getKey();
 		cNode->setCorrespondingNarratorNodeGroup(this);
+		lowestIndex=cNode->getIndex();
+		highestIndex=lowestIndex;
 	}
 
 	virtual NarratorNodeIfc & getChild(int index1,int ) {
@@ -453,21 +460,9 @@ public:
 		assert(index>=0 && index<size());//check redundant will be done inside [] for QList
 		return *list[index];
 	}
-	void addChainNode(NarratorGraph */*g*/,ChainNarratorNode & nar) { //we dont check for duplicates here
-	#if 0
-		for (int i=0;i<size();i++)
-		{
-			ChainNarratorNode * n2=&(*this)[i];
-			ChainNarratorNode * n1=&nar;
-			if (n1==n2)
-				return;
-			//assert (n1!=n2); //TODO...
-		}
-	#endif
-		//assert(nar.graphNode==NULL);
-		list.append(&nar);
-		nar.setCorrespondingNarratorNodeGroup(this);
-	}
+	void addChainNode(NarratorGraph * g,ChainNarratorNode & nar);  //we dont check for duplicates here
+	int getLowestIndex() {return lowestIndex;}
+	int getHighestIndex() {return highestIndex; }
[…]

### 2011-08-01 trunk/ATSarf/src/case/narrator_abstraction.h

@@ -206,9 +206,9 @@ public:
 		for (int i=0;i<m_narrator.count();i++) {
 			if (m_narrator[i]->getString().isEmpty())
 				continue;
-			if (m_narrator[i]->isNamePrim())
+			if (m_narrator[i]->isNamePrim()){
 				names[j].append(m_narrator[i]);
-			else {
+			} else {
 				NameConnectorPrim * c=(NameConnectorPrim*)m_narrator[i];
 				if (c->isPossessive()) {
 					possessives.append(c);

### 2011-08-04 trunk/tex/hadith.ijcnlp.tex
changed the biography clustering code (now calculates the center also)

@@
 \begin{itemize}
-\item We automate the segmentation of hadith books into separate narrations and extract chains of narrators into a partial order relation graph using a knowledge based approach with morphological analysis.
+\item We automate the segmentation of hadith books into separate narrations and extract chains of narrators 
+%into a partial order relation graph 
+using a knowledge based approach with morphological analysis.
+\item We optimize our analysis through utilizing punctuation and learning previously unknown names.
+\item We build from different chains of narrators a partial order relation graph based on merging common narrators.
 \item We introduce a specialized distance metric for Arabic names used in Arabic literature. 
-\item We provide a working and efficient implementation of
-our work. 
+\item We highly outperform the currently available narrator extraction tool. 
+\item We provide a working and efficient implementation of our work. 
 \end{itemize}
@@
 
+(Figure~\ref{f:chain} is confusing since it misses some information uses other symbols like 
+those of $\theta$ and is redundant with what follows)
+
 \subsection{The hadith controller}
@@
 between narrator connectors or contained name connectors.
+
+(explain better and merged information present about them in "Results" section)
 
@@
 
+\subsection{Speedup Optimizations for Large Datasets}
+
+\subsubsection{MultiKey Hashing}
+
+(explain what are the keys used for each value and why suitable)
+
 \section{Related work }
@@
 
-\transfalse
-\novocalize
-\begin{figure*}[tb]
-\center{
-\resizebox{1.8\columnwidth}{!}
-%{ \input{iTreeMohamada}}
-{\includegraphics{iTreeMohamada}}
-\caption{Output of iTree fails to detect \RL{m_hmdA}.}
-\label{f:mohamada}
-}
-\end{figure*}
-\transtrue
+%\transfalse
+%\novocalize
+%\begin{figure*}[tb]
+%\center{
+%\resizebox{1.8\columnwidth}{!}
+%%{ \input{iTreeMohamada}}
+%{\includegraphics{iTreeMohamada}}
+%\caption{Output of iTree fails to detect \RL{m_hmdA}.}
+%\label{f:mohamada}
+%}
+%\end{figure*}
+%\transtrue
 
@@
 The numbers we present in the tables are the result of a 
-manual verification of the correctness of the outputs of 
+manual verification of the correctness of %the outputs of 
 10\% of the narrations in the books under consideration.
+
+(discuss annotation and verification process and mention smg about inter-annotator agreement)
 
@@
 an Ikhbar word (\noVocRL{AdaT al-A_hbAar}) such as
-\noVocRL{qAl}, \noVocRL{.hadda_t}, etc...; 
+\noVocRL{qAl} ``said'', \noVocRL{.hadda_t} ``narrated'', etc...; 
 however, the Kafi and Istibsar books followed a flexible style to express the narration 
@@
 
-\subsection{On Unorganized hadith books}
-
-(report results on such hadith books)
+%\subsection{On Unorganized hadith books}
+%
+%(report results on such hadith books)
 
@@
 
-(show detailed statistics and compare against that of eNarrator)
+(show detailed statistics and compare against that of eNarrator which uses a similarity string comparison appraoch)
 
@@
 (report recall and precision on the chainnodes per graphnode, that is all what retains to correctness resulting from algorithm)
+
+\section{Significance \& Extensions -- Named Entity Extraction in Other Contexts}
+
+\subsection{Genealogical Lists in the Bible}
+
+(explain what will be equivalent of NMC and NRC and how it will work)
+
+(Here is a sophisticated example:
+``This is the book of Adam’s history.
+In the day of God’s creating Adam he made him in the likeness of God.
+Male and female he created them.
+After that he blessed them and called their name Man in the day of their being created.
+And Adam lived on for a hundred and thirty years.
+Then he became father to a son in his likeness, in his image, and called his name Seth.
+And the days of Adam after his fathering Seth came to be eight hundred years.
+Meanwhile he became father to sons and daughters.
+So all the days of Adam that he lived amounted to nine hundred and thirty years and he died.
+And Seth lived on for a hundred and five years.
+Then he became father to E′nosh.
+And after his fathering E′nosh Seth continued to live eight hundred and seven years.
+Meanwhile he became father to sons and daughters.
+So all the days of Seth amounted to nine hundred and twelve years and he died.
+And E′nosh lived on for ninety years.
+Then he became father to Ke′nan.
+And after his fathering Ke′nan E′nosh continued to live eight hundred and fifteen years.
+Meanwhile he became father to sons and daughters.
+So all the days of E′nosh amounted to nine hundred and five years and he died...'' (Genesis 5:1-11)
+
+Here is a simple example:
+
+``Adam, Seth, E′nosh,  Ke′nan, Ma‧hal′a‧lel, Ja′red,  E′noch, Me‧thu′se‧lah, La′mech,  Noah, Shem, Ham and Ja′pheth.
+The sons of Ja′pheth were Go′mer and Ma′gog and Ma′da‧i and Ja′van and Tu′bal and Me′shech and Ti′ras.
+And the sons of Go′mer were Ash′ke‧naz and Ri′phath and To‧gar′mah.
+And the sons of Ja′van were E‧li′shah and Tar′shish, Kit′tim and Ro′da‧nim.
+The sons of Ham were Cush and Miz′ra‧im, Put and Ca′naan.'' (1 Chronicles 1:1-8)
+)
+
+In many cases, such lists cover overlapping sections and when they do, they sometimes exhibit some seeming contradictions 
+due to nicknames and other reasons. Detecting those conflicts is a task that has been performed by countless 
+number of Bible students, theologians and historians. 
+Their results have been published in articles and books; however, almost all of them are in written format 
+and rarely displayed the genealogical graphs to make it easier for the reader to pinpoint those conflicts and visualize them.
+
+In addition, being able to create a comprehensive genealogical graph from all biblical accounts would greatly enhance its study,
+since readers would be able to better understand a character's genealogical context and life span. Such comprehensive graphs can be tagged
+with time durations between a father and his giving birth of his sons as is common in many biblical accounts...
+
+\subsection{Organizing Quoted Email Replies/Forwards}
+
+(we can use the "FROM:" and punctuation to extract the sender and the body of the email and organize a 
+received email into individual chains of replies/forwards.
+Also the Inbox can be merged into a POR of contacts with whom a user communicates and the history of their communication, as in the case of the hadith...)
+
+\subsection{Place Names in Travel Itineraries}
+
+(check if it can be applicable)
+
+(Here is an example from wikipedia ($http://wikitravel.org/en/Silk_Road$):
+
+-Northern Silk Route \\
+Dunhuang to Hami \\
+Hami to Turfan \\ 
+Turfan to Korla \\
+Korla to Kashgar
+
+-Middle Silk Route \\ 
+Dunhuang to Cherchen \\
+Cherchen to Korla \\
+Korla to Kashgar
+
+-Southern (Jade) Silk Route \\
+Dunhuang to Cherchen \\
+Cherchen to Khotan \\
+Khotan to Yarkand \\
+Yarkand to Kashgar \\
+)
+
+In many cases, these itineraries contain some common places and merging them in a POR is much more illustrative than having them in text format.
+Also, when electronic maps are available, such textual itineraries can automatically be turned into emphasized routes on maps.
+
+\subsection{Extending Arabic People Name Gazetteers}
+
+(briefly explain how this specific application has general value through contributing to the general task of entity extraction [learned names])
 

### 2011-08-04 trunk/ATSarf/src/case/biographies.h

@@ -39,8 +39,8 @@ public:
 		parse=new QPushButton("&Parse Biographies",this);
 		colorBiography=new QPushButton("Color &Biography",this);
 		colorNarrators=new QPushButton("Color &Narrators",this);
-		nmc_label=new QLabel("NMC min", this);
-		nrc_label=new QLabel("NRC min",this);
+		nmc_label=new QLabel("NMC max", this);
+		nrc_label=new QLabel("NRC max",this);
 		narr_label=new QLabel("NARR min",this);
 		reachability_label=new QLabel("REACH max",this);
 		nmc_max=new QTextEdit("1",this);

### 2011-08-04 trunk/ATSarf/src/case/graph.h

@@ -21,7 +21,7 @@ protected:
 	friend class GraphVisitorController;
 public:
 	virtual void initialize()=0;
-	virtual void visit(NarratorNodeIfc & n)=0;
+	virtual bool visit(NarratorNodeIfc & n)=0;
 	virtual void visit(NarratorNodeIfc & n1,NarratorNodeIfc & n2, int child_num)=0;
 	virtual void finish()=0;
 
@@ -171,7 +171,7 @@ public:
 			}
 		}
 	}
-	void visit(NarratorNodeIfc & n)
+	bool visit(NarratorNodeIfc & n)
 	{
 		if (keep_track_of_nodes && graph!=NULL) {
 			n.resetVisited(finishIndex);
@@ -185,16 +185,17 @@ public:
 			else
 				g->fillParents();
 		}
-		visitor->visit(n);
+		bool retVal=visitor->visit(n);
 
 		if (!keep_track_of_nodes)
-			return ;
+			return retVal;
 		if (graph!=NULL) {
 			n.setVisited(visitIndex);
 		}
 		else {
 			(*visitedMap)[&n]=VISITED;
 		}
+		return retVal;
 	}
 	bool isFinished( NarratorNodeIfc & node)
 	{
@@ -343,7 +344,7 @@ public:
 		QString s1=getAndInitializeDotNode(n1), s2=getAndInitializeDotNode(n2);
 		d_out<<s1<<"->"<<s2<<";\n";
 	}
-	virtual void visit(NarratorNodeIfc & n) //this is enough
+	virtual bool visit(NarratorNodeIfc & n) //this is enough
 	{
 		QString s=getAndInitializeDotNode(n);
 		displayChainNumsEndingJustAfter(n,s);
@@ -351,6 +352,7 @@ public:
 		if (n.isGraphNode())
 			out<<n.toString()<<"\n";
 	#endif
+		return true;
 	}
 	virtual void finishVisit(NarratorNodeIfc & ){}
 	virtual void detectedCycle(NarratorNodeIfc & n)
@@ -492,7 +494,7 @@ public:
 	}
 	void initialize(){ hash->clear();}
 	virtual void visit(NarratorNodeIfc & ,NarratorNodeIfc &, int ){	}
-	virtual void visit(NarratorNodeIfc & n) {
+	virtual bool visit(NarratorNodeIfc & n) {
 		assert(!n.isGroupNode());
 		if (n.isGraphNode()) {
[…]

### 2011-08-04 trunk/ATSarf/src/case/graph_nodes.h

@@ -130,6 +130,14 @@ public:
 					<<"<"<<n->CanonicalName()<<">:"<<i;*/
 		}
 	}
+	int getBiographIndexCount() {
+		NarratorNodeIfc & n=getCorrespondingNarratorNode();
+		return (n.indicies.size());
+	}
+	int getBiographyIndex(int i) {
+		NarratorNodeIfc & n=getCorrespondingNarratorNode();
+		return (n.indicies[i]);
+	}
 	bool hasBiographyIndex(int i) {
 		NarratorNodeIfc & n=getCorrespondingNarratorNode();
 		/*qDebug()<<"---"<<(long)&n<<"<"<<CanonicalName()<<">---";
@@ -137,7 +145,13 @@ public:
 			qDebug()<<n.indicies[j];*/
 		return (n.indicies.contains(i));
 	}
-
+	bool hasSomeBiographyIndex() {
+		NarratorNodeIfc & n=getCorrespondingNarratorNode();
+		return (!n.indicies.isEmpty());
+	}
+	void clearBiographyIndicies() {
+		indicies.clear();
+	}
 	void serialize(QDataStream &chainOut,NarratorGraph & graph) const {
 		chainOut<<isNull();
 		chainOut<<isGraphNode();

### 2011-08-04 ATSarf/src/case/graph_nodes.cpp
changed the biography clustering code (now calculates the center also)

@@ -366,9 +366,10 @@ public:
 	void initialize() {
 		assert(list.isEmpty());
 	}
-	void visit(NarratorNodeIfc & n) {
+	bool visit(NarratorNodeIfc & n) {
 		if (&n!=node)
 			list.append(&n);
+		return true;
 	}
 	virtual void visit(NarratorNodeIfc & ,NarratorNodeIfc & , int ){}
 	virtual void finish(){}

### 2011-08-04 ATSarf/src/case/hadithCommon.h

@@ -547,13 +547,15 @@
 	}
 
 #if 1
-	typedef struct stateData_ {
+	typedef struct StateData_ {
 		long  mainStructureStartIndex;
-		long narratorCount,narratorStartIndex,narratorEndIndex,nrcStartIndex,nrcEndIndex;
-		long  nmcCount, nrcCount,nameStartIndex,nmcStartIndex;
-		bool nmcValid;
-		bool ibn_or_3abid;
-		bool nrcPunctuation;
+		long narratorStartIndex,narratorEndIndex,nrcStartIndex,nrcEndIndex,nameStartIndex,nmcStartIndex;
+		int  nmcCount, nrcCount,narratorCount, bio_nrcCount ; //bio_nrcCount needed bc dont want finding an NRC to change the tolerance breaking count
+
+		bool nmcValid:1;
+		bool ibn_or_3abid:1;
+		bool nrcPunctuation:1;
+		int unused:29;
 
 		void initialize() {
 			nmcCount=0;
@@ -568,12 +570,13 @@
 			mainStructureStartIndex=0;
 			nrcStartIndex=0;
 			nrcEndIndex=0;
+			bio_nrcCount=0;
 		#ifdef PUNCTUATION
 			nrcPunctuation=false;
 		#endif
 		}
 
-	} stateData;
+	} StateData;
 	class HadithData {
 	public:
 		QString * text;
@@ -636,7 +639,7 @@
 
 
 #ifndef BUCKWALTER_INTERFACE
-	bool proceedInStateMachine(StateInfo &  stateInfo,HadithData *structures, stateData & currentData ) ;
+	bool proceedInStateMachine(StateInfo &  stateInfo,HadithData *structures, StateData & currentData ) ;
 #endif
 
 #endif

### 2011-09-17 tex/hadith.ijcnlp.tex
completed some graphs for the email section

@@
 \usepackage{fancyvrb}
-\usepackage[colorlinks=false]{hyperref}
+\usepackage[colorlinks=false,draft]{hyperref}
 
@@
 (Figure~\ref{f:chain} is confusing since it misses some information uses other symbols like 
-those of $\theta$ and is redundant anyways)
+those of $\theta$ and is redundant with what follows)
 
@@
 
-(explain better and merged information present about them in "Results" section)
-
 In addition to the controller, 
@@
 
-(explain why suitable)
-
-\subsubsection{Three-Level Hierarchy Structure for Graph Nodes}
-
-(possibly omit this info, since is complicated)
+(explain what are the keys used for each value and why suitable)
 
@@
 
-(explain what will be equivalent of NMC and NRC and how it will work
+Bible contains numerous genealogical lists which are of interest to Bible scholars, historians and critics. 
+The genealogies in the Bible trace key figures especially kings of the nation of Israel back to Adam; 
+the genealogy of Jesus is also recorded in the New testament by two of his followers: Matthew and Luke.
+
+%In many cases, 
+%there is an overlap between the many individual lists available across the Bible. When such an overlap occurs 
+%meticulous readers can detect some seeming contradictions, in many cases due to conflicts or omissions.
+%Detecting those conflicts is a task that has been performed by countless 
+%number of Bible students, theologians and historians.
+
+The importance of such lists was profound for the old nation of Israel. They helped establish the kingship lineage, determining 
+those eligible to work as priests and in the temple. The whole concept of land ownership was also tied to descent. Most importantly,
+genealogical links were preserved due to the possible privilege of being the ancestor of the awaited Messiah.~\cite{catholicEncyclopedia:Online}
+
+For many believers, genealogies are used as the main indicators of chronology especially when calculating the age of mankind~\cite{Thomas:92}.
+They where always used to defend the historicity of the events and characters recorded in the Bible in contrast to those present in other religious 
+texts of old nations which contain much fables and myths. However, it is known that when different genealogical lists overlap, they are
+not always identical and contain some seeming contradictions. Those conflicts are in many cases due to nicknames and omissions and need to be examined
+carefully and justified if possible.
+
+%In many cases, such lists cover overlapping sections and when they do, they sometimes exhibit some seeming contradictions 
+%due to nicknames and other reasons.
+Detecting those conflicts is a task that has been performed by countless 
+number of Bible students, theologians and historians~\cite{catholicEncyclopedia:Online,Belote:Online,completeBibleGenealogy:Online}. 
+Their results have been published in articles and books; however, most of them are in written format 
+and rarely display the genealogical graphs to make it easier for the reader to pinpoint those conflicts and visualize them.
+
+In addition, being able to create a comprehensive genealogical graph from all biblical accounts greatly enhances its study,
+since readers would be able to better understand a character's genealogical context and life span. Such comprehensive graphs can be tagged
+with time durations between a father and his giving birth of his sons as is common in many biblical accounts. 
+Attempts have been made for such endeavor, but resulted in manually performed unappealing pseudo-complete graphs as in \cite{Belote:Online,SoulLiberty:Online}.
+The main difficulty that hindered this task was the huge size of the resulting graphs whose layout becomes
+challenging. This difficulty is also present in the hadith problem we undertook in this paper; however, we were able to tackle this issue by 
+allowing for partial layout to be drawn on the fly as the user navigates in the POR which is already computed, 
+instead of precomputed complete layouting of the graph on which plain scrolling is used to navigate. (check the accompanying software).
+In all cases, it is noteworthy, that all genealogical trees/lineages available in any literature today are the product of manual work. 
+In what follows, we explain how NaGEMA can be easily manipulated to be utilized to generate this information automatically.
+
+%is in establishing the historicity of characters under consideration,
+%Such lists had and still have an utmost importance for Jews because heritage
+%according to their law dictates the land's ownership according to their inheritance which cannot be sold perminantly, they also
+
+Before presenting the necessary modifications for the state-machine used by NaGEMA, we present two representative biblical sections that
+contain genealogical lists. These sections contain overlapping linages, however stated with different styles: \\
+
+``This is the written account of {\bf Adam}'s line...
+
+%When God created man, he made him in the likeness of God. He created them male and female and blessed them. 
+%And when they were created, he called them `man'.
+
+When {\bf Adam} had lived 130 years, he had a son in his own likeness, in his own image; and he named him {\bf Seth}. 
+After Seth was born, {\bf Adam} lived 800 years and had other sons and daughters. 
+Altogether, {\bf Adam} lived 930 years, and then he died.
+
+When {\bf Seth} had lived 105 years, he became the father of {\bf Enosh}. 
+And after he became the father of {\bf Enosh}, {\bf Seth} lived 807 years and had other sons and daughters. 
+Altogether, {\bf Seth} lived 912 years, and then he died.
+
+When {\bf Enosh} had lived 90 years, he became the father of {\bf Kenan}. 
+And after he became the father of {\bf Kenan}, {\bf Enosh} lived 815 years and had other sons and daughters. 
+Altogether, {\bf Enosh} lived 905 years, and then he died.'' (Genesis 5:1-11) \\
+
+``{\bf Adam}, {\bf Seth}, {\bf Enosh}, {\bf Kenan}, {\bf Mahalalel}, {\bf Jared}, {\bf Enoch}, 
+{\bf Methuselah}, {\bf Lamech}, {\bf Noah}. 
+The sons of {\bf Noah}: {\bf Shem}, {\bf Ham} and {\bf Japheth}.
+The sons of {\bf Japheth}: {\bf Gomer}, {\bf Magog}, {\bf Madai}, {\bf Javan}, {\bf Tubal}, {\bf Meshech} and {\bf Tiras}.
+The sons of {\bf Gomer}: {\bf Ashkenaz}, {\bf Riphath} and {\bf Togarmah}.
+The sons of {\bf Javan}: {\bf Elishah}, {\bf Tarshish}, the {\bf Kittim} and the {\bf Rodanim}.
+The sons of {\bf Ham}: {\bf Cush}, {\bf Mizraim}, {\bf Put} and {\bf Canaan}.
+The sons of {\bf Cush}: {\bf Seba}, {\bf Havilah}, {\bf Sabta}, {\bf Raamah} and {\bf Sabteca}.
+The sons of {\bf Raamah}: {\bf Sheba} and {\bf Dedan}. 
+{\bf Cush} was the father of {\bf Nimrod}, who grew to be a mighty warrior on earth. ...'' (1 Chronicles 1:1-8) \\
+
+\begin{figure}[tb]
+\center{
+\resizebox{1\columnwidth}{!}
+%{\includegraphics{figs/narrator_chain_output.pdf}}
+{\includegraphics{figs/controller_bible.pdf}}
+\caption{Controller of biblical genealogies}
+\label{f:bible} 
+}
+\end{figure}
+
+The problem at hand can be stated as follows: Given the whole text of the Bible detect all sections that have genealogical lists
+and extract from them the names structured in a genealogical tree. The resemblance between this problem and that of hadith is evident; 
+but still there are some differences that should be highlighted. 
+First, in biblical genealogies, the individual names are simple in contrast to those present in hadith; 
+consequently, the concept of a `name connector (NMC)' is unnecessary in the problem at hand.
+Second, genealogies are not a flat sequence of narrators as in hadith but a branching tree of decedents; 
+for this reason, it is necessary to detect not just the names in the text but also the node to which they are to be attached.
+Those two differences are treated in the state-machine presented in Figure~\ref{f:bible} that contains all changes necessary to adapt NAGEMA to detect 
+biblical genealogies.
+
+The controller has three states that correspond to 
+the sentence structure under consideration. 
+State $\mathit{TEXT\_S}$ is the initial state and denotes 
+that the controller is outside the context of a genealogical list.
+State $\mathit{NAME\_S}$ denotes that the controller is in
+the middle of a list and has last met some names. 
+State $\mathit{SONS\_S}$ denotes that the controller is in
+the middle of sentence that lists the sons of a previously named ancestor such as sentences that appeared in the second quoted biblical texts.
+
+The conditions that appear in the controller, are as based on the following symbols:
+\begin{itemize}
+\item {\bf NEW\_NAME} denotes that the word at had is a name not present in the
+	currently built tree.
+\item {\bf LEAF\_NAME} denotes a name that appears as a leaf in the tree (i.e. has no
+	descendants attached to it).
+\item {\bf CORE\_NAME} denotes a name that has already descendants attached to it (i.e. 
+	non-leaf node).
+\item {\bf DC} denotes a descent connector such as "the son of", "became father to",
+	"begot", etc... that help indicate -- in addition to the names -- that the current text
+	is a genealogical list. They also have the role of indicating the type of link and its 
+	direction (father-son or son-father); such role is omitted in the state-machine for clarity.
+\item {\bf OTHER} denotes any other word.
+\end{itemize}
+The threshold $\theta_{\mathit{o}}$ 
+corresponds to the number of tolerated words that are not names 
+or descent connectors (DC); while 
+$i_{\mathit{o}}$ corresponds to their current count in the state-machine. 
+Another threshold which is used is $\mathit{N_{min}}$ which denotes the number of minimum
+levels that are necessarily available in the tree to be considered a valid genealogical
+tree instead of an erroneous one generated from a group of coincidental names. The
+concept of tolerance and minimum number of structures have been inspired from the
+hadith controller and are suitable for detecting texts that contain lists of similar
+type structures.
+
+The function $\mathit{n_1.setFather(n_2)}$ creates a node in the tree 
+for $\mathit{n_1}$ and creates
+a link between $\mathit{n_1}$ and the already present node $\mathit{n_2}$. 
+The function $\mathit{getLevels()}$ 
+returns the number of generations already present in the tree. This number
+assesses whether the currently built tree is an acceptable one or not upon
+transitioning from sate $\mathit{NAME\_S}$ to $\mathit{TEXT\_S}$
+
+We used the values of 3 and 20 for $\theta_{\mathit{o}}$ and 
+$N_{\mathit{min}}$ respectively, to run the experiments 
+we report in what follows.
+
+(report preliminary results on bible genealogy)
+
+\subsection{Organizing Quoted Email Replies/Forwards}
+
+\begin{figure}[b]
+\center{
+\resizebox{1\columnwidth}{!}
+%{\includegraphics{figs/narrator_chain_output.pdf}}
+{\includegraphics{figs/email1.pdf}}
+\caption{Thread tree generated from flat emails}
+\label{f:email1} 
+}
+\end{figure}
+
+\begin{figure}[tb]
+\center{
+\resizebox{0.7\columnwidth}{!}
+%{\includegraphics{figs/narrator_chain_output.pdf}}
+{\includegraphics{figs/email2.pdf}}
+\caption{Graph of contacts and other people known through the threads available}
+\label{f:email2} 
+}
+\end{figure}
+
+Emails have been one of the most important tool for communication in business, distant family members and friends.
+In many cases, one email may trigger a thread of replies and forwards that may circulate between different people. It may even
+end up after several replies being sent to someone new to the whole thread and he/she might consider navigating through
+the lengthy list of quoted replies/forwards to get a better understanding of the conversations which have been going prior to his
+receiving the email. However, the lengthy lists of conversations with some added indentation or symbols each possibly in a different
+format may be a cumbersome task to do. That's why email clients such as gmail try organizing each conversation in a separate tab for
+better use. Nonetheless, such clients are restricted to emails that are already available in the recipient's `Inbox' or `Sent Mails'.
+In other words, even if the email client receives an email which has 
+tens of separate replies/forwards but that are not previously available for the client, it 
+fails to organize them and instead keeps them as one list of quoted conversations.
+
+In hadith, we have a similar concept: a narration from the prophet that has been orally transmitted and ends up being written
+at the end with all the narrators in between being quoted by the last narrator who receives the narration. 
+It is interesting to be able to solve the `reincarnated' version of
+the problem that appears in our technological world.
+
+It is noteworthy that the problem at had is easier than its historical counterpart, due to the semi-structured nature of emails.
+For instance, the senders, co-recipients, subject, body and so on of the email follow clear rules. Since in our case, we need not 
+to mine the unstructured content of the email, the problem is clearly much easier than the ones presented earlier. 
+Consequently by considering the `From', `To', `CC' and `BCC' fields one can construct the `narrator' chains for conversation
+threads, each tagged with the body of the the individual reply/forward conversation. Consider a sample email tree that
+can be generated as shown in Figure~\ref{f:email1}. Notice that the four flat emails shown to the right of the figure,
+three of which lie in the `Inbox' and one of which in the `Sent Mails' were used to generate the more instructive structure
+to the left. Although in literature~\cite{Smith:00,Venolia:03} 
+alternative tree-like structures where proposed, however, they were restricted to emails
+available in the Inbox/Sent-Mails; consequently they will be unable to recover the top three levels of the tree generated through 
+the quoted previously unknown emails. % (search more extensively literature, if this is true)
+
+Furthermore, it can be envisioned that the Inbox or subset of it, can be merged into a graph structure of people with 
+whom a user communicates and the
+history of their communication as in the case of the hadith. 
+%This structure might be useful to extract other relations
+%such as the importance and priority of emails sent by certain contacts and for predicting any possibly forgotten recipient of an email to be sent.
+Consider a sample graph depicted in Figure~\ref{f:email2} 
+that shows the people involved in the thread of figure~\ref{f:email1} along with others of previous threads.
+With such a graph, the user can get to learn about people that are not known to him and are possible contacts because they are involved 
+in threads that reach him/her.
+
+%(we can use the "FROM:" and punctuation to extract the sender and the body of the email and organize a 
+%received email into individual chains of replies/forwards.
+Also the Inbox can be merged into a POR of contacts with whom a user communicates and the history of their communication, as in the case of the hadith...)
 
@@
 (check if it can be applicable)
+
+(Here is an example from wikipedia (\url{http://wikitravel.org/en/Silk_Road}):
+
+-Northern Silk Route \\
+Dunhuang to Hami \\
+Hami to Turfan \\ 
+Turfan to Korla \\
+Korla to Kashgar
+
+-Middle Silk Route \\ 
+Dunhuang to Cherchen \\
+Cherchen to Korla \\
+Korla to Kashgar
+
+-Southern (Jade) Silk Route \\
+Dunhuang to Cherchen \\
+Cherchen to Khotan \\
+Khotan to Yarkand \\
+Yarkand to Kashgar \\
+)
+
+In many cases, these itineraries contain some common places and merging them in a POR is much more illustrative than having them in text format.
+Also, when electronic maps are available, such textual itineraries can automatically be turned into emphasized routes on maps.
 
@@
 {\small \bibliography{fzAr}}
-
+\nocite{completeBibleGenealogy:Online}
 \end{document}

### 2011-09-17 ATSarf/src/case/bibleGeneology.h

@@ -1,0 +1,11 @@
+#ifndef BIBLEGENEOLOGY_H
+#define BIBLEGENEOLOGY_H
+
+#include <QString>
+#include "common.h"
+#include "ATMProgressIFC.h"
+
+int geneologyHelper(QString input_str,ATMProgressIFC *prgs);
+void geneology_initialize();
+
+#endif // BIBLEGENEOLOGY_H

### 2011-09-19 trunk/tex/hadith.ijcnlp.tex
completed some text in the section of extending gazetteers

@@
 
-(explain what will be equivalent of NMC and NRC and how it will work)
-
-(Here is a sophisticated example:
-``This is the book of Adam’s history.
-In the day of God’s creating Adam he made him in the likeness of God.
-Male and female he created them.
-After that he blessed them and called their name Man in the day of their being created.
-And Adam lived on for a hundred and thirty years.
-Then he became father to a son in his likeness, in his image, and called his name Seth.
-And the days of Adam after his fathering Seth came to be eight hundred years.
-Meanwhile he became father to sons and daughters.
-So all the days of Adam that he lived amounted to nine hundred and thirty years and he died.
-And Seth lived on for a hundred and five years.
-Then he became father to E′nosh.
-And after his fathering E′nosh Seth continued to live eight hundred and seven years.
-Meanwhile he became father to sons and daughters.
-So all the days of Seth amounted to nine hundred and twelve years and he died.
-And E′nosh lived on for ninety years.
-Then he became father to Ke′nan.
-And after his fathering Ke′nan E′nosh continued to live eight hundred and fifteen years.
-Meanwhile he became father to sons and daughters.
-So all the days of E′nosh amounted to nine hundred and five years and he died...'' (Genesis 5:1-11)
-
-Here is a simple example:
-
-``Adam, Seth, E′nosh,  Ke′nan, Ma‧hal′a‧lel, Ja′red,  E′noch, Me‧thu′se‧lah, La′mech,  Noah, Shem, Ham and Ja′pheth.
-The sons of Ja′pheth were Go′mer and Ma′gog and Ma′da‧i and Ja′van and Tu′bal and Me′shech and Ti′ras.
-And the sons of Go′mer were Ash′ke‧naz and Ri′phath and To‧gar′mah.
-And the sons of Ja′van were E‧li′shah and Tar′shish, Kit′tim and Ro′da‧nim.
-The sons of Ham were Cush and Miz′ra‧im, Put and Ca′naan.'' (1 Chronicles 1:1-8)
-)
-
-In many cases, such lists cover overlapping sections and when they do, they sometimes exhibit some seeming contradictions 
-due to nicknames and other reasons. Detecting those conflicts is a task that has been performed by countless 
-number of Bible students, theologians and historians. 
-Their results have been published in articles and books; however, almost all of them are in written format 
-and rarely displayed the genealogical graphs to make it easier for the reader to pinpoint those conflicts and visualize them.
-
-In addition, being able to create a comprehensive genealogical graph from all biblical accounts would greatly enhance its study,
+Bible contains numerous genealogical lists which are of interest to Bible scholars, historians and critics. 
+The genealogies in the Bible trace key figures especially kings of the nation of Israel back to Adam; 
+the genealogy of Jesus is also recorded in the New testament by two of his followers: Matthew and Luke.
+
+%In many cases, 
+%there is an overlap between the many individual lists available across the Bible. When such an overlap occurs 
+%meticulous readers can detect some seeming contradictions, in many cases due to conflicts or omissions.
+%Detecting those conflicts is a task that has been performed by countless 
+%number of Bible students, theologians and historians.
+
+The importance of such lists was profound for the old nation of Israel. They helped establish the kingship lineage, determining 
+those eligible to work as priests and in the temple. The whole concept of land ownership was also tied to descent. Most importantly,
+genealogical links were preserved due to the possible privilege of being the ancestor of the awaited Messiah.~\cite{catholicEncyclopedia:Online}
+
+For many believers, genealogies are used as the main indicators of chronology especially when calculating the age of mankind~\cite{Thomas:92}.
+They where always used to defend the historicity of the events and characters recorded in the Bible in contrast to those present in other religious 
+texts of old nations which contain much fables and myths. However, it is known that when different genealogical lists overlap, they are
+not always identical and contain some seeming contradictions. Those conflicts are in many cases due to nicknames and omissions and need to be examined
+carefully and justified if possible.
+
+%In many cases, such lists cover overlapping sections and when they do, they sometimes exhibit some seeming contradictions 
+%due to nicknames and other reasons.
+Detecting those conflicts is a task that has been performed by countless 
+number of Bible students, theologians and historians~\cite{catholicEncyclopedia:Online,Belote:Online,completeBibleGenealogy:Online}. 
+Their results have been published in articles and books; however, most of them are in written format 
+and rarely display the genealogical graphs to make it easier for the reader to pinpoint those conflicts and visualize them.
+
+In addition, being able to create a comprehensive genealogical graph from all biblical accounts greatly enhances its study,
 since readers would be able to better understand a character's genealogical context and life span. Such comprehensive graphs can be tagged
-with time durations between a father and his giving birth of his sons as is common in many biblical accounts...
+with time durations between a father and his giving birth of his sons as is common in many biblical accounts. 
+Attempts have been made for such endeavor, but resulted in manually performed unappealing pseudo-complete graphs as in \cite{Belote:Online,SoulLiberty:Online}.
+The main difficulty that hindered this task was the huge size of the resulting graphs whose layout becomes
+challenging. This difficulty is also present in the hadith problem we undertook in this paper; however, we were able to tackle this issue by 
+allowing for partial layout to be drawn on the fly as the user navigates in the POR which is already computed, 
+instead of precomputed complete layouting of the graph on which plain scrolling is used to navigate. (check the accompanying software).
+In all cases, it is noteworthy, that all genealogical trees/lineages available in any literature today are the product of manual work. 
+In what follows, we explain how NaGEMA can be easily manipulated to be utilized to generate this information automatically.
+
+%is in establishing the historicity of characters under consideration,
+%Such lists had and still have an utmost importance for Jews because heritage
+%according to their law dictates the land's ownership according to their inheritance which cannot be sold perminantly, they also
+
+Before presenting the necessary modifications for the state-machine used by NaGEMA, we present two representative biblical sections that
+contain genealogical lists. These sections contain overlapping linages, however stated with different styles: \\
+
+``This is the written account of {\bf Adam}'s line...
+
+%When God created man, he made him in the likeness of God. He created them male and female and blessed them. 
+%And when they were created, he called them `man'.
+
+When {\bf Adam} had lived 130 years, he had a son in his own likeness, in his own image; and he named him {\bf Seth}. 
+After Seth was born, {\bf Adam} lived 800 years and had other sons and daughters. 
+Altogether, {\bf Adam} lived 930 years, and then he died.
+
+When {\bf Seth} had lived 105 years, he became the father of {\bf Enosh}. 
+And after he became the father of {\bf Enosh}, {\bf Seth} lived 807 years and had other sons and daughters. 
+Altogether, {\bf Seth} lived 912 years, and then he died.
+
+When {\bf Enosh} had lived 90 years, he became the father of {\bf Kenan}. 
+And after he became the father of {\bf Kenan}, {\bf Enosh} lived 815 years and had other sons and daughters. 
+Altogether, {\bf Enosh} lived 905 years, and then he died.'' (Genesis 5:1-11) \\
+
+``{\bf Adam}, {\bf Seth}, {\bf Enosh}, {\bf Kenan}, {\bf Mahalalel}, {\bf Jared}, {\bf Enoch}, 
+{\bf Methuselah}, {\bf Lamech}, {\bf Noah}. 
+The sons of {\bf Noah}: {\bf Shem}, {\bf Ham} and {\bf Japheth}.
+The sons of {\bf Japheth}: {\bf Gomer}, {\bf Magog}, {\bf Madai}, {\bf Javan}, {\bf Tubal}, {\bf Meshech} and {\bf Tiras}.
+The sons of {\bf Gomer}: {\bf Ashkenaz}, {\bf Riphath} and {\bf Togarmah}.
+The sons of {\bf Javan}: {\bf Elishah}, {\bf Tarshish}, the {\bf Kittim} and the {\bf Rodanim}.
+The sons of {\bf Ham}: {\bf Cush}, {\bf Mizraim}, {\bf Put} and {\bf Canaan}.
+The sons of {\bf Cush}: {\bf Seba}, {\bf Havilah}, {\bf Sabta}, {\bf Raamah} and {\bf Sabteca}.
+The sons of {\bf Raamah}: {\bf Sheba} and {\bf Dedan}. 
+{\bf Cush} was the father of {\bf Nimrod}, who grew to be a mighty warrior on earth. ...'' (1 Chronicles 1:1-8) \\
+
+\begin{figure}[tb]
+\center{
+\resizebox{1\columnwidth}{!}
+%{\includegraphics{figs/narrator_chain_output.pdf}}
+{\includegraphics{figs/controller_bible.pdf}}
+\caption{Controller of biblical genealogies}
+\label{f:bible} 
+}
+\end{figure}
+
+The problem at hand can be stated as follows: Given the whole text of the Bible detect all sections that have genealogical lists
+and extract from them the names structured in a genealogical tree. The resemblance between this problem and that of hadith is evident; 
+but still there are some differences that should be highlighted. 
+First, in biblical genealogies, the individual names are simple in contrast to those present in hadith; 
+consequently, the concept of a `name connector (NMC)' is unnecessary in the problem at hand.
+Second, genealogies are not a flat sequence of narrators as in hadith but a branching tree of decedents; 
+for this reason, it is necessary to detect not just the names in the text but also the node to which they are to be attached.
+Those two differences are treated in the state-machine presented in Figure~\ref{f:bible} that contains all changes necessary to adapt NAGEMA to detect 
+biblical genealogies.
+
+The controller has three states that correspond to 
+the sentence structure under consideration. 
+State $\mathit{TEXT\_S}$ is the initial state and denotes 
+that the controller is outside the context of a genealogical list.
+State $\mathit{NAME\_S}$ denotes that the controller is in
+the middle of a list and has last met some names. 
+State $\mathit{SONS\_S}$ denotes that the controller is in
+the middle of sentence that lists the sons of a previously named ancestor such as sentences that appeared in the second quoted biblical texts.
+
+The conditions that appear in the controller, are as based on the following symbols:
+\begin{itemize}
+\item {\bf NEW\_NAME} denotes that the word at had is a name not present in the
+	currently built tree.
+\item {\bf LEAF\_NAME} denotes a name that appears as a leaf in the tree (i.e. has no
+	descendants attached to it).
+\item {\bf CORE\_NAME} denotes a name that has already descendants attached to it (i.e. 
+	non-leaf node).
+\item {\bf DC} denotes a descent connector such as "the son of", "became father to",
+	"begot", etc... that help indicate -- in addition to the names -- that the current text
+	is a genealogical list. They also have the role of indicating the type of link and its 
+	direction (father-son or son-father); such role is omitted in the state-machine for clarity.
+\item {\bf OTHER} denotes any other word.
+\end{itemize}
+The threshold $\theta_{\mathit{o}}$ 
+corresponds to the number of tolerated words that are not names 
+or descent connectors (DC); while 
+$i_{\mathit{o}}$ corresponds to their current count in the state-machine. 
+Another threshold which is used is $\mathit{N_{min}}$ which denotes the number of minimum
+levels that are necessarily available in the tree to be considered a valid genealogical
+tree instead of an erroneous one generated from a group of coincidental names. The
+concept of tolerance and minimum number of structures have been inspired from the
+hadith controller and are suitable for detecting texts that contain lists of similar
+type structures.
+
+The function $\mathit{n_1.setFather(n_2)}$ creates a node in the tree 
+for $\mathit{n_1}$ and creates
+a link between $\mathit{n_1}$ and the already present node $\mathit{n_2}$. 
+The function $\mathit{getLevels()}$ 
+returns the number of generations already present in the tree. This number
+assesses whether the currently built tree is an acceptable one or not upon
+transitioning from sate $\mathit{NAME\_S}$ to $\mathit{TEXT\_S}$
+
+We used the values of 3 and 20 for $\theta_{\mathit{o}}$ and 
+$N_{\mathit{min}}$ respectively, to run the experiments 
+we report in what follows.
+
+(report preliminary results on bible genealogy)
 
@@
 
-(we can use the "FROM:" and punctuation to extract the sender and the body of the email and organize a 
-received email into individual chains of replies/forwards.
+\begin{figure}[b]
+\center{
+\resizebox{1\columnwidth}{!}
+%{\includegraphics{figs/narrator_chain_output.pdf}}
+{\includegraphics{figs/email1.pdf}}
+\caption{Thread tree generated from flat emails}
+\label{f:email1} 
+}
+\end{figure}
+
+\begin{figure}[tb]
+\center{
+\resizebox{0.7\columnwidth}{!}
+%{\includegraphics{figs/narrator_chain_output.pdf}}
+{\includegraphics{figs/email2.pdf}}
+\caption{Graph of contacts and other people known through the threads available}
+\label{f:email2} 
+}
+\end{figure}
+
+Emails have been one of the most important tool for communication in business, distant family members and friends.
+In many cases, one email may trigger a thread of replies and forwards that may circulate between different people. It may even
+end up after several replies being sent to someone new to the whole thread and he/she might consider navigating through
+the lengthy list of quoted replies/forwards to get a better understanding of the conversations which have been going prior to his
+receiving the email. However, the lengthy lists of conversations with some added indentation or symbols each possibly in a different
+format may be a cumbersome task to do. That's why email clients such as Gmail try organizing each conversation in a separate tab for
+better use. Nonetheless, such clients are restricted to emails that are already available in the recipient's `Inbox' or `Sent Mails'.
+In other words, even if the email client receives an email which has 
+tens of separate replies/forwards but that are not previously available for the client, it 
+fails to organize them and instead keeps them as one list of quoted conversations.
+
+In hadith, we have a similar concept: a narration from the prophet that has been orally transmitted and ends up being written
+at the end with all the narrators in between being quoted by the last narrator who receives the narration. 
+It is interesting to be able to solve the `reincarnated' version of
+the problem that appears in our technological world.
+
+It is noteworthy that the problem at had is easier than its historical counterpart, due to the semi-structured nature of emails.
+For instance, the senders, co-recipients, subject, body and so on of the email follow clear rules. Since in our case, we need not 
+to mine the unstructured content of the email, the problem is clearly much easier than the ones presented earlier. 
+Consequently by considering the `From', `To', `CC' and `BCC' fields one can construct the `narrator' chains for conversation
+threads, each tagged with the body of the the individual reply/forward conversation. Consider a sample email tree that
+can be generated as shown in Figure~\ref{f:email1}. Notice that the four flat emails shown to the right of the figure,
+three of which lie in the `Inbox' and one of which in the `Sent Mails' were used to generate the more instructive structure
+to the left. Although in literature~\cite{Smith:00,Venolia:03} 
+alternative tree-like structures where proposed, however, they were restricted to emails
+available in the Inbox/Sent-Mails; consequently they will be unable to recover the top three levels of the tree generated through 
+the quoted previously unknown emails. % (search more extensively literature, if this is true)
+
+Furthermore, it can be envisioned that the Inbox or subset of it, can be merged into a graph structure of people with 
+whom a user communicates and the
+history of their communication as in the case of the hadith. 
+%This structure might be useful to extract other relations
+%such as the importance and priority of emails sent by certain contacts and for predicting any possibly forgotten recipient of an email to be sent.
+Consider a sample graph depicted in Figure~\ref{f:email2} 
+that shows the people involved in the thread of figure~\ref{f:email1} along with others of previous threads.
+With such a graph, the user can get to learn about people that are not known to him and are possible contacts because they are involved 
+in threads that reach him/her.
+
+%(we can use the "FROM:" and punctuation to extract the sender and the body of the email and organize a 
+%received email into individual chains of replies/forwards.
 Also the Inbox can be merged into a POR of contacts with whom a user communicates and the history of their communication, as in the case of the hadith...)
@@
 
-(check if it can be applicable)
-
-(Here is an example from wikipedia ($http://wikitravel.org/en/Silk_Road$):
-
--Northern Silk Route \\
-Dunhuang to Hami \\
-Hami to Turfan \\ 
-Turfan to Korla \\
-Korla to Kashgar
-
--Middle Silk Route \\ 
-Dunhuang to Cherchen \\
-Cherchen to Korla \\
-Korla to Kashgar
-
--Southern (Jade) Silk Route \\
-Dunhuang to Cherchen \\
-Cherchen to Khotan \\
-Khotan to Yarkand \\
-Yarkand to Kashgar \\
-)
-
-In many cases, these itineraries contain some common places and merging them in a POR is much more illustrative than having them in text format.
-Also, when electronic maps are available, such textual itineraries can automatically be turned into emphasized routes on maps.
+Travel Itineraries are plans of travel or routes. They are usually used as guides for travelers. They list different
+places through which the route to the destination passes. 
+In many cases, those guides are provided in written format
+It is also common that alternative routes be suggested to the same destination.
+Consider the example below for the itineraries known as the `Silk Road'\footnote{from \url{http://wikitravel.org/en/Silk_Road}}:
+
+{\setlength{\parindent}{20pt} 
+\setlength{\parskip}{1ex}
+\indent-Northern Silk Route \\
+\indent \indent Dunhuang to Hami \\
+\indent \indent Hami to Turfan \\ 
+\indent \indent Turfan to Korla \\
+\indent \indent Korla to Kashgar
+
+\indent -Middle Silk Route \\ 
+\indent \indent Dunhuang to Cherchen \\
+\indent \indent Cherchen to Korla \\
+\indent \indent Korla to Kashgar
+
+\indent -Southern (Jade) Silk Route \\
+\indent \indent Dunhuang to Cherchen \\
+\indent \indent Cherchen to Khotan \\
+\indent \indent Khotan to Yarkand \\
+\indent \indent Yarkand to Kashgar \\ }
+
+In a similar manner to extracting hadith, a controller that looks for aggregations of Place names
+with the NRC equivalent being phrases such as `to', `pass in', `stop at', etc... 
+Afterwards, itineraries can be merged into a POR by merging common locations.
+Clearly, such a display is much more illustrative than having them in text format.
+Also, when electronic maps are available, POR's can automatically be turned into emphasized routes on maps.
 
@@
 
-(briefly explain how this specific application has general value through contributing to the general task of entity extraction [learned names])
+The problem at hand has significant contribution to the general task of Named Entity Recognition (NER). It has been noted that
+gazetteers play an important role in the accuracy of knowledge-based approaches and bootstrapping learning
+approaches~\cite{Kozareva:06,Carlson:09,Toral:06,Benajiba:07,Kazama:08,Nadeau:06}. Nonetheless, although gazetteers result in high accuracy,
+they are hard to obtain and mantain. % They are also language and domain-dependent. 
+Consequently, many automatic methods have been proposed to generate extensive gazetteers instead of relying on the time-consuming manual work
+to compile such lists. Proposed methods include: writing some rules for the context of the named entities targeted such as propositions that
+introduce them, also known as trigger gazetteers~\cite{Kozareva:06,Toral:06}; utilizing free online encyclopedias such as Wikipedia which are 
+continually mantained~\cite{Benajiba:07,Kazama:08}; in addition to reliance on web search engines starting from a small list of words and extracting
+similar that appear in the extracted pages~\cite{Nadeau:06}.
+
+As previously noted, when solving the hadith problem, we were able to learn much previously unknown names using their context. We have used methods
+relating to those found in~\cite{Traboulsi:09,Kozareva:06}; however, our learned lists appear to be more accurate than their counterparts due to the
+the more characteristics of the hadith corpus and the wider context on which we rely on. Particularly, we learn only names that appear in the hadith sanad
+detected through the presence of an aggregation of previously known names in its neighborhood along with name connectors such as narration verbs and 
+prepositions. In Table X, we compare the results of our learning methods to evaluate the role of the sand context. Clearly, the sanad context provides 
+a higher confidence in the accuracy of our extracted names. Accordingly, our generated gazetteers are superior than those extracted otherwise and 
+enrich the task of Arabic Named Entity Extraction resources. The same applies to that of Place names that can be discovered in the context of travel
+itineraries and the people names that appear in biblical genealogical lists. Since the latter sources are available in multitude of languages, they
+can be a rich source for building gazetteers for resource-poor languages.
+
+(present the necessary table and expand upon its significance)
 

### 2011-09-19 trunk/tex/hadith.ijcnlp.tex
updated genealogy code, but still some issue in corpus present to be solved

@@
 introduce them, also known as trigger gazetteers~\cite{Kozareva:06,Toral:06}; utilizing free online encyclopedias such as Wikipedia which are 
-continually mantained~\cite{Benajiba:07,Kazama:08}; in addition to reliance on web search engines starting from a small list of words and extracting
+continually maintained~\cite{Benajiba:07,Kazama:08}; in addition to reliance on web search engines starting from a small list of words and extracting
 similar that appear in the extracted pages~\cite{Nadeau:06}.

### 2011-09-19 ATSarf/src/case/hadith.h
updated genealogy code, but still some issue in corpus present to be solved

@@ -62,6 +62,8 @@
 #define SAVE_MERGED_GRAPH_ONLY
 #define DESERIALIZE_POR
 //#define DEBUG_BIOGRAPHY_CLUSTERING
+#define BIBLE
+
 
 int hadithHelper(QString input_str,ATMProgressIFC *prgs);
 void hadith_initialize();

### 2011-09-19 ATSarf/src/gui/test.cpp

@@ -11,6 +11,7 @@
 #include "text_handling.h"
 #include "diacritics.h"
 #include "timeRecognizer.h"
+#include "bibleGeneology.h"
 #include <sys/time.h>
 
 
@@ -55,8 +56,7 @@ int morphology(QString input_str,ATMProgressIFC *) {
 	return 0;
 }
 int hadith(QString input_str,ATMProgressIFC * prg) {
-	for (int i=0;i<REPETITIONS;i++)
-	{
+	for (int i=0;i<REPETITIONS;i++) {
 		timeval tim;
 		gettimeofday(&tim,NULL);
 		double t1=tim.tv_sec+(tim.tv_usec/1000000.0);
@@ -81,6 +81,9 @@ int test(QString inputString,ATMProgressIFC * prg) {
 		return -1;
 #elif 0
 	if (biographyHelper(inputString,prg))
+		return -1;
+#elif defined(BIBLE)
+	if (genealogyHelper(inputString,prg))
 		return -1;
 #elif defined(DESERIALIZE_POR)
 	if (deserializeGraph(inputString,prg))

### 2011-09-21 tex/hadith.ijcnlp.tex
some updates for code of bible genealogies

@@
 
-Bible contains numerous genealogical lists which are of interest to Bible scholars, historians and critics. 
-The genealogies in the Bible trace key figures especially kings of the nation of Israel back to Adam; 
-the genealogy of Jesus is also recorded in the New testament by two of his followers: Matthew and Luke.
-
-%In many cases, 
-%there is an overlap between the many individual lists available across the Bible. When such an overlap occurs 
-%meticulous readers can detect some seeming contradictions, in many cases due to conflicts or omissions.
-%Detecting those conflicts is a task that has been performed by countless 
-%number of Bible students, theologians and historians.
-
-The importance of such lists was profound for the old nation of Israel. They helped establish the kingship lineage, determining 
-those eligible to work as priests and in the temple. The whole concept of land ownership was also tied to descent. Most importantly,
-genealogical links were preserved due to the possible privilege of being the ancestor of the awaited Messiah.~\cite{catholicEncyclopedia:Online}
-
-For many believers, genealogies are used as the main indicators of chronology especially when calculating the age of mankind~\cite{Thomas:92}.
-They where always used to defend the historicity of the events and characters recorded in the Bible in contrast to those present in other religious 
-texts of old nations which contain much fables and myths. However, it is known that when different genealogical lists overlap, they are
-not always identical and contain some seeming contradictions. Those conflicts are in many cases due to nicknames and omissions and need to be examined
-carefully and justified if possible.
-
-%In many cases, such lists cover overlapping sections and when they do, they sometimes exhibit some seeming contradictions 
-%due to nicknames and other reasons.
-Detecting those conflicts is a task that has been performed by countless 
-number of Bible students, theologians and historians~\cite{catholicEncyclopedia:Online,Belote:Online,completeBibleGenealogy:Online}. 
-Their results have been published in articles and books; however, most of them are in written format 
-and rarely display the genealogical graphs to make it easier for the reader to pinpoint those conflicts and visualize them.
-
-In addition, being able to create a comprehensive genealogical graph from all biblical accounts greatly enhances its study,
-since readers would be able to better understand a character's genealogical context and life span. Such comprehensive graphs can be tagged
-with time durations between a father and his giving birth of his sons as is common in many biblical accounts. 
-Attempts have been made for such endeavor, but resulted in manually performed unappealing pseudo-complete graphs as in \cite{Belote:Online,SoulLiberty:Online}.
-The main difficulty that hindered this task was the huge size of the resulting graphs whose layout becomes
-challenging. This difficulty is also present in the hadith problem we undertook in this paper; however, we were able to tackle this issue by 
-allowing for partial layout to be drawn on the fly as the user navigates in the POR which is already computed, 
-instead of precomputed complete layouting of the graph on which plain scrolling is used to navigate. (check the accompanying software).
-In all cases, it is noteworthy, that all genealogical trees/lineages available in any literature today are the product of manual work. 
-In what follows, we explain how NaGEMA can be easily manipulated to be utilized to generate this information automatically.
-
-%is in establishing the historicity of characters under consideration,
-%Such lists had and still have an utmost importance for Jews because heritage
-%according to their law dictates the land's ownership according to their inheritance which cannot be sold perminantly, they also
-
-Before presenting the necessary modifications for the state-machine used by NaGEMA, we present two representative biblical sections that
-contain genealogical lists. These sections contain overlapping linages, however stated with different styles: \\
-
+Biblical genealogical lists describe family trees up to Adam and are of interest to Bible scholars including historians and critics. 
+They trace key figures such as Israelite kings and prophets. Figure~\ref{f:gene1} shows an excerpt from Genesis 5:1-11.
+
+\begin{figure}\centering
+\resizebox{0.9\columnwidth}{!}{
+\parbox[t]{1\columnwidth}{
+\tiny
 ``This is the written account of {\bf Adam}'s line...
@@
 And after he became the father of {\bf Kenan}, {\bf Enosh} lived 815 years and had other sons and daughters. 
-Altogether, {\bf Enosh} lived 905 years, and then he died.'' (Genesis 5:1-11) \\
-
+Altogether, {\bf Enosh} lived 905 years, and then he died.''
+}}
+\caption{Genealogical list from Genesis 5:1-11}
+\end{figure}
+
+\begin{figure}\centering
+\resizebox{0.9\columnwidth}{!}{
+\parbox[t]{1\columnwidth}{
+\tiny
 ``{\bf Adam}, {\bf Seth}, {\bf Enosh}, {\bf Kenan}, {\bf Mahalalel}, {\bf Jared}, {\bf Enoch}, 
@@
 The sons of {\bf Raamah}: {\bf Sheba} and {\bf Dedan}. 
-{\bf Cush} was the father of {\bf Nimrod}, who grew to be a mighty warrior on earth. ...'' (1 Chronicles 1:1-8) \\
+{\bf Cush} was the father of {\bf Nimrod}, who grew to be a mighty warrior on earth. ...'' }}
+\caption{Genealogical list from 1 Chronicles 1:1-8}
+\end{figure}
+
+%In many cases, 
+%there is an overlap between the many individual lists available across the Bible. When such an overlap occurs 
+%meticulous readers can detect some seeming contradictions, in many cases due to conflicts or omissions.
+%Detecting those conflicts is a task that has been performed by countless 
+%number of Bible students, theologians and historians.
+
+%The importance of such lists was profound for the old nation of Israel. They 
+The genealogical lists established the kingship lineage and determined
+rights and privileges in the temple. Genealogical lists also determined descent with all its implications in biblical times
+such as the concept of reverted ownership of sold real estate; every fifty years ownership of sold land reverted to the 
+original owners and their inheritors. Genealogical lists were preserved due to the possible privilege of being the ancestor of the awaited Messiah.~\cite{catholicEncyclopedia:Online}
+
+For many believers, genealogies are used as the main indicators of chronology especially when calculating the age of mankind~\cite{Thomas:92}.
+Believers used them to defend the historicity of Biblical events and characters against other accounts from other 
+texts thought of as fables and myths. 
+However, separate genealogical lists that overlap suffer from consistency issues. 
+For example, Matthew and Luke report two different genealogies of Jesus. 
+Several biblical researchers studied the consistency of the lists~\cite{catholicEncyclopedia:Online,Belote:Online,completeBibleGenealogy:Online}, 
+and published articles and books on the topic. Up to our knowledge very few include graphical illustrations such as family trees and graphs 
+which makes it easier for the reader to visualize the differences. 
+Our methodology can automatically create such illustrations.
+
+create a comprehensive genealogical graph from all biblical accounts greatly enhances its study,
+since readers would be able to better understand a character's genealogical context and life span. Such comprehensive graphs can be tagged
+with time durations between a father and his giving birth of his sons as is common in many biblical accounts.
+
+Work such as~\cite{Belote:Online,SoulLiberty:Online} constructed partial genealogical graphs manually from several lists.
+This work is short from being comprehensive.
+
+In what follows, we explain how NaGEMA can be easily manipulated to be utilized 
+to generate this information automatically.
+
+%The main difficulty that hindered this task was the huge size of the resulting graphs whose layout becomes
+%challenging. This difficulty is also present in the hadith problem we undertook in this paper; however, we were able to tackle this issue by 
+%allowing for partial layout to be drawn on the fly as the user navigates in the POR which is already computed, 
+%instead of precomputed complete layouting of the graph on which plain scrolling is used to navigate. (check the accompanying software).
+%In all cases, it is noteworthy, that all genealogical trees/lineages available in any literature today are the product of manual work.
+
+%is in establishing the historicity of characters under consideration,
+%Such lists had and still have an utmost importance for Jews because heritage
+%according to their law dictates the land's ownership according to their inheritance which cannot be sold perminantly, they also
+
+%Before presenting the necessary modifications for the state-machine used by NaGEMA, we present two representative biblical sections that
+%contain genealogical lists. 
+%NaGEMA can handle 
+%These sections contain overlapping linages, however stated with different styles: \\
 
@@
 
-The problem at hand can be stated as follows: Given the whole text of the Bible detect all sections that have genealogical lists
-and extract from them the names structured in a genealogical tree. The resemblance between this problem and that of hadith is evident; 
+Given the whole text of the Bible, NaGEMA detects all sections that have genealogical lists
+and extract from them the names structured in a genealogical DAG. The resemblance between this problem and that of hadith is evident; 
 but still there are some differences that should be highlighted. 
@@
 end up after several replies being sent to someone new to the whole thread and he/she might consider navigating through
-the lengthy list of quoted replies/forwards to get a better understanding of the conversations which have been going prior to his
+the lengthy list of quoted replies/forwards to get a better understanding of the conversations which have been going prior to 
 receiving the email. However, the lengthy lists of conversations with some added indentation or symbols each possibly in a different
-format may be a cumbersome task to do. That's why email clients such as gmail try organizing each conversation in a separate tab for
+format may be a cumbersome task to do. That's why email clients such as Gmail try organizing each conversation in a separate tab for
 better use. Nonetheless, such clients are restricted to emails that are already available in the recipient's `Inbox' or `Sent Mails'.
@@
 
-(check if it can be applicable)
-
-(Here is an example from wikipedia (\url{http://wikitravel.org/en/Silk_Road}):
-
--Northern Silk Route \\
-Dunhuang to Hami \\
-Hami to Turfan \\ 
-Turfan to Korla \\
-Korla to Kashgar
-
--Middle Silk Route \\ 
-Dunhuang to Cherchen \\
-Cherchen to Korla \\
-Korla to Kashgar
-
--Southern (Jade) Silk Route \\
-Dunhuang to Cherchen \\
-Cherchen to Khotan \\
-Khotan to Yarkand \\
-Yarkand to Kashgar \\
-)
-
-In many cases, these itineraries contain some common places and merging them in a POR is much more illustrative than having them in text format.
-Also, when electronic maps are available, such textual itineraries can automatically be turned into emphasized routes on maps.
+Travel Itineraries are plans of travel or routes. They are usually used as guides for travelers. They list different
+places through which the route to the destination passes. 
+In many cases, those guides are provided in written format
+It is also common that alternative routes be suggested to the same destination.
+Consider the example below for the itineraries known as the `Silk Road'\footnote{from \url{http://wikitravel.org/en/Silk_Road}}:
+
+{\setlength{\parindent}{20pt} 
+\setlength{\parskip}{1ex}
+\indent-Northern Silk Route \\
+\indent \indent Dunhuang to Hami \\
+\indent \indent Hami to Turfan \\ 
+\indent \indent Turfan to Korla \\
+\indent \indent Korla to Kashgar
+
+\indent -Middle Silk Route \\ 
+\indent \indent Dunhuang to Cherchen \\
+\indent \indent Cherchen to Korla \\
+\indent \indent Korla to Kashgar
+
+\indent -Southern (Jade) Silk Route \\
+\indent \indent Dunhuang to Cherchen \\
+\indent \indent Cherchen to Khotan \\
+\indent \indent Khotan to Yarkand \\
+\indent \indent Yarkand to Kashgar \\ }
+
+In a similar manner to extracting hadith, a controller that looks for aggregations of Place names
+with the NRC equivalent being phrases such as `to', `pass in', `stop at', etc... 
+Afterwards, itineraries can be merged into a POR by merging common locations.
+Clearly, such a display is much more illustrative than having them in text format.
+Also, when electronic maps are available, POR's can automatically be turned into emphasized routes on maps.
 
@@
 
-(briefly explain how this specific application has general value through contributing to the general task of entity extraction [learned names])
+The problem at hand has significant contribution to the general task of Named Entity Recognition (NER). It has been noted that
+gazetteers play an important role in the accuracy of knowledge-based approaches and bootstrapping learning
+approaches~\cite{Kozareva:06,Carlson:09,Toral:06,Benajiba:07,Kazama:08,Nadeau:06}. Nonetheless, although gazetteers result in high accuracy,
+they are hard to obtain and mantain. % They are also language and domain-dependent. 
+Consequently, many automatic methods have been proposed to generate extensive gazetteers instead of relying on the time-consuming manual work
+to compile such lists. Proposed methods include: writing some rules for the context of the named entities targeted such as propositions that
+introduce them, also known as trigger gazetteers~\cite{Kozareva:06,Toral:06}; utilizing free online encyclopedias such as Wikipedia which are 
+continually maintained~\cite{Benajiba:07,Kazama:08}; in addition to reliance on web search engines starting from a small list of words and extracting
+similar that appear in the extracted pages~\cite{Nadeau:06}.
+
+As previously noted, when solving the hadith problem, we were able to learn much previously unknown names using their context. We have used methods
+relating to those found in~\cite{Traboulsi:09,Kozareva:06}; however, our learned lists appear to be more accurate than their counterparts due to the
+the more characteristics of the hadith corpus and the wider context on which we rely on. Particularly, we learn only names that appear in the hadith sanad
+detected through the presence of an aggregation of previously known names in its neighborhood along with name connectors such as narration verbs and 
+prepositions. In Table X, we compare the results of our learning methods to evaluate the role of the sand context. Clearly, the sanad context provides 
+a higher confidence in the accuracy of our extracted names. Accordingly, our generated gazetteers are superior than those extracted otherwise and 
+enrich the task of Arabic Named Entity Extraction resources. The same applies to that of Place names that can be discovered in the context of travel
+itineraries and the people names that appear in biblical genealogical lists. Since the latter sources are available in multitude of languages, they
+can be a rich source for building gazetteers for resource-poor languages.
+
+(present the necessary table and expand upon its significance)
 

### 2011-09-21 ATSarf/src/case/bibleGeneology.cpp

@@ -20,8 +20,6 @@
 #include "textParsing.h"
 #include "common.h"
 
-#define GET_WAW
-
 
 enum WordType { NEW_NAME,CORE_NAME, LEAF_NAME, DC,ENDING_PUNC, OTHER};
 enum StateType { TEXT_S , NAME_S, SONS_S};
@@ -75,17 +73,6 @@ inline void display(QString t) {
 	#define display(c)
 #endif
 
-class GeneologyParameters {
-public:
-	unsigned int theta_0:10;
-	unsigned int N_min:10;
-	unsigned int unused:12;
-	GeneologyParameters() {
-		theta_0=20;
-		N_min=3;
-	}
-};
-
 GeneologyParameters geneologyParameters;
 QList<int> bits_gene_NAME;
 
@@ -127,8 +114,7 @@ public:
 #ifndef COMPARE_TO_BUCKWALTER
 	bool on_match()	{
 		solution_position * S_inf=Stem->computeFirstSolution();
-		do
-		{
+		do {
 			stem_info=Stem->solution;
 		#ifdef GET_AFFIXES_ALSO
 			solution_position * p_inf=Prefix->computeFirstSolution();
@@ -180,12 +166,8 @@ public:
 		if (info.finish>info.start) { //more than one letter to be tested for being a name
 			int bitsNamesSize=bits_gene_NAME.size();
 			for (int i=0;i<bitsNamesSize;i++) {
-				if (stem_info->abstract_categories.getBit(bits_gene_NAME[i])
-					#ifdef REFINEMENTS
-						&& Suffix->info.finish-Suffix->info.start<0 && Stem->info.finish>Stem->info.start) //i.e. has no suffix and stem > a letter
-					#else
-						)
-					#endif
+				if (stem_info->abstract_categories.getBit(bits_gene_NAME[i]))
+						//&& Suffix->info.finish-Suffix->info.start<0 && Stem->info.finish>Stem->info.start) //i.e. has no suffix and stem > a letter
 				{
 					name=true;
 					if (info.finish>finish_pos) {
@@ -205,22 +187,18 @@ public:
 };
 
 void geneology_initialize() {
-
-	long abstract_COMPOUND_NAMES=database_info.comp_rules->getAbstractCategoryID("Compound Names");
-	int bit_COMPOUND_NAMES=database_info.comp_rules->getAbstractCategoryBitIndex(abstract_COMPOUND_NAMES);
[…]

### 2011-09-21 ATSarf/src/case/bibleGeneology.h

@@ -5,6 +5,23 @@
 #include "common.h"
 #include "ATMProgressIFC.h"
 
+#define GET_WAW
+#define TRUST_OLD
+#define GENEOLOGYDEBUG
+
+class GeneologyParameters {
+public:
+	unsigned int theta_0:10;
+	unsigned int N_min:10;
+	unsigned int unused:12;
+	GeneologyParameters() {
+		theta_0=30;
+		N_min=5;
+	}
+};
+
+extern GeneologyParameters geneologyParameters;
+
 int genealogyHelper(QString input_str,ATMProgressIFC *prgs);
 void geneology_initialize();
 

### 2011-09-21 ATSarf/src/gui/mainwindow.cpp

@@ -56,11 +56,11 @@ void MainWindow::tag(int start, int length,QColor color, bool textcolor){
 		c.movePosition(QTextCursor::Left,QTextCursor::MoveAnchor,-diff);
 	c.movePosition(QTextCursor::Right,QTextCursor::KeepAnchor,length);
 #else
-	if (length>200) {
+	/*if (length>200) {
 		start=start+length-1;
 		length=5;
 		color=Qt::red;
-	}
+	}*/
 	c.setPosition(start,QTextCursor::MoveAnchor);
 	c.setPosition(start+length,QTextCursor::KeepAnchor);
 #endif
@@ -131,7 +131,9 @@ void MainWindow::on_pushButton_clicked(){
 	QString error_str,output_str,hadith_str;
 	bool v1,v2,v3,v4,v5,v6;
 	hadithParameters.narr_min=m_ui->NARRATOR->toPlainText().toInt(&v1);
+	geneologyParameters.N_min=hadithParameters.narr_min;
 	hadithParameters.nmc_max=m_ui->NMC->toPlainText().toInt(&v2);
+	geneologyParameters.theta_0=hadithParameters.nmc_max;
 	hadithParameters.nrc_max=m_ui->NRC->toPlainText().toInt(&v3);
 	hadithParameters.equality_delta=m_ui->EQ_delta->toPlainText().toDouble(&v4);
 	hadithParameters.equality_radius=m_ui->EQ_radius->toPlainText().toInt(&v5);

### 2011-09-21 ATSarf/src/gui/test.h

@@ -7,7 +7,8 @@
 
 //#define AUGMENT_DICTIONARY
 //#define JUST_BUCKWALTER
-#define INSERT_ONLY_TIME
+//#define INSERT_ONLY_TIME
+#define INSERT_ONLY_NAMES
 #define REPETITIONS 1
 
 int morphology(QString input_str,ATMProgressIFC * prg);

### 2011-10-12 tex/hadith.ijcnlp.tex
support for min, max boundaries statistics, solved issue in max graph statistics

@@
 
-create a comprehensive genealogical graph from all biblical accounts greatly enhances its study,
+Creating a comprehensive genealogical graph from all biblical accounts greatly enhances its study,
 since readers would be able to better understand a character's genealogical context and life span. Such comprehensive graphs can be tagged
@@
 
+%TODO:
+%\begin{itemize}
+%\item segmentation, boundary (recall and precision)
+%\item local graph min boundary (metrics)
+%\item local graph max boundary (metrics)
+%\item global graph(metrics)
+%
+%\end{itemize}
+
 \subsection{Organizing Quoted Email Replies/Forwards}

### 2011-10-12 trunk/ATSarf/src/case/bibleGeneology.h
support for min, max boundaries statistics, solved issue in max graph statistics

@@ -413,8 +413,9 @@ private:
 		for (int i=0;i<node->children.size();i++) {
 			if (node->children[i]!=NULL)
 				visit(node,node->children[i]->name,false);
+		}
+		for (int i=0;i<node->children.size();i++) {
 			visit(node->children[i]);
-
 		}
 
 	}

### 2011-10-18 trunk/tex/cicling.tex
added necessary code for comparing accuracy and recall of learning names

--- 2011-10-12 tex/hadith.ijcnlp.tex
@@
 % Contact: nasmith@cs.cmu.edu
-\documentclass[11pt]{article}
-\usepackage{ijcnlp2011}
+\documentclass{llncs}
 \usepackage{times}
@@
 \usepackage{rotating}
-\usepackage{amsthm}
+%\usepackage{amsthm}
 \usepackage{booktabs}
@@
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Automatic Hadith Graphs using Arabic Morphological Analysis}
+\title{Knowledge-based Detection of Arabic Entity Graphs using Morphology and Graph Transformations}
 
@@
 \begin{abstract}
-\end{list}
 A hadith in Islamic literature is a narration from the 
@@
 \section{Introduction}
+ Arabic morphological rich
+ entity detection in Arabic
+ entity graphs
+ locality of entities enables detection of entities and aggregating local clusters of entities into general graphs
+ applications
+ 	bible genealogy 
+ 	hadith
+ 	bibliography
+	travel itineraries
+	others...
+ learning and enriching gazetteers
+
+\section{The Method}
+
+Our method takes a set of documents and builds 
+a graph where entities are nodes and edges are relations between the nodes. 
+Our target graph has a predefined set of relations expressed as edge labels such as ``child'' and ``spouse''.
+We use morphological analysis with finite-state machines to detect local entities and
+relations between them.
+We then use graph merging and comparison techniques to build the final general graphs.
+
+In the sequel, we automatically learn entities and append them to our lexicon as new tags. 
+We automatically learn lexicon items and infer their semantical meanings as relations in the 
+graph merging phase. We express the inferred semantic meaning as a sequence of graph transformations
+that removes the edge with the lexicon item and replaces it with edges with equivalent predefined labels.
+We also interactively learn the semantics of other lexicon items that we detect as relations but could not infer their meaning 
+in terms of the predefined set of relations.
+The entities and relations are detected
+
+Our method has two folds. At the first level, we detect entities and edges that
+Given a lexicon that can distinguish entities and connections between them, and several documents that contain these entities,
+we hypothesize that entities that occur next to each other are relevant entities even if their connections do not belong to the
+connection lexicon. We also hypothesize that entities occur after connection are relevant entities.
+
+-importance of cluster (Min number of entities in context)
+-importance of tolerance 
+-entities using lexicons
+-clusters
+-
+
+\section{Application to Hadith}
 
@@
 300 thousand lines of text. 
-Al-Azami\shortcite{Al-Azami-91} cites more than eleven books
+Al-Azami\cite{Al-Azami-91} cites more than eleven books
 of digitized tradition books each of several volumes, and a dozen
@@
 Current morphological analyzers such as 
-Buckwalter~\shortcite{Buckwalter:02},
-Beesley~\shortcite{Beesley:01},
+Buckwalter~\cite{Buckwalter:02},
+Beesley~\cite{Beesley:01},
 SAMA~\cite{Kulick:10},
@@
 Sarf builds on and extends the lexicon of 
-Buckwalter~\shortcite{Buckwalter:02} 
+Buckwalter~\cite{Buckwalter:02} 
 with proper and location names extracted from different online 
@@
 %
-%Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
+%Sarf builds upon the lexicon of Buckwalter\cite{Buckwalter:02}.
 %It differs from Buckwalter in that it defines a shorter list of affixes
@@
 %
-%Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
+%Similar to Sarf, Beesley~\cite{Beesley:01} uses
 %finite state Automata (FSA). 
@@
 %in parentheses as~\cite{Gusfield:97} or, if the author's name appears in
-%the text itself, as Gusfield~\shortcite{Gusfield:97}. 
+%the text itself, as Gusfield~\cite{Gusfield:97}. 
 %Append lowercase letters to the year in cases of ambiguities.  
@@
 %\bibliographystyle{naaclhlt2010}
-\bibliographystyle{acl}
+\bibliographystyle{splncs03}
+
 {\small \bibliography{fzAr}}
+
 \nocite{completeBibleGenealogy:Online}
+
 \end{document}

### 2011-10-18 trunk/ATSarf/src/case/bibleGeneology.h

@@ -5,6 +5,7 @@
 #include <QDebug>
 #include <QFile>
 #include "common.h"
+#include "textParsing.h"
 #include "text_handling.h"
 #include "ATMProgressIFC.h"
 
@@ -55,7 +56,9 @@ inline bool equalNames(const QString & n1,const QString & n2) {
 		return true;
 	if (equal_withoutLastDiacritics(n2,israel) && equal_withoutLastDiacritics(n1,ya3coub))
 		return true;
-	return equal_withoutLastDiacritics(n1,n2);
+	QString s1=withoutAL(n1),s2=withoutAL(n2);
+	assert(!equal_withoutLastDiacritics(n1,n2) || equal_withoutLastDiacritics(s1,s2));
+	return equal_withoutLastDiacritics(s1,s2);
 }
 
 
@@ -74,6 +77,8 @@ private:
 	long start,end;
 	QString * text;
 	bool male:1;
+public:
+	QString edgeText;
 private:
 	friend QDataStream &operator>>(QDataStream &in, GeneNode &t);
 	Name() {}
@@ -102,6 +107,7 @@ public:
 	QString getString() const {
 		return text->mid(start,end-start+1);
 	}
+	QString * getTextPointer() { return text;}
 	long getStart() const{return start;}
 	long getLength() const{return end-start+1;}
 	long getEnd() const { return end;}
@@ -110,7 +116,13 @@ public:
 	}
 };
 inline int qHash(const Name & n) {
-	return qHash(removeDiacritics(n.getString()));
+	if (equal_withoutLastDiacritics(n.getString(),abram))
+		return qHash(ibrahim);
+	if (equal_withoutLastDiacritics(n.getString(),saray))
+		return qHash(sarah);
+	if (equal_withoutLastDiacritics(n.getString(),israel))
+		return qHash(ya3coub);
+	return qHash(withoutAL(removeDiacritics(n.getString())));
 }
 
 class GeneTree;
@@ -186,7 +198,6 @@ public:
 	QList<GeneNode *> children;
 	int height;
 	bool ignoreInSearch;
-	QString edgeText;
 
 	GeneNode(Name n, GeneNode * parent ):name(n) {
 		assert(parent!=this);
@@ -303,6 +314,7 @@ public:
 };
 class GeneTree {
 public:
+	class MergeVisitor;
 	class GraphStatistics {
 	public:
 		double	foundRecall, foundPrecision,
[…]

### 2011-10-18 trunk/ATSarf/src/case/bibleManualTagger.cpp

@@ -1,6 +1,148 @@
 #include "bibleManualTagger.h"
 #include <QtAlgorithms>
 
+
+BibleTaggerDialog::BibleTaggerDialog(QString filename):QMainWindow() {
+	this->filename=filename;
+	text=new QTextBrowser(this);
+	tagGenealogy=new QPushButton("Tag &Genealogy",this);
+	unTagGenealogy=new QPushButton("&Un-Tag Genealogy",this);
+	tagName=new QPushButton("Tag &Name",this);
+	unTagName=new QPushButton("Un-Ta&g Name",this);
+	save=new QPushButton("&Save",this);
+	modifyGraph=new QPushButton("&Modify Graph",this);
+	scrollArea=new QScrollArea(this);
+	forceWordNames=new QCheckBox("Full Word Names");
+	resetGlobalGraph=new QPushButton("Reset Global Graph");
+	resetGlobalGraph->setEnabled(false);
+	isGlobalGraph=new QPushButton("Global Graph");
+	isGlobalGraph->setCheckable(true);
+	graphArea=new QScrollArea(this);
+	treeText=new QTextBrowser(this);
+	treeText->setReadOnly(false);
+	grid=new QGridLayout(scrollArea);
+	grid->addWidget(tagGenealogy,0,0);
+	grid->addWidget(unTagGenealogy,0,1);
+	grid->addWidget(forceWordNames,0,2);
+	grid->addWidget(tagName,0,3);
+	grid->addWidget(unTagName,0,4);
+	grid->addWidget(save,0,5);
+	grid->addWidget(text,1,0,4,2);
+	grid->addWidget(treeText,1,2,1,1);
+	grid->addWidget(modifyGraph,2,2);
+	grid->addWidget(isGlobalGraph,3,2);
+	grid->addWidget(resetGlobalGraph,4,2);
+	grid->addWidget(graphArea,1,3,4,3);
+	grid->setColumnStretch(0,2);
+	grid->setColumnStretch(1,2);
+	grid->setColumnStretch(2,4);
+	grid->setColumnStretch(3,2);
+	grid->setColumnStretch(4,2);
+	grid->setColumnStretch(5,2);
+	graph=new QLabel(graphArea);
+	graphArea->setWidgetResizable(true);
+	graphArea->setAlignment(Qt::AlignCenter);
+	graphArea->setWidget(graph);
+	forceWordNames->setChecked(true);
+	setCentralWidget(scrollArea);
[…]

### 2011-10-18 trunk/ATSarf/src/case/genealogyItem.cpp

@@ -1,0 +1,141 @@
+#include "genealogyItem.h"
+
+QVariant GeneItemModel::headerData(int section, Qt::Orientation orientation, int role ) const{
+	if (orientation == Qt::Horizontal && role == Qt::DisplayRole)
+	{
+		if (section == 0)
+			return QVariant("Name");
+		if (section == 1)
+			return QVariant("Inferred");
+		if (section == 2)
+			return QVariant("Edge Label");
+		if (section == 3)
+			return QVariant("Children Count");
+	}
+	return QVariant();
+}
+
+int GeneItemModel::columnCount(const QModelIndex &parent ) const{
+	return 4;
+}
+
+int
+GeneItemModel::rowCount(const QModelIndex &parent ) const {
+	if (!parent.isValid())
+		return 1;
+	GeneNode * node = (GeneNode *)parent.internalId();
+	if (node == NULL) {
+		// this means that it is a pre/post or behavior that does not exist!
+		return 0;
+	}
+	return node->children.size()+node->spouses.size();
+}
+
+QModelIndex GeneItemModel::parent(const QModelIndex &index) const
+{
+	if (!index.isValid())
+		return QModelIndex();
+	GeneNode * node = (GeneNode*)index.internalId();
+	if (node == NULL){
+		return QModelIndex();
+	}
+	GeneNode * parent = node->parent;
+	int i=0;
+	if (parent != NULL) {
+		GeneNode * ancestor = parent->parent;
+		if (ancestor != NULL) {
+			int j = 0;
+			//for (;j < ancestor->spouses.size(); j++) // can never happen
+			for (;j < ancestor->children.size(); j++) {
+				if (ancestor->children[j]==parent) {
+					i=j+ancestor->spouses.size();
+					break;
+				}
+			}
+			assert (j<ancestor->children.size());
+		}
+	}
+	void * p=(void *)parent;
+	return createIndex(i, 0, p);
+}
+
+Qt::ItemFlags GeneItemModel::flags(const QModelIndex &index) const
+{
+	if (!index.isValid())
+		return 0;
+
+	return Qt::ItemIsEnabled | Qt::ItemIsSelectable;
+}
+
+
+QVariant GeneItemModel::data(const QModelIndex &index, int role) const
+{
+	if (!index.isValid())
+		return QVariant();
+
+	if (role != Qt::DisplayRole)
+		return QVariant();
+	if (index.column() > 3)
+		return QVariant();
+
[…]

### 2011-10-18 trunk/ATSarf/src/case/hadith.h

@@ -57,12 +57,14 @@
 #define SEGMENT_AFTER_PROCESSING_ALL_BIOGRAPHY
 
 //#define DISPLAY_BIOGRAPHY_GRAPH
-//#define TAG_HADITH
+#define TAG_HADITH
 
 #define SAVE_MERGED_GRAPH_ONLY
 #define DESERIALIZE_POR
 //#define DEBUG_BIOGRAPHY_CLUSTERING
 #define BIBLE
+
+#define NONCONTEXT_LEARNING
 
 
 int hadithHelper(QString input_str,ATMProgressIFC *prgs);

### 2011-10-18 trunk/ATSarf/src/case/hadithCommon.h

@@ -99,7 +99,8 @@
 		bool _3an:1;
 		bool learnedName:1;
 		bool nrcIsPunctuation:1; //if state is NRC and that is caused soley by punctuation
-		int unused:12;
+		bool nrcPreviousType:1;
+		int unused:11;
 		PunctuationInfo previousPunctuationInfo,currentPunctuationInfo;
 		void resetCurrentWordInfo()	{familyNMC=false;ibn=false;_2ab=false;_2om=false;_3abid=false;possessivePlace=false;number=false;isWaw=false;_3an=false;learnedName=false;currentPunctuationInfo.reset();}
 		bool familyConnectorOr3abid() { return familyNMC || _3abid;}
@@ -111,7 +112,7 @@
 	extern QString alrasoul,abyi;
 #endif
 	extern QString hadath,abid,alrasoul,abyi,_3an,_2ama,_3ama;
-	extern int bit_POSSESSIVE, bit_PLACE,bit_CITY,bit_COUNTRY,bit_NOUN_PROP;
+	extern int bit_POSSESSIVE, bit_PLACE,bit_CITY,bit_COUNTRY,bit_NOUN_PROP, bit_ENARRATOR_NAMES;
 	extern QList<int> bits_NAME;
 
 #ifdef PREPROCESS_DESCRIPTIONS
@@ -164,6 +165,7 @@
 	#endif
 	public:
 		long finish_pos;
+		int numSolutions;
 		bool name:1, nrc:1, nmc,possessive:1, familyNMC:1,ibn:1,_2ab:1,_2om:1,_3abid:1,stopword:1;
 	#ifdef GET_WAW
 		bool has_waw:1,is3an:1;
@@ -202,6 +204,7 @@
 			_3abid=false;
 			stopword=false;
 			finish_pos=start;
+			numSolutions=0;
 		#ifdef REFINEMENTS
 			tryToLearnNames=false;
 			learnedName=false;
@@ -300,6 +303,7 @@
 
 		bool analyze()
 		{
+			numSolutions++;
 		#ifdef STATS
 			QString temp_stem=removeDiacritics(info.text->mid(Stem->starting_pos, Suffix->info.start-Stem->starting_pos));//removeDiacritics(stem_info->raw_data);
 		#endif
@@ -428,6 +432,8 @@
 				if (tryToLearnNames)
 					bits_NAME.removeLast();
 			#endif
[…]

### 2011-10-18 trunk/ATSarf/src/case/timeRecognizer.cpp

@@ -84,7 +84,7 @@ public:
 };
 
 QDataStream &operator<<(QDataStream &out, const TimeEntity &entity) {
-	out<<(int)entity.getStart()<<(int)entity.getEnd()+1;
+	out<<(int)entity.getStart()<<(int)entity.getEnd();
 	return out;
 }
 
@@ -552,7 +552,7 @@ int calculateStatistics(QString filename){
 	int i=0,j=0;
 
 	while (i<tags.size() && j<timeVector->size()) {
-		int start1=tags[i].first,end1=tags[i].second-1,
+		int start1=tags[i].first,end1=tags[i].second,
 			start2=(*timeVector)[j].getStart(),end2=(*timeVector)[j].getEnd();
 		if (overLaps(start1,end1,start2,end2)) {
 			TimeTaggerDialog::Selection overlap=overLappingPart(start1,end1,start2,end2);
@@ -590,7 +590,7 @@ int calculateStatistics(QString filename){
 	}
 #ifdef DETAILED_DISPLAY
 	while (i<tags.size()) {
-		int start1=tags[i].first,end1=tags[i].second-1;
+		int start1=tags[i].first,end1=tags[i].second;
 		displayed_error	<<time_text->mid(start1,end1-start1+1)<<"\t"
 						<<"-----\n";
 		i++;

### 2011-10-18 trunk/ATSarf/src/gui/test.cpp

@@ -86,17 +86,37 @@ int biography(QString inputString,ATMProgressIFC *prg) {
 		return -1;
 	return 0;
 }
-int annotation(QString inputString,ATMProgressIFC *) {
-#ifdef BIBLE
-	if (bibleTagger(inputString))
-#else
+int simple_annotation(QString inputString,ATMProgressIFC *) {
 	if (timeTagger(inputString))
-#endif
+		return -1;
+	return 0;
+}
+
+int bible_annotation(QString inputString,ATMProgressIFC *) {
+	if (bibleTagger(inputString))
 		return -1;
 	return 0;
 }
 
 int test(QString inputString,ATMProgressIFC * prg) {
+	long i=0;
+	skipAL(&inputString,i);
+	out<<i<<"\n";
+	out<<startsWithAL(inputString)<<"\n";
+	out<<withoutAL(inputString)<<"\n";
+	out<<removeAL(inputString)<<"\n";
+	i=0;
+	skipAL(&inputString,i);
+	out<<i<<"\n";
+	skipOneLetter(&inputString,i);
+	out<<i<<"\n";
+	skipOneLetter(&inputString,i);
+	out<<i<<"\n";
+	skipOneLetter(&inputString,i);
+	out<<i<<"\n";
+	skipOneLetter(&inputString,i);
+	out<<i<<"\n";
+	return 0;
 #ifdef AUGMENT_DICTIONARY
 	if (augment()<0)
 		return -1;

### 2011-10-18 trunk/ATSarf/src/util/text_handling.h

@@ -145,20 +145,51 @@ inline bool startsWith(const QString &text,const QString &substring) // is diacr
 	return startsWith(text.rightRef(-1),substring,dummy_int);
 }
 
-inline bool startsWithAL(const QString & word)//does not take in account cases were Diacritics may be present on the alef and lam of "al"
-{
-	if (word.length()<=2)
-		return false;
-	if (word[0]==alef && word[1]==lam)//TODO: whenever it is changed to include diactrics in-between it must notice that only some combinations of diactrics imply AL (definite article) else is part of the word
-		return true;
+
+inline void skipOneLetter(const QString * text,long & currentPos) {
+	if (currentPos<text->size()) {
+		currentPos++;
+		while (currentPos<text->size() && isDiacritic(text->at(currentPos)))
+			currentPos++;
+	}
+}
+
+inline bool skipAL(const QString * text, long & currPos) { //just moves currPos
+	if (currPos<text->size()) {
+		QChar l=text->at(currPos);
+		if (l==alef) {
+			long newPos=currPos;
+			skipOneLetter(text,newPos);
+			if (newPos<text->size()) {
+				l=text->at(newPos);
+				if (l==lam) {
+					skipOneLetter(text,newPos);
+					currPos=newPos;
+					return true;
+				}
+			}
+		}
+	}
 	return false;
 }
-inline bool removeAL( QString &word)//does not take in account cases were Diacritics may be present on the alef and lam of "al"
+
+inline bool startsWithAL(const QString & word)//does not take in account cases were Diacritics may be present on the alef and lam of "al"
 {
-	if (!startsWithAL(word))
+	long i=0;
+	return (skipAL(&word,i)); //will not affect word
+}
+inline bool removeAL( QString &word)
+{
+	long i=0;
+	if (!skipAL(&word,i))
 		return false;
-	word=word.right(word.length()-2);
+	word=word.mid(i,word.length()); //word.length() is more than needed but no problem will be ignored supposedly
 	return true;
+}
+inline QString withoutAL( QString word)
+{
+	removeAL(word);
+	return word;
 }
 
 #endif // _TEXT_HANDLING_H

### 2011-10-18 ATSarf/src/case/bibleManualTagger.h
added necessary code for comparing accuracy and recall of learning names

@@ -22,20 +22,13 @@
 #include "ATMProgressIFC.h"
 #include "letters.h"
 #include "bibleGeneology.h"
+#include "abstractAnnotator.h"
 
 using namespace std;
 
-class BibleTaggerDialog:public QMainWindow, public ATMProgressIFC{
+class BibleTaggerDialog:public QMainWindow, public ATMProgressIFC,public AbstractAnnotator{
 	Q_OBJECT
 public:
-	enum SelectionMode {SELECTION_OUTSIDE,SELECTION_INSIDE,SELECTION_OUTSIDEOVERLAP};
-	inline static bool isConsistentWithSelectionCondidtion(int start, int end, int tagStart,int tagEnd,SelectionMode selectionMode=SELECTION_OUTSIDE) {
-		bool con1= selectionMode==SELECTION_OUTSIDE && start<=tagStart && end>=tagEnd,
-			 con2= selectionMode==SELECTION_INSIDE && start>=tagStart && end<=tagEnd,
-			 con3= selectionMode==SELECTION_OUTSIDEOVERLAP && ((start<=tagStart && end>=tagStart) ||
-														(start<=tagEnd && end>=tagEnd));
-		return con1 || con2 || con3;
-	}
 	class Selection {
 		friend QDataStream &operator<<(QDataStream &out, const BibleTaggerDialog::Selection &t);
 		friend QDataStream &operator>>(QDataStream &in, BibleTaggerDialog::Selection &t);
@@ -94,8 +87,8 @@ public:
 		void addName( const Name & name) {
 			addName(name.getStart(),name.getEnd());
 		}
-		int getMainStart() { return main.first;}
-		int getMainEnd() {return main.second;}
+		int getMainStart() const { return main.first;}
+		int getMainEnd() const {return main.second;}
 		bool operator <(const Selection & second) const {
 			return main<second.main;
 		}
@@ -328,77 +321,7 @@ public:
 	};
 	typedef QList<Selection> SelectionList;
 public:
-	BibleTaggerDialog(QString filename):QMainWindow() {
-		this->filename=filename;
-		text=new QTextBrowser(this);
-		tagGenealogy=new QPushButton("Tag &Genealogy",this);
-		unTagGenealogy=new QPushButton("&Un-Tag Genealogy",this);
-		tagName=new QPushButton("Tag &Name",this);
[…]

### 2011-10-18 ATSarf/src/case/graph_nodes.cpp

@@ -111,7 +111,7 @@ NarratorNodeIfc & GroupNode::getCorrespondingNarratorNode() {
 	if (graphNode!=NULL)
 		return *graphNode;
 	else { //in case we have a group that has just one node
-		assert (size()==1);
+		//assert (size()==1);
 		return operator [](0);
 	}
 }

### 2011-10-18 ATSarf/src/case/narrator_abstraction.h

@@ -8,6 +8,7 @@
 #include <QFile>
 #include <QDebug>
 #include "logger.h"
+#include "text_handling.h"
 
 #define SHOW_AS_TEXT
 
@@ -48,11 +49,11 @@ public:
 
 class NarratorPrim: public ChainNarratorPrim {
 public:
-    long long m_start;
-    long long m_end;
+	int m_start;
+	int m_end;
 
 	NarratorPrim(QString * hadith_text);
-	NarratorPrim(QString * hadith_text,long long m_start);
+	NarratorPrim(QString * hadith_text,int m_start);
 
  //   virtual CNPIterator first () const=0;
  //   virtual CNPIterator end () const=0;
@@ -85,7 +86,7 @@ class NamePrim :public NarratorPrim {
 public:
 	bool learnedName;
 	NamePrim(QString * hadith_text);
-	NamePrim(QString * hadith_text,long long m_start);
+	NamePrim(QString * hadith_text,int m_start);
 
    // virtual CNPIterator first () const{
    // return QList<ChainNarratorPrim*> ().begin();
@@ -98,7 +99,14 @@ public:
 	virtual void serialize(QDataStream &chainOut) const;
 	virtual void deserialize(QDataStream &chainOut);
 	virtual void serialize(QTextStream &chainOut) const;
+	bool operator ==(const NamePrim & name) const {
+		return equal_withoutLastDiacritics(getString(),name.getString());
+	}
 };
+inline uint qHash(const NamePrim & name) {
+	return qHash(removeDiacritics(name.getString()));
+}
+
 class NameConnectorPrim :public NarratorPrim {
 private:
 	enum Type {POS,IBN,AB,OM,FAMILY_OTHER, OTHER};
@@ -117,7 +125,7 @@ public:
 	void setOther(){type=OTHER;}
 	bool isOther(){return type==OTHER;}
 	NameConnectorPrim(QString * hadith_text);
-	NameConnectorPrim(QString * hadith_text,long long m_start);
+	NameConnectorPrim(QString * hadith_text,int m_start);
 
 //    virtual CNPIterator first () const{
 //    return QList<ChainNarratorPrim*> ().begin();
@@ -133,10 +141,10 @@ public:
 };
 class NarratorConnectorPrim :public ChainPrim {
 public:
-    long long m_start;
-    long long m_end;
+	int m_start;
+	int m_end;
 	NarratorConnectorPrim(QString * hadith_text);
[…]

### 2011-10-18 ATSarf/src/case/timeManualTagger.h

@@ -1,55 +1,22 @@
 #ifndef TIMEMANUALTAGGER_H
 #define TIMEMANUALTAGGER_H
 
-#include "ATMProgressIFC.h"
 #include <QMainWindow>
-#include <QFile>
-#include <QTextStream>
-#include <QStringList>
 #include <QPushButton>
-#include <QGridLayout>
 #include <QTextBrowser>
 #include <QScrollArea>
-#include "logger.h"
-#include <iostream>
-#include <QList>
-#include <QPair>
+#include <QGridLayout>
+
+
+#include "ATMProgressIFC.h"
+#include "abstractAnnotator.h"
 
 using namespace std;
 
-class TimeTaggerDialog:public QMainWindow{
+class TimeTaggerDialog:public QMainWindow,public AbstractAnnotator{
 	Q_OBJECT
 public:
-	TimeTaggerDialog(QString filename):QMainWindow() {
-
-		this->filename=filename;
-		text=new QTextBrowser(this);
-		tag=new QPushButton("&Tag",this);
-		unTag=new QPushButton("&Un-Tag",this);
-		save=new QPushButton("&Save",this);
-		scrollArea=new QScrollArea(this);
-		grid=new QGridLayout(scrollArea);
-		grid->addWidget(tag,0,0);
-		grid->addWidget(unTag,0,1);
-		grid->addWidget(save,0,2);
-		grid->addWidget(text,1,0,1,3);
-		setCentralWidget(scrollArea);
-	#ifdef ERRORS_TIME
-		errors=new QTextBrowser(this);
-		errors->resize(errors->width(),50);
-		errors_text=new QString();
-		grid->addWidget(errors,2,0,1,3);
-		displayed_error.setString(errors_text);
-		errors->setText(*errors_text);
-	#endif
-		connect(tag,SIGNAL(clicked()),this,SLOT(tag_clicked()));
-		connect(unTag,SIGNAL(clicked()),this,SLOT(unTag_clicked()));
-		connect(save,SIGNAL(clicked()),this,SLOT(save_clicked()));
-		open_action();
-		setWindowTitle(filename);
-		this->resize(700,700);
-	}
-
+	TimeTaggerDialog(QString filename);
 public slots:
 	void tag_clicked() {
 		tag_action();
@@ -66,7 +33,10 @@ private:
 	void unTag_action();
 	void save_action();
 	void open_action();
-
[…]

### 2011-10-18 ATSarf/src/util/textParsing.h

@@ -2,6 +2,7 @@
 #define TEXTPARSING_H
 #include "hadith.h"
 #include "letters.h"
+#include "diacritics.h"
 
 class PunctuationInfo {
 public:
@@ -110,5 +111,7 @@ inline bool isNumber(QString * text,long & currentPos,long & finish) {
 	finish=i; //means (still number and text finished) or did not enter the loop at all
 	return ret_val;
 }
+
+
 
 #endif // TEXTPARSING_H

### 2011-11-04 tex/cicling.tex
updated some issues to make SUBMISSION version suitable for use with no reliance on database and corrected some minor issues

--- 2011-10-18 trunk/tex/cicling.tex
@@
 \caption{Results comparing between NaGEMA and iTree.}
-\resizebox{2.1\columnwidth}{!}{
-\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+\resizebox{1\columnwidth}{!}{
+\begin{tabular}{p{2.1cm}lp{.1cm}cp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
 %\multicolumn{13}{c}{NaGEMA with all refinements} \\ \cline{3-13}
-& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
-& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{NaGEMA }&musnad & & 0.999 & 0.999 & 0.999 & & 0.993 & 0.990 & 0.991 & & 0.999 & 0.998 & 0.999 \\
-&kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.988 & 0.990 \\
-&istibsar & & 0.999 & 0.973 & 0.986 & & 0.980 & 0.950 & 0.965 & & 0.996 & 0.998 & 0.997 \\ \cline{2-14}
-& & & 0.999 & 0.964 & 0.982 & & 0.970 & 0.970 & 0.970 & & 0.996 & 0.995 & 0.995 \\ \hline \hline
+& & & hadith & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & count & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-16}
+\multirow{3}{*}{NaGEMA }
+&musnad & & 212 & & 0.986 & 0.976 & 0.981 & & 0.969 & 0.967 & 0.968 & & 0.999 & 0.996 & 0.998 \\
+&kafi & & 199 & & 0.970 & 0.909 & 0.938 & & 0.934 & 0.954 & 0.944 & & 0.999 & 0.998 & 0.998 \\
+&istibsar  & & 189 & & 0.968 & 0.973 & 0.971 & & 0.946 & 0.945 & 0.945 & & 0.999 & 0.998 & 0.999 \\ \cline{2-16}
+& & & 600 & & 0.975 & 0.953 & 0.964 & & 0.950 & 0.956 & 0.953 & & 0.999 & 0.997 & 0.998 \\ \hline \hline
 %& & &  &  &  & &  &  &  & &  &  & \\
 %&\multicolumn{13}{c}{iTree} \\ \cline{3-13}
-& & & \multicolumn{3}{c}{accept as hadith} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
-& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{iTree} &musnad & & 0.999 & NA & 0.999 & & 0.997 & 0.977 & 0.987 & & 0.998 & 0.999 & 0.999  \\
-&kafi & & 0.043 & NA & 0.083 & & 0.999 & 0.313 & 0.476 & & 0.999 & 0.999 & 0.999 \\
-&istibsar & & 0.028 & NA & 0.054 & & 0.875 & 0.583 & 0.700 & & 0.999 & 0.999 & 0.999 \\ \cline{2-14}
-& & & 0.357 & NA & 0.379 & & 0.957 & 0.624 & 0.721 & & 0.999 & 0.999 & 0.999 \\ 
+& & & hadith & & \multicolumn{3}{c}{accept as hadith} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+& & & count & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-16}
+\multirow{3}{*}{iTree} 
+&musnad && 62 && 0.999 & NA & NA & & 0.997 & 0.977 & 0.987 & & 0.998 & 0.999 & 0.999  \\
+&kafi && 23 && 0.043 & NA & NA & & 0.999 & 0.313 & 0.476 & & 0.999 & 0.999 & 0.999 \\
+&istibsar && 36 && 0.028 & NA & NA & & 0.875 & 0.583 & 0.700 & & 0.999 & 0.999 & 0.999 \\ \cline{2-16}
+& & & 124 && 0.357 & NA & NA & & 0.957 & 0.624 & 0.721 & & 0.999 & 0.999 & 0.999 \\ 
  \end{tabular}                 
@@
 \centering                                            
-\caption{Results without punctuation and without learning. }
-\resizebox{2.1\columnwidth}{!}{
-\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
-& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
-& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{no punctuation} 
-& musnad  & & 0.999 & 0.999 & 0.999 & & 0.982 & 0.955 & 0.968 & & 0.999 & 0.998 & 0.999 \\ 
-& kafi & & 0.999 & 0.793 & 0.885 & & 0.944 & 0.967 & 0.955 & & 0.996 & 0.938 & 0.966 \\
-& istibsar & & 0.999 & 0.999 & 0.999 & & 0.980 & 0.931 & 0.955 & & 0.994 & 0.985 & 0.990 \\  \cline{2-14}
-& &  & 0.999 & 0.931 & 0.962 & & 0.969 & 0.951 & 0.960 & & 0.997 & 0.974 & 0.985 \\ \hline \hline
-%& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
-%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{no learning}
-& musnad & & 0.999 & 0.999 & 0.999 & & 0.894 & 0.985 & 0.937 & & 0.995 & 0.999 & 0.998\\
-& kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.991 & 0.992\\
-& istibsar & & 0.999 & 0.973 & 0.986 & & 0.977 & 0.950 & 0.963 & & 0.996 & 0.998 & 0.997\\ \cline{2-14}
- & &  & 0.999 & 0.964 & 0.982 & & 0.936 & 0.968 & 0.951 & & 0.995 & 0.996 & 0.995 \\
+\caption{Inter-annotator Agreement}
+\resizebox{1\columnwidth}{!}{
+\begin{tabular}{p{2cm}p{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+\multirow{2}{*}{Test Case} & &  \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad/genealogy boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+ & &  recall & precision & average & & recall & precision & average & & recall & precision & average \\ \cline{2-13}
+Hadith &  & 0.976 & 1.000 & 0.988 & & 0.991 & 0.971 & 0.981 & & 1.000 & 0.997 & 0.999 \\ 
+Genealogy &  &  1.000 & 0.390 & 0.695 & & 0.969 & 0.954 & 0.812 & & NA & NA & NA\\
  \end{tabular}                 
@@
 \normalsize                   
-\label{t:nopunctlearnresults} 
+\label{t:interannotator} 
+\end{table*}
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{POR and Name Learning Results}
+\resizebox{1\columnwidth}{!}{
+\begin{tabular}{cp{1cm}c}
+	\begin{tabular}{p{2.5cm}p{.1cm}ccc}
+								& & recall & precision & F-score \\ \cline{2-5}
+	All Narrators & &  0.632 & 0.771 & 0.694\\
+	Detected Narrators & & 0.673 & 0.824 & 0.741\\
+	\multicolumn{5}{c}{(a) Narrator POR Accuracy} \\
+	\end{tabular}  
+&  &
+	\begin{tabular}{p{2cm}p{.1cm}ccc}
+								& & recall & precision & F-score \\ \cline{2-5}
+	Sanad Names & &  0.649 & 0.940 & 0.768\\
+	All Names & & 0.794 & 0.733 & 0.762\\
+	\multicolumn{5}{c}{(b) Name Learning Accuracy} \\
+	\end{tabular}  \\
+ \end{tabular}               
+}                             
+\normalsize                   
+\label{t:PORnLearning} 
+\end{table*}
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Results comparing Narrator Detection Accuracy in Hadith and Biography}
+\resizebox{1\columnwidth}{!}{
+
+\begin{tabular}{cp{0.5cm}c}
+
+	\begin{tabular}{p{2.5cm}cp{.1cm}cccp{.1cm}ccc}
+	& & & \multicolumn{3}{c}{detection} & &  \multicolumn{3}{c}{boundary} \\
+	& reported narrators& & recall & precision & F-score & & recall & precision & F-score \\ 
+	\cline{2-2} \cline{4-10}
+	\multirow{2}{*}{Hadith (musnad)}
+	& of sanad & & 0.923 & 0.999 & 0.960 & & 0.996 & 0.999 & 0.998  \\
+	& all & & 0.954 & 0.961 & 0.958 & & 0.996 & 0.987 & 0.992 \\ 
+	\hline \hline
+	\multirow{2}{*}{Biography (khoei)} 
+	& found in POR & & 0.654 & 0.909 & 0.760 & & 0.933 & 0.960 & 0.946 \\
+	& all & & 0.943 & 0.812 & 0.873 & & 0.407 & 0.945 & 0.569 \\
+	 \end{tabular}   
+&  &
+	\begin{tabular}{cc}
+	\multicolumn{2}{c}{narrator statistics} \\
+	concentration & aggregation \\ \hline
+	\multirow{2}{*}{0.217} &\multirow{2}{*}{ 2.64}\\ \\
+	\hline \hline
+	\multirow{2}{*}{0.489} & \multirow{2}{*}{3.54}\\
+	\\
+	\end{tabular}  \\
+\end{tabular}          
+}                             
+\normalsize                   
+\label{t:narratorDifference} 
+\end{table*}
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Biography Detection Results}
+\resizebox{0.25\columnwidth}{!}{
+	\begin{tabular}{p{2.5cm}p{.1cm}c}
+	Exact Recall & &  0.6 \\
+	Partial Recall & & 0.8\\
+	Search Count & & 10 \\
+	\end{tabular}             
+}                             
+\normalsize                   
+\label{t:biography} 
+\end{table*}
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Results for Biblical Genealogy Parsing}
+\resizebox{0.7\columnwidth}{!}{
+\begin{tabular}{p{2.5cm}cccp{.1cm}ccc}
+ & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{names extraction} \\
+ &  recall & precision & F-score & & recall & precision & F-score \\ \cline{2-8}
+basic controller & 0.796 & 0.708 & 0.749 & & 0.883 & 0.906 & 0.894 \\
+refined controller & 0.796 & 0.742 & 0.768 & & 0.887 & 0.910 & 0.898 \\
+ \end{tabular}                 
+}                             
+\normalsize                   
+\label{t:genealogyParsing} 
+\end{table*}
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Results for Biblical Genealogy Kinship Graphs}
+\resizebox{1\columnwidth}{!}{
+\begin{tabular}{p{2.5cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+& & & \multicolumn{3}{c}{available nodes} & &  \multicolumn{3}{c}{neighborhood} & &  \multicolumn{3}{c}{context} \\
+& & &  recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{4-14}
+\multirow{3}{*}{Basic Controller}
+&Local Graphs && 0.583 & 0.898 & 0.707 && 0.778 & 0.760 & 0.769 && 0.552 & 0.588 & 0.570 \\
+&Local Graphs (Aligned) && 0.849 & 0.868 & 0.858 && 0.777 & 0.738 & 0.757 && 0.554 & 0.590 & 0.572 \\
+&Global Graph && 0.784 & 0.800 & 0.792 && 0.652 & 0.652 & 0.652 && 0.536 & 0.573 & 0.554  \\
+\hline \hline
+\multirow{3}{*}{Refined Controller}
+&Local Graphs && 0.586 & 0.900 & 0.710 && 0.780 & 0.725 & 0.751 && 0.557 & 0.577 & 0.567 \\
+&Local Graphs (Aligned) && 0.854 & 0.871 & 0.862 && 0.785 & 0.723 & 0.753 && 0.582 & 0.600 & 0.591 \\
+&Global Graph && 0.820 & 0.810 & 0.815 && 0.701 & 0.708 & 0.704 && 0.580 & 0.630 & 0.604  \\
+ \end{tabular}                 
+}                             
+\normalsize                   
+\label{t:genealogyGraphs} 
+\end{table*}      .
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Statistics for Biblical Genealogy Kinship Graphs}
+\resizebox{0.7\columnwidth}{!}{
+\begin{tabular}{p{1.5cm}lp{.1cm}ccccp{.1cm}c}
+& & & \multicolumn{6}{c}{Node Count} \\ \cline{4-9}
+& & & \multicolumn{4}{c}{Local Graphs} & & \multicolumn{1}{c}{Global Graph}\\
+& Controller & &  minimum & average & maximum & total size & & size \\ \cline{2-9}
+\multirow{2}{*}{Extracted}
+&Basic && 3 & 7.89 & 35 & 513 && 245 \\
+&Refined && 3 & 8.00 & 35 & 496 && 253 \\
+ \hline
+\multirow{1}{*}{Annotation}
+&  && 2 & 8.80 & 77 & 475 && 250 \\
+ \end{tabular}                 
+}                             
+\normalsize                   
+\label{t:genealogyGraphStatistics} 
 \end{table*}

### 2011-11-04 ATSarf/src/cache/compatibility_rules.h

@@ -102,6 +102,11 @@ class compatibility_rules
 			else
 				return (*i);
 		}
+		int getAbstractCategoryID(int bitIndex) {
+			if (bitIndex<0 || bitIndex>absCatIDForBits.size())
+				return -1;
+			return absCatIDForBits[bitIndex];
+		}
 		int getAbstractCategoryID(const QString & name) //-1 is invalid, -2 not abstract
 		{
 		#if 1

### 2011-11-04 ATSarf/src/case/bibleGeneology.h

@@ -14,7 +14,7 @@
 #define GET_WAW
 //#define TRUST_OLD
 #define SINGULAR_DESCENT
-#define GENEOLOGYDEBUG
+//#define GENEOLOGYDEBUG
 //#define DISPLAY_INDIVIDUAL
 #define REDUCE_AFFIX_SEARCH
 //#define SHOW_MERGING_ERRORS
@@ -23,11 +23,13 @@
 class GeneologyParameters {
 public:
 	unsigned int theta_0:14;
-	unsigned int L_min:10;
-	unsigned int N_min:10;
+	unsigned int L_min:9;
+	unsigned int N_min:9;
 	unsigned int C_max:10;
 	unsigned int radius:10;
 	unsigned int step:10;
+	bool detailed_statistics:1;
+	bool refined:1;
 	GeneologyParameters() {
 		theta_0=45;
 		L_min=3;
@@ -35,6 +37,7 @@ public:
 		C_max=2;
 		radius=3;
 		step=10;
+		detailed_statistics=true;
 	}
 };
 

### 2011-11-04 ATSarf/src/case/hadithCommon.cpp

@@ -49,7 +49,7 @@ HadithParameters hadithParameters;
 
 
 #ifdef PREPROCESS_DESCRIPTIONS
-void readFromDatabasePreProcessedDescriptions() {
+void readFromDatabasePreProcessedHadithDescriptions() {
 	Retrieve_Template nrc_s("description","id","name='said' OR name='say' OR name='notify/communicate' OR name LIKE '%/listen' OR name LIKE 'listen/%' OR name LIKE 'listen %' OR name LIKE '% listen' OR name = 'listen' OR name LIKE '%/inform' OR name LIKE 'inform/%' OR name LIKE 'inform %' OR name LIKE '% inform' OR name = 'inform' OR name LIKE '%from/about%' OR name LIKE '%narrate%'");
 	while (nrc_s.retrieve())
 		NRC_descriptions.insert(nrc_s.get(0).toULongLong(),true);
@@ -81,9 +81,9 @@ void readFromDatabasePreProcessedDescriptions() {
 	else
 		error <<"Unexpected Error: Unable to write PreProcessed Descriptions to file\n";
 }
-void readFromFilePreprocessedDescriptions() {
+void readFromFilePreprocessedHadithDescriptions() {
 #ifndef LOAD_FROM_FILE
-	readFromDatabasePreProcessedDescriptions();
+	readFromDatabasePreProcessedHadithDescriptions();
 #else
 	QFile file(preProcessedDescriptionsFileName.toStdString().data());
 	if (file.open(QIODevice::ReadOnly))
@@ -97,7 +97,7 @@ void readFromFilePreprocessedDescriptions() {
 		file.close();
 	}
 	else
-		readFromDatabasePreProcessedDescriptions();
+		readFromDatabasePreProcessedHadithDescriptions();
 #endif
 }
 #endif
@@ -181,7 +181,7 @@ void hadith_initialize() {
 	stat.narrator_per_chain.clear();
 #endif
 #ifdef PREPROCESS_DESCRIPTIONS
-	readFromFilePreprocessedDescriptions();
+	readFromFilePreprocessedHadithDescriptions();
 #endif
 }
 
@@ -1833,27 +1833,31 @@ inline bool result(WordType t, StateInfo &  stateInfo,HadithData *currentChain, 
 			QSet<Name> detectedOnlyInNonContext=commonNonContextNames;
 			detectedOnlyInNonContext.subtract(commonContextNames);
 
[…]

### 2011-11-04 ATSarf/src/case/hadithInterAnnotatorAgreement.h

@@ -19,9 +19,10 @@ private:
 						* nodeDetected;
 	int countCommon;
 	double mergesRecallALL,mergesPrecisionALL,mergesRecallNew,mergesPrecisionNew;
+	bool displayPORstatistics;
 #endif
 public:
-	HadithInterAnnotatorAgreement(QString * text,QString fileName,AbstractGraph * generatedGraph, OutputDataList & generatedTags);
+	HadithInterAnnotatorAgreement(QString * text,QString fileName,AbstractGraph * generatedGraph, OutputDataList & generatedTags, bool displayPORstatistics=false);
 	void displayStatistics();
 private:
 	virtual int equalNames(QString * text, int start1, int end1, int start2, int end2);

### 2011-11-04 ATSarf/src/case/narratorEqualityComparision.cpp

@@ -42,8 +42,8 @@ int narrator_equality_comparision(QString input_str, ATMProgressIFC *) {
 			countAllEqualeNarrator++;
 		if (eq && eNarratorEqual)
 			countCorrectEqualeNarrator++;
-		Narrator * narr1=HadithChainGraph::getNarrator(n1);
-		Narrator * narr2=HadithChainGraph::getNarrator(n2);
+		Narrator * narr1=HadithChainGraph::getNarrator(n1,false);
+		Narrator * narr2=HadithChainGraph::getNarrator(n2,false);
 		double equalScore=equal(*narr1,*narr2);
 		bool equalOur=equalScore>hadithParameters.equality_threshold;
 		if (equalOur)
@@ -52,10 +52,12 @@ int narrator_equality_comparision(QString input_str, ATMProgressIFC *) {
 			countCorrectEqualOur++;
 		delete narr1;
 		delete narr2;
-	#ifndef DETAILED_DISPLAY
-		if (eq ==!(equalOur || eNarratorEqual))
-	#endif
-			displayed_error<<n1<<"\t"<<n2<<"\t"<<eq<<"\t"<<equalOur<<"\t"<<eNarratorEqual<<"\n";
+		if (hadithParameters.detailed_statistics) {
+		#ifndef DETAILED_DISPLAY
+			if (eq ==!(equalOur || eNarratorEqual))
+		#endif
+				displayed_error<<n1<<"\t"<<n2<<"\t"<<eq<<"\t"<<equalOur<<"\t"<<eNarratorEqual<<"\n";
+		}
 	}
 	displayed_error <<	"eNarrator:\n"
 					<<  "\trecall:\t"<<countCorrectEqualeNarrator<<"/"<<countAnnotationEqual<<"=\t"

### 2011-11-04 ATSarf/src/gui/test.cpp

@@ -2,6 +2,7 @@
 #include <QStringList>
 #include <QString>
 #include <QDateTime>
+#include <QMessageBox>
 #include "test.h"
 #include "logger.h"
 #include "hadith.h"
@@ -106,25 +107,12 @@ int hadith_annotation(QString inputString,ATMProgressIFC *) {
 }
 
 int test(QString inputString,ATMProgressIFC * prg) {
-	long i=0;
-	skipAL(&inputString,i);
-	out<<i<<"\n";
-	out<<startsWithAL(inputString)<<"\n";
-	out<<withoutAL(inputString)<<"\n";
-	out<<removeAL(inputString)<<"\n";
-	i=0;
-	skipAL(&inputString,i);
-	out<<i<<"\n";
-	skipOneLetter(&inputString,i);
-	out<<i<<"\n";
-	skipOneLetter(&inputString,i);
-	out<<i<<"\n";
-	skipOneLetter(&inputString,i);
-	out<<i<<"\n";
-	skipOneLetter(&inputString,i);
-	out<<i<<"\n";
-	return 0;
-#ifdef AUGMENT_DICTIONARY
+#ifdef SUBMISSION
+	QMessageBox msg;
+	msg.setWindowTitle("Sarf");
+	msg.setText("Test Options Unavailable for submission version of software.");
+	msg.exec();
+#elif  defined(AUGMENT_DICTIONARY)
 	if (augment()<0)
 		return -1;
 #elif 0

### 2011-11-04 trunk/ATSarf/src/case/AbstractTwoLevelAgreement.cpp
updated some issues to make SUBMISSION version suitable for use with no reliance on database and corrected some minor issues

@@ -2,6 +2,7 @@
 #include "logger.h"
 #include "text_handling.h"
 #include "Math_functions.h"
+#include "hadithCommon.h"
 
 AbstractTwoLevelAgreement::AbstractTwoLevelAgreement(QString * text,QString fileName,AbstractGraph * generatedGraph, OutputDataList & generatedTags)
 	:outputList(generatedTags) {
@@ -144,11 +145,11 @@ void AbstractTwoLevelAgreement::overLapMainFinished(int i,int j,const SelectionL
 		namePrecisionList.append(0);
 	}
 	if (underComputation) { //under computation only
-	#ifdef DETAILED_DISPLAY
-		displayed_error	<</*text->mid(start1,end1-start1+1)*/i<<"\t"
-						<</*text->mid(start2,end2-start2+1)*/j<<"\t"
-						<<countCommon<<"/"<<countCorrect<<"\t"<<allCommonCount<<"/"<<countDetected<<"\n";
-	#endif
+		if (hadithParameters.detailed_statistics) {
+			displayed_error	<</*text->mid(start1,end1-start1+1)*/i<<"\t"
+							<</*text->mid(start2,end2-start2+1)*/j<<"\t"
+							<<countCommon<<"/"<<countCorrect<<"\t"<<allCommonCount<<"/"<<countDetected<<"\n";
+		}
 	}
 }
 
@@ -171,6 +172,7 @@ bool AbstractTwoLevelAgreement::readAnnotations() {
 	annotatedGraph=newGraph(true);
 	if (!readAnnotation(fileName+".tags",tags,annotatedGraph)){
 		error << "Annotation File does not exist\n";
+	#ifndef SUBMISSION
 		QFile file(QString("%1.tags").arg(fileName).toStdString().data());
 		if (file.open(QIODevice::WriteOnly)) {
 			QDataStream out(&file);   // we will serialize the data into the file
@@ -179,6 +181,7 @@ bool AbstractTwoLevelAgreement::readAnnotations() {
 			file.close();
 			error << "Annotation File has been written from current detected expressions, Correct it before use.\n";
 		}
+	#endif
 		return false;
 	}
 	return true;
@@ -257,10 +260,10 @@ int AbstractTwoLevelAgreement::calculateStatisticsHelper() {
 				outputNamesOverLap.clear();
 			}
[…]

### 2011-11-04 trunk/ATSarf/src/case/bibleGeneology.cpp

@@ -99,6 +99,10 @@ inline void display(QString t) {
 	#define display(c)
 #endif
 
+#ifdef PREPROCESS_DESCRIPTIONS
+	QString preProcessedGenealogyDescriptionsFileName= ".GenealogyPreProcessedDescriptions";
+#endif
+
 class DescentConnectors;
 class DescentConnectorGroup {
 private:
@@ -106,8 +110,24 @@ private:
 	QList<long> descriptions;
 	DescentDirection dir;
 	bool plural;
+private:
 	long operator[](int i) { return descriptions[i]; }
 	int size() { return descriptions.size(); }
+	friend QDataStream & operator << (QDataStream &  out, const DescentConnectorGroup & group) {
+		out<<group.descriptions;
+		out<<(int)group.dir;
+		out<<group.plural;
+		return out;
+	}
+	friend QDataStream & operator >> (QDataStream &  in, DescentConnectorGroup & group) {
+		in>>group.descriptions;
+		int dir;
+		in>>dir;
+		group.dir=(DescentDirection)dir;
+		in>>group.plural;
+		return in;
+	}
+
 public:
 	DescentConnectorGroup(DescentDirection dir, bool plural=false) {
 		this->dir=dir;
@@ -130,6 +150,15 @@ public:
 class DescentConnectors {
 private:
 	QList<DescentConnectorGroup> groups;
+
+	friend QDataStream & operator << (QDataStream &  out, const DescentConnectors & conn) {
+		out<<conn.groups;
+		return out;
+	}
+	friend QDataStream & operator >> (QDataStream &  in, DescentConnectors & conn) {
+		in>>conn.groups;
+		return in;
+	}
 public:
 	void addDescentGroup(DescentConnectorGroup group) {
 		groups<<group;
@@ -320,20 +349,8 @@ public:
 	}
 };
 
-void geneology_initialize() {
-	long abstract_NAME=database_info.comp_rules->getAbstractCategoryID("Hebrew Bible Names");
-	int bit_NAME=database_info.comp_rules->getAbstractCategoryBitIndex(abstract_NAME);
-	bits_gene_NAME.append(bit_NAME);
-#if 0
-	abstract_NAME=database_info.comp_rules->getAbstractCategoryID("christian names");
[…]

### 2011-11-04 trunk/ATSarf/src/case/biographies.h

@@ -198,53 +198,57 @@ public slots:
 	}
 	void colorBiography_clicked() {
 	#ifdef DISPLAY_BIOGRAPHY_GRAPH
-		int num=biographyNum->currentText().toInt();
-		setCurrentAction("Display Graph");
-		report(0);
-		DisplayNodeVisitorColoredBiography visitor(num);
-		GraphVisitorController c(&visitor,graph,true,true);
-		graph->DFS_traverse(c);
-		setCurrentAction("Completed");
-		report(100);
-		try{
-			system("dot -Tsvg graph.dot -o graph.svg");
-			pic->setPixmap(QPixmap("./graph.svg"));
-			subScrollArea->setWidget(pic);
-		}catch(...) {}
-	#ifdef ERRORS_BIO
-		errors->setText(*errors_text);
-		errors_text->clear();
-	#endif
+		if (graph->size()<MAX_DISPLAYABLE_SIZE) {
+			int num=biographyNum->currentText().toInt();
+			setCurrentAction("Display Graph");
+			report(0);
+			DisplayNodeVisitorColoredBiography visitor(num);
+			GraphVisitorController c(&visitor,graph,true,true);
+			graph->DFS_traverse(c);
+			setCurrentAction("Completed");
+			report(100);
+			try{
+				system("dot -Tsvg graph.dot -o graph.svg");
+				pic->setPixmap(QPixmap("./graph.svg"));
+				subScrollArea->setWidget(pic);
+			}catch(...) {}
+		#ifdef ERRORS_BIO
+			errors->setText(*errors_text);
+			errors_text->clear();
+		#endif
+		}
 	#endif
 	}
 	void colorNarrators_clicked() {
 	#ifdef DISPLAY_BIOGRAPHY_GRAPH
-		QList<QTableWidgetSelectionRange>  selection=narratorListDisplay->selectedRanges();
-		ColorNarratorsAction::DetectedNodesMap map;
-		ColorNarratorsAction action(map);
-		for (int i=0;i<selection.size();i++) {
-			int topRow=selection[i].topRow();
-			for (int j=0;j<selection[i].rowCount();j++) {
-				int row=topRow+j;
-				Narrator * n=narratorList[row];
-				graph->performActionToAllCorrespondingNodes(n,action);
+		if (graph->size()<MAX_DISPLAYABLE_SIZE) {
[…]

### 2011-11-04 trunk/ATSarf/src/case/graph.h

@@ -2089,6 +2089,7 @@ inline int test_GraphFunctionalities(ChainsContainer &chains, ATMProgressIFC *pr
 		graph->setFileName(text,fileName);
 	}
 	QString allName=fileName;
+#ifndef SUBMISSION
 	QFile file(fileName.remove(".txt")+".por");
 	file.remove();
 	if (!file.open(QIODevice::ReadWrite))
@@ -2096,6 +2097,7 @@ inline int test_GraphFunctionalities(ChainsContainer &chains, ATMProgressIFC *pr
 	QDataStream fileStream(&file);
 	graph->serialize(fileStream);
 	file.close();
+#endif
 	prg->setCurrentAction("Completed");
 	prg->report(100);
 	calculateStatisticsOrAnotate(chains,graph,text,allName);

### 2011-11-04 trunk/ATSarf/src/case/hadith.h

@@ -56,13 +56,13 @@
 #define TAG_BIOGRAPHY
 #define SEGMENT_AFTER_PROCESSING_ALL_BIOGRAPHY
 
-//#define DISPLAY_BIOGRAPHY_GRAPH
+#define DISPLAY_BIOGRAPHY_GRAPH
+#define MAX_DISPLAYABLE_SIZE 3000
 #define TAG_HADITH
 
 #define SAVE_MERGED_GRAPH_ONLY
 #define DESERIALIZE_POR
 //#define DEBUG_BIOGRAPHY_CLUSTERING
-#define BIBLE
 
 #define NONCONTEXT_LEARNING
 #define INTER_ANNOTATION_AGREEMENT

### 2011-11-04 trunk/ATSarf/src/case/hadithChainGraph.cpp

@@ -48,11 +48,13 @@ int HadithChainGraph::read(ChainsContainer & chains, ATMProgressIFC * prg, QStri
 				if (this_->narr==NULL) {
 					this_->narr=n;
 				} else {
-					QMessageBox msgBox;
-					msgBox.setIcon(QMessageBox::Information);
-					msgBox.setWindowTitle("Warning Processing Graph");
-					msgBox.setText(QString("More than one narrator was found in supposedly one text. The others will be ignored"));
-					msgBox.exec();
+					if (this_->displayWarnings) {
+						QMessageBox msgBox;
+						msgBox.setIcon(QMessageBox::Information);
+						msgBox.setWindowTitle("Warning Processing Graph");
+						msgBox.setText(QString("More than one narrator was found in supposedly one text. The others will be ignored"));
+						msgBox.exec();
+					}
 					break;
 				}
 			}
@@ -61,13 +63,14 @@ int HadithChainGraph::read(ChainsContainer & chains, ATMProgressIFC * prg, QStri
 	return 0;
 }
 
-Narrator * HadithChainGraph::getNarrator(QString &text) {
+Narrator * HadithChainGraph::getNarrator(QString &text, bool displayWarnings) {
 	Name n(&text,0,text.size()-1);
-	return getNarrator(n);
+	return getNarrator(n,displayWarnings);
 }
 
-Narrator * HadithChainGraph::getNarrator(const Name & name) {
+Narrator * HadithChainGraph::getNarrator(const Name & name,bool displayWarnings) {
 	NarratorReader r;
+	r.displayWarnings=displayWarnings;
 	segmentNarrators(name.getTextPointer(),name.getStart(),name.getEnd(),&read,&r);
 	bool addEdges=false;
 	if (r.narr==NULL) {

### 2011-11-04 trunk/ATSarf/src/case/hadithRefactored.cpp

@@ -411,8 +411,8 @@ private:
 						<<stemmings<<"\n";
 	#endif
 		//delete text;
-		if (currentChain!=NULL)
-			delete currentChain;
+		/*if (currentChain!=NULL)
+			delete currentChain;*/
 		return 0;
 	}
 

### 2011-11-04 trunk/ATSarf/src/gui/mainwindow.cpp

@@ -19,6 +19,12 @@ MainWindow::MainWindow(QWidget *parent) :
 #ifdef EQUAL_NEW
 	m_ui->EQ_threshold->setText("0.1");
 #endif
+#ifdef SUBMISSION
+	m_ui->chk_AffixBreaker->setVisible(false);
+	m_ui->chk_verification->setVisible(false);
+#else
+	//m_ui->chk_detailed_Statistics->setVisible(false);
+#endif
 }
 
 MainWindow::~MainWindow() {
@@ -85,7 +91,6 @@ void MainWindow::startTaggingText(QString & text){
 	taggedBox->setTextBackgroundColor(Qt::white);
 	taggedBox->setTextColor(Qt::black);
 	taggedBox->setText(text);
-
 }
 
 void MainWindow::finishTaggingText(){
@@ -166,7 +171,7 @@ void MainWindow::on_chk_bible_toggled(bool checked){
 
 void MainWindow::on_pushButton_clicked(){
 	QString error_str,output_str,hadith_str;
-	bool v1,v2,v3,v4,v5,v6;
+	bool v1,v2,v3,v4,v5,v6=true;
 	int narr_min=m_ui->NARRATOR->toPlainText().toInt(&v1);
 	int nmc_max=m_ui->NMC->toPlainText().toInt(&v2);
 	int nrc_max=m_ui->NRC->toPlainText().toInt(&v3);
@@ -183,13 +188,15 @@ void MainWindow::on_pushButton_clicked(){
 		geneologyParameters.theta_0=nmc_max;
 		geneologyParameters.C_max=nrc_max;
 		geneologyParameters.radius=equality_radius;
+		geneologyParameters.refined=m_ui->chk_genealogyRefined->isChecked();
 	}
 	hadithParameters.equality_threshold=m_ui->EQ_threshold->toPlainText().toDouble(&v6);
 	hadithParameters.display_chain_num=m_ui->chk_chainNum->isChecked();
 	hadithParameters.break_cycles=m_ui->chk_breakCycles->isChecked();
+	hadithParameters.detailed_statistics=geneologyParameters.detailed_statistics=m_ui->chk_detailed_Statistics->isChecked();
 	sarfParameters.enableRunonwords=m_ui->chk_runon->isChecked();
 	if ((m_ui->chk_hadith->isChecked() || m_ui->chk_bible->isChecked()) && (!v1 || !v2 || !v3 || !v4 || !v5 || !v6)){
[…]

### 2011-11-24 trunk/ATSarf/src/case/hadith.h
updated code of biography segmenter (with 2 algorithms: the practical one is using lookups, the other is a true segmentation algorithm)

@@ -6,7 +6,12 @@
 #include "ATMProgressIFC.h"
 
 
-//#define SUBMISSION //Dont forget to do so when u want to compile for submission to account for change in filename path
+#define SUBMISSION //Dont forget to do so when u want to compile for submission to account for change in filename path
+#define WRITE_POR
+#define BIOGRAPHY_SEGMENT
+//#define SEGMENT_BIO_ALGORITHM
+#define SEGMENT_AFTER_PROCESSING_ALL_BIOGRAPHY
+//#define FAST_NEIGHBOR_CHECK
 
 //#define DEBUG_BFS_TRAVERSAL
 //#define DEBUG_DFS_TRAVERSAL
@@ -54,7 +59,6 @@
 #define SEGMENT_BIOGRAPHY_USING_POR
 #define HASH_GRAPH_BUILDING
 #define TAG_BIOGRAPHY
-#define SEGMENT_AFTER_PROCESSING_ALL_BIOGRAPHY
 
 #define DISPLAY_BIOGRAPHY_GRAPH
 #define MAX_DISPLAYABLE_SIZE 3000

### 2011-11-24 trunk/ATSarf/src/case/narratorHash.h

@@ -20,7 +20,7 @@ public:
 	class FoundAction: public Action {
 	public:
 		virtual void action(const QString & searchKey, GroupNode * node, double similarity) =0;
-		virtual bool isFoundAction() {return true;}
+		bool isFoundAction() {return true;}
 	};
 	class BiographyAction: public Action {
 	public:

### 2011-11-24 trunk/ATSarf/src/case/narrator_abstraction.cpp

@@ -9,6 +9,7 @@
 #include "hadith_utilities.h"
 #include "narratorHash.h"
 #include "graph.h"
+#include "biographyGraphUtilities.h"
 
 #ifdef EQUALITYDEBUG
 	inline void display(QString t) {
@@ -714,7 +715,7 @@ inline double equalNew(const Narrator & n1,const Narrator & n2) {
 				name1[0]=alef;
 			if (name2.size()>0 && alefs.contains(name2[0]))
 				name2[0]=alef;
-			if (name1!=name2)
+			if (name1!=name2) //TODO: check if this works when diacritics are available???
 				return 0;
 			else
 				equal=true;
@@ -818,39 +819,6 @@ void Biography::serialize(QTextStream &chainOut) const
 }
 
 #ifdef SEGMENT_BIOGRAPHY_USING_POR
-class RealNarratorAction:public NarratorHash::FoundAction {
-private:
-	Biography::NarratorNodeList & list;
-	bool found;
-public:
-	RealNarratorAction(Biography::NarratorNodeList & nodeList):list(nodeList) {}
-	virtual void action(const QString & , GroupNode * node, double similarity){
-		if (similarity>hadithParameters.equality_threshold) {
-			found =true;
-		#ifndef SEGMENT_AFTER_PROCESSING_ALL_BIOGRAPHY
-			NarratorNodeIfc *n=&node->getCorrespondingNarratorNode();
-		#if 1
-			if(n==NULL) //is a chain node
-				list.append(Biography::MatchingNode(&(*node)[0],similarity));
-			else 
-		#else
-			assert(n!=NULL);
-		#endif
-		#else
-			NarratorNodeIfc *n=node;
-		#endif
-			int index;
-			if (!Biography::MatchingNode::contains(list,n,similarity,index)) {
-				list.append(Biography::MatchingNode(n,similarity));
-			} else if (index>=0) { //i.e. found but with a smaller similarity
-				list[index]=Biography::MatchingNode(n,similarity);
-			}
-		}
-	}
-	void resetFound() {found=false;}
-	bool isFound() {return found;}
-};
-
 bool Biography::isRealNarrator(Narrator * n) {
 	int i;
 	if (!nodeGroups.isEmpty()) {
@@ -863,13 +831,11 @@ bool Biography::isRealNarrator(Narrator * n) {
 		i=0;
 	}
 
-	RealNarratorAction v(nodeGroups[i]);
[…]

### 2011-11-24 trunk/ATSarf/src/case/split_affixes.cpp

@@ -53,6 +53,7 @@ void SplitDialog::split_action() {
 						error<<"Inconsistent Reverse Description.\n";
 						continue;
 					}
+				#if 0 //seems wrong logic
 					if (reverse_description){
 						originalAffixList->setItem(rowIndex2,6,new QTableWidgetItem(QString("1")));
 						QString category2=originalAffixList->item(rowIndex2,2)->text();
@@ -60,6 +61,7 @@ void SplitDialog::split_action() {
 						KEEP_OLD=false;
 						insert_item(t,affix2,raw_data2,category2,source_id,new QList<long>,description2,pos2,"","1");
 					}
+				#endif
 				}
 				QString category2=originalAffixList->item(rowIndex2,2)->text();
 				remove_item(t,affix_id1,raw_data1,database_info.comp_rules->getCategoryID(category1),database_info.descriptions->indexOf(description1),pos1);

### 2011-11-24 tex/cicling.tex
updated code of biography segmenter (with 2 algorithms: the practical one is using lookups, the other is a true segmentation algorithm)

--- 2011-10-12 tex/hadith.ijcnlp.tex
@@
 \usepackage{rotating}
-\usepackage{amsthm}
-\usepackage{booktabs}
+%\usepackage{amsthm}
+%\usepackage{booktabs}
 
 \usepackage{relsize}
-\usepackage{fancyvrb}
+%\usepackage{fancyvrb}
 \usepackage[colorlinks=false,draft]{hyperref}
-
+%\usepackage[colorlinks=false]{hyperref}
+
+%\allowarab{\fbox}
 \usepackage{utf8}
@@
 %\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
-\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
+\newcommand{\true}{\ensuremath{\mathit{true}}}
+\newcommand{\false}{\ensuremath{\mathit{false}}}
+
+\newcommand{\cci}[1]{{\small\texttt{#1}}}
+\newcommand{\chci}[1]{\cci{#1}}
+
 \newcommand{\frl}[1]{\fbox{\RL{#1}}} 
@@
 \newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}  
+\newcommand{\utfRL}[1]{\setcode{utf8} \RL{#1} \setcode{standard}}
+\newcommand{\noTrnoVocutfRL}[1]{\setcode{utf8} \transfalse\novocalize\RL{#1}\vocalize \transtrue \setcode{standard}}
+\newcommand{\noTrutfRL}[1]{\setcode{utf8} \transfalse\RL{#1}\transtrue \setcode{standard}}
+\newcommand{\noArutfRL}[1]{\setcode{utf8} \arabfalse \RL{#1} \arabtrue \setcode{standard}}
 
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Automatic Hadith Graphs using Arabic Morphological Analysis}
+\title{Detection of Arabic Entity Graphs using Morphology, Finite State Machines, and Graph Transformations}
 
@@
 %}
-
 \date{}
@@
 \begin{document}
-
 \maketitle
@@
 \begin{abstract}
-\end{list}
-A hadith in Islamic literature is a narration from the 
-prophet Mohammad related by multiple narrators.
-Establishing the authenticity of a hadith is an important task
-based on which Islamic scholars issue rules that affect the life
-of people around the world. 
-The literature lacks an automated and exhaustive mechanism of 
-authentication checks. 
-This is currently manual work prone to error and almost
-impossible to complete in the life time of a scholar due to the
-large literature available. 
-In this paper, we present a tool that uses 
-an Arabic morphological analyzer 
-to successfully automate the
-analysis of three books of hadith selected 
-arbitrarily and abstract 
-each book into a graph of narrators related to each other. 
+Research on automatic recognition of named entities
+%events and relations amongst them 
+from Arabic text uses techniques that work well for the 
+Latin based languages such as local grammars, 
+statistical learning models, pattern matching, 
+and rule-based techniques. 
+These techniques boost their results by using application 
+specific 
+corpora, parallel language corpora,
+and morphological stemming analysis. 
+We propose a method for extracting entities, events, 
+and relations amongst them
+from Arabic text using 
+a hierarchy of finite state machines driven by morphological 
+features such as part of speech and gloss tags,
+and graph transformation algorithms. 
+We evaluated our method on three natural language processing 
+applications.
+We automated the extraction of narrators and 
+narrator relations from 
+several corpora of Islamic narration books (hadith). 
+We automated the extraction of genealogical family trees 
+from Biblical texts. 
+Finally, we automated locating and extracting
+individual biographies from historical biography books.
+In all applications, our method reports high precision and recall
+and learns lemmas about phrases that improve results. 
 \end{abstract}
@@
 \section{Introduction}
-
-A \RL{.hady_t}~\footnote{In this document, 
-we use the default ArabTeX transliteration style ZDMG.}
-(tradition) 
+Named entity recognition (NER) is harder for Arabic text than 
+it is for Latin. 
+Arabic does not have upper case letters which distinguish the 
+start and end of named entities (NE).
+Arabic is morphologically rich, 
+more ambiguous than Latin languages,
+and its short vowels are frequently omitted~\cite{Debili1998}.
+
+Industrial tools that perform NE extraction for Arabic exist~\cite{Sak09,IdentiFinder,basis:06,ANEE:07},
+but little is said about the techniques behind them and no formal evaluation of the tools exist. 
+Researchers proposed and evaluated 
+local grammars with morphological stemming~\cite{ZAGHOUANI10,Traboulsi:09},
+rule-based systems powered with local grammars~\cite{ShaalanR09} 
+and stop words ~\cite{Abuleil04},
+statistical analysis boosted with Arabic-English parallel corpora~\cite{BenajibaZDR10}, 
+statistical analysis boosted with task specific gazetteers~\cite{ANERSys:07,BenajibaDR08},
+and pattern matching powered with morphological stemming~\cite{jumaily2011,TAGARAB98}
+to perform Arabic NER.
+We discuss these techniques and compare to them in 
+the ``Related work'' Section.
+
+In this paper, we consider the problem of NER
+in the context of more complex natural language
+processing (NLP) tasks that look for NEs and 
+relations amongst the detected NEs and build entity graphs.
+An {\em entity graph} is a graph where NEs are nodes and relational entities (RE) are edges.
+For example, consider the tasks of extracting a time table
+from a document describing the events of a conference, 
+extracting genealogical family trees based on person names 
+from biblical text,  and
+extracting routes and maps from travel itineraries.
+Also consider applications with cross-document NER
+such as extracting narrator authentication graphs based on narration 
+documents that include narrator sequences, and biography documents 
+that include credibility information about the narrators. 
+We hypothesize that NEs of the same kind happen 
+at proximity of each other, and we base our method on that 
+working hypothesis.
+We target NLP applications and corpora where our working hypothesis
+obviously holds.
+These applications help build databases of NEs
+that we can later use in other related contexts.
+
+We propose a method for extracting NEs and 
+relations amongst entities using 
+morphological features such as part of speech (POS) and gloss tags, 
+a hierarchy of finite state machines, 
+and graph transformations such as node merge, node split, and replace edge transformations. 
+Morphological analysis helps reduce ambiguities and resolves lexical items with their morphological variations.
+For each application, we augment the lexicon of an inhouse morphological 
+analyzer with lexical elements and tags that describe categories relevant to the application. 
+We also build finite state transducers (FST) that take 
+morphological tags as input and detect the NEs.
+Manually built FSTs can express rules that are not intuitive (and sometimes impossible) 
+to express with local grammars using regular expressions.
+Consider, for example, counting the number of tags up to a threshold, or capturing left and right recursive structures 
+which are common in Arabic since an Arabic phrase-sentence can start with a name or a verb.
+We finally infer relations between the NEs based on their context.
+
+Our method takes text and computes morphological features that match, precede, follow, or connect NEs
+and uses the features to learn unknown NEs.
+Our method also computes morphological features that correspond to relations between NEs, and learns semantics
+for them. 
+For example, consider a graph with two predefined edge labels $l_s=$``spouse'' and $l_p=$``parent'', 
+with names of persons $n_1, n_2,$ and $n_3$ as nodes,
+and with $(n_1,n_2,l_s)$, $(n_2,n_3,l_p)$, and $(n_1,n_3,l_p)$ as edges inferred from a document.
+Our method can learn that the expression \utfRL{ولدت له} (gave him a child) connecting 
+$n_1$ to $n_2$ and $n_3$ means that $n_2$ is the spouse of $n_1$
+and the parent of $n_3$.
+The learned semantics for the expression can be expressed as a graph transformation that leads to a graph with only 
+$l_s$ and $l_p$ as labels.
+
+We evaluated our method and used it to solve three 
+NLP applications including extracting genealogical family
+trees from Biblical text, extracting a graph of 
+narrators from hadith books, and extracting biographies from biography books.
+Our results show high precision and recall metrics and report
+learning interesting named entities and relational lemmas.
+In this paper we make the following contributions.
+\begin{itemize}
+\item We present a method for named entity and entity graph
+extraction from Arabic text using morphology, state 
+machines, and graph transformation algorithms.
+\item We present an Arabic lexicon augmented 
+with (1) person names collected from several resources as well as
+learned from our corpora, and (2) tags suitable to extract complex names,
+kinship, and narration sequence relations. 
+\item We learn semantics for morphological features associated 
+with detected phrases that are equivalent to predefined features
+modulo one or two graph transformations. 
+The learned semantics improved recall and precision by 
+5 and 6\%, respectively in the genealogy application.
+\item We perform Arabic cross-document named entity and 
+relational entity extraction using graph algorithms. 
+The use of a graph generated from the hadith application 
+improved precision in the biography application from 40\% to 93\%.
+%\item We present a novel equality metric to compare Arabic 
+%names based on name structure and morphological features. 
+\end{itemize}
+
+\section{The method}
+
+Our method takes a set of Arabic text documents and builds 
+a graph where named entities are nodes and edges are relations 
+between the nodes. 
+Our target graph has a predefined set of relations expressed 
+as edge labels such as ``parent'' and ``spouse'' where the nodes
+are person names.
+We use an inhouse morphological analyzer augmented with 
+a list of lexicon elements relevant to the application.
+We also augment the analyzer with
+a set of tags that classify the 
+elements into several categories 
+relevant to the targeted named entities and edge labels.
+
+The method passes the input text string to the morphological 
+analyzer. The analyzer processes the input text and whenever
+it identifies a morpheme, it calls the application morpheme
+classifier with the morpheme, the tags of the morpheme, 
+and the current solution context.
+Note that several morphological solutions may exist for the same
+input string.
+The morpheme classifier either stores the morpheme in a list of
+unresolved morphemes, 
+or resolves the unresolved morphemes and produces
+a category as input to the application FST.
+The application FST is manually built to
+detect {\bf (1)} named entities,  
+{\bf (2)} sequences of named entities,  
+{\bf (3)} predefined edges connecting named entities, 
+{\bf (4)} previously unknown named entities occurring between predefined edges, 
+and {\bf (5)} previously unknown edge labels occurring between named entities.
+%Note that there might be more than one level of finite state
+%transducers to handle an NLP application.
+The detected named entities and edges form an entity graph.
+
+Finally, the method uses a morphological and structural distance
+function to compute equality between the named entities and
+to consequently merge the nodes of the graph that represent 
+equal named entities.
+%When nodes are merged, the edge labels are augmented to be the 
+%concatenation of old edges.
+The method then learns semantics for the discovered edge labels 
+and suggests edge replacement algorithms to transform the 
+discovered edge labels into the predefined edges. 
+The method also interactively learns the semantics of other 
+detected edge labels that it could not infer 
+in terms of the predefined set of relations.
+
+%-importance of cluster (Min number of entities in context)
+%-importance of tolerance 
+%-entities using lexicons
+%-clusters
+%-
+
+\section{The hadith application}
+A \RL{.hady_t} (tradition) 
 is a narration related to the prophet Mohammad
@@
 The collections of traditions are the second source of
-jurisprudence after the \RL{qor'An} for all Islamic schools of thought. 
+jurisprudence after the \RL{qor'An} for all Islamic 
+schools of thought. 
 Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
@@
 The second part of the \noArRL{.hady_t} is the 
-\RL{matn} (content) and constitutes the actual content 
-of the tradition.
-
-Due to religious and political reasons, 
-writing the traditions was forbidden 
-until the days of the eighth Umayad Calif,
+\RL{matn} (content) and constitutes its actual content.
+
+Due to religious and political reasons, writing the traditions was forbidden 
+until the days of the Umayad Calif,
 \novocalize
-\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after 
-the death of the prophet~\cite{AlAskari}. 
-Consequently, many inconsistencies were
-introduced to the literature which necessitated 
-a thorough authentication study of a tradition before its use in 
-jurisprudence.
+\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after the prophet. 
+Many inconsistencies were introduced to the literature which necessitates 
+a thorough authentication study before the use of a tradition in jurisprudence.
 \vocalize
@@
 %jurisprudence. 
-While different Islamic schools of thought differ on 
-how to interpret the content, they almost all agree
-that if a \noArRL{sanad} lacks authenticity, 
-then scholars can not use the tradition for jurisprudence.
+%While different Islamic schools of thought differ on 
+%how to interpret the content, they almost all agree
+All Islamic schools of thought agree
+that scholars can not use a tradition whose sanad lacks authenticity for jurisprudence.
 
@@
 of existing traditions and tradition books. 
-Hadithopedia~\cite{Hadithopaedia:08}
-reports that the tradition and biography
-books for one of the Islamic sects amounts to more than 
-300 thousand lines of text. 
-Al-Azami\shortcite{Al-Azami-91} cites more than eleven books
+%Hadithopedia~\cite{Hadithopaedia:08}
+%reports that a non-comprehensive collection of tradition 
+%and biography books amounts to more than 300 thousand lines of 
+%text. 
+Al-Azami\cite{Al-Azami-91} cites more than eleven books
 of digitized tradition books each of several volumes, and a dozen
@@
 
+%In this paper, we present the hadith {\em narrator graph
+%extractor using morphological analysis} (NaGEMA), 
+%a novel technique that automates
+%extracting the \noArRL{sanad} from the books of traditions 
+%into a diagram of sequences of narrators. NaGEMA segments
+%a book into narrations, each narration into its \noArRL{matn} 
+%and its \noArRL{sanad}, 
+%and each sanad into its separate narrators. 
+%This work will lead to 
+%automate an exhaustive \noArRL{.hady_t} authentication 
+%effort in the future as we plan to 
+%automatically analyze the biography books.
+
+%Automated analysis of Arabic data sets, including texts, 
+%publications, records and digital media is essential
+%with the huge digital Arabic content available nowadays. 
+%\subsection{ Arabic morphological analysis}
+%Arabic morphological analysis is key to our analysis. 
+%Current morphological analyzers~\cite{Sughaiyer:04}
+%use concatenative analysis when
+%considering the internal structure 
+%of an Arabic word and
+%composing it into several {\em morphemes}. 
+%A morpheme can be a {\em stem}, or an {\em affix}.
+%An affix can be a {\em prefix, suffix, } or an {\em infix}.
+%%The analysis of one word may lead to several possible
+%%morphological solutions.
+%\vocalize
+%The word \RL{'a.hmadH}
+%may have two valid morphological analyses. 
+%The letter \RL{'a} may be a prefix and the word means 
+%``I praise him'', or 
+%%The letter \RL{'a} may also be 
+%part of the stem \RL{'a.hmad} (a proper noun)
+%and the word means ``his Ahmad''.
+%
+%The morphological analysis of the word \RL{.hd_tnA} 
+%``narrated to us'' that starts the hadith in Figure~\ref{f:exhadith}
+%returns the stem \RL{.hd_t} ``narrated''.
+%The stem  \RL{.hd_t} is also the stem of other 
+%words such as \RL{.hd_tny} ``narrated to me'' and \RL{.hd_thm} ``narrated to them''.
+%The stem also shares similar part of speech (POS)
+%and English meaning gloss annotations with words
+%such as \RL{qaal} ``said'' and \RL{'a_hbar} ``told''. 
+%The use of morphological analysis allows us to 
+%subsume listing these stems and all of their morphological
+%derivations with one category that matches the gloss and 
+%POS tags. 
+%Similarly the words \RL{ibn} ``the son of'' and \RL{bn} 
+%return the same morphological analysis.
+%
+%
+%\subsection{Narrator sequence extraction }
+
+%\begin{figure*}
+%\center{
+%\resizebox{.9\columnwidth}{!}
+%{\includegraphics{figs/detailed_submachine.pdf}
+%}
+%\caption{The sequence of narrators extractor diagram}
+%\label{f:sequence}
+%}
+%\end{figure*}
+
 \novocalize
@@
 \center{
-\resizebox{.9\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
+\resizebox{.95\columnwidth}{!}
+{ \input{figs/exhadith2cols.pdftex_t}}
+\caption{Hadith narrator sequence example.}
 \label{f:exhadith}
@@
 \end{figure}
-% the FSMs for the three words
 \transtrue
@@
 
-In this paper, we present the hadith {\em narrator graph
-extractor using morphological analysis} (NaGEMA), 
-a novel technique that automates
-extracting the \noArRL{sanad} from the books of traditions 
-into a diagram of chains of narrators. NaGEMA segments
-a book into narrations, each narration into its \noArRL{matn} 
-and its \noArRL{sanad}, 
-and each sanad into its separate narrators. 
-This work will lead to 
-automate an exhaustive \noArRL{.hady_t} authentication 
-effort in the future as we plan to 
-automatically analyze the biography books.
-
-%Automated analysis of Arabic data sets, including texts, 
-%publications, records and digital media is essential
-%with the huge digital Arabic content available nowadays. 
-\subsection{ Arabic morphological analysis}
-Arabic morphological analysis is key to our analysis. 
-Current morphological analyzers~\cite{Sughaiyer:04}
-use concatenative analysis when
-considering the internal structure 
-of an Arabic word and
-composing it into several {\em morphemes}. 
-A morpheme can be a {\em stem}, or an {\em affix}.
-An affix can be a {\em prefix, suffix, } or an {\em infix}.
-%The analysis of one word may lead to several possible
-%morphological solutions.
-\vocalize
-The word \RL{'a.hmadH}
-may have two valid morphological analyses. 
-The letter \RL{'a} may be a prefix and the word means 
-``I praise him'', or 
-%The letter \RL{'a} may also be 
-part of the stem \RL{'a.hmad} (a proper noun)
-and the word means ``his Ahmad''.
-
-The morphological analysis of the word \RL{.hd_tnA} 
-``narrated to us'' that starts the hadith in Figure~\ref{f:exhadith}
-returns the stem \RL{.hd_t} ``narrated''.
-The stem  \RL{.hd_t} is also the stem of other 
-words such as \RL{.hd_tny} ``narrated to me'' and \RL{.hd_thm} ``narrated to them''.
-The stem also shares similar part of speech (POS)
-and English meaning gloss annotations with words
-such as \RL{qaal} ``said'' and \RL{'a_hbar} ``told''. 
-The use of morphological analysis allows us to 
-subsume listing these stems and all of their morphological
-derivations with one rule that matches the gloss and 
-POS tags. 
-Similarly the words \RL{ibn} ``the son of'' and \RL{bn} 
-return the same morphological analysis.
-
-\novocalize
-%The accuracy of the solutions suffer due to inherent difficulties
-%of morphological analysis of the Arabic language. 
-%For example, it is common practice to write Arabic text
-%without short vowels. 
-%This greatly increases the ambiguity of Arabic text. 
-%Arabic letters can have up to 
-%four different forms
-%corresponding to their position in a word, i.e, beginning,
-%middle, end of word and separate forms. 
-%This allows the phrase \transfalse
-%\RL{il_A\nospace almdrsT} \transtrue
-%to be visually recognizable
-%as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
-%without the need of a space in between. 
-%The reason is the first word \RL{il_A} ends with
-%\RL{_A} a non-connecting letter. 
-%These words,
-%referred to as ``run-on'' words~\cite{Buckwalter:04},
-%occur often, and greatly increase the
-%difficulty of tokenization.
-
-Current morphological analyzers such as 
-Buckwalter~\shortcite{Buckwalter:02},
-Beesley~\shortcite{Beesley:01},
-SAMA~\cite{Kulick:10},
-and ElixirFM~\cite{Otakar:07} 
-take as input white space delimited tokens~\cite{Kulick:10},
-consider them as words,
-and enumerate all possible solutions. 
-This approach has several problems. 
-First, a white space delimited token may have 
-more than one word, referred to as ``run-on'' 
-words~\cite{Buckwalter:04}.
-Second, the exhaustive enumeration may not be efficient and may
-not be necessary or appropriate
-in some applications as noted in~\cite{Maamouri:10}. 
-%Other morphological analyzers such as 
-%Amira~\cite{Diab:07,Benajiba:07},
-%MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
-%use machine learning and support vector machines (SVM) 
-%to disambiguate the morphological analysis at the expense 
-%of efficiency. 
-%Xerox rules can be compiled into specialized finite state
-%machine (FSM) based analyzers as described in~\cite{Beesley:03}.
-%However, the efficiency of the resulting analyzer depends on the
-%way the Xerox rules are written. 
-%This requires deep knowledge and insight from the user
-%in compilation techniques, context free grammars, 
-%and morphological analysis.
-%context free grammar rules that can be automatically compiled
-%into efficient non-deterministic finite state Automata (NDFSA).
-
-%\subsection{Sarf}
-%\label{sec:intro:sarf}
-
-{\bf Sarf.~~~}
-Sarf~\footnote{Sarf (name modified for blind review) }
-is 
-an {\em application-specific efficient
-morphological analyzer} based on
-non-deterministic finite state Automata 
-driven by an application-specific controller customized 
-by the user. 
-%Each machine in Sarf takes one letter at a time as input
-%and represents a valid morphological analysis.
-The application-specific controller prunes false positives
-early in the run by making a decision at every input letter
-to control the machines.
-Sarf builds on and extends the lexicon of 
-Buckwalter~\shortcite{Buckwalter:02} 
-with proper and location names extracted from different online 
-sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
-\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
-as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
-
-%Each alive machine in Sarf takes as input one letter at a time 
-%from a text stream. 
-Sarf does not assume that the word at hand is a token and
-performs tokenization on the fly based on morphological correctness
-and thus can handle 
-the ``run-on'' words problem. 
-Sarf also uses diacritics if they exist to disambiguate 
-morphological solutions.
-
-In this paper we make the following contributions. 
-\begin{itemize}
-\item We automate the segmentation of hadith books into separate narrations and extract chains of narrators 
-%into a partial order relation graph 
-using a knowledge based approach with morphological analysis.
-\item We optimize our analysis through utilizing punctuation and learning previously unknown names.
-\item We build from different chains of narrators a partial order relation graph based on merging common narrators.
-\item We introduce a specialized distance metric for Arabic names used in Arabic literature. 
-\item We highly outperform the currently available narrator extraction tool. 
-\item We provide a working and efficient implementation of our work. 
-\end{itemize}
-
-The rest of this paper is organized as follows. 
-We describe our approach to narrator chain extraction
-in Section~\ref{sec:islamic}.
-We describe how we build the partial order relation
-graph in Section~\ref{sec:graph}.
-We discuss related work in Section~\ref{sec:related}.
-We present our results in Section~\ref{sec:results} and 
-we conclude in Section~\ref{sec:future}.
-
-%The hadith 
-%application specific controller that controls Sarf considers 
-%proper names and ignores text where
-%proper names occur rarely.
-%It also considers words that relate names to each other 
-%such as \RL{bin} or ``the son of''
-%and words that mean narrate such as \RL{qaal} or ``said''.
-
-%With Sarf, we successfully automated the analysis of 
-%three books of Islamic narrations~\cite{IbnHanbal,AlTousi,AlKulayni}
-%and extracted from each one of them a vector of complex structures 
-%that are three level deep in hierarchy.
-%We also merged these vectors into a graph of narrators where 
-%we represented identical narrators from separate traditions 
-%in one single node.
-
-%The rest of this paper is structured as follows. 
-%In Section~\ref{sec:sarf}
-%we discuss Sarf and illustrate it with a running example.
-%We also explain how non-deterministic finite state Automata
-%work. 
-%In Section~\ref{sec:islamic} we present the hadith application. 
-%We present our results in Section~\ref{sec:results}.
-%In Section~\ref{sec:related}
-%we compare NaGEMA to related work. 
-%We conclude with future work in Section~\ref{sec:future}.
-
-%\section{Background}
-% TODO: nondeterministic finite-state automata
-
-%The survey in~\cite{Sughaiyer:04} compares
-%several morphological analyzers. 
-%Analyzers such as~\cite{Khoja:01,Darwish:02} 
-%target specific applications in the 
-%analyzer itself or use a specific set of POS tags
-%as their reference.
-%Sarf differs in that it is a general morphological 
-%analyzer that reports all possible solutions. 
-%It is case-based in the sense that a controller 
-%prunes these solutions. 
-%
-%Sarf builds upon the lexicon of Buckwalter\shortcite{Buckwalter:02}.
-%It differs from Buckwalter in that it defines a shorter list of affixes
-%and a longer list of affix compatibility rules to allow 
-%affixes to be defined recursively
-%so that we can have prefix-prefix an suffix-suffix 
-%concatenations.
-%This allows Sarf to better maintain and propagate 
-%the POS tags associated with affixes. 
-%%This information is highly 
-%%redundant with Buckwalter.% and that may lead to consistency issues. 
-%Sarf keeps all the possible analyses alive where each analysis
-%corresponds to a set of decisions and positions in the text. 
-%Buckwalter produces a set of solutions for each token 
-%and the several analysis threads of the text need to be 
-%generated as a product of the token solutions. 
-%
-%Similar to Sarf, Beesley~\shortcite{Beesley:01} uses
-%finite state Automata (FSA). 
-%It reports that the number of Automata generated by a compiler
-%for Xerox rules can not be controlled and also reports a difficulty 
-%to compose the FSAs into a single framework. 
-%Sarf however constructs an NDFSA from
-%the controller and the prefix, stem, and suffix deterministic 
-%transducers to solve the problem.
-%Sarf also differs in that it uses recursive affixes in the 
-%affix transducers, and in that the transducers consider
-%input from the controller to modify their behavior. 
-%%With Sarf it is easy to control the FSA with a case-based
-%%controller. 
-%Beesley requires recompilation of the Xerox rules.
-%
-%Similar to The MADA+TOKAN toolkit, Sarf thinks of
-%the several valid morphological analyses as a richness that 
-%should be exploited at a higher abstract level rather than
-%an inaccuracy that should be corrected. 
-%Sarf differs in that it provides an interface for the 
-%case study to interfere in judging the analyses much earlier and
-%at a finer granularity level. 
-%In Sarf, there is no need for a separate tokenizer such as
-%TOKAN since each solution keeps a stack of positions
-%that partition text into morphemes.
-%%The case-based controller can build the trace 
-%%when needed. 
-%
-%%MAGEAD 
-%%Maamouri, Mohamed and \bgroup et al.\egroup },
-%
-%Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
-%SAMA was refined to interact with
-%the ATB~\cite{Maamouri:04} project after the addition of a large 
-%new corpus. 
-%%We find supporting evidence in this decision to our hypothesis
-%%about the utility of a case-based analyzer. 
-%The algorithmic changes in SAMA were
-%done manually and worked in integration with the ATB format. 
-%Sarf is a case-based controller approach that can modify 
-%the behavior of the analyzer on the fly and can interact
-%with any case study. 
-%%In Sarf, the analyzer remains general, and the 
-%%case-based controller specializes the analysis.
-%
-%The AMIRA~\cite{Diab:07} analyzer uses 
-%a language independent SVM based analysis. 
-%The SVM analysis learns features from the input text 
-%and does not make use of any information 
-%in the target query.
-%
-%Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
-%of the Buckwalter analyzer. 
-%Sarf also uses deterministic parsing with tries 
-%to implement the affix and stem transducers. 
-%We think that the inferential-realizational approach 
-%of ElixirFM
-%that is highly compatible with the Arabic linguistic 
-%description~\cite{Badawi:04}
-%can benefit from many features unique to the Arabic language.
-%Sarf leaves implementing that to the case-based controller
-%since in many cases the abstraction needs only a partial 
-%linguistic model of the Arabic language. 
-%%We can compare the case-based controller in Sarf to a high level
-%%component in the ElixirFM domain specific language. 
-%
-
-%Sarf encodes the prefix, suffix and stem lexicons into
-%three deterministic transducers and 
-%uses a non-deterministic construction
-%of the three to compute all valid morphological analyses.
-%We first present a running example of Sarf.
-%The running example explains how Sarf works,
-%implements recursive affixes, and
-%handles ``run-on'' words.
-%Then we discuss how Sarf builds and composes the different FSAs.
-
-\section{Narrator chain extraction }
-\label{sec:islamic}
-
-\begin{figure*}
-\center{
-\resizebox{1.9\columnwidth}{!}
-{\includegraphics{figs/detailed_submachine.pdf}
-}
-\caption{The chain of narrators extractor diagram}
-\label{f:chain}
-}
-\end{figure*}
-
-%Several researchers~\cite{Hadithopaedia:08} attempted to automate 
-%the analysis of the hadith literature. 
-NaGEMA uses Sarf to automate the
-analysis of three books of hadith selected 
-arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
-  the digitized books from online sources such as 
+Several researchers~\cite{iTree,Hadithopaedia:08} attempted to automate 
+the analysis of the hadith literature. 
+We fully automate the analysis of three books of hadith selected 
+arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}~\footnote{We obtained
+  the digitized books from 
   \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
   \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
-%  Our understanding is that those books are historical documents with 
- % public IP.}.
-%\setcode{utf8}
-%\begin{arabtext}
-%حدثنا \emphasize{هشيم}، أخبرنا \emphasize{إسماعيل بن أبي خالد}،
-% عن أبي إسحاق، عن سعيد بن جبير، قال كنا
-%    مع ابن عمر رضي الله عنه حيث أفاض من
-%    عرفات إلى جمع فصلى بنا المغرب ومضى
-%\end{arabtext}
-%\setcode{standard}
-NaGEMA accepts a book as input
-and segments it into a vector of hadiths. 
-NaGEMA segments each hadith into its
-sanad and matn parts. 
-NaGEMA 
-detects the chain of narrators in the sanad and 
+We accept a book as input
+and segment it into a vector of hadiths,
+and segment each hadith into its sanad and matn parts. 
+We detect the sequence of narrators in the sanad and 
 the relation that links each narrator to his ancestor and 
-predecessor in the chain. 
-It also detects the full name of each narrator that is
+predecessor in the sequence. 
+We also detect the full name of each narrator that is
 composed of several proper names with connectors
@@
 
-To do that NaGEMA queries Sarf for the following morphological 
-categories.
+%NRC=19 descriptions= 91 stems
+%family nmc connectors (all)=15 descriptions=33stems
+%       ibn=1 descriptions=3 stems
+%       2ab =2 descriptions=11 stems
+%       2om=4 descriptions =5stems
+
+To do that, we augmented our lexicon with person names obtained from online and Biblical sources,
+and augmented the tag set of our lexicon with the following categories.
 \begin{itemize}
-\item $\mathit{NAME}$: a proper name of a person such as \RL{'a.hmad} .
-\item $\mathit{IBN}$: a special name connector meaning the son of such as \RL{ibn}.
-\item $\mathit{NMC}$: other name connectors that may be any word appearing amongst the other interesting categories. 
-\item $\mathit{NISBA}$: a possessive adjective that qualifies a person such 
-as \RL{alma.sriy} ``the Egyptian''. 
+\item $\mathit{NAME}$: a proper name of a person such as \RL{'a.hmad}. (2,500 male and 3,327 female lexicon items). 
+\item $\mathit{IBN}$: a special name connector meaning ``the child of'' such as \RL{ibn}. (3 lexicon items expressing child,
+11 expressing father, and 5 expressing mother relations).
+\item $\mathit{NMC}$: name connectors that may be any word appearing amongst the other name connector categories. (33 lexicon items).
+\item $\mathit{NISBA}$: a possessive adjective that qualifies a person such as \RL{alma.sriy} ``the Egyptian''. (4,698 possessive city or town, 82 country, and 4 continent items). 
 \item $\mathit{NRC}$: a narrator connector such as
-\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
-\RL{'a_hbar} ``told'' or one of their morphological equivalents. 
+\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', \RL{'a_hbar} ``told'' or one of their morphological equivalents. (91 lexicon items).
 \end{itemize}
-
-The diagram in Figure~\ref{f:chain} describes 
-the chain extraction process. 
-We pass the hadith to the morphological analyzer 
-and check the returned words and category pairs 
-for name categories.
-Next we compute the number of words between every two names
-and compare that to a threshold $\theta_1$.
-If the name is within the threshold we add it to a temporary 
-chain of names. 
-Otherwise, we 
-{\bf 1.} tabulate the chain into a temporary table containing 
-the list of chains and
-{\bf 2.} Create a new chain with the last element.
-Finally, we filter out those chains whose cardinality 
-is less than some threshold  $\theta_2$.
-
-The process also benefits from name connector 
-words and narrator connector words to enhance the
-accuracy of the chain extraction and to segment the 
-chain accurately into narrators.
-Once we meet a name connector or a narrator 
-connector word we reset the word counters that we
-compare against the thresholds.
-
-(Figure~\ref{f:chain} is confusing since it misses some information uses other symbols like 
-those of $\theta$ and is redundant with what follows)
-
-\subsection{The hadith controller}
-\label{sec:controller}
-% FSM for the controller
-
-We implemented our hadith chain extraction 
-in a finite state machine
-controller that interfaces with Sarf.
-Intuitively, the controller reports a valid chain 
+The categories can also be inferred from the concatenation of the tagged morphemes.
+
+\subsection{The hadith finite state transducer}
+\label{sec:FST}
+% FSM for the FST
+
+We implemented our hadith sequence extraction 
+in an FST that reports a valid sequence
 of narrators when a sequence of names
@@
 %such as \RL{`bd alr.hmn} often appearing as ``run-on'' words.
-The controller marks the beginning of the hadith with 
-the beginning of 
-the current proper name sequence,
-and marks the end of the hadith with the beginning of the 
-next proper name sequence. 
-It detects the chain of narrators as the sequence itself.
-
-%The controller also targets words that mean ``narrate'' when
+The FST marks the beginning and the end of the hadith with 
+the beginning of the current name sequence
+and the beginning of the next name sequence, respectively.
+%It detects the sequence of narrators as the sequence itself.
+
+%The FST also targets words that mean ``narrate'' when
 %they appear between proper names. 
-We illustrate the hadith controller 
-in Figure~\ref{f:hadith}. 
-The transitions are labeled using expressions
-that operate on the output tags 
-generated from the Sarf morphological analyzer.
-
-The controller has four states that correspond to 
-a position in text relative to the next sanad. 
+We illustrate the hadith FST
+in Figure~\ref{f:hadith}(a). 
+The transitions are  labeled with expressions
+that operate on the category tags 
+generated from the morphological analyzer.
+The FST has four abstract states.
 State $\mathit{TEXT\_S}$ is the initial state and denotes 
-that the controller is outside the context of a sanad.
-State $\mathit{NAME\_S}$ denotes that the controller is in
-the middle of a chain and has last met a name. 
-State $\mathit{NMC\_S}$ denotes that the controller is in
+that the FST is outside the context of a sanad.
+State $\mathit{NAME\_S}$ denotes that the FST is in
+the middle of a sequence and has met a name. 
+State $\mathit{NMC\_S}$ denotes that the FST is in
 the middle of a narrator name.
-State $\mathit{NRC\_S}$ denotes that the controller is 
-in the context of a narrator name and the controller 
+State $\mathit{NRC\_S}$ denotes that the FST is 
+in the context of a narrator name and the FST 
 has met a narrator connector and is ready to meet 
@@
 The symbol $\mathit{LIST}_{\mathit{nmc}}$ corresponds to the list 
-of name connectors collected since the controller
+of name connectors collected since the FST
 started looping in the state $\mathit{NMC\_S}$.
@@
 that corresponds to a relaxed tolerance measure that
-the controller resorts to in case the words separating
+the FST resorts to in case the words separating
 two names were longer than $\theta_{\mathit{nmc}}$ but 
@@
 
-The controller moves to state $\mathit{NAME\_S}$ on
+The FST moves to state $\mathit{NAME\_S}$ on
 $\mathit{NAME}$.
@@
 State $\mathit{NMC\_S}$
-indicates that the controller expects a name to appear within 
+indicates that the FST expects a name to appear within 
 a tolerance threshold expressed by 
@@
 signaling that the last collection of names met do not qualify
-as a narrator or a chain of narrators.
-
-Similarly, $\mathit{NRC\_S}$ tolerates $\theta_{\mathit{nrc}}$ words 
+as a narrator or a sequence of narrators.
+
+Similarly, $\mathit{NRC\_S}$ 
+tolerates $\theta_{\mathit{nrc}}$ words 
 before it gives up on its expectations. 
@@
 $\mathit{NAME\_S}$ state.
+
+Whenever the FST transitions to $\mathit{NRC\_S}$ from either
+$\mathit{NAME\_S}$ or $\mathit{NMC\_S}$, the FST reports 
+a narrator detected and adds it to $\mathit{LIST}_{\mathit{narr}}$.
+Whenever the FST transitions back to $\mathit{TEXT\_S}$,
+the FST reports a sanad detected if 
+$|\mathit{LIST}_{\mathit{narr}}| \ge N_{min}$ where 
+$\mathit{LIST}_{\mathit{narr}}$ is the list 
+of detected narrators and $N_{min}$ is a threshold denoting 
+the minimum number of narrators required in a sanad.
+
 We used the values of 3, 5 and 5 for $\theta_{\mathit{nmc}}$, 
 $\lambda_{\mathit{nmc}}$, and 
-$\theta_{\mathit{nrc}}$ respectively to run the experiments 
-we report in Section~\ref{sec:results}.
+$\theta_{\mathit{nrc}}$, respectively to obtain the results in 
+Table~\ref{t:hadithresults}.
 
@@
 \center{
-\resizebox{.9\columnwidth}{!}
+\begin{tabular} {cc}
+\resizebox{.45\columnwidth}{!}
 { \input{figs/hadith.pdftex_t}} 
-\caption{The controller of the hadith application.}
+&
+\resizebox{.55\columnwidth}{!}
+{\includegraphics{figs/DAG_narrator.pdf}} \\
+(a) FST & (b) Extracted narrator graph 
+\end{tabular}
+\caption{Hadith application.}
 \label{f:hadith}
@@
 
+%\begin{figure}[tb!]
+%\center{
+%\resizebox{.6\columnwidth}{!}
+%{ \input{figs/hadith.pdftex_t}}
+%\caption{The FST of the hadith application.}
+%\label{f:hadith}
+%}
+%\end{figure}
+
 %The definition of input labels such as NAME and IBN depends on the 
 %morphological analyzer. 
-%However, our case-based controller approach was able to perform well
+%However, our case-based FST approach was able to perform well
 %under both Sarf and a refined version of Sarf.
 
-\subsection{Optimizations}
-
-In addition to the controller, 
-NaGEMA uses punctuation marks to refine its analysis and 
-increase the accuracy of name detection, narration detection and narrator boundaries. 
-It also learns narrator names from simple patterns that 
-happen within narrator connectors and that are not 
-reported by Sarf as names. 
-NaGEMA learns names of narrators that consist of words missing from 
-the Sarf lexicon.
-NaGEMA learns the narrator names based on their context as they occurred 
-between narrator connectors or contained name connectors.
-
-(explain better and merged information present about them in "Results" section)
-
-\section{The narrator graph}
+%\subsection{Optimizations}
+%
+%In addition to the FST, 
+%NaGEMA uses punctuation marks to refine its analysis and 
+%increase the accuracy of name detection, narration detection and narrator boundaries. 
+%It also learns narrator names from simple patterns that 
+%happen within narrator connectors and that are not 
+%reported by Sarf as names. 
+%NaGEMA learns names of narrators that consist of words missing from 
+%the Sarf lexicon.
+%NaGEMA learns the narrator names based on their context as they occurred 
+%between narrator connectors or contained name connectors. 
+%
+%(explain better and merged information present about them in "Results" section)
+%
+\subsection{Extracting the narrator graph using node merge and split transformations}
 \label{sec:graph}
 
-\begin{figure}[tb]
-\center{
-\resizebox{1.1\columnwidth}{!}
-%{\includegraphics{figs/narrator_chain_output.pdf}}
-{\includegraphics{figs/DAG_narrator.pdf}}
-\caption{Narrator directed acyclic graph 
-    extracted using NaGEMA }
-\label{f:narrators} 
-}
-\end{figure}
-
-The directed acyclic graph (DAG) 
-in Figure~\ref{f:narrators} shows a partial order 
-relation (POR) between narrators.
-We automatically extracted the POR from the three books of 
-hadith~\cite{IbnHanbal,AlKulayni,AlTousi}.
+%\begin{figure}[tb]
+%\center{
+%\resizebox{.8\columnwidth}{!}
+%%{\includegraphics{figs/narrator_sequence_output.pdf}}
+%{\includegraphics{figs/DAG_narrator.pdf}}
+%\caption{Narrator directed acyclic graph 
+%    extracted using NaGEMA }
+%\label{f:narrators} 
+%}
+%\end{figure}
+
+The hadith FST generates sequences of hadith narrators
+such as $\langle n_1,n_2,n_3,n_4,n_5 \rangle$ shown in 
+Figure~\ref{f:exhadith}. 
+Our target is to automate the extraction of 
+narrator graphs from one or more narration books
+similar to the directed acyclic graph (DAG) 
+in Figure~\ref{f:hadith}(b).
+%We automatically extracted the POR from the three books of 
+%hadith~\cite{IbnHanbal,AlKulayni,AlTousi}.
 The nodes in boxes are the \RL{matn} of the hadith, 
 and the other nodes are the narrators.
-Given a set of chains of narrators similar to 
-Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
-that merged equal names in the chains.
-
-\subsection{Motivation}
-
-The hadith POR is instrumental to automate
-localization of narrators in biography documents and
-to segment biography documents in future work.
-Consider a biography that discusses a narrator.
-It mentions his professors and students.
-We plan to segment the biographies with a graph 
-coloring algorithm
-that parses the text looking for names. 
-Once a name is found it locates it in the POR graph
-and colors the POR node corresponding
-to it.
-The narrator predecessors in the POR are supposed to be
-his professors and the narrator successors in the POR 
-are supposed to be his students. 
-Consequently, whenever a name moves away from the 
-colored local cluster in the POR graph, 
-we know that we are within a new biography.
-
-Once the biographies are analyzed, we can annotate
-the POR with qualifiers of the narrators that reflect
-their authenticity. 
-We can also annotate the POR with the locations and 
-the time the narrators lived in. 
-Then we can check for time and location overlap
-to discover inconsistent narrations.
-This motivates our interest to automatically generate 
-the POR graph.
-
-\subsection{Graph merge Algorithm}
-
-(modify according to hash)
-
-\begin{table}[tb]
-%\centering
-%\resizebox{.9\columnwidth}{!}{
-\begin{tabular} {p{6cm}}
-\begin{Verbatim}[fontsize=\relsize{-2},
-frame=topline,framesep=4mm,label=\fbox{Merge Algorithm},
-commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
-Merge(Chain $C$[])
-  Chain $C_c$[]=\{\}
-  foreach Chain $c_1$ in $C$ 
-    $C_c = C_c \cup c_1$
-    foreach Chain $c_2$ in $C-C_c$
-      foreach Narrator $n_1$ in $c_1$
-        start = max( $n_1$.index - radius, 0)
-        end = min($n_1$.index + radius, $n_2$.size)
-        $n_2$ = $c_2$[start]
-        while (true)
-          if (Distance($n_1$, $n_2$) < $\theta$)
-            mergeNodes($n_1$, $n_2$)
-          if ($n_2$ = $c_2$[end])
-            break
-          $n_2$ = $c_2$.next($n_2$)        
-\end{Verbatim}
+
+Our method uses two graph transformations to build the narrator
+graph. 
+It considers the graph with the disconnected sequence components, 
+and performs a sequence of {\em merge} transformation 
+for nodes that
+represent equivalent narrators. 
+A merge transformation takes two equivalent nodes $n_1$ and $n_2$,
+appends the edge list of $n_2$ to that of $n_1$,
+directs the incoming edges of $n_2$ to $n_1$,
+and finally removes $n_2$ from the graph.
+
+Then the method checks for cycles in the graph
+introduced by the merge transformations and breaks the cycles
+using node split transformations. 
+The method also reports the cycles that require splitting nodes 
+with high equivalency values to the user since such cycles may
+indicate inconsistencies or fabricated narrations.
+
+Scholars can use the narrator graph to perform interesting queries 
+that were not possible before such as 
+{\bf (1)} find important narrators who happen to be 
+articulation nodes in the narrator graph,  and
+{\bf (2)} conduct what-if analysis by simulating 
+the effect of removing a narrator on the hadith literature
+as several hadith with the same content 
+may have several different narration sequences.
+
+%\subsection{Graph merge Algorithm}
+%
+%(modify according to hash)
+%
+%\begin{table}[tb]
+%%\centering
+%%\resizebox{.9\columnwidth}{!}{
+%\begin{tabular} {p{6cm}}
+%\begin{Verbatim}[fontsize=\relsize{-2},
+%frame=topline,framesep=4mm,label=\fbox{Merge Algorithm},
+%commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
+%Merge(sequence $C$[])
+%  sequence $C_c$[]=\{\}
+%  foreach sequence $c_1$ in $C$ 
+%    $C_c = C_c \cup c_1$
+%    foreach sequence $c_2$ in $C-C_c$
+%      foreach Narrator $n_1$ in $c_1$
+%        start = max( $n_1$.index - radius, 0)
+%        end = min($n_1$.index + radius, $n_2$.size)
+%        $n_2$ = $c_2$[start]
+%        while (true)
+%          if (Distance($n_1$, $n_2$) < $\theta$)
+%            mergeNodes($n_1$, $n_2$)
+%          if ($n_2$ = $c_2$[end])
+%            break
+%          $n_2$ = $c_2$.next($n_2$)        
+%\end{Verbatim}
+%\end{tabular}
+%%}
+%\label{t:merge}
+%\end{table}
+%
+
+%The \CodeIn{Merge} Algorithm 
+%iterates over the sequences of narrators looking 
+%for equivalent narrators and merges the equivalent
+%narrators.
+%The \CodeIn{radius} parameter refines the search 
+%to only consider narrators who belong to the same 
+%generation. 
+%For example, a narrator relating directly to the prophet
+%is definitely not equivalent to a narrator separated
+%from the prophet by ten or more narrators.
+
+%
+%\begin{table}[tb]
+%%\centering
+%%\resizebox{.9\columnwidth}{!}{
+%\begin{tabular} {p{6cm}}
+%\begin{Verbatim}[fontsize=\relsize{-2},
+%frame=topline,framesep=4mm,label=\fbox{Narrator distance metric },
+%commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
+%Distance(Narrator n1, Narrator n2)
+%  $N_1$ = stems(n1.names)
+%  $N_2$ = stems(n2.names)
+%
+%  $E_n$ = getEqualNames($N_1$, $N_2$)
+%  if ( $E_n$.isEmpty ) 
+%      return $\infty$
+%  if ( $E_n$.size $\ge$ min($N_1$.size,$N_2$.size))
+%    dist -= $\delta_1$
+%  if ($E_n$.equalNamesInOrder())  
+%    dist -= $\delta_2$
+%
+%  $I_1$ = extractIbn( stems(n1.connectors) )
+%  $I_2$ = extractIbn( stems(n2.connectors) )
+%  $C_1$ = stems(n1.connectors - $I_1$)
+%  $C_2$ = stems(n2.connectors - $I_2$)
+%  if ($C_1 \cap$ Places $\not= C_2 \cap$ Places)
+%    dist += $\delta_3$
+%  else
+%    dist -= $\delta_3$
+%
+%  $E_c$ = getEqualConnectors($C_1$, $C_2$)
+%  dist -= $E_c$.size * $\delta_4$ 
+%
+%  if (correctIbnCorrespondence($I_1$, $I_2$))
+%      dist -= $\delta_5$
+%  return (MAX - dist)
+%\end{Verbatim}
+%\end{tabular}
+%%}
+%\end{table}
+
+\subsection{Narrator distance metric}
+%MultiKey Hashing
+%explain what are the keys used for each value and why suitable
+% and say that we use the computed keys also conduct the 
+% biography lookup
+Two narrator names may refer to the same person but may 
+differ in text and in structure. 
+For example, the names \noTrutfRL{عبد الله بن عمر}, 
+\noTrutfRL{ابن عمر}, and \noTrutfRL{ابن عمر رضي الله عنه}
+refer to the same person. 
+Our metric uses the \noTrutfRL{ابن} (son) name connector and 
+its morphological equivalents to order the person names
+that appear within a narrator entity. 
+The metric compares the name hierarchy and if the hierarchy 
+does not match, as in the case of 
+\noTrutfRL{عبيد الله بن عبد الله بن عمر}  and his parent 
+\noTrutfRL{عبد الله بن عمر},
+the metric returns \cci{false}.
+The metric also compares name qualifiers and if two name 
+qualifiers of the same morphological POS tag,  
+such as the place possessive qualifiers 
+\utfRL{العراقي} (the Iraqi) and \utfRL{المصري} (the Egyptian) 
+conflict, the metric returns \cci{false}.
+The metric then compares the morphological stems of the person 
+names that are in order, ignores the ones that are not in order,
+and computes the fraction of the matching names.
+
+Both our metric and the Levenshtein distance metric 
+used in~\cite{Azmi-2010}
+achieved 100\% precision on an arbitrarily selected set that 
+included 208 narrator pairs with 102 equal narrators.
+Our metric achieved 95\% recall versus 72\% recall for the 
+Levenshtein distance metric. 
+Which means that both metrics do not announce names equal if they
+are not, but our metric is better at detecting equal names.
+We believe that our metric is adequate to compare 
+complex Arabic names including contemporary ones in general.
+
+\section{The biography application}
+Biography books include one or more short biography 
+for each hadith narrator. 
+The short biography may contain information 
+about the dates and locations of birth and death, 
+the professors and the students of the narrator,
+and one or more evaluations of the credibility of the narrator.
+
+We use a similar approach to the hadith application
+to locate narrator names within the biography books. 
+A biography text may contain the name of the narrator described
+in the biography as well as the names of other narrators who happen
+to be the professors, students, or the persons who evaluated the 
+narrator.
+Our method uses the narrator graph 
+extracted from several hadith books 
+as described earlier to segment the biographies
+and detect the owner of each biography.
+
+Once our method detects a narrator, it colors the node in  the 
+narrator graph corresponding to it. 
+The professors and students of the narrator
+are the neighbors of the 
+node corresponding to the narrator in the graph. 
+Consequently, whenever a detected narrator entity 
+moves away from the 
+colored local cluster in the narrator graph, 
+we change the color and infer that we moved to a separate
+biography.
+The owner of each biography is the narrator at the center of the
+colored graph.
+
+The naive implementation of this approach is computationally
+expensive as it requires searching for each discovered narrator
+in the narrator graph. 
+We implemented a multi-hash narrator technique to 
+improve narrator localization in the graph and in the biography 
+books where we associate
+each narrator with all narrator names that can be generated 
+from its structure and compute a hash key and a relevance score 
+for that association. 
+We omit the discussion for brevity and we will report on that 
+elsewhere.
+
+Scholars can use the located biographies to annotate
+each narrator with corresponding biographies thus automating
+the task of hadith authentication checks. 
+They can also annotate the narrator graph narrator graph with the locations and 
+the lifetime of the narrators and 
+check for time and location overlap that are necessary conditions
+for the consistency of two 
+neighboring narrators. 
+Scholars can also use the method to check the correspondence
+of different biography and hadith books.
+
+We believe that this is the first work that applies
+{\em Arabic cross-document NLP}, and we believe
+that this approach can be extended to several applications
+such as the cross analysis of (1) investigation versus 
+interrogation reports, and (2) insurance claims versus 
+accident police reports to find inconsistencies.
+
+\section{Genealogical family tree application}
+
+\begin{figure}
+\center{
+%\resizebox{0.9\columnwidth}{!}{
+\tiny
+\begin{tabular}{cp{.1cm}c}
+\parbox[b]{.45\columnwidth}{
+\utfRL{وَهذِهِ مَوَالِيدُ 
+\emph{تَارَحَ}: وَلَدَ \emph{تَارَحُ} \emph{أَبْرَامَ} وَ \emph{نَاحُورَ} 
+وَ \emph{هَارَانَ}. 
+وَوَلَدَ \emph{هَارَانُ} \emph{لُوطًا}. 
+وَمَاتَ \emph{هَارَانُ} قَبْلَ \emph{تَارَحَ} أَبِيهِ فِي أَرْضِ مِيلاَدِهِ فِي 
+أُورِ الْكَلْدَانِيِّينَ. 
+وَاتَّخَذَ \emph{أَبْرَامُ} وَ \emph{نَاحُورُ} لأَنْفُسِهِمَا امْرَأَتَيْنِ: 
+اسْمُ امْرَأَةِ \emph{أَبْرَامَ} \emph{سَارَايُ}، 
+وَاسْمُ امْرَأَةِ \emph{نَاحُورَ} \emph{مِلْكَةُ} بِنْتُ \emph{هَارَانَ}، 
+أَبِي \emph{مِلْكَةَ} وَأَبِي \emph{يِسْكَةَ}}.
+
+And the following are the heirs of Terah. 
+Terah became the father of Abram, Nahor and Haran; and Haran became the father of Lot. Haran before died before the death of 
+his father Terah in his mother land, in Ur of the Chaldeans. 
+Abram and Nahor took wives for themselves: the name of Abram’s wife was Sarai; and the name of Nahor’s wife was Milcah, the daughter of Haran, the father of Milcah and Iscah.
+}
+& &
+\resizebox{.55\columnwidth}{!}
+{\includegraphics{figs/exgenesis.pdf}} \\
+(a) Genesis 11:27-29 & & (b) Extracted family tree
 \end{tabular}
+\normalsize
+\caption{Genealogy example.}
+\label{f:genelists}
+}
 %}
-\label{t:merge}
+\end{figure}
+
+%In many cases, 
+%there is an overlap between the many individual lists available across the Bible. When such an overlap occurs 
+%meticulous readers can detect some seeming contradictions, in many cases due to conflicts or omissions.
+%Detecting those conflicts is a task that has been performed by countless 
+%number of Bible students, theologians and historians.
+
+%The importance of such lists was profound for the old nation of Israel. They
+
+%The genealogical lists established the kingship lineage and determined
+%rights and privileges in the temple. Genealogical lists also determined descent with all its implications in biblical times
+%such as the concept of reverted ownership of sold real estate; every fifty years ownership of sold land reverted to the 
+%original owners and their inheritors. Genealogical lists were preserved due to the possible privilege of being the ancestor of the awaited Messiah.~\cite{catholicEncyclopedia:Online}
+%
+%For many believers, genealogies are used as the main indicators of chronology especially when calculating the age of mankind~\cite{Thomas:92}.
+%Believers used them to defend the historicity of Biblical events and characters against other accounts from other 
+%texts thought of as fables and myths. 
+Biblical genealogical lists 
+trace key biblical figures such as 
+Israelite kings and prophets.
+The example in 
+Figure~\ref{f:genelists}(a)
+from Genesis that traces the
+heirs of \utfRL{تارح} (Terah).
+%They contain person names
+%and kinship relations between the names. 
+Separate genealogical lists that overlap may have consistency 
+%issues~\cite{catholicEncyclopedia:Online,Belote:Online,completeBibleGenealogy:Online}. 
+issues~\cite{completeBibleGenealogy:Online}. 
+For example, Matthew and Luke report two different 
+genealogies of Jesus. 
+Scholars constructed partial genealogical graphs 
+manually from several lists
+and their work is short from being 
+comprehensive~\cite{Belote:Online,SoulLiberty:Online}.
+Creating a comprehensive genealogical graph from all biblical 
+accounts greatly enhances the study of the bible. 
+Such comprehensive graphs can be tagged
+with time durations between a parent and the birth of 
+its children as is common in many biblical accounts.
+
+Our method takes as input Arabic text with a genealogical
+list such as the genealogical list in
+Figure~\ref{f:genelists}(a) 
+and automatically extracts the genealogical family tree 
+where person names are entities and edges represent 
+``spouse'' and ``child'' labels such as 
+Figure~\ref{f:genelists}(b).
+We marked the person names with bars above them, and we 
+filled the spouse nodes with grey. 
+To perform this task, our method first passes the text to
+the morphological analyzer that reports detected morphemes 
+with their corresponding tags.
+
+We augmented the tag set of our morphological analyzer with 
+the following categories relevant to the genealogical 
+application. 
+\begin{itemize}
+\item {\em NAME} a person name that can be either male or female. 509 lexicon items. 
+\item {\em CHILD} a connector with a descent semantics such as \utfRL{ابن} (son). 12 lexicon items. 
+\item {\em CHILDREN} a connector with a plural descent semantics such as \utfRL{ابناء} (children). 16 lexicon items. 
+\item {\em PARENT} a connector with parenthood semantics such as \utfRL{أباه} (his father). 7 lexicon items. 
+\item {\em SPOUSE} a connector with marital partnership semantics such as \utfRL{سريّته} (his concubine). 11 lexicon items. 
+\item {\em SPOUSES} a connector with plural partnership semantics such as \utfRL{سراري} (his concubines). 7 lexicon items. 
+\item {\em SIBLING} a connector with sisterhood semantics such as \utfRL{أخيه} (his brother). 14 lexicon items.
+\item {\em SIBLINGS} a connector with plural sisterhood semantics such as \utfRL{اخوته} (his siblings). 18 lexicon items. 
+\end{itemize}
+Note that (1) some lexicon items might be categorized into 
+more than one of the above categories, and 
+(2) the method infers the gender from POS and gloss tags 
+of the person name as well as from those of the 
+other categories. 
+In case of conflict, the method gives precedence to the 
+gender inferred from the person names. 
+Note also that the phrases
+\utfRL{اسحاق بن يعقوب} (Izhac - son of - Jacob)
+and 
+\utfRL{اسحاق ابنه يعقوب} (Izhac - his son - Jacob)
+contain the same named entities, and are identical 
+as far as stemming is concerned. 
+However, the analysis 
+of the suffix morphemes of \utfRL{ابنه} results in inferring
+a parent category in the second phrase.
+
+The method considers the detected categories and passes 
+them to a genealogy FST.
+The genealogy FST distinguishes detected names as follows. 
+\begin{itemize}
+\item {\em NAME} denotes a newly detected name that is not 
+a node in the current tree. 
+\item {\em LEAF} denotes a name that is a leaf in the current tree.
+\item {\em NODE} denotes a name that is an internal node in the current tree.
+\end{itemize}
+The rest of the categories are presented to the FST as is and
+are all denoted with {\em EDGE}.
+
+\begin{table}[bt]
+\centering
+\caption{State transitions and actions for genealogy application.}
+\begin{tabular}{cp{.1cm}p{1.1in}cp{2.4in}}
+State & & Input & Transition & Actions and output \\ \hline
+\multirow{3}{*}{\em Text} 
+&  & $n=\mbox{NAME}$                   & {\em Parent} & addNode(n), $n_i = n, w = 0, e_i=\mbox{CHILD}$ \\ \cline{3-5}
+&  & $n=\mbox{LEAF} \vee \mbox{NODE}$ & {\em Parent} & $n_i = n, w = 0, e_i=\mbox{CHILD}$ \\ \cline{3-5}
+&  & $e=\mbox{EDGE}$                   & {\em Parent} & $n_i = \empty, w = 0, e_i=e$ \\ \hline
+\multirow{5}{*}{\em Parent} 
+&  & $n=\mbox{NAME}$                   & {\em Parent} & addNode($n$), addEdge($n_i$,$n$,$e_i$), $n_i = n, w = 0$ \\ \cline{3-5}
+&  & $n=\mbox{LEAF}$                   & {\em Child}  & $n_i = n$ \\ \cline{3-5}
+&  & $e=\mbox{EDGE}$                   & {\em Parent} & $w = 0, e_i=e$ \\ \cline{3-5}
+&  & otherwise $\wedge w \le \theta$   & {\em Parent} & $w=w+1$, add unclassified edges and nodes \\ \cline{3-5}
+&  & otherwise $\wedge w > \theta$     & {\em Text}   & local tree detected \\ \hline
+\multirow{5}{*}{\em Child} 
+&  & $n=\mbox{NAME}$                   & {\em Child}  & addNode($n$), addEdge($n_i$,$n$,$e_i$) \\ \cline{3-5}
+&  & $n=\mbox{LEAF}$                   & {\em Child}  & $n_i = n$ \\ \cline{3-5}
+&  & $e=\mbox{EDGE}$                   & {\em Parent} & $w = 0, e_i=e$ \\ \cline{3-5}
+&  & $\mbox{ENDLINE}$                  & {\em Parent} & $w = 0$ \\ \cline{3-5}
+&  & otherwise                         & {\em Child}  & add unclassified edges and nodes \\ \hline
+\end{tabular}
+\label{t:genefst}
 \end{table}
 
-The \CodeIn{Merge} Algorithm 
-iterates over the chains of narrators looking 
-for equivalent narrators and merges the equivalent
-narrators.
-The \CodeIn{radius} parameter refines the search 
-to only consider narrators who belong to the same 
-generation. 
-For example, a narrator relating directly to the prophet
-is definitely not equivalent to a narrator separated
-from the prophet by ten or more narrators.
-
-\begin{table}[tb]
-%\centering
-%\resizebox{.9\columnwidth}{!}{
-\begin{tabular} {p{6cm}}
-\begin{Verbatim}[fontsize=\relsize{-2},
-frame=topline,framesep=4mm,label=\fbox{Narrator distance metric },
-commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
-Distance(Narrator n1, Narrator n2)
-  $N_1$ = stems(n1.names)
-  $N_2$ = stems(n2.names)
-
-  $E_n$ = getEqualNames($N_1$, $N_2$)
-  if ( $E_n$.isEmpty ) 
-      return $\infty$
-  if ( $E_n$.size $\ge$ min($N_1$.size,$N_2$.size))
-    dist -= $\delta_1$
-  if ($E_n$.equalNamesInOrder())  
-    dist -= $\delta_2$
-
-  $I_1$ = extractIbn( stems(n1.connectors) )
-  $I_2$ = extractIbn( stems(n2.connectors) )
-  $C_1$ = stems(n1.connectors - $I_1$)
-  $C_2$ = stems(n2.connectors - $I_2$)
-  if ($C_1 \cap$ Places $\not= C_2 \cap$ Places)
-    dist += $\delta_3$
-  else
-    dist -= $\delta_3$
-
-  $E_c$ = getEqualConnectors($C_1$, $C_2$)
-  dist -= $E_c$.size * $\delta_4$
-
-  if (correctIbnCorrespondence($I_1$, $I_2$))
-      dist -= $\delta_5$
-  return (MAX - dist)
-\end{Verbatim}
-\end{tabular}
-%}
-\end{table}
-
-\subsection{Narrator morphological distance metric}
-
-The \CodeIn{Distance} Algorithm 
-takes as input two narrators and returns 
-their morphological and order distance
-from each other. 
-The lower the distance is, the more
-equivalent the narrators are.
-
-Each narrator has several names and several name and 
-narrator connectors. 
-\CodeIn{Distance} computes the names of the two narrators
-and stores them in $N_1$ and $N_2$ 
-and checks whether they are morphologically equivalent. 
-If they are not then the distance is set to $\infty$. 
-Otherwise the distance is decremented by $\delta_1$. 
-Then \CodeIn{Distance} checks whether the morphologically 
-equivalent names appear in the same order and 
-decrements the distance if so.
-
-\CodeIn{Distance}  pays attention to the $\mathit{NISBA}$ 
-words that relate to places and checks on that to make sure
-that the two narrators considered equivalent are actually 
-from the same place. 
-It also checks whether one place such as Baghdad 
-is geographically 
-contained in the other such as Iraq 
-(we skip that in the Algorithm for clarity). 
-\CodeIn{Distance} also checks for the equivalency of the
-name connectors and the correspondence of the \RL{ibn}
-connectors.
-
-The distance metric of NaGEMA performs better than       
-existing metrics~\cite{Azmi-2010} as it can detect that  
-\RL{`bd al-lah bn `mr} and \RL{ibn `mr} are the same person
-while other distance metrics based on simple
-string similarity techniques can not.
-
-\subsection{Speedup Optimizations for Large Datasets}
-
-\subsubsection{MultiKey Hashing}
-
-(explain what are the keys used for each value and why suitable)
+Table~\ref{t:genefst} described the states, input, transitions and the actions and outputs of the genealogy FST. 
+The FST keeps track of the last detected {\em NODE} and {\em EDGE} in the state elements $n_i$ and $e_i$.
+It also keeps track of the number of the words in sequence that do not match a category in $w$. 
+Once $w$ passes the threshold $\theta$, and if the FST has any nodes and edges detected, the tree is reported as output. 
+The FST has three abstract states where {\em Text} 
+is the initial state and denotes that the FST is outside the context of a genealogical list.
+State {\em Parent} denotes that the FST is in the middle of a genealogical list and has met some names. 
+State {\em Child} denotes that the FST is in the middle of a genealogical list and in the context of names of children of a discovered 
+name. 
+When a {\em NAME} is met, the FST adds a node and possibly and edge to the tree. 
+When a {\em NODE} or a {\em LEAF} is met, the FST adds
+an edge to the tree. 
+When words that do not belong to the specified categories are met between {\em EDGE} words or 
+between {\em NAME, LEAF}, or {\em NODE} words the FST adds them as labels to unclassified edges. 
+The FST uses these unclassified names and edges to learn edge 
+semantics that can help in a subsequent run to resolve 
+ambiguities; connected nodes with un-classified edges.
+
+The FST produces a set of genealogical trees. The method takes those trees and merges equivalent 
+nodes to build a global genealogical tree. 
+The method merges nodes with the same names and whose merger does not cause a cycle in the genealogical tree with only the 
+``child'' and ``spouse'' edges considered.
 
 \section{Related work }
-\label{sec:related}
+\label{s:related}
 
@@
 
-The closest work to NaGEMA is 
-iTree~\cite{Azmi-2010,iTree}, a 
-tool that uses a grammar based approach to solve 
+The  iTree~\cite{Azmi-2010,iTree} tool
+uses a context free grammar (CFG) approach to solve 
 the narrator extraction problem. 
-The iTree tool preprocesses the text to remove 
-punctuation and diacritics and to normalize white
-spaces. 
-Then iTree uses a similarity-based approach
+It preprocesses the text to remove punctuation and diacritics 
+and to normalize white spaces, and uses a similarity-based approach
 to memory based learning~\cite{Azmi-2010} 
-in order to perform 
-shallow parsing of the preprocessed text. 
-The iTree tool uses a parser based on a context
-free grammar that enumerates several ways to express
-the name of the prophet, several possible words that 
-might precede the name of the prophet, several possible
-words that might separate two narrators, and several
-words that might separate parts of the name of 
-the narrators.
+in order to perform shallow parsing of the preprocessed text. 
+The iTree CFG enumerates several stop phrases that surround
+names narrators.
 The iTree tool fails when a morphological variation of 
-one of the 
-listed words and not an exact form of it
-happens in the text. 
-For example, Figure~\ref{f:mohamada} extracted 
-using iTree fails to detect \RL{m_hmdA} as a name of the
-prophet and does not stop at it when computing the chain 
-of the hadith
-\transfalse
-\setcode{utf8}
-%\begin{arabtext}
-\RL{
-حدثنا وكيع، حدثني سعيد بن السائب، عن داود بن أبي عاصم الثقفي، قال
-سألت ابن عمر عن الصلاة، بمنى فقال هل سمعت {\bf محمدا}، صلى الله عليه وسلم قلت
-نعم وآمنت فاهتديت به قال فإنه كان يصلي بمنى ركعتين.
-%\end{arabtext}
-}
-\setcode{standard}
-\transtrue
-\vocalize
-
-%In their grammar they consider a name to be any sequence 
-%of 
-%legal Arabic characters.
-
-NaGEMA does not need a prepossessing step since 
-it uses morphological analysis to process the text. 
-While iTree bases its analysis on the patterns 
-of words that surround the narrator names, 
-NaGEMA bases its analysis on the fact that narrator
-names are likely to happen in localities with a high
-number of proper names. 
-NaGEMA uses the narrator connectors as helpers and
-uses morphological analysis to compare the stems of
-those words and thus NaGEMA avoids the problem of 
-``parser noise words'' that iTree faces.
-
-The iTree tool extracts successfully 86.7 percent of
-the narration chains of 90 selected traditions with 
+one of the stop phrases occurs.
+%For example, Figure~\ref{f:mohamada} extracted 
+%using iTree fails to detect \RL{m_hmdA} as a name of the
+%prophet and does not stop at it when computing the sequence 
+%of the hadith
+%\transfalse
+%\setcode{utf8}
+%%\begin{arabtext}
+%\RL{
+%حدثنا وكيع، حدثني سعيد بن السائب، عن داود بن أبي عاصم الثقفي، قال
+%سألت ابن عمر عن الصلاة، بمنى فقال هل سمعت {\bf محمدا}، صلى الله عليه وسلم قلت
+%نعم وآمنت فاهتديت به قال فإنه كان يصلي بمنى ركعتين.
+%%\end{arabtext}
+%}
+%\setcode{standard}
+%\transtrue
+%\vocalize
+We differ in that we do not preprocess the text,
+we use morphological analysis,
+and we do not base our method
+on stop phrases that surround narrator names. 
+Rather, we assume that narrator names are likely to happen in 
+localities with a high number of person names. 
+We use narrator connectors and their
+morphological equivalents and thus we do not suffer from 
+the ``parser noise words'' problem of iTree. 
+The iTree tool extracts successfully 86.7\% of
+the narration sequences of 90 selected traditions with 
 34 simple cases and 56 hard cases. 
-We ran NaGEMA against the five hadith that ship with
-the tool and the two hadith that are listed in the iTree
-paper and NaGEMA was able to extract all narration
-chains with no mistakes even on the examples where iTree
+We ran our method against the five hadith that ship with
+iTree and the two hadith from the iTree
+paper and we were able to extract all narration
+sequences even on the examples where iTree
 reports a ``parser noise'' problem.
 
-As we do not have access to the selected
-90 traditions iTree reports on, we claim
-that NaGEMA outperforms iTree as it works on
-the full text of the presented hadith books
-and reports a higher success rate.
-
-Our approach to detect and relate names to verbs
-is similar to a local grammar based 
-approach~\cite{Traboulsi:09} that detects names by
-associating them with verbs 
-such as \RL{qaal} ``said'' and \RL{'a_hbar} ``told''. 
-The local grammars in~\cite{Traboulsi:09} are also
-built into transducers that generate the names and
-the learning phase benefits from morphological analysis. 
-We differ in that we do not use a local grammar, rather
-our tolerance thresholds allow a flexible pattern that is
-not easy to capture in a typical regular expression.
-
-Another attempt to Arabic named entity recognition
-uses an n-gram approach and maximum 
-entropy~\cite{ANERSys:07}. 
-The authors use their own corpora and gazetteers to boost 
-and train the system and ignore POS tags. 
-They achieve major improvements when they use lexical and
+%As we do not have access to the selected
+%90 traditions iTree reports on, we claim
+%that NaGEMA outperforms iTree as it works on
+%the full text of the presented hadith books
+%and reports a higher success rate.
+
+Our NE detection method differs from 
+local grammar based approaches~\cite{ZAGHOUANI10,Traboulsi:09}
+in that they enumerate several local grammar rules
+to capture the desired named entities, 
+and use techniques to automatically compute finite state 
+machines that detect the structures expressed in the 
+local grammars.
+The work in~\cite{ZAGHOUANI10} uses morphological stemming
+for Arabic with local grammars developed for Latin languages 
+and suited to fit Arabic and reports 87\% precision and 
+66\% recall for person names. 
+Similarly, the work in~\cite{jumaily2011,TAGARAB98} 
+uses morphological stemming and pattern matching. 
+We differ in that we are not limited to the use of stems 
+and we use the augmented
+set of morphological tags for all prefixes and suffixes as well.
+This allows us to lift the restriction of the stop words 
+used in local grammars and in~\cite{Abuleil04}.
+We also differ in that we directly build the
+FST to detect the named entities as well 
+as the relational entities, 
+we do not enumerate all the structures that express our targeted
+entities, and we are not restricted to the expressive power of 
+a local grammar language where it is often not intuitive 
+(and in cases not possible) to express thresholds using regular 
+expressions. 
+In the future, we plan to develop a local
+grammar language that has that expressive power.
+
+ANERSys boosts its capabilities to detect named entities by using 
+task specific corpora and gazetteers and ignores 
+POS tags~\cite{ANERSys:07,BenajibaDR08}. 
+We are similar to ANERSys in that we augment the lexicon of our
+morphological analyzer with application relevant lexicon items
+and tags. 
+We differ in that they use statistical analysis models such as
+n-gram and maximum entropy measures.
+In other work, they
+achieve major improvements when they use lexical and
 morphosyntactic features and a parallel Arabic and English 
-corpora to bootstrap noisy features~\cite{Benajiba:2010}.
+corpora to bootstrap noisy features~\cite{BenajibaZDR10}.
+
+We differ from rule-based systems with local 
+grammars~\cite{ShaalanR09} in that we our system 
+uses FSTs instead of rules, and in that it can learn named 
+entities and infer semantics
+for entities that are equivalent within one or two graph 
+transformations to predefined graph edge labels.
 
@@
 %Narrators  & 2,623 & & 5,767& & 1,755 \\ 
-%%Narrators/Chain  & 4.84 & & 4.76 & &4.05 \\
-%Chains  & 542 &  & 1,211& & 433 \\ 
+%%Narrators/sequence  & 4.84 & & 4.76 & &4.05 \\
+%sequences  & 542 &  & 1,211& & 433 \\ 
 %Ignored names  & 6,400 &  & 3,348 & & 642 \\ \hline
 %Segmentation accuracy  & 96\%& & 96\%& & 92\%\\ 
-%Chain accuracy & 99\%&  & 99\%& & 97\% \\ 
+%sequence accuracy & 99\%&  & 99\%& & 97\% \\ 
 %%Narrator accuracy  & & & & &  & & & \\ 
@@
 %
+%
+%\begin{table*}[bt]                                     
+%\centering                                            
+%\caption{Hadith results compared with iTree. }
+%\resizebox{.95\columnwidth}{!}{
+%\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+%%\multicolumn{13}{c}{NaGEMA with all refinements} \\ \cline{3-13}
+%& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+%\multirow{3}{*}{NaGEMA }&musnad & & 0.999 & 0.999 & 0.999 & & 0.993 & 0.990 & 0.991 & & 0.999 & 0.998 & 0.999 \\
+%&kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.988 & 0.990 \\
+%&istibsar & & 0.999 & 0.973 & 0.986 & & 0.980 & 0.950 & 0.965 & & 0.996 & 0.998 & 0.997 \\ \cline{2-14}
+%& & & 0.999 & 0.964 & 0.982 & & 0.970 & 0.970 & 0.970 & & 0.996 & 0.995 & 0.995 \\ \hline \hline
+%%& & &  &  &  & &  &  &  & &  &  & \\
+%%&\multicolumn{13}{c}{iTree} \\ \cline{3-13}
+%& & & \multicolumn{3}{c}{accept as hadith} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+%\multirow{3}{*}{iTree} &musnad & & 0.999 & NA & 0.999 & & 0.997 & 0.977 & 0.987 & & 0.998 & 0.999 & 0.999  \\
+%&kafi & & 0.043 & NA & 0.083 & & 0.999 & 0.313 & 0.476 & & 0.999 & 0.999 & 0.999 \\
+%&istibsar & & 0.028 & NA & 0.054 & & 0.875 & 0.583 & 0.700 & & 0.999 & 0.999 & 0.999 \\ \cline{2-14}
+%& & & 0.357 & NA & 0.379 & & 0.957 & 0.624 & 0.721 & & 0.999 & 0.999 & 0.999 \\ 
+% \end{tabular}                 
+%}                             
+%\normalsize                   
+%\label{t:hadithresults} 
+%\end{table*}
+
+%\input{cicling/tables.tex}
+
+We evaluated our method against three books of 
+hadith~\cite{IbnHanbal,AlTousi,AlKulayni},
+the Genesis~\cite{},
+and one book of biographies with several volumes~\cite{khoei}. 
+We define our metrics, present our evaluation reference with interannotation agreement results, 
+and then present the results for the hadith
+biography, and genealogical list applications.
+
+{\bf Metrics.}~~
+Segmentation recall refers to the ratio of the narrations correctly detected against the
+total number of narrations present. 
+Segmentation precision refers to the fraction of correctly detected narration sequences compared to all the extracted
+narration sequences. 
+Similarly, sanad boundary recall measures whether we detected all narrators
+in the sanad; while sanad boundary precision measures whether we did so without introducing false positives. 
+The same concept applies to the narrator boundary where we count the number of words constituting 
+a valid narrator.
+For the genealogy application, segmentation refers to the ratio of the genealogical lists correctly detected against the total 
+number of lists present.
+Segmentation precision refers to the fraction of correctly detected genealogical lists compared to all the extracted lists. 
+Recall for genealogical boundary detection measures whether we detected all person names in the lists; 
+precision measures whether we did so without introducing false positives.
+
+{\bf Interannotation.}~~
+We developed a GUI tool and asked two volunteers to tag the sanad boundaries and the narrator names therein for 10\% of the hadith text, 
+and to tag the lists, the person names therein, and to edit the tree structure within the genealogy list for the Genesis.
+%We computed the interannotation agreement metrics presented in Table~\ref{t:interannotator}.
+The annotators agreed with above 98\% precision and recall on hadith segmentation, sanad boundaries and name boundaries.
+The annotators agreed with 99\% and 97\% recall on genealogy segmentation and list boundaries.
+They also scored 95\% precision on list boundaries.
+For the genealogy segmentation, the tags of the first annotator were a subset of the tags of the second and therefore they scored  
+39\% precision as we measured the second against the first.
+The Genesis contains texts that are intended to describe genealogy, 
+and also contains person names with mentions of family relations that are not intended to be genealogical lists.
+The second annotator aggressively selected both types of genealogical lists, while the first annotator selected the first type and was 
+very conservative in selecting the second type. 
+Some genealogy lists may contain more than one family tree and our GUI was limited to allow the annotators to select one of the trees.
+This resulted in a 65\% precision score for genealogy list boundaries agreement. 
+The two annotators agreed on a reference tag set that we used to compute the following accuracy results.
+
+%
+%\begin{table*}[bt]                                     
+%\centering                                            
+%\caption{Interannotator Agreement }
+%\resizebox{1\columnwidth}{!}{
+%\begin{tabular}{p{1.3cm}p{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+%%\multirow{2}{*}{Test Case} 
+%& &  \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad/genealogy boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+% & &  recall & precision & average & & recall & precision & average & & recall & precision & average \\ \cline{3-5} \cline{7-9} \cline{11-13}
+%Hadith &  & 0.976 & 0.999 & 0.988 & & 0.991 & 0.971 & 0.981 & & 0.999 & 0.997 & 0.998 \\ 
+%Genealogy &  &  0.999 & 0.390 & 0.695 & & 0.969 & 0.954 & 0.812 & & NA & NA & NA\\
+% \end{tabular}                 
+%}                             
+%\normalsize                   
+%\label{t:interannotator} 
+%\end{table*}
+
+{\bf Hadith application results.}~~
+The results in Table~\ref{t:hadithresults} compare the results of our method against iTree~\cite{iTree} for the hadith application.
+We augmented iTree with the list of names used by our method to conduct a fair assessment.
+The iTree tool does not tackle the problem of segmentation as it
+takes as input an isolated hadith. 
+%In many cases iTree fails to recognize the structure of
+%the hadith and does not extract any useful information from the narration. 
+%This happens for example when a slight variation of the iTree grammar rules 
+%happens first in the hadith.
+In Table~\ref{t:hadithresults}, instead of the segmentation columns 
+we report the percentage of times iTree succeeded to recognize the valid hadith we pass.
+This measure is equivalent to recall; nonetheless, precision in this context is not applicable.
+
+On average, our method acheived a 97\% segmentation recall rate compared to 35.7\% for iTree.
+The difference depended highly on the book under consideration. 
+For Musnad, both methods performed well;
+however iTree reported less than 5\% recall for the Kafi and Istibsar books due
+to the less structured nature of the narrations in both books.
+%For example, iTree's grammar is defined such that a hadith starts by 
+%an Ikhbar word (\noVocRL{AdaT al-A_hbAar}) such as
+%\noVocRL{qAl} ``said'', \noVocRL{.hadda_t} ``narrated'', etc...; 
+%however, the Kafi and Istibsar books followed a flexible style to express the narration 
+For example, in a significant number of narrations a narrator name preceded narration words 
+(\noVocRL{qAl} ``said'', \noVocRL{.hadda_t} ``narrated'', etc...) contrary to what iTree expects.
+Note that it is not trivial (if even possible) to take care of this variation using local grammars 
+and CFGs as this introduces a high level of ambiguity.
+For the hadith that iTree succeeded to recognize, iTree had a lower sanad boundary precision (62.4\% on the average).
+This is mainly due to the stop phrases used to detect the termination of sanad boundaries.
+Sanad termination is often implicit in Kafi and Istibsar causing iTree to extend the boundary of the sanad until the next stop phrase.
+Our method detects the end of the sanad when it meets a sequence of words of length $> \theta$ with no names.
+The narrator boundary accuracy showed similar results between iTree and 
+our method since we augmented iTree with the names from the lexicon of Sarf.
 
@@
 \centering                                            
-\caption{Results comparing between NaGEMA and iTree.}
-\resizebox{2.1\columnwidth}{!}{
-\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+\caption{Hadith results compared with iTree.}
+\resizebox{1\columnwidth}{!}{
+\begin{tabular}{lp{.1cm}cp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
 %\multicolumn{13}{c}{NaGEMA with all refinements} \\ \cline{3-13}
-& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
-& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{NaGEMA }&musnad & & 0.999 & 0.999 & 0.999 & & 0.993 & 0.990 & 0.991 & & 0.999 & 0.998 & 0.999 \\
-&kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.988 & 0.990 \\
-&istibsar & & 0.999 & 0.973 & 0.986 & & 0.980 & 0.950 & 0.965 & & 0.996 & 0.998 & 0.997 \\ \cline{2-14}
-& & & 0.999 & 0.964 & 0.982 & & 0.970 & 0.970 & 0.970 & & 0.996 & 0.995 & 0.995 \\ \hline \hline
+{\bf our} & & hadith & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{narrator name boundary} \\
+{\bf method} & & count & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ 
+    \cline{1-3} \cline{5-7} \cline{9-11} \cline{13-15}
+musnad & & 212 & & 0.991 & 0.976 & 0.983 & & 0.969 & 0.967 & 0.968 & & 0.999 & 0.996 & 0.998 \\
+kafi & & 199 & & 0.970 & 0.909 & 0.938 & & 0.934 & 0.954 & 0.944 & & 0.999 & 0.998 & 0.998 \\
+istibsar  & & 189 & & 0.968 & 0.973 & 0.971 & & 0.946 & 0.945 & 0.945 & & 0.999 & 0.998 & 0.999 \\ \hline 
+ & & 600 & & 0.976 & 0.953 & 0.964 & & 0.950 & 0.956 & 0.953 & & 0.999 & 0.997 & 0.998 \\ 
+% \hline \\ \hline
+ \\
 %& & &  &  &  & &  &  &  & &  &  & \\
 %&\multicolumn{13}{c}{iTree} \\ \cline{3-13}
-& & & \multicolumn{3}{c}{accept as hadith} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
-& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{iTree} &musnad & & 0.999 & NA & 0.999 & & 0.997 & 0.977 & 0.987 & & 0.998 & 0.999 & 0.999  \\
-&kafi & & 0.043 & NA & 0.083 & & 0.999 & 0.313 & 0.476 & & 0.999 & 0.999 & 0.999 \\
-&istibsar & & 0.028 & NA & 0.054 & & 0.875 & 0.583 & 0.700 & & 0.999 & 0.999 & 0.999 \\ \cline{2-14}
-& & & 0.357 & NA & 0.379 & & 0.957 & 0.624 & 0.721 & & 0.999 & 0.999 & 0.999 \\ 
+{\bf iTree} & & hadith& & \multicolumn{3}{c}{accept as hadith}& & \multicolumn{3}{c}{sanad boundary} & & \multicolumn{3}{c}{narrator name boundary} \\
+ & & count & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ 
+    \cline{1-3} \cline{5-7} \cline{9-11} \cline{13-15}
+musnad && 62 && 0.999 & NA & NA & & 0.997 & 0.977 & 0.987 & & 0.998 & 0.999 & 0.999  \\
+kafi && 23 && 0.043 & NA & NA & & 0.999 & 0.313 & 0.476 & & 0.999 & 0.999 & 0.999 \\
+istibsar && 36 && 0.028 & NA & NA & & 0.875 & 0.583 & 0.700 & & 0.999 & 0.999 & 0.999 \\ \hline
+ & & 124 && 0.357 & NA & NA & & 0.957 & 0.624 & 0.721 & & 0.999 & 0.999 & 0.999 \\ 
  \end{tabular}                 
@@
 
-To evaluate our approach, we ran NaGEMA against three books of 
-hadith~\cite{IbnHanbal,AlTousi,AlKulayni} and compared the
-results to iTree. 
-The results are detailed in 
-Tables~\ref{t:hadithresults} and \ref{t:nopunctlearnresults}. 
-The numbers we present in the tables are the result of a 
-manual verification of the correctness of %the outputs of 
-10\% of the narrations in the books under consideration.
-
-(discuss annotation and verification process and mention smg about inter-annotator agreement)
-
-We used recall and precision as our evaluation metrics.
-We report results on three levels of accuracy: segmentation, sanad boundary and
-narrator boundary. 
-Segmentation accuracy refers to the success of NaGEMA in
-detecting in the plain text of a hadith book the locations of the different
-narrations therein.
-The sanad boundary accuracy measures the
-correctness of individual narrators inside the sanad. 
-Finally, the narrator boundary accuracy measures
-the correctness of narrator full names.
-
-Precisely defined, segmentation recall refers to the ratio of the narrations correctly detected against the
-total number of narrations present. 
-Segmentation precision refers to the fraction of correctly detected narration chains compared to all the extracted
-narration chains. 
-Similarly, sanad boundary recall measures whether we detected all narrators
-in the sanad; while sanad boundary precision measures whether we did so without introducing false positives. 
-The same concept applies to the narrator boundary where we count the number of words constituting 
-a valid narrator.
-
-\subsection{NaGEMA versus iTree}
-
-iTree does not tackle the problem of hadith segmentation. 
-In contrast to NaGEMA, it does not work on a book level, 
-instead it takes as input an isolated hadith. 
-It assumes the hadith book is already segmented.
-
-Consequently, for iTree to extract the narrators from the sanad of a narration, 
-the user has to extract and pass the hadith manually.
-Still in many cases iTree fails to recognize the structure of
-the hadith and does not extract any useful information from the narration. 
-This happens for example when a slight variation of the iTree grammar rules 
-happens first in the hadith.
-In Table~\ref{t:hadithresults}, instead of the segmentation columns 
-we report the percentage of times iTree succeeded to recognize the text passed
-as a hadith.
-This measure is equivalent to the concept of recall; nonetheless, precision
-in this context is not applicable since by default the input is required 
-to represent correctly extracted narrations.
-
-\subsubsection{On Test-suite Of iTree}
-
-(report on the 90 hadith used in paper of iTree)
-
-\subsubsection{On More Representative Test-suite}
-
-In order to evaluate the approaches used by iTree and NaGEMA without bias,
-we augmented the list of names used by 
-iTree with those available in the lexicon of Sarf.
-%'s lexicon on top of which NaGEMA operates.
-
-First, we compare the recall of the segmentation 
-in NaGEMA to that of hadith acceptance rates in iTree. 
-Although the task of segmentation is more complex than that of
-hadith recognition/acceptance, we still notice that iTree has on the average 35.7\% recall rate compared to 99.9\% for 
-NaGEMA. 
-However, we notice that the difference in these numbers depended highly on the books under consideration. 
-For instance,
-for the book of musnad, both approaches correctly detected  practically all narrations under consideration; 
-however iTree suffers a lot in 
-the books of both Kafi and Istibsar, resulting in a recall of less than 5\%. 
-This is due to the less structured nature of the narrations of those books.
-For example, iTree's grammar is defined such that a hadith starts by 
-an Ikhbar word (\noVocRL{AdaT al-A_hbAar}) such as
-\noVocRL{qAl} ``said'', \noVocRL{.hadda_t} ``narrated'', etc...; 
-however, the Kafi and Istibsar books followed a flexible style to express the narration 
-and most narrations started directly by a narrator name without a preceding Ikhbar word. 
-This is a strong evidence of the superiority and tolerance of NaGEMA 
-and its threshold based approach
-compared to the strict grammar approach used in iTree. 
-Clearly, NaGEMA does not fail on such cases, since what it detects as a hadith is based 
-on a concentration of names.
-Similarly, iTree had a lower sanad boundary precision (62.4\% on the average).
-This is mainly due to the grammar that ends a sanad only when a predefined prophet phrase is met.
-Prophet phrases are phrases that point to the prophet such as 
-\noVocRL{rasowl al-lah} and 
-\noVocRL{.sal_A `alayh al-lah wasalam}.
-
-However, especially in the books of Kafi and Istibsar, the prophet is not explicitly mentioned but 
-implicitly understood. In those cases, iTree will continue reporting narrators whenever it reaches 
-a word that it predefines as a name even if it appears
-in the matn of the hadith. 
-On the other hand, NaGEMA considers that a sanad ends whenever names become less frequent and measures 
-that with a threshold even if it does not reach a prophet
-indicative phrase.
-
-Narrator boundary accuracy showed similar results between iTree and 
-NaGEMA since we augmented iTree with the names from the lexicon of Sarf.
-
-%\subsection{On Unorganized hadith books}
-%
-%(report results on such hadith books)
-
-\subsection{Effect of punctuation and learning}
+{\bf Narrator graph.}~~
+The results in Table~\ref{t:PORnLearning}(a) report the precision and recall metrics for generating 
+the full narrator graph via merging the detected hadith sequences.
+For each narrator node, we computed the number of nodes it correctly merges with against the number of nodes it should merge with and 
+averaged that across all nodes to compute the recall metric.
+The precision metric denotes the average of the
+ratio of correctly merged nodes against the number of merged 
+nodes for each narrator node.
+
+The ``detected'' row includes only the narrator nodes detected by our method and indicates the accuracy of the graph building routine 
+isolated from the rest of the method.
+The ``all'' row includes all narrators and sanads in the text.
+The recall rate was 63\% mainly due to the conservative cycle breaking transformation that splits merged narrator nodes even if 
+they were equivalent. 
+We expected higher precision values than 82\% and 77\% and upon examining the results, we discovered two main issues. 
+The first issue is in the reference graph since the manual annotator decided to keep nodes split whenever he was faced with
+confusing cases such as the narrator nodes \utfRL{سعيد بن السائب}, \noTrutfRL{سعيد}, and \utfRL{سعيد بن جبير}. 
+The text included several of these examples such as \utfRL{سليمان التيمي}, \utfRL{سليمان بن رزين}, and \utfRL{ سليمان بن موسى}.
+The automated merging will make a decision to merge two of the three names in each case 
+and uses other helper information such as the rank of the nodes in the graph to make a decision.
+
+The second issue has to do with the degree of equality checks that the merging algorithm performs when deciding to merge narrator nodes.
+Consider three narrator nodes merged into one node $n_{merged}$.
+Now consider merging a new narrator node $n_*$ with $n_{merged}$. 
+We currently compare $n_*$ with only one of the constituents of 
+$n_{merged}$ and make the simplifying assumption that our equality
+metric is transitive. 
+Comparing to more narrators in $n_{merged}$ helps improve precision.
+
+%In the POR, we have 2 ways of computing the accuracy, just as it is (all narrators), second (accounting for missing narrators in the trees in the first place which will result in having the errors of parsing accumulate over the merging of nodes (so in the detected narrators we calculate the results without expecting to find the undetected nodes (i.e.
+in recall we dont count nodes that are missing due to not having them in the first place)
+
+ %           In all cases, there will still be accumulated errors in the POR due to errors in parsing, for example (boundaries of detected narrators will hurt accuracy) also if we have missing narrators or extra narrators in our POR we might get extra cycles and thus have to break nodes that should not be breaked.
+so in all cases, POR results have accumulated errors.
+But still the detected narrators measure is fairer.
 
@@
 \centering                                            
-\caption{Results without punctuation and without learning. }
-\resizebox{2.1\columnwidth}{!}{
-\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
-& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
-& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{no punctuation} 
-& musnad  & & 0.999 & 0.999 & 0.999 & & 0.982 & 0.955 & 0.968 & & 0.999 & 0.998 & 0.999 \\ 
-& kafi & & 0.999 & 0.793 & 0.885 & & 0.944 & 0.967 & 0.955 & & 0.996 & 0.938 & 0.966 \\
-& istibsar & & 0.999 & 0.999 & 0.999 & & 0.980 & 0.931 & 0.955 & & 0.994 & 0.985 & 0.990 \\  \cline{2-14}
-& &  & 0.999 & 0.931 & 0.962 & & 0.969 & 0.951 & 0.960 & & 0.997 & 0.974 & 0.985 \\ \hline \hline
-%& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
-%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
-\multirow{3}{*}{no learning}
-& musnad & & 0.999 & 0.999 & 0.999 & & 0.894 & 0.985 & 0.937 & & 0.995 & 0.999 & 0.998\\
-& kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.991 & 0.992\\
-& istibsar & & 0.999 & 0.973 & 0.986 & & 0.977 & 0.950 & 0.963 & & 0.996 & 0.998 & 0.997\\ \cline{2-14}
- & &  & 0.999 & 0.964 & 0.982 & & 0.936 & 0.968 & 0.951 & & 0.995 & 0.996 & 0.995 \\
+\caption{Narrator graph and name learning results.}
+\resizebox{1\columnwidth}{!}{
+\begin{tabular}{cp{1cm}c}
+	\begin{tabular}{p{1.5cm}p{.1cm}ccc}
+    narrators & & recall & precision & F-score \\ \cline{1-1} \cline{3-5}
+	detected  & & 0.673 & 0.824 & 0.741\\
+	all       & &  0.632 & 0.771 & 0.694\\
+	\multicolumn{5}{c}{(a) Narrator graph accuracy} \\
+	\end{tabular}  
+&  &
+	\begin{tabular}{p{1.5cm}p{.1cm}ccc}
+     names    & & recall & precision & F-score \\ \cline{1-1} \cline{3-5}
+	 sanad    & &  0.649 & 0.940 & 0.768\\
+	 all      & & 0.794 & 0.733 & 0.762\\
+	\multicolumn{5}{c}{(b) Name learning accuracy} \\
+	\end{tabular}  \\
  \end{tabular}               
@@
 \normalsize                   
-\label{t:nopunctlearnresults} 
+\label{t:PORnLearning} 
 \end{table*}
 
-We are interested in studying the effect of the punctuation 
-and the learning optimizations we implemented in NaGEMA.
-We utilized punctuation marks to increase the accuracy of narration detection and narrator boundaries. 
-We also learned names of narrators that consisted of words missing from 
-our lexicon.
-We learned the narrator names based on their context as they occurred 
-between narrator connectors or contained name connectors.
-
-In Table~\ref{t:nopunctlearnresults}, we report the results of running NaGEMA, 
-disabling each of the optimizations at a time.
-
-We notice that punctuations are useful for increasing all precision measures: hadith segmentation, sanad boundary and
-narrator boundary. Those measures are enhanced by about 
-3\%, 2\%, and 1\% respectively as punctuation related 
-information is utilized.  
-We observed that punctuation marks helped eliminate confusion that resulted from a 
-concentration of names and words that 
-can be interpreted falsely as narrator connectors. 
-A period or a paragraph mark separated some name concentrations 
-dropping each part below the acceptable sanad threshold. 
-Punctuation marks also help cleaning the sanad of a narration from 
-names that coincidentally appear at the 
-end of the matn of a previous narration. 
-Names from a previous narration maybe mistaken to belong to the sanad of the current 
-narration if they were close enough. 
-Punctuation marks reduce such confusions.
-
-Although simple, our name learning optimization enhanced the recall 
-of sanad boundaries by about 3.4\% on the average.
-This means that we correctly report more narrators.
-This is a significant improvement as the task of hadith authentication requires high accuracy. 
-In particular, sometimes the presence of a specific narrator in the chain can render a 
-hadith as not trustworthy since the narrator is not considered credible.
-%That's why it is important not to lose any of the narrators that appear in the sanad.
-
-\subsection{Distance metric}
-Finally, the distance metric had 0.993 precision,
-0.999 recall and an F-score of 0.997.
-
-(explain why successful on this problem)
-
-(show detailed statistics and compare against that of eNarrator which uses a similarity string comparison appraoch)
-
-\subsection{POR Correctness}
-
-(report recall and precision of nodes missing or those added, bc edges cannot be mistaken other than that)
-
-(report recall and precision on the chainnodes per graphnode, that is all what retains to correctness resulting from algorithm)
-
-\section{Significance \& Extensions -- Named Entity Extraction in Other Contexts}
-
-\subsection{Genealogical Lists in the Bible}
-
-Biblical genealogical lists describe family trees up to Adam and are of interest to Bible scholars including historians and critics. 
-They trace key figures such as Israelite kings and prophets. Figure~\ref{f:gene1} shows an excerpt from Genesis 5:1-11.
-
-\begin{figure}\centering
-\resizebox{0.9\columnwidth}{!}{
-\parbox[t]{1\columnwidth}{
-\tiny
-``This is the written account of {\bf Adam}'s line...
-
-%When God created man, he made him in the likeness of God. He created them male and female and blessed them. 
-%And when they were created, he called them `man'.
-
-When {\bf Adam} had lived 130 years, he had a son in his own likeness, in his own image; and he named him {\bf Seth}. 
-After Seth was born, {\bf Adam} lived 800 years and had other sons and daughters. 
-Altogether, {\bf Adam} lived 930 years, and then he died.
-
-When {\bf Seth} had lived 105 years, he became the father of {\bf Enosh}. 
-And after he became the father of {\bf Enosh}, {\bf Seth} lived 807 years and had other sons and daughters. 
-Altogether, {\bf Seth} lived 912 years, and then he died.
-
-When {\bf Enosh} had lived 90 years, he became the father of {\bf Kenan}. 
-And after he became the father of {\bf Kenan}, {\bf Enosh} lived 815 years and had other sons and daughters. 
-Altogether, {\bf Enosh} lived 905 years, and then he died.''
-}}
-\caption{Genealogical list from Genesis 5:1-11}
-\end{figure}
-
-\begin{figure}\centering
-\resizebox{0.9\columnwidth}{!}{
-\parbox[t]{1\columnwidth}{
-\tiny
-``{\bf Adam}, {\bf Seth}, {\bf Enosh}, {\bf Kenan}, {\bf Mahalalel}, {\bf Jared}, {\bf Enoch}, 
-{\bf Methuselah}, {\bf Lamech}, {\bf Noah}. 
-The sons of {\bf Noah}: {\bf Shem}, {\bf Ham} and {\bf Japheth}.
-The sons of {\bf Japheth}: {\bf Gomer}, {\bf Magog}, {\bf Madai}, {\bf Javan}, {\bf Tubal}, {\bf Meshech} and {\bf Tiras}.
-The sons of {\bf Gomer}: {\bf Ashkenaz}, {\bf Riphath} and {\bf Togarmah}.
-The sons of {\bf Javan}: {\bf Elishah}, {\bf Tarshish}, the {\bf Kittim} and the {\bf Rodanim}.
-The sons of {\bf Ham}: {\bf Cush}, {\bf Mizraim}, {\bf Put} and {\bf Canaan}.
-The sons of {\bf Cush}: {\bf Seba}, {\bf Havilah}, {\bf Sabta}, {\bf Raamah} and {\bf Sabteca}.
-The sons of {\bf Raamah}: {\bf Sheba} and {\bf Dedan}. 
-{\bf Cush} was the father of {\bf Nimrod}, who grew to be a mighty warrior on earth. ...'' }}
-\caption{Genealogical list from 1 Chronicles 1:1-8}
-\end{figure}
-
-%In many cases, 
-%there is an overlap between the many individual lists available across the Bible. When such an overlap occurs 
-%meticulous readers can detect some seeming contradictions, in many cases due to conflicts or omissions.
-%Detecting those conflicts is a task that has been performed by countless 
-%number of Bible students, theologians and historians.
-
-%The importance of such lists was profound for the old nation of Israel. They 
-The genealogical lists established the kingship lineage and determined
-rights and privileges in the temple. Genealogical lists also determined descent with all its implications in biblical times
-such as the concept of reverted ownership of sold real estate; every fifty years ownership of sold land reverted to the 
-original owners and their inheritors. Genealogical lists were preserved due to the possible privilege of being the ancestor of the awaited Messiah.~\cite{catholicEncyclopedia:Online}
-
-For many believers, genealogies are used as the main indicators of chronology especially when calculating the age of mankind~\cite{Thomas:92}.
-Believers used them to defend the historicity of Biblical events and characters against other accounts from other 
-texts thought of as fables and myths. 
-However, separate genealogical lists that overlap suffer from consistency issues. 
-For example, Matthew and Luke report two different genealogies of Jesus. 
-Several biblical researchers studied the consistency of the lists~\cite{catholicEncyclopedia:Online,Belote:Online,completeBibleGenealogy:Online}, 
-and published articles and books on the topic. Up to our knowledge very few include graphical illustrations such as family trees and graphs 
-which makes it easier for the reader to visualize the differences. 
-Our methodology can automatically create such illustrations.
-
-Creating a comprehensive genealogical graph from all biblical accounts greatly enhances its study,
-since readers would be able to better understand a character's genealogical context and life span. Such comprehensive graphs can be tagged
-with time durations between a father and his giving birth of his sons as is common in many biblical accounts.
-
-Work such as~\cite{Belote:Online,SoulLiberty:Online} constructed partial genealogical graphs manually from several lists.
-This work is short from being comprehensive.
-
-In what follows, we explain how NaGEMA can be easily manipulated to be utilized 
-to generate this information automatically.
-
-%The main difficulty that hindered this task was the huge size of the resulting graphs whose layout becomes
-%challenging. This difficulty is also present in the hadith problem we undertook in this paper; however, we were able to tackle this issue by 
-%allowing for partial layout to be drawn on the fly as the user navigates in the POR which is already computed, 
-%instead of precomputed complete layouting of the graph on which plain scrolling is used to navigate. (check the accompanying software).
-%In all cases, it is noteworthy, that all genealogical trees/lineages available in any literature today are the product of manual work.
-
-%is in establishing the historicity of characters under consideration,
-%Such lists had and still have an utmost importance for Jews because heritage
-%according to their law dictates the land's ownership according to their inheritance which cannot be sold perminantly, they also
-
-%Before presenting the necessary modifications for the state-machine used by NaGEMA, we present two representative biblical sections that
-%contain genealogical lists. 
-%NaGEMA can handle 
-%These sections contain overlapping linages, however stated with different styles: \\
-
-\begin{figure}[tb]
-\center{
-\resizebox{1\columnwidth}{!}
-%{\includegraphics{figs/narrator_chain_output.pdf}}
-{\includegraphics{figs/controller_bible.pdf}}
-\caption{Controller of biblical genealogies}
-\label{f:bible} 
+{\bf Learning names.}~~
+Table~\ref{t:PORnLearning}(b) reports the precision and recall of the names that our method learned from the 
+books of hadith. 
+Within the sanad segments, our method learned 65\% of the names it should learn with 94\% accuracy.
+It learned 79\% of the names it should learn from the whole text at a 73\% precision. 
+The precision is higher within the sanad since names are more frequent in the sanad context.
+The recall is higher for the whole text since inside the sanad our method is more likely to resolve ambiguous tags 
+as name and narrator connectors than if outside the sanad.
+
+{\bf Biography narrator detection.}~~
+Table~\ref{t:narratorDifference} shows the accuracy results for the narrator detection and the biography boundary detection
+for the Khoei biography book.
+The first row shows the results computed by considering the biography book alone without the help of the narrator graph
+extracted from the hadith books. 
+Without the narrator graph, our method detected 94\% of the narrators mentioned in the book with 81\% precision. 
+The method without the narrator graph, detected correctly (94\% precision) the boundaries of the biographies.
+
+With the help of the narrator graph, our method detected 65\% of the narrator names in the biography book. 
+The rest of the names were simply not found in the hadith books that we used to generate the narrator graph. 
+With the use of the narrator graph and the coloring algorithm, our method detected the boundaries of the biographies
+with 93\% precision. 
+We believe that an exhaustive graph that includes narrator names from more books of hadith will improve the 65\% narrator 
+detection recall metric.
+In all cases, narrator names that exist in the biography and not in the hadith books are not of interest to the narrator
+graph annotation problem.
+
+%The numbers in the right hand side of Table~\ref{t:narratorDifference}  may explain why 
+Narrator detection in biography books is harder than in hadith books. 
+For example, the ratio of narrator names to the rest of the text is 22\% in Musnad while it is 49\% in Khoei,
+and the average length of a narrator sequence in Musnad is 2.6 narrators while it is 3.5 narrators in Khoei.
+With the use of the narrator graph extracted from the hadith books, our method was able to improve biography boundary detection
+from 41\% to 93\%.
+This is evidence that cross-document entity detection helps to accurately find named entities in harder NLP tasks.
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Accuracy results for the biography application.}
+\resizebox{.7\columnwidth}{!}{
+
+%\begin{tabular}{cp{0.3cm}c}
+
+	\begin{tabular}{cp{.1cm}cccp{.1cm}ccc}
+	          & & \multicolumn{3}{c}{narrator detection} & &  \multicolumn{3}{c}{biography boundary} \\
+	 & & recall & precision & F-score & & recall & precision & F-score \\ 
+ \cline{3-5} \cline{7-9}
+%hadith	 & sanad & & 0.923 & 0.999 & 0.960 & & 0.996 & 0.999 & 0.998  \\
+%(musnad) & all & & 0.954 & 0.961 & 0.958 & & 0.996 & 0.987 & 0.992 \\ 
+%	\hline \hline
+w/o graph & & 0.943 & 0.812 & 0.873 & & 0.407 & 0.945 & 0.569 \\
+with graph & & 0.654 & 0.909 & 0.760 & & 0.933 & 0.960 & 0.946 \\ % found in narrator graph
+	 \end{tabular}   
+%&  &
+%	\begin{tabular}{ccc}
+%	& names    & nar. \\
+%    & per text & elements \\ \cline{2-3}
+%	\multirow{2}{*}{0.217} &\multirow{2}{*}{ 2.64}\\ \\
+%	\hline \hline
+%	\multirow{2}{*}{0.489} & \multirow{2}{*}{3.54}\\
+%    musnad & 0.217 & 2.64 \\ 
+%    biography & 0.489 & 3.54 \\
+%	\end{tabular}  \\
+%\end{tabular}          
 }                             
-\end{figure}
-
-Given the whole text of the Bible, NaGEMA detects all sections that have genealogical lists
-and extract from them the names structured in a genealogical DAG. The resemblance between this problem and that of hadith is evident; 
-but still there are some differences that should be highlighted. 
-First, in biblical genealogies, the individual names are simple in contrast to those present in hadith; 
-consequently, the concept of a `name connector (NMC)' is unnecessary in the problem at hand.
-Second, genealogies are not a flat sequence of narrators as in hadith but a branching tree of decedents; 
-for this reason, it is necessary to detect not just the names in the text but also the node to which they are to be attached.
-Those two differences are treated in the state-machine presented in Figure~\ref{f:bible} that contains all changes necessary to adapt NAGEMA to detect 
-biblical genealogies.
-
-The controller has three states that correspond to 
-the sentence structure under consideration. 
-State $\mathit{TEXT\_S}$ is the initial state and denotes 
-that the controller is outside the context of a genealogical list.
-State $\mathit{NAME\_S}$ denotes that the controller is in
-the middle of a list and has last met some names. 
-State $\mathit{SONS\_S}$ denotes that the controller is in
-the middle of sentence that lists the sons of a previously named ancestor such as sentences that appeared in the second quoted biblical texts.
-
-The conditions that appear in the controller, are as based on the following symbols:
-\begin{itemize}
-\item {\bf NEW\_NAME} denotes that the word at had is a name not present in the
-	currently built tree.
-\item {\bf LEAF\_NAME} denotes a name that appears as a leaf in the tree (i.e. has no
-	descendants attached to it).
-\item {\bf CORE\_NAME} denotes a name that has already descendants attached to it (i.e. 
-	non-leaf node).
-\item {\bf DC} denotes a descent connector such as "the son of", "became father to",
-	"begot", etc... that help indicate -- in addition to the names -- that the current text
-	is a genealogical list. They also have the role of indicating the type of link and its 
-	direction (father-son or son-father); such role is omitted in the state-machine for clarity.
-\item {\bf OTHER} denotes any other word.
-\end{itemize}
-The threshold $\theta_{\mathit{o}}$ 
-corresponds to the number of tolerated words that are not names 
-or descent connectors (DC); while 
-$i_{\mathit{o}}$ corresponds to their current count in the state-machine. 
-Another threshold which is used is $\mathit{N_{min}}$ which denotes the number of minimum
-levels that are necessarily available in the tree to be considered a valid genealogical
-tree instead of an erroneous one generated from a group of coincidental names. The
-concept of tolerance and minimum number of structures have been inspired from the
-hadith controller and are suitable for detecting texts that contain lists of similar
-type structures.
-
-The function $\mathit{n_1.setFather(n_2)}$ creates a node in the tree 
-for $\mathit{n_1}$ and creates
-a link between $\mathit{n_1}$ and the already present node $\mathit{n_2}$. 
-The function $\mathit{getLevels()}$ 
-returns the number of generations already present in the tree. This number
-assesses whether the currently built tree is an acceptable one or not upon
-transitioning from sate $\mathit{NAME\_S}$ to $\mathit{TEXT\_S}$
-
-We used the values of 3 and 20 for $\theta_{\mathit{o}}$ and 
-$N_{\mathit{min}}$ respectively, to run the experiments 
-we report in what follows.
-
-(report preliminary results on bible genealogy)
-
+\normalsize                   
+\label{t:narratorDifference} 
+\end{table*}
+
+% last set of tables
+
+%Table~\ref{t:biography} reports the accuracy of the results computed when we looked up the biographies using the 
+%narrator graph multi-hash method: 80\% hit matches. 
+%\begin{table*}[bt]                                     
+%\centering                                            
+%\caption{Biography Detection Results}
+%\resizebox{0.25\columnwidth}{!}{
+%	\begin{tabular}{p{2.5cm}p{.1cm}c}
+%	Exact Recall & &  0.6 \\
+%	Partial Recall & & 0.8\\
+%	Search Count & & 10 \\
+%	\end{tabular}             
+%}                             
+%\normalsize                   
+%\label{t:biography} 
+%\end{table*}     
+%
+
+\begin{table*}[bt]                                     
+\centering                                            
+\caption{Results for Biblical genealogy family tree extraction. }
+\resizebox{1\columnwidth}{!}{
+\begin{tabular}{p{1.2cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+& & & \multicolumn{3}{c}{nodes} & &  \multicolumn{3}{c}{neighbors} & &  \multicolumn{3}{c}{context} \\
+& trees & &  recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score 
+  \\ \cline{2-2} \cline{4-6} \cline{8-10} \cline{12-14}
+%\multirow{3}{*}{\bf Before learning }
+{\bf Before} &local && 0.583 & 0.898 & 0.707 && 0.778 & 0.760 & 0.769 && 0.552 & 0.588 & 0.570 \\
+{\bf learning} &aligned && 0.849 & 0.868 & 0.858 && 0.777 & 0.738 & 0.757 && 0.554 & 0.590 & 0.572 \\
+&full && 0.784 & 0.800 & 0.792 && 0.652 & 0.652 & 0.652 && 0.536 & 0.573 & 0.554  \\
+\hline \hline
+%\multirow{3}{*}{Refined Controller}
+{\bf After} &local && 0.586 & 0.900 & 0.710 && 0.780 & 0.725 & 0.751 && 0.557 & 0.577 & 0.567 \\
+{\bf learning} &aligned&& 0.854 & 0.871 & 0.862 && 0.785 & 0.723 & 0.753 && 0.582 & 0.600 & 0.591 \\
+&full && 0.820 & 0.810 & 0.815 && 0.701 & 0.708 & 0.704 && 0.580 & 0.630 & 0.604  \\
+ \end{tabular}                 
+}                             
+\normalsize                   
+\label{t:genealogyGraphs} 
+\end{table*}
+
+{\bf Genealogical family tree detection.}~~ 
+We report the precision and recall of the detected names in the ``nodes'' columns of Table~\ref{t:genealogyGraphs},
+the accuracy of their neighboring nodes in the ``neighbors'' column, 
+and the accuracy of the semantics of the edge labels (``spouse'' and ``child'' edge labels) 
+in the ``context'' columns.
+The table shows the results before and after learning the semantics of unclassified edges.
+Learned edge semantics improved the recall of the full merged graph 
+by 5\% and the precision by 6\%.
+
+The local row shows the accuracy results without aligning the local detected lists to the local annotated lists. 
+For example, if the manual annotation has a list that spanned two automatically detected lists, we compute the accuracy for each 
+detected list against the manually detected list alone.
+The aligned row shows the accuracy results with maximum boundary inclusion. In this case we merge the automatically detected lists
+and compute the metrics.
+The full row shows the accuracy results for the tree automatically generated from merging all automatically detected lists.
+%We manually generated the reference tree.
+
+Our method extracted names with 77\% recall and 91\% precision, and segmented genealogical lists with an 80\% recall
+and 74\% precision. 
+The manually generated full graph had 250 nodes, while our method extracted a graph with 245 nodes
+before learning, and 253 nodes after learning. 
+The largest manually annotated local graph contained 77 nodes, while the largest extracted local graph
+using our method had 35 nodes.
+
+%
+%\begin{table*}[bt]                                     
+%\centering                                            
+%\caption{Results for Biblical Genealogy Parsing}
+%\resizebox{0.7\columnwidth}{!}{
+%\begin{tabular}{p{2.5cm}cccp{.1cm}ccc}
+% & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{names extraction} \\
+% &  recall & precision & F-score & & recall & precision & F-score \\ \cline{2-8}
+%basic controller & 0.796 & 0.708 & 0.749 & & 0.883 & 0.906 & 0.894 \\
+%refined controller & 0.796 & 0.742 & 0.768 & & 0.887 & 0.910 & 0.898 \\
+% \end{tabular}                 
+%}                             
+%\normalsize                   
+%\label{t:genealogyParsing} 
+%\end{table*}     
+%
+
+%\begin{table*}[bt]                                     
+%\centering                                            
+%\caption{Statistics for Biblical Genealogy Kinship Graphs}
+%\resizebox{0.7\columnwidth}{!}{
+%\begin{tabular}{p{1.5cm}lp{.1cm}ccccp{.1cm}c}
+%& & & \multicolumn{6}{c}{Node Count} \\ \cline{4-9}
+%& & & \multicolumn{4}{c}{Local Graphs} & & \multicolumn{1}{c}{Global Graph}\\
+%& Controller & &  minimum & average & maximum & total size & & size \\ \cline{2-9}
+%\multirow{2}{*}{Extracted}
+%&Basic && 3 & 7.89 & 35 & 513 && 245 \\
+%&Refined && 3 & 8.00 & 35 & 496 && 253 \\
+% \hline
+%\multirow{1}{*}{Annotation}
+%&  && 2 & 8.80 & 77 & 475 && 250 \\
+% \end{tabular}                 
+%}                             
+%\normalsize                   
+%\label{t:genealogyGraphStatistics} 
+%\end{table*}     
+%
+
+%\subsection{On Unorganized hadith books}
+%
+%(report results on such hadith books)
+%
+%\subsection{Effect of punctuation and learning}
+%
+%\begin{table*}[bt]                                     
+%\centering                                            
+%\caption{Results without punctuation and without learning. }
+%\resizebox{.95\columnwidth}{!}{
+%\begin{tabular}{p{2.1cm}lp{.1cm}cccp{.1cm}cccp{.1cm}ccc}
+%& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+%\multirow{3}{*}{no punctuation} 
+%& musnad  & & 0.999 & 0.999 & 0.999 & & 0.982 & 0.955 & 0.968 & & 0.999 & 0.998 & 0.999 \\ 
+%& kafi & & 0.999 & 0.793 & 0.885 & & 0.944 & 0.967 & 0.955 & & 0.996 & 0.938 & 0.966 \\
+%& istibsar & & 0.999 & 0.999 & 0.999 & & 0.980 & 0.931 & 0.955 & & 0.994 & 0.985 & 0.990 \\  \cline{2-14}
+%& &  & 0.999 & 0.931 & 0.962 & & 0.969 & 0.951 & 0.960 & & 0.997 & 0.974 & 0.985 \\ \hline \hline
+%%& & & \multicolumn{3}{c}{segmentation} & &  \multicolumn{3}{c}{sanad boundary}  & &  \multicolumn{3}{c}{name boundary} \\
+%%& & & recall & precision & F-score & & recall & precision & F-score & & recall & precision & F-score \\ \cline{2-14}
+%\multirow{3}{*}{no learning}
+%& musnad & & 0.999 & 0.999 & 0.999 & & 0.894 & 0.985 & 0.937 & & 0.995 & 0.999 & 0.998\\
+%& kafi & & 0.999 & 0.920 & 0.958 & & 0.937 & 0.970 & 0.953 & & 0.993 & 0.991 & 0.992\\
+%& istibsar & & 0.999 & 0.973 & 0.986 & & 0.977 & 0.950 & 0.963 & & 0.996 & 0.998 & 0.997\\ \cline{2-14}
+% & &  & 0.999 & 0.964 & 0.982 & & 0.936 & 0.968 & 0.951 & & 0.995 & 0.996 & 0.995 \\
+% \end{tabular}                 
+%}                             
+%\normalsize                   
+%\label{t:nopunctlearnresults} 
+%\end{table*}                   
+%
+%
+%We are interested in studying the effect of the punctuation 
+%and the learning optimizations we implemented in NaGEMA.
+%We utilized punctuation marks to increase the accuracy of narration detection and narrator boundaries. 
+%We also learned names of narrators that consisted of words missing from 
+%our lexicon.
+%We learned the narrator names based on their context as they occurred 
+%between narrator connectors or contained name connectors. 
+%
+%In Table~\ref{t:nopunctlearnresults}, we report the results of running NaGEMA, 
+%disabling each of the optimizations at a time.
+%
+%We notice that punctuations are useful for increasing all precision measures: hadith segmentation, sanad boundary and
+%narrator boundary. Those measures are enhanced by about 
+%3\%, 2\%, and 1\% respectively as punctuation related 
+%information is utilized.  
+%We observed that punctuation marks helped eliminate confusion that resulted from a 
+%concentration of names and words that 
+%can be interpreted falsely as narrator connectors. 
+%A period or a paragraph mark separated some name concentrations 
+%dropping each part below the acceptable sanad threshold. 
+%Punctuation marks also help cleaning the sanad of a narration from 
+%names that coincidentally appear at the 
+%end of the matn of a previous narration. 
+%Names from a previous narration maybe mistaken to belong to the sanad of the current 
+%narration if they were close enough. 
+%Punctuation marks reduce such confusions.
+%
+%
+%Although simple, our name learning optimization enhanced the recall 
+%of sanad boundaries by about 3.4\% on the average.
+%This means that we correctly report more narrators.
+%This is a significant improvement as the task of hadith authentication requires high accuracy. 
+%In particular, sometimes the presence of a specific narrator in the sequence can render a 
+%hadith as not trustworthy since the narrator is not considered credible.
+%%That's why it is important not to lose any of the narrators that appear in the sanad. 
+%
+%\subsection{POR Correctness}
+%
+%(report recall and precision of nodes missing or those added, bc edges cannot be mistaken other than that)
+%
+%
+%\section{Significance \& Extensions -- Named Entity Extraction in Other Contexts}
+%(report preliminary results on bible genealogy)
+%
 %TODO:
@@
 
-\subsection{Organizing Quoted Email Replies/Forwards}
-
-\begin{figure}[b]
-\center{
-\resizebox{1\columnwidth}{!}
-%{\includegraphics{figs/narrator_chain_output.pdf}}
-{\includegraphics{figs/email1.pdf}}
-\caption{Thread tree generated from flat emails}
-\label{f:email1} 
-}
-\end{figure}
-
-\begin{figure}[tb]
-\center{
-\resizebox{0.7\columnwidth}{!}
-%{\includegraphics{figs/narrator_chain_output.pdf}}
-{\includegraphics{figs/email2.pdf}}
-\caption{Graph of contacts and other people known through the threads available}
-\label{f:email2} 
-}
-\end{figure}
-
-Emails have been one of the most important tool for communication in business, distant family members and friends.
-In many cases, one email may trigger a thread of replies and forwards that may circulate between different people. It may even
-end up after several replies being sent to someone new to the whole thread and he/she might consider navigating through
-the lengthy list of quoted replies/forwards to get a better understanding of the conversations which have been going prior to 
-receiving the email. However, the lengthy lists of conversations with some added indentation or symbols each possibly in a different
-format may be a cumbersome task to do. That's why email clients such as Gmail try organizing each conversation in a separate tab for
-better use. Nonetheless, such clients are restricted to emails that are already available in the recipient's `Inbox' or `Sent Mails'.
-In other words, even if the email client receives an email which has 
-tens of separate replies/forwards but that are not previously available for the client, it 
-fails to organize them and instead keeps them as one list of quoted conversations.
-
-In hadith, we have a similar concept: a narration from the prophet that has been orally transmitted and ends up being written
-at the end with all the narrators in between being quoted by the last narrator who receives the narration. 
-It is interesting to be able to solve the `reincarnated' version of
-the problem that appears in our technological world.
-
-It is noteworthy that the problem at had is easier than its historical counterpart, due to the semi-structured nature of emails.
-For instance, the senders, co-recipients, subject, body and so on of the email follow clear rules. Since in our case, we need not 
-to mine the unstructured content of the email, the problem is clearly much easier than the ones presented earlier. 
-Consequently by considering the `From', `To', `CC' and `BCC' fields one can construct the `narrator' chains for conversation
-threads, each tagged with the body of the the individual reply/forward conversation. Consider a sample email tree that
-can be generated as shown in Figure~\ref{f:email1}. Notice that the four flat emails shown to the right of the figure,
-three of which lie in the `Inbox' and one of which in the `Sent Mails' were used to generate the more instructive structure
-to the left. Although in literature~\cite{Smith:00,Venolia:03} 
-alternative tree-like structures where proposed, however, they were restricted to emails
-available in the Inbox/Sent-Mails; consequently they will be unable to recover the top three levels of the tree generated through 
-the quoted previously unknown emails. % (search more extensively literature, if this is true)
-
-Furthermore, it can be envisioned that the Inbox or subset of it, can be merged into a graph structure of people with 
-whom a user communicates and the
-history of their communication as in the case of the hadith. 
-%This structure might be useful to extract other relations
-%such as the importance and priority of emails sent by certain contacts and for predicting any possibly forgotten recipient of an email to be sent.
-Consider a sample graph depicted in Figure~\ref{f:email2} 
-that shows the people involved in the thread of figure~\ref{f:email1} along with others of previous threads.
-With such a graph, the user can get to learn about people that are not known to him and are possible contacts because they are involved 
-in threads that reach him/her.
-
-%(we can use the "FROM:" and punctuation to extract the sender and the body of the email and organize a 
-%received email into individual chains of replies/forwards.
-Also the Inbox can be merged into a POR of contacts with whom a user communicates and the history of their communication, as in the case of the hadith...)
-
-\subsection{Place Names in Travel Itineraries}
-
-Travel Itineraries are plans of travel or routes. They are usually used as guides for travelers. They list different
-places through which the route to the destination passes. 
-In many cases, those guides are provided in written format
-It is also common that alternative routes be suggested to the same destination.
-Consider the example below for the itineraries known as the `Silk Road'\footnote{from \url{http://wikitravel.org/en/Silk_Road}}:
-
-{\setlength{\parindent}{20pt} 
-\setlength{\parskip}{1ex}
-\indent-Northern Silk Route \\
-\indent \indent Dunhuang to Hami \\
-\indent \indent Hami to Turfan \\ 
-\indent \indent Turfan to Korla \\
-\indent \indent Korla to Kashgar
-
-\indent -Middle Silk Route \\ 
-\indent \indent Dunhuang to Cherchen \\
-\indent \indent Cherchen to Korla \\
-\indent \indent Korla to Kashgar
-
-\indent -Southern (Jade) Silk Route \\
-\indent \indent Dunhuang to Cherchen \\
-\indent \indent Cherchen to Khotan \\
-\indent \indent Khotan to Yarkand \\
-\indent \indent Yarkand to Kashgar \\ }
-
-In a similar manner to extracting hadith, a controller that looks for aggregations of Place names
-with the NRC equivalent being phrases such as `to', `pass in', `stop at', etc... 
-Afterwards, itineraries can be merged into a POR by merging common locations.
-Clearly, such a display is much more illustrative than having them in text format.
-Also, when electronic maps are available, POR's can automatically be turned into emphasized routes on maps.
-
-\subsection{Extending Arabic People Name Gazetteers}
-
-The problem at hand has significant contribution to the general task of Named Entity Recognition (NER). It has been noted that
-gazetteers play an important role in the accuracy of knowledge-based approaches and bootstrapping learning
-approaches~\cite{Kozareva:06,Carlson:09,Toral:06,Benajiba:07,Kazama:08,Nadeau:06}. Nonetheless, although gazetteers result in high accuracy,
-they are hard to obtain and mantain. % They are also language and domain-dependent. 
-Consequently, many automatic methods have been proposed to generate extensive gazetteers instead of relying on the time-consuming manual work
-to compile such lists. Proposed methods include: writing some rules for the context of the named entities targeted such as propositions that
-introduce them, also known as trigger gazetteers~\cite{Kozareva:06,Toral:06}; utilizing free online encyclopedias such as Wikipedia which are 
-continually maintained~\cite{Benajiba:07,Kazama:08}; in addition to reliance on web search engines starting from a small list of words and extracting
-similar that appear in the extracted pages~\cite{Nadeau:06}.
-
-As previously noted, when solving the hadith problem, we were able to learn much previously unknown names using their context. We have used methods
-relating to those found in~\cite{Traboulsi:09,Kozareva:06}; however, our learned lists appear to be more accurate than their counterparts due to the
-the more characteristics of the hadith corpus and the wider context on which we rely on. Particularly, we learn only names that appear in the hadith sanad
-detected through the presence of an aggregation of previously known names in its neighborhood along with name connectors such as narration verbs and 
-prepositions. In Table X, we compare the results of our learning methods to evaluate the role of the sand context. Clearly, the sanad context provides 
-a higher confidence in the accuracy of our extracted names. Accordingly, our generated gazetteers are superior than those extracted otherwise and 
-enrich the task of Arabic Named Entity Extraction resources. The same applies to that of Place names that can be discovered in the context of travel
-itineraries and the people names that appear in biblical genealogical lists. Since the latter sources are available in multitude of languages, they
-can be a rich source for building gazetteers for resource-poor languages.
-
-(present the necessary table and expand upon its significance)
+%\subsection{Place Names in Travel Itineraries}
+%
+%Travel Itineraries are plans of travel or routes. They are usually used as guides for travelers. They list different
+%places through which the route to the destination passes. 
+%In many cases, those guides are provided in written format
+%It is also common that alternative routes be suggested to the same destination.
+%Consider the example below for the itineraries known as the `Silk Road'\footnote{from \url{http://wikitravel.org/en/Silk_Road}}:
+%
+%{\setlength{\parindent}{20pt} 
+%\setlength{\parskip}{1ex}
+%\indent-Northern Silk Route \\
+%\indent \indent Dunhuang to Hami \\
+%\indent \indent Hami to Turfan \\ 
+%\indent \indent Turfan to Korla \\
+%\indent \indent Korla to Kashgar  
+%
+%\indent -Middle Silk Route \\ 
+%\indent \indent Dunhuang to Cherchen \\
+%\indent \indent Cherchen to Korla \\
+%\indent \indent Korla to Kashgar 
+%
+%\indent -Southern (Jade) Silk Route \\
+%\indent \indent Dunhuang to Cherchen \\
+%\indent \indent Cherchen to Khotan \\
+%\indent \indent Khotan to Yarkand \\
+%\indent \indent Yarkand to Kashgar \\ }
+%
+%
+%In a similar manner to extracting hadith, a FST that looks for aggregations of Place names
+%with the NRC equivalent being phrases such as `to', `pass in', `stop at', etc... 
+%Afterwards, itineraries can be merged into a POR by merging common locations.
+%Clearly, such a display is much more illustrative than having them in text format.
+%Also, when electronic maps are available, nnPOR's can automatically be turned into emphasized routes on maps.
+%
+%\subsection{Extending Arabic People Name Gazetteers}
+%
+%The problem at hand has significant contribution to the general task of Named Entity Recognition (NER). It has been noted that
+%gazetteers play an important role in the accuracy of knowledge-based approaches and bootstrapping learning
+%approaches~\cite{Kozareva:06,Carlson:09,Toral:06,Benajiba:07,Kazama:08,Nadeau:06}. Nonetheless, although gazetteers result in high accuracy,
+%they are hard to obtain and mantain. % They are also language and domain-dependent. 
+%Consequently, many automatic methods have been proposed to generate extensive gazetteers instead of relying on the time-consuming manual work
+%to compile such lists. Proposed methods include: writing some rules for the context of the named entities targeted such as propositions that
+%introduce them, also known as trigger gazetteers~\cite{Kozareva:06,Toral:06}; utilizing free online encyclopedias such as Wikipedia which are 
+%continually maintained~\cite{Benajiba:07,Kazama:08}; in addition to reliance on web search engines starting from a small list of words and extracting
+%similar that appear in the extracted pages~\cite{Nadeau:06}.
+%
+%As previously noted, when solving the hadith problem, we were able to learn much previously unknown names using their context. We have used methods
+%relating to those found in~\cite{Traboulsi:09,Kozareva:06}; however, our learned lists appear to be more accurate than their counterparts due to the
+%the more characteristics of the hadith corpus and the wider context on which we rely on. Particularly, we learn only names that appear in the hadith sanad
+%detected through the presence of an aggregation of previously known names in its neighborhood along with name connectors such as narration verbs and 
+%prepositions. In Table X, we compare the results of our learning methods to evaluate the role of the sand context. Clearly, the sanad context provides 
+%a higher confidence in the accuracy of our extracted names. Accordingly, our generated gazetteers are superior than those extracted otherwise and 
+%enrich the task of Arabic Named Entity Extraction resources. The same applies to that of Place names that can be discovered in the context of travel
+%itineraries and the people names that appear in biblical genealogical lists. Since the latter sources are available in multitude of languages, they
+%can be a rich source for building gazetteers for resource-poor languages.
+%
+%(present the necessary table and expand upon its significance)
 
@@
 
-We presented NaGEMA, a narrator chain extractor
-using morphological analysis. 
-NaGEMA automatically extracted narrator chains with 
-above 90\% accuracy
-from three different hadith books and generated 
-directed acyclic graphs relating the narrators 
-in a partial order relation.
-
-Up to our knowledge, NaGEMA is the first 
-tool that automates hadith segmentation and builds
-a comprehensive POR graph.
-It uses morphological analysis coupled with 
-and a knowledge based FSM to  detect the hadith 
-sanad.
-
-The partial order graph is instrumental to automate
-localization of narrators in biography documents and
-segmenting biography documents in the future.
-Once the biographies are analyzed, we can annotate
-the POR with qualifiers of the narrators that reflect
-their authenticity. 
-We also plan to 
-enhance the accuracy of hadith detection and the 
-narrator merging algorithms. 
+We presented a method to detect named entities and relations amongst them from 
+Arabic text using morphological analysis, FSTs, and 
+graph transformation algorithms. 
+We implemented our method and evaluated it using three NLP applications.
+Our method extracted named entities with high accuracy in all application.
+Our method extracted relations amongst the named entities with high accuracy in the hadith and biography applications, 
+and with acceptable accuracy in the genealogical lists application. 
+Our method learns named entities and semantics of named entities in terms of graph transformations that 
+replace a detected unclassified edge in a graph with predefined edges. 
+In the future, we plan to extend the use of our method to other applications. 
+We also plan to enhance the learning capabilities of our method.
+%We are exploring building a language that enables users to intuitively design and build flexible FSTs similar to the ones presented in this paper.
+
 %We can also annotate the POR with the locations and 
@@
 %\bibliographystyle{naaclhlt2010}
-\bibliographystyle{acl}
+%\clearpage
+\bibliographystyle{splncs03}
+
 {\small \bibliography{fzAr}}
+
 \nocite{completeBibleGenealogy:Online}
+
 \end{document}

### 2011-11-24 ATSarf/src/build/functions.cpp

@@ -245,12 +245,12 @@ int insert_buckwalter()
 					assert (insert_category(resCat,(rule[j]==AA?PREFIX:SUFFIX),source_id,false));
 				}
 			}
-#ifdef ONLY_SUFFIXES
+//#ifdef ONLY_SUFFIXES
 			else {
 				assert (insert_category(cat1,(rule[j]==AB ||rule[j]==AC ?PREFIX:STEM),source_id,false)); //if not AB or AC => BC
 				assert (insert_category(cat2,(rule[j]==AC ||rule[j]==BC?SUFFIX:STEM),source_id,false));  ////if not AC or BC => AB
 			}
-#endif
+//#endif
 			if (insert_compatibility_rules(rule[j],cat1,cat2,resCat,source_id)<0)
 			{
 				out<<"Error at line "<<line_num<<": '"<<line<<"'\n";

### 2011-11-24 ATSarf/src/case/biographyGraphUtilities.cpp

@@ -1,0 +1,9 @@
+#include "biographyGraphUtilities.h"
+#include "graph.h"
+
+bool isRealNarrator(NarratorGraph * graph, Narrator * n,Biography::NarratorNodeList & list) {
+	RealNarratorAction v(list);
+	v.resetFound();
+	graph->performActionToAllCorrespondingNodes(n,v);
+	return v.isFound();
+}

### 2011-11-24 ATSarf/src/case/graph.cpp

@@ -214,7 +214,7 @@ int deserializeGraph(QString fileName,ATMProgressIFC * prg) {
 	graph->serialize(fileStream2);
 	file2.close();
 	delete graph;
-#elif 0
+#elif !defined(SEGMENT_AFTER_PROCESSING_ALL_BIOGRAPHY) || defined(BIOGRAPHY_SEGMENT)
 	biographies(graph);
 #else
 	localizedDisplay(graph);
@@ -285,13 +285,7 @@ int calculateStatisticsOrAnotate(ChainsContainer &generatedChains, NarratorGraph
 		sel.setGraph(localGraph);
 		outputList.append(sel);
 	}
-	HadithInterAnnotatorAgreement h(text,fileName,generatedGraph,outputList
-			#ifdef SUBMISSION
-				,true
-			#else
-				,false
-			#endif
-									);
+	HadithInterAnnotatorAgreement h(text,fileName,generatedGraph,outputList, true);
 	if (h.calculateStatisticsOrAnotate()==0)
 		h.displayStatistics();
 

### 2011-11-24 ATSarf/src/case/graph.h

@@ -2089,7 +2089,7 @@ inline int test_GraphFunctionalities(ChainsContainer &chains, ATMProgressIFC *pr
 		graph->setFileName(text,fileName);
 	}
 	QString allName=fileName;
-#ifndef SUBMISSION
+#if !defined(SUBMISSION) || defined(WRITE_POR)
 	QFile file(fileName.remove(".txt")+".por");
 	file.remove();
 	if (!file.open(QIODevice::ReadWrite))

### 2011-11-24 ATSarf/src/case/hadithInterAnnotatorAgreement.cpp

@@ -323,7 +323,7 @@ int hadith_agreement(QString input_str,ATMProgressIFC *prg) {
 	HadithInterAnnotatorAgreement::OutputDataList tags2;
 	HadithDagGraph * graph2=new HadithDagGraph();
 	h2.readAnnotation(newFileName,tags2,graph2);
-	HadithInterAnnotatorAgreement h(text,fileName,graph2,tags2);
+	HadithInterAnnotatorAgreement h(text,fileName,graph2,tags2,false);
 	h.calculateStatisticsOrAnotate();
 	h.displayStatistics();
 	return 0;

### 2011-11-24 ATSarf/src/case/suffix_verification.cpp

@@ -143,8 +143,7 @@ private:
 	QTextStream * dout;
 	#define d_out *(dout)
 
-	void traverse(node * n,QString affix,QString raw_data, QString description, QString POS)
-	{
+	void traverse(node * n,QString affix,QString raw_data, QString description, QString POS) {
 		if (!n->isLetterNode()) {
 			result_node * r=(result_node* )n;
 			for (int i=0; i<r->raw_datas.size();i++) {
@@ -159,6 +158,8 @@ private:
 					if (type==SUFFIX){
 						bool r=isReverseDirection(d);
 						//qDebug()<<raw_data<<" "<<r;
+						if (description[0]=='[' && description.size()>0 && description[description.size()-1]==']' && !added_desc.isEmpty())
+							description="";
 						if (description.contains("%1"))
 							desc=description.arg(added_desc+(added_desc=="" || description=="%1"?"":" "));
 						else {

### 2011-11-24 ATSarf/src/common/tree.cpp

@@ -41,6 +41,10 @@ inline bool isAcceptState(item_types type,long cat_r_id) {
 			}
 		}
 	} else {
+		QString s=database_info.comp_rules->getCategoryName(cat_r_id);
+		/*qDebug()<<s<<"\t"<<cat_r_id;
+		if (cat_r_id==168)
+			qDebug()<<"\t["<<s<<"]";*/
 		Retrieve_Template existACcheck("compatibility_rules","COUNT(*)",QString("category_id2=%1 AND type=%2").arg(cat_r_id).arg((int)AC));
 		if (existACcheck.retrieve() && existACcheck.get(0).toInt()>0) {
 			Retrieve_Template existBCcheck("compatibility_rules","COUNT(*)",QString("category_id2=%1 AND type=%2").arg(cat_r_id).arg((int)BC));
@@ -461,7 +465,7 @@ int tree::build_affix_tree(item_types type)
 			bool isAccept=isAcceptState(type,inf.category_id);
 			if (isAccept || hasCompatibleAffixes(type,inf.category_id)) {
 				node * next=addElement(name,affix_id1,inf.category_id,inf.category_id,isAccept,inf.raw_data,base);
-				if (type!=STEM)
+				//if (type!=STEM)
 					build_helper(type,inf.category_id,6-name.length(),next);
 			}
 		}

### 2011-11-24 ATSarf/src/sarf/stemmer.cpp

@@ -207,28 +207,32 @@ bool Stemmer::on_match()
 		out<<"]";
 		out <<")-";
 	}
-	if (called_everything || type==SUFFIX)
-	{
+	if (called_everything || type==SUFFIX) {
 		QString later_part="";
 		out<< "-(";
 		count=0;
-		for (int i=0;i<suffix_infos->count();i++)
-		{
+		for (int i=0;i<suffix_infos->size();i++) {
 			//qDebug()<< "{"<<suffix_infos->at(i).POS<<"}";
+			QString desc=suffix_infos->at(i).description();
 			if (count>0)
-					out << " ";//" + ";
+				out << " ";//" + ";
+			else {// (count==0)
+				if (suffix_infos->size()>1 && desc[0]=='[' && desc.size()>0 && desc[desc.size()-1]==']') {
+					desc="";
+				}
+			}
 			count++;
 			if (later_part=="" && suffix_infos->count()>i+1 && isReverseDirection(suffix_infos->at(i).abstract_categories)) {
-				later_part=suffix_infos->at(i).description();
+				later_part=desc;
 				//out<< "{"<<suffix_infos->at(i).POS<<"}";
 				count =0;
 				continue;
 			} else {
 				later_part="";
 			}
-			if (suffix_infos->at(i).description().isEmpty())
+			if (desc.isEmpty())
 				count=0;
-			out<</*Suffix->sub_positionsOFCurrentMatch[i]<<" "<<*/ suffix_infos->at(i).description()<<later_part;
+			out<</*Suffix->sub_positionsOFCurrentMatch[i]<<" "<<*/ desc<<later_part;
 			//out<< "{"<<suffix_infos->at(i).POS<<"}";
 		}
 		out <<")";
@@ -236,10 +240,10 @@ bool Stemmer::on_match()
 	if (called_everything)
 	{
 		QString word;
-		for (int i=0;i<prefix_infos->count();i++)
+		for (int i=0;i<prefix_infos->size();i++)
 			word.append(prefix_infos->at(i).raw_data);
 		word.append(stem_info->raw_data);
-		for (int i=0;i<suffix_infos->count();i++)
+		for (int i=0;i<suffix_infos->size();i++)
 			word.append(suffix_infos->at(i).raw_data);
 		//QString suff;
 		//for (int i=0;i<suffix_infos->count();i++)

### 2012-07-31 tex/Manual.tex
Manual.tex file added

\documentclass{article}
\usepackage{graphicx}
%\usepackage{arabtex}

\title{Stemmer Tool Manual}
\author{Ameen Jaber}
\usepackage{float}
\usepackage{arabtex}
\usepackage{fancyhdr} % Custom headers and footers
\usepackage{fancyvrb}
\pagestyle{fancyplain} % Makes all pages in the document conform to the custom headers and footers
\fancyhead{} % No page header - if you want one, create it in the same way as the footers below
\fancyfoot[L]{} % Empty left footer
\fancyfoot[C]{} % Empty center footer
\fancyfoot[R]{\thepage} % Page numbering for right footer
\renewcommand{\headrulewidth}{0pt} % Remove header underlines
\renewcommand{\footrulewidth}{0pt} % Remove footer underlines
\setlength{\headheight}{12pt} % Customize the height of the header
\usepackage{hyperref}

%\usepackage{utf8}
%\fullvocalize
%\transtrue
%\arabtrue

\setlength\parindent{0pt}
%----------------------------------------------------------------------------------------
%	TITLE SECTION
%----------------------------------------------------------------------------------------

\newcommand{\horrule}[1]{\rule{\linewidth}{#1}} % Create horizontal rule command with 1 argument of height

\title{	
\normalfont \normalsize 
\textsc{American University of Beirut} \\ [25pt]
\horrule{0.5pt} \\[0.4cm] % Thin top horizontal rule
\huge AUB Sarf\\
Arabic Morphological Analyzer\\
User Guide
\horrule{2pt} \\[0.5cm] % Thick bottom horizontal rule
}

\author{Sarf Team: Jad Makhlouta, Hamza Harkous, and Ameen Jaber\\Advisor: Dr. Fadi Zaraket}

\date{\normalsize\today}

\begin{document}

\maketitle

\newpage

\section{Introduction}
\paragraph{}
Sarf is an Arabic Morphological Analyzer. Morphology is the analysis of the structure of a word into its morphemes identified as stems and affixes, including prefixes and suffixes. Moreover, types are associated with those morphemes such as gloss and part of speech (POS) tags. Sarf takes a word such as ``\RL{wyAkolohA}'' and returns its morphological solutions. Figure~\ref{tab:SarfSample} below shows a sample output of Sarf's Analysis.

%\begin{figure}[H]
%\centering
%  \includegraphics[width=4in]{Sample.png}
%  \caption[Sample morphological solution of ``\RL{wyAkolohA}'']
%   {Sample morphological solution of ``\RL{wyAkolohA}''}
%   \label{fig:Sample}
%\end{figure}
%For more information on the morphological features of Sarf, please refer to {\bf PAPER}.

\begin{table}[H]
\centering
\begin{tabular}{ | lcl | }
\hline
\textbf{ALTERNATIVE}:& \fullvocalize\RL{waya'kolhA} & \\
\textbf{PREFIX:}\novocalize\RL{w} & \vocalize\RL{wa}	& \textbf{"[part.]"	wa/PART}\\
\textbf{PREFIX:}\novocalize\RL{y} & \vocalize\RL{ya}	& \textbf{"he/it"	ya/IV3MS}\\
\textbf{STEM:}\novocalize\RL{'kl}  & \fullvocalize\RL{'kol} & \textbf{"eat/consume"}\\
\textbf{$>$okul/VERB\_IMPERFECT} & & \\
\textbf{SUFFIX:} & \textbf{$^{o}$} &   \textbf{"[jus.]"	o/IVSUFF\_MOOD:J}\\
\textbf{SUFFIX:}\novocalize\RL{hA}& \vocalize\RL{hA}	& \textbf{"it/them/her"}\\
\textbf{hA/IVSUFF\_DO:3FS} & & \\
\hline
\end{tabular}
\caption{Sample morphological solution of ``\fullvocalize\RL{waya'kolhA}''}
\label{tab:SarfSample}
\end{table}

\paragraph{}
The rest of this document describes how to get the source code of Sarf, build it, and use it through three simple examples.

\section{Get the Source Code}
\label{sec:sourcecode}

Sarf is an open source tool. The code is currently hosted on google code under the project name ``atmine''. To download the code, you can use the svn interface:\\
\begin{verbatim}
svn checkout http://atmine.googlecode.com/svn/trunk/ atmine-read-only.
\end{verbatim}
Preferablly, you should be able to get a stable version of Sarf from the download section under:\\
\begin{verbatim}
http://code.google.com/p/atmine/downloads/list
\end{verbatim}
\paragraph{}
From the following link, download a zipped file named ``ATSarf-0.1.tar.bz2'' which contains all the files required to run the morphological analyzer along with all the applications implemented based on it. After downloading the compressed folder, extract it and you should have a folder named ``ATSarf'', which is the root folder of the Sarf.\\
You also need the encoded lexicon in the ``Bama-Public.zip'' file to be extracted in your running directory. This zip file is found in the original compressed file introduced above or in ``ATSarf/bin'' directory if you are using svn to download the code.

\section{Build the Tool}
\subsection{Dependencies}
In order to be able to build and run Sarf, the following dependencies must be installed:
\begin{itemize}
\item mysql-client
\item mysql-server
\item libmysqlcppconn
\item libmysqlcppconn-dev
%\item subversion
\item qt-sdk
%\item qtcreator
\item g++
%\item graphviz
\end{itemize}
The above tools can be downloaded and installed using the ``Synaptic Package Manager'' in ubuntu systems or through the ``apt-get'' command. You can use similar commands such as yum in other linux distributions.
\subsubsection{datrie}
You also need to download the double-array trie library called ``datrie''. Datrie can be downloaded from the following link:\\
\url{http://linux.thai.net/~thep/datrie/datrie.html#Download }\\
After downloading the trie structure to a destination you specify, go to {\bf datrie/libdatrie-0.2.3} and apply the following commands to configure and build.
\begin{enumerate}
\item ./configure
\item make
\end{enumerate}
Note that the name of the folder ``libdatrie-0.2.3'' includes the version of the library downloaded. So you could have a number in your download referring to a different version, which shouldn't cause any problem.

\subsection{links}
The directory ``ATSarf/third/'' in the extracted Sarf code should contain links to the dependencies as follows.

To create the links, you can use the ``ln'' command as follows.\\
\begin{verbatim}
ln -s TARGET LINK_NAME
\end{verbatim}
where TARGET is the dependency we want to point to and LINK\_NAME is the name of the link to be created.

Given that you put the datrie file in the same directory as the root folder ``ATSarf'', you can set the link to it as follows:
\begin{verbatim}
ln -s ../datrie/libdatrie-0.2.3/ datrie
\end{verbatim}

Moreover, you need to set a link to the Qt library as well using the same procedure used for the preceding link. 
The following command should do the trick if you are using an Ubuntu 10.10 version of linux:
\begin{verbatim}
ln -s /usr/share/qt4 qt
\end{verbatim}
Otherwis, you'll have to find the qt4 files on your system and create a link for them.

\subsection{Build Database}

\paragraph{}
In case you downloaded the tool from google code using the svn interface, you can either use the files found in the compressed folder ``Bama-Public.tar.bz2'' as instructed in section~\ref{sec:sourcecode} or you'll have to build the database.\\
In order to build the database, open a terminal and execute the following comands:

\begin{enumerate}
\item sudo mysql -u root
\item create database atm;
\item exit;
\item cd atmine/src/sql design
\item sudo mysql atm $<$ atm\_filled.sql
\end{enumerate}

\paragraph{}
The first command opens ``mysql'' database with root access where the password is empty. In case you had different settings for ``mysql'', you'll have to insert them. Inside ``mysql'', use the second command to create the database required, then command 3 to close it.\\
Now move to the ``atmine/src/sql design'' directory as shown in line 4, then execute command 5. If everything works normal, the database should be created and filled by now.

\subsection{Build}
Now we are ready to build Sarf.
Go to the ``ATSarf/'' directory and build the tool using the following command:
\begin{verbatim}
make
\end{verbatim}

\subsubsection{Windows-cygwin}

To build and run Sarf on windows, you need to have cygwin installed along with the packages shown previously. Also, you need to install x11 which is the windowing server of cygwin.
%{\bf TODO: Explain how to do this on windows}

\section{Run Sarf}

\subsection{Try existing applications}
Now that you have built Sarf, you can try existing applications that use it. To do that launch the GUI from ``ATSarf/bin'' by double clicking the executable output file called ``ATSarf''.
To test Sarf, first press the button with the label ``fill'', which fills the structures required for analysis. Then, you insert a word in the textbox present to the right of the label ``input'' in the upper left of the GUI and press ``GO!''. You can see the result in the textbox present under the label ``Output''.

The GUI includes other applications based on the arabic morphological analyzer such as Hadith Analysis that extracts the so called ``matn'' and ``sanad'' automatically and builds a combined gragh out of it.

%{\bf TODO: Describe how to try them with sample files}

\subsection{Build your first Simple Sarf application - Hello World Example}
\label{subsec:hwex}

Other than using the provided GUI, you can create you own program based on the libraries to either use Sarf functionality as it is or extend it.

\paragraph{}
The following example explains the steps of using the basic function of the morphological analyzer through a simple program. Hence, we will build a simple stemmer that takes an Arabic word and returns the different morphological solutions similar to the example shown in the introduction of the manual.

\subsubsection{Default}
\label{subsubsec:default}

\paragraph{}
We start by creating a folder in ``ATSarf/src'' and name it ``ex0''. Inside this folder, we create a file and call it ``ex0main.cpp'' which implements the main of our program. Avoid naming your files with existing file names in the code of Sarf. Inside the file ``ex0main.cpp'', create a function called ``stemmerExampleDefault()'' which reads the input, triggers the tool, and writes the output solution to the terminal along with any errors if present. This function returns an integer of 0 value if successful, else -1.

\begin{Verbatim}[numbers=left]
#include <iostream>
#include <QFile>
#include <sarf.h>
#include <stemmer.h>

int stemmerExampleDefault() {
	//code goes here
\end{Verbatim}

\paragraph{}
The header file ``sarf.h'' in the code above (line 3) is included in order to use the ``Sarf'' class and its functions. This class is used to initialize Sarf tool, customize it, and terminate it when done. Also, the header ``stemmer.h'' (line 4) is included to use ``Stemmer'' which is the main class of Sarf tool to be used.. Throughout this example, We will explain how to perform the following tasks:

\begin{itemize}
\item Delcare {\em sarf}  an instance of the Sarf class
\item Call {sarf.start } method: initializes the tool
\item Process the input
\item Call {sarf.exit } method: Exit the tool
\end{itemize}

\paragraph{}
\begin{Verbatim}[numbers=left,firstnumber=8]
    Sarf srf;
    bool all_set = srf.start();

    Sarf::use(&srf);

    if(!all_set) {
        error<<"Can't Set up Project";
        return -1;
    }
    else {
        cout<<"All Set"<<endl;
    }
\end{Verbatim}

\paragraph{}
Inside the ``stemmerExampleDefault'' function, we start by declaring an instance of the class Sarf as shown in the code above (line 8). Then, we call call the ``start'' function at line 9 to initialize the tool by reading the required data from the lexicon databases and initialize internal variables. The method start returns a boolean of true value if successful, else false.\\

By calling start without parameters, the output will be written on the terminal, both tool output and any error that might be encountered.

\paragraph{}
After this step, we call the static ``use'' method of Sarf (line 10) and pass a reference to the declared Sarf instance as a parameter. This is used to set a global variable internally. The ``use'' function allows you to have more than 1 sarf instance.

\begin{Verbatim}[numbers=left,firstnumber=20]
    char filename[100];
    cout << "please enter a file name: " << endl;
    cin >> filename;

    QFile Ifile(filename);
    if (!Ifile.open(QIODevice::ReadOnly | QIODevice::Text)) {
       cerr << "error opening file." << endl;
       return -1;
    }

    QTextStream in(&Ifile);
    while (!in.atEnd()) {
        QString line = in.readLine();
        run_process(line);
    }
    srf.exit();
    return 0;
}
\end{Verbatim}

\paragraph{}
Line 21 above asks the user to enter the input file name and line 25 tests if opening the file is successful or not. Then in each iteration of the while loop at Line 31, we read a line from the 
input file (line 32) and pass it to the function ``run\_process'' declared inside the ex0main.cpp file.

\paragraph{}
Then,  we call the ``srf.exit'' (line 35) to exit the program gracefully and clean the sarf structure which deletes any open query along with the database. The last step in building this example is to implement the ``run\_Process'' function used in line 33.

\begin{Verbatim}[numbers=left]
void run_process(QString & input) {

    QStringList list = input.split(' ', QString::SkipEmptyParts);
    for(int i=0; i<list.size(); i++) {
        QString * inString = &(list[i]);
        Stemmer stemmer(inString,0);
        stemmer();
    }
}
\end{Verbatim}

\paragraph{}
The ``run\_process'' function shown above takes a reference to a QString input representing a line of words, and splits it based on the space character as shown in line 3. Then, it iterates over all the words in the QStringList ``list'', and passes them one by one to the constructor of ``Stemmer'' (line 6).

\paragraph{}
The ``Stemmer'' class is the main structure in Sarf. The result of this morphological analysis is printed to the console as explained previously. Thus, an instance of the Stemmer class is declared in line 6 taking the following parameters:
\begin{itemize}
\item A pointer of type QString pointing to the input word to analyze.
\item A starting index for the input word analysis to start from.
\end{itemize}
In line 7, we call the bracket operator which triggers the tool to start the analysis.

\begin{Verbatim}[numbers=left]
int ex0_main(int argc, char *argv[]) {

    int test;

    test = stemmerExampleDefault();
    if(!test) {
        cout<<"The example without interface is successful\n";
    }
    else {
        cout<<"The example without interface failed\n";
    }
\end{Verbatim}

\paragraph{}
The last step is to create the main in which we call the ``stemmerExampleDefault()'', where the code is shown above. Name your main function ``ex0\_main'' to respect the linking conventions in ATSarf, and we will deal with this issue later in the configuration step.

\begin{center}
\textbf{Configuration}
\end{center}

\paragraph{}
Now that we are done with the coding part, we still need to configure the environment for compilation. Thus, the first step is to add a link to the makefile template which has rules to build an archive (lib\%.a) from a directory. The makefile is called ``archiveMakefile'' found  in ``ATSarf/scripts''. Hence, in order to create the link, we use the following command line given that we are in the directory of the example:

\begin{Verbatim}
ln -s ../../scripts/archiveMakefile makefile
\end{Verbatim}

Next, we add a file in ``ATSarf/bin'' called ex0.cpp and inside this file we add the following lines of code:
\begin{Verbatim}[numbers=left]
extern int ex0_main(int argc, char ** argv);

int
main(int argc, char ** argv) {
  return ex0_main(argc, argv);
}
\end{Verbatim}

\paragraph{}
In the lines of code shown above, the extern keyword in line 1 declares ``ex0\_main'' and specifies that it has external linkage. Note that ``ex0\_main'' is the function that we used in place of the main in the file ``ex0main.cpp'' to respect the linking conventions of the project. Then, in lines 3-6 we write our main that calles the ``ex0\_main'' function previously declared.

\paragraph{}
The last step left is to modify the makefile found in ``ATSarf/bin'' directory to create an executable program out of the libraries created in compilation. Thus, we add the following rules to the makefile to create the ex0 executable.

\begin{Verbatim}[numbers=left]
EX0_LIBDIR = ../lib/libex0.a ../lib/libbuild.a ../lib/libcache.a
 ../lib/libcommon.a ../lib/liblog.a ../lib/libsarf.a ../lib/libsql.a
  ../lib/libutil.a
EX0_LIBS_LINKFLAGS=-L ../lib -lex0 -lbuild -lcache -lcommon -llog
 -lsarf -lsql -lutil

ex0: $(EX0_LIBDIR)
	$(CXX) -o $@ ex0.cpp $(LDDFLAGS) $(EX0_LIBS_LINKFLAGS) \
		$(EX0_LIBS_LINKFLAGS) \
		$(QT_LIBS_LDDFLAGS) $(DATRIE_LIB_LDDFLAGS)
\end{Verbatim}

\paragraph{}
In line 1 above, we define the libraries from the lib directory that our program depends on. Also, in line 4 we specify the flags referring to those libraries for dependency resolving. Last in line 7, we introduce the target and dependecies along with the compilation command. The following can be used as a template for any other program.

\paragraph{}
Now that all the steps of implementation and configuration are done, we only need to compile the project. For that purpose, move to the directory ``ATSarf'' in the terminal and run the makefile through the command make. If all the steps are done correctly, you should have an executable with the name ``ex0'' in the ``bin'' directory.

\subsubsection{Interface}
\label{subsubsec:interface}

\paragraph{}
Sarf can be used with an interface which states a destination for the output analysis, error, and implementation for a structure internally used showing the progress of internal work along with other features.

\paragraph{}
Similar to the Default example explained above, we start by creating the program folder ``exo'' in ``ATSarf/src'' directory. We introduce an additional header file called ``myprogressifc.h'' which includes the implementation of the class ``MYProgressIFC''. This class implements functions that are internally used in the tool, and returns data that the user might find useful about the tool progress.

\label{vrbtm:myprogressifc}
\begin{Verbatim}[numbers=left]
class MyProgressIFC : public ATMProgressIFC {

public:
    virtual void report(int value) {
        cout<<"Progress is "<<value<<'.'<<endl;
    }

    virtual void startTaggingText(QString & text) {
    }

    virtual void tag(int start, int length,QColor color, bool textcolor=true) {
    }

    virtual void finishTaggingText() {
    }

    virtual void setCurrentAction(const QString & s) {
    }

    virtual void resetActionDisplay() {
    }

    virtual QString getFileName() {
        return "";
    }
};
\end{Verbatim}

\paragraph{}
Thus, we create a file called ``myprogressifc.h'' in the example folder ``ex0''. Inside of it, we include the implementation of the class as shown above, where ``MyProgressIFC'' inherets from a predefined structure called ``ATMProgressIFC''. The class inhereted from is abstract, thus all its functions should be implemented even if not used.

\paragraph{}
Some of the functions of the abstract class ``ATMProgressIFC'' are related to other applications than the morphological analyzer, thus we don't need to discuss them at the moment. However, the relevant functions that could be of benefit are the following:

\begin{itemize}
\item report(int value): This function reports the value of progress at a scale from 0-100 at different steps of the initialization of the tool or execution. This value is being printed to the terminal in the code above lines 4-6.
\item setCurrentAction(const QString \& s): This function reports the current action being performed by the tool through the QString parameter. Usually, this parameter can refer to an action being performed in one of different structures found in the analyzer.
\item resetActionDisplay(): This function resets the action settings explained above. So in case we had a progress bar as is the case in a GUI, this function can be used to reset the progress bar.
\end{itemize}

\paragraph{}
The next steps are almost the same as the default example in \ref{subsubsec:default}. Thus, we create the file ``ex0main.cpp'' and proceed to implement the relevant functions. The main function added is ``stemmerExamplewithInterface()'' which is very similar to ``stemmerExampleDefault()''.

\begin{Verbatim}[numbers=left]
int stemmerExamplewithInterface() {

    QFile Ofile("output.txt");
    QFile Efile("error.txt");
    Ofile.open(QIODevice::WriteOnly);
    Efile.open(QIODevice::WriteOnly);
    MyProgressIFC * pIFC = new MyProgressIFC();

    Sarf srf;
    bool all_set = srf.start(&Ofile,&Efile, pIFC);

    Sarf::use(&srf);

    if(!all_set) {
        error<<"Can't Set up Project";
    }
    else {
        cout<<"All Set"<<endl;
    }

    char filename[100];
    cout << "please enter a file name: " << endl;
    cin >> filename;

    QFile Ifile(filename);
    if (!Ifile.open(QIODevice::ReadOnly | QIODevice::Text)) {
       cerr << "error opening file." << endl;
       return -1;
    }

    QTextStream in(&Ifile);
    while (!in.atEnd()) {
        QString line = in.readLine();
        run_process(line);
    }
    srf.exit();
    return 0;
}
\end{Verbatim}

\paragraph{}
As shown in the code above, the steps are exactly the same as the default function except for the parameters passed to the ``start'' function. In lines 3-6, we declare two files named ``output.txt'' for the analysis output and ``error.txt'' for any possible errors., then we set the properties of the files as write only. Then in line 7, we initialize a pointer pIFC of the previously defined class ``MYProgressIFC''.

\paragraph{}
The sarf instance is declared as before, however the start function now takes three variables as follows:
\begin{itemize}
\item QFile * \_out: This is a pointer to a QFile used to print the result in.
\item QFile * \_displayed\_error: This is a pointer to a QFile used to print any errors that could occur in.
\item ATMProgressIFC * pIFC: This is a pointer to a structure that inherets from ATMProgressIFC as we did in the above class ``MyProgressIFC''.
\end{itemize}

\paragraph{}
The rest of the function proceeds exactly same as the default case. Also, we can use the same ``run\_process'' function implemented previously, then call the function ``stemmerExamplewithInterface()'' in ``ex0\_main'' function.\\
The configuration steps are not required if we decide to include those modifications inside the same example, else the same configuration procedure explained in \ref{subsubsec:default} should be applied here.

\section{Examples}
\label{sec:examples}
\paragraph{}
This section explains the steps required to implement two examples examples in which the user can implement a specific functionality based on the Sarf. These steps can be generalized to implement any other functionality the user wants.

\subsection{Example-Verb Extraction}
\paragraph{}
This example extracts all the words from an input file that have a ``verb'' part of speech (POS) tag. In other words, the output is a list of all the words extracted from the input file that could be interpreted as a verb, since an arabic word can possess multiple meanings and POS.

\paragraph{}
The following example takes a specific case to implement, however you can search for other POS tags. The following is a list of those tags:
\begin{itemize}
\item VERB: regular verb.
\item VERB\_PERFECT: verb in perfect form.
\item NOUN: regular noun.
\item NOUN\_PROP: proper noun.
\item ADJ: adjective.
\item PREP: preposition.
\end{itemize}

\paragraph{}
We start by creating a folder for our program in the directory ``ATSarf/src'' and name it ``ex1''. In this folder, we will create the following three files:
\begin{itemize}
\item POSVerb.h
\item POSVerb.cpp
\item ex1main.cpp
\end{itemize}

\begin{Verbatim}[numbers=left]
#include<stemmer.h>
#include<QTextStream>

class POSVerb : public Stemmer
{
private:
    QString text;
    QString RelatedW;

public:
    POSVerb(QString * text);
    bool on_match();
};
\end{Verbatim}

\paragraph{}
In the file ``POSVerb.h'', we will create a class called ``POSVerb'' as shown in the code above. This class inherets from the class ``Stemmer'' which is the main structure used for the arabic morphological analysis. For that purpose, we add the include ``stemmer.h'' shown in line 1.\\

Inside the class defined, we add two private variables:
\begin{itemize}
\item text (line 7): String that holds the input text
\item RelatedW (line 8): String that holds the previous solutions which were accepted as verbs in order to prevent repitition since a single word can have different solutions from the analyzer.
\end{itemize}

Moreover, we define two public functions:
\begin{itemize}
\item POSVerb (line 11): This is the constructor of the POSVerb class and it takes a QString as the input text to be processed.
\item on\_match (line 12): This function is called by the tool upon finding a possible solution for an input word and is defined as a virtual function in the stemmer class. By implementing this method, we control the accepted solutions that meet our functionality required.
\end{itemize}

\begin{Verbatim}[numbers=left]
POSVerb::POSVerb(QString *text) : Stemmer(text,0)
{
    this->text = *text;
};
\end{Verbatim}

\paragraph{}
Now that we have defined our class along with the required functions and variables, we have to implement the functions in the ``POSVerb.cpp'' file.\\
The constructor takes the input text and sets its text variable to its value as shown in the code above (line 3). Also, it calls the constructor of the Stemmer class which takes the input text along with the starting index as its parameters. In our case, the starting index is 0.

\paragraph{}
The on\_match() function is triggered by the tool upon finding a solution for a word. A solution refers to a combination of prefix, stem , and suffix along with descriptions of those parts including the POS and gloss. In our program, we require the POS field to include a verb tag.

\begin{Verbatim}[numbers=left]
bool POSVerb::on_match()
{
    if(stem_info->POS.contains("VERB", Qt::CaseInsensitive) )
    {
        if (!(RelatedW.contains(this->text,Qt::CaseInsensitive)))
        {
            RelatedW += this->text;
            theSarf->out<<this->text<<endl;
        }
    }
    return true;
};
\end{Verbatim}

\paragraph{}
The implementation of this function is shown above. In line 3, we check if the stem's POS includes the ``VERB'' tag by checking the POS variable of the stem\_info structure.\\

stem\_info is a variable of the structure ``minimal\_item\_info'' which holds the solution details. For further details about the ``minimal\_item\_info structure'', refer to section\ref{sec:classes}.\\

If the tag is ``VERB'', we check in line 5 if this word is preiously caught as a correct solution by searching for it in the RelatedW variable. If not, we add it to the RelatedW variable and write it to a QTextStream called out. This variable is found in the global instance of the Sarf class introduced previously.

\paragraph{}
The on\_match function returns a boolean which is caught by the tool, if that boolean is true the tool will continue to find all the possible solutios till no more are found, else if false is returned the tool will stop.

\paragraph{}
Now, we need to build the file ``ex1main.cpp'' which is the same as the one explained in the ``Hello World Example''\ref{subsec:hwex}. The only difference is in the implementation of the ``run\_process'' where we should create an instance of the POSVerb class instead of Stemmer as shown in the code below line 6. Then, we trigger the tool by calling the bracket operator (line 7).

\begin{Verbatim}[numbers=left]
void run_process(QString & input) {

    QStringList list = input.split(' ', QString::SkipEmptyParts);
    for(int i=0; i<list.size(); i++) {
        QString * inString = &(list[i]);
        POSVerb posverb(inString);
	posverb();
    }
}
\end{Verbatim}

\paragraph{}
If we want to implement the example with an interface, we should include a class implementation similar to the one introduced in ``mprogressifc.h''\ref{vrbtm:myprogressifc} and pass it to the Sarf start function along with the output and error files. Moreover, the required headers should be included in the file ``ex1main.cpp'' such as ``POSVerb.h''.\\
With all the required files implemented, we prepare the example for compilation as explained before in example\ref{subsec:hwex}; configuration step.

\subsection{Example-Word related Extraction}
\paragraph{}
In this example, our aim is to extract every word in a text file input that is related in gloss
to a list of predefined english words. Hence, any word that has a solution with a gloss tag including one of the predefined words will be considered a valid solution and is extracted.

\paragraph{}
We need a structure similar to the one in the previous example and the configuration process is exactly the same. Hence, in this example we will only focus on implementing the functionality of the program and not the details which are previously explained. The main files that will be explained are the following:

\begin{itemize}
\item GlossSFR.h
\item GlossSFR.cpp
\end{itemize}

\begin{Verbatim}[numbers=left]
class GlossSFR : public Stemmer
{
private:
    QString text;
    QString RelatedW;
    QStringList _LIST;

public:
    GlossSFR(QString * text);
    bool on_match();
};
\end{Verbatim}

\paragraph{}
In ``GlossSFR.h'', we create the needed class. The code for that is shown above where we use similar functions and variables to the previous example. The variables defined are as follows:

\begin{itemize}
\item text (line 4): QString that holds the input word
\item RelatedW (line 5): QString that holds the previous solutions which were accepted in order to prevent repitition since a single word can have different solutions from the analyzer.
\item \_LIST (line 6): QStringList that holds the english words which we will search the gloss tag of each solution for.
\end{itemize}

\paragraph{}
As for the functions, we define two functions. The first of which is the constructor of the class and the second is the ``on\_match'' function explained earlier. Line 9 is the constructor of the class and it takes a pointer to the input word as a parameter. As for the ``on\_match'' function (line 10), we need to implement it to select the solutions that meet with our specifications.

\begin{Verbatim}[numbers=left]
GlossSFR::GlossSFR(QString *text) : Stemmer(text,0)
{
    this->text = *text;
     _LIST<<"flower"<<"nose"<<"smell"<<"rose";
};
\end{Verbatim}

\paragraph{}
Next, we need to implement the previously defined functions in the file ``GlossSFR.cpp''. The first function is the constructor where we set the internal text variable to the input and fill ``\_LIST'' with the gloss words we intend to look for in Sarf solutions. Similar to what is done before, we call the constructor of the Stemmer class as well passing the input string and starting index.

\begin{Verbatim}[numbers=left]
bool GlossSFR::on_match()
{
    bool RELATED=false;

    for(int i =0; i<_LIST.size(); i++)
    {
        QString _desc = stem_info->description();
        if(_desc.contains(_LIST[i],Qt::CaseInsensitive))
        {
            RELATED = true;
            break;
        }
    }
    bool Present = RelatedW.contains(this->text,Qt::CaseInsensitive);
    if(RELATED && !Present)
    {
        RelatedW += this->text;
        theSarf->out<<this->text<<'\n';
    }
    return true;
};
\end{Verbatim}

\paragraph{}
Moreover, the code for ``on\_match'' function is shown above. In the lines 5-13, we loop over the strings in ``\_LIST'' and check whether the description field in ``stem\_info'' contains that gloss. The description fields in ``stem\_info'' refers to the gloss field of the solution.\\
Thus, we extract the description field in line 7 and check in line 8 whether this field contains the string in ``\_LIST''.  If true, we set the boolean ``RELATED'' to true and exit the loop since we only care if the input is related to any of the base words.

\paragraph{}
Then, we check if the input word has been detected as a solution in previous matches by searching for it in the variable ``RelatedW''. If it is not found and the RELATED boolean is true, then we add the input word to RelatedW and out variables which are explained before.

\paragraph{}
With those two files implemented, we need to add ``ex2main.cpp'' then configure the program for compilation as previously done.

\section{Classes}
\label{sec:classes}

\paragraph{}
The following section explains a list of the important and useful structures and classes needed in order to use the tool.

\subsection{Stemmer}
\paragraph{}
This class implements the methods that extract the different morphological features of an arabic string input such as the stem, prefix, and suffix along with features for the stem such as the POS and Gloss.

\subsection{Prefix}
This class implements the methods that extract the prefix morpheme of the input arabic word along with the relevant POS and gloss tags.

\subsection{Stem}
This class implements the methods that extract the stem morpheme of the input arabic word along with the relevant POS and gloss tags. This machine is triggered after finding a correct possible prefix for the word which can vary from zero to multiple prefixes.

\subsection{Suffix}
This class implements the methods that extract the suffix morpheme of the input arabic word along with the relevant POS and gloss tags. This stage of the analysis is triggered after finding a possible prefix/stem solution. Based on this solution, the Suffix machine is triggered to detect the suffix.

\subsection{minimal\_item\_info}
\paragraph{}
The following defines a class type including the minimal information of a match which could be stem or an affix. This structure includes the following variables:
\begin{itemize}
\item QString desc: This includes the description of a matching prefix/stem/suffix which is the gloss.
\item QString raw\_data: This string holds the lexicon or stem of the input string.
\item QString POS: This string holds the part of speech of the match.
\item item\_types type: This variable specifies the type of the solution which can be PREFIX, STEM, or SUFFIX. ``item\_types'' is an enum so PREFIX has a 1 value, STEM has a 2 value, and SUFFIX has a 3 value.
\end{itemize}

%\section{Functions}

%\section{Important Global Variables}

%\section{Frequently Asked Questions}

\section{How to Get Help}
\paragraph{}
If you need more details about the code,you can find the documentation in ``ATSarf/doc/html'', and open it by double clicking on the file ``index.html''. We used doxygen to automatically generate the documentation of Sarf.

\paragraph{}
For any further questions, contact us at the following email: \textbf{aaj15@aub.edu.lb}

\end{document}

### 2012-08-24 tex/Manual.tex
Maunal Updated

@@
 \paragraph{}
-We start by creating a folder in ``ATSarf/src'' and name it ``ex0''.
+We start by creating a folder in ``ATSarf/apps'' and name it ``ex0''.
 Inside this folder, we create a file and call it ``ex0main.cpp'' which implements the main of our program.
@@
 \begin{Verbatim}[numbers=left]
-int ex0_main(int argc, char *argv[]) {
+int main(int argc, char *argv[]) {
 
@@
 The last step is to create the main in which we call the ``stemmerExampleDefault()'', where the code is shown above.
-Name your main function ``ex0\_main'' to respect the linking conventions in ATSarf, and we will deal with this issue later in the configuration step.
 
@@
 
-Next, we add a file in ``ATSarf/bin'' called ex0.cpp and inside this file we add the following lines of code:
+\paragraph{}
+We have to modify the makefile found in ``ATSarf/bin'' directory to create an executable program out of the libraries created in compilation.
+Thus, we add the following rules to the makefile to create the ex0 executable.
+
 \begin{Verbatim}[numbers=left]
-extern int ex0_main(int argc, char ** argv);
+EX0_LIBDIR = $(TOP)/lib/libex0.a $(ATSarf_LIBS)
+EX0_LIBS_LINKFLAGS=-L$(TOP)/lib -lex0 $(ATSarf_LIBS_LINKFLAGS4)
 
-int
-main(int argc, char ** argv) {
-  return ex0_main(argc, argv);
-}
+ex0: $(EX0_LIBDIR)
+	$(CXX) -o $@ $(LDDFLAGS) $(EX0_LIBS_LINKFLAGS) \
+		$(THIRD_LIBS_LINKFLAGS)
 \end{Verbatim}
@@
 \paragraph{}
-In the lines of code shown above, the extern keyword in line 1 declares ``ex0\_main'' and specifies that it has external linkage.
-Note that ``ex0\_main'' is the function that we used in place of the main in the file ``ex0main.cpp'' to respect the linking conventions of the project.
-Then, in lines 3-6 we write our main that calles the ``ex0\_main'' function previously declared.
+In line 1 above, we define the macro ``EX0\_LIBDIR'' which includes the libraries from the ``lib'' directory that our program depends on.
+We have to add our application archive library to this macro in addition to the ATSarf libraries included in the predefined macro ``ATSarf\_LIBS''.
+Also, in line 2 we specify the flags referring to those libraries for dependency resolving.
+Last in line 4, we introduce the target and dependecies along with the compilation command.
+The following can be used as a template for any other program.
+
+\begin{verbatim}
+opt 32 64 cyg ATSarf ex0: all
+\end{verbatim}
 
 \paragraph{}
-The last step left is to modify the makefile found in ``ATSarf/bin'' directory to create an executable program out of the libraries created in compilation.
-Thus, we add the following rules to the makefile to create the ex0 executable.
-
-\begin{Verbatim}[numbers=left]
-EX0_LIBDIR = ../lib/libex0.a ../lib/libbuild.a ../lib/libcache.a
- ../lib/libcommon.a ../lib/liblog.a ../lib/libsarf.a ../lib/libsql.a
-  ../lib/libutil.a
-EX0_LIBS_LINKFLAGS=-L ../lib -lex0 -lbuild -lcache -lcommon -llog
- -lsarf -lsql -lutil
-
-ex0: $(EX0_LIBDIR)
-	$(CXX) -o $@ ex0.cpp $(LDDFLAGS) $(EX0_LIBS_LINKFLAGS) \
-		$(EX0_LIBS_LINKFLAGS) \
-		$(QT_LIBS_LDDFLAGS) $(DATRIE_LIB_LDDFLAGS)
-\end{Verbatim}
-
-\paragraph{}
-In line 1 above, we define the libraries from the lib directory that our program depends on.
-Also, in line 4 we specify the flags referring to those libraries for dependency resolving.
-Last in line 7, we introduce the target and dependecies along with the compilation command.
-The following can be used as a template for any other program.
+Last in the makefile present in the directory ``ATSarf'', you have to add the name of your application to the last line as shown above.
+The name has to be the same as the taget name in the makefile added to the makefile in ``ATSarf/bin'' directory.
 
@@
 Now that all the steps of implementation and configuration are done, we only need to compile the project.
-For that purpose, move to the directory ``ATSarf'' in the terminal and run the makefile through the command make.
+Move to the directory ``ATSarf'' in the terminal and run the makefile through the command ``make ex0''.
 If all the steps are done correctly, you should have an executable with the name ``ex0'' in the ``bin'' directory.
@@
 \paragraph{}
-Similar to the Default example explained above, we start by creating the program folder ``exo'' in ``ATSarf/src'' directory.
+Similar to the Default example explained above, we start by creating the program folder ``exo'' in ``ATSarf/apps'' directory.
 We introduce an additional header file called ``myprogressifc.h'' which includes the implementation of the class ``MYProgressIFC''.
@@
 The rest of the function proceeds exactly same as the default case.
-Also, we can use the same ``run\_process'' function implemented previously, then call the function ``stemmerExamplewithInterface()'' in ``ex0\_main'' function.\\
+Also, we can use the same ``run\_process'' function implemented previously, then call the function ``stemmerExamplewithInterface()'' in the main.\\
 The configuration steps are not required if we decide to include those modifications inside the same example, else the same configuration procedure explained in \ref{subsubsec:default} should be applied here.
@@
 \paragraph{}
-We start by creating a folder for our program in the directory ``ATSarf/src'' and name it ``ex1''. In this folder, we will create the following three files:
+We start by creating a folder for our program in the directory ``ATSarf/apps'' and name it ``ex1''. In this folder, we will create the following three files:
 \begin{itemize}

### 2012-08-27 trunk/tex/Manual.tex
updated Manual

--- 2012-08-24 tex/Manual.tex
@@
 \item libmysqlcppconn-dev
+%\item libqjson-dev
 %\item subversion
@@
 
+\subsubsection{QJson}
+QJson is a qt-based library that maps JSON data to QVariant objects: JSON arrays will be mapped to QVariantList instances, while JSON objects will be mapped to QVariantMap.
+This library can be downloaded from the following link:\\
+\url{http://sourceforge.net/projects/qjson/files/}\\
+After downloading QJson to the desired destination, follow the instructions found in the ``INSTALL'' file found in the root directory ``qjson'' to install the library.
+
 \subsection{links}
@@
 
-Given that you put the datrie file in the same directory as the root folder ``ATSarf'', you can set the link to it as follows:
+Given that you put the datrie file in the same directory as the root folder ``ATSarf'' and you are in the ``third'' directory, you can set the link to it as follows:
 \begin{verbatim}
-ln -s ../datrie/libdatrie-0.2.3/ datrie
+ln -s ../../datrie/libdatrie-0.2.3/ datrie
+\end{verbatim}
+
+Similarly, given that you add the QJson library to the same directory as the root folder ``ATSarf'', add a link to it using the following command:
+\begin{verbatim}
+ln -s ../../qjson/ qjson
 \end{verbatim}

### 2012-09-02 tex/Manual.tex
manual updated

@@
 \usepackage{hyperref}
+\hypersetup{%
+    pdfborder = {0 0 0}
+}
+\bibliographystyle{plain}
 
@@
 
+\tableofcontents
+
+\newpage
+
 \section{Introduction}
@@
 Moreover, types are associated with those morphemes such as gloss and part of speech (POS) tags.
-Sarf takes a word such as ``\RL{wyAkolohA}'' and returns its morphological solutions.
+Sarf takes a word such as ``\fullvocalize\RL{waya'kolhA}'' and returns its morphological solutions.
 Figure~\ref{tab:SarfSample} below shows a sample output of Sarf's Analysis.
@@
 \end{verbatim}
-Preferablly, you should be able to get a stable version of Sarf from the download section under:\\
+Preferablly, get a stable version of Sarf from the download section under:\\
 \begin{verbatim}
@@
 \paragraph{}
-From the following link, download a zipped file named ``ATSarf-0.1.tar.bz2'' which contains all the files required to run the morphological analyzer along with all the applications implemented based on it.
-After downloading the compressed folder, extract it and you should have a folder named ``ATSarf'', which is the root folder of the Sarf.\\
-You also need the encoded lexicon in the ``Bama-Public.zip'' file to be extracted in your running directory.
-This zip file is found in the original compressed file introduced above or in ``ATSarf/bin'' directory if you are using svn to download the code.
+From the link above, download a zipped file named ``ATSarf-0.1.tar.bz2'' which contains all the files required to run the morphological analyzer along with all the applications implemented based on it.
+After downloading the compressed folder, extract it and you should have a folder named ``ATSarf-0.1''.
+Inside this folder, you'll find ``ATSarf''; the top folder of Sarf.\\
+You also need the encoded lexicon in the zip file ``Bama-Public.zip''.
+This zip file is found in ``ATsarf-0.1'' introduced above or in ``ATSarf/bin'' directory if you are using svn to download the code.
+You have to extract this file in your running directory in order for the tool to work correctly.
 
 \section{Build the Tool}
+The following are the steps required to build Sarf on a linux machine.
+
 \subsection{Dependencies}
+\label{subsec:dependencies}
 In order to be able to build and run Sarf, the following dependencies must be installed:
@@
 \item mysql-server
-\item libmysqlcppconn
+\item libmysqlcppconn4
 \item libmysqlcppconn-dev
+%\item libqjson-dev
 %\item subversion
@@
 \end{enumerate}
-Note that the name of the folder ``libdatrie-0.2.3'' includes the version of the library downloaded.
-So you could have a number in your download referring to a different version, which shouldn't cause any problem.
+Note that the name of the folder ``libdatrie-0.2.3'' includes the version of the library downloaded ``0.2.3''.If you download another version of this library, it will have a different name referring to a different version.
+Using any of the other versions of datrie shouldn't lead to anu problem, however it is preffered to use the same version as the one introduced above.
+
+\subsubsection{QJson- Only Needed by Tagger Application}
+QJson is a qt-based library that maps JSON data to QVariant objects: JSON arrays will be mapped to QVariantList instances, while JSON objects will be mapped to QVariantMap.
+This library can be downloaded from the following link:\\
+\url{http://sourceforge.net/projects/qjson/files/}\\
+After downloading QJson to the desired destination, follow the instructions in the ``INSTALL'' file found in the top directory ``qjson'' to install the library.
 
 \subsection{links}
-The directory ``ATSarf/third/'' in the extracted Sarf code should contain links to the dependencies as follows.
+In the directory ``ATSarf/third'', we have to create symbolic links to target dependencies.
+Those dependencies are third party software such as datrie and qjson downloaded and installed in subsection~\ref{subsec:dependencies}.\\
 
@@
 \end{verbatim}
-where TARGET is the dependency we want to point to and LINK\_NAME is the name of the link to be created.
+where TARGET is the dependency we want to point to and LINK\_NAME is the name of the link to be created.\\
 
-Given that you put the datrie file in the same directory as the root folder ``ATSarf'', you can set the link to it as follows:
+\subsubsection{datrie link}
+Given that you put the datrie file in the same directory as the top folder ``ATSarf'' and you are in the ``third'' directory, you can use the following command line to set the link:
 \begin{verbatim}
-ln -s ../datrie/libdatrie-0.2.3/ datrie
+ln -s ../../datrie/libdatrie-0.2.3/ datrie
 \end{verbatim}
 
+\subsubsection{qjson link}
+Similarly, given that you add the QJson library to the same directory as the top folder ``ATSarf'', add a link to it using the following command:
+\begin{verbatim}
+ln -s ../../qjson/ qjson
+\end{verbatim}
+
+Alternatively, you can install qjson using the command line:
+\begin{verbatim}
+apt-get install libqjson-dev
+\end{verbatim}
+
+Given that you are in third directory, create the directories qjson/build/lib as follows :
+\begin{verbatim}
+mkdir -p qjson/build/lib
+\end{verbatim}
+
+The last step would be to locate the shared library of qjson in your system and create a link to it as follows:
+\begin{verbatim}
+cd qjson/build/lib
+ln -f QJSONLIB_PATH/libqjson.so
+\end{verbatim}
+where QJSONLIB\_PATH is the path of the qjson library in your system.\\
+
+\subsubsection{Qt link}
 Moreover, you need to set a link to the Qt library as well using the same procedure used for the preceding link. 
@@
 \paragraph{}
-The first command opens ``mysql'' database with root access where the password is empty.
+The first command opens ``mysql'' database with mysql root access where the password is empty.
 In case you had different settings for ``mysql'', you'll have to insert them.
@@
 \subsection{Try existing applications}
-Now that you have built Sarf, you can try existing applications that use it. To do that launch the GUI from ``ATSarf/bin'' by double clicking the executable output file called ``ATSarf''.
+Now that you have built Sarf, you can try existing applications that use it. To do that launch the GUI from ``ATSarf/bin'' by double clicking the executable output file called ``ATSarfGui''.
+
+\subsubsection{Morphology - Sarf}
 To test Sarf, first press the button with the label ``fill'', which fills the structures required for analysis.
-Then, you insert a word in the textbox present to the right of the label ``input'' in the upper left of the GUI and press ``GO!''.
-You can see the result in the textbox present under the label ``Output''.
+Then, type an arabic word in the textbox present to the right of the label ``input'' in the upper left of the GUI and press ``GO!''.
+You can see the results in the output textbox.\\
 
-The GUI includes other applications based on the arabic morphological analyzer such as Hadith Analysis that extracts the so called ``matn'' and ``sanad'' automatically and builds a combined gragh out of it.
+\subsubsection{Hadith}
+The GUI includes other applications based on the arabic morphological analyzer (Sarf) such as Hadith Analysis that extracts the so called ``matn'' and ``sanad'' automatically and builds a combined graph out of it.
+To test this application follow the following steps:
+\begin{enumerate}
+\item Press Fill
+\item Select Hadith Radio Button
+\item Press Browse and go to ATSarf/tests
+\item Select the file hadith/kafi1.txt
+\item Press Go!
+\end{enumerate}
 
+At this stage, the output will be the input hadith text shown in the textbox to the right with tagged narrators referring to what is called ``sanad''.
+Also, the start of each hadith and end of its sanad are shown in the left textbox.\\
+
+Moreover, you can pass biography text file to the ``biographies view'' that opens upon finishing the previous step. To do so, follow these steps:
+\begin{enumerate}
+\item Press Browse and go to ATSarf/tests/Kafi
+\item Select khoei.txt
+\item Press ``Parse Biographies''
+\end{enumerate}
+
+% narrator/biography pair
+The output will be the list of narrators from the hadith with the corresponding biographies. For further explanation about this application, refer to \cite{zaraket2012arabic}.
 %{\bf TODO: Describe how to try them with sample files}
 
-\subsection{Build your first Simple Sarf application - Hello World Example}
+\section{Build your first Simple Sarf application - Hello World Example}
 \label{subsec:hwex}
@@
 
+\bibliography{manual}
+
 \end{document}

### 2016-01-21 tex/Manual.tex
update with cmake and qt5 instructions

@@
 Sarf is an open source tool. The code is currently hosted on google code under the project name ``atmine''. To download the code, you can use the svn interface:\\
+
 \begin{verbatim}
-svn checkout http://atmine.googlecode.com/svn/trunk/ atmine-read-only.
+git clone http://github.com/codelogicanalysis/atmine 
 \end{verbatim}
-Preferablly, get a stable version of Sarf from the download section under:\\
-\begin{verbatim}
-http://code.google.com/p/atmine/downloads/list
-\end{verbatim}
+
+%\begin{verbatim}
+%svn checkout http://atmine.googlecode.com/svn/trunk/ atmine-read-only.
+%\end{verbatim}
+%Preferablly, get a stable version of Sarf from the download section under:\\
+%\begin{verbatim}
+%http://code.google.com/p/atmine/downloads/list
+%\end{verbatim}
+
 \paragraph{}
@@
 %\item subversion
-\item qt-sdk
+\item qt-sdk (Qt 5.2 minimum) 
 %\item qtcreator
@@
 \end{itemize}
+
 The above tools can be downloaded and installed using the ``Synaptic Package Manager'' in ubuntu systems or through the ``apt-get'' command.
@@
 \end{enumerate}
+
 Note that the name of the folder ``libdatrie-0.2.3'' includes the version of the library downloaded ``0.2.3''.If you download another version of this library, it will have a different name referring to a different version.
@@
 \begin{verbatim}
-ln -s /usr/share/qt4 qt
+ln -s /usr/share/qt5 qt
 \end{verbatim}
@@
 \begin{verbatim}
-make
+%make
+1. Checkout the source code
+2. mkdir build
+3. cd build
+4. cmake ../atmine/ATSarf
+5. make
 \end{verbatim}

### 2018-03-13 tex/merf/TAL/conclusion.tex
adding merf

In this work, we present a morphology-based entity and relational entity extraction framework for Arabic text.
\framework provides a friendly interface where the user defines tag types 
and associates them with regular expressions defined over Boolean formulae.
The Boolean formulae are in turn defined over matches of Arabic morphological features and
a novel extended synonymy feature ($Syn^k$).
\framework allows the user to associate code actions with each regular sub-expression 
and to define semantic relations between sub-expressions. 
%\framework uses Sarf, an Arabic morphological analyzer, 
%to compute morphological and thereafter
%regular expression matches, and relational entities. 
We evaluate \framework with several case studies and compare with existing application-specific 
techniques.
The results show that \framework requires shorter development time and effort compared 
to existing techniques and produces reasonably accurate results within a reasonable 
overhead in run time. 
In the future, \framework will support user-defined cross-reference predicates, 
and will infer morphological features from relevant example words to express a concept.

%Currently, \framework supports one built in cross-reference predicate based on the $Syn^2$ feature. 
%In the future, \framework will support user-defined cross-reference predicates. 
%Currently the user selects the morphological features to specify the MBF. 
%We will explore techniques that can infer the features from example words that the user judges as relevant to the basic concept in question.

### 2018-03-13 tex/merf/TAL/gui.tex
adding merf

\framework~provides a user friendly interface to specify the 
atomic terms, the MBFs, the MREs, 
the tag types, and the legends. 
The GUI also allows the user to modify and correct the 
tag set $R$. 
The GUI allows the user also to compute accuracy results 
that compare different tag sets and that can serve well as 
inter annotation agreement results
when the tag sets come from two human annotators, 
or as evaluation results when comparing with reference tag sets.

\framework saves the tags and the tag types in a user friendly format using
the JavaScript Object Notation (JSON) format~\cite{nolan2014javascript}.
The tag file keeps the paths of the separate text and tag type files. 
It also contains a list of tags with their tag type identifiers,
position in text, length, and word index.
The tag type file contains MBF and MRE tag types. 
%Each object has an array value containing the user-defined tag types. 
Each tag type contains the name, description, expression, and the visualization legends.

%Figure~\ref{fig:mbftagger} shows the \framework GUI. 
%%\framework's GUI allows the user to create, manipulate, and analyze a project. 
%The {\em File} menu is used to create, close, or open an existing project. 
%The {\em Tags} menu enables the user to manually define and edit tag types. 
%The user can define and edit the MBF and MRE based tag types and 
%start the corresponding automatic morphology-based simulators from the {\em Tagtypes} menu. 
%The user performs comparison and analysis of two tag sets using the {\em Analyse} menu. 
%{\em View} menu enables the user to switch visualization modes of the tags.

\subsection{Tag type Boolean formula editor}
The user writes MBF tag types with the tag type editor introduced in~\cite{JaZaMatar}. %shown in Figure~\ref{f:bfe}.
First the user specifies atomic terms by selecting a feature from ${\cal F}$. 
%The pattern filters the feature values.
The user can also choose whether to require an exact match using the
\cci{isA} predicate, or a substring match using the
\cci{contains} predicate option.

The user can add and remove feature values to the atomic terms 
using push buttons. 
A check box in the ``Feature'' column allows negating the term, and the
``Relation'' column switches the predicate between 
\cci{isA} and \cci{contains}. 
The list of feature and value pairs is interpreted as a disjunction to form
the MBF. 
A right pane shows a description of the tag type and a set of legend 
descriptors. 
When the stem or gloss features are selected, the user has the option to 
use the $Syn^k$ feature. 
%The user selects the check box in the lower left of the editor and sets the order in the spin box.

In the direction extraction task example, the user specifies four MBF-based 
tag types with labels 
$N$, $P$, $R$, and $U$ with  ``name of person'', ``name of place'', 
``relative position'', and ``numerical term'' descriptions, respectively. 
For each MBF, the user selects the morphological features, 
specifies the constant value $CF$, and adds it to the Boolean formula editor. 
%The user also assigns legend descriptors to each defined tag type. 
%
%Figure~\ref{f:bfe} shows the $U$ MBF which requires the gloss tag of 
%the stem to be an exact match of either {\tt first}, {\tt second}, \dots, or {\tt tenth}. 
%The legend descriptors set foreground and background colors to {\tt orangered}, 
%and {\tt gray}, respectively, and the font size to {\tt 12}.
%, normal font weight, and unitalicized.

\begin{figure}[tb!]
  \centering
  {\setlength{\fboxsep}{0pt}%
  \setlength{\fboxrule}{0.5pt}%
  \fbox{\includegraphics[width=\textwidth]{figures/mreedit}}}
  \caption{\framework tag type regular expression editor.}
  \label{f:sfe}
\end{figure}

\subsection{MBF match visualization}

%After defining the MBFs of the direction extraction task, 
%the user calls the MBF simulator from the {\tt Tagtypes} menu. 
%The snapshot in Figure~\ref{fig:mbftagger} shows 
The MBF match visualizer shows color sensitive text view, the tag list view, and the tag description view. 
%The color-sensitive text view shows the text with visualization of the MBF tag type matches. 
%The tag list view shows all tags.
%that are automatically or manually applied to the text. 
The tag description view presents the details of the selected tag along with 
the relevant tag type information.
%
The user can edit the tags using a context sensitive menus. 
\framework GUI also allows manual tag types and corresponding tags 
that are not based on morphological features.
This enables building reference corpora without 
help from the morphological analyzer.

\vspace{-1em}
\subsection{Tag type regular expression editor}

After interacting with the MBF editor, the user moves to 
specify the regular expressions. 
The MRE editor of Figure~\ref{f:sfe} allows the 
definition of an MRE tag type in a user friendly manner. 
The user first adds the required MBF formulae 
by selecting a label from ${\cal T}$ under MBFs. 
The Boolean formula of a highlighted tag type is shown in the table on the lower left pane. 
Each selected MBF is associated with an automatic name. 
The user can nest the MRE expression using a tree view of the MRE operations. 
The tree features the name, MBF, and operation for each sub-expression. 
%Moreover, a defined MRE is directly added to the MBF list enabling substitution and recursion 
%in formula definition.

%{\tt NONE} is a special tag type with no Boolean formula. 
%It tags all the words that are not tagged with any of the morphology-based Boolean tag types. 
%{\tt NONE} is defined by default and can be used to introduce flexibility and noise tolerance into the formula. 
%We previously defined this tag type in Section~\ref{sec:framework} and referred to it by $O$ (other).

%The regular expression editor enables the user to apply operations to the selected expressions. 
To specify a binary operation the user selects two sub-expressions and clicks the corresponding
operation button. 
%To do so, the user selects one or two expressions then selects an operation from the ones shown 
%in the lower left of the view. 
The operations include disjunction, conjunction, zero or one, sequence, zero or more, 
one or more, and up to a user defined constant.
The right pane shows a description of the tag type and a set of legend 
descriptors.

%\[
%	(P|N)+~\mathit{O}?~R~\mathit{O^{\wedge}2}~(P|N|U)+
%\]
%In the direction task, we aim to build the expression shown above. 
%Thus we start by adding the MBFs $P$, $N$, $\textit{NONE}$, $R$, $\textit{NONE}$, $P$, $N$, and $U$. 
%We apply the required operations incrementally to build the final expression. 
%For example, we apply the disjunction ($|$) operation to the first $P$ and $N$ added, 
%then we apply the one or more ($+$) operation to the resulting sub-expression.
%We proceed as such to build the regular expression shown in Figure~\ref{f:sfe}. 
%Similar to the MBF tag type editor, the user specifies a set of legend descriptors for each MRE tag type.
%

\begin{figure}[tb!]
  \centering
  \includegraphics[width=\textwidth]{figures/treegraph}
  \caption{MRE annotated Text, MRE match tree, and entity-relation graph}
  \label{fig:treegraph}
\end{figure}

\vspace{-1em}
\subsection{MRE match visualization}

While specifying an MRE the user can interact with the visualization and editor views
to make sure the MRE expresses the intent. 
The color-sensitive text view in Figure~\ref{fig:treegraph} shows 
the highlighted tag matches after the user called the MRE simulator using 
the {\tt Tagtypes} menu.

The match tree view shows the selected match in a graph view.
Figure~\ref{fig:treegraph} shows the MRE match tree of the direction task 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
%Leaf nodes show the match text, internal nodes show the operations, and edge labels shows the MBF tag type name.

%\begin{figure}[tb]
%  \centering
%  \includegraphics[width=0.5\textwidth]{figures/merfsreditor.png}
%\caption{\framework~relation editor}
%  \label{fig:relationeditor}
%\end{figure}

\subsection{User defined relation editor}

After the user is satisfied with the MRE matches, 
the user moves to define relations and code actions. 
The relation editor allows the user to define relations 
by specifying $\langle e_1,e_2,r\rangle$ tuples, 
where $e_1$ and $e_2$ denote source and destination entities, and $r$ denotes 
the label.
The editor shows the MRE tree and allows the user to select the sub 
expressions and select features 
of the matches of the sub expressions to define the three components of the relation.

A snapshot of the GUI in 
Figure~\ref{fig:treegraph} shows in an interactive graph view
the entity-relation graph of the match of the user defined relation 
extracted from the match tree of the MRE.

In the computational action editor, an advanced user can 
enter C++ code and use the \framework API to program and process 
subexpression matches. 
%RE matches In order to add computational actions to an MRE sub-expression, 
%the user selects to edit the actions in the MRE tag type editor. 
%The view allows the user to specify the actions in a user friendly manner. 
%\framework provides the user with an API to access the sub-expression match features easily and define the pre-match actions. 
%The match features include the text, position, length, and numerical value if applicable, and morphological features for MBF matches. 
%\framework also enables the user to add C++ declarations and library includes.

\subsection{Analysis}

In the analysis view, the user provides 
two tag sets $R_1$ and $R_2$ and 
two tag type sets ${\cal T}_1$ and ${\cal T}_2$ as input. 
%The analysis view provides views of the differences between the tag types and 
%the tag sets.
%
The tag type difference view shows the text annotated in three panes: 
1) the common tag types ${\cal T}_1 \cap {\cal T}_2$,
2) the tag types in ${\cal T}_1$ but not in ${\cal T}_2$, 
and 3) the tag types in ${\cal T}_2$ and not in ${\cal T}_1$.

Similarly, the tag difference view shows $R_1\cap R_2$, $R_1/R_2$ and $R_2/R_1$
in addition to precision, recall and F-measure values. 
The user selects a predicate to compute the metrics from the following predicates:
(1) ``Intersection'': a tag from $R_1$ intersects in text with a tag in $R_2$,
(2) ``Exact'': a tag from $R_1$ exactly matches a tag in $R_2$,
(3) ``A includes B'': a tag from $R_1$ contains a tag from $R_2$, and
(4) ``B includes A'': a tag from $R_2$ contains a tag from $R_1$.

### 2018-03-13 tex/merf/TAL/motivation.tex
adding merf

We motivate \framework with the example of Figure~\ref{fig:intromotiv} 
that shows how to extract direction entities from sample Arabic text. 
The text includes directions to a shopping mall and the 
figure presents a transliteration and an English translation of the text.
%The text contains some details that are not interesting to the direction extraction task. 
Interesting entities, such as names, places, relative directions, 
and numerical terms are highlighted with boxes in the text. 
We are interested in detecting those entities and presenting 
the direction as relational entities as 
shown in Figure~\ref{fig:intromotiv}.

%We demonstrate how to do that using \framework. 
%The user interacts with the \framework interface to specify Boolean 
%formulae, regular expressions, and semantic relations. 
%\framework will automatically extract formulae matches 
%using an in-house Arabic morphological analyzer (Sarf)
%\cite{ZaMaColing2012DemosSarf},
%extract regular expression matches, and construct the 
%semantic relations.

%For example, row 1 in the table shown in Figure~\ref{fig:motiv} 
%lists the matches of names of persons as $n_1$, $n_2$, and $n_3$.

{\bf Entity detection.~}
As described in the table of Figure~\ref{fig:motiv}, the user denotes 
the ``name of person'' entities with formula $N$ 
which requires the {\em category} feature in the morphological solution of a word
to be {\tt name of person}. 
The entities $n_1,n_2,$ and $n_3$ are matches of the formula $N$. 
Similarly, the user specifies formula $P$ to denote ``name of place'' entities. 
The user specifies formula 
$R$ to denote ``relative position'' entities, and requires the stem feature to belong 
to a selected list of 
stems containing \RL{fy} and \RL{qrb}. 
Similarly, $U$ denotes numerical terms and is a disjunction of constraints 
requiring the stem feature to belong to a set of stems such as
\RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).

\framework calls an inhouse open source morphological 
analyzer~\cite{ZaMaColing2012DemosSarf} and computes matches of 
the formulae $N$, $P$, $R$, and $U$. 
We refer to all {\em other} words in the text that do not match 
a user defined formula as {\em null} words and we denote them by $O$. 
The resulting matches are illustrated with boxes and superscripts 
in Figure~\ref{fig:intromotiv} and are listed in the table of 
Figure~\ref{fig:motiv}.

%{\bf Direction regular expressions.~}
The user now interacts with the regular expression editor to 
specify the direction entities and relations.
Intuitively, 
the directions are names of places ($P$) related to each other with 
positional propositions ($R$). 
A place name can be a tabulated place name, a street named after 
a person ($N$), or a numbered street ($U$). 
The text containing the directions might also include words that are not 
necessary to indicate directions ($O$)
but are necessary to complete the sentence.

The user tries several sequences of the above entities in the 
editor and checks their matches in the visualizer. 
Finally, the user is satisfied with the matches of an expression such as 
$(P|N)\!+~O?~R~O^\wedge\!2~(P|N|U)+$ where 
$|,+,?,$ and $^\wedge k$ denote disjunction, one or more, zero or one, and
up to $k$ matches, respectively.
The expression specifies a sequence of places or names of persons, 
optionally followed by a null word, 
followed by one relative position, followed by up to two possible null words, 
followed by one or more match of name of place, name of person, or numerical term. 
$O?$ and $O^\wedge 2$ are used in the expression to allow for flexible matches.
The user reaches the satisfying matches by experimenting with the visualizer and the expression editor which do not require knowledge and expertise in regular expressions.

The match trees in Figure~\ref{fig:motiv} illustrate two matches of the expression 
computed by \framework.
The first match tree refers to the text 
\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection). 
The second match tree refers to the text 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
The leaf nodes of the trees are entities and the edges and internal nodes are text, 
morphology-based, 
and word distance based relational entities. 
%The sequence and structure gives us the text of a parent node from the children nodes. 
%The same sequence can also give us the interesting (matching) morphological features. 
%The word distance is defined and abstracted by the matches of the regular expression 
%operators (internal nodes).

{\bf User defined direction relations.~}
The user now uses the relation editor to declare user defined relations that relate 
parts of the matches of the expression with each other. 
%The aim of the semantic relations to be defined is to construct the entity-relation 
%graph shown in Figure~\ref{fig:intromotiv}. 
Intuitively, 
the user wants to create relations between places, names, and numerical entities. 
A relation between two entities can be a prepositional entity.
For example, Figure~\ref{fig:intromotiv} shows the entities \RL{dby mwl}(Dubai Mall) and 
\RL{Almbn_A}(the building) related by the preposition \RL{mqrbT}(near).

Let $e_1,o_1,r,o_2,$ and $e_2$ be the 
$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$ subexpressions, respectively.
The user selected $(P|N)+$ to be an entity after noticing in the visualizer  
that it happens to capture 
non-separated sequences of place and name entities denoting a single 
entity such as \cci{Khalifa tower}. 
\cci{Relation($e_1$,$e_2$,$r$)} creates the 
edge labeled \cci{next to} between \cci{intersection 1} and 
\cci{Khalifa tower} nodes in match 1 of Figure~\ref{fig:motiv}(b), and the 
edge labeled \cci{near} between \cci{Dubai Mall} and \cci{the building} nodes in 
match 2 of Figure~\ref{fig:motiv}(b).

\cci{Relation($r$,$e_1$,$o_1$)} creates the edge labeled \cci{prep} 
between \cci{Dubai Mall} and \cci{near} nodes in match 2 of Figure~\ref{fig:motiv}(b). 
\cci{Relation($r$,$e_2$,$o_2$)} creates the edge labeled \cci{from} 
between \cci{intersection 1} and \cci{next to} nodes in 
match 1 of Figure~\ref{fig:motiv}(b), and the 
edge labeled \cci{from this} between \cci{near} and \cci{the building} 
nodes in match 2 of Figure~\ref{fig:motiv}(b).

After constructing the relations, 
the user is interested to relate the discovered entities and relational entities
that express the same concept; in particular, the same place. 
\framework provides the \cci{isA} predicate as a default cross-reference
relation which creates the edge between the nodes \cci{Khalifa Tower} and 
\cci{The building} in  Figure~\ref{fig:intromotiv}.

### 2018-03-13 tex/merf/TAL/results.tex
adding merf

In this section we compare \framework implementations of 
the narrator chain, temporal entity, and genealogy entity extration 
tasks to the task specific techniques proposed to solve them
in ANGE~\cite{ZaMaFlairs2012HadithBio}, 
ATEEMA~\cite{ZaMa2012IJCLATime},  and
GENTREE~\cite{ZaMaHaCicling2012Entity}, respectively. 
We also compare a \framework number normalization task to 
a task specific implementation. 
In the online appendix%
~\footnote{available at ~\url{-ommited-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
, we report on eight additional \framework case studies.

Table~\ref{tab:results} reports the development time,
extraction runtime, recall and precision 
of the output MRE tags, 
the size of the task in lines of code or in number of \framework rules, 
for both the standalone task specific and the \framework implementations. 
%Table~\ref{tab:mbfer} reports the accuracy of the output MBF tags and the user-defined relation 
%construction in \framework for each task.

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
\caption{\framework compared to task specific applications.}
  \resizebox{0.85\columnwidth}{!}{
    \begin{tabular}{l|l|l|ll|l}
     \toprule
     \multirow{2}{*}{Task} & Development& Run & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
     & time & time(s) & Recall & \multicolumn{1}{c|}{Precision} & \\
    \midrule
    ANGE~\cite{ZaMaFlairs2012HadithBio}
    & 2 months & 1.79 & 0.99 & 0.99 & 3000+ lines of code\\
    \framework & 3 hours & 7.24 & 0.99 & 0.93  & 8 MBFs and 4 MREs\\
    \midrule
    ATEEMA~\cite{ZaMa2012IJCLATime}
    & 1.5 months & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
    \framework & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
    \midrule
    Genealogy tree~\cite{ZaMaHaCicling2012Entity} 
    & 3 weeks & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
    \framework & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
    \midrule
    NUMNORM & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
    \framework & 1 hour & 1.53 & 0.91 & 0.90 & 3 MBFs/1 MRE/57 lines\\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:results}%
\end{table}%

%Recall refers to the fraction of the entities correctly detected against 
%the total number of entities available. 
%Precision refers to the fraction of correctly detected entities against the 
%total number of extracted entities. 
%Intuitively, precision denotes whether the system generated false positives.

For the temporal and number normalization cases, 
we evaluated the techniques against arbitrary text from issues of 
the Lebanese Assafir and Al-Akhbar newspapers
\footnote{available at \url{http://www.assafir.com} and 
\url{http://www.al-akhbar.com}.}. 
For the narrator chain case, we used
Musnad Ahmad, a hadith book, for evaluation. 
For the genealogical tree extraction we used 
an extract from the Genesis biblical text.

%\subsection{Comparison}
Table~\ref{tab:results} shows that \framework has a clear advantage over 
task specific techniques in the effort required to develop the application at 
a reasonable cost in terms of accuracy and run time. 
Developers needed three hours, three hours, four hours, and one hour 
to develop the narrator chain, temporal entity, genealogy, and number 
normalization case studies using \framework, respectively. 
However, the developers of ANGE, ATEEMA, GENTREE, and 
NUMNORM needed two months, one and a half months, 
three weeks, and one week, respectively. 
\framework needed eight MBFs and four MREs for narrator chain, 
three MBFs and 2 MREs for temporal entity, three MBFs and three MREs for 
genealogy, and three MBFs, one MRE, and 57 lines of code actions for the number normalization tasks. 
However, ANGE, ATEEMA, GENTREE, and NUMNORM required 
3,000+, 1,000+, 3,000+, and 500 lines of code, respectively.

\framework required reasonably more runtime than the task specific 
implementations and reported acceptable and 
slightly less precision metrics with around
the same recall.
%a runtime of 7.29, 1.53, 3.14, and 2.28 seconds for 
%the narrator chain, temporal, genealogy, and number normalization tasks, respectively. 
%However, ANGE, ATEEMA, GENTREE, and NUMNORM required 1.79, 2.53, 0.74, and 0.32 seconds, respectively. 
%\framework required a runtime of 7.29, 1.53, 3.14, and 2.28 seconds for 
%the narrator chain, temporal, genealogy, and number normalization tasks, respectively. 
%However, ANGE, ATEEMA, GENTREE, and NUMNORM required 1.79, 2.53, 0.74, and 0.32 seconds, respectively. 
%In narrator chain, \framework scored 0.99\% recall and 0.93\% precision while ANGE scored 0.99\% recall and 0.99\% precision. 
%In temporal entity, \framework scored 0.91\% recall and 0.81\% precision while ATEEMA scored 0.88\% recall and 0.89\% precision. 
%In genealogy, \framework scored 0.84\% recall and 0.93\% precision while GENTREE scored 0.96\% recall and 0.98\% precision. 
%In number normalization, \framework scored 0.91\% recall and 0.90\% precision while NUMNORM scored 0.91\% recall and 0.93\% precision.

\setcode{utf8}
\setarab
\begin{table}[tb!]
  \centering
  \caption{Narrator chain example.}
  \begin{Verbatim}[xleftmargin=1.5cm,fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
name:   PN ((MEAN)? PN)*;
nar:    name ((NONE)^3 FAM (NONE)^3 name)*;
pbuh:   BLESS GOD UPONHIM GREET;
nchain: ($s_1=$TOLD $s_2=$nar)+ ((PN|FAM|NONE)^8 pbuh)?
\end{Verbatim}
\resizebox{0.8\columnwidth}{!}{
    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
    \toprule 
    \notrrl{القعقاع} & \notrrl{بن} & \notrrl{عمارة} & \notrrl{عن} & \notrrl{جرير} & \notrrl{حدثنا} & \notrrl{سعيد} & \notrrl{بن} & \notrrl{قتيبة} & \notrrl{حدثنا} \\
    \midrule 
    \noarrl{القعقاع} & \noarrl{بن} & \noarrl{عمارة} & \noarrl{عن} & \noarrl{جرير} & \noarrl{حدثنا} & \noarrl{سعيد} & \noarrl{بن} & \noarrl{قتيبة} & \noarrl{حدثنا} \\
    \midrule
    PN    & FAM   & PN    & TOLD  & PN    & TOLD  & PN    & FAM   & PN    & TOLD \\
    \midrule
    name & & name & & name & & name &  & name & \\
    \midrule
    \multicolumn{3}{|c|}{nar} &       & nar   &       & \multicolumn{3}{c|}{nar} &  \\
    \midrule
    \multicolumn{10}{|c|}{nchain}
    \\
    \bottomrule
    \end{tabular}%
}
  \label{tab:nchain}%
\end{table}%
\setcode{standard}

\subsection{Narrator chain case study}
A narrator chain is a sequence of narrators referencing each other. 
%A sample narrator chain is shown in Table~\ref{tab:nchain}. 
The chain includes proper nouns, paternal entities, and referencing entities. 
ANGE uses Arabic morphological analysis, finite state machines, and graph transformations 
to extract named entities and relations including narrator chains~\cite{ZaMaFlairs2012HadithBio}.

\transfalse
MBF \cci{PN} checks the abstract category {\tt Name of Person}. 
MBF \cci{FAM} denotes ``family connector'' and checks the stem gloss ``son''. 
MBF \cci{TOLD} denotes referencing between narrators and checks the disjunction of 
the stems \RL{.hd_t}(spoke to), \RL{`n}(about), \RL{sm`}(heard), \RL{'_hbr}(told), and \RL{'nb-'}(inform). 
MBF \cci{MEAN} checks the stem \RL{`ny}(mean). 
MBFs \cci{BLESS}, \cci{GOD}, \cci{UPONHIM}, and \cci{GREET} check the 
stems \RL{.sll_A}, \RL{Al-ll_ah}, \RL{`ly}, and \RL{sllm}, respectively. 
\transtrue

Table~\ref{tab:nchain} presents the defined MREs. 
MRE {\em name} is one or more \cci{PN} tags optionally followed 
with a \cci{MEAN} tag. 
MRE \cci{nar} denotes narrator which is a complex Arabic name
composed as a sequence of Arabic names (\cci{name}) 
connected with family indicators (\cci{FAM}). 
The \cci{NONE} tags in \cci{nar} allow for unexpected words 
that can occur between names. 
%a complex Arabic name that is constructed
%a \cci{name} tag followed by zero or more 
%sequences of \cci{FAM} tag followed by \cci{name} tag 
%with up to three \cci{NONE} tags. 
MRE \cci{pbuh} denotes a praise phrase often associated with 
the end of a hadith (peace be upon him), 
and is the satisfied by the sequence of
\cci{BLESS}, \cci{GOD}, \cci{UPONHIM}, and \cci{GREET} tags. 
MRE \cci{nchain} denotes narrator chain, 
and is a sequence of narrators (\cci{nar})
separated with \cci{TOLD} tags, and optionally followed
by a \cci{pbuh} tag. 
%nal sequence of up to eight \cci{PN}, \cci{FAM}, or 
%\cci{NONE} tags followed by a \cci{pbuh} tag.

The first row in Table~\ref{tab:nchain} is an example narrator chain,
the second is the transliteration, the third 
shows the MBF tags. Rows 4, 5, and 6 show the 
matches for \cci{name}, \cci{nar}, and \cci{nchain},
respectively.
%
\framework assigns the symbols $s_1$ and $s_2$ for the 
MRE sub-expressions \cci{TOLD} and \cci{nar}, respectively. 
We define the relation $\langle s_2,s_2',s_1\rangle$ 
to relate sequences of narrators with edges labelled by the tags of \cci{TOLD} where 
$s_2'$ denotes the next match of \cci{nar} in the one or more MRE subexpression.
%The narrators in the example shown in Table~\ref{tab:nchain} are \RL{qtybT bn s`yd}, \RL{jryr}, and \RL{`mArT bn Alq`qA`}. 
%\transfalse
%The edges relating the entities are labeled by the word set \{\RL{.hdd_tnA}, \RL{.hdd_tnA}, \RL{`n}\} which contains all the %matches of the MBF \cci{TOLD}.
%\transtrue

Table~\ref{tab:mbfer} shows that \framework detected almost all the MBF matches 
with 99\% recall and 85\% precision and 
extracted user-defined relations with 98\% recall and 99\% precision.

For brevity, we omit the details of \framework temporal entity extraction, 
genealogy tree, and number normalization case studies, describe them shortly
below and provide a full description in the online Appendix.

\begin{table}[tb!]
  \centering
  \caption{\framework MBF and user-defined relation accuracy }
  \resizebox{0.65\columnwidth}{!}{
    \begin{tabular}{l|c|c|c|c}
     \toprule
     \multirow{2}{*}{Task} & \multicolumn{2}{c|}{MBF accuracy} & \multicolumn{2}{c}{relation accuracy}\\
     & Recall & Precision & Recall & Precision \\
    \midrule
    Narrator chain & 0.99 & 0.85 & 0.99 & 0.98 \\
    Number normalization & 0.99 & 0.99 & 0.97 & 0.95 \\
    Temporal entity & 0.99 & 0.52 & 0.98 & 0.89 \\
    Genealogy tree & 0.99 & 0.75 & 0.81 & 0.96 \\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:mbfer}%
\end{table}%

\vspace{-1em}
\subsection{Temporal entity extraction}

Temporal entities are text chunks that express temporal information. 
Some represent absolute time such as \RL{Al_hAms mn 'Ab 2010}. 
Others represent relative time such as \RL{b`d _hmsT 'ayAm}, and quantities 
such as \RL{14 ywmA}. 
{\em ATEEMA} presents a temporal entity detection technique for the Arabic language using 
morphological analysis and finite state transducers~\cite{ZaMa2012IJCLATime}.

Table~\ref{tab:mbfer} shows that \framework detected almost all the MBF matches with 99\% recall, 
however it shows low precision (52\%). 
As for the semantic relation construction, \framework presents a 98\% recall and 89\% precision.

\vspace{-1em}
\subsection{Genealogy tree}

Biblical genealogical lists trace key biblical figures such as Israelite kings and
prophets with family relations. 
The family relations include wife and parenthood. 
A sample genealogical chunk of text is \RL{w wld hArAn lw.tA} 
meaning ``and Haran became the father of Lot''.

GENTREE~\cite{ZaMaHaCicling2012Entity} 
automatically extracts the genealogical family trees using morphology, 
finite state machines, and graph transformations. 
Table~\ref{tab:mbfer} shows that \framework detected 
MBF matches with 99\% recall, and 75\% precision, and
extracted user-defined relations with 81\% recall and 96\% precision.

\vspace{-1em}
\subsection{Number normalization}
\label{sec:sec:number}

We implemented a number normalization extractor using \framework and 
compared it with {\em NUMNORM}, a 
C++ implementation for number normalization. 
First, we defined the MBFs \cci{DT}, \cci{H}, and \cci{TMB}
to denote (1) digits and tens, (2) hundreds, and (3) 
thousands, millions, and billions, respectively.
The \cci{num} MRE 
\cci{(DT|TMB|H)+} is one or more \cci{DT}, \cci{TMB}, or \cci{H} tags. 
\framework assigns the symbols $s_1$, $s_2$, and $s_3$ 
for the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H}, respectively. 
The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
%Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.
The actions use \framework API to access features of the matches such 
as text (\cci{\$s1.text}) and numeric 
value (\cci{\$s1.number}) of literal numbers such as numbers from one to ten.

Table~\ref{tab:mbfer} shows high accuracy in MBF tagging with 99\% recall and 99\% precision, and
high accuracy in relation extraction with 97\% recall and 95\% precision.

%\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
%\begin{figure}[tb!]
%\centering
%  \begin{tabular}{p{5.5cm}p{5.5cm}}
%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=TMB algorithm,commandchars=\\\[\]] 
%cout << \fvtextcolor[red][$s1.text];
%if(isHundred) {
%  if(current != 0) {
%    previous += current;
%  }
%  current = currentH * \fvtextcolor[red][$s1.number];
%  currentH = 0;
%  isHundred = false;
%  isKey = true;
%} else if(current == 0) {
%  
%  current = \fvtextcolor[red][$s1.number];
%  isKey = true;
%} else if(!isKey) {
%
%  isKey = true;
%  current = current * \fvtextcolor[red][$s1.number];
%} else {
%  previous += current;
%  current = \fvtextcolor[red][$s1.number];
%}
%\end{Verbatim}
%&
%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=DT algorithm,commandchars=\\\[\]] 
%if(isHundred) {
%  currentH += \fvtextcolor[red][$s0.number];
%} else if(current == 0) {
%  current = \fvtextcolor[red][$s0.number];
%} else if(isKey) {
%  previous += current;
%  current = \fvtextcolor[red][$s0.number];
%} else {
%  current += \fvtextcolor[red][$s0.number]; }
%isKey = false;
%\end{Verbatim}
%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=H algorithm,commandchars=\\\[\]] 
%isHundred = true;
%if(current == 0)  {
%  currentH = \fvtextcolor[red][$s2.number];
%} else if(!isKey) {
%  currentH = current * \fvtextcolor[red][$s2.number];
%  current = 0;
%} else {
%  currentH = \fvtextcolor[red][$s2.number];}
%isKey = false;
%\end{Verbatim}
%\\ 
%\end{tabular}
%\caption{Actions for TMB, DT, and H MRE expressions.}
%\label{fig:numnormalgo}
%\end{figure}

\vspace{-1em}
\subsection{Discussion}
\label{subsec:discuss}
\input{discussion}

### 2018-03-13 tex/merf/TAL/sarf.tex
adding merf

\vocalize

\def\pp{\ensuremath{{\cal P}}} % prefix
\def\ss{\ensuremath{{\cal S}}} % stem
\def\xx{\ensuremath{{\cal X}}} % suffix
\def\PP{\ensuremath{\mathit{POS}}} % pos
\def\GG{\ensuremath{\mathit{GLOSS}}} % gloss
\def\AC{\ensuremath{\mathit{CAT}}} % category

%~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
%{\bf Morphological analyzer (Sarf).~}
Morphological analysis is key to Arabic NLP~\cite{arabicmorph} 
due to the exceptional degree of ambiguity in writing, 
the rich morphology, and the complex word derivation system. 
Short vowels, also known as diacritics, are typically omitted in Arabic text
and inferred by readers~\cite{habash2006arabic}. 
For example, the word \RL{bn}%
can be interpreted as \RL{bon} {\tt ``coffee''} with a {\em damma} diacritic on the 
letter \utfrl{بـ} or 
\vocalize \RL{bin} (son of) with a 
{\em kasra} diacritic on the letter \utfrl{بـ}.

Morphological analysis is required even for tokenization 
of Arabic text. 
The position of an Arabic letter in a word 
(beginning, middle, end, and standalone) changes
its visual form.
Some letters have non-connecting end forms which allows visual
word separation without the need of a white space separator. 
For example, the word \utfrl{ياسمين} can be interpreted as
the ``Jasmine'' flower, 
as well as \notrutfrl{يا} (the calling word) followed by
the word \notrutfrl{سمين} (obese). 
Consider the sentence 
%\utfrl{ذهب الولدإلى المدرسة} 
\notrrl{AlmdrsT}\notrrl{_dhb alwald-il_A}
\arabfalse \RL{_dhb alwald-il_A almdrsT} \arabtrue
(the kid went to school). 
The letters \notrutfrl{د} and \notrutfrl{ى} have 
non-connecting end of word forms and the words 
\notrutfrl{الولد},\notrutfrl{الى}, and\notrutfrl{المدرسة} 
are visually separable, 
yet there is no space character in between.
Newspaper articles with text justification requirements, 
SMS messages, and automatically digitized documents
are examples where such problems occur.

\framework is integrated with {\em Sarf}, 
an in-house open source Arabic morphological analyzer based on 
finite state transducers~\cite{ZaMaColing2012DemosSarf}. 
Given an Arabic word, Sarf returns 
a set of morphological solutions. 
A word might have more than one solution 
due to multiple possible segmentations and multiple tags associated 
with each word. 
A morphological solution is the internal structure of the word 
composed of several morphemes including 
{\em affixes} ({\em prefixes} and {\em suffixes}), and a
{\em stem}, where each morpheme is associated with tags such as 
POS, gloss, and category tags~\cite{arabicmorph,habash2010introduction}.

Prefixes attach before the stem and a word can have multiple prefixes. 
Suffixes attach after the stem and a word can have multiple suffixes. 
Infixes are inserted inside the stem to form a new stem. 
In this work we consider a set of stems that includes infix morphological changes. 
The part-of-speech tag, referred to as POS, 
assigns a morpho-syntactic tag for a morpheme. 
The gloss is a brief semantic notation of morpheme in English. 
A morpheme might have multiple glosses as it could stand for multiple meanings. 
The category is a custom tag that we assign to multiple morphemes. 
For example, we define the {\tt Name of Person} category to include proper names.
%The category is a user defined tag that the user can assign to several 
%morphemes.
%For example, the user can define a temporal category to include the 
%prefix \utfrl{س} (will) and the time unit \utfrl{ساعة} (hour).

We denote by 
\ss,
\pp,
\xx,
\PP,
\GG, and 
\AC, the set of 
all stems,
prefixes,
suffixes,
POS,
gloss, 
and user defined category tags, respectively. 
Let $T=\langle t_1,t_2,\ldots,t_M\rangle$ be a set of Arabic words denoting the 
text documents. 
\framework uses Sarf to compute a set of morphological solutions $M(t)=\{m_1,m_2,\ldots,m_N\}$
for each word $t\in T$. 
Each morphological solution $m\in M(t)$ is a tuple of the form 
$\langle p,s,x,P,G,C\rangle\in \pp \times  \ss\times \xx\times \PP\times\GG \times \AC$.

%\setarab
\begin{table}[tb!]
  \centering
  \caption{Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
  \resizebox{0.7\columnwidth}{!}{
    \begin{tabular}{|r|c|c|c|c|c|}
          & \multicolumn{3}{c|}{\textbf{Prefixes}} & \textbf{Stem} & \textbf{Suffix} \\
    \textbf{Data} & \utfrl{فَ} & \utfrl{سَ} & \utfrl{يَ} & \utfrl{أْكُل} & \utfrl{ها} \\
    \textbf{POS} & CONJ+ & FUT+ & IV3MS+ & VERB\_IMPERFECT & IVSUFF\_DO:3FS \\
    \textbf{Gloss} & and/so & will & he/it & eat/consume & it/them/her \\
    \textbf{index} & \multicolumn{3}{c|}{10} & 13 & 16 \\
    \textbf{length} & \multicolumn{3}{c|}{3} & 3 & 2 \\
    \end{tabular}%
    }
  \label{tab:samplesolution}%
\end{table}%

Table~\ref{tab:samplesolution} shows the morphological analysis
of the word \notrutfrl{فَسَيَأْكُلها}. 
The word is composed of the prefix morphemes 
\utfrl{فَ}, \utfrl{سَ}, and \utfrl{يَ}, followed by the 
stem \utfrl{أْكُل}, and then followed by the suffix morpheme
\utfrl{ها}. 
Each morpheme is associated with a number of morphological features.
The {\tt CONJ},
{\tt FUT}, 
{\tt IV3MS} 
{\tt VERB\_IMPERFECT}, and 
{\tt IVSUFF\_DO:3FS} POS tags indicate
conjunction, 
future, 
third person masculine singular subject pronoun,
an imperfect verb, and 
and a third person feminine singular object pronoun, respectively.
The POS and gloss notations follow the Buckwalter notation~\cite{Buckwalter:02}.

\input{MERFOverview}

### 2018-03-13 tex/merf/TAL/tool.tex
adding merf

The \framework framework is illustrated in the flow diagram of
Figure~\ref{f:overview}.
The Arabic text and the reference tag chunks are the primary inputs to \framework. 
Solutions, morphology-based Boolean formulae, tags, 
morphology-based regular expressions, 
tag chunks, relation and action definitions, and data structures expressing entities 
and relations are input and output data to processes. 
The morphological analyzer (Sarf), $Syn^k$ detector, 
GUI for Boolean formulae definition, 
visualization annotator, GUI for regular expression and action definition, 
Boolean formula simulator, regular expression simulator, 
relation extraction and action execution, and difference and statistical analyzer 
are processes.

%The user interacts with \framework~through a user-friendly interfaces to 
%specify morphology-based Boolean formulae and regular expressions 
%and associate them with tag types that are in turn associated with
%visualization legends such as 
%foreground and background colors and fonts. 
%The user also associates code actions to parts of the regular expressions.
%The user can also declare relations between matches using the relation
%editor.

\subsection{The extended synonymy feature $Syn^k$}

The sets $E, A,$ and $L$ denote all English words, Arabic words, 
and Arabic lexicon words, respectively. 
We have $\GG \subset E$ and $\ss \subset L \subset A$. 
Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
subsets of related English glosses. %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.

Given a word $w\in L$, 
$Sy(w)=\{u\mid u\in \ss \land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$
is the set of Arabic stems 
directly related to $w$ through the gloss map.

Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
$Sy^{1}(w) = Sy(w)$ and
$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.

The example in Figure~\ref{fig:introsynEx} illustrates the computation. 
The word \RL{mA'}is related to \RL{n.d.h}through the gloss intersection \cci{water}. 
The $Syn^2$ feature relates 
the words \RL{mA'} and ``\RL{r^s^s}'' since \RL{n.d.h} and \RL{r^s^s} 
have the gloss intersection \cci{spray}.

\begin{figure}[tb!]
\setcode{utf8}
\begin{center}
  \resizebox{0.7\columnwidth}{!}{ 
  	{\relsize{-2} \input{figures/syn2ex.tex}} }
\setcode{standard}
\caption{$Syn^2($\RL{mA'}$)$}
\label{fig:introsynEx}
\end{center}
\end{figure}

\subsection{MRE: Morphology-based regular expressions}

Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
set of morphological features. 
Given a word $w$, a morphological feature $A\in{\cal F}$, 
a user defined constant feature value $CF\in A$, and an integer 
$k, 1\le k\le 7$, 
the following are morphology-based atomic terms (MAT). 
\begin{itemize}
  \item $a(w):= \exists m \in M(w). m=\langle p,s,x,P,G,C\rangle. r \circ CF$
where $\circ \in {\cal O}$, $r \in \{p,s,x,P,G,C\}$, and $r\in A$.
Informally, a solution vector of $w$ exists with
a feature containing (or exactly matching) $CF$. 
\item $a(w) := w \in Syn^k(CF), CF \in \ss$.
  Informally, this checks $w$ is an extended synonym of a stem $CF$. 
\end{itemize}

A morphology-based Boolean formula (MBF) is of the following form. 
\begin{itemize}
  \item $a$ and $\neg a$ are MBF formulae where $a$ is a MAT and $\neg$ is the negation operator. 
  \item $(f \vee g)$ is an MBF where $f$ and $g$ are MBF formulae, 
    and $\vee$ is the disjunction (union) operator. 
\end{itemize}
Formula $N$ in Figure~\ref{fig:motiv} checks whether a solution has a category
feature matching category {\tt Name\_of\_Person}.
Formula $R$ is the disjunction of MAT terms that check for
solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).

A morphology-based regular expression (MRE) is one of the following. 
\begin{itemize}
\item $m$ is an MRE where $m$ is an MBF.
\item $fg$ is an MRE denoting a concatenation operation
  where $f$ and $g$ are both MRE expressions. This is 
  satisfied by a match of $f$ followed by a match of $g$. 
\item $f*,f+,f$\textasciicircum$x,$ and $f?$ are MRE expressions where $f$ is an MRE,
  and are satisfied by zero or more matches, one of more matches, up to $x$ matches, 
  and zero or only one match of $f$, respectively.
\item $f\& g,$ (conjunction) and $f|g$ (disjunction) are  MRE expressions where 
  $f$ and $g$ are MRE expressions, and are satisfied by a 
  a match of both $f$ and $g$, 
  and a match of either $f$ or $g$, respectively. 
\end{itemize}
We denote by $\ldbrack f\rdbrack$ the set of matches of an MRE $f$.

\subsection{User-defined relations and actions}

A user-defined relation $R$ defined by a tuple 
$\langle e_1,e_2,r\rangle$ where 
$e_1,e_2,$ and $r$ are identifiers associated with MRE 
subexpressions in as expression $f$ is a set of labeled binary edges
where matches of $e_1$ and $e_2$ are the source and destination nodes
and matches of $r$ are the edge labels.
We refer to a member of $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ 
as a user defined relational entity.

\framework allows advanced users to write C++ code snippets 
to process matches of MRE subexpressions. 
Users can use these actions to compute statistical features, 
store intermediate results, or apply intelligent entity inference techniques
as we show in the numerical extraction example of Section~\ref{sec:sec:number}. 
\framework provides an API that enrich the actions with detailed access to
all solution features of an MRE or an MBF match including 
text, position, length, equivalent numerical value if applicable, 
and morphological features.

Once \framework computes all match trees, it traverses each match to 
execute the user defined \cci{pre-match} actions in pre-order manner
and the \cci{on-match} actions in post-order manner.

\subsection{\framework simulator}

The set of tag types ${\cal T}$ contains tuples of the form $\langle l,f,d\rangle$ 
where $l$ is a text label with a descriptive name, 
$f$ is an MRE, and $d$ is a visualization legend 
with font and color information.

For each word $t_i\in T, 0\le i < n, n=|T|$ 
\framework computes a Boolean value 
($\{\mathit{true}, \mathit{false}\}$)
for all MBFs. 
Then, it computes the set of MBF tags
$R_i=\{(t_i,tt)| tt=\langle l,f,d\rangle \wedge
f~\mathit{is~an~MBF} \wedge f(t_i)\} \subseteq T \times {\cal T}$
which tags a word $t_i$ with $\mathit{tt}$ 
iff the MBF $f$ associated with
tag type $\mathit{tt}$ is true for $t_i$.

The MBF evaluation results in a sequence of tag sets 
$\langle R_0, R_1, \ldots, R_{n-1}\rangle$.
If a word $t_o$ has no tag type match, 
its tag set $R_o$ is by default the singleton $O=\{\mathit{NONE}\}$
and $t_o$ is referred to as a {\em null} word.

%\subsection{MRE and action simulation}

For each MRE, 
\framework generates its equivalent non-deterministic finite automaton (NFA) 
in the typical manner~\cite{sipser2006introduction}. 
%Each MRE operation has its equivalent representation in an NFA. 
We support the upto operation ($f$\textasciicircum$x$), which is not directly 
supported in ~\cite{sipser2006introduction}, by 
%it is equivalent to $f?|ff|\dots|\underbrace{f}_{x \text{ times}}$ which has an NFA mapping. 
%we can expand it into a standard regular expression form, for example
expanding it into a regular expression form; for example 
$f$\textasciicircum$3$ is equivalent to $f?|ff|fff$. 
Consider the example of Figure~\ref{fig:motiv} and the
MRE $\mathit{dir}= (P|N)+~O?~R~O\wedge 2~(P|N|U)+$. 
Figure~\ref{fig:nfaEx} shows part of the corresponding NFA where
$q_8, q_9, \dots, q_{13}$ represent NFA states,
and edges are transitions based labeled with MBF tags such as 
$P,$ and $N$. Edges labeled with the 
empty string $\epsilon$ are non-deterministic.
The expression uses operations $|$, $?$, $+$, and $\wedge$ to 
relate places {\tt P}, names of persons {\tt N}, 
relative positions {\tt R}, numerical terms {\tt U}, and other {\tt O}.

\framework simulates the generated NFA over the sequence of tag sets 
matching the MBF formulae.
A simulation match $m$ of an expression $f$ is a tree where the root is the 
MRE expression, the internal nodes are the MRE and MBF operations, and the 
leaves are matches of the MAT terms of $f$.
The leave matches form a vector of tags
$\langle r_k,r_{k+1},\dots,r_j\rangle$ 
corresponding to the text sequence 
$\langle t_k,t_{k+1},\dots,t_j\rangle$ where $r_{\ell}\in R_{\ell},0\le k\le \ell \le j < n$.
%
%If the simulation has a match $\langle r_m,r_{m+1},\dots,r_n\rangle$ 
%where $0\le m\le n$, 
%the next simulation starts at $R_{n+1}$. 
%This disallows overlap of matches for the same MRE. 
%
%In case the NFA simulation has no match, 
%the next simulation starts at $R_{m+1}$. 
If we have more than one match %starting at $R_k$ where $0\le k\le n$, 
\framework returns the longest. 
%
Figure~\ref{fig:motiv}(b) shows two match trees of $\mathit{dir}$ extracted 
from the text of Figure~\ref{fig:intromotiv}.
\RL{dby} and \RL{mwl} are leaf nodes referring to name of place tags ($P$). 
The $+$, $|$, and $?$ MRE operations
are internal nodes.

%\framework maintains a function $\phi\subset Q\times\Phi$, 
%where $\mathit{Q}$ is the set of states in the NFA and $\Phi$ is the set of sub-expressions in the MRE.
%So $(q,f)\in \phi$ iff state $q\in Q$ was generated by \framework to correspond to sub-expression $f\in \Phi$. 
%\framework uses $\phi$ to compute a match tree with respect to the MRE regular expression. 
%It also uses $\phi$ and the match sequence to compute the sequence of computational actions of an MRE match.

\setcode{utf8}
\setarab
\transfalse

\begin{figure}[tb!]
\centering
\resizebox{0.7\columnwidth}{!}{
	\relsize{-1} \input{figures/reNFA.tex}
}
  \caption{Equivalent NFA of direction expression}
  \label{fig:nfaEx}
\end{figure}
\transtrue
\setcode{standard}

%\begin{figure}[tb!]
%{ \relsize{-1}
%\begin{framed}
%\begin{multicols}{2}
%\begin{itemize}
%\item Annotated Expression: \\
%\\
%{ \relsize{-2.5}
%%\begin{itemize}
%%  \item 
%  $\stackrel{e_1}{(P|N)+}~\stackrel{o_1}{\mathit{O}?}~\stackrel{r}{R}~\stackrel{o_2}{\mathit{O^{\wedge}2}}~\stackrel{e_2}{(P|N|U)+}$
%%\end{itemize}
%}
%\end{itemize}
%\begin{itemize}
%\item User defined semantic \\relations:
%\begin{itemize}
%\item $\langle e_1,e_2,r\rangle$
%\item $\langle r,e_1,o_1\rangle$
%\item $\langle r,e_2,o_2\rangle$
%\end{itemize}
%\end{itemize}
%\columnbreak
%\setcode{utf8}
%\transfalse
%\resizebox{0.9\columnwidth}{!}{
%	\relsize{+1.5} \input{figures/ergraphEx.tex}
%}
%\transtrue
%\setcode{standard}
%\end{multicols}
%\end{framed}
%}
%\caption{User-defined semantic relation example}
%\label{fig:srEx}
%\end{figure}

\framework computes the relational entities in a user defined 
relation $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$ 
to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
with the smallest nonzero positive distance between the source and the destination
where the distance is the number of words between the matches.

%\framework supports the following configurations for the subexpressions of $e_i$ and $e_j$:
%\begin{itemize}
%\item $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert=1$: Relate $e_i$ match to $e_j$ match with edge labelled by $r$.
%\item $i=j\wedge\left\vert{e_i}\right\vert>1$: Relate sequences of $e_i$ matches with edges labelled by $r$.
%\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert=1$: Relate each $e_i$ match to the single $e_j$ match with an edge labelled by $r$. The reverse applies for $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert>1$.
%\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert>1$: Relate each $e_i$ match to each $e_j$ match with an edge labelled by $r$.
%\end{itemize}
%
In Figure~\ref{fig:motiv}(b), \framework names the subexpressions 
$(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, 
as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
The user defines the semantic relations
$\langle e_1, e_2, r\rangle$, 
$\langle r, e_1, o_1\rangle$, and 
$\langle r,e_2,o_2\rangle$.

The matches of $e_1,~e_2,~o_1,~o_2,$ and $r$ from the second match tree 
of Figure~\ref{fig:motiv}(b) are \RL{dby mwl}(Dubai Mall), 
\RL{Almbn_A}(the building), \RL{`l_A}(prep), \RL{mn h_dA}(from this), 
and \RL{mqrbT}(near), respectively. 
\framework~constructs the semantic relation matches and builds the lower part of the 
entity-relation graph shown in Figure~\ref{fig:intromotiv}.

For the first match \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'-wl},
the matches of $e_1,~e_2,~o_2,$ and $r$ are 
\RL{brj _hlyfT}(Khalifa tower), \RL{AltqA.t` Al-'-wl}(intersection 1), 
\RL{mn}(from), and \RL{bAlqrb}(next to), respectively. 
\framework~doesn't construct the relation $\langle r, e_1, o_1\rangle$ 
since $o_1$ has no match. 
Therefore, we get the upper part of the entity-relation graph shown 
in Figure~\ref{fig:intromotiv}.

After computing the relational entities, 
\framework~computes a cross-reference relation between the extracted entities
using a second order synonymy feature ($Syn^2$).
The \cci{isA} edge in the 
graph of Figure~\ref{fig:intromotiv} shows the cross-reference relation 
between \RL{brj _hlyfT}(Khalifa tower) from the first match 
with \RL{Almbn_A}(the building) from the second match.

### 2018-03-13 tex/merf/TALMajor/abstract.tex
adding merf

Rule-based techniques and tools to extract {\em entities} and {\em relational entities} from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.  These techniques and tools require expertise in linguistics and programming and lack support of Arabic {\em morphological analysis} which is key to process Arabic text.
In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.  \framework provides a friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines {\em tag types} and interactively associates them with regular expressions defined over Boolean formulae. Boolean formulae range over matches of Arabic morphological features, and synonymity features. Users define relations with tuples of subexpression matches and can associate code actions with subexpressions. \framework computes feature matches, regular expression matches, and constructs entities and relational entities from the user-defined relations. We evaluated our work with several case studies and compared with existing application-specific techniques. The results show that \framework requires shorter development time and effort compared to existing techniques and produces reasonably accurate results within a reasonable overhead in run time.

### 2018-03-13 tex/merf/TALMajor/answer/Reviewer1.comments.tex
adding merf

\section*{Reviewer 1 comments}
\textit{Comments to the Author}\\

The paper describes a tool (MERF) for entity and relation 
extraction in Arabic text.
The tool allows to describe regular expressions which atomic 
elements are morphological descriptions of Arabic words.
The regular expressions are compiled as finite state 
automata that are used for pattern matching.

MERF uses a morphological analyzer for Arabic Sarf that 
is itself based on finite state transducers.
MERF proposes several GUI to edit rules and analyze results.
The paper is clearly written.

\answer{
  Thank you. 
}

\begin{enumerate}[leftmargin=0mm,label=\bfseries CommentR1.\arabic*]

  \item \label{Review1.1} 
The authors chose to use formal notation to describe the 
objects manipulated by the software. 
This make things rigouros but the notations are sometimes heavy.
\answer{
  We clarified the following notations.... 
  The new notations appear in Sectioon AA in teh paper. 
}

\item \label{Review1.2} 
I did not understand what is the exact status of the predicate 
isA and how it is used, the authors should be a bit 
 clearer about that.

\answer{
We modified the description of isA to .... 
}

\item \label{Review1.3} 
The system is evaluated and compared to four other systems for 
tasks of Information Extraction.
the data used for the evaluation is not clearly described, 
did you have a manual gold annotation of all the data ?
\answer{
We have to look at the test data for the paper. 
}

\item \label{Review1.4} 
what is the size of the evaluation data ? 
did you have a train and a test data set ?
\answer{
No training... testing yes. 
}

\item \label{Review1.5} 
how did you compute the time for the implementation of rules 
with the other system (one of the tasks needed 2 month of work 
!! is it really 2 acutal months of work or is it the period of 
time during which the development has been done)
\answer{
A research assistant 6 hours course work and 14 hours 
teaching duties performed the task in two months. 
}

\item \label{Review1.5} 
more should be said about the evaluation methodology.
\answer{
We added a methodology section. 
}

\item \label{Review1.6}
you say nothing about the availability of the system, 
is it distributed ? free ? what kind of licence ?
\answer{
Available online on github, with GPL license. 
}

\end{enumerate}

### 2018-03-13 tex/merf/TALMajor/answer/Reviewer2.comments.tex
adding merf

\section*{Reviewer 2 comments}
\textit{Comments to the Author}

The submitted paper is dealing with an interesting topic 
of knowledge extraction in context of presenting a tool 
for extracting Entities and Relationships from Arabic texts. 
The tool is mainly based on an intensive interactive process, 
where a user has to define tag types, and associates them 
with regular expressions defined over Boolean formulae, 
which are in turn defined over matches of Arabic morphological 
features extended by additional synonyms. 
A major feature of this system is providing the user with a 
user-friendly GUI, however it does not support classification 
tasks.
\answer{
  Thank you for finding our work interesting. 
}

Despite the interesting motivation and the importance of 
this topic, the presentation shows some shortcomings in the 
following sense:

\begin{enumerate}[leftmargin=0mm,label=\bfseries CommentR2.\arabic*]

\item \label{Review.2.1} 
The conceptual and implementation aspects of the framework 
were introduced in not clear and differentiated form. 
The authors are advised to separate these issues in formal 
definitions and to present the resulted software tool in 
terms of more clear software techniques.
\answer{
We introduced the following formal definitions...

We clarify the implementation in the following manner. 
Tranformative algorithms: 1. RE to FSM
                          2. Entities and relations to garphs
Generative algorithms: 1. Actions to code
... 
}

\item \label{Review.2.2} 
Despite the motivation, that framework is proceeding from a 
morphological aspect, this feature came to short in the 
presentation; e.g. sets of morphological solutions.

\answer{
  Section XXX discussed morphological analysis in Arabic. 
  We extend that in ...
  We highlight that by referring to the Section in the 
  Introduction. 
}

\item \label{Review.2.3} 
Finally, I believe this framework should be also expressed 
in terms of the complexity and decidability problems, 
particularly when evaluating the required time of different 
development tasks. 
\answer{
  The problem is decidable: RE to FSM. 
  The problem might be not tractable: NDFSM to FSM might 
  be exponential. 
  In practice Table XXX provides runtime results... 
  We modified the paper to include the above. 
}

\item \label{Review.2.4} 
Providing the system with GUI is an advantage, 
However the system requires probably to some extend 
good linguistic and formal languages expertise to optimize 
the benefit of the implements tool.
\answer{
We discuss the target audience of the MERF tool in Section XXX. 
We prepared a set of 45 minutes slides based on the regular
expression chapter in Martin Jurafsky's book to... 
}

\end{enumerate}

### 2018-03-13 tex/merf/TALMajor/answer/Reviewer3.comments.tex
adding merf

\section*{Reviewer 3 comments} 
\textit{Comments to the Author}

This work is an improvement step in a continuous previous work 
(Saref,MATAr) which leads to a complete project that can be 
used as a tool for other researchers in the field to help 
improving the Arabic language computations and processings.

It depends on the morphological analyser (Saref) results 
to define tags for flexible corpus annotation that is automatic 
but allow the users to correct and fix the tagging. 
In additions it extract entitiy and relational entityies from 
the input document.

\begin{enumerate}[leftmargin=0mm,label=\bfseries CommentR3.\arabic*]
\item \label{Review.3.1}
As a reader, It wasn't very easy to followup with the writing 
style. I needed to return back to ``Saref'' and ``MATAr'' papers 
to understand the terminology used.
\answer{
  We added a background section with definitions... 
}

First, I think we can ask 
"What would a researcher reading this paper need to know?", the answer would be:

\item \label{Review.3.2}
1- ``What is new in MERF?''. and this was clear for the reader 
    which is mainly
    a)(entity and relational entity extraction).
    b) defining tag of words based on a synonymic relation
\answer{
  ...
}

\item \label{Review.3.3}
2- ``How to use MERF?''
(you can have more explanations in your appendex and could have 
``help'' button in your API. This part was also relatively 
clear for the reader.
\answer{
  ... 
}

\item \label{Review.3.4}
3- The most important part is ``how MERF was built 
(methodology)?''. but I found this part was not very clear.
\answer{
  We added a methodology section...
}

4-and for sure ``how it was evaluated and good results discussion''. hich needs more attention.
\answer{
  We added subsection Results.X to discuss the evaluation 
  strategy. 
  We also added extended the discussion section... 
}

The following are some detailed recomendations:

\item \label{Review.3.5}
1-In the introduction, I think you need to omit ``We discuss the importance of the morphological featureotivs supported in MERF terms in Section 3; in brief, morphological preprocessing iskey to Arabic NLP.'' , in page 2 since it will be repeated in page 5.
\answer{
  Done. 
}

\item \label{Review.3.6}
2- in the introduction you illusterated a target example, 
so why don you need the Motivation section. 
What about if combine both.
\answer{
  Thank you. We combine both as motivation and we include that
  as a subsection of the introduction. 
}

\item \label{Review.3.7}
In addition, at page 9 ``the edges and internal nodes are text, 
morphology-based, and word distance based relational entities'' 
is not clear.
\answer{
  We rephrase the sentence to read ... 
}

\item \label{Review.3.8}
please always define your terminology before using it and 
provide examples. suggest the ``formula" was not so clear. 
I suggest using regular grammar insted since you are using 
regular expression and the match tree can be considered as 
parse tree.
\answer{
  We use formula to define predicates that range over 
  the morphological features. 
  We consider the matches of the formulae as tokens/tags (maybe
  we call them tag types.)
  We then use regular expressions (maybe we should use rules). 
  ... 
  We modify the paper by adopting the suggested terminology: 
  rule for regular expression and parse tree for match tree. 
}

\item \label{Review.3.9}
2- Section ``Existing annotation and entity extraction tools” 
need to be with section 6 or section 6 need to be combined 
with it in as one section but leaving the comparision to 
the analysis section.
\answer{
  We thank the reviewer for the comment. 
  We moved Section 1.1 to Section 6. 
}

\item \label{Review.3.10}
3- Section Background:Morphological analysis, 
the definition explaned in page 9 did not match table 1 at page 7.
\answer{
  We fix the definition to reflect that a prefix (and an suffix)
  could be a concatenation of other prefixes. 
}

\item \label{Review.3.11}
4- Section 5 can be combined with section 2. 
you can introduce the user friendly interface using the 
specified example. 
Section 2 seemed not very imporatnt as stand alone section.
\answer{
  We merged Section 2 with Section 1. 
  We move Section 5 to become Section 2 so that it follows 
  the motivation directly. 
}

5- Section MERF/

\item \label{Review.3.12}
a) figure 3 did not make it clear for the reader to 
understand the process. Also, the figure location is far 
away from its explanation.

\answer{
We split Figure 3 into three Figures and provide explanations
for both. 
a. Arabic Text (Consider MBBF as input coming fro another figure) ----> Tags.
b. Rest of the figure... 
c. RTC + Diff. + Visualizatiin 
}

\item \label{Review.3.13}
b) Syn$^k$ is nicely explaned but I did not recognize it as 
your own idea. if this is your idea to use English translation 
as a pivot to extract Arabic synonyms then it needs to have 
more focus because I think this is new.
\answer{
Up to our knowledge, we introduced the idea. 
We highlight the concept more in the paper. 
}

\item \label{Review.3.14}
what is $2^{gloss}$ and $2^{s}$?
\answer{
  The power sets of... We clarify that in the paper. 
}

\item \label{Review.3.15}
please to make it easier for the reader to followup, 
let the example in figure 4 also be defined as the formula 
variables. w,{u},..so on
\answer{
  We do that now.
}

\item \label{Review.3.16}
c) In MRE section,page 10, $k <=7$ why? 
\answer{
  Computational reasons ... say it in the paper. 
}

\item \label{Review.3.17}
also, A belongs-to F , A is a morphological feature, 
then how CF belongs-to A. A is defined as one item not a set.
\answer{
Thank you. This is a mistake. We corrected that in teh paper to 
read as follows... 
}

\item \label{Review.3.18}
MRT definition is not clear.
\answer{
We clarify the definition of MRE in the paper. 
}

\item \label{Review.3.19}
"MRE" and "user defined relations and actions " sections 
were hard to be understood.
\answer{
We rewrote both sections... 
}

\item \label{Review.3.20}
MREF simulator: please provide examples soon after the 
definitions to make it easier for the reader. 
The reader can be lost so fast within the variable definitions. 
\answer{
We provide running examples with the definitions....
}

\item \label{Review.3.21}
for each word you are computing all the formulas values, 
so the comlexity is O(text length X number of formulas), 
what is the range of number of formulas ?
\answer{
  ... 
  The complexity is (\#solutions for all words) times 
  (\#number of user defined formulae). 
  We provide no limit on that.
  Practically, users define up to 10 formula. 
  We include this discussion in the paper .. 
}

\item \label{Review.3.22}
you are using R as tuple defining relation, 
then bellow you used it as set of feature vectors for the text. 
This is confusing the reader. 
I needed to read the definition in MATAr paper to understand 
what R is.
\answer{
  We cleaned the definitions... 
}

\item \label{Review.3.23}
d) In general, what was the data structure used? 
how would the annotated text be saved?
\answer{
  We used XXX data structures... 
}

\item \label{Review.3.24}
6- At section Results, comparing the development time 
did not strength the work. 
This is my first time to see researchers compare their 
algorithm in term of line of code and developing time. 
I would suggest comparing the algorithm complexity 
(linear , exponential, etc), and comparing how fast a user can 
accomplish similar jobs in the different applications. 
This can be done by assiging similar corpus annotation job 
and information extraction while computeing time needed by 
the user 
``as a survey like comparison for qualitative evaluation''. 
you can search for ``user interface system evaluation''. 
There is huristic evaluation where you give weight for job 
complexity and record time needed to complete it.

\answer{
By development time we meant the time it required the 
users to build an automated annotator similar to 
the MERF based annotator. 
We clarify that in the paper ...

We conducted a survey as requested and we compare the 
manual annotation time without MERF to the annotation
time with MERF. 
}

\item \label{Review.3.24}
Also, comparison with previous work done by (Zaraket ) 
in different vesion seems like comparing with your previous work 
only and not with others. 
It would be nice to compare others work too because user 
interface could be introduced in different view.
\answer{
 ?????   
}

\item \label{Review.3.24}
What is the data size of your evaluation? 
I couldn't find how presion and recal was counted. 
you mentioned books but did you covered all the text in those 
books ? and how long was it? you can define it in term of 
number of relation extracted , tags added, expesion processed, 
sentences,..etc
\answer{
  We include more information about the data and 
  the evaluation process... 
}

\item \label{Review.3.24}
7- At section Discussion, I strongly urge you to discuss 
WHY the precission and recall have these values.
\answer{
We extend the discussion section to discuss ... 
}

\item \label{Review.3.25}
Clarity:  
It was not very clear how the work was done 
``Methodology part''. 
Having ``Existing annotation and entity extraction tools'' 
under unspecified section is misleading the reader.

\item \label{Review.3.25}
unnumbered subsections was misleading me as a reader. 
\answer{
  Fixed. CLS file was not compatible with arabtex package. 
}

\item \label{Review.3.25}
table 1 at page 7 while referring to it only at page 9. 
I think it need to be closer to where it was mentioned in the text. 
\answer{
  Done.
}

\item \label{Review.3.25}
``MERF regular expressions support operators such as concatenation, zero or one, zero or more, one or more, up to $M$” what is M? 
it could be for example 0<=M<=|chunk| Section 4 “MERF”please add “Methedology”
\answer{
  $M$ is ... we clarified that in the paper. 
}

\item \label{Review.3.25}
Correctness: 
1- in the abstract, the sentence ``These techniques and 
tools require expertise in linguistics and programming and 
lack support of Arabic morphological analysis which is key to 
process Arabic text''
gives the meaning that ``it require lack support'' so I think it 
needs to be reformulated 
\answer{
  We rephrased the sentence as requested. 
}

\item \label{Review.3.25}
2-``defines tag types and''.// unify the font here and 
many other places in the paper. 
for example look at words(,correct,commercial)
\answer{
  Emphasize ... 
}

\item \label{Review.3.25}
Originality:  
This work will contribute in facilitating Arabic corpus 
annotation, if it was made available for other researchers in 
the field to access and use it, although there is no 
referencing to where it can be accessed. 
this is an improvement of a per-existing work ``MATAr: Morphology-based Tagger for Arabic''
\answer{
  The work is available on github... 
}

\end{enumerate}

### 2018-03-13 tex/merf/TALMajor/answer/Reviewer4.comments.tex
adding merf

\section*{Reviewer 4 comments}
\textit{Comments to the Author}\\

This work presents MERF; a GUI-enabled regular expressions 
based tool for entity and relational entity extraction, 
enhanced for Arabic text by enabling morphology-based matching.

\begin{enumerate}[leftmargin=0mm,label=\bfseries CommentR2.\arabic*]

\item \label{Review.4.1} 
Clarity:   
It is not clear how to associate code actions with 
subexpressions!
\answer{
  We clarify that... 
}

\item \label{Review.4.1} 
Correctness :  
- $Syn^k$ introduces a lot of noise especially if $k$ is large. 
\answer{
  7 is a limit... 
}

\item \label{Review.4.1} 
In the results section, there is no standard evaluation. 
The evaluation is based on comparisons with non-standard 
task-specific tools. 
Also, the reported precision and recall are very sensitive 
to the quality of the annotator!
\answer{
  Same answer to reviewer 3 results... 
}

\end{enumerate}

### 2018-03-13 tex/merf/TALMajor/answer/additionalComments.tex
adding merf

\section*{Additional Comments:} 
Please consider the following related efforts:

Y Benajiba et al. 2007. Anersys: An arabic named entity recognition system based on maximum entropy

Y Benajiba et al. 2008. Arabic named entity recognition using optimized feature sets.

A Pasha, et al. 2014. MADAMIRA: A Fast, Comprehensive Tool for Morphological Analysis and Disambiguation of Arabic.

A Shahrour et al. 2016. CamelParser: A System for Arabic Syntactic Analysis and Morphological Disambiguation.

\answer{
We discuss the work in the related section. 
}

### 2018-03-13 tex/merf/TALMajor/answer/answer.tex
adding merf

\documentclass[11pt]{article}
\usepackage[letterpaper,margin=1.5in,left=35mm,right=20mm, top=20mm, bottom=20mm ]{geometry}
\input{packages}

\tcbuselibrary{skins}
\usetikzlibrary{shadings}
\input{answerMacros}
\input{answerPackages}

\title{MERF: Morphology-based Entity and Relational Entity Extraction Framework for Arabic
 \\
~~ \\
  Response to Major Revision }
\date{}

\begin{document}
\maketitle

{\noindent Dear Editor-in-Chief, Associate Editor and expert Reviewers,}

We would like to thank you for your recommendations, valuable comments, 
and constructive observations. 
%
In this revised version of the paper we made every effort to address the 
concerns brought up in the reviews. 
%
We highlighted the changed parts in the paper with color blue. 
This document discusses the comments of the Reviewers 
in details, provides answers for them, and points to sections of the revised 
paper where the comments were addressed. 
\\
\noindent
Thank you.
\newline
\newline
\hrule
\newpage

\input{Reviewer1.comments}
~\newline
\hrule

\input{Reviewer2.comments}
~\newline
\hrule

\input{Reviewer3.comments}
~\newline
\hrule

\input{Reviewer4.comments}
~\newline
\hrule

\input{additionalComments}

%\input{other_notes}

%\newsavebox\mytempbib
%\savebox\mytempbib{\parbox{\textwidth}{
%\input{psqBMCaig.bbl}
%}}

\end{document}

### 2018-03-13 tex/merf/TALMajor/answer/editorComments.tex
adding merf

\section*{Editor-in-Chief Letter}

Dear Amin Jaber,

We are happy to inform you that your submission to the journal 
``Traitement automatique de l'arabe et des langues apparentées (TAL 58-3)''
has been accepted but requires major revisions.

You are invited to make major revisions to be considered for publication in 
the second review round.

Please submit your revised article before April 2, 2018 (midnight Paris time).

The reference of your submission is:
MERF: Morphology-based Entity and Relational Entity Extraction Framework for Arabic, 
A. Jaber [et al.] (sciencesconf.org:tal-58-3:171885)

Best Regards,
tal-58-3 team

https://tal-58-3.sciencesconf.org

### 2018-03-13 tex/merf/TALMajor/directionFigure.tex
adding merf

\transtrue
\setcode{utf8}
\setarab
\transfalse
\begin{figure}[tb]
\begin{center}
\relsize{-0.5}
%\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{\columnwidth}{!}{
    \begin{tabular}{|C{9.5cm}C{8cm}|} 
      \hline
  \begin{tabular}{R{9.5cm}}
  \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من} 
  \\ ~ \\
	$\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك} \RL{تقود} 
  \\ ~ \\
  $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا} $\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي} 
  \\
  \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  
  \\
  \multicolumn{1}{L{9.5cm}}{
   \arabfalse \transtrue
   \RL{من المستحيل ألا تلاحظ برج خليفة بالقرب من التقاطع الأول وأنت تقود سيارتك في شارع الشيخ زايد، حتّى وإن كانت هذه المرّة الأولى التي تسلك فيها هذا الطريق؛ يقع دبيّ مول على مقربة من هذا المبنى الذي يُعد الأطول في العالم.}
   \arabtrue \transfalse} 
  \\
  \multicolumn{1}{L{9.5cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.} 
  \end{tabular}
&
\resizebox{0.6\columnwidth}{!}{
  \input{figures/ergraphEx.tex}}
 \\ \hline
%\multicolumn{1}{c}{\multirow{1}{*}{\textbf{(a) Text with directions}}} &
%\multicolumn{1}{c}{\textbf{(b) Formula, matches, and entity-relation graph}} \\
    \end{tabular}
}
\end{center}
  \caption{Direction example with Arabic text, annotated with entities, transliteration, translation, and extracted relational entities in a graph} 
  \label{fig:intromotiv}
\end{figure}
\transtrue
\setcode{standard}

### 2018-03-13 tex/merf/TALMajor/discussion.tex
adding merf

The results show that \framework 
provides a friendly environment to develop entity and relational
entity extraction tasks with acceptable 
accuracy and runtime overheads compared to task specific applications. 
%
\framework requires the user to understand and interact with 
basic linguistic concepts such as readable values of morphological 
features, sequences, repetitions, and bounded repetitions. 
The user interacts with the MBF editor to specify basic concepts
and visualize their matches over highlighted text. 
%
Then, the user interacts with the MRE editor to specify 
sequences of the concepts and visualize the matches
in a graph, in conjunction with the highlighted text.

The two levels of interaction allow the user to separate between concepts 
that relate to word features, and more sophisticated entities 
that relate to sequences and context. 
%
The MBF, MRE, and user defined relations 
can be used to generate large annotated corpora in a fast manner. 
\framework visualization can be used later to edit the corpora 
and fix the annotations.

%\subsection{ Threats to validity}
%As with any GUI tool, it takes time for the user to get acquainted with the 
%user friendly interface. 
%Relating the subexpressions to their suggested names by \framework might 
%not be directly intuitive to the user. 
%We address the above two concerns with providing the users with 
%a short tutorial video on how to use the tool.

### 2018-03-13 tex/merf/TALMajor/implementation.tex
adding merf

In this Section, we describe the different implementation aspects of \framework. 
We cover the data model used to store the MBF and MRE based tag types, the MBF and MRE tags, and the generated actions. 
We explain the data structures used to hold the data from the files including MBF and MRE tag types, MBF and MRE matches, code action, and relations. 
Then, we describe the simulation of the MBFs and MREs to compute the matches, the construction of relations matches, and the execution of actions. 
We also describe how the interface is written with Qt and how the match tree and entity-relation graphs are visualized.

\subsection{Data Model}

\framework saves the tag types and the tags in a user friendly format using
the JavaScript Object Notation (JSON) data-interchange format~\cite{nolan2014javascript}.
The JSON interface is built on two structures.
The first structure is a collection of name and value pairs. An instance
of the structure is referred to as an object. 
The second structure defines ordered lists of values as arrays. 
The name in a name value pair is an identifier string 
and the value can be a string, a number, a Boolean (true or false), an object, 
or an array. 
%The file format is selected to make it easier to compute statistical metrics 
%necessary for CL and NLP tasks such as frequency and distance.

\subsubsection{Tag file}

The tag file keeps the paths of the separate text and tag type files. 
It also contains a list of tags with their tag type identifiers as well 
as their position in text in terms of number of characters from the beginning 
of text, length of the tag, and word index. 
In addition to the previous features, 
the MRE matches contain an object representing the match tree with 
the identifier ``match''.

\begin{figure}[h*]
\begingroup
    \fontsize{8pt}{12pt}\selectfont
\begin{verbatim}
{ "TagArray":
  [{"length": 3,"pos": 152,"source": 1,"type": "P","wordIndex": 29},...],
  "simulationTags":
  [{"formula": "direction","length": 33,"match": {...},"pos": 22,"source": 1},...],
  "TagTypeFile":"directions.stt.json",
  "file":"directions.txt",
  "textchecksum":211
}
\end{verbatim}
\endgroup
\caption{Tag file format}
\label{fig:tagfile}
\end{figure}

A sample tag file is shown in Figure~\ref{fig:tagfile}. 
The tag file includes the tag type file pointer(TagTypeFile), 
text file pointer (file), the list of MBF tags (TagArray), 
the list of MRE tags (simulationTags), 
and {\em textchecksum} which is used to ensure that the 
text is not corrupt. 
An MBF tag is defined by its position in the text (pos), length of the tag (length), 
tag type (type), word index (wordIndex), and the source of the tag (source). 
A simulation tag, referring to an MRE match, is defined by the formula (MRE), length, position and the match tree (match).

\subsubsection{Tag type file}

The tag type file contains two objects referring to MBF and MRE based tag types. 
Each object has an array value containing the user-defined tag types. 
Each tag type is stored as an object and contains the name, description, formula or expression, and the visualization legends.

\begin{figure}[h!]
\begingroup
    \fontsize{8pt}{12pt}\selectfont
\begin{multicols}{2}
\begin{verbatim}
{ "TagTypeSet":
  [{"Description":"numerical term",
   "Features":[{"Negation": "",
                "Relation": "isA",
                "Stem-Gloss": "first"},
                ... 
              ],
   "Tag": "U",
   "background_color": "grey",
   "bold": false,
   "font": 12,
   "foreground_color": "orangered",
   "italic": false,
   "underline": false
   },...
  ],

  "MSFs": [{
      "name": "direction","description": "",
  	  "MSF": {... {"MBF": "U","actions": "",
  	               "init": "","name": "s7",
  	               "parent": "s13","type": "mbf"} 
  	               ... 
  	         },
  	  "Relations": [...
  	           { "entity1": "s9","e1Label": "text",
  	             "entity2": "s14","e2Label": "text",
  	             "edge": "s3","edgeLabel": "text",
  	             "name": "r1"} 
  	             ... ],
      "bgcolor": "#9acd32","fgcolor": "#f0f8ff",
      "includes": "","members": "",
      "delimiter": true
     }]
}
\end{verbatim}
\end{multicols}
\endgroup
\caption{Tag type file format}
\label{fig:tagtypefile}
\end{figure}

Figure~\ref{fig:tagtypefile} shows a sample tag type file. 
The file includes a list of defined MBF tag types (TagTypeSet). 
Each tag type is defined by the fields name (Tag), description, 
morphological features (Features), foreground color, 
background color, bold, italic, underline, font size (font), and id. 
The morphological features are stored in an array of objects where each object refers to an MAT. 
An MAT object contains the fields negation, relation (isA or contains), and feature name and value pair.

The file also includes a list of defined MRE tag types and refers to them with the
notation Morphology-based Sequential Formulae (MSFs); the tool name of MRE. 
Each tag type is defined by the fields name, description, expression, 
semantic relations (Relations), code action information (includes and members), 
delimiter, and foreground and background color. 
The includes and members contain the libraries and global variables added to the actions by the user respectively.

The expression is a tree of objects and arrays whose depends on the definition of the user. 
For example, a sequence of sub-expressions is expressed as an array of objects. 
A sample MBF added to the expression is shown in the Figure	defined by the name (MBF), pre-match actions (init), on-match actions (actions), unique name assigned by \framework(name), name of parent expression (parent), and expression type (type) which is specified based on operation applied. 
The delimiter entry specifies whether the simulation should stop on a full stop or any punctuation mark.

Relations are saved in an array where each relation is defined by its name, the first entity (entity1) and its label (e1Label), second entity (entity2) and its label (e2Label), and the edge and its label (edgeLabel).

The user-defined actions (pre-match and on-match) are written into a cpp output file. 
The file includes the libraries included, global members declared, pre-match and on-match functions, and the function calls. 
A sample of the generated file is shown in Figure~\ref{fig:generatedactions}. 
The Figure shows a sample on-match function for a sub-expression named {\tt s2} (s2\_Match). 
The function takes an integer as a parameter. 
The integer refers to the numerical value of the match of {\tt s2} (s2\_number). 
The second function named {\tt number\_actions} contains the list of function calls. 
A sample call of the s2\_Match is shown with the parameter value 200.

\begin{figure}[h!]
%\begingroup
%    \fontsize{8pt}{12pt}\selectfont
\begin{multicols}{2}
\begin{Verbatim}[fontsize=\relsize{-2}] 
extern "C" void s2_onMatch(int s2_number) {
   isHundred = true;
   if(current == 0)  {
      currentH=s2_number;
   }
   else {
      if(!isKey) {
         currentH= current * s2_number;
         current = 0;
      }
      else {
         currentH = s2_number;
      }
   }
   isKey = false;
}
extern "C" void number_actions() {
   s5_preMatch();
   s4_preMatch();
   s2_preMatch();
   s2_onMatch(200);
   s4_onMatch();
   s4_preMatch();
   s3_preMatch();
   s0_preMatch();
   s0_onMatch(9);
   .
   .
   .
}

\end{Verbatim}
\end{multicols}
%\endgroup
\caption{sample generated actions file}
\label{fig:generatedactions}
\end{figure}

\subsection{Data structures}

In this section, we explain the data structures we used to store the text, MBF-based tag types, 
MRE-based tag types, relations, MBF matches, MRE match trees, and relations constructed.

\subsubsection{Arabic document}

\framework supports Arabic documents with UTF-8 encoding only. 
The input Arabic text is saved as a string. 
We process the text to build a word position to word index hash, 
a set of words that end by punctuation, and a set of words that end with full stop. 
The two sets are used in MRE simulation.

\subsubsection{MBF tag types}

We store the MBF tag types in a vector of {\tt TagType} class. 
Each tag type contains strings to store the name, description, foreground color, and background color. 
Boolean variables are used to track the bold and italic properties. 
The Boolean formula is stored as a vector of quadruples where each quadruple contains the feature, feature value, negation option, and predicate (isA or contains).

\subsubsection{MRE tag types}

The MRE tag types are stored in a vector of {\tt MSFormula} class. 
The regular expression is stored in a tree structure where each node refers to a sub-expression. 
Each sub-expression is represented by a class derived from a base class called {\tt MSF}. 
{\tt MSF} defines the common attributes of all sub-expressions including name, pre-match code actions, and on-match code actions. 
It also contains the common functions required to be implemented by all derived classes.

An MRE tag type is stored in a class called MSFormula. 
This class contains strings for description of the tag type, foreground color, 
background color, included libraries by user, and action global variables. 
It also contains a vector of MSF pointers referring to a sequence of sub-expressions. 
MSFormula contains a formula that maps the name of each sub-expression to a pointer of the relevant structure instance.

The expressions with the operations zero or one ($?$), 
zero or more ($*$), one or more ($+$), and zero up to x ($f$\textasciicircum$x$) 
are stored in a class called {\tt UNARYF}. 
This class holds the operation and an MSF pointer referring to the sub-expression subject to the operation. 
The expressions with the disjnuction ($|$) or conjunction ($\&$) operations 
are stored in a class called {\tt BINARYF}. 
This class holds the operation along with two MSF pointers referring to the two sub-expressions subject to conjunction or disjunction. 
The sequential expression is stored in a class called {\tt SequentialF}. 
It contains a vector of MSF pointers referring to a sequence of sub-expressions. 
An MBF-based sub-expression is stored in a class called {\tt MBF} which contains the name of the MBF tag type. 
All the classes introduced above are derived from the base class {\tt MSF}.

Moreover, the MRE tag type stores the relevant semantic relations defined by the user. 
The relations are stored in a vector and each relation is represented by a class {\tt Relation}. 
This class holds MSF pointers that refer to the three entity sub-expressions, and strings to identify the name of the relation and the labels of the three entities.

\subsubsection{MBF tags}

The MBF tags are stored in a multi-hash based on wordindex. 
We use a multi-hash because a single word can have multiple tags as stated before. 
Each tag is represented in an instance of the {\tt Tag} class. 
This class holds a pointer to the tag type, and integers for word index, position, and length. 
It also contains a flag to differentiate the automatic tag from a user one.

\subsubsection{MRE tags}
\label{subsubsec:mretag}

We store the MRE tag type matches in a vector of {\tt Match} class. 
In order to preserve the relation between the expression and the match tags, we use similar classes to store the match in a tree structure as well. 
Hence, we define the classes {\tt KeyM}, {\tt UnaryM}, {\tt BinaryM}, and {\tt SequentialM} referring to MBF, UNARYF, BINARYF, and SequentialF matches, respectively. 
All those classes inherit from base class {\tt Match}.

The base class contains common data including pointer to relevant MSF, 
operation if present, and a flag to differentiate the automatic tag from a user one. 
KeyM holds the matching word, tag name (key), position, and length. 
UnaryM contains a vector of {\tt Match} pointers and an integer to the limit in UPTO operation if applicable. 
BinaryM contains two {\tt Match} pointers. 
Both pointers are required in the conjunction case, but only one is used in the disjunction. 
The SequentialM class holds a vector of {\tt Match} pointers.

\subsection{MBF simulation}

The Arabic morphological analyzer, Sarf, provides a pure virtual function called $on\_match$. 
This function is triggered by the analyzer for every morphological solution computed for an input word. 
Through this function, we can access the morphological features of the solution including the stem, affixs, POS and gloss tags, and categories.

We derive the class {\tt SarfTag} from {\tt Stemmer}, the main class of Sarf, and implement the $on\_match$ function. 
For each solution found, we iterate over the MBFs and check whether the solution matches any of the MATs defined by the user. 
In case there is a match, we add a tag to the multi-hash referring to the word index. 
The tag contains a pointer to the tag type that refers to the matching MBF.

In order to minimize the computational cost of the $Syn^k$ feature, we compute the sets prior to the tag computation process. 
For each $CS$ (constant stem) and $k$ (order) pair, we compute the set of Arabic stems that are synonyms of the stem $CS$ of order $k$. 
When we find a $Syn^k$ based MAT in the tag computation process, we detect a match if a solution stem belongs to the relevant set.

\subsection{MRE simulation}

In this Section, we describe how we generate the NFA equivalent to the MRE. 
Then we describe how we simulate the NFA and construct the match trees.

\subsubsection{NFA generation}

In order to simulate the regular expression, we first generate the equivalent non-deterministic finite automata (NFA). 
We represent the generated NFA with the {\tt NFA} class which contains strings representing start and accept states, a multi-map for transitions, and a map from NFA state to relevant MSF. 
The state name is represented by a string of the form $q_0$, $q_1$, \dots
The transition multi-hash takes a concatenation of current state and label (epsilon or tag type name). 
It the transition exists, the hash returns one or more transition states. 
A sample transition input is $q_1|NONE$ and a sample value would be $q_2$ if the transition exists. 
The MSF map is used to track the source sub-expression of each state and whether the state was generated with a pre-match or on-match property. 
This information is important in the simulation of the NFA, and computation of the match tree. 
Thus, the map takes the state name as input and returns a pair of MSF pointer and string indicating the pre-match or on-match property.

\begin{figure}[h!]
\centering
\resizebox{0.7\columnwidth}{!}{
	\relsize{-3} \input{figures/starNFA.tex}
}
\caption{Star expression NFA}
\label{fig:starnfa}
\end{figure}

Figure~\ref{fig:starnfa} shows a sample NFA generated for a zero or more expression ($*$). 
The dotted block represents the sub-expression subject to the zero or more operation. 
$q_i$ is the start state of this expression and $q_j$ is the end state. 
The map entry of the start state is $s_k$ concatenated with $pre$. 
$s_k$ is the name of the stared expression and $pre$ refers to pre-match. 
Similarly, $on$ in the $q_j$ entry refers to on-match.

\subsubsection{NFA simulation}

We simulate the generated NFA in a recursive function. 
The function takes as parameters a pointer to the NFA, current state, and current word index and returns a {\tt Match} instance. 
At each stage, we check for possible transitions from current state based on empty strings ($\epsilon$) or MBF tags. 
Note that we can get the tags for current word index using the multi-hash used to store the MBF tag based on word index.

When we reach the accept state, we build the match tree backwards. 
At each stage, we check if the current state refers to a sub-expression pre-match or on-match property. 
On an on-match property, we generate the relevant match node compatible with the sub-expression type (Subsubsection~\ref{subsubsec:mretag}) and add it as a child node to returned {\tt Match} structure. 
In case of a pre-match property, we return the parent node of the {\tt Match} structure.

\subsection{Code action execution}

In this Section, we describe how we generate the code action file and then compile, link, and run it.

\subsubsection{Action file generation}

In order to execute the user-defined code actions, we first generate a C++ file. 
This file contains the includes, global declared variables, pre-match and on-match functions, and the sequence of function calls.

The libraries and global variables are directly inserted at the beginning of the file as entered by the user. 
Then, we generate the pre-match and on-match functions referring to each sub-expression in the MRE tree. 
In case the user uses the feature access API, we process the feature API and pass the relevant variable as a parameter to the function. 
For example, \cci{\$s0.number} allows the user to access the numerical value of the match of the sub-expression $s_0$. 
This call is processed and transformed into the form \cci{s0\_number}, and it is added as an integer parameter of the calling function.

After the MRE simulation stage, we traverse the match trees and generate the sequence of pre-match and on-match calls. 
The function calls are listed in a function named after the MRE tag type name such as \cci{number\_actions() \{\dots\}}. 
At each node in a match tree, we generate the pre-match call of the current node, call the generation method of children {\tt Match}, then generate the on-match call of the current node.

\subsubsection{Action file execution}

In order to execute the generated C++ file, we use shared object libraries and dynamic loading. 
The shared object library allows us to compile, load, and run the action file at run-time using dynamic linking loader system functions. 
The process requires us to create the shared library, load it, then use it by calling the target functions.

We create the library using a system call of the form: 
\newline
\cci{/usr/bin/g++ -fPIC -shared cppFile -o cppFile\_Path lib\_MREName.so}.
\newline
The \cci{cppFile} is the name of the action file generated, \cci{cppFile\_Path} is the path of the action file, and \cci{lib\_MRENAME.so} is the name of the output shared library.

We load the library using the \cci{dlopen} function provided by the dynamic library. 
The function takes the library name as input and returns a library pointer. 
As previously noted, the function calls are listed in a single function. 
In order to call this function, we use \cci{dlsym} which is a content extraction function. 
This function takes the library pointer and the name of the target method as parameters.

\subsection{\framework~interface}

We implemented \framework and its GUI as a C++ desktop application 
with the Qt GUI toolkit. 
The \framework GUI uses standard intuitive tagging facilities such as 
the application menus, the context menus, the mouse selection mechanism,
and the drag and drop mechanism to perform tagging and editing operation. 
The \framework GUI allows the user to create, manipulate, and analyze a tagging project.

In order to provide a user-friendly interface, we use {\tt QDockWidget} in \framework main window. 
This class provides the concept of dock widgets which can be moved into new areas and removed by the user. 
The main window contains 4 widgets referring to text visualization, 
tag list, tag description, and match tree and entity-relation graph. 
The main window displays the text with rich colors in a text browser pane. 
A companion pane shows a list of all the tags. 
A secondary companion pane shows details about the selected tag. 
A tabbed widget visualizes the match tree and entity-relation graph.

As for the MBF and MRE tag type editors, we use grids to organize the layout. 
The MBF and MRE are represented using the tree widget provided by Qt. 
We use combo box, label, edit text, and push button classes provided by Qt to 
add the other objects in the editor.

\subsubsection{Tree and graph visualization}

We use graphics scene and graphics view in order to visualize the match tree and entity-relation graph. 
The graphics scene provides a surface to manage 2D graphical items such as lines, text, and shapes. 
The graphics view provides a widget to display the content of the graphics scene. 
Our visualization method is based on an example provided by Qt showing how to implement nodes, and edges between nodes in a graph~\footnote{\url{http://qt-project.org/doc/qt-4.8/graphicsview-elasticnodes.html}}. 
However, Qt doesn't calculate the layout automatically but requires the user to position the elements.

In order to generate the tree or graph layout, we use the graphviz library. 
This library provides a variety of software for drawing attributed graphs and computes the layout using a set of common graph layout algorithms such as dot. 
First, we define our match tree or entity-relation graph using this library. 
We use commands such as \cci{agopen} to create a graph, \cci{agsafeset} to set graph attributes, 
\cci{agnode} to create a graph node, and \cci{agedge} to create a graph edge. 
After creating the graph, we compute a layout using the {\em dot} layout engine. 
This task is performed by calling the method \cci{gvLayoutJobs}. 
We use the generated coordinates of the nodes in the graphviz graph to position the nodes in \framework's main window graphics scene.

\subsection{Open source tool}

\framework is an open source tool present on google code under \cci{atmine} repository (\url{https://code.google.com/p/atmine/}). 
In addition to \framework, the repository contains entity extraction tasks developed by our research group. 
People are welcome to download and use the tools. 
We appreciate any feedback that we get and we try to improve the tools accordingly.

### 2018-03-13 tex/merf/TALMajor/introduction.tex
adding merf

{\em Computational Linguistics} (CL) is concerned with building 
accurate linguistic computational models. 
{\em Natural Language Processing} (NLP) is concerned with automating the 
understanding of natural language. 
CL and NLP tasks range from simple ones such as spell checking and typing 
error correction to more 
complex tasks including %text summarization, 
{\em named entity recognition} (NER), {\em cross-document analysis}, 
{\em machine translation}, 
and {\em relational entity extraction}~\cite{linckels2011natural,ferilli2011natural}.
Entities are elements of text that are of interest to an NLP task. 
Relational entities are elements that connect entities. 
{\em Annotations} relate chunks of text to {\em labels} denoting
semantic values such as entities or relational entities.
We refer to annotations and labels as {\em tags} and {\em tag types}, respectively, 
in the sequel.

Supervised and unsupervised empirical learning techniques tackle NLP and CL tasks. 
They employ machine learning without the need to manually encode the requisite 
knowledge~\cite{soudi2007arabic}. 
Supervised learning techniques require training corpora annotated 
with {\em correct} tags to learn a computational model. 
Supervised and unsupervised techniques require annotated reference corpora to 
evaluate the accuracy of the technique using metrics such as precision and 
recall~\cite{englishtreebank,arabictreebank,chinesetreebank}.

Researchers build %the required %seed 
training and reference corpora 
either manually, incrementally using learning techniques, or using knowledge-based 
annotation techniques that recognize and extract entities and relational 
entities from text. 
Knowledge-based techniques use linguistic and 
rhetorical domain specific knowledge encoded into sets of rules 
to extract entities and relational entities~\cite{soudi2007arabic}.
While existing annotation, entity, and relational entity 
extraction tools exist
\cite{chiticariu2010systemt% SystemT
,atzmueller2008rule%Textmarker
,urbain2012user%
,settles2011closing% dualist
,mmax2%MMAX
,brat% BRAT
},
most of them lack Arabic language support, and almost all of them
lack Arabic morphological analysis support
% that is key to Arabic NLP due to the rich morphological nature of Arabic
~\cite{habash2006arabic}. 
Fassieh~\cite{attia2009fassieh} is a {\em commercial} Arabic annotation tool
with morphological analysis support and text factorization. 
However, this tool lacks support for entity and relational entity extraction.

\input{directionFigure}

%These tasks take as input digital text documents from sources including literature, books, news, business reports, chat messages, and emails. 
%Some documents are originally typed in digital format such as emails. 
%Other documents are automatically digitized using techniques such as {\em optical character recognition} (OCR) and 
%{\em automatic speech recognition}~\cite{margner2012guide,rashwan2012robust}.

In this paper, 
we present \framework, a {\em morphology-based entity and relational entity
extraction framework for Arabic text }. 
\framework provides a user-friendly interface where the user defines tag types 
and associates them with \framework formulae that are regular expressions over 
\framework Boolean formulae. 
Boolean formulae are terms, negations of terms, and disjunctions of terms. 
Terms are matches to Arabic morphological features including 
prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym tags, 
and semantic categories. 
We discuss the importance of the morphological features supported in 
\framework terms in Section~\ref{sec:morph}; in brief, morphological preprocessing
is key to Arabic NLP.% due to the richness of Arabic morphology.

We illustrate the target of \framework using a simple example. 
Consider the text in Figure~\ref{fig:intromotiv} which contains
directions to Dubai Mall~\footnote{text taken from the Dubai Mall website
~\url{http://www.thedubaimall.com/ar/}}.
The framed words in the text are entities referring to names of people 
($n_1,n_2,n_3$), 
names of places ($p_1,\dots,p_7$), 
relative positions ($r_1,\dots,r_4$), 
and numerical terms ($u_1,u_2$). 
We would like to extract those entities, and then extract
the relational entities forming the graph in Figure~\ref{fig:intromotiv} 
where vertices express entities, 
and edges represent the relational entities.
\framework allows a regular user to do that interactively
in four phases that include morphological analysis, 
entity extraction based on morphological features, 
relational entity extraction, and entity cross-reference.

\framework regular expressions support operators such as concatenation, 
zero or one, zero or more, one or more, up to $M\in\mathbb{N}^+$, and logical 
conjunction and disjunction. 
An editor allows the user to associate an action with each sub-expression.
The user specifies the action with C++ code and uses an API to access information related to the matches such as text, position, length, morphological features, and numerical value.

\framework takes an Arabic text, a set of user-defined \framework Boolean 
formulae, and a set of user-defined \framework regular expressions. 
\framework computes the morphological solutions of the words in the 
input text then computes matches to the Boolean formulae. 
\framework then generates a {\em non-deterministic finite state 
automata} (NDFSA) for each expression and simulates it with the 
sequence of Boolean formulae matches to compute the regular 
expression matches. 
\framework generates executable code for the actions associated with
the regular expressions, 
compiles, links, and executes the generated code 
as shared object libraries. 
Finally, \framework constructs the semantic relations and 
cross-reference between entities. 
\framework also provides visualization tools to present the matches, 
and estimate their accuracy with respect to reference tags.

\subsection{Existing annotation and entity extraction tools}

Researchers proposed and evaluated empirical and 
knowledge-based techniques to extract entities and relational entities
from text. 
We briefly review them here and we discuss them and compare to them
in Section~\ref{sec:related}. 
The work in~\cite{ekbal2010named} presents a language independent 
approach for NER extraction using {\em support vector machines}. 
The work in~\cite{abdelrahman2010integrated} integrates 
a semi-supervised bootstrapping pattern recognition technique, 
and a supervised classifier based on {\em conditional random fields}
to solve NER problems.

Knowledge-based techniques such 
as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars 
with morphological stemming to perform NER. 
%The work in~
\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
events, and relations amongst them from Arabic text using a hierarchy of manually built 
finite state machines driven by morphological features and graph 
transformation algorithms. 
Such techniques require advanced linguistic and programming expertise. 
QARAB is a question answering system that takes 
Arabic natural language queries and provides short answers~\cite{hammo2002qarab}.

Researchers also proposed systems for automatic IE based 
on user specifications. 
CPSL is a common pattern specification language for finite-state 
grammar~\cite{appelt1998common}. 
%\framework is an extension to CPSL with action execution and relation construction. 
%It defines three sections for declaration, rule definition, and macros. 
The work in \cite{chiticariu2010systemt} presents SystemT, a system based on 
an algebraic Approach to Declarative information extraction (IE). 
%It uses a declarative rule language and an optimizer. 
TEXTMARKER is a rule-based IE system designed to extract structured data from 
text \cite{atzmueller2008rule}. 
\cite{urbain2012user} presents a user-driven relational model 
requiring a user natural language query to extract entities and relational entities.

This work significantly extends the position 
paper~\cite{JaZaMatar} that allows for manual, and 
basic morphology annotation.
\framework enables a regular user to incrementally create 
complex annotations for Arabic text based on automatic 
extraction of morphological tags through a user friendly interactive interface. 
\framework has the following advantages.
\begin{itemize}
  \item \framework provides a novel and intuitive visual interface to build Boolean formulae over morphological features, 
    build regular expressions over the resulting Boolean formulae, and thereafter compute automatic tags.
  \item To our knowledge, this morphology-based framework  is the first for Arabic entity and relational entity extraction.
  \item \framework provides the user with the ability to rapidly create annotated Arabic text corpora with sophisticated morphology-based tags.
\end{itemize}

In \framework, we make the following contributions.
\begin{itemize}
  \item \framework enables the user to define relations in a simple manner
    and automatically detects relational entities matching the user defined relations. 
  \item \framework enables the user to associate subexpressions
    with code actions, and executes the code action 
    when a corresponding match is found. 
    It also provides an API to enable access to match 
    features such as text, position, length, numerical value, and morphological features.
  \item \framework enables the user to tag words based on a synonymic relation using the $Syn^k$ feature.
\end{itemize}

\input{directionMBFMRE}

In the rest of this paper,
Section~\ref{sec:motivation} presents a motivational example. 
Section~\ref{sec:morph} introduces Arabic morphological analysis 
and its important role in Arabic NLP. 
%We present preliminaries necessary to understand the work 
%in Section~\ref{sec:preliminaries}. 
Section~\ref{sec:framework} explains the \framework methodology. 
Section~\ref{sec:gui} presents the visual and user friendly interface of \framework. 
Section~\ref{sec:related} presents and discusses related work.
Section~\ref{sec:results} presents the evaluation results. 
Finally, we conclude and discuss future work in Section~\ref{sec:conclusion}.

### 2018-03-13 tex/merf/TALMajor/macros.tex
adding merf

\documentclass[english,utf8x]{article-hermes}

%%%%%%%% packages %%%%%%%%%%
\usepackage{xspace}
\usepackage{arabtex}
\usepackage{array}
\usepackage{tikz}
\usepackage{hyperref}
\hypersetup{
    colorlinks=false,
    pdfborder={0 0 0},
}
\usepackage{booktabs}
\usepackage[ruled]{algorithm2e}
\usepackage{etex}
\usepackage{graphicx}
\usepackage{framed}
\usepackage{graphics}
\usepackage{booktabs}
\usepackage{multirow}
\usepackage{amssymb}
\usepackage{amsmath}
\usepackage{savesym}
\savesymbol{bigtimes}
\usepackage{mathabx}
\restoresymbol{TXF}{bigtimes}
\usepackage{utf8}
\usepackage{subfigure}
\usepackage{todonotes}
\usetikzlibrary{shapes,arrows}
\usepackage{color,soul}
\usepackage{verbatim}
\usepackage{array}
\usepackage{multicol}
\usepackage{fancyvrb}
\usepackage{relsize}
\usepackage{wasysym}
\usepackage{pbox}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%%%%%%%%%% Macros %%%%%%%%%%
\def\framework{\textsc{MERF}\xspace}
\newcommand{\itodo}[1]{\todo[inline,color=green]{\tiny #1}}
\newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{C}[1]{>{\centering\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcolumntype{R}[1]{>{\raggedleft\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
\newcommand{\TinyRL}[1]{{\RL{#1}}}
\newcommand{\cci}[1]{{\small \texttt{#1}}}
%%% RL commands
\newcommand{\utfrl}[1]{\setcode{utf8}\RL{#1}\setcode{standard}}
\newcommand{\notrutfrl}[1]{\transfalse\setcode{utf8}\RL{#1}\setcode{standard}\transtrue}
\newcommand{\notrrl}[1]{\transfalse\RL{#1}\transtrue}
\newcommand{\noarrl}[1]{\arabfalse\RL{#1}\arabtrue}
\def\showvrb#1{%
\texttt{\detokenize{#1}}%
}
%%% define cal
\makeatletter
\DeclareRobustCommand*\cal{\@fontswitch\relax\mathcal}
\makeatother
% done
%other 
% correct bad hyphenation here
\hyphenation{op-tical net-works semi-conduc-tor}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\submitted[15/10/2017]{TAL~58-03}{R}

\title[\framework]{\framework: Morphology-based Entity and Relational Entity Extraction Framework for Arabic}

%\author{Firstname$_1$ Lastname$_1$\fup{*} \andauthor Firstname$_2$ Lastname$_2$\fup{**}  } 
%\address{%
%\fup{*} First affiliation\\
%\fup{**} Second affiliation
%}

\author{}
%\author{Amin A. Jaber \andauthor Fadi A. Zaraket}

\address{}
%\address{American University of Beirut, Beirut 1107 2020, Lebanon}

\abstract{
\input{abstract}
}

\keywords{
Arabic,
information extraction,
natural language processing,
entity,
relational entity,
information visualization,
tagging.
}

\motscles{
%mot-cl\'e$_1$,
%mot-cl\'e$_2$,
%mot-cl\'e$_3$.
}

\resume{
%Le r\'esum\'e est \`a \'ecrire ici.
}

\begin{document}

\maketitlepage

### 2018-03-13 tex/merf/TALMajor/relatedwork.tex
adding merf

In this section we review the literature on 
entity and relation IE and on automatic and manual annotation
techniques and compare to \framework.

%SystemT\cite{chiticariu2010systemt}, 
%TEXTMARKER\cite{atzmueller2008rule}, 
%Urbain\cite{urbain2012user}, 
%QARAB\cite{hammo2002qarab}

\begin{table}[tb!]
\caption{Comaprison of \framework with SystemT, 
TEXTMARKER, 
Urbain, 
QARAB}
\resizebox{\columnwidth}{!}{
\begin{tabular}{l|c|c|c|c|c}
%{p{2.7cm}|p{.7cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}}
Features & \framework & SystemT & TEXT\linebreak MARKER & Urbain & QARAB\\ \hline
Query type & MRE & AQL & matching \linebreak rules & natural language & natural language \\ \hline
Morphology support& $\checked$ & - & - & OpenNLP & Parser \\ \hline
Relations & $\checked$ & - & - & $\checked$ & - \\ \hline
Actions & $\checked$ & - & - & - & - \\ \hline
Editor & $\checked$ & - & $\checked$ & - & - \\ \hline
Tag visualization & $\checked$ & - & $\checked$ & - & - \\ \hline
Graph visualization & $\checked$ & - & - & - & - \\
\end{tabular}
}
\label{tab:iecomp}
\end{table}

{\bf Information Extraction.~}
The common pattern specification language (CPSL) targets system independent 
IE specifications. 
CPSL consists of 
(1) a declaration part specifying names and labels, 
(2) a rule definition part specifying patterns, regular expressions and associated actions, and 
(3) and macro text substitution part~\cite{appelt1998common}.
\framework extends CPSL with Arabic morphological features, code actions, and user defined relations.

SystemT~\cite{chiticariu2010systemt}
aims to overcome the performance and expressivity limitations of CPSL.
It is based on an algebraic approach to declarative information extraction,
uses the declarative annotation query language (AQL), and uses
an optimizer to generate high performance execution plans for the AQL rules. 
%The AQL query is translated into an algebraic expression and an optimizer selects an execution plan. 
%\framework overcomes the limitations described by SystemT. 
\framework supports multiple tags per word, and supports the MRE conjunction operator
which allows for overcoming the overlapping annotation problem. 
%es the overfor each word and this overcomes the lossy sequencing caused 
%by random annotation dropping. 
%\framework introduces the conjunction operation in the MRE to overcome the expressivity limitations of expressing rules with overlapping annotations.

TEXTMARKER is a semi-automatic rule-based IE system 
for structured data acquisition\cite{atzmueller2008rule}.
%TEXTMARKER requires user defined rules that consider features of the text 
%However, constructing the rules requires manual work of a domain specialist. 
Both TEXTMARKER and \framework provide the user with GUI editor and result visualizer. 
%\framework also differs in that it provides graph visualization for regular expression and user defisemantic relation matches.

The work in~\cite{urbain2012user} presents a user 
driven relational model and targets entity and relation extraction. 
The user enters a natural language query, and uses the OpenNLP toolkit to 
extract tags and relations from the query. 
Then it extends the extracted entities into a query model,
and uses the query model to extract matches from the document. 
Similar to \framework, the system constructs entities and relations.

QARAB is an Arabic question answering system that 
takes an Arabic natural language query and provides short answers for it~\cite{hammo2002qarab}. 
QARAB uses traditional information retrieval techniques and an outdated Arabic NLP 
analyzer that computes limited features of Arabic words compared 
to the morphological analysis of \framework. 
QARAB then classifies the query against a set of known question types.

Table~\ref{tab:iecomp} summarizes the comparison between \framework and
other systems. \framework differs in that it provides 
code actions, user defined relations, and an interactive 
graph visualization of the relational entities. 
It also differs in that it fully supports Arabic morphological analysis
while only QARAB supports Arabic linguistic features using a parser, and 
the work in ~\cite{urbain2012user} uses OpenNLP that currently lacks full
support for Arabic morphological features. 
Similar to TEXTMARKER, \framework has the advantage of providing 
a user friendly interactive interface to edit the entity and relational 
specifications and visualize the results.

DUALIST is an annotation system for quickly building classifiers for 
text processing tasks using active learning and semi-supervised learning~\cite{settles2011closing}. 
The classifier is interactive as it queries the annotator 
for entity detetction correction and annotation correction. 
\framework doesn't support classification tasks. 
However, \framework provides an interactive GUI where the user can edit MBF and MRE tags. 
This interactive environment contributes to the regular expression 
extraction and semantic relation construction which increases the overall accuracy.

WordNet is a lexical reference system that mimics human lexical memory 
%inspired by psycholinguistic theories 
and that relates words based on their semantic values and their functional 
categories: nouns, verbs, adjectives, adverbs, and function words~\cite{intro}. 
The $Syn^k$ feature in \framework is inspired by WordNet.

{\bf Annotation tools.~}
An overview of annotation tools and their Arabic-English word alignment issues 
concludes with a set of rules and guidelines needed in an Arabic annotation alignment tool
~\cite{kholidy2010towards}. 
The work in \cite{dukes2011supervised} presents a collaborative effort towards morphological 
and syntactic annotation of the Quran. 
\cite{dorr2010interlingual} presents a framework for interlingual 
annotation of parallel text corpora with multi-level representations. 
\cite{kulick2010consistent} presents the integration of the 
Standard Arabic Morphological Analyzer (SAMA) into the workflow of the Arabic Treebank.

MMAX2 is a manual multi-level linguistic annotation tool with an XML 
based data model~\cite{mmax2}. 
It enables the user to create, browse, visualize, and query annotations
and may be able to resolve coreference tags. 
BRAT~\cite{brat} and WordFreak~\cite{wordfreak} are manual 
multi-lingual user friendly web-based annotators that allow the construction 
of entity and relation annotation corpora~\cite{brat}. 
They can be extended through plug-ins to enable automatic annotators 
and customized annotation visualization specifications.
Knowtator~\cite{ogren2006knowtator} is a general purpose incremental text annotation tool 
implemented as a Prot\'eg\'e~\cite{gennari2003evolution} plug-in. 
Prot\'eg\'e is an open-source platform with a suite of tools to construct domain 
models and knowledge-based applications with ontologies. 
However, it doesn't support the Arabic language.

\framework differs from MMAX2, BRAT, WordFreak, and Knowtator in that it is 
an automatic annotator that allows manual corrections 
and sophisticated tag type and relation specifications over
Arabic morphological features.

The work in \cite{smrz2004morphotrees} presents %the extension of TrEd, 
a customizable general purpose tree editor, with the Arabic MorphoTrees annotations. 
The MorphoTrees present the morphological analyses in a hierarchical organization 
based on common features.

Task specific annotation tools such as~\cite{alrahabi2006semantic}
uses enunciation semantic maps to automatically annotate 
directly reported Arabic and French speech. 
AraTation is another task specific tool for semantic annotation of 
Arabic news using web ontology based semantic maps~\cite{saleh2009aratation}.
We differ in that \framework is general, and not task specific, and it uses 
morphology-based features as atomic terms.
Fassieh is a commercial Arabic text annotation tool that enables the production of large 
Arabic text corpora~\cite{attia2009fassieh}. 
The tool supports Arabic text factorization including morphological analysis, POS tagging, 
full phonetic transcription, and lexical semantics analysis in an automatic mode. 
Fassieh is not directly accessible to the research community and requires commercial licensing. 
\framework is open source and differs in that it allows the user to build tag types 
and extract entities and relations from text.

### 2018-03-13 tex/merf/merf/abstract.tex
adding merf

\vspace{0.5cm} {\em Natural Language Processing} is concerned with automating the understanding of natural language. 
Morphological analysis is key to Arabic natural language processing due to the morphological richness of Arabic.
Researchers proposed and evaluated knowledge-based and empirical techniques to extract entities and relations from text. 
Knowledge-based techniques require advanced linguistic and programming expertise. 
Empirical techniques require large training and reference corpora to learn and evaluate computational models respectively.

In this work, we present a morphology-based entity and relational entity information extraction framework for Arabic text. 
The framework provides a user-friendly interface where the user defines tag types 
and associates them with regular expressions defined over Boolean formulae. 
Boolean formulae are terms, negations of terms, and disjunctions of terms where terms are matches to Arabic morphological features. 
The framework introduces a semantic feature that relates words based on synonymity. 
The framework allows the user to associate an action with each regular sub-expression and to define semantic relations. 
The framework uses an in-house Arabic morphological analyzer to compute morphological matches, computes regular expression matches, and then builds the relations across matches. 
We evaluated our work with several case studies and compared with existing application-specific techniques.

### 2018-03-13 tex/merf/merf/gui.tex
adding merf

\begin{figure*}[tb]
  \centering
  \includegraphics[width=0.9\textwidth]{figures/mbfedit}
  \caption{\framework tag type Boolean formula editor.}
  \label{f:bfe}
\end{figure*}

\framework provides a user friendly interface to specify the 
atomic terms, the \framework Boolean formulae, the \framework regular expressions, 
the tag types, and the legends. 
The \framework GUI also allows the user to modify and correct the 
resulting tag set $R$.
The \framework GUI allows the user also to compute accuracy results 
that compare different tag sets. 
The accuracy results serve well as inter annotation agreement results
when the tag sets come from two human annotators, 
or as evaluation results for learning and information extraction techniques.

The snapshot in Figure~\ref{f:tagger} shows the \framework GUI with 
the tag type color sensitive view, the tag list view, the tag description view, 
and the match tree. 
In the shown example, the user specified three MBF-based tag types with labels ``HUNDRED'',
``KEY'', and ``DIGITSTENS''. 
The ``HUNDRED'' tag type is based on a simple MBF formula that inspects the gloss tag of 
the stem of the word and checks whether it is {\tt hundred}. 
The ``KEY'' tag type is built on an MBF formula that computes a disjunction 
between the gloss tags of the stems being {\tt thousand/s}, {\tt million/s}, and {\tt billion/s}. 
The ``DIGITSTENS'' tag type is based on a disjunction formula between the gloss tags of the stems 
being {\tt one}, {\tt two},\dots, {\tt ten}, and {\tt twenty}, {\tt thirty},\dots, {\tt ninety}. 
The user also defined an MRE-based tag type with the label ``number''. 
The ``number'' tag type is based on the formula $\left(\textit{HUNDRED}|\textit{KEY}|\textit{DIGITSTENS}\right)+$. 
In words, the formula defines the occurence of one or more consequent tags of the three tags.

The context sensitive menus in Figure~\ref{f:tagger} allow the user to tag 
a selected word or chunk differently or to entirely remove the tag.

The \framework GUI also allows manual tag types that are not based on morphological features.
These tags enable the users to build their own reference corpora without 
help from the morphological analyzer.

\begin{figure*}[tb]
  \centering
  \includegraphics[width=0.9\textwidth]{figures/msfedit}
  \caption{\framework tag type regular expression editor.}
  \label{f:sfe}
\end{figure*}

\subsection{Tag type Boolean formula editor}
The morphology based tag type editor shown in Figure~\ref{f:bfe}
allows the user to write a tag type Boolean formula in a user friendly manner. 
The user first specifies atomic terms by selecting 
a feature from ${\cal F}$. 
The pattern is a regular expression that filters the feature values.
This implicitly helps the user specify 
an {\em exact} versus a {\em partial} match of the  desired 
morphological feature value.

Then the user can add the selected feature values to the atomic terms 
under the tag type name. 
The feature column has a context sensitive menu that allows negating 
the term.
The value column has a context sensitive menu that can switch the 
operation between the values in 
${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$.
The right pane shows a description of the tag type and a set of legend 
descriptors.

For example, the user can select the feature {\em category} and associate it 
with the value {\em temporal unit}. 
Multiple feature/value pairs can be included in a single tag type 
definition with a disjunction semantics.

\subsection{Tag type regular expression editor}
The morphology based tag type editor shown in Figure~\ref{f:sfe} allows the 
user to define a tag type regular expression in a user friendly manner. 
The user first adds the required MREs to the formula under the tag type name 
by selecting a label from ${\cal T}$ under MBFs. 
The Boolean formula of a highlighted tag type is shown in the table on the lower left pane. 
Each selected label is associated with an automatic name. 
The regular expression tree features the name, label, and operation for each member MRE.

{\tt NONE} is a special tag type with no Boolean formula. 
It tags all the words that are not tagged with any of the morphology based tag types. 
{\tt NONE} is defined by default and can be used to introduce flexibility and noise tolerance into the formula. 
Moreover, a defined MRE is directly added to the MRE list enabling recursion 
in formula definition.

Then, the user can construct the formula by applying operations to the selected MREs. 
To do so, the user selects one or two MREs then selects an operation from the ones shown 
in the lower left of the view. 
Without adding operations, the formula is a sequence of the selected MREs. 
The right pane shows a description of the tag type and a set of legend 
descriptors.

For example, the user can add the {\tt DIGITSTENS}, {\tt KEY}, and {\tt HUNDRED} MREs to the formula. 
Then, he/she selects the {\tt DIGITSTENS} and {\tt KEY} MREs and apply the disjunction ( $|$ ) 
operation to them. 
Similarly, he/she can apply the disjunction operation to the disjunction MRE formed before and {\tt HUNDRED}. 
Last, the user selects the disjunction formula and applies the Kleene star. 
Thus, we get the regular expression shown in Figure~\ref{f:sfe}.

\begin{figure}[h!]
  \centering
  \includegraphics[width=0.45\textwidth]{figures/actions}
  \caption{\framework MRE action sample.}
  \label{f:actions}
\end{figure}

In order to add computational actions to an MRE, 
the user selects to edit the actions in the MRE tag type editor. 
The view shown in Figure~\ref{f:actions} allows the user to specify 
the actions in a user friendly manner. 
\framework enables the user to access the MBF match features easily and define the pre-match actions. 
\framework also enables the user to edit the formula declarations and library includes.

\subsection{Analysis}
In addition to automatic and manual tagging, \framework 
allows comparing tag sets and tag types applied to the
same input text. 
The \framework comparator takes as input two tag sets $R_1$ and $R_2$ and 
two tag type sets ${\cal T}_1$ and ${\cal T}_2$. 
It produces a difference view for the tag types and 
a difference view for the tag sets. 
The tag type difference view shows the common tag types ${\cal T}_1 \cap {\cal T}_2$,
the tag types in ${\cal T}_1$ and not in ${\cal T}_2$, 
and the tag types in ${\cal T}_2$ and not in ${\cal T}_1$.

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{figures/diff}
  \caption{\framework comparison and accuracy results view.}
  \label{f:diff}
\end{figure*}

Similarly, the tag set difference view shows $R_1\cap R_2$, $R_1/R_2$ and $R_2/R_1$. 
The tag set difference view, as shown in Figure~\ref{f:diff} also shows the precision, 
recall and F-measure between the two sets. 
The metrics can be computed based on several predicates. 
The ``Intersection'' predicate returns true if a tag from $R_1$ intersects in text $T$ 
with a tag in $R_2$. 
The ``Exact'' predicate returns true if a tag from $R_1$ exactly matches a tag
in $R_2$. 
The ``A includes B'' predicate returns true if a tag from $R_1$ contains a tag from $R_2$. 
Finally, the ``B includes A'' predicate returns true if a tag from $R_2$ contains 
a tag from $R_1$.

In the difference view panes, the user can select a difference tag and accept it, or reject
it to build a corrected corpora.

### 2018-03-13 tex/merf/merf/implementation.tex
adding merf

We implemented \framework and its GUI as a C++ desktop application 
with the Qt GUI toolkit. 
Figure~\ref{f:tagger} presents the main window of the \framework GUI.

The \framework GUI uses standard intuitive tagging facilities such as 
the application menus, the context menus, the mouse selection mechanism,
and the drag and drop mechanism to perform tagging and editing operation.

The \framework GUI allows the user to create, manipulate, and analyze a tagging project. 
The {\em File} menu is used to create, close, or open an existing project. 
The {\em tags} menu enables the user to manually define and edit general purpose tag types.
The user can define and edit the MBF and MSF based tag types and 
trigger the automatic morphology based simulator from the {\em sarf} menu. 
The user performs comparison and analysis of two tag sets 
using the {\em Analyse} menu.

The main window displays the text with rich colors in a pane. 
A companion pane shows a list of all the tags. 
A secondary companion pane shows details about the selected tag.

\subsection{Data Model}

The \framework tool saves the tag types and the tags in a user friendly format using
the JavaScript Object Notation (JSON) format.
The JSON interface uses a tree format based on name value pairs, where the name 
is an identifier and the value is either a sequence of values of a nested set of values. 
The nested set of values is easily defined using nested open close braces.

The tag files keep pointers to the separate text and tag type files. 
Then they contain a list of tags with their tag type identifiers as well as their location
in text in terms of characters from the beginning of text, length of the tag, 
and in terms of word counts.

The file formats are selected to make it easier to compute statistical metrics 
necessary for CL and NLP tasks such as frequency and distance.

\begin{figure}[h*]
\begingroup
    \fontsize{8pt}{12pt}\selectfont
\begin{verbatim}
{ "TagArray":
  [{"length":5,"pos":0,"source":1,"type":"Place"},...],
  "TagTypeFile":"/home/autoexample.stt.json",
  "file":"/home/politics.txt",
  "textchecksum":696
}
\end{verbatim}
\endgroup
\caption{Tag File Format}
\label{fig:tagfile}
\end{figure}

A sample tag file is shown in Figure~\ref{fig:tagfile}. 
The tag file includes the list of tags (TagArray), tag type file (TagTypeFile), 
text file (file), and {\em textchecksum} which is used to ensure that the 
text is not corrupt. 
A tag is defined by its position in the text (pos), length of the tag (length), 
tag type (type), and the source of the tag (source).

\begin{figure}[h*]
\begingroup
    \fontsize{8pt}{12pt}\selectfont
\begin{verbatim}
{"TagTypeSet":
  [{"Description":"Detects all the adjectives",
   "Features":[{"Stem-POS" : "ADJ" },... ],
   "Tag" : "Adjectives",
   "background_color":"#f08080",
   "bold" : false,"font" : 12,
   "foreground_color":"#778899",
   "id" : 3,"italic" : false,
   "underline" : false},...
  ]
}
\end{verbatim}
\endgroup
\caption{Tag Type File Format}
\label{fig:tagtypefile}
\end{figure}

Figure~\ref{fig:tagtypefile} shows a sample tag type file. 
The file includes the list of defined tag types (TagTypeSet). 
Each tag type is defined by the fields name (Tag), description, 
morphological features (Features), foreground color, 
background color, bold, italic, underline, font size (font), and id. 
The morphological features are stored in an array of name/value pairs.

### 2018-03-13 tex/merf/merf/introduction.tex
adding merf

\transtrue
\IEEEPARstart{C}{\em omputational Linguistics} (CL) is concerned with building accurate linguistic computational models. 
{\em Natural Language Processing} (NLP) is concerned with automating the understanding of natural language. 
CL and NLP tasks range from simple tasks such as spell checking and typo-error correction to more complex tasks including text summarization, {\em named entity recognition} (NER), {\em cross-document analysis}, {\em machine translation}, and {\em entity-relational information extraction}~\cite{linckels2011natural,ferilli2011natural}.
These tasks take as input digital text documents from sources including literature, books, news, business reports, chat messages, and emails. 
Some documents are originally typed in digital format such as emails. 
Other documents are automatically digitized using techniques such as {\em optical character recognition} (OCR) and {\em automatic speech recognition}~\cite{margner2012guide,rashwan2012robust}.

{\em Entities} are elements of text that are of interest to an NLP task. 
{\em Relational entities} are elements of text that connect entities. 
{\em Annotations} (referred to as {\em tags} in the sequel) relate a chunk of text to a {\em label} (a {\em tag type}) denoting a semantic value such as an entity or a relational entity.

\setcode{utf8}
\setarab
\transfalse
\begin{figure*}[tb]
%  \begin{center}
%\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{2\columnwidth}{!}{
    \begin{tabular}{C{12cm} C{12cm}}
  \begin{tabular}{R{12cm}}
  \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من} 
  \\
	$\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك} \RL{تقود}  
  \\
  $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا} $\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي} 
  \\
  \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  
  \\
  \multicolumn{1}{L{12cm}}{
   \arabfalse \transtrue
   \RL{من المستحيل ألا تلاحظ برج خليفة بالقرب من التقاطع الأول وأنت تقود سيارتك في شارع الشيخ زايد، حتّى وإن كانت هذه المرّة الأولى التي تسلك فيها هذا الطريق؛ يقع دبيّ مول على مقربة من هذا المبنى الذي يُعد الأطول في العالم.}
   \arabtrue \transfalse} 
  \\
  \multicolumn{1}{L{12cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.} 
  \\
  \end{tabular}
&
\resizebox{0.8\columnwidth}{!}{
  \relsize{-1} \input{figures/ergraphEx.tex}}
  \\
\multicolumn{1}{C{12cm}}{\multirow{1}{*}{\textbf{(a) Text with directions}}} &
\multicolumn{1}{c}{\textbf{(b) Formula, matches, and entity-relation graph}} \\
    \end{tabular}
}
%  \end{center}
  \caption{Text, formula, and match \framework example}
  \label{fig:intromotiv}
\end{figure*}
\transtrue
\setcode{standard}

In this work, we address entity and relational entity extraction from Arabic text using morphological analysis. 
For example, given the text in Figure~\ref{fig:intromotiv}(a), we would like to extract entities and relations that form the graph in Figure~\ref{fig:intromotiv}(b). 
The text in Figure~\ref{fig:intromotiv}(a) contains directions to Dubai Mall taken from the mall website~\footnote{~\url{http://www.thedubaimall.com/ar/}}. 
The framed words in the text are target entities referring to names of people, 
names of places, relative position, and numerical terms. 
From those entities, we extract the graph shown in Figure~\ref{fig:intromotiv}(b) where vertices express entities, and edges represent the relations between those entities. 
This task is accomplished in four phases that include morphological analysis, entity extraction based on morphological features, relation construction, and entity cross-reference.

\subsection*{Morphological analysis}

The Arabic language is morphologically rich. 
Short vowels, also known as diacritics, are almost always omitted in Arabic text
and inferred by human readers~\cite{habash2006arabic}. 
For example, the word \RL{'sd}%
                              ~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
can be interpreted as \RL{'sad} {\tt ``lion''} with a {\em fatha} on the letter \RL{s} or 
\vocalize \RL{'sodd} (I block) with a 
{\em damma} on the letter \RL{s} and {\em shadda} on the letter \RL{d}.
Consequently, {\em morphological analysis} is key to Arabic CL and NLP 
even for simple tasks 
such as tokenization and stemming\cite{arabicmorph}.
Tokenization requires morphological analysis in Arabic because a subset of the Arabic letters are 
non-connecting letters and do not require a space after them to provide a visual separation 
from the next letter. 
For example, in the sentence \setcode{utf8}\RL{ذهب الولدإلى المدرسة} (the kid went to the school)
\transfalse
the letter \RL{ى} is non-connecting and the two words 
\RL{إلى} and \RL{الولد}
are visually separable, 
yet there is no space character between them. 
\setcode{standard}
\transtrue

Given an Arabic word delimited by white space and punctuation, 
a {\em morphological analyzer} returns the internal structure of the word. 
The word structure is composed of several morphemes including 
{\em affixes} ({\em prefixes} and {\em suffixes}), and 
{\em stems}~\cite{arabicmorph}. 
The analyzer returns a set of morphological solution vectors with features
such as {\em prefix, stem, suffix}, and part of speech (POS), gloss, and category tags. 
\vocalize
For example, for the word \RL{fsyl`bwn},
the analyzer may return 
\setcode{utf8}\RL{فسيـ}\setcode{standard} 
as a prefix morpheme with the POS tag {\tt fa/CONJ+sa/FUT+ya/IV3MD}
and with gloss tag {\tt and + will + they (both)},
\RL{l`b} as a stem with POS tag {\tt loEab/VERB\_IMPERFECT}
and with gloss tag {\tt ``play''},
and 
\setcode{utf8}\RL{ـون}\setcode{standard}
as a suffix with POS tag {\tt uwna/IVSUFF\_SUBJ:MP\_MOOD:I}
and with gloss tag {\tt [MASC.PL.]}.
%Additionally, affixes sometimes produce word forms that are homographic with other words. 
\novocalize

\subsection*{Knowledge-based entity extraction}

Researchers proposed and evaluated empirical and knowledge-based techniques to extract entities and relations from text. 
Knowledge-based techniques target a task based on solid linguistic or structural grounds~\cite{soudi2007arabic}.
Knowledge-based techniques such as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars 
with morphological stemming to perform NER. 
The work in~\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
events, and relations amongst them from Arabic text using a hierarchy of manually built 
finite state machines driven by morphological features and graph transformation algorithms. 
Such techniques require advanced linguistic and programming expertise. 
Our method provides a user-friendly interface that enables an average user to define 
target entities and relations.

\subsection*{Empirical entity extraction}

Supervised and unsupervised empirical techniques employ machine learning techniques to 
automatically extract entities without the need to manually encode the requisite 
knowledge~\cite{soudi2007arabic}. 
Supervised learning techniques require training text that is annotated with correct tags to learn a computational model. 
Supervised and unsupervised techniques require reference text that is annotated with correct tags to evaluate the 
accuracy of the technique in terms of metrics such as precision and recall~\cite{englishtreebank,arabictreebank,chinesetreebank}. 
The work in~\cite{ekbal2010named} presents a language independent approach for NER extraction using {\em support vector machines}. 
The work in~\cite{abdelrahman2010integrated} integrates 
a semi-supervised bootstrapping pattern recognition technique, 
and a supervised classifier based on {\em conditional random fields}
to solve NER problems. 
Our method enables the user to incrementally create complex annotations for Arabic text based on automatic 
extraction of morphological tags through a user friendly interactive interface.
%Researchers proposed and evaluated techniques to target different tasks in the fields of NLP and CL. 
%The target tasks include named entity recognition, temporal entity recognition, and cross-document analysis.
%Researchers propose local grammars with morphological stemming to perform NER\cite{zaghouani2010adapting,traboulsi2009arabic}. 
%The work in \cite{shaalan2009nera} suggests rule-based systems powered with local grammars. 
%The work in \cite{ZaMaHaCicling2012Entity} propose a method for extracting entities, events, and relations amongst them from Arabic text using a hierarchy of finite state machines driven by morphological features and graph transformation algorithms. However, building of the computational models of these tasks requires expertise and manual coding; something not everyone is capable of.

\subsection*{\framework}

In this work, we present a {\em morphology-based entity and relational information extraction framework for Arabic text} (\framework). 
\framework provides a user-friendly interface where the user defines tag types 
and associates them with \framework formulae that are regular expressions over 
\framework Boolean formulae. 
Boolean formulae are terms, negations of terms, and disjunctions of terms. 
Terms are matches to Arabic morphological features including 
prefix, stem, suffix, POS tags, gloss tags, and semantic categories. 
In addition to the morphological features, \framework 
introduces $Syn^k$ as a feature that relates two words $w_1$ and $w_2$ iff
$w_1$ is a synonym of $w_2$ or $w_1$ is a $Syn^{k-1}$ with one of the synonyms of $w_2$. 
Consider the example shown in Figure~\ref{fig:introsynEx}. 
Given the Arabic words \RL{.t`Am}, \RL{'kl}, and \RL{'t`b} and their glosses \{food\}, \{food, eat, make tired\}, and \{make tired, bother, drink\} respectively. 
We can relate \RL{.t`Am} to \RL{'kl} based on the gloss intersection \cci{food}. 
Moreover, we can relate \RL{.t`Am} to ``\RL{'t`b}'' since \RL{'kl} and \RL{'t`b} have the gloss intersection \cci{make tired}.

\begin{figure}[h]
\setcode{utf8}
\begin{center}
  \resizebox{0.6\columnwidth}{!}{ 
  	{\relsize{-3} \input{figures/syn2ex.tex}}
  }
  \setcode{standard}
  \caption{$Syn^2($\RL{.t`Am}$)$}
  \label{fig:introsynEx}
\end{center}
\end{figure}

\framework regular expressions support operators such as concatenation, zero or one, zero or more, one or more, up to $M$, and logical conjunction and disjunction operations. 
\framework editor allows the user to associate an action with each \framework sub-expression. 
The user specifies the action with C++ code and uses the \framework API to access information related to 
the matches such as text, position, length, morphological features, and numerical values.

\framework takes an Arabic text, a set of user-defined \framework Boolean formulae, and a set of user-defined \framework regular expressions. \framework computes the morphological solutions of the words in the input text then computes matches to the Boolean formulae. 
\framework then generates a {\em non-deterministic finite state automata} (NDFSA) for each expression and simulates it with the sequence of Boolean formulae matches to compute the regular expression matches. 
\framework generates, complies, links, and executes the actions of the corresponding regular expressions as shared object libraries. 
Finally, \framework constructs the semantic relations and cross-reference between entities. 
\framework also provides visualization tools to present the matches, and estimate their accuracy with respect to reference tags.

\framework has the following advantages:

\begin{itemize}
  \item \framework provides a novel and intuitive visual interface to build Boolean formulae over morphological features, 
    build regular expressions over the resulting Boolean formulae, and thereafter compute automatic tags.
  \item Up to our knowledge, \framework  is the first morphology-based framework for Arabic entity and relational entity extraction.
  \item \framework provides the user with the ability to rapidly create annotated Arabic text corpora with sophisticated morphology-based tags.
\end{itemize}

In \framework, we make the following contributions:

\begin{itemize}
  \item \framework enables the user to define semantic relations and automatically construct matches of these relations.
  \item \framework enables the user to associate code actions with sub-expressions with API access to 
    match features including text, position, length, numerical value, and morphological features.
  \item \framework enables the user to tag words based on a synonymic relation using the $Syn^k$ feature.
\end{itemize}

\subsection*{Related work}

Researchers proposed systems for automatic information extraction based on user specifications. 
CPSL is a common pattern specification language for finite-state grammar~\cite{appelt1998common}. 
It defines three sections for declaration, rule definition, and macros. 
\framework is an extension to CPSL with action execution and relation construction. 
The work in \cite{chiticariu2010systemt} presents SystemT, a system based on an algebraic Approach to Declarative information extraction (IE). 
It uses a declarative rule language and an optimizer. 
TEXTMARKER is a rule-based IE system designed to extract structured data from text \cite{atzmueller2008rule}.  
The work in~\cite{urbain2012user} presents a user-driven relational model targeting entity-relation extraction. 
In this model, the user enters a natural language query. 
QARAB is a question answering system supporting the Arabic language~\cite{hammo2002qarab}. 
It takes Arabic natural language query and attempts to provide short answers for it. 
We discuss related work and compare to it in details in Section~\ref{sec:related}.

### 2018-03-13 tex/merf/merf/merf.tex
adding merf

\include{macros}
\begin{document}
%
% paper title
% can use linebreaks \\ within to get better formatting as desired
\title{\framework: Morphology-based Entity and Relational Entity Extraction Framework for Arabic}
%~\IEEEmembership{Member,~IEEE,}
\author{Ameen~Jaber,
        Fadi~A.~Zaraket
\IEEEcompsocitemizethanks{\IEEEcompsocthanksitem A. Jaber and F. Zaraket are with the Department
of Electrical and Computer Engineering, American University of Beirut, Beirut,
Lebanon.\protect\\
E-mail: \{aaj15,fz11\}@aub.edu.lb}}

\IEEEcompsoctitleabstractindextext{%
\begin{abstract}
\input{abstract}
\end{abstract}

\begin{IEEEkeywords}
Tagging, Arabic, Natural language processing, Morphological analysis, Computational linguistics, Information extraction, Entity, Relational Entity.
\end{IEEEkeywords}}

% make the title area
\maketitle

% For peerreview papers, this IEEEtran command inserts a page break and
% creates the second title. It will be ignored for other modes.
\IEEEpeerreviewmaketitle

\section{Introduction}
\label{sec:introduction}
\input{introduction}
%\hfill mds
%\hfill August 15, 2013
\section{Motivation}
\label{sec:motivation}
\input{motivation}

\section{Overview}
\label{sec:overview}
\input{overview}

\section{Preliminaries}
\label{sec:preliminaries}
\input{preliminaries}

\section{Related Work}
\label{sec:related}
\input{relatedwork}

\section{\framework}
\label{sec:framework}
\input{tool}

\section{\framework GUI}
\label{sec:gui}
\input{gui}

\section{Implementation}
\label{sec:implementation}
\input{implementation}

\section{Results}
\label{sec:results}
\input{results}

\section{Conclusion}
\label{sec:conclusion}
\input{conclusion}

\bibliography{references}
\end{document}

### 2018-03-13 tex/merf/merf/motivation.tex
adding merf

\setcode{utf8}
\setarab
\transfalse

\begin{figure*}[tb]
  \begin{center}
\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{1.9\columnwidth}{!}{
    \begin{tabular}{R{12cm}L{12cm}}
     \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من} &
     \resizebox{\columnwidth}{!} {
     %[$\showvrb{cout<<$R.text}$]
     Expression = $(P|N)+~\mathit{O}?~R~\mathit{O^{\wedge}2}~(P|N|U)+$}\\
     $\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك} \RL{تقود}&
     \multirow{2}{*}[20pt]{
           \resizebox{1.5\columnwidth}{!}{ 
				\relsize{-4} \input{figures/motivTree.tex}}
          }\\
     $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا} $\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي}& \\
     \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}& \\
	  & \multicolumn{1}{c}{\multirow{1}{*}[-15pt]{\textbf{(b) Formula, matches, and entity-relation graph}}}\\
     \multicolumn{1}{L{12cm}}{
     \arabfalse
     \transtrue
     \RL{من المستحيل ألا تلاحظ برج خليفة بالقرب من التقاطع الأول وأنت تقود سيارتك في شارع الشيخ زايد، حتّى وإن كانت هذه المرّة الأولى التي تسلك فيها هذا الطريق؛ يقع دبيّ مول على مقربة من هذا المبنى الذي يُعد الأطول في العالم.}
     \arabtrue
     \transfalse}
     & \\
     \multicolumn{1}{L{12cm}}{It is impossible not to notice $\stackrel{n_1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.} &
     \multirow{2}{*}[20pt]{
     \resizebox{1.3\columnwidth}{!}{
                   \begin{tabular}{c|c|c|c}
                   MBF & description & formula & matches\\ \hline
                   N & name of person & $category~=~Name\_of\_Person$ & $n_1,n_2,n_3$ \\ \hline
                   P & name of place & $category~=~Name\_of\_Place$ & $p_1,p_2,...,p_7$ \\ \hline
                   R & relative position & $stem\in$ \{\RL{قرب},\RL{في},\dots\} & $r_1,r_2,r_3,r_4$ \\ \hline
                   U & numerical term & $stem\in$ \{\RL{أول},\RL{ثاني},\dots\} & $u_1,u_2$ \\
                   \end{tabular}
          }}\\
      \multicolumn{1}{c}{\multirow{1}{*}[-10pt]{\textbf{(a) Text with directions}}} & \multicolumn{1}{c}{\multirow{1}{*}[-10pt]{\textbf{(c) Tag types with Boolean formulae}}}\\
    \end{tabular}
}
  \end{center}
  \caption{Text, formula, and match \framework example}
  \label{fig:motiv}
\end{figure*}
\transtrue
\setcode{standard}

We motivate \framework with a direction extraction example. 
Consider the example in Figure~\ref{fig:motiv}\footnote{text taken from the directions page of Dubai Mall \url{http://www.thedubaimall.com/ar/}.}.
We are interested in the extraction of the entities and relations illustrated in Figure~\ref{fig:motiv}(b)
from the Arabic text in Figure~\ref{fig:motiv}(a). Figure~\ref{fig:motiv}(a) 
also presents a transliteration, and an English translation. 
The user interacts with \framework to specify her/his intent, 
and \framework automates the extraction process.

Using our tool, the user specifies the $N$, $P$, $R$, and $U$ tag types with Boolean formulae based on morphological features.
As described in Figure~\ref{fig:motiv}(c), $N$ denotes names of persons and is specified by a constraint on its category feature,
$P$ denotes places and is specified similarly, $R$ denotes relative positions and is specified by belonging to selected list of 
stems, and $U$ denotes numerical terms and is specified by a disjunction of a category constraint and an stem constraint.

\framework calls an inhouse morphiological analyzer~\cite{sarf} and computes matches to $N$, $P$, $R$, and $U$. 
Stemming is required in the case of \RL{bAlqrb} in order to detect the position key word \RL{qrb}. 
We refer to all {\em other} words in the text that did not match a user defined tag type as {\em null} words and we denote them
by $O$. 
The resulting matches are illustrated with boxes and superscripts in Figure~\ref{fig:motiv}(a) and are also listed 
in Figure~\ref{fig:motiv}(c).

The user now defines the direction entities and relations with the regular expression 
$(P|N)+~O?~R~O\wedge 2~(P|N|U)+$ shown in Figure~\ref{fig:motiv}(b) where 
$|,+,?,$ and $\wedge k$ denote disjunction, one or more, zero or one, and
up to $k$ matches, respectively.
The expression specifies a sequence of places or names of persons, 
optionally followed by a null word, 
followed by one relative position, followed by up two possible null words, 
followed by one or more match of name of place, name of person, or numerical term. 
$O$ and $\wedge 2$ are used in the expression to allow for flexible matches. 
\framework computes the matches of the expression in the text. 
The match trees in Figure~\ref{fig:motiv}(b) illutstrates two of them.
The nodes of the trees are entities and the edges and internal nodes are text, morphology-based, 
and word distance based relational entities. 
% the sequence and structure gives me the text of a parent node from the children nodes
% the same sequence can also give me the interesting (matching) morphological features
% the word distance is defined and abstracted by the matches of the re operators (internal nodes).

The user declares semantic relations that relate 
parts of the matches of the expression with each other. 
For example, in Figure~\ref{fig:motiv}(b), let $e_1,o_1,r,o_2,$ and $e_2$ be the 
$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$ parts of the expression, respectively;
the declaration \cci{Relation($r$,$e_1$,$e_2$)} creates the 
edge labeled with \cci{next to} between \cci{intersection 1} and \cci{Khalifa tower} nodes in match 1, and the
edge labeled with \cci{near} between the \cci{Dubai Mall} and \cci{the building} nodes in match 2. 
The conditional relation \cci{if($o_1$,Relation($o_1$,$r$,$e_1$))} creates 
the edge labeled with \cci{prep} between the \cci{Dubai Mall} and \cci{near} nodes in match 2.
The conditional relation \cci{if($o_2$,Relation($o_2$,$r$,$e_2$))} creates 
he edge labeled with \cci{from} between the \cci{intersection 1} and \cci{next to} nodes in match 1, and the 
edge labeled with \cci{from this} between the \cci{near} and \cci{the building} nodes in match 2.

In addition, the user specifies the \cci{isA} relation as a cross reference relation, which creates
the edge between the \cci{Khalifa Tower} and the \cci{The building} nodes.

%The user can add c++ code action to each term in the expression with access to match properties. 
%A sample action is shown in the expression where the user prints the text of the relative position match. 
%The last step for the user is to run the smilulator to extract the expression matches and simulate the actions. 
%The tree in Table~\ref{fig:motiv}(b) shows the match trees of the directions extracted from the text based on the user-defined expression.

%The existence of a morphological analyzer is crucial to the analysis. 
%In order to extract the directions, \framework requires the user to first define those tag types as stated above.

### 2018-03-13 tex/merf/merf/overview.tex
adding merf

%\begin{figure}[h!]
%  \centering
%  \includegraphics[width=0.45\textwidth]{figures/overview}
%  \caption{\framework Overview}
%  \label{f:overview}
%\end{figure}

\begin{figure}[h!]
\centering
\resizebox{\columnwidth}{!}
{\tiny \input{figures/overview.tex}}
\caption{\framework flow diagram.}
\label{f:overview}
\end{figure}

Figure~\ref{f:overview} shows the flow diagram of \framework.
\framework takes as input the Arabic text. 
The user interacts with \framework through a user-friendly interface to specify the
Morphology-based Boolean formulae, the Morphology-based regular expressions, 
the relation declarations, and the actions.
\framework analyzes the text using an inhouse Arabic morphological anaylzer~\cite{sarf} and computes morphological solutions
for each word in the Arabic text.
\framework passes the solutions to the Boolean formula simulator including the K-Synonym detector 
which computes the matches and produces a tag set for each word.
\framework then passes the sequence of tag sets to the regular expression simulator which computes 
chunks; i.e. sequences of words, whose tag sets match the expressions. 
\framework then evaluates the semantic relation declarations against the matches to compute the semantic relational entities, 
and executes the corresponding user-defined actions. 
Finally, \framework uses a cross-referrence relation such as the \cci{isA} relation to create relations across chunks and 
documents.

\framework presents the resulting tags to the user incrementally in the form of 
text annotated with style and color legends, match trees, and graphs.
\framework also provides statistical analysis tools that help compare sets of tags to reference tags,
compute standard accuracy measures, and correct generated annotations.

### 2018-03-13 tex/merf/merf/preliminaries.tex
adding merf

\setcode{utf8}
\setarab

\subsection{Morphological features}

Arabic is a morphologically rich language. 
Due to the rich Arabic morphology, a single Arabic word 
might replace a complete sentence in English. 
For example, the word \RL{يَأْكُله}
stands for the English 
statement ``he/it is eating him/it''.

Morphological analysis decomposes an Arabic word into
several morphemes,
where a morpheme is the smallest linguistic unit 
that has a meaning and fulfills a grammatical function. 
A morpheme can be a {\em stem}, or an {\em affix}. 
An affix can be a {\em prefix}, a {\em suffix}, or an {\em infix}. 
Each morpheme is associated with other morphological features 
including {\em POS}, {\em gloss}, {\em lemma}, and {\em category} tags. 
The gloss is a brief notation of the semantic meaning of a word. 
The category is a user defined tag that includes several morphemes of 
one kind. 
For example, the user can define a temporal category to include the 
prefix \RL{س} (will) and the time unit \RL{ساعة} (hour).

For example, the word \RL{يَأْكُله} is composed of three morphemes 
with associated morphological features as shown in Table~\ref{tab:samplesolution}.
The {\tt IV3MS} POS tag indicates a third person masculine singular subject pronoun 
attached to a verb, and the {\tt IVSUFF\_DO:3MS} POS tag
indicates a third person masculine
singular pronoun attached as an object to an action verb.
The {\tt VERB\_IMPERFECT} tag indicates an imperfect verb.
The notation for the gloss and POS tags is taken from the Buckwalter 
morphological analyzer~\cite{Buckwalter:02}.

%In particular, the word belongs to the category {\em Verb Imperfect}.

\begin{table}[h!]
  \centering
  \caption{Sample Solution Vector for the text \RL{يَأْكُله}.}
    \begin{tabular}{|r|c|c|c|}
          & \textbf{Prefix} & \textbf{Stem} & \textbf{Suffix} \\
    \textbf{Data} & \RL{يَ} & \RL{أْكُل} & \RL{ه} \\
    \textbf{POS} & IV3MS+ & VERB\_IMPERFECT & IVSUFF\_DO:3MS \\
    \textbf{Gloss} & he/it & eat/consume & him/it \\
    \end{tabular}%
  \label{tab:samplesolution}%
\end{table}%
\setcode{standard}

\subsection{Morphological Analyzer}
An Arabic morphological analyzer takes a word in Arabic
and returns a set of morphological solutions. 
Each solution splits the word into its morphemes and 
associates each morpheme with corresponding tags. 
Multiple solutions can be the result of multiple valid 
segmentations of the word into morphemes, or the multiple 
possible tags associated with a morpheme.

\framework is integrated with an in-house open source Arabic morphological analyzer
based on finite state transducers~\cite{ZaMaColing2012DemosSarf}.

\subsection{Finite State Transducers}
A finite state transducer (FST) is a finite state machine with an input and an output tape. 
FSTs differ from finite state automata (FSA) in that
they have an output tape while FSAs have accept states instead.
Formally, an FST is a tuple $M = (S,S_{0},\sigma,\varGamma,\delta)$ where $S$ is the set of states,
$S_{0} \subset S$ is the set of initial states, 
$\sigma$ is the input alphabet, 
$\varGamma$ is the output alphabet, 
and $\delta \subseteq S \times (\sigma \times \{\epsilon\}) \times (\varGamma \times \{\epsilon\}) \times S$ is the transition relation. 
FSTs have been used extensively in text mining applications where the input is the text and the output is the
delimiters of a chunk of text with an associated class~\cite{beesley2001finite}. 
FSTs are attractive due to their efficiency and ease of use.

\subsection{Classes and labels}
A {\em class} is a semantic decision that the NLP of CL task tries to make. 
Parts of text that belong to the desired class are all assigned the same class {\em label}. 
For example, {\em temporal unit} is a label of a class that encompasses explicit temporal 
information in the sentence related to time such as 
\RL{dqyqT} (minute) and \RL{sA`T} (hour).

The user may wish to define a simple class as an abstract category that contains a 
set of morphemes. 
For more sophisticated classes, the user can use the visual interface in \framework to 
define the class through a Boolean formulae with morphology based atomic terms. 
The annotation of text with the labels of the classes can later be used in 
CL and NLP tasks for learning, testing, and validation.

### 2018-03-13 tex/merf/merf/relatedwork.tex
adding merf

The work in \cite{dukes2011supervised} presents a collaborative effort towards morphological 
and syntactic annotation of the Quran. 
The task is held through online supervised collaboration using a multi-stage approach that 
includes automatic POS tagging, manual verification, and online supervised collaborative 
proofreading. 
Moreover, the work in \cite{dorr2010interlingual} presents a framework for interlingual 
annotation of parallel text corpora with multi-level representations. 
This corpora is important for NLP tasks including text summarization, information retrieval, 
and machine translation.
An overview of annotation tools and their Arabic-English word alignment issues 
concludes with a set of rules and guidelines needed in an Arabic annotation alignment tool
~\cite{kholidy2010towards}.
The work in \cite{kulick2010consistent} presents the integration of the Standard Arabic Morphological 
Analyzer (SAMA) into the annotation workflow of the Arabic Treebank.
Such tasks motivated us to build \framework, a morphology based open source annotation tool 
for the Arabic language.

MMAX2 is a manual multi-level linguistic annotation tool with an XML 
based data model ~\cite{mmax2}. 
It enables the user to create, browse, visualize, and query annotations
and may be able to resolve coreference tags. 
BRAT is a multi-lingual user friendly manual web-based annotator that allows the construction 
of entity and relation annotation corpora~\cite{brat}. 
BRAT provides an API for automated annotators to provide annotations.
WordFreak is similar to BRAT. It supports Arabic text and can be extended through a plug-in 
architecture to integrate with NLP and CL tasks. 
The plug-in API may enable the use of automatic annotators along with customized visualization 
and annotation specifications~\cite{wordfreak}. 
AGTK is a toolkit for the development of text and speech annotation tools~\cite{agtk}. 
It provides import APIs from other data and graphical user interface (GUI) components. 
The work in \cite{smrz2004morphotrees} presents the extension of TrEd, a customizable 
general purpose tree editor, with the Arabic MorphoTrees annotation. The MorphoTrees present 
the morphological analyses in a hierarchical organization based on common features.

\framework differs from MMAX2, BRAT, WordFreak, AGTK, and TrEd in that it allows the user to specify 
sophisticated tag types using Boolean formulae of Arabic morphological features. 
These are key in CL and NLP for Arabic text. 
Fassieh is a commercial Arabic text annotation tool that enables the production of large 
Arabic text corpora~\cite{attia2009fassieh}. 
The tool supports Arabic text factorizations including morphological analysis, POS tagging, 
full phonetic transcription, and lexical semantics analysis in an automatic mode. 
Fassieh is not directly accessible to the research community and requires commercial licensing. 
\framework is open source and differs in that it allows the user to build tag types 
using Boolean formulae of several atomic terms and define Sequential formulae based on the Boolean formulae.

Task specific annotation tools such as~\cite{alrahabi2006semantic}
uses enunciation semantic maps to automatically annotate 
directly reported Arabic and French speech. 
AraTation is another task specific tool for semantic annotation of 
Arabic news using web ontology based semantic maps~\cite{saleh2009aratation}.
We differ in that \framework is general, and not task specific, and it uses 
morphology based features as atomic terms.

### 2018-03-13 tex/merf/merf/tool.tex
adding merf

\framework takes a sequence of Arabic words 
$T=\langle t_1,t_2,\ldots,t_M\rangle$ as input text, 
a set of tag types ${\cal T}$ each with its Boolean or sequential formulae. 
\framework is integrated with {\em Sarf}, 
an open source Arabic morphological analyzer based on finite state transducers~\cite{ZaMaColing2012DemosSarf}. 
\framework uses Sarf to compute a set of morphological solutions $M(t_i)=\{m_1,m_2,\ldots,m_N\}$
for each word $t_i, 1\le i \le M$.

\def\pp{\ensuremath{{\cal P}}} % prefix
\def\ss{\ensuremath{{\cal S}}} % stem
\def\xx{\ensuremath{{\cal X}}} % suffix
\def\PP{\ensuremath{\mathit{POS}}} % pos
\def\GG{\ensuremath{\mathit{GLOSS}}} % gloss
\def\AC{\ensuremath{\mathit{CAT}}} % category

Let \pp, \ss, \xx, \PP, \GG, and \AC~be the set of 
prefixes, stems, suffixes, POS tags, gloss tags, and abstract category tags
in our system.
Each morphological solution $m$ is of the form $\langle p,s,x,P,G,C\rangle$ 
where $p\in \pp, s\in \ss, x\in \xx, P \in \PP, G \in \GG,$ and $C \in \AC$.

\subsection{$\{k\}Syn$}
Let E and A be the sets of all English and Arabic words respectively. 
Let G $\subset$ E be the set of English glosses. 
Let L $\subset$ A be the set of Arabic words in a given lexicon. 
Let S $\subset$ L be the set of stems in the given Arabic lexicon.
We define the following functions.\\
Let function $\alpha$: S $\rightarrow$ $2^G$, map input Arabic Lexicon stems to subsets of related English glosses; e.g. g$_{s} = \alpha(s)\subset$2$^{G}$.\\
Let function $\gamma$: L $\rightarrow$ $2^S$, map input Arabic Lexicon words to subsets of relevant Arabic stems; e.g. s$_{l} = \gamma(l)\subset$2$^{S}$.

Given an Arabic word w$\in$L, we define Syn(w) to be the set of Arabic words directly related to w under gloss map.\\
$Sy(w)=\{u\mid u\in S\land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$

We define Sy$^{i}$(w) to denote stems related to w using gloss map of order i recursively.\\
$Sy^{1}(w) = Sy(w)$.\\
$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.

We define $\{k\}Syn(w)$ to be the union of Sy$^{i}(w)$ for i=$1\dots k$.\\
Formally, $\{k\}Syn(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$.

\subsection{Morphology-based Atomic terms (MAT)}
Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
set of morphological features.

A MAT $a(w)$ in a Boolean tag type formula takes a word $w$ 
and a constant value of a morphological features $CF$ as input and is of the form. 
\[
  a(w) := \exists m \in M(w). m=\langle p,s,x,P,G,C\rangle. r \circ CF
\]

where $\circ \in {\cal O}$, 
$r \in \{p,s,x,P,G,C\}$, $\exists A \in{\cal F}.r\in A,CF \in A$.

Informally, a MAT indicates that a solution vector exists where 
a feature from the solution contains or exactly matches a constant value for the 
feature specified by the user.

Another form of an MAT is based on the K-Synonym feature. 
It takes a word $w$, a constant stem value $CS$ and a constant integer $k$ and checks whether a stem of w belongs to $\{k\}Syn$(CS). Formally,
\[
  a(w) := \exists s\in \gamma(w).s\in \{k\}Syn(CS)
\]

where $k\in\{1,\dots,7\}$

\subsection{Morphology-based Boolean formula (MBF)}

The \framework Boolean formula is of the following form.
\begin{itemize}
  \item $a$ is an MBF where $a$ is an MAT.
  \item $\neg f$ is an MBF where $f$ is an MAT. This is interpreted
    as the negation (complement) of words matching $f$. 
  \item $f \vee g$ where $f$ and $g$ are MBF. This is interpreted as 
    the disjunction (union) of words matching $f$ with the words matching $g$. 
\end{itemize}

\subsection{Morphology-based regular expression (MRE)}
The \framework sequential formula is of the following form.
\begin{itemize}
\item $m$ is an MRE where $m$ is an MBF.
\item $fg$ is an MRE where $f$ and $g$ is an MRE. This is interpreted as
the sequence MRE $f$ followed by MRE $g$.
\item $f*$ is an MRE where $f$ is an MRE. 
This operation refers to the Kleene star interpreted as
zero or more occurrences of MRE $f$.
\item $f+$ is an MRE where $f$ is an MRE. 
This operation is interpreted as one or more occurrences of MRE $f$.
\item $f$\textasciicircum$x$ is an MRE where $f$ is an MRE and $x \in \mathbb{N^*}$. 
This is interpreted as zero or more occurrences of MRE $f$ up to $x$.
\item $f?$ is an MRE where $f$ is an MRE. 
This is interpreted as zero or one occurrence of MRE $f$.
\item $f\& g$ is an MRE where $f$ is an MRE and $g$ is an MRE. 
This is interpreted as the occurence of MRE $f$ and MRE $g$.
\item $f|g$ is an MRE where $f$ is an MRE and $g$ is an MRE. 
This is interpreted as the occurence of MRE $f$ or MRE $g$.
\end{itemize}

\subsection{Computational Actions}
The computational actions are C++ based with access to word features referring to 
an MBF match. 
The features include text, position in text, length of word (count of characters), 
and the equivalent integer number if applicable.

Each MRE is associated with pre-match and on-match actions. 
Pre-match actions are executed before a match of the MRE and on-match actions 
are executed upon finding a match for the MRE. 
The regular expression is associated with global declarations and library includes.

\subsection{Tag type}
The set of tag types ${\cal T}$ contains tuples of the form $\langle l,f,d\rangle$ 
where $l$ is a text label with a descriptive name of the tag type, 
$f$ is a \framework MBF or MRE, and $d$ is a visualization legend. 
The visualization legend describes how the tags should be displayed to the user 
and contains information such as the foreground and background color, the
font size, weight, and style.

\begin{figure*}[t]
  \centering
  \includegraphics[width=0.9\textwidth]{figures/tagger}
  \caption{\framework main window with annotated text, tag descriptions, tag type legend properties, and manual tag edition menus. }
  \label{f:tagger}
\end{figure*}

\subsection{MBF Evaluation}
For each word $t_i \in T$, \framework computes a Boolean value 
($\{\mathit{true}, \mathit{false}\}$)
for all atomic terms. 
Then \framework computes Boolean values for all Boolean formulae. 
Then \framework computes the set of tags 
$R_i\subseteq T \times {\cal T}$  such that 
$(t_i,\mathit{tt}_j) \in R_i$ if and only if the Boolean formula $F_j$ associated with
tag type $\mathit{tt}_j$ is true for $t_i$.

\subsection{MRE and Action Simulation}
For each MRE, 
\framework generates its equivalent non-deterministic finite automaton (NFA)~\cite{sipser2006introduction}. 
Each MRE operation has its equivalent representation in an NFA. 
As for the upto operation ($f$\textasciicircum$x$), 
it is equivalent to $f?|ff|\dots|\underbrace{f}_{n \text{ times}}$ which has an NFA mapping.

Based on MBF evaluation stage, each word $t_i \in T$ is tagged by a tag set $R_i$. 
\framework simulates an NFA over the tag sets $R_i$ starting with $i=0,\dots,n$. 
A simulation match $m$ is a vector of the form $\langle r_i,r_{i+1},\dots,r_j\rangle$ where $0\le i\le j\le n~\text{and}~r_i\in R_i$. 
This tag match is equivalent to the text sequence match $\angle t_i,t_{i+1},\dots,t_j\rangle$ where $t_i\in T$.

If the simulation has a match $\langle r_m,r_{m+1},\dots,r_n\rangle$ where $0\le m\le n\le n$, 
the next simulation starts at $R_{n+1}$. 
In case the NFA simulation has no match, the next simulation starts at $R_{m+1}$. 
If we have more than one match starting at $R_k$ where $0\le k\le n$, 
\framework returns the longest one.

\framework maintains a function $\phi\subset Q\times\Phi$, 
where $\mathit{Q}$ is the set of states in the NFA and $\Phi$ is the set of MREs in a formula. 
So $(q_i,f_j)\in\phi$ if state $q_i\in Q$ is generated by \framework in equivalence to $f_j\in\Phi$. 
\framework uses $\phi$ to retrieve the match tree with respect to the MRE formula. 
It also uses the map to execute the pre-match and on-match computational actions of an MRE based on current match.

### 2018-03-14 tex/merf/OldSubmission/Apendix.tex
old submission files

--- 2018-03-13 tex/merf/merf/gui.tex
@@
-\begin{figure*}[tb]
-  \centering
-  \includegraphics[width=0.9\textwidth]{figures/mbfedit}
-  \caption{\framework tag type Boolean formula editor.}
-  \label{f:bfe}
-\end{figure*}
+\subsection{Finite state transducers}
 
-\framework provides a user friendly interface to specify the 
-atomic terms, the \framework Boolean formulae, the \framework regular expressions, 
-the tag types, and the legends. 
-The \framework GUI also allows the user to modify and correct the 
-resulting tag set $R$.
-The \framework GUI allows the user also to compute accuracy results 
-that compare different tag sets. 
-The accuracy results serve well as inter annotation agreement results
-when the tag sets come from two human annotators, 
-or as evaluation results for learning and information extraction techniques.
+A {\em finite state transducer} (FST) is a finite state machine with input and output tapes. 
+FSTs differ from {\em finite state automata} (FSA) in that
+they have an output tape while FSAs have accept states instead.
 
-The snapshot in Figure~\ref{f:tagger} shows the \framework GUI with 
-the tag type color sensitive view, the tag list view, the tag description view, 
-and the match tree. 
-In the shown example, the user specified three MBF-based tag types with labels ``HUNDRED'',
-``KEY'', and ``DIGITSTENS''. 
-The ``HUNDRED'' tag type is based on a simple MBF formula that inspects the gloss tag of 
-the stem of the word and checks whether it is {\tt hundred}. 
-The ``KEY'' tag type is built on an MBF formula that computes a disjunction 
-between the gloss tags of the stems being {\tt thousand/s}, {\tt million/s}, and {\tt billion/s}. 
-The ``DIGITSTENS'' tag type is based on a disjunction formula between the gloss tags of the stems 
-being {\tt one}, {\tt two},\dots, {\tt ten}, and {\tt twenty}, {\tt thirty},\dots, {\tt ninety}. 
-The user also defined an MRE-based tag type with the label ``number''. 
-The ``number'' tag type is based on the formula $\left(\textit{HUNDRED}|\textit{KEY}|\textit{DIGITSTENS}\right)+$. 
-In words, the formula defines the occurence of one or more consequent tags of the three tags.
+Formally, an FST is a tuple $M = (S,S_{0},\sigma,\varGamma,\delta)$ where $S$ is the set of states,
+$S_{0} \subset S$ is the set of initial states, 
+$\sigma$ is the input alphabet, 
+$\varGamma$ is the output alphabet, 
+and $\delta \subseteq S \times (\sigma \times \{\epsilon\}) \times (\varGamma \times \{\epsilon\}) \times S$ is the transition relation.
 
-The context sensitive menus in Figure~\ref{f:tagger} allow the user to tag 
-a selected word or chunk differently or to entirely remove the tag.
+FSTs have been used extensively in text mining applications where the input is the text and the output is the
+delimiters of a chunk of text with an associated class~\cite{beesley2001finite}. 
+FSTs are attractive in NLP tasks due to their efficiency and ease of use.
 
-The \framework GUI also allows manual tag types that are not based on morphological features.
-These tags enable the users to build their own reference corpora without 
-help from the morphological analyzer.
+\subsection{Classes, labels, and tag types}
 
-\begin{figure*}[tb]
-  \centering
-  \includegraphics[width=0.9\textwidth]{figures/msfedit}
-  \caption{\framework tag type regular expression editor.}
-  \label{f:sfe}
-\end{figure*}
+A {\em class} is a semantic decision that the NLP of CL task tries to make. 
+Parts of text that belong to the desired class are all assigned the same class {\em label}. 
+For example, {\em temporal unit} is a label of a class that encompasses explicit temporal 
+information in the sentence related to time such as 
+\RL{dqyqT} (minute) and \RL{sA`T} (hour).
 
-\subsection{Tag type Boolean formula editor}
-The morphology based tag type editor shown in Figure~\ref{f:bfe}
-allows the user to write a tag type Boolean formula in a user friendly manner. 
-The user first specifies atomic terms by selecting 
-a feature from ${\cal F}$. 
-The pattern is a regular expression that filters the feature values.
-This implicitly helps the user specify 
-an {\em exact} versus a {\em partial} match of the  desired 
-morphological feature value.
-
-Then the user can add the selected feature values to the atomic terms 
-under the tag type name. 
-The feature column has a context sensitive menu that allows negating 
-the term.
-The value column has a context sensitive menu that can switch the 
-operation between the values in 
-${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$.
-The right pane shows a description of the tag type and a set of legend 
-descriptors.
-
-For example, the user can select the feature {\em category} and associate it 
-with the value {\em temporal unit}. 
-Multiple feature/value pairs can be included in a single tag type 
-definition with a disjunction semantics.
-
-\subsection{Tag type regular expression editor}
-The morphology based tag type editor shown in Figure~\ref{f:sfe} allows the 
-user to define a tag type regular expression in a user friendly manner. 
-The user first adds the required MREs to the formula under the tag type name 
-by selecting a label from ${\cal T}$ under MBFs. 
-The Boolean formula of a highlighted tag type is shown in the table on the lower left pane. 
-Each selected label is associated with an automatic name. 
-The regular expression tree features the name, label, and operation for each member MRE.
-
-{\tt NONE} is a special tag type with no Boolean formula. 
-It tags all the words that are not tagged with any of the morphology based tag types. 
-{\tt NONE} is defined by default and can be used to introduce flexibility and noise tolerance into the formula. 
-Moreover, a defined MRE is directly added to the MRE list enabling recursion 
-in formula definition.
-
-Then, the user can construct the formula by applying operations to the selected MREs. 
-To do so, the user selects one or two MREs then selects an operation from the ones shown 
-in the lower left of the view. 
-Without adding operations, the formula is a sequence of the selected MREs. 
-The right pane shows a description of the tag type and a set of legend 
-descriptors.
-
-For example, the user can add the {\tt DIGITSTENS}, {\tt KEY}, and {\tt HUNDRED} MREs to the formula. 
-Then, he/she selects the {\tt DIGITSTENS} and {\tt KEY} MREs and apply the disjunction ( $|$ ) 
-operation to them. 
-Similarly, he/she can apply the disjunction operation to the disjunction MRE formed before and {\tt HUNDRED}. 
-Last, the user selects the disjunction formula and applies the Kleene star. 
-Thus, we get the regular expression shown in Figure~\ref{f:sfe}.
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.45\textwidth]{figures/actions}
-  \caption{\framework MRE action sample.}
-  \label{f:actions}
-\end{figure}
-
-In order to add computational actions to an MRE, 
-the user selects to edit the actions in the MRE tag type editor. 
-The view shown in Figure~\ref{f:actions} allows the user to specify 
-the actions in a user friendly manner. 
-\framework enables the user to access the MBF match features easily and define the pre-match actions. 
-\framework also enables the user to edit the formula declarations and library includes.
+The user may wish to define a simple class as an abstract category that contains a 
+set of morphemes. 
+For more sophisticated classes, the user can use the visual interface in \framework to 
+define the class through Boolean formulae with morphology-based atomic terms. 
+The annotation of text with the labels of the classes can later be used in 
+CL and NLP tasks for learning, testing, and validation.
 
 \subsection{Analysis}
+
 In addition to automatic and manual tagging, \framework 
@@
 and the tag types in ${\cal T}_2$ and not in ${\cal T}_1$.
-
-\begin{figure*}[t]
-  \centering
-  \includegraphics[width=0.9\textwidth]{figures/diff}
-  \caption{\framework comparison and accuracy results view.}
-  \label{f:diff}
-\end{figure*}
 
@@
 it to build a corrected corpora.
+
+\begin{figure*}[tb!]
+  \centering
+  \includegraphics[width=\textwidth]{figures/diff}
+  \caption{\framework comparison and accuracy results view.}
+  \label{f:diff}
+\end{figure*}

### 2018-03-14 tex/merf/OldSubmission/Appendix/Appendix.tex
old submission files

\include{macros}
\begin{document}

\begin{appendices}

In Appendix~\ref{a:motivation} we include a full version of the direction motivation example. 
In Appendices~\ref{a:temporal}, \ref{a:gen}, and \ref{a:numnorm} we describe the temporal entity
extraction, the genealogy tree extraction, and the number normalization tasks that we discuss 
in the paper, respectively. 
In Appendix\ref{a:cases}, we discuss other case studies we developed with \framework.

%\tableofcontents

%\newpage

\section{Motivation}
\label{a:motivation}

\setcode{utf8}
\setarab
\transfalse

\begin{figure}[tb!]
%  \begin{center}
%\renewcommand{\arraystretch}{1.5}% Wider
\resizebox{\columnwidth}{!}{
\begin{tabular}{R{10cm} C{10cm}}
\multicolumn{2}{C{20cm}}{
\resizebox{1.7\columnwidth}{0.4\columnwidth} {
\begin{tabular}{R{10cm} C{10cm}}
    \begin{tabular}{R{10cm}}
     \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من}
     \\ ~ \\
     $\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك} \RL{تقود}
     \\ ~ \\
     $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا} $\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي}\\
     \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي} 
     \end{tabular} &
     \multicolumn{1}{L{10cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.} \\ 
     \multicolumn{1}{C{10cm}}{\multirow{1}{*}{\textbf{(a) Text with directions}}} &
     \multicolumn{1}{C{10cm}}{\multirow{1}{*}{\textbf{(b) English translation}}} \\ 
     & \\ 
     \multicolumn{1}{L{10cm}}{
     \arabfalse
     \transtrue
     \RL{من المستحيل ألا تلاحظ برج خليفة بالقرب من التقاطع الأول وأنت تقود سيارتك في شارع الشيخ زايد، حتّى وإن كانت هذه المرّة الأولى التي تسلك فيها هذا الطريق؛ يقع دبيّ مول على مقربة من هذا المبنى الذي يُعد الأطول في العالم.}
     \arabtrue
     \transfalse
     } &
\resizebox{0.85\columnwidth}{!}{
	\begin{tabular}{c|c|c|c}
    	MBF & description & formula & matches\\ \hline
        N & name of person & $category~=~Name\_of\_Person$ & $n_1,n_2,n_3$ \\ \hline
        P & name of place & $category~=~Name\_of\_Place$ & $p_1,p_2,...,p_7$ \\ \hline
        R & relative position & $stem\in$ \{\RL{قرب},\RL{في},\dots\} & $r_1,r_2,r_3,r_4$ \\ \hline
        U & numerical term & $stem\in$ \{\RL{أول},\RL{ثاني},\dots\} & $u_1,u_2$ \\ \hline
    \end{tabular}
} \\ 
& \\
\multicolumn{1}{C{10cm}}{\multirow{1}{*}{\textbf{(c) Arabic text transliteration}}} &
     \multicolumn{1}{C{10cm}}{\multirow{1}{*}{\textbf{(d) Tag types with Boolean formulae}}} \\   
\end{tabular}
}
} \\ 
 & \\
\hline
\begin{tabular}{C{10cm}}
\resizebox{0.9\columnwidth}{!}{
	\relsize{-4} \input{figures/match1.tex}} \\
\resizebox{0.9\columnwidth}{!}{
	\relsize{-4} \input{figures/match2.tex}}
\end{tabular} &
\begin{tabular}{C{10cm}}
\resizebox{0.8\columnwidth}{!}{
	\relsize{-1} \input{figures/ergraphEx.tex}}
\end{tabular} \\
 & \\
\multicolumn{1}{c}{\textbf{\relsize{1} (f) MRE match trees}} &
\multicolumn{1}{c}{\textbf{\relsize{1} (g) Entity-relation graph}} \\
%\multicolumn{1}{C{10cm}}{
     %[$\showvrb{cout<<$R.text}$]
%     Expression = $(P|N)+~\mathit{O}?~R~\mathit{O^{\wedge}2}~(P|N|U)+$} \\
%\resizebox{0.8\columnwidth}{!}{
%	\relsize{-4} \input{figures/motivTree.tex}} \\
%\multicolumn{1}{c}{\textbf{(c) Formula, matches, and entity-relation graph}} \\
\end{tabular}
}
%  \end{center}
  \caption{Text, formula, and match \framework example}
  \label{fig:motiv}
\end{figure}
\transtrue
\setcode{standard}

We motivate \framework with an example that extracts directions from a sample Arabic text. 
Consider the text in Figure~\ref{fig:motiv}(a) which includes directions to a shopping mall. 
Figure~\ref{fig:motiv}(b) presents an English translation of the Arabic text and a transliteration. 
The text contains some details that are not interesting to the direction extraction task. 
Interesting entities, such as names, places, relative directions, 
and numerical terms are highlighted in the text. 
For example, row 1 in the table shown in Figure~\ref{fig:motiv}(d) 
lists the matches of names of persons as $n_1$, $n_2$, and $n_3$. 
We are also interested in presenting the directions in a relational diagram as shown in Figure ~\ref{fig:motiv}(g).

We demonstrate how to do that using \framework. 
The user interacts with the \framework interface to specify Boolean 
formulae, regular expressions, and semantic relations. 
\framework will automatically extract formulae matches 
using an in-house Arabic morphological analyzer (Sarf)
\cite{ZaMaColing2012DemosSarf},
extract regular expression matches, and construct the 
semantic relations.

\subsection*{Direction morphology-based formulae}

Using \framework, the user specifies the $N$, $P$, $R$, and $U$ tag 
types with Boolean formulae based on morphological features. 
As described in the table in Figure~\ref{fig:motiv}(d), the user denotes 
names of persons with formula $N$ and specifies it with a constraint 
requiring the category feature in the morphological solution to be {\tt name of person}. 
Similarly, the user specifies formula $P$ with a {\tt name of place} category. 
The user specifies formula 
$R$ to denote relative positions and requires the stem feature to belong to a selected list of 
stems containing \RL{fy} and \RL{qrb}. 
$U$ denotes numerical terms and is specified by a disjunction of constraints 
requiring the stem feature to belong to a set of stems such as
\RL{'wl}(first), \RL{_tAny}(second), \RL{_tAl_t}(third), 
\dots, \RL{tAs`}(ninth), and \RL{`A^sr}(tenth).

\subsection*{Direction morphology-based Boolean formulae matches}

The existence of a morphological analyzer is crucial 
to the Boolean formulae match extraction. 
Stemming is required in the case of \RL{bAlqrb} in order to detect 
the relative position tag match  based on \RL{qrb}. 
Thus, \framework calls an in-house morphological 
analyzer~\cite{ZaMaColing2012DemosSarf} and computes matches of 
the tag types $N$, $P$, $R$, and $U$. 
We refer to all {\em other} words in the text that do not match 
a user defined tag type as {\em null} words and we denote them by $O$.

The resulting matches are illustrated with boxes and superscripts 
in Figure~\ref{fig:motiv}(a) and are also listed in 
the fourth column of the table in Figure~\ref{fig:motiv}(d). 
So words with superscripts $n_1$, $n_2$, and $n_3$ are matches of tag type $N$. 
Words with superscripts $p_1$, $p_2$, \dots, $p_7$ are matches of tag type $P$. 
Words with superscripts $r_1$, $r_2$, $r_3$, and $r_4$ are matches of tag type $R$. 
Words with superscripts $u_1$, and $u_2$ are matches of tag type $U$.

\subsection*{Direction regular expressions}

The user now interacts with the regular expression editor to 
specify the direction entities and relations.
Intuitively, 
the directions are names of places ($P$) related to each other with 
positional propositions ($R$). 
A place name can be a tabulated place name, a street named after 
a person ($N$), or a numbered street ($U$). 
The text containing the directions might also include words that are not 
necessary to indicate directions ($O$)
but are necessary to complete the sentence.

The user tries several sequences of the above entities in the 
editor and checks their matches in the visualizer. 
Finally, the user is satisfied with an expression such as 
$(P|N)+~O?~R~O\wedge 2~(P|N|U)+$ where 
$|,+,?,$ and $\wedge k$ denote disjunction, one or more, zero or one, and
up to $k$ matches, respectively. 
The expression specifies a sequence of places or names of persons, 
optionally followed by a null word, 
followed by one relative position, followed by up to two possible null words, 
followed by one or more match of name of place, name of person, or numerical term. 
$O$ and $\wedge 2$ are used in the expression to allow for flexible matches.

The user writes the expression by experimenting with the visualizer and 
the expression editor which does not
require knowledge and expertise in regular expressions.

\subsection*{Direction regular expression matches}

\framework computes the matches of the expression in the text. 
The match trees in Figure~\ref{fig:motiv}(f) illustrate two of them. 
The first match tree refers to the text 
\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to 
the first intersection). 
The second match tree refers to the text 
\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).

The nodes of the trees are entities and the edges and internal nodes are text, morphology-based, 
and word distance based relational entities. 
The sequence and structure gives us the text of a parent node from the children nodes. 
The same sequence can also give us the interesting (matching) morphological features. 
The word distance is defined and abstracted by the matches of the regular expression operators (internal nodes).

\subsection*{Direction semantic relations}

The user now uses the semantic relation editor to declare semantic relations that relate 
parts of the matches of the expression with each other. 
The aim of the semantic relations to be defined is to construct the entity-relation 
graph shown in Figure~\ref{fig:motiv}(g). 
Intuitively, 
the user wants to create relations between place, name, and numerical entities. 
A relation between two entities can be a prepositional entity occurring between them. 
For example, the entities \RL{dby mwl}(Dubai Mall) and \RL{Almbn_A}(the building) are 
related by the preposition \RL{mqrbT}(near) as shown in Figure~\ref{fig:motiv}(g).

Let $e_1,o_1,r,o_2,$ and $e_2$ be the 
$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$ parts of the expression, respectively.
The user selected $(P|N)+$ to be an entity after noticing in the visualizer  that it happens to capture 
non-separated sequences of place and name entities denoting a single entity such as \cci{Khalifa tower}.

The declaration of the relation \cci{Relation($e_1$,$e_2$,$r$)} creates the 
edge labeled with \cci{next to} between \cci{intersection 1} and \cci{Khalifa tower} nodes in match 1, and the 
edge labeled with \cci{near} between the \cci{Dubai Mall} and \cci{the building} nodes in match 2. 
Those relations will be constructed if $e_1$, $e_2$, and $r$ exist as match entities.

The relation \cci{Relation($r$,$e_1$,$o_1$)} creates 
the edge labeled with \cci{prep} between the \cci{Dubai Mall} and \cci{near} nodes in match 2. 
This relation will be constructed if $r$, $e_1$, and $o_1$ exist as match entities.

The relation \cci{Relation($r$,$e_2$,$o_2$)} creates 
the edge labeled with \cci{from} between the \cci{intersection 1} and \cci{next to} nodes in match 1, and the 
edge labeled with \cci{from this} between the \cci{near} and \cci{the building} nodes in match 2. 
This relation will be constructed if $r$, $e_2$, and $o_2$ exist as match entities.

\subsection*{Cross-reference}

After constructing the semantic relations, the user is interested to relate the entities across relations to each other. 
The user aims in this step to relate entities that point to the same concept or thing. 
In the directions task, the user aims to relate the entities referring to the same place.

\framework provides the \cci{isA} relation as a default cross-reference relation. 
This relation relates two entities if they are directly synonymous or one entity is a synonym of a synonym of the other entity. 
As shown in Figure~\ref{fig:motiv}(g), the cross-reference relation creates the edge between the \cci{Khalifa Tower} and the \cci{The building} nodes.

\newpage

\section{Temporal entity extraction case}
\label{a:temporal}

Temporal entities are text chunks that express temporal information. 
Some of these entities represent absolute time and dates such as \RL{Al_hAms mn 'Ab 2010}. 
Others represent relative time such as \RL{b`d _hmsT 'ayAm} and others represent quantities 
such as \RL{14 ywmA}. 
{\em ATEEMA} presents a temporal entity detection technique for the Arabic language using 
morphological analysis and a finite state transducer~\cite{ZaMa2012IJCLATime}. 
Hence, we design our expressions and compare the tag results with ATEEMA.

The MBFs defined for this purpose are introduced in Table~\ref{tab:tbtags}. 
\cci{TIME} defines explicit temporal words. 
\cci{NUM} denotes numerals and includes digits or words referring to numbers. 
As for \cci{TIMEPREP}, it denotes temporal prepositions which precedes or follows a time expression.

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
  \caption{Temporal MBFs}
  \resizebox{!}{!}{
    \begin{tabular}{|r|r|r|}
    \toprule
    MBF &  & Examples \\
    \midrule
    \multicolumn{1}{|c|}{\multirow{5}[0]{*}{TIME}} & unit  & \RL{dqyqT} (minute) \\
    \multicolumn{1}{|c|}{} & relative & \RL{.gd} (tomorrow) \\
    \multicolumn{1}{|c|}{} & range & \RL{rby`} (spring) \\
    \multicolumn{1}{|c|}{} & nominal & \RL{'.hd} (Sunday) \\
    \multicolumn{1}{|c|}{} & events & \RL{hjry} \\
    \midrule
    \multicolumn{1}{|c|}{\multirow{3}[0]{*}{NUM}} & digit & \RL{3} (3) \\
    \multicolumn{1}{|c|}{} & word  & \RL{_tmAnyn} (eighty) \\
    \multicolumn{1}{|c|}{} & range & \RL{_tmAnynyAt} (eighties) \\
    \midrule
    \multicolumn{1}{|c|}{\multirow{4}[0]{*}{TIMEPREP}} & point & \RL{fy} (in) \\
    \multicolumn{1}{|c|}{} & relative & \RL{qbl} (before) \\
    \multicolumn{1}{|c|}{} & approximate & \RL{n.hw} (about) \\
    \multicolumn{1}{|c|}{} & range & \RL{xlAl} (during) \\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:tbtags}%
\end{table}%

The \framework MREs constructed are shown below. 
MRE {\em mts}, denotes {\tt maybe time start}, and is a disjunction of \cci{NUM} tag or \cci{TIMEPREP} tag. 
MRE \cci{dt}, denotes {\tt definitetime}, and is zero or more sequence of \cci{mts} tags and up to two 
\cci{NONE} tags, \cci{TIME} tag, then an optional sub-expression. 
The sub-expression is a sequence of zero or more sequence of up to two \cci{NONE} tags followed by \cci{mts} tag or \cci{TIME} tag followed by up to two \cci{NONE} tags and \cci{TIME} tag or \cci{NUM} tag. 
The last sub-expression \cci{TIME|NUM} is introduced to ensure that the temporal match ends with a \cci{TIME} tag or \cci{NUM} tag and not \cci{TIMEPREP} tag.

\begin{Verbatim}[fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
mts : NUM | TIMEPREP;
dt : ( $\stackrel{s_1}{mts}$ NONE^2)* $\stackrel{s_2}{TIME}$
       ( (NONE^2 $\stackrel{s_3}{( mts|TIME )}$)* 
         NONE^2 $\stackrel{s_4}{( TIME|NUM )}$ )?;
\end{Verbatim}

\framework assigns the symbols $s_1$, $s_2$, $s_3$, and $s_4$ for the sub-expressions \cci{mts}, \cci{TIME}, \cci{mts|TIME}, and \cci{TIME|NUM}, respectively. 
These symbols are also shown in the regular expression defined above. 
Using these symbols, we define the semantic relations $r_1=\langle s_1,s_2,\epsilon\rangle$, $r_2=\langle s_2,s_3,\epsilon\rangle$, and $r_3=\langle s_3,s_4,\epsilon\rangle$. 
$r_1$ relates each entity representing a match of \cci{mts} to the entity representing the match of TIME. 
$r_2$ relates each entity representing a match of \cci{mts|TIME} to the entity representing the match of TIME. 
$r_3$ relates each entity representing a match of \cci{mts|TIME} to the entity representing the match of \cci{TIME|NUM}.

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
  \caption{Temporal entity example}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{|r|r|r|r|r|r|r|r|}
    \toprule
    \RL{AlsnwAt} & \RL{'.s`b} & \RL{mn} & \RL{w_hms} & \RL{`jAf} & \RL{'^shr} & \RL{'rb`T} & \RL{b`d} \\
    \midrule
    TIME  & NONE  & NONE  & NUM   & NONE  & TIME  & NUM   & TIMEPREP \\
    \midrule
          &       &       & mts   &       &       & mts   & mts \\
    \midrule
    \multicolumn{8}{|c|}{definitetime} \\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:tempex}%
\end{table}%

Table~\ref{tab:tempex} shows a sample temporal entity match. 
The first row in Table~\ref{tab:tempex} is a temporal entity Arabic text. 
Each word is denoted with MBF tags based on its morphological features. 
The matches of the MRE {\tt mts} and {\tt dt} are shown as tags in rows 3 and 4, respectively.

\begin{table}[tb!]
  \centering
  \caption{MBF and semantic relation accuracy of temporal entity extraction in \framework}
  %\resizebox{\columnwidth}{!}{
    \begin{tabular}{c|c|c|c}
     \toprule
     \multicolumn{2}{c|}{MBF accuracy} & \multicolumn{2}{c}{Semantic relation accuracy}\\
     Recall & Precision & Recall & Precision \\
    \midrule
    1 & 0.52 & 0.98 & 0.89 \\
    \bottomrule
    \end{tabular}%
   % }
  \label{tab:tmbfer}%
\end{table}%

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
  \caption{\framework~against ATEEMA: Time, accuracy, and ease of composition}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l|l|l|ll|l}
     \toprule
     & \multirow{2}{*}{Build Time} & \multirow{2}{*}{Run Time(s)} & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
     & & & Recall & \multicolumn{1}{c|}{Precision} & \\
    \midrule
    ATEEMA & 1-2 month & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
    framework & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:tresults}%
\end{table}%

We evaluated ATEEMA and \framework against articles chosen arbitrarily from the political section of 
Al-Akhbar 17/6/2011 and 27/12/2013 issues and sports section of Al-Akhbar 24/12/2013 issue.

We present the accuracy of the MBF match extraction and semantic relation construction of \framework in Table~\ref{tab:tmbfer}. 
\framework detects all the correct MBF matches with a 100\% recall, however it shows a low precision of 52\%. 
As for the semantic relation construction, \framework presents a 98\% recall and 89\% precision. 
We also report the build time, runtime, accuracy, and ease of composition as our evaluation metrics for ATEEMA and \framework.

Table~\ref{tab:tresults} shows the metrics for both techniques. 
The time metric shows an advantage of \framework over ATEEMA with three hours compared to one to two month. 
The same advantage for \framework approach appears in the ease of composition.

ATEEMA contains more than one thousand lines of code while \framework requires the definition of three MBFs and two MREs. 
However, ATEEMA has a lower runtime with 2.53 seconds compared to 3.14 seconds for \framework. 
\framework shows higher recall (91\%)  compared to ATEEMA (88\%) which means 
that \framework detected more temporal entities than ATEEMA.

However, ATEEMA presents higher precision (89\%) than \framework (81\%). 
This indicates than the MRE in \framework is underfitting which leads to incorrect temporal matches. 
An incorrect temporal match that \framework detects is \RL{`qd} which has has the two meanings {\tt contract} and {\tt decade}.

\newpage

\section{Genealogy tree case}
\label{a:gen}

Biblical genealogical lists trace key biblical figures such as Israelite kings and
prophets with family relations. 
The family relations include wife and parenthood. 
A sample genealogical chunk of text is \RL{wwld hArAn lw.tA} 
meaning ``and Haran became the father of Lot
''. 
Our aim is to extract the names {\em Haran} and {\em Lot} and be able to relate them with parenthood relation.

GENTREE~\cite{ZaMaHaCicling2012Entity} automatically 
extracts the genealogical family tree using morphology, 
finite state machines, and graph transformations. 
We build our model in \framework by defining the required MBFs first, then construct 
the regular expression and semantic relations.

In this application, we need to extract the possible male names, female names, 
and family connectors. 
Thus, we define the MBFs \cci{MN}, \cci{FN}, and \cci{CONN}. 
\cci{MN} denotes male name and is defined by the {\tt male name} category. 
\cci{FN} denotes female name and is defined by the {\tt female name} category. 
\cci{CONN} denotes connection and is defined by the stem \RL{wld} (give birth to).

A genealogical chunk in the text has one of two forms which are 
defined in the expressions \cci{fform} and \cci{sform}. 
\cci{fform} denotes first form and it is defined by a male name (\cci{MN} tag) followed by an optional  female name (\cci{FN} tag) representing the male's wife followed by a one or more sequences of connectors and male sons (\cci{CONN} and \cci{MN} tags).

\begin{Verbatim}[fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
fform : $\stackrel{s_1}{MN}$ NONE^4 $\stackrel{s_2}{FN}$? (NONE^10 $\stackrel{s_3}{CONN}$ NONE^4 $\stackrel{s_4}{MN}$)+
\end{Verbatim}

\cci{sform} denotes second form and it is defined by a connector (\cci{CONN} tag) followed by a male (\cci{MN} tag) or female name (\cci{FN} tag) then a male name (\cci{MN} tag).

\begin{Verbatim}[fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
sform : $\stackrel{s_5}{CONN}$ NONE^4 $\stackrel{s_6}{(MN|FN)}$ NONE^4 $\stackrel{s_7}{MN}$
\end{Verbatim}

We insert \cci{NONE} in both forms to introduce flexibility to the expression. 
Last, we define the expression \cci{geneinfo} which denotes genealogical information and is defined by a disjunction of the previously defined two forms.

\begin{Verbatim}[fontsize=\relsize{-1}]
geneinfo : fform | sform
\end{Verbatim}

\framework assigns the symbols $s_1$, $s_2$, $s_3$, $s_4$, $s_5$, $s_6$, and $s_7$ for the MRE sub-expressions \cci{MN}, \cci{FN}, \cci{CONN}, \cci{MN}, \cci{CONN}, \cci{MN|FN}, and \cci{MN}, respectively. 
These symbols are shown over the sub-expressions in the defined MREs.

Based on those symbols, we define the user-defined relations $r_1=\langle s_1,s_4,s_3\rangle$, $r_2=\langle s_1,s_2,``spouse''\rangle$, $r_3=\langle s_2,s_4,s_3\rangle$, and $r_4=\langle s_6,s_7,s_5\rangle$. 
$r_1$ relates $s_1$ matches to $s_4$ matches with an edge labeled with $s_3$ matches. 
This relation indicates that entity $s_1$ is the father of $s_4$. 
$r_2$ relates $s_2$ matches to $s_1$ matches with an edge labeled by {\tt spouse}. 
This relation indicates that the female is the male's spouse. 
$r_3$ relates $s_4$ matches to $s_2$ matches using an edge labeled by $s_3$. 
This relation indicates that the male is the female's child. 
$r_4$ relates $s_6$ matches to $s_7$ matches using an edge labeled by $s_5$ matches. 
This relation indicates that $s_7$ match is the child of $s_6$.

\begin{table}[tb!]
  \centering
  \caption{MBF and semantic relation accuracy of genealogy extraction in \framework}
    \begin{tabular}{c|c|c|c}
     \toprule
     \multicolumn{2}{c|}{MBF accuracy} & \multicolumn{2}{c}{Semantic relation accuracy}\\
     Recall & Precision & Recall & Precision \\
    \midrule
    1 & 0.75 & 0.81 & 0.96 \\
    \bottomrule
    \end{tabular}%
  \label{tab:gmbfer}%
\end{table}%

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
  \caption{\framework~against manual coding of genealogy extraction: Time, accuracy, and ease of composition}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l|l|l|ll|l}
     \toprule
     & \multirow{2}{*}{Build Time} & \multirow{2}{*}{Run Time(s)} & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
     & & & Recall & \multicolumn{1}{c|}{Precision} & \\
    \midrule
    Manual technique & 1-2 month & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
    framework & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:gresults}%
\end{table}%

We evaluated the work in~\cite{ZaMaHaCicling2012Entity} and \framework against parts of the genesis chosen randomly. 
We present the accuracy of the MBF match extraction and semantic relation construction of \framework in Table~\ref{tab:gmbfer}. 
\framework detects all the correct MBF matches with a 100\% recall, but has a precision of 75\% which indicates that there are some incorrect matches. 
As for the semantic relation construction, \framework presents an 81\% recall and 96\% precision. 
We also report the build time, runtime, accuracy, and ease of composition as our evaluation metrics for both techniques. 
Table~\ref{tab:gresults} shows the metrics for both techniques. 
Similar to the previous cases, \framework has an advantage over the manual technique when it comes to the time required to build the application where the user needs four hours using \framework compared to one to two month. 
The same advantage for \framework approach appears in the ease of composition. 
The other technique contains more than three thousand lines of code while \framework requires the definition of three MBFs and two MREs. 
However, the manual technique has a lower runtime with 0.74 seconds compared to 2.28 seconds for \framework. 
\framework shows an 84\% recall and 93\% precision of MRE extraction compared to a 96\% recall and 98\% precision for the other technique.

\newpage

\section{Number normalization}
\label{a:numnorm}

In this application, we are interested in number normalization in Arabic text. 
Number normalization is the process of converting numbers in text into real numbers. 
For example the normalization of \RL{`^srwn alf} (twenty thousand) would be 20,000. 
Our target is to extract text chunks that express numbers and normalize them. 
We implemented a number normalization extractor using \framework and 
compared it with {\em NUMNORM}, a 
C++ implementation for number normalization.

NUMNORM reads an Arabic text as input, extracts the text chunks with numerical information, and normalizes them. 
NUMNORM uses the in-house morphological analyzer, {\em Sarf}, to identify the words referring to numbers. 
Sarf provides a {\tt Number} category that tags all words with numerical information. 
In the normalization algorithm, NUMNORM identifies three different categories for numbers;
each with different behavior in normalization. 
The categories are:
\begin{itemize}
\item DT, denoting digits and tens
\item H, denoting hundred
\item TMB, denoting thousand, million, and billion
\end{itemize}

NUMNORM retrieves the number referring to a word through a fixed map. 
This map takes the word gloss as input and returns the matching number if found. 
For example, the map returns 100 for the input gloss {\tt hundred}. 
As for the normalization of more than one word, such as \RL{'lfyn w-'rb`T `^sr} (2014), 
the normalization algorithm used for each category is shown in Figure~\ref{fig:numnormalgo}.  
The algorithm uses Boolean variables and three integer variables. 
The three integer variables are previous, current, and currentH for hundred. 
The main steps in the algorithm are as follows. 
Hundred is multiplied to previous digit if found, else saved. 
Digits and tens are added to previous hundred if found, else added to current. 
TMB match is multiplied by the previous hundred or DT if found, 
else saved in current.

\begin{figure}[tb!]
\centering
  \begin{tabular}{p{3.6cm}p{3.6cm}p{3.6cm}}
\begin{Verbatim}[fontsize=\relsize{-3},frame=single,label=TMB algorithm]

  if(isHundred) {
    if(current != 0) {
      previous += current;
    }
    current = currentH * matchNumber;
    currentH = 0;
    isHundred = false;
    isKey = true;
  } else if(current == 0) {
    current = matchNumber;
    isKey = true;
  } else if(!isKey) {
    isKey = true;
    current = current * matchNumber;
  } else {
    previous += current;
    current = matchNumber;
  }
\end{Verbatim}
&
\begin{Verbatim}[fontsize=\relsize{-3},frame=single,label=DT algorithm]

  if(isHundred) {
    currentH += matchNumber;
  } else if(current == 0) {
    current = matchNumber;
  } else if(isKey) {
    previous += current;
    current = matchNumber;
  } else {
    current += matchNumber;
  }
isKey = false;

\end{Verbatim}
&
\begin{Verbatim}[fontsize=\relsize{-3},frame=single,label=H algorithm]

  isHundred = true;
  if(current == 0)  {
    currentH = matchNumber;
  } else if(!isKey) {
    currentH= current * matchNumber;
    current = 0;
  } else {
    currentH = matchNumber;
  }
  isKey = false;

\end{Verbatim}
\\ 
\end{tabular}
\caption{NUMNORM algorithm for TMB, DT, and H.}
\label{fig:numnormalgo}
\end{figure}

Using \framework , we designed a morphology-based regular expression to detect the numerical chunks in the text. 
Based on the previously introduced algorithm, we added actions to this MRE to normalize each match. 
First, we defined the MBFs \cci{DT}, \cci{H}, and \cci{TMB}. 
The MBF 
\cci{DT} denotes digits and tens, and is defined by a disjunction formula of the stem glosses 
{\tt one}, {\tt two}, \dots , {\tt ten}, {\tt twenty}, \dots , and {\tt ninety}. 
The MBF \cci{H} is defined by the stem gloss {\tt hundred}. 
The MBF \cci{TMB} is defined by a disjunction formula of the stem glosses {\tt thousand}, {\tt million}, and {\tt billion}.

The \cci{number} MRE is defined as one or more matches of \cci{DT}, \cci{TMB}, or \cci{H}. 
The actions associated with each MBF are shown in Figure~\ref{fig:numnormalgo}.
We access the text and numerical values of the matches using the API provided by \framework. 
For example, we access the numerical value of the match of $s_1$ in the TMB algorithm using the API \cci{\$s1.number}. 
We also declare global variables such as \cci{current} and \cci{isHundred} to store intermediate values.

\begin{Verbatim}[numbers=left]
number : (((DT)|(TMB))|(H))+;
\end{Verbatim}

\framework also assigns the symbol $s_3$ for the sub-expression \cci{DT|TMB|H}. 
Accordingly, we define the user-defined relation $\langle s3,s3',\epsilon\rangle$. 
This relation identifies the matches of $s_3$ as entities and relates them successively with non-labelled edges.

\begin{table}[tb!]
  \centering
  \caption{MBF and semantic relation accuracy of number normalization in \framework}
  %\resizebox{\columnwidth}{!}{
    \begin{tabular}{c|c|c|c}
     \toprule
     \multicolumn{2}{c|}{MBF accuracy} & \multicolumn{2}{c}{Semantic relation accuracy}\\
     Recall & Precision & Recall & Precision \\
    \midrule
    0.99 & 1 & 0.97 & 0.95 \\
    \bottomrule
    \end{tabular}%
   % }
  \label{tab:nnmbfer}%
\end{table}%

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[tb!]
  \centering
  \caption{\framework~against NUMNORM: Time, accuracy, and ease of composition}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{l|l|l|ll|l}
     \toprule
     & \multirow{2}{*}{Build Time} & \multirow{2}{*}{Run Time(s)} & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
     & & & Recall & \multicolumn{1}{c|}{Precision} & \\
    NUMNORM & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
    framework & 1 hour & 1.53 & 0.91 & 0.90  & 3 MBFs/1 MRE/57 lines\\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:nnresults}%
\end{table}%

We evaluated NUMNORM and \framework against an article chosen arbitrarily from the economical section 
of Assafir newspaper 25/12/2013 issues. 
Table~\ref{tab:nnmbfer} shows the accuracy of the MBF match extraction and semantic relation construction in \framework. 
\framework presents high accuracy in extracting the MBF matches with a 99\% recall and 100\% precision. 
\framework also shows a high accuracy in constructing the semantic relations with a 97\% recall and 95\% precision. 
We report the build time, runtime, accuracy, and ease of composition as our evaluation metrics for NUMNORM and \framework. 
Table~\ref{tab:nnresults} shows the metrics for both techniques. 
Using \framework to perform normalization takes 1 hour while NUMNORM implementation required one to two month. 
NUMNORM takes 0.32 seconds to perform the task while the \framework model needs 1.53 seconds. 
As for ease of composition, \framework requires the construction of 3 MBFs and 1 
MRE with code actions of 57 lines while NUMNORM requires more than 500 lines of code. 
Both techniques have the same recall (91\%) which means that both detected the same correct number of matches. 
However, NUMNORM has a higher precision (93\%) than \framework (90\%) which means that NUMNORM has less false matches.

\newpage 
\section{Additional case studies}
\label{a:cases}

\end{appendices}

\bibliography{merf}

\end{document}

### 2018-03-14 tex/merf/OldSubmission/CoverLetter/coverletter.tex
old submission files

\documentclass[10pt,stdletter,dateno,sigleft]{newlfm}

\usepackage{charter}

\newsavebox{\Luiuc}\sbox{\Luiuc}{\parbox[b]{1.75in}{\vspace{0.5in}
\includegraphics[width=1.2\linewidth]{aub-logo.jpg}}} % Company/institution logo at the top left of the page
\makeletterhead{Uiuc}{\Lheader{\usebox{\Luiuc}}}

\newlfmP{sigsize=50pt} % Slightly decrease the height of the signature field
\newlfmP{addrfromphone} % Print a phone number under the sender's address
\newlfmP{addrfromemail} % Print an email address under the sender's address
\PhrPhone{Phone} % Customize the "Telephone" text
\PhrEmail{Email} % Customize the "E-mail" text

\lthUiuc % Print the company/institution logo

%----------------------------------------------------------------------------------------
%	YOUR NAME AND CONTACT INFORMATION
%----------------------------------------------------------------------------------------

\namefrom{Ameen Jaber} % Name

\addrfrom{
\today\\[12pt] % Date
American University of Beirut\\
P.O.Box 11-3875\\
Riad El-Solh / Beirut 1107 2020
}

\phonefrom{(+961)70-962096} % Phone number

\emailfrom{aaj15@aub.edu.lb} % Email address

%----------------------------------------------------------------------------------------
%	ADDRESSEE AND GREETING/CLOSING
%----------------------------------------------------------------------------------------

\greetto{Dear Jian Pei, editor in cheif, }% Greeting text
\closeline{Sincerely yours,} % Closing text

%\nameto{Mrs. Jane Smith} % Addressee of the letter above the to address

%\addrto{
%Recruitment Officer \\ % To address
%The Corporation \\
%123 Pleasant Lane \\
%City, State 12345
%}

%----------------------------------------------------------------------------------------

\begin{document}
\begin{newlfm}

%----------------------------------------------------------------------------------------
%	LETTER CONTENT
%----------------------------------------------------------------------------------------

I am Ameen Jaber, PhD candidate,  working with Fadi Zaraket, Assistant professor, 
from the Electrical and Computer Engineering department at the American University of Beirut, 
on Natural Language Processing (NLP) techniques for the Arabic language. 
We are submitting an original manuscript for consideration of publication at 
the IEEE Transactions on Knowledge and Data Engineering journal.

The title of the manuscript is ``MERF: Morphology-based Entity and Relational Entity Extraction Framework for Arabic''.
The manuscript present a set of techniques with a supporting tool to allow user to quickly develop entity and relational
entity extraction application for the Arabic language based on morphology features. 
Existing tools and techniques lack proper support for Arabic morphology analysis which is key for Arabic NLP.
The tool also presents a user friendly interface to help users build their applications.

The paper presents novel specification techniques for Arabic relational entity extraction as it allows 
user defined relations, and code actions based on Arabic morphological analysis.

We highly appreciate your consideration of our manuscript for publication.

%----------------------------------------------------------------------------------------

\end{newlfm}
\end{document}

### 2018-03-14 tex/merf/OldSubmission/Difference/Difference.tex
old submission files

\documentclass[10pt,stdletter,dateno,sigleft]{newlfm}

\usepackage{charter}

%\newsavebox{\Luiuc}\sbox{\Luiuc}{\parbox[b]{1.75in}{\vspace{0.5in}
%\includegraphics[width=1.2\linewidth]{aub-logo.jpg}}} % Company/institution logo at the top left of the page
\makeletterhead{Uiuc}{\Lheader{\usebox{\Luiuc}}}

\newlfmP{sigsize=50pt} % Slightly decrease the height of the signature field
\newlfmP{addrfromphone} % Print a phone number under the sender's address
\newlfmP{addrfromemail} % Print an email address under the sender's address
\PhrPhone{Phone} % Customize the "Telephone" text
\PhrEmail{Email} % Customize the "E-mail" text

%\lthUiuc % Print the company/institution logo

%----------------------------------------------------------------------------------------
%	YOUR NAME AND CONTACT INFORMATION
%----------------------------------------------------------------------------------------

\namefrom{Ameen Jaber} % Name

\addrfrom{
\today\\[12pt] % Date
American University of Beirut\\
P.O.Box 11-3875\\
Riad El-Solh / Beirut 1107 2020
}

\phonefrom{(+961)70-962096} % Phone number

\emailfrom{aaj15@aub.edu.lb} % Email address

%----------------------------------------------------------------------------------------
%	ADDRESSEE AND GREETING/CLOSING
%----------------------------------------------------------------------------------------

\greetto{Dear Jian Pei, editor in cheif, }% Greeting text
\closeline{Sincerely yours,} % Closing text

\begin{document}
\begin{newlfm}

The submitted manuscript 
``MERF: Morphology-based Entity and Relational Entity Extraction Framework for Arabic'', 
significantly extends the work in the attached position 
paper entitled 
``MATAr: Morphology-based Tagger for Arabic'' (AICCSA, 2013). 
MERF is an entity and relational entity extraction framework that 
uses the output of a morphological analyzer to extract entities and 
relations amongst them from text documents. 
MATAr is an Arabic morphology tagger that tags words in text documents 
with their morphological solutions.

In addition, MATAr only supports Boolean formulae defined over morphological features on the word level, 
and MERF extends the support to 
\begin{enumerate}
\item User defined regular expressions over Boolean formulae.
\item User defined relations between subexpressions of the regular expressions. 
\item User defined code actions associated with matches of the subexpressions. 
\item User friendly interfaces to edit the formulae, regular expressions, relations, and actions. 
\item MERF API that allows programmatic access to match morphological features in the code actions. 
\item Automatic cross-reference relation inference that relate entities and relational entities across documents. 
\item User friendly visualizations aids to report the results and facilitate incremental and interactive editing using:
\begin{enumerate}
\item Color and font style sensitive text tagging. 
\item Graph based visualization of morphology and regular expression match trees.
\item Graph based visualization of relational entities and cross reference relations. 
  \end{enumerate}
\end{enumerate}

%----------------------------------------------------------------------------------------

\end{newlfm}
\end{document}

### 2018-03-14 tex/merf/OldSubmission/abstract.tex
old submission files

--- 2018-03-13 tex/merf/TALMajor/abstract.tex
@@
 These techniques and tools require expertise in linguistics and programming and lack support of Arabic {\em morphological analysis} which is key to process Arabic text.
-In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.
-\framework provides a friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines {\em tag types} and interactively associates them with regular expressions defined over Boolean formulae.
+
+In this work, we present MERF; a morphology-based entity and relational entity extraction framework for Arabic text.
+MERF provides a user-friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines {\em tag types } and interactively associates them with regular expressions defined over Boolean formulae.
 Boolean formulae range over matches of Arabic morphological features, and synonymity features.
-Users define relations with tuples of subexpression matches and can associate code actions with subexpressions.
-\framework computes feature matches, regular expression matches, and constructs entities and relational entities from the user-defined relations.
+Users define user defined relations with tuples of subexpression matches and can associate code actions with subexpressions.
+MERF computes feature matches, regular expression matches, and constructs entities and relational entities from user defined relations.
 We evaluated our work with several case studies and compared with existing application-specific techniques.
-The results show that \framework requires shorter development time and effort compared to existing techniques and produces reasonably accurate results within a reasonable overhead in run time.
+The results show that MERF requires shorter development time and effort compared to existing techniques and produces reasonably accurate results within a reasonable overhead in run time.

### 2018-03-14 tex/merf/OldSubmission/conclusion.tex
old submission files

In this work, we presented a morphology-based entity and relational entity information extraction framework for Arabic text; \framework . 
\framework provides a user-friendly interface where the user defines tag types 
and associates them with regular expressions defined over Boolean formulae. 
The Boolean formulae are terms, negations of terms, and disjunctions of terms where terms are matches to Arabic morphological features. 
In addition to the morphological features, \framework introduces $Syn^k$; a semantic feature that relates words based on synonymity. 
The editor allows the user to associate code actions with each regular sub-expression and to define semantic relations between sub-expressions. 
\framework uses an in house Arabic morphological analyzer to compute morphological matches. \framework then computes regular expression matches, 
and then builds the relations. 
We evaluated our work with several case studies comparing the results with existing application-specific techniques. 
The results show good accuracy for \framework compared to manually-coded techniques.

Currently, \framework supports one built in cross-reference predicate 
based on the $Syn^2$ feature. 
In the future, \framework will support user-defined cross-reference predicates. 
Currently the user selects the morphological features to specify the MBF. 
We will explore techniques that can infer the features from example words 
that the user judges as relevant to the basic concept in question.

### 2018-03-14 tex/merf/OldSubmission/directionFigure.tex
old submission files

--- 2018-03-13 tex/merf/TALMajor/directionFigure.tex
@@
 \transfalse
-\begin{figure}[tb]
-\begin{center}
-\relsize{-0.5}
+\begin{figure*}[tb]
+%  \begin{center}
 %\renewcommand{\arraystretch}{1.5}% Wider
-\resizebox{\columnwidth}{!}{
-    \begin{tabular}{|C{9.5cm}C{8cm}|} 
+\resizebox{2\columnwidth}{!}{
+    \begin{tabular}{|C{10cm}C{10cm}|} 
       \hline
-  \begin{tabular}{R{9.5cm}}
+  \begin{tabular}{R{10cm}}
   \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من}
@@
   \\
-  \multicolumn{1}{L{9.5cm}}{
+  \multicolumn{1}{L{10cm}}{
    \arabfalse \transtrue
@@
   \\
-  \multicolumn{1}{L{9.5cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.}
+  \multicolumn{1}{L{10cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.}
   \end{tabular}
 &
-\resizebox{0.6\columnwidth}{!}{
+\resizebox{0.9\columnwidth}{!}{
   \input{figures/ergraphEx.tex}}
@@
 }
-\end{center}
+%  \end{center}
   \caption{Direction example with Arabic text, annotated with entities, transliteration, translation, and extracted relational entities in a graph} 
   \label{fig:intromotiv}
-\end{figure}
+\end{figure*}
 \transtrue

### 2018-03-14 tex/merf/OldSubmission/discussion.tex
old submission files

--- 2018-03-13 tex/merf/TALMajor/discussion.tex
@@
 %
-Then, the user interacts with the MRE editor to specify 
+Then the user interacts with the MRE editor to specify 
 sequences of the concepts and visualize the matches
-in a graph, in conjunction with the highlighted text.
+in a graph form, in conjunction with the highlighted text.
 
@@
 
-%\subsection{ Threats to validity}
-%As with any GUI tool, it takes time for the user to get acquainted with the 
-%user friendly interface. 
-%Relating the subexpressions to their suggested names by \framework might 
-%not be directly intuitive to the user. 
-%We address the above two concerns with providing the users with 
-%a short tutorial video on how to use the tool.
+\subsection{ Threats to validity}
+As with any GUI tool, it takes time for the user to get acquainted with the 
+user friendly interface. 
+Relating the subexpressions to their suggested names by \framework might 
+not be directly intuitive to the user. 
+We address the above two concerns with providing the users with 
+a short tutorial video on how to use the tool.

### 2018-03-14 tex/merf/OldSubmission/gui.tex
old submission files

--- 2018-03-13 tex/merf/TAL/gui.tex
@@
+\begin{figure*}[tb!]
+  \centering
+  {\setlength{\fboxsep}{0pt}%
+  \setlength{\fboxrule}{0.5pt}%
+  \fbox{\includegraphics[width=0.8\textwidth]{figures/mbfedit}}}
+  \caption{\framework tag type Boolean formula editor.}
+  \label{f:bfe}
+\end{figure*}
+
+\begin{figure*}[tb!]
+  \centering
+  \includegraphics[width=0.85\textwidth]{figures/mbfmatchvis}
+  \caption{\framework main window with MBF match annotated text, tag type properties, and manual tag editing menus.}
+  \label{fig:mbftagger}
+\end{figure*}
+
+%In this Section, we present the user friendly interface of \framework. 
+%We show how to build the direction extraction task shown in 
+%Section~\ref{sec:motivation}. 
+%The task requires the user to use the \framework interactively to 
+%define Boolean formulae, define regular expressions, 
+%declare the semantic relations.
+
 \framework~provides a user friendly interface to specify the 
-atomic terms, the MBFs, the MREs, 
+atomic terms, the \framework~Boolean formulae, the~\framework regular expressions, 
 the tag types, and the legends. 
-The GUI also allows the user to modify and correct the 
+The \framework~GUI also allows the user to modify and correct the 
 tag set $R$. 
-The GUI allows the user also to compute accuracy results 
-that compare different tag sets and that can serve well as 
+The \framework~GUI allows the user also to compute accuracy results 
+that compare different tag sets and that can sever well as 
 inter annotation agreement results
@@
 \framework saves the tags and the tag types in a user friendly format using
-the JavaScript Object Notation (JSON) format~\cite{nolan2014javascript}.
+the JavaScript Object Notation (JSON) data-interchange 
+format~\cite{nolan2014javascript}.
 The tag file keeps the paths of the separate text and tag type files. 
@@
 
-%Figure~\ref{fig:mbftagger} shows the \framework GUI. 
-%%\framework's GUI allows the user to create, manipulate, and analyze a project. 
-%The {\em File} menu is used to create, close, or open an existing project. 
-%The {\em Tags} menu enables the user to manually define and edit tag types. 
-%The user can define and edit the MBF and MRE based tag types and 
-%start the corresponding automatic morphology-based simulators from the {\em Tagtypes} menu. 
-%The user performs comparison and analysis of two tag sets using the {\em Analyse} menu. 
-%{\em View} menu enables the user to switch visualization modes of the tags.
+Figure~\ref{fig:mbftagger} shows the \framework GUI. 
+%\framework's GUI allows the user to create, manipulate, and analyze a project. 
+The {\em File} menu is used to create, close, or open an existing project. 
+The {\em Tags} menu enables the user to manually define and edit tag types. 
+The user can define and edit the MBF and MRE based tag types and 
+start the correspoding automatic morphology-based simulators from the {\em Tagtypes} menu. 
+The user performs comparison and analysis of two tag sets using the {\em Analyse} menu. 
+{\em View} menu enables the user to switch visualization modes of the tags.
 
 \subsection{Tag type Boolean formula editor}
-The user writes MBF tag types with the tag type editor introduced in~\cite{JaZaMatar}. %shown in Figure~\ref{f:bfe}.
+The user writes MBF tag types with the tag type editor shown in Figure~\ref{f:bfe}.
 First the user specifies atomic terms by selecting a feature from ${\cal F}$. 
-%The pattern filters the feature values.
-The user can also choose whether to require an exact match using the
-\cci{isA} predicate, or a substring match using the
+The pattern filters the feature values.
+The user can also chose whether to require an exact match using the
+\cci{isA} predicate from the combobox, or a substring match using the
 \cci{contains} predicate option.
 
-The user can add and remove feature values to the atomic terms 
-using push buttons. 
-A check box in the ``Feature'' column allows negating the term, and the
-``Relation'' column switches the predicate between 
+The user can add and remove selected feature values to the atomic terms 
+using the push buttons. 
+The check box in the ``Feature'' column allows negating the term, and the
+``Relation'' column menu switches the predicate between 
 \cci{isA} and \cci{contains}. 
 The list of feature and value pairs is interpreted as a disjunction to form
-the MBF. 
-A right pane shows a description of the tag type and a set of legend 
+the MBF of the tag type. 
+The right pane shows a description of the tag type and a set of legend 
 descriptors. 
@@
 %
-%Figure~\ref{f:bfe} shows the $U$ MBF which requires the gloss tag of 
-%the stem to be an exact match of either {\tt first}, {\tt second}, \dots, or {\tt tenth}. 
-%The legend descriptors set foreground and background colors to {\tt orangered}, 
-%and {\tt gray}, respectively, and the font size to {\tt 12}.
+Figure~\ref{f:bfe} shows the $U$ MBF which requires the gloss tag of 
+the stem to be an exact match of either {\tt first}, {\tt second}, \dots, or {\tt tenth}. 
+The legend descriptors set foreground and background colors to {\tt orangered}, 
+and {\tt gray}, respectively, and the font size to {\tt 12}.
 %, normal font weight, and unitalicized.
-
-\begin{figure}[tb!]
-  \centering
-  {\setlength{\fboxsep}{0pt}%
-  \setlength{\fboxrule}{0.5pt}%
-  \fbox{\includegraphics[width=\textwidth]{figures/mreedit}}}
-  \caption{\framework tag type regular expression editor.}
-  \label{f:sfe}
-\end{figure}
 
@@
 
-%After defining the MBFs of the direction extraction task, 
-%the user calls the MBF simulator from the {\tt Tagtypes} menu. 
-%The snapshot in Figure~\ref{fig:mbftagger} shows 
-The MBF match visualizer shows color sensitive text view, the tag list view, and the tag description view. 
+After defining the MBFs of the direction extraction task, 
+the user calls the MBF simulator from the {\tt Tagtypes} menu. 
+The snapshot in Figure~\ref{fig:mbftagger} shows 
+the color sensitive text view, the tag list view, and the tag description view. 
 %The color-sensitive text view shows the text with visualization of the MBF tag type matches. 
-%The tag list view shows all tags.
+The tag list view shows all tags.
 %that are automatically or manually applied to the text. 
@@
 %
-The user can edit the tags using a context sensitive menus. 
+The context sensitive menus allow the user to tag 
+a selected word or chunk differently or to entirely remove the tag. 
 \framework GUI also allows manual tag types and corresponding tags 
@@
 
-\vspace{-1em}
+\begin{figure*}[t]
+  \centering
+  {\setlength{\fboxsep}{0pt}%
+  \setlength{\fboxrule}{0.5pt}%
+  \fbox{\includegraphics[width=0.8\textwidth]{figures/mreedit}}}
+  \caption{\framework tag type regular expression editor.}
+  \label{f:sfe}
+\end{figure*}
+
 \subsection{Tag type regular expression editor}
@@
 
-\begin{figure}[tb!]
+\begin{figure*}[tb!]
   \centering
   \includegraphics[width=\textwidth]{figures/treegraph}
+%  \begin{tabular}{ccc}
+%  \textbf{\relsize{-2} (a) Arabic text{\color{white}{....................................}}} &
+%  \textbf{\relsize{-2} (b) Match Tree{\color{white}{....................................}}} &
+%  \textbf{\relsize{-2} (c) Entity-relation graph} \\
+%  \end{tabular}
   \caption{MRE annotated Text, MRE match tree, and entity-relation graph}
   \label{fig:treegraph}
-\end{figure}
+\end{figure*}
 
-\vspace{-1em}
 \subsection{MRE match visualization}
@@
 to make sure the MRE expresses the intent. 
+%The user calls the MRE simulator from the {\tt Tagtypes} menu to compute the matches of the MRE expressions. 
 The color-sensitive text view in Figure~\ref{fig:treegraph} shows 
@@
 
+%The tag list view shows all the tag matches that are automatically or manually applied to the text. 
 The match tree view shows the selected match in a graph view.
-Figure~\ref{fig:treegraph} shows the MRE match tree of the direction task 
+%tree view using 
+%graphviz library. 
+%This library provides a variety of software for drawing attributed graphs and computes the layout using a set of common graph layout algorithms such as dot. 
+%We use the generated coordinates of the nodes in the graphviz graph and Qt's graphics items to draw the match tree in \framework's main window.
+%
+Figure~\ref{fig:treegraph} shows the match tree of the direction task 
+regular expression match 
 \RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
@@
 
-%\begin{figure}[tb]
-%  \centering
-%  \includegraphics[width=0.5\textwidth]{figures/merfsreditor.png}
-%\caption{\framework~relation editor}
-%  \label{fig:relationeditor}
-%\end{figure}
+\begin{figure}[tb]
+  \centering
+  \includegraphics[width=0.5\textwidth]{figures/merfsreditor.png}
+\caption{\framework~relation editor}
+  \label{fig:relationeditor}
+  \vspace{-1em}
+\end{figure}
 
 \subsection{User defined relation editor}
+%\vspace{-2em}
 
@@
 of the matches of the sub expressions to define the three components of the relation. 
+%The user first adds a relation, then assigns the values for the entities and relational entity (edge). 
+%For each element in the tuple, the user specifies the sub-expression match and relevant label. 
+%The label can be a match text, position, length, or numerical value, or a user-defined label.
 
-A snapshot of the GUI in 
-Figure~\ref{fig:treegraph} shows in an interactive graph view
+%Recall the semantic relations we defined in the direction extraction task; namely
+%relations $\langle e_1,e_2,r\rangle$, $\langle r,e_1,o_1\rangle$, and $\langle r,e_2,o_2\rangle$. 
+Figure~\ref{fig:relationeditor} shows the declaration of the first relation 
+$\langle s_9, s_{14},s_{3}\rangle$ where the 
+labels of the nodes and the edges in the relational entities 
+are defined as the text of the subexpression matches;
+the editor allows the user to select the labels to be one of the match features. 
+%was selected to be the entity node and relational entity edge labels. 
+%Similarly, we define the other relations 
+%$\langle s_{3},s_9,s_{10}\rangle$, and $\langle s_{3},s_{14},s_{11}\rangle$ 
+%corresponding to $\langle r,e_1,o_1\rangle$, and $\langle r,e_2,o_2\rangle$, respectively.
+
+%After specifying the semantic relations, we call the semantic relation constructor. 
+A snapshot of the \framework GUI in 
+Figure~\ref{fig:treegraph} shows shows in an interactive graph view
 the entity-relation graph of the match of the user defined relation 
 extracted from the match tree of the MRE. 
+%The entity-relation graph is visualized when the user selects an MRE match tag from the tag list.
+%The graph in Figure~\ref{fig:treegraph} shows the entity-relation graph of the match tree shown 
+%in Figure~\ref{fig:treegraph}(b). 
+%The interactive graph view enables the user to move the nodes.
+
+%\subsection{Computational action editor}
 

### 2018-03-14 tex/merf/OldSubmission/introduction.tex
old submission files

--- 2018-03-13 tex/merf/TALMajor/introduction.tex
@@
-{\em Computational Linguistics} (CL) is concerned with building 
+\IEEEPARstart{C}{\em omputational Linguistics} (CL) is concerned with building 
 accurate linguistic computational models. 
@@
 {\em named entity recognition} (NER), {\em cross-document analysis}, 
-{\em machine translation}, 
+machine translation, 
 and {\em relational entity extraction}~\cite{linckels2011natural,ferilli2011natural}.
@@
 knowledge~\cite{soudi2007arabic}. 
-Supervised learning techniques require training corpora annotated 
+Supervised learning techniques require training corpora that are annotated 
 with {\em correct} tags to learn a computational model. 
-Supervised and unsupervised techniques require annotated reference corpora to 
-evaluate the accuracy of the technique using metrics such as precision and 
+Supervised and unsupervised techniques require reference corpora that are
+annotated with correct tags to evaluate the 
+accuracy of the technique in terms of metrics such as precision and 
 recall~\cite{englishtreebank,arabictreebank,chinesetreebank}.
 
-Researchers build %the required %seed 
+Researchers build the required %seed 
 training and reference corpora 
@@
 most of them lack Arabic language support, and almost all of them
-lack Arabic morphological analysis support
-% that is key to Arabic NLP due to the rich morphological nature of Arabic
-~\cite{habash2006arabic}. 
+lack Arabic morphological analysis support that is key to Arabic NLP
+due to the rich morphological nature of Arabic~\cite{habash2006arabic}. 
 Fassieh~\cite{attia2009fassieh} is a {\em commercial} Arabic annotation tool
 with morphological analysis support and text factorization. 
-However, this tool lacks support for entity and relational entity extraction.
+Fassieh lacks support for entity and relational entity extraction.
 
@@
 In this paper, 
-we present \framework, a {\em morphology-based entity and relational entity
-extraction framework for Arabic text }. 
+we present a {\em morphology-based entity and relational entity
+extraction framework for Arabic text } (\framework). 
 \framework provides a user-friendly interface where the user defines tag types 
@@
 We discuss the importance of the morphological features supported in 
-\framework terms in Section~\ref{sec:morph}; in brief, morphological preprocessing
-is key to Arabic NLP.% due to the richness of Arabic morphology.
+\framework terms in Section~\ref{s:s:morph}; in breif, morphological preprocessing
+is key to Arabic NLP due to the richness of Arabic morphology.
 
-We illustrate the target of \framework using a simple example. 
-Consider the text in Figure~\ref{fig:intromotiv} which contains
+We illusrate the target of \framework using a simple example. 
+Given the text in Figure~\ref{fig:intromotiv} that contains
 directions to Dubai Mall~\footnote{text taken from the Dubai Mall website
@@
 \framework regular expressions support operators such as concatenation, 
-zero or one, zero or more, one or more, up to $M\in\mathbb{N}^+$, and logical 
-conjunction and disjunction. 
-An editor allows the user to associate an action with each sub-expression.
-The user specifies the action with C++ code and uses an API to access information related to the matches such as text, position, length, morphological features, and numerical value.
+zero or one, zero or more, one or more, up to $M$, and logical 
+conjunction and disjunction operations. 
+\framework editor allows the user to associate an action with each 
+\framework sub-expression. 
+The user specifies the action with C++ code and uses the 
+\framework API to access information related to 
+the matches such as text, position, length, morphological features, 
+and numerical values.
 
@@
 as shared object libraries.
+
 Finally, \framework constructs the semantic relations and 
@@
 
-\subsection{Existing annotation and entity extraction tools}
+\subsection*{Existing annotation and entity extraction tools}
 
@@
 with morphological stemming to perform NER. 
-%The work in~
-\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
+The work in~\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
 events, and relations amongst them from Arabic text using a hierarchy of manually built 
@@
 QARAB is a question answering system that takes 
-Arabic natural language queries and provides short answers~\cite{hammo2002qarab}.
+an Arabic natural language query and provides short answers for it~\cite{hammo2002qarab}.
 
@@
 grammar~\cite{appelt1998common}. 
-%\framework is an extension to CPSL with action execution and relation construction. 
 %It defines three sections for declaration, rule definition, and macros. 
@@
 text \cite{atzmueller2008rule}. 
-\cite{urbain2012user} presents a user-driven relational model 
+The work in~\cite{urbain2012user} presents a user-driven relational model 
 requiring a user natural language query to extract entities and relational entities.
 
-This work significantly extends the position 
-paper~\cite{JaZaMatar} that allows for manual, and 
-basic morphology annotation.
+\framework is an extension to CPSL with action execution and relation construction. 
 \framework enables a regular user to incrementally create 
@@
     build regular expressions over the resulting Boolean formulae, and thereafter compute automatic tags.
-  \item To our knowledge, this morphology-based framework  is the first for Arabic entity and relational entity extraction.
+  \item Up to our knowledge, \framework  is the first morphology-based framework for Arabic entity and relational entity extraction.
   \item \framework provides the user with the ability to rapidly create annotated Arabic text corpora with sophisticated morphology-based tags.
@@
     when a corresponding match is found. 
-    It also provides an API to enable access to match 
+    \framework provides an API to enable access to match 
     features such as text, position, length, numerical value, and morphological features.
@@
 
-\input{directionMBFMRE}
-
-In the rest of this paper,
+The work in this paper significantly extends the position 
+paper~\cite{JaZaMatar} that allows for manual, and 
+basic morphology annotation. 
+The rest of this paper is organized as follows. 
 Section~\ref{sec:motivation} presents a motivational example. 
-Section~\ref{sec:morph} introduces Arabic morphological analysis 
-and its important role in Arabic NLP. 
+Then, we present an overview of \framework in Section~\ref{sec:overview}. 
 %We present preliminaries necessary to understand the work 
 %in Section~\ref{sec:preliminaries}. 
-Section~\ref{sec:framework} explains the \framework methodology. 
-Section~\ref{sec:gui} presents the visual and user friendly interface of \framework. 
-Section~\ref{sec:related} presents and discusses related work.
-Section~\ref{sec:results} presents the evaluation results. 
-Finally, we conclude and discuss future work in Section~\ref{sec:conclusion}.
+Section~\ref{sec:framework} explains the methodology behind \framework. 
+Section~\ref{sec:gui} presents the visual and user friendly aspects of \framework. 
+In Section~\ref{sec:results}, we present our result in comparison to other tools. 
+In Section~\ref{sec:related}, we present and discuss related work. 
+Finally, we present our conclusion and future work in Section~\ref{sec:conclusion}.

### 2018-03-14 tex/merf/OldSubmission/motivation.tex
old submission files

--- 2018-03-13 tex/merf/TAL/motivation.tex
@@
+\setcode{utf8}
+\setarab
+\transfalse
+\begin{figure*}[tb!]
+\renewcommand{\arraystretch}{1.5}% Wider
+\resizebox{2\columnwidth}{!}{
+\begin{tabular}{cc}
+\resizebox{\columnwidth}{!}{
+	\begin{tabular}{c|c|c|c}
+		MBF & description & formula & matches\\ \hline
+        N & name of person & $category~=~Name\_of\_Person$ & $n_1,n_2,n_3$ \\ \hline
+        P & name of place & $category~=~Name\_of\_Place$ & $p_1,p_2,...,p_7$ \\ \hline
+        R & relative position & $stem\in$ \{\RL{قرب},\RL{في},\dots\} & $r_1,r_2,r_3,r_4$ \\ \hline
+        U & numerical term & $stem\in$ \{\RL{أول},\RL{ثاني},\dots\} & $u_1,u_2$ \\
+    \end{tabular}
+} 
+&
+\resizebox{\columnwidth}{!}{
+\begin{tabular}{cc}
+\relsize{-2} \input{figures/match1.tex} &
+\relsize{-2} \input{figures/match2.tex} \\
+    \end{tabular}
+} \\
+\end{tabular}
+}
+  \caption{\framework Boolean formulae, and matches of regular expression $(P|N)+~O?~R~O\wedge 2~(P|N|U)+$}
+  \label{fig:motiv}
+\end{figure*}
+\transtrue
+\setcode{standard}
+
 We motivate \framework with the example of Figure~\ref{fig:intromotiv} 
@@
 the direction as relational entities as 
-shown in Figure~\ref{fig:intromotiv}.
+shown in Figure~\ref{fig:intromotiv} using \framework.
 
@@
 As described in the table of Figure~\ref{fig:motiv}, the user denotes 
-the ``name of person'' entities with formula $N$ 
-which requires the {\em category} feature in the morphological solution of a word
+Using \framework, the user denotes the ``name of person'' entities with formula $N$ 
+which requires the category feature in the morphological solution of a word
 to be {\tt name of person}. 
@@
 requiring the stem feature to belong to a set of stems such as
-\RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).
+\RL{'wl}(first), \RL{_tAny}(second), \dots, \RL{tAs`}(ninth), and \RL{`A^sr}(tenth).
 
 \framework calls an inhouse open source morphological 
-analyzer~\cite{ZaMaColing2012DemosSarf} and computes matches of 
+analyzer based on finite state transducers~\cite{ZaMaColing2012DemosSarf} and computes matches of 
 the formulae $N$, $P$, $R$, and $U$. 
@@
 The resulting matches are illustrated with boxes and superscripts 
-in Figure~\ref{fig:intromotiv} and are listed in the table of 
+in Figure~\ref{fig:intromotiv} and are also listed in the table of 
 Figure~\ref{fig:motiv}.
@@
 Finally, the user is satisfied with the matches of an expression such as 
-$(P|N)\!+~O?~R~O^\wedge\!2~(P|N|U)+$ where 
-$|,+,?,$ and $^\wedge k$ denote disjunction, one or more, zero or one, and
+$(P|N)+~O?~R~O\wedge 2~(P|N|U)+$ where 
+$|,+,?,$ and $\wedge k$ denote disjunction, one or more, zero or one, and
 up to $k$ matches, respectively.
@@
 followed by one or more match of name of place, name of person, or numerical term. 
-$O?$ and $O^\wedge 2$ are used in the expression to allow for flexible matches.
-The user reaches the satisfying matches by experimenting with the visualizer and the expression editor which do not require knowledge and expertise in regular expressions.
+$O$ and $\wedge 2$ are used in the expression to allow for flexible matches. 
+The user reaches the satisfying matches by experimenting with the visualizer and 
+the expression editor which do not
+require knowledge and expertise in regular expressions.
 
-The match trees in Figure~\ref{fig:motiv} illustrate two matches of the expression 
-computed by \framework.
+\framework computes the matches of the expression in the text. 
+The match trees in Figure~\ref{fig:motiv} illustrate two of them. 
 The first match tree refers to the text 
-\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection). 
+\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to 
+the first intersection). 
 The second match tree refers to the text 
 \RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
-The leaf nodes of the trees are entities and the edges and internal nodes are text, 
+The nodes of the trees are entities and the edges and internal nodes are text, 
 morphology-based, 
@@
 A relation between two entities can be a prepositional entity.
-For example, Figure~\ref{fig:intromotiv} shows the entities \RL{dby mwl}(Dubai Mall) and 
-\RL{Almbn_A}(the building) related by the preposition \RL{mqrbT}(near).
+For example, the entities \RL{dby mwl}(Dubai Mall) and \RL{Almbn_A}(the building) are 
+related by the preposition \RL{mqrbT}(near) as shown in Figure~\ref{fig:intromotiv}.
 
@@
 entity such as \cci{Khalifa tower}.
-\cci{Relation($e_1$,$e_2$,$r$)} creates the 
+
+The declaration of \cci{Relation($e_1$,$e_2$,$r$)} creates the 
 edge labeled \cci{next to} between \cci{intersection 1} and 
@@
 match 2 of Figure~\ref{fig:motiv}(b). 
-
+%Those relations will be constructed if $e_1$, $e_2$, and $r$ exist as match entities.
+%
 \cci{Relation($r$,$e_1$,$o_1$)} creates the edge labeled \cci{prep} 
 between \cci{Dubai Mall} and \cci{near} nodes in match 2 of Figure~\ref{fig:motiv}(b). 
+%This relation will be constructed if $r$, $e_1$, and $o_1$ exist as match entities.
+%
 \cci{Relation($r$,$e_2$,$o_2$)} creates the edge labeled \cci{from} 
@@
 nodes in match 2 of Figure~\ref{fig:motiv}(b). 
+%This relation will be constructed if $r$, $e_2$, and $o_2$ exist as match entities.
 
-After constructing the relations, 
+%\subsection*{Cross-reference}
+
+After constructing the user defined relations, 
 the user is interested to relate the discovered entities and relational entities
@@
 \cci{The building} in  Figure~\ref{fig:intromotiv}.
+%The \cci{isA} predicate relates an entity $w_21$ to an entity $w_2$ two entities if they are direct synonyms 
+%synonymous or one entity is a synonym of a synonym of the other entity.

### 2018-03-14 tex/merf/OldSubmission/oldResults.tex
old submission files

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}[htbp]
  \centering
  \caption{\framework~against manually-coded applications: Time, accuracy, and ease of composition}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{ll|l|l|ll|l}
     \toprule
    \multicolumn{2}{c|}{\multirow{2}{*}{Task}} & \multirow{2}{*}{Build Time} & \multirow{2}{*}{Run Time(s)} & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
    \multicolumn{2}{c|}{} & & & Recall & \multicolumn{1}{c|}{Precision} & \\
    \midrule
    \multirow{2}[0]{*}{Narrator Chain} & ANGE & 1-2 month & 1.79 & 1 & 1 & 3000+ lines of code\\
          & framework & 3 hours & 7.24 & 1 & 0.93  & 8 MBFs and 4 MREs\\
    \midrule
    \multirow{2}[0]{*}{Number Normalization} & NUMNORM & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
          & framework & 1 hour & 1.53 & 0.91 & 0.90  & 3 MBFs/1 MRE/57 lines\\
    \midrule    
    \multirow{2}[0]{*}{Temporal Entity} & ATEEMA & 1-2 month & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
          & framework & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
    \bottomrule
    \end{tabular}%
    }
  \label{tab:results}%
\end{table}%

### 2018-03-14 tex/merf/OldSubmission/overview.tex
old submission files

\begin{figure}[tb!]
\centering
\resizebox{\columnwidth}{!}{
	\relsize{+1} \input{figures/overview.tex}
}
\caption{\framework flow diagram.}
\label{f:overview}
\end{figure}

Figure~\ref{f:overview} shows the flow diagram of \framework. 
%A box node represents a \framework process and 
%an ellipse node represents input and output data objects.
The Arabic text and the reference tag chunks are the primary inputs to \framework.
Solutions, morphology-based Boolean formulae, tags, 
morphology-based regular expressions, 
tag chunks, relation and action definitions, and data structures expressing entities 
and relations are input and output data to \framework processes. 
The morphological analyzer (Sarf), $Syn^k$ detector, 
GUI for Boolean formulae definition, 
visualization annotator, GUI for regular expression and action definition, 
Boolean formula simulator, regular expression simulator, 
relation extraction and action execution, and difference and statistical analyzer 
are \framework processes.

\def\pp{\ensuremath{{\cal P}}} % prefix
\def\ss{\ensuremath{{\cal S}}} % stem
\def\xx{\ensuremath{{\cal X}}} % suffix
\def\PP{\ensuremath{\mathit{POS}}} % pos
\def\GG{\ensuremath{\mathit{GLOSS}}} % gloss
\def\AC{\ensuremath{\mathit{CAT}}} % category

%~\footnote{In this document, we use the default ArabTeX transliteration style ZDMG.}
{\bf Morphological analyzer (Sarf).~}
\label{s:s:morph}
Morphological analysis is key to Arabic NLP~\cite{arabicmorph}.
Short vowels, also known as diacritics, are typically omitted in Arabic text
and inferred by readers~\cite{habash2006arabic}. 
For example, the word \RL{'sd}%
can be interpreted as \RL{'sad} {\tt ``lion''} with a {\em fatha} diacritic on the 
letter \utfrl{سـ} or 
\vocalize \RL{'sodd} (I block) with a 
{\em damma} diacritic on the letter \utfrl{سـ} and {\em shadda} on the letter \RL{d}.

In addition, the position of an Arabic letter in a word 
(beginning, middle, end, and standalone) changes
its visual form.
Some letters have non-connecting end forms which allows visual
word separation without the need of a white space separator. 
This requires morphological analysis for even the simple tokenization 
task of Arabic text.
%For example, the word \utfrl{ياسمين} can be interpreted as
%the ``Jasmine'' flower, 
%as well as \notrutfrl{يا} (the calling word) followed by
%the word \notrutfrl{سمين} (obese). 
For example, consider the sentence 
%\utfrl{ذهب الولدإلى المدرسة} 
\notrrl{AlmdrsT}\notrrl{_dhb alwald-il_A}
\arabfalse \RL{_dhb alwald-il_A almdrsT} \arabtrue
(the kid went to school). 
The letters \notrutfrl{د} and \notrutfrl{ى} have 
non-connecting end of word forms and the words 
\notrutfrl{الولد}, \notrutfrl{الى}, and  \notrutfrl{المدرسة} 
are visually separable, 
yet there is no space character in between.
Newspaper articles with text justification requirements, 
SMS messages, and automatically digitized documents
are examples where such problems often occur.

\framework is integrated with {\em Sarf}, 
an in-house open source Arabic morphological analyzer based on 
finite state transducers~\cite{ZaMaColing2012DemosSarf}. 
Given an Arabic word, Sarf returns 
a set of morphological solutions. 
A word might have more than one solution 
due to multiple possible segmentations and multiple tags associated 
with each word. 
A morphological solution is the internal structure of the word 
composed of several morphemes including 
{\em affixes} ({\em prefixes} and {\em suffixes}), and a
{\em stem}, where each morpheme is associated with tags such as 
POS, gloss, and category tags~\cite{arabicmorph,habash2010introduction}. 
%POS, gloss,  {\em lemma}, and category tags~\cite{arabicmorph,habash2010introduction}. 
%A morpheme is the smallest linguistic unit 
%that has a meaning and fulfills a grammatical function.

Prefixes attach before the stem and a word can have multiple prefixes. 
Suffixes attach after the stem and a word can have multiple suffixes. 
Infixes are inserted inside the stem to form a new stem. 
In this work we consider a set of stems that includes infix morphological changes. 
The part-of-speech tag, referred to as POS, 
assigns a morpho-syntactic tag for a morpheme. 
The gloss is a brief semantic notation of morpheme in English. 
A morpheme might have multiple glosses as it could stand for multiple meanings. 
The category is a user defined tag that the user can assign to several 
morphemes.
For example, the user can define a temporal category to include the 
prefix \utfrl{س} (will) and the time unit \utfrl{ساعة} (hour). 
We denote by 
\ss,
\pp,
\xx,
\PP,
\GG, and 
\AC, the set of 
all stems,
prefixes,
suffixes,
POS,
gloss, 
and user defined category tags, respectively.

%The lemma is a convention choice that assigns a core meaning of word forms that share similar stems. 
%For example, the words \RL{byt}(house), 
%\RL{llbyt}(for the house), 
%and \RL{bywt}(houses) are mapped to the singular noun form \RL{byt} as a lemma.

\vocalize

%\setarab
\begin{table}[tb!]
  \centering
  \caption{Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
  \resizebox{\columnwidth}{!}{
    \begin{tabular}{|r|c|c|c|c|c|}
          & \multicolumn{3}{c|}{\textbf{Prefixes}} & \textbf{Stem} & \textbf{Suffix} \\
    \textbf{Data} & \utfrl{فَ} & \utfrl{سَ} & \utfrl{يَ} & \utfrl{أْكُل} & \utfrl{ها} \\
    \textbf{POS} & CONJ+ & FUT+ & IV3MS+ & VERB\_IMPERFECT & IVSUFF\_DO:3FS \\
    \textbf{Gloss} & and/so & will & he/it & eat/consume & it/them/her \\
    \textbf{index} & \multicolumn{3}{c|}{10} & 13 & 16 \\
    \textbf{length} & \multicolumn{3}{c|}{3} & 3 & 2 \\
    \end{tabular}%
    }
    \vspace{-3em}
  \label{tab:samplesolution}%
\end{table}%

Table~\ref{tab:samplesolution} shows the morphological analysis
of the \utfrl{فَسَيَأْكُلها}. 
The word is composed of the prefix morphemes 
\utfrl{فَ}, \utfrl{سَ}, and \utfrl{يَ}, followed by the 
stem \utfrl{أْكُل}, and then followed by the suffix morpheme
\utfrl{ها}. 
Each morpheme is associated with a number of morphological features.
The {\tt CONJ},
{\tt FUT}, 
{\tt IV3MS} 
{\tt VERB\_IMPERFECT}, and 
{\tt IVSUFF\_DO:3FS} POS tags indicate
conjunction, 
future, 
third person masculine singular subject pronoun,
an imperfect verb, and 
and a third person feminine singular object pronoun, respectively.
The POS and gloss notations follow the Buckwalter notation~\cite{Buckwalter:02}.

%A morpheme can be a stem or an affix. 
%Each morpheme is associated with other morphological features 
%including {\em POS}, {\em gloss}, {\em lemma}, and {\em category} tags.

%A stem can be a root word, a {\em templatic}, or {\em non-templatic} stem. 
%Templatic stems are formed from roots using template morphological rules. 
%For example, the stem
%\RL{kAtb} (writer) is a subject noun formed from the root verb \RL{ktb} (write). 
%Non-templatic stems tend to be foreign names such as \RL{واشنطن} (Washington). 
%A root is an atomic word with three, four, and rarely five letters.

%An affix can be a {\em prefix}, a {\em suffix}, or an {\em infix}. 
%The lemma is a convention choice that assigns a core meaning of word forms that share similar stems. 
%For example, the words \RL{byt}(house), 
%\RL{llbyt}(for the house), 
%and \RL{bywt}(houses) are mapped to the singular noun form \RL{byt} as a lemma.

%We introduce this feature in detail in Section~\ref{sec:framework}.

The user interacts with \framework~through a user-friendly interfaces to 
specify morphology-based Boolean formulae and regular expressions 
and associate them with tag types that are in turn associated with
visualization legends such as 
foreground and background colors and fonts.

{\bf Boolean formula simulator.~}
\framework~uses Sarf to compute morphological solutions for the Arabic text, 
and passes the solutions along with the user-defined tag types to the 
Boolean formula simulator.
The simulator interacts with the $Syn^k$ detector, 
computes tag type matches, and produces a tag set for each word. 
%A word might have multiple tags as its morphological solutions could 
%match multiple Boolean formulae. 
%Each tag contains information about the matching word and the relevant tag type name. 
%Word information include word index relative to the text, character position, and text.
%We formally define the MBF and explain its simulation in Section~\ref{sec:framework}.

{\bf Regular expression simulator.~}
\framework~passes the sequence of tag sets and the user-defined 
tag types with morphology based regular expressions to the regular expression 
simulator. 
The simulator computes tag chunks; i.e. sequences of tags sets, that match the 
regular expressions. 
%Each tag chunk contains information about the matching sequence of words and the tag type with the matching regular expression (MRE). 
%We formally define the MRE and explain its simulation in Section~\ref{sec:framework}.

{\bf Relation extraction and action execution.~}
\framework~enables the user to associate code actions to parts of the regular
expressions. 
The user can use an API to access information such as 
text, position, length, and morphological features of the tag chunks that 
match the sub-expression. 
\framework~also allows the user to declare relations between
matches using the relation editor. 
Each relation is specified by a source, destination, and a label tuple. 
The user selects a feature of a match of sub-expressions to specify the 
tuple entities.

\framework~executes the user-defined actions corresponding to each 
sub-expression match in a tag chunk. 
It also evaluates the semantic relation declarations against the 
tag chunks to compute the relational matches. 
Finally, \framework~uses the \cci{isA} cross-reference relation 
to create relations across tag type matches. 
%The output of this process expresses entities and relations among them. 
%We formally define the semantic relation and explain its construction in Section~\ref{sec:framework}.

{\bf Visualization annotator.~}
\framework presents the resulting tags to the user incrementally in the form of 
text annotated with style and color legends, match trees, and graphs expressing
the relational entities. 
The annotation can be edited by the user in a user-friendly interface. 
Match trees present the match text associated with the relevant tags 
and regular expression structure. 
The graphs are the result of the user defined relations.
%We present \framework's interface in Section~\ref{sec:gui}.

{\bf Agreement and statistical analysis.~}
\framework provides statistical analysis tools that help compare sets 
of tags to reference tags and compute standard agreement and accuracy measures. 
\framework~ provides criteria for comparison including exact match and intersection. 
This comparison tool can be used to edit the automatically generated annotations. 
%\framework~provides the user with an interactive interface to 
%build the reference tags. 
%We explain the analysis process and its interface in Section~\ref{sec:gui}.

### 2018-03-14 tex/merf/OldSubmission/preliminaries.tex
old submission files

\setcode{utf8}
\setarab

\def\pp{\ensuremath{{\cal P}}} % prefix
\def\ss{\ensuremath{{\cal S}}} % stem
\def\xx{\ensuremath{{\cal X}}} % suffix
\def\PP{\ensuremath{\mathit{POS}}} % pos
\def\GG{\ensuremath{\mathit{GLOSS}}} % gloss
\def\AC{\ensuremath{\mathit{CAT}}} % category

In this Section we define and discuss terms important to our method such as 
morphological features, finite state transducers, morphological analyzer, 
classes, labels, and tag types.

\subsection{Morphological features}

Arabic is a morphologically rich language. 
Due to the rich Arabic morphology, a single Arabic word 
might replace a complete sentence in English. 
For example, the word \RL{يَأْكُله}
stands for the English 
statement ``he/it is eating him/it''.

Form-based morphological analysis decomposes an Arabic word into
several morphemes~\cite{habash2010introduction}.
A morpheme is the smallest linguistic unit 
that has a meaning and fulfills a grammatical function. 
A morpheme can be a stem or an affix. 
Each morpheme is associated with other morphological features 
including {\em POS}, {\em gloss}, {\em lemma}, and {\em category} tags.

A stem can be a root word, a {\em templatic}, or {\em non-templatic} stem. 
Templatic stems are formed from roots using template morphological rules. 
For example, the stem
\RL{kAtb} (writer) is a subject noun formed from the root verb \RL{ktb} (write). 
Non-templatic stems tend to be foreign names such as \RL{واشنطن} (Washington). 
A root is an atomic word with three, four, and rarely five letters.
We denote the set of all stems by the symbol \ss.

An affix can be a {\em prefix}, a {\em suffix}, or an {\em infix}. 
Prefixes attach before the stem and a word can have multiple prefixes. 
We denote the set of all prefixes by the symbol \pp. 
Suffixes attach after the stem and a word can have multiple suffixes. 
We denote the set of all suffixes by the symbol \xx.
Infixes are inserted inside the stem to form a new stem. 
In this work we consider a set of stems that includes infix morphological changes.

The part-of-speech tag, referred to as POS, assigns a morpho-syntactic tag for a morpheme. 
Sample POS tags are {\tt NOUN} and {\tt VERB\_PERFECT}. 
We denote the set of all POS tags by the symbol \PP.

The gloss is a brief notation of the semantic meaning of a morpheme in English. 
A word might have multiple glosses attached to it as it could stand for multiple meanings. 
We denote the set of all glosses by the symbol \GG.

The lemma is a convention choice that assigns a core meaning of word forms that share similar stems. 
For example, the words \RL{byt}(house), 
\RL{llbyt}(for the house), 
and \RL{bywt}(houses) are mapped to the singular noun form \RL{byt} as a lemma.

The category is a user defined tag that the user can assign to several 
morphemes.
For example, the user can define a temporal category to include the 
prefix \RL{س} (will) and the time unit \RL{ساعة} (hour). 
We denote the set of all categories by the symbol \AC.

\subsection{Morphological analyzer}

An Arabic morphological analyzer takes an Arabic word 
and returns a set of morphological solutions. 
Each solution splits the word into its morphemes and 
associates each morpheme with corresponding tags. 
Multiple solutions can be the result of multiple valid 
segmentations of the word into morphemes, multiple 
possible tags associated with a morpheme.

\framework is integrated with {\em Sarf}, 
an in-house open source Arabic morphological analyzer based on 
finite state transducers~\cite{ZaMaColing2012DemosSarf}. 
We use Sarf to extract the morphological features of words present in a text. 
Sarf takes an Arabic word as input and returns a set of morphological solutions. 
Each solution splits the word into its morphemes and associates each morpheme with the corresponding tags. 
We refer to the morphemes and associated tags as features. 
These features are used to tag each input word with one or more user-defined tags. 
Sarf processes a word $w_{i}$ in a text and returns the relevant solution vectors. 
%Multiple solutions are the result of multiple valid segmentations of the word into morphemes, 
%or the multiple possible tags associated with a morpheme. 
%Also, a word might have multiple solution vectors as it can have multiple interpretations and meanings.
%The contribution of Sarf over previous work is that it considers the Arabic affixes as agglutinative, i.e. composed of one or more morphemes. 
%Hence, it introduces fusional compatibility rules for affix-affix concatenations. 
%This contribution results in a lexicon which is smaller in size, has less redundancy, and resolves a number of inconsistencies. 
%Figure~\ref{fig:sarfFSM} shows the finite state machine of Sarf. 
%Boxes denote legal affixes and stems, and circles denote regular nodes. 
%The edges are transitions and the labels correspond to the input letters. 
%$\epsilon$ represents an empty string and is the source of non-determinism.
%The transitions between the prefix \pp, stem \ss, and suffix \xx~sub-machines are non-deterministic to compute all valid morphological analyses.

%\transfalse
%\begin{figure}[h!]
%\begin{center}
%\resizebox{\columnwidth}{!}{
%	\input{figures/FSM.pdftex_t}
%}
%\caption{FSM of Sarf}
%\label{fig:sarfFSM}
%\end{center}
%\end{figure}
%\transtrue

%An example is shown in Table~\ref{tab:table1} below:

% Table generated by Excel2LaTeX from sheet 'Sheet1'
%\begin{table}[h]
%\setcode{utf8}
%\setarab
%  \centering
%    \begin{tabular}{|r|r|}
%    \toprule
%    \RL{وَلَد} & child/son \\
%    \midrule
%    \RL{وَلَّدَ}& generate \\
%    \midrule
%    \RL{وُلِدَ}& be born \\
%    \bottomrule
%    \end{tabular}%
%    \caption{Word with different interpretations}
%    \label{tab:table1}%
%\setcode{standard}
%\end{table}%

A sample solution vector of Sarf is shown in Table~\ref{tab:table2}. 
$w_{i}$-index shows the index of the word $w_{i}$ in the text. 
The terms pref and suf in the features refer to prefix and suffix respectively. 
The features pref-data, stem-data, and suf-data presents the prefix, stem, and suffix morphemes with diacritics. 
Sarf provides an additional feature called stem-ac. 
This feature shows a list of the abstract categories relevant to this stem. 
For example, ``Country'' is an abstract category for ``Lebanon''.

%By the end of the morphological analysis stage, Sarf returns a sequence of solution vectors. 
%These vectors contain all the possible morphological analyses of the words in the input text.
\setcode{standard}

### 2018-03-14 tex/merf/OldSubmission/relatedwork.tex
old submission files

--- 2018-03-13 tex/merf/TALMajor/relatedwork.tex
@@
 
-%SystemT\cite{chiticariu2010systemt}, 
-%TEXTMARKER\cite{atzmueller2008rule}, 
-%Urbain\cite{urbain2012user}, 
-%QARAB\cite{hammo2002qarab}
-
 \begin{table}[tb!]
-\caption{Comaprison of \framework with SystemT, 
-TEXTMARKER, 
-Urbain, 
-QARAB}
+\caption{Comaprison of IE system}
 \resizebox{\columnwidth}{!}{
-\begin{tabular}{l|c|c|c|c|c}
-%{p{2.7cm}|p{.7cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}}
-Features & \framework & SystemT & TEXT\linebreak MARKER & Urbain & QARAB\\ \hline
+\begin{tabular}{p{1.4cm}|p{.7cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}|p{1.2cm}}
+Features & \framework & SystemT\cite{chiticariu2010systemt} & TEXT\cite{atzmueller2008rule} \linebreak MARKER
+& Urbain\cite{urbain2012user} & QARAB\cite{hammo2002qarab}\\ \hline
 Query type & MRE & AQL & matching \linebreak rules & natural language & natural language \\ \hline
@@
 {\bf Information Extraction.~}
-The common pattern specification language (CPSL) targets system independent 
+The common pattern specification grammar language (CPSL) targets system independent 
 IE specifications. 
@@
 This interactive environment contributes to the regular expression 
-extraction and semantic relation construction which increases the overall accuracy.
+extraction and semantic relation construction which increases the overall accuracy of the framework.
 
-WordNet is a lexical reference system that mimics human lexical memory 
-%inspired by psycholinguistic theories 
-and that relates words based on their semantic values and their functional 
+WordNet is a lexical reference system that mimics human lexical memory inspired 
+by psycholinguistic theories and that relates words based on their semantic values and their functional 
 categories: nouns, verbs, adjectives, adverbs, and function words~\cite{intro}. 
+%The lexicon is composed of five categories: nouns, verbs, adjectives, adverbs, and function words; 
+%each category with its distinct organization to better represent the psychological complexity of lexical knowledge~\cite{nouns,adjectives}. 
 The $Syn^k$ feature in \framework is inspired by WordNet.
@@
 and syntactic annotation of the Quran. 
-\cite{dorr2010interlingual} presents a framework for interlingual 
+%The task is held through online supervised collaboration using a multi-stage approach that 
+%includes automatic POS tagging, manual verification, and online supervised collaborative 
+%proofreading. 
+The work in \cite{dorr2010interlingual} presents a framework for interlingual 
 annotation of parallel text corpora with multi-level representations. 
-\cite{kulick2010consistent} presents the integration of the 
+%This corpora is important for NLP tasks including text summarization, information retrieval, 
+%and machine translation. 
+The work in \cite{kulick2010consistent} presents the integration of the 
 Standard Arabic Morphological Analyzer (SAMA) into the workflow of the Arabic Treebank.
+%Such tasks motivated us to build \framework, a morphology-based open source annotation tool 
+%for the Arabic language.
 
@@
 of entity and relation annotation corpora~\cite{brat}. 
+%BRAT provides an API for automated annotators to provide annotations. 
 They can be extended through plug-ins to enable automatic annotators 
 and customized annotation visualization specifications.
+%AGTK is a toolkit for the development of text and speech annotation tools~\cite{agtk}. 
+%It provides import APIs from other data and graphical user interface (GUI) components. 
 Knowtator~\cite{ogren2006knowtator} is a general purpose incremental text annotation tool 
 implemented as a Prot\'eg\'e~\cite{gennari2003evolution} plug-in. 
-Prot\'eg\'e is an open-source platform with a suite of tools to construct domain 
-models and knowledge-based applications with ontologies. 
-However, it doesn't support the Arabic language.
+Prot\'eg\'e is a free, open-source platform with a suite of tools to construct domain 
+models and knowledge-based applications with ontologies~\cite{gennari2003evolution}. 
+However, Prot\'eg\'e doesn't support the Arabic language.
 
@@
 an automatic annotator that allows manual corrections 
-and sophisticated tag type and relation specifications over
+and sophisticated tag type and relation specifications using regular expressions over
 Arabic morphological features.
@@
 and extract entities and relations from text.
+%using Boolean formulae of several atomic terms and define regular expressions based on the Boolean formulae.

### 2018-03-14 tex/merf/OldSubmission/results.tex
old submission files

--- 2018-03-13 tex/merf/TAL/results.tex
@@
 In the online appendix%
-~\footnote{available at ~\url{-ommited-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
+~\footnote{Appendix available at ~\url{http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf}}%
 , we report on eight additional \framework case studies.
@@
 Table~\ref{tab:results} reports the development time,
-extraction runtime, recall and precision 
+entity extraction runtime, recall and precision accuracy metrics
 of the output MRE tags, 
@@
 \caption{\framework compared to task specific applications.}
-  \resizebox{0.85\columnwidth}{!}{
+  \resizebox{\columnwidth}{!}{
     \begin{tabular}{l|l|l|ll|l}
@@
 
+\begin{table}[tb!]
+  \centering
+  \caption{\framework MBF and user-defined relation accuracy }
+  \resizebox{\columnwidth}{!}{
+    \begin{tabular}{l|c|c|c|c}
+     \toprule
+     \multirow{2}{*}{Task} & \multicolumn{2}{c|}{MBF accuracy} & \multicolumn{2}{c}{relation accuracy}\\
+     & Recall & Precision & Recall & Precision \\
+    \midrule
+    Narrator chain & 0.99 & 0.85 & 0.99 & 0.98 \\
+    Number normalization & 0.99 & 0.99 & 0.97 & 0.95 \\
+    Temporal entity & 0.99 & 0.52 & 0.98 & 0.89 \\
+    Genealogy tree & 0.99 & 0.75 & 0.81 & 0.96 \\
+    \bottomrule
+    \end{tabular}%
+    }
+  \label{tab:mbfer}%
+\end{table}%
+
 %Recall refers to the fraction of the entities correctly detected against 
@@
 %\subsection{Comparison}
+
 Table~\ref{tab:results} shows that \framework has a clear advantage over 
@@
 a reasonable cost in terms of accuracy and run time. 
-Developers needed three hours, three hours, four hours, and one hour 
+Developers with \framework needed three hours, three hours, four hours, and one hour 
 to develop the narrator chain, temporal entity, genealogy, and number 
@@
 \framework required reasonably more runtime than the task specific 
-implementations and reported acceptable and 
-slightly less precision metrics with around
-the same recall.
+implementations and reported also reasonably acceptable and 
+slightly less preision metrics with around
+the same recall values. 
 %a runtime of 7.29, 1.53, 3.14, and 2.28 seconds for 
@@
 
-\setcode{utf8}
-\setarab
-\begin{table}[tb!]
-  \centering
-  \caption{Narrator chain example.}
-  \begin{Verbatim}[xleftmargin=1.5cm,fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
-name:   PN ((MEAN)? PN)*;
-nar:    name ((NONE)^3 FAM (NONE)^3 name)*;
-pbuh:   BLESS GOD UPONHIM GREET;
-nchain: ($s_1=$TOLD $s_2=$nar)+ ((PN|FAM|NONE)^8 pbuh)?
-\end{Verbatim}
-\resizebox{0.8\columnwidth}{!}{
-    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
-    \toprule 
-    \notrrl{القعقاع} & \notrrl{بن} & \notrrl{عمارة} & \notrrl{عن} & \notrrl{جرير} & \notrrl{حدثنا} & \notrrl{سعيد} & \notrrl{بن} & \notrrl{قتيبة} & \notrrl{حدثنا} \\
-    \midrule 
-    \noarrl{القعقاع} & \noarrl{بن} & \noarrl{عمارة} & \noarrl{عن} & \noarrl{جرير} & \noarrl{حدثنا} & \noarrl{سعيد} & \noarrl{بن} & \noarrl{قتيبة} & \noarrl{حدثنا} \\
-    \midrule
-    PN    & FAM   & PN    & TOLD  & PN    & TOLD  & PN    & FAM   & PN    & TOLD \\
-    \midrule
-    name & & name & & name & & name &  & name & \\
-    \midrule
-    \multicolumn{3}{|c|}{nar} &       & nar   &       & \multicolumn{3}{c|}{nar} &  \\
-    \midrule
-    \multicolumn{10}{|c|}{nchain}
-    \\
-    \bottomrule
-    \end{tabular}%
-}
-  \label{tab:nchain}%
-\end{table}%
-\setcode{standard}
+\vspace{-1em}
+\subsection{Narrator chain case study}
 
-\subsection{Narrator chain case study}
+%In this application, our target is to detect the narrator chains. 
 A narrator chain is a sequence of narrators referencing each other. 
-%A sample narrator chain is shown in Table~\ref{tab:nchain}. 
-The chain includes proper nouns, paternal entities, and referencing entities. 
+A sample narrator chain is shown in Table~\ref{tab:nchain}. 
+The chain includes proper nouns (names), paternal entities, and referencing entities. 
 ANGE uses Arabic morphological analysis, finite state machines, and graph transformations 
-to extract named entities and relations including narrator chains~\cite{ZaMaFlairs2012HadithBio}.
+to extract named entities and relations including the narrator chains~\cite{ZaMaFlairs2012HadithBio}.
 
-\transfalse
-MBF \cci{PN} checks the abstract category {\tt Name of Person}. 
+First, the MBF \cci{PN} checks the abstract category {\tt Name of Person}. 
 MBF \cci{FAM} denotes ``family connector'' and checks the stem gloss ``son''. 
@@
 stems \RL{.sll_A}, \RL{Al-ll_ah}, \RL{`ly}, and \RL{sllm}, respectively.
-\transtrue
 
-Table~\ref{tab:nchain} presents the defined MREs. 
-MRE {\em name} is one or more \cci{PN} tags optionally followed 
+Table~\ref{tab:nchain} presents the defined MRE expressions. 
+MRE {\em name} is satisfied by one or more \cci{PN} tags optionally followed 
 with a \cci{MEAN} tag. 
@@
 %\cci{NONE} tags followed by a \cci{pbuh} tag.
+
+\setcode{utf8}
+\setarab
+\begin{table}[tb!]
+  \centering
+  \caption{Narrator chain example.}
+\resizebox{\columnwidth}{!}{
+    \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|}
+    \toprule 
+    \notrrl{القعقاع} & \notrrl{بن} & \notrrl{عمارة} & \notrrl{عن} & \notrrl{جرير} & \notrrl{حدثنا} & \notrrl{سعيد} & \notrrl{بن} & \notrrl{قتيبة} & \notrrl{حدثنا} \\
+    \midrule 
+    \noarrl{القعقاع} & \noarrl{بن} & \noarrl{عمارة} & \noarrl{عن} & \noarrl{جرير} & \noarrl{حدثنا} & \noarrl{سعيد} & \noarrl{بن} & \noarrl{قتيبة} & \noarrl{حدثنا} \\
+    \midrule
+    PN    & FAM   & PN    & TOLD  & PN    & TOLD  & PN    & FAM   & PN    & TOLD \\
+    \midrule
+    name & & name & & name & & name &  & name & \\
+    \midrule
+    \multicolumn{3}{|c|}{nar} &       & nar   &       & \multicolumn{3}{c|}{nar} &  \\
+    \midrule
+    \multicolumn{10}{|c|}{nchain}
+    \\
+    \bottomrule
+    \end{tabular}%
+}
+\begin{Verbatim}[fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
+name:   PN ((MEAN)? PN)*;
+nar:    name ((NONE)^3 FAM (NONE)^3 name)*;
+pbuh:   BLESS GOD UPONHIM GREET;
+nchain: ($s_1=$TOLD $s_2=$nar)+ ((PN|FAM|NONE)^8 pbuh)?
+\end{Verbatim}
+  \label{tab:nchain}%
+  \vspace{-3em}
+\end{table}%
+\setcode{standard}
 
@@
 
-\begin{table}[tb!]
-  \centering
-  \caption{\framework MBF and user-defined relation accuracy }
-  \resizebox{0.65\columnwidth}{!}{
-    \begin{tabular}{l|c|c|c|c}
-     \toprule
-     \multirow{2}{*}{Task} & \multicolumn{2}{c|}{MBF accuracy} & \multicolumn{2}{c}{relation accuracy}\\
-     & Recall & Precision & Recall & Precision \\
-    \midrule
-    Narrator chain & 0.99 & 0.85 & 0.99 & 0.98 \\
-    Number normalization & 0.99 & 0.99 & 0.97 & 0.95 \\
-    Temporal entity & 0.99 & 0.52 & 0.98 & 0.89 \\
-    Genealogy tree & 0.99 & 0.75 & 0.81 & 0.96 \\
-    \bottomrule
-    \end{tabular}%
-    }
-  \label{tab:mbfer}%
-\end{table}%
-
 \vspace{-1em}
@@
 morphological analysis and finite state transducers~\cite{ZaMa2012IJCLATime}.
+
+%Table~\ref{tab:tbtags} shows the temporal MBFs. 
+%MBF \cci{TIME} is explicit temporal words. 
+%MBF \cci{NUM}, denotes numerals, and is digits or words referring to numbers. 
+%MBF \cci{TIMEPREP} denotes temporal prepositions.
+
+%% Table generated by Excel2LaTeX from sheet 'Sheet1'
+%\begin{table}[tb!]
+%  \centering
+%  \caption{Temporal MBFs}
+%  \resizebox{0.8\columnwidth}{!}{
+%    \begin{tabular}{|r|r|r|}
+%    \toprule
+%    MBF & & Examples \\
+%    \midrule
+%    \multicolumn{1}{|c|}{\multirow{5}[0]{*}{TIME}} & unit  & \RL{dqyqT} (minute) \\
+%    \multicolumn{1}{|c|}{} & relative & \RL{.gd} (tomorrow) \\
+%    \multicolumn{1}{|c|}{} & range & \RL{rby`} (spring) \\
+%    \multicolumn{1}{|c|}{} & nominal & \RL{'.hd} (Sunday) \\
+%    \multicolumn{1}{|c|}{} & events & \RL{hjry} \\
+%    \midrule
+%    \multicolumn{1}{|c|}{\multirow{3}[0]{*}{NUM}} & digit & \RL{3} (3) \\
+%    \multicolumn{1}{|c|}{} & word  & \RL{_tmAnyn} (eighty) \\
+%    \multicolumn{1}{|c|}{} & range & \RL{_tmAnynyAt} (eighties) \\
+%    \midrule
+%    \multicolumn{1}{|c|}{\multirow{4}[0]{*}{TIMEPREP}} & point & \RL{fy} (in) \\
+%    \multicolumn{1}{|c|}{} & relative & \RL{qbl} (before) \\
+%    \multicolumn{1}{|c|}{} & approximate & \RL{n.hw} (about) \\
+%    \multicolumn{1}{|c|}{} & range & \RL{xlAl} (during) \\
+%    \bottomrule
+%    \end{tabular}%
+%    }
+%  \label{tab:tbtags}%
+%\end{table}%
+%
+%MRE {\em mts}, denotes {\tt maybe time start}, and is a disjunction of \cci{NUM} tag or \cci{TIMEPREP} tag. 
+%MRE \cci{dt}, denotes {\tt definitetime}, and is zero or more sequence of \cci{mts} tags and up to two 
+%\cci{NONE} tags, \cci{TIME} tag, then an optional sub-expression. 
+%The sub-expression is a sequence of zero or more sequence of up to two \cci{NONE} tags followed by \cci{mts} tag or \cci{TIME} tag followed by up to two \cci{NONE} tags and \cci{TIME} tag or \cci{NUM} tag.
+%The last sub-expression \cci{TIME|NUM} is introduced to ensure that the temporal match ends with a \cci{TIME} tag or \cci{NUM} tag and not \cci{TIMEPREP} tag.
+%
+%\begin{Verbatim}[fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
+%mts : NUM | TIMEPREP;
+%dt : ( $\stackrel{s_1}{mts}$ NONE^2)* $\stackrel{s_2}{TIME}$
+%       ( (NONE^2 $\stackrel{s_3}{( mts|TIME )}$)* 
+%         NONE^2 $\stackrel{s_4}{( TIME|NUM )}$ )?;
+%\end{Verbatim}
+%
+%\framework assigns the symbols $s_1$, $s_2$, $s_3$, and $s_4$ for the sub-expressions \cci{mts}, \cci{TIME}, \cci{mts|TIME}, and \cci{TIME|NUM}, respectively. 
+%The symbols are shown in the regular expression defined above. 
+%Using these symbols, we define the user-defined relations $r_1=\langle s_1,s_2,\epsilon\rangle$, $r_2=\langle s_2,s_3,\epsilon\rangle$, and $r_3=\langle s_3,s_4,\epsilon\rangle$. 
+%$r_1$ relates \cci{mts} matches to TIME matches. 
+%$r_2$ relates \cci{mts|TIME} matches to TIME matches. 
+%$r_3$ relates \cci{mts|TIME} matches to \cci{TIME|NUM} matches.
+%
+%% Table generated by Excel2LaTeX from sheet 'Sheet1'
+%\begin{table}[tb!]
+%  \centering
+%  \caption{Temporal entity example}
+%  \resizebox{\columnwidth}{!}{
+%    \begin{tabular}{|r|r|r|r|r|r|r|r|}
+%    \toprule
+%    \RL{AlsnwAt} & \RL{'.s`b} & \RL{mn} & \RL{w_hms} & \RL{`jAf} & \RL{'^shr} & \RL{'rb`T} & \RL{b`d} \\
+%    \midrule
+%    TIME  & NONE  & NONE  & NUM   & NONE  & TIME  & NUM   & TIMEPREP \\
+%    \midrule
+%          &       &       & mts   &       &       & mts   & mts \\
+%    \midrule
+%    \multicolumn{8}{|c|}{definitetime} \\
+%    \bottomrule
+%    \end{tabular}%
+%    }
+%  \label{tab:tempex}%
+%\end{table}%
+%
+%The first row in Table~\ref{tab:tempex} is a temporal entity Arabic text. 
+%Each word is denoted with MBF tags based on its morphological features. 
+%The matches of the MRE {\tt mts} and {\tt dt} are shown as tags in rows 3 and 4, respectively.
 
@@
 finite state machines, and graph transformations. 
+%
+%We defined the MBFs \cci{MN}, \cci{FN}, and \cci{CONN}. 
+%MBF \cci{MN}, denotes male name, and is defined by the {\tt male name} category. 
+%MBF \cci{FN}, denotes female name, and is defined by the {\tt female name} category. 
+%MBF \cci{CONN}, denotes connection, and is defined by the stem \RL{wld} (give birth to).
+%
+%We define the MREs \cci{fform} and \cci{sform}. 
+%MRE \cci{fform}, denotes first form, and is \cci{MN} tag followed by an optional \cci{FN} tag 
+%followed by a one or more sequences of \cci{CONN} and \cci{MN} tags.
+%
+%\begin{Verbatim}[fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
+%fform : $\stackrel{s_1}{MN}$ NONE^4 $\stackrel{s_2}{FN}$? (NONE^10 $\stackrel{s_3}{CONN}$ NONE^4 $\stackrel{s_4}{MN}$)+
+%\end{Verbatim}
+%
+%MRE \cci{sform}, denotes second form, and is \cci{CONN} tag followed by \cci{MN} or \cci{FN} tag then \cci{MN} tag. 
+%
+%\begin{Verbatim}[fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
+%sform : $\stackrel{s_5}{CONN}$ NONE^4 $\stackrel{s_6}{(MN|FN)}$ NONE^4 $\stackrel{s_7}{MN}$
+%\end{Verbatim}
+%
+%MRE \cci{geneinfo}, denotes genealogical information, and is \cci{fform} tag or \cci{sform} tag.
+%
+%\begin{Verbatim}[fontsize=\relsize{-1}]
+%geneinfo : fform | sform
+%\end{Verbatim}
+%
+%\framework assigns the symbols $s_1$, $s_2$, $s_3$, $s_4$, $s_5$, $s_6$, and $s_7$ for the MRE sub-expressions \cci{MN}, \cci{FN}, \cci{CONN}, \cci{MN}, \cci{CONN}, \cci{MN|FN}, and \cci{MN}, respectively.
+%These symbols are shown over the sub-expressions in the defined MREs.
+%
+%Based on those symbols, we define the user-defined relations $r_1=\langle s_1,s_4,s_3\rangle$, $r_2=\langle s_1,s_2,``spouse''\rangle$, $r_3=\langle s_2,s_4,s_3\rangle$, and $r_4=\langle s_6,s_7,s_5\rangle$.
+%$r_1$ relates $s_1$ matches to $s_4$ matches with an edge labeled with $s_3$ matches. 
+%This relation indicates that entity $s_1$ is the father of $s_4$. 
+%$r_2$ relates $s_2$ matches to $s_1$ matches with an edge labeled by {\tt spouse}. 
+%This relation indicates that the female is the male's spouse. 
+%$r_3$ relates $s_4$ matches to $s_2$ matches using an edge labeled by $s_3$. 
+%This relation indicates that the male is the female's child. 
+%$r_4$ relates $s_6$ matches to $s_7$ matches using an edge labeled by $s_5$ matches. 
+%This relation indicates that $s_7$ match is the child of $s_6$.
+%
 Table~\ref{tab:mbfer} shows that \framework detected 
@@
 
-\vspace{-1em}
 \subsection{Number normalization}
@@
 C++ implementation for number normalization.
-First, we defined the MBFs \cci{DT}, \cci{H}, and \cci{TMB}
+
+%NUMNORM reads an Arabic text as input, extracts the text chunks with numerical information, and normalizes them. 
+%NUMNORM uses the in-house morphological analyzer, {\em Sarf}, to identify the words referring to numbers. 
+%Sarf provides a {\tt Number} category that tags all words with numerical information. 
+%In the normalization algorithm, NUMNORM identifies three different categories for numbers;
+%each with different behavior in normalization. 
+%The categories are:
+%\begin{itemize}
+%\item DT, denoting digits and tens
+%\item H, denoting hundred
+%\item TMB, denoting thousand, million, and billion
+%\end{itemize}
+%
+%NUMNORM retrieves the number referring to a word through a fixed map. 
+%This map takes the word gloss as input and returns the matching number if found. 
+%For example, the map returns 100 for the input gloss {\tt hundred}. 
+%As for the normalization of more than one word, such as \RL{'lfyn w-'rb`T `^sr} (2014), 
+%the normalization algorithm used for each category is shown in Figure~\ref{fig:numnormalgo}.  
+%The algorithm uses Boolean variables and three integer variables. 
+%The three integer variables are previous, current, and currentH for hundred. 
+%The main steps in the algorithm are as follows. 
+%Hundred is multiplied to previous digit if found, else saved. 
+%Digits and tens are added to previous hundred if found, else added to current. 
+%TMB match is multiplied by the previous hundred or DT if found, 
+%else saved in current.
+
+\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
+
+\begin{figure}[tb!]
+\centering
+  \begin{tabular}{p{3.5cm}p{3.5cm}}
+\begin{Verbatim}[fontsize=\relsize{-3.5},frame=single,label=TMB algorithm,commandchars=\\\[\]] 
+cout << \fvtextcolor[red][$s1.text];
+if(isHundred) {
+  if(current != 0) {
+    previous += current;
+  }
+  current = currentH*\fvtextcolor[red][$s1.number];
+  currentH = 0;
+  isHundred = false;
+  isKey = true;
+} else if(current == 0) {
+
+  current=\fvtextcolor[red][$s1.number];
+  isKey = true;
+} else if(!isKey) {
+
+  isKey = true;
+  current = current*\fvtextcolor[red][$s1.number];
+} else {
+  previous += current;
+  current = \fvtextcolor[red][$s1.number];
+}
+\end{Verbatim}
+&
+\begin{Verbatim}[fontsize=\relsize{-3.5},frame=single,label=DT algorithm,commandchars=\\\[\]] 
+if(isHundred) {
+  currentH +=\fvtextcolor[red][$s0.number];
+} else if(current == 0) {
+  current=\fvtextcolor[red][$s0.number];
+} else if(isKey) {
+  previous += current;
+  current = \fvtextcolor[red][$s0.number];
+} else {
+  current += \fvtextcolor[red][$s0.number]; }
+isKey = false;
+\end{Verbatim}
+
+\begin{Verbatim}[fontsize=\relsize{-3.5},frame=single,label=H algorithm,commandchars=\\\[\]] 
+isHundred = true;
+if(current == 0)  {
+  currentH=\fvtextcolor[red][$s2.number];
+} else if(!isKey) {
+  currentH = current*\fvtextcolor[red][$s2.number];
+  current = 0;
+} else {
+  currentH = \fvtextcolor[red][$s2.number];}
+isKey = false;
+\end{Verbatim}
+\\ 
+\end{tabular}
+\caption{Actions for TMB, DT, and H MRE expressions.}
+\label{fig:numnormalgo}
+  \vspace{-1em}
+\end{figure}
+
+%Using \framework, we designed an MRE to detect the numerical entities. 
+%We added actions to the MRE to normalize matches. 
+First, we defined the MBF formulae \cci{DT}, \cci{H}, and \cci{TMB}
 to denote (1) digits and tens, (2) hundreds, and (3) 
 thousands, millions, and billions, respectively.
+
 The \cci{num} MRE 
@@
 for the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H}, respectively. 
-The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
-%Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.
+Figure~\ref{fig:numnormalgo} shows the actions associated with the 
+\cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute 
+the numeric value of the numeric expression match.
 The actions use \framework API to access features of the matches such 
-as text (\cci{\$s1.text}) and numeric 
-value (\cci{\$s1.number}) of literal numbers such as numbers from one to ten.
+as the text (\cci{\$s1.text}) and the numeric 
+value (\cci{\$s1.number}). 
+of literal numbers such as digits and numbers from one to ten.
+
+\framework also assigns the symbol $s_3$ for the sub-expression \cci{DT|TMB|H}. 
+Accordingly, we define the user-defined relation $\langle s3,s3',\epsilon\rangle$. 
+This relation identifies the matches of $s_3$ as entities and relates them successively with non-labelled edges.
 
 Table~\ref{tab:mbfer} shows high accuracy in MBF tagging with 99\% recall and 99\% precision, and
-high accuracy in relation extraction with 97\% recall and 95\% precision.
-
-%\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
-%\begin{figure}[tb!]
-%\centering
-%  \begin{tabular}{p{5.5cm}p{5.5cm}}
-%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=TMB algorithm,commandchars=\\\[\]] 
-%cout << \fvtextcolor[red][$s1.text];
-%if(isHundred) {
-%  if(current != 0) {
-%    previous += current;
-%  }
-%  current = currentH * \fvtextcolor[red][$s1.number];
-%  currentH = 0;
-%  isHundred = false;
-%  isKey = true;
-%} else if(current == 0) {
-%  
-%  current = \fvtextcolor[red][$s1.number];
-%  isKey = true;
-%} else if(!isKey) {
-%
-%  isKey = true;
-%  current = current * \fvtextcolor[red][$s1.number];
-%} else {
-%  previous += current;
-%  current = \fvtextcolor[red][$s1.number];
-%}
-%\end{Verbatim}
-%&
-%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=DT algorithm,commandchars=\\\[\]] 
-%if(isHundred) {
-%  currentH += \fvtextcolor[red][$s0.number];
-%} else if(current == 0) {
-%  current = \fvtextcolor[red][$s0.number];
-%} else if(isKey) {
-%  previous += current;
-%  current = \fvtextcolor[red][$s0.number];
-%} else {
-%  current += \fvtextcolor[red][$s0.number]; }
-%isKey = false;
-%\end{Verbatim}
-%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=H algorithm,commandchars=\\\[\]] 
-%isHundred = true;
-%if(current == 0)  {
-%  currentH = \fvtextcolor[red][$s2.number];
-%} else if(!isKey) {
-%  currentH = current * \fvtextcolor[red][$s2.number];
-%  current = 0;
-%} else {
-%  currentH = \fvtextcolor[red][$s2.number];}
-%isKey = false;
-%\end{Verbatim}
-%\\ 
-%\end{tabular}
-%\caption{Actions for TMB, DT, and H MRE expressions.}
-%\label{fig:numnormalgo}
-%\end{figure}
-
-\vspace{-1em}
-\subsection{Discussion}
-\label{subsec:discuss}
-\input{discussion}
+high accuracy in user-defined relation extraction with 97\% recall and 95\% precision.

### 2018-03-14 tex/merf/OldSubmission/tool.tex
old submission files

--- 2018-03-13 tex/merf/TAL/tool.tex
@@
-The \framework framework is illustrated in the flow diagram of
-Figure~\ref{f:overview}.
-The Arabic text and the reference tag chunks are the primary inputs to \framework. 
-Solutions, morphology-based Boolean formulae, tags, 
-morphology-based regular expressions, 
-tag chunks, relation and action definitions, and data structures expressing entities 
-and relations are input and output data to processes. 
-The morphological analyzer (Sarf), $Syn^k$ detector, 
-GUI for Boolean formulae definition, 
-visualization annotator, GUI for regular expression and action definition, 
-Boolean formula simulator, regular expression simulator, 
-relation extraction and action execution, and difference and statistical analyzer 
-are processes.
+This section assumes that the user passed a set of text documents and 
+specified a set of tag types using the interactive user friendly
+interface of \framework. 
+Let $T=\langle t_1,t_2,\ldots,t_M\rangle$ be a set of Arabic words denoting the 
+text documents, and let ${\cal T}$  be the 
+set of tag types each with an associated formula. 
+\framework uses Sarf~\cite{ZaMaColing2012DemosSarf},
+to compute a set of morphological solutions $M(t)=\{m_1,m_2,\ldots,m_N\}$
+for each word $t\in T$.
 
-%The user interacts with \framework~through a user-friendly interfaces to 
-%specify morphology-based Boolean formulae and regular expressions 
-%and associate them with tag types that are in turn associated with
-%visualization legends such as 
-%foreground and background colors and fonts. 
-%The user also associates code actions to parts of the regular expressions.
-%The user can also declare relations between matches using the relation
-%editor.
+Recall from Section~\ref{sec:overview} 
+that \pp, \ss, \xx, \PP, \GG, and \AC~denote the set of prefixes, stems, 
+suffixes, POS, gloss, and category tags, respectively. 
+Each morphological solution $m\in M(t)$ is a tuple of the form 
+$\langle p,s,x,P,G,C\rangle\in \pp \times  \ss\times \xx\times \PP\times\GG \times \AC$.
+%where $p\in \pp, s\in \ss, x\in \xx, P \in \PP, G \in \GG,$ and $C \in \AC$.
+
+In what follows we formally present 
+the extended synonymy feature, 
+the morphology-based atomic terms (MAT), 
+the morphology-based Boolean formulae (MBF), 
+the morphology-based regular expressions (MRE), 
+the computational actions, and the user defined relational entities.
+
+%\begin{table}[tb!]
+%\centering
+%\caption{Example of computing $Sy(l_{5})$ and $Sy^{1}(l_{5})$}
+%\begin{tabular}{|c|c|c|c|c|c|}
+%\hline
+%$s$ & $\alpha(s) $ & $\alpha(l_{1})\cap\alpha(s)$ & $\alpha(l_{2})\cap\alpha(s)$ & $\alpha(l_{3})\cap\alpha(s)$ \\
+%\hline
+%$l_{1}$ & $e_{1}$,$e_{2}$ & $e_{1}$,$e_{2}$ & $e_{2}$ & $e_{2}$ \\
+%\hline
+%$l_{2}$ & $e_{2}$ & $e_{2}$ & $e_{2}$ & $e_{2}$ \\
+%\hline
+%$l_{3}$ & $e_{2},e_{3}$ & $e_{2}$ & $e_{2}$ & $e_{2},e_{3}$ \\ 
+%\hline
+%$l_{4}$ & $e_{3},e_{4}$ & $\emptyset$ & $\emptyset$ & $e_{3}$ \\
+%\hline 
+%\end{tabular}
+%\label{table:syex}
+%\end{table}
+%
+
+\begin{figure}[tb!]
+\setcode{utf8}
+\begin{center}
+  \resizebox{\columnwidth}{!}{ 
+  	{\relsize{-3} \input{figures/syn2ex.tex}} }
+\setcode{standard}
+\vspace{-2em}
+\caption{$Syn^2($\RL{mA'}$)$}
+\vspace{-3em}
+\label{fig:introsynEx}
+\end{center}
+\end{figure}
 
@@
 Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
-subsets of related English glosses. %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
+subsets of related English glosses; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
 Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
-of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.
+of relevant Arabic stems; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.
 
@@
 directly related to $w$ through the gloss map.
+For example, consider % the example of Table~\ref{table:syex} where
+$\ss$ = $\{l_{1}, l_{2}, l_{3}, l_{4}\}$, 
+$L$ = $\ss \cup \{l_{5}, l_{6}\}$,
+$\GG$ = $\{e_{1}, e_{2}, e_{3}, e_{4}\}$, 
+$E$ = $\GG \cup \{e_{5}, e_{6}\}$, 
+$\alpha$ = $\{ \mathbf{(l_{1}, \{e_{1}, e_{2}\})}, 
+\mathbf{(l_{2}, \{e_{2}\})}, (l_{3}, \{e_{2}, e_{3}\}),
+(l_{4}, \{e_{3}, e_{4}\})\}$,  and 
+$\gamma$ = $\{ (l_{1}, \{l_{1}\}),  
+  (l_{2}, \{l_{2}\}),  (l_{3},  \{l_{3}\}), (l_{4},  \{l_{4}\}), \\
+\mathbf{(l_{5}, \{l_{1}, l_{2}\})},  (l_{6}, \{l_{4}\})\}$.
+
+Given the Arabic word $l_{5}$ with stems
+$\gamma(l_5) = \{l_{1},l_{2}\}$ and related glosses
+$\alpha(l_{1}) = \{e_{1},e_{2}\} \cup \alpha(l_{2}) = \{e_{2}\}$. 
+%Table~\ref{table:syex} illustrates the process of computing 
+%$Sy(l_5)$ is the stems whose glosses have
+%some intersection with the glosses related to $l_{5}$. 
+Stem $l_3$ is in $Sy(l_5)$ since $\alpha(l_{3})\cap\alpha(l_{1}) = \{e_{2}\}\not=\emptyset$;
+similarly, for  $l_1$ and $l_2$. However, $l_4$ is not in $Sy(l_5)$ since 
+$\alpha(l_{4})\cap\alpha(l_{5}) = \emptyset$.
+Consequently, $Sy(w) = \{l_{1},l_{2},l_{3}\}$.
 
@@
 $Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
-Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
+For example, $Sy^{2}(l_{5})$ includes  $l_{4}$ since $(\alpha(l_{3})\cap\alpha(l_{4}))\neq\emptyset$. 
+Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$
+to be the union of $Sy^{i}(w)$ for $i\in[1 \ldots k]$.
 
 The example in Figure~\ref{fig:introsynEx} illustrates the computation. 
-The word \RL{mA'}is related to \RL{n.d.h}through the gloss intersection \cci{water}. 
+%Words \RL{mA'}, \RL{n.d.h}, and \RL{r^s^s} and their glosses
+%\{water\}, \{water, leak, spray\}, and \{spray, splatter\}, respectively,
+%shown in the example of Figure~\ref{fig:introsynEx}. 
+The word \RL{mA'} is related to \RL{n.d.h} through the gloss intersection \cci{water}. 
 The $Syn^2$ feature relates 
@@
 
-\begin{figure}[tb!]
-\setcode{utf8}
-\begin{center}
-  \resizebox{0.7\columnwidth}{!}{ 
-  	{\relsize{-2} \input{figures/syn2ex.tex}} }
-\setcode{standard}
-\caption{$Syn^2($\RL{mA'}$)$}
-\label{fig:introsynEx}
-\end{center}
-\end{figure}
+%\begin{figure}[tb!]
+%\setcode{utf8}
+%\begin{center}
+%  \resizebox{0.8\columnwidth}{!}{ 
+%  	{\relsize{-3} \input{figures/syn2ex.tex}}
+%  }
+%  \setcode{standard}
+%  \caption{$Syn^2($\RL{.t`Am}$)$}
+%  \label{fig:synEx}
+%\end{center}
+%\end{figure}
 
+\vspace{-3em}
 \subsection{MRE: Morphology-based regular expressions}
@@
 Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
-predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
+predicates, and ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
 set of morphological features. 
@@
 \item $a(w) := w \in Syn^k(CF), CF \in \ss$.
+%\item $a(w) := \exists s\in \gamma(w).s\in Syn^k(CF)
+%  \textit{where} k\in\{1,\dots,7\}$. 
   Informally, this checks $w$ is an extended synonym of a stem $CF$. 
@@
 
+%Consider the motivational example introduced in Section~\ref{sec:motivation}. 
+%Some of the MATs defined are:
+
+%\begin{itemize}
+%\item $t_1=C~isA~Name\_of\_Person$
+%\item $t_2=s~isA~\RL{qrb}$
+%\item $t_3=s~isA~\RL{fy}$
+%\end{itemize}
+
+%$t_1$ checks whether a word has the category {\tt Name\_of\_Person}.  
+%$t_2$ checks whether the stem of a word is \RL{qrb} (near). 
+%$t_3$ checks whether the stem of a word is \RL{fy} (in). 
+%Note that a word can be tagged with more than one MAT.
+
 A morphology-based Boolean formula (MBF) is of the following form. 
 \begin{itemize}
-  \item $a$ and $\neg a$ are MBF formulae where $a$ is a MAT and $\neg$ is the negation operator. 
+  \item $a$ and $\neg a$ are MBF formulae where $a$ an MAT, where
+    $\neg$ is the negation (complement) operator. 
   \item $(f \vee g)$ is an MBF where $f$ and $g$ are MBF formulae, 
@@
 \item $m$ is an MRE where $m$ is an MBF.
-\item $fg$ is an MRE denoting a concatenation operation
+\item $fg$ is an MRE denoting a concatenation or a sequencing operation
   where $f$ and $g$ are both MRE expressions. This is 
-  satisfied by a match of $f$ followed by a match of $g$. 
+  satisfied by a match of $f$ followed directly by a match of $g$. 
 \item $f*,f+,f$\textasciicircum$x,$ and $f?$ are MRE expressions where $f$ is an MRE,
@@
   and zero or only one match of $f$, respectively.
+%  This refers to the {\em Kleene-star} operation, and is satisfied by 
+%  zero or more matches of $f$.
+%\item $f+$ is an MRE where $f$ is an MRE. 
+%This operation is interpreted as one or more matches of $f$.
+%\item $f$\textasciicircum$x$ is an MRE where $f$ is an MRE and $x \in \mathbb{N^*}$. 
+%This is interpreted as zero or more matches of $f$ up to $x$ matches.
+%\item $f?$ is an MRE where $f$ is an MRE. 
+%This is interpreted as zero or one occurrence of MRE $f$.
 \item $f\& g,$ (conjunction) and $f|g$ (disjunction) are  MRE expressions where 
@@
 
+%\itodo{handle the following}
+%Moreover, \framework~defines the 
+%{\em default cross-reference} relation between two 
+%entities $e_1$ and $e_2$ as
+%$cr(e_1,e_2) := e_1.text \in Syn^2(e_2.text)$.
+%\itodo{end of handle the following (:)}
+
 \framework allows advanced users to write C++ code snippets 
@@
 Once \framework computes all match trees, it traverses each match to 
-execute the user defined \cci{pre-match} actions in pre-order manner
-and the \cci{on-match} actions in post-order manner.
+execute the user defined \cci{pre-match} actions in a pre-order manner
+and the \cci{on-match} actions in a post-order manner.
+% as illustrated in the code of Figure~\ref{fig:actions-traverse}.
+
+%In addition the user can also specify code headers to associate the expressions 
+%with proprietary code such as declaring global variables and including user libraries.
+
+%\begin{figure}[tb!]
+
+%\begin{Verbatim}[fontsize=\relsize{-2},frame=lines,label=Match tree traverse to execute actions,commandchars=\\\[\]]
+%traverse (node) {    //node is the match tree
+%  pre-match(node)    //call user defined pre-match function
+%  foreach child in node.children
+%    traverse(child)  //recursively traverse children in tree
+%  on-match(node)}    //call on-match function 
+%\end{Verbatim}
+
+%\begin{Verbatim}[fontsize=\relsize{-2},frame=lines,label=\cci{On-match} example,commandchars=\\\[\]]
+%cout << \fvtextcolor[red][$s2.text];
+%isHundred = true;
+%if(current == 0)  {
+%  currentH=\fvtextcolor[red][$s2.number];
+%} else if(!isKey) {
+%  currentH= current * \fvtextcolor[red][$s2.number];
+%  current = 0;
+%} else {
+%  currentH = \fvtextcolor[red][$s2.number]; }
+%isKey = false;
+%\end{Verbatim}
+
+%\caption{User-defined computational actions} 
+%%(a) match tree traverse & (b) example \cci{on-match} action
+%\label{fig:actionEx}
+%\label{fig:actions-traverse}
+%\end{figure}
+
+%Consider the sample code action shown in Figure~\ref{fig:numnormalgo} (TMB algorithm). 
+%This code action is used in the number normalization case study of Section~\ref{sec:results}. 
+%The variables \cci{isHundred}, \cci{isKey}, \cci{current}, and \cci{currentH} are global variables declared by the user. 
+%The use of \cci{cout} requires the user to include the \cci{iostream} library. 
+%In the code the user can use the API with the \$ sign with an expression name to retrieve features of the match expressions. 
+%In this example, the user prints the sub-expression text and processes its numerical value.
+%
+%The use can access the morphological features with the API {\tt \$sub\_expression\_name.matches}.
+%The API returns a vector of morphological solutions where each solution contains the stem, 
+%arrays of affixes (prefixes and suffixes), stem gloss, arrays for affix glosses, 
+%stem POS, and arrays for affix POS tags.
 
@@
 where $l$ is a text label with a descriptive name, 
-$f$ is an MRE, and $d$ is a visualization legend 
-with font and color information.
+$f$ is an MRE expression, and $d$ is a visualization legend
+%The visualization legend describes how the tags should be displayed to the user 
+that contains font and color information.
+%such as the foreground and background colors, the
+%font size, weight, and style.
 
-For each word $t_i\in T, 0\le i < n, n=|T|$ 
+For each word $t_i \in T, 0\le i < n, n=|T|$ 
 \framework computes a Boolean value 
 ($\{\mathit{true}, \mathit{false}\}$)
-for all MBFs. 
-Then, it computes the set of MBF tags
+for all MBF formulae. 
+and computes the set of MBF tags
 $R_i=\{(t_i,tt)| tt=\langle l,f,d\rangle \wedge
@@
 which tags a word $t_i$ with $\mathit{tt}$ 
-iff the MBF $f$ associated with
+iff the MBF formula $f$ associated with
 tag type $\mathit{tt}$ is true for $t_i$.
@@
 corresponding to the text sequence 
-$\langle t_k,t_{k+1},\dots,t_j\rangle$ where $r_{\ell}\in R_{\ell},0\le k\le \ell \le j < n$.
+$\langle t_k,t_{k+1},\dots,t_j\rangle$ where 
+where $r_{\ell}\in R_{\ell},0\le k\le \ell \le j < n$.
 %
@@
 \centering
-\resizebox{0.7\columnwidth}{!}{
+\resizebox{0.8\columnwidth}{!}{
 	\relsize{-1} \input{figures/reNFA.tex}
@@
   \caption{Equivalent NFA of direction expression}
+\vspace{-2em}
   \label{fig:nfaEx}
@@
 \framework computes the relational entities in a user defined 
-relation $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
-$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$ 
+relation $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack
+\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$ 
 to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
@@
 
-For the first match \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'-wl},
+Similarly, for the  
+first match \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'-wl},
 the matches of $e_1,~e_2,~o_2,$ and $r$ are 

### 2018-03-16 tex/merf/TALMajor/abstract.tex
In progress: merging motivation with section 3

@@
-Rule-based techniques and tools to extract {\em entities} and {\em relational entities} from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.
-These techniques and tools require expertise in linguistics and programming and lack support of Arabic {\em morphological analysis} which is key to process Arabic text.
+Rule-based techniques and tools to extract entities and relational entities from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.
+These techniques and tools require expertise in linguistics and programming and they lack support of Arabic {\em morphological analysis} which is key to process Arabic text.
 In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.

### 2018-03-16 tex/merf/TALMajor/introduction.tex
In progress: merging motivation with section 3

@@
 understanding of natural language. 
-CL and NLP tasks range from simple ones such as spell checking and typing 
-error correction to more 
-complex tasks including %text summarization, 
+CL and NLP tasks range from simple ones such as spell checking and typing error correction to more complex tasks including %text summarization, 
 {\em named entity recognition} (NER), {\em cross-document analysis}, 
-{\em machine translation}, 
-and {\em relational entity extraction}~\cite{linckels2011natural,ferilli2011natural}.
+{\em machine translation}, and {\em relational entity extraction}~\cite{linckels2011natural,ferilli2011natural}.
 Entities are elements of text that are of interest to an NLP task.
@@
 semantic values such as entities or relational entities.
-We refer to annotations and labels as {\em tags} and {\em tag types}, respectively, 
-in the sequel.
+We refer to annotations and labels as {\em tags} and {\em tag types}, respectively, in the sequel.
 
 Supervised and unsupervised empirical learning techniques tackle NLP and CL tasks.
-They employ machine learning without the need to manually encode the requisite 
-knowledge~\cite{soudi2007arabic}. 
+They employ machine learning without the need to manually encode the requisite knowledge~\cite{soudi2007arabic}. 
 Supervised learning techniques require training corpora annotated 
@@
 In this paper, 
-we present \framework, a {\em morphology-based entity and relational entity
-extraction framework for Arabic text }. 
+we present \framework, a morphology-based entity and relational entity
+extraction framework for Arabic text.
 \framework provides a user-friendly interface where the user defines tag types 
-and associates them with \framework formulae that are regular expressions over 
-\framework Boolean formulae. 
-Boolean formulae are terms, negations of terms, and disjunctions of terms. 
+and associates them with
+%\framework formulae that are 
+regular expressions over 
+%\framework
+Boolean formulae.
+A Boolean formula is defined by a term, negation of a term, or disjunction of terms.
 Terms are matches to Arabic morphological features including 
-prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym tags, 
-and semantic categories. 
-We discuss the importance of the morphological features supported in 
-\framework terms in Section~\ref{sec:morph}; in brief, morphological preprocessing
-is key to Arabic NLP.% due to the richness of Arabic morphology.
+prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym tags, and semantic categories.
+We often refer to regular expressions and Boolean formulae as expressions and formulae, respectively, for short.
+%We discuss the importance of the morphological features supported in 
+%\framework terms in Section~\ref{sec:morph}; in brief, morphological preprocessing
+%is key to Arabic NLP.% due to the richness of Arabic morphology.
 
@@
 ~\url{http://www.thedubaimall.com/ar/}}.
+The figure also presents a transliteration and an English translation of the Arabic text.
 The framed words in the text are entities referring to names of people 
@@
 \framework regular expressions support operators such as concatenation, 
-zero or one, zero or more, one or more, up to $M\in\mathbb{N}^+$, and logical 
-conjunction and disjunction. 
+zero or one, zero or more, one or more, up to $M\in\mathbb{N}^+$, and logical conjunction and disjunction.
 An editor allows the user to associate an action with each sub-expression.
@@
 
-\framework takes an Arabic text, a set of user-defined \framework Boolean 
-formulae, and a set of user-defined \framework regular expressions. 
+\framework takes an Arabic text, a set of user-defined Boolean 
+formulae, and a set of user-defined regular expressions.
 \framework computes the morphological solutions of the words in the 
@@
 
-\subsection{Existing annotation and entity extraction tools}
-
-Researchers proposed and evaluated empirical and 
-knowledge-based techniques to extract entities and relational entities
-from text. 
-We briefly review them here and we discuss them and compare to them
-in Section~\ref{sec:related}. 
-The work in~\cite{ekbal2010named} presents a language independent 
-approach for NER extraction using {\em support vector machines}. 
-The work in~\cite{abdelrahman2010integrated} integrates 
-a semi-supervised bootstrapping pattern recognition technique, 
-and a supervised classifier based on {\em conditional random fields}
-to solve NER problems.
-
-Knowledge-based techniques such 
-as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars 
-with morphological stemming to perform NER. 
-%The work in~
-\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
-events, and relations amongst them from Arabic text using a hierarchy of manually built 
-finite state machines driven by morphological features and graph 
-transformation algorithms. 
-Such techniques require advanced linguistic and programming expertise. 
-QARAB is a question answering system that takes 
-Arabic natural language queries and provides short answers~\cite{hammo2002qarab}.
-
-Researchers also proposed systems for automatic IE based 
-on user specifications. 
-CPSL is a common pattern specification language for finite-state 
-grammar~\cite{appelt1998common}. 
-%\framework is an extension to CPSL with action execution and relation construction. 
-%It defines three sections for declaration, rule definition, and macros. 
-The work in \cite{chiticariu2010systemt} presents SystemT, a system based on 
-an algebraic Approach to Declarative information extraction (IE). 
-%It uses a declarative rule language and an optimizer. 
-TEXTMARKER is a rule-based IE system designed to extract structured data from 
-text \cite{atzmueller2008rule}. 
-\cite{urbain2012user} presents a user-driven relational model 
-requiring a user natural language query to extract entities and relational entities.
-
-This work significantly extends the position 
-paper~\cite{JaZaMatar} that allows for manual, and 
-basic morphology annotation.
+This work significantly extends the position paper~\cite{JaZaMatar} that allows for manual, and basic morphology annotation.
 \framework enables a regular user to incrementally create 
@@
 \begin{itemize}
-  \item \framework provides a novel and intuitive visual interface to build Boolean formulae over morphological features, 
-    build regular expressions over the resulting Boolean formulae, and thereafter compute automatic tags.
+  \item \framework provides a novel and intuitive visual interface to build formulae over morphological features, 
+    build regular expressions over the resulting formulae, and thereafter compute automatic tags.
   \item To our knowledge, this morphology-based framework is the first for Arabic entity and relational entity extraction.
@@
 
-\input{directionMBFMRE}
-
-In the rest of this paper,
-Section~\ref{sec:motivation} presents a motivational example. 
+The rest of the paper is structured as follows.
+%In the rest of this paper,
+%Section~\ref{sec:motivation} presents a motivational example. 
 Section~\ref{sec:morph} introduces Arabic morphological analysis 

### 2018-03-16 tex/merf/TALMajor/relatedwork.tex
In progress: merging motivation with section 3

@@
 and extract entities and relations from text.
+
+\begin{comment}
+\subsection{Existing annotation and entity extraction tools}
+
+Researchers proposed and evaluated empirical and 
+knowledge-based techniques to extract entities and relational entities
+from text. 
+We briefly review them here and we discuss them and compare to them
+in Section~\ref{sec:related}. 
+The work in~\cite{ekbal2010named} presents a language independent 
+approach for NER extraction using {\em support vector machines}. 
+The work in~\cite{abdelrahman2010integrated} integrates 
+a semi-supervised bootstrapping pattern recognition technique, 
+and a supervised classifier based on {\em conditional random fields}
+to solve NER problems.
+
+Knowledge-based techniques such 
+as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars 
+with morphological stemming to perform NER. 
+%The work in~
+\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
+events, and relations amongst them from Arabic text using a hierarchy of manually built 
+finite state machines driven by morphological features and graph 
+transformation algorithms. 
+Such techniques require advanced linguistic and programming expertise. 
+QARAB is a question answering system that takes 
+Arabic natural language queries and provides short answers~\cite{hammo2002qarab}.
+
+Researchers also proposed systems for automatic IE based 
+on user specifications. 
+CPSL is a common pattern specification language for finite-state 
+grammar~\cite{appelt1998common}. 
+%\framework is an extension to CPSL with action execution and relation construction. 
+%It defines three sections for declaration, rule definition, and macros. 
+The work in \cite{chiticariu2010systemt} presents SystemT, a system based on 
+an algebraic Approach to Declarative information extraction (IE). 
+%It uses a declarative rule language and an optimizer. 
+TEXTMARKER is a rule-based IE system designed to extract structured data from 
+text \cite{atzmueller2008rule}. 
+\cite{urbain2012user} presents a user-driven relational model 
+requiring a user natural language query to extract entities and relational entities.
+\end{comment}

### 2018-03-16 tex/merf/TALMajor/sarf.tex
In progress: merging motivation with section 3

--- 2018-03-13 tex/merf/TAL/sarf.tex
@@
 Each morphological solution $m\in M(t)$ is a tuple of the form 
-$\langle p,s,x,P,G,C\rangle\in \pp \times  \ss\times \xx\times \PP\times\GG \times \AC$.
+$\langle p^k,s,x^l,P,G,C\rangle\in \pp^k \times  \ss\times \xx^l\times \PP\times\GG \times \AC$ where $k>0$ and $l>0$ indicate possibly multiple prefixes and suffixes, respectively.
 
@@
 The POS and gloss notations follow the Buckwalter notation~\cite{Buckwalter:02}.
-
-\input{MERFOverview}

### 2018-03-16 tex/merf/TALMajor/tool.tex
In progress: merging motivation with section 3

--- 2018-03-13 tex/merf/TAL/tool.tex
@@
+\input{MERFOverview}
+
 The \framework framework is illustrated in the flow diagram of
@@
 
+The motivation for $Syn^k$ is to construct a light Arabic ontology based on the lexicon of Sarf.
 The sets $E, A,$ and $L$ denote all English words, Arabic words, 
 and Arabic lexicon words, respectively.
+Recall that $\GG$ and $\ss$ denote the set of glosses and stems in the morphological analyzer, respectively.
 We have $\GG \subset E$ and $\ss \subset L \subset A$. 
@@
 The example in Figure~\ref{fig:introsynEx} illustrates the computation.
-The word \RL{mA'}is related to \RL{n.d.h}through the gloss intersection \cci{water}. 
-The $Syn^2$ feature relates 
-the words \RL{mA'} and ``\RL{r^s^s}'' since \RL{n.d.h} and \RL{r^s^s} 
-have the gloss intersection \cci{spray}.
+The input Arabic word \RL{mA'}, denoted $w$, shares the gloss \cci{water} with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
+%The input word \RL{mA'} shares the gloss \RL{n.d.h} through the gloss intersection \cci{water}. 
+Next, the stem \RL{r^s^s}, denoted $s_2$, shares the gloss \cci{spray} with $s_1$, i.e. $s_2\in Sy^{1}(s1)\subset Sy^{2}(w)$.
+Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.
 
@@
 predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
-set of morphological features. 
+set of morphological features where each entry is in turn a set of values for that feature.
 Given a word $w$, a morphological feature $A\in{\cal F}$, 
@@
 $k, 1\le k\le 7$, 
-the following are morphology-based atomic terms (MAT). 
+the following are morphology-based atomic terms (MAT), {\em terms} for short.
 \begin{itemize}
-  \item $a(w):= \exists m \in M(w). m=\langle p,s,x,P,G,C\rangle. r \circ CF$
-where $\circ \in {\cal O}$, $r \in \{p,s,x,P,G,C\}$, and $r\in A$.
+  \item $a(w):= \exists m \in M(w). m=\langle p^k,s,x^l,P,G,C\rangle. r \circ CF$
+where $\circ \in {\cal O}$, $r \in \{p^k,s,x^l,P,G,C\}$, and $r\in A$.
 Informally, a solution vector of $w$ exists with
-a feature containing (or exactly matching) $CF$. 
+a feature containing or exactly matching $CF$.
+The predicates $\mathit{isA}$ and $\mathit{contains}$ denote the exact match and contain criteria, respectively.
 \item $a(w) := w \in Syn^k(CF), CF \in \ss$.
-  Informally, this checks $w$ is an extended synonym of a stem $CF$. 
+  Informally, this checks if $w$ is an extended synonym of a stem $CF$. 
 \end{itemize}
@@
 \end{itemize}
-Formula $N$ in Figure~\ref{fig:motiv} checks whether a solution has a category
-feature matching category {\tt Name\_of\_Person}.
-Formula $R$ is the disjunction of MAT terms that check for
+
+Moreover, \framework provides $O$ to be a default Boolean formula that tags all {\em other} words in the text that do not match a user defined formula.
+We also refer to those words as {\em null} words.
+
+\input{directionMBF}
+
+Consider the task we discussed in the introduction (Figure~\ref{fig:intromotiv}) 
+and recall that we are interested in identifying names of people, names of places, relative positions, and numerical terms.
+The table in Figure~\ref{fig:taskMBF} presents the defined formulae.
+The user denotes the ``name of person'' entities with formula $N$ 
+which requires the {\em category} feature in the morphological solution of a word to be {\tt Name\_of \_Person}.
+The entities $n_1$, $n_2$, and $n_3$ are matches of the formula $N$ in the text.
+Similarly, the user specifies formula $P$ to denote ``name of place'' entities. 
+The user specifies formula $R$ to denote ``relative position'' entities, 
+and defines it as a disjunction of MAT terms that check for 
 solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).
+%requires the stem feature of a word to belong to a selected list of stems containing \RL{fy} and \RL{qrb}. a
+Similarly, $U$ denotes numerical terms and is a disjunction of constraints 
+requiring the stem feature to belong to a set of stems such as
+\RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).
 
-A morphology-based regular expression (MRE) is one of the following. 
+%Formula $N$ in Figure~\ref{fig:motiv} checks whether a solution has a category feature matching category {\tt Name\_of\_Person}.
+%Formula $R$ is the disjunction of MAT terms that check for solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).
+
+Next, we define a morphology-based regular expression (MRE) as follows.
 \begin{itemize}
@@
 We denote by $\ldbrack f\rdbrack$ the set of matches of an MRE $f$.
+
+Back to the example in Figure~\ref{fig:intromotiv}.
+We use the formulae defined in Figure~\ref{fig:directionMBF} to construct an MRE such as
+$(P|N)\!+~O?~R~O^\wedge\!2~(P|N|U)+$ where 
+$|,+,?,$ and $^\wedge k$ denote disjunction, one or more, zero or one, and
+up to $k$ matches, respectively.
+The expression specifies a sequence of places or names of persons, 
+optionally followed by a null word, 
+followed by one relative position, followed by up to two possible null words, 
+followed by one or more match of name of place, name of person, or numerical term.
+$O?$ and $O^\wedge 2$ are used in the expression to allow for flexible matches.
+%The user reaches the satisfying matches by experimenting with the visualizer and the expression editor which do not require knowledge and expertise in regular expressions.
+
+\input{directionMRE}
+
+The match trees in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
+The first match tree refers to the text 
+\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection).
+The second match tree refers to the text 
+\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
+The leaf nodes of the trees are entities and the edges and internal nodes are text, 
+morphology-based, 
+and word distance based relational entities.
 

### 2018-03-17 tex/merf/TALMajor/introduction.tex
In progress: New methodology section ...

@@
     and automatically detects relational entities matching the user defined relations. 
-  \item \framework enables the user to associate subexpressions
+  \item \framework enables the user to associate sub-expressions
     with code actions, and executes the code action 
@@
 %in Section~\ref{sec:preliminaries}. 
-Section~\ref{sec:framework} explains the \framework methodology. 
+Section~\ref{sec:framework} presents the \framework components.
 Section~\ref{sec:gui} presents the visual and user friendly interface of \framework. 
+Section~\ref{sec:methodology} explains the methodology used to design \framework.
 Section~\ref{sec:related} presents and discusses related work.

### 2018-03-17 tex/merf/TALMajor/methodology.tex
In progress: New methodology section ...

--- 2018-03-13 tex/merf/TAL/tool.tex
@@
-The \framework framework is illustrated in the flow diagram of
-Figure~\ref{f:overview}.
-The Arabic text and the reference tag chunks are the primary inputs to \framework. 
-Solutions, morphology-based Boolean formulae, tags, 
-morphology-based regular expressions, 
-tag chunks, relation and action definitions, and data structures expressing entities 
-and relations are input and output data to processes. 
-The morphological analyzer (Sarf), $Syn^k$ detector, 
-GUI for Boolean formulae definition, 
-visualization annotator, GUI for regular expression and action definition, 
-Boolean formula simulator, regular expression simulator, 
-relation extraction and action execution, and difference and statistical analyzer 
-are processes.
-
-%The user interacts with \framework~through a user-friendly interfaces to 
-%specify morphology-based Boolean formulae and regular expressions 
-%and associate them with tag types that are in turn associated with
-%visualization legends such as 
-%foreground and background colors and fonts. 
-%The user also associates code actions to parts of the regular expressions.
-%The user can also declare relations between matches using the relation
-%editor.
-
-\subsection{The extended synonymy feature $Syn^k$}
-
-The sets $E, A,$ and $L$ denote all English words, Arabic words, 
-and Arabic lexicon words, respectively. 
-We have $\GG \subset E$ and $\ss \subset L \subset A$. 
-Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
-subsets of related English glosses. %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
-Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
-of relevant Arabic stems. %; e.g. $s_{l} = \gamma(l)\subset 2^{\ss}$.
-
-Given a word $w\in L$, 
-$Sy(w)=\{u\mid u\in \ss \land\exists s\in \gamma(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$
-is the set of Arabic stems 
-directly related to $w$ through the gloss map.
-
-Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
-$Sy^{1}(w) = Sy(w)$ and
-$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
-Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
-
-The example in Figure~\ref{fig:introsynEx} illustrates the computation. 
-The word \RL{mA'}is related to \RL{n.d.h}through the gloss intersection \cci{water}. 
-The $Syn^2$ feature relates 
-the words \RL{mA'} and ``\RL{r^s^s}'' since \RL{n.d.h} and \RL{r^s^s} 
-have the gloss intersection \cci{spray}.
-
-\begin{figure}[tb!]
-\setcode{utf8}
-\begin{center}
-  \resizebox{0.7\columnwidth}{!}{ 
-  	{\relsize{-2} \input{figures/syn2ex.tex}} }
-\setcode{standard}
-\caption{$Syn^2($\RL{mA'}$)$}
-\label{fig:introsynEx}
-\end{center}
-\end{figure}
-
-\subsection{MRE: Morphology-based regular expressions}
-
-Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
-predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
-set of morphological features. 
-Given a word $w$, a morphological feature $A\in{\cal F}$, 
-a user defined constant feature value $CF\in A$, and an integer 
-$k, 1\le k\le 7$, 
-the following are morphology-based atomic terms (MAT). 
-\begin{itemize}
-  \item $a(w):= \exists m \in M(w). m=\langle p,s,x,P,G,C\rangle. r \circ CF$
-where $\circ \in {\cal O}$, $r \in \{p,s,x,P,G,C\}$, and $r\in A$.
-Informally, a solution vector of $w$ exists with
-a feature containing (or exactly matching) $CF$. 
-\item $a(w) := w \in Syn^k(CF), CF \in \ss$.
-  Informally, this checks $w$ is an extended synonym of a stem $CF$. 
-\end{itemize}
-
-A morphology-based Boolean formula (MBF) is of the following form. 
-\begin{itemize}
-  \item $a$ and $\neg a$ are MBF formulae where $a$ is a MAT and $\neg$ is the negation operator. 
-  \item $(f \vee g)$ is an MBF where $f$ and $g$ are MBF formulae, 
-    and $\vee$ is the disjunction (union) operator. 
-\end{itemize}
-Formula $N$ in Figure~\ref{fig:motiv} checks whether a solution has a category
-feature matching category {\tt Name\_of\_Person}.
-Formula $R$ is the disjunction of MAT terms that check for
-solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).
-
-A morphology-based regular expression (MRE) is one of the following. 
-\begin{itemize}
-\item $m$ is an MRE where $m$ is an MBF.
-\item $fg$ is an MRE denoting a concatenation operation
-  where $f$ and $g$ are both MRE expressions. This is 
-  satisfied by a match of $f$ followed by a match of $g$. 
-\item $f*,f+,f$\textasciicircum$x,$ and $f?$ are MRE expressions where $f$ is an MRE,
-  and are satisfied by zero or more matches, one of more matches, up to $x$ matches, 
-  and zero or only one match of $f$, respectively.
-\item $f\& g,$ (conjunction) and $f|g$ (disjunction) are  MRE expressions where 
-  $f$ and $g$ are MRE expressions, and are satisfied by a 
-  a match of both $f$ and $g$, 
-  and a match of either $f$ or $g$, respectively. 
-\end{itemize}
-We denote by $\ldbrack f\rdbrack$ the set of matches of an MRE $f$.
-
-\subsection{User-defined relations and actions}
-
-A user-defined relation $R$ defined by a tuple 
-$\langle e_1,e_2,r\rangle$ where 
-$e_1,e_2,$ and $r$ are identifiers associated with MRE 
-subexpressions in as expression $f$ is a set of labeled binary edges
-where matches of $e_1$ and $e_2$ are the source and destination nodes
-and matches of $r$ are the edge labels.
-We refer to a member of $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ 
-as a user defined relational entity.
-
-\framework allows advanced users to write C++ code snippets 
-to process matches of MRE subexpressions. 
-Users can use these actions to compute statistical features, 
-store intermediate results, or apply intelligent entity inference techniques
-as we show in the numerical extraction example of Section~\ref{sec:sec:number}. 
-\framework provides an API that enrich the actions with detailed access to
-all solution features of an MRE or an MBF match including 
-text, position, length, equivalent numerical value if applicable, 
-and morphological features.
-
-Once \framework computes all match trees, it traverses each match to 
-execute the user defined \cci{pre-match} actions in pre-order manner
-and the \cci{on-match} actions in post-order manner.
-
-\subsection{\framework simulator}
-
 The set of tag types ${\cal T}$ contains tuples of the form $\langle l,f,d\rangle$ 
@@
 
-For each word $t_i\in T, 0\le i < n, n=|T|$ 
+For each word $t_i\in T, 0\le i < M, M=|T|$.
 \framework computes a Boolean value 
@@
 tag type $\mathit{tt}$ is true for $t_i$. 
-
 The MBF evaluation results in a sequence of tag sets 
 $\langle R_0, R_1, \ldots, R_{n-1}\rangle$.
-If a word $t_o$ has no tag type match, 
-its tag set $R_o$ is by default the singleton $O=\{\mathit{NONE}\}$
-and $t_o$ is referred to as a {\em null} word.
+If a word $t_i$ has no tag type match, 
+its tag set $R_i$ is by default the singleton $O=\{\mathit{NONE}\}$.
+%and $t_i$ is referred to as a {\em null} word.
 
@@
 For each MRE, 
-\framework generates its equivalent non-deterministic finite automaton (NFA) 
-in the typical manner~\cite{sipser2006introduction}. 
+\framework generates its equivalent non-deterministic finite automaton (NFA) in the typical manner~\cite{sipser2006introduction}. 
 %Each MRE operation has its equivalent representation in an NFA. 
-We support the upto operation ($f$\textasciicircum$x$), which is not directly 
+We support the upto operation ($f$\^{}$x$), which is not directly 
 supported in ~\cite{sipser2006introduction}, by 
@@
 expanding it into a regular expression form; for example 
-$f$\textasciicircum$3$ is equivalent to $f?|ff|fff$. 
-Consider the example of Figure~\ref{fig:motiv} and the
-MRE $\mathit{dir}= (P|N)+~O?~R~O\wedge 2~(P|N|U)+$. 
+$f$\^{}$3$ is equivalent to $f?|ff|fff$. 
+Consider the example of Figure~\ref{fig:intromotiv} and the
+corresponding expression $(P|N)\!+~O?~R~O^\wedge 2~(P|N|U)+$. 
 Figure~\ref{fig:nfaEx} shows part of the corresponding NFA where
 $q_8, q_9, \dots, q_{13}$ represent NFA states,
-and edges are transitions based labeled with MBF tags such as 
-$P,$ and $N$. Edges labeled with the 
-empty string $\epsilon$ are non-deterministic.
-The expression uses operations $|$, $?$, $+$, and $\wedge$ to 
-relate places {\tt P}, names of persons {\tt N}, 
-relative positions {\tt R}, numerical terms {\tt U}, and other {\tt O}.
+and edges are transitions based on MBF tags such as 
+$P,$ and $N$.
+Edges labeled with the empty string $\epsilon$ are non-deterministic.
+%The expression uses operations $|$, $?$, $+$, and $\wedge$ to relate places {\tt P}, names of persons {\tt N}, relative positions {\tt R}, numerical terms {\tt U}, and other {\tt O}.
 
-\framework simulates the generated NFA over the sequence of tag sets 
-matching the MBF formulae.
+\framework simulates the generated NFA over the sequence of tag sets matching the MBF formulae.
 A simulation match $m$ of an expression $f$ is a tree where the root is the 

### 2018-03-17 tex/merf/TALMajor/results.tex
In progress: New methodology section ...

--- 2018-03-13 tex/merf/TAL/results.tex
@@
 
+\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
+\begin{figure}[tb!]
+\centering
+  \begin{tabular}{p{5.5cm}p{5.5cm}}
+\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=TMB algorithm,commandchars=\\\[\]] 
+cout << \fvtextcolor[red][$s1.text];
+if(isHundred) {
+  if(current != 0) {
+    previous += current;
+  }
+  current = currentH * \fvtextcolor[red][$s1.number];
+  currentH = 0;
+  isHundred = false;
+  isKey = true;
+} else if(current == 0) {
+
+  current = \fvtextcolor[red][$s1.number];
+  isKey = true;
+} else if(!isKey) {
+
+  isKey = true;
+  current = current * \fvtextcolor[red][$s1.number];
+} else {
+  previous += current;
+  current = \fvtextcolor[red][$s1.number];
+}
+\end{Verbatim}
+&
+\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=DT algorithm,commandchars=\\\[\]] 
+if(isHundred) {
+  currentH += \fvtextcolor[red][$s0.number];
+} else if(current == 0) {
+  current = \fvtextcolor[red][$s0.number];
+} else if(isKey) {
+  previous += current;
+  current = \fvtextcolor[red][$s0.number];
+} else {
+  current += \fvtextcolor[red][$s0.number]; }
+isKey = false;
+\end{Verbatim}
+\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=H algorithm,commandchars=\\\[\]] 
+isHundred = true;
+if(current == 0)  {
+  currentH = \fvtextcolor[red][$s2.number];
+} else if(!isKey) {
+  currentH = current * \fvtextcolor[red][$s2.number];
+  current = 0;
+} else {
+  currentH = \fvtextcolor[red][$s2.number];}
+isKey = false;
+\end{Verbatim}
+\\ 
+\end{tabular}
+\caption{Actions for TMB, DT, and H MRE expressions.}
+\label{fig:numnormalgo}
+\end{figure}
+
 \vspace{-1em}
@@
 for the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H}, respectively. 
-The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
-%Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.
+%The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
+Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.
 The actions use \framework API to access features of the matches such 
@@
 
-%\newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
-%\begin{figure}[tb!]
-%\centering
-%  \begin{tabular}{p{5.5cm}p{5.5cm}}
-%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=TMB algorithm,commandchars=\\\[\]] 
-%cout << \fvtextcolor[red][$s1.text];
-%if(isHundred) {
-%  if(current != 0) {
-%    previous += current;
-%  }
-%  current = currentH * \fvtextcolor[red][$s1.number];
-%  currentH = 0;
-%  isHundred = false;
-%  isKey = true;
-%} else if(current == 0) {
-%  
-%  current = \fvtextcolor[red][$s1.number];
-%  isKey = true;
-%} else if(!isKey) {
-%
-%  isKey = true;
-%  current = current * \fvtextcolor[red][$s1.number];
-%} else {
-%  previous += current;
-%  current = \fvtextcolor[red][$s1.number];
-%}
-%\end{Verbatim}
-%&
-%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=DT algorithm,commandchars=\\\[\]] 
-%if(isHundred) {
-%  currentH += \fvtextcolor[red][$s0.number];
-%} else if(current == 0) {
-%  current = \fvtextcolor[red][$s0.number];
-%} else if(isKey) {
-%  previous += current;
-%  current = \fvtextcolor[red][$s0.number];
-%} else {
-%  current += \fvtextcolor[red][$s0.number]; }
-%isKey = false;
-%\end{Verbatim}
-%\begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=H algorithm,commandchars=\\\[\]] 
-%isHundred = true;
-%if(current == 0)  {
-%  currentH = \fvtextcolor[red][$s2.number];
-%} else if(!isKey) {
-%  currentH = current * \fvtextcolor[red][$s2.number];
-%  current = 0;
-%} else {
-%  currentH = \fvtextcolor[red][$s2.number];}
-%isKey = false;
-%\end{Verbatim}
-%\\ 
-%\end{tabular}
-%\caption{Actions for TMB, DT, and H MRE expressions.}
-%\label{fig:numnormalgo}
-%\end{figure}
-
 \vspace{-1em}

### 2018-03-17 tex/merf/TALMajor/tool.tex
In progress: New methodology section ...

@@
 Back to the example in Figure~\ref{fig:intromotiv}.
-We use the formulae defined in Figure~\ref{fig:directionMBF} to construct an MRE such as
-$(P|N)\!+~O?~R~O^\wedge\!2~(P|N|U)+$ where 
+We use the formulae defined in Figure~\ref{fig:taskMBF} to construct an MRE such as
+$(P|N)\!+O?~R~O^\wedge\!2~(P|N|U)+$ where 
 $|,+,?,$ and $^\wedge k$ denote disjunction, one or more, zero or one, and
@@
 $O?$ and $O^\wedge 2$ are used in the expression to allow for flexible matches.
-%The user reaches the satisfying matches by experimenting with the visualizer and the expression editor which do not require knowledge and expertise in regular expressions.
+%Note that the user reaches the satisfying matches by experimenting with the visualizer and the expression editor which do not require knowledge and expertise in regular expressions.
 
@@
 \RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
-The leaf nodes of the trees are entities and the edges and internal nodes are text, 
-morphology-based, 
-and word distance based relational entities.
+%The leaf nodes of the trees are entities and the edges and internal nodes are text, 
+%morphology-based, 
+%and word distance based relational entities. 
+The leaf nodes of the match trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
+For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the sub-expression $(P|N)+$.
 
@@
 
-A user-defined relation $R$ defined by a tuple 
+A relation $R$ is defined by the user as a tuple 
 $\langle e_1,e_2,r\rangle$ where 
-$e_1,e_2,$ and $r$ are identifiers associated with MRE 
-subexpressions in as expression $f$ is a set of labeled binary edges
+$e_1,e_2,$ and $r$ are identifiers associated with 
+sub-expressions of an expression $f$.
+Matches of $R$ are a set of labeled binary edges
 where matches of $e_1$ and $e_2$ are the source and destination nodes
 and matches of $r$ are the edge labels.
-We refer to a member of $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ 
-as a user defined relational entity.
+We denote $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ to be the set of matches of $R$, and refer to them as user defined relational entities.
 
-\framework allows advanced users to write C++ code snippets 
-to process matches of MRE subexpressions. 
+We are interested in constructing the relational entity graph in Figure~\ref{fig:intromotiv}.
+Let $e_1,o_1,r,o_2,$ and $e_2$ be identifiers to the sub-expressions 
+$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$, respectively.
+We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
+The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (Khalifa Tower), \RL{bAlqrb} (next), \RL{mn} (to), and \RL{AltqA.t` Al-'wl} (first intersection).
+Note that there is no match to the optional $O$ formula in match 1.
+Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second match tree are
+\RL{dby mwl} (Dubai Mall), \RL{`l_A} (is located), \RL{mqrbT} (near), \RL{mn h_dA} (this), and \RL{Almbn_A} (building), respectively.
+
+\cci{Relation($e_1$,$e_2$,$r$)} creates the 
+edge labeled \cci{next to} between \cci{Khalifa tower} and 
+\cci{intersection 1} nodes from match 1, and the 
+edge labeled \cci{near} between \cci{Dubai Mall} and \cci{the building} nodes from match 2.
+\cci{Relation($r$,$e_1$,$o_1$)} creates the edge labeled \cci{prep} 
+between \cci{Dubai Mall} and \cci{near} nodes from match 2. 
+\cci{Relation($r$,$e_2$,$o_2$)} creates the edge labeled \cci{from} 
+between \cci{intersection 1} and \cci{next to} nodes in 
+match 1, and the 
+edge labeled \cci{from this} between \cci{near} and \cci{the building} 
+nodes in match 2.
+
+After the relation construction phase, we are interested to relate the discovered entities
+% and relational entities 
+that express the same concept.
+\framework provides the extended synonym feature of second order as a default cross-reference relation ($Syn^2$).
+%\framework provides the \cci{isA} predicate as a default cross-reference relation.
+In Figure~\ref{fig:intromotiv}, triggering this feature creates the edge between the nodes \cci{Khalifa Tower} and \cci{The building}.
+The edge is labeled with \cci{isA}.
+
+Moreover, \framework allows advanced users to write C++ code snippets 
+to process matches of sub-expressions.
+Each sub-expression can be associated with two computational actions: \cci{pre-match} and \cci{on-match}.
 Users can use these actions to compute statistical features, 
-store intermediate results, or apply intelligent entity inference techniques
-as we show in the numerical extraction example of Section~\ref{sec:sec:number}. 
-\framework provides an API that enrich the actions with detailed access to
-all solution features of an MRE or an MBF match including 
-text, position, length, equivalent numerical value if applicable, 
+store intermediate results, 
+or apply intelligent entity inference techniques
+as we show later in the numerical extraction example of Section~\ref{sec:sec:number}.
+\framework provides an API that enriches the actions with detailed access to
+all solution features of an expression or a formula match including 
+text, position, length, equivalent numerical value when applicable, 
 and morphological features.
-
-Once \framework computes all match trees, it traverses each match to 
+Once \framework computes all match trees, it traverses each tree to 
 execute the user defined \cci{pre-match} actions in pre-order manner
@@
 
+\begin{comment}
 \subsection{\framework simulator}
@@
 with \RL{Almbn_A}(the building) from the second match.
+\end{comment}

### 2018-03-17 tex/merf/TALMajor/methodology.tex
intro subsection moved to related work, edited methodology

@@
 We support the upto operation ($f$\^{}$x$), which is not directly 
-supported in ~\cite{sipser2006introduction}, by 
+supported in~\cite{sipser2006introduction}, by 
 %it is equivalent to $f?|ff|\dots|\underbrace{f}_{x \text{ times}}$ which has an NFA mapping. 
@@
 \framework simulates the generated NFA over the sequence of tag sets matching the MBF formulae.
-A simulation match $m$ of an expression $f$ is a tree where the root is the 
-MRE expression, the internal nodes are the MRE and MBF operations, and the 
-leaves are matches of the MAT terms of $f$.
-The leave matches form a vector of tags
-$\langle r_k,r_{k+1},\dots,r_j\rangle$ 
+%A simulation match $m$ of an expression $f$ is a tree where the root is the MRE expression, the internal nodes are the MRE and MBF operations, and the leaves are matches of the MAT terms of $f$.
+A simulation match $m$ of an expression $f$ is a tree where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g.
+Fig.~\ref{fig:taskMRE}.
+The sequence of leaf matches forms a vector of tags $\langle r_k,r_{k+1},\dots,r_j\rangle$ 
 corresponding to the text sequence 
@@
 %the next simulation starts at $R_{m+1}$. 
-If we have more than one match %starting at $R_k$ where $0\le k\le n$, 
+If we have more than one match for an expression, 
+%starting at $R_k$ where $0\le k\le n$, 
 \framework returns the longest.
-%
-Figure~\ref{fig:motiv}(b) shows two match trees of $\mathit{dir}$ extracted 
-from the text of Figure~\ref{fig:intromotiv}.
-\RL{dby} and \RL{mwl} are leaf nodes referring to name of place tags ($P$). 
-The $+$, $|$, and $?$ MRE operations
-are internal nodes.
+%Figure~\ref{fig:taskMRE} shows two match trees of $\mathit{dir}$ extracted from the text of Figure~\ref{fig:intromotiv}.
+%\RL{dby} and \RL{mwl} are leaf nodes referring to name of place tags ($P$).
+%The $+$, $|$, and $?$ MRE operations are internal nodes.
 
@@
 
-\framework computes the relational entities in a user defined 
-relation $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
-$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$ 
-to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
-with the smallest nonzero positive distance between the source and the destination
-where the distance is the number of words between the matches.
+Next, \framework computes the relational entities corresponding to each user defined relation $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
+$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$.
+Note that the number of matches for $e_i$ ($\ldbrack e_i \rdbrack$) depends on the corresponding sub-expression.
+For instance, we could have multiple matches if we associate $e_i$ with the sub-expression $P|N$ while we have $(P|N)+$.
+%to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
+%with the smallest nonzero positive distance between the source and the destination
+%where the distance is the number of words between the matches. 
+Hence, \framework supports the following configurations for $e_i$ and $e_j$ based on the corresponding matches of sub-expressions:
+\begin{itemize}
+\item $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert=1$: Relate $e_i$ match to $e_j$ match with edge labeled by $r$.
+\item $i=j\wedge\left\vert{e_i}\right\vert>1$: Relate sequences of $e_i$ matches with edges labeled by $r$.
+\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert=1$: Relate each $e_i$ match to the single $e_j$ match with an edge labeled by $r$.
+The reverse applies for $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert>1$.
+\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert>1$: Relate each $e_i$ match to each $e_j$ match with an edge labeled by $r$.
+\end{itemize}
 
-%\framework supports the following configurations for the subexpressions of $e_i$ and $e_j$:
-%\begin{itemize}
-%\item $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert=1$: Relate $e_i$ match to $e_j$ match with edge labelled by $r$.
-%\item $i=j\wedge\left\vert{e_i}\right\vert>1$: Relate sequences of $e_i$ matches with edges labelled by $r$.
-%\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert=1$: Relate each $e_i$ match to the single $e_j$ match with an edge labelled by $r$.
-The reverse applies for $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert>1$.
-%\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert>1$: Relate each $e_i$ match to each $e_j$ match with an edge labelled by $r$.
-%\end{itemize}
-%
-In Figure~\ref{fig:motiv}(b), \framework names the subexpressions 
-$(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, 
-as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
-The user defines the semantic relations
-$\langle e_1, e_2, r\rangle$, 
-$\langle r, e_1, o_1\rangle$, and 
-$\langle r,e_2,o_2\rangle$.
+%In Figure~\ref{fig:taskMRE}, \framework names the sub-expressions $(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
+%The user defines the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
 
-The matches of $e_1,~e_2,~o_1,~o_2,$ and $r$ from the second match tree 
-of Figure~\ref{fig:motiv}(b) are \RL{dby mwl}(Dubai Mall), 
-\RL{Almbn_A}(the building), \RL{`l_A}(prep), \RL{mn h_dA}(from this), 
-and \RL{mqrbT}(near), respectively. 
-\framework~constructs the semantic relation matches and builds the lower part of the 
-entity-relation graph shown in Figure~\ref{fig:intromotiv}.
+%The matches of $e_1,~e_2,~o_1,~o_2,$ and $r$ from the second match tree of Figure~\ref{fig:motiv}(b) are \RL{dby mwl}(Dubai Mall), \RL{Almbn_A}(the building), \RL{`l_A}(prep), \RL{mn h_dA}(from this), and \RL{mqrbT}(near), respectively.
+%\framework~constructs the semantic relation matches and builds the lower part of the entity-relation graph shown in Figure~\ref{fig:intromotiv}.
 
-For the first match \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'-wl},
-the matches of $e_1,~e_2,~o_2,$ and $r$ are 
-\RL{brj _hlyfT}(Khalifa tower), \RL{AltqA.t` Al-'-wl}(intersection 1), 
-\RL{mn}(from), and \RL{bAlqrb}(next to), respectively. 
-\framework~doesn't construct the relation $\langle r, e_1, o_1\rangle$ 
-since $o_1$ has no match. 
-Therefore, we get the upper part of the entity-relation graph shown 
-in Figure~\ref{fig:intromotiv}.
+%For the first match \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'-wl}, the matches of $e_1,~e_2,~o_2,$ and $r$ are \RL{brj _hlyfT}(Khalifa tower), \RL{AltqA.t` Al-'-wl}(intersection 1), \RL{mn}(from), and \RL{bAlqrb}(next to), respectively.
+%\framework doesn't construct the relation $\langle r, e_1, o_1\rangle$ since $o_1$ has no match. 
+%Therefore, we get the upper part of the entity-relation graph shown in Figure~\ref{fig:intromotiv}.
 
 After computing the relational entities, 
-\framework~computes a cross-reference relation between the extracted entities
-using a second order synonymy feature ($Syn^2$).
-The \cci{isA} edge in the 
-graph of Figure~\ref{fig:intromotiv} shows the cross-reference relation 
-between \RL{brj _hlyfT}(Khalifa tower) from the first match 
-with \RL{Almbn_A}(the building) from the second match.
+\framework computes cross-reference relations between the extracted entities using a second order synonymy feature ($Syn^2$).
+%The \cci{isA} edge in the graph of Figure~\ref{fig:intromotiv} shows the cross-reference relation between \RL{brj _hlyfT}(Khalifa tower) from the first match with \RL{Almbn_A}(the building) from the second match.

### 2018-03-17 tex/merf/TALMajor/relatedwork.tex
In progress: New methodology section ...
intro subsection moved to related work, edited methodology

@@
-In this section we review the literature on 
-entity and relation IE and on automatic and manual annotation
-techniques and compare to \framework.
+In this section we review the literature on entity and relation IE 
+and on automatic and manual annotation techniques 
+and compare to \framework.
 
@@
 %QARAB\cite{hammo2002qarab}
+
+{\bf Information Extraction.~}
+The common pattern specification language (CPSL) targets system independent IE specifications. 
+CPSL consists of 
+(1) a declaration part specifying names and labels, 
+(2) a rule definition part specifying patterns, regular expressions and associated actions, and 
+(3) macro text substitution part~\cite{appelt1998common}.
+\framework extends CPSL with Arabic morphological features, code actions, and user defined relations.
+
+SystemT~\cite{chiticariu2010systemt}
+aims to overcome the performance and expressivity limitations of CPSL.
+It is based on an algebraic approach to declarative information extraction, 
+uses the declarative annotation query language (AQL), 
+and uses an optimizer to generate high performance execution plans for the AQL rules.
+%The AQL query is translated into an algebraic expression and an optimizer selects an execution plan. 
+%\framework overcomes the limitations described by SystemT. 
+\framework supports multiple tags per word, and supports the MRE conjunction operator which overcomes the overlapping annotation problem discussed in SystemT.
+%es the overfor each word and this overcomes the lossy sequencing caused 
+%by random annotation dropping. 
+%\framework introduces the conjunction operation in the MRE to overcome the expressivity limitations of expressing rules with overlapping annotations.
+
+TEXTMARKER is a semi-automatic rule-based IE system 
+for structured data acquisition\cite{atzmueller2008rule}.
+%TEXTMARKER requires user defined rules that consider features of the text 
+%However, constructing the rules requires manual work of a domain specialist. 
+Both TEXTMARKER and \framework provide the user with GUI editor and result visualizer.
+%\framework also differs in that it provides graph visualization for regular expression and user defisemantic relation matches.
+
+The work in~\cite{urbain2012user} presents a user 
+driven relational model and targets entity and relation extraction. 
+The user enters a natural language query, and uses the OpenNLP toolkit to 
+extract tags and relations from the query. 
+Then, it extends the extracted entities into a query model,
+and uses the query model to extract matches from the document. 
+Similar to \framework, the system constructs entities and relations.
+
+QARAB is an Arabic question answering system that 
+takes an Arabic natural language query and provides short answers for it~\cite{hammo2002qarab}. 
+QARAB uses traditional information retrieval techniques and an outdated Arabic NLP 
+analyzer that computes limited features of Arabic words compared 
+to the morphological analysis of \framework. 
+QARAB then classifies the query against a set of known question types.
 
@@
 
-{\bf Information Extraction.~}
-The common pattern specification language (CPSL) targets system independent 
-IE specifications. 
-CPSL consists of 
-(1) a declaration part specifying names and labels, 
-(2) a rule definition part specifying patterns, regular expressions and associated actions, and 
-(3) and macro text substitution part~\cite{appelt1998common}.
-\framework extends CPSL with Arabic morphological features, code actions, and user defined relations.
-
-SystemT~\cite{chiticariu2010systemt}
-aims to overcome the performance and expressivity limitations of CPSL.
-It is based on an algebraic approach to declarative information extraction,
-uses the declarative annotation query language (AQL), and uses
-an optimizer to generate high performance execution plans for the AQL rules. 
-%The AQL query is translated into an algebraic expression and an optimizer selects an execution plan. 
-%\framework overcomes the limitations described by SystemT. 
-\framework supports multiple tags per word, and supports the MRE conjunction operator
-which allows for overcoming the overlapping annotation problem. 
-%es the overfor each word and this overcomes the lossy sequencing caused 
-%by random annotation dropping. 
-%\framework introduces the conjunction operation in the MRE to overcome the expressivity limitations of expressing rules with overlapping annotations.
-
-TEXTMARKER is a semi-automatic rule-based IE system 
-for structured data acquisition\cite{atzmueller2008rule}.
-%TEXTMARKER requires user defined rules that consider features of the text 
-%However, constructing the rules requires manual work of a domain specialist. 
-Both TEXTMARKER and \framework provide the user with GUI editor and result visualizer. 
-%\framework also differs in that it provides graph visualization for regular expression and user defisemantic relation matches.
-
-The work in~\cite{urbain2012user} presents a user 
-driven relational model and targets entity and relation extraction. 
-The user enters a natural language query, and uses the OpenNLP toolkit to 
-extract tags and relations from the query. 
-Then it extends the extracted entities into a query model,
-and uses the query model to extract matches from the document. 
-Similar to \framework, the system constructs entities and relations.
-
-QARAB is an Arabic question answering system that 
-takes an Arabic natural language query and provides short answers for it~\cite{hammo2002qarab}. 
-QARAB uses traditional information retrieval techniques and an outdated Arabic NLP 
-analyzer that computes limited features of Arabic words compared 
-to the morphological analysis of \framework. 
-QARAB then classifies the query against a set of known question types.
-
 Table~\ref{tab:iecomp} summarizes the comparison between \framework and
@@
 text processing tasks using active learning and semi-supervised learning~\cite{settles2011closing}. 
-The classifier is interactive as it queries the annotator 
-for entity detetction correction and annotation correction. 
+%The classifier is interactive as it queries the annotator for entity detection correction and annotation correction.
 \framework doesn't support classification tasks. 
 However, \framework provides an interactive GUI where the user can edit MBF and MRE tags.
-This interactive environment contributes to the regular expression 
-extraction and semantic relation construction which increases the overall accuracy.
+This interactive environment contributes to the regular expression extraction 
+and semantic relation construction which increases the overall accuracy.
+
+Another track in the literature targets specific tasks such as NER using statistical machine learning techniques \cite{ekbal2010named,abdelrahman2010integrated}.
+%The work in~\cite{ekbal2010named} presents a language independent approach for NER extraction using {\em support vector machines}. 
+%The work in~\cite{abdelrahman2010integrated} integrates a semi-supervised bootstrapping pattern recognition technique, and a supervised classifier based on {\em conditional random fields} to solve NER problems.
+Knowledge-based techniques such 
+as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars 
+with morphological stemming to perform NER. 
+%The work in~
+\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities and events, and relations amongst them from Arabic text using a hierarchy of manually built 
+finite state machines driven by morphological features 
+and graph transformation algorithms. 
+Such techniques require advanced linguistic and programming expertise.
 
@@
 %inspired by psycholinguistic theories 
-and that relates words based on their semantic values and their functional 
+and relates words based on their semantic values and their functional 
 categories: nouns, verbs, adjectives, adverbs, and function words~\cite{intro}.
@@
 {\bf Annotation tools.~}
-An overview of annotation tools and their Arabic-English word alignment issues 
-concludes with a set of rules and guidelines needed in an Arabic annotation alignment tool
-~\cite{kholidy2010towards}. 
-The work in \cite{dukes2011supervised} presents a collaborative effort towards morphological 
-and syntactic annotation of the Quran. 
-\cite{dorr2010interlingual} presents a framework for interlingual 
-annotation of parallel text corpora with multi-level representations. 
-\cite{kulick2010consistent} presents the integration of the 
-Standard Arabic Morphological Analyzer (SAMA) into the workflow of the Arabic Treebank.
-
 MMAX2 is a manual multi-level linguistic annotation tool with an XML 
 based data model~\cite{mmax2}.
-It enables the user to create, browse, visualize, and query annotations
-and may be able to resolve coreference tags. 
+%It enables the user to create, browse, visualize, and query annotations and may be able to resolve coreference tags. 
 BRAT~\cite{brat} and WordFreak~\cite{wordfreak} are manual 
 multi-lingual user friendly web-based annotators that allow the construction 
-of entity and relation annotation corpora~\cite{brat}. 
-They can be extended through plug-ins to enable automatic annotators 
-and customized annotation visualization specifications.
+of entity and relation annotation corpora.
+%They can be extended through plug-ins to enable automatic annotators and customized annotation visualization specifications.
 Knowtator~\cite{ogren2006knowtator} is a general purpose incremental text annotation tool 
 implemented as a Prot\'eg\'e~\cite{gennari2003evolution} plug-in. 
-Prot\'eg\'e is an open-source platform with a suite of tools to construct domain 
-models and knowledge-based applications with ontologies. 
+Prot\'eg\'e is an open-source platform with a suite of tools to construct domain models and knowledge-based applications with ontologies. 
 However, it doesn't support the Arabic language.
@@
 
+\cite{kholidy2010towards} present an overview of annotation tools 
+%and their Arabic-English word alignment issues 
+and concludes with a set of rules and guidelines needed in an Arabic annotation alignment tool.
+The work in \cite{dukes2011supervised} presents a collaborative effort towards morphological and syntactic annotation of the Quran.
+\cite{dorr2010interlingual} presents a framework for inter-lingual 
+annotation of parallel text corpora with multi-level representations.
+\cite{kulick2010consistent} presents the integration of the 
+Standard Arabic Morphological Analyzer (SAMA) into the workflow of the Arabic Treebank.
+
 The work in \cite{smrz2004morphotrees} presents %the extension of TrEd, 
 a customizable general purpose tree editor, with the Arabic MorphoTrees annotations. 
-The MorphoTrees present the morphological analyses in a hierarchical organization 
-based on common features.
+The MorphoTrees present the morphological analyses in a hierarchical organization based on common features.
 
 Task specific annotation tools such as~\cite{alrahabi2006semantic}
-uses enunciation semantic maps to automatically annotate 
+use enunciation semantic maps to automatically annotate 
 directly reported Arabic and French speech.
@@
 morphology-based features as atomic terms.
-Fassieh is a commercial Arabic text annotation tool that enables the production of large 
-Arabic text corpora~\cite{attia2009fassieh}. 
+Fassieh is a commercial Arabic text annotation tool that enables the production of large Arabic text corpora~\cite{attia2009fassieh}.
 The tool supports Arabic text factorization including morphological analysis, POS tagging, 
@@
 Fassieh is not directly accessible to the research community and requires commercial licensing.
-\framework is open source and differs in that it allows the user to build tag types 
-and extract entities and relations from text.
+\framework is open source and differs in that it allows the user to build tag types and extract entities and relations from text.
 
@@
 Such techniques require advanced linguistic and programming expertise.
+
 QARAB is a question answering system that takes 

### 2018-03-17 tex/merf/TALMajor/tool.tex
intro subsection moved to related work, edited methodology

@@
 
-A relation $R$ is defined by the user as a tuple 
+A relation is defined by the user as a tuple 
 $\langle e_1,e_2,r\rangle$ where 
@@
 sub-expressions of an expression $f$.
-Matches of $R$ are a set of labeled binary edges
+Matches of the relation are a set of labeled binary edges
 where matches of $e_1$ and $e_2$ are the source and destination nodes
 and matches of $r$ are the edge labels.
-We denote $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ to be the set of matches of $R$, and refer to them as user defined relational entities.
+We denote $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$ to be the set of matches of the corresponding relation, 
+and we refer to them as relational entities.
 
@@
 $(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$, respectively.
-We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
+%We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
+We define the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
 The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (Khalifa Tower), \RL{bAlqrb} (next), \RL{mn} (to), and \RL{AltqA.t` Al-'wl} (first intersection).
@@
 
-\cci{Relation($e_1$,$e_2$,$r$)} creates the 
+Relation $\langle e_1,e_2,r\rangle$ creates the 
 edge labeled \cci{next to} between \cci{Khalifa tower} and 
@@
 edge labeled \cci{near} between \cci{Dubai Mall} and \cci{the building} nodes from match 2.
-\cci{Relation($r$,$e_1$,$o_1$)} creates the edge labeled \cci{prep} 
+Relation $\langle r,e_1,o_1\rangle$ creates the edge labeled \cci{prep} 
 between \cci{Dubai Mall} and \cci{near} nodes from match 2. 
-\cci{Relation($r$,$e_2$,$o_2$)} creates the edge labeled \cci{from} 
+Relation $\langle r,e_2,o_2\rangle$ creates the edge labeled \cci{from} 
 between \cci{intersection 1} and \cci{next to} nodes in 

### 2018-03-18 tex/merf/TALMajor/relatedwork.tex
updated related work

@@
 {\bf Information Extraction.~}
-The common pattern specification language (CPSL) targets system independent IE specifications. 
-CPSL consists of 
-(1) a declaration part specifying names and labels, 
-(2) a rule definition part specifying patterns, regular expressions and associated actions, and 
-(3) macro text substitution part~\cite{appelt1998common}.
+The common pattern specification language (CPSL) targets system independent IE specifications~\cite{appelt1998common}. 
+%CPSL consists of 
+%(1) a declaration part specifying names and labels, 
+%(2) a rule definition part specifying patterns, regular expressions and associated actions, and 
+%(3) macro text substitution part~\cite{appelt1998common}.
 \framework extends CPSL with Arabic morphological features, code actions, and user defined relations.
-
-SystemT~\cite{chiticariu2010systemt}
-aims to overcome the performance and expressivity limitations of CPSL.
+SystemT~\cite{chiticariu2010systemt} aims to overcome the 
+%performance and expressivity 
+limitations of CPSL. 
 It is based on an algebraic approach to declarative information extraction, 
@@
 extract tags and relations from the query. 
-Then, it extends the extracted entities into a query model,
-and uses the query model to extract matches from the document. 
+%Then, it extends the extracted entities into a query model, 
+%and uses the query model to extract matches from the document. 
 Similar to \framework, the system constructs entities and relations.
@@
 QARAB uses traditional information retrieval techniques and an outdated Arabic NLP 
-analyzer that computes limited features of Arabic words compared 
+analyzer with limited features of Arabic words compared 
 to the morphological analysis of \framework. 
-QARAB then classifies the query against a set of known question types.
+%QARAB then classifies the query against a set of known question types.
 
@@
 
-DUALIST is an annotation system for quickly building classifiers for 
-text processing tasks using active learning and semi-supervised learning~\cite{settles2011closing}. 
+DUALIST is an annotation system for building classifiers for 
+text processing tasks using machine learning techniques 
+%active learning and semi-supervised learning
+\cite{settles2011closing}. 
 %The classifier is interactive as it queries the annotator for entity detection correction and annotation correction.
@@
 
-Another track in the literature targets specific tasks such as NER using statistical machine learning techniques \cite{ekbal2010named,abdelrahman2010integrated}.
+Another track in the literature targets specific tasks such as NER using statistical machine learning techniques \cite{benajiba2007anersys,benajiba2008arabic,ekbal2010named,abdelrahman2010integrated}.
 %The work in~\cite{ekbal2010named} presents a language independent approach for NER extraction using {\em support vector machines}. 
@@
 Knowledge-based techniques such 
-as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars 
-with morphological stemming to perform NER. 
+as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars with morphological stemming.% to perform NER. 
 %The work in~
-\cite{ZaMaHaCicling2012Entity} presents a method for extracting entities and events, and relations amongst them from Arabic text using a hierarchy of manually built 
-finite state machines driven by morphological features 
+\cite{ZaMaHaCicling2012Entity} present a method for extracting entities and events, and relations amongst them from Arabic text using a hierarchy of manually built 
+finite state machines driven by morphological features, 
 and graph transformation algorithms. 

### 2018-03-18 tex/merf/TALMajor/tool.tex
added availability

@@
 
+\subsection{Framework Availability}
+\framework is an open source tool present on google code under atmine repository (\url{https://code.google.com/p/atmine/}).
+In addition to \framework, the repository contains entity extraction tasks developed by our research group.
+People are welcome to download and use the tools.
+We appreciate any feedback that we get and we try to improve the tools accordingly.
+
 \begin{comment}

### 2018-04-12 tex/cicling.tex
changes to answer, colors, and some reviewers addressed

@@
 %\title{Instructions for NAACL HLT 2010 Proceedings\Thanks{This...}}
-\title{Detection of Arabic Entity Graphs using Morphology, Finite State Machines, and Graph Transformations}
-
-%\author{ Jad Makhlouta \\\And
-%Hamza Harkous \\
-%  American University of Beirut \\
-%  {\tt \{jem04, hhh20, fz11\}@aub.edu.lb } \\\And 
-%  Fadi Zaraket 
-%}
+\title{Arabic Entity Graph Extraction using Morphology, Finite State Machines, and Graph Transformations}
+
+\author{ Jad Makhlouta\inst{1} \and
+    Fadi Zaraket\inst{1} \and 
+Hamza Harkous\inst{1} }
+\institute{ American University of Beirut \\
+  \email{ \{jem04, fz11, hhh20\}@aub.edu.lb } }
 \date{}
@@
 and graph transformation algorithms. 
-We evaluated our method on three natural language processing 
+We evaluated our method on two natural language processing 
 applications.
@@
 narrator relations from 
-several corpora of Islamic narration books (hadith). 
+several corpora of Islamic narration books. 
 We automated the extraction of genealogical family trees 
 from Biblical texts. 
-Finally, we automated locating and extracting
-individual biographies from historical biography books.
+%Finally, we automated locating and extracting
+%individual biographies from historical biography books.
 In all applications, our method reports high precision and recall
@@
 statistical analysis boosted with Arabic-English parallel corpora~\cite{BenajibaZDR10}, 
-statistical analysis boosted with task specific gazetteers~\cite{ANERSys:07,BenajibaDR08},
+statistical analysis boosted with task specific gazetteers~\cite{Benajiba:07,Benajiba:08},
 and pattern matching powered with morphological stemming~\cite{jumaily2011,TAGARAB98}
@@
 An {\em entity graph} is a graph where NEs are nodes and relational entities (RE) are edges.
-For example, consider the tasks of extracting a time table
-from a document describing the events of a conference, 
+For example, consider the tasks of 
 extracting genealogical family trees based on person names 
-from biblical text,  and
-extracting routes and maps from travel itineraries.
-Also consider applications with cross-document NER
-such as extracting narrator authentication graphs based on narration 
-documents that include narrator sequences, and biography documents 
-that include credibility information about the narrators. 
+from biblical text, 
+extracting routes and maps from travel itineraries,
+and 
+extracting narrator graphs based on narration 
+documents that include narrator sequences.
+%, and biography documents 
+%that include credibility information about the narrators. 
 We hypothesize that NEs of the same kind happen 
@@
 Morphological analysis helps reduce ambiguities and resolves lexical items with their morphological variations.
-For each application, we augment the lexicon of an inhouse morphological 
-analyzer with lexical elements and tags that describe categories relevant to the application. 
+For each application, we augment the lexicon of Sarf~\cite{ATMine09}, an inhouse morphological analyzer, 
+with lexical elements and tags that describe categories relevant to the application. 
 We also build finite state transducers (FST) that take 
@@
 
-We evaluated our method and used it to solve three 
+We evaluated our method and used it to solve two
 NLP applications including extracting genealogical family
-trees from Biblical text, extracting a graph of 
-narrators from hadith books, and extracting biographies from biography books.
+trees from Biblical text, and extracting a graph of 
+narrators from hadith books.%, and extracting biographies from biography books.
 Our results show high precision and recall metrics and report
@@
 5 and 6\%, respectively in the genealogy application.
-\item We perform Arabic cross-document named entity and 
+\item We perform Arabic named entity and 
 relational entity extraction using graph algorithms. 
-The use of a graph generated from the hadith application 
-improved precision in the biography application from 40\% to 93\%.
+%The use of a graph generated from the hadith application 
+%improved precision in the biography application from 40\% to 93\%.
 %\item We present a novel equality metric to compare Arabic 
@@
 are person names.
-We use an inhouse morphological analyzer augmented with 
+We use Sarf, an inhouse morphological analyzer, augmented with 
 a list of lexicon elements relevant to the application.
-We also augment the analyzer with
+We also augment Sarf with 
 a set of tags that classify the 
@@
 
-The method passes the input text string to the morphological 
-analyzer. The analyzer processes the input text and whenever
+The method passes the input text string to Sarf.
+Sarf processes the input text and whenever
 it identifies a morpheme, it calls the application morpheme
@@
 
-Finally, the method uses a morphological and structural distance
+Finally, the method uses a distance
 function to compute equality between the named entities and
@@
 
-Due to religious and political reasons, writing the traditions was forbidden 
-until the days of the Umayad Calif,
-\novocalize
-\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after the prophet. 
-Many inconsistencies were introduced to the literature which necessitates 
-a thorough authentication study before the use of a tradition in jurisprudence.
-\vocalize
-%An Islamic hadith scholar studies the authenticity of 
-%the \noArRL{sanad} of a 
-%set of related traditions before using these traditions 
-%jurisprudence. 
-%While different Islamic schools of thought differ on 
-%how to interpret the content, they almost all agree
-All Islamic schools of thought agree
-that scholars can not use a tradition whose sanad lacks authenticity for jurisprudence.
-
-The authenticity of a \noArRL{.hady_t} depends on 
-the credibility of the narrators as reported in 
-separate biography books. 
-The study of \noArRL{.hady_t} authentication is 
-currently manual and error prone due to the huge number
-of existing traditions and tradition books. 
+%Due to religious and political reasons, 
+%writing the traditions was forbidden 
+%until the days of the Umayad Calif,
+%\novocalize
+%\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after the prophet. 
+%Many inconsistencies were introduced to the literature which necessitates 
+%a thorough authentication study before the use of a tradition in jurisprudence.
+%\vocalize
+%%An Islamic hadith scholar studies the authenticity of 
+%%the \noArRL{sanad} of a 
+%%set of related traditions before using these traditions 
+%%jurisprudence. 
+%%While different Islamic schools of thought differ on 
+%%how to interpret the content, they almost all agree
+%All Islamic schools of thought agree
+%that scholars can not use a tradition whose sanad lacks authenticity for jurisprudence.
+%
+%The authenticity of a \noArRL{.hady_t} depends on 
+%the credibility of the narrators as reported in 
+%separate biography books. 
+%The study of \noArRL{.hady_t} authentication is 
+%currently manual and error prone due to the huge number
+%of existing traditions and tradition books. 
 %Hadithopedia~\cite{Hadithopaedia:08}
@@
 Our target is to automate the extraction of 
-narrator graphs from one or more narration books
+narrator graphs,
 similar to the directed acyclic graph (DAG) 
-in Figure~\ref{f:hadith}(b).
+in Figure~\ref{f:hadith}(b),
+from one or more narration books
 %We automatically extracted the POR from the three books of 
@@
 
-\subsection{Narrator distance metric}
+%\subsection{Narrator distance metric}
 %MultiKey Hashing
@@
 % biography lookup
-Two narrator names may refer to the same person but may 
-differ in text and in structure. 
-For example, the names \noTrutfRL{عبد الله بن عمر}, 
-\noTrutfRL{ابن عمر}, and \noTrutfRL{ابن عمر رضي الله عنه}
-refer to the same person. 
-Our metric uses the \noTrutfRL{ابن} (son) name connector and 
-its morphological equivalents to order the person names
-that appear within a narrator entity. 
-The metric compares the name hierarchy and if the hierarchy 
-does not match, as in the case of 
-\noTrutfRL{عبيد الله بن عبد الله بن عمر}  and his parent 
-\noTrutfRL{عبد الله بن عمر},
-the metric returns \cci{false}.
-The metric also compares name qualifiers and if two name 
-qualifiers of the same morphological POS tag,  
-such as the place possessive qualifiers 
-\utfRL{العراقي} (the Iraqi) and \utfRL{المصري} (the Egyptian) 
-conflict, the metric returns \cci{false}.
-The metric then compares the morphological stems of the person 
-names that are in order, ignores the ones that are not in order,
-and computes the fraction of the matching names.
-
-Both our metric and the Levenshtein distance metric 
-used in~\cite{Azmi-2010}
-achieved 100\% precision on an arbitrarily selected set that 
-included 208 narrator pairs with 102 equal narrators.
-Our metric achieved 95\% recall versus 72\% recall for the 
-Levenshtein distance metric. 
-Which means that both metrics do not announce names equal if they
-are not, but our metric is better at detecting equal names.
-We believe that our metric is adequate to compare 
-complex Arabic names including contemporary ones in general.
-
-\section{The biography application}
-Biography books include one or more short biography 
-for each hadith narrator. 
-The short biography may contain information 
-about the dates and locations of birth and death, 
-the professors and the students of the narrator,
-and one or more evaluations of the credibility of the narrator.
-
-We use a similar approach to the hadith application
-to locate narrator names within the biography books. 
-A biography text may contain the name of the narrator described
-in the biography as well as the names of other narrators who happen
-to be the professors, students, or the persons who evaluated the 
-narrator.
-Our method uses the narrator graph 
-extracted from several hadith books 
-as described earlier to segment the biographies
-and detect the owner of each biography.
-
-Once our method detects a narrator, it colors the node in  the 
-narrator graph corresponding to it. 
-The professors and students of the narrator
-are the neighbors of the 
-node corresponding to the narrator in the graph. 
-Consequently, whenever a detected narrator entity 
-moves away from the 
-colored local cluster in the narrator graph, 
-we change the color and infer that we moved to a separate
-biography.
-The owner of each biography is the narrator at the center of the
-colored graph.
-
-The naive implementation of this approach is computationally
-expensive as it requires searching for each discovered narrator
-in the narrator graph. 
-We implemented a multi-hash narrator technique to 
-improve narrator localization in the graph and in the biography 
-books where we associate
-each narrator with all narrator names that can be generated 
-from its structure and compute a hash key and a relevance score 
-for that association. 
-We omit the discussion for brevity and we will report on that 
-elsewhere.
-
-Scholars can use the located biographies to annotate
-each narrator with corresponding biographies thus automating
-the task of hadith authentication checks. 
-They can also annotate the narrator graph narrator graph with the locations and 
-the lifetime of the narrators and 
-check for time and location overlap that are necessary conditions
-for the consistency of two 
-neighboring narrators. 
-Scholars can also use the method to check the correspondence
-of different biography and hadith books.
-
-We believe that this is the first work that applies
-{\em Arabic cross-document NLP}, and we believe
-that this approach can be extended to several applications
-such as the cross analysis of (1) investigation versus 
-interrogation reports, and (2) insurance claims versus 
-accident police reports to find inconsistencies.
+%Two narrator names may refer to the same person but may 
+%differ in text and in structure. 
+%For example, the names \noTrutfRL{عبد الله بن عمر}, 
+%\noTrutfRL{ابن عمر}, and \noTrutfRL{ابن عمر رضي الله عنه}
+%refer to the same person. 
+%Our metric uses the \noTrutfRL{ابن} (son) name connector and 
+%its morphological equivalents to order the person names
+%that appear within a narrator entity. 
+%The metric compares the name hierarchy and if the hierarchy 
+%does not match, as in the case of 
+%\noTrutfRL{عبيد الله بن عبد الله بن عمر}  and his parent 
+%\noTrutfRL{عبد الله بن عمر},
+%the metric returns \cci{false}.
+%The metric also compares name qualifiers and if two name 
+%qualifiers of the same morphological POS tag,  
+%such as the place possessive qualifiers 
+%\utfRL{العراقي} (the Iraqi) and \utfRL{المصري} (the Egyptian) 
+%conflict, the metric returns \cci{false}.
+%The metric then compares the morphological stems of the person 
+%names that are in order, ignores the ones that are not in order,
+%and computes the fraction of the matching names.
+%
+%Both our metric and the Levenshtein distance metric 
+%used in~\cite{Azmi-2010}
+%achieved 100\% precision on an arbitrarily selected set that 
+%included 208 narrator pairs with 102 equal narrators.
+%Our metric achieved 95\% recall versus 72\% recall for the 
+%Levenshtein distance metric. 
+%Which means that both metrics do not announce names equal if they
+%are not, but our metric is better at detecting equal names.
+%We believe that our metric is adequate to compare 
+%complex Arabic names including contemporary ones in general.
+
+%\section{The biography application}
+%Biography books include one or more short biography 
+%for each hadith narrator. 
+%The short biography may contain information 
+%about the dates and locations of birth and death, 
+%the professors and the students of the narrator,
+%and one or more evaluations of the credibility of the narrator. 
+%
+%We use a similar approach to the hadith application
+%to locate narrator names within the biography books. 
+%A biography text may contain the name of the narrator described
+%in the biography as well as the names of other narrators who happen
+%to be the professors, students, or the persons who evaluated the 
+%narrator.
+%Our method uses the narrator graph 
+%extracted from several hadith books 
+%as described earlier to segment the biographies
+%and detect the owner of each biography. 
+%
+%Once our method detects a narrator, it colors the node in  the 
+%narrator graph corresponding to it. 
+%The professors and students of the narrator
+%are the neighbors of the 
+%node corresponding to the narrator in the graph. 
+%Consequently, whenever a detected narrator entity 
+%moves away from the 
+%colored local cluster in the narrator graph, 
+%we change the color and infer that we moved to a separate
+%biography.
+%The owner of each biography is the narrator at the center of the
+%colored graph.
+%
+%The naive implementation of this approach is computationally
+%expensive as it requires searching for each discovered narrator
+%in the narrator graph. 
+%We implemented a multi-hash narrator technique to 
+%improve narrator localization in the graph and in the biography 
+%books where we associate
+%each narrator with all narrator names that can be generated 
+%from its structure and compute a hash key and a relevance score 
+%for that association. 
+%We omit the discussion for brevity and we will report on that 
+%elsewhere. 
+%
+%Scholars can use the located biographies to annotate
+%each narrator with corresponding biographies thus automating
+%the task of hadith authentication checks. 
+%They can also annotate the narrator graph narrator graph with the locations and 
+%the lifetime of the narrators and 
+%check for time and location overlap that are necessary conditions
+%for the consistency of two 
+%neighboring narrators. 
+%Scholars can also use the method to check the correspondence
+%of different biography and hadith books. 
+%
+%We believe that this is the first work that applies
+%{\em Arabic cross-document NLP}, and we believe
+%that this approach can be extended to several applications
+%such as the cross analysis of (1) investigation versus 
+%interrogation reports, and (2) insurance claims versus 
+%accident police reports to find inconsistencies.
 
@@
 To perform this task, our method first passes the text to
-the morphological analyzer that reports detected morphemes 
+Sarf, the morphological analyzer, that reports detected morphemes 
 with their corresponding tags.
 
-We augmented the tag set of our morphological analyzer with 
+We augmented the tag set of Sarf with 
 the following categories relevant to the genealogical 
@@
 task specific corpora and gazetteers and ignores 
-POS tags~\cite{ANERSys:07,BenajibaDR08}. 
-We are similar to ANERSys in that we augment the lexicon of our
-morphological analyzer with application relevant lexicon items
-and tags. 
+POS tags~\cite{Benajiba:07,Benajiba:08}. 
+We are similar to ANERSys in that we augment the lexicon Sarf
+with application relevant lexicon items and tags. 
 We differ in that they use statistical analysis models such as
@@
 hadith~\cite{IbnHanbal,AlTousi,AlKulayni},
-the Genesis~\cite{},
-and one book of biographies with several volumes~\cite{khoei}. 
+and the book of Genesis~\cite{}.
+%and one book of biographies with several volumes~\cite{khoei}. 
 We define our metrics, present our evaluation reference with interannotation agreement results, 
 and then present the results for the hadith
-biography, and genealogical list applications.
+and genealogical list applications.
 
@@
 
-{\bf Biography narrator detection.}~~
-Table~\ref{t:narratorDifference} shows the accuracy results for the narrator detection and the biography boundary detection
-for the Khoei biography book.
-The first row shows the results computed by considering the biography book alone without the help of the narrator graph
-extracted from the hadith books. 
-Without the narrator graph, our method detected 94\% of the narrators mentioned in the book with 81\% precision. 
-The method without the narrator graph, detected correctly (94\% precision) the boundaries of the biographies.
-
-With the help of the narrator graph, our method detected 65\% of the narrator names in the biography book. 
-The rest of the names were simply not found in the hadith books that we used to generate the narrator graph. 
-With the use of the narrator graph and the coloring algorithm, our method detected the boundaries of the biographies
-with 93\% precision. 
-We believe that an exhaustive graph that includes narrator names from more books of hadith will improve the 65\% narrator 
-detection recall metric.
-In all cases, narrator names that exist in the biography and not in the hadith books are not of interest to the narrator
-graph annotation problem.
-
-%The numbers in the right hand side of Table~\ref{t:narratorDifference}  may explain why 
-Narrator detection in biography books is harder than in hadith books. 
-For example, the ratio of narrator names to the rest of the text is 22\% in Musnad while it is 49\% in Khoei,
-and the average length of a narrator sequence in Musnad is 2.6 narrators while it is 3.5 narrators in Khoei.
-With the use of the narrator graph extracted from the hadith books, our method was able to improve biography boundary detection
-from 41\% to 93\%.
-This is evidence that cross-document entity detection helps to accurately find named entities in harder NLP tasks.
-
-\begin{table*}[bt]                                     
-\centering                                            
-\caption{Accuracy results for the biography application.}
-\resizebox{.7\columnwidth}{!}{
-
-%\begin{tabular}{cp{0.3cm}c}
-
-	\begin{tabular}{cp{.1cm}cccp{.1cm}ccc}
-	          & & \multicolumn{3}{c}{narrator detection} & &  \multicolumn{3}{c}{biography boundary} \\
-	 & & recall & precision & F-score & & recall & precision & F-score \\ 
- \cline{3-5} \cline{7-9}
-%hadith	 & sanad & & 0.923 & 0.999 & 0.960 & & 0.996 & 0.999 & 0.998  \\
-%(musnad) & all & & 0.954 & 0.961 & 0.958 & & 0.996 & 0.987 & 0.992 \\ 
-%	\hline \hline
-w/o graph & & 0.943 & 0.812 & 0.873 & & 0.407 & 0.945 & 0.569 \\
-with graph & & 0.654 & 0.909 & 0.760 & & 0.933 & 0.960 & 0.946 \\ % found in narrator graph
-	 \end{tabular}   
-%&  &
-%	\begin{tabular}{ccc}
-%	& names    & nar. \\
-%    & per text & elements \\ \cline{2-3}
-%	\multirow{2}{*}{0.217} &\multirow{2}{*}{ 2.64}\\ \\
-%	\hline \hline
-%	\multirow{2}{*}{0.489} & \multirow{2}{*}{3.54}\\
-%    musnad & 0.217 & 2.64 \\ 
-%    biography & 0.489 & 3.54 \\
-%	\end{tabular}  \\
-%\end{tabular}          
-}                             
-\normalsize                   
-\label{t:narratorDifference} 
-\end{table*}
+%
+%{\bf Biography narrator detection.}~~
+%Table~\ref{t:narratorDifference} shows the accuracy results for the narrator detection and the biography boundary detection
+%for the Khoei biography book.
+%The first row shows the results computed by considering the biography book alone without the help of the narrator graph
+%extracted from the hadith books. 
+%Without the narrator graph, our method detected 94\% of the narrators mentioned in the book with 81\% precision. 
+%The method without the narrator graph, detected correctly (94\% precision) the boundaries of the biographies. 
+%
+%With the help of the narrator graph, our method detected 65\% of the narrator names in the biography book. 
+%The rest of the names were simply not found in the hadith books that we used to generate the narrator graph. 
+%With the use of the narrator graph and the coloring algorithm, our method detected the boundaries of the biographies
+%with 93\% precision. 
+%We believe that an exhaustive graph that includes narrator names from more books of hadith will improve the 65\% narrator 
+%detection recall metric.
+%In all cases, narrator names that exist in the biography and not in the hadith books are not of interest to the narrator
+%graph annotation problem.
+%
+%%The numbers in the right hand side of Table~\ref{t:narratorDifference}  may explain why 
+%Narrator detection in biography books is harder than in hadith books. 
+%For example, the ratio of narrator names to the rest of the text is 22\% in Musnad while it is 49\% in Khoei,
+%and the average length of a narrator sequence in Musnad is 2.6 narrators while it is 3.5 narrators in Khoei.
+%With the use of the narrator graph extracted from the hadith books, our method was able to improve biography boundary detection
+%from 41\% to 93\%.
+%This is evidence that cross-document entity detection helps to accurately find named entities in harder NLP tasks. 
+%
+%\begin{table*}[bt]                                     
+%\centering                                            
+%\caption{Accuracy results for the biography application.}
+%\resizebox{.7\columnwidth}{!}{
+%
+%%\begin{tabular}{cp{0.3cm}c}
+%
+%	\begin{tabular}{cp{.1cm}cccp{.1cm}ccc}
+%	          & & \multicolumn{3}{c}{narrator detection} & &  \multicolumn{3}{c}{biography boundary} \\
+%	 & & recall & precision & F-score & & recall & precision & F-score \\ 
+% \cline{3-5} \cline{7-9}
+%%hadith	 & sanad & & 0.923 & 0.999 & 0.960 & & 0.996 & 0.999 & 0.998  \\
+%%(musnad) & all & & 0.954 & 0.961 & 0.958 & & 0.996 & 0.987 & 0.992 \\ 
+%%	\hline \hline
+%w/o graph & & 0.943 & 0.812 & 0.873 & & 0.407 & 0.945 & 0.569 \\
+%with graph & & 0.654 & 0.909 & 0.760 & & 0.933 & 0.960 & 0.946 \\ % found in narrator graph
+%	 \end{tabular}   
+%%&  &
+%%	\begin{tabular}{ccc}
+%%	& names    & nar. \\
+%%    & per text & elements \\ \cline{2-3}
+%%	\multirow{2}{*}{0.217} &\multirow{2}{*}{ 2.64}\\ \\
+%%	\hline \hline
+%%	\multirow{2}{*}{0.489} & \multirow{2}{*}{3.54}\\
+%%    musnad & 0.217 & 2.64 \\ 
+%%    biography & 0.489 & 3.54 \\
+%%	\end{tabular}  \\
+%%\end{tabular}          
+%}                             
+%\normalsize                   
+%\label{t:narratorDifference} 
+%\end{table*}    
+%
 
@@
 %(present the necessary table and expand upon its significance)
+\section{Acknoledgements}
+\label{sec:acc}
+This work is supported by grant number 113030522236 from the Lebanese National Council for Scientific Research (LNCRS).
 
@@
 graph transformation algorithms. 
-We implemented our method and evaluated it using three NLP applications.
-Our method extracted named entities with high accuracy in all application.
-Our method extracted relations amongst the named entities with high accuracy in the hadith and biography applications, 
+We implemented our method and evaluated it using two NLP applications.
+Our method extracted named entities with high accuracy in all applications.
+Our method extracted relations amongst the named entities with high accuracy in the hadith application, 
 and with acceptable accuracy in the genealogical lists application. 
@@
 
-%\section{Acknoledgements}
-%\label{sec:acc}
-%We thank the Lebanese National Council for Scientific Research (LNCSR) for funding this research.
-
 %For reasons of uniformity, Adobe's {\bf Times Roman} font should be

### 2018-04-12 tex/merf/TALMajor/abstract.tex
changes to answer, colors, and some reviewers addressed

@@
 Rule-based techniques and tools to extract entities and relational entities from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.
-These techniques and tools require expertise in linguistics and programming and they lack support of Arabic {\em morphological analysis} which is key to process Arabic text.
+These techniques and tools require expertise in linguistics and programming.
+\major{
+They lack support of Arabic {\em morphological analysis} which is key to 
+process Arabic text.
+}
 In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.
-\framework provides a friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines {\em tag types} and interactively associates them with regular expressions defined over Boolean formulae.
+\framework provides a friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines tag types and interactively associates them with regular expressions defined over Boolean formulae.
 Boolean formulae range over matches of Arabic morphological features, and synonymity features.

### 2018-04-12 tex/merf/TALMajor/answer/Reviewer1.comments.tex
In progress: merging motivation with section 3
changes to answer, colors, and some reviewers addressed

@@
 
-\answer{
+\noindent\answer{
   Thank you. 
@@
 This make things rigouros but the notations are sometimes heavy.
+
 \answer{
   We clarified the following notations.... 
-  The new notations appear in Sectioon AA in teh paper. 
+  The new notations appear in Section AA in the paper. 
 }
@@
 
+\itodo{
+Add two sentences in the paper to say what isA does. 
+}
+
 \item \label{Review1.3} 
@@
 did you have a manual gold annotation of all the data ?
+
 \answer{
-We have to look at the test data for the paper. 
+Manual annotators inspected the outcome of the tools and provided 
+corrections where tools made mistakes. 
+The corrections form the manual gold annotation that we
+compared against. 
+We indicate that in the Results Section. 
+}
+\itodo{
+  Add teh above to teh results section. 
 }
@@
 did you have a train and a test data set ?
+
 \answer{
-No training... testing yes. 
+  \framework does not require training. 
+  It is an interactive tool where the user provides the formulae, expressions,
+    relation tuples and the associated code. 
+  The user inspects the results and refines the model accordingly. 
+  The testing set is available with the tool. 
+}
+
+\itodo{
+  Ameen will get the size from teh files and provide. 
 }
@@
 time during which the development has been done)
+more should be said about the evaluation methodology.
+
 \answer{
@@
 }
+\itodo{
+Add this to the results/evaluation
+}
 
 \item \label{Review1.5}
-more should be said about the evaluation methodology.
-\answer{
-We added a methodology section. 
-}
-
-\item \label{Review1.6}
 you say nothing about the availability of the system, 
 is it distributed ? free ? what kind of licence ?
+
 \answer{
-Available online on github, with GPL license. 
+\framework is open source and available online. We modified the
+contributions in the Introduction Section to indicate that. 
 }

### 2018-04-12 tex/merf/TALMajor/answer/Reviewer3.comments.tex
changes to answer, colors, and some reviewers addressed

@@
 As a reader, It wasn't very easy to followup with the writing 
-style. I needed to return back to ``Saref'' and ``MATAr'' papers 
+style. 
+I needed to return back to ``Saref'' and ``MATAr'' papers 
 to understand the terminology used.
+
 \answer{
@@
 
+\item \label{Review.3.2}
 First, I think we can ask 
-"What would a researcher reading this paper need to know?", the answer would be:
+``What would a researcher reading this paper need to know?", the answer would be:
 
-\item \label{Review.3.2}
 1- ``What is new in MERF?''. and this was clear for the reader 
@@
     b) defining tag of words based on a synonymic relation
+
 \answer{
-  ...
+  Thank you. 
 }
@@
 clear for the reader.
+
 \answer{
-  ... 
+  Thank you. We added a tutorial/video demo... 
 }
@@
 (methodology)?''. but I found this part was not very clear.
+
 \answer{
@@
 
-4-and for sure ``how it was evaluated and good results discussion''. hich needs more attention.
+\itodo{
+  We should the following sections.
+    1. \framework methodology: diagram with its description, and the 
+    reference to the example +...
+    2. \framework supporting tools: the current \framework section
+    3. \framework GUI: remain as is
+
+  Look for high level stuff in the (2) and (3) sections that can move
+    into (1).
+
+    When this is done go over the answers to make sure the references
+    to answers in the paper are accurate.
+}
+
+4-and for sure ``how it was evaluated and good results discussion''. 
+which needs more attention.
+
 \answer{
@@
 
-The following are some detailed recomendations:
+\itodo{same answer from Reviewer 1}
+
+The following are some detailed recommendations:
 
 \item \label{Review.3.5}
-1-In the introduction, I think you need to omit ``We discuss the importance of the morphological featureotivs supported in MERF terms in Section 3; in brief, morphological preprocessing iskey to Arabic NLP.'' , in page 2 since it will be repeated in page 5.
+1-In the introduction, I think you need to omit 
+    ``We discuss the importance of the morphological featureotivs supported in 
+    MERF terms in Section 3; in brief, morphological preprocessing iskey to 
+    Arabic NLP.'' , in page 2 since it will be repeated in page 5.
 \answer{
@@
 \item \label{Review.3.6}
-2- in the introduction you illusterated a target example, 
+2- in the introduction you illustrated a target example, 
 so why don you need the Motivation section. 
 What about if combine both.
+
 \answer{
@@
 is not clear.
+
 \answer{
-  We rephrase the sentence to read ... 
+  We rephrased the sentence in the paper to read as follows. 
+  \begin{quote}
+The leaf nodes of the match trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
+  \end{quote}
 }
@@
 provide examples. suggest the ``formula" was not so clear. 
-I suggest using regular grammar insted since you are using 
+I suggest using regular grammar instead since you are using 
 regular expression and the match tree can be considered as 
 parse tree.
+
 \answer{
@@
   We then use regular expressions (maybe we should use rules).
-  ... 
+
   We modify the paper by adopting the suggested terminology: 
   rule for regular expression and parse tree for match tree. 
+}
+
+\itodo{
+Lets use regular grammar and introduce that with similarity to antler.
+MRE becomes a macro maybe a good acronym is morphology based regular grammar rule 
+MRGR. 
 }
@@
 need to be with section 6 or section 6 need to be combined 
-with it in as one section but leaving the comparision to 
+with it in as one section but leaving the comparison to 
 the analysis section.
+
 \answer{
-  We thank the reviewer for the comment. 
-  We moved Section 1.1 to Section 6. 
+We merged the content of Subsection 1.1 with the Related Work 
+Section~\ref{sec:related}. 
 }
@@
 3- Section Background:Morphological analysis, 
-the definition explaned in page 9 did not match table 1 at page 7.
+the definition explained in page 9 did not match table 1 at page 7.
+
 \answer{
-  We fix the definition to reflect that a prefix (and an suffix)
-  could be a concatenation of other prefixes. 
+  We modified the definition to indicate that the prefix and suffix 
+  and the corresponding tags maybe concatenations of smaller prefix,
+  suffix and tags.  
 }
@@
 Section 2 seemed not very imporatnt as stand alone section.
+
 \answer{
-  We merged Section 2 with Section 1. 
-  We move Section 5 to become Section 2 so that it follows 
-  the motivation directly. 
+  We introduced a new Methodology Section~\ref{sec:methodology} and
+  we moved the content of old Section 2 to it. 
 }
-
-5- Section MERF/
 
@@
 \answer{
-We split Figure 3 into three Figures and provide explanations
-for both. 
-a. Arabic Text (Consider MBBF as input coming fro another figure) ----> Tags.
+We split Figure 3 (the flow diagram figure) 
+into three Figures and provide explanations
+for each. 
+a. Arabic Text (Consider MBF as input coming from another figure) ----> Tags.
 b. Rest of the figure... 
@@
 
+\itodo{
+  Edit the figure and split it  and refer to the (a), (b) and (c) in the paper. 
+}
+
 \item \label{Review.3.13}
-b) Syn$^k$ is nicely explaned but I did not recognize it as 
+b) Syn$^k$ is nicely explanied but I did not recognize it as 
 your own idea. if this is your idea to use English translation 
@@
 more focus because I think this is new.
+
 \answer{
 Up to our knowledge, we introduced the idea. 
-We highlight the concept more in the paper. 
+We claim that now as a novelty and a contribution in the introduction
+and we rephrase the beginning of Section~\ref{sec:synk}. 
 }
@@
 what is $2^{gloss}$ and $2^{s}$?
+
 \answer{
-  The power sets of... We clarify that in the paper. 
+The notation $2^s$ denotes the power set of set $s$ which is 
+the set of all subsets of $s$. 
+We clarify that in the paper. 
 }
@@
 variables. w,{u},..so on
+
 \answer{
-  We do that now.
+We modified the paragraph explaining the example to refer
+to the formula variables. 
 }
@@
 c) In MRE section,page 10, $k <=7$ why?
+
 \answer{
-  Computational reasons ... say it in the paper. 
+  We answer this by introducing the following in the paper.
+  \begin{quote}
+  We limit $k$ to a maximum of $7$ since we practically noticed that 
+  (1) values above $7$ introduce significant semantic noise and
+  (2) the computation is expensive without a bound. 
+  \end{quote} 
 }
@@
 then how CF belongs-to A. A is defined as one item not a set.
+
 \answer{
-Thank you. This is a mistake. We corrected that in teh paper to 
+This is a mistake. We corrected that in the paper to 
 read as follows... 
+}
+
+\itodo{
+  Ameen please find where you fixed this and say it here. 
 }
@@
 MRT definition is not clear.
+
 \answer{
 We clarify the definition of MRE in the paper. 
+}
+
+\itodo{
+  Clarify the definition. 
 }
@@
 were hard to be understood.
+
 \answer{
 We rewrote both sections... 
+}
+
+\itodo{
+  Figure our something to do/say.
 }
@@
 The reader can be lost so fast within the variable definitions.
+
 \answer{
 We provide running examples with the definitions....
+}
+\itodo{
+  Color them and say where they are. 
 }
@@
 for each word you are computing all the formulas values, 
-so the comlexity is O(text length X number of formulas), 
+so the complexity is O(text length X number of formulas), 
 what is the range of number of formulas ?
+
 \answer{
-  ... 
-  The complexity is (\#solutions for all words) times 
-  (\#number of user defined formulae). 
-  We provide no limit on that.
-  Practically, users define up to 10 formula. 
-  We include this discussion in the paper .. 
+  The complexity is the number of morphological solutions 
+  times the number of user defined formulae. 
+  We provide no limit on the number of user defined formulae.
+  Practically, in our case studies we did not need to defined
+  more than 10 formulae per case study. 
+  We include this discussion in the Results Section of the paper. 
+}
+
+\itodo{
+  Add that to the Results. 
 }
@@
 what R is.
+
 \answer{
   We cleaned the definitions... 
+}
+
+\itodo{
+  Ameen: highlight the changes and refer to them please. 
 }
@@
 how would the annotated text be saved?
+
 \answer{
   We used XXX data structures... 
+}
+
+\itodo{
+  Ameen: JSON add that as a paragraph in the methodology section. 
 }
@@
 accomplish similar jobs in the different applications. 
-This can be done by assiging similar corpus annotation job 
-and information extraction while computeing time needed by 
+This can be done by assigning similar corpus annotation job 
+and information extraction while computing time needed by 
 the user 
@@
 you can search for ``user interface system evaluation''. 
-There is huristic evaluation where you give weight for job 
+There is heuristic evaluation where you give weight for job 
 complexity and record time needed to complete it.
@@
 \answer{
-By development time we meant the time it required the 
-users to build an automated annotator similar to 
-the MERF based annotator. 
-We clarify that in the paper ...
+Our evaluation is a survey based evaluation.
+Users used \framework to define their \framework based annotators. 
+Users also built finite state transducer based annotators to 
+perform the same tasks on the same corpora. 
+We measured and reported the time needed by both.
 
-We conducted a survey as requested and we compare the 
-manual annotation time without MERF to the annotation
-time with MERF. 
+We also performed manual annotation to construct the correctness 
+benchmark.
+We did that by correcting the automated annotations of both annotators
+for the benchmark corpora. 
+That alone took much more time than the time it took to develop and run
+the \framework annotators.
+
+We clarify that in the beginning of the Results Section ~\ref{sec:results}.
+}
+
+\itodo{
+  Ameen: lets add the above to the section. 
 }
@@
 interface could be introduced in different view.
+
 \answer{
- ?????   
+  \framework targets entity and 
+  relational entity extraction. 
+  We are not aware and do not have access to other 
+  previous work that performs morphology based relational entity 
+  extraction for Arabic.
+  We are open to perform such experiments in case the reviewers 
+  can point us to such accessible tools. 
 }
@@
 sentences,..etc
+
 \answer{
@@
 
+\itodo{
+  Answer.. 
+}
+
 \item \label{Review.3.24}
 7- At section Discussion, I strongly urge you to discuss 
-WHY the precission and recall have these values.
+WHY the precision and recall have these values.
+
 \answer{
@@
 ``Methodology part''. 
+Section 4 ``MERF'' please add ``Methedology''.
+
+\answer{
+  We added a methodology section. 
+}
+
+\item \label{Review.3.25}
 Having ``Existing annotation and entity extraction tools'' 
 under unspecified section is misleading the reader. 
+unnumbered subsections was misleading me as a reader.
 
-\item \label{Review.3.25}
-unnumbered subsections was misleading me as a reader. 
 \answer{
@@
 table 1 at page 7 while referring to it only at page 9. 
-I think it need to be closer to where it was mentioned in the text. 
+I think it needs to be closer to where it was mentioned in the text.
+
 \answer{
@@
 
+\itodo{
+  Make sure this is moved. 
+}
+
 \item \label{Review.3.25}
 ``MERF regular expressions support operators such as concatenation, zero or one, zero or more, one or more, up to $M$” what is M? 
-it could be for example 0<=M<=|chunk| Section 4 “MERF”please add “Methedology”
+it could be for example 0<=M<=|chunk|
+
 \answer{
-  $M$ is ... we clarified that in the paper. 
+  We clarify the 
+  notation ``up to $M$'' in the paper to be as follows. 
+  \begin{quote}
+   up to $M$ repititions where $M$ is 
+   a non-zero positive integer 
+  \end{quote}
 }
@@
 needs to be reformulated
+
 \answer{
-  We rephrased the sentence as requested. 
+  We rephrased the sentence.
 }
@@
 for example look at words(,correct,commercial)
+
 \answer{
-  Emphasize ... 
+  We use italics to emphasize important 
+  concepts that we introduce for the first time. 
+  We reduced the use of emphasize in the paper. 
 }
@@
 this is an improvement of a per-existing work ``MATAr: Morphology-based Tagger for Arabic''
+
 \answer{
-  The work is available on github... 
+\framework is open source and available online. We modified the
+contributions in the Introduction Section to indicate that. 
 }

### 2018-04-12 tex/merf/TALMajor/introduction.tex
changes to answer, colors, and some reviewers addressed

@@
 ~\url{http://www.thedubaimall.com/ar/}}.
-The figure also presents a transliteration and an English translation of the Arabic text.
+\major{
+The figure also presents a transliteration and an English translation of 
+the Arabic text.
+}
 The framed words in the text are entities referring to names of people 
@@
 \framework regular expressions support operators such as concatenation, 
-zero or one, zero or more, one or more, up to $M\in\mathbb{N}^+$, and logical conjunction and disjunction.
+zero or one, zero or more, one or more, up to $M$ \major{repititions where $M$ is 
+a non-zero positive integer}, and logical conjunction and disjunction.
 An editor allows the user to associate an action with each sub-expression.
@@
     features such as text, position, length, numerical value, and morphological features.
-  \item \framework enables the user to tag words based on a synonymic relation using the $Syn^k$ feature.
+  \item \framework enables the user to tag words based on a 
+    \major{
+    novel ligh Arabic wordnet relation that leverages the 
+    synonym $Syn^k$ feature.
+  }
+
+  \item \major{
+\framework is open source and available online for the research community 
+    under on (\url{https://github.com/codelogicanalysis/atmine}).}
 \end{itemize}

### 2018-04-12 tex/merf/TALMajor/relatedwork.tex
changes to answer, colors, and some reviewers addressed

@@
 
-Another track in the literature targets specific tasks such as NER using statistical machine learning techniques \cite{benajiba2007anersys,benajiba2008arabic,ekbal2010named,abdelrahman2010integrated}.
+\major{
+Another track in the literature targets specific tasks such as NER using 
+statistical and machine learning techniques such as maximum entropy, 
+optimized feature sets and conditional random fields~\cite{benajiba2007anersys,benajiba2008arabic,ekbal2010named,abdelrahman2010integrated}.
+}
 %The work in~\cite{ekbal2010named} presents a language independent approach for NER extraction using {\em support vector machines}. 
@@
 %It defines three sections for declaration, rule definition, and macros. 
-The work in \cite{chiticariu2010systemt} presents SystemT, a system based on 
+The work in~\cite{chiticariu2010systemt} presents SystemT, a system based on 
 an algebraic Approach to Declarative information extraction (IE). 

### 2018-04-12 tex/merf/TALMajor/sarf.tex
updated related work
changes to answer, colors, and some reviewers addressed

--- 2018-03-13 tex/merf/TAL/sarf.tex
@@
 %{\bf Morphological analyzer (Sarf).~}
-Morphological analysis is key to Arabic NLP~\cite{arabicmorph} 
-due to the exceptional degree of ambiguity in writing, 
-the rich morphology, and the complex word derivation system. 
+\major{
+Morphological analysis is key to Arabic 
+NLP
+due its to the exceptional degree of ambiguity in writing, 
+the rich morphology, and the complex word derivation 
+system~\cite{arabicmorph,shahrour2016camelparser,pasha2014madamira}.
+}
 Short vowels, also known as diacritics, are typically omitted in Arabic text
@@
 
-Morphological analysis is required even for tokenization 
-of Arabic text. 
+Morphological analysis is required even for tokenization of Arabic text. 
 The position of an Arabic letter in a word 
@@
 for each word $t\in T$. 
+%
+\major{
 Each morphological solution $m\in M(t)$ is a tuple of the form 
-$\langle p,s,x,P,G,C\rangle\in \pp \times  \ss\times \xx\times \PP\times\GG \times \AC$.
+$\langle p,s,x,P,G,C\rangle \in \pp \times \ss \times \xx \times \PP \times \GG \times \AC$ 
+where $p=p_1 \ldots p_{|p|}$, $x=x_1 \ldots x_{|x|}$, 
+$P=P_{p_1}\ldots P_{p_{|p|}} P_s P_{x_1}\ldots P_{x_{|x|}}$,
+$G=G_{p_1}\ldots G_{p_{|p|}} G_s G_{x_1}\ldots G_{x_{|x|}}$, and
+$C=C_{p_1} \ldots C_{p_{|p|}} C_s C_{x_1}\ldots C_{x_{|x|}}$. 
+$P_{p_i}, G_{p_i},$ and $C_{p_i},1\le i \le |p|$ are the POS, Gloss and 
+category tags of prefix $p_i$. 
+$P_{x_j}, G_{x_j},$ and $C_{x_j},1\le j \le |x|$ are the POS, Gloss and 
+category tags of suffix $x_i$. 
+$P_s, G_s,$ and $C_s$ are 
+$P_s, G_s,$ and $C_s$ are the POS, Gloss and 
+category tags of stem $s$. 
+%
+Intuitively, 
+$p,x,P,G$ and $C$ are concatenations of prefix, suffix, POS, Gloss and
+Category values, respectively.
+}
 

### 2018-04-12 tex/merf/TALMajor/tool.tex
changes to answer, colors, and some reviewers addressed

--- 2018-03-17 tex/merf/TALMajor/tool.tex
@@
 \subsection{The extended synonymy feature $Syn^k$}
+\label{sec:synk}
 
-The motivation for $Syn^k$ is to construct a light Arabic ontology based on the lexicon of Sarf.
+\major{
+Up to our knowledge, 
+$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.
+}
 The sets $E, A,$ and $L$ denote all English words, Arabic words, 
@@
 Function $\alpha: \ss \rightarrow 2^{\GG}$ maps Arabic stems to 
-subsets of related English glosses. %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
+subsets of related English glosses, %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
+\major{
+where $2^{\GG}$ denote the power set of $\GG$ which is the set of all subsets 
+of $\GG$. 
+}
 Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
@@
 directly related to $w$ through the gloss map.
-
+%
 Let $Sy^{i}(w)$ denote stems related to $w$ using the gloss map of order $i$ recursively such that
@@
 Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
-
+%
+\major{
 The example in Figure~\ref{fig:introsynEx} illustrates the computation.
@@
 Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.
+}
 
@@
 \begin{itemize}
-  \item $a(w):= \exists m \in M(w). m=\langle p^k,s,x^l,P,G,C\rangle. r \circ CF$
-where $\circ \in {\cal O}$, $r \in \{p^k,s,x^l,P,G,C\}$, and $r\in A$.
+  \item $a(w):= \exists m \in M(w).~m=\langle p,s,x,P,G,C\rangle. r \circ CF$
+where $\circ \in {\cal O}$, $r \in \{p,s,x,P,G,C\}$, and $r\in A$.
 Informally, a solution vector of $w$ exists with
@@
   Informally, this checks if $w$ is an extended synonym of a stem $CF$. 
+\major{
+  We limit $k$ to a maximum of $7$ since we practically noticed that 
+  (1) values above $7$ introduce significant semantic noise and
+  (2) the computation is expensive without a bound. 
+}
 \end{itemize}
@@
 %and word distance based relational entities. 
+\major{
 The leaf nodes of the match trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
 For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the sub-expression $(P|N)+$.
+}
 

### 2018-04-12 tex/sarf.tex
changes to answer, colors, and some reviewers addressed

@@
 Name false positives & & 29\% & 22\%&  & 14\%& 13\% & & 36\% & 32\% \\ \hline
-Running time (secs.)& & 6.65 & 12$\times$60& &4.55 & 14$\times$60 & & .66 & 6$\times$60 
+Running time (secs.)& & 6.65 & 12$\times$60& &4.55 & 2.3$\times$3600 & & .66 & 29.2$\times$60 
 \end{tabular}
@@
 
-In this paper we o
-
 In the future, we plan to complete the hadith case study 

### 2018-04-12 ATSarf/src/sql/sql_queries.cpp

@@ -224,16 +224,19 @@ bool start_connection(ATMProgressIFC *p_ifc) { //and do other initializations
     theSarf->db.setHostName("localhost");
     theSarf->db.setDatabaseName("atm");
     theSarf->db.setUserName("root");
-    theSarf->db.setPassword("root");
+    theSarf->db.setPassword("");
+    theSarf->db.setConnectOptions("UNIX_SOCKET=/opt/lampp/var/mysql/mysql.sock");
     bool ok = theSarf->db.open();
 
     if (ok) {
+      std::cout << "Opening the database worked"  << (const char *) theSarf->db.lastError().text().toLatin1();
         theSarf->db.exec("SET NAMES 'utf8'");
         QSqlQuery temp(theSarf->db);
         theSarf->query = temp;
         check_for_staleness();
         return 0;
     } else {
+      std::cerr << "Error opening atm db: " << (const char *) theSarf->db.lastError().text().toLatin1();
         if (!tried_once) {
             system("mysql --user=\"root\" --password=\"\" -e \"create database atm\"");
             system(string("mysql --user=\"root\" --password=\"\" atm <\"" + databaseFileName.toStdString() + "\"").c_str());

### 2018-04-18 tex/merf/TALMajor/answer/Reviewer1.comments.tex
changes in reviewer comments

@@
 }
+\itodo{
+Pending...
+}
 
@@
 \answer{
-We modified the description of isA to .... 
+We modified the description of isA to address this issue. 
+Basically, isA is used to detect the `exact match' between a user-chosen value for a morphological feature and the corresponding feature value in a morphological solution of a word.
 }
 
-\itodo{
-Add two sentences in the paper to say what isA does. 
+\done{
+Add two sentences in the paper to say what isA does.\\ 
+Changes are highlighted in red at beginning of Subsection~\ref{subsec:grammar}.
 }
@@
 }
-\itodo{
-  Add teh above to teh results section. 
+\done{
+  Add the above to the results section.\\ 
+Done.
 }
@@
 \itodo{
-  Ameen will get the size from teh files and provide. 
+  Ameen will get the size from the files and provide. 
 }
@@
 teaching duties performed the task in two months. 
+We clarify this issue in the results Section.
 }
-\itodo{
-Add this to the results/evaluation
+
+\done{
+Add this to the results/evaluation.\\
+Done
 }

### 2018-04-18 tex/merf/TALMajor/answer/Reviewer2.comments.tex
In progress: merging motivation with section 3
changes in reviewer comments

@@
 tasks.
-\answer{
+
+\noindent\answer{
   Thank you for finding our work interesting. 
@@
 terms of more clear software techniques.
+
 \answer{
@@
 ... 
+}
+
+\itodo{
+Pending...
 }
@@
 development tasks.
+
 \answer{
-  The problem is decidable: RE to FSM. 
-  The problem might be not tractable: NDFSM to FSM might 
+  The problem is decidable: regular grammar rules to FSM. 
+  However, the problem might be not tractable: NDFSM to FSM might 
   be exponential. 
-  In practice Table XXX provides runtime results... 
+  In practice Table~\ref{tab:results} provides runtime results... 
   We modified the paper to include the above.
+}
+
+\done{
+Done.
 }
@@
 the benefit of the implements tool.
+
 \answer{
@@
 
+\done{
+You note is accurate.
+An advantage of the GUI is that it enables a user with mediocre 
+linguistic skills to perform information extraction tasks and 
+benefit from the framewrok.
+}
+
 \end{enumerate}

### 2018-04-18 tex/merf/TALMajor/answer/Reviewer3.comments.tex
changes in reviewer comments

@@
   We added a background section with definitions... 
+}
+
+\done{
+Pending...
 }
@@
     Arabic NLP.'' , in page 2 since it will be repeated in page 5.
+
 \answer{
@@
   as a subsection of the introduction. 
+}
+
+\done{
+  We combine the Motivation section with \framework section (Now Section~\ref{sec:framework}).
 }
@@
 We claim that now as a novelty and a contribution in the introduction
-and we rephrase the beginning of Section~\ref{sec:synk}. 
+and we rephrase the beginning of Section~\ref{subsec:synk}. 
 }
@@
 \answer{
-This is a mistake. We corrected that in the paper to 
-read as follows... 
+F is the set of morphological features where each element is in turn a set of the values of that feature.
+Hence, A is one such set corresponding to a morphological feature, and CF is a feature value in A.
+We clarify this point in Subsection~\ref{subsec:grammar}.
 }
 
-\itodo{
-  Ameen please find where you fixed this and say it here. 
+\done{
+  DONE: Ameen please find where you fixed this and say it here. 
 }
@@
 
-\itodo{
-  Add that to the Results. 
+\done{
+  Add that to the Results.\\
+  Added the above explanation in the Results section, needs to be checked.
 }
@@
 \answer{
-  We cleaned the definitions... 
+  Thank you for pointing it out. We cleaned the definitions.
 }
 
-\itodo{
-  Ameen: highlight the changes and refer to them please. 
+\done{
+  highlight the changes and refer to them please.\\
+  We use R first to denote the set of user-defined relations.
+  Then, we use it to denote the set of MBF tags which is contradicting.
+  I got rid of the first case as we don't need it.
 }
@@
 
-\itodo{
-  Make sure this is moved. 
+\done{
+  Done: Make sure this is moved. 
 }

### 2018-04-18 tex/merf/TALMajor/answer/Reviewer4.comments.tex
changes to answer, colors, and some reviewers addressed
changes in reviewer comments

@@
 subexpressions!
+
 \answer{
-  We clarify that... 
+  We elaborate the use of code actions in Subsection~\ref{subsec:numnorm} 
+    by introducing Figure~\ref{fig:numnormalgo} and discussing it 
+    in the corresponding case study. 
+}
+
+\done{
+  Discussion of the code actions is pending...
 }
@@
 - $Syn^k$ introduces a lot of noise especially if $k$ is large.
+
 \answer{
-  7 is a limit... 
+  Your note is accurate, this is why we limit the degree (k) to 7.
+  We further elaborate on this by introducing the following in the paper; Subsection~\ref{subsec:grammar}.
+  \begin{quote}
+  We limit $k$ to a maximum of $7$ since we practically noticed that 
+  (1) values above $7$ introduce significant semantic noise and
+  (2) the computation is expensive without a bound. 
+  \end{quote}
+}
+
+\done{
+  Done: Use the same answer for Reviewer 3, and say why you used 7. 
 }
@@
 to the quality of the annotator!
+
 \answer{
@@
 
+\itodo{
+Say something about clarifying the evaluation process. 
+}
+
 \end{enumerate}

### 2018-04-18 tex/merf/TALMajor/results.tex
changes to answer, colors, and some reviewers addressed
changes to results and tool

@@
 In this section we compare \framework implementations of 
-the narrator chain, temporal entity, and genealogy entity extration 
+the narrator chain, temporal entity, and genealogy entity extraction 
 tasks to the task specific techniques proposed to solve them
@@
 , we report on eight additional \framework case studies.
-
+\major{
+Manual annotators inspected the outcome of the tools and provided corrections where tools made mistakes.
+The corrections form the manual gold annotation that we compared against.}
 Table~\ref{tab:results} reports the development time,
@@
 for both the standalone task specific and the \framework implementations.
+
+\major{
+The development time presents the period of time during which the development was done.
+For instance, ANGE task~\cite{ZaMaFlairs2012HadithBio} was performed in two months by a research assistant with a load of six hours of course work and 14 hours of teaching duties.
+
+Another aspect to evaluate is the complexity of running \framework simulator for both MBFs and MRGRs over an input text.
+The complexity of the MBF simulation is the total number of morphological solutions for all the words multiplied by the number of user-defined MBFs.
+Note that we don't provide a limit on the number of formulae defined by the user.
+Practically, we did not need to define more than 10 formulae per case study.
+As for the complexity of MRGR simulation, converting the rules into non-deterministic finite state machines (NDFSM) is decidable.
+However, simulating an NDFSM over the MBF tags might not be tractable as it could be exponential.
+Table~\ref{tab:results} provides run time results of \framework which presents a practical evaluation of the complexity of MBF and MRGR simulation jointly.}
 %Table~\ref{tab:mbfer} reports the accuracy of the output MBF tags and the user-defined relation 
@@
 \subsection{Number normalization}
-\label{sec:sec:number}
+\label{subsec:numnorm}
 

### 2018-04-18 tex/merf/TALMajor/tool.tex
changed some white space
changes to results and tool

@@
 \subsection{The extended synonymy feature $Syn^k$}
-\label{sec:synk}
+\label{subsec:synk}
 
@@
 Up to our knowledge, 
-$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.
-}
+$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.}
 The sets $E, A,$ and $L$ denote all English words, Arabic words, 
@@
 \major{
-where $2^{\GG}$ denote the power set of $\GG$ which is the set of all subsets 
-of $\GG$. 
-}
+where $2^{\GG}$ denotes the power set of $\GG$ which is the set of all subsets 
+of $\GG$.}
 Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
@@
 The example in Figure~\ref{fig:introsynEx} illustrates the computation.
-The input Arabic word \RL{mA'}, denoted $w$, shares the gloss \cci{water} with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
+Let $w$ denote an input Arabic word \RL{mA'}, which has the gloss \cci{water}, i.e. \cci{water} $\in\alpha(w)$.
+$w$ shares this gloss with the stem \RL{n.d.h}, denoted $s_1$, i.e. $s_1\in Sy^{1}(w)$.
 %The input word \RL{mA'} shares the gloss \RL{n.d.h} through the gloss intersection \cci{water}. 
@@
 \setcode{standard}
+\vspace{-2.5em}
 \caption{$Syn^2($\RL{mA'}$)$}
+\vspace{-3em}
 \label{fig:introsynEx}
@@
 \subsection{MRE: Morphology-based regular expressions}
+\label{subsec:grammar}
 
+\major{
 Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
-predicates, and let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
+predicates, where $\mathit{isA}$ and $\mathit{contains}$ denote exact match and containment, respectively.
+Also, let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
 set of morphological features where each entry is in turn a set of values for that feature.
-Given a word $w$, a morphological feature $A\in{\cal F}$, 
-a user defined constant feature value $CF\in A$, and an integer 
-$k, 1\le k\le 7$, 
+Given a word $w$, a morphological feature set $A\in{\cal F}$,}
+a user defined constant feature value $CF\in A$,
+and an integer $k, 1\le k\le 7$, 
 the following are morphology-based atomic terms (MAT), {\em terms} for short.
@@
 Informally, a solution vector of $w$ exists with
-a feature containing or exactly matching $CF$.
-The predicates $\mathit{isA}$ and $\mathit{contains}$ denote the exact match and contain criteria, respectively.
+a feature containing or exactly matching \major{the user-chosen feature value} $CF$.
 \item $a(w) := w \in Syn^k(CF), CF \in \ss$.
@@
 or apply intelligent entity inference techniques
-as we show later in the numerical extraction example of Section~\ref{sec:sec:number}.
+as we show later in the numerical extraction example of Subsection~\ref{subsec:numnorm}.
 \framework provides an API that enriches the actions with detailed access to

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer1.comments.tex
done till 3.19

@@
 \answer{
-  We clarified the following notations.... 
-  The new notations appear in Section AA in the paper. 
-}
-\itodo{
-Pending...
+  We added a methodology section that precedes the formal 
+    notation. 
+  We also embedded a running example along with the formal
+    notations. 
 }
@@
 \answer{
-We modified the description of isA to address this issue. 
-Basically, isA is used to detect the `exact match' between a user-chosen value for a morphological feature and the corresponding feature value in a morphological solution of a word.
-}
+  We clarified the \cci{isA} predicate in Section~\ref{subsec:grammar}.
+  Basically, isA is used to detect the `exact match' 
+  between a user-chosen value for a morphological feature and the 
+  corresponding feature value in a morphological solution of a word.
 
-\done{
-Add two sentences in the paper to say what isA does.\\ 
-Changes are highlighted in red at beginning of Subsection~\ref{subsec:grammar}.
+  We also replaced a mistake that referenced the \cci{isA} predicate
+  in Figure~\ref{fig:intromotiv} by the correct predicate \cci{isSyn}. 
 }
@@
 \answer{
+  We added description of the case studies in the Results Section~\ref{sec:results}. 
+  The reference chunks are manually created gold annotations.
   Manual annotators inspected the outcome of the tools and provided 
@@
 We indicate that in the Results Section. 
-}
-\done{
-  Add the above to the results section.\\ 
-Done.
 }
@@
 \answer{
+  We added a column to Table~\ref{tab:results} to refer to the size 
+  of the annotated data for each case study and we discussed the numbers
+  in the Results Section~\ref{sec:results}. 
   \framework does not require training. 
@@
   The testing set is available with the tool. 
-}
-
-\itodo{
-  Ameen will get the size from the files and provide. 
 }
@@
 \answer{
+  We clarify this issue in the Results Section~\ref{sec:results}.
 A research assistant 6 hours course work and 14 hours 
 teaching duties performed the task in two months. 
-We clarify this issue in the results Section.
-}
-
-\done{
-Add this to the results/evaluation.\\
-Done
 }
@@
 \framework is open source and available online. We modified the
-contributions in the Introduction Section to indicate that. 
+contributions in the Introduction Section~\ref{sec:introduction} to indicate that. 
 }

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer2.comments.tex
done till 3.19

--- 2018-03-13 tex/merf/TALMajor/answer/Reviewer2.comments.tex
@@
 terms of more clear software techniques.
+
 \answer{
-We introduced the following formal definitions...
+We introduced a Methodology Section~\ref{sec:methodology} to separate
+the conceptual aspects from the implementation aspects of the
+framework.
 
-We clarify the implementation in the following manner. 
-Tranformative algorithms: 1. RE to FSM
-                          2. Entities and relations to garphs
-Generative algorithms: 1. Actions to code
-... 
+We refer to the collection of regular expressions as a grammar, 
+we also provide information about the data model used to store
+the user defined tag types. 
+In the results Section~\ref{sec:results} we discuss the 
+complexity of the simulators.
+
+We now relate the action code API to corresponding software design patterns in Section~\ref{sec:relations-actions}, 
+%
+We discuss the translation from grammar to NDFSA. We also 
+discuss the generative nature of the action code as a shared 
+object library. 
 }
@@
 \answer{
+  We provide a Methodology Section~\ref{sec:methodology} 
+    that benefits directly from the preceding Background.
+  We introduced new Figure~\ref{f:overview} to show that 
+  the morphological analysis is the first preprocessing 
+  step in the data flow of \framework. 
+  Section~\ref{sec:morph} that discusses morphological 
+    analysis with solution examples in Table~\ref{tab:samplesolution}.
   Section XXX discussed morphological analysis in Arabic. 
@@
 development tasks.
+
 \answer{
-  The problem is decidable: RE to FSM. 
-  The problem might be not tractable: NDFSM to FSM might 
-  be exponential. 
-  In practice Table XXX provides runtime results... 
-  We modified the paper to include the above. 
+  We added discussion in the Results Section~\ref{sec:results}
+    to discuss and comment on the complexity of the 
+    regular expression to non-deterministic finite state
+    automata (NDFSA), and the simulation of the NDFSA  
+    with the morphology based Boolean formulae tags. 
 }
@@
 the benefit of the implements tool.
+
 \answer{
-We discuss the target audience of the MERF tool in Section XXX. 
-We prepared a set of 45 minutes slides based on the regular
-expression chapter in Martin Jurafsky's book to... 
+We added a statement at the end of the Discussion 
+Subsection~\ref{sec:discuss} to reflect the comment of the reviewer. 
 }

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer3.comments.tex
done till 3.19

@@
 \answer{
-  We added a background section with definitions... 
-}
-
-\done{
-Pending...
+  We cleared notations in the paper that might have led to 
+    the confusion.
+    For example, we cleared (1) the notation for morphological solutions,
+    (2) the use of symbol $R$ as a both a relation tuple and a 
+    tag set. 
+    We clarified the \cci{isA} and \cci{contains} predicates. 
+    We define the power set notation $2^A$ where $A$ is a set
+    and $2^A$ is the set of all subsets of $A$.
 }
@@
 \answer{
-  Thank you. We added a tutorial/video demo... 
+  Thank you. 
 }
@@
 \answer{
-  We added a methodology section...
-}
-
-\itodo{
-  We should the following sections.
-    1. \framework methodology: diagram with its description, and the 
-    reference to the example +...
-    2. \framework supporting tools: the current \framework section
-    3. \framework GUI: remain as is
-
-  Look for high level stuff in the (2) and (3) sections that can move
-    into (1).
-
-    When this is done go over the answers to make sure the references
-    to answers in the paper are accurate.
+  We added a methodology Section~\ref{sec:methodology} with a 
+    new Figure~\ref{f:overview} that illustrates the four
+    processes involved in the \framework Methodology. 
 }
@@
 \answer{
-  We added subsection Results.X to discuss the evaluation 
-  strategy. 
-  We also added extended the discussion section... 
+  We added more discussions to the results Section~\ref{sec:results}
+    including case study size, development time evaluation, 
+    complexity, and a discussion on required expertise. 
 }
-
-\itodo{same answer from Reviewer 1}
 
@@
 \answer{
-  Thank you. We combine both as motivation and we include that
-  as a subsection of the introduction. 
-}
-
-\done{
-  We combine the Motivation section with \framework section (Now Section~\ref{sec:framework}).
+  Thank you for the note. 
+  We removed the motivation Section and embedded its content as 
+  a running example in the \framework Components Section~\ref{sec:tool}. 
 }
@@
   \begin{quote}
-The leaf nodes of the match trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
+The leaf nodes of the trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
   \end{quote}
@@
 \answer{
-  We use formula to define predicates that range over 
-  the morphological features. 
-  We consider the matches of the formulae as tokens/tags (maybe
-  we call them tag types.)
-  We then use regular expressions (maybe we should use rules).
-
-  We modify the paper by adopting the suggested terminology: 
-  rule for regular expression and parse tree for match tree. 
-}
-
-\itodo{
-Lets use regular grammar and introduce that with similarity to antler.
-MRE becomes a macro maybe a good acronym is morphology based regular grammar rule 
-MRGR. 
+  We modified the paper to refer to the collection of the regular
+  expressions as a local grammar. 
+  We also use matching parse trees instead of match trees in the paper now. 
+  We added a sentence in the Introduction Section to 
+  clarify that ``formula'' across the paper denotes a morphology
+  based Boolean formula (MBF). 
+  We do the same for regular expressions and rules. 
 }
@@
 \answer{
-  We introduced a new Methodology Section~\ref{sec:methodology} and
-  we moved the content of old Section 2 to it. 
+  We removed the Motivation Section (2 in the old version) and
+  embedded its contents in the \framework Components Section (4 in the
+  old version and \ref{sec:tool} now). 
+  We also introduced a Methodology Section~\ref{sec:methodology}. 
 }
@@
 \answer{
-We split Figure 3 (the flow diagram figure) 
-into three Figures and provide explanations
-for each. 
-a. Arabic Text (Consider MBF as input coming from another figure) ----> Tags.
-b. Rest of the figure... 
-c. RTC + Diff. + Visualizatiin 
-}
-
-\itodo{
-  Edit the figure and split it  and refer to the (a), (b) and (c) in the paper. 
+  We replaced the figure with Figure~\ref{f:overview}.
+  Figure~\ref{f:overview} is 
+  a clearer and is composed of four
+  separate processes. 
+  We discuss the new Figure in the Methodology Section~\ref{sec:methodology}.
 }
@@
 \item \label{Review.3.13}
-b) Syn$^k$ is nicely explanied but I did not recognize it as 
+b) Syn$^k$ is nicely explained but I did not recognize it as 
 your own idea. if this is your idea to use English translation 
@@
 \answer{
-F is the set of morphological features where each element is in turn a set of the values of that feature.
-Hence, A is one such set corresponding to a morphological feature, and CF is a feature value in A.
-We clarify this point in Subsection~\ref{subsec:grammar}.
+F is a sets of sets. It is a set of morphological feature sets.
+Each morphological feature set is in turn a set of values of that feature.
+CF belongs to A denotes that CF is a value of a morphological feature
+set  A. 
+We clarify this point in SubSection~\ref{subsec:grammar}.
 }

### 2018-04-20 tex/merf/TALMajor/gui.tex
done till 3.19

--- 2018-03-13 tex/merf/TAL/gui.tex
@@
 
-\framework saves the tags and the tag types in a user friendly format using
-the JavaScript Object Notation (JSON) format~\cite{nolan2014javascript}.
-The tag file keeps the paths of the separate text and tag type files. 
-It also contains a list of tags with their tag type identifiers,
-position in text, length, and word index.
-The tag type file contains MBF and MRE tag types. 
-%Each object has an array value containing the user-defined tag types. 
-Each tag type contains the name, description, expression, and the visualization legends.
+%\framework saves the tags and the tag types in a user friendly format using
+%the JavaScript Object Notation (JSON) format~\cite{nolan2014javascript}.
+%The tag file keeps the paths of the separate text and tag type files. 
+%It also contains a list of tags with their tag type identifiers,
+%position in text, length, and word index.
+%The tag type file contains MBF and MRE tag types. 
+%%Each object has an array value containing the user-defined tag types. 
+%Each tag type contains the name, description, expression, and the visualization legends.
 
@@
   \includegraphics[width=\textwidth]{figures/treegraph}
-  \caption{MRE annotated Text, MRE match tree, and entity-relation graph}
+  \caption{MRE annotated Text, MRE \major{matching parse tree}, and entity-relation graph}
   \label{fig:treegraph}
@@
 
-The match tree view shows the selected match in a graph view.
-Figure~\ref{fig:treegraph} shows the MRE match tree of the direction task 
+The \major{matching parse tree} view shows the selected match in a graph view.
+Figure~\ref{fig:treegraph} shows the \major{matching parse tree} of the direction task 
 \RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
@@
 the entity-relation graph of the match of the user defined relation 
-extracted from the match tree of the MRE.
-
+extracted from the \major{matching parse tree} of the MRE. 
+%
 In the computational action editor, an advanced user can 
@@
 The tag type difference view shows the text annotated in three panes: 
-1) the common tag types ${\cal T}_1 \cap {\cal T}_2$,
-2) the tag types in ${\cal T}_1$ but not in ${\cal T}_2$, 
-and 3) the tag types in ${\cal T}_2$ and not in ${\cal T}_1$.
-
+(i) the common tag types ${\cal T}_1 \cap {\cal T}_2$,
+(ii) the tag types in ${\cal T}_1$ but not in ${\cal T}_2$, 
+and (iii) the tag types in ${\cal T}_2$ and not in ${\cal T}_1$.
+%
 Similarly, the tag difference view shows $R_1\cap R_2$, $R_1/R_2$ and $R_2/R_1$

### 2018-04-20 tex/merf/TALMajor/introduction.tex
done till 3.19

@@
 semantic values such as entities or relational entities.
-We refer to annotations and labels as {\em tags} and {\em tag types}, respectively, in the sequel.
+We refer to annotations and labels as {\em tags} and {\em tag types}, 
+respectively, in the sequel.
 
@@
 
+\major{
+Figure~\ref{fig:intromotiv} illustrates the target of \framework using 
+the directions to Dubai Mall example~\footnote{text taken from the Dubai 
+Mall website ~\url{http://www.thedubaimall.com/ar/}}.
+The figure also presents a transliteration and an English translation of 
+the Arabic text.
+}
+The framed words in the text are entities referring to names of people 
+($n_1,n_2,n_3$), 
+names of places ($p_1,\dots,p_7$), 
+relative positions ($r_1,\dots,r_4$), 
+and numerical terms ($u_1,u_2$). 
+We would like to extract those entities, and then extract
+the relational entities forming the graph in Figure~\ref{fig:intromotiv} 
+where vertices express entities, 
+and edges represent the relational entities.
+
 %These tasks take as input digital text documents from sources including literature, books, news, business reports, chat messages, and emails. 
@@
 Boolean formulae.
+%
 A Boolean formula is defined by a term, negation of a term, or disjunction of terms.
 Terms are matches to Arabic morphological features including 
-prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym tags, and semantic categories.
-We often refer to regular expressions and Boolean formulae as expressions and formulae, respectively, for short.
+prefix, stem, suffix, part of speech (POS) tags, gloss tags, extended synonym 
+tags, and semantic categories.
+\major{
+For example, entity $p_1$ in Figure~\ref{fig:intromotiv} has a place semantic 
+category. 
+}
+\framework regular expressions support operators such as concatenation, 
+zero or one, zero or more, one or more, up to $M$ \major{repititions where $M$ is 
+a non-zero positive integer}, and logical conjunction and disjunction.
+\major{
+For example, the sequence between $p_1$ and $p_2$ matches a regular expression $re$ 
+that requires two semantic place categories with a place-preposition POS 
+tag ($r_1$) in between. 
+}
+
+An editor allows the user to associate an action with each sub-expression.
+The user specifies the action with C++ code and uses an API to access information 
+related to the matches such as text, position, length, morphological features, 
+and numerical value.
+%
+\major{
+Each regular expression is associated with a named identifier to form a 
+{\em local grammar} like structure~\cite{traboulsi2009arabic}. 
+%
+A relation definition GUI allows the user to provide relational tuples 
+where each tuple has a 
+source, a destination and an edge label.
+The user uses the regular expression identifiers to define the relational 
+tuple elements.
+For example, the relation between $e_1, e_2$ and $r$ shown in 
+Figure~\ref{fig:intromotiv} is a match of a relational tuple 
+over the components of $re$. 
+%
+We refer to regular expressions and Boolean formulae as expressions and formulae, respectively.
+We also refer to expressions as rules
+when used in a grammar context; e.g.  when used with an identifier. 
 %We discuss the importance of the morphological features supported in 
@@
 %is key to Arabic NLP.% due to the richness of Arabic morphology.
+}
 
-We illustrate the target of \framework using a simple example.
-Consider the text in Figure~\ref{fig:intromotiv} which contains
-directions to Dubai Mall~\footnote{text taken from the Dubai Mall website
-~\url{http://www.thedubaimall.com/ar/}}.
-\major{
-The figure also presents a transliteration and an English translation of 
-the Arabic text.
-}
-The framed words in the text are entities referring to names of people 
-($n_1,n_2,n_3$), 
-names of places ($p_1,\dots,p_7$), 
-relative positions ($r_1,\dots,r_4$), 
-and numerical terms ($u_1,u_2$). 
-We would like to extract those entities, and then extract
-the relational entities forming the graph in Figure~\ref{fig:intromotiv} 
-where vertices express entities, 
-and edges represent the relational entities.
-\framework allows a regular user to do that interactively
-in four phases that include morphological analysis, 
-entity extraction based on morphological features, 
-relational entity extraction, and entity cross-reference.
-
-\framework regular expressions support operators such as concatenation, 
-zero or one, zero or more, one or more, up to $M$ \major{repititions where $M$ is 
-a non-zero positive integer}, and logical conjunction and disjunction.
-An editor allows the user to associate an action with each sub-expression.
-The user specifies the action with C++ code and uses an API to access information related to the matches such as text, position, length, morphological features, and numerical value.
-
-\framework takes an Arabic text, a set of user-defined Boolean 
-formulae, and a set of user-defined regular expressions.
-\framework computes the morphological solutions of the words in the 
-input text then computes matches to the Boolean formulae.
+\framework takes an Arabic text and the local grammar defined by the 
+Boolean formulae and the regular expressions.
+\framework computes the morphological solutions of the 
+input text then computes matches to the Boolean formulae therein.
 \framework then generates a {\em non-deterministic finite state 
@@
 
-This work significantly extends the position paper~\cite{JaZaMatar} that allows for manual, and basic morphology annotation.
+This work significantly extends the position paper~\cite{JaZaMatar} 
+that allows for manual, and basic morphology annotation.
 \framework enables a regular user to incrementally create 

### 2018-04-20 tex/merf/TALMajor/methodology.tex
done till 3.19

--- 2018-03-17 tex/merf/TALMajor/methodology.tex
@@
-The set of tag types ${\cal T}$ contains tuples of the form $\langle l,f,d\rangle$ 
-where $l$ is a text label with a descriptive name, 
-$f$ is an MRE, and $d$ is a visualization legend 
-with font and color information.
+Figure~\ref{f:overview} illustrates the four processes involved in 
+\framework methodology. 
+The first process takes the Arabic text input and provides the user 
+with a morphology-based Boolean (MB) formulae GUI. 
+The user interactively composes MB-formulae using the GUI and the output 
+of the simulator and the $Syn^k$ detector. 
+The simulator and the detector apply the formulae over the morphological solutions 
+of the Arabic text and produce the MB-formulae tags.
 
-For each word $t_i\in T, 0\le i < M, M=|T|$.
-\framework computes a Boolean value 
-($\{\mathit{true}, \mathit{false}\}$)
-for all MBFs. 
-Then, it computes the set of MBF tags
-$R_i=\{(t_i,tt)| tt=\langle l,f,d\rangle \wedge
-f~\mathit{is~an~MBF} \wedge f(t_i)\} \subseteq T \times {\cal T}$
-which tags a word $t_i$ with $\mathit{tt}$ 
-iff the MBF $f$ associated with
-tag type $\mathit{tt}$ is true for $t_i$. 
-The MBF evaluation results in a sequence of tag sets 
-$\langle R_0, R_1, \ldots, R_{n-1}\rangle$.
-If a word $t_i$ has no tag type match, 
-its tag set $R_i$ is by default the singleton $O=\{\mathit{NONE}\}$.
-%and $t_i$ is referred to as a {\em null} word.
+The second process takes the MB-formulae tags and the Arabic text and provides
+the user with a morphology-based grammar rule GUI. 
+The user interactively composes MB-grammar rules using the GUI and the output 
+of the MB-grammar rule simulator. 
+The grammar rule simulator applies the rules over the MB-formulae tags and 
+produces the MB-grammar rule tags.
 
-%\subsection{MRE and action simulation}
+The third process takes the MB-grammar rule tags and provides the user with 
+a relation and action GUI. 
+The user interactively provides (1) the relation definitions and 
+(2) the actions in terms of identifiers from the MB-grammar 
+rules.
+The relation extraction produces the target entities and relational
+entities. 
+The action execution enriches the entities and the relational entities
+with powerful semantics.
+For example, users can utilize actions to compute statistical features, 
+store intermediate results, 
+or apply intelligent entity inference techniques
+as we show later in the numerical extraction example of 
+Subsection~\ref{subsec:numnorm}.
+%
+Finally, in the fourth process the user compares the results with golden
+reference chunks and visualizes the difference. 
+This allows the user to refine the formulae, rules, relations and actions.
 
-For each MRE, 
-\framework generates its equivalent non-deterministic finite automaton (NFA) in the typical manner~\cite{sipser2006introduction}. 
-%Each MRE operation has its equivalent representation in an NFA. 
-We support the upto operation ($f$\^{}$x$), which is not directly 
-supported in ~\cite{sipser2006introduction}, by 
-%it is equivalent to $f?|ff|\dots|\underbrace{f}_{x \text{ times}}$ which has an NFA mapping. 
-%we can expand it into a standard regular expression form, for example
-expanding it into a regular expression form; for example 
-$f$\^{}$3$ is equivalent to $f?|ff|fff$. 
-Consider the example of Figure~\ref{fig:intromotiv} and the
-corresponding expression $(P|N)\!+~O?~R~O^\wedge 2~(P|N|U)+$. 
-Figure~\ref{fig:nfaEx} shows part of the corresponding NFA where
-$q_8, q_9, \dots, q_{13}$ represent NFA states,
-and edges are transitions based on MBF tags such as 
-$P,$ and $N$.
-Edges labeled with the empty string $\epsilon$ are non-deterministic.
-%The expression uses operations $|$, $?$, $+$, and $\wedge$ to relate places {\tt P}, names of persons {\tt N}, relative positions {\tt R}, numerical terms {\tt U}, and other {\tt O}.
+After the relation construction phase, we are interested to relate the 
+discovered entities that express the same concept.
+\framework provides the extended synonym feature of second order as 
+a default cross-reference relation ($Syn^2$).
+In Figure~\ref{fig:intromotiv}, triggering this feature creates the 
+edge labels with \cci{isSyn} 
+between the nodes \cci{Khalifa Tower} and \cci{The building}.
 
-\framework simulates the generated NFA over the sequence of tag sets matching the MBF formulae.
-A simulation match $m$ of an expression $f$ is a tree where the root is the 
-MRE expression, the internal nodes are the MRE and MBF operations, and the 
-leaves are matches of the MAT terms of $f$.
-The leave matches form a vector of tags
-$\langle r_k,r_{k+1},\dots,r_j\rangle$ 
-corresponding to the text sequence 
-$\langle t_k,t_{k+1},\dots,t_j\rangle$ where $r_{\ell}\in R_{\ell},0\le k\le \ell \le j < n$.
-%
-%If the simulation has a match $\langle r_m,r_{m+1},\dots,r_n\rangle$ 
-%where $0\le m\le n$, 
-%the next simulation starts at $R_{n+1}$. 
-%This disallows overlap of matches for the same MRE. 
-%
-%In case the NFA simulation has no match, 
-%the next simulation starts at $R_{m+1}$. 
-If we have more than one match %starting at $R_k$ where $0\le k\le n$, 
-\framework returns the longest. 
-%
-Figure~\ref{fig:motiv}(b) shows two match trees of $\mathit{dir}$ extracted 
-from the text of Figure~\ref{fig:intromotiv}.
-\RL{dby} and \RL{mwl} are leaf nodes referring to name of place tags ($P$). 
-The $+$, $|$, and $?$ MRE operations
-are internal nodes.
+\input{MERFOverview}
 
-%\framework maintains a function $\phi\subset Q\times\Phi$, 
-%where $\mathit{Q}$ is the set of states in the NFA and $\Phi$ is the set of sub-expressions in the MRE.
-%So $(q,f)\in \phi$ iff state $q\in Q$ was generated by \framework to correspond to sub-expression $f\in \Phi$. 
-%\framework uses $\phi$ to compute a match tree with respect to the MRE regular expression. 
-%It also uses $\phi$ and the match sequence to compute the sequence of computational actions of an MRE match.
-
-\setcode{utf8}
-\setarab
-\transfalse
-
-\begin{figure}[tb!]
-\centering
-\resizebox{0.7\columnwidth}{!}{
-	\relsize{-1} \input{figures/reNFA.tex}
-}
-  \caption{Equivalent NFA of direction expression}
-  \label{fig:nfaEx}
-\end{figure}
-\transtrue
-\setcode{standard}
-
-%\begin{figure}[tb!]
-%{ \relsize{-1}
-%\begin{framed}
-%\begin{multicols}{2}
-%\begin{itemize}
-%\item Annotated Expression: \\
-%\\
-%{ \relsize{-2.5}
-%%\begin{itemize}
-%%  \item 
-%  $\stackrel{e_1}{(P|N)+}~\stackrel{o_1}{\mathit{O}?}~\stackrel{r}{R}~\stackrel{o_2}{\mathit{O^{\wedge}2}}~\stackrel{e_2}{(P|N|U)+}$
-%%\end{itemize}
-%}
-%\end{itemize}
-%\begin{itemize}
-%\item User defined semantic \\relations:
-%\begin{itemize}
-%\item $\langle e_1,e_2,r\rangle$
-%\item $\langle r,e_1,o_1\rangle$
-%\item $\langle r,e_2,o_2\rangle$
-%\end{itemize}
-%\end{itemize}
-%\columnbreak
-%\setcode{utf8}
-%\transfalse
-%\resizebox{0.9\columnwidth}{!}{
-%	\relsize{+1.5} \input{figures/ergraphEx.tex}
-%}
-%\transtrue
-%\setcode{standard}
-%\end{multicols}
-%\end{framed}
-%}
-%\caption{User-defined semantic relation example}
-%\label{fig:srEx}
-%\end{figure}
-
-\framework computes the relational entities in a user defined 
-relation $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
-$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$ 
-to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
-with the smallest nonzero positive distance between the source and the destination
-where the distance is the number of words between the matches.
-
-%\framework supports the following configurations for the subexpressions of $e_i$ and $e_j$:
-%\begin{itemize}
-%\item $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert=1$: Relate $e_i$ match to $e_j$ match with edge labelled by $r$.
-%\item $i=j\wedge\left\vert{e_i}\right\vert>1$: Relate sequences of $e_i$ matches with edges labelled by $r$.
-%\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert=1$: Relate each $e_i$ match to the single $e_j$ match with an edge labelled by $r$.
-The reverse applies for $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert>1$.
-%\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert>1$: Relate each $e_i$ match to each $e_j$ match with an edge labelled by $r$.
-%\end{itemize}
-%
-In Figure~\ref{fig:motiv}(b), \framework names the subexpressions 
-$(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, 
-as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
-The user defines the semantic relations
-$\langle e_1, e_2, r\rangle$, 
-$\langle r, e_1, o_1\rangle$, and 
-$\langle r,e_2,o_2\rangle$.
-
-The matches of $e_1,~e_2,~o_1,~o_2,$ and $r$ from the second match tree 
-of Figure~\ref{fig:motiv}(b) are \RL{dby mwl}(Dubai Mall), 
-\RL{Almbn_A}(the building), \RL{`l_A}(prep), \RL{mn h_dA}(from this), 
-and \RL{mqrbT}(near), respectively. 
-\framework~constructs the semantic relation matches and builds the lower part of the 
-entity-relation graph shown in Figure~\ref{fig:intromotiv}.
-
-For the first match \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'-wl},
-the matches of $e_1,~e_2,~o_2,$ and $r$ are 
-\RL{brj _hlyfT}(Khalifa tower), \RL{AltqA.t` Al-'-wl}(intersection 1), 
-\RL{mn}(from), and \RL{bAlqrb}(next to), respectively. 
-\framework~doesn't construct the relation $\langle r, e_1, o_1\rangle$ 
-since $o_1$ has no match. 
-Therefore, we get the upper part of the entity-relation graph shown 
-in Figure~\ref{fig:intromotiv}.
-
-After computing the relational entities, 
-\framework~computes a cross-reference relation between the extracted entities
-using a second order synonymy feature ($Syn^2$).
-The \cci{isA} edge in the 
-graph of Figure~\ref{fig:intromotiv} shows the cross-reference relation 
-between \RL{brj _hlyfT}(Khalifa tower) from the first match 
-with \RL{Almbn_A}(the building) from the second match.
+The user may inspect and refine the defined formulae, rules and relations and 
+the corresponding formulae tags, rule tags, entities and relational entities
+either using the GUI or directly through readable output files. 
+The files are in the javascript object notation (JSON)~\cite{nolan2014javascript} 
+format that is intuitive to read and modify. 
+\framework separates the user defined formulae, rules, actions and relations in
+a \framework tag type file and the 
+matching tags in a tags files. 
+The separation serves the user to apply the tag types to multiple case studies 
+and to obtain a separate resulting tags file for each case study.

### 2018-04-20 tex/merf/TALMajor/results.tex
done till 3.19

--- 2018-03-17 tex/merf/TALMajor/results.tex
@@
+% Table generated by Excel2LaTeX from sheet 'Sheet1'
+\begin{table}[tb!]
+  \centering
+  \caption{\framework compared to task specific applications.}
+  \label{tab:results}%
+  \resizebox{\columnwidth}{!}{
+    \begin{tabular}{l|l|l|l|ll|l}
+     \toprule
+      \multirow{2}{*}{Task} & Size & Development& Run & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
+      & (words) & time & time(s) & Recall & \multicolumn{1}{c|}{Precision} & \\
+    \midrule
+    %ANGE~\cite{ZaMaFlairs2012HadithBio}
+      ANGE & 1,865  
+    & 2 months & 1.79 & 0.99 & 0.99 & 3000+ lines of code\\
+      \framework &  & 3 hours & 7.24 & 0.99 & 0.93  & 8 MBFs and 4 MREs\\
+    \midrule
+    %ATEEMA~\cite{ZaMa2012IJCLATime}
+      ATEEMA & 1,677 & 1.5 months & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
+      \framework & & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
+    \midrule
+    %Genealogy tree~\cite{ZaMaHaCicling2012Entity} 
+      Genealogy tree & 1,227 
+    & 3 weeks & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
+      \framework &  & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
+    \midrule
+      NUMNORM &  1,399 & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
+      \framework &  & 1 hour & 1.53 & 0.91 & 0.90 & 3 MBFs/1 MRE/57 lines\\
+    \bottomrule
+    \end{tabular}%
+    }
+\vspace{-1em}
+\end{table}%
+
 In this section we compare \framework implementations of 
-the narrator chain, temporal entity, and genealogy entity extration 
+the narrator chain, temporal entity, and genealogy entity extraction 
 tasks to the task specific techniques proposed to solve them
@@
 a task specific implementation. 
+We evaluated ANGE with
+Musnad Ahmad, a hadith book, where we constructed an annotated 
+golden reference containing 1,865 words.
+We evaluated ATEEMA 
+with articles from issues of  the Lebanese 
+Al-Akhbar newspaper where we constructed 
+an annotated golden reference containing 1,677 words. 
+For the genealogical tree extraction we used 
+an extract from the Genesis biblical text with 1,227 words.
+Finally, we used an annotated article from the Lebanese Assafir
+newspaper with 1,399 words to evaluate the NUMNORM case study
+\footnote{available at \url{http://www.assafir.com} and 
+\url{http://www.al-akhbar.com}.}. 
+%
 In the online appendix%
@@
 , we report on eight additional \framework case studies.
+\major{
+Manual annotators inspected the outcome and provided corrections where tools 
+made mistakes.
+The corrections form the manual gold annotation that we compared against.
+}
 
@@
 for both the standalone task specific and the \framework implementations.
+\major{
+The development time measures the time required for 
+developing the case study.
+For instance, ANGE~\cite{ZaMaFlairs2012HadithBio} required two 
+months of development by a research assistant with six and 14 hours of 
+course work and teaching duties, respectively.
+}
+
+\major{
+Table~\ref{tab:results} provides runtime results of \framework 
+compared to the task specific implementations while running 
+MBF and MRE simulations jointly.
+This is a rough estimate of the complexity of the \framework simulator. 
+%
+The complexity of the MBF simulation is the total number of morphological 
+solutions for all the words multiplied by the number of user-defined MBFs.
+We do not provide a limit on the number of user defined formulae.
+In practice, we did not encounter more than ten formulae per case study.
+As for the complexity of MRE simulation, converting the rules into 
+non-deterministic finite state machines (NDFSM) is done once. 
+Simulating an NDFSM over the MBF tags is potentially exponential. 
+In practice all our case studies terminated within a predetermined 
+time bound of less than 30 minutes. 
+\framework required reasonably more runtime than the task specific 
+implementations and reported acceptable and 
+slightly less precision metrics with around
+the same recall.
+}
+
 %Table~\ref{tab:mbfer} reports the accuracy of the output MBF tags and the user-defined relation 
 %construction in \framework for each task.
-
-% Table generated by Excel2LaTeX from sheet 'Sheet1'
-\begin{table}[tb!]
-  \centering
-\caption{\framework compared to task specific applications.}
-  \resizebox{0.85\columnwidth}{!}{
-    \begin{tabular}{l|l|l|ll|l}
-     \toprule
-     \multirow{2}{*}{Task} & Development& Run & \multicolumn{2}{c|}{Accuracy} & \multirow{2}{*}{Ease of Composition}\\
-     & time & time(s) & Recall & \multicolumn{1}{c|}{Precision} & \\
-    \midrule
-    ANGE~\cite{ZaMaFlairs2012HadithBio}
-    & 2 months & 1.79 & 0.99 & 0.99 & 3000+ lines of code\\
-    \framework & 3 hours & 7.24 & 0.99 & 0.93  & 8 MBFs and 4 MREs\\
-    \midrule
-    ATEEMA~\cite{ZaMa2012IJCLATime}
-    & 1.5 months & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
-    \framework & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
-    \midrule
-    Genealogy tree~\cite{ZaMaHaCicling2012Entity} 
-    & 3 weeks & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
-    \framework & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
-    \midrule
-    NUMNORM & 1 week & 0.32 & 0.91 & 0.93  & 500 lines of code\\
-    \framework & 1 hour & 1.53 & 0.91 & 0.90 & 3 MBFs/1 MRE/57 lines\\
-    \bottomrule
-    \end{tabular}%
-    }
-  \label{tab:results}%
-\end{table}%
 
@@
 %Intuitively, precision denotes whether the system generated false positives.
-
-For the temporal and number normalization cases, 
-we evaluated the techniques against arbitrary text from issues of 
-the Lebanese Assafir and Al-Akhbar newspapers
-\footnote{available at \url{http://www.assafir.com} and 
-\url{http://www.al-akhbar.com}.}. 
-For the narrator chain case, we used
-Musnad Ahmad, a hadith book, for evaluation. 
-For the genealogical tree extraction we used 
-an extract from the Genesis biblical text.
 
@@
 
-\framework required reasonably more runtime than the task specific 
-implementations and reported acceptable and 
-slightly less precision metrics with around
-the same recall.
 %a runtime of 7.29, 1.53, 3.14, and 2.28 seconds for 
@@
 ANGE uses Arabic morphological analysis, finite state machines, and graph transformations 
-to extract named entities and relations including narrator chains~\cite{ZaMaFlairs2012HadithBio}.
+to extract entities and relations including narrator chains~\cite{ZaMaFlairs2012HadithBio}.
 
 \transfalse
+Table~\ref{tab:nchain} presents the MREs for the narrator chain case study. 
 MBF \cci{PN} checks the abstract category {\tt Name of Person}. 
@@
 
-Table~\ref{tab:nchain} presents the defined MREs. 
 MRE {\em name} is one or more \cci{PN} tags optionally followed 
@@
 %\transtrue
-
+%
 Table~\ref{tab:mbfer} shows that \framework detected almost all the MBF matches 
@@
 %The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
+\major{
 Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.
@@
 value (\cci{\$s1.number}) of literal numbers such as numbers from one to ten.
-
+}
+%
 Table~\ref{tab:mbfer} shows high accuracy in MBF tagging with 99\% recall and 99\% precision, and
@@
 \label{subsec:discuss}
+\label{sec:discuss}
 \input{discussion}

### 2018-04-20 tex/merf/TALMajor/tool.tex
done till 3.19

@@
-\input{MERFOverview}
-
-The \framework framework is illustrated in the flow diagram of
-Figure~\ref{f:overview}.
-The Arabic text and the reference tag chunks are the primary inputs to \framework. 
-Solutions, morphology-based Boolean formulae, tags, 
-morphology-based regular expressions, 
-tag chunks, relation and action definitions, and data structures expressing entities 
-and relations are input and output data to processes. 
-The morphological analyzer (Sarf), $Syn^k$ detector, 
-GUI for Boolean formulae definition, 
-visualization annotator, GUI for regular expression and action definition, 
-Boolean formula simulator, regular expression simulator, 
-relation extraction and action execution, and difference and statistical analyzer 
-are processes.
-
 %The user interacts with \framework~through a user-friendly interfaces to 
@@
 Also, let ${\cal F} = \{ \pp, \ss, \xx, \PP, \GG, \AC\}$ be the 
-set of morphological features where each entry is in turn a set of values for that feature.
-Given a word $w$, a morphological feature set $A\in{\cal F}$,}
+set of morphological features where each morphological feature $A\in {\cal F}$ is in 
+turn a set of morphological feature values.}
+Given a word $w$,
 a user defined constant feature value $CF\in A$,
@@
 
+\major{
 Consider the task we discussed in the introduction (Figure~\ref{fig:intromotiv}) 
@@
 The user specifies formula $R$ to denote ``relative position'' entities, 
-and defines it as a disjunction of MAT terms that check for 
+and defines it as a disjunction of terms that check for 
 solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).
@@
 \RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).
+}
 
@@
 \item $m$ is an MRE where $m$ is an MBF.
-\item $fg$ is an MRE denoting a concatenation operation
-  where $f$ and $g$ are both MRE expressions. This is 
-  satisfied by a match of $f$ followed by a match of $g$. 
-\item $f*,f+,f$\textasciicircum$x,$ and $f?$ are MRE expressions where $f$ is an MRE,
-  and are satisfied by zero or more matches, one of more matches, up to $x$ matches, 
-  and zero or only one match of $f$, respectively.
-\item $f\& g,$ (conjunction) and $f|g$ (disjunction) are  MRE expressions where 
-  $f$ and $g$ are MRE expressions, and are satisfied by a 
-  a match of both $f$ and $g$, 
-  and a match of either $f$ or $g$, respectively. 
+\item $fg$ is an MRE 
+  \major{
+  where $f$ and $g$ are both MRE expressions. 
+    A match of $f$ followed by a match of $g$ satisfies this concatenation operation.}
+\item $f*,f+,f$\textasciicircum$x,$ and $f?$ \major{ are MRE } where $f$ is an MRE,
+  and are satisfied by zero or more, \major{ one or }more, up to $x$ matches, 
+    and an \major{optional single match} of $f$, respectively.
+  \item $f\& g,$ (conjunction) and $f|g$ (disjunction) are \major{MRE} where 
+    $f$ and $g$ are \major{MRE}, and are satisfied by 
+    \major{the intersection of $f$ and $g$ matches, 
+    and the union of the $f$ and $g$ matches}, respectively. 
 \end{itemize}
@@
 
+\major{
 Back to the example in Figure~\ref{fig:intromotiv}.
@@
 %Note that the user reaches the satisfying matches by experimenting with the visualizer and the expression editor which do not require knowledge and expertise in regular expressions. 
+}
 
@@
 
-The match trees in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
-The first match tree refers to the text 
+\major{
+The \major{matching parse trees} in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
+The first tree refers to the text 
 \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection).
-The second match tree refers to the text 
+The second tree refers to the text 
 \RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
+}
 %The leaf nodes of the trees are entities and the edges and internal nodes are text, 
@@
 \major{
-The leaf nodes of the match trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
+The leaf nodes of the trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
 For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the sub-expression $(P|N)+$.
@@
 \subsection{User-defined relations and actions}
+\label{sec:relations-actions}
 
@@
 
+\major{
 We are interested in constructing the relational entity graph in Figure~\ref{fig:intromotiv}.
@@
 Note that there is no match to the optional $O$ formula in match 1.
-Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second match tree are
+Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second \major{matching parse tree} are
 \RL{dby mwl} (Dubai Mall), \RL{`l_A} (is located), \RL{mqrbT} (near), \RL{mn h_dA} (this), and \RL{Almbn_A} (building), respectively.
+}
 
+\major{
 Relation $\langle e_1,e_2,r\rangle$ creates the 
@@
 nodes in match 2.
-
-After the relation construction phase, we are interested to relate the discovered entities
-% and relational entities 
-that express the same concept.
-\framework provides the extended synonym feature of second order as a default cross-reference relation ($Syn^2$).
-%\framework provides the \cci{isA} predicate as a default cross-reference relation.
-In Figure~\ref{fig:intromotiv}, triggering this feature creates the edge between the nodes \cci{Khalifa Tower} and \cci{The building}.
-The edge is labeled with \cci{isA}.
+}
 
@@
 to process matches of sub-expressions.
-Each sub-expression can be associated with two computational actions: \cci{pre-match} and \cci{on-match}.
-Users can use these actions to compute statistical features, 
-store intermediate results, 
-or apply intelligent entity inference techniques
-as we show later in the numerical extraction example of Subsection~\ref{subsec:numnorm}.
+Each sub-expression can be associated with two computational 
+actions: \cci{pre-match} and \cci{on-match}.
+%
 \framework provides an API that enriches the actions with detailed access to
@@
 and morphological features.
-Once \framework computes all match trees, it traverses each tree to 
+\major{
+The API follows a decorator pattern in that it incrementally adds the 
+action results to the matching entities. 
+}
+Once \framework computes all \major{matching parse trees}, it traverses each tree to 
 execute the user defined \cci{pre-match} actions in pre-order manner
 and the \cci{on-match} actions in post-order manner.
+\major{
+This follows an observer pattern that notifies listeners with each
+produced match. 
+}
 
-\begin{comment}
-\subsection{\framework simulator}
+\subsection{\framework simulators}
 
@@
 
-%\subsection{MRE and action simulation}
-
 For each MRE, 
@@
 %\end{figure}
+%
+Next, \framework computes the relational entities corresponding to each user defined relation $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
+$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$.
 
-\framework computes the relational entities in a user defined 
-relation $R=\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
-$\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$ 
-to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
-with the smallest nonzero positive distance between the source and the destination
-where the distance is the number of words between the matches.
+\begin{comment} 
+Note that the number of matches for $e_i$ ($\ldbrack e_i \rdbrack$) depends on the corresponding sub-expression.
+For instance, we could have multiple matches if we associate $e_i$ with the sub-expression $P|N$ while we have $(P|N)+$.
+%to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
+%with the smallest nonzero positive distance between the source and the destination
+%where the distance is the number of words between the matches. 
+Hence, \framework supports the following configurations for $e_i$ and $e_j$ based on the corresponding matches of sub-expressions:
+\begin{itemize}
+\item $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert=1$: Relate $e_i$ match to $e_j$ match with edge labeled by $r$.
+\item $i=j\wedge\left\vert{e_i}\right\vert>1$: Relate sequences of $e_i$ matches with edges labeled by $r$.
+\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert=1$: Relate each $e_i$ match to the single $e_j$ match with an edge labeled by $r$.
+The reverse applies for $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert>1$.
+\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert>1$: Relate each $e_i$ match to each $e_j$ match with an edge labeled by $r$.
+\end{itemize}
+\end{comment}
 
-%\framework supports the following configurations for the subexpressions of $e_i$ and $e_j$:
-%\begin{itemize}
-%\item $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert=1$: Relate $e_i$ match to $e_j$ match with edge labelled by $r$.
-%\item $i=j\wedge\left\vert{e_i}\right\vert>1$: Relate sequences of $e_i$ matches with edges labelled by $r$.
-%\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert=1$: Relate each $e_i$ match to the single $e_j$ match with an edge labelled by $r$.
-The reverse applies for $\left\vert{e_i}\right\vert=1\wedge\left\vert{e_j}\right\vert>1$.
-%\item $i\neq j\wedge\left\vert{e_i}\right\vert>1\wedge\left\vert{e_j}\right\vert>1$: Relate each $e_i$ match to each $e_j$ match with an edge labelled by $r$.
-%\end{itemize}
-%
-In Figure~\ref{fig:motiv}(b), \framework names the subexpressions 
-$(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, 
-as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
-The user defines the semantic relations
-$\langle e_1, e_2, r\rangle$, 
-$\langle r, e_1, o_1\rangle$, and 
-$\langle r,e_2,o_2\rangle$.
+%In Figure~\ref{fig:taskMRE}, \framework names the sub-expressions $(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
+%The user defines the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
 
-The matches of $e_1,~e_2,~o_1,~o_2,$ and $r$ from the second match tree 
-of Figure~\ref{fig:motiv}(b) are \RL{dby mwl}(Dubai Mall), 
-\RL{Almbn_A}(the building), \RL{`l_A}(prep), \RL{mn h_dA}(from this), 
-and \RL{mqrbT}(near), respectively. 
-\framework~constructs the semantic relation matches and builds the lower part of the 
-entity-relation graph shown in Figure~\ref{fig:intromotiv}.
+%The matches of $e_1,~e_2,~o_1,~o_2,$ and $r$ from the second match tree of Figure~\ref{fig:motiv}(b) are \RL{dby mwl}(Dubai Mall), \RL{Almbn_A}(the building), \RL{`l_A}(prep), \RL{mn h_dA}(from this), and \RL{mqrbT}(near), respectively.
+%\framework~constructs the semantic relation matches and builds the lower part of the entity-relation graph shown in Figure~\ref{fig:intromotiv}.
 
-For the first match \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'-wl},
-the matches of $e_1,~e_2,~o_2,$ and $r$ are 
-\RL{brj _hlyfT}(Khalifa tower), \RL{AltqA.t` Al-'-wl}(intersection 1), 
-\RL{mn}(from), and \RL{bAlqrb}(next to), respectively. 
-\framework~doesn't construct the relation $\langle r, e_1, o_1\rangle$ 
-since $o_1$ has no match. 
-Therefore, we get the upper part of the entity-relation graph shown 
-in Figure~\ref{fig:intromotiv}.
+%For the first match \RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'-wl}, the matches of $e_1,~e_2,~o_2,$ and $r$ are \RL{brj _hlyfT}(Khalifa tower), \RL{AltqA.t` Al-'-wl}(intersection 1), \RL{mn}(from), and \RL{bAlqrb}(next to), respectively.
+%\framework doesn't construct the relation $\langle r, e_1, o_1\rangle$ since $o_1$ has no match. 
+%Therefore, we get the upper part of the entity-relation graph shown in Figure~\ref{fig:intromotiv}.
 
-After computing the relational entities, 
-\framework~computes a cross-reference relation between the extracted entities
-using a second order synonymy feature ($Syn^2$).
-The \cci{isA} edge in the 
-graph of Figure~\ref{fig:intromotiv} shows the cross-reference relation 
-between \RL{brj _hlyfT}(Khalifa tower) from the first match 
-with \RL{Almbn_A}(the building) from the second match.
-\end{comment}
+%The \cci{isA} edge in the graph of Figure~\ref{fig:intromotiv} shows the cross-reference relation between \RL{brj _hlyfT}(Khalifa tower) from the first match with \RL{Almbn_A}(the building) from the second match.

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer3.comments.tex
left with 3.3 3.20 3.24 3.27 and 4.3

@@
 
-\done{
-  DONE: Ameen please find where you fixed this and say it here. 
-}
-
 \item \label{Review.3.18}
@@
 \answer{
-We clarify the definition of MRE in the paper. 
-}
-
-\itodo{
-  Clarify the definition. 
+  We modified the definition. We also 
+  followed the MRE definition with a running illustrating example.
 }
@@
 \answer{
-We rewrote both sections... 
-}
-
-\itodo{
-  Figure our something to do/say.
+  We provided a running example across both sections and rewrote parts of the sections. 
 }
@@
 We provide running examples with the definitions....
-}
-\itodo{
-  Color them and say where they are. 
 }
@@
 
-\done{
-  Add that to the Results.\\
-  Added the above explanation in the Results section, needs to be checked.
-}
-
 \item \label{Review.3.22}
@@
 
-\done{
-  highlight the changes and refer to them please.\\
-  We use R first to denote the set of user-defined relations.
-  Then, we use it to denote the set of MBF tags which is contradicting.
-  I got rid of the first case as we don't need it.
-}
-
 \item \label{Review.3.23}
@@
 \answer{
-  We used XXX data structures... 
-}
-
-\itodo{
-  Ameen: JSON add that as a paragraph in the methodology section. 
+  We included descriptions of the storage format 
+  in the Methodology Section~\ref{sec:methodology}. 
 }
@@
 
-We clarify that in the beginning of the Results Section ~\ref{sec:results}.
+We clarify that in the beginning of the Results 
+Section ~\ref{sec:results}.
 }
@@
 
-\item \label{Review.3.24}
+\item \label{Review.3.25}
 Also, comparison with previous work done by (Zaraket ) 
-in different vesion seems like comparing with your previous work 
+in different version seems like comparing with your previous work 
 only and not with others. 
@@
 
-\item \label{Review.3.24}
+\item \label{Review.3.26}
 What is the data size of your evaluation? 
-I couldn't find how presion and recal was counted. 
+I couldn't find how precision and recall was counted. 
 you mentioned books but did you covered all the text in those 
 books ? and how long was it? you can define it in term of 
-number of relation extracted , tags added, expesion processed, 
+number of relation extracted , tags added, expression processed, 
 sentences,..etc
@@
   We include more information about the data and 
-  the evaluation process... 
+  the evaluation process.
+  We include the size of the annotated data sets that 
+  we used for accuracy evaluation in Table~\ref{tab:results}. 
+  We discuss how the golden reference was constructed. 
+  We also discuss the workload of the research assistant who built
+  the task specific case studies. 
 }
 
-\itodo{
-  Answer.. 
-}
-
-\item \label{Review.3.24}
+\item \label{Review.3.27}
 7- At section Discussion, I strongly urge you to discuss 
@@
 
-\item \label{Review.3.25}
+\item \label{Review.3.28}
 Clarity:  
@@
 
-\item \label{Review.3.25}
+\item \label{Review.3.29}
 Having ``Existing annotation and entity extraction tools'' 
@@
 \answer{
-  Fixed. CLS file was not compatible with arabtex package. 
+  We fixed the non-numbered subsections.
+  We removed Subsection ``Existing annotation and entity extraction tools''
+  as requested in \ref{Review.3.9}.
 }
 
-\item \label{Review.3.25}
+\item \label{Review.3.30}
 table 1 at page 7 while referring to it only at page 9. 
@@
 
-\done{
-  Done: Make sure this is moved. 
-}
-
-\item \label{Review.3.25}
+\item \label{Review.3.31}
 ``MERF regular expressions support operators such as concatenation, zero or one, zero or more, one or more, up to $M$” what is M? 
@@
 
-\item \label{Review.3.25}
+\item \label{Review.3.32}
 Correctness: 
@@
 
-\item \label{Review.3.25}
+\item \label{Review.3.33}
 2-``defines tag types and''.// unify the font here and 
@@
 \answer{
-  We use italics to emphasize important 
+  We used italics to emphasize important 
   concepts that we introduce for the first time. 
-  We reduced the use of emphasize in the paper. 
+  We reduced the use of emphasid text in the paper. 
 }
 
-\item \label{Review.3.25}
+\item \label{Review.3.34}
 Originality:  

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer4.comments.tex
left with 3.3 3.20 3.24 3.27 and 4.3

@@
 
-\done{
-  Discussion of the code actions is pending...
-}
-
-\item \label{Review.4.1} 
+\item \label{Review.4.2} 
 Correctness :  
@@
 
-\done{
-  Done: Use the same answer for Reviewer 3, and say why you used 7. 
-}
-
-\item \label{Review.4.1} 
+\item \label{Review.4.3} 
 In the results section, there is no standard evaluation. 
@@
 
-\itodo{
-Say something about clarifying the evaluation process. 
-}
-
 \end{enumerate}

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer3.comments.tex
done 25 pages

@@
 \answer{
-  Thank you. 
+  We will provide a GUI tutorial file on the tool's website. 
 }
@@
 \answer{
-We provide running examples with the definitions....
+We now provide running examples for the definitions in the 
+MEF simulator Section. 
 }
@@
 \answer{
-Our evaluation is a survey based evaluation.
-Users used \framework to define their \framework based annotators. 
-Users also built finite state transducer based annotators to 
-perform the same tasks on the same corpora. 
-We measured and reported the time needed by both.
 
-We also performed manual annotation to construct the correctness 
-benchmark.
+We address the comment by adding the following at the beginning
+of the Results Section~\ref{sec:results}.
+
+\begin{quote}
+In this section we evaluate \framework with four case studies. 
+We perform a survey like evaluation where developers manually 
+built task specific information extraction tools for the case studies 
+and other developers built equivalent \framework tools. 
+The aim of the comparison is to showcase that \framework enables 
+fast development of linguistic applications with similar accuracy 
+and a reasonable affordable overhead in computational time. 
+We report development time,
+size of developed code versus size of grammar,
+running time, and 
+precision-recall as metrics
+of cost, complexity, overhead, and accuracy,
+respectively. 
+\end{quote}
+
+Developers implemented the same tasks and ran on the same corpora. 
+%
+We also performed manual annotation to construct the golden reference. 
 We did that by correcting the automated annotations of both annotators
 for the benchmark corpora. 
-That alone took much more time than the time it took to develop and run
-the \framework annotators.
-
-We clarify that in the beginning of the Results 
-Section ~\ref{sec:results}.
-}
-
-\itodo{
-  Ameen: lets add the above to the section. 
+We clarify that in the Results Section~\ref{sec:results}.
 }
@@
 \answer{
+  We address this in the Results Section~\ref{sec:results}
+  as follows. 
+  \begin{itemize} 
+    \item 
   We include more information about the data and 
   the evaluation process.
+\item 
   We include the size of the annotated data sets that 
   we used for accuracy evaluation in Table~\ref{tab:results}. 
-  We discuss how the golden reference was constructed. 
+\item 
+  We now discuss how the golden reference was constructed. 
+\item 
+  We now discuss how precision and recall are computed as ratios of 
+  entities correctly detected against extracted entities, and 
+  entities correctly detected against the total number of entities, 
+  respectively. 
+\item 
   We also discuss the workload of the research assistant who built
   the task specific case studies. 
+  \end{itemize} 
 }
@@
 \answer{
-We extend the discussion section to discuss ... 
+  We added  the following to the discussion Section. 
+  \begin{quote}
+We notice that ANGE, ATEEMA, and Genealogy tree report higher 
+    precision than \framework. 
+This is mainly due to their capacity 
+to learn words and relations that may not have a match in the 
+morphological analyzer based on co-occurence relations. 
+For example, the sequence $p_1 t_1 p_2$ where $p_1$ and 
+$p_2$ are persons and $t_1$ is a tell relationship helps
+indicate that $x$ is a tell relationship in $p_1 x p_2$ 
+even if the morphological analyzer did not return the required
+feature for $x$ to match a tell relationship. 
+\framework does not have that capacity yet unless it is
+encoded in the C++ actions. 
+  \end{quote} 
 }

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer4.comments.tex
done 25 pages

@@
 task-specific tools.
+
+\answer{
+  We modified the Results Section~\ref{sec:results} to 
+    clarify our evaluation method as follows.
+    \begin{itemize}
+      \item We compare two different approaches trying to achieve the 
+        same results using the same input corpora and the same
+        golden reference. 
+      \item We use standard metrics to report the results: development
+        time, size of solution, runtime, precision and recall.
+      \item We use the reported task-specific implementations because, 
+        up to our knowledge, no other accessible relational entity 
+        extraction tools exist for Arabic. 
+    \end{itemize} 
+}
+
+\item \label{Review.4.4} 
 Also, the reported precision and recall are very sensitive 
@@
 \answer{
-  Same answer to reviewer 3 results... 
+  This is correct. We added a precision discussion paragraph
+    to reflect that in the Discussion Section~\ref{sec:discuss}. 
 }

### 2018-04-20 tex/merf/TALMajor/discussion.tex
done till 3.19
done 25 pages

@@
 and fix the annotations.
+%
+\major{
+The case studies showed that 
+\framework requires some linguistic expertise to successfully execute the 
+tasks. 
+In contrast, the case specific implementations require more sophisticated 
+linguistic and programming expertise to attain similar results.
+
+We notice that ANGE, ATEEMA, and Genealogy tree report higher precision than \framework. 
+This is mainly due to their capacity 
+to learn words and relations that may not have a match in the 
+morphological analyzer based on co-occurence relations. 
+For example, the sequence $p_1 t_1 p_2$ where $p_1$ and 
+$p_2$ are persons and $t_1$ is a tell relationship helps
+indicate that $x$ is a tell relationship in $p_1 x p_2$ 
+even if the morphological analyzer did not return the required
+feature for $x$ to match a tell relationship. 
+\framework does not have that capacity yet unless it is
+encoded in the C++ actions. 
+}
 

### 2018-04-20 tex/merf/TALMajor/introduction.tex
left with 3.3 3.20 3.24 3.27 and 4.3
done 25 pages

@@
 
-This work significantly extends the position paper~\cite{JaZaMatar} 
-that allows for manual, and basic morphology annotation.
-\framework enables a regular user to incrementally create 
-complex annotations for Arabic text based on automatic 
+This work significantly extends~\cite{JaZaMatar} 
+that allows for manual, and morphology annotation.
+\framework enables a user to incrementally create 
+complex annotations for Arabic based on automatic 
 extraction of morphological tags through a user friendly interactive interface. 
@@
 %in Section~\ref{sec:preliminaries}. 
-Section~\ref{sec:framework} presents the \framework components.
-Section~\ref{sec:gui} presents the visual and user friendly interface of \framework. 
-Section~\ref{sec:methodology} explains the methodology used to design \framework.
+Section~\ref{sec:methodology} explains the methodology of \framework.
+Section~\ref{sec:framework} presents \framework components.
+Section~\ref{sec:gui} presents \framework GUI. 
 Section~\ref{sec:related} presents and discusses related work.

### 2018-04-20 tex/merf/TALMajor/methodology.tex
done 25 pages

@@
 \framework methodology. 
-The first process takes the Arabic text input and provides the user 
+The first process takes Arabic text and provides the user 
 with a morphology-based Boolean (MB) formulae GUI. 
@@
 
-After the relation construction phase, we are interested to relate the 
-discovered entities that express the same concept.
+After relation extraction, we are interested to relate 
+entities that express the same concept.
 \framework provides the extended synonym feature of second order as 
@@
 The separation serves the user to apply the tag types to multiple case studies 
-and to obtain a separate resulting tags file for each case study.
+and to obtain a separate file of resulting tags for each.

### 2018-04-20 tex/merf/TALMajor/results.tex
done 25 pages

@@
+In this section we evaluate \framework with four case studies. 
+We perform a survey like evaluation where developers manually 
+built task specific information extraction tools for the case studies 
+and other developers built equivalent \framework tools. 
+The aim of the comparison is to showcase that \framework enables 
+fast development of linguistic applications with similar accuracy 
+and a reasonable affordable overhead in computational time. 
+We report development time,
+size of developed code versus size of grammar,
+running time, and 
+precision-recall as metrics
+of cost, complexity, overhead, and accuracy,
+respectively.
+
+We survey three case studies from the literature: 
+(1) narrator chain, (2) temporal entity, and (3) genealogy entity extraction 
+tasks and we use the reported development time for the task specific 
+techniques proposed in
+ANGE~\cite{ZaMaFlairs2012HadithBio}, 
+ATEEMA~\cite{ZaMa2012IJCLATime},  and
+GENTREE~\cite{ZaMaHaCicling2012Entity}, respectively. 
+%
+We also compare a \framework number normalization task to 
+a task specific implementation.
+
+We evaluated ANGE with
+Musnad Ahmad, a hadith book, where we constructed an annotated 
+golden reference containing 1,865 words.
+We evaluated ATEEMA 
+with articles from issues of  the Lebanese 
+Al-Akhbar newspaper where we constructed 
+an annotated golden reference containing 1,677 words. 
+For the genealogical tree extraction we used 
+an extract from the Genesis biblical text with 1,227 words.
+Finally, we used an annotated article from the Lebanese Assafir
+newspaper with 1,399 words to evaluate the NUMNORM case study
+\footnote{available at \url{http://www.assafir.com} and 
+\url{http://www.al-akhbar.com}.}. 
+%
+In the online appendix%
+~\footnote{available at ~\url{-ommited-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
+, we report on eight additional \framework case studies.
+\major{
+Manual annotators inspected the outcome and provided corrections where tools 
+made mistakes.
+The corrections form the manual gold annotation that we compared against.
+}
+
 % Table generated by Excel2LaTeX from sheet 'Sheet1'
@@
 
-In this section we compare \framework implementations of 
-the narrator chain, temporal entity, and genealogy entity extraction 
-tasks to the task specific techniques proposed to solve them
-in ANGE~\cite{ZaMaFlairs2012HadithBio}, 
-ATEEMA~\cite{ZaMa2012IJCLATime},  and
-GENTREE~\cite{ZaMaHaCicling2012Entity}, respectively. 
-We also compare a \framework number normalization task to 
-a task specific implementation. 
-We evaluated ANGE with
-Musnad Ahmad, a hadith book, where we constructed an annotated 
-golden reference containing 1,865 words.
-We evaluated ATEEMA 
-with articles from issues of  the Lebanese 
-Al-Akhbar newspaper where we constructed 
-an annotated golden reference containing 1,677 words. 
-For the genealogical tree extraction we used 
-an extract from the Genesis biblical text with 1,227 words.
-Finally, we used an annotated article from the Lebanese Assafir
-newspaper with 1,399 words to evaluate the NUMNORM case study
-\footnote{available at \url{http://www.assafir.com} and 
-\url{http://www.al-akhbar.com}.}. 
-%
-In the online appendix%
-~\footnote{available at ~\url{-ommited-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
-, we report on eight additional \framework case studies.
-\major{
-Manual annotators inspected the outcome and provided corrections where tools 
-made mistakes.
-The corrections form the manual gold annotation that we compared against.
-}
-
 Table~\ref{tab:results} reports the development time,
@@
 course work and teaching duties, respectively.
+Recall refers to the fraction of the entities correctly detected against the
+total number of entities. 
+Precision refers to the fraction of correctly
+detected entities against the total number of extracted entities. 
 }
@@
   \label{tab:nchain}%
+  \vspace{-1.5em}
 \end{table}%
@@
 
-For brevity, we omit the details of \framework temporal entity extraction, 
-genealogy tree, and number normalization case studies, describe them shortly
-below and provide a full description in the online Appendix.
+%For brevity, we omit the details of \framework temporal entity extraction, 
+%genealogy tree, and number normalization case studies, describe them shortly
+%below and provide a full description in the online Appendix.
 
@@
   \label{tab:mbfer}%
+  \vspace{-1.5em}
 \end{table}%
 
+\vspace{-1.5em}
+\subsection{Temporal entity extraction}
 \vspace{-1em}
-\subsection{Temporal entity extraction}
 
@@
 morphological analysis and finite state transducers~\cite{ZaMa2012IJCLATime}. 
-
+%
 Table~\ref{tab:mbfer} shows that \framework detected almost all the MBF matches with 99\% recall, 
@@
 
+\vspace{-2em}
+\subsection{Genealogy tree}
 \vspace{-1em}
-\subsection{Genealogy tree}
 
@@
 meaning ``and Haran became the father of Lot''.
-
+%
 GENTREE~\cite{ZaMaHaCicling2012Entity} 
@@
 MBF matches with 99\% recall, and 75\% precision, and
-extracted user-defined relations with 81\% recall and 96\% precision.
+extracted relations with 81\% recall and 96\% precision.
 
@@
 \centering
-  \begin{tabular}{p{5.5cm}p{5.5cm}}
+  \begin{tabular}{p{5.5cm}p{5.8cm}}
 \begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=TMB algorithm,commandchars=\\\[\]] 
@@
 } else if(current == 0) {
-
   current = \fvtextcolor[red][$s1.number];
@@
 } else if(!isKey) {
-
   isKey = true;
@@
   previous += current;
-  current = \fvtextcolor[red][$s1.number];
-}
+  current = \fvtextcolor[red][$s1.number];}
 \end{Verbatim}
@@
 \begin{Verbatim}[fontsize=\relsize{-1},frame=single,label=DT algorithm,commandchars=\\\[\]] 
-if(isHundred) {
-  currentH += \fvtextcolor[red][$s0.number];
+if(isHundred) {currentH += \fvtextcolor[red][$s0.number];
 } else if(current == 0) {
@@
   current = \fvtextcolor[red][$s0.number];
-} else {
-  current += \fvtextcolor[red][$s0.number]; }
+} else {current += \fvtextcolor[red][$s0.number]; }
 isKey = false;
@@
   current = 0;
-} else {
-  currentH = \fvtextcolor[red][$s2.number];}
+} else {currentH = \fvtextcolor[red][$s2.number];}
 isKey = false;
@@
 \end{tabular}
+  \vspace{-2em}
 \caption{Actions for TMB, DT, and H MRE expressions.}
+  \vspace{-1.5em}
 \label{fig:numnormalgo}
@@
 %
-Table~\ref{tab:mbfer} shows high accuracy in MBF tagging with 99\% recall and 99\% precision, and
-high accuracy in relation extraction with 97\% recall and 95\% precision.
+Table~\ref{tab:mbfer} shows high accuracy in MBF tagging and relation extraction 
+with 99\% and 97\% recall and 99\% and 95\% precision, respectively.
 
-\vspace{-1em}
+\vspace{-1.5em}
 \subsection{Discussion}

### 2018-04-20 tex/merf/TALMajor/tool.tex
done 25 pages

@@
 with font and color information.
+\major{
+For the example of Figure~\ref{fig:intromotiv},
+$l$ is ``direction'',
+$f$ is $(P|N)\!+~O?~R~O^\wedge 2~(P|N|U)+$,
+and $d$ is italic. 
+}
 
-For each word $t_i\in T, 0\le i < M, M=|T|$.
+For each word $t_i\in T, 0\le i < |T|$.
 \framework computes a Boolean value 
-($\{\mathit{true}, \mathit{false}\}$)
 for all MBFs. 
+\major{
+For example, \RL{brj} matches MBF $P$. 
+}
 Then, it computes the set of MBF tags
@@
 its tag set $R_i$ is by default the singleton $O=\{\mathit{NONE}\}$.
-%and $t_i$ is referred to as a {\em null} word.
+\major{
+For example, $R$ for the text in Figure~\ref{f:overview} follows 
+$\{ \{NONE\}, \{NONE\}, \{NONE\}, \{NONE\}, \{($\RL{brj}$,P)\}, \{($\RL{xlyfT}$,N)\},\ldots\}$. 
+}
 
@@
 corresponding expression $(P|N)\!+~O?~R~O^\wedge 2~(P|N|U)+$. 
+\major{
 Figure~\ref{fig:nfaEx} shows part of the corresponding NFA where
@@
 $P,$ and $N$.
-Edges labeled with the empty string $\epsilon$ are non-deterministic.
+Edges labeled with the empty string $\epsilon$ are non-deterministic.}
 %The expression uses operations $|$, $?$, $+$, and $\wedge$ to relate places {\tt P}, names of persons {\tt N}, relative positions {\tt R}, numerical terms {\tt U}, and other {\tt O}.

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer1.comments.tex
done spell checking

@@
 objects manipulated by the software. 
-This make things rigouros but the notations are sometimes heavy.
+This make things rigorous but the notations are sometimes heavy.
 
@@
 with the other system (one of the tasks needed 2 month of work 
-!! is it really 2 acutal months of work or is it the period of 
+!! is it really 2 actual months of work or is it the period of 
 time during which the development has been done)
@@
 you say nothing about the availability of the system, 
-is it distributed ? free ? what kind of licence ?
+is it distributed ? free ? what kind of license ?
 

### 2018-04-20 tex/merf/TALMajor/answer/Reviewer3.comments.tex
done spell checking

@@
 used as a tool for other researchers in the field to help 
-improving the Arabic language computations and processings.
+improving the Arabic language computations and processing.
 
@@
 but allow the users to correct and fix the tagging. 
-In additions it extract entitiy and relational entityies from 
+In additions it extract entity and relational entities from 
 the input document.
@@
 2- ``How to use MERF?''
-(you can have more explanations in your appendex and could have 
+(you can have more explanations in your appendix and could have 
 ``help'' button in your API. This part was also relatively 
@@
 specified example. 
-Section 2 seemed not very imporatnt as stand alone section.
+Section 2 seemed not very important as stand alone section.
 
@@
 to learn words and relations that may not have a match in the 
-morphological analyzer based on co-occurence relations. 
+morphological analyzer based on co-occurrence relations. 
 For example, the sequence $p_1 t_1 p_2$ where $p_1$ and 
@@
 ``Methodology part''. 
-Section 4 ``MERF'' please add ``Methedology''.
+Section 4 ``MERF'' please add ``Methodology''.
 
@@
   \begin{quote}
-   up to $M$ repititions where $M$ is 
+   up to $M$ repetitions where $M$ is 
    a non-zero positive integer 
@@
   concepts that we introduce for the first time. 
-  We reduced the use of emphasid text in the paper. 
+  We reduced the use of emphasized text in the paper. 
 }

### 2018-04-20 tex/merf/TALMajor/introduction.tex
done spell checking

@@
 \framework regular expressions support operators such as concatenation, 
-zero or one, zero or more, one or more, up to $M$ \major{repititions where $M$ is 
+zero or one, zero or more, one or more, up to $M$ \major{repetitions where $M$ is 
 a non-zero positive integer}, and logical conjunction and disjunction.
@@
     \major{
-    novel ligh Arabic wordnet relation that leverages the 
+    novel light Arabic wordnet relation that leverages the 
     synonym $Syn^k$ feature.

### 2018-04-20 tex/merf/TALMajor/relatedwork.tex
done spell checking

@@
 \begin{table}[tb!]
-\caption{Comaprison of \framework with SystemT, 
+\caption{Comparison of \framework with SystemT, 
 TEXTMARKER, 
@@
 %The work in~
-\cite{ZaMaHaCicling2012Entity} present a method for extracting entities and events, and relations amongst them from Arabic text using a hierarchy of manually built 
+\cite{ZaMaHaCicling2012Entity} present a method for extracting entities and events, and relations among them from Arabic text using a hierarchy of manually built 
 finite state machines driven by morphological features, 
@@
 implemented as a Prot\'eg\'e~\cite{gennari2003evolution} plug-in. 
-Prot\'eg\'e is an open-source platform with a suite of tools to construct domain models and knowledge-based applications with ontologies. 
+Prot\'eg\'e is an open-source platform with a suite of tools to construct domain models and knowledge-based applications with ontology. 
 However, it doesn't support the Arabic language.
@@
 \cite{ZaMaHaCicling2012Entity} presents a method for extracting entities, 
-events, and relations amongst them from Arabic text using a hierarchy of manually built 
+events, and relations among them from Arabic text using a hierarchy of manually built 
 finite state machines driven by morphological features and graph 

### 2018-04-20 tex/merf/TALMajor/results.tex
done spell checking

@@
 In the online appendix%
-~\footnote{available at ~\url{-ommited-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
+~\footnote{available at ~\url{-omitted-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
 , we report on eight additional \framework case studies.
@@
 We define the relation $\langle s_2,s_2',s_1\rangle$ 
-to relate sequences of narrators with edges labelled by the tags of \cci{TOLD} where 
+to relate sequences of narrators with edges labeled by the tags of \cci{TOLD} where 
 $s_2'$ denotes the next match of \cci{nar} in the one or more MRE subexpression.

### 2018-04-20 tex/merf/TALMajor/methodology.tex
minor fixes

@@
 In Figure~\ref{fig:intromotiv}, triggering this feature creates the 
-edge labels with \cci{isSyn} 
+edge labeled with \cci{isSyn} 
 between the nodes \cci{Khalifa Tower} and \cci{The building}.

### 2018-04-20 tex/merf/TALMajor/results.tex
minor fixes

@@
+\major{
 In this section we evaluate \framework with four case studies. 
@@
 (1) narrator chain, (2) temporal entity, and (3) genealogy entity extraction 
-tasks and we use the reported development time for the task specific 
+tasks, and we use the reported development time for the task specific 
 techniques proposed in
@@
 , we report on eight additional \framework case studies.
-\major{
 Manual annotators inspected the outcome and provided corrections where tools 

### 2018-04-20 tex/merf/TALMajor/sarf.tex
done till 3.19
minor fixes

@@
 NLP
-due its to the exceptional degree of ambiguity in writing, 
+due to the exceptional degree of ambiguity in writing, 
 the rich morphology, and the complex word derivation 
@@
 category tags of suffix $x_i$. 
-$P_s, G_s,$ and $C_s$ are 
 $P_s, G_s,$ and $C_s$ are the POS, Gloss and 
@@
   \caption{Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
+  \label{tab:samplesolution}%
   \resizebox{0.7\columnwidth}{!}{
@@
     }
-  \label{tab:samplesolution}%
+  \vspace{-2em}
 \end{table}%
@@
 an imperfect verb, and 
-and a third person feminine singular object pronoun, respectively.
+a third person feminine singular object pronoun, respectively.
 The POS and gloss notations follow the Buckwalter notation~\cite{Buckwalter:02}.

### 2018-04-20 tex/merf/TALMajor/tool.tex
minor fixes

@@
 $Sy^{1}(w) = Sy(w)$ and
-$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\phi\}$.
+$Sy^{i+1}(w)=\{u\mid u\in S\land\exists s\in Sy^{i}(w)\land~\alpha(u)\cap\alpha(s)\neq\emptyset\}$.
 Formally, $Syn^k(w) = \bigcup\limits_{i=1}^{k} Sy^{i}(w)$ for $i\in[1 \ldots k]$.
@@
 The user denotes the ``name of person'' entities with formula $N$ 
-which requires the {\em category} feature in the morphological solution of a word to be {\tt Name\_of \_Person}.
+which requires the {\em category} feature in the morphological solution of a word to be {\tt Name\_of\_Person}.
 The entities $n_1$, $n_2$, and $n_3$ are matches of the formula $N$ in the text.
@@
 $e_1,e_2,$ and $r$ are identifiers associated with 
-sub-expressions of an expression $f$.
+sub-expressions of an MRE $f$.
 Matches of the relation are a set of labeled binary edges
@@
 %We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
-We define the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
 The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (Khalifa Tower), \RL{bAlqrb} (next), \RL{mn} (to), and \RL{AltqA.t` Al-'wl} (first intersection).
@@
 \major{
+We define the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
 Relation $\langle e_1,e_2,r\rangle$ creates the 
@@
 \major{
-For example, $R$ for the text in Figure~\ref{f:overview} follows 
+For example, the tag sets for the text in Figure~\ref{f:overview} follows 
 $\{ \{NONE\}, \{NONE\}, \{NONE\}, \{NONE\}, \{($\RL{brj}$,P)\}, \{($\RL{xlyfT}$,N)\},\ldots\}$. 
@@
 %A simulation match $m$ of an expression $f$ is a tree where the root is the MRE expression, the internal nodes are the MRE and MBF operations, and the leaves are matches of the MAT terms of $f$.
-A simulation match $m$ of an expression $f$ is a tree where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g.
+A simulation match $m$ of an expression $f$ is a \major{parse tree} where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g.
 Fig.~\ref{fig:taskMRE}.
@@
 %
-Next, \framework computes the relational entities corresponding to each user defined relation $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
+Finally, \framework computes the relational entities corresponding to each user defined relation $\ldbrack \langle e_1,e_2,r\rangle \rdbrack$
 $\subseteq \ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$.

### 2018-06-22 tex/merf/TALMajor/abstract.tex
format issues ficed: figure and table caption style

@@
 They lack support of Arabic {\em morphological analysis} which is key to 
-process Arabic text.
-}
+process Arabic text.}
 In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.

### 2018-06-22 tex/merf/TALMajor/macros.tex
changes to answer, colors, and some reviewers addressed
format issues ficed: figure and table caption style

@@
 
-%%%%%%%% packages %%%%%%%%%%
+%%%%%%%%% packages %%%%%%%%%%
 \usepackage{xspace}
@@
 \usepackage{tikz}
-\usepackage{hyperref}
-\hypersetup{
-    colorlinks=false,
-    pdfborder={0 0 0},
-}
+\usepackage{url}
+%\usepackage{hyperref}
+%\hypersetup{
+%    colorlinks=false,
+%    pdfborder={0 0 0},
+%}
 \usepackage{booktabs}
 \usepackage[ruled]{algorithm2e}
-\usepackage{etex}
+%\usepackage{etex}
 \usepackage{graphicx}
@@
 \usepackage{pbox}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%% Macros %%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%%%%%%%%%%% Macros %%%%%%%%%%
+\newcommand{\major}[1]{{\color{black}#1}}
 \def\framework{\textsc{MERF}\xspace}
@@
 \hyphenation{op-tical net-works semi-conduc-tor}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\author{}
-%\author{Amin A. Jaber \andauthor Fadi A. Zaraket}
+\author{Amin Jaber \andauthor Fadi Zaraket}
 
-\address{}
-%\address{American University of Beirut, Beirut 1107 2020, Lebanon}
+\address{American University of Beirut, Beirut 1107 2020, Lebanon}
 
@@
 \motscles{
-%mot-cl\'e$_1$,
-%mot-cl\'e$_2$,
-%mot-cl\'e$_3$.
+Arabe,
+extraction d'information,
+traitement du langage naturel,
+entité,
+entité relationnelle,
+visualisation d'information,
+marquage.
 }
@@
 \resume{
-%Le r\'esum\'e est \`a \'ecrire ici.
+Les techniques et outils basés sur des règles pour extraire des entités et des entités relationnelles des documents permettent aux utilisateurs de spécifier les entités souhaitées en utilisant des questions de langage naturel, des automates à états finis, des expressions régulières, des instructions de langage de requête structurées ou des scripts propriétaires.
+Ces techniques et outils nécessitent une expertise en linguistique et en programmation.
+Ils manquent de soutien de l'analyse morphologique arabe qui est la clé de
+traiter le texte arabe.
+Dans ce travail, nous présentons \framework; une entité morphologique et un cadre d'extraction d'entité relationnelle pour le texte arabe.
+\framework fournit une interface conviviale où l'utilisateur, avec une connaissance de base des caractéristiques linguistiques et des expressions régulières, définit les types de balises et les associe de manière interactive aux expressions régulières définies sur des formules booléennes.
+Les formules booléennes couvrent des correspondances d'entités morphologiques arabes et des caractéristiques de synonymie.
+Les utilisateurs définissent des relations avec des tuples de correspondance de sous-expression et peuvent associer des actions de code à des sous-expressions.
+\framework calcule des correspondances de fonctions, des correspondances d'expressions régulières et construit des entités et des entités relationnelles à partir des relations définies par l'utilisateur.
+Nous avons évalué notre travail avec plusieurs études de cas et les avons comparées avec des techniques spécifiques à une application existante.
+Les résultats montrent que \framework nécessite un temps et un effort de développement plus courts par rapport aux techniques existantes et produit des résultats raisonnablement précis dans une surcharge raisonnable en temps d'exécution.
 }

### 2018-06-22 tex/merf/TALMajor/methodology.tex
format issues ficed: figure and table caption style

@@
-Figure~\ref{f:overview} illustrates the four processes involved in 
-\framework methodology. 
+\input{MERFOverview}
+
+Figure~\ref{f:overview} illustrates the four processes involved in \framework methodology. 
 The first process takes Arabic text and provides the user 
@@
 
-\input{MERFOverview}
-
-The user may inspect and refine the defined formulae, rules and relations and 
+The user may refine the defined formulae, rules and relations and 
 the corresponding formulae tags, rule tags, entities and relational entities

### 2018-06-22 tex/merf/TALMajor/relatedwork.tex
format issues ficed: figure and table caption style

@@
 
-{\bf Information Extraction.~}
+{\bf Information Extraction.}
 The common pattern specification language (CPSL) targets system independent IE specifications~\cite{appelt1998common}. 
@@
 
-\begin{table}[tb!]
-\caption{Comparison of \framework with SystemT, 
-TEXTMARKER, 
-Urbain, 
-QARAB}
+\begin{table}[tb]
 \resizebox{\columnwidth}{!}{
@@
 }
-\label{tab:iecomp}
+\caption{\label{tab:iecomp}Comparison of \framework with SystemT, 
+TEXTMARKER, 
+Urbain, 
+QARAB}
 \end{table}
@@
 
-{\bf Annotation tools.~}
+{\bf Annotation tools.}
 MMAX2 is a manual multi-level linguistic annotation tool with an XML 

### 2018-06-22 tex/merf/TALMajor/tool.tex
format issues ficed: figure and table caption style

@@
 
+\begin{figure}[tb]
+\setcode{utf8}
+\begin{center}
+  \resizebox{0.75\columnwidth}{!}{ 
+  	{\relsize{-2} \input{figures/syn2ex.tex}} }
+\setcode{standard}
+\vspace{-3em}
+\caption{$Syn^2($\RL{mA'}$)$}
+\vspace{-2em}
+\label{fig:introsynEx}
+\end{center}
+\end{figure}
+
 Given a word $w\in L$, 
@@
 
-\begin{figure}[tb!]
-\setcode{utf8}
-\begin{center}
-  \resizebox{0.7\columnwidth}{!}{ 
-  	{\relsize{-2} \input{figures/syn2ex.tex}} }
-\setcode{standard}
-\vspace{-2.5em}
-\caption{$Syn^2($\RL{mA'}$)$}
-\vspace{-3em}
-\label{fig:introsynEx}
-\end{center}
-\end{figure}
-
+\vspace{-1em}
 \subsection{MRE: Morphology-based regular expressions}
@@
   Informally, this checks if $w$ is an extended synonym of a stem $CF$. 
-\major{
-  We limit $k$ to a maximum of $7$ since we practically noticed that 
+\major{We limit $k$ to a maximum of $7$ since we practically noticed that 
   (1) values above $7$ introduce significant semantic noise and
-  (2) the computation is expensive without a bound. 
-}
+  (2) the computation is expensive without a bound.}
 \end{itemize}
@@
 and recall that we are interested in identifying names of people, names of places, relative positions, and numerical terms.
-The table in Figure~\ref{fig:taskMBF} presents the defined formulae.
+Table~\ref{tab:taskMBF} presents the defined formulae.
 The user denotes the ``name of person'' entities with formula $N$ 
@@
 Back to the example in Figure~\ref{fig:intromotiv}.
-We use the formulae defined in Figure~\ref{fig:taskMBF} to construct an MRE such as
+We use the formulae defined in Table~\ref{tab:taskMBF} to construct an MRE such as
 $(P|N)\!+O?~R~O^\wedge\!2~(P|N|U)+$ where 
@@
 
+\vspace{-2em}
 \subsection{User-defined relations and actions}
@@
 
+\vspace{-1em}
 \subsection{\framework simulators}
@@
 }
+
+\setcode{utf8}
+\setarab
+\transfalse
+\begin{figure}[tb]
+\centering
+\resizebox{0.75\columnwidth}{!}{
+	\relsize{-1} \input{figures/reNFA.tex}
+}
+\vspace{-1em}
+  \caption{\label{fig:nfaEx}Equivalent NFA of direction expression.}
+  \vspace{-1em}
+\end{figure}
+\transtrue
+\setcode{standard}
 
@@
 %It also uses $\phi$ and the match sequence to compute the sequence of computational actions of an MRE match.
-
-\setcode{utf8}
-\setarab
-\transfalse
-
-\begin{figure}[tb!]
-\centering
-\resizebox{0.7\columnwidth}{!}{
-	\relsize{-1} \input{figures/reNFA.tex}
-}
-  \caption{Equivalent NFA of direction expression}
-  \label{fig:nfaEx}
-\end{figure}
-\transtrue
-\setcode{standard}
 

### 2018-06-23 tex/merf/TALMajor/abstract.tex
minor changes

@@
 These techniques and tools require expertise in linguistics and programming.
-\major{
-They lack support of Arabic {\em morphological analysis} which is key to 
-process Arabic text.}
+They lack support of Arabic morphological analysis which is key to 
+process Arabic text.
 In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.

### 2018-06-23 tex/merf/TALMajor/directionFigure.tex
done till 3.19
format issues ficed: figure and table caption style
minor changes

@@
 \resizebox{\columnwidth}{!}{
-    \begin{tabular}{|C{9.5cm}C{8cm}|} 
+    \begin{tabular}{|C{10cm}C{8cm}|} 
       \hline
-  \begin{tabular}{R{9.5cm}}
-  \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من}
+  \begin{tabular}{R{10cm}}
+\RL{تقود} \RL{وأنت} $\stackrel{u1}{\framebox[1.2\width]{\RL{الأول}}}$ $\stackrel{p_2}{\framebox[1.2\width]{\RL{التقاطع}}}$ \RL{من} $\stackrel{r_1}{\framebox[1.2\width]{\RL{بالقرب}}}$ $\stackrel{n_1}{\framebox[1.2\width]{\RL{خليفة}}}$ $\stackrel{p_1}{\framebox[1.2\width]{\RL{برج}}}$ \RL{تلاحظ} \RL{ألا} \RL{المستحيل} \RL{من}
   \\ ~ \\
-	$\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك} \RL{تقود}
+$\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي} $\stackrel{u_2}{\framebox[1.2\width]{\RL{الأولى}}}$ \RL{المرّة} \RL{هذه} \RL{كانت} \RL{وإن} \RL{حتّى} $\stackrel{n_3}{\framebox[1.2\width]{\RL{زايد،}}}$ $\stackrel{n_2}{\framebox[1.2\width]{\RL{الشيخ}}}$ $\stackrel{p_3}{\framebox[1.2\width]{\RL{شارع}}}$ $\stackrel{r_2}{\framebox[1.2\width]{\RL{في}}}$ \RL{سيارتك}
   \\ ~ \\
-  $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا} $\stackrel{r_3}{\framebox[1.2\width]{\RL{فيها}}}$ \RL{تسلك} \RL{التي}
-  \\
-  \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  
-  \\
-  \multicolumn{1}{L{9.5cm}}{
+\RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  $\stackrel{p_7}{\framebox[1.2\width]{\RL{المبنى}}}$ \RL{هذا} \RL{من} $\stackrel{r_4}{\framebox[1.2\width]{\RL{مقربة}}}$ \RL{على} $\stackrel{p_6}{\framebox[1.2\width]{\RL{مول}}}$ $\stackrel{p_5}{\framebox[1.2\width]{\RL{دبيّ}}}$ \RL{يقع} $\stackrel{p_4}{\framebox[1.2\width]{\RL{الطريق؛}}}$ \RL{هذا}
+  \\ ~ \\
+%  \RL{العالم.} \RL{في} \RL{الأطول} \RL{يُعد} \RL{الذي}  
+%  \\
+  \multicolumn{1}{L{10cm}}{
    \arabfalse \transtrue
@@
    \arabtrue \transfalse} 
-  \\
-  \multicolumn{1}{L{9.5cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.}
+  \\ ~ \\
+  \multicolumn{1}{L{10cm}}{It is impossible not to notice $\stackrel{n1}{\framebox[1.2\width]{Khalifa}}$ $\stackrel{p_1}{\framebox[1.2\width]{Tower}}$ $\stackrel{r_1}{\framebox[1.2\width]{next to}}$ the $\stackrel{u_1}{\framebox[1.2\width]{first}}$ $\stackrel{p_2}{\framebox[1.2\width]{intersection}}$ while you are driving your car $\stackrel{r_2}{\framebox[1.2\width]{on}}$ $\stackrel{n_2}{\framebox[1.2\width]{Sheikh}}$ $\stackrel{n_3}{\framebox[1.2\width]{Zayed}}$ $\stackrel{p_3}{\framebox[1.2\width]{Road}}$, even if this was $\stackrel{u_2}{\framebox[1.2\width]{the first}}$ time that you take this $\stackrel{p_4}{\framebox[1.2\width]{road}}$; $\stackrel{p_5}{\framebox[1.2\width]{Dubai}}$ $\stackrel{p_6}{\framebox[1.2\width]{Mall}}$ is located $\stackrel{r_4}{\framebox[1.2\width]{near}}$ this $\stackrel{p_7}{\framebox[1.2\width]{building}}$, which is the longest in the world.}
   \end{tabular}
@@
 \end{center}
+\vspace{-1em} 
   \caption{Direction example with Arabic text, annotated with entities, transliteration, translation, and extracted relational entities in a graph} 
   \label{fig:intromotiv}
+\vspace{-1em} 
 \end{figure}

### 2018-06-23 tex/merf/TALMajor/gui.tex
format issues ficed: figure and table caption style
minor changes

@@
 
-\begin{figure}[tb!]
-  \centering
-  {\setlength{\fboxsep}{0pt}%
-  \setlength{\fboxrule}{0.5pt}%
-  \fbox{\includegraphics[width=\textwidth]{figures/mreedit}}}
-  \caption{\framework tag type regular expression editor.}
-  \label{f:sfe}
-\end{figure}
-
 \subsection{MBF match visualization}
@@
 
+\begin{figure}[tb]
+  \centering
+  {\setlength{\fboxsep}{0pt}%
+  \setlength{\fboxrule}{0.5pt}%
+  \fbox{\includegraphics[width=0.995\textwidth]{figures/mreedit}}}
+  \vspace{-2em}
+  \caption{\label{f:sfe}\framework tag type regular expression editor.}
+\end{figure}
+
+\subsection{Tag type regular expression editor}
 \vspace{-1em}
-\subsection{Tag type regular expression editor}
 
@@
 
-\begin{figure}[tb!]
-  \centering
-  \includegraphics[width=\textwidth]{figures/treegraph}
-  \caption{MRE annotated Text, MRE \major{matching parse tree}, and entity-relation graph}
-  \label{fig:treegraph}
-\end{figure}
-
 \vspace{-1em}
@@
 
-The \major{matching parse tree} view shows the selected match in a graph view.
-Figure~\ref{fig:treegraph} shows the \major{matching parse tree} of the direction task 
+The matching parse tree view shows the selected match in a graph view.
+Figure~\ref{fig:treegraph} shows the matching parse tree of the direction task 
 \RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
 %Leaf nodes show the match text, internal nodes show the operations, and edge labels shows the MBF tag type name.
+
+\begin{figure}[tb]
+  \centering
+  \includegraphics[width=\textwidth]{figures/treegraph}
+  \vspace{-2em}
+  \caption{\label{fig:treegraph}MRE annotated Text, MRE matching parse tree, and entity-relation graph.}
+\end{figure}
 
@@
 the entity-relation graph of the match of the user defined relation 
-extracted from the \major{matching parse tree} of the MRE. 
+extracted from the matching parse tree of the MRE. 
 %

### 2018-06-23 tex/merf/TALMajor/introduction.tex
format issues ficed: figure and table caption style
minor changes

--- 2018-04-20 tex/merf/TALMajor/introduction.tex
@@
 
-\major{
 Figure~\ref{fig:intromotiv} illustrates the target of \framework using 
@@
 the Arabic text.
-}
 The framed words in the text are entities referring to names of people 
@@
 tags, and semantic categories.
-\major{
 For example, entity $p_1$ in Figure~\ref{fig:intromotiv} has a place semantic 
 category.
-}
 \framework regular expressions support operators such as concatenation, 
-zero or one, zero or more, one or more, up to $M$ \major{repititions where $M$ is 
-a non-zero positive integer}, and logical conjunction and disjunction.
-\major{
+zero or one, zero or more, one or more, up to $M$ repetitions where $M$ is 
+a non-zero positive integer, and logical conjunction and disjunction.
 For example, the sequence between $p_1$ and $p_2$ matches a regular expression $re$ 
@@
 tag ($r_1$) in between.
-}
 
@@
 %
-\major{
 Each regular expression is associated with a named identifier to form a 
@@
 %is key to Arabic NLP.% due to the richness of Arabic morphology.
-}
 
@@
     features such as text, position, length, numerical value, and morphological features.
-  \item \framework enables the user to tag words based on a 
-    \major{
-    novel ligh Arabic wordnet relation that leverages the 
-    synonym $Syn^k$ feature.
-  }
-
-  \item \major{
-\framework is open source and available online for the research community 
-    under on (\url{https://github.com/codelogicanalysis/atmine}).}
+  \item \framework enables the user to tag words based on a novel light Arabic wordnet relation that leverages the synonym $Syn^k$ feature.
+  \item \framework is open source and available online for the research community 
+    under on (\url{https://github.com/codelogicanalysis/atmine}).
 \end{itemize}

### 2018-06-23 tex/merf/TALMajor/relatedwork.tex
minor changes

@@
 
-\major{
 Another track in the literature targets specific tasks such as NER using 
@@
 optimized feature sets and conditional random fields~\cite{benajiba2007anersys,benajiba2008arabic,ekbal2010named,abdelrahman2010integrated}.
-}
 %The work in~\cite{ekbal2010named} presents a language independent approach for NER extraction using {\em support vector machines}. 

### 2018-06-23 tex/merf/TALMajor/results.tex
format issues ficed: figure and table caption style
minor changes

@@
-\major{
 In this section we evaluate \framework with four case studies. 
@@
 The corrections form the manual gold annotation that we compared against.
-}
 
-% Table generated by Excel2LaTeX from sheet 'Sheet1'
-\begin{table}[tb!]
+\begin{table}[tb]
   \centering
-  \caption{\framework compared to task specific applications.}
-  \label{tab:results}%
   \resizebox{\columnwidth}{!}{
@@
     }
-\vspace{-1em}
+   \caption{\label{tab:results}\framework compared to task specific applications.}
+%\vspace{-1em}
 \end{table}%
@@
 for both the standalone task specific and the \framework implementations.
-\major{
 The development time measures the time required for 
@@
 detected entities against the total number of extracted entities.
-}
 
-\major{
 Table~\ref{tab:results} provides runtime results of \framework 
@@
 the same recall.
-}
 
@@
   \centering
-  \caption{Narrator chain example.}
   \begin{Verbatim}[xleftmargin=1.5cm,fontsize=\relsize{-1},commandchars=\\\{\},codes={\catcode`$=3 \catcode`_=8}]
@@
 }
-  \label{tab:nchain}%
-  \vspace{-1.5em}
-\end{table}%
+\caption{\label{tab:nchain}Narrator chain example.}
+ \vspace{-1em}
+\end{table}
 \setcode{standard}
@@
 
-\begin{table}[tb!]
+\begin{table}[tb]
   \centering
-  \caption{\framework MBF and user-defined relation accuracy }
-  \resizebox{0.65\columnwidth}{!}{
+  \resizebox{0.8\columnwidth}{!}{
     \begin{tabular}{l|c|c|c|c}
@@
     }
-  \label{tab:mbfer}%
+  \caption{\label{tab:mbfer}\framework MBF and user-defined relation accuracy.}
  \vspace{-1.5em}
@@
 
-\vspace{-2em}
+\vspace{-1.5em}
 \subsection{Genealogy tree}
@@
 
+\vspace{-1em}
+\subsection{Number normalization}
+\label{subsec:numnorm}
+
 \newcommand*{\fvtextcolor}[2]{\textcolor{#1}{#2}}
-\begin{figure}[tb!]
+\begin{figure}[tb]
 \centering
@@
 
-\vspace{-1em}
-\subsection{Number normalization}
-\label{subsec:numnorm}
-
 We implemented a number normalization extractor using \framework and 
@@
 %The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
-\major{
 Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.
@@
 value (\cci{\$s1.number}) of literal numbers such as numbers from one to ten.
-}
 %

### 2018-06-23 tex/merf/TALMajor/sarf.tex
format issues ficed: figure and table caption style
minor changes

@@
 %{\bf Morphological analyzer (Sarf).~}
-\major{
 Morphological analysis is key to Arabic 
@@
 system~\cite{arabicmorph,shahrour2016camelparser,pasha2014madamira}.
-}
 Short vowels, also known as diacritics, are typically omitted in Arabic text
@@
 %
-\major{
 Each morphological solution $m\in M(t)$ is a tuple of the form 
@@
 Category values, respectively.
-}
 
 %\setarab
-\begin{table}[tb!]
+\begin{table}[tb]
   \centering
-  \caption{Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
-  \label{tab:samplesolution}%
-  \resizebox{0.7\columnwidth}{!}{
+  \resizebox{0.85\columnwidth}{!}{
     \begin{tabular}{|r|c|c|c|c|c|}
@@
     }
-  \vspace{-2em}
+    \vspace{-1em}
+  \caption{\label{tab:samplesolution}Sample solution vector for \utfrl{فَسَيَأْكُلها}.}
 \end{table}%

### 2018-06-23 tex/merf/TALMajor/tool.tex
minor changes

@@
 
-\major{
 Up to our knowledge, 
-$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.}
+$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.
 The sets $E, A,$ and $L$ denote all English words, Arabic words, 
@@
 subsets of related English glosses, %; e.g. $g_{s} = \alpha(s)\subset 2^{\GG}$.
-\major{
 where $2^{\GG}$ denotes the power set of $\GG$ which is the set of all subsets 
-of $\GG$.}
+of $\GG$.
 Function $\gamma: L \rightarrow 2^{\ss}$ maps Arabic lexicon words to subsets 
@@
 %
-\major{
 The example in Figure~\ref{fig:introsynEx} illustrates the computation.
@@
 Therefore, $Syn^2(w)$ relates the words \RL{mA'} and \RL{r^s^s}.
-}
 
@@
 
-\major{
 Let ${\cal O} = \{ \mathit{isA}, \mathit{contains}\}$ be the set of atomic term 
@@
 set of morphological features where each morphological feature $A\in {\cal F}$ is in 
-turn a set of morphological feature values.}
+turn a set of morphological feature values.
 Given a word $w$,
@@
 Informally, a solution vector of $w$ exists with
-a feature containing or exactly matching \major{the user-chosen feature value} $CF$.
+a feature containing or exactly matching the user-chosen feature value $CF$.
 \item $a(w) := w \in Syn^k(CF), CF \in \ss$.
   Informally, this checks if $w$ is an extended synonym of a stem $CF$.
-\major{We limit $k$ to a maximum of $7$ since we practically noticed that 
+  We limit $k$ to a maximum of $7$ since we practically noticed that 
   (1) values above $7$ introduce significant semantic noise and
-  (2) the computation is expensive without a bound.}
+  (2) the computation is expensive without a bound.
 \end{itemize}
@@
 
-\major{
 Consider the task we discussed in the introduction (Figure~\ref{fig:intromotiv}) 
@@
 \RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).
-}
 
@@
 \item $m$ is an MRE where $m$ is an MBF.
-\item $fg$ is an MRE 
-  \major{
-  where $f$ and $g$ are both MRE expressions. 
-    A match of $f$ followed by a match of $g$ satisfies this concatenation operation.}
-\item $f*,f+,f$\textasciicircum$x,$ and $f?$ \major{ are MRE } where $f$ is an MRE,
-  and are satisfied by zero or more, \major{ one or }more, up to $x$ matches, 
-    and an \major{optional single match} of $f$, respectively.
-  \item $f\& g,$ (conjunction) and $f|g$ (disjunction) are \major{MRE} where 
-    $f$ and $g$ are \major{MRE}, and are satisfied by 
-    \major{the intersection of $f$ and $g$ matches, 
-    and the union of the $f$ and $g$ matches}, respectively. 
+\item $fg$ is an MRE where $f$ and $g$ are both MRE expressions.
+A match of $f$ followed by a match of $g$ satisfies this concatenation operation.
+\item $f*,f+,f$\textasciicircum$x,$ and $f?$ are MRE where $f$ is an MRE,
+  and are satisfied by zero or more, one or more, up to $x$ matches, 
+    and an optional single match of $f$, respectively.
+  \item $f\& g,$ (conjunction) and $f|g$ (disjunction) are MRE where 
+    $f$ and $g$ are MRE, and are satisfied by the intersection of $f$ and $g$ matches, 
+    and the union of the $f$ and $g$ matches, respectively. 
 \end{itemize}
@@
 
-\major{
 Back to the example in Figure~\ref{fig:intromotiv}.
@@
 %Note that the user reaches the satisfying matches by experimenting with the visualizer and the expression editor which do not require knowledge and expertise in regular expressions.
-}
 
@@
 
-\major{
-The \major{matching parse trees} in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
+The matching parse trees in Figure~\ref{fig:taskMRE} illustrate two matches of the expression computed by \framework.
 The first tree refers to the text 
@@
 \RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
-}
 %The leaf nodes of the trees are entities and the edges and internal nodes are text, 
@@
 %and word distance based relational entities. 
-\major{
 The leaf nodes of the trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
 For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the sub-expression $(P|N)+$.
-}
 
@@
 
-\major{
 We are interested in constructing the relational entity graph in Figure~\ref{fig:intromotiv}.
@@
 Note that there is no match to the optional $O$ formula in match 1.
-Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second \major{matching parse tree} are
-\RL{dby mwl} (Dubai Mall), \RL{`l_A} (is located), \RL{mqrbT} (near), \RL{mn h_dA} (this), and \RL{Almbn_A} (building), respectively.
-}
+Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second matching parse tree are \RL{dby mwl} (Dubai Mall), \RL{`l_A} (is located), \RL{mqrbT} (near), \RL{mn h_dA} (this), and \RL{Almbn_A} (building), respectively.
 
-\major{
 We define the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.
@@
 nodes in match 2.
-}
 
@@
 and morphological features.
-\major{
 The API follows a decorator pattern in that it incrementally adds the 
 action results to the matching entities.
-}
-Once \framework computes all \major{matching parse trees}, it traverses each tree to 
+Once \framework computes all matching parse trees, it traverses each tree to 
 execute the user defined \cci{pre-match} actions in pre-order manner
 and the \cci{on-match} actions in post-order manner.
-\major{
 This follows an observer pattern that notifies listeners with each
 produced match.
-}
 
@@
 with font and color information.
-\major{
 For the example of Figure~\ref{fig:intromotiv},
@@
 and $d$ is italic.
-}
 
 For each word $t_i\in T, 0\le i < |T|$.
-\framework computes a Boolean value 
-for all MBFs. 
-\major{
+\framework computes a Boolean value for all MBFs.
 For example, \RL{brj} matches MBF $P$.
-}
 Then, it computes the set of MBF tags
@@
 its tag set $R_i$ is by default the singleton $O=\{\mathit{NONE}\}$.
-\major{
 For example, the tag sets for the text in Figure~\ref{f:overview} follows 
 $\{ \{NONE\}, \{NONE\}, \{NONE\}, \{NONE\}, \{($\RL{brj}$,P)\}, \{($\RL{xlyfT}$,N)\},\ldots\}$.
-}
 
@@
 corresponding expression $(P|N)\!+~O?~R~O^\wedge 2~(P|N|U)+$.
-\major{
 Figure~\ref{fig:nfaEx} shows part of the corresponding NFA where
@@
 $P,$ and $N$.
-Edges labeled with the empty string $\epsilon$ are non-deterministic.}
+Edges labeled with the empty string $\epsilon$ are non-deterministic.
 %The expression uses operations $|$, $?$, $+$, and $\wedge$ to relate places {\tt P}, names of persons {\tt N}, relative positions {\tt R}, numerical terms {\tt U}, and other {\tt O}.
@@
 %A simulation match $m$ of an expression $f$ is a tree where the root is the MRE expression, the internal nodes are the MRE and MBF operations, and the leaves are matches of the MAT terms of $f$.
-A simulation match $m$ of an expression $f$ is a \major{parse tree} where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g.
+A simulation match $m$ of an expression $f$ is a parse tree where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g.
 Fig.~\ref{fig:taskMRE}.

### 2018-06-26 tex/sarf/Response_rev1.tex
sarf to ACLing draft added

\documentclass[11pt]{article}
\usepackage[dvipsnames]{xcolor}
\usepackage[letterpaper,margin=1.5in,left=35mm,right=20mm, top=20mm, bottom=20mm ]{geometry}
\usepackage{url}
%\usepackage{todonotes}
\usepackage{enumitem}
\usepackage{amssymb}
\usepackage{arabtex}
\usepackage{hyperref}
\hypersetup{          
  colorlinks=false, 
  pdfborder={0 0 0},
}                     
% used for cross reference from nlesarf.tex
\usepackage{xr} 
\externaldocument{out/nlesarf}

\usepackage{csquotes}

% You ou can also add a prefix: 
%\externaldocument[I-]{chapterI}
% then you can reference it as \ref{I-a}

\newcommand{\cci}[1]{\mbox{\textup{\small \texttt{#1}}}}
\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}

\newcommand{\chci}[1]{`\cci{#1}'}

\newcommand{\noTrnoVocRL}[1]{\transfalse\novocalize\noTrRL{#1}\vocalize\transtrue}  
\newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}  
\newcommand{\utfRL}[1]{\setcode{utf8} \RL{#1} \setcode{standard}}
\newcommand{\noTrutfRL}[1]{\setcode{utf8} \transfalse\RL{#1}\transtrue \setcode{standard}}
\newcommand{\noArutfRL}[1]{\setcode{utf8} \arabfalse \RL{#1} \arabtrue \setcode{standard}}

\definecolor{shadecolor}{named}{gray}
\definecolor{DarkGray}{gray}{0.30}

\usepackage{utf8}
\setarab
%\fullvocalize
\transtrue
\arabtrue

\RequirePackage{xspace}
\linespread{1.5}

%\newcommand{\itodo}[1]{\todo[inline,color=green]{#1}}

\title{Sarf: Fast and Application Customizable Arabic Morphological Analyzer
 \\
~~ \\
  Revision summary }
\date{}

\begin{document}
\maketitle

{\noindent Dear Editor-in-Chief and Expert Reviewers,}

We would like to thank you for your recommendations, valuable comments, 
and constructive observations. 
In this revised version of the paper we made every effort to address the 
concerns brought up in your review. 
This document includes a summary of our changes and lists each of the 
reviewers’ comments followed by a description of how it was addressed.

For the convenience of the reviewers, we also provide online 
(\url{http://research-fadi.aub.edu.lb/nlesarf/nlesarf-rev1-annotated.pdf})  
a version of the revised paper with annotations that point to changes and 
revisions and relate them to the concerns of the reviewers.

Thank you,
\newline
\hrule

\input{editorComments.tex}
~\newline
\hrule

\input{Reviewer1.comments}
~\newline
\hrule

\input{Reviewer2.comments}
~\newline
\hrule

\input{Reviewer3.comments}
~\newline
\hrule

%\newsavebox\mytempbib
%\savebox\mytempbib{\parbox{\textwidth}{
%\input{psqBMCaig.bbl}
%}}

\end{document}

### 2018-06-26 tex/sarf/Reviewer1.comments.tex
sarf to ACLing draft added

\section*{Reviewer 1 comments}
This paper presents Sarf, an application customizable morphological analyzer for Arabic.

\begin{enumerate}[leftmargin=0mm,label=\bfseries CommentR1.\arabic*]
\item \label{Review.1.1}
While Sarf's API specification is new in this paper, the most significant research contributions, namely, Sarf's use of agglutinative morphemes and fusional concatenation rules (Section 4) along with the relevant experimental results (Section 7) had already been published in COLING 2012 cited in the paper.

More specifically, among the five contributions noted on page 2,  and repeated below, only the first contribution is new to this paper, and all others have significant overlaps with the previously published  ``Arabic morphological analyzer with agglutinative affix morphemes and fusional concatenation rules''.

1)      Sarf provides an application customizable morphological analyzer where the developer can control and refine analysis\\
2)      Sarf is a novel Arabic morphological analyzer with agglutinative affixes and fusional affix concatenation rules based on textbook Arabic morphological rules and on the concatenation rules of existing analyzers\\
3)      Sarf solves inconsistencies in existing affix lexicons of BAMA \& SAMA\\
4)      Sarf solves correspondence between morphological solution and the morphological segmentation of the original text problem\\
5)      Sarf is fully implemented and available online as an open source tool\\

Both customizable API and agglutinative morphology sections need to be substantiated with more technical details and content.

{\bf Response}\\
We modified the contribution list in Section \ref{sec:intro} 
to reflect the following. 
The paper provides several additional contributions to the short paper published in COLING 2012 including the following.
\begin{itemize}
  \item The paper presents the application specific API. The API allows NLP application developers to do the following.
    \begin{itemize}
      \item Control the morphological analysis by dynamically accepting and rejecting solutions. 
      \item Refine the analysis by dynamically selecting the morphological features to compute. 
      \item Refine the analysis by dynamically prioritizing what features to compute first. 
    \end{itemize}
  \item Presents and describes in detail the engineering behind the efficient implementation of Sarf including the following.
    \begin{itemize}
      \item The organization of the lexical structures into root index tries during runtime. 
      \item The traversal structures that allows multiple solutions to be considered at once.
      \item The computation of the solution features in a tree based structure instead of computing all features and filtering later on.
      \item The manual optimization of the traversal of the lexicon tries to reduce the typical excessive non-determinism exhibited in 
        the finite state machine based solutions. 
    \end{itemize}
  \item The solution to the ``run-on'' words problem. 
  \item The optional use of partial diacritics to disambiguate the analysis. 
  \item Lists and discusses the inconsistencies found in BAMA and SAMA in details. This might be helpful for users of the two tools. 
  \item Presents additional experimental results with discussions. 
\end{itemize}
$\blacksquare$

\subsection*{Detailed Comments:}
\item\label{Review.1.2}
Page 5. In the paragraph above ``2 Overview'', mention of `Section 7 Results' is missing.

\textbf{Response}\\
Fixed.

$\blacksquare$

\item\label{Review.1.3}
Page 6. What's the syntax of feature priority and feature selection rules?

\textbf{Response}\\
The rules are added using C++ APIs. We added more details about the API in Section ~\ref{sec:api}.
$\blacksquare$

\item\label{Review.1.4}
Page 7. Regarding Fig 2, how are the DAG's for affixes generated in the first place? If possible, provide an algorithm for affix DAG generation given a list of prefixes and suffixes in the language?

\textbf{Response}\\
%Note that relations and category are not referred to in the text. \\
%\\
%In overview: ``construct Sarf structures process takes as input the lexicon, the fusional and agglutinative
%rules, and the use diacritic option and constructs directed acyclic graph (DAG) structures
%that encode the affixes, and a root index trie structure that encodes the stems (Aoe, 1989)."\\
%Suggested modification in page 7: The diagram in Figure 2 illustrates the Sarf data structures. Subfigures (a), (b), and (c) in
%Figure 2 represent P the prefix DAG, S the stem trie, and X the suffix DAG, respectively. \textit{The DAG's are generated using  lexicons, fusional and agglutinative
%rules, and use diacritics (Aoe, 1989).} Boxes and circles \textit{in the DAG's} denote morpheme versus non-morpheme node.\\
%Or\\
We added the ConstructSarfDAGs algorithm in Section~\ref{sec:overview} 
and modified the name of the section to 
be Sarf. 
$\blacksquare$

\item\label{Review.1.5}
Page 7. What is the speed-up gained by adopting affix DAGs and stem trie? How does the speed of Sarf with affix DAGs and stem trie compare with  BAMA and SAMA. Provide a table comparing the speed of Sarf and BAMA and SAMA, analogous to Table 10 for lexicon size.

\textbf{Response}\\
Table ~\ref{t:reduction} provides a comparison to SAMA and ElixirFM using the Hadith application. SAMA subsumes BAMA.
We also performed an experiment where we computed all solutions with SAMA and Sarf and we report on that in text in 
Section ~\ref{ss:perform}.
$\blacksquare$

\item\label{Review.1.6}
Page 9.  Line 3. Regarding the statement ``Otherwise, it moves to a regular node or dies.'' 
What is a regular node? How does it differ from “morpheme”, “non-morpheme” nodes introduced 
on page 6 in relation to Fig.2 (4th line from the bottom)?

\textbf{Response}\\
This was a typo. We meant to say non-morpheme node.
$\blacksquare$

\item\label{Review.1.7}
Page 9. 3rd paragraph regarding run-on word problem, what are the possible structures of run-on words? 
Fig. 2 illustrates a traversal from the suffix node X to the prefix node P. 
How are the run-on words consisting of only stem sequences or prefix-stem sequences handled?

\textbf{Response}\\
Run-on words are words that are separated by a non-connecting letter rather than a delimiter space. 
In case of a string with only stem sequences, empty transitions between 
${\cal S}$ and ${\cal X}$, ${\cal X}$ and ${\cal P}$, and ${\cal P}$ and ${\cal S}$ to accept nodes allows parsing the string. 
We added this note at the end of the ``Running example'' section. \\
$\blacksquare$

\item\label{Review.1.8}
Page 10. Is it possible to provide the complete list of partial affix morphemes?

\textbf{Response}\\
The complete list of partial affix morphemes that are generated from BAMA is available online in the open source tool. 
The same algorithm applies for extracting morphemes from SAMA.  
Our understanding is that our license covers only
using SAMA and limits publishing an exhaustive list. 
$\blacksquare$

\item\label{Review.1.9}
Page 12.  Regarding the statement ``The first five rows in Table 2 can be replaced with three atomic affix morphemes and one partial affix morpheme in Lp and three rules to generate compound morphemes in Rpp'',  what are the three atomic affix morphemes? What are the three rules to generate  compound morphemes in Rpp?

\textbf{Response}\\
The two atomic affix morphemes are \RL{f} and \RL{y}, and the partial morpheme is \RL{s}. The three rules are the rules that connects the category of \RL{f} to that of \RL{y}, the category of \RL{s} to that of \RL{y}, and the category of \RL{f} to that of \RL{s}. We added the morphemes and rules to the text.
$\blacksquare$

\item\label{Review.1.10}
Page 12.  Regarding the statement ``With Sarf, the equivalent addition of ya requires only two rules in Rpp'',   what are the two rules?

\textbf{Response}\\
The two rules are those that connect the category of \RL{f} to that of \RL{y}, and the category of \RL{s} to that of \RL{y}. 
We modified the text to reflect this. 
$\blacksquare$

\item\label{Review.1.11}
Page 15.  Regarding diacritics, how does one determine whether or not a particular diacritic reduces the ambiguity of an input word? 
Is it algorithmic?

\textbf{Response}\\
We explain the use of partial diacritics to disambiguate the solutions in Section~\ref{sec:diac}. 
$\blacksquare$

\item\label{Review.1.12}
Page 18-19.  Regarding the statement ``Sarf differs in that it is a general morphological analyzer that reports all possible solutions'': 1) Is there any score associated with each solution? 2) Is it possible to restrict the number of solutions, analogous to $n$-best hypotheses in statistical models? 3) Is it possible to reduce the ambiguities on the basis of surrounding word contexts, i.e., word $n$-gram contexts?

\textbf{Response}\\
Sarf is a morphological analyzer that works on a word level. It does not consider context to disambiguate solutions and it currently does not 
provide a score with the solution.
$\blacksquare$

\item\label{Review.1.13}
Page 20.  7.1. Regarding Sarf used in the evaluation, how were the prefix, suffix, stem lexicons constructed? Have they been derived from ATBv3.1? 99.991\% agreement indicates that there is no out-of-vocabulary, which is almost impossible unless the lexicons are derived from the same data as the evaluation data.

\textbf{Response}\\
The lexicons of SAMA (and BAMA) are already updated according to the ATB.\\
Sarf refines and extends the lexicons of SAMA (and BAMA) and applies some modifications according to affix rules derived from Arabic textbooks on morphology and syntax (AlRajehi, 2000a, 2000b; Mosaad, 2009) as discussed in Section 5.1.

The Arabic TreeBank (ATB) originally bases its morphological analysis on BAMA where acceptable morphological
analysis reaches a minimum of 98.87\%. 
Also, SAMA was updated according to ATB according to ``Consistent and Flexible Integration of Morphological Annotation in the Arabic Treebank'', S. Kulick \& M. Maamouri/

%\mayatodo{Maya: Read paper of SAMA on segmentation to understand the \% of it (check references -Maamouri et al. 2008)}
%\textbf{Remove in final rev version:}\\
%The ATB uses a level of annotation more accurately
%described as morphological analysis than as part-of-
%speech tagging. \\
%%%%%%
%The source token is run through the
%Buckwalter Arabic Morphological Analyzer (Buck-
%walter, 2004). \\
%%%%%%
%This provides all possible analyses of
%the token, each analysis consisting of (1) breaking a
%word into its various segments (prefixes, suffixes, gen-
%der and Case endings) and (2) providing a vocalization
%for each segment, with the diacritics appropriate for
%that solution. The annotator picks one such analysis
%as the proper solution for this token. We refer to this
%solution as the POS token, which includes the break-
%down of the source token into its various segments,
%each one vocalized.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%Maamouri 2004  Developing an Arabic Treebank: Methods, Guidelines, Procedures, and Tools
%The automatic Arabic morphological
%analysis and part-of-speech tagging was performed
%with the Buckwalter Arabic Morphological
%Analyzer, an open-source software package
%distributed by the Linguistic Data Consortium
%(LDC catalog number LDC2002L49).
%\\
%The Arabic Treebank: Part 2 corpus contains
%125,698 Arabic-only word tokens (prior to the
%separation of clitics), of which 124,740 (99.24\%)
%were provided with an acceptable morphological
%analysis and POS tag by the morphological parser,
%and 958 (0.76\%) were items that the morphological
%parser failed to analyze correctly.\\
%Items with solution 124740 99.24\% \\
%Items with no solution 958 0.76\% \\
%Total 125698 100.00\% \\
%Table 1. Buckwalter lexicon coverage, UMAAH
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%Consistent and Flexible Integration of Morphological Annotation in the Arabic Treebank
%In this paper we have discussed how we have overcome
%these problems with consistent and more precise
%categorization of each Treebank token for its relationship
%with SAMA. We also discussed how we have improved
%the creation of alternative forms of the tokens used in the
%treebank structures.
$\blacksquare$

\item\label{Review.1.14}
Page 20-21.  Regarding the statement ``Since Sarf preserves correspondence when performing segmentation,  it is capable of generating a vocalized tag in the ATB 'after' dataset, which carries more information in 15.47\% of the time than the counterpart POS derived vocabulary entry",  what does it mean to carry more information?

\textbf{Response}\\
%No vocalized types in ATB. Sarf has vocalized types. 
By carrying more information, we mean adding  diacritics on the source token and not on modified tokens. \\
We added the following to the article for illustration.

In some cases, the POS derived ATB vocalization processes the source token by splitting it into multiple unvocalized POS tokens which are then converted into vocalized ATB tokens. 
For example, 
the word \utfRL{للقضاء} (for/to the justice/judiciary) is split into \cci{li/PREP + Al/DET + qaDA'/NOUN} tokens. 
In such cases, the POS tokens differ from the source tokens. 
$\blacksquare$

\item\label{Review.1.15}
Page 21. The transition from SAMA to BAMA $\rightarrow$ should be from BAMA to SAMA.

\textbf{Response}\\
Fixed.
$\blacksquare$

\item\label{Review.1.16}
Add Sarf lexicon size to Table 10  for side-by-side comparisons with BAMA and SAMA

\textbf{Response}\\
We rewrote the table to clearly compare BAMA and SAMA to Sarf in terms of the size of the lexicons. 
$\blacksquare$

\item\label{Review.1.17}
Page 23. Table 11. Why is the precision so much lower for Proper Names that that of Tell connectors and Family connectors? What is the number of truth, i.e. denominator in recall calculation for Proper Names, Tell connectors and Family connectors?

\textbf{Response}\\
Some of the proper names occur in the text outside the narrator chains. While they are correct proper names, they are not 
necessarily wanted proper names for the application. We modified the paper to reflect that. 
$\blacksquare$

\end{enumerate}

### 2018-06-26 tex/sarf/Reviewer2.comments.tex
sarf to ACLing draft added

\section*{Reviewer 2 comments}

\subsection*{Overview}
The work is a constructive attempt by the authors to build a tool facilitating the applicability of morphological analysis customizable for applications compared to existing tools which present  less flexible output as solution.

It identifies some of the shortcomings and discrepancies of existing morphological analyzers.

The main features advocated by the author about the tool are 
\begin{itemize}
\item[-] the characteristic of choice of detail output for a particular analysis resulting in performance gains. 
\item [-] Concept of  ‘agglutinative’ and ‘fusional’ affixes refining affix combination rules resulting in  reducing lexicon size and speedier analysis.
This aspect is an inclusion from their earlier work, which is previously published.
\item[-] Identify numerous `inconsistencies’ in some existing tools due to error in the encoding of the lexicons for affixes.
\item[-] They introduce certain  other features such as unprecedented analysis of  `Run-­‐on word’; using information from partially diacriticized text which have been used by one other tool
\end{itemize}
Paper mainly describes the technical aspect and functioning of the Sarf Application to choose appropriate analysis based on user defined choices.  The main novelty  aspect of the techniques for analyzing words based on `agglutinative and fusional’ affixes has been covered by the author in previous publication.

\begin{enumerate}[leftmargin=0mm,label=\bfseries CommentR2.\arabic*]
\item \label{Review.2.1} 
(A) Paper doesn't provide conceptually and linguistically  motivated reasoning  behind the application of a certain methods to obtain their solution. \\
(B) Not much theoretical and descriptive insight about morphological concepts

\textbf{Response}\\
The paper informally introduces the necessary concepts for morphological analysis. 
We reorganized the paper and the background and motivation section is now a motivation section. 
The paper is already at or beyond page limit. 
If the reviewers see fit, we may include the textbook theoretical morphological concepts in a separate Appendix. 
$\blacksquare$

\item \label{Review.2.2}
Whereas on the one hand  they improve upon and combine many of the advantages of existing tools they didn't include a key aspect of morphological analysis which is disambiguation of the analysis provided by some existing analysers such as MADAAMIRA.

\textbf{Response}\\
%Sarf provides the option to uses existing partial diacritics for disambiguation. It considers
%the diacritics at morpheme boundaries, to generate only the diacritic matching solutions.
%\\
%Maya: what's the difference of this and morphological disambiguation. In literature review we state that we don't do morphological disambiguation. (shadda example)
%
MADAMIRA does not support application customization, run-on words, and partial diacritics. 
Sarf does, but does not support disambiguation. 
An NLP developer can benefit from the MADAMIRA disambiguation features in the API. 
In the future, we will work on integrating the disambiguation features from MADAMIRA into Sarf. 
$\blacksquare$

\item \label{Review.2.3}
Paper needs to improve upon the overall presentation. 
It lacks professionalism with a clutter of spontaneous examples, which could be more organized. 
Also includes some mundane details such as about the architecture of the system, which could be omitted.

\textbf{Response}\\
We split and renamed the sections to improve the organization of the paper. 
We also embedded examples in place as suggested in Comment R2.8.
Regarding details about the system architecture of Sarf, 
other reviewers requested that we expand on the engineering aspect 
of Sarf.
$\blacksquare$

\subsection*{Specifics:}
\item \label{Review.2.4}
Paragraphs don't sometimes connect and lack coherence such as in section 2.2

\textbf{Response}\\
We fixed this by editing and rewriting some of the paragraphs in the section.
$\blacksquare$

\item \label{Review.2.5}
Poor presentation of the Arabic script and its Romanized form: quite unprofessional.

\textbf{Response}\\
We are using the ArabTex package for Arabic text editing and transliteration as recommended in the ``Introduction to Arabic Natural Language Processing'' book by Nizar Habash.
We added a footnote indicating the used package and citing the book recommending it.
$\blacksquare$

\item \label{Review.2.6}
Poor formatting of the section heading without variation in font sizes for subsections.

\textbf{Response}\\
We are using the Natural Language Engineering Latex template. The template defines the used font sizes for the sections/subsections. 
$\blacksquare$

\item \label{Review.2.7}
Also poor organization of section heading: Describing the various
aspects of their tool should be organized under one large section with various
subsections for each feature described.\\
\textbf{Response}\\
We expanded the overview section and we renamed it to be Sarf as suggested. 
$\blacksquare$

\item \label{Review.2.8}
Related works section should after introduction rather than towards the end.

\textbf{Response}\\
We modified this. The Related Works section is now after the Introduction.
$\blacksquare$

\item \label{Review.2.9}
Lack of illustrative examples; only refers back to the same table to explain
their point of view which remains unclear

\textbf{Response}\\
We embedded the examples in place in most of the cases as suggested. 
$\blacksquare$

\item \label{Review.2.10}
Diagrams missing key details and lack illustrative power to explain the
concept clearly.

\textbf{Response}\\
We added details to the diagram of Figure~\ref{fig:flowdiagram}. The labels of the diagram components now
better relate to the text. 
We also highlighted the traversal path in Figure~\ref{f:example} with shaded nodes. 
$\blacksquare$

\item \label{Review.2.11}
`Solution construction' section (2.2) is very ambiguous and unclear. Lacks
clarity in explaining the working of their application. E.g. it lacks clarity as to
how the solution is reached from the DAG and Trie structure. Paragraph 2
needs rephrasing.

\textbf{Response}\\
We edited Section~\ref{ss:solconstr} to make it clearer. 
It now explains extracting the morpheme sequence from the DAG and trie structures. 
It also explains processing the input sequence to construct the morpheme solutions. 
$\blacksquare$

\item \label{Review.2.12}
Missing gloss for the Arabic script in the given examples in text makes
grasping the concept difficult

\textbf{Response}\\
We added glosses for most of the examples unless redundant in the same paragraph. 
$\blacksquare$

\item \label{Review.2.13}
The text has many typos e.g. `the was' instead of `there was' page 9 line 9;
`Sarfover' missing space in page 20; ``and compares that'' on page 23 Sec 7.5,
instead of ``and compared that''; to name a few examples.

\textbf{Response}\\
Thank you. We fixed the typos.
$\blacksquare$

\item \label{Review.2.14}
English used is sometimes improper and not of a quality standard. E.g. the
sentence p 15, para 5, ``It checks that the sequence of non-diacritic letters,
ignoring the diacritics between them, are equal''

\textbf{Response}\\
Fixed. 
$\blacksquare$

\item \label{Review.2.15}
The purpose of the ``diacritic-aware consistency check is unclear''. No
procedure is outlined how it is used to discard certain diacritic as invalid.

\textbf{Response}\\
We added a paragraph at the end of Section~\ref{sec:diac} to explain how the check is used
to eliminate solutions while parsing. 
$\blacksquare$

\item \label{Review.2.16}
Tables distant from the referenced text.
\textbf{Response}\\
We fixed this for all tables.
$\blacksquare$

\item \label{Review.2.17}
Conclusion quite brief: needs more detail and conclusive deductions about the
work

\textbf{Response}\\
We added a brief on the results of using Sarf to the conclusion. 
$\blacksquare$

\end{enumerate}

### 2018-06-26 tex/sarf/Reviewer3.comments.tex
sarf to ACLing draft added

\section*{Reviewer 3 comments}

\subsection*{Comments to the Author}
In general, I like the approach the authors take, and they make a good case that it is a better way to maintain the
SAMA information. The API looks useful as well.

\subsection*{Most serious points:}
\begin{enumerate}[leftmargin=0mm,label=\bfseries CommentR3.\arabic*]
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.1}
A. The authors need to make it clear that they are using the morphological categories and ``compatibility" tables from SAMA, which as far as I can tell they are doing. (\#3 below)

\textbf{Response}\\
We already mentioned in the paper that Sarf refines the lexicons of SAMA and BAMA and encodes the compatibility rules into agglutinative
and fusional rules. 
We made several modifications to stress this point including.
\begin{itemize}
  \item In the abstract and conclusion, we stressed that Sarf extends and refines the lexicons of SAMA and BAMA. 
\item At the end of the Section~\ref{sec:motivation}, we clarified that Sarf starts from the lexicons of SAMA and BAMA and 
  detailed the changes from BAMA and Sarf in terms of lexical content. 
\end{itemize}
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.2}
B. For comparison with SAMA, there needs to be discussion of the preprocessing SAMA for non-standard characters (\#5 below)

\textbf{Response}\\
We respond to this comment in \ref{Review.3.11} below. 
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.3}
C. The empirical justification for using the partial diacritics is not adequate (\#6 below).

\textbf{Response}\\
We respond to this comment in \#6 (\ref{Review.3.13}) below. 
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.4}
D. Evaluation should not be done with ATB3v3.1.  It's not even a public release, and was an intermediate step. 
It should be done with ATB3 v3.2.  (\#9 below)

\textbf{Response}\\
This was a typo. We use ATB3v3.2 for evaluation. 
$\blacksquare$

\subsection*{Points:}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.5}
1. There seems to be an ongoing confusing with regard to ``segmentation'' or ``tokenization" of SAMA-type solutions.
This is first seen on page 5, with  ``The analysis of the example  li/PREP+Al/DET+qaDA'/NOUN is segmented into two tokens li/PREP and Al/DET+qaDA'/NOUN."

This is not relevant for the paper.  That segmentation is done for  purposes of treebanking, so that ``li" can be a separate leaf on  the syntactic tree.  It is using the output of SAMA of course, but this segmentation is not an issue for SAMA.  SAMA produces li/PREP+Al/DET+qaDA'/NOUN.  The issue of the alignment and correspondence between the original word and its morphemes is already there regardless of the later usage of the SAMA output.  It takes a particular form in the ATB because of this separation for treebanking, but it is an independent problem.

\textbf{Response}\\
We thank the reviewer for the comment. We clarified that the problem is not a SAMA problem and that it is pertinent to the 
applications that use the SAMA output later. 
That said, the fusional rules help solve this problem and hence we claim it as an advantage for Sarf.
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.6}
Section 6 has several references to TOKAN.  
TOKAN is concerned with, also, splitting up the parts of a SAMA solution (e.g. li,Al,qaDA' above) into different ``segments'', as above, such as ``li" and ``Al+qaDA'", with options for doing it different ways, e.g. to split off the future particle or not, etc.., depending on the application. It's a separate issue from the output of SAMA (or of Sarf, as far as I can tell), so a comment such as ``In Sarf, there is no need for a separate segmentor such as TOKAN..." don't make sense, or is at best very unclear.

\textbf{Response}\\
We modified the sentence to read as follows.

\blockquote{
Sarf keeps a stack of the positions that partition text into morphemes
as part of the solution construction process. 
Therefore, the text segments corresponding to the solution features is preserved
with no need for further post-processing. }
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.7}
Section 7.1 has a confusing reference to ``the same experiment using TOKAN toolkit. What does this mean?  What were the option settings for TOKAN?  What relevance does that have for the problem of matching sequences in the input string to the parts of the morphological solution?

\textbf{Response}\\
We used the ATB scheme alias with TOKAN. We see the point of the reviewer that the relevance is hard to argue for. 
We removed the paragraph.
$\blacksquare$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\item \label{Review.3.8}
2. p. 7, section 2.2. The point about the priority rules is unclear. In either case, with POS as the priority or the gloss, you end up with six solutions in Fig.3, don't you?  It's unclear what the utility is of this.  This seems to be the same mechanism whereby later it's explained that the API can be modified to augment or filter solutions as desired?  Is so, could a reference be made to that section later, or the later section refer back to this one, to make that explicit?

\textbf{Response}\\
We further explained the benefit of priority features and referred to the API for priority rules in Section \ref{ss:solconstr}. 
$\blacksquare$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\item \label{Review.3.9}
3. It's referred to several times that the ``prefix-prefix", etc. compatibility rules are used. (eg. page 7).  
Are these taken directly from SAMA, which implies that the morphological categories taken from SAMA are also used?  
It should be stated clearly that you are taking these categories from SAMA, which after all already involves a 
certain amount of analysis.

\textbf{Response}\\
We addressed this issue in the response to \ref{Review.3.1}.
$\blacksquare$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.10}
4. p. 11, last paragraph before section 4.1. ``Row 4 illustrates the use of the wild character..."  There should be a concrete example of the application of this expression with *.  It's too confusing otherwise.

\textbf{Response}\\
We added an example just before Section~\ref{ss:affix-affix}.
$\blacksquare$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\item \label{Review.3.11}
5. p. 15, (a) - ``This is resolved in SAMA for standalone Alef prefixes via preprocessing tokens...".   More needs to be said about this. As I am sure the authors are aware, the Sama Perl engine does some preprocessing of various character, allowing for certain common  alternative spellings (the function get\_variants in SAMA.pm). Since this is done by preprocessing, they are not included in  the SAMA tables.  How does Sarf handle these cases?  To what extent do they matter?

\textbf{Response}\\
We address this concern by the separation of prefixes into partial and atomic variants, and by 
introducing agglutinative and fusional rules. 
The rules construct compound prefixes from partial and atomic prefixes. 
This, therefore, enables the introduction of all prefixes that include Alef, with any of its three alternative spellings, 
by only adding the Alef alternatives into the atomic prefix set. In contrast, SAMA lists all prefixes with all 
possible spelling alternatives (e.g. \RL{sA}, \utfRL{سأ}, \utfRL{سآ}), or does a preprocessing to map all alternatives into one.

We illustrate this point in the paper in Section ~\ref{s:s:inc} when describing the SAMA inconsistencies. 
$\blacksquare$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.12}
5. pp. 13,14 - lots of inconsistencies are listed, but they almost all have to do with the glosses.  Some summary statement should be made of this. This seems to be confirmed by the later Table 9, and perhaps a forward reference should be made.  A reasonable assumption is that  for the Arabic Treebank, ensuring gloss consistency was a lower priority than other aspects of the analysis, and I was surprised that there were so few non-gloss inconsistencies found.  It's obviously better to have the glosses consistent as well, of course.

\textbf{Response}\\
Sure. We indicated this in Section ~\ref{s:s:inc} (Inconsistencies) and added a forward reference to the table.
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.13}
6. p. 15, section 5.  The authors write ``We inspected the ATB v3.2 corpus for diacritics and we found that 1.364\% of the words were partially diacriticized and those diacritics eventually reduced morphological ambiguity."

This is not adequate.  What we need to know is not just that morphological ambiguity was reduced using the diacritics (wouldn't that almost have to be the case?) but also that the diacritics *were consistent with the annotation assigned to the word in the ATB morphological annotation*.  That is, the diacritic could simply be wrong, and so reduce the morphological ambiguity, but also eliminate the use of the diacritic that should have been used - the point (2) that they could be unreliable.  The authors don't answer that.

\textbf{Response}\\
We agree with the reviewer. 
We are not making the argument for the use of diacritics all the time for disambiguation. 
We think this should be optional for the following reasons.

In theory, partial diacritics help reduce ambiguity. 
Unfortunately, sometimes writers miss-place the diacritics for several reasons. 
\begin{itemize}
  \item They make mistakes (mostly syntax),
  \item They use a pronunciation of the MSA word that is common in their local Arabic dialect,
  \item They place the diacritic in an incorrect textual position, however, font effects such as kasheeda and 
    ligatures compensate for that and the diacritic shows visually correct, and
  \item They use diacritics as ornaments to decorate their text!
\end{itemize}
As for the ATB, the most common diacritics are tanween and shadda. These significantly and effectively reduce ambiguity. 
They are followed by dhamma associated with verbs to mark the passive tense. Which is also effective to reduce ambiguity.
We refer to ``Diacritization: A Challenge to Arabic Treebank Annotation and Parsing '' Maamouri-2006 since we found out they
report the same results.
$\blacksquare$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.14}
7. p. 15, section 5 - ``The Diacritic-aware consistency check algorithm takes as input two words w\_1 and w\_2...". I didn't understand this. What are the words being input?  Is this filtering the possible solutions for a word ignoring the diacritics with the diacritics that are actually present?  This needs to be made clearer.

\textbf{Response}\\
The algorithm takes two chunks (not necessarily full words). 
It is called several times in the traversal when a morpheme is found. 
It eliminates solutions that do not match the partial diacritics during the traversal and does not filter them later.
We added a paragraph at the end of Section~\ref{sec:diac} to clarify the issue. 
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.15}
8. p. 18 - comparison with Beesley\&Kartunnen.  One of the questions I had reading section 2 is how this is different than finite-state machines. This sort of discusses it, but it doesn't seem to me like a fair comparison.   The authors claim that the Xerox approach ``requires deep knowledge..." etc. while Sarf ``constructs a framework of efficient
structures that encode the stems...".   
But Sarf relies on categories and compatibility tables taken from SAMA! (as best as I can tell).

\textbf{Response}\\
Sarf encodes the extended tables and rules from SAMA into structures that can be traversed linear in the length of the input string.
Those are comparable to finite state machines since the nodes are states and the edges represent transitions. 
The main difference is that Sarf structures are manually optimized to minimize (1) number of states (storage) and (2) non-deterministic transitions that are limited to transitions between ${\cal P}$, ${\cal S}$, ${\cal X}$.
No control on non-determinism is possible with Xerox and the result is left to the compilers. 
We further clarified this point. 
$\blacksquare$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.16}
9. Some questions arise about the evaluation in section 7.1.

a. The evaluation should not be done with v3.1 of ATB3.  It's not even a public release. ATB3v3.2 should be used, which is referenced elsewhere in the paper.

b. Does the 99.991\% segmentation agreement with the ``oracle ATB segmentation" not include the words for which SAMA did not have a correct morphological solution?  Does ``correct morphological solution" include the gloss, which
seems irrelevant for this (and is not part of such statistics in the  ATB documentation.)?

c. first sentence, p. 21 - ``..generating a vocalized tag in the ATB 'after' dataset, which carries more information in 15.47\% of the time than the counterpart POS derived vocalized entry. The vocalized entry in the 'after' dataset was dropped because of maintenance and segmentation issues."

I am very familiar with this data, and I do not understand what the authors are referring to.  An example is badly needed.

\textbf{Response}\\
We used ATB3.2. 3.1 was a typo.  The 99.991\% is over the words that have a SAMA correct solution and it does not include the gloss. We clarify that in the text.

We thank the reviewer for bringing up the issue with the vocalized tag in the after dataset. 
We could not directly recall where this claim came from. 
We must have meant the UNSPLITVOC entry that relates directly to the word and not the vocalization that is built
from concatenating the vocalized segment (VOCALIZED tag in the 'after' data set).

After inspection, we think it is a misunderstanding (maybe due to several rewrites of that paragraph) 
on our part to the following quote from the ATBv3\_2 documentation discussing the 
challenges and difficulties associated with defining and computing the INPUT\_STRING and UNVOCALIZED tags.

\blockquote{
... We plan
for future releases to continue utilizing the present improved creation of the
INPUT STRING tree tokens, as is done in this release.  However, this is not 
part of the annotation process itself, as explained above, and it is possible 
that future releases either will not include extensive checking on the
creation of these INPUT STRING tree tokens, or will {\bf leave out completely
such tokens}.
}

We removed the claim ``The vocalized entry in the 'after' dataset was dropped because of maintenance and segmentation issues" 
from the paper.
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.17}
11. p. 22, Augmentation - this is a good point, but at the same time, SAMA did not want to generate all possible solutions because it might overwhelm the annotator.  This is less of an issue if a tagger such as MADA, ranking the SAMA output, is used.

\textbf{Response}\\
We agree.
$\blacksquare$
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.18}
References:
1. The reference for SAMA is not the best one.  It's okay to refer to (Kulick, Bies \& Maamouri 2010a), but the actual release should also be cited, with the citation information listed here: https://catalog.ldc.upenn.edu/LDC2010L01

\textbf{Response}\\
We used the indicated reference. 
$\blacksquare$

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\item \label{Review.3.19}
Typos/confusing language
p. 2, bottom ``cosistency"
p. 6, top - ``and the use diacritic option and constructs..."
This is confusing. Either put quotes around the ``use diacritic option" or make it something like ``And the option to use existing diacritics, and  constructs..."
p. 14, (k) - ``...if each belongs to an one In SAMA..."

\textbf{Response}\\
Fixed. 
$\blacksquare$

\end{enumerate}

### 2018-06-26 tex/sarf/abstract.tex
sarf to ACLing draft added

The rich nature of Arabic morphology makes morphological analysis 
key for Arabic natural language processing applications. 
%
Arabic morphological analyzers return several
morphological solutions for a given Arabic word. 
Each solution consists of several morphological features
such as part of speech and gloss description
tags. 
%
Often times, applications need only few of those features.
%
This paper presents Sarf, an application customizable 
morphological analyzer for Arabic. 
Sarf provides an interface that allows application developers to 
(1) control and prioritize the analysis, 
(2) refine solution features, and
(3) define categories and associate them with existing morphemes.

Sarf uses agglutinative and fusional morphemes for
affix representation,
and extends and refines the morpheme lexicons of SAMA and BAMA.
This reduces redundant morphemes, and subsequently inconsistent 
morpheme tags in the lexicons.
It also solves the segmentation correspondence problem between 
an input word and the several parts of the associated morphological 
solution. 
It uses diacritics to refine solutions, and 
solves the `run-on words' problem. 
%
The implementation of Sarf efficiently encodes
the morpheme lexicons.
Sarf was used in several NLP 
applications for information extraction and provided more accurate
solutions than existing solvers with faster running time.

### 2018-06-26 tex/sarf/affixtable.tex
sarf to ACLing draft added

--- 2011-04-19 trunk/tex/tp.tex
@@
-\documentclass[a4,12pt]{report}
-\usepackage[Bjornstrup]{fncychap}
-\usepackage{graphicx}
-\usepackage{epsf}
-\usepackage{a4}
-\usepackage{fancyhdr}
-\usepackage{setspace}
-\usepackage{amsmath}
-\usepackage{amsfonts}
-\usepackage{amssymb}
-\usepackage{enumerate}
-
-\usepackage{caption}
-\usepackage{epsfig}
-\usepackage{subfigure}
-\usepackage{colortbl}
-\usepackage{color}
-\usepackage{rotate}
-\usepackage{rotating}
-\usepackage{amsthm}
-\usepackage{booktabs}
-\usepackage{slashbox}
-
-\usepackage{relsize}
-\usepackage{fancyvrb}
-\usepackage[colorlinks=false]{hyperref}
-
-\usepackage{url}
-\usepackage{arabtex}
-
-\usepackage{utf8}
-\setarab
-\fullvocalize
-\transtrue
-\arabtrue
-
-%\newcommand{\CharCodeIn}[1]{`\CodeIn{#1}'}
-\newcommand{\CodeIn}[1]{{\small\texttt{#1}}}
-\newcommand{\frl}[1]{\fbox{\RL{#1}}} 
-\newcommand{\noArRL}[1]{\arabfalse\RL{#1}\arabtrue} 
-\newcommand{\noTrRL}[1]{\transfalse\RL{#1}\transtrue}
-\newcommand{\noTrnoVocRL}[1]{\transfalse\novocalize\noTrRL{#1}\vocalize\transtrue}  
-\newcommand{\noVocRL}[1]{\transtrue\novocalize\RL{#1}\vocalize}
-
-\newcommand{\X}{\cellcolor[gray]{0.5}}
-\newcommand{\x}{\cellcolor[gray]{0.8}}
-
-%Header and Footer
-\fancyhead{} \fancyhead[LO,RE]{\scriptsize\sc\leftmark}
-%\fancyhead[R]{Jad Makhlouta}
-\fancyfoot[l]{ \scriptsize \textit{AUB, Spring 2010-2011}}
-\fancyfoot[R]{ \scriptsize \textit{ Thesis Proposal Report}}
-
-%-------BEGIN----------------------------------------------------
-\begin{document}
-
-%----------------Report Cover--------------------------------------------------
-%\include{parts/cover}
-\begin{titlepage}
-%\vspace{1cm}
-\begin{center}
-
-% \vspace{1cm}
-
-\large{Department of Electrical and Computer Engineering\\
-Faculty of Engineering and Architecture \\
-American University of Beirut\\}
-\vspace{3cm}
-%\begin{figure}[h]
-%    \begin{center}
-%       \includegraphics[scale=0.25]{Fig/aublogo.eps}
-%    \end{center}
-%\end{figure}
- \LARGE \textbf{Application-Specific Arabic Morphological Analyzer and \\Automatic Hadith Authentication\\}
-\vspace{1.5cm}
-\LARGE {Thesis Proposal \\ Spring 2010-2011}\\
-\end{center}
-\vspace{3.5cm}
-\begin{tabbing}
-
-Submitted by: \quad\quad \quad \quad \=Jad Makhlouta, 200700856 \\
-\\
-Thesis Supervisor:  \> Dr. Fadi Zaraket \\
-
-Committee Member: \> Dr. Hasan Artail  \\
-
-Committee Member:  \> Dr. Ibrahim Abou Faycal \\
-
-\\
-\\
-\\
-Submitted on: \> \today
-\end{tabbing}
-\end{titlepage}
-
-% Select arabic numbering style
-\pagestyle{fancy}
- \pagenumbering{arabic}
-\setcounter{page}{1}
-
-\begin{abstract}
-Text Mining 
-%can be considered as a subfield of both Natural Language Processing (NLP) and Data Mining.
-is interested in automatically extracting useful information and relations from text documents.
-%Extracted information is hard to discover manually or previously unknown
-%Natural Language Processing applications perform 
-%to discover useful information that is otherwise unknown.
-The complexity of natural languages introduces interesting challenges 
-to automatic analysis of underlying textual data.
-Morphological analysis is key to many Natural Language Processing Applications.
-General morphological analyzers such as Buckwalter and ElixirFM 
-generate all morphological solutions.
-However, this exhaustive enumeration may not be needed or appropriate for the
-application at hand.
-We propose developing Sarf, an Arabic application-specific morphological analyzer whose functionality can be customized 
-based on the specific needs of the application. 
-%This flexibility has been motivated by 
-%the broad spectrum of text mining applications, each with its specific characteristics and corpora, which
-%necessitate a special treatment.
-Sarf encodes its lexicon into a non-deterministic composition of three finite automata.
-It also solves the issue of `run-on' words, supports disambiguation using diacritics, 
-handles multi-word expressions and performs online tokenization.
-In our thesis, we will use Sarf to solve two useful Natural Language Processing applications: 
-(1) the extraction of time entities from Arabic documents, and 
-(2) the automation of the process of hadith authentication.
-
-\end{abstract}
-
-\chapter{Introduction}
-\label{s:intro}
-
-The world we live in overwhelms us with electronic data. There is an ever-increasing gap between the amount
-of information available and our ability to understand and analyze it. ``Omnipresent computers make
-it too easy to save things that previously we would have trashed. Inexpensive disks
-and online storage make it too easy to postpone decisions about what to do with all
-this stuff -- we simply get more memory and keep it all.''~\cite{Witten:11}
-This explosion in available data has motivated the automation of the task 
-of extracting information of interest for a certain usage. 
-Manual extraction is practically unfeasible due to
-our limited precious time recourses compared to the magnitude of the task.
-In addition, it was shown that automatic methods have been able to discover previously unknown 
-information through relating separate datasets~\cite{JNi06}.
-
-%Consequently, \textit{Data Mining} as a field of study emerged
-%as an attempt to tackle this problem which is manually unfeasible.
-%Data Mining in general is defined as the task of `extraction of interesting 
-%(non-trivial, implicit, previously unknown and potentially useful) patterns 
-%or knowledge from huge amounts of data'~\cite{Witten:00}.
-
-Clearly, a small percentage of the data available worldwide 
-is stored in databases in structured form; instead, the greater amount of information is available in the 
-form of electronic text files and documents such as news articles, publications, governmental records, etc. ...
-Even old paper documents are being ported to digital form to facilitate their use. 
-Text Mining emerged as an attempt to automate the extraction of interesting, 
-previously unknown and potentially useful information
-from the huge amounts of text documents~\cite{Witten:08}.
-Information Extraction (IE) from natural languages, however, is challenged
-by the difficulty of getting structured 
-entities from the underlying unstructured textual data.
-
-%Information Extraction (IE) from Natural Language Texts has always been an important 
-%problem in Data mining. What distinguishes text mining from the 
-%other subfields of data mining is mainly due to the challenge of getting structured 
-%entities from the underlying unstructured textual data. Clearly, a small percentage of the data available worldwide 
-%is stored in databases in structured form; instead the greater amount of information is available in the 
-%form of electronic text files and documents such as news articles, publications, governmental records, etc. ...
-%Even old paper documents are being ported to digital form to facilitate their use.
-
-Natural Language Processing (NLP)
-as a field of research is interested in achieving a computational model of the language
-and understanding its underlying meaning~\cite{sharp:01}. It is known that this task
-is very complex and possibly not completely achievable. However, in most applications it 
-is not needed to grasp the whole meaning of the text but to extract some useful entities.
-This Information Extraction(IE) problem can be performed through some statistical approaches 
-and/or partial NLP techniques.
-
-A subproblem of NLP is morphology.
-Morphological analyzers%~\cite{Sughaiyer:04}
-are interested in understanding the internal structure of a word;
-It considers words as being composed of 
-several {\em morphemes}. 
-A morpheme is a {\em stem} or an {\em affix}.
-An affix is a {\em prefix, suffix,} or an {\em infix}.
-Prefixes and suffixes are attached to the beginning and end of the word respectively; 
-while infixes introduce changes in the stem itself. 
-Morphological analyzers have been extensively used as building blocks 
-in many text mining applications~\cite{Sou07}.  In Arabic, many morphological analyzers exist including:  
-Buckwalter~\cite{Buckwalter:02},
-Beesley~\cite{Beesley:01},
-SAMA ~\cite{Kulick:10},
-ElixirFM~\cite{Otakar:07}, 
-AMIRA~\cite{Diab:07,Benajiba:07},
-MAGEAD~\cite{Habash:05}, 
-and MADA+TOKAN~\cite{Habash:09}.
-%Arabic however has distinctive characteristics that
-%complicate its analysis; and consequently, much remains to be performed in the field.
-
-In this thesis, we propose to develop Sarf, an application-specific 
-morphological analyzer for the Arabic language 
-which is capable of benefiting
-from the specific characteristics of the application at hand. 
-In addition, it supports disambiguation using diacritics and 
-handles `run-on' words. We also propose to use Sarf to solve two interesting .
-text mining applications: time entity extraction and hadith authentication.
-
-\section{Applications of Text Mining}
-\label{sec:Applications}
-
-%Text Mining has gained much interest lately especially due to its wide range of applications.
-%However, these applications have substantially differing corpora (terminology, 
-%sentence structure, etc...). In this section, we present some of the 
-%previously explored applications of text mining.
-%
-%Research in text mining started in the mid 1980s when Swanson 
-%realized that slicing and combining seemingly unrelated medical 
-%articles led to the discovery of new 
-%hypotheses~\cite{JNi06}. 
-%Lately we saw applications of text mining in the advertisement, 
-%political campaigning, and other businesses.
-
-Research in text mining started in the mid 1980s when Swanson 
-realized that slicing and combining seemingly unrelated medical 
-articles led to the discovery of new 
-hypotheses~\cite{JNi06}. 
-Text Mining has gained much interest lately due to its wide range of applications,
-including its use in the areas of advertisement, 
-political campaigning, and other businesses.
-However, these applications have substantially differing corpora (terminology, 
-sentence structure, etc...). In this section, we present some of the 
-previously explored applications of text mining.
-
-\subsection{Sentiment Analysis} 
-	This field -- also known as Opinion Mining' -- attempts classifying the polarity of sentiments of people 
-	or even their emotional state ~\cite{Pang:08}. The inputs for such analyzers include product and movie reviews, 
-	Internet blogs, electronic social networks and possibly newspapers. Early research was conducted on review sites
-	to automate the task classifying whether the review was positive and negative. Consequently, users can get concise helpful
-	sentiment summaries about the product, movie, etc...~\cite{Pang:02, Turney:02}. Such abilities motivated research targeting 
-	the following applications:
-	\subsubsection{Marketing Applications} 
-		The strength of blogs in disseminating opinions is a two-edged sword that can 
-		``make or break a company's reputation''~\cite{hoffman:08a}. Consequently, analyzing customers' opinions as portrayed 
-		on online discussion forums should be detected as soon as possible. This input is vital in the company's policy for its
-		reputation management. It can accordingly take actions to retain current customers, attract new ones and win back former 
-		customers~\cite{hoffman:08b}. Sentiment Mining has also been used to predict customer churns~\cite{marketing:07,marketing:08}
-		to take appropriate actions. It can help also in producing more effective advertisements based on the interests of the market.
-	\subsubsection{Political Applications} 
-		Political experts and analysts can benefit also. Since opinions extracted from textual data 
-		can reveal much even without the need of performing any explicit surveys. According to ~\cite{Gryc:10}, it ``opens up many avenues
-		for more accurate and naturalistic large-scale political analysis.'' Some research has been performed on classifying and predicting
-		political inclinations~\cite{Durant:07, Yu:08}. This field can help political leaders to better understand the views of their people 
-		and possibly perform better campaigns.
-
-\subsection{Academic Applications} 
-	In scientific disciplines where written papers and publications contain technical information, the need to retrieve
-	specific information that a user needs, aids in \textbf{\em developing advanced search engines}. 
-	Text mining plays a great role toward this end through the following:
-	\subsubsection{Text Summarization} 
-		When an application is interested in mining a text for the most important and informative portions,
-		the application is considered to be tackling the field of `Text Summarization'. Summaries may have different forms such as being 
-		a group of representative words, to a coherent abstraction of the text in novel terms and phrases that capture its main meaning. 
-		The main motivation
-		for such a research is to enhance search engines. By capturing the most informative concepts in a document, it aids determining a 
-		better metric for a document's relevance to a search query. It also helps search engines display a shortened preview of the document for returned
-		results. This reduces the overwhelming amount of information a user needs to read before deciding weather the whole document deserves
-		comprehensive study. It is also helpful when the editorial staff of a magazine or newspaper need to present an article in the limited space.
-		Similarly, summarization is vital when the content of 
-		news articles needs to be dramatically shortened to be sent in an SMS form to subscribed PDA users. 
-		Useful surveys of current research on text Summarization can be
-		found in ~\cite{Radev:02,Das:07}.
-	\subsubsection{Text Categorization and Clustering} 
-		In many cases, it is very useful to be able to search for documents by theme, topic, domain, etc...
-		This ability is achieved if documents can be automatically clustered into different categories over which a user can search. It has been also 
-		proposed to cluster documents based on biographical data~\cite{Kostoff:99}. Technically speaking the difference between Categorization and
-		Clustering is that the former organizes documents into predefined categories where in the latter the categories arise from the document content
-		themselves~\cite{sharp:01}. Many approaches exist in this field as described in~\cite{Berry:03, Aas:99, MUNTEANU:05}.
-		\begin{enumerate}
-		\item \textbf{Email Classification:} It is noteworthy to mention that it is text categorization can be used for emails 
-			-- such as urgent, work, friends, spam, etc... Consequently, it can be used for spam detection and email 
-			filtering~\cite{Mock:01,bekkerman:04}.
-		\item \textbf{Parental Control:} It has also been proposed to use categorization for parental control by
-			automatically detecting inappropriate content for their children~\cite{Kontostathis:09}.
-		\end{enumerate}
-	\subsubsection{Entity Extraction:} Named Entity Extraction/Recognition (NER) targets being able to detect phrases that convey some general concept
-		and/or belong to a predefined set of categories such as persons, places, time, etc... It has been argued that having the luxury of searching
-		for entities enhances search abilities instead of being restricted to a list of words appearing in a text~\cite{basis:06}. 
-		It seems that humans think of their
-		queries in terms of entities/concepts and then try translating them to some list of words which are most likely to appear in a document that
-		is relevant to their search query. In general, a user does not ``know all the search terms that will return all instances of the desired
-		information''~\cite{basis:06}. A survey of research in the field is presented in ~\cite{Nadeau:07}.
-
-\subsection{Security Applications} 
-	It has been proposed that text mining of emails can be used for detecting money laundering and 
-	terrorist plans. However the effectiveness of data mining in this context is still debatable~\cite{Jonas:06,Federici:07}.
-
-	We also propose another security application which can be used to be used when investigating in criminal cases:
-	Given two sets of reports consisting of crime scene investigation and suspect interrogation reports, 
-	an officer may query use a specialized text mining application to query for relations between suspects in terms of 
-	criminal action, locality, or third party persons unknown to the officer.
-	It can be envisioned that such an application will return two convicted people who had no direct relation
-	in the reports, but who are related only through a third person unknown to the investigator. 
-	Such a query is very hard to answer manually.
-
-\subsection{Medical Applications} 
-	Medical records and clinical reports are studied to detect
-	patterns of occurrence of certain symptoms and diseases. It attempts relate those patterns with
-	other biological factors such as certain 
-	genes or protein interaction~\cite{blaschke:99, Krallinger:05, Rebholz-Schuhmann:07, Cohen:08, Tsuruoka:08}.
-	They are also used to analyze factors correlating with medical accidents to minimize their occurrence~\cite{Kimura:08}.
-	Till present, most of the interest in text mining comes from biological sciences~\cite{Giannopoulou:08}.
-
-\subsection{Hadith Authentication}
-
-In this thesis, we will present and propose to solve an interesting application 
-for text mining in the field of Islamic Studies. 
-A \noVocRL{.hady_t}~\footnote{In this document, 
-we use the default ArabTeX transliteration style ZDMG.}
-is a narration related to the prophet Mohammad
-through a \noVocRL{sanad} or a sequence of narrators. 
-The collections of traditions are the second source of
-jurisprudence after the \noVocRL{qor'An} for all Islamic schools of thought. 
-Figure~\ref{f:exhadith} shows an example \noArRL{.hady_t} in 
-Arabic with its transliteration and translation. 
-We show proper names in boxes connected
-to form complex names of narrators. 
-For example, 
-\novocalize
-\noTrRL{qtybT} is the first name 
-of narrator $n_1$, and 
-\noTrRL{s`yd} is the name of his father as 
-the word \noTrRL{bn} (son of) indicates. 
-The sequence of names from $n_1$ to $n_5$ 
-constitutes the \noArRL{sanad} 
-of the \noArRL{.hady_t}. 
-The second part of the \noArRL{.hady_t} is the 
-\RL{matn} (content) and constitutes the actual content 
-of the tradition.
-
-Due to religious and political reasons, 
-writing the traditions was forbidden 
-until the days of the eighth Umayad Calif,
-\novocalize
-\RL{`mr bn `bd al`zyz}(717-720 AC), seventy or so years after 
-the death of the prophet~\cite{AlAskari}. 
-Consequently, many inconsistencies were
-introduced to the literature which necessitated 
-a thorough authentication study of a tradition before its use in 
-jurisprudence.
 \vocalize
-An Islamic hadith scholar studies the authenticity of 
-the \noArRL{sanad} of a 
-set of related traditions before using these traditions 
-jurisprudence. 
-While different Islamic schools of thought differ on 
-how to interpret the content, they almost all agree
-that if a \noArRL{sanad} lacks authenticity, 
-then scholars can not use the tradition for jurisprudence.
-
-\novocalize
-
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{.9\columnwidth}{!}
-{ \input{figs/exhadith.pdftex_t}}
-\caption{Hadith abstraction example.}
-\label{f:exhadith}
-}
-\end{figure}
-% the FSMs for the three words
-\transtrue
-\vocalize
-
-The authenticity of a \noArRL{.hady_t} depends on 
-the credibility of the narrators as reported in 
-separate biography books. 
-The study of \noArRL{.hady_t} authentication is 
-currently manual and error prone due to the huge number
-of existing traditions and tradition books. 
-Hadithopedia~\cite{Hadithopaedia:08}
-reports that the tradition and biography
-books for one of the Islamic sects amounts to more than 
-300 thousand lines of text. 
-Al-Azami\cite{Al-Azami-91} cites more than eleven books
-of digitized tradition books each of several volumes, and a dozen
-other biography and secondary authentication books such
-as a geographical dictionary of places in hadith.
-
-In this our thesis, we present an application
-for authenticating inputted hadith based on inputted biography books.
-
-\section{Arabic Text Mining}
-
-Morphological analysis is key in current automated 
-analysis techniques for Arabic text.
-The almost comprehensive book 
-``Arabic Computational Morphology''~\cite{Sou07}
-covers substantial academic research work on Arabic text mining.
-It summarizes recent work, and presents strong evidence of 
-shortage of historic research. It also highlights  an emerging 
-interest with a specific application to machine 
-translation from Arabic to English after 2001.
-The research community showed little interest in 
-automating Arabic text analysis that may serve
-the Arabic user.
-While machine translation is of general interest, 
-an Arabic user is more interested in 
-information hard to discover manually.
-Examples are the relation between clinical records, pharmaceutical 
-prescriptions and sales, and a specific sickness,
-the relation
-between stock prices and political news, or the 
-relation between interrogation reports with suspects
-and investigative reports from the crime scenes.
-
-\subsection{Distinctive Features of the Arabic Language}
-\label{sec:ArabicDistinctive}
-
-In our work, we will be focus on morphological analyzers. 
-However it is noteworthy to mention that 
-Arabic analyzers face many difficulties
-inherent to the morphological analysis of Arabic. 
-In what follows, we consider some distinctive features that
-render Arabic morphology more complicated than that of Latin languages
-such as English.
-
-\subsubsection{\ref*{sec:ArabicDistinctive}.1 High degree of Ambiguity}
-
-	Morphological ambiguity is a `notorious' problem for Arabic language~\cite{Kiraz:98}.
-	Ambiguity refers to the fact that one word may lead to several possible solutions.
-	As discussed in~\cite{Attia:08a}, the following are the reasons of genuine ambiguity\footnote{
-	Here we are discussing ambiguity that is inherent to the language. Other sources of ambiguity
-	may be due to over-approximations and relaxed compatibility rules in the morphological analyzer.} 
-	in the language:
-	\begin{enumerate}
-		\item \textbf{Orthographic alternation operations:} These refer to alterations from the stem
-			original letters upon inflection, which can make it similar to other distinct words.
-			For instance, \noVocRL{ya`id} can result in 5 different solutions, 2 of which are
-			inflections of \RL{A`id} (bring back) and \RL{`od} (come back) as in \RL{lam ya`od}.
-			~\cite{Attia:08a}
-		\item \textbf{Prefixes and Suffixes that are homomorphic:} For example, 
-			``the prefix {\em ta-} can indicate 3rd person feminine or 2nd person masculine''~\cite{Attia:08a}.
-			Consider \noVocRL{tadros} which can mean `she studies' or `you (masculine) study'.
-		\item \textbf{``Coincidental Identity'':} In this case, complex forms (i.e. containing affixes) coincidentally
-			are homomorphic to a simple full word ~\cite{Kamir:02}. 
-			%An example is in the homomorphism of \noVocRL{Akala}
-			%he ate) and \noVocRL{Akalla} [\noVocRL{A} + \noVocRL{kalla}] (he made tired).
-			\vocalize 
-			For instance, the word \RL{'a.hmadH} may have two valid morphological analyses. 
-			The word means ``I praise him'' when the letter \RL{'a} is a prefix and  it means
-			``his Ahmad'' when the letter \RL{'a} is part of the proper noun stem \RL{'a.hmad}.
-		\item \textbf{Natural Homomorphism in Un-inflected Forms:} This occurs in \noVocRL{_dahaba} which can mean
-			`went' as a verb and `gold' as a noun.~\cite{Attia:08a}
-		\item \textbf{Omitted Diacritizations} (will be discussed next)
-	\end{enumerate}
-
-	The statistical measure of ambiguity is the average count of valid solutions per word. According to our results performed
-	on 3 books of hadith~\cite{IbnHanbal,AlKulayni,AlTousi}, the resulting ambiguity score was 2.62 using an extended corpus to that of
-	Buckwalter Analyzer~\cite{Buckwalter:02}\footnote{This analyzer will be considered in Section \ref{sec:literature}.}.
-	Those results where verified by ~\cite{Attia:06b} which reported a similar score (2.6) and a score of 1.76 for English.
-	This difference can be used to argue that Arabic is more ambiguous than English. Still however, the author of ~\cite{Attia:06b}, Attia,
-	claims that this ambiguity resulted from some grammatical over-approximations and inaccuracies in the lexicon of Buckwalter along with
-	its inclusion of classical terms that are rarely used. By this he means that the score does not reflect the intrinsic Arabic ambiguity and 
-	they report a score of 1.75 for their tool which used tighter grammar and lexical rules and omitted classical words. Although, Dr. Attia
-	has a point relating to exaggerated numbers due to over-approximations, the reduction of the lexicon is not a correct approach since it does
-	not reflect the ambiguity of Arabic but of the subset of Arabic needed for some applications (mainly modern text documents such as news).
-	In all cases, this is a question open to future investigations by NLP researchers and Arabic linguists...
-
-\subsubsection{\ref*{sec:ArabicDistinctive}.2 Diacritics}
-
-	It is common practice to write Arabic text without short vowels (diacritics) 
-	which greatly increases its ambiguity. For instance, \noVocRL{jzr} can have very different interpretations based on 
-	the fashion in which it is vocalized. Consequently, 
-	\RL{jazar} is a noun that means carrots; 
-	\RL{jozor} is a noun that means islands; 
-	\RL{jaz"r} is a noun that means sink/ebb typically referring to the state of the sea as in \noVocRL{Al-mad wAl-jazr};
-	\RL{jazara} is a verb in the past tense that means slaughtered/butchered.
-
-	Since most Arabic documents are unvocalized, the ambiguity of the morphological analyses is significantly increased.
-	For instance, in many cases, it is impossible to distinguish between active and passive verbs on the word level without
-	appropriate diacritics such as the undiacriticized versions of \RL{`abada} ($<$subject$>$ worshiped) and \RL{`obida} ($<$subject$>$ was worshiped).
-
-	Also, removing the shadda which is commonly assumed to be a diacritization removes practically a repeated letter as in \RL{`allama} (taught)
-	and results in ambiguation with other words having a totally different meaning such as \RL{`alima} (learned)~\cite{Attia:08a}.
-
-\subsubsection{\ref*{sec:ArabicDistinctive}.3 Run-on words}
-\label{sec:ArabicDistinctive:runon}
-
-	Arabic letters can have up to 
-	four different forms
-	corresponding to their position in a word, i.e. beginning,
-	middle, end of word and separate forms. 
-	This allows the phrase
-	\noTrnoVocRL{il_A\nospace almdrsT}  with no delimiter within
-	to be visually recognizable
-	as two separate words \noVocRL{il_A} (to) and \noVocRL{almdrsT} (the school).
-	This happens because the first word \noVocRL{il_A} ends with
-	\noVocRL{_A} a non-connecting letter. 
-	The non-connecting letters are \RL{A}, \noVocRL{'a}, \noVocRL{'i},
-		\noVocRL{'A},\noVocRL{_A}, \noVocRL{w}, \noVocRL{|u'|}, \noVocRL{|i'|},	
-		\noVocRL{d}, \noVocRL{_d}, \noVocRL{r}, \noVocRL{z} and \noVocRL{T}.
-	These words occur often in text and greatly increase the
-	difficulty of tokenization and are referred to as 
-	``run-on'' words~\cite{Buckwalter:04}.
-	Such difficulties are inherent in the 
-	morphological analysis of the Arabic language and
-	render the solutions inaccurate.
-	Consequently, tokenization of Arabic text is not a trivial task of 
-	detecting delimiters as is the case in other languages.
-
-\subsubsection{\ref*{sec:ArabicDistinctive}.4 Multi-word Expressions} % not totally distinctive
-
-	Multiword Expressions (MWEs) are defined in~\cite{MWE} as
-	``idiosyncratic interpretations
-	that cross word boundaries (or spaces)''. 
-	By this term, we refer to frequent cases when a group of words form together a single
-	item with a semantical meaning not captured by its individuals words compositionally. 
-	We focus on the example of compound names such as 
-	\noVocRL{`bid alkarym}. 
-	In current analyzers, this phrase is analyzed word by word and in best cases is 
-	considered to contain two names, instead of being understood to be a single compound name.
-
-\subsubsection{\ref*{sec:ArabicDistinctive}.5 Inflectional Nature}
-
-	It is known that ``Arabic is a highly inflectional language, which makes the morphological
-	analysis complicated'' ~\cite{Attia:08a}. In other words, instead of forming words from stems, the
-	structure of the Arabic language is built from roots. That's why Arabic dictionaries are indexed by the roots.
-	The Arabic stems are formed by applying some inflectional patterns that modify the meaning of the word.
-	For instance, the root \noVocRL{drs} can derive the following stems: \RL{darras}, \RL{dAris}, 
-	\RL{modarris}, \RL{madrows}, \RL{madrasaT}, etc..
-	However, in Latin languages the inflectional forms don't transform the meaning completely and can generally be treated
-	as prefixes and suffixes. In such inflectional languages such as Arabic, concatenative analysis sometimes fails to capture the
-	word structure especially when occur inside the stem itself caused by infixes. In this our thesis, we will not focus on this problem,
-	however we have mentioned it for completeness.
-
-\section{Proposal Organization} %TODO:
-
-The rest of the proposal is organized as follows. In Section 2, we present the problem(s) to be attempted. 
-In Section 3, we review related literature. 
-Then, in Section 4, we describe our preliminary results. 
-Finally, in Section 5, a time-schedule and a research plan are presented.
-
-\chapter{Problem Description}
-
-In what follows we describe the problems that we would like to tackle in our thesis:
-
-\section{Enhance Morphological Analysis Methods}
-
-We propose developing `Sarf', a novel morphological analyzer that introduces many enhancements to the
-field of morphological analysis. As noted in Section~\ref{sec:ArabicDistinctive}, 
-Arabic poses many difficulties which currently 
-have not been tackled in a satisfactory way. Of these we focus on the issues of `run-on' words and
-the use of diacritics in morphological analysis -- two issues which have been under-looked by researchers in the field.
-We will also work on speeding morphological analysis and 
-enhancing affix representations to better reflect the semantics of underlying word structure.
-Morover, we target supporting for interaction between the different text mining applications 
-and the morphological analyzer, 
-allowing the applications to drive the analysis and prune it on the fly. 
-We will start first by justifying the need for the latter flexibility.
-
-\subsection{Application-Specific Analysis}
-
-Current Analyzers are unable to benefit from the characteristics of the application in which the Analyzer is used.
-Their output is the same regardless of the needs of the application used and the characteristics of the corpus
-being analyzed. We noted in Section~\ref{sec:Applications}, that
-there is a broad spectrum of applications to which Text Mining is applied. In most cases, morphology 
-``plays an essential role in any higher-level understanding and processing of Arabic text"~\cite{Sou07}.
-
-Our focus when building Sarf, was not on improving the semantical rules and 
-expanding the morphological alternatives, instead we worked on adding support 
-for higher level control by the specific application utilizing the lower level 
-morphological analyzer. In other words, the application-specific controller, has 
-full control over and directs the morphological analysis of Sarf. 
-% (sentence should be reworded differently, but is needed to explain why we mention 
-% higher and lower accuracy hypothesis, which seems vague otherwise)
-
-We justify the above enhancement as follows:
-
-\subsubsection{Reduction in the Required Analysis}
-
-We have noted that many NLP case studies need the 
-morphological analyzer to answer queries that do not need 
-high accuracy at the morphological analysis level.
-The analysis at the higher levels 
-(i.e. answering the application-specific query)
-can often compensate for 
-tolerable inaccuracy at lower morphological levels. 
-For example, if a query concerns proper names and the 
-analyzer is considering a 
-prefix that connects only to a verb,
-%the analyzer can answer without considering the 
-%rest of the word.
-the analyzer does not need to further analyze the rest of the word 
-to provide a negative answer.
-If another query is interested in extracting time entities, the only 
-information that it needs to extract from the morphological analyzer 
-is whether the word under consideration has a stem which reflects 
-time structure; and whenever a positive answer is verified by one of the analysis,
-the analyzer can stop without computing the other alternative solutions.
-
-In other words, if the analyzer can be controlled at every step by an application-specific
-controller, the analysis can be enhanced and reduced significantly.
-
-\subsubsection{Evidence from Related Works}
-
-\begin{enumerate}[a)]
-
-\item We find evidence to our hypothesis in~\cite{Maamouri:10}.
-The addition of a new corpus to the Arabic Tree Bank 
-(ATB)~\cite{Maamouri:04}
-%required compliance with the DARPA GALE program 
-%including several NLP tasks such as English translation 
-%annotations and word-level alignment of Arabic and English. 
-%The new data 
-challenged the existing processing and 
-annotation guidelines. 
-This led to refining 
-the analyzer SAMA~\cite{Kulick:10} with 
-alternative guidelines. 
-%This refinement was achieved only after the NLP 
-%task interacted with the morphological analyzer,
-%a flexibility our analyzer is designed to provide for the application-specific controller on the run. %(check if this sentence makes sense)
-It reports that a refined version of SAMA~\cite{Kulick:10} was achieved when the NLP 
-task interacted with the morphological analyzer, which is
-a flexibility our analyzer is designed to provide for the application-specific controller on the run.
-
-\item In addition, 
-the work of Habash~\cite{Habash:06} reports that different 
-types of morphological analyses behave differently on the same %case study. 
-data set.
-The analyzers may report different solutions or 
-may present the solutions in different orders. 
-Clearly, some of 
-the presented solutions and their order of presentation may not be 
-appropriate for the %case study.
-application needs, which is an issue that Sarf tackles. % (maybe reworded)
-\end{enumerate}
-
-\subsubsection{Possible Reorder of Analyzer Structures}
-
-Analyzers provide different levels of features for the alternative solutions.
-Commonly, those features include: the vocalized version corresponding to the analysis,
-its Part of Speech (POS), its gloss (i.e. semantical meaning) and other abstract categories to which
-it belongs (e.g. Proper Name, Place, Time, etc...)\footnote{For example, the analyzer Attia 
-provides the following information for Nouns: Gender, Number, Case, Humanness, its possible
-categorization as proper noun (subdivided into person, place, organization), etc..~\cite{Attia:08a}.}.
-If it is known beforehand by the application if one feature is its primary interest, using which
-early termination of a word analysis can be interpreted, the analyzer can reorder its structures to support for
-faster lookup for this feature.
-
-For example, by default in Sarf, the features are ordered in the following hierarchy: 
-vocalized entry, abstract categories, gloss, then POS. In this way, Sarf will store corresponding vocalized entries per 
-solution and for each of them will report the different abstract categories that pertain to this (solution, vocalization) tuple,
-and so on...
-In this case, if some analysis is interested in solely the part of speech of the words considered, an advanced 
-application-specific analyzer can re-order its internal maps, trees, and lookup tables to report 
-that information first, which will significantly speed up its retrieval.
-
-\subsubsection{Refine Reported Solutions}
-
-We argue also that by introducing an application specific controller that is capable
-of intervening with the morphological analyzer at every decision, the reported query results 
-can be refined significantly. This task can be much more difficult otherwise.
-
-For example, assume we are
-building an application that verifies that the output of an OCR document is correct by checking
-that the scanned words are morphologically correct and exist in an exhaustive Arabic lexicon.
-To make the problem more interesting, we will assume that our application can benefit from the morphological analyzer to 
-improve its interpretation of the scanned glitch by refining its interpretation on the fly, restricting its output
-to words that exist in the lexicon.\footnote{This application was inspired from the works of~\cite{Badr:95,Obaid:98} that
-used benefited from the lexicon while performing the OCR; their methods, however, is not totally applicable to our
-hypothetical application.}
-We can envision that for a particular font used in some documents, 
-the default OCR engine misrepresents a character 
-by another with a high probability. Also it may be the case that word 
-boundaries are confused in some other way than `run-on' words, possibly splitting a word into two \cite{Kukich:92}.
-
-In such cases, if the analyzer's actual functionality cannot be altered, the application cannot 
-benefit much from the output of the analyzer. On the other hand, if an application-specific controller
-can be written to represent those peculiarities in the application at hand, the analysis will become 
-of high utility. The interventions need not be complicated. For instance, in the above examples, all that might 
-be needed is to intervene whenever any of the frequently confused letters is encountered in the analysis 
-to continue the analysis with either of the possibly confused letters assumed as matching.
-%In other words, refine the state of the machines to assume continue as if the alternative letter is encountered.
-% (reword this example, very bad structure)
-Another check might be needed when a space is encountered to solve the issue of a split word.
-In this case, the controller upon detecting a white-space can drive another analysis of the expression by directing 
-the analyzer to ignore the white-space and check if a valid solution will be generated.
-
-In some cases, the application would consider the high number of run-on words that can be generated as possible solutions
-as an unnecessary over-approximation. In that case, it can modify its controller to report only the `run-on' words it considers
-interesting or most frequent. This can be done by possibly applying some intelligent heuristics such as ignoring all run-on words
-which contains a sub-word of size: equal to one except for the clitic \noVocRL{w}; equal to two except for \noVocRL{mA} 
-and \noVocRL{lA} which appear frequently in run-on words and so on... Such statistical refinements, when handled early on during
-the analysis of the subparts of the word through the application-specific controller, are clearly much more affective than 
-a later post-processing step.
-
-~\\
-
-We hypothesize that in many NLP applications, the morphological analyzer may
-benefit from the characteristics of the query at hand, 
-to greatly enhance the analysis.
-Consequently, we believe that 
-an application-specific 
-controller that intervenes at every decision to guide, use, order, and 
-refine the morphological analysis is of high utility.
-
-\subsection{Solving the Issue of `Run-on' Words}
-
-`Run-on' words as described in Section~\ref{sec:ArabicDistinctive:runon} represent multiple words
-that are not separated by a delimiter but are visually conceived as non-connected separate words.
-
-This problem is especially important due to the fact that many of the valuable 
-Arabic texts available for text mining have been originally ported from
-paper documents. OCR is often used to automate this task; however, this task
-results in mistakes that are often escape manual inspection.
-In particular, run-on words are admitted 
-to be hard to detect especially when the `glyph design (i.e., the
-perceived width) of the non-connecting letter' seems comparable
-to that of a white-space~\cite{Sou07}. In his PhD thesis, Mohammad S. M. Khorsheed
-explained the OCR difficulty that arises: ``Arabic scripts are inherently cursive...
-Since not all characters connect, word boundary location becomes an interesting problem, 
-as spacing may not only separate words but also certain characters within a word.''~\cite{Khorsheed:00}
-
-OCR and manual errors of typing are not the only source for such mistakes. As the field
-of speech mining gains interest and many systems that translate speech to text for parsing it
-are getting prevalent, such errors are getting more prevalent. For instance, a study in~\cite{Kukich:92}
-has shown that 13\% of nonword-spelling errors which have been generated when transforming
-conversations to text are `run-on' words. Although this study was performed on English-like concept of
-`run-on' words\footnote{such as `{\em willsee, weare, iam, onthe, yesthisis}'~\cite{Kukich:92}} 
-which are unarguably assumed as errors, in Arabic readers generally don't assume such
-expressions as being spelling mistakes~\cite{Buckwalter:04}.
-
-The importance of this characteristic in Arabic morphology can be appreciated 
-through the frequency statistics reported in~\cite{Sou07}. It reports some statistics
-from Google Search Engine. Obviously, many ``run-on'' words have a high frequency in online 
-Arabic texts exceeding the count of thousands. Another noteworthy observation that is reflected
-in the collected numbers is that those frequencies increased in magnitudes from 2004 to 2006; this 
-suggests that the solving this difficulty is a problem that deserves research attention, instead of 
-one that is related to some old restricted group of documents. Also as noted in~\cite{Buckwalter:04},
-lexicalized collocations which are spelled without intervening space -- such as \noVocRL{lAyzAl}, 
-\noVocRL{lAbod}, etc... -- are ``just as frequent as their spelling in concatenated form".
-
-This difficulty in tokenization has been reported by Buckwalter in~\cite{Buckwalter:04} because of its 
-frequency in three corpus of the Arabic Tree Bank on which some experiments were performed. 
-That paper was concluded that ``the problem of run-on words in Arabic calls for
-a reassessment of current tokenization strategies,
-including the definition of `word token' itself. It
-should be assumed that each input string represents
-one or more potential word tokens, each of which
-needs to be submitted individually for morphology
-analysis.'' In this thesis, we reassess current tokenization strategies and solve the issue
-of run-on words in a very efficient way to be described later in Section~\ref{sec:runon}.
-
-\subsection{Disambiguation using Diacritics}
-
-Up to our knowledge, currently all Arabic morphological analyzers have ignored diacritics due to the complexity 
-of their analysis and the tendency of Arabic writers to omit them.
-Nonetheless, many Arabic documents especially religious texts are fully vocalized.
-Even in documents written without diacritics, we often notice 
-a couple of words which are partially vocalized
-to reduce the difficulty of their disambiguation by the reader. Clearly, if native readers
-need the aid of short vowels, automatic morphological analyzers which are context-free should not lose this precious piece of information.
-
-\subsection{Support for Multi-Word Expressions} % maybe ommit %TODO
-
-Many Arabic proper names are compound names which are composed of multiple words such as 
-\noVocRL{bayt la.hem}, \noVocRL{`bd al-razAq}, etc...
-It is important for the morphological analyzer to analyze them as one unit to be able to
-correctly reflect their part of speech information. Especially for the hadith case study,
-where the extraction of names of people is an essential task, having Sarf supportive of
-multi-word expressions will significantly enhance the accuracy of the results.
-
-\subsection{Enhancing Affix Representation}
-
-\begin{table}[tb]
-\centering
-\caption{Illustrative Subset of the Prefix Lexicon of Buckwalter's Analyzer}
-\resizebox{1\columnwidth}{!}{
-\begin{tabular}{|c|c|c|c|c|} \hline
- ~ & \textbf{Vocalized} & \textbf{Compatibility}  & ~ & ~ \\
-\textbf{Prefix} & \textbf{Prefix} & \textbf{Category} & \textbf{Gloss} & \textbf{POS data} \\ \hline \hline
-\RL{w} & \RL{wa} & Pref-Wa & and & wa/CONJ+ \\
+\begin{table}[!tb]
+\begin{minipage}{\textwidth}
+\relsize{-1.5} {
+\begin{tabular}{ccccc} 
+\hline \hline
+\textbf{Prefix} & \textbf{Vocalized} & \textbf{Category} & \textbf{Gloss} & \textbf{POS data} \\ \hline
+\RL{f} & \RL{fa} & Pref-Wa & and/so & fa/CONJ+ \\
 \RL{y} & \RL{ya} & IVPref-hw-ya & he/it & ya/IV3MS+ \\
-\RL{wy} & \RL{waya} & IVPref-hw-ya & and + he/it & wa/CONJ+ya/IV3MS+ \\
 \RL{fy} & \RL{faya} & IVPref-hw-ya & and/so + he/it & fa/CONJ+ya/IV3MS+ \\
 \RL{sy} & \RL{saya} & IVPref-hw-ya & will + he/it & sa/FUT+ya/IV3MS+ \\
-\RL{wsy} & \RL{wasaya} & IVPref-hw-ya & and + will + he/it & wa/CONJ+sa/FUT+ya/IV3MS+ \\
-\RL{fsy} & \RL{fasaya} & IVPref-hw-ya & and/so + will + he/it & fa/CONJ+sa/FUT+ya/IV3MS+ \\
+\RL{fsy} & \RL{fasaya} & IVPref-hw-ya & and/so + will + he/it & fa/CONJ+sa/FUT+ya/IV3MS+ \\ \\
 
 \RL{y} & \RL{ya} & IVPref-hmA-ya & they (both) & ya/IV3MD+ \\
-\RL{wy} & \RL{waya} & IVPref-hmA-ya & and + they (both) & wa/CONJ+ya/IV3MD+ \\
 \RL{fy} & \RL{faya} & IVPref-hmA-ya & and/so + they (both) & fa/CONJ+ya/IV3MD+ \\
 \RL{sy} & \RL{saya} & IVPref-hmA-ya & will + they (both) & sa/FUT+ya/IV3MD+ \\
+\RL{fsy} & \RL{fasaya} & IVPref-hmA-ya & and/so + will + they (both) & fa/CONJ+sa/FUT+ya/IV3MD+ \\  \\
+
+\RL{w} & \RL{wa} & Pref-Wa & and & wa/CONJ+ \\
+\RL{wy} & \RL{waya} & IVPref-hw-ya & and + he/it & wa/CONJ+ya/IV3MS+ \\
+\RL{wsy} & \RL{wasaya} & IVPref-hw-ya & and + will + he/it & wa/CONJ+sa/FUT+ya/IV3MS+ \\
+\RL{wy} & \RL{waya} & IVPref-hmA-ya & and + they (both) & wa/CONJ+ya/IV3MD+ \\
 \RL{wsy} & \RL{wasaya} & IVPref-hmA-ya & and + will + they (both) & wa/CONJ+sa/FUT+ya/IV3MD+ \\
-\RL{fsy} & \RL{fasaya} & IVPref-hmA-ya & and/so + will + they (both) & fa/CONJ+sa/FUT+ya/IV3MD+ \\ \hline
+\hline \hline
 \end{tabular}
 }
+\end{minipage}
+\caption{Partial BAMA v1.2 prefix lexicon}
+%\vspace{-2cm}
 \label{t:affixes}
 \end{table}
-
-Current affix representations are compound and contain highly redundant information. 
-These issues can be seen in Table~\ref{t:affixes}
-which contains a subset of the Prefix Lexicon of Buckwalter's Analyzer.
-The column for `Compatibility Category' is used to later validate
-if the affix is compatible with other parts of the word. The `Gloss'
-reflects the possibly compound description of the prefix. The `POS data'
-conveys the part of speech information.
-
-We call two morphemes compatible if their concatenation
-forms a legal morpheme. 
-Buckwalter~\cite{Buckwalter:02} keeps separate lists 
-of affix and stem morphemes and assigns categories to
-morphemes. 
-The relation between morphemes and categories is one 
-to many. 
-A compatibility rule is a pair of categories 
-$\langle c_1, c_2\rangle$  stating that $c_1$ morphemes
-can be concatenated with $c_2$ morphemes.
-Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
-of compatibility rules relating
-prefixes to stems, stems to suffixes, and prefixes to suffixes
-respectively. 
-Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
-a prefix, $\beta$ is a stem, and $\gamma$ is a suffix;
-$\alpha\beta\gamma$ is a 
-valid morphological analysis iff
-$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
-$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
-$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
-$c()$ returns the category of the morpheme.
-
-It is clear that in current analyzers, the list of affixes is much larger than needed
-due to the explosion of possible valid combinations that result from the most basic 
-prefixes. Note also that in current analyzers, sometimes to add just one prefix to the lexicon, it is required to
-manually generate all suitable combinations with those previously in the dictionary.
-For instance in Table~\ref{t:affixes}, the two syntactically and semantically different but homomorphic prefixes for \RL{ya}
-repeat the same combinations with other prefixes.
-
-Consider the case of the missing prefix from the lexicon of Buckwalter detected by Prof Attia. It is the 
-``clitic question morpheme which can be prefixed to verbs and nouns''(\RL{'a})~\cite{Attia:06b}.
-If we are to modify Buckwalter to support for its insertion that would result in reproducing most of the current
-prefix entries to concatenate to them this possible clitic. Having prefixes been recursive as in Sarf, just a single 
-prefix will be inserted along with some compatibility rules.
-
-As noted in Table~\ref{t:affixes}, many affixes are composed of other affixes. For example,
-the prefix \noTrnoVocRL{wsy} is composed of three other prefixes
-namely \noTrnoVocRL{w}, \noTrnoVocRL{s}, and \noTrnoVocRL{y}.
-The POS tag for \noTrnoVocRL{wsy} is a concatenation
-%of the POS tags of each of the three morphemes. 
-of those of the three morphemes. 
-Keeping all valid morphemes in a list
-is redundant and may lead to consistency issues when
-users try to add their custom morphemes.
-We introduce $L_{pp}$ and
-$L_{xx}$ as prefix-prefix and suffix-suffix 
-compatibility rules respectively.
-This allows us to keep a shorter list of affixes. 
-We also introduce a {\em resulting category}
-for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
-The resulting category in all practical cases
-is that of the second morpheme in the pair.
-
-Sarf only keeps atomic affixes and their categories.
-By doing so, the affix representation better reflects the rules
-which govern their formation. It also is able to better capture what each part of the compound prefix
-contributed to the gloss of the word and its POS. For instance, it is much more instructive to analyze
-and classify words based on their atomic affixes related POS and gloss instead of parsing string
-delimited entries.
-
-More than that, the fine scale of POS tagging and word division better suits any higher level attempt of semantical
-parsing at the sentence structure and better aligns with tagged corpora which commonly tag atomic affixes such as clitic 
-parts which are generally attached as prefixes. 
-%(check for some citation or paper evidence?)
-
-\subsection{Speeding the Analysis} %TODO:
-
-As will be shown later in Section~\ref{sec:timeefficiency}, Sarf has outperformed commonly used
-morphological analyzers. That has been made possible through many optimizations to the functioning
-of the analysis methods and implementation, of which we enumerate the following:
-
-\begin{enumerate}
-\item \textbf{Linear Traversal of the input text through the use of finite state automata:}
-	This will be described in Section~\ref{sec:Automata}.
-\item \textbf{Efficient use of data structures:} Lexicon entries are loaded into main memory 
-	when Sarf is started in appropriate structures that support efficient look-up. For instance,
-	for stem lookup, we used a double array trie (datrie)~\cite{Aoe:89}. We also extensively used
-	maps and hash-tables when needed. Also whenever an array reverse lookup was needed we stored
-	that lookup array for constant time access instead of a linear search in the original array.
-	Also when using a bit vector to access a specific bit, binary operations are used instead of
-	multiplication and division...
-\item \textbf{Reduced analysis and computations when not needed:} As discussed above, the 
-	application-specific controller can enhance performance by some early analysis aborting, and
-	non-complete retrieval of needed information for the analysis. For example, if for the application 
-	at hand, only stems are needed, affix information such as gloss and POS need not be retrieved.
-	Many other reductions can be envisioned suitable for particular applications of Sarf. 
-\end{enumerate}
-\section{Building a Time Entity Extractor}
-
-One of the applications we propose to build using Sarf is a time entity extractor. This application
-recognizes time expressions in text document and transforms them to time information of the default format
-that can be stored in a database and on which computations and analysis can be performed.
-
-Time structures which are conveyed in the content of a document instead of its meta-data 
-have always been an interesting problem for the research community \cite{GROSS:93,Llido:01,DSTO:05,patent:09}. 
-Li et al. working at the 
-Microsoft have been awarded a patent for having devised an application for extracting authorship 
-dates from documents~\cite{patent:09}. The importance of their work is justified by the fact that 
-in many cases the modification date of a document is not representative of its date of generation due 
-to many reasons. For example, ``when documents are uploaded or copied to
-collaborative websites, the date meta-data gets changed from the last modification date to the 
-upload date, which is rarely a significant or helpful date''~\cite{patent:09}. It is pointed also that any change in
-other meta-data fields results in automatically changing the modification date of the document although its content would have 
-remained intact. For such reasons, it is important to better approximate the authorship date 
-using the actual content of the document. This date can be used later in the search engines, when the user is interested 
-in documents that have been written in some time span. Obviously, a necessary capability for performing this analysis, 
-is being able 
-to extract temporal entities from documents.
-
-Apart from authorship dates,~\cite{Llido:01} considers the problem of automatic assignment of event-time periods.
-For many documents such as newspaper articles, medical reports and legal texts which contain many temporal 
-references, time periods of the events under consideration in the document may be of much higher importance than 
-the date of the document writing. This extracted information would facilitate the task of document retrieval.
-As proposed and partially implemented in~\cite{Koen:00}, extraction of time references from news articles can help
-augment those articles with personalized visualizations. Of those visualizations is a timeline containing all dates
-extracted from the article with the ability of to navigate to sentences from which a selected date from the timeline 
-is extracted. This tool also supports for augmentations which relate the events under consideration to
-other news articles about incidents that are temporally
-related to the one being read. Additionally, it provides an ability to display the historical context of an event and its relation
-to the personal context of the user reading the article. For instance, it is proposed that the tool
-be able to providing answers to questions like: `Where was I when that event took place?', `What other events took place
-at the same time worldwide or in my neighborhood?', etc...
-
-An interesting application based on time entity extraction was developed by a group of German researchers~\cite{Busemann:97}.
-They developed a system for Appointment Scheduling via Emails (COSMA). This system automates the time-consuming task of
-scheduling meeting between different parties having their own free/busy electronic schedule. The process of agreeing on
-a suitable time for all corresponding participants takes more than one round of communication. The automation of this task
-saves this time which is lost on such a routinely task.
-
-In addition, when time entities are extracted, other data mining techniques can help discover useful relationships.
-\cite{Berlanga:98} considers the problem of recognizing frequent temporal patterns in newspapers. While~\cite{Swan:99}
-considered extracting time varying features from the news corpus to discover interesting events. Also, the problem of detecting
-new evolving events in the news and tracking them was considered by~\cite{Allan:98}. One can also envision other usages in 
-sentiment mining such as the development and advancement of the opinions of people living in some country as conveyed by the 
-news articles across time.
-
-Other applications also include comparison of parallel accounts and narrations of the same events. For example, many history 
-books have been written to describe events that took place in the same periods of time. With event-time extraction, correspondence
-between such events can be achieved. Afterwards detection of inconsistencies can be facilitated. Also, sorting events from the 
-different accounts in partial-order time-line can help merging complementary accounts where one may describe events which are 
-missed in another and vise versa. This analysis used to be performed all manually -- 
-such as manually merging the different narrations of 
-the Bible.
-
-Finally, We would like to mention that the main motivation of this project is its usefulness in hadith authentication. 
-In this field, one criterion for criticizing a given narrator
-in the sequence of those having relayed the narration is based on 
-time consistency. In other words, there should exist an overlap in the lifespans 
-between a student narrating from his presumed 
-teacher~\cite{Al-Azami-91}. In the absence of such an overlap, the relayed hadith loses its credibility.
-A prerequisite to such consistency checks is being able to extract dates of birth and death from biography books and relate
-them to individual narrators.
-
-\section{Automating the Process of Hadith Authentication}
-
-We also consider the case study of hadith authentication, a very important problem
-that has historically been a totally manual work; however, authentication can clearly be 
-enhanced by automatic aids.
-
-We propose the following process:
-\begin{enumerate}
-\item \textbf{Segmenting Hadith Literature into Chains of Narrators:}
-
-	A hadith book is divided into its different narrations and locating the sanad of each,
-	subdivided into its chain of narrators. This can be performed by detecting sequences of narrators
-	connected by narration connectors. This task has been completed and is presented in detail in Secion~
-	\ref{sec:hadithSegmentation}.
-\item \textbf{Building a Partial Order Graph of Narrators:}
-
-	At this stage, the different extracted chains of Narrators are merged into a 
-	partial order graph which shows the relation
-	between the narrators and their contribution in relaying the hadith literatures. An example of such a 
-	partial order graph
-	is depicted in Figure~\ref{f:narrators}. 
-	It has been extracted by our preliminary tool for the hadith case study introduced in Section~\ref{sec:hadith}.
-
-\item \textbf{Localization of Narrators in Biograph Books:}
-
-	This partial order graph is instrumental to automate
-	localization of narrators in biography documents and
-	segmenting biography documents in the future.
-	Given that a biography discusses a narrator 
-	and mentions his professors and students,
-	we plan to segment the biographies with a graph 
-	coloring algorithm
-	that traverses the text and colors the graph node whenever
-	a name is found.
-
-\item \textbf{Graph Annotation and Consistency Checks:}
-
-	Once the biographies are analyzed, we can annotate
-	the POR with qualifiers of the narrators that reflect
-	their authenticity. 
-	We can also annotate the POR with the locations and 
-	the time the narrators lived in. 
-	Then we can compute a time and location overlap
-	check to discover inconsistent narrations.
-	We can also perform several interesting checks using 
-	this POR on its own, such as checking for the effect of
-	one narrator on the hadith literature.
-\end{enumerate}
-
-\chapter{Literature Review}
-\label{sec:literature}
-
-The survey in~\cite{Sughaiyer:04} compares
-several morphological analyzers. 
-Analyzers such as~\cite{Khoja:01,Darwish:02} 
-target specific applications 
-%in the analyzer itself or use a specific set of POS tags as their reference
-or limit themselves to a specific set of POS tags. 
-Sarf differs in that it is a general morphological 
-analyzer that reports all possible solutions. 
-It is application-specific in the sense that a controller 
-prunes these solutions.
-
-Buckwalter~\cite{Buckwalter:02},
-Beesley~\cite{Beesley:01},
-SAMA ~\cite{Kulick:10},
-and ElixirFM~\cite{Otakar:07} 
-take as input white space delimited tokens~\cite{Kulick:10},
-consider them as words,
-and enumerate all possible solutions. 
-%This approach has several problems. 
-However, we note the following points that deserve consideration:
-%(We need not to mention especially point 2 as a "problem" according to reviewer comments)
-First, a white space delimited token may have 
-more than one word or just a part of a MWE.
-%Second, the exhaustive enumeration may hurt performance and may
-%not be necessary or appropriate
-%in some case studies as noted in~\cite{Maamouri:10}. 
-Second, in many applications, the exhaustive enumeration may not be needed or appropriate as 
-noted in~\cite{Maamouri:10}; consequently, adding unnecessary running time.
-
-Other morphological analyzers such as 
-Amira~\cite{Diab:07,Benajiba:07},
-MAGEAD~\cite{Habash:05}, and MADA+TOKAN~\cite{Habash:09} 
-use machine learning and support vector machines (SVM) 
-to enhance the accuracy of the morphological analysis at the expense 
-of efficiency.
-Xerox rules can be compiled into specialized finite state
-finite state Automata (FSA)~\cite{Beesley:03}.
-However, the efficiency of the resulting analyzer depends on the
-way the Xerox rules are written. 
-This requires deep knowledge and insight from the user
-in compilation techniques, context free grammars, 
-and morphological analysis.
-%context free grammar rules that can be automatically compiled
-%into efficient non-deterministic finite state Automata (NDFSA). 
-Beesley reports that the number of Automata generated by a compiler
-for Xerox rules can not be controlled and also reports a difficulty 
-to compose the FSAs into a single framework. 
-%Beesley requires recompilation of the Xerox rules. 
-% (I am not sure if it is ok to have the above removed)
-Sarf however constructs an NDFSA from 
-%the controller and 
-the prefix, stem, and suffix deterministic 
-transducers to solve the problem.
-Sarf also differs in that 
-%it uses recursive affixes in the affix transducers, and in that 
-the transducers consider
-input from the controller to modify their behavior.
-The user can express the controller,not necessarily an FSA, in a programming language such as C++.
-
-Sarf builds upon the lexicon of Buckwalter\cite{Buckwalter:02}.
-It differs from Buckwalter in that it defines a shorter list of affixes
-and a longer list of 
-compatibility rules to define 
-affixes recursively and allow 
-prefix-prefix and suffix-suffix 
-concatenations.
-This better maintains and propagates 
-the tags associated with affixes.  
-This information is highly 
-redundant with Buckwalter and may lead to consistency issues. 
-%Sarf keeps all the possible analyses alive where each analysis
-%corresponds to a set of decisions and positions in the text. 
-%Buckwalter produces a set of solutions for each token 
-%and the several analysis threads of the text need to be 
-%generated as a product of the token solutions. 
-Similar to Sarf, Attia uses FSAs and handles more and qualitatively different MWEs
-~\cite{Attia:06,Attia:10}.
-
-%Similar to %the 
-%MADA+TOKAN, Sarf uses
-%the several valid morphological analyses as 
-%%a richness that should 
-%richness at a higher abstract level, and does not view them as 
-%inaccuracies to be corrected. 
-Sarf differs from MADA+TOKAN in that it provides an interface for the 
-application to interfere in judging the analyses much earlier and
-at a finer granularity level. 
-%In Sarf, there is no need for a separate tokenizer such as
-Sarf does not need a separate tokenizer such as
-TOKAN since each solution keeps a stack of positions
-that partition text into morphemes.
-%The application-specific controller can build the trace 
-%when needed.
-
-Similar to The MADA+TOKAN toolkit, Sarf thinks of
-the several valid morphological analyses as a richness that 
-should be exploited at a higher abstract level rather than
-an inaccuracy that should be corrected. 
-Sarf differs in that it provides an interface for the 
-application to interfere in judging the analyses much earlier and
-at a finer granularity level. 
-In Sarf, there is no need for a separate tokenizer such as
-TOKAN since each solution keeps a stack of positions
-that partition text into morphemes.
-%The application-specific controller can build the trace 
-%when needed.
-
-%MAGEAD 
-%Maamouri, Mohamed and \bgroup et al.\egroup },
-
-Our approach is close to the refinement of SAMA~\cite{Maamouri:10}.
-SAMA was refined to interact with
-the ATB~\cite{Maamouri:04} project after the addition of a large 
-new corpus. 
-We find supporting evidence in this decision to our hypothesis
-about the utility of a application-specific analyzer. 
-The algorithmic changes in SAMA were
-done manually and worked in integration with the ATB format. 
-Sarf is an application-specific controller approach that can modify 
-the behavior of the analyzer on the fly and can interact
-with any application as needed. 
-In Sarf, the analyzer remains general, and the 
-application-specific controller specializes the analysis.
-
-The AMIRA~\cite{Diab:07} analyzer uses 
-a language independent SVM based analysis. 
-The SVM analysis learns features from the input text 
-and does not make use of any information 
-in the target query.
-
-Like ElixirFM~\cite{Otakar:07}, Sarf builds on the lexicon
-of the Buckwalter analyzer. 
-Sarf also uses deterministic parsing with tries 
-to implement the affix and stem transducers. 
-We think that the inferential-realizational approach 
-of ElixirFM
-that is highly compatible with the Arabic linguistic 
-description~\cite{Badawi:04}
-can benefit from many features unique to the Arabic language.
-Sarf leaves implementing that to the application-specific controller
-since in many cases the abstraction needs only a partial 
-linguistic model of the Arabic language. 
-We can compare the application-specific controller in Sarf to a high level
-component in the ElixirFM domain specific language.
-
-\chapter{Preliminary Results}
-
-In what follows, we present our preliminary results which have been performed throughout the course of the past 2 years
-which support our hypothesis and provide evidence for the effectiveness of our methods. We will divide our presentation into
-results pertaining to the functionality of Sarf and results pertaining to the hadith case study in which Saf was a fundamental component.
-
-\section{Application-Specific Morphological Analyzer (Sarf)}
-
-Sarf extends the lexicon of Buckwalter~\cite{Buckwalter:02} with
-proper names and location names extracted from different online 
-sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
-\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
-as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
-
-Sarf encodes the prefix, suffix and stem lexicons into
-three deterministic transducers and 
-uses a non-deterministic construction
-of the three to compute all valid morphological analyses.
-We first present a running example 
-%of Sarf. The running example 
-that 
-explains how Sarf works,
-implements recursive affixes, and
-handles ``run-on'' words.
-%Then we discuss how Sarf builds and composes the different FSAs.
-
-%\subsection{Sarf example}
-%\label{sec:example}
-
-\transfalse
-\begin{figure}[tb]
-\center{
-\resizebox{\columnwidth}{!}
-{ \input{figs/FSM.pdftex_t}}
-\caption{Example affix and stem transducers.}
-\label{f:example}
-}
-\end{figure}
-% the FSMs for the three words
-%\transtrue
-
-\subsubsection{Sarf example}
-The diagram in Figure~\ref{f:example}
-illustrates the operation of $\Phi$, the Sarf FSA,
-when parsing the
-string \noTrnoVocRL{wsyl`b-h-aa\nospace al--lA`bwn} \noArRL{wsyl`b-h-aa al--lA`bwn}.
-%\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
-\footnote{ 
-%\noVocRL{n}%
-%\noVocRL{w}
-%\noVocRL{b}
-%\noVocRL{`}
-%\noVocRL{a} 
-%\noVocRL{l}
-%\noVocRL{l}
-%\noVocRL{a} %
-%\noVocRL{a}
-%\noVocRL{h}
-%\noVocRL{b}
-%\noVocRL{`}
-%\noVocRL{l}
-%\noVocRL{y}
-%\noVocRL{s}
-%\noVocRL{w}
-\noTrnoVocRL{n}%
-\noTrnoVocRL{w}
-\noTrnoVocRL{b}
-\noTrnoVocRL{`}
-\noTrnoVocRL{a} 
-\noTrnoVocRL{l}
-\noTrnoVocRL{l}
-\noTrnoVocRL{a} %
-\noTrnoVocRL{a}
-\noTrnoVocRL{h}
-\noTrnoVocRL{b}
-\noTrnoVocRL{`}
-\noTrnoVocRL{l}
-\noTrnoVocRL{y}
-\noTrnoVocRL{s}
-\noTrnoVocRL{w}
-in separate form to ease following the example
-}
-
-Boxes and circles denote accept and regular
-states respectively.
-The edges are transitions with labels corresponding to
-the input letters.
-The reject states are omitted for clarity and are denoted
-by the absence of corresponding transitions. 
-Subfigures (a), (b), and (c) in Figure~\ref{f:example}
-represent ${\cal P}$ the prefix ,
-${\cal S}$ the stem, and ${\cal X}$ the suffix
-transducers respectively. 
-The symbol $\epsilon$ represents an empty string and is 
-the source of non-determinism.
-
-We start with $\Phi$ at the root accept state 
-in ${\cal P}$. 
-%The $\epsilon$-edge connects ${\cal P}$
-%to ${\cal S}$ and transitions from any accept state
-%in ${\cal P}$ to the root state in ${\cal S}$.
-%The $\epsilon$-edge that connects ${\cal S}$ to ${\cal X}$
-%follows the same behavior. 
-The $\epsilon$-edges connect ${\cal P}$
-to ${\cal S}$  and ${\cal S}$ to ${\cal X}$.
-They transition from any accept state
-in the current transducer to the root state of the target transducer.
-
-When there are two valid transitions such as \noTrnoVocRL{w} 
-and $\epsilon$ in the start case, 
-$\Phi$ spawns an exact copy of itself $\Psi$. 
-$\Phi$ makes one transition, and $\Psi$ makes the other. 
-Each of $\Phi$ and $\Psi$ represent a valid analysis so far. 
-An FSA %($\Phi$ or $\Psi$) 
-dies when it reaches a reject state.
-In our example, if there were no stems that start 
-with the letter \noTrnoVocRL{w}, $\Psi$ will die. 
-In reality, $\Psi$ will die when the input 
-is at \noTrnoVocRL{wsyl`}.
-
-Note that the affix transducers allow recursive 
-affixes. 
-For example, \noTrnoVocRL{w} 
-%will result 
-results 
-in an accept state
-that transitions to ${\cal S}$.
-When \noTrnoVocRL{s} follows, we move to another accept state in 
-${\cal P}$ corresponding to 
-%the 
-prefix \noTrnoVocRL{ws}. 
-%The same applies to ${\cal X}$.
-
-Lets consider $\Phi$ after it consumed \noTrnoVocRL{wsy} 
-and transitioned into the root node of ${\cal S}$.
-Now $\Phi$ will transition with \noTrnoVocRL{l`b} to reach an accept 
-state.
-
-Before moving with the letter \noTrnoVocRL{b} to the accept state,
-$\Phi$ needs to make sure that %the 
-stem \noTrnoVocRL{l`b} is compatible
-with %the 
-prefix \noTrnoVocRL{wsy}. 
-Sarf keeps compatibility information as part
-of the accept states. 
-%Sarf considers the value of the category as part of the state.
-Thus each accept state in Figure~\ref{f:example} represents
-more than one concrete state. 
-If the categories of \noTrnoVocRL{l`b} and \noTrnoVocRL{wsy} are compatible
-then \noTrnoVocRL{b} will move $\Phi$ to an accept state. 
-Otherwise, it will move it to a regular or a reject state.
-
-Since $\Phi$ is now in an accept state in ${\cal S}$, it 
-spawns $\Xi$ which transitions to the root of ${\cal X}$. 
-After it consumes \noTrnoVocRL{h--} $\Phi$ dies since there is no
-\noTrnoVocRL{h--}-edge from the current state.
-$\Xi$ represents a valid analysis and proceeds.
-
-Note that $\Xi$ reports a full word at any accept state
-at ${\cal X}$, whenever a white space or delimiter is encountered.
-We left that out in Figure~\ref{f:example} for clarity. 
-However, non-connecting letters result in spawning a new FSA 
-using the $\epsilon$ transition
-to ${\cal P}$, to be able to analyze the other `sub-word' of 
-the possible `run-on' word.
-
-\subsection{Recursive affixes}
-\label{sec:recaffix}
-
-We call two morphemes compatible if their concatenation
-forms a legal morpheme. 
-Buckwalter~\cite{Buckwalter:02} keeps separate lists 
-of affix and stem morphemes and assigns categories to
-morphemes. 
-%The relation between morphemes and categories is one 
-%to many. 
-A compatibility rule is a pair of categories 
-$\langle c_1, c_2\rangle$  stating that $c_1$ morphemes
-can be concatenated with $c_2$ morphemes.
-Buckwalter keeps three lists $L_{ps}, L_{sx},$ and $L_{px}$ 
-of compatibility rules relating
-prefixes to stems, stems to suffixes, and prefixes to suffixes
-respectively. 
-Consider a string $s=\alpha\beta\gamma$ where $\alpha$ is 
-a prefix, $\beta$ is a stem, and $\gamma$ is a suffix;
-$\alpha\beta\gamma$ is a 
-valid morphological analysis iff
-$\langle c(\alpha),c(\beta)\rangle \in L_{ps}$ and
-$\langle c(\beta),c(\gamma)\rangle \in L_{sx}$ and
-$\langle c(\alpha),c(\gamma)\rangle \in L_{px}$, where
-$c()$ returns the category of the morpheme.
-
-Many affixes are composed of other affixes. For example,
-the prefix \noTrnoVocRL{wsy} is composed of three other prefixes
-namely \noTrnoVocRL{w}, \noTrnoVocRL{s}, and \noTrnoVocRL{y}.
-The POS tag for \noTrnoVocRL{wsy} is a concatenation
-%of the POS tags of each of the three morphemes. 
-of those of the three morphemes. 
-Keeping all valid morphemes in a list
-is redundant and may lead to consistency issues when
-users try to add their custom morphemes.
-We introduce $L_{pp}$ and
-$L_{xx}$ as prefix-prefix and suffix-suffix 
-compatibility rules respectively.
-%This allows us to keep a shorter list of affixes. 
-We also introduce a {\em resulting category}
-for the $L_{pp}$ and  $L_{xx}$ resulting morphemes.
-The resulting category in all practical cases
-is that of the second morpheme in the pair.
-
-Consequently, the above representation allows Sarf to keep 
-only atomic affixes, their categories and compatibility rules.
-
-\subsection{Finite State Automata Implementation}
-\label{sec:Automata}
-
-In what follows, we detail the implementation and functionality of Sarf's
-automata for both stems and affixes.
-
-\subsubsection{Affix transducers}
-\label{sec:affixFSA}
-
-We analyzed the prefixes and suffixes of 
-Buckwalter~\cite{Buckwalter:02}
-and constructed the recursive affixes
-into directed acyclic graphs (DAG) similar to 
-those in Figures~\ref{f:example}(a) and~\ref{f:example}(c).
-The transducers are deterministic and thus 
-their traversal is linear.
-This is computationally superior to the 
-approach of Buckwalter where the analyzer considers
-all possible substrings %of the word in question
-and looks them up in the affix tables.
-
-The affix transducers encode $L_{pp}$ and
-$L_{xx}$.
-An accept state in the transducer corresponds to the last letter 
-of an affix in the affix lists.
-The accept states produce as output the POS and other tags
-associated with the state.
-
-\subsubsection{Stem transducer}
-\label{sec:stemFSA}
-
-We built our stem lexicon using the stem lexicon of 
-Buckwalter. 
-We augmented the lexicon with a set of proper names and
-a set of location names which we 
-obtained from online and biblical sources.
-
-%The stems share lots of substrings. 
-Since stems share lots of substrings. 
-We encoded them in
-an efficient double array trie (DAT) structure~\cite{Aoe:89}. 
-The structure is deterministic and free of cycles where the 
-accept states are the terminal nodes corresponding to the last 
-letters in stems. 
-Note that DAT is a suffix trie implementation which may not 
-necessary work on letter-by-letter states, but also compacts
-common substrings.
-This approach is superior to Buckwalter since it consists of
-a linear walk in the trie while with Buckwalter we need
-a number of hash lookups in the order of all possible partitions
-of the string.
-
-\subsubsection{Construction of Sarf from transducers}
-\label{sec:ndfsa}
-
-The diagram in Figure~\ref{f:composition} shows an 
-abstraction of the non-deterministic construction of Sarf
-from the prefix, stem and suffix transducers. 
-The circle represents regular states, the square
-represents accept states and the triangle represents
-reject states. 
-Note that the concrete prefix, stem and suffix transducers
-are all deterministic and cycle free as discussed 
-in previous subsections.
-%in Sections~\ref{sec:affixFSA} and~\ref{sec:stemFSA}. (bc will result in same refernce bc are subsubsections
-The cycles are introduced in the abstract machine
-for pedagogical purposes.
-
-%Note that Sarf does not rely on implementing its machines as 
-%equivalent regular expressions due to  the following reasons ...
-% (not necessary)
-
-\begin{figure}[tb]
-\center{
-\resizebox{\columnwidth}{!}
-{ \input{figs/abstract_machine.pdftex_t}}
-\caption{Sarf NDFSA.}
-\label{f:composition}
-}
-\end{figure}
-
-The algorithm \CodeIn{NDSarf} (detailed below)
-takes as input a text string $L$ and a application-specific controller
-$A$. 
-It starts with one machine $\Phi_1$ similar to the one in 
-Figure~\ref{f:example} and inserts it into a collection
-$M$. 
-The algorithm reads a letter $c$ at a time from $L$
-and iterates over all machines in $M$. 
-For each $\Phi$ in $M$,
-if $\Phi$ is in an accept state and it is in the suffix
-phase, then the algorithm checks whether it should report
-a full word. 
-This happens when $c$ is a white space, a delimiter, 
-or when the preceding letter %leading to the current state
-was a non-connecting letter.
-
-In all cases, if $\Phi$ is in an accept state, 
-it spawns another machine $\Psi$ and adds it to $M$. 
-Notice that spawning a new machine requires only keeping
-the state of the machine since all machines share the
-structures of the three transducers.
-If $c$ leads to a reject state, then $\Phi$ dies 
-and we remove it from $M$. 
-Otherwise, $\Phi$ transitions using the $c$ edge.
-After iterating over all machines in $M$, we invoke the 
-controller $A$
-and grant it full access and control over
-all machines in $M$.
-
-The arrows in Figure~\ref{f:composition} that transition
-from prefix to stem, stem to suffix, and suffix to prefix
-require compatibility constraints for 
-the transition.
-
-\begin{table}[tb]
-\centering
-%\resizebox{.9\columnwidth}{!}{
-\begin{tabular} {p{8.5cm}}
-\begin{Verbatim}[fontsize=\relsize{-2},
-frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
-commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}] 
-$$
-NDSarf(string $L$, Controller $A$) 
-  MachineVec $M$; -- collection of machines
-  Machine $\Phi_1$;
-  $M$.insert($\Phi_1$);
-  foreach $c$ in $L$ \{
-    foreach $\Phi$ in $M$ \{
-      if ( $\Phi$.state.isAccept() ) \{
-        if ( $\Phi$.isSuffix() ) \{
-          if ( $c$.isWhite() )  
-            $A$.report();
-          if ($\Phi$.lastChar().isNonConnecting()) -- run-on
-            $\Phi$.transition($\epsilon$); -- back to Prefix Machine
-        \}
-        $\Psi$ = $\Phi$.clone();
-        $M$.insert($\Psi$); 
-      \}
-
-      if ( $\Phi$.isWalkable($c$) ) \{
-        $\Phi$.transition($c$);
-      \} else \{
-        $M$.remove($\Phi$);
-        $\Phi$.die();
-      \} 
-    \} 
-    $A$.control($M$, $c$); 
-  \}
-\end{Verbatim}
-\end{tabular}
-%}
-\label{a:ndsarf}
-\end{table}
-
-\subsection{Application-Specific Controller}
-
-A valuable aspect of Sarf is that its machines are driven by 
-an application-specific controller that can refine the computed
-analyses as needed.
-
-The controller can interfere at any point in the 
-analysis to make a decision like ignoring a solution that 
-is not interesting. 
-It may also decide to correct one solution
-based on the decision of the others. 
-The controller uses the output of the transducers as input
-to make 
-%its decision. 
-refinement decisions.
-%%It does not produce external output and therefore 
-%%the three transducers with the 
-%%controller compose an FSA.
-
-In a sense, Sarf can be treated as a customizable morphological analyzer which can be
-programmed to take into account the characteristics of the application at hand.
-
-\subsection{Disambiguation Using Diacritics}
-\label{sec:diacritics}
-
-Sarf is the first in the field to support diacritic-based disambiguation. Consequently, 
-Sarf does not confuse \RL{bon} (coffee beans) with \RL{bin} (child)
-or \RL{`ilm} (science) with \RL{`lim} (knew), regardless of the specific letters which are vocalized.
-The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
-On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
-to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.
-
-The lexicon of buckwalter contains the vocalized equivalent of each entry, however makes no use of their power. 
-On the other hand, in Sarf's transducers after each accept state an (optional) linear-time diacritic based equality function is run
-to check if the state should be reported to the controller. Accordingly, much unnecessary ambiguity is reduced.
-
-\subsection{Handling Run-on Words}
-\label{sec:runon}
-
-As depicted in Figure~\ref{f:example}, there exists an $\epsilon$ 
-transition from ${\cal X}$ to ${\cal P}$.
-A cloned state of the machines spawns this transition 
-only when a non-connection letter is the last encountered letter.
-In this way, the remaining part of the word is analyzed as if a 
-complete word consisting of Prefix, Stem, Suffix concatenated constituents.
-Note that upon performing this transition, the analysis of the previous subpart is
-not reported since it may turn that this structure does not result in a valid run-on word.
-Results are only reported when a white space or delimiter is encountered
-and the word is at some accept state in ${\cal X}$; otherwise, the corresponding
-analysis dies.
-
-If results were reported at at any connecting letter, many false positive 
-run-on words will be generated. For instance, \noVocRL{al-madrasaT} would have its 
-\noVocRL{al-mad} subpart reported as a valid run-on constituent although the analysis
-will die finally since the remaining part \noVocRL{rasaT} is invalid. Such errors are 
-avoided by restricting the result reportage to the encounter of a white space or delimiter.
-In the case of a run-on word, the complete analysis is reported consisting of a valid analysis 
-to multiple `words'.
-
-\subsection{Support for Multi-word Expressions}
-\label{sec:MWE}
-
-Current analyzers, with the exception of Attia, fail to recognize MWE such as \noTrnoVocRL{`bid alkarym}, and 
-report it as two consecutive names. 
-Sarf includes such expressions in the stem transducer (with one letter being the space) and 
-detects them as one unit since tokenization
-is performed on the fly. The multiword stems
-can accept some compatible affixes such as \RL{w} (and) and \RL{f-} (but).
-This approach works for proper names but not for ``Syntactically-Flexible Expressions'' which are
-MWEs that can undergo lexical inflections for their constituent words~\cite{MWE}.
-
-The strength of our multi-word expression Support is that it can be designed to support for
-multi-word expressions which are written as one word in a similar fashion to run-on words such as
-\noVocRL{`abidAl-ra.hyim}. A correct analysis can be achieved since spaces that exist inside stems 
-that exist in the lexicon are allowed to match with an empty string when the previous letter is a
-non-connecting one. If such a flexibility was not supported, it would have been required to add
-space-containing multi-word versions along with run-on-like multi-word versions of those expressions,
-which would be highly redundant.
-
-\subsection{Time Efficiency Comparison Results}
-\label{sec:timeefficiency}
-
-\begin{table}[tb]
-\centering
-\caption{Results of Sarf, Buckwalter and ElixirFM.}
-\resizebox{0.9\columnwidth}{!}{
-%\begin{tabular}{lp{.01cm}cp{.01cm}cp{.01cm}c}
-\begin{tabular}{lccc}
-\hline
- & Al Kafi & Al Istibsar & Ibn Hanbal \\ \cline{2-4} 
-Word Count & 98,943 & 103,835 & 20,354 \\
-Sarf (sec) & 2.70 & 2.69 & 0.236 \\
-Sarf/Controller (sec) & 1.32 & 1.31 & 0.0960 \\
-Buckwalter (sec) & 3.32 & 5.41 & 0.547 \\
-ElixirFM (sec) & 167$\times60$ & 138$\times60$ & 29.2$\times$60 \\
-\hline
-\end{tabular}
-}
-\normalsize
-\label{t:efficeincyresults}
-\end{table}
-
-%In this section we compare Sarf to existing 
-%morphological analyzers such as Buckwalter and ElixirFM in terms of running time.
-We ran Sarf without a controller then with a controller that only 
-focused on proper names and words that are related to the word ``narrated".
-We compared the running times to those of Buckwalter and ElixirFM.
-We used three books of Islamic narrations \RL{.hady_t} ~\cite{IbnHanbal,AlTousi,AlKulayni} as our data set.
-We report our results in Table~\ref{t:efficeincyresults}.
-
-With a controller, Sarf performed much better since the controller pruned several solutions irrelevant to the application. 
-Sarf, with and without a controller, was more efficient than both Buckwalter and ElixirFM. 
-Buckwalter performed much better than ElixirFM in 
-terms of running time. 
-The difference may be due to the sophisticated formal 
-functional morphological analysis of ElixirFM. 
-%and its reliance on additional annotations extracted from 
-%both ATB~\cite{Maamouri:04} and the Prague Arabic Dependency Treebank~\cite{Prague04}.
-
-\section{Hadith Case Study}
-\label{sec:hadith}
-
-Several researchers~\cite{Hadithopaedia:08,Azmi-2010} attempted to automate 
-the analysis of the hadith literature. 
-In this our thesis, we use Sarf to successfully automate the
-analysis of three books of hadith selected 
-arbitrarily~\cite{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained
-  the digitized books from online sources such as 
-  \href{http://www.yasoob.com/}{http://www.yasoob.com/} and 
-  \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }.
-
-In what follows, we present our preliminary tool:
-the hadith {narrator chain 
-extractor using morphological analysis} (NaCEMA).
-It employs novel techniques for automating
-the extraction of the \noArRL{sanad} from the books of traditions 
-into a diagram of chains of narrators. NaCEMA segments
-a book into narrations, each narration into its \noArRL{matn} 
-and its \noArRL{sanad}, 
-and each sanad into its separate narrators. 
-This work will lead to 
-automate an exhaustive \noArRL{.hady_t} authentication 
-effort in the future as we plan to 
-automatically analyze the biography books.
-
-\subsection{Hadith Segmentation and Abstraction}
-\label{sec:hadithSegmentation}
-
-NaCEMA accepts a book as input
-and segments it into a vector of hadiths. 
-NaCEMA segments each hadith into its
-sanad and matn parts. 
-NaCEMA 
-detects the chain of narrators in the sanad and 
-the relation that links each narrator to his ancestor and 
-predecessor in the chain. 
-It also detects the full name of each narrator that is
-composed of several proper names with connectors
-in between.
-
-\subsection{The hadith controller}
-\label{sec:controller}
-% FSM for the controller
-
-Intuitively, the controller reports a valid chain 
-of narrators when a sequence of names
-connected by narrator connectors appears. 
-This includes the task of finding multi-word names
-such as \RL{`bd alr.hmn} often appearing as ``run-on'' words.
-The controller marks the beginning of the hadith with the beginning of 
-the current proper name sequence,
-and marks the end of the hadith with the beginning of the 
-next proper name sequence. 
-It detects the chain of narrators as the sequence itself.
-
-%The controller also targets words that mean ``narrate'' when
-%they appear between proper names. 
-We illustrate the hadith controller 
-in Figure~\ref{f:hadith}. 
-The transitions are labeled by outputs generated from 
-the Sarf morphological analyzer that correspond to 
-part of speech (POS) tags. 
-A NAME label denotes a proper name.
-An NRC label denotes a narrator connector such as
-\RL{`an} ``on behalf of'', \RL{.hada_t} ``narrated'', \RL{qaal} ``said'', 
-\RL{'a_hbar} ``told'' or one of their morphological equivalents. 
-An IBN label denotes the word \RL{ibn} ``the son of'' and is a name connector.
-A NISBA is a possessive adjective that qualifies a person such 
-as \RL{alma.sriy} ``the Egyptian''.
-
-The symbol $\tau_{\mbox{NMC}}$ is a threshold
-that corresponds to the number of tolerated name connectors 
-that may occur between two names. %3
-The symbol LIST$_{\mbox{NMC}}$ corresponds to the list 
-of name connectors collected since the controller
-started looping in the state NMC\_S. 
-The symbol $\lambda_{\mbox{NMC}}$ is a parameter 
-that corresponds to a relaxed tolerance measure that
-the controller resorts to in case the words separating
-two names were longer than $\tau_{\mbox{NMC}}$ but 
-contained a name connector word such as IBN or NISBA.
-
-The controller has four states that correspond to 
-a position in text relative to the next sanad. 
-State TEXT\_S is the initial state and denotes that
-the controller is outside the context of a sanad.
-The controller moves to state NAME\_S on
-NAME.
-State NRC\_S means that the controller is in the context
-of a narrator name after Sarf reports an NRC.
-State NMC\_S
-indicates that the controller expects a name to appear within 
-a tolerance threshold expressed by 
-$\tau_{\mbox{NMC}}$ and $\lambda_{\mbox{NMC}}$.
-
-Similarly, the NRC\_S state tolerates $\tau_{\mbox{NRC}}$ words 
-before it gives up on its expectations. 
-We reach the NAME\_S state only when a
-valid NAME is detected and we leave when no more NAME's are detected.
-The NRC\_S state can only be reached if an NRC is detected.
-NMC\_S can only be reached from a NAME\_S state.
-We used the values of 3, 5 and 5 for $\tau_{\mbox{NMC}}$, 
-$\lambda_{\mbox{NMC}}$, and $\tau_{\mbox{NRC}}$ respectively.
-
-\begin{figure}[tb!]
-\center{
-\resizebox{.9\columnwidth}{!}
-{ \input{figs/hadith.pdftex_t}}
-\caption{The controller of the hadith application.}
-\label{f:hadith}
-}
-\end{figure}
-
-%The definition of input labels such as NAME and IBN depends on the 
-%morphological analyzer. 
-%However, our case-based controller approach was able to perform well
-%under both Sarf and a refined version of Sarf.
-
-\subsection{Building Narrator Partial Order Graphs}
-\label{sec:POR}
-
-\begin{figure}[tb]
-\center{
-\resizebox{1.1\columnwidth}{!}
-%{\includegraphics{figs/narrator_chain_output.pdf}}
-{\includegraphics{figs/DAG_narrator.pdf}}
-\caption{Narrator directed acyclic graph 
-    extracted using NaCEMA }
-\label{f:narrators} 
-}
-\end{figure}
-
-The directed acyclic graph (DAG) 
-in Figure~\ref{f:narrators} shows a partial order relation (POR) 
-between narrators.
-We automatically extracted the POR from the three books of 
-hadith~\cite{IbnHanbal,AlKulayni,AlTousi}.
-The nodes in boxes are the \RL{matn} of the hadith, 
-and the other nodes are the narrators.
-Given a set of chains of narrators similar to 
-Figure~\ref{f:exhadith} we formed the DAG using graph algorithms 
-that merged equal names in the chains. 
-We compared the narrators together using a morphological
-distance function. 
-We considered each narrator to have several morphological 
-components and we rewarded every morphological match with 
-a small $\delta$. 
-Then we applied a threshold $\tau$ to the sum of rewards
-and decided the equality of the narrators. 
-We omitted the details of the distance function for brevity; 
-%
-in all cases, we are considering their revisit and re-implementation
-in ways more suitable to the hadith case study.
-
-\subsection{Accuracy Results}
-
-\begin{table}[bt]
-\centering
-\caption{Results of the hadith case study with Sarf.}
-%\begin{tabular}{|p{1.5cm}||c|c||c|c||c|c|} \hline
-\resizebox{1.1\columnwidth}{!}{
-\begin{tabular}{lcp{.2cm}cp{.2cm}c} %\cline{2-10}
- &  AlKafi & & AlIstibsar & &IbnHanbal \\ \cline{1-6}
-Word count  &98,943 & & 103,835 & & 20,354 \\ 
- Names       & 12,060  & & 14,613& & 3,013\\
-%Names/Narrator & 1.97 & & 1.84& & 1.25 \\
-Narrators  & 2,623 & & 5,767& & 1,755 \\ 
-%Narrators/Chain  & 4.84 & & 4.76 & &4.05 \\
-Chains  & 542 &  & 1,211& & 433 \\ 
-Ignored names  & 6,400 &  & 3,348 & & 642 \\ \hline
-Segmentation accuracy  & 96\%& & 96\%& & 92\%\\ 
-Chain accuracy & 99\%&  & 99\%& & 97\% \\ 
-%Narrator accuracy  & & & & &  & & & \\ 
-Narrator accuracy  & 91\%& & 90\% & & 90\% \\ \hline
-%Name false positives  & 7\%&  & 4\% & & 4\% \\ \hline
-Running time (secs.)& 1.32 & & 1.31 & & .096\\ \hline 
-\end{tabular}
-}
-\normalsize
-\label{t:hadithresallresults}
-\end{table}
-
-We ran our experiments using a dual core 2.66 Ghz 64-bit processor 
-with 4GB of memory using a Linux operating system. 
-We report our results in Table~\ref{t:hadithresallresults} on the 
-three hadith books.
-NaCEMA detected narrator chains and segmented the books 
-correctly with a chain accuracy above 97\% and a narrator name
-accuracy above 90\%
-in all cases and across all the 1986 traditions in the three 
-different books with different formats.
-
-The row ``word count'' is an indicator of the size of the data.
-The %numbers in 
-rows ``names'', ``narrators'', and ``chains'' denote the total
-number of the corresponding entities per book.
-The row ``ignored names'' reports the number of names detected 
-outside the chains and thus are irrelevant to
-the application. 
-%In addition, the rows ``names/narrator'' and ``narrators/chain'' 
-%report the average
-%number of names per narrator and 
-%the average number of narrators per chain respectively.
-
-The ``segmentation accuracy'' reports on the accuracy of 
-segmentation of the book into separate hadiths.
-We computed the metric manually by marking the beginning 
-and end of a representative sample of hadiths in each 
-book and compared that to the NaCEMA results. 
-We used the same approach to compute all other accuracy measures.
-The ``chain accuracy'' reflects the percentage of correct narrators 
-covered by extracted chains.
-The ``narrator name accuracy'' row reflects the percentage of 
-correct proper names detected within
-narrator structures.
-
-%The row ``name false positives'' reports the percentage of words 
-%recognized as names while they should not.
-%We notice that this metric did not affect the accuracy
-%of narrator and chain segmentation since detecting
-%a narrator or a chain with the refined analysis without substantially affecting the 
-%high abstract measures such as chain and segmentation
-%accuracy.
-
-%\subsection{Future Suggestions}
-
-Although our results reflect a high accuracy, we plan to 
-enhance the accuracy of hadith detection and the 
-narrator merging algorithms. We also plan to make use of the name connector and narrator
-connector stop phrases that our case study extracted to create a 
-learning system that can learn more names. 
-%We also plan on developing an Arabic linguistic computational model 
-%similar to that of ElixirFM~\cite{Otakar:07} but that can be 
-%explored partially based on the case study.
-
-\chapter{Thesis Plan}
-
-The main tasks of the thesis are listed below: \\
-\begin{enumerate}[1)]
-\item \textbf{Build and Enhance Sarf}
-	\begin{enumerate}[\ref*{s:intro}.1)]
-	\item Port Buckwalter into C++ and mysql
-	\item Implement the Finite State Automata Framework
-	\item Implement an application-specific controller interface
-	\item Support for disambiguation using diacritics
-	\item Support for multi-word expressions
-	\item Solve the Issue of `run-on' words
-	\item Optimize performance
-	\item Support for Running Sarf Under the Windows Operating System
-	\item Refactor the code
-	\item Write a user Manual
-	\item Write a developer manual and documentation
-	\end{enumerate}
-\item \textbf{Build a time entity extraction application}
-\item \textbf{Automate the Hadith Authentication Case Study}
-	\begin{enumerate}[3.1)]
-	\item Extend the Lexicon of Buckwalter with proper names 
-	\item Implement the hadith segmentation tool
-	\item Write the Narrator Equality Function
-	\item Build Parial Order Graphs
-	\item Correct Graph mistakes that result from Ambiguous narrator names which introduce false cycles to the graph
-	\item Segment biography books using a graph coloring algorithm
-	\item Extract Annotations from the Biography books
-	\item Assign annotations to graph nodes
-	\item Perform time and possibly location consistency checks
-	\item Perform Experiments on the effects of different narrators in Hadith Literature
-	\item Write a user Manual
-	\item Distribute the tool to willing religious scientists and report their findings
-	\end{enumerate}
-\item \textbf{Write thesis Report and defend the thesis}
-\end{enumerate}
-
-\section{Time Table}
-
-We anticipate that the work can be performed according to the time table presented in
-Table~\ref{t:timetable}. \\
-
-~\\
-
-\begin{table}[bt]
-\centering
-\caption{Thesis Time Table}
-\resizebox{1\columnwidth}{!}{
-\begin{tabular}{|c|c||c|c||c|c||c||c|c|c|c||c|c|c|c||c|} \bottomrule[0.1em]
-\textbf{Year} & \multicolumn{9}{|c|}{\textbf{2011}} & \multicolumn{6}{|c|}{\textbf{2012}} \\ \hline
-\textbf{Semester} & ~ & \multicolumn{2}{|c||}{\textbf{\tiny Spring}} & \multicolumn{2}{|c||}{\textbf{\tiny Summer}} 
-		& ~ & \multicolumn{4}{|c||}{\textbf{\tiny Fall}} & \multicolumn{4}{|c||}{\textbf{ \tiny Spring}} & ~ \\ \hline
-\backslashbox{Task}{Month}
-			& \textbf{Done} & \textbf{5} & \textbf{6} & \textbf{7} & \textbf{8} & \textbf{9} & \textbf{\small 10} & 
-			 \textbf{\small 11} & \textbf{\small 12} &
-			 \textbf{1} & \textbf{2} & \textbf{3} & \textbf{4} & \textbf{5} &
-				\begin{tabular}{c}
-				\textbf{\small Future} \\
-				\textbf{\small Work}
-				\end{tabular}
-			 \\ \toprule[0.1em] \bottomrule[0.1em]
-\textbf{1.1)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{1.2)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
-\textbf{1.3)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{1.4)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~  & ~ &\\ \hline
-\textbf{1.5)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{1.6)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~  & ~ &\\ \hline
-\textbf{1.7)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{1.8)} & \x & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ &\\ \hline
-\textbf{1.9)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & ~ & \\ \hline
-\textbf{1.10)}& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X  & ~ & \\ \hline
-\textbf{1.11)}& ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \toprule[0.1em] \bottomrule[0.1em]
-\textbf{2)}   & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \toprule[0.1em] \bottomrule[0.1em]
-\textbf{3.1)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.2)} & \x & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.3)} & \x & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.4)} & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.5)} & ~ & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.6)} & ~ & ~ & ~ & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.7)} & ~ & ~ & ~ & ~ & ~ & \X & \X & \X & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.8)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & ~ & ~ & ~ & ~ & ~ & \\ \hline
-\textbf{3.9)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & \X & ~ & ~ & ~ & \X \\ \hline
-\textbf{3.10)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & \X & ~ & ~ & \X \\ \hline
-\textbf{3.11)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X & ~ & \\ \hline
-\textbf{3.12)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X \\ \toprule[0.1em] \bottomrule[0.1em]
-\textbf{4)} & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & ~ & \X &  \\ \hline
-\end{tabular}
-}
-\normalsize
-\label{t:timetable}
-\end{table}
-
-\large\underline{\textbf{Important Dates}}\\\\
-\normalsize
-The important dates are as follows:
-\begin{itemize}
-\item \textbf{Comprehensive exam:} Mid May 2011
-\item\textbf{Thesis defense:} Mid May 2012
-\end{itemize}
-
-% Include Bibliography
-\bibliographystyle{IEEEtran}
-{\small \bibliography{jm}}
-%\nocite{*}
-
-\end{document}
-%-------END----------------------------------------------------

### 2018-06-26 tex/sarf/aggaffix.tex
sarf to ACLing draft added

%\vocalize
Sarf considers three types of affixes: 
%\vspace{-2em}
\begin{itemize}
  \item {\em Atomic affix morphemes } such as \noTrutfRL{يـ} can be affixes on their own and can directly connect to stems using the $R_{ps}$ and $R_{sx}$ rules. 
  \item {\em Partial affix morphemes} such as \noTrutfRL{سـ} can not be affixes on their own and need to connect to other affixes before they connect to a stem.
  \item {\em Compound affixes} are concatenations of atomic and partial affix morphemes as well as other smaller compound affixes. 
    They can connect to stems according to the $R_{ps}$ and $R_{sx}$ rules. 
\end{itemize}

Sarf forms compound affixes from atomic and partial affix morphemes using newly introduced prefix-prefix $R_{pp}$ and suffix-suffix $R_{xx}$ concatenation rules. 
%
Sarf considers $L_p$ and $L_x$ to be lexicons of atomic and partial affix morphemes associated with their tags. 
Sarf forms agglutinative affixes using prefix-prefix $R_{pp}$ and suffix-suffix $R_{xx}$ concatenation or agglutination rules.
A rule $r \in R_{pp} \cup R_{xx}$ takes the compatibility category tags of affixes $a_1$ and $a_2$ and checks whether they can be concatenated. 
If so, the rule takes $a_1$ and $a_2$ and their tags and generates the
affix $a=r(a_1,a_2)$ with its associated tags according to substitution rules based on regular expressions. 
The rules are fusional in the sense that they modify the orthography and the semantics of the resulting 
affixes by more than simple concatenation.

\begin{table}[!tb]
\begin{minipage}{0.95\textwidth}
\relsize{-1.5} {
  \begin{tabular}{lp{4.4cm}cp{3.3cm}}
\hline \hline
{\bf Category 1} & {\bf Category 2} & {\bf Resulting Category}  & {\bf Substitution rules} \\ \hline
        NPref-Bi & NPref-Al & NPref-BiAl &  \\[3pt]
        NPref-Li & NPref-Al & NPref-LiAl & \em r$//$\noTrRL{Al} $||$\noTrRL{l}$\backslash\backslash$ \\[4pt]
 Pref-Wa & none of \{ Pref-0, NPref-La, PVPref-La\} & \$2 & \\[4pt]
%Pref-sa- & \{IVPref* AND NOT IVPref-li-\} & \{\$2\} \\
IVPref-li- & IVPref-*-y* & IVPref-(@1)-liy(@2)  & { \em d$//$he/$||$him/$\backslash\backslash$, ~ d$//$they$||$them$\backslash\backslash\ldots$ ~ d$//$(+2)$||$ to$\backslash\backslash$} \\ 
\hline \hline
\end{tabular}
}
\end{minipage}
\caption{Example rules from $R_{pp}$ }
%\vspace{-1cm}
\label{t:rules}
\end{table}
%
We illustrate this with the example rules in Table~\ref{t:rules}.
Row 1 presents a simple rule that allows the concatenation of prefixes with category \cci{NPref-Bi} such as \RL{bi-} and \RL{ka-}
to prefixes with category \cci{NPref-Al} such as \noTrRL{Al}, the result is the compound prefix with category \cci{ NPref-BiAl}. 
Since no substitution rule is specified, the tags of the resulting prefix are simple concatenations.

Row 2 presents a rule that takes prefixes with category 
\cci{NPref-Li} such as \RL{li-} and prefixes with category \cci{NPref-Al} such as \noTrRL{Al}.
The substitution rule replaces the \noTrRL{Al-} with \noTrRL{l-} resulting in \noTrRL{lil-}. 
The syntax of the substitution rule for the affix form is \cci{r$//$(substring)$||$(replacement)$\backslash\backslash$}.

The rule in the Row 3 states that prefixes of category \cci{Pref-Wa} can be concatenated with 
prefixes with categories that are neither of 
\cci{Pref-0}, \cci{NPref-La}, and \cci{PVPref-La} categories. 
%The Boolean operators allowed are \cci{AND}, \cci{OR}, and \cci{NOT}.
The resulting category is denoted with $\$2$ %$
which means the category of the second prefix.

Row 4 illustrates the use of the wild character \chci{*} to capture sub-strings of length zero or more in the second category,
and refers to the captured sub-strings in the resulting category using the \chci{@} operator. 
The \chci{@} operator is always followed by a number that denotes the captured \chci{*} expression.
Row 4 has also an example of substitution rules for the gloss (description) tag that start with the letter $d$.
The \cci{+2} pattern in the last substitution rule means that the \cci{to} partial gloss description should be appended after the gloss 
of the second affix.
Substitution rules for POS tags start with the letter $p$.

\todo{R3.10}
The rule introduced in row 4 replaces 24 prefix category rules in BAMA. 
Given the word \utfRL{ليأكل} (for him/it to eat/consume), 
Sarf returns the solution with the prefixes \noTrutfRL{ل} (for) and \noTrutfRL{ي} (him/it), 
and the stem \noTrutfRL{أكل} (eat/consume). 
The prefixes \noTrutfRL{ل} and \noTrutfRL{ي} have the categories 
\cci{IVPref-li-} and \cci{IVPref-hw-ya}, respectively. 
The category \cci{IVPref-li-} is the same as category 1 of the rule in row 4 and 
the category \cci{IVPref-hw-ya} matches the pattern of category 2 of the same rule. 
The first wild character within the pattern captures the sub-string \cci{hw} 
while the second wild character captures the sub-string \cci{a}. 
Sarf recognizes that those two affixes are compatible using this rule, 
and applies the corresponding expression to get the resulting category 
\cci{IVPref-hw-liya} for the compound affix \noTrutfRL{لي}. 
Then, Sarf matches the result category of the compound affix 
with the category \cci{IV\_no-Pref-A} of the stem \noTrutfRL{أكل} 
and returns a morphological solution.

%\vspace{-3em}
\subsection{Building $R_{pp}$ and $R_{xx}$}
\label{ss:affix-affix}

Our method is in line with native Arabic textbooks on morphology and syntax~\citep*{Mosad09,Abd00,Abd001} 
where only atomic and partial affixes are introduced.
The textbooks also list rules to concatenate the affixes and discuss the syntax, semantic, and phonological forms of the
resulting affixes. 
%This representation is more suited for linguists.
For example, the fourth rule in Table~\ref{t:rules} is derived from a textbook rule that states 
\cci{IVPref-li-} prefixes connect to all imperfect verb prefixes and transform the subject pronoun in the gloss 
to an object pronoun.

The method built the rules in four steps: 
%\vspace{-2em}
\begin{enumerate}
    \item In the first step, we encoded textbook morphological rules into patterns.
    \item In the second step, we inspected the BAMA and SAMA affix lexicons and extracted the atomic and partial affixes from them.
    \item Then, we grouped the rest of the BAMA and SAMA affixes into the rules we collected from the textbooks. 
    \item We refined the rules wherever necessary, and we grouped rules that shared the same patterns. 
\end{enumerate}
%\vspace{-1em}

We validated our work by generating all possible affixes and compared them against the BAMA and SAMA affix lexicons. 
The comparison resulted in discovering the BAMA and SAMA inconsistencies listed in Tables ~\ref{t:incBAMA} and ~\ref{t:incSAMA}.

### 2018-06-26 tex/sarf/api.tex
sarf to ACLing draft added

\label{s:api}
Sarf provides the developer with an application programming interface (API) 
  that allows to 
  (1) define developer categories and associate them with specific morphemes, 
  (2) provide rules that prioritize and filter the solution features, and 
  (3) control and refine the morphological analysis on the fly at solution {\em control points}.
%
The control points are (1) agglutinative prefix matches, (2) stem matches, 
(3) agglutinative suffix matches, and 
(4) full solution matches.

\todo{R1.3}
To use the API, an NLP application developer is required to inherit from 
one or more of the prefix, suffix, stem and general 
stemmer C++ classes and implement the \textit{OnMatch} interface. 
The developer configures the feature priority and selection rules 
by using the \textit{setSolutionSetting} method to push the desired features 
into a feature configuration vector. 
The computation returns the features present in vector in the solutions ordered 
by the same order in the vector. 
If the vector is empty, all features are considered.

The developer implements the \textit{OnMatch} interface that Sarf calls at the 
control points. 
The developer processes the solution features provided at the control point and returns 
a value that instructs Sarf to 
(1) proceed with the analysis ignoring the current solution, 
(2) accept the solution and continue considering other solutions, 
and (3) accept the solution and stop the analysis.
%
The developer can inspect at each control point 
the agglutinative morphemes and their compatibility category tags, 
the VMF, gloss, POS, lemma, and the developer-defined category tags.

Consider the task that aims to detect words with possible VERB POS tags. 
The API can be implemented to reject the analysis at the stem control point if the POS tag is not a VERB. 
This prevents the analyzer from computing insignificant 
full morphological solutions at early stages of the analysis.
The developer can also use the feature selection filter API to 
disregard features such as VMF, gloss, and categories. 
This reduces the size of the solution trees and 
their corresponding traversal time. 
%
For example, given the word \RL{'kl}, 
with gloss tags such as `he/it eat' and `fruit',
Sarf typically returns nine possible morphological solutions 
with different VMF, POS tags, and gloss tags. 
With the feature selection filter API, Sarf considers only two solutions with 
two solutions with the {\tt VERB\_PERFECT} and {\tt NOUN} POS tags.

Sarf also provides the developer with the ability to alter 
the structure of the morphological solutions so 
that the traversal of the solutions is application specific. 
The developer can provide a factorization order 
of the solution features using the API priority rules. 
Sarf uses the priority rules to build the structures. 
This allows the developer to dismiss analysis earlier at the control points. 
Consider the task with an aim is to detect adjectives 
with positive sentiment. 
Since the POS feature is more limited than the gloss feature, 
it might give priority to the POS over the gloss. 
Hence, it can filter invalid solutions early in 
the analysis.

Moreover, Sarf enables the application developer to define 
categories and associate them with existing morphemes. 
Consider the task of detecting words that indicate 
family relations such as \RL{ibn} (son), \RL{'b} (father), 
and \RL{'m} (mother). 
The developer can define a category called `family connections' 
and associate it with the stems \RL{ibn}, \RL{'b}, \RL{'m} 
or with their relevant glosses. 
The user defined categories are attached to the tags in the 
morpheme nodes as auxiliary tags that can be looked up in constant time.

%During the analysis, Sarf calls the API at the control points and 
%the API detect the morphemes that fall under the defined category.

### 2018-06-26 tex/sarf/appendix/appendix.tex
sarf to ACLing draft added

\include{macros}
\begin{document}

\doublespacing

\begin{appendices}

%\tableofcontents

%\newpage

\section{Morphology}

Morphology is the study of the internal structure of an Arabic word. 
There are two types of approaches to morphology. 
The first approach is \textit{form-based morphology}. 
It is concerned with the form of the units constructing a word, 
the interactions between them, and how they form the overall structure of a word. 
The second approach is \textit{functional morphology}. 
It is concerned with the function of the units inside a word and the units' effect 
on the work syntactically and semantically.

\subsection{Form-based morphology}
The \textit{morpheme} is a central concept in form-based morphology.
Under this approach, we can distinguish \textit{concatenative} and \textit{templatic} morphemes.

Concatenative morphology has three types of morphemes: \textit{stems}, \textit{affixes}, and \textit{clitics}. 
Stems is the base of every word, affixes attach to the beginning and the end of the stem, 
and clitics attach to the affixes. 
Affixes include \textit{prefixes}, \textit{suffixes}, and \textit{circumfixes}. 
Prefixes attach to the beginning of the stem, suffixes attach to the end of the stem, and 
circumfixes surround the stem. 
Clitics include \textit{proclitics} and \textit{enclitics}. 
The terms prefix and suffix are sometimes used to refer to proclitics and 
enclitics, respectively.

The stem can be templatic or non-templatic. 
Templatic stems are stems formed using templatic morphemes, 
while non-templatic stems can't be derived using templatic morphemes. 
Templatic morphology is concerned with the construction of the templatic stems. 
This morphology has three base morphemes that are required to construct a word templatic stem. 
The morphemes are \textit{root}, \textit{pattern}, and \textit{vocalim}. 
The root morpheme is a sequence of 3-5 consonants and 
it implies an abstract meaning common to all the derivations. 
The pattern morpheme is an abstract template used to insert both the root and vocalism. 
The vocalism morpheme specifies the short vowels that are used with a pattern.

\subsection{Functional morphology}
In functional morphology, we analyze words in terms of their 
morpho-syntactic and morpho-semantic behavior. 
This morphology has three functional operations: 
\textit{derivation}, \textit{inflection}, and \textit{cliticization}.
Derivation morphology is concerned with the creating new words from other words 
based on lexical relations such as location or time. 
Inflections provide obligatory extensions to words limited to 
a set of features such as gender and numbers. 
Cliticization is similar to inflection, 
however stands for optional features in clitics.

\section{Morphological Analysis}

Morphological analysis is the process by which we compute all the morphological solutions of a word. 
Each solution includes the choice of a single core part-of-speech for the stem or root. 
The analysis can be either form-based or functional.
\textit{Sarf} is a morphological analyzer that 
computes the morphological solutions of an input word 
based on form-based morphology and concatenative morphemes. 
It also refers to the proclitics and enclitics as prefixes and suffixes, respectively.

\end{appendices}

%\bibliography{merf}

\end{document}

### 2018-06-26 tex/sarf/bamainctable.tex
sarf to ACLing draft added

\begin{table}[!tbp]
\begin{minipage}{0.95\textwidth}
\relsize{-1.5} {
  \begin{tabular}{p{2.5cm}ccc}
\hline \hline
   & {\bf Affix } & \textbf{Vocalized} & {\bf Inconsistent tag} \\ 
\hline
\multirow{2}{2.5cm}{ {\bf \ref{inc:plus})}
missing plus in {\bf gloss} tag  of {\bf prefix}}
 & \RL{fl} & \RL{fali} & and/so \fbox{+} for/to + the  \\
% & \RL{fl--l} & \RL{falil} & and/so \fbox{+} for/to + the \\
 & \RL{wbAl} & \RL{wabiAl} & and + with/by \fbox{+} the  \\[6pt]
\multirow{3}{2.5cm}{{\bf \ref{inc:so})} 
  missing alternative {\bf gloss} %for fa
  in {\bf prefix} }
 & \RL{f} & \RL{fa} & and/so  \\ 
 & \RL{fb} & \RL{fabi} & and \fbox{/so} + with/by          \\
 & \RL{fk} & \RL{faka} & and \fbox{/so} + like/such as     \\[6pt]
% & \RL{fl} & \RL{fali} & and \fbox{/so} + for/to           \\
% & \RL{fl} & \RL{fali} & and \fbox{/so} + for/to + the    \\
\multirow{4}{2.5cm}{ {\bf \ref{inc:num_gender_subject}) } 
  gender/number qualifier omitted in {\bf gloss} of subject {\bf suffix} }
 & \RL{n} & \RL{n} & they [fem.pl.] $<$verb$>$                     \\
 & \RL{nhm} & \RL{nahom} & they \fbox{[fem.pl.]} $<$verb$>$ them \\
 & \RL{At} & \RL{At} & [fem.pl.]                                \\
 & \RL{Atk} & \RL{Atka} & \fbox{[fem.pl.]} your                \\[6pt]
% & \multicolumn{4}{c}{48 total inconsistencies} \\ 
\multirow{2}{2.5cm}{
  {\bf \ref{inc:num_gender_object}) } 
  also in {\bf gloss} of object {\bf suffix} }
 & \noVocRL{AnhmA}&  \RL{AnihimA}& them \fbox{(both)}                      \\
 & \noVocRL{tkm}& \utfRL{|Bَتْكُم}& it/they/she $<$verb$>$ you \fbox{(pl.)}  \\[6pt]
% & \multicolumn{4}{c}{39 total inconsistencies} \\ 
\multirow{3}{2.5cm}{
  {\bf \ref{inc:both}) } 
  different ways to express them (both) in {\bf gloss} of {\bf suffix} }
 &\noVocRL{thmA}&  \utfRL{|Bَتْهُما} & it/they/she $<$verb$>$ them (both)    \\
 & \noVocRL{AhmA} &  \utfRL{|BBاهُما} & we $<$verb$>$ 
 \colorbox{shadecolor}{\color{white} (both of)} them \fbox{(both)}  \\
\vocalize
 & \RL{nAhmA} &  \utfRL{nAhُmA} & we $<$verb$>$ 
\colorbox{shadecolor}{\color{white} (both of)} them \fbox{(both)}   \\[6pt]
% & \RL{nhmA} &  \utfRL{nَhُmA} & they $<$verb$>$ 
%        \colorbox{shadecolor}{\color{white} (both of)} them \fbox{(both)}  \\
% & \RL{hmA} &  \utfRL{|BBهُما} & they $<$verb$>$ 
%        \colorbox{shadecolor}{\color{white} (both of)} them \fbox{(both)}  \\
\multirow{2}{2.5cm}{
  {\bf \ref{inc:dot}) } 
  `.' omitted  after pl in {\bf gloss} }
 & \utfRL{م} & \utfRL{|Bُّم} & you [masc.pl.] $<$verb$>$              \\
% & \utfRL{وا} & \utfRL{|Bُوا} & you [masc.pl\fbox{.}] $<$verb$>$     \\ 
 & \utfRL{ونا} & \utfRL{|Bُونا} & you [masc.pl\fbox{.}] $<$verb$>$ us \\[6pt]
% & \multicolumn{4}{c}{8 total  inconsistencies } \\        
\multirow{2}{2.5cm}{
  {\bf \ref{inc:pos}) } 
  {\bf POS} tag is not same as vocalized 
}
 & \noVocRL{th} & \utfRL{thِ}     & +ti/PVSUFF\_SUBJ:2FS  \\
 &              &                & +\colorbox{shadecolor}{\color{white} hu/}\fbox{hi/}PVSUFF\_DO:3MS \\
% & \utfRL{ونا}  & \utfRL{|Bَوْنا} & +\colorbox{shadecolor}{\color{white} uw/}\fbox{aw/}PVSUFF\_SUBJ:3MP  \\ 
% &              &               & +nA/PVSUFF\_DO:1P \\
%% & \multicolumn{4}{c}{38 total inconsistencies} \\         
\hline \hline
\end{tabular}
}
\end{minipage}
\caption{Sample BAMA inconsistencies}
\label{t:incBAMA}
\end{table}

### 2018-06-26 tex/sarf/conclusion.tex
sarf to ACLing draft added

This paper presents Sarf, an application customizable Arabic morphological analyzer. 
NLP applications can implement the Sarf API to 
to refine the morphological analysis on the fly by 
(1) selecting the interesting features, (2) prioritizing the features, and
(3) accepting and rejecting solutions on the fly based on partial features reported 
so far. 
Sarf extends the SAMA and BAMA lexicons and 
represents affixes using agglutinative and fusional morphemes which 
(1) significantly reduces the size of the lexicons needed to represent Arabic morphology,(2) fixes inconsistencies in morphological features corresponding to morphemes,
(3) simplifies the maintenance and augmentation of the affix morpheme lexicons, and
(4) solves the segmentation correspondence problem between the morphological solution 
and the original text. 
Sarf also allows the NLP application to use partial diacritics for morphological solution
disambiguation, and solves the `run-on words' problem.
Sarf is available online as an open source tool. 
\todo{R2.17}
We evaluated Sarf in terms of effectiveness and performance with respect to 
other tools such as SAMA, BAMA and ElixirFM. Sarf outperforms other analyzers in 
run-time and slightly increases recall rates in morphological analysis due to the 
extended lexicons. 
We used Sarf application specific API in several NLP applications 
for information extraction from Arabic documents and the applications
provided more accurate results than existing analyzers with faster running time.

In the future, we plan to improve Sarf by allowing root analysis of stems,
supporting inflectional stems, and providing a graphical user interface to 
allow the users to edit the affix and stem morpheme lexicons of Sarf.

### 2018-06-26 tex/sarf/cover.tex
sarf to ACLing draft added

\documentclass[11pt,a4paper]{letter}
\usepackage{graphicx}
\usepackage{multirow}

\oddsidemargin -0.25in
\evensidemargin -0.25in
\topmargin -1in
%\topsep 0in
\textheight 10.25in
\textwidth 6.75in
%\parskip 6pt
\longindentation 0.5\textwidth
\parindent 0.4in

\begin{document}

\address{
    \begin{tabular}{p{4.2cm}p{6cm}r}
%    \multirow{3}{*} {
%    \includegraphics[width=4cm]{../AUB-concise-logo-color.jpg} } & &
& & Fadi A. Zaraket \\
& & AUB Faculty of Eng. \& Arch. \\
& & ECE Department \\
%& P.O. Box: 11-0236\\
%& Riad El-Solh, Beirut 1107 2020\\
%& Beirut, Lebanon\\
& & Tel: (+961) 1 350 000, ext: 3484  \\ \hline
& & Email: {\tt fadi.zaraket@aub.edu.lb} \\
\end{tabular}
}

\begin{letter}{}

\signature{Fadi A. Zaraket \\ Assistant Professor
  of Electrical and Computer Engineering}

\opening{Dear concerned at NLE,}

We appreciate your consideration of our submission entitled {\em Sarf: Application Customizable Efficient Arabic Morphological Analyzer}. 
The paper describes an Arabic morphological analyzer that has been used effectively and efficiently in several Arabic NLP 
applications for information extraction from Arabic text. 
The paper describes the contributions of Sarf as compared with existing work. 
Sarf provides application customizable analysis through an API that allows the NLP application developer to control and refine the 
analysis on the fly. 
Sarf refines the lexicons of existing analyzers, reduces the required size and maintenance by using agglutinative and fusional concatenation rules and solves morphological feature inconsistencies in the existing analyzers. 
Sarf also solves the token segmentation problem which is a hindrance to machine learning with current analyzers.

We look forward for NLE to consider our submission as Sarf bridges the gap between computational linguistics and practical applications, 
provides an efficient engineering implementation for the Arabic morphological analysis problem, and is similar to 
morphological analyzers and lexical resources for other semitic languages such as Hebrew that NLE published before.

\closing{Best regards,
    \\~
}
\vspace{-2in}

\end{letter}

\end{document}

\end{document}

### 2018-06-26 tex/sarf/diacritics.tex
sarf to ACLing draft added

Diacritics are short vowels that
are often omitted in Arabic text and inferred by readers from context. 
Their omission adds to the ambiguity problem of Arabic morphological analysis. 
\fullvocalize
The diacritics \RL{|Ba} ({\em fatha}),  \RL{|Bu} ({\em damma}), and 
\transfalse\RL{|B}\transtrue represent
and appears above the letter. 
`a' vowel,  `o' vowel, and consonant, respectively,  and appear above the letter. 
The diacritic \RL{|Bi} represents a `y' vowel and appears below the letter. 
The diacritics 
\RL{|BaN}, \RL{|BuN}, and \RL{|BiN} represent the `a', `o', and `y' vowels followed 
by a phonetically stressed \transtrue \RL{n} consonant.

The shadda \RL{|BB} mark is not a 
diacritic but is treated typographically as one, and is also often omitted in Arabic text. 
It denotes  a repeated letter, first as consonant, and second 
as vocalized.
Arabic forbids two letters with a consonant diacritic to follow each other. 
%A shadda can be followed by all diacritics except \transfalse\RL{|B}\transtrue.

Analyzers such as BAMA and SAMA ignore input partial diacritics because 
they consider them to be (1) rare in common corpora, and (2) 
unreliable because of dialect diversity and human errors~\citep*{Elkateb:06,Attia:06b}. 
\todo{R3.13}
The diacritics can be miss-placed for several reasons including: 
(1) writers use a local Arabic dialect pronunciation of the word, 
(2) font effects such as ligatures compensate for miss-placing diacritics, 
and
(3) writers sometimes use diacritics to decorate their text. 
%
However, the work in \citep{Chaaben:10,Attia:00,Beesley:01} considers 
partial diacritics to decrease morphological ambiguity.
The work in~\cite{Maamouri:06} inspected the ATB v3.2 corpus~\citep*{Maamouri:04}
for diacritics and found that 1.364\% of the words were partially diacriticized and 
those diacritics eventually and {\em effectively} reduced morphological ambiguity.
Most of the diacritics we found were tanween marks which indicate indefiniteness, 
followed in frequency by shadda, and then by dhamma that often indicates 
passive tense when associated with verbs. 
Hence, we decided not to force the use of partial diacritics to disambiguate the 
solutions and to provide it as an option.

%Figure~\ref{t:consistent} presents the Diacritic-aware consistency check Algorithm.
Key to partial diacritic analysis is a diacritic-aware consistency check that replaces
standard string matching checks. 
The \cci{ Diacritic-aware consistency check} algorithm takes as input 
two text chunks $c_1$ and $c_2$. 
\todo{R2.14}It checks if the sequence of letters in $c1$ is the same as that in $c2$. 
It also checks the consistency of the sequence of diacritics in $c1$ and in $c2$. 
%It first checks if the sequence of letters, without diacritics, in $c1$ is consistent with that in $c2$. 
%Then, it checks if the sequence of diacritics of letters in  $c1$ is consistent with that of $c2$.
%
%It checks that the sequence of non-diacritic letters, ignoring the 
%diacritics between them, are equal. 
%It also checks that all sequences of diacritics occurring between 
%non-diacritic letters are consistent.

Two sequences of diacritics are consistent iff:
%\vspace{-3em}
\begin{enumerate}
  \item Both are equal, or
  \item One of the sequences is empty, or
%  \item don't contain different diacritics
  \item If one has a shadda, then the other has no sukoun, or
  \item If one has a shadda and the other has no shadda, then the rest of the diacritics are compared recursively. 
\end{enumerate}

If both checks were positive, then $c1$ and $c2$ are said to be equal with diacritic-aware consistency. \todo{R2.14}

\input{diaceqtable}

Table~\ref{t:correspondence} illustrates the 
\cci{diacritic-aware consistency check} as compared to the standard
string comparison with an example. 
Part (a) shows two diacritic-consistent words 
\vocalize
\utfRL{أَكّل} and \RL{'kal} as a fatha \RL{|Ba} is compatible with an empty diacritic 
and a shadda \RL{|BB} is compatible with a fatha \RL{|Ba}. 
%In fact, a shadda is compatible with all diacritics except 
%for the sukun.
%and consequently its disambiguation value is mainly measured 
%through its presence/absence and not through its compatibility. 
Part (b) illustrates two inconsistent diacritizations 
since \RL{|Bi} is incompatible with \RL{|Ba} next to the letter \noVocRL{k}.
%Inconsistencies can be detected when different short vowels are encountered as in 
%part~\hyperref[ex:valid_correspondence]{(b)} of Table~\ref{t:correspondence}
%in which \fullvocalize

\todo{R2.15, R3.14}
Sarf takes an Arabic word as input with possible partial diacritics. 
Then, it traverses ${\cal P}$, ${\cal S}$, and ${\cal X}$ structures 
to compute the morphological solutions. 
For each matching morpheme in the structures, 
Sarf uses the \cci{diacritic-aware consistency check} algorithm 
to check for the consistency of the fully diacriticized morpheme
with the corresponding partially diacriticized chunk of the input word.
The traversal path corresponding to the morpheme dies if the consistency 
check is negative (inconsistent).

### 2018-06-26 tex/sarf/editorComments.tex
sarf to ACLing draft added

\section*{Editor Comments to the Author}
Dear Prof. Zaraket,

We have now received all reviewers’ reports on your paper ID NLE-ARTC-15-0011 entitled ``Sarf: Efficient and Application Customizable Arabic Morphological Analyzer".

As you can see from the reviews enclosed, the reviewers identified a number of issues in your article.
As a consequence, we cannot accept your paper as it stands; however, WE ENCOURAGE YOU TO REVISE YOUR ARTICLE according to reviewers' feedback and RESUBMIT IT. (The comments of the reviewers are included at the bottom of this letter).

If you decide to resubmit, please enclose a list of the points that have been addressed, and the criticisms which they answer.

Once we have received your resubmission, it will be checked and will also undergo additional reviews which may require further modification. This is to ensure the paper meets the standards of the journal.

I would be grateful if you could let me know for administrative purposes whether you are prepared to make the suggested amendments, and, if so, when I can expect to receive the revised version.

To revise your manuscript, log into \url{https://mc.manuscriptcentral.com/nle} and enter your Author Center, where you will find your manuscript under ``Manuscripts with Decisions."  Under "Actions," click on "Create a Revision."  Your manuscript number will be appended to denote a revision. You may also click this link to start your revision:

\url{https://mc.manuscriptcentral.com/nle?URL_MASK=f5a4558b88a24181a82b5db451930250}

When submitting your revised manuscript, you will be able to respond to the comments made by the reviewers in the space provided.  Please use this space to document any changes you make to the original manuscript.

Could you please acknowledge receipt of this e-mail?

We hope that the reviewers' feedback will be useful for you in improving your work.

Thank you again for submitting your article to the NLE.

Best regards,

### 2018-06-26 tex/sarf/intro.tex
sarf to ACLing draft added

{\em Natural language processing} (NLP) applications such as 
{\em machine translation} (MT) and {\em information extraction} (IE)
require {\em morphological analysis} to preprocess Arabic text
due to the rich morphological nature of the Arabic language
~\citep*{Benajiba:07,Habash:06}. 
%
%Given a white space and punctuation delimited Arabic word,
Arabic morphological analyzers
return the internal structure of a given Arabic word composed of 
several {\em morphemes} including {\em affixes} ({\em prefixes} 
and {\em suffixes}), {\em clitics} ({\em proclitics} and {\em enclitics}), 
and {\em stems}~\citep*{Sughaiyer:04}. 
The morphological solution consists also of several {\em morphological features}
(tags) associated with the word and its constituent morphemes 
such as {\em part of speech} (POS), transliteration, {\em gloss}, 
and {\em vocalized morpheme form} (VMF) tags (diacriticized form of the morpheme).
%
%The stem is the core of a word. 
The prefix and suffix attach before and after the stem, respectively. 
Clitics are special affixes that attach to the stem to form a word, 
and differ from regular affixes in that 
they play a syntactic role of another word (often omitted) 
%while being bound to the word in question
~\citep*{habash2010introduction}. 
%
%However, some lexicons such as Buckwalter~\citep*{Buckwalter:04} use the terms prefix and suffix to refer to  proclitic and enclitic, respectively.

This work presents Sarf, a \todo{R2.13} {\em fast and application 
customizable morphological analyzer} for Arabic
that is used in several applications for information extraction 
from Arabic text~\citep*{JaZaMatar,ZaMaFlairs2012HadithBio,ZaMa2012IJCLATime,ZaMaHaCicling2012Entity}. 
Sarf provides NLP application developers with an application programming interface (API) 
to control and refine morphological analysis on the fly. 
The developer implements the interfaces in the application.
Sarf calls the interfaces 
on control points such as prefix, stem, suffix, and full solution matches.
%We distinguish between the NLP application developer 
%who uses Sarf to build an NLP application and the end user who provides the text
%and inspects the solutions.
The Sarf API allows the application to 
(1) control and prioritize the analysis, 
(2) refine the solution features, and
(3) define developer categories and associate them with existing morphemes.

%Users of the developed application pass Arabic text as input and the application uses 
%Sarf to perform customizable morphological analysis on the text.

Sarf is a significant extension of the work in~\citep*{ZaMaColing2012DemosSarf}. 
It represents Arabic affixes as 
agglutinative affix morphemes with fusional affix concatenation rules. 
Simpler agglutinative affix morphemes can be concatenated to form a more complex 
affix~\citep*{Vajda}. 
%Fusional models are similar to agglutinative with the difference that 
%morpehemes which make up the word fuse together several different features meanings~\citep*{Spencer:91}. 
Fusional affix concatenation rules specify affix pairs and use 
regular expressions \todo{R2.13 removed in} as substitution rules to 
compose the resulting orthographic and semantic tags from the 
tags of the original morphemes~\citep*{Spencer:91}.
The Sarf substitution rules are in sync with 
rules and examples on morpheme concatenative properties from
Arabic morphology textbooks~\citep*{Abd00,Abd001}.
This representation resolves consistency, maintenance, and segmentation 
issues of the current approaches in BAMA and SAMA. 
%Sarf also provides the option to use partial diacritics 
%in disambiguating the morphological solutions of a partially diacritized word.

\todo{R1.1}
In this paper, we make several additional contributions including the following. 
\begin{itemize}
    \item Sarf provides an application customizable morphological analyzer where the developer can control and refine the analysis. 
    \item Sarf structures and computations are engineered to reduce computation cost as follows. 
      \begin{itemize}
        \item The organization of the lexical structures into root index tries during runtime.
        \item The traversal structures that allows multiple solutions to be considered at once.
        \item The computation of the solution features in a tree based structure instead of computing all features and filtering later on
        \item The manual optimization of the traversal of the lexicon tries to reduce the typical excessive non-determinism exhibited in the finite state machine based solutions.
      \end{itemize}
    %\item Sarf is a novel Arabic morphological analyzer with agglutinative affixes and fusional affix concatenation rules based on textbook Arabic morphological rules and on the concatenation rules of existing analyzers. 
    \item Sarf solves the ``run-on'' words problem and optionally uses partial diacritics for solution disambiguation. 
    %\item Sarf solves inconsistencies in existing affix lexicons of BAMA and SAMA. In this paper we list and discuss the inconsistencies.
    %\item Sarf solves the correspondence between the morphological solution and the morphological segmentation of the original text problem.
%    \item We evaluate our approach using the ATB data set and report on the effect of our corrections on the annotations.
    \item Sarf is fully implemented and available online as an open source tool.\footnote{\label{fn:online} \url{http://research-fadi.aub.edu.lb/carla/doku.php?id=sarf} }
\end{itemize}

We evaluated Sarf for %segmentation correspondence, lexicon size, lexicon consistency\todo{R3.19}, 
accuracy and runtime efficiency.
%Our results show that Sarf lexicons are smaller than lexicons of existing analyzers and provide coverage for more morphological solutions. 
Sarf %simplifies the lexicon maintenance task, 
provides better accuracy 
and improves run time efficiency as compared to existing analyzers.  
%
\todo{R2.14}We evaluated Sarf also within applications that use its API 
~\citep{JaZaMatar,ZaMaFlairs2012HadithBio,ZaMa2012IJCLATime,ZaMaHaCicling2012Entity}. 
The carried case studies show that Sarf performs better than existing Arabic morphological analyzers in terms of running time. 
They also demonstrate the efficiency and the utility of the Sarf 
application customizable API.

The rest of this paper is structured as follows. 
\todo{R2.1A} In Section~\ref{sec:motivation}, we motivate Sarf and discuss how it deals with existing morphological analysis challenges. 
In Section~\ref{sec:related}, we compare Sarf to related work. 
In Section~\ref{sec:overview}, we present the structre of Sarf. 
In Section~\ref{sec:api}, we present the interface provided by Sarf for the 
application developer to control and refine the morphological analysis. 
%In Section~\ref{sec:aggaffix}, we present our method to build agglutinative affix morphemes with fusional affix concatenation rules. 
%%In Section~\ref{sec:motivation}, we discuss existing redundancy and inconsistencies in the lexicons of existing morphological analyzers.
In Section~\ref{sec:diac}, we present our method of using partial diacritics to reduce the morphological ambiguity.
\todo{R1.2}In Section~\ref{sec:results}, we discuss the use of Sarf with several NLP tasks and 
present the results of comparing Sarf to other analyzersin terms of speed and accuracy.%, and consistency.
Finally, we conclude and discuss future work in Section~\ref{sec:conclusion}.

%\vspace{-1cm}

### 2018-06-26 tex/sarf/lstdefinestyle.tex
sarf to ACLing draft added

\lstdefinestyle{c-code}{%
  language=C,                % choose the language of the code
  basicstyle=\tiny,       % the size of the fonts that are used for the code
  numberstyle=\tiny,      % the size of the fonts that are used for the line-numbers
  frame=Ltbr,
  rulesep=.4pt,
  mathescape, % Allows escaping to (La)TeX mode within $$,
  stringstyle=\color{red},% \color{redb}
  %keywords=[1]{elseif, return, else, int, bool, if, Boolean, do-together, foreach, endfor, endif, while, to, Let},
%keywords=[1]{declaration,program,list-of-statements,statement,conditional,loop,sync,assignment,target,bexpr,expr,return-statement,type,variable-decl,function-decl,property-decl,precondition,postcondition,property,quantifier,range,term,id,access,function-call,call-argument,argument-list,call-arg-list},
 % keywords=[2]{@dotogether,@pre,@post,if,while,return,forall,exists,int,bool,num,else,constant},
  showstringspaces = false,
  escapeinside={/*@}{@*/},
  basicstyle=\scriptsize\ttfamily,
  keywords=[1]{foreach,endfor,in,if,elseif,endif,to,else,let,be, end},
  keywords=[2]{append,next-list,v-next-list,v-expr,pc-next-list,decl-list,init-list,a-index,a-expr,a-next-statement},
  keywordstyle=[3]\color{darkbrown}\bf,
  keywordstyle=[2]\color{darkred}\bf,
  keywordstyle=[1]\color{darkblue}\bf,
  commentstyle=\color{darkgreen},
  alsoletter={-0123456789},
  captionpos=b, %
  breaklines=true,
  backgroundcolor=\color{background},
}

%End Added by Maya

### 2018-06-26 tex/sarf/motivation.tex
sarf to ACLing draft added

\todo{R2.1A}
In this section, we discuss issues that %currently
challenge morphological analysis and explain how Sarf resolves them.

\subsection{Exhaustive enumeration and performance:}

Current morphological analyzers such as 
%Buckwalter~\citep*{Buckwalter:02},
BAMA~\citep*{Buckwalter:02}, 
SAMA~\citep*{SAMAmain, Kulick:10}, \todo{R3.18} 
Beesley~\citep*{Beesley:01}, 
MADAMIRA~\citep*{pasha2014madamira}, 
and ElixirFM~\citep*{Otakar:07} 
take as input white space delimited tokens,
consider them as words,
and enumerate all possible morphological solutions for each word.

\input{solutiontable}

For example, given the word \utfRL{وسيلعبونها}\todo{R2.5}\footnote{The paper uses ArabTex package \citep{arabtex} for Arabic text editing and transliteration. ArabTex package is recommended and used by \citep{IntroNHabash}.} (and they will play it), 
an analyzer may return the solution presented in Table~\ref{t:solexample} with 
a prefix, a stem, a suffix, and their corresponding 
POS, transliteration, and gloss tags. 
%\noTrutfRL{وسيـ} with POS tag \cci{wa/CONJ+sa/FUT+ya/IV3MP} 
%and gloss tag \cci{and they will}, 
%the stem 
%\noTrutfRL{لعبـ} with POS tag \cci{loEab/VERB\_IMPERFECT}
%and gloss tag  \cci{play},
%and 
%the suffix \noTrutfRL{ونها} with POS tag \cci{uwna/IVSUFF\_SUBJ:MP\_MOOD:I+hA/IVSUFF\_DO:3FS}
%and gloss tag \cci{[MASC.PL.]+it/them/her}. 
The prefix 
\noTrutfRL{وسيـ} can be further segmented into 
(1) the proclitic \noTrutfRL{و} with POS tag \cci{CONJ} and gloss tag \cci{and},
(2) the proclitic \noTrutfRL{سـ} with POS tag \cci{FUT} and gloss tag \cci{will},
and (3) the prefix \noTrutfRL{ـيـ} with POS tag \cci{IV3MP} and gloss tag \cci{they(people)}.
Similarly, the suffix 
\noTrutfRL{ونها} 
can be segmented into
(1) the suffix \noTrutfRL{ونـ}, forming a circumfix with \noTrutfRL{ـيـ}, 
with POS tag \cci{IVSUFF\_SUBJ:MP\_MOOD:I}
and gloss tag \cci{[MASC.PL.]}, 
and (2) the enclitic \noTrutfRL{ـها} with POS tag \cci{IVSUFF\_DO:3FS}
and gloss tag \cci{it/them/her}. 
%Some analyzers may refer to \noTrutfRL{و} and \noTrutfRL{سـ} as proclitics, 
%and may refer to \noTrutfRL{ـها} as an enclitic.
\novocalize

The exhaustive enumeration of all solutions may hurt performance and may
not be necessary or appropriate
in some applications~\citep*{Maamouri:10}. 
Thus the need of a customizable morphological analyzer that adapts to 
application specific requirements.
%\vspace{-2.35cm}

\subsection{Diacritics and accuracy}

The accuracy of morphological analysis %, with respect to the context,
suffers from inherent difficulties
%of morphological analysis 
of the Arabic language such as omitted diacritics and position dependent letter
forms. 
%
\fullvocalize
Diacritics, i.e. short vowels, such as 
fatha (\RL{|Ba}), damma (\RL{|Bu}), kasra (\RL{|Bi}), 
tanween (i.e. doubled diacritic including \RL{|BaN}, \RL{|BuN}, \RL{|BiN}), 
\transfalse and sokun (\RL{|B}) 
are almost always omitted in written Arabic text as they can be 
inferred by human readers.
The mark \novocalize shadda (\RL{|BB}) \transtrue denotes the repetition
of the marked character and is also often omitted. 
Partial diacritics can help disambiguate solutions. 
%This is a source of disambiguity. 
Consider the unvocalized word \RL{'kl} with nine morphological solutions.  
Its partially vocalized version \utfRL{أكِل} has only two solutions; 
VMF \utfRL{أَكِل} with gloss \cci{I+trust/put in charge}, and 
VMF \utfRL{أُكِلّ} with gloss \cci{I+make tired/wear out}.

Analyzers such as BAMA and SAMA ignore partial diacritics while 
other analyzers such as \citep*{Beesley:01,Chaaben:10,Attia:00} make use of 
the partial diacritics to reduce ambiguity. 
Sarf provides an option that enables the use of existing diacritics for disambiguation, 
and considers the diacritics at morpheme boundaries to generate only the diacritic 
matching solutions, rather than generating all morphological solutions then 
filtering them.

%\vspace{-1.2cm}

\subsection{Problem of `run-on' words:}

Arabic letters have up to four different forms
corresponding to their position in a word, i.e, beginning,
middle, end, and separate word forms. 
This allows the phrase \transfalse
\RL{il_A\nospace almdrsT} \transtrue \todo{R2.12} (to the school)
to be visually recognizable
as two separate words \RL{il_A} (to) and \RL{almdrsT} (the school) 
without the need of a delimiter space in between. 
The reason is the first word \RL{il_A} ends with
\RL{_A} a non-connecting letter. 
These words, referred to as `run-on' words~\citep*{Buckwalter:04},
occur regularly, and greatly increase the
difficulty of tokenization. 
\todo{2.1A}None of existing morphological analyzers, up to our knowledge, 
resolve this issue. Sarf on the other hand does by the structures and rules it introduces, which are further discussed in the next issue.

\begin{comment}
\input{affixtable}

\subsection {Structure of existing analyzers and related problems:}

Concatenative morphological analyzers~\citep{Buckwalter:02,Kulick:10} %BAMA,SAMA
are based on lexicons of prefixes $L_p$, 
stems $L_s$, and suffixes $L_x$. 
As shown in Table~\ref{t:affixes}, each entry in a lexicon includes the morpheme, 
its vocalized form with diacritics, 
a {\em concatenation compatibility category} (CCC) rule, 
a gloss tag, and a POS tag.
Separate CCC rules specify the compatibility of prefix-stem 
$R_{ps}$, stem-suffix $R_{sx}$, and prefix-suffix (circumfix) $R_{px}$ concatenations. 
The affixes \noTrutfRL{و} and 
\noTrutfRL{يـ} in the before mentioned example are valid standalone prefixes, 
and can be concatenated to the stem \noTrutfRL{لعب} \todo{R2.12} (he plays) to form 
\noTrutfRL{ولعب} (and he plays) and
\noTrutfRL{يلعب} (he is playing), respectively. 
%In addition, the morpheme \noTrutfRL{سـ} can connect to \noTrutfRL{يلعب} to form 
%\noTrutfRL{سيلعب}. 
The $L_p$ and $L_x$ lexicons contain also all final forms of concatenated affixes
as shown in Table~\ref{t:affixes} for sample morphemes.

This is the source of several problems: 
\begin{itemize}
%\vspace{-2cm}
    \item $L_p$ and $L_x$ contain redundant entries which result \todo{R2.13} in maintenance and consistency issues~\citep*{LRECMaamouriKB08,lrecKulickBM10}.
    \item Augmenting $L_p$ and $L_x$ with additional morphemes, such as \utfRL{أَ} (the question glottal hamza), 
      may result in a quadratic increase in the size of the lexicons~\citep*{Hunspell}.
      The additional morpheme may attach to exiting morphemes. 
      Currently, this results in adding all the resulting morphemes to the BAMA and SAMA lexicons. 
    \item The $L_p$ and $L_x$ lexicons are larger than needed especially that 
      they have to account 
      for several forms of a morpheme with varying diacritics.
    \item The concatenated forms in $L_p$ and $L_x$ contain concatenated POS and 
      other tags. 
%        The segmentation correspondence between the prefix concatenated from several morphemes and the tags associated with it is lost. 
%        In several cases, this leads to missing correspondence between the tokens of the morphological 
%        solution and the segmentation of the original word.
The alignment and correspondence between the original word and its
morphemes with the tags of its morphological solution 
are essential to the success of NLP tasks such as MT 
and IE~\citep*{Regina2011,ELRASemmar08}.
The analysis of the example \utfRL{للقضاء} (for/to the justice/judiciary\todo{R2.12 gloss}) using SAMA,
generates the morphological solution \cci{li/PREP + Al/DET + qaDA'/NOUN}.
The issue of the alignment and correspondence between the original word \noTrutfRL{للقضاء} 
and its morphemes arises as the concatenation of the unvocalized prefix morphemes 
\cci{li} and \cci{Al} is not the same as the corresponding prefix 
segment \noTrutfRL{لل} in the original word. 
\todo{R.3.5}
This is not a problem for SAMA, but rather for the applications that 
use the SAMA output such as treebanking that is interested in \cci{li} be present
as a separate leaf in the syntactic tree. 
Sarf provides a general solution for the segmentation correspondence problem 
since the valid compound affixes preserve the input text segmentation. 
In particular, a partial affix \noTrutfRL{ل} \cci{Al/DET}
connects to the atomic affix \noTrutfRL{ل} \cci{li/PREP} and resolves the problem.

%is segmented into two tokens:
%\cci{li/PREP} and \cci{Al/DET + qaDA'/NOUN}~\citep{LRECMaamouriKB08}. 
%The best approximation of the unvocalized entry 
%of each token is
%\noTrnoVocRL{li} and \noTrnoVocRL{Alqa.dA'}, respectively, 
%with an extra letter \RL{A}. 
%This is not a faithful representation of the original text data 
%and the segmentation does not correspond with that of the
%input text.
\end{itemize}

\todo{R3.1}
Alternatively, 
Sarf extends and refines the lexicons and compatibility category rules of BAMA and SAMA 
such that:
\begin{itemize}
  \item it represents only atomic affix morphemes in the lexicons,
  \item it adds prefix-prefix $R_{pp}$ and suffix-suffix $R_{xx}$ agglutinative and 
    fusional rules,
  \item it generates compound affixes from the atomic ones using $R_{pp}$ and $R_{xx}$, and
  \item it extends the stem lexicon with additional words necessary for the NLP applications it considers.
\end{itemize}

Considering the entries of Table \ref{t:affixes} as example, Sarf represents them using 
only five atomic affix morphemes and five 
prefix-prefix rules.\todo{R2.4} 
\end{comment}

### 2018-06-26 tex/sarf/overview.tex
sarf to ACLing draft added

\begin{figure}
\centering
\resizebox{.8\textwidth}{!} {
 \input{out/sarf_flow_diagram.pdftex_t} 
}
 \caption{Sarf flow diagrams: construction and traversal of morpheme structures, and generation of solutions.}
 \label{fig:flowdiagram}
\end{figure}
\todo{R2.10 fig1}
The flow diagram in Figure~\ref{fig:flowdiagram} illustrates the components of Sarf. 
%{\tt Construct Sarf structures} takes the morpheme {\tt lexicons} 
%including the prefix, stem, and suffix constituents. 
The stem lexicon of Sarf $L_s$ extends the lexicon of 
Buckwalter~\citep{Buckwalter:02} with 
proper and location names extracted from different online 
sources as well as biblical sources.
\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/},
Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}
The fusional and agglutinative affix rules 
%($L_p, L_x, R_{pp}, R_{ps}, R_{sx}, R_{xx}, R_{px}$)
encode affix-stem compatibility rules which are also taken from Buckwalter~\citep{Buckwalter:02} \todo{R.3.9}
Additionally, they encode morpheme-morpheme concatenation rules to implement 
agglutinative morphemes and they are associated with fusional rules expressed in 
regular expressions where necessary.
%The {\tt fusional/agglutinative rules} are used to check for the compatibility between 
%the prefix, stem, and suffix morphemes and construct the compound affixes. 
The use of diacritics to disambiguate morphological solutions is optional. 
%The {\tt diacritic on} option enables the user to disambiguate and filter the morphological 
%solutions based on possible input diacritics.

The {\em construct Sarf structures} process takes as input 
the lexicons, the fusional and agglutinative rules, and 
the ``use diacritic'' \todo{R3.19} option 
and constructs directed acyclic graph (DAG) structures that encode the 
affixes, and a root index trie structure that encodes the stems~\citep*{Aoe:89}. 
%Those structures are compact and deterministic representations and allow linear
%time traversal in each of them.

%Figure~\ref{fig:flowdiagram}(right) presents the traversal of the constructed
%structures for application specific analysis. 
%The Sarf structures serve as input to 
The {\em traverse structures} process
reads the user-provided Arabic text in question one character at a time and traverses 
the Sarf structures accordingly. 
The traversal produces a sequence of morphemes each with its morpheme features,
applies the fusional and agglutinative rules on each morpheme, 
checks for concatenation compatibility,
and constructs morphological solution feature trees. 
The construction of the solution trees calls the \cci{OnMatch API} 
at every morpheme match (control point) and takes into account 
the feature priority and selection rules 
defined by the NLP application developer. 
A \cci{feature priority rule} is an order on features that is 
followed when constructing the morphological solution tree. 
Features with higher priority appear at higher levels in the solution tree. 
\cci{A feature selection rule} defines the morphological features to be included
in the morphological solution tree; other features are ignored in the 
construction of the solution. 
The traversal reports the constructed solution trees to a \cci{filter} specified by 
the developer that either accepts or rejects the reported solutions.

%\vspace{-1.5em}
\subsection{Sarf structures}

Sarf represents affix lexicons and rules using directed acyclic graphs as shown in Figure~\ref{f:example}. 
This provides compactness as well as linear time traversal with respect to the input text.
%
\todo{R1.4}
Algorithm \cci{ConstructSarfDAGs} in Figure~\ref{f:dag-alg} describes how the 
structures are constructed from the lexicons and the rules. 
First, the algorithm creates efficient double array trie structures~\citep{Aoe:89} to
represent the lexicons and benefit from the common sub-strings. 
A morpheme is represented as an accept-node in the trie structure. 
Then for each relation tuple in $R_{pp}, R_{ps}, R_{sx},$ and $R_{xx}$ Sarf adds 
corresponding edges between the corresponding morphemes. 
Note that this transforms the affix trie structures into DAGs and keeps the 
stem structure as a trie since no stem-stem relations exist.

The \cci{AddEdges} algorithm iterates over all relation tuples in $R$, and for each
relation tuple $r=\langle c_1, c_2\rangle$ it creates edges for all morphemes 
$\langle m_1,m_2\rangle$ with compatibility categories $c_1$ and $c_2$, respectively.

In contrast, the Buckwalter analyzer considers all possible sub-strings and looks 
them up in affix hash tables, 
and performs several hash lookups in the stem hash tables in the order of all 
possible partitions of the input string. 
Sarf saves a binary version of the affix and stem structures which allows a 
fast loading time and 
only regenerates them if one of the lexicons 
and agglutinative and fusional rules are modified.

\transfalse
\begin{figure}
\centering
\resizebox{.8\textwidth}{!}{ 
 \input{FSM.pdftex_t}
}
\caption{Example affix DAGs and stem trie. 
The traversal of the shaded nodes corresponds to the final solution of \RL{wsyl`b-h-aa}.}
%\vspace{-2cm}
\label{f:example}
\end{figure}
\transtrue

\todo{R2.10 fig2}
The diagram in Figure~\ref{f:example}
illustrates the Sarf data structures. 
Subfigures (a), (b), and (c) in Figure~\ref{f:example}
represent ${\cal P}$ the prefix DAG,
${\cal S}$ the stem trie, and ${\cal X}$ the suffix
DAG, respectively. 
Boxes and circles denote morpheme versus non-morpheme nodes, 
respectively. 
The edges between nodes are labeled with input letters. 
Morpheme nodes contain the morphological features of the matching morpheme.
These features include the VMF, gloss, POS, and compatibility 
information for morpheme concatenation.

When we reach a morpheme node in ${\cal P}$, ${\cal S}$, 
or ${\cal X}$, we proceed with the traversal in the next data structure. 
We use the symbol $\epsilon$ to refer to this transition. 
Invalid moves from a current node given an input letter denote 
the absence of a valid solution through this traversal path.

\begin{lrbox}{0}
  \begin{tabular}{p{2.7in}p{0.02in}p{1.9in}}
\begin{lstlisting}
/*@\textbf{AddEdges}@*/($R$, $T_1$, $T_2$)
 foreach $r=(c_1,c_2)\in R$,
  foreach $m_1\in T_1$ such that category($m_1$)= $c_1$
   foreach $m_2\in T_2$ such that category($m_2$)= $c_2$
    createEdge($m_1$, $m_2$);
   end
  end
 end
\end{lstlisting}
& & 
\begin{lstlisting}
/*@\textbf{ConstructSarfDAGs}@*/($L_p$, $L_s$, $L_x$,
          $R_{pp}$,$R_{ps}$,$R_{sx}$,$R_{xx}$)
$T_p$ = makeTrie($L_p$)
$T_x$ = makeTrie($L_s$)
$T_s$ = makeTrie($L_x$)
AddEdges($R_{pp}$, $T_p$, $T_p$)
AddEdges($R_{ps}$, $T_p$, $T_s$)
AddEdges($R_{sx}$, $T_s$, $T_x$)
AddEdges($R_{xx}$, $T_x$, $T_x$)
\end{lstlisting}
\end{tabular}
\end{lrbox}

\begin{figure}[tb]
\resizebox{.9\textwidth}{!}{\usebox0}
\caption{Algorithm to construct Sarf structures.}
\label{f:dag-alg}
\end{figure}

%\begin{figure}[h]
%\begin{lstlisting}
%/*@\textbf{ConstructSarfDAGs}@*/($R_{ps}$, $T_p$, $T_s$ )
%  foreach $r=(pc_i, sc_j)\in R_{ps}$, //pc: prefix category, sc: stem category
%    foreach $p\in T_p$ such that category(p)= $pc_i$
%      foreach $s\in T_s$ such that category(s)= $sc_j$
%        createEdge(p, s);
%      end
%    end
%  end
%\end{lstlisting}
%\caption{Algorithm to construct edges between prefix and stem lexical nodes.}
%\label{f:dag-alg}
%\end{figure}
%
%\vspace{-1.5cm}

\transtrue

\subsection{Running example: Sarf traversal}

%In what follows, we present a running example of Sarf. 
%The running example explains how Sarf works,
In this section, we illustrate how Sarf constructs morphological solutions from affix DAGs and stem trie using an example string \RL{wsyl`b-h-aa al-lA`bwn}
\footnote{ 
\noTrRL{n}%
\noTrRL{w}
\noTrRL{b}
\noTrRL{`}
\noTrRL{a} 
\noTrRL{l}
\noTrRL{l}
\noTrRL{a} %
\noTrRL{a}
\noTrRL{h}
\noTrRL{b}
\noTrRL{`}
\noTrRL{l}
\noTrRL{y}
\noTrRL{s}
\noTrRL{w}
in separate form to ease following the example
}
 (and they will play it) \todo{R2.12}. The solution construction process also reveals how Sarf implements agglutinative and fusional affixes and how it handles `run-on' words.
Figure \ref{f:example} highlights the node sequences of morphemes in \RL{wsyl`b-h-aa al-lA`bwn} in the affix DAGs and stem trie. A square node (box) signifies reaching a complete morpheme and is referred to as a morpheme node, and an $\epsilon$-edge is a transition move from one tree structure to another or a starting move to ${\cal P}$. It connects any morpheme node in the source tree to the root node in the destination tree.\\

In our example, Sarf parses the string \RL{wsyl`b-h-aa al-lA`bwn}
%\RL{al-lA`bwn}\RL{wsyl`b-h-aa}.
and traverses the tree structures using a traversal path $\Phi$. 
$\Phi$ first uses the starting $\epsilon$-edge and moves to the root square node in ${\cal P}$. 
A square node implies that an $\epsilon$-edge from ${\cal P}$ to ${\cal S}$ is a valid transition, i.e. the morpheme sequence can start as a stem without a prefix \todo{R1.7}. All edges in ${\cal P}$ starting from the root node with the current parsed character 
\RL{w} are also valid edges. 
Therefore, there are two valid moves, \RL{w} and $\epsilon$.  
$\Phi$ hence spawns an exact copy of itself $\Psi$. 
$\Phi$ proceeds with \RL{w} in ${\cal P}$, and $\Psi$ moves to ${\cal S}$ through $\epsilon$. 
Each of $\Phi$ and $\Psi$ represents a valid analysis path so far. 
A traversal path ($\Phi$ or $\Psi$) dies when it reaches a node of no valid moves.
In our example, if there were no stems that start 
with the letter \RL{w}, $\Psi$ dies. 
With the complete stem trie, $\Psi$ actually dies when the input 
is at \RL{wsyl`} as no stem word in the Arabic language is of such morpheme sequence. 
%
The Sarf DAGs allow agglutinative and fusional affixes. 
In our example, after the \RL{w} move, $\Phi$ reaches a square morpheme node. A copy of $\Phi$ gets created and proceeds with $\epsilon$, and $\Phi$ proceeds with \RL{s}, in ${\cal P}$, being the next parsed character.
The move of $\Phi$ leads to a round node signifying that $\epsilon$ move is invalid. $\Phi$ proceeds with \RL{y} and reaches a square node corresponding to compound prefix \RL{wsy}.

Now $\Phi$ transitions along $\epsilon$, being the only valid move, into the root node of ${\cal S}$. Similar to its traversal in ${\cal P}$, $\Phi$ traverses moves corresponding to  \RL{l`} in ${\cal S}$. 
%
Before proceeding with \RL{b} and reaching a morpheme node, $\Phi$ checks if stem \RL{l`b} is compatible with the prefix \RL{wsy}. 
If the category of \RL{l`b} is compatible with the category of
\RL{wsy} then $\Phi$ moves to a morpheme node. 
\todo{R1.6}Otherwise, it moves to a non-morpheme node or dies. 
Sarf enables this by keeping compatibility category values as part of the morpheme nodes.%square morpheme 
Thus a morpheme node represents the whole morpheme sequence path leading to it within its tree.  
\\
Since $\Phi$ is now in a morpheme node in ${\cal S}$, it continues 
to traverse ${\cal S}$ and spawns a copy $\Xi$ that moves to root node of ${\cal X}$. 
$\Phi$ dies when \RL{h--} is parsed since no
\noTrRL{h--}-edge from the current node exists in ${\cal S}$.
On the other hand $\Xi$ moves along \RL{h--} in ${\cal X}$ and reaches a valid analysis morpheme node.
%which represents 
%a valid analysis.
\\\\\\
A Sarf traversal considers a full word at any morpheme node in ${\cal X}$ 
and continues the traversal using an $\epsilon$ transition
to the root node of ${\cal P}$.
This solves the `run-on' words problem.
%We left this transition out of Figure~\ref{f:example} for clarity. 
Consider there was no space between words 
\noTrRL{al-lA`bwn}
and 
\noTrRL{wsyl`b-h-aa}.
The traversal will transition to the root of ${\cal P}$ when the word 
\noTrRL{wsyl`b-h-aa} is fully consumed, and then the traversal of 
\noTrRL{al-lA`bwn} will resume. 
As for the other transitions from ${\cal X}$ morpheme nodes to the root of ${\cal P}$
before the completion of 
\noTrRL{wsyl`b-h-aa}, they will result in dead traversals and 
they will not be reported. 
%
Sarf reaches a solution and reports a valid traversal, including the morphological solutions of 
the `run-on' words, when it reaches text delimiters, 
such as white space and punctuation, with valid segmentation of the input string. \\
\todo{R1.7}In case of string with only stem sequences, empty transitions $\epsilon$ between ${\cal S}$ and ${\cal X}$, ${\cal X}$ and ${\cal P}$, 
and ${\cal P}$ and ${\cal X}$ to accept nodes allow parsing the string.
%The application developer can interfere in the analysis 
%at the control points to make decisions like ignoring 
%a possible solution path that is not interesting. 
%The developer may also decide to correct one traversal
%based on the decision of the others. 
%The developer uses the output of the traversals as input
%to make decision.

\subsection{Morphological solution construction}
\label{ss:solconstr}

%Given an input string, the traversal results in sequences of 
%morpheme nodes. 
%Let $M=\langle m_1,\dots , m_n\rangle$ represent a sequence of solution morphemes.

\itodo{R2.4.}This section describes how Sarf  constructs the morphological solutions of an input string. We use \RL{w-'klh} as an example input for illustration throughout the section.
As a first step, Sarf processes the string input by the traversal algorithm which results in two sequences of morpheme nodes where each sequence 
refers to a valid segmentation. The first sequence is $\langle$\RL{w},\RL{'},\RL{kl},\RL{h}$\rangle$ and 
the second sequence is $\langle$\RL{w},\RL{'kl},\RL{h}$\rangle$. 
Figures~\ref{f:strees} (a) to (d) show the first sequence and figures~\ref{f:strees} (a),(e),(d) show the second sequence. Below, we describe how the morphological solution tree of every sequence is generated.

\itodo{R2.11}
For every sequence, Sarf calls the developer-defined API at each morpheme node
and constructs its solution  tree according to the selected features and priority rules. 
Each path from the root of a morpheme solution tree to one of its leaves forms a
morpheme solution path. 
Figures~\ref{f:strees}(a) to (e) show the constructed solution feature trees 
of the morphemes \RL{w}, \RL{'}, \RL{kl}, \RL{h}, and \RL{'kl}, respectively. 
The figures show that the API selects the gloss, POS, and VMF 
features with decreasing priority. Hence, the gloss is at first level of the tree, then POS follows, and finally VMF is at leaf level. 
Figure~\ref{f:strees}(f) illustrates an alternative solution feature tree of 
the morpheme \RL{'kl} with POS at highest priority followed by \todo{R2.13}VMF and gloss. 
The comparison between the solution trees (e) and (f) shows 
how the priority rules can lead to the construction of smaller 
trees depending on the application at hand. 
For example, the solution feature tree (f) is more efficient 
if the developer is interested in the POS first. In such case, if \cci{Noun} POS is required, only the right hand side branch of \RL{'kl} solution tree (f) is considered. This would be much simpler than the case of (e) where every single branch has to be checked to determine whether to include it or not in the solution. More details on the API are provided in Section \ref{s:api}. \todo{3.8}\\
Finally, Sarf composes the set of valid morphological solutions for the input string from solution paths
that match the prefix-prefix, prefix-stem, stem-suffix, 
suffix-suffix, and prefix-suffix compatibility rules.
%
%Let $T=\langle t_1,\dots , t_n\rangle$ be the sequence of solution feature 
%trees referring to a solution morphemes. 
%Let ${\cal T}(t_i)=\{\cci{root to leaf paths of tree }t_i\}~1\leq i\leq n~$. 
%The set of full morphological solutions is 
%${\cal Q}\subseteq{\cal T}(t_1)\times {\cal T}(t_2)\times \dots\times {\cal T}(t_n)$. 
%The valid full morphological solutions are 
%subject to the prefix-prefix, prefix-stem, stem-suffix, 
%suffix-suffix, and prefix-suffix compatibility rules 
%of the morpheme solutions.
For example, the first paths (paths to the leftmost leaves) of each of the 
solution trees in Figures~\ref{f:strees}(a), (b), (c), and (d) are compatible.
The resulting morphological solution has the 
prefix \RL{w} with POS tag \cci{CONJ+}, 
gloss tag \cci{and}, and VMF tag \RL{wa}, 
the prefix \RL{'} with POS tag \cci{IVIS+}, 
gloss tag \cci{I}, and VMF tag \RL{'a},
the stem \noTrutfRL{kl} with POS tag \cci{VERB\_IMPERFECT},
gloss tag  \cci{trust/put in charge}, and VMF tag \RL{kil}, 
and the suffix \noTrutfRL{h} with POS tag \cci{PVSUFF:DO:3MS}, 
gloss tag \cci{him/it}, and VMF \RL{ho}.

\transfalse
\begin{figure}[tb]
\centering
\resizebox{.9\textwidth}{!}{ 
 \input{sftrees.pdftex_t}
}
\caption{\label{f:strees}Sample solution feature trees.}
%\vspace{-.5cm}
\end{figure}

### 2018-06-26 tex/sarf/placestable.tex
sarf to ACLing draft added

% Table generated by Excel2LaTeX from sheet 'Sheet1'
\begin{table}
  \begin{minipage}{\textwidth}
%  \relsize{-2} {
    \begin{tabular}{l|ccc|ccc|ccc}
    \hline \hline
    \multicolumn{1}{c}{} & \multicolumn{3}{c}{Kafi} & \multicolumn{3}{c}{Istibsar} & \multicolumn{3}{c}{Ibn Hanbal} \\
    \hline
          & correct & detected & required & correct & detected & required & correct & detected & required \\
    Places & 20    & 213   & 32    & 26    & 113   & 29    & 11    & 127   & 44 \\
    Possessive & 6     & 21    & 7     & 15    & 20    & 16    & 3     & 18    & 8 \\
    vicinity of name & 8     & 180   & 13    & 18    & 99    & 22    & 9     & 111   & 20 \\
    \hline \hline
    \end{tabular}%
 %   }
    \end{minipage}
  \caption{Sarf results of extracting places, places with possessive suffix, and places in vicinity of proper names.}
  \label{tab:placesRes}%
\end{table}%

### 2018-06-26 tex/sarf/redinc.tex
sarf to ACLing draft added

\subsection{Redundancy} 
\label{s:s:redundant}

Consider the partial lexicon of prefixes in Table~\ref{t:affixes}. 
\todo{R1.9.}The first five rows can be replaced with two atomic affix morphemes ( \RL{f} and \RL{y})  
and one partial affix morpheme (\RL{s}) in $L_{p}$, 
and three rules to generate compound morphemes in $R_{pp}$ (r (\RL{f}, \RL{y}), r (\RL{f}, \RL{s}), r (\RL{s}, \RL{y})). 
Representing prefix \RL{ya} (them/both) required four entries, three of them only differ
in their dependency on the added \RL{ya}. 
Representing prefix \RL{w} required the addition of five entries.
\todo{1.10}
With Sarf, the equivalent addition of \RL{ya} (them/both) requires only two rules in $R_{pp}$ that relate the categories of \RL{f} and \RL{s} to that of \RL{y}. 
The addition of \RL{w} requires only one additional entry (\RL{w}) in $L_p$.
The difference is much larger when we consider the full lexicon as will be shown 
in Section~\ref{sec:results}.

\subsection{Inconsistencies} 
\label{s:s:inc}

The entries in Tables~\ref{t:incBAMA} and ~\ref{t:incSAMA} list examples of the 197 and 208 inconsistencies detected in the affix 
lexicons of BAMA version 1.2 and SAMA version 3.1, respectively.
We found a small number of these inconsistencies manually and we computed the full list via comparing $L_p$ and $L_x$ with 
their counterparts computed using our agglutinative affixes. 
Most of the inconsistencies are direct results of partially redundant entries with erroneous tags. Experiments described later in section \ref{s:s:lexicon_consistency} shows that most of the detected inconsistencies are specifically related to gloss (Table~\ref{t:comp:SAMA}). \todo{3.12}
We note that SAMA corrected several BAMA inconsistencies, but also introduced 
several new ones when modifying existing entries to meet new standards and when introducing new entries.

%%%%%%%%%%%% BAMA inconsistencies table %%%%%%%%%%%%%%%%
\input{bamainctable}

The following describes the BAMA inconsistencies illustrated in Table~\ref{t:incBAMA}:

\begin{enumerate}[(a)]
    %%%%%%%%%%%%%%%%%%% BAMA prefixes
\item \label{inc:plus}  $L_p$ omits a plus (+) symbol that indicates boundaries in compound prefixes. 
\item \label{inc:so}    $L_p$ omits the (so) alternative gloss that corresponds to \RL{f-} in several compound prefixes.
    %%%%%%%%%%%%%%%%%%% BAMA suffixes
\item \label{inc:num_gender_subject} $L_x$ omits gender and number qualifiers 
  that appear within within square brackets
    from several glosses of subject suffixes.
                % not very precise seems sometimes also uses () for subjects like:
                %اكن    اكُنَّ  PVSuff-Ah       they (both) <verb> you (women)  +A/PVSUFF_SUBJ:3MD+kun~a/PVSUFF_DO:2FP  
\item \label{inc:num_gender_object} $L_x$ omits gender and number qualifiers 
    from several glosses of subject suffixes that appear within parenthesis.
\item \label{inc:both} $L_x$
    expresses the dual quantifier as `them (both)' in the majority of the entries, and 
    as `(both of) them' in several entries.
\item \label{inc:dot} $L_x$ omits the dot (`.') symbol from the gloss abbreviation of plural.
\item \label{inc:pos} $L_x$ contains POS tags that are not consistent with the semantics of the vocalized tags for compound affixes.
\end{enumerate}
\

%%%%%%%%%%%% SAMA inconsistencies table %%%%%%%%%%%%%%%%
\input{samainctable}

The following describes the SAMA inconsistencies illustrated in Table~\ref{t:incSAMA}:
\todo{R3.11 item (\ref{inc:relaxed)} below}

\arabtrue

\begin{enumerate}[(a)]
    %%%%%%%%%%%%%%%%%%% SAMA prefixes
\item \label{inc:relaxed}
  $L_p$ misses entries for Alef prefixes with omitted hamza or madda due to relaxed
    writing standards which are common in many documents. 
    This is resolved in SAMA for standalone Alef prefixes via preprocessing tokens and flipping all forms of Alef into one form. 
    We report it here since compound prefix entries with Alef are all listed, 
    e.g. \RL{sA}, 
    \utfRL{سأ}, \utfRL{سآ}, and 
    the standalone prefixes are available but commented out.
Sarf addresses this concern by the separation of prefixes into partial and atomic variants, and by 
introducing agglutinative and fusional rules. The rules construct compound prefixes from partial and atomic prefixes. 
This, therefore, enables the introduction of all prefixes that include Alef, with any of its alternative spellings, 
by only adding the Alef alternatives into the atomic prefix set. 
\item \label{inc:by} $L_p$ contains an additional erroneous alternative gloss for \RL{wa-} in only one compound prefix~\RL{wa-}; while correctly not included elsewhere.
\item \label{inc:fa} $L_p$  contains stray spaces in the vocalized tags of one of the \RL{fa-} alternatives.
\item \label{inc:wafa} $L_p$  contains an entry that supports the concatenation of \RL{wa-} and \RL{fa-} conjunctions. 
    This entry is erroneous and is illegal in Standard Arabic.
    %%%%%%%%%%%%%%%%%%% SAMA suffixes
\item \label{inc:def} $L_x$  omits the definite indicator in the gloss of several suffixes.
\item \label{inc:num_gender} 
    $L_x$  omits gender and number qualifiers that appear within square brackets 
    from the gloss tags of several suffixes.
\item \label{inc:both:sama} 
    $L_x$  
    expresses the dual quantifier as `them (both)' in the majority of the entries, and 
    as `(both of) them' in several other entries.
\item \label{inc:num_gender_differences} 
    $L_x$  
    contains number indicators in the gloss tags still expressed in BAMA style.% square brackets versus parentheses
\item \label{inc:jus} 
    $L_x$  contains entries with an 
    {\em indicative} gloss mood and a {\em jussive} POS mood.
\item \label{inc:dot:SAMA} 
    $L_x$  omits dot (`.') for the abbreviation of plural in several gloss tags.
\item \label{inc:pos:SAMA} 
    $L_x$  
    represents a repeated consonant by a shadda in the POS tag where it should not. 
    SAMA POS tags should spell out the repeated consonants if each belongs to one.\todo{R3.19}  
    In SAMA, the repeated consonant (of the shadda) is spelled out
    whenever the consonants has its separated partial POS tag. 
\end{enumerate}
%\vspace{-1em}

In addition, 53 BAMA and 27 SAMA minor differences exist between $L_p$ and $L_x$ of BAMA and SAMA and
their counterparts computed using our agglutinative affixes. 
For example, the BAMA gloss tags for prefixes that contain
`bi/PREP' report `with/by' in some entries and its reverse `by/with' in others. 
In addition, we detected several entries in $L_p$ of SAMA with no category compatibility 
rules in $R_{ps}$, $R_{sx}$, and $R_{px}$.

### 2018-06-26 tex/sarf/related.tex
sarf to ACLing draft added

\begin{table}[tb]
%\caption{Comaprison of Sarf with SAMA~\citep{Kulick:10}, 
%ElixirFM~\citep{Otakar:07}, 
%MADA+TOKAN~\citep{},
%MADAMIRA~\citep{pasha2014madamira}, 
%Beesley~\citep{}, and
%Xerox~\citep{Beesley:01}
%}
\centering
\begin{minipage}{\textwidth}
\caption{\label{tab:comparison}Comparison of Sarf with SAMA, 
ElixirFM, MADA+TOKAN, MADAMIRA, Beesley, and Fassieh}
\relsize{-1} {
  \begin{tabular}{lccccccc}\toprule
             & Sarf & SAMA & ElixirFM & MADA+TOKAN & MADAMIRA & Beesley & Fassieh\\ \colrule
Application customizable & $\checkmark$ & - & - & - &  - & - & - \\[4pt]
Feature selection & $\checkmark$ & - & - &  - & $\checkmark$ & - & - \\[4pt]
Run-on words & $\checkmark$ & - & - & - & - & - & - \\[4pt]
Partial diacritics & $\checkmark$ & - & - & - & - & $\checkmark$ & - \\[4pt]
Affix segmentation & $\checkmark$ & - & functional & tokenization schemes & statistical & - & - \\[4pt]
Root-Pattern & - & - & $\checkmark$ & - & - & $\checkmark$ & $\checkmark$ \\[4pt]
Automated disambiguation & - & - & - & SVM & SVM & - & maximum  aposteriori  \\ 
\botrule
\end{tabular}
}
\end{minipage}
\end{table}

In this section, we review work related to Arabic morphological analyzers,
segmentation correspondence, 
partial diacritics, and application specific analyzers and compare it to Sarf.

Table~\ref{tab:comparison} summarizes the comparison between Sarf 
and related Arabic morphological analyzers. 
Only ElixirFM~\citep{Otakar:07}, Beesley~\citep*{Beesley:03}, and Fassieh~\citep*{attia2009fassieh} 
provide root-pattern analysis of the stem. 
ElixirFM, MADAMIRA~\citep{pasha2014madamira}, and MADA+TOKAN~\citep*{Habash:09} are 
based on BAMA and SAMA and use functional and statistical techniques to 
address the segmentation problem by reverse engineering the multiple tags of the 
affixes. 
Sarf differs in that the segmentation is an output of the morphological
analysis and not a reverse engineering of the multi-tag affixes. 
Sarf is the only analyzer that addresses the 'run-on words' problem 
and solves it while performing the analysis. 
MADA+TOKAN, MADAMIRA, and Fassieh apply morphological disambiguation 
using support vector machines (SVM), 
and {\em maximum a posteriori} (MAP) estimation, respectively.
Beesley and Sarf consider partial diacritics to eliminate morphological 
solutions that are not in agreement with the partial diacritization. 
Sarf provides an application customizable analyzer that enables 
the developer to control and refine the analysis on the fly 
and filter the solution features. 
MADA+TOKAN and MADAMIRA provide \todo{R2.13} partial control over the output, and not the 
analysis, where MADA+TOKAN allows the user to select \todo{R2.13 } from several segmentation schemes and 
MADAMIRA enables the user to select solution features.

Sarf builds upon the lexicon of Buckwalter\citep{Buckwalter:02}.
SAMA is an updated version of BAMA with increased lexicon coverage 
and additional POS tags~\citep*{maamouri2010ldc}.
Sarf differs from Buckwalter and SAMA in that it defines 
agglutinative and fusional affixes using a shorter list of affixes
and a list of concatenation compatibility rules that allow 
prefix-prefix and suffix-suffix concatenations.
This allows Sarf to better maintain %and propagate 
the morphological tags associated with the affixes.

Buckwalter\citep{Buckwalter:02} and SAMA~\citep{Maamouri:10} 
produce a set of segmentation solutions for
a word, compute the morphological solutions for each segment, 
compute the product of the solutions, eliminate \todo{R2.13 } the 
incompatible solutions, and then report \todo{R2.13 } the valid solutions. 
Sarf traverses the affix and stem structures with the input word
character by character and keeps a stack of morpheme nodes. 
When a morpheme node in a structure is met, 
it is checked for compatibility with the stack of nodes. 
Consequently, Sarf generates only the solutions with valid 
segmentation, and reports only those with compatible stem and affix 
concatenation.

SAMA was refined to interact with the Arabic Treebank(ATB)~\citep{Maamouri:04}  %Arabic Treebank at the Linguistic Data Consortium 
project after the addition of a large new corpus. 
The algorithmic changes in SAMA were done manually 
to integrate with the ATB format. 
Sarf API allows for customizable refinements 
and allows Sarf to interact with any 
application on the fly
without the modification of the morphological engine itself
as was reportedly required with SAMA~\citep{Maamouri:10}

Like ElixirFM~\citep{Otakar:07}, Sarf builds on the lexicon
of the Buckwalter analyzer. 
Sarf also uses deterministic parsing with tries and DAGs
to implement the affix and stem structures. 
We think that the inferential-realizational approach 
of ElixirFM that is highly compatible with the 
Arabic linguistic description~\citep*{Badawi:04}
can benefit from many features unique to the Arabic language.
Sarf leaves implementing that to the developer customization 
through the API since in several cases the NLP application that uses the morphological analyzer
needs only a partial linguistic model of Arabic.

MADA+TOKAN~\citep{Habash:09} is a toolkit for Arabic tokenization, diacritization, 
morphological disambiguation, POS tagging, stemming, and lemmatization. 
Sarf performs all those tasks except for morphological disambiguation 
where MADA uses SVM.
\todo{R3.6}
Sarf keeps a stack of the positions that partition text into morphemes 
as part of the solution construction process. 
Therefore, the text segments corresponding to the solution features is preserved 
with no need for further post-processing. 
%a separate segmentor such as TOKAN is not required.

%Similar to The MADA+TOKAN toolkit, 
%Also, Sarf thinks of the several valid morphological analyses as 
%a richness that should be exploited at a higher abstract level 
%rather than an inaccuracy that should be corrected.

MADAMIRA is a tool for Arabic morphological analysis and disambiguation 
that is based on the general design of MADA, 
an Arabic morphological analyzer and disambiguator, with additional components 
inspired by AMIRA~\citep{pasha2014madamira}, a language independent SVM based analyzer. 
MADAMIRA improves upon the two systems and returns information selectively 
upon the request of the user. 
Sarf provides means for the developer to control and adapt 
the morphological analysis according to application needs. 
Moreover, the API enables the developer to implement high level 
applications such as NER which is provided by AMIRA.

%Xerox rules can be compiled into specialized finite state
%machine (FSM) based analyzers as described in~\citep{Beesley:03}.
%However, the efficiency of the resulting analyzer depends on the
%way the Xerox rules are written. 
%Writing application specific Xerox grammars and rules, or modifying the existing ones, 
%requires the NLP application developer to have deep knowledge 
%of compilation techniques, context free grammars, 
%and morphological analysis. 
%%Other morphological analyzers such as 
%%Amira~\citep{Diab:07,Benajiba:07},
%%and MAGEAD~\citep{Habash:05}%, and MADA+TOKAN~\citep{Habash:09} 
%%use machine learning and support vector machines (SVM) 
%%to enhance the accuracy of the morphological analysis at the expense 
%%of performance.

Beesley~\citep{Beesley:03} compiles 
Xerox rules into specialized finite state
machine (FSM) based morphological analyzer.
The number of machines 
generated by a compiler for Xerox rules cannot be controlled by the developer of the analyzer, 
and the composition of the FSMs into a single framework is a difficult task 
~\citep{Beesley:01}.
Consequently the efficiency of the resulting analyzer depends on the
way the Xerox rules are written. 
Writing application specific Xerox grammars and rules, or modifying the existing ones, 
requires deep knowledge and insight from the NLP application developer
in compilation techniques, context free grammars, 
and morphological analysis. 
Sarf constructs a framework of efficient structures %including 
%a trie and two DAGs 
that encode the stems and the agglutinative and fusional affixes, respectively.
\todo{R.3.15}
The structures are traversed in a manner similar to the Xerox finite state machines, 
however, they are manually optimized to reduce the number of states and the 
number of non-deterministic transitions. 
Control on non-determinism is left to the compilers with Xerox.
Sarf also provides an application customizable API that allows the developer 
to control the analysis. 
Doing the equivalent with Beesley 
requires the modification of the Xerox rules and 
the recompilation of the analyzer.
Unlike Sarf, Beesley provides a root-pattern analysis of the stem.

Fassieh is a commercial Arabic text annotation tool that 
enables the production of large Arabic text corpora~\citep{attia2009fassieh}.
The tool supports Arabic text factorization including 
morphological analysis, POS tagging, full phonetic transcription, 
and lexical semantics analysis in an automatic mode. 
Unlike Sarf, Fassieh provides morphological 
disambiguation and root-pattern analysis. 
However, Fassieh does not provide segmentation of the affix 
and reports it as a whole unit. 
This tool is not directly accessible to the research community 
and requires commercial licensing. 
Sarf differs in that it is an open-source application 
customizable tool that solves the affix segmentation and 'run-on words' problems.

The work in~\citep*{Attia:10} addresses the detection of 
Arabic Multi-word Expressions (MWE). 
They define MWEs as 'idiosyncratic interpretations 
that cross word boundaries or spaces'. 
Sarf adopts a similar approach for specific entities such as person names 
and place names.

Several researchers stress the importance of 
correspondence
between the input string and the tokens of the morphological 
solutions. 
Some work uses POS tags and a syntactic morphological agreement
hypothesis to refine 
syntactic boundaries within words~\citep{Regina2011}.
The work in~\citep*{LIMASemmar05}\citep{ELRASemmar08}
uses an extensive lexicon 
with 3,164,000 stems, stem rewrite rules~\citep*{Darwish:02}, 
syntax analysis, proclitics, and enclitics to address the 
same problem. 
Parallel traversal of the input string and the tokens 
of the morphological solution, while accounting for all possible
SAMA normalizations, partially solves the problem as 
reported in~\citep{LRECMaamouriKB08,lrecKulickBM10}.
Later notes in the documentation of the ATB~\citep*{ATB32LDCNum}
indicate that extensive manual work is still required and that
later versions may drop the input tokens. 
\citep{Regina2011} uses syntactic analysis to resolve the same problem. 
%A significant amount of the literature on Arabic NLP uses the Arabic Tree Bank 
%(ATB)~\citep{Maamouri:04} with tags 
%generated from BAMA and SAMA for learning and 
%evaluation~\citep{ShaalanMF10,Benajiba:07,jumaily2011}.

The survey in~\citep{Sughaiyer:04} compares
several morphological analyzers. 
Analyzers such as~\citep*{Khoja:01}\citep{Darwish:02} 
target specific applications in the analyzer itself 
or use a specific set of POS tags as their reference.
Sarf differs in that it is a general morphological 
analyzer that reports all possible solutions. 
It is application customizable in the sense that the API is used to 
control and prioritize the analysis, refine the solution features, 
and associate morphemes with developer-defined categories.

The work in \citep{Chaaben:10,Attia:00,Beesley:01} 
considers partial diacritics and performs \todo{R2.13 } morphological 
disambiguation by filtering the full morphological solutions 
and excluding inconsistent ones. 
This approach constructs several solutions that will be excluded later. 
Sarf considers diacritic consistency at the morpheme level instead of the final 
solution level. 
It checks for diacritic consistency %using Algorithm~\ref{t:consistent} 
between the input morpheme and the candidate VMF features at every 
accept node during the traversal of Sarf structures.
Sarf analysis proceeds with the consistent VMFs
and terminates the inconsistent ones.

\citep{Beesley:01}, \citep{Chaaben:10}, and \citep{Attia:00} present 
analyzers that consider partial diacritics for morphological disambiguation. 
They filter the output morphological analysis based on 
compatibility with input diacritics if found.
Sarf differs in that it considers the diacritics at morpheme boundaries 
to generate only the diacritic matching solutions, rather than 
generating all morphological solutions then filtering them.

### 2018-06-26 tex/sarf/results.tex
sarf to ACLing draft added

In this section we present and discuss the results of evaluating Sarf and compare it to existing morphological 
analyzers such as SAMA and ElixirFM.%BAMA, SAMA, MADA+TOKAN, and ElixirFM.
%
%We compared the segmentation capabilities of Sarf to that of SAMA and MADA+TOKAN under the ATB v3.2. 
%We also evaluated the presence of the BAMA and SAMA inconsistencies of Tables~\ref{t:incBAMA} and \ref{t:incSAMA} in the ATB v3.2 Part 3. 
%Results show that the annotations of Sarf nearly perfectly agree with the manual annotations of the ATB v3.2. 
%The results also show that Sarf can automatically correct 0.76\% of the ATB annotations due to lexicon inconsistencies. 
%
%Moreover, we evaluated the efficiency of Sarf by measuring the size of the lexicons, lexicon augmentation cost, 
%and the runtime and accuracy performance  of the analyzer. 
%
%We compared the cost of augmenting Sarf with the question clitic (hamza \RL{'}) to that of BAMA and SAMA. 
We conducted two experiments to evaluate the performance of Sarf compared to SAMA and ElixirFM. 
The experiments show the advantage of Sarf over the other analyzers in terms of performance. 
%We also present results that compare Sarf to existing morphological analyzers such as Buckwalter and ElixirFM. 
We also evaluate our hypothesis of the utility of a customizable Arabic morphological analyzer.

%\subsection{Inconsistency}

\begin{comment}
%\vspace{-1.5cm}
\subsection{Segmentation correspondence}

We evaluated the segmentation correspondence capabilities of 
Sarf under the segmentation guidelines of the ATB v3.2.
For each entry in the `before' section of ATB v3.2 
Part 3 with a correct SAMA solution,
we automatically computed the segmentation using Sarf 
and compared the result to the segmentation in
the `after' entries that are manually validated by the LDC.

SAMA had a correct morphological 
solution for 273,618 words out of 393,201 ATB words.
Those required later segmentation and manual validation. 
Our automatically generated segmentation agree with 
99.991\% with the oracle ATB segmentation for the 273,618 words. 
%Our automatically generated segmentation agree with 99.991\% 
%of 273,618 words out of the total 393,710 ATB words 
%for which SAMA had a correct morphological solution and that 
%required later segmentation and manual validation.
We inspected the 25 entries for which our segmentation 
disagreed with ATB and found that both segmentations were valid. 
For example, the entry \utfRL{منّا} is formed of \noTrutfRL{من} 
\cci{min/PREP} (from) and \noTrutfRL{نا} \cci{nA/PRON\_1P} (us). 
Our segmentation is \noTrutfRL{من}+\noTrRL{|BBA} 
while that of ATB is \noTrutfRL{م+نّا}. 
When the morphemes are concatenated, the two \noTrRL{-n-} consonants 
can be fused into a single one with a shadda (\noTrRL{|BB}).
Since it is common to omit the shadda, we are left with a single consonant
at the boundary, which can correspond to either morpheme.

%\input{tokantable}

\todo{R1.14} Since Sarf preserves correspondence when performing segmentation, it is capable of
generating a vocalized tag in the ATB `after' dataset which carries more information in
15.47\% of the time than the counterpart POS derived vocalized entry. 
The POS derived ATB vocalization processes, in some cases, the source token by splitting it into multiple unvocalized POS tokens which are then converted into vocalized ATB tokens. For example, 
the word \utfRL{للقضاء} (for/to the justice/judiciary) is split into \cci{li/PREP + Al/DET + qaDA'/NOUN} tokens. In such cases, the POS tokens differ from the source token. 
%The vocalized entry in the `after' dataset was therefore dropped. 
This issue does not exist in Sarf and the vocalized entry is maintained. 
%Since Sarf preserves correspondence when performing segmentation,
%it is capable of generating a vocalized tag in the ATB `after' dataset,
%which carries more information in 15.47\% of the time 
%than the counterpart POS derived vocalized entry.
%The vocalized entry in the `after' dataset was dropped 
%because of maintenance and segmentation issues.
%With Sarf that entry can be maintained.

\subsection{Lexicon consistency}
\label{s:s:lexicon_consistency}
We evaluated the presence of the inconsistencies of Tables~\ref{t:incBAMA} and \ref{t:incSAMA} 
in the ATB v3.2 Part 3.
The first experiment considered the ATB entries that adopted the SAMA solution.
The rest of the entries have manually entered solutions.
The gloss inconsistencies affect 0.76\% of those entries.

\begin{table}[!tbp]
\begin{minipage}{0.6\textwidth}
\begin{tabular}{p{3cm}cc} 
\hline \hline
Conflict       & $\Delta_{word}$  & $\Delta_{analysis}$ \\ \hline
POS            & 0.206\%        & 0.016\% \\
Gloss          & 8.317\%        & 3.226\% \\
WaFa           & 0.251\%        & 0.022\% \\
%Question clitic & 0.100\% & 0.033\% \\ 
{\bf Total } & {\bf 8.774\%} & {\bf 3.264\%}\\
\hline \hline
\end{tabular}
\end{minipage}
\caption{Effect of lexicon inconsistencies.}
%\vspace{-1.5cm}
\label{t:comp:SAMA}
\end{table}
The second experiment considered all tokens in the ATB with a SAMA solution. 
The $\Delta_{word}$ column of Table~\ref{t:comp:SAMA}
reports the ratio of the affected ATB words, 
and the $\Delta_{analysis}$ columns reports the ratio 
of the conflicting morphological analyses.
One word might have several morphological 
solutions which explains the difference.
The rows report the effect of the POS and gloss tags, 
and that of the wrong prefix entry~\ref{inc:wafa} of Table~\ref{t:incSAMA}.
In total 8.774\% of the words and 3.264\% of the 
morphological solutions are affected.
Sarf automatically solves all these conflicts. 
%in addition, Sarf also automatically solves 
%the 7 ATB occurrences of the question clitic.
%\subsection{Efficiency}

\begin{table}
\begin{minipage}{\textwidth}
\begin{tabular}{lccp{0.05cm}ccp{0.05cm}cc}
\hline \hline
 & $|L_p|$ & $|R_{pp}|$ & & $|L_x|$ & $|R_{xx}|$ & & $\Delta_L^{\noTrutfRL{ء}}$ & $\Delta_R^{\noTrutfRL{ء}}$  \\
 \hline
%    \\ \cline{2-3} \cline{5-6} \cline{8-9}
{\bf BAMA}    & 299 & --& &618 & --  & & 295 & -- \\
{\bf Sarf}: Agglutinative only & 70      & 89& &181 & 123 & & 1   & 32 \\
{\bf Sarf}: Aggl. \& fusional only &                43  & 89& &146 & 128 & & 1   & 32 \\
{\bf Sarf}: Aggl., fusional, \& grouping & 41      & 7     & &146 & 32      & & 1   & 1  \\  \hline
{\bf SAMA } &    1325 & --      & &945  & -- & & 1,296 & -- \\
{\bf Sarf}: Agglutinative only & 107 & 129       & & 221 & 188& & 1     & 38     \\ 
{\bf Sarf}: Aggl. \& fusional only & 56      & 129 & & 188   & 194& & 1     & 38 \\ 
{\bf Sarf}: Aggl., fusional, \& grouping & 53      & 18 & & 188    & 64 & & 1     & 1  \\
\hline \hline
\end{tabular}
\end{minipage}
\caption{Comparison of lexicon size between BAMA, SAMA, and Sarf. Sarf is represented as BAMA or SAMA with Agglutinative, fusional, \& grouping.}
%\vspace{-2cm}
\label{t:reduction}
\end{table}

%\vspace{-2cm}
\subsection{Lexicon size.}~~
The $|L_p|$, $|L_x|$, $|R_{pp}|$, and $|R_{xx}|$ 
entries in Table~\ref{t:reduction} report 
the size of the prefix lexicon, the size of the suffix lexicon,
the number of prefix-prefix concatenation rules , 
and the number of suffix-suffix concatenation rules , respectively for BAMA and SAMA.
The entries also report the effect of using 
agglutinative affixes, fusional rules,
and grouping of rules with similar patterns using wildcards on reducing the size of the lexicon.

\todo{R1.16}
The size of the affix lexicons of Sarf is obtained by applying agglutinative affixes, fusional rules,
and grouping of rules on BAMA and SAMA affix lexicons. 
Sarf only requires 226 entries to represent the 917 entries 
of BAMA affixes with inconsistencies corrected. 
Sarf entries include 41 prefixes, seven prefix-prefix rules, 
146 suffixes, and 32 suffix-suffix rules 
while BAMA affixes include 299 prefixes and 618 suffixes.
Similarly, Sarf only requires 323 entries to represent 2,270 entries 
of SAMA affixes with inconsistencies corrected.

The transition from BAMA to SAMA required the addition of 1,353 entries to 
the lexicons of BAMA.\todo{R1.15.} 
Sarf only required the addition of one order of magnitude less entries 
to accommodate an equivalent change. 
The 136 entries consisted of 12 more entries in $L_p$, 
42 in $L_x$, 18 rules in $R_{pp}$, and 64 in $R_{xx}$.
%We also note that we detect most of the inconsistencies automatically 
%and only needed to validate our corrections in textbooks and corpora.

{\bf Augmentation.}~~
The question clitic, denoted by the glottal sign 
(hamza \utfRL{أ}), is missing in BAMA and SAMA as noted by ~\citep{Attia:06b}.
The $\Delta_L^{\noTrutfRL{ء}}$ and $\Delta_R^{\noTrutfRL{ء}}$ entries 
in Table~\ref{t:reduction} show the difference in the number of 
additional affixes and rules needed to accommodate 
for the addition of the question clitic.
Our method only requires the addition of one 
atomic affix and one fusional concatenation rule.
Whereas BAMA and SAMA need 295 and 1,296 additional entries to their lexicons, respectively, 
Moreover, the process requires manual intervention 
with a possibility of inducing inconsistencies in the process.
This is evidence that our method is better for 
the consistency and the maintenance of the lexicons.

\tododone{the table below should include the atomic features that sarf is concerned with. 
proper names, family connectors, tell-connectors, places, place prepositions, possessive.
need to recompute the values using sarf.}
\end{comment}
%\input{performanceTable}

%\input{performanceTableNew}

%\vspace{-1.2cm}
\subsection{Performance}
\label{ss:perform}

\todo{R1.5}
We evaluated the speed of Sarf and compared it to that of SAMA and ElixirFM. 
We selected a random article from an Arabic newspaper containing 1,122 words and 
computed the morphological solutions of those words using the three analyzers. 
Sarf with no application specific API implemented generates all the solutions of 
all words in 0.72 seconds while SAMA does 
the task in 2.46 seconds. 
However, ElixirFM requires much more time with 34 minutes and 19 seconds.

We also evaluated the accuracy and runtime efficiency of 
Sarf and compared that to SAMA and ElixirFM
with one of the applications~\citep{ZaMaFlairs2012HadithBio}
that used Sarf as a back-end morphological analyzer. 
The hadith extraction application~\citep{ZaMaFlairs2012HadithBio} is concerned
with analyzing a book of traditions
related to prophet Mohammad through a chain of narrators. 
%
Narrators are identified by composite proper person names
that are connected with family connectors. 
For example, the chain of narrators \RL{.hdd_tnA qtybT bn s`yd `n ^gryr}, starts with the first narrator 
\RL{qtybT bn s`yd} where \RL{qtybT} and \RL{s`yd} 
are {\tt proper person names}.
The word \RL{bn} (son of) indicates a parental 
relation that we will refer to as {\tt family-connectors}.
The name \RL{^gryr} is a proper person name denoting 
the second narrator, and the words 
\RL{.hdd_tnA} (told us) and \RL{`n} (from/about) indicate 
a narration relation and we refer to them as 
{\tt tell-connectors}. 
Key to narrator detection are morphological features that point to 
places such as names and location prepositions.

%We calculated the execution time of the analyzers and 
%evaluated the accuracy of the tags in terms of precision and recall.
\input{exp1Results}

Table~\ref{tab:exp1Res} reports the results of detecting morphological features 
that define proper names, tell-connectors, and family connectors and compares the 
results with SAMA and ElixirFM in terms of accuracy and running time. 
%
The table considers 
three books of hadith selected arbitrarily~\citep*{IbnHanbal,AlKulayni,AlTousi}\footnote{We obtained the digitized books from online sources such as \href{http://www.yasoob.com/}{http://www.yasoob.com/} and \href{http://www.al-eman.com/}{http://www.al-eman.com/}. }. 
All experiments used a Linux operating system running on a 
dual core 2.66 Ghz 64-bit processor with 4GB of memory.

%For each book, the table reports the accuracy of the extracted 
%proper names, tell-connectors, and family connectors in terms 
%of precision and recall for each of the compared analyzers. 
%The table also reports on the execution time 
%required by each analyzer for each book.

Sarf scored higher recall for all the features and approximately 
similar precision across the three books. 
%The precision and recall measures of the tell-connectors 
%are very close between the analyzers across the three books. 
The precision and recall measures of the family connectors 
in Sarf and SAMA are close, unlike ElixirFM which reports 
lower accuracy measures. 
\todo{R1.17}
The precision for proper names is low since some proper names exist 
in the books but outside the context of a chain of narrator names
and are hence not wanted. 
After analyzing the results, it turned out that  ElixirFM
misses some gloss tags such as the `son' tag associated with the stem `\RL{bn}'.
%This token is an important and frequent tag in our application 
%which explains the low accuracy. 
Sarf produces significantly higher proper name recall measure 
compared to SAMA and ElixirFM. This mainly due to augmenting the stem lexicon of 
Sarf with proper names as explained in Section~\ref{sec:overview}.

Sarf outperformed both SAMA and ElixirFM in running time even without the 
use of the feature priority and the feature selection API. 
SAMA performed better than ElixirFM. 
%This result was expected for Sarf as it uses structures 
%that enables the traversal in an efficient way.

%\input{placestable}

%ElixirFM outperformed Buckwalter in terms of accuracy.
%We think the big difference is due to the sophistication 
%of the formal functional morphological analysis of ElixirFM 
%and its reliance on additional annotations extracted from 
%both ATB~\citep{Maamouri:04} and the 
%Prague Arabic Dependency Treebank~\citep{Prague04}.

\subsection{Sarf API}

\input{apitable}

We evaluated the application customizable Sarf API with several applications that
used Sarf. 
\todo{R2.13}
Table~\ref{t:sarfapi} reports the number of morphological solutions per word 
and the runtime of Sarf for several applications before and after using the application customizable Sarf API.
%to refine the analysis and compares that with the same applications 
%without the use of the Sarf API. 
The numbers show the utility of the Sarf API at improving the runtime and the efficiency of 
NLP applications by an order or magnitude across the four applications.

In what follows, we shortly describe the applications. 
Detailed results including accuracy 
measures can be found in the corresponding papers.

The temporal entity extraction application uses finite state machines driven by
morphological features that indicate temporal units, intervals, quantities, and
prepositions as input~\citep{ZaMa2012IJCLATime}.
The application processed 43 articles arbitrarily selected from local newspapers. 
The average number of solutions that Sarf reported per word without the use of the API
was 4.33 solutions per word. 
%The use of the API indicating that 
The application specific refinements of the analysis implemented using the
Sarf API eliminated solutions that the application is not 
interested in and thus the number of solutions per word that Sarf ended up
reporting was 1.74 and that in turn resulted in a substantial improvement in runtime.

The hadith segmentation and extraction application extracts 
chains of narrators from hadith books using finite state machines that take 
morphological features such as gloss and POS tags that indicate proper names, 
family relations, tell-connections, places, and possessive nouns as input
~\citep{ZaMaFlairs2012HadithBio}.
The application processed a total of 41 books with a total of 196,171 narrations 
to build a graph where narrators are nodes and their relation to each other are edges. 
Similarly to above, the number of solutions per word improved from 5.41 to 1.82 
and the runtime improved by more than half.

The biography application matches a narrator extracted from the hadith application 
to corresponding biographies in biography books and extracts entities such as 
birth and death dates, location, students, professors, and authentication 
qualifications
~\citep{ZaMaFlairs2012HadithBio}.
The biography application uses narrator extraction and temporal extraction, 
and in addition
it uses morphological features that indicate qualifying adjectives that relate 
to authenticity. 
The application used the graph generated from the hadith application and 
processed 15 books of biographies with a total of 79,946 biographies 
to (1) segment the biographies, (2) extract the narrators in the biographies, and 
(3) annotate the narrator nodes in the graph with qualifiers extracted from the 
corresponding biographies. 
The use of the Sarf API improved the solutions per word ratio from 5.61 to 2.12
and improved the run time by almost a factor of 4.

The genealogy extraction application extracts a family tree and learns words
that indicate family relations from biblical texts.
It uses morphological features that indicate proper names, family relations, places, 
and professions~\citep{ZaMaHaCicling2012Entity}. 
The application processed the book of Genesis with fifty verses. 
21,385 words. 
The use of Sarf API improved the solutions per word ratio from 4.63 to 2.44 and improved run time substantially.

%\input{exp2Results}
%
%The concept of having an application customizable analyzer is based 
%on the hypothesis that many NLP case studies need the 
%morphological analyzer to answer queries that do not need 
%high accuracy at the low morphological analysis level.
%We conducted another experiment to validate this hypothesis.
%Thus, we extended the first application with several 
%low-level morphological refinements to Sarf such as disambiguation of 
%female proper names so that the application can ignore them. 
%We also added a refinement that detected compound proper names 
%such as \RL{`bd alr.hmn} as they are common in narrator names. 
%We inspect the effect of low level refinements on the high 
%level output and report our results in Table~\ref{tab:exp2Res}.
%
%The `narrator name' row reflects the accuracy of 
%proper names detected within narrator structures. 
%The `narrator chain' reflects the accuracy of 
%narrators covered by extracted chains. 
%The `hadith segmentation' reports on the accuracy of 
%the hadiths extracted from the book. 
%The metrics (precision and recall) are computed 
%manually over a representative sample of each book.
%
%In general, we observe that the refinements 
%slightly improved performance and accuracy at
%the lowest level of abstraction. 
%However, the refinements had little to no effect on the high 
%accuracy measures at higher levels of abstraction.
%We believe that this is a strong supporting evidence to our hypothesis.

### 2018-06-26 tex/sarf/rev1.tex
sarf to ACLing draft added

\documentclass[11pt]{article}
\usepackage[letterpaper,margin=1.5in,left=35mm,right=20mm, top=20mm, bottom=20mm ]{geometry}

\usepackage{url}
\usepackage{todonotes}
\usepackage{enumitem}
\usepackage{amssymb}

% used for cross reference from psq2aig.tex
\usepackage{xr} 
\externaldocument{psqBMCaig}
% You ou can also add a prefix: 
%\externaldocument[I-]{chapterI}
% then you can reference it as \ref{I-a}

\RequirePackage{xspace}
\linespread{1.5}

\newcommand{\cci}[1]{\mbox{\textup{\small \texttt{#1}}}}
\newcommand{\psqlanguage}{\ensuremath{\mathit{J}}\xspace}
\newcommand{\psqcore}{\ensuremath{\mathit{Jcore}}\xspace}
\newcommand{\psqtool}{\ensuremath{\{\cal P\}{\cal S}\{\cal Q\}}\xspace}
\newcommand{\oneloop}{\ensuremath{\mathit{\cal O}lp}\xspace}
\newcommand{\true}{\ensuremath{\mathit{true}}\xspace}
\newcommand{\false}{\ensuremath{\mathit{false}}\xspace}
\newcommand{\itodo}[1]{\todo[inline,color=green]{#1}}
\newcommand{\Pm}{\ensuremath{\mathcal{S}}\xspace}         % program

\title{Sarf: Fast and Application Customizable Arabic Morphological Analyzer
 \\
~~ \\
  Revision summary }
\date{}

\begin{document}
\maketitle

{\noindent Dear Editor-in-Chief and Expert Reviewers,}

We would like to thank you for your recommendations, valuable comments, 
and constructive observations. 
In this revised version of the paper we made every effort to address the 
concerns brought up in your review. 
This document includes a summary of our changes and lists each of the 
reviewers’ comments followed by a description of how it was addressed.

For the convenience of the reviewers, we also provide online 
(\url{http://research-fadi.aub.edu.lb/psqaigfiles/psqaig-tse-revised-annotated.pdf})  
a version of the revised paper with annotations that point to changes and 
revisions and relate them to the concerns of the reviewers.

Thank you,
\newline
\hrule

\input{editorComments.tex}
~\newline
\hrule

\input{Reviewer1.comments}
~\newline
\hrule

\input{Reviewer2.comments}
~\newline
\hrule

\input{Reviewer3.comments}
~\newline
\hrule

\newsavebox\mytempbib
\savebox\mytempbib{\parbox{\textwidth}{
%\input{psqBMCaig.bbl}
}}

\end{document}

### 2018-06-26 tex/sarf/samainctable.tex
sarf to ACLing draft added

\begin{table}[!tbp]
\resizebox{0.95\textwidth}{!}{
%\begin{minipage}{0.8\textwidth}
\relsize{-1.5} {
\begin{tabular}{p{2.5cm}cccc}
\hline \hline
 & {\bf Affix } & \textbf{Vocalized} & {\bf Inconsistent tag} & \\ \hline
\multirow{2}{2.5cm}{
  {\bf \ref{inc:relaxed})} 
  missing standalone alef with no hamza {\bf prefix}  forms
}
% & \RL{A} & \utfRL{أَ} & I \\ 
 & \RL{A} & \utfRL{آ} & I  \\ 
% & \RL{A} & \utfRL{أُ} & I  \\
 & \RL{sA} & \utfRL{سَأُ} & I \\[6pt]
\multirow{2}{2.5cm}{
  {\bf \ref{inc:by}) } 
   additional by in {\bf gloss} 
}
 & \RL{w} & \RL{wa} & with      \\ 
 & \RL{wAl-} & \RL{waAl-} & with\colorbox{shadecolor}{\color{white} /by} + the \\[6pt]
\multirow{1}{2.5cm}{ {\bf \ref{inc:fa}) } 
  additional space in vocalized form 
}
& \RL{f-} & \colorbox{shadecolor}{~}\RL{fa-}\colorbox{shadecolor}{~} \\[6pt]
\multirow{1}{2.5cm}{ {\bf \ref{inc:wafa})} 
   wrong {\bf prefix}
}
& \RL{wf-}& \RL{wafa-} & and + so/and \\[6pt]
\multirow{3}{2.5cm}{ {\bf \ref{inc:def}) } 
  missing definite indicator in {\bf suffix gloss} 
}
 & \utfRL{آت} & \utfRL{آتِ} & [fem.pl.] + [def.acc.]                               \\
 & \utfRL{اتك} & \utfRL{اتِكِ} & [fem.pl.] + [def.acc.] + your [fem.sg.]            \\
 & \utfRL{آتك} & \utfRL{  آتِكِ} & [fem.pl.] + [\fbox{def.}acc.] + your [fem.sg.]  \\[6pt]
% & \multicolumn{4}{c}{33 other inconsistencies} \\ 
\multirow{2}{2.5cm}{ {\bf \ref{inc:num_gender}) }
  omitted gender/num  qualifier in {\bf gloss}
}
 & \utfRL{اك}&  \utfRL{|BBاكِ}& we [verb] + you \fbox{[fem.sg.]}          \\
 & \utfRL{نهم}&  \utfRL{|Bْنَهُم}& they \fbox{[fem.pl.]} [verb] + them      \\[6pt]
% & \multicolumn{4}{c}{72 total inconsistencies} \\ 
\multirow{3}{2.5cm }{  {\bf \ref{inc:both:sama}) } 
   different ways to express them (both) in {\bf gloss} of {\bf suffix} 
}
 & \noVocRL{thmA}&  \utfRL{|Bَتْهُما} & it/they/she [verb] them (both)   \\
 & \noVocRL{AhmA} &  \noTrutfRL{|BBاهُما} \noArutfRL{ّاهُما}& we [verb] 
                \colorbox{shadecolor}{\color{white} (both of)} them \fbox{(both)}   \\
 & \noVocRL{nAhmA} &  \noVocRL{nAhomA} & we [verb] 
                \colorbox{shadecolor}{\color{white} (both of)} them \fbox{(both)}   \\[6pt]
% & \noVocRL{nhmA} &  \noVocRL{nahomA} & they [verb]
%                \colorbox{shadecolor}{\color{white} (both of)} them \fbox{(both)}  \\
% & \noVocRL{hmA} &  \utfRL{|BBهُما} & they [verb[ 
%\colorbox{shadecolor}{\color{white} (both of)} them \fbox{(both)}  \\
\multirow{3}{2.5cm}{ {\bf \ref{inc:num_gender_differences}) } 
  leftover BAMA style tags  in {\bf gloss}
}
 & \utfRL{كم} & \utfRL{كُم} & he/it [verb] + you [masc.pl.]        \\
 & \utfRL{تكم} & \utfRL{تُكُم} & I [verb] + you 
                \colorbox{shadecolor}{\color{white}(pl.)} \fbox{[masc.pl.]}  \\
 & \utfRL{كن} & \utfRL{|Bُّكُنَّ}  & I [verb] + you 
                \colorbox{shadecolor}{\color{white}(women) } \fbox{[fem.pl.]}   \\[6pt]
% & \multicolumn{4}{c}{13 other inconsistencies } \\                
\multirow{3}{2.5cm}{ {\bf \ref{inc:jus}) } 
  indicative {\bf gloss} with jussive {\bf POS}
}
% & ~ & ~ & [jus.] ~ ~ ~ (null)/IVSUFF\_MOOD:J  \\ 
 & \utfRL{نا} & \utfRL{نا} & 
       IVSUFF\_MOOD:J,
       \colorbox{shadecolor}{\color{white}[ind.]} \fbox{[jus.]} + us  \\
 & \utfRL{ني} & \utfRL{نِي} & 
       IVSUFF\_MOOD:J,
       \colorbox{shadecolor}{\color{white}[ind.]} \fbox{[jus.]} + me \\[6pt]
\multirow{3}{2.5cm}{ {\bf \ref{inc:dot:SAMA}) } 
   omitted `.'  in {\bf gloss} 
}
 & & \utfRL{|Bُ} & [def.nom.] & \\ 
 & \utfRL{ه} & \utfRL{|Bُهُ} & [def.nom \fbox{.}]  \\
 & \utfRL{تكن} & \utfRL{|Bَتُكُنَّ} & 
[fem.sg.] + [def.nom \fbox{.}] + your [fem.pl \fbox{.}]  \\[6pt]
% & \multicolumn{4}{c}{36 other inconsistencies } \\
\vocalize
\multirow{2}{2.5cm} { {\bf \ref{inc:pos:SAMA})} 
  shadda inconsistent in {\bf POS} 
}
 & \utfRL{ي} & \utfRL{يَ} & ya/POSS\_PRON\_1S \\
 & \utfRL{ي} & \utfRL{|Bِyya} & 
 % iy/NSUFF\_MASC\_PL\_NOM  +\colorbox{shadecolor}{\color{white}$\sim$a/}\fbox{ya/}POSS\_PRON\_1S  \\
 $\ldots$ +\colorbox{shadecolor}{\color{white}$\sim$a/}\fbox{ya/}POSS\_PRON\_1S  \\
\hline \hline
\end{tabular}
}
}
%\end{minipage}
\caption{Sample SAMA inconsistencies}
\vspace{-2em}
\label{t:incSAMA}
\end{table}

### 2018-06-26 tex/sarf/sarf.tex
sarf to ACLing draft added

In this Section, we describe the methodology used to construct the affix and stem data structures. 
We also discuss the advantage of choosing these data structures over the ones used in the literature. 
We also describe the construction of the system of Sarf by combining the different data structures.

\subsection{Affix and stem data structures}
%\label{subsec:dfsas}

We analyzed the prefixes and suffixes of 
Buckwalter~\shortcite{Buckwalter:02}
and constructed the agglutinative affixes
into directed acyclic graphs (DAG) similar to 
those in Figures~\ref{f:example}(a) and~\ref{f:example}(c).
The use of the DAG data structure provides compactness as well as 
linear time traversal.
This is computationally superior to the 
approach of Buckwalter where the analyzer considers
all possible substrings and looks them up in the affix tables.

The affix structures encode $L_{pp}$ and $L_{xx}$. 
An accept node in the DAG corresponds to the last letter 
of an affix in the affix lists. 
The accept nodes produce as output the POS and other tags
associated with the node. 
Consider the affix \noTrutfRL{وسيـ} of the word \RL{wsyl`bhA}. 
Possible accept nodes for this input in ${\cal P}$ are for 
\noTrutfRL{و} and \noTrutfRL{يـ}. 
However, only the accept node of \noTrutfRL{يـ} will yield a valid 
solution for the input word \RL{wsyl`bhA}.

%\subsection{Stem transducer}
%\label{subsec:stemFSA}

We built our stem lexicon using the stem lexicon of 
Buckwalter. 
Sarf extends the lexicon of Buckwalter~\shortcite{Buckwalter:02} with
proper names and location names extracted from different online 
sources~\footnote{\href{http://alasmaa.net/}{http://alasmaa.net/ }, 
\href{http://ar.wikipedia.org/}{http://ar.wikipedia.org/}}
as well as biblical sources~\footnote{Genesis 4:17-23; 5:1-32; 9:28-10:32; 11:10-32; 25:1-4, 12-18; 36:1-37:2; Exodus 6:14-25; Ruth 4:18-22; 1 Samuel 14:49-51; 1 Chronicles 1:1-9:44; 14:3-7; 24:1; 25:1-27:22; Nehemiah 12:8-26; Matthew 1:1-16; Luke 3:23-38}.
%We augmented the lexicon with a set of proper names and
%a set of location names which we 
%obtained from online and biblical sources.

The stems share lots of sub-strings. We encoded them in
an efficient double array trie structure~\citep{Aoe:89}. 
The accept nodes are the terminal nodes corresponding to the last 
letters in stems. 
This approach is superior to Buckwalter since it consists of
a linear walk in the trie while with Buckwalter we need 
a number of hash lookups in the order of all possible partitions
of the string. 
For example given the word \RL{'kl}, our approach requires three 
steps to check that this is a valid stem. 
However, buckwalter needs to partition this word into six different partitions 
where each partition contains possible prefix, stem, and suffix.

\subsection{Combined system}
%\label{subsec:combined}

\begin{figure}
\begin{minipage}{0.6\textwidth}
\begin{Verbatim}[fontsize=\relsize{-2},
frame=topline,framesep=4mm,label=\fbox{NDSarf algorithm},
commandchars=\\\{\}, codes={\catcode`$=3\catcode`_=8}]
NDSarf(string $L$, Interface $A$) 
  TraversalVec $M$; -- collection of traversals
  Traversal $\Phi_1$;
  $M$.insert($\Phi_1$);
  foreach $c$ in $L$ \{
    foreach $\Phi$ in $M$ \{
      if ( $\Phi$.node.isAccept() ) \{
        if ($\Phi$.isSuffix())
          if ($c$.isWhite()) or 
            $A$.report();
          if ($\Phi$.lastChar().isNonConnecting())
            $A$.report();
        $\Psi$ = $\Phi$.clone();
        $M$.insert($\Psi$); \}

      if ( $\Phi$.isWalkable($c$) ) \{
        $\Phi$.child($c$);
      \} else \{
        $M$.remove($\Phi$);
        $\Phi$.die();
      \} \} 
    $A$.control($M$, $c$); \}
\end{Verbatim}
\end{minipage}
\caption{Sarf algorithm}
\label{f:ndsarf}
\end{figure}

The algorithm \CodeIn{NDSarf} shown in Figure~\ref{f:ndsarf}
takes as input a text string $L$ and a user-implemented interface
$A$. 
It starts with one traversal $\Phi_1$ and inserts it into a collection $M$. 
The algorithm reads a letter $c$ at a time from $L$
and iterates over all the traversals in $M$. 
For each $\Phi$ in $M$, 
if $\Phi$ is in an accept node and it is in the suffix
phase, then the algorithm checks whether it should report
a full word. 
This happens when $c$ is a white space, a delimiter, 
or when the preceding letter %leading to the current state
was a non-connecting letter.

In all cases, if $\Phi$ is in an accept node, 
it spawns another traversal $\Psi$ and adds it to $M$. 
Notice that spawning a new traversal requires only keeping
the current node of the traversal since all traversals share the
structures of the three transducers.
If $c$ leads to a reject, then $\Phi$ dies 
and we remove it from $M$. 
Otherwise, $\Phi$ proceeds using the $c$ edge.
After iterating over all traversals in $M$, we invoke the 
user-implemented interface $A$
and grant it full access and control over
all traversals in $M$.

### 2018-06-26 tex/sarf/solutiontable.tex
sarf to ACLing draft added

\begin{table}[tb]
\centering
\begin{minipage}{.9\textwidth}
\caption{\label{t:solexample}Morphological solution for \utfRL{وسيلعبونها}(and they will play it)}
\begin{tabular}{lccc} \hline
  & suffix  \noTrutfRL{ونها} & stem  \noTrutfRL{لعبـ}  & prefix \noTrutfRL{وسيـ} \\ \hline
       POS & \cci{IVSUFF\_SUBJ:MP\_MOOD:I+IVSUFF\_DO:3FS} & \cci{VERB\_IMPERFECT} & \cci{CONJ+FUT+IV3MP}  \\
  Transliteration & \cci{uwnahA} & \cci{loEab} & \cci{wa+sa+ya}  
  \\
           Gloss  & \cci{[MASC.PL.]+it/them/her}.  & \cci{play} & \cci{and they will} \\
  \hline% \hline
\end{tabular}
\end{minipage}
\end{table}

### 2018-06-26 tex/sarf/tokantable.tex
sarf to ACLing draft added

\begin{table}[!tbp]
\begin{minipage}{\textwidth}
\begin{tabular}{lccc} 
\hline \hline
Reason & {\bf ATB}  & {\bf TOKAN} & {\bf Frequency} \\ \hline
Dropping diacritics & \utfRL{اميركياً} & \utfRL{اميركيا} &
1.456\% \\
Hamza normalization & \utfRL{أنقرة} & \utfRL{انقرة} &
2.799\% \\
Other normalizations & \utfRL{مغادرت+ه} & \utfRL{مغادرة+ه} &
5.450\% \\
Removing letters & \utfRL{لكن+ني} & \utfRL{لكن+ي} &
0.034\% \\
Adding letters & \utfRL{ل+لتحقيق} & \utfRL{ل+التحقيق} &
1.318\% \\                                                                                                              
{\bf Total} & & &{\bf 11.058\%} \\
\hline \hline
\end{tabular}
\end{minipage}
\caption{ATB-TOKAN segmentation disagreement examples}
{%\vspace{-0.5cm}
}
\label{t:tokan}
\end{table}

When we performed the same experiment using TOKAN toolkit 
(with the ATB scheme alias)
we got a total of 88.942\% agreement with ATB. 
When analyzing the inconsistent instances,
we noticed that TOKAN disregarded input diacritics. 
It also performed its segmentation based on the POS tags
of the morphological solutions in a similar approach 
to that mentioned in~\citep{LRECMaamouriKB08}. 
Table~\ref{t:tokan} shows examples of the disagreement instances.

### 2018-06-27 tex/merf/TALMajor/abstract.tex
french translation and shrink first page

@@
-Rule-based techniques and tools to extract entities and relational entities from documents allow users to specify desired entities using natural language questions, finite state automata, regular expressions, structured query language statements, or proprietary scripts.
-These techniques and tools require expertise in linguistics and programming.
-They lack support of Arabic morphological analysis which is key to 
-process Arabic text.
-In this work, we present \framework; a morphology-based entity and relational entity extraction framework for Arabic text.
-\framework provides a friendly interface where the user, with basic knowledge of linguistic features and regular expressions, defines tag types and interactively associates them with regular expressions defined over Boolean formulae.
-Boolean formulae range over matches of Arabic morphological features, and synonymity features.
-Users define relations with tuples of subexpression matches and can associate code actions with subexpressions.
-\framework computes feature matches, regular expression matches, and constructs entities and relational entities from the user-defined relations.
-We evaluated our work with several case studies and compared with existing application-specific techniques.
-The results show that \framework requires shorter development time and effort compared to existing techniques and produces reasonably accurate results within a reasonable overhead in run time.
+Rule-based techniques to extract relational entities from 
+documents allow users to specify desired entities with 
+natural language questions, finite state automata, regular expressions and
+structured query language. 
+They require linguistic and programming expertise and 
+lack support for Arabic morphological analysis.
+We present a morphology-based entity and relational entity extraction 
+framework for Arabic (\framework).
+\framework requires
+basic knowledge of linguistic features and regular expressions, 
+and provides the ability to interactively specify
+Arabic morphological and synonymity features, 
+tag types associated with regular expressions, and
+relations and code actions defined over matches of subexpressions. 
+\framework constructs entities and relational entities from matches 
+of the specifications.
+We evaluated \framework with several case studies.
+The results show that \framework requires shorter development time and effort compared to existing application specific techniques and produces 
+reasonably accurate results within a reasonable overhead in run time.

### 2018-06-27 tex/merf/TALMajor/macros.tex
minor changes
french translation and shrink first page

--- 2018-03-13 tex/merf/TALMajor/macros.tex
@@
 \usepackage{tikz}
-\usepackage{hyperref}
-\hypersetup{
-    colorlinks=false,
-    pdfborder={0 0 0},
-}
-\usepackage{booktabs}
-\usepackage[ruled]{algorithm2e}
-\usepackage{etex}
-\usepackage{graphicx}
-\usepackage{framed}
-\usepackage{graphics}
+\usepackage{url}
+%\usepackage{hyperref}
+%\hypersetup{
+%    colorlinks=false,
+%    pdfborder={0 0 0},
+%}
+%\usepackage{booktabs}
+%\usepackage[ruled]{algorithm2e}
+%\usepackage{etex}
+%\usepackage{graphicx}
+%\usepackage{framed}
+%\usepackage{graphics}
 \usepackage{booktabs}
 \usepackage{multirow}
-\usepackage{amssymb}
-\usepackage{amsmath}
-\usepackage{savesym}
-\savesymbol{bigtimes}
+%\usepackage{amssymb}
+%\usepackage{amsmath}
+%\usepackage{savesym}
+%\savesymbol{bigtimes}
 \usepackage{mathabx}
-\restoresymbol{TXF}{bigtimes}
+%\restoresymbol{TXF}{bigtimes}
 \usepackage{utf8}
-\usepackage{subfigure}
-\usepackage{todonotes}
-\usetikzlibrary{shapes,arrows}
-\usepackage{color,soul}
+%\usepackage{subfigure}
+%\usetikzlibrary{shapes,arrows}
+%\usepackage{color,soul}
 \usepackage{verbatim}
-\usepackage{array}
-\usepackage{multicol}
+%\usepackage{multicol}
 \usepackage{fancyvrb}
@@
 \usepackage{wasysym}
-\usepackage{pbox}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
-
-%%%%%%%%%% Macros %%%%%%%%%%
+%\usepackage{pbox}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%
+%%%%%%%%%%% Macros %%%%%%%%%%
+%\newcommand{\major}[1]{{\color{black}#1}}
 \def\framework{\textsc{MERF}\xspace}
-\newcommand{\itodo}[1]{\todo[inline,color=green]{\tiny #1}}
 \newcolumntype{L}[1]{>{\raggedright\let\newline\\\arraybackslash\hspace{0pt}}m{#1}}
@@
 \newcommand{\noarrl}[1]{\arabfalse\RL{#1}\arabtrue}
-\def\showvrb#1{%
-\texttt{\detokenize{#1}}%
-}
+%\def\showvrb#1{%
+%\texttt{\detokenize{#1}}%
+%}
 %%% define cal
-\makeatletter
-\DeclareRobustCommand*\cal{\@fontswitch\relax\mathcal}
-\makeatother
+%\makeatletter
+%\DeclareRobustCommand*\cal{\@fontswitch\relax\mathcal}
+%\makeatother
 % done
@@
 % correct bad hyphenation here
-\hyphenation{op-tical net-works semi-conduc-tor}
-%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+%\hyphenation{op-tical net-works semi-conduc-tor}
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-\title[\framework]{\framework: Morphology-based Entity and Relational Entity Extraction Framework for Arabic}
+\title[\framework]{Morphology-based Entity and Relational Entity Extraction Framework for Arabic}
 
@@
 
-\author{}
-%\author{Amin A. Jaber \andauthor Fadi A. Zaraket}
+\author{Amin Jaber \andauthor Fadi A. Zaraket}
 
-\address{}
-%\address{American University of Beirut, Beirut 1107 2020, Lebanon}
+\address{American University of Beirut, Beirut 1107 2020, Lebanon}
 
@@
 natural language processing,
-entity,
-relational entity,
-information visualization,
 tagging.
@@
 \motscles{
-%mot-cl\'e$_1$,
-%mot-cl\'e$_2$,
-%mot-cl\'e$_3$.
+Arabe,
+extraction d'information,
+traitement du langage naturel,
+marquage.
 }
@@
 \resume{
-%Le r\'esum\'e est \`a \'ecrire ici.
+Les techniques à base de règles pour extraire des entités permettent 
+de spécifier les entités souhaitées en utilisant des questions 
+de langage naturel, des automates à états finis, des expressions régulières, et des 
+instructions d'extraction de données.
+Ils nécessitent des expertises en linguistique et en programmation et
+ne soutiennent pas l'analyse morphologique d'Arabe.
+On présente pour l'Arabe 
+un cadre d'extraction d'entité 
+renforcé par l'analyse morphologique (\framework). 
+Il exige des connaissances be base des 
+caractéristiques linguistiques et des expressions régulières, 
+et fournit la pouvoire de spécifier de façon interactive 
+des fonctionnalités de morphologie et synonymie Arabe,
+des types de tag associés avec des expressions régulières, 
+et des relations et actions de code définies sur les correspondances de sous-expressions.
+\framework construit des entités relationnelles à partir des correspondances
+des spécifications.
+On évalue \framework avec des études de cas.
+Les résultats montrent que \framework nécessite un effort de développement plus courts par rapport aux techniques existantes 
+et produit des résultats raisonnablement précis avec une surcharge raisonnable en temps d'exécution.
 }

### 2018-06-27 tex/merf/TALMajor/tool.tex
merf: fit into 25 pages

@@
 For each MRE, 
-\framework generates its equivalent non-deterministic finite automaton (NFA) in the typical manner~\cite{sipser2006introduction}.
+\framework generates its equivalent non-deterministic finite automaton (NFA) in the typical manner~\cite{sipser2012introduction}.
 %Each MRE operation has its equivalent representation in an NFA. 
 We support the upto operation ($f$\^{}$x$), which is not directly 
-supported in~\cite{sipser2006introduction}, by 
+supported in~\cite{sipser2012introduction}, by 
 %it is equivalent to $f?|ff|\dots|\underbrace{f}_{x \text{ times}}$ which has an NFA mapping. 

### 2018-07-08 tex/merf/TALMajor/results.tex
final version

@@
 In the online appendix%
-~\footnote{available at ~\url{-omitted-for-anonymity-}}%http://webfea.fea.aub.edu.lb/fadi/pdfs/merfappendix.pdf
+~\footnote{available at ~\url{http://research-fadi.aub.edu.lb/pdfs/merfappendix.pdf}}
 , we report on eight additional \framework case studies.

### 2018-09-07 tex/merf/TALMajor/gui.tex
applied recommended edits; almost done

@@
 \subsection{Tag type Boolean formula editor}
-The user writes MBF tag types with the tag type editor introduced in~\cite{JaZaMatar}. %shown in Figure~\ref{f:bfe}.
+The user writes MBF tag types with the tag type editor introduced in~\citename{JaZaMatar}~\citeyear{JaZaMatar}. %shown in Figure~\ref{f:bfe}.
 First the user specifies atomic terms by selecting a feature from ${\cal F}$. 
@@
 \subsection{Tag type regular expression editor}
-\vspace{-1em}
+%\vspace{-1em}
 
@@
 The MRE editor of Figure~\ref{f:sfe} allows the 
-definition of an MRE tag type in a user friendly manner. 
+definition of an MRE tag type in a user-friendly manner. 
 The user first adds the required MBF formulae 
@@
 The user can nest the MRE expression using a tree view of the MRE operations. 
-The tree features the name, MBF, and operation for each sub-expression. 
+The tree features the name, MBF, and operation for each subexpression. 
 %Moreover, a defined MRE is directly added to the MBF list enabling substitution and recursion 
@@
 %The regular expression editor enables the user to apply operations to the selected expressions. 
-To specify a binary operation the user selects two sub-expressions and clicks the corresponding
+To specify a binary operation the user selects two subexpressions and clicks the corresponding
 operation button. 
@@
 %For example, we apply the disjunction ($|$) operation to the first $P$ and $N$ added, 
-%then we apply the one or more ($+$) operation to the resulting sub-expression.
+%then we apply the one or more ($+$) operation to the resulting subexpression.
 %We proceed as such to build the regular expression shown in Figure~\ref{f:sfe}. 
@@
 
-While specifying an MRE the user can interact with the visualization and editor views
+While specifying an MRE, the user can interact with the visualization and editor views
 to make sure the MRE expresses the intent. 
@@
 Figure~\ref{fig:treegraph} shows the matching parse tree of the direction task 
-\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building). 
+\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(``Dubai Mall is located near this building''). 
 %Leaf nodes show the match text, internal nodes show the operations, and edge labels shows the MBF tag type name.
@@
 the label.
-The editor shows the MRE tree and allows the user to select the sub 
-expressions and select features 
-of the matches of the sub expressions to define the three components of the relation.
+The editor shows the MRE tree and allows the user to select the subexpressions and select features 
+of the matches of the subexpressions to define the three components of the relation.
 
@@
 subexpression matches. 
-%RE matches In order to add computational actions to an MRE sub-expression, 
+%RE matches In order to add computational actions to an MRE subexpression, 
 %the user selects to edit the actions in the MRE tag type editor. 
 %The view allows the user to specify the actions in a user friendly manner. 
-%\framework provides the user with an API to access the sub-expression match features easily and define the pre-match actions. 
+%\framework provides the user with an API to access the subexpression match features easily and define the pre-match actions. 
 %The match features include the text, position, length, and numerical value if applicable, and morphological features for MBF matches. 

### 2018-09-07 tex/merf/TALMajor/introduction.tex
final version
applied recommended edits; almost done

@@
 Figure~\ref{fig:intromotiv} illustrates the target of \framework using 
-the directions to Dubai Mall example~\footnote{text taken from the Dubai 
-Mall website ~\url{http://www.thedubaimall.com/ar/}}.
+the directions to Dubai Mall example~\footnote{Text taken from the Dubai 
+Mall website ~\url{http://www.thedubaimall.com/ar/}.}.
 The figure also presents a transliteration and an English translation of 
@@
 
-An editor allows the user to associate an action with each sub-expression.
+An editor allows the user to associate an action with each subexpression.
 The user specifies the action with C++ code and uses an API to access information 
@@
 
-This work significantly extends~\cite{JaZaMatar} 
+This work significantly extends~\citename{JaZaMatar}~\citeyear{JaZaMatar}
 that allows for manual, and morphology annotation.
@@
 complex annotations for Arabic based on automatic 
-extraction of morphological tags through a user friendly interactive interface. 
+extraction of morphological tags through a user-friendly interactive interface. 
 \framework has the following advantages.
@@
     and automatically detects relational entities matching the user defined relations. 
-  \item \framework enables the user to associate sub-expressions
+  \item \framework enables the user to associate subexpressions
     with code actions, and executes the code action 
@@
     features such as text, position, length, numerical value, and morphological features.
-  \item \framework enables the user to tag words based on a novel light Arabic wordnet relation that leverages the synonym $Syn^k$ feature.
+  \item \framework enables the user to tag words based on a novel light Arabic WordNet relation that leverages the synonym $Syn^k$ feature.
   \item \framework is open source and available online for the research community 
-    under on (\url{https://github.com/codelogicanalysis/atmine}).
+    under \url{https://github.com/codelogicanalysis/atmine}.
 \end{itemize}

### 2018-09-07 tex/merf/TALMajor/relatedwork.tex
applied recommended edits; almost done

@@
 TEXTMARKER is a semi-automatic rule-based IE system 
-for structured data acquisition\cite{atzmueller2008rule}.
+for structured data acquisition~\cite{atzmueller2008rule}.
 %TEXTMARKER requires user defined rules that consider features of the text 
@@
 
-The work in~\cite{urbain2012user} presents a user 
-driven relational model and targets entity and relation extraction. 
+The work in~\citename{urbain2012user}~\citeyear{urbain2012user} presents a user-driven relational model and targets entity and relation extraction. 
 The user enters a natural language query, and uses the OpenNLP toolkit to 
@@
 Urbain, 
-QARAB}
+QARAB.}
 \end{table}
@@
 while only QARAB supports Arabic linguistic features using a parser, and 
-the work in ~\cite{urbain2012user} uses OpenNLP that currently lacks full
+the work in~\citename{urbain2012user}~\citeyear{urbain2012user} uses OpenNLP that currently lacks full
 support for Arabic morphological features. 
 Similar to TEXTMARKER, \framework has the advantage of providing 
-a user friendly interactive interface to edit the entity and relational 
+a user-friendly interactive interface to edit the entity and relational 
 specifications and visualize the results.
@@
 Another track in the literature targets specific tasks such as NER using 
-statistical and machine learning techniques such as maximum entropy, 
+statistical and machine-learning techniques such as maximum entropy, 
 optimized feature sets and conditional random fields~\cite{benajiba2007anersys,benajiba2008arabic,ekbal2010named,abdelrahman2010integrated}.
@@
 Knowledge-based techniques such 
-as~\cite{zaghouani2010adapting,traboulsi2009arabic} propose local grammars with morphological stemming.% to perform NER. 
-%The work in~
-\cite{ZaMaHaCicling2012Entity} present a method for extracting entities and events, and relations among them from Arabic text using a hierarchy of manually built 
-finite state machines driven by morphological features, 
+as~\citename{zaghouani2010adapting}~\citeyear{zaghouani2010adapting} and~\citename{traboulsi2009arabic}~\citeyear{traboulsi2009arabic} propose local grammars with morphological stemming.
+% to perform NER.
+The work in~
+\citename{ZaMaHaCicling2012Entity}~\citeyear{ZaMaHaCicling2012Entity} extract entities and events, and relations among them, from Arabic text using a hierarchy of manually built finite state machines driven by morphological features,
 and graph transformation algorithms.
@@
 BRAT~\cite{brat} and WordFreak~\cite{wordfreak} are manual 
-multi-lingual user friendly web-based annotators that allow the construction 
+multi-lingual user-friendly web-based annotators that allow the construction 
 of entity and relation annotation corpora.
@@
 
-\cite{kholidy2010towards} present an overview of annotation tools 
+\citename{kholidy2010towards}~\citeyear{kholidy2010towards} present an overview of annotation tools 
 %and their Arabic-English word alignment issues 
 and concludes with a set of rules and guidelines needed in an Arabic annotation alignment tool.
-The work in \cite{dukes2011supervised} presents a collaborative effort towards morphological and syntactic annotation of the Quran.
-\cite{dorr2010interlingual} presents a framework for inter-lingual 
-annotation of parallel text corpora with multi-level representations.
-\cite{kulick2010consistent} presents the integration of the 
-Standard Arabic Morphological Analyzer (SAMA) into the workflow of the Arabic Treebank.
+The work in~\citename{dukes2013supervised}~\citeyear{dukes2013supervised} presents a collaborative effort towards morphological and syntactic annotation of the Quran.
+\citename{dorr2010interlingual}~\citeyear{dorr2010interlingual} present a framework for interlingual annotation of parallel text corpora with multi-level representations.
+\citename{kulick2010consistent}~\citeyear{kulick2010consistent} presents the integration of the Standard Arabic Morphological Analyzer (SAMA) into the workflow of the Arabic Treebank.
 
-The work in \cite{smrz2004morphotrees} presents %the extension of TrEd, 
+The work in~\citename{smrz2004morphotrees}~\citeyear{smrz2004morphotrees} presents %the extension of TrEd, 
 a customizable general purpose tree editor, with the Arabic MorphoTrees annotations. 
@@
 
-Task specific annotation tools such as~\cite{alrahabi2006semantic}
-use enunciation semantic maps to automatically annotate 
-directly reported Arabic and French speech.
+Task specific annotation tools such as~\citename{alrahabi2006semantic}~\citeyear{alrahabi2006semantic} use enunciation semantic maps to automatically annotate directly reported Arabic and French speech.
 AraTation is another task specific tool for semantic annotation of 

### 2018-09-07 tex/merf/TALMajor/results.tex
applied recommended edits; almost done

@@
 In this section we evaluate \framework with four case studies. 
-We perform a survey like evaluation where developers manually 
+We perform a survey-like evaluation where developers manually 
 built task specific information extraction tools for the case studies 
@@
 We evaluated ANGE with
-Musnad Ahmad, a hadith book, where we constructed an annotated 
+\textit{Musnad Ahmad}, a hadith book, where we constructed an annotated 
 golden reference containing 1,865 words.
@@
 with articles from issues of  the Lebanese 
-Al-Akhbar newspaper where we constructed 
+\textit{Al-Akhbar} newspaper where we constructed 
 an annotated golden reference containing 1,677 words. 
@@
 an extract from the Genesis biblical text with 1,227 words.
-Finally, we used an annotated article from the Lebanese Assafir
+Finally, we used an annotated article from the Lebanese \textit{Assafir}
 newspaper with 1,399 words to evaluate the NUMNORM case study
-\footnote{available at \url{http://www.assafir.com} and 
+\footnote{Available at \url{http://www.assafir.com} and 
 \url{http://www.al-akhbar.com}.}. 
 %
-In the online appendix%
-~\footnote{available at ~\url{http://research-fadi.aub.edu.lb/pdfs/merfappendix.pdf}}
-, we report on eight additional \framework case studies.
+In the online appendix\footnote{Available at ~\url{http://research-fadi.aub.edu.lb/pdfs/merfappendix.pdf}.},we report on eight additional \framework case studies.
 Manual annotators inspected the outcome and provided corrections where tools 
@@
       ANGE & 1,865  
-    & 2 months & 1.79 & 0.99 & 0.99 & 3000+ lines of code\\
+    & 2 months & 1.79 & 0.99 & 0.99 & 3,000+ lines of code\\
       \framework &  & 3 hours & 7.24 & 0.99 & 0.93  & 8 MBFs and 4 MREs\\
@@
     %ATEEMA~\cite{ZaMa2012IJCLATime}
-      ATEEMA & 1,677 & 1.5 months & 2.53 & 0.88  & 0.89  & 1000+ lines of code  \\
+      ATEEMA & 1,677 & 1.5 months & 2.53 & 0.88  & 0.89  & 1,000+ lines of code  \\
       \framework & & 3 hours & 3.14 & 0.91  & 0.81  & 3 MBFs and 2 MREs\\
@@
       Genealogy tree & 1,227 
-    & 3 weeks & 0.74 & 0.96 & 0.98 & 3000+ lines of code\\
+    & 3 weeks & 0.74 & 0.96 & 0.98 & 3,000+ lines of code\\
       \framework &  & 4 hours & 2.28 & 0.84 & 0.93  & 3 MBFs and 3 MREs\\
@@
 Simulating an NDFSM over the MBF tags is potentially exponential. 
-In practice all our case studies terminated within a predetermined 
+In practice, all our case studies terminated within a predetermined 
 time bound of less than 30 minutes. 
@@
 \framework needed eight MBFs and four MREs for narrator chain, 
-three MBFs and 2 MREs for temporal entity, three MBFs and three MREs for 
+three MBFs and two MREs for temporal entity, three MBFs and three MREs for 
 genealogy, and three MBFs, one MRE, and 57 lines of code actions for the number normalization tasks. 
@@
 MBF \cci{TOLD} denotes referencing between narrators and checks the disjunction of 
-the stems \RL{.hd_t}(spoke to), \RL{`n}(about), \RL{sm`}(heard), \RL{'_hbr}(told), and \RL{'nb-'}(inform). 
-MBF \cci{MEAN} checks the stem \RL{`ny}(mean). 
+the stems \RL{.hd_t}(``spoke to''), \RL{`n}(``about''), \RL{sm`}(``heard''), \RL{'_hbr}(``told''), and \RL{'nb-'}(``inform''). 
+MBF \cci{MEAN} checks the stem \RL{`ny}(``mean''). 
 MBFs \cci{BLESS}, \cci{GOD}, \cci{UPONHIM}, and \cci{GREET} check the 
@@
 MRE \cci{pbuh} denotes a praise phrase often associated with 
-the end of a hadith (peace be upon him), 
+the end of a hadith (``peace be upon him''), 
 and is the satisfied by the sequence of
@@
 \framework assigns the symbols $s_1$ and $s_2$ for the 
-MRE sub-expressions \cci{TOLD} and \cci{nar}, respectively. 
+MRE subexpressions \cci{TOLD} and \cci{nar}, respectively. 
 We define the relation $\langle s_2,s_2',s_1\rangle$ 
@@
 Temporal entities are text chunks that express temporal information.
-Some represent absolute time such as \RL{Al_hAms mn 'Ab 2010}. 
-Others represent relative time such as \RL{b`d _hmsT 'ayAm}, and quantities 
-such as \RL{14 ywmA}. 
+Some represent absolute time such as~\RL{Al_hAms mn 'Ab 2010}$\!\!$.
+Others represent relative time such as~\RL{b`d _hmsT 'ayAm}$\!\!\!$, and quantities such as~\RL{14 ywmA}$\!$.
 {\em ATEEMA} presents a temporal entity detection technique for the Arabic language using 
@@
 \framework assigns the symbols $s_1$, $s_2$, and $s_3$ 
-for the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H}, respectively. 
-%The actions associated with the sub-expressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
+for the subexpressions \cci{DT}, \cci{TMB}, and \cci{H}, respectively. 
+%The actions associated with the subexpressions \cci{DT}, \cci{TMB}, and \cci{H} are presented in the appendix.
 Figure~\ref{fig:numnormalgo} shows the actions associated with the \cci{DT}, \cci{TMB}, and \cci{H} subexpressions that cumulatively compute the numeric value of the numeric expression match.

### 2018-09-07 tex/merf/TALMajor/sarf.tex
applied recommended edits; almost done

@@
 For example, the word \RL{bn}%
-can be interpreted as \RL{bon} {\tt ``coffee''} with a {\em damma} diacritic on the 
+can be interpreted as \RL{bon} ``coffee'' with a {\em damma} diacritic on the 
 letter \utfrl{بـ} or 
-\vocalize \RL{bin} (son of) with a 
+\vocalize \RL{bin} (``son of'') with a 
 {\em kasra} diacritic on the letter \utfrl{بـ}.
@@
 \arabfalse \RL{_dhb alwald-il_A almdrsT} \arabtrue
-(the kid went to school). 
+(``the kid went to school''). 
 The letters \notrutfrl{د} and \notrutfrl{ى} have 
@@
 $C=C_{p_1} \ldots C_{p_{|p|}} C_s C_{x_1}\ldots C_{x_{|x|}}$. 
-$P_{p_i}, G_{p_i},$ and $C_{p_i},1\le i \le |p|$ are the POS, Gloss and 
+$P_{p_i}, G_{p_i},$ and $C_{p_i},1\le i \le |p|$ are the POS, gloss and 
 category tags of prefix $p_i$. 
-$P_{x_j}, G_{x_j},$ and $C_{x_j},1\le j \le |x|$ are the POS, Gloss and 
+$P_{x_j}, G_{x_j},$ and $C_{x_j},1\le j \le |x|$ are the POS, gloss and 
 category tags of suffix $x_i$. 
-$P_s, G_s,$ and $C_s$ are the POS, Gloss and 
+$P_s, G_s,$ and $C_s$ are the POS, gloss and 
 category tags of stem $s$. 
@@
 Intuitively, 
-$p,x,P,G$ and $C$ are concatenations of prefix, suffix, POS, Gloss and
-Category values, respectively.
+$p,x,P,G$ and $C$ are concatenations of prefix, suffix, POS, gloss and
+category values, respectively.
 

### 2018-09-07 tex/merf/TALMajor/tool.tex
final version
applied recommended edits; almost done

@@
 Up to our knowledge, 
-$Syn^k$ provides the first light Arabic wordnet based on the lexicon of Sarf.
+$Syn^k$ provides the first light Arabic WordNet based on the lexicon of Sarf.
 The sets $E, A,$ and $L$ denote all English words, Arabic words, 
@@
 \vspace{-3em}
-\caption{$Syn^2($\RL{mA'}$)$}
+\caption{$Syn^2($\RL{mA'}$)$.}
 \vspace{-2em}
@@
 and defines it as a disjunction of terms that check for 
-solutions matching stems such as \RL{qrb} (near) and \RL{fy} (in).
+solutions matching stems such as \RL{qrb} (``near'') and \RL{fy} (``in'').
 %requires the stem feature of a word to belong to a selected list of stems containing \RL{fy} and \RL{qrb}. a
@@
 requiring the stem feature to belong to a set of stems such as
-\RL{'wl}(first), \RL{_tAny}(second), \dots \RL{`A^sr}(tenth).
+\RL{'wl}(``first''), \RL{_tAny}(``second''), \dots \RL{`A^sr}(``tenth'').
 
@@
 The first tree refers to the text 
-\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(Khalifa Tower next to the first intersection).
+\RL{brj _hlyfT bAlqrb mn AltqA.t` Al-'wl}(``Khalifa Tower next to the first intersection'').
 The second tree refers to the text 
-\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(Dubai Mall is located near this building).
+\RL{dby mwl `l_A mqrbT mn h_dA Almbn_A}(``Dubai Mall is located near this building'').
 %The leaf nodes of the trees are entities and the edges and internal nodes are text, 
@@
 %and word distance based relational entities. 
-The leaf nodes of the trees are matches to formulae and the internal nodes represent roots to sub-expression matches.
-For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the sub-expression $(P|N)+$.
+The leaf nodes of the trees are matches to formulae and the internal nodes represent roots to subexpression matches.
+For instance, \RL{brj _hlyfT} in match 1 tree corresponds to the subexpression $(P|N)+$.
 
-\vspace{-2em}
+\vspace{-1em}
 \subsection{User-defined relations and actions}
@@
 $e_1,e_2,$ and $r$ are identifiers associated with 
-sub-expressions of an MRE $f$.
+subexpressions of an MRE $f$.
 Matches of the relation are a set of labeled binary edges
@@
 We are interested in constructing the relational entity graph in Figure~\ref{fig:intromotiv}.
-Let $e_1,o_1,r,o_2,$ and $e_2$ be identifiers to the sub-expressions 
-$(P|N)+$,$O?$,$R$,$O\wedge 2$, and $(P|N|U)+$, respectively.
+Let $e_1$, $o_1$, $r$, $o_2,$, and $e_2$ be identifiers to the subexpressions 
+$(P|N)+$, $O?$, $R$, $O\wedge 2$, and $(P|N|U)+$, respectively.
 %We select $(P|N)+$ to denote an entity as it captures a composite place name such as \cci{Khalifa tower}.
-The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (Khalifa Tower), \RL{bAlqrb} (next), \RL{mn} (to), and \RL{AltqA.t` Al-'wl} (first intersection).
+The matches to $e_1$, $r$, $o_2$, and $e_2$ in match 1 (Fig.~\ref{fig:taskMRE}) are \RL{brj _hlyfT} (``Khalifa Tower''), \RL{bAlqrb} (``next''), \RL{mn} (``to''), and \RL{AltqA.t` Al-'wl} (``first intersection'').
 Note that there is no match to the optional $O$ formula in match 1.
-Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second matching parse tree are \RL{dby mwl} (Dubai Mall), \RL{`l_A} (is located), \RL{mqrbT} (near), \RL{mn h_dA} (this), and \RL{Almbn_A} (building), respectively.
+Similarly, the matches to $e_1$, $o_1$, $r$, $o_2$, and $e_2$ in the second matching tree are \RL{dby mwl} (``Dubai Mall''), \RL{`l_A} (``is located''), \RL{mqrbT} (``near''), \RL{mn h_dA} (``this''), and \RL{Almbn_A} (``building''), respectively.
 
@@
 Moreover, \framework allows advanced users to write C++ code snippets 
-to process matches of sub-expressions.
-Each sub-expression can be associated with two computational 
+to process matches of subexpressions.
+Each subexpression can be associated with two computational 
 actions: \cci{pre-match} and \cci{on-match}.
@@
 We support the upto operation ($f$\^{}$x$), which is not directly 
-supported in~\cite{sipser2012introduction}, by 
+supported in~\citename{sipser2012introduction}~\citeyear{sipser2012introduction}, by 
 %it is equivalent to $f?|ff|\dots|\underbrace{f}_{x \text{ times}}$ which has an NFA mapping. 
@@
 %A simulation match $m$ of an expression $f$ is a tree where the root is the MRE expression, the internal nodes are the MRE and MBF operations, and the leaves are matches of the MAT terms of $f$.
-A simulation match $m$ of an expression $f$ is a parse tree where the root spans the expression, the internal nodes are roots to sub-expressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g.
-Fig.~\ref{fig:taskMRE}.
+A simulation match $m$ of an expression $f$ is a parse tree where the root spans the expression, the internal nodes are roots to subexpressions of $f$, and the leaves are matches of the MBF formulae of $f$, e.g.
+Figure~\ref{fig:taskMRE}.
 The sequence of leaf matches forms a vector of tags $\langle r_k,r_{k+1},\dots,r_j\rangle$ 
@@
 %\framework maintains a function $\phi\subset Q\times\Phi$, 
-%where $\mathit{Q}$ is the set of states in the NFA and $\Phi$ is the set of sub-expressions in the MRE.
-%So $(q,f)\in \phi$ iff state $q\in Q$ was generated by \framework to correspond to sub-expression $f\in \Phi$. 
+%where $\mathit{Q}$ is the set of states in the NFA and $\Phi$ is the set of subexpressions in the MRE.
+%So $(q,f)\in \phi$ iff state $q\in Q$ was generated by \framework to correspond to subexpression $f\in \Phi$. 
 %\framework uses $\phi$ to compute a match tree with respect to the MRE regular expression. 
@@
 \begin{comment} 
-Note that the number of matches for $e_i$ ($\ldbrack e_i \rdbrack$) depends on the corresponding sub-expression.
-For instance, we could have multiple matches if we associate $e_i$ with the sub-expression $P|N$ while we have $(P|N)+$.
+Note that the number of matches for $e_i$ ($\ldbrack e_i \rdbrack$) depends on the corresponding subexpression.
+For instance, we could have multiple matches if we associate $e_i$ with the subexpression $P|N$ while we have $(P|N)+$.
 %to be the elements of $\ldbrack e_1 \rdbrack \times \ldbrack e_2 \rdbrack \times \ldbrack r \rdbrack$
@@
 %where the distance is the number of words between the matches. 
-Hence, \framework supports the following configurations for $e_i$ and $e_j$ based on the corresponding matches of sub-expressions:
+Hence, \framework supports the following configurations for $e_i$ and $e_j$ based on the corresponding matches of subexpressions:
 \begin{itemize}
@@
 
-%In Figure~\ref{fig:taskMRE}, \framework names the sub-expressions $(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
+%In Figure~\ref{fig:taskMRE}, \framework names the subexpressions $(P|N)+$, $(P|N|U)+$, $O?$, $O\wedge 2$, and $R$, as $e_1, e_2, o_1, o_2,$ and $r$, respectively. 
 %The user defines the semantic relations $\langle e_1, e_2, r\rangle$, $\langle r, e_1, o_1\rangle$, and $\langle r,e_2,o_2\rangle$.

### 2018-09-10 tex/merf/TALMajor/macros.tex
fixed journal tag
fixed journal tag: 2017 not 2018
merf edits done

@@
 
-\submitted[15/10/2017]{TAL~58-03}{R}
+%\submitted[15/10/2017]{TAL~58-03}{R}
+\journal{TAL. Volume 58 -- n° 3/2017}{1}{25}
 
@@
 de spécifier les entités souhaitées en utilisant des questions 
-de langage naturel, des automates à états finis, des expressions régulières, et des 
+de langage naturel, des automates à états finis, des expressions régulières et des 
 instructions d'extraction de données.
-Ils nécessitent des expertises en linguistique et en programmation et
-ne soutiennent pas l'analyse morphologique d'Arabe.
-On présente pour l'Arabe 
+Ils nécessitent des expertises en linguistique et en programmation, et
+ne soutiennent pas l'analyse morphologique de l'arabe.
+On présente pour l'arabe 
 un cadre d'extraction d'entité 
 renforcé par l'analyse morphologique (\framework). 
-Il exige des connaissances be base des 
+Il exige des connaissances de base des 
 caractéristiques linguistiques et des expressions régulières, 
-et fournit la pouvoire de spécifier de façon interactive 
-des fonctionnalités de morphologie et synonymie Arabe,
+et fournit la possibilité de spécifier de façon interactive 
+des fonctionnalités de morphologie et synonymie arabes,
 des types de tag associés avec des expressions régulières, 
@@
 On évalue \framework avec des études de cas.
-Les résultats montrent que \framework nécessite un effort de développement plus courts par rapport aux techniques existantes 
+Les résultats montrent que \framework nécessite un effort de développement plus court par rapport aux techniques existantes 
 et produit des résultats raisonnablement précis avec une surcharge raisonnable en temps d'exécution.

### 2018-09-10 tex/merf/TALMajor/results.tex
merf edits done

@@
 
-We evaluated ANGE with
-\textit{Musnad Ahmad}, a hadith book, where we constructed an annotated 
-golden reference containing 1,865 words.
-We evaluated ATEEMA 
-with articles from issues of  the Lebanese 
-\textit{Al-Akhbar} newspaper where we constructed 
-an annotated golden reference containing 1,677 words. 
-For the genealogical tree extraction we used 
-an extract from the Genesis biblical text with 1,227 words.
-Finally, we used an annotated article from the Lebanese \textit{Assafir}
-newspaper with 1,399 words to evaluate the NUMNORM case study
-\footnote{Available at \url{http://www.assafir.com} and 
+We evaluated ANGE with \textit{Musnad Ahmad}, a hadith book, where we constructed an annotated golden reference containing 1,865 words.
+We evaluated ATEEMA with articles from issues of  the Lebanese \textit{Al-Akhbar} newspaper where we constructed an annotated golden reference containing 1,677 words. 
+For the genealogical tree extraction we used an extract from the Genesis biblical text with 1,227 words.
+Finally, we used an annotated article from the Lebanese \textit{Assafir} newspaper with 1,399 words to evaluate the NUMNORM case study\footnote{Available at~\url{http://www.assafir.com} and 
 \url{http://www.al-akhbar.com}.}. 
 %
-In the online appendix\footnote{Available at ~\url{http://research-fadi.aub.edu.lb/pdfs/merfappendix.pdf}.},we report on eight additional \framework case studies.
+In the online appendix\footnote{Available at~\url{http://research-fadi.aub.edu.lb/pdfs/merfappendix.pdf}.}, we report on eight additional \framework case studies.
 Manual annotators inspected the outcome and provided corrections where tools 
@@
 For instance, ANGE~\cite{ZaMaFlairs2012HadithBio} required two 
-months of development by a research assistant with six and 14 hours of 
+months of development by a research assistant with 6 and 14 hours of 
 course work and teaching duties, respectively.
